### 2021-01-05 analysis/paper/paper.Rmd
Second commit

---
title: "Title Goes Here"
author:
  - Myfirstname Mylastname:
      email: fl@oneplace.edu
      institute: [UofO]
      correspondence: true
  - Yourfirstname Yourlastname:
      email: fl@another.edu
      institute: [UofA]
      correspondence: false
institute:
  - UofO: University of One Place
  - UofA: University of Another Place
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::pdf_document2:
    fig_height: 7.6
    fig_width: 12.8
  powerpoint_presentation:
    fig_height: 7.6
    fig_width: 12.8
    reference_doc: ../templates/template.pptx
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
header-includes:
  - \usepackage{subfig}
# abstract: |
#   Text of abstract
# keywords: |
#   keyword 1; keyword 2; keyword 3
# highlights: |
#   These are the highlights. 
---

<!-- cambiarn nomenclatura. Imaginario / real -> pasarlo a nombrar la fase -->

<!-- Chequear con ncep -->

<!-- The following code chunk defines some general settings how code chunks should behave. -->

```{r setup, echo = FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  cache = TRUE,
  comment = "#>",
  cache.extra = 43,
  cache.path = "../cache/paper/",
  fig.path = "../figures/",
  dpi = 300
)

library(metR)
library(data.table)
library(magrittr)
library(ggplot2)
library(tagger)
library(ggperiodic)
library(patchwork)
library(shceof)

theme_set(theme_shceof(base_size = 16))

grid <- theme(panel.grid = element_line(color = "gray10", size = 0.4, linetype = 3))

guide_colorstrip_bottom <- function(width = 25, height = 0.5, ...) {
  guide_colorstrip(title.position = "top", title.hjust = 0.5,
                   barheight = height,
                   barwidth = width, ...)
}

guide_colorsteps_bottom <- function(width = 25, height = 0.5, ...) {
  guide_colorsteps(title.position = "top", title.hjust = 0.5,
                   barheight = height,
                   barwidth = width, ...)
}

axis_labs_smol <- theme(axis.text = element_text(size = 6))

pattern_crosshatch <- ggplot2::ggproto("GeomCrosshatch", ggpattern::GeomPolygonPattern)

ggplot2::update_geom_defaults(pattern_crosshatch, 
                              list(pattern = "crosshatch",
                                   pattern_colour = "black",
                                   pattern_fill = "black",
                                   pattern_density = 0.01,
                                   pattern_alpha = 0.2,
                                   pattern_spacing = 0.02))

lab_lev <- AddSuffix(" hPa")
lab_cplx <- c(R = "Real", 
              I = "Imaginary")
combine_lists <- function(...) {
  # browser(expr = length(x) != length(y))
  X <- list(...)
  names <- names(X[[1]])
  names(names) <- names
  lapply(names, function(n) {
    unlist(lapply(X, function(x) x[[n]]))
  })
}
```

```{r read-data}
era5 <- ReadNetCDF(ERA5(), vars = c(hgt = "z",
                                    t = "t"),
                   subset = list(level = list(850, 700, 200, 50),
                                 latitude = -90:10)) %>% 
  normalise_coords() %>% 
  .[season(time) == "SON"] %>%
  .[, .(hgt = mean(hgt), 
        t = mean(t)), by = .(lev, lon, lat, time = seasonally(time))]
```

# Introduction

# Data & Methods

ERA5, ERA5 BE, ERA 20C

If

# Results

```{r}
eof_naive <- era5[lev %in% c(50, 200) & lat %between% c(-90, -20)] %>% 
  # .[, hgt := Anomaly(hgt), by = .(season(time), lon, lat, lev)] %>%
  .[, hgt  := Anomaly(hgt)*sqrt(cos(lat*pi/180)), by = .(time, lev, lat)] %>% 
  .[, .(eof = list(EOF(hgt ~ time | lon + lat, n = 1:4, data = .SD, suffix = "EOF"))), by = lev]

explained <- eof_naive[, eof[[1]]$sdev, by = lev] %>% 
  .[order(lev, EOF)] %>% 
  .[, scales::percent(r2, accuracy = .1)]

```

(ref:cap-eof-naive) Leading 4 EOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2009.

(ref:subcap-eof-naive) Spatial patterns (arbitrary units). The numbers at the top-left of each panel is the variance explained by each EOF.

(ref:subcap-cor-eof-naive) Coefficient of determination of the temporal index of each EOF between levels.

```{r cor-eof-naive, fig.cap = "(ref:cap-eof-naive)", fig.subcap = c("(ref:subcap-eof-naive)", "(ref:subcap-cor-eof-naive)"), fig.ncol = 1}
eof_naive[, eof[[1]]$right, by = lev] %>% 
  periodic(lon = c(0, 360)) %>% 
  ggplot(aes(lon, lat)) +

  geom_contour_fill(aes(z = hgt, fill = ..level..), 
                    breaks = AnchorBreaks(0, 0.01, exclude = 0)) +
  geom_contour_tanaka(aes(z = hgt), breaks = AnchorBreaks(0, 0.01, exclude = 0)) +

  geom_qmap(~.x[lat <= -20]) +
  scale_fill_divergent_discretised(guide = "none")  +
  scale_y_latitude() +
  scale_x_longitude() +
  coord_polar() +
  facet_grid(lev~EOF, labeller = labeller(lev = lab_lev))  +
  axis_labs_smol +
  tagger::tag_facets(tag = "panel", tag_pool = explained, tag_suffix = "")

eof_naive[, eof[[1]]$left, by = lev] %>% 
  .[, item := paste(lev, EOF, sep = "_")] %>% 
  widyr::pairwise_cor(item, time, hgt) %>% 
  setDT() %>%
  # .[]
  .[tstrsplit(item1, "_")[[1]] != tstrsplit(item2, "_")[[1]], ] %>% 
  .[tstrsplit(item1, "_")[[1]] != 50] %>% 
  .[, item1 := tstrsplit(item1, "_")[[2]]] %>%
  .[, item2 := tstrsplit(item2, "_")[[2]]] %>%
  ggplot(aes(item1, item2)) +
  geom_raster(aes(fill = correlation^2)) +
  shadowtext::geom_shadowtext(aes(label = round(correlation^2, 2)), 
                              color = "black", bg.colour = "white", bg.r = 0.2) +
  scale_fill_gradient(guide = "none", low = "white", high = scales::muted("red"), limits = c(0, 1)) +
  scale_x_discrete("200 hPa") +
  scale_y_discrete("50 hPa") +
  plot_layout(ncol = 1)
```

Figure \@ref(fig:cor-eof-naive-1) show the spatial pattern of the 4 leading EOF of zonal anomalies of geopotential height for 50 hPa and 200 hPa computed separately. The percentages XXX . At 50 hPa, EOF1 and EOF2 on one hand, and EOF3 and EOF4 on the other are clearly pairs of zonal waves with wave numbers 1 and 2-3 respectively shifted by 1/4 wavelength. That is, each group of two EOFs represents the same zonal wave structure that changes in magnitude and location. Similarly, at 200 hPa EOF2 and EOF3 represent the same zonal wave structure with wave number ~3.

Furthermore, the wave 1 pattern represented by the first two EOFs at 50 hPa is similar to the wave 1 pattern shown as the leading EOF at 200 hPa, and the wave 2-3 pattern represented by the third and fourth EOF at 50 hPa is similar to the wave 3 pattern present in EOF2 and EOF3 at 200 hPa. These similarities suggest some level of joint variability across levels. Figure \@ref(fig:cor-eof-naive-2) shows the coefficient of determination between temporal series of each EOF at each level. EOF3 and EOF4 at 50 hPa are highly correlated with EOF2 and EOF3 at 200 hPa. Not only each pair of EOFs represent a single structure within each level, but they represent a coherent pattern of variability between levels. Similarly, the leading EOF of each level are relatively highly correlated, again suggesting an unique mode of joint variability.

```{r compute-ceof}
invisible(
  ReadNetCDF(ERA5(), vars = c(hgt = "z"),
             subset = list(latitude = -90:10,
                           level = list(200, 50))) %>%
    normalise_coords() %>%
    .[season(time) == "SON"] %>%
    .[, .(hgt = mean(hgt)), by = .(lev, lon, lat, time = seasonally(time))] %>%
    .[, season := season(time)] %>%
    {
      ceof <<- .[, .(eof = list(compute_ceof(.SD))), by = .(season)]
      ceof_splitted <<-  .[, levs := lev] %>%
        .[, .(eof = list(compute_ceof(.SD, n = 1:3))), by = .(season, levs)]

    }
)
```

(ref:cap-splitted) Coefficient of determination of the absolute magnitude of complex EOFs between 200 hPa and 50 hPa computing EOF separatedly for each level. The high correlation between levels for the two leading EOFs justifies treating the pattern as a mode of covariability between the stratosphere and the troposphere and, thus, computing the EOFs using both levels at the same time. (probably not to be shown.)

```{r corr-ceof-splitted, fig.cap = "(ref:cap-splitted)"}
ceof_splitted[, eof[[1]]$left, by = .(lev = levs)] %>%
  copy() %>%
  .[, mag := abs(hgt)] %>%
  .[, item := paste(lev, PC, sep = "_")] %>%
  widyr::pairwise_cor(item, time, mag) %>%
  setDT() %>%
  # .[]
  .[tstrsplit(item1, "_")[[1]] != tstrsplit(item2, "_")[[1]], ] %>%
  .[tstrsplit(item1, "_")[[1]] != 50] %>%
  .[, item1 := tstrsplit(item1, "_")[[2]]] %>%
  .[, item2 := tstrsplit(item2, "_")[[2]]] %>%
  ggplot(aes(item1, item2)) +
  geom_raster(aes(fill = correlation^2)) +
  shadowtext::geom_shadowtext(aes(label = round(correlation^2, 2)),
                              color = "black", bg.colour = "white", bg.r = 0.2) +
  scale_fill_gradient(guide = "none", low = "white", high = scales::muted("red")) +
  scale_x_discrete("200 hPa") +
  scale_y_discrete("50 hPa")
```

```{r}
plot_ceof <- function(data, n = 1:2, formula = lev  ~  PC) {

  gdata <- data[, cut(eof[[1]], n)$right, by = .(season)]

  var <- data[, cut(eof[[1]], n)$sdev, by = .(season)] %>%
    .[, setNames(paste0(PC, " (", scales::percent(r2), ")"),
                 PC)]

  gdata %>%
    ggperiodic::periodic(lon = c(0, 360)) %>%
    ggplot(aes(lon, lat)) +

    geom_contour_fill(aes(z = Re(hgt), fill = ..level..),
                      breaks = AnchorBreaks(0, 0.005, exclude = 0)) +
    geom_contour2(aes(z = Im(hgt),  linetype = factor(sign(..level..))),
                  breaks = AnchorBreaks(0, 0.005, exclude = 0)) +

    scale_fill_divergent_discretised(guide = "none") +

    geom_qmap(~ .x[lat <= -20]) +
    scale_x_longitude() +
    scale_y_latitude() +
    scale_linetype_manual(values = c("1" = 1, "-1" = 2), labels = c("1" = "+", "-1" = "-"),
                          guide = "none") +
    facet_grid(formula,
               labeller = labeller(lev = lab_lev,
                                   PC = var)) +
    coord_quickmap(ylim = c(NA, -20)) +
    theme(legend.title = element_blank())
}
```

```{r ceofs, fig.cap = "Leading 2 EOFs of zonal anomalies of geopotential height in 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2009.", fig.subcap = c("Spatial patterns. Real part in shading, imaginary part in contours.", "Temproal series."), fig.ncol = 1}
ceof[season == "SON"] %>%
  plot_ceof() +
  coord_polar() +
  axis_labs_smol +
  tag_facets(tag = "cr")

ceof %>%
  .[, eof[[1]]$left, by = season] %>%
  sep_ReIm(hgt) %>%
  .[season == "SON"] %>%
  # .[, hgt := scale(hgt), by = .(part, PC)] %>%
  ggplot(aes(time, hgt)) +
  geom_hline(yintercept = 0, size = 0.1, color = "gray30") +
  geom_line(aes(color = part)) +
  scale_x_datetime(NULL) +
  scale_y_continuous("EOF") +
  scale_color_brewer(NULL, palette = "Dark2") +
  facet_grid(PC ~ .) +
  theme(panel.spacing = grid::unit(1.5, "lines"))  +
  ggthemes::geom_rangeframe(sides = "lr", size = 0.1, color = "gray30")
```
Both observations motivate the decision of performing complex EOF jointly between levels. The computation of the EOFs was carried out using data from both levels at the same time, therefore, each complex EOF has a spatial part that depends on longitude, latitude and level, an a temporal part that depends only on time. Figures \@ref(fig:ceofs-1) and \@ref(fig:ceofs-2) show, respectively, the spatial and temporal parts of the first two leading complex EOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa. In the spatial patterns in Figure \@ref(fig:ceofs-1), the real (in shading) and the imaginary parts (in contour) are in quadrature by construction, so that each EOF describe a single wave-like pattern whose amplitude and position (i.e. phase) is controlled by the magnitude and argument of the complex temporal EOF.

The wave patterns described by these complex EOFs match the patterns seen in the *navie* EOFs of Figure \@ref(fig: cor-eof-naive-1): The first is a wave 1, while the second is a wave 3. Note that in Figure \@ref(fig:ceofs-2), the real part of EOF1 has non zero mean. This is due to th fact that the geopotential fields that enter into the algorithm are anomalies with respect to the zonal mean, not the time mean. The real part of the first EOF is, therefore, capturing the mean zonally anomalous field *and the variability that projects onto it*.

```{r, fig.cap = "Correlación entre el las anomalías zonales de geopotential de cada mes y el campo medio de anomalías zonales de geopotencial (negro) y correlación (inversa) entre anomalías zonales de altura geopotencial y la parte real del EOF1 en (rojo) entre 80S y 20S."}
# cor_mean <- era5 %>% 
#   .[lev %in% c(50, 200)] %>% 
#   .[lat <= -20] %>% 
#   # .[lat >= -80] %>% 
#   .[, hgt := Anomaly(hgt), by = .(time, lat, lev)] %>% 
#   .[, hgt_mean := mean(hgt), by = .(lat, lon, lev)] %>% 
#   .[, .(mean = cor(hgt_mean, hgt)), by = .(lev, time)] %>%
#   identity()
# 
# 
# ceof$eof[[1]] %>% 
#   .$right %>% 
# 
#   .[PC == "PC1"] %>% 
#   sep_ReIm(format = "wide") %>% 
#   .[, hgt := NULL] %>% 
#   .[, PC := NULL] %>% 
#   era5[., on = .NATURAL] %>% 
#   .[, hgt := Anomaly(hgt), by = .(time, lat, lev)] %>% 
#   .[, .(hgt = -cor(hgt, R)), by = .(lev, time)] %>% 
# 
#   ggplot(aes(time, hgt))  +
#   geom_line(color = "red") +
#   geom_line(data = cor_mean, aes(y = mean)) +
#   facet_grid(lev~.)

```

(ref:extended-series-cap) Temporal series extended using ERA5 Back extended preliminary edition (period 1950 -- 1978), ERA5 (period 1979 -- 2019) and ERA20C (period 1900 -- 2010). Each series is computed by projecting monthly geopotential height zonal anomalies standardised by level south of 20ºS onto the corresponding spatial pattern. In numbers, the correlation between the ERA20C-based series and the two ERA5-based series. The ERA5 + ERABC and the ERA20C series are centred and scaled using the mean and standard deviation for their period of overlap (1950 -- 2010).

```{r extended-series, fig.cap = "(ref:extended-series-cap)"}

pattern <- ceof$eof[[1]]$right %>%
  sep_ReIm() %>%
  setnames("hgt", "EOF")

era20c <- ERA20C() %>%
  ReadNetCDF(vars = c(hgt = "z"),
             subset = list(latitude = range(pattern$lat),
                           level = list(200, 50))) %>%
  normalise_coords() %>%
  .[season(time) == "SON"] %>%
  .[, .(hgt = mean(hgt)), by = .(lev, lon, lat, time = seasonally(time))] %>%
  .[, season := season(time)] %>%
  .[, hgt := hgt/sd(hgt), by = .(lev)] %>%
  .[pattern, on = .NATURAL, allow.cartesian = TRUE] %>%
  .[, .(hgt = weighted.mean(hgt*EOF, w = cos(lat*pi/180))),  by = .(part, PC, time)]

era_more <- ERA5_BE() %>%
  ReadNetCDF(vars = c(hgt = "z"),
             subset = list(latitude = range(pattern$lat),
                           level = list(200, 50))) %>%
  rbind(.,
        ReadNetCDF(ERA5(), vars = c(hgt = "z"),
                   subset = list(latitude = range(pattern$lat),
                                 level = list(200, 50)))) %>%
  normalise_coords() %>%
  .[season(time) == "SON"] %>%
  .[, .(hgt = mean(hgt)), by = .(lev, lon, lat, time = seasonally(time))] %>%
  .[, season := season(time)] %>%
  .[, hgt := hgt/sd(hgt), by = .(lev)] %>%
  .[pattern, on = .NATURAL, allow.cartesian = TRUE] %>%
  .[, .(hgt = weighted.mean(hgt*EOF, w = cos(lat*pi/180))),  by = .(part, PC, time)] %>%
  era20c[., on = c("part", "PC", "time")] %>%
  .[, hgt := (i.hgt - mean(i.hgt))/sd(i.hgt)*sd(hgt, na.rm = TRUE) + mean(hgt, na.rm = TRUE),
    by = .(part, PC)] %>%
  .[, i.hgt := NULL]

extended <- rbindlist(list(era20c = era20c,
               era5 = era_more),  use.names = TRUE, idcol = "dataset") %>%
  .[, hgt := (hgt - mean(hgt[dataset == "era5"], na.rm = TRUE))/sd(hgt[dataset == "era5"], na.rm = TRUE),
    by = .(part, PC)]

correlations <- extended %>% 
  dcast(time + PC + part ~ dataset, value.var = "hgt") %>% 
  na.omit() %>% 
  .[, cor(era20c, era5), 
    by = .(PC, part, time = ifelse(year(time) >= 1979, 2005, 1960))] %>% 
  .[, time := lubridate::as_datetime(paste0(time,  "-01-01"))] %>% 
  .[, dataset := ifelse(year(time) >= 1979, "era5", "era5be")]

extended %>% 
  copy() %>% 
  .[dataset == "era5", dataset := ifelse(year(time) < 1979, "era5be", "era5")] %>% 
  ggplot(aes(time, hgt)) +
  geom_hline(yintercept = 0, size = 0.1, color = "gray30") +
  geom_line(aes(color = dataset)) +
  geom_text(data = correlations, aes(color = dataset, y = 2.3, label = signif(V1, 2))) +
  scale_x_datetime(NULL) +
  scale_y_continuous("Standarised EOF") +
  scale_color_brewer("Data", palette = "Dark2", labels = c(era5 = "ERA5",
                                                           era20c = "ERA20C",
                                                           era5be = "ERA5 BE"),
                     guide = guide_legend(override.aes = list(size = 1), order = 1)) +
  ggnewscale::new_scale_colour() +
  geom_smooth(data = ~.x[dataset == "era20c"][, dataset := dataset == "era20c"], 
              method = "lm", se = FALSE,  aes(color = dataset)) +
  geom_smooth(data = ~copy(.x)[, dataset := dataset == "era20c"][time %between% common_range(time, dataset)],
              aes(color = dataset), method = "lm", se = FALSE) +

  scale_color_brewer("Trend", palette = "Dark2", direction = -1,
                     guide = guide_legend(override.aes = list(size = 1), order = 99),
                     labels = c("FALSE" = "ERA5+ERA5 BE",
                                "TRUE" = "ERA2CO")) +
  facet_grid(part ~ PC) +
  theme(panel.spacing = grid::unit(1.5, "lines"))  +
  ggthemes::geom_rangeframe(sides = "lr", size = 0.1, color = "gray30") +
  tag_facets("cr")
```

Figure \@ref(fig:extended-series) shows temporal series of the two complex EOFs extended beyond the satellite era using the preliminary ERA5 back extension going back to 1950 and using ERA20C for the period 1900 -- 2010. The hybrid ERA5 + ERA5 BE series and the ERE20C series agree relatively well during the period of overlap, although correlations are higher in the satellite period. The ERA20C time-series has a significant upward linear trend in the imaginary part of EOF1. However, there is no trend in the corresponding hybrid modern dataset during the period of overlap and the correlations between the old and modern series is much poorer. This suggests that the observed trends is not robust.

## Regression

```{r}
regr <- ceof[, eof[[1]]$left, by = season] %>%
  .[season == "SON"] %>%
  sep_ReIm(format = "wide") %>%
  .[, hgt := NULL] %>%
  .[, `:=`(R = R/sd(R), I = I/sd(I)), by = .(season, PC)] %>%
  dcast(time ~ PC, value.var = c("R", "I")) %>%
  .[melt(era5, id.vars = c("time", "lon", "lat", "lev")), on = "time", allow.cartesian = TRUE] %>%
  .[lev %in% c(850, 700, 200, 50)] %>%
  # .[variable == "hgt"] %>%
  .[, FitLm(value, I_PC1, R_PC1, I_PC2, R_PC2, se = TRUE), by = .(lev, lon, lat, variable)] %>%
  .[, c("term", "PC") := tstrsplit(term, "_")] %>%
  rm_intercept()
```

```{r}
plot_regr <- function(data, which_PC, which_levs, which_var) {
  data <- data %>%
    .[ PC %in% which_PC] %>%
    .[variable == which_var]

  if (length(which_PC) > 1) {
    facet <- ggh4x::facet_nested(PC ~  term, labeller = labeller(term = lab_cplx))
  } else {
    facet <- ggh4x::facet_nested(lev ~  term, labeller = labeller(term = lab_cplx,
                                                                  lev = lab_lev))
  }

  if ("lev" %in% colnames(data)) {
    data <- data[lev %in% which_levs]
  } else {
    data[, lev := which_levs]
  }
  data[lat <= -20] %>%
    # .[term == "I" & lev == 50] %>%
    .[, p_val := Pvaluate(estimate, std.error, df, "fdr"), by = .(lev, PC, term)] %>%
    periodic(lon = c(0, 360)) %>%
    ggplot(aes(lon, lat)) +
    ditto(
      geom_contour_fill(aes(z = estimate), breaks = AnchorBreaks(exclude = 0), global.breaks = FALSE) +
        geom_contour_tanaka() +
        scale_fill_divergent(NULL, guide = guide_colorstrip_bottom())
    ) +

    geom_contour2(aes(z = p_val), breaks = 0.05, size = 0.1) +
    stat_subset(aes(subset = p_val <= 0.05 & is.cross(lon, lat, 1)),
                geom = "point", size = 0.2, alpha = 0.1) +
    geom_qmap(~ .x[lat <= -20]) +

    scale_x_longitude() +
    scale_y_latitude() +
    facet +
    coord_polar() +
    axis_labs_smol
}

```

```{r, fig.cap = "Regresiones con altura geopotencial PC1"}
plot_regr(regr, "PC1", c(200, 50), "hgt")
```

```{r, fig.cap = "Regresiones con temperatura y PC1"}
plot_regr(regr, "PC1", c(200, 50), "t")
```

```{r, fig.cap = "Regresiones con altura geopotencial y PC2"}
plot_regr(regr, "PC2", c(200, 50), "hgt")
```

```{r, fig.cap = "Regresiones con temperatura y PC2"}
plot_regr(regr, "PC2", c(200, 50), "t")
```

```{r}
regr_simpler <- ceof[, eof[[1]]$left, by = season] %>%
  .[season == "SON"] %>%
  sep_ReIm(format = "wide") %>%
  .[, hgt := NULL] %>%
  .[, `:=`(R = R/sd(R), I = I/sd(I)), by = .(season, PC)] %>%
  .[era5, on = "time", allow.cartesian = TRUE] %>%
  .[lev %in% c(200, 50)] %>%
  .[, FitLm(hgt, I, R, se = TRUE), by = .(lev, lon, lat, PC, season)] %>%
  rm_intercept()
```

```{r, fig.cap = "Varianza temporal de altura geopotencial explicada por la variación temporal de cada componente principal."}
regr_simpler %>%
  # .[term == "R"] %>%
  .[lat <= -20] %>%
  periodic(lon = c(0, 360)) %>%
  ggplot(aes(lon, lat)) +

  geom_contour_fill(aes(z = r.squared, fill = ..level..),
                    breaks = MakeBreaks(0.1, exclude = 0)) +
  scale_fill_divergent_discretised(NULL, guide = guide_colorsteps_bottom())  +

  geom_qmap(~.x[lat <= -20]) +
  scale_x_longitude() +
  scale_y_latitude() +
  ggh4x::facet_nested(lev ~ PC, labeller = labeller(lev = lab_lev)) +
  coord_polar() +
  axis_labs_smol
```

```{r, fig.cap = "Amplitud de ondas zonales de la regresión con altura geopotencial (figuras anteriores)"}
regr %>%
  .[lat <= -20] %>%
  .[lev %in% c(200, 50)] %>%
  # .[PC == "PC1"] %>%
  .[, FitWave(estimate, 1:3), by = .(lev, PC, term, lat)] %>%
  ggplot(aes(lat, amplitude)) +
  geom_line(aes(color = factor(k))) +
  ggh4x::facet_nested(lev + PC ~ term, scales = "free") +
  scale_color_brewer("Wavenumber", palette = "Dark2") +
  scale_y_continuous("Amplitude") +
  scale_x_latitude() +
  coord_flip()
```

Interpretaciones:

* Parte real de EOF1: variabilidady que se proyecta en la dirección del campo medio

* Parte imaginaria de EOF1: variabilidad asociada a la parte simétrica del SAM

* EOF2: PSA1 y PSA2.  
  - Parte imaginaria: proyecta sobre la parte asimétrica del SAM (PSA2)
  - Parte real: relativamente independiente del SAM

El primer EOF es esencialmente una medida de la variabilidad proyectada en la dirección el campo medio. La parte importante, de todas formas, es la Real.

## Relationship with other indices

```{r}
sam <- "/home/elio/Documents/CONICET/asymsam/analysis/data/derived_data/indexes.Rds" %>%
  readRDS() %>%
  .[, .(time, sam = -estimate_norm, lev, symmetry = term)] %>%
  .[, sam := sam/sd(sam), by = .(lev, symmetry)] %>%
  .[season(time) == "SON"] %>%
  .[, .(sam = mean(sam)), by = .(symmetry, lev, time = seasonally(time))]

enso <- rsoi::download_oni(TRUE, data_path("raw", "oni.csv")) %>%
  as.data.table() %>%
  .[, .(time = lubridate::as_datetime(Date), oni = ONI)] %>%
  na.omit() %>%
  .[season(time) == "SON"] %>%
  .[, .(oni = mean(oni)), by = .(time = seasonally(time))]

plot_with_index <- function(ceof, index) {
  index_col <- colnames(index)[2]

  ceof[, eof[[1]]$left] %>%
    sep_ReIm() %>%
    .[, hgt := scale(hgt), by = .(PC, part)] %>%
    index[., on = "time"] %>%
    na.omit() %>%
    .[, index := get(index_col)] %>%
    ggplot(aes(hgt, index)) +
    geom_point(aes(color = part)) +
    geom_smooth(aes(color = part), method = "lm") +
    shadowtext::geom_shadowtext(data = ~.x[, .(cor = cor(index, hgt),
                                               x = -1.2,
                                               y = sign(-cor(index, hgt))*.1 + predict(lm(index ~ hgt),
                                                                                       newdata = data.frame(hgt = -1.2))),
                                           by = .(PC, part)],
                                bg.colour = "white", bg.r = 0.2,
                                size = 5,
                                aes(x, y, color = part, label = paste0("r² = ", round(cor^2, 2)))) +
    scale_x_continuous("EOF") +
    scale_y_continuous(toupper(index_col)) +
    scale_color_brewer("Part", palette = "Dark2") +
    facet_grid(part~PC)
}
```

```{r, fig.cap = "Relación del EOF (estandarizado) con SAM"}
plot_with_index(ceof, sam[lev == 700 & symmetry == "full"][, .(time, sam)])
```

```{r}

time <- ceof[, eof[[1]]$left] %>%
  sep_ReIm() %>%
  .[, hgt := scale(hgt), by = .(PC, part)] %>%
  .[]

"/home/elio/Documents/CONICET/asymsam/analysis/data/derived_data/indexes.Rds" %>%
  readRDS() %>%
  # .[lev %in% c(200, 50)] %>%
  # .[term != "full"] %>%
  rm_singleton() %>%
  .[season(time) == "SON"] %>%
  .[, .(estimate_norm = mean(estimate_norm)), by = .(lev, term, time = seasonally(time))] %>%
  .[time, on =.NATURAL, allow.cartesian = TRUE] %>%
  na.omit() %>%
  .[, cor(hgt, estimate_norm)^2, by = .(lev, term, part, PC)] %>%
  ggplot(aes(lev, V1)) +
  geom_line(aes(color = term, size = term)) +
  scale_x_level() +
  scale_y_continuous("r²") +
  scale_size_manual(guide = "none",
                    values = c(full = 1,
                               asym = 0.5,
                               sym = 0.5)) +
  scale_color_brewer(NULL, palette = "Dark2", labels = c(full = "SAM",
                                                         asym = "Asymmetric SAM",
                                                         sym = "Symmetric SAM")) +
  facet_grid(part ~ PC) +
  coord_flip()

```

```{r, fig.cap = "Idem anterior pero con ENSO"}
plot_with_index(ceof, enso)
```

::: notes

En la tropósfera, la pate imaginaria del PC2 captura la parte asimétrica del SAM.
La parte real es muy parecida en estructura, pero como está desfasada, no coincide con el SAM y entonces no proyecta fuerte en él.

La Parte imaginaria está muy influenciada por ENSO, como se ve en la relación con el íncide y en la regresión con la SST. La fase Real parece no tener influencia del océano tropical. Sin embargo, los flujos de acción de onda.

A su vez, la parte imaginaria del PC1 captura también parte del SAM, pero principalmente la parte simétrica.

:::

```{r}
psi <- ReadNetCDF(ERA5(), vars = "vo",
                  subset = list(level = c(50, 200),
                                time = c("1979-01-01", "2019-12-01"))) %>%
  normalise_coords() %>%
  .[season(time) == "SON"] %>%
  .[, .(vo = mean(vo)), by = .(lon, lat, lev,time = seasonally(time))] %>%
  .[, solve_poisson(vo, lon, lat), by = .(lev, time)] %>%
  .[, psi.z := Anomaly(value), by = .(time, lev, lat)] %>%
  .[, value := NULL]
```

```{r}
regr_psi <- ceof[, eof[[1]]$left, by = season] %>%
  sep_ReIm(hgt, "wide") %>%
  .[, hgt := NULL] %>%
  .[, `:=`(R = R/sd(R), I = I/sd(I)), by = .(season, PC)] %>%
  .[psi, on = "time", allow.cartesian = TRUE] %>%
  .[season(time) == "SON"] %>%
  .[, FitLm(psi.z, R, I, se = TRUE),by = .(lon, lat, lev, PC, season)] %>%
  rm_intercept() %>%
  .[, psi.z := estimate] %>%
  .[, c("fx", "fy") := shceof::WaveFlux(.SD, p = 200), by = .(lev, PC, season, term)]
```

```{r}

plot_psi <- function(data, which_pc, which_lev) {
  data %>%
    .[PC == which_pc] %>%
    .[lev == which_lev] %>%
    # .[abs(estimate) > 1, estimate := NA] %>%
    .[lat %between% c(-90, 10)] %>%
    .[, p_val := Pvaluate(estimate, std.error, df, "fdr"), by = .(PC, term)] %>%
    periodic(lon = c(0, 360)) %>%
    ggplot(aes(lon, lat)) +
    geom_contour_fill(aes(z = estimate), breaks = AnchorBreaks(exclude = 0)) +

    geom_contour2(aes(z = p_val), breaks = 0.05, size = 0.1) +
    stat_subset(aes(subset = p_val <= 0.05 & is.cross(lon, lat, 2)),
                geom = "point", size = 0.2, alpha = 0.5) +
    geom_streamline(aes(dx = fx, dy = fy), min.L = 3, L = 10, n = 30,
                    size = 0.2, res = 2) +
    geom_qmap(keep = .015) +
    scale_fill_divergent(guide = "none") +
    scale_x_longitude() +
    scale_y_latitude() +
    scale_mag(guide = "none") +
    facet_grid(term ~ PC, labeller = labeller(term = lab_cplx)) +
    coord_quickmap(ylim = c(NA, 10)) +
    # ggh4x::facet_nested(term ~ PC) +
    axis_labs_smol
}

```

```{r}
sst <- ERA5_SST() %>%
  ReadNetCDF(vars = c("t" = "sst")) %>%
  normalise_coords() %>%
  .[season(time) == "SON"] %>%
  .[, .(t = mean(t)), by = .(lon, lat, time = seasonally(time))]
```

```{r}
regr_sst <- ceof[, eof[[1]]$left, by = season] %>%
  sep_ReIm(hgt, "wide") %>%
  .[, hgt := NULL] %>%
  .[, `:=`(R = R/sd(R), I = I/sd(I)), by = .(season, PC)] %>%
  .[sst, on = "time", allow.cartesian = TRUE] %>%
  .[season(time) == "SON"] %>%
  .[, FitLm(t, I, R, se = TRUE), by = .(lon, lat, PC, season)] %>%
  rm_intercept()
```

```{r}
plot_sst <- function(data, which_pc) {
  data %>%
    .[PC == which_pc] %>%
    # .[abs(estimate) > 1, estimate := NA] %>%
    .[lat %between% c(-90, 10)] %>%
    .[, p_val := Pvaluate(estimate, std.error, df, "fdr"), by = .(PC, term)] %>%
    periodic(lon = c(0, 360)) %>%
    ggplot(aes(lon, lat)) +
    ditto(
      geom_contour_fill(aes(z = estimate), breaks = AnchorBreaks(exclude = 0)) +
        geom_contour_tanaka(aes(z = estimate), breaks = AnchorBreaks(exclude = 0)) +
        scale_fill_divergent(NULL, guide = guide_colorstrip_bottom(width = 15))
    ) +

    geom_contour2(aes(z = p_val), breaks = 0.05, size = 0.1) +
    stat_subset(aes(subset = p_val <= 0.05 & is.cross(lon, lat, 2)),
                geom = "point", size = 0.2, alpha = 0.5) +
    geom_qmap(keep = .015) +

    scale_x_longitude() +
    scale_y_latitude() +
    facet_grid(term ~ PC, labeller = labeller(term = lab_cplx)) +
    axis_labs_smol +
    coord_quickmap(ylim = c(-90, 10))
}

```

```{r, fig.cap = "Regresión con PSI en 200 hPa y sst, PC1"}
plot_psi(regr_psi, "PC1", 200) +
  plot_sst(regr_sst, "PC1")
```

```{r, fig.cap = "Regresión con PSI en 200 hPa y sst, PC2"}
plot_psi(regr_psi, "PC2", 200) +
  plot_sst(regr_sst, "PC2")
```

### Precipitation

```{r}
cmap <- ReadNetCDF(CMAP(), vars = c(pp = "precip"),
                   subset = list(lat = -90:10)) %>%
  normalise_coords() %>%
  .[, .(pp = mean(pp, na.rm = TRUE)), by = .(lon, lat, time = seasonally(time))]
```

```{r}
regr_pp <- ceof[, eof[[1]]$left, by = season] %>%
  sep_ReIm(hgt, "wide") %>%
  .[, hgt := NULL] %>%
  .[, `:=`(R = R/sd(R), I = I/sd(I)), by = .(season, PC)] %>%
  .[melt(cmap, id.vars = c("lon", "lat", "time")), on = "time", allow.cartesian = TRUE] %>%
  .[season == "SON"] %>%
  .[, FitLm(value, I, R, se = TRUE), by = .(lon, lat, PC, variable, season)] %>%
  rm_intercept()
```

```{r}

continents <- list(africa  = list(lon = c(0, 60),
                                  lat = c(-40, 0)),
                   oceania = list(lon = c(100, 180),
                                  lat = c(-50, -5)),
                   america = list(lon = c(265, 325),
                                  lat = c(-60, -10)))

add_continent <- function(dt) {
  dt[, continent := fcase(lon %between% continents$africa$lon &
                            lat %between% continents$africa$lat, "africa",
                          lon %between% continents$oceania$lon &
                            lat %between% continents$oceania$lat, "oceania",
                          lon %between% continents$america$lon &
                            lat %between% continents$america$lat, "america")] %>%
    .[, continent := factor(continent, levels = c("africa", "oceania", "america"), ordered = TRUE)] %>%
    .[]
}

```

```{r}
plot_pp <- function(data, which_pc, which_term = c("R", "I"), which_continent) {
  data %>%
    .[PC %in% which_pc] %>%
    .[term %in% which_term] %>%
    .[, lev := "sfc"] %>%
    add_continent() %>%
    .[continent == which_continent] %>%
    .[, p_val := Pvaluate(estimate, std.error, df, "fdr"), by = .(lev, PC, term)] %>%
    periodic(lon = c(0, 360)) %>%
    ggplot(aes(lon, lat)) +
    ditto(
      geom_contour_fill(aes(z = estimate), breaks = AnchorBreaks(exclude = 0)) +
        geom_contour_tanaka() +
        scale_fill_divergent(NULL, guide = guide_colorstrip_bottom(15))
    ) +

    geom_contour2(aes(z = p_val), breaks = 0.05, size = 0.1) +
    stat_subset(aes(subset = p_val <= 0.05),
                geom = "point", size = 1) +
    geom_qmap(~.x %>% add_continent() %>% .[continent == which_continent], keep = 1) +

    scale_x_longitude(ticks = 15) +
    scale_y_latitude(ticks = 15) +
    ggh4x::facet_nested(term ~ PC, labeller = labeller(term = lab_cplx)) +
    axis_labs_smol +
    coord_quickmap()
}
```

```{r, fig.cap = "Regresión de cada PC con la precipitación"}
plot_pp(regr_pp, "PC1", which_continent =  "america")  +
  plot_pp(regr_pp, "PC2", which_continent =  "america")
```

```{r, fig.cap = "Regresión de cada PC con la precipitación"}
plot_pp(regr_pp, "PC1", which_continent = "africa") +
  plot_pp(regr_pp, "PC2", which_continent = "africa")
```

```{r, fig.cap = "Regresión de cada PC con la precipitación"}
plot_pp(regr_pp, "PC1", which_continent =  "oceania") +
  plot_pp(regr_pp, "PC2", which_continent =  "oceania")
```

# Conclusions

# References

<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

### 2021-01-05 analysis/paper/psi-test.Rmd
Second commit

---
title: "Pescaditos"
author: "Elio Campitelli"
date: "02/11/2020"
output: bookdown::html_document2
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)

library(metR)
library(data.table)
library(magrittr)
library(ggplot2)
library(shceof)
```

Este documento documenta el resultado de usar `solve_poisson()` para obtener la función corriente (en particular, su anomalías zonal) a partir de la vorticidad.

Dado que ERA provee vorticidad pero no función corriente, tenía que usar NCEP. Pero NCEP tiene función corriente en (algunos) niveles sigma, por lo cual si yo hago el análisis usando ERA5 en niveles de presión, la comparación con la función corriente tiene ya dos incompatibilidades (aunque tampoco tan grandes, preliminarmente, los resultados no varían mucho usando la función corriente de NCEP en sigma 0.2101 o la función corriente derivada de ERA5 en 200 hPa.).

En cualquier caso, antes de usar `solve_poisson()` quiero chequear que funciona correctamente. Lo que voy a hacer es usar la vorticidad de NCEP en sigma 0.2101 para obtener la función corriente y comparar el resultado con el valor descargado de NCEP.

Como detalle, si a la función corriente se le suma y multiplica constantes, sigue siendo solución de la ecuación de Poisson, por lo que la solución obtenida por `solve_poisson()` puede tener cualquier escala que nada que ver. Para resolver esto, voy a escalar la función corriente derivada con constantes obtenidas a partir de hacer `lm(psi ~ ps_derivada)`.

```{r}
psi <- NCEP_PSI() %>% 
  ReadNetCDF(subset = list(level =  0.2101, 
                           time = c("1979-01-01", NA))) %>% 
  rm_singleton()

vor <-  NCEP_VOR() %>% 
  ReadNetCDF(subset = list(level =  0.2101, 
                           time = c("1979-01-01", NA))) %>% 
  rm_singleton()

ncep <- vor[psi, on = .NATURAL] %>% 
  na.omit()
rm(psi, vor)
```

```{r}
psi_derived <- ncep[, solve_poisson(vor, lon, lat), by = time]
setnames(psi_derived, "value", "psi_derived")
```

```{r}
ncep <- ncep[psi_derived, on = .NATURAL]

scaling <- ncep[, FitLm(psi, psi_derived)]
ncep[, psi_derived := psi_derived*scaling$estimate[2] + scaling$estimate[1]]
ncep[, `:=`(psi         = Anomaly(psi),
           psi_derived = Anomaly(psi_derived)), by = .(lat, time)] 
```

La Figura \@ref(fig:comparacion) muestra cuatro campos de anomalía zonal de geopotencial tomados al azar. La función corriente derivada con es casi indistinguible de la función corriente "real". La correlación global durante todo el período entre la función corriente real y la derivada es de `r ncep[, cor(psi, psi_derived)]`. Prácticamente perfecto.

```{r comparacion, fig.cap = "Anomalía zonal de función corriente provista por NCEP (panel izquierdo) y derivada a partir de la vorticidad (panel derecho)"}
set.seed(42)
ncep[time %in% sample(unique(time), 4)] %>% 
  melt(id.vars = c("lon", "lat", "time"), measure.vars = patterns("^psi")) %>% 
  ggplot(aes(lon, lat)) +
  geom_contour_fill(aes(z = value)) +
  scale_fill_divergent() +
  facet_grid(factor(time)~variable)
```
La correlación es igual de alta para todos los campos (\@ref(fig:cor-tiempo)) y latitudes (\@ref(fig:cor-lat)).

```{r, cor-tiempo, fig.cap = "Correlación entre función corriente derivada y obtenida por NCEP  travès del tiempo."}
ncep %>% 
  .[, cor(psi, psi_derived), by = .(time)] %>% 
  ggplot(aes(time, V1)) +
  geom_line() +
  scale_y_continuous(limits = c(0, 1))
```

```{r cor-lat, fig.cap = "Correlación entre función corriente derivada y obtenida por NCEP para cada latitud"}
ncep %>% 
  .[, cor(psi, psi_derived), by = .(lat)] %>% 
  ggplot(aes(lat, V1)) +
  geom_line() +
  scale_y_continuous(limits = c(0, 1))
```

## Conclusión

`solve_poisson` funciona perfecto para obtener función corriente a partir de vorticidad! 🥳

### 2021-01-05 analysis/paper/trabajo.Rmd
Second commit

---
title: ""
author: "Elio Campitelli"
output: 
   # pdf_document:
   #    latex_engine: xelatex
   powerpoint_presentation:
    # ioslides_presentation:
        fig_height: 7.6
        fig_width: 12.8
        reference_doc: template.pptx
urlcolor: blue
header_includes:
   - \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
# Notification
start.time <- unclass(Sys.time())
min.time <- 10
knit_doc <- knitr::knit_hooks$get("document")

knitr::knit_hooks$set(document = function(x) {
   took <- unclass(Sys.time()) - start.time
   if (unclass(Sys.time()) - start.time >= min.time) {
      notify("Done knitting!", 
             paste0("Took ", round(took), " seconds"),
             time = 5)
   }  
   knit_doc(x)
})

name <- tools::file_path_sans_ext(knitr::current_input())
knitr::opts_chunk$set(echo = FALSE,
                      cache = TRUE,
                      cache.extra = 42, 
                      dpi = 150,
                      warning = FALSE, message = FALSE,
                      # out.extra = "",
                      fig.width = 19.88/2.45,
                      fig.height=11.84/2.45,
                      cache.path = paste0("cache/", name, "/"),
                      fig.path = paste0("fig/", name, "/")
                      # fig.align = "center"
)

# knitr::opts_chunk$set(fig.width = 13, 
#                       fig.height = 7)

library(metR)
library(data.table)
library(ggplot2)
library(magrittr)
library(patchwork)
library(here)

here <- here::here

lev.breaks <- c(1000, 500, 300, 200, 100, 50, 10)

theme_elio <- theme_minimal(base_size = 14) +
   theme(strip.background = element_rect(fill = NA, color = "gray30"),
         # text = element_text(family = font_rc),
         legend.position = "bottom", legend.box = "vertical",
         panel.spacing.y = unit(5, "mm"),
         panel.spacing.x = unit(5, "mm"),
         legend.spacing = unit(2, "mm"), 
         panel.border = element_rect(colour = "black", fill = NA),
         plot.margin = grid::unit(rep(3, 4), "mm"),
         # legend.title = element_blank(),
         legend.box.spacing = unit(3, "mm"),
         legend.margin = margin(t = -5),
         panel.grid = element_line(color = "gray10", size = 0.4, linetype = 3),
         panel.ontop = TRUE)
theme_set(theme_elio)
# theme_set(hrbrthemes::theme_ipsum_rc() + theme(panel.ontop = TRUE))

guide_colorstrip_bottom <- function(width = 25, height = 0.5, ...) {
   guide_colorstrip(title.position = "top", title.hjust = 0.5,
                    barheight = height,
                    barwidth = width, ...)
}

subset_data <- list(lat = c(-90, 0), 
                    time = lubridate::as_datetime(c("1985-01-01",
                                                    "2014-12-01")))
breaks_zero <- AnchorBreaks(0, NULL, 0)

makeActiveBinding(".", function(value) .Last.value, .GlobalEnv)

setnames <- function(x, ...) {
   names <- c(...)
   # print(names)
   data.table::setnames(x, names(names), unname(names))
}

ReIm <- function(complex) {
   list(R = Re(complex), I = Im(complex))
}

Pvaluate <- function(estimate, std.error, df, adjustement = "none") {
   p.adjust(2*pt(abs(estimate)/std.error, df, lower.tail = FALSE), method = "fdr")
}

source(here::here("scripts", "facet_nested.R"))
source(here::here("scripts", "helperfun.R"))

map_simple <- function(wrap = c(0, 360), out = "sf") {
   map <- maps::map("world", fill = TRUE, 
                    col = "transparent", plot = FALSE, wrap = wrap)
   IDs <- vapply(strsplit(map$names, ":"), function(x) x[1], 
                 "")
   proj <- sp::CRS("+proj=longlat +datum=WGS84")
   map <- maptools::map2SpatialPolygons(map, IDs = IDs, 
                                        proj4string = proj)

   simple <- rmapshaper::ms_simplify(map, keep = 0.015)
   simple
}

map_data <- subset(fortify(map_simple()), lat <= 0)

map <- function(subset = NULL, color = "black", size = 0.2, fill = NA, wrap = c(0, 360), ...) {
   data <- fortify(map_simple(wrap = wrap))
   subset <- eval(substitute(subset), envir = data)
   if (is.null(subset)) subset <- TRUE

   geom_polygon(data = data[subset, ], 
                aes(long, lat, group = group), 
                color = color, 
                size = size, 
                fill = fill,
                ...)
}

sep_ReIm <- function(dt, col, longer = TRUE) {
   names <- c("R", "I")
   expr <- quote(copy(dt)[, (names) := ReIm(col)])
   expr  <-  do.call(substitute, list(expr, 
                                      list(col = substitute(col))))
   data <- eval(expr)

   if (isTRUE(longer)) {
      data[, deparse(substitute(col)) := NULL]
      data <- setDT(tidyr::pivot_longer(data, R:I, names_to = "part", values_to = deparse(substitute(col))))
   }

   return(data[])
}

geom_zero <- function(dir = "h", ...) {
   switch(dir,
          h = geom_hline(yintercept = 0, ...),
          v = geom_vline(xintercept = 0, ...))
}

```

```{r read-data}
subset <- list(latitude = -90:0, 
               level = list(50, 200))

files <- c(era   = here("datos", "ERA-Interim", "erai.mon.mean.nc"),
           era20 = here("datos", "ERA-20C", "era20c.mon.mean.nc"),
           ncep  = here("datos", "NCEP Reanalysis", "hgt.mon.mean.nc"))
var <- c("z", "z", "hgt")
labs_datasets <- c(era20 = "ERA20C", era = "ERA-I", ncep = "NCEP")
names(var) <- files
subsets <- list(
   era = subset,
   era20 = subset,
   ncep = list(lat = -90:0,
               level = list(50, 200)))
names(subsets) <- files

read_nc <- function(f) {
   nc <- ReadNetCDF(f, c(hgt = unname(var[f])), subset = subsets[[f]])

   if (unname(var[f]) == "z") {
      nc[, hgt := hgt/9.8] %>% 
         setnames(longitude = "lon", latitude = "lat", level = "lev") %>% 
         .[]
   }

   if (unname(var[f]) == "hgt") {
      setnames(nc, level = "lev") %>% 
         .[]
   }
   return(nc)
}

datos <- lapply(files, read_nc) %>% 
   rbindlist(idcol = "dataset") %>% 
   .[, year := year(time[1]), by = time] %>% 
   .[month(time) == 12, year := year + 1] %>%  # december counts to the next year
   .[, .(hgt = mean(hgt), n = .N), by = .(dataset, lev, lat, lon, year, season(time))] %>% 
   .[n == 3] %>%  # keep only full seasons
   .[, n := NULL] %>% 
   .[]

# datos <- rbind(datos, ipsl, use.names = TRUE)

modern <- range(datos[dataset == "era", year])
# Surface mask
pres <- ReadNetCDF(here("datos", "NCEP Reanalysis", "pres.mon.mean.nc"), 
                   subset = list(lat = -90:0))
pres.mean <- pres[, .(pres = median(pres)), by = lat]
pres.mean <- rbind(data.table(lat = 0.0, pres = Inf), 
                   pres.mean, 
                   data.table(lat = -90.0, pres = Inf))

censor_ground <- function(df, col = "underground") {
   df[, c(col) := lev > pres.mean[lat == .BY$lat, pres] , by = lat][]
}

surface <- function(lats = c(-90:0)) {
   geom_polygon(data = pres.mean[lat %between% range(lats)], aes(y = pres), fill = "white", 
                alpha = 0.9, color = "gray10", size = 0.5)
}

labs_part <- c(R = "Real", I = "Imaginary", M = "Module")
```

<!-- ::: notes -->
<!-- Vamos a separar las anomalías de geopogential $Z'$ en su partes zonalmente simétricas y asimétricas -->

<!-- $$ -->
<!-- Z' = Z^{'*} + \left [Z' \right] -->
<!-- $$ -->

<!-- La varianza de $Z'$ entonces puede descomponserse en: -->

<!-- $$ -->
<!-- \mathrm{var}\left (Z' \right ) = \mathrm{var}\left (Z^{'*} \right) + \mathrm{var}\left ( \left [Z' \right ] \right ) + 2 \mathrm{cov}\left ( Z^{'*}, \left [ Z' \right ] \right) -->
<!-- $$ -->
<!-- ::: -->

<!-- ```{r} -->
<!-- datos <- datos %>% -->
<!--    .[, hgt_z := Anomaly(hgt), by = .(lat, lev, dataset, season, year)] %>%  -->
<!--    .[, prime := Anomaly(hgt), by = .(lon, lat, lev, dataset, season)] %>%  -->
<!--    .[, prime_mean := mean(prime), by = .(lat, lev, dataset, year, season)] %>%  -->
<!--    .[, prime_star := prime - prime_mean] -->
<!-- ``` -->

<!-- ```{r} -->
<!-- varianzas <- datos %>%  -->
<!--    .[, .(prime      = var(prime),  -->
<!--          prime_star = var(prime_star), -->
<!--          prime_mean = var(prime_mean), -->
<!--          cov2       = 2*cov(prime_star, prime_mean)),  -->
<!--      by = .(dataset, lev, lon, lat, season)] %>%  -->
<!--    # .[, prime_z := Anomaly(prime), by = .(dataset, season, lev, lat)] %>% -->
<!--    melt(id.vars = c("dataset", "lev", "lon", "lat", "season")) %>%  -->
<!--    .[, variable := factor(variable,  -->
<!--                           levels = c("prime", "prime_star", "prime_mean", "cov2"), -->
<!--                           labels = c("Z'", "Z'*", "[Z']", "2*cov(Z'*, [Z'])"))] -->
<!-- ``` -->

<!-- ```{r} -->
<!-- plot_var <- function(season, lev, dataset = "era") { -->
<!--    dataset_out <- dataset -->
<!--    out_season  <- season -->
<!--    out_lev     <- lev -->
<!--    varianzas %>% -->
<!--       ggperiodic::periodic(lon = c(0, 360)) %>%  -->
<!--       # .[lat < -15] %>%  -->
<!--       .[dataset == dataset_out] %>%  -->
<!--       .[season %in% out_season] %>% -->
<!--       .[lev %in% out_lev] %>% -->
<!--       ggplot(aes(lon, lat)) + -->
<!--       geom_contour_fill(aes(z = value), breaks = AnchorBreaks(0, NULL, 0)) + -->
<!--       map(lat < 0) + -->
<!--       # geom_contour(aes(z = value), color = "black") + -->
<!--       # geom_contour2(aes(z = value)) + -->
<!--       scale_fill_divergent(guide = guide_colorstrip_bottom()) + -->
<!--       scale_y_latitude(limits = c(-90, -20)) + -->
<!--       scale_x_longitude() + -->
<!--       coord_quickmap(ylim = c(NA, -30)) + -->
<!--       facet_nested(season ~ variable) + -->
<!--       # theme(legend.position = "right") -->
<!--       NULL -->
<!-- } -->

<!-- plot_vars <- function(lev, dataset = "era") { -->
<!--    seasons <- levels(varianzas$season) -->

<!--    lapply(seasons, plot_var, lev = lev, dataset = dataset) %>%  -->
<!--       Reduce("+", .) + -->
<!--       plot_layout(ncol = 1) + -->
<!--       plot_annotation(title = paste0(lev, "hPa")) -->
<!-- } -->

<!-- ``` -->

<!-- # División de varianza -->

<!-- --- -->

<!-- ```{r, fig.cap = "Descomposición de la varianza de Z' en 50hPa."} -->
<!-- # plot_vars(50) -->

<!-- ``` -->

<!-- ::: notes -->
<!-- La varianza de la parte zonalmente simétrica de geopotencial tiene siempre la misma estructura de máxima varianza en los polos. Esto creo que hay que agarrarlo con pinzas porque podría deberse más que nada a la dependencia con el parámetro de coriolis. Podría estar señalándo que es más razonable manejarse con la función corriente. Lo malo es que no encuentro dónde bajar la función corriente en ERA.  -->

<!-- En cualquier caso, la varianza de Z'* es máxima en latitudes polares y longitudinalmente magimiza al sur de sudamérica y al sur del Índic\o en todas las estaciones, prácticamente. Sin embargo, el campo de varianza total no tiene ese doble máximo salvo en JJA y la razón está en la estructura de la covarianza. En todas las estaciones hay un máximo de covariana negativa cerca de la península antártica que cancela el máximo de varianza de Z'\*. Esta estructura "bipolar" es interesante pero hay que tener cuidado. Por las fórmulas matemáticas, el promedio zonal de la covarianza de Z'\* y [Z^\*] tiene que ser nulo, por lo que es una certeza matemática que zonas con covarianza positiva tienen que estar compensadas por áreascon covarianza negativa.  -->
<!-- ::: -->

<!-- ```{r, fig.cap = "Descomposición de la varianza de Z' en 200hPa"} -->
<!-- # plot_vars(200) -->
<!-- ``` -->

<!-- ::: notes -->
<!-- En 200hPa hay algunos cambios. La estructura de varianza es tiene escalas espaciales más pequeñas. Esto es de esperarse dado que como las ondas cortas no se propagan verticalmente tanto como las cortas, toda la estructura de la estratósfera es más "suave" que la de la tropósfera. La varianza de [Z'] sigue siendo máxima en el polo sur, salvo en verano, que tiene un máximo en latitudes medias que probablemente corresponda a la variabilidad del jet subtropical. La varianza de Z*' maximiza en latitudes plares igual que en 5hPa, pero en vez de tener dos máximos, tiene uno solo que coincide con la Baja de Amundsen. En cuanto a la covarianza, sigue tieniendo un máximo negativo en el mar de Weddell.  -->

<!-- Para analizar con más detalle la relación entre Z y [Z] pero sin los probelmas matemáticos de la covarianza mostrada arriba, abajo muestro directamente la correlación entre Z (sin ningún filtro) y [Z]. -->
<!-- ::: -->

<!-- # EOF complejo `r paste(with(datos, common_range(year, dataset)), collapse = "-")` -->

<!-- --- -->

<!-- ```{r} -->
<!-- lats.eof <- c(-80, -20) -->
<!-- # ceof <- datos[lev %in% c(50, 200) & lat %between% lats.eof] %>% -->
<!-- #    # .[dataset != "ncep"] %>%  -->
<!-- #    .[year %between% common_range(year, dataset)] %>% -->
<!-- #    .[, hgt := Anomaly(hgt), -->
<!-- #      by  = .(lat, season, year, lev, dataset)] %>% -->
<!-- #    # .[, hgt := Anomaly(hgt), by = .(lon, lat, season, lev, dataset)] %>%  -->
<!-- #    .[, hgt := hgt/sd(hgt), by = .(lev, season, dataset)] %>% -->
<!-- #    .[, hgt := hgt*sqrt(cos(lat*pi/180))] %>% -->
<!-- #    .[, hgt.cpx := spectral::analyticFunction(hgt), -->
<!-- #      by = .(lat, season, year, dataset, lev)] %>%  -->
<!-- #    .[, hgt := hgt.cpx] %>% -->
<!-- #    .[, list(ceof = list(EOF(hgt ~ year | lon + lat, n = 1:2, suffix = "PC"))), -->
<!-- #      by = .(dataset, season, lev)] -->
<!-- ``` -->

<!-- ```{r} -->
<!-- plot_ceof <- function(gdata) { -->
<!--    gdata %>%     -->
<!--       ggperiodic::periodic(lon = c(0, 360)) %>%  -->
<!--       ggplot(aes(lon, lat)) + -->
<!--       geom_contour_fill(aes(z = abs(hgt)), binwidth = 0.005) + -->
<!--       map(lat < 10) + -->
<!--       geom_contour2(data =  ggperiodic::periodic(sep_ReIm(gdata, hgt), lon = c(0, 360)), -->
<!--                     breaks = AnchorBreaks(0, 0.01, 0), -->
<!--                     aes(z = hgt, linetype = factor(sign(..level..)), color = part), size = 0.2) + -->
<!--       scale_fill_distiller(palette = "Oranges", direction = 1,  -->
<!--                            guide = guide_colorstrip_bottom(),  -->
<!--                            breaks = MakeBreaks(binwidth = 0.005)) + -->
<!--       scale_x_longitude() + -->
<!--       scale_y_latitude() + -->
<!--       scale_linetype_manual(values = c("1" = 1, "-1" = 2), labels = c("1" = "+", "-1" = "-")) + -->
<!--       scale_color_manual(values = c(R = "black", I = "blue"), guide = "none") + -->
<!--       facet_nested(season ~ lev + PC, -->
<!--                    labeller = labeller(lev = AddSuffix(" hPa"))) + -->
<!--       coord_quickmap(ylim = c(NA, 0)) -->
<!-- } -->
<!-- # gdata <- ceof[, ceof[[1]]$right, by = .(dataset, season, lev)] %>%  -->
<!-- # .[dataset == "era"]  -->
<!-- # plot_ceof(gdata) -->
<!-- ``` -->

<!-- ::: notes -->
<!-- El PC1 en 50 hPa "va" con el PC2 en 200 tanto en estructura como en series temporales. Abajo se muestra las correlaciónes entre las series temporales de cada componente principal y cada nivel y para la parte imaginaria y real respectivamente. El PC2 en 50hPa tiene una alta correlación con el PC1 en 200hPa en JJA y SON. Entonce está claro que es el mismo modo que tiene una estructura barotrópica equivalente y que se propaga hacia o desde la estratósfera. En DJF y MAM la sincronización entre niveles se da más bien entre el PC1 de ambos niveles (aunque en menor medida). -->
<!-- ::: -->

<!-- ```{r, fig.cap = "Correlaciones entre componentes principales."} -->
<!-- # ceof[, ceof[[1]]$left, by = .(dataset, season, lev)] %>%  -->
<!-- #    .[dataset == "era"] %>%  -->
<!-- #    .[, c("R", "I") := ReIm(hgt)] %>%  -->
<!-- #    tidyr::pivot_longer(R:I) %>%  -->
<!-- #    as.data.table() %>%  -->
<!-- #    .[, item :=  interaction(lev, PC)] %>%  -->
<!-- #    .[, widyr::pairwise_cor(.SD, item, year, value), by = .(season, name)] %>% -->
<!-- #    ggplot(aes(item1, item2)) + -->
<!-- #    geom_raster(aes(fill = correlation)) + -->
<!-- #    geom_text(aes(label = format(correlation, digits = 1, nsmall = 1))) + -->
<!-- #    scale_fill_divergent(guide = "none") + -->
<!-- #    facet_nested(season ~ name)  -->
<!-- ``` -->

<!-- ::: notes -->
<!-- Esto motiva hacer el EOF con ambos niveles juntos de manera de obtener modos de oscilación conjuntos. Lo bueno de esto es que reduce la dimensión del problema ya que se reducen las series temporales a la mitad (porque hay una sola serie temporal para cada componente principal para los dos niveles).  -->
<!-- ::: -->

# EOF complejo conjunto

---

```{r define-funs}
compute_ceof <- function(dt, temporal = FALSE, lats.eof = c(-80, -20), n = 1:2) {
   # browser()
   dt <- dt[lev %in% c(50, 200) & lat %between% lats.eof] %>%
      .[, hgt := Anomaly(hgt),
        by  = .(lat, year, lev)]

   if (temporal) {
      dt <- dt[, hgt := Anomaly(hgt), by = .(lon, lat, lev)]
   }

   dt %>% 
      .[, hgt := hgt/sd(hgt), by = .(lev)] %>%
      .[, hgt := hgt*sqrt(cos(lat*pi/180))] %>%
      .[, hgt.cpx := spectral::analyticFunction(hgt),
        by = .(lat, year, lev)] %>% 
      .[, hgt := hgt.cpx] %>%
      EOF(hgt ~ year | lon + lat + lev, n = n, suffix = "PC", data = .)
}

plot_ceof <- function(data, eof = 1:2, formula = lev  ~  PC) {

   gdata <- data[, cut(ceof[[1]], eof)$right, by = .(dataset, season)]

   var <- data[, cut(ceof[[1]], eof)$sdev, by = .(dataset, season)] %>% 
      .[, setNames(paste0(PC, " (", scales::percent(r2), ")"), 
                   PC)]

   gdata %>%    
      ggperiodic::periodic(lon = c(0, 360)) %>% 
      ggplot(aes(lon, lat)) +
      geom_contour_fill(aes(z = Re(hgt)),
                        breaks = AnchorBreaks(0, 0.005, exclude = 0)) +
      geom_contour2(aes(z = Im(hgt),  linetype = factor(sign(..level..))),
                    breaks = AnchorBreaks(0, 0.005, exclude = 0)) +
      map(lat < 10) +
      scale_fill_divergent(guide = guide_colorstrip_bottom(35), 
                           breaks = AnchorBreaks(0, 0.005, exclude = 0),
                           limits = pm(0.039)) +
      scale_x_longitude() +
      scale_y_latitude() +
      scale_linetype_manual(values = c("1" = 1, "-1" = 2), labels = c("1" = "+", "-1" = "-"), 
                            guide = "none") +
      facet_nested(formula,
                   labeller = labeller(lev = AddSuffix(" hPa"),
                                       PC = var,
                                       dataset = labs_datasets)) +
      coord_quickmap(ylim = c(NA, -20)) +
      theme(legend.title = element_blank())
}

```

```{r calcula-ceof}
# Sacado de James
H <- 6.4*1000
lats.eof <- c(-80, -20)

year_range <- with(datos, common_range(year, dataset))

ceof <- datos %>% 
   .[year %between% year_range] %>%
   .[, list(ceof = list(compute_ceof(.SD, temporal = FALSE, n = 1:2))), by = .(dataset, season)]

long_ceof <- datos %>% 
   .[dataset == "era20" & season == "SON"] %>% 
   compute_ceof(temporal = FALSE)
```

```{r hilbert-example}
datos[dataset == "era" & season == "SON" & year == 1986] %>% 
   .[, hgt := Anomaly(hgt), by = .(lat, lev)] %>% 
   .[, H := spectral::analyticFunction(hgt), by = .(lat, lev)] %>%
   .[lev == 200] %>% 
   sep_ReIm(H, FALSE) %>% 
   ggperiodic::periodic(lon = c(0, 360)) %>% 
   ggplot(aes(lon, lat)) +
   geom_contour_fill(aes(z = R), 
                     breaks = AnchorBreaks(0, 25, 0)) +
   geom_contour2(aes(z = I, linetype = factor(sign(..level..))),
                 breaks = AnchorBreaks(0, 25, 0)) +
   map(lat <= 0) +
   scale_fill_divergent(guide = guide_colorstrip_bottom(), 
                        breaks = AnchorBreaks(0, 25, exclude = 0)) +
   scale_linetype_manual(values = c("1" = 1, "-1" = 2), labels = c("1" = "+", "-1" = "-"), 
                         guide = "none") +
   scale_x_longitude() +
   scale_y_latitude() +
   coord_quickmap() +
   theme(legend.title = element_blank())
```

```{r eof-espacial, fig.cap = "cEOF calculado usando la anomalía zonal. Parte real en sombreado, parte imaginaria en contornos (dataset = ERA Interim)."}
# gdata <- ceof[, ceof[[1]]$right, by = .(dataset, season)] %>% 
#    .[dataset == "era"] 
plot_ceof(ceof[season == "SON" & dataset == "era"]) 
```

```{r eof-espacial_comparacion, fig.cap = "PC2 calculado usando la anomalía zonal para distintos reanálisis. Parte real en sombreado, parte imaginaria en contornos."}
plot_ceof(ceof[season == "SON"], eof = 2, lev + PC ~  dataset) 
```

```{r ondas-planetarias, fig.cap = "Amplitud de las ondas zonales de cada componente principal."}
# ceof[, denormalize(ceof[[1]], "right"), by = .(dataset, season)] %>% 
ceof[, ceof[[1]]$right, by = .(dataset, season)] %>% 
   .[dataset == "era"] %>% 
   .[season == "SON"] %>% 
   # .[PC == "PC2"] %>% 
   .[, FitWave(Re(hgt), 1:3), by = .(lat, lev, dataset, PC, season)] %>% 
   ggplot(aes(lat, amplitude)) +
   geom_line(aes(color = factor(k))) +
   scale_y_continuous("Amplitude", expand = expansion(mult = c(0, 0.05))) +
   scale_x_latitude(ticks = 15) +
   scale_color_brewer("Wavenumber", palette = "Set1") +
   facet_nested(lev  ~  PC, 
                labeller = labeller(lev = AddSuffix(" hPa"))) +
   coord_flip()
```

```{r ceof-temporal-largo, fig.cap = "EOF temporal para ERA20", fig.width=18.7/2.45, fig.height=11.13/2.45}
long_ceof$left %>% 
   sep_ReIm(hgt) %>% 
   .[, hgt := scale(hgt), by = .(part, PC)] %>% 
   .[, part := forcats::fct_rev(part)] %>% 
   ggplot(aes(year, hgt)) +
   geom_zero(linetype = 1) +
   geom_line() +
   geom_smooth(span = 30/109, se = FALSE, method.args = list(degree = 1), n = 300,
               method = "loess", color = "#FD005F") +
   # geom_smooth(method = "lm", linetype = 2, color = "#FD005F", se = TRUE) +
   scale_color_brewer(palette = "Set1") +
   scale_x_continuous("Years", breaks = AnchorBreaks(1900, 20)) +
   scale_y_continuous("") +
   facet_nested(part ~ PC, labeller = labeller(anomaly = AddPreffix("Anomalía\n"),
                                               part = labs_part))
```

```{r ceof-temporal-largo-raw, fig.cap = "EOF temporal para ERA20", fig.width=18.7/2.45, fig.height=11.13/2.45}
long_ceof$left %>% 
   sep_ReIm(hgt) %>%
   # .[, hgt := scale(hgt), by = .(part, PC)] %>% 
   .[, part := forcats::fct_rev(part)] %>% 
   ggplot(aes(year, hgt)) +
   geom_zero(linetype = 1) +
   geom_line() +
   geom_smooth(span = 30/109, se = FALSE, method.args = list(degree = 1), n = 300,
               method = "loess", color = "#FD005F") +
   # geom_smooth(method = "lm", linetype = 2, color = "#FD005F", se = TRUE) +
   scale_color_brewer(palette = "Set1") +
   scale_x_continuous("Years", breaks = AnchorBreaks(1900, 20)) +
   scale_y_continuous("") +
   facet_nested(part ~ PC, labeller = labeller(anomaly = AddPreffix("Anomalía\n"),
                                               part = labs_part))
```

```{r ceof-temporal-largo-regr, fig.cap = "EOF temporal para ERA20", fig.width=18.7/2.45, fig.height=11.13/2.45}

long_ceof$left %>% 
   sep_ReIm(hgt) %>% 
   # .[, hgt := scale(hgt), by = .(part, PC)] %>% 
   .[, part := forcats::fct_rev(part)] %>% 
   # .[, hgt := scale(hgt), by = .(PC, part, anomaly)] %>% 
   ggplot(aes(year, hgt)) +
   geom_zero(linetype = 1) +
   geom_line(alpha = 0.3) +
   # geom_smooth(span = 30/109, se = FALSE, method.args = list(degree = 1), n = 300,
   #             method = "loess", color = "#FD005F") +
   geom_smooth(method = "lm", linetype = 1, color = "#7239b3", se = TRUE) +
   scale_color_brewer(palette = "Set1") +
   scale_x_continuous("Years", breaks = AnchorBreaks(1900, 20)) +
   scale_y_continuous("") +
   facet_nested(part ~ PC, labeller = labeller(anomaly = AddPreffix("Anomalía\n"),
                                               part = labs_part))
```

```{r, fig.width=18.7/2.45, fig.height=11.13/2.45}
linear_ceof <- copy(long_ceof)

linear_ceof$left <- linear_ceof$left %>% 
   sep_ReIm(hgt) %>% 
   .[, hgt := predict(lm(hgt ~ year)), by = .(PC, part)] %>% 
   dcast(year + PC ~ part) %>% 
   .[, hgt := complex(real = R, imaginary = I)] %>% 
   .[, `:=`(R = NULL, I = NULL)] %>% 
   .[]

trends <- lapply(list(PC1 = 1, PC2 = 2, "PC1+2" = 1:2), predict, object = linear_ceof) %>% 
   rbindlist(idcol = "PC")

long_sd <- datos[dataset == "era20" & season == "SON"] %>% 
   .[, hgt_a := Anomaly(hgt), by  = .(lat, year, lev)] %>% 
   .[, .(hgt_sd = sd(hgt_a)), by = .(lev)]

trends <- trends %>% 
   .[long_sd, on = .(lev)] %>% 
   .[, hgt := hgt*hgt_sd/sqrt(cos(pi/180*lat))]

plot_reconstruction <- function(pc = "PC1", binwidth) {
   trends[year %in% range(year)] %>% 
      # .[lev == 200] %>% 
      .[PC == pc] %>%
      ggperiodic::periodic(lon = c(0, 360)) %>% 
      ggplot(aes(lon, lat)) +
      geom_contour_fill(aes(z = Re(hgt)),
                        breaks = AnchorBreaks(0, binwidth, exclude = 0)) +
      # geom_contour2(aes(z = Re(hgt)), breaks = 30) +
      map(lat < 10) +
      scale_fill_divergent(guide = guide_colorstrip_bottom(),
                           breaks = AnchorBreaks(0, binwidth, exclude = 0)) +
      scale_x_longitude() +
      scale_y_latitude() +
      scale_linetype_manual(values = c("1" = 1, "-1" = 2), labels = c("1" = "+", "-1" = "-")) +
      facet_nested(lev  ~ PC + year,
                   labeller = labeller(lev = AddSuffix(" hPa"),
                                       PC = AddPreffix("Reconstruction with linear "),
                                       dataset = labs_datasets)) +
      coord_quickmap(ylim = c(NA, -20)) +
      theme(legend.title = element_blank())
}

plot_reconstruction("PC1", 50) 
```

```{r,  fig.width=18.7/2.45, fig.height=11.13/2.45}
plot_reconstruction("PC2", 5)

```

```{r}

trends_full <- datos %>% 
   .[dataset == "era20" & season == "SON"]  %>% 
   .[, hgt := Anomaly(hgt), by = .(lat, lev, year)] %>% 
   # .[, hgt := hgt/sd(hgt), by = .(lev)] %>%
   .[, FitLm(hgt, year, se = TRUE), by = .(lon, lat, lev)] %>% 
   .[term == "year"] %>% 
   .[, p.val := Pvaluate(estimate, std.error, df, "fdr"), by = .(lev)] %>% 
   .[, PC := "Full Data"]

trends_full %>% 
   .[lev == 200] %>% 
   ggplot(aes(lon, lat)) +
   geom_contour_fill(aes(z = estimate*109),
                     breaks = AnchorBreaks(0, NULL, exclude = 0)) +

   map(lat < 10) +
   scale_fill_divergent(guide = guide_colorstrip_bottom()) +
   # breaks = AnchorBreaks(0, 0.005, exclude = 0) +
   scale_x_longitude() +
   scale_y_latitude() +
   # scale_linetype_manual(values = c("1" = 1, "-1" = 2), labels = c("1" = "+", "-1" = "-")) +
   facet_nested(lev  ~ .,
                labeller = labeller(lev = AddSuffix(" hPa"),
                                    PC = AddPreffix("Reconstruction with lienar "),
                                    dataset = labs_datasets)) +
   coord_quickmap(ylim = c(NA, -20)) +
   theme(legend.title = element_blank())
```

```{r}
long_ceof$left %>% 
   sep_ReIm(hgt) %>% 
   .[, hgt := scale(hgt), by = .(part, PC)] %>% 
   .[, broom::tidy(lm(hgt ~year)), by = .(part, PC)]
```

```{r}
long_ceof$left %>% 
   copy() %>% 
   .[, `:=`(hgt = abs(hgt), 
            part = "M")] %>% 
   # .[, hgt := scale(hgt), by = .(part, PC)] %>%
   # .[, hgt := scale(hgt), by = .(PC, part, anomaly)] %>% 
   ggplot(aes(year, hgt)) +
   geom_zero(linetype = 1) +
   geom_line() +
   geom_smooth(span = 30/109, se = TRUE, method.args = list(degree = 1), n = 300,
               method = "loess") +
   scale_color_brewer(palette = "Set1") +
   facet_nested(part ~ PC, labeller = labeller(anomaly = AddPreffix("Anomalía\n"),
                                               part = labs_part))
```

```{r}
long_ceof$left %>% 
   copy() %>% 
   .[, `:=`(hgt = Arg(hgt), 
            part = "A")] %>% 
   # .[, hgt := scale(hgt), by = .(part, PC)] %>%
   # .[, hgt := scale(hgt), by = .(PC, part, anomaly)] %>% 
   ggplot(aes(year, hgt)) +
   geom_zero(linetype = 1) +
   geom_line() +
   geom_smooth(span = 30/109, se = TRUE, method.args = list(degree = 1), n = 300,
               method = "loess") +
   scale_color_brewer(palette = "Set1") +
   facet_nested(part ~ PC, labeller = labeller(anomaly = AddPreffix("Anomalía\n"),
                                               part = labs_part))
```

```{r}
long_ceof$left %>% 
   copy() %>% 
   .[, `:=`(hgt = abs(hgt), 
            part = "M")] %>% 
   .[, hgt := scale(hgt), by = .(part, PC)] %>% 
   .[, broom::tidy(lm(hgt ~year)), by = .(part, PC)] 
```

```{r}

my_spectrum <- function(data, spans = NULL, R = 1000, ..., probs = 0.95) {
   mtm <- spec.pgram(data, spans = spans, ..., plot = FALSE)

   out <- as.data.table(mtm[c("freq", "spec")])

   out[, boot_null := null_spec(data, spans = spans, R = R, ..., probs = probs)]

   return(out[])
}

null_spec <- memoise::memoise(function(data, spans, R = 1000, ..., probs = 0.95) {

   b <- boot::boot(data, function(d, i) spec.pgram(d[i], spans = spans, 
                                                   ...,
                                                   plot = FALSE)$spec, 
                   R = R)

   apply(b$t, 2, quantile, probs = probs)
}
)

long_ceof$left %>% 
   # .[PC == "PC2"] %>% 
   sep_ReIm(hgt) %>%
   .[, hgt := scale(hgt), by = .(PC, part)] %>%
   .[, my_spectrum(hgt, span = 5, R = 10000, detrend = TRUE, demean = TRUE),
     by = .(PC, part)] %>% 
   ggplot(aes(1/freq, spec, color = part)) +
   geom_line() +
   # geom_point() +
   geom_line(aes(y = boot_null), linetype = 2) +
   scale_x_log10("Period (years)", breaks = c(1:10, (2:10)*10), expand = c(0, 0),
                 guide = guide_axis(check.overlap = TRUE, n.dodge = 2)) +
   scale_y_continuous("Spectrum", expand = expansion(mult = c(0, 0.05), add = 0)) +
   annotation_logticks(sides = "b") +
   scale_color_brewer("Part", palette = "Set1", labels = labs_part) +
   facet_nested( ~ PC) 
```

```{r eof-temporales, fig.cap = "Series temporales de eofs"}
ceof[, ceof[[1]]$left, by = .(dataset, season)] %>% 
   .[season == "SON"] %>% 
   sep_ReIm(hgt) %>% 
   # .[, hgt := scale(hgt), by = .(PC, part, anomaly)] %>% 
   ggplot(aes(year, hgt)) +
   geom_zero(linetype = 1) +
   geom_line(aes(color = dataset)) +
   # geom_smooth(aes(color = dataset), span = 15/100, se = F, method.args = list(degree = 1), n = 100) +
   scale_color_brewer(palette = "Set1") +
   facet_nested(part ~ PC, labeller = labeller(anomaly = AddPreffix("Anomalía\n")))
```

```{r pc1}
erai <- copy(ceof) %>% 
   .[dataset == "era" & season == "SON"] %>% 
   .[, ceof[[1]]]
```

```{r read-temp-psi}
air <- ReadNetCDF(here::here("datos", "ERA-Interim", "erai.mon.mean.nc"),
                  vars = "t",
                  subset = list(latitude = -90:20,
                                level = list(850, 50))) %>% 
   setnames(longitude = "lon", latitude = "lat", level = "lev") %>% 
   .[, year := year(time[1]), by = time] %>% 
   .[month(time) == 12, year := year + 1L] %>%  # december counts to the next year
   .[, .(t = mean(t), n = .N), by = .(lev, lat, lon, year, season(time))] %>% 
   .[n == 3] %>%  # keep only full seasons
   .[, n := NULL] %>%
   .[season == "SON"]

psi <- ReadNetCDF("datos/NCEP Reanalysis/psi.mon.mean.nc",
                  subset = list(lat = c(-90, 20),
                                level = c(0.2101),
                                time = c("1979-01-01", "2019-01-01"))) %>% 
   setnames(level = "lev") %>% 
   .[, year := year(time[1]), by = time] %>% 
   .[month(time) == 12, year := year + 1L] %>%  # december counts to the next year
   .[, .(psi = mean(psi), n = .N), by = .(lev, lat, lon, year, season(time))] %>% 
   .[n == 3] %>%  # keep only full seasons
   .[, n := NULL] %>% 
   .[, psi.z := Anomaly(psi), by = .(lat, season, year, lev)]

sst <- ReadNetCDF("~/DATOS/ERA-Interim/erai-sfc.mon.mean.nc", 
                  vars = c("sst"), 
                  subset = list(latitude = -90:20)) %>% 
   na.omit() %>% 
   .[, year := year(time[1]), by = time] %>% 
   .[month(time) == 12, year := year + 1L] %>%  # december counts to the next year
   .[, .(sst = mean(sst), n = .N), by = .(latitude, longitude, year, season(time))] %>% 
   .[n == 3] %>%  # keep only full seasons
   .[, n := NULL] %>% 
   setnames(longitude = "lon", latitude = "lat") %>% 
   .[, sst_a := Anomaly(sst), by = .(lon, lat, season)]

```

```{r}

binwidths <- c(5e5, 1e6)
arrows <- c(0.01, 0.03)

```

```{r air-hgt-regr}
air_reg <- erai$left %>% 
   sep_ReIm(hgt)  %>% 
   .[, hgt := scale(hgt), by = .(part, PC)] %>% 
   .[air[season == "SON"], on = c("year"), allow.cartesian = TRUE] %>% 
   na.omit() %>%
   .[, FitLm(t, hgt, se = TRUE),  
     by = .(lon, lat, PC, part, lev)] %>% 
   .[term == "hgt"] %>% 
   .[, p.val := Pvaluate(estimate, std.error, df, "fdr"), by = .(PC, part, lev)]

air_mean <- air %>% 
   .[ lev == 50 & season == "SON"] %>% 
   .[, .(t = mean(t)), by = .(lon, lat, lev)]

hgt_reg <- erai$left %>% 
   copy() %>% 
   setnames(hgt = "EOF") %>% 
   sep_ReIm(EOF) %>% 
   .[, EOF := scale(EOF), by = .(part, PC)] %>%

   .[datos[dataset == "era" & season == "SON"], on = c("year"), allow.cartesian = TRUE] %>%
   .[lev %in% c(50, 200)] %>% 
   na.omit() %>% 
   .[, FitLm(hgt, EOF, se = TRUE), 
     by = .(lon, lat, lev, season, dataset, PC, part)] %>% 
   .[term == "EOF"] %>% 
   .[, p.val := Pvaluate(estimate, std.error, df, "fdr"), by = .(PC, part, lev)]

hgt_mean <- datos %>% 
   .[dataset == "era" & season == "SON"] %>% 
   .[, .(hgt = mean(hgt)), by = .(lon, lat, lev)]

psi_regr <- ceof[dataset == "ncep"] %>% 
   # .[, denormalize(ceof[[1]], "left"), by = .(season)] %>% 
   .[, ceof[[1]]$left, by = .(season)] %>%

   # .[PC == "PC2"] %>% 
   sep_ReIm(hgt) %>%
   .[, hgt := scale(hgt), by = .(part, PC, season)] %>% 
   .[psi, on = c("year", "season"), allow.cartesian = TRUE] %>% 
   na.omit() %>% 
   .[season == "SON"] %>% 
   .[, FitLm(psi.z, hgt = hgt/sd(hgt), se = TRUE), by = .(lat, lon, season, PC, part, lev)] %>% 
   .[term == "hgt"] %>%
   .[, psi.z := estimate] %>%
   .[, p.val := Pvaluate(estimate, std.error, df, "fdr"), by = .(part, PC, lev)] %>% 
   .[, c("f.lon", "f.lat") := WaveFlux(.SD), by = .(season, PC, part, lev)]

sst_regr <- ceof[season == "SON" & dataset == "era"] %>% 
   .[, ceof[[1]]$left] %>% 
   sep_ReIm(hgt) %>% 
   .[, hgt := scale(hgt), by = .(part, PC)] %>% 
   .[sst[season == "SON"], on = c("year"), allow.cartesian = TRUE] %>% 
   na.omit() %>% 
   .[, FitLm(sst, hgt, se = TRUE), by = .(lon, lat, part, PC)] %>% 
   .[term == "hgt"] %>% 
   .[, p.val := Pvaluate(estimate, std.error, df, "fdr"), by = .(PC, part)]

plot_reg <- function(reg, variable = "", component = TRUE, 
                     binwidth = NULL, geom_map = map(lat %between% c(-90, 0))) {
   reg$var <-  variable

   if (component) {
      component <- paste0(reg$PC[1], " - ", labs_part[reg$part[1]], " part")
   } else {
      component <- reg$PC[1]
   }

   ggperiodic::periodic(reg, lon = c(0, 360)) %>% 
      ggplot(aes(lon, lat)) +
      geom_contour_fill(aes(z = estimate), na.fill = TRUE, 
                        breaks = AnchorBreaks(0, binwidth, exclude = 0)) +
      geom_contour2(aes(z = p.val), breaks = 0.05, size = 0.2) +
      stat_subset(aes(subset = p.val <= 0.05 & is.cross(lon, lat, 1)), geom = "point", size = 0.01) +
      scale_x_longitude() +
      scale_y_latitude() +
      scale_fill_divergent(component, guide = guide_colorstrip_bottom(20),
                           labels = function(x) signif(x, 2),
                           breaks = AnchorBreaks(0, binwidth, exclude = 0)) +
      facet_nested(lev ~ var,
                   labeller = labeller(lev = AddSuffix(" hPa"))) +
      geom_map +
      coord_quickmap() 
   # theme(legend.title = element_blank())
}
```

# PC1

```{r PC1-regr-R, fig.height=11.13/2.45, fig.width = 23/2.45}
air_reg %>% 
   .[PC == "PC1"] %>% 
   .[part == "R"] %>% 
   .[lat <= 0] %>% 
   plot_reg("Temperature", binwidth = 0.5) +
   # geom_contour2(data = air_mean[lat <= 0], aes(z = t)) +

   hgt_reg %>% 
   .[PC == "PC1"] %>% 
   .[part == "R"] %>% 
   plot_reg("Z", binwidth = 20) 
# geom_contour2(data = hgt_m0an[lat <= 0], aes(z = hgt))
```

```{r PC1-regr-I, fig.height=11.13/2.45, fig.width = 23/2.45}
air_reg %>% 
   .[PC == "PC1"] %>% 
   .[part == "I"] %>% 
   .[lat <= 0] %>% 
   plot_reg("Temperature", binwidth = 0.5) +

   hgt_reg %>% 
   .[PC == "PC1"] %>% 
   .[part == "I"] %>% 
   plot_reg("Z", binwidth = 20)
```

```{r PC1-regr-psi, fig.height=11.13/2.45, fig.width = 23/2.45}
binwidth <- 5e5
psi_regr %>% 
   .[PC == "PC1"] %>% 
   .[, part := forcats::fct_rev(part)] %>% 
   .[, estimate := estimate/1e4] %>% 
   plot_reg(variable = "Streamfunction/1e4", binwidth = 30, geom_map = map(lat <= 20), 
            component = FALSE) +
   facet_nested(part ~ var,
                labeller = labeller(lev = AddSuffix(" hPa"),
                                    part = labs_part)) +
   geom_vector(aes(dx = f.lon, dy = f.lat,
                   x = ifelse(is.cross(lon, lat, 2), lon, NA)),
               skip = 0, min.mag = 0, size = 0.3, arrow.angle = 12) +
   scale_mag(max_size = 1, guide = "none")  +

   sst_regr %>% 
   .[PC == "PC1"] %>% 
   .[, part := forcats::fct_rev(part)] %>% 
   plot_reg(variable = "SST", binwidth = 0.1, geom_map = map(lat <= 20, fill = "white"),
            component = FALSE) +
   facet_nested(part ~ var,
                labeller = labeller(lev = AddSuffix(" hPa"),
                                    part = labs_part))

```

```{r ampl-eof-r2}
datos[season == "SON" & dataset == "era"] %>% 
   .[, FitWave(hgt, 1:3), by = .(dataset, lat, year, season, lev)] %>% 
   .[ceof[season == "SON" & dataset == "era", ceof[[1]]$left], on = c("year"), allow.cartesian = TRUE] %>% 
   .[, hgt := abs(hgt)] %>% 
   .[, cor(amplitude, hgt), by = .(lat, k, PC, lev)] %>%
   # .[PC == "PC2"] %>%
   ggplot(aes(lat, V1^2)) +
   # geom_zero() +
   annotate("rect", xmin = min(lats.eof), xmax = max(lats.eof), ymin = -Inf, ymax = Inf,
            color = NA, alpha = 0.1) +
   geom_line(aes(color = factor(k))) +
   # geom_line(aes(color = part)) +
   scale_x_latitude() +
   scale_y_continuous("R2 amplitud PC") +
   scale_color_brewer(palette = "Set1") +
   coord_flip() +
   # facet_nested(PC + lev ~ k) +
   facet_nested(lev ~ PC, labeller = labeller(lev = AddSuffix(" hPa")), scales = "free") +
   NULL
```

# PC2

```{r PC2-regr-R, fig.height=11.13/2.45, fig.width = 23/2.45}
air_reg %>% 
   .[PC == "PC2"] %>% 
   .[part == "R"] %>% 
   .[lat <= 0] %>% 
   plot_reg("Temperature", binwidth = 0.25) +

   hgt_reg %>% 
   .[PC == "PC2"] %>% 
   .[part == "R"] %>% 
   plot_reg("Z", binwidth = 10)
```

```{r PC2-regr-I, fig.height=11.13/2.45, fig.width = 23/2.45}
air_reg %>% 
   .[PC == "PC2"] %>% 
   .[part == "I"] %>% 
   .[lat <= 0] %>% 
   plot_reg("Temperature", binwidth = 0.1)  +

   hgt_reg %>% 
   .[PC == "PC2"] %>% 
   .[part == "I"] %>% 
   plot_reg("Z", binwidth = 10)
```

```{r PC2-regr-psi, fig.height=11.13/2.45, fig.width = 23/2.45}
psi_regr %>% 
   .[PC == "PC2"] %>% 
   .[, part := forcats::fct_rev(part)] %>% 
   .[, estimate := estimate/1e5] %>% 
   plot_reg(variable = "Streamfunction/1e5", binwidth = 10, geom_map = map(lat <= 20),
            component = TRUE) +
   facet_nested(part ~ var,
                labeller = labeller(lev = AddSuffix(" hPa"),
                                    part = labs_part)) +
   geom_vector(aes(dx = f.lon, dy = f.lat,
                   x = ifelse(is.cross(lon, lat, 2), lon, NA)),
               skip = 0, min.mag = 0, size = 0.2, arrow.angle = 12) +
   scale_mag(max_size = 1, guide = "none") +

   sst_regr %>% 
   .[, part := forcats::fct_rev(part)] %>% 
   .[PC == "PC2"] %>% 
   plot_reg(variable = "SST", binwidth = 0.25, geom_map = map(lat <= 20, fill = "white"),
            component = TRUE) +
   facet_nested(part ~ var,
                labeller = labeller(lev = AddSuffix(" hPa"),
                                    part = labs_part)) 
```

```{r}
rsoi::download_enso("oni") %>%
   as.data.table() %>% 
   .[, .(enso = mean(dSST3.4)), by = .(year(Date), season(Date))] %>% 
   .[season == "SON"] -> enso34

 erai$left %>% 
   sep_ReIm(hgt)  %>% 
    .[, part := forcats::fct_rev(part)] %>% 
   .[, hgt := scale(hgt), by = .(part, PC)] %>% 
   .[enso34, on = c("year"), allow.cartesian = TRUE] %>% 
   na.omit() %>% 
    ggplot(aes(hgt, enso)) +
    geom_point(aes(color = part)) +
    geom_smooth(aes(color = part), method = "lm") +
    scale_color_brewer(palette = "Set1") +
    scale_x_continuous("PC") +
    scale_y_continuous("ENSO34") +
    facet_grid(part ~ PC, labeller = labeller(part = labs_part)) 
```

```{r}

 erai$left %>% 
   sep_ReIm(hgt)  %>% 
   .[, hgt := scale(hgt), by = .(part, PC)] %>% 
   .[enso34, on = c("year"), allow.cartesian = TRUE] %>% 
   na.omit() %>% 
   .[, cor(hgt, enso)^2, by = .(PC, part)]
```

```{r}

# f <- cimadata::cmip_available("~/DATOS/CMIP6/")[-2, ]
# cmip_eof <- lapply(seq_len(nrow(f)), function(i) {
#    message("Processing ", i)
#    cmip <- ReadNetCDF(f[i, "file"], c(hgt = "zg"),
#                       subset = list(ensemble = 1,
#                                     lat = lats.eof,
#                                     time = c(paste0(year_range[1], "-01-01"), 
#                                              paste0(year_range[2], "-12-31")),
#                                     plev = list(50*100, 200*100))) %>% 
#       .[season(time) == "SON"] %>% 
#       data.table::setnames("plev", "lev") %>% 
#       .[, lev := lev/100] %>% 
#       .[, year := year(time[1]), by = time] %>% 
#       .[month(time) == 12, year := year + 1] %>%  # december counts to the next year
#       .[, .(hgt = mean(hgt), n = .N), by = .(lev, lat, lon, year, season(time))] %>% 
#       .[n == 3] %>%  # keep only full seasons
#       .[, n := NULL] %>% 
#       .[]
#    
#    cmip_eof <- cmip %>% compute_ceof()
#    remove(cmip)
#    
#    list(list(ceof = cmip_eof))
# }) 
# 
# names(cmip_eof) <- f[["source_id"]]
# 
# cmip_eof <- rbindlist(cmip_eof, idcol = "dataset")
# setnames(cmip_eof, V1 = "ceof")
# cmip_eof[, season := "SON"]
# 
# lapply(unique(cmip_eof$dataset), function(d) {
#    plot_ceof(cmip_eof[dataset == d])  +
#       labs(subtitle = d)
# }) %>% 
#    Reduce("+", .) +
#    plot_layout(ncol = 2)

```

---

#sdfsdf

```{r, fig.cap = "Regresión entre sst eofs (dataset = ERAI)."}

sst <- ReadNetCDF("~/DATOS/ERA-Interim/erai-sfc.mon.mean.nc", 
                  vars = c("sst"), 
                  subset = list(latitude = -90:20)) %>% 
   na.omit() %>% 
   .[, year := year(time[1]), by = time] %>% 
   .[month(time) == 12, year := year + 1] %>%  # december counts to the next year
   .[, .(sst = mean(sst), n = .N), by = .(latitude, longitude, year, season(time))] %>% 
   .[n == 3] %>%  # keep only full seasons
   .[, n := NULL] %>% 
   setnames(longitude = "lon", latitude = "lat") %>% 
   .[, sst_a := Anomaly(sst), by = .(lon, lat, season)]

ceof[season == "SON" & dataset == "era"] %>% 
   .[, ceof[[1]]$left] %>% 
   sep_ReIm(hgt) %>% 
   .[sst[season == "SON"], on = c("year"), allow.cartesian = TRUE] %>% 
   na.omit() %>% 
   .[, FitLm(sst, hgt, se = TRUE), by = .(lon, lat, part, PC)] %>% 
   .[term == "hgt"] %>% 
   .[, p.val := Pvaluate(estimate, std.error, df, "fdr"), by = .(PC, part)] %>% 
   .[] %>% 
   .[PC == "PC2"] %>% 
   ggplot(aes(lon, lat)) +
   # geom_raster(aes(fill = estimate)) +
   geom_contour_fill(aes(z = estimate), na.fill = TRUE, breaks = AnchorBreaks(0, NULL, exclude = 0)) +
   geom_contour2(aes(z = p.val), breaks = 0.05, size = 0.2) +
   stat_subset(aes(subset = p.val <= 0.05 & is.cross(lon, lat, 1)), geom = "point", size = 0.01) +
   # scale_fill_divergent(limits = c(-1, 1)) +
   scale_x_longitude() +
   scale_y_latitude() +
   scale_fill_divergent(guide = guide_colorstrip_bottom()) +
   facet_nested(part ~ PC,
                labeller = labeller(lev = AddSuffix(" hPa"))) +
   map(lat <= 20, fill = "white") +
   coord_quickmap()
```

```{r, fig.cap = "Composiciones de anomalías de SST para cada tercil del PC2"}
ceof[season == "SON" & dataset == "era"] %>% 
   .[, ceof[[1]]$left] %>% 
   .[PC == "PC1"] %>% 
   sep_ReIm(hgt) %>% 
   .[, hgt_cut := cut_number(hgt, 3, labels = c("Positive", "Neutral", "Negative")), by = .(part)] %>% 
   .[] %>% 
   .[sst[season == "SON"], on = c("year"), allow.cartesian = TRUE] %>% 
   .[] %>% 
   na.omit() %>% 
   .[, t.test(sst_a)[c("p.value", "estimate")], by = .(lon, lat, part, hgt_cut)] %>% 
   ggplot(aes(lon, lat)) +
   # geom_raster(aes(fill = sst)) +
   geom_contour_fill(aes(z = estimate), na.fill = TRUE, breaks = AnchorBreaks(0, 0.25, 0)) +
   geom_contour2(aes(z = p.value), breaks = 0.05, size = 0.2) +
   stat_subset(aes(subset = p.value <= 0.05 & is.cross(lon, lat, 1)), geom = "point", size = 0.01) +
   scale_x_longitude() +
   scale_y_latitude() +
   scale_fill_divergent(guide = guide_colorstrip_bottom(), breaks = AnchorBreaks(0, 0.25, 0)) +
   facet_nested(part ~ hgt_cut,
                labeller = labeller(lev = AddSuffix(" hPa"))) +
   map(lat <= 20, fill = "white") +
   coord_quickmap()
```

```{r, fig.cap = "Composiciones de anomalías de altura geopotential para cada tercil del PC2 en 200hPa y 50hPa."}
ceof[season == "SON" & dataset == "era"] %>% 
   .[, ceof[[1]]$left] %>% 
   .[PC == "PC2"] %>% 
   sep_ReIm(hgt) %>% 
   .[, eof_cut := cut_number(hgt, 3, labels = c("Positive", "Neutral", "Negative")), by = .(part)] %>% 
   .[, hgt := NULL] %>% 
   .[datos[season == "SON" & dataset == "era"], on = c("year"), allow.cartesian = TRUE] %>% 
   na.omit() %>% 
   .[, hgt := Anomaly(hgt), by = .(lon, lat, lev)] %>% 
   .[, t.test(hgt)[c("p.value", "estimate")], by = .(lon, lat, part, eof_cut, lev)] %>% 
   # .[lev == 200] %>%
   ggplot(aes(lon, lat)) +
   # geom_raster(aes(fill = sst)) +
   geom_contour_fill(aes(z = estimate), na.fill = TRUE, breaks = AnchorBreaks(0, NULL, 0)) +
   geom_contour2(aes(z = p.value), breaks = 0.05, size = 0.2) +
   stat_subset(aes(subset = p.value <= 0.05 & is.cross(lon, lat, 1)), geom = "point", size = 0.01) +
   scale_x_longitude() +
   scale_y_latitude() +
   scale_fill_divergent(guide = guide_colorstrip_bottom()) +
   facet_nested(lev + part ~ eof_cut,
                labeller = labeller(lev = AddSuffix(" hPa"))) +
   map(lat <= 20) +
   coord_quickmap()
```

```{r, fig.cap = "Composiciones de anomalías temporales de anomalía zonal de función corriente para cada tercil del PC2. "}
ceof[season == "SON" & dataset == "era"] %>% 
   .[, ceof[[1]]$left] %>% 
   .[PC == "PC2"] %>% 
   sep_ReIm(hgt) %>% 
   .[, eof_cut := cut_number(hgt, 3, labels = c("Positive", "Neutral", "Negative")), by = .(part)] %>% 
   .[, hgt := NULL] %>% 
   .[psi[season == "SON"], on = c("year"), allow.cartesian = TRUE] %>% 
   na.omit() %>% 
   .[, psi := Anomaly(psi.z), by = .(lon, lat)] %>% 
   .[, t.test(psi)[c("p.value", "estimate")], by = .(lon, lat, part, eof_cut)] %>% 
   .[, psi.z := estimate] %>%
   .[, c("f.lon", "f.lat") := WaveFlux(.SD), by = .(part, eof_cut)] %>%
   ggperiodic::periodic(lon = c(0, 360))  %>% 
   ggplot(aes(lon, lat)) +
   # geom_raster(aes(fill = sst)) +
   geom_contour_fill(aes(z = estimate), na.fill = TRUE, breaks = AnchorBreaks(0, NULL, 0)) +
   geom_contour2(aes(z = p.value), breaks = 0.05, size = 0.2) +
   stat_subset(aes(subset = p.value <= 0.05 & is.cross(lon, lat, 1)), geom = "point", size = 0.01) +
   geom_vector(aes(dx = f.lon, dy = f.lat,
                   x = ifelse(is.cross(lon, lat, 2), lon, NA)),
               skip = 0, min.mag = 0, size = 0.2, arrow.angle = 12) +
   scale_mag(max_size = 1) + 
   scale_x_longitude() +
   scale_y_latitude() +
   scale_fill_divergent(guide = guide_colorstrip_bottom()) +
   facet_nested( part ~ eof_cut,
                 labeller = labeller(lev = AddSuffix(" hPa"))) +
   map(lat <= 20) +
   coord_quickmap()
```

Ideas:

El PC2 está influenciado por enso, obviamente, pero sólo la parte Real. La parte imaginaria parece insensible al sst o la temperatura tropical.

Entonces: tenemos un patrón espacial de ondas planetarias tipo PSA que en una fase (la fase asociada a la parte Real) está asociada al enso y en la otra (la fase asociada a la parte imaginaria), no. La imagen qu ese me viene a la mente, entonces, es que el PSA (o al menos algo parecido al PSA) es relativamente independiente del ENSO cuando está en "fase imaginaria", pero el ENSO positivo o negativo mueve ese patrón a la "fase real".
Las composiciones de geopotencial parecen mostrar que las ondas planetarias son mucho más zonales en la fase imaginaria y tienen más propagación meridional durante la fase real. Aunque las composición de función corriente contradice eso.

# Extender antes de 1979

¿Se puede extender antes de 1979?

---

```{r}
start_years <- c(1979, 1950, 1900)
names(start_years) <- paste0(start_years, "-", max(datos[dataset == "era20" & season == "SON"]$year))

ceofs <- lapply(start_years, function(y) {
   datos[dataset == "era20" & season == "SON"] %>% 
      # .[, hgt := Detrend(hgt), by = .(lon, lat, lev)] %>% 
      .[year >= y] %>% 
      .[, .(ceof = list(compute_ceof(.SD, temporal = TRUE)))]
}) %>% 
   rbindlist(idcol = "period")
```

```{r, fig.cap = "Series temporales del EOFs obtenido a partir de distintos períodos."}
ceofs %>% 
   .[, ceof[[1]]$left, by = .(period)] %>% 
   sep_ReIm(hgt) %>% 
   ggplot(aes(year, hgt)) +
   geom_zero() +
   # geom_line(data = ceof[, ceof[[1]]$left, by = .(dataset, season)] %>%
   #              .[dataset == "era"] %>%
   #              .[season == "SON"] %>%
   #              # .[PC == "PC2"] %>%
   #              .[, .(year, hgt, PC, period = "ERA-I")] %>%
   #              sep_ReIm(hgt),
   #           aes(color = period), color = "black") +
   geom_line(aes(color = period)) +
   scale_x_continuous("Year") +
   scale_y_continuous("HGT") +
   scale_color_brewer(palette = "Set1") +
   facet_grid(part ~ PC)
```

```{r}
ceofs %>% 
   .[, ceof[[1]]$left, by = .(period)] %>% 
   copy() %>% 
   .[, `:=`(hgt = abs(hgt), part = "M")] %>% 
   ggplot(aes(year, hgt)) +
   geom_zero() +
   # geom_line(data = ceof[, ceof[[1]]$left, by = .(dataset, season)] %>%
   #              .[dataset == "era"] %>%
   #              .[season == "SON"] %>%
   #              # .[PC == "PC2"] %>%
   #              .[, .(year, hgt, PC, period = "ERA-I")] %>%
   #              sep_ReIm(hgt),
   #           aes(color = period), color = "black") +
   geom_line(aes(color = period)) +
   scale_x_continuous("Year") +
   scale_y_continuous("HGT") +
   scale_color_brewer(palette = "Set1") +
   facet_grid(part ~ PC)
```

```{r, fig.cap = "Patrones espaciales de las componentes principal para distintos períodos."}
ceofs %>% 
   .[, ceof[[1]]$right, by = .(period)] %>% 
   sep_ReIm(hgt, FALSE) %>% 
   ggplot(aes(lon, lat)) +
   # geom_contour_fill(aes(z = Re(hgt)), breaks = MakeBreaks(0.005)) +
   geom_contour_fill(aes(z = R),
                     breaks = AnchorBreaks(0, 0.005, exclude = 0)) +
   geom_contour2(aes(z = I,  linetype = factor(sign(..level..))),
                 breaks = AnchorBreaks(0, 0.005, exclude = 0)) +
   map(lat <= -20) +
   scale_fill_divergent(guide = guide_colorstrip_bottom(), breaks = AnchorBreaks(0, 0.005, exclude = 0)) +
   scale_x_longitude() +
   scale_y_latitude() +
   scale_linetype_manual(values = c("1" = 1, "-1" = 2)) +
   facet_nested(lev + period ~ PC,
                labeller = labeller(lev = AddSuffix(" hPa"))) +
   coord_quickmap()
```

```{r, fig.cap = "Serie tempral de eofs (dataset = ERA-20C)"}
long_eof <- copy(ceofs[period %like% "1900", ceof[[1]]])

long_eof$left %>% 
   sep_ReIm(hgt) %>% 
   # .[, hgt := scale(hgt), by = .(PC, part)] %>% 
   # .[, pw.lin := predict(SiZer::piecewise.linear(year, hgt)$model), by = .(PC, part)] %>% 
   ggplot(aes(year, hgt)) +
   geom_line() +
   # geom_line(aes(y = pw.lin)) +
   geom_smooth(method = "lm") +
   scale_x_continuous("Year") +
   scale_y_continuous("Hgt") +
   scale_color_brewer(palette = "Set1") +
   facet_nested(part ~ PC, scales = "free") 
```

```{r}
long_eof$left %>% 
   copy() %>% 
   .[, `:=`(hgt = Mod(hgt), part = "M")] %>% 
   # .[, hgt := scale(hgt), by = .(PC, part)] %>% 
   ggplot(aes(year, hgt)) +
   geom_line(aes(color = part)) +
   geom_smooth(method = "lm", aes(color = part)) +
   scale_x_continuous("Year") +
   scale_y_continuous("Hgt") +
   scale_color_brewer(palette = "Set1") +
   facet_nested(PC~., scales = "free") 
```

```{r, fig.cap = "Tendencia lineal para los campos de geopotencial reconstruidos a partir de la PC1, PC2 y ambas combinadas."}

long_sd <- datos[dataset == "era20" & season == "SON"] %>% 
   .[, hgt_a := Anomaly(hgt), by  = .(lat, year, lev)] %>% 
   .[, .(hgt_sd = sd(hgt_a)), by = .(lev)]

trends_full <- datos %>% 
   .[dataset == "era20" & season == "SON"]  %>% 
   .[, hgt := Anomaly(hgt), by = .(lat, lev, year)] %>% 
   .[, hgt := hgt/sd(hgt), by = .(lev)] %>%
   .[, FitLm(hgt, year, se = TRUE), by = .(lon, lat, lev)] %>% 
   .[term == "year"] %>% 
   .[, p.val := Pvaluate(estimate, std.error, df, "fdr"), by = .(lev)] %>% 
   .[, PC := "Full Data"]

trends <- lapply(list(PC1 = 1, PC2 = 2, "PC1+2" = 1:2), predict, object = long_eof) %>% 
   rbindlist(idcol = "PC") %>% 
   # .[long_sd, on = "lev"] %>% 
   .[, hgt := Re(hgt)/sqrt(cos(lat*pi/180))] %>% 
   .[, FitLm(hgt, year, se = TRUE), by = .(lon, lat, lev, PC)] %>% 
   .[term == "year"] %>% 
   .[, p.val := Pvaluate(estimate, std.error, df, "fdr"), by = .(PC, lev)] %>% 
   rbind(trends_full) %>% 
   .[, PC := factor(PC, levels = c("PC1", "PC2", "PC1+2", "Full Data"))]

trends %>% 
   .[lat %between% lats.eof] %>% 
   # .[PC != "PC1+2"] %>% 
   ggperiodic::periodic(lon = c(0, 360)) %>%
   ggplot(aes(lon, lat)) +
   geom_contour_fill(aes(z = estimate), breaks = AnchorBreaks(0, 0.002, 0)) + 
   map(lat <= -20) +
   # stat_subset(aes(subset = p.val <= 0.05 & is.cross(lon, lat)), geom = "point", size = 0.1) +
   scale_fill_divergent(guide = guide_colorstrip_bottom(), 
                        breaks = AnchorBreaks(0, 0.002, 0)) +
   scale_x_longitude() +
   scale_y_latitude() +
   coord_quickmap() +
   facet_nested(lev ~ PC, labeller = labeller(lev = AddSuffix(" hPa")))

# breaks = AnchorBreaks(0, 0.001, 0), 
```

```{r}

```

```{r, fig.cap = "Espectro de eof (dataset = ERA-20C). En linea punteada, el valor de 95% de significancia. "}
long_eof$left %>% 
   # .[PC == "PC2"] %>% 
   sep_ReIm(hgt) %>%
   .[, hgt := scale(hgt), by = .(PC, part)] %>%
   .[, my_spectrum(hgt, span = 3, R = 10000, detrend = FALSE, demean = FALSE),
     by = .(PC, part)] %>% 
   ggplot(aes(1/freq, spec*freq, color = part)) +
   geom_line() +
   # geom_point() +
   geom_line(aes(y = boot_null*freq), linetype = 2) +
   scale_x_log10("Period (years)", breaks = c(1:10, (2:10)*10), expand = c(0, 0)) +
   scale_y_continuous(expand = c(0, 0)) +
   annotation_logticks(sides = "b") +
   scale_color_brewer(palette = "Set1") +
   facet_nested( ~ PC)
```

```{r}
long_eof$left %>% 
   copy() %>% 
   # .[PC == "PC2"] %>% 
   .[, hgt := abs(hgt)] %>% 
   .[, part := "M"] %>% 
   .[, hgt := scale(hgt), by = .(PC, part)] %>%
   .[, my_spectrum(hgt, spans = c(5), R = 10000, detrend = TRUE, demean = FALSE),
     by = .(PC, part)] %>% 
   ggplot(aes(1/freq, spec*freq, color = part)) +
   geom_line() +
   # geom_point() +
   geom_line(aes(y = boot_null*freq), linetype = 2) +
   scale_x_log10("Period (years)", breaks = c(1:10, (2:10)*10), expand = c(0, 0)) +
   scale_y_continuous(expand = c(0, 0)) +
   annotation_logticks(sides = "b") +
   scale_color_brewer(palette = "Set1") +
   facet_nested( ~ PC)
```

```{r}
long_eof$left %>% 
   # .[PC == "PC2"] %>% 
   sep_ReIm(hgt) %>%
   .[, hgt := scale(hgt), by = .(PC, part)]  %>% 
   .[PC == "PC1" & part == "R" , hgt]  -> s
```

### 2021-01-05 R/0-download.R

@@ -1,0 +1,118 @@
+download_cds <- function(request) {
+  force(request)
+  function(file) {
+    user <- Sys.getenv("CDSUSER", NA)
+    key <- Sys.getenv("CDSKEY", NA)
+    filename <- basename(file)
+
+    if (is.na(user) | is.na(key)) {
+      stop("Please set the environmental variables CDSUSER and CDSKEY")
+    }
+
+    request$target <- filename
+
+    ecmwfr::wf_set_key(user = user,
+                       key = key,
+                       service = "cds")
+
+    files <- ecmwfr::wf_request(request = request,
+                                user = user,
+                                time_out = 5*3600,
+                                path = dirname(file))
+    file
+  }
+}
+
+
+download_webapi <- function(request) {
+  force(request)
+  function(file) {
+    user <- Sys.getenv("WEBAPIUSER", NA)
+    key <- Sys.getenv("WEBAPIKEY", NA)
+    filename <- basename(file)
+
+    if (is.na(user) | is.na(key)) {
+      stop("Please set the environmental variables CDSUSER and CDSKEY")
+    }
+
+    request$target <- filename
+
+    ecmwfr::wf_set_key(user = user,
+                       key = key,
+                       service = "webapi")
+
+    files <- ecmwfr::wf_request(request = request,
+                                user = user,
+                                time_out = 5*3600,
+                                path = dirname(file))
+    file
+  }
+}
+
+
+request_era5be <- list(
+  format = "netcdf",
+  product_type = "reanalysis-monthly-means-of-daily-means",
+  variable = c("geopotential", "temperature", "vorticity"),
+  pressure_level = c("1", "2", "3", "5", "7", "10", "20", "30", "50", "70", "100", "125", "150", "175", "200", "225", "250", "300", "350", "400", "450", "500", "550", "600", "650", "700", "750", "775", "800", "825", "850", "875", "900", "925", "950", "975", "1000"),
+  year = as.character(1950:1978),
+  month = c("01", "02", "03", "04", "05", "06", "07", "08", "09", "10", "11", "12"),
+  time = "00:00",
+  grid = c("2.5", "2.5"),
+  dataset_short_name = "reanalysis-era5-pressure-levels-monthly-means-preliminary-back-extension"
+)
+
+
+request_era5 <- list(
+  format = "netcdf",
[…]

### 2021-01-05 R/ReIm.R

@@ -1,0 +1,46 @@
+#' Splits complex numbers into real and imaginary vectors
+#'
+#' @param complex vector of complex values
+#' @param data a data.table
+#' @param column unquoted name of the column to split into real and imaginary part
+#' @param format Whether to return a long or wide dataset.
+#'
+#' @export
+ReIm <- function(complex) {
+  list(Real = Re(complex), Imaginary = Im(complex))
+}
+
+#' @export
+#' @rdname ReIm
+sep_ReIm <- function(data, column, format = c("longer", "wider")) {
+  R <- part <- I <- NULL
+  names <- c("R", "I")
+
+
+  if (missing(column)) {
+    complex <- vapply(data, function(x) inherits(x, "complex"), TRUE)
+    if (sum(complex) > 1) {
+      stop("`column` missing and more than one complex column found")
+    }
+    if (sum(complex) == 0) {
+      warning("`column` missing and no complex column found. Returning unchanged data")
+      return(data)
+    }
+
+    col <- colnames(data)[complex]
+  } else {
+    col <- deparse(substitute(column))
+  }
+
+
+  data <- data.table::copy(data)[, (names) := ReIm(get(col))]
+
+
+  if (format[1] == "longer") {
+    data[, c(col) := NULL]
+    data <- data.table::setDT(tidyr::pivot_longer(data, R:I, names_to = "part", values_to = col))
+    data[, part := factor(part, levels = c("R", "I"), labels = c("Real", "Imaginary"))]
+  }
+
+  return(data[])
+}

### 2021-01-05 R/WaveFlux.R

@@ -1,0 +1,14 @@
+#' @export
+WaveFlux <- function(psi, p = 250, a = 6371000) {
+  k <- p*100/(a^2*2000)
+  psi <- copy(psi)
+  psi[, c("psi.dlon", "psi.dlat") := Derivate(psi.z ~ lon + lat,
+                                              cyclical = c(TRUE, FALSE))] %>%
+    .[, psi.ddlon := Derivate(psi.z ~ lon, cyclical = TRUE, order = 2),
+      by = lat] %>%
+    .[, psi.dlondlat := Derivate(psi.dlon ~ lat),
+      by = lon] %>%
+    .[, `:=`(f.lon = k/cos(lat*pi/180)*(psi.dlon^2 - psi.z*psi.ddlon),
+             f.lat = k*(psi.dlon*psi.dlat - psi.z*psi.dlondlat))]
+  list(f.lon = psi$f.lon, f.lat = psi$f.lat)
+}

### 2021-01-05 R/common-range.R

@@ -1,0 +1,15 @@
+#' @export
+common_range <- function(x, groups) {
+  group_order <- grouping(groups)
+  group_bounds <- c(0, attr(group_order, "ends"))
+
+  x_ordered <- x[group_order]
+  common_range <- c(-Inf, Inf)
+  for (i in seq_along(group_bounds)[-1]) {
+    new_range <- range(x_ordered[(group_bounds[i-1] + 1):group_bounds[i]])
+    common_range <- c(max(common_range[1], new_range[1]),
+                      min(common_range[2], new_range[2]))
+
+  }
+  common_range
+}

### 2021-01-05 R/data_locations.R

@@ -1,0 +1,87 @@
+#' @export
+print.nc_file <- function(x, ...) {
+   cat("NetCDF file.\nLocation:",
+       as.character(x), "\n\nDetails:\n")
+   print(metR::GlanceNetCDF(x))
+}
+
+#' Data locations
+#'
+#' Functions that return the location of different datasets. They check that
+#' relevant file exist and then return it with a class that has a neat print method.
+#'
+#'
+#' @param type character indicating the type of data. The resulting base will be type_data
+#' @param ... characters to be contactenadted with [file.path()]
+#' @export
+#' @rdname data_locations
+data_path <- function(type = c("raw", "derived"), ...) {
+   here::here("analysis", "data", paste0(type[1], "_data"), ...)
+}
+
+
+data_nc_fun <- function(file, download) {
+   force(file)
+
+   if (is.character(download)) {
+      download <- simple_download(download)
+   }
+
+   function(force_download = FALSE) {
+      if (force_download || !file.exists(file)) {
+         file <- download(file)
+      }
+
+      checkmate::assert_access(file, access = "r")
+      class(file) <- c("nc_file", class(file))
+      return(file)
+   }
+}
+
+
+#' @param force_download Logical indicating wether to force download of data.
+#' @export
+#' @rdname data_locations
+ERA5 <- data_nc_fun(file = data_path("raw", "era5.mon.mean.nc"),
+                    download = download_cds(request_era5))
+
+
+#' @param force_download Logical indicating wether to force download of data.
+#' @export
+#' @rdname data_locations
+ERA5_BE <- data_nc_fun(file = data_path("raw", "era5be.mon.mean.nc"),
+                       download = download_cds(request_era5be))
+
+
+#' @export
+#' @rdname data_locations
+ERA5_SST <- data_nc_fun(file = data_path("raw", "era5.sst.mon.mean.nc"),
+                        download = download_cds(request_era5_sst))
+
+#' @export
+#' @rdname data_locations
+ERA20C <- data_nc_fun(file = data_path("raw", "era20c.mon.mean.nc"),
+                      download = download_webapi(request_era20c))
+
+
+#' @export
+#' @rdname data_locations
+NCEP_PSI <- data_nc_fun(file = data_path("raw", "psi.mon.mean.nc"),
[…]

### 2021-01-05 R/facet_nested.R

@@ -1,0 +1,416 @@
+library(grid)
+library(gtable)
+#' @title Layout panels in a grid with nested strips
+#'
+#' @description \code{facet_nest()} form a matrix of panels defined by row and
+#'   column faceting variables and nests grouped facets.
+#'
+#' @param
+#' rows,cols,scales,space,shrink,labeller,as.table,switch,drop,margins,facets
+#' see \code{\link[ggplot2]{facet_grid}}
+#' @param nest_line a \code{logical} vector of length 1, indicating whether to
+#'   draw a nesting line to indicate the nesting of variables.
+#' @param resect  a \code{unit} vector of length 1, indicating how much the
+#'   nesting line should be shortened.
+#' @param bleed a \code{logical} vector of length 1, indicating wether merging
+#'   of lower-level variables is allowed when the higher-level variables are
+#'   seperate. See details.
+#'
+#' @details Unlike \code{facet_grid()}, this function only automatically expands
+#'   missing variables when they have no variables in that direction, to allow
+#'   for unnested variables. It still requires at least one layer to have all
+#'   faceting variables.
+#'
+#'   Hierarchies are inferred from the order of variables supplied to
+#'   \code{rows} or \code{cols}. The first variable is interpreted to be the
+#'   outermost variable, while the last variable is interpreted to be the
+#'   innermost variable. They display order is always such that the outermost
+#'   variable is placed the furthest away from the panels. Strips are
+#'   automatically grouped when they span a nested variable.
+#'
+#'   The \code{bleed} argument controls wether lower-level variables are allowed
+#'   to be merged when higher-level are different, i.e. they can bleed over
+#'   hierarchies. Suppose the \code{facet_grid()} behaviour would be the
+#'   following:
+#'
+#'   \code{[_1_][_2_][_2_]} \cr \code{[_3_][_3_][_4_]}
+#'
+#'   In such case, the default \code{bleed = FALSE} argument would result in the
+#'   following:
+#'
+#'   \code{[_1_][___2____]} \cr \code{[_3_][_3_][_4_]}
+#'
+#'   Whereas \code{bleed = TRUE} would allow the following:
+#'
+#'   \code{[_1_][___2____]} \cr \code{[___3____][_4_]}
+#'
+#' @export
+#'
[…]

### 2021-01-05 R/factories.R

@@ -1,0 +1,18 @@
+#' Function factories for adding preffixes or Suffixes
+#'
+#' @param suffix,preffix character that will be pasted pefore or after the string
+#'
+#' @export
+AddSuffix <- function(suffix = "") {
+  function(string) {
+    paste0(string, suffix)
+  }
+}
+
+#' @export
+#' @rdname AddSuffix
+AddPreffix <- function(preffix = "") {
+  function(string) {
+    paste0(preffix, string)
+  }
+}

### 2021-01-05 R/filter.R

@@ -1,0 +1,9 @@
+#' @export
+is.sa <- function(data, lons = c(265, 325), lats = c(-80, -10)) {
+  data[, lon %between% lons & lat %between% lats]
+}
+
+#' @export
+filter_sa <- function(data, lons = c(265, 325), lats = c(-60, -10)) {
+  data[is.sa(data, lons = lons, lats = lats)]
+}

### 2021-01-05 R/geom-tstudent.R

@@ -1,0 +1,66 @@
+#' Draws a violin plot representing a T-Student distribution
+#'
+#'
+#'
+#' @export
+geom_tstudent <- function (mapping = NULL, data = NULL, stat = StatTStudent, position = "dodge",
+                           na.rm = FALSE, show.legend = NA, inherit.aes = TRUE, orientation = NA,
+                           df = 30,...)
+{
+  layer(data = data, mapping = mapping, stat = stat, geom = GeomViolin,
+        position = position, show.legend = show.legend, inherit.aes = inherit.aes,
+        params = list(na.rm = na.rm, df = df, orientation = orientation, ...))
+}
+
+
+#' @export
+StatTStudent <- ggproto("StatTStudent", Stat,
+                        required_aes = c("x|y", "mean", "sd"),
+                        optional_aes = c("df"),
+                        extra_params = c("na.rm", "orientation"),
+                        setup_params = function(data, params) {
+                          params$flipped_aes <- ggplot2::has_flipped_aes(data, params, main_is_orthogonal = FALSE,
+                                                                         main_is_continuous = TRUE)
+                          has_x <- !(is.null(data$x) && is.null(params$x))
+                          has_y <- !(is.null(data$y) && is.null(params$y))
+                          if (!has_x && !has_y) {
+                            stop("stat_tstudent() requires an x or y aesthetic.")
+                          }
+
+                          params
+                        },
+                        compute_group = function(self, data, scales, df = 30, confidence = .95, flipped_aes = FALSE) {
+
+                          if (is.null(data$df)) {
+                            data$df <- df
+                          }
+
+                          alpha <- (1-confidence)/2
+                          q <- qt(c(alpha, 1-alpha), data$df)
+                          q <- seq(min(q), max(q), length.out = 40)
+
+                          density <- dt(q, data$df)
+
+                          data.frame(y = q*data$sd + data$mean, x = data$x,
+                                     density = density/data$sd,
+                                     scaled = density/max(density))
+
+
+                        },
+                        compute_panel = function(self, data, scales, df = 30, confidence = .95, flipped_aes = FALSE) {
+
+                          data <- ggplot2::flip_data(data, flipped_aes)
+                          data <- ggproto_parent(Stat, self)$compute_panel(
+                            data, scales, df = df, confidence = confidence, flipped_aes = flipped_aes
+                          )
+
+                          # choose how violins are scaled relative to each other
[…]

### 2021-01-05 R/invert-laplacian.R

@@ -1,0 +1,33 @@
+#' @export
+invert_laplacian <- function(v, x, y) {
+  a <- 6371000
+  field <- data.table::data.table(x, y, v)
+
+  field[, c("k", "l", "v.hat") := fft2d(v, x, y)]
+  field[, lp := l*2*pi/a]
+  field[, kp := k*2*pi/(a*cos(y*pi/180))]
+
+  field[, v.hat := -v.hat/(k^2 + l^2)]
+  # field[, v.hat := -v.hat/(kp^2 + lp^2)]
+  field[, v.hat := c(0, v.hat[2:.N])]
+
+  field[, c("k", "l", "v.inv") := fft2d(v.hat, k, l, inverse = TRUE)]
+
+  return(Re(field$v.inv))
+}
+
+fft2d <- function(v, x, y, inverse = FALSE) {
+
+  field <- as.matrix(reshape2::dcast(data.table::data.table(v, x, y),
+                                     y ~ x, value.var = "v"))
+
+  field.fft <- fft(field[, -1], inverse = inverse)
+
+  dimnames(field.fft) <- list(
+    k = 1:nrow(field.fft) - 1,
+    l = 1:ncol(field.fft) - 1)
+
+  field.fft <- setDT(melt(field.fft, value.name = "f"))
+
+  with(field.fft, list(k, l, f))
+}

### 2021-01-05 R/jacobi.R

@@ -1,0 +1,241 @@
+if (FALSE) {
+lapl <- function(y, div)   {
+
+  ddx <- t(apply(y, 1, function(x) metR:::.derv(x, matrix$coldims$lat,
+                                                order = 2, cyclical = FALSE, fill = TRUE)))
+  ddy <- apply(y, 2, function(x) metR:::.derv(x, matrix$rowdims$lon, order = 2, cyclical = TRUE))
+  lap <- ddx + ddy + div
+
+
+  return(dY)
+}
+library(metR)
+data <- "~/DATOS/NCEP Reanalysis/psi.mon.mean.nc" %>%
+  ReadNetCDF(subset = list(level = 0.1682,
+                           time = c("2020-01-01"))) %>%
+  .[, psi := Anomaly(psi), by = lat]
+
+
+
+
+laplacian <- function(x, lon, lat) {
+  xbk <- x
+  x <- x$matrix
+  prev <- function(N) c(N, seq_len(N)[-N])
+  nex <- function(N) c(seq_len(N)[-1], 1)
+
+  lon <- lon*pi/180
+  lat <- (lat+90)*pi/180
+
+  dlon <- diff(lon)[1]
+  dlat <- diff(lat)[1]
+
+  Nrow <- nrow(x)
+  Ncol <- ncol(x)
+
+  sinlat <- sin(lat)
+  sinplus <- sin(lat + dlat/2)
+  sinminus <- sin(lat - dlat/2)
+
+  lap <- 1/dlat^2*(sinplus*(x[nex(Nrow), ] - x) - sinminus*(x - x[prev(Nrow), ])) +
+    1/(dlon^2*sinlat^2)*(-2*x + x[, nex(Ncol)] + x[, prev(Ncol)])
+
+
+  # lap[c(1, Nrow), ] <- lap[c(2, Nrow-1), ]
+  xbk$matrix <- lap
+  xbk
+}
+
+jacob_laplacian <- function(x, lon, lat, forcing) {
+  xbk <- x
+  x <- x$matrix
+  prev <- function(N) c(N, seq_len(N)[-N])
+  nex <- function(N) c(seq_len(N)[-1], 1)
+
+  lon <- lon*pi/180
+  lat <- (lat+90)*pi/180
+
+  dlon <- diff(lon)[1]
+  dlat <- diff(lat)[1]
+
+  Nrow <- nrow(x)
+  Ncol <- ncol(x)
+
[…]

### 2021-01-05 R/lm_lite.R

@@ -1,0 +1,85 @@
+#' Fast linear regression
+#'
+#' Computes regression coefficient quickly for the case
+#' of simple lienar regression using the simple
+#' covariance/variance formula.
+#'
+#' @param y dependent variable
+#' @param x independent variable
+#' @param weights vector of weights
+#' @param r2 logical indicating wether to compute r.quared
+#'
+#' @export
+lm_lite <- function(y, x, weights = rep(1, length(y)), r2 = FALSE) {
+  cov <- stats::cov.wt(cbind(y, x), wt = weights, cor = r2)
+  estimate <- cov$cov[1, 2]/var.wt(x, w = weights)
+  out <- list(term = deparse(substitute(x)),
+              estimate = estimate)
+
+  if (r2) {
+    out$r.squared <-  cov$cor[1, 2]^2
+    out$adj.r.squared <- cov$cor[1, 2]^2   # Not real!!
+  }
+  return(out)
+}
+
+
+#' Weighted vaiance
+#' @keywords internal
+var.wt <- function(x, w, na.rm = FALSE) {
+  if (na.rm) {
+    w <- w[i <- !is.na(x)]
+    x <- x[i]
+  }
+  sum.w <- sum(w)
+  (sum(w*x^2) * sum.w - sum(w*x)^2) / (sum.w^2 - sum(w^2))
+}
+
+#' @export
+Detrend <- function(y, x) {
+   nas <- is.na(y)
+   m <- mean(y, na.rm = TRUE)
+   if (!hasArg(x)) x <- seq_along(y)
+   y[!nas] <- .lm.fit(cbind(1, x[!nas]), y[!nas])$residuals
+   return(y + m)
+}
+
+#' Compute p.values
+#'
+#' Computes p.values from the result of [metR::FitLm()] (or any assumed student-t-distributed
+#' statistic) with optional adjustments.
+#'
+#' @param estimate estimate of the regression (or other fit)
+#' @param std.error standard error
+#' @param df degrees of freedom
+#' @param adjustment method for adjustment, see [stats::p.adjust()].
+#'
+#' @export
+Pvaluate <- function(estimate, std.error, df, adjustment = "none") {
+  stats::p.adjust(2*stats::pt(abs(estimate)/std.error, df, lower.tail = FALSE), method = adjustment)
+}
+
+
+#' Remove intercept
+#'
+#' Removes intercept term from the result of [metR::FitLm()]. More to the point,
[…]

### 2021-01-05 R/maps.R

@@ -1,0 +1,56 @@
+#' Quick maps
+#'
+#'
+#' @param subset a function that will be applied to the map data.
+#' @param color,size,fill parameters passed to [ggplot2::geom_polygon()]
+#' @param wrap vector of length 2 interpreted as the longitude range of a global map (see [maps::map()])
+#' @param keep proportions of point to retain (passed to [rmapshaper::ms_simplify()])
+#' @param ... other arguments passed to [ggplot2::geom_polygon()]
+#'
+#' @export
+geom_qmap <- function(subset = identity, color = "gray50", size = 0.2,
+                      fill = NA, wrap = c(0, 360), weighting = 0.7,
+                      keep = 0.015, ...) {
+  lon <- lat <- group <- NULL
+  data <- map_simple(wrap = wrap, keep  = keep, weighting = weighting)
+
+  data <- data %>%
+    fortify() %>%
+    data.table::as.data.table() %>%
+    .[, c("long", "lat", "group")] %>%
+    data.table::setnames("long", "lon")
+  subset <- purrr::as_mapper(subset)
+  data <- subset(data)
+
+  ggplot2::geom_polygon(data = data,
+                        ggplot2::aes(lon, lat, group = group),
+                        color = color,
+                        size = size,
+                        fill = fill,
+                        ...)
+}
+
+map_simple <- function(wrap = c(0, 360), keep = 0.015, weighting = 0.7) {
+  map <- maps::map("world", fill = TRUE,
+                   col = "transparent", plot = FALSE, wrap = wrap)
+  IDs <- vapply(strsplit(map$names, ":"), function(x) x[1],
+                "")
+  proj <- sp::CRS("+proj=longlat +datum=WGS84")
+  map <- maptools::map2SpatialPolygons(map, IDs = IDs,
+                                       proj4string = proj)
+
+  if (keep != 1) {
+    map <- rmapshaper::ms_simplify(map, keep = keep, weighting = weighting)
+  }
+  map
+}
+
+fortify <- ggplot2::fortify
+
+# zzz.R
+.onLoad <- function(pkgname, libname) {
+  map_simple <<- memoise::memoise(map_simple)
+  fortify <<- memoise::memoise(fortify)
+}
+
+

### 2021-01-05 R/partial_cor.R

@@ -1,0 +1,28 @@
+#' Partial correlation between x1 and y and x2 and y.
+#'
+#' @param y variable to which partial correlation will be computed
+#' @param x1,x2 variables that will be correlated with y
+# #' @param weights vector of weights
+#'
+#' @export
+partial_cor <- function(y, x1, x2) {
+
+
+  estimate1 <- ppcor::pcor.test(y, x1, x2)
+  estimate2 <- ppcor::pcor.test(y, x2, x1)
+
+
+  # rho_x1x2 <- stats::cov.wt(cbind(x1, x2), wt = weights, cor = TRUE)$cor[1, 2]
+  # rho_x1y <- stats::cov.wt(cbind(x1, y), wt = weights, cor = TRUE)$cor[1, 2]
+  # rho_x2y <- stats::cov.wt(cbind(x2, y), wt = weights, cor = TRUE)$cor[1, 2]
+  #
+  # estimate1 <- (rho_x1y - rho_x2y*rho_x1x2 ) / (sqrt(1 - rho_x2y^2) * sqrt(1 - rho_x1x2^2))
+  # estimate2 <- (rho_x2y - rho_x1y*rho_x1x2 ) / (sqrt(1 - rho_x1y^2) * sqrt(1 - rho_x1x2^2))
+
+  list(term = c(deparse(substitute(x1)),
+                deparse(substitute(x2))),
+       partial_correlation = c(estimate1$estimate, estimate2$estimate),
+       p.value = c(estimate1$p.value, estimate2$p.value)
+  )
+}
+

### 2021-01-05 R/solve-poisson.R

@@ -1,0 +1,85 @@
+#' @useDynLib shceof hwsssp_
+#' @export
+solve_poisson <- function(forcing, lon, lat) {
+  data <- data.table::data.table(forcing = forcing,
+                                 lon = lon,
+                                 lat = lat)
+
+  forcing <- metR:::.tidy2matrix(data, lat ~ lon, value.var = "forcing")
+  pi <- 4*atan(1)
+
+  f <- forcing$matrix
+
+  colat <- (90 - forcing$rowdims$lat)*pi/180
+  lon <- forcing$coldims$lon*pi/180
+
+  M <- length(colat) - 1L
+  N <- length(lon) - 1L
+
+  TS <- as.single(min(colat))
+  TF <- as.single(max(colat))
+
+  TF_is_pi <- isTRUE(all.equal(as.single(pi), TF, check.attributes = FALSE))
+  TS_is_zero <- isTRUE(all.equal(0, TS, check.attributes = FALSE))
+
+  if (TF_is_pi & TS_is_zero) {
+    # Unspecified boudnary conditions
+    MBDCND <- 9L
+  } else if (TF_is_pi & !TS_is_zero) {
+    # BC unspecified at TF and derivative specified at TS
+    MBDCND <- 8L
+  } else if (!TF_is_pi & TS_is_zero) {
+    # BC unspecified at TS and derivative specified at TF
+    MBDCND <- 6L
+  } else {
+    # Derivative of the solution specifided at TS and TF
+    MBDCND <- 3L
+  }
+
+  # Derivatives at the poles
+  BDTS <- as.single(rep(0., N+1))
+  BDTF <- as.single(rep(0., N+1))
+
+  PS <- as.single(min(lon))
+  PF <- as.single(max(lon))
+
+  # Periodic boundary condition
+  NBDCND <- 0L
+
+  BDPS <- 1L  # dummy value
+  BDPF <- 1L  # dummy value
+
+  # lalbda = 0 to get poisson equation
+  ELMBDA <- as.single(0.)
+
+  f <- as.single(f)
+
+
+  IDIMF <- M+1L
+
+  # Working memory
+  W <- rep(0., 4*(N+1) + (16 + (ceiling(log2(N+1))))*(M+1))
+
+  PERTRB <- 1L
+  IERROR <- 0L
+
+
+  result <- .Fortran("hwsssp_",
+                     TS,
+                     TF, M,MBDCND,BDTS,BDTF,
[…]

### 2021-01-05 R/spectrum.R

@@ -1,0 +1,57 @@
+#' FFT spectrum
+#'
+#' A light wrapper around [stats::spec.pgram()] and [stats::ar()]. Returns the spectrum of
+#' the signal and the spectrum of the fitted autoregressiv model.
+#'
+#' @param x numeric vector
+#' @param B number of bootstrap samples.
+#' @param spans vector of odd integers giving the widths of modified Daniell
+#' smoothers to be used to smooth the periodogram
+#' @param ... other arguments passed to [stats::spec.pgram()]
+#'
+#' @export
+fftspectrum <- function(x, spans = NULL, B = 10000, ..., probs = 0.95) {
+  mtm <- spec.pgram(x, spans = spans, ..., plot = FALSE)
+
+  out <- as.data.table(mtm[c("freq", "spec")])
+
+  ar <- ar(ts(x))
+  # rho <- a$ar
+  # var <- a$var.pred
+  out[, ar_spectrum := arspectrum(mtm$freq, ar$ar, ar$var.pred)]
+  out[, c(scales::percent(probs)) := null_ar_spectrum(B = B, length(x), ar, spans = spans, ..., probs = probs)]
+
+  return(out[])
+}
+
+arspectrum <- function(freq, rho, var) {
+  k <- seq_along(rho)
+  e <- vapply(freq, function(f) sum(rho * exp(-2*1i*pi*f*k)), complex(1))
+  var / (Mod(1 - e))^2
+}
+
+
+
+null_ar_spectrum_ <- function(B = 100, n, ar, spans = NULL, ..., probs = 0.95) {
+  y <- as.vector(arima.sim(model = list(ar = ar$ar), n = n))*sqrt(ar$var.pred)
+  nfreq <- length(spec.pgram(y, spans = spans, ..., plot = FALSE)$spec)
+  boots <- vapply(seq_len(B), function(b) {
+    y <- as.vector(arima.sim(model = list(ar = ar$ar), n = n))*sqrt(ar$var.pred)
+    spec.pgram(y, spans = spans, plot = FALSE)$spec
+
+  }, numeric(nfreq))
+  data.table::as.data.table(t(apply(boots, 1, quantile, probs = probs)))
+}
+
+null_ar_spectrum <- memoise::memoise(null_ar_spectrum_)
+
+null_spec <- memoise::memoise(function(x, spans, B = 1000, ..., probs = 0.95) {
+
+  b <- boot::boot(x, function(d, i) spec.pgram(d[i], spans = spans,
[…]

### 2021-01-05 R/theme.R

@@ -1,0 +1,28 @@
+#' Main theme used in the paper
+#'
+#' Based on [ggplot2::theme_minimal()] but with minimal (wink) modifications.
+#'
+#'
+#' @param base_size base font size, given in pts
+#'
+#' @export
+theme_shceof <- function(base_size = 16) {
+  ggplot2::theme_minimal(base_size = base_size) +
+    ggplot2::theme(
+      # strip.background = ggplot2::element_rect(fill = NA, color = "gray30"),
+      # text = element_text(family = font_rc),
+      legend.position = "bottom", legend.box = "vertical",
+      tagger.panel.tag.background = ggplot2::element_rect(color = NA),
+      strip.text = ggplot2::element_text(size = 12),
+      # panel.spacing.y = grid::unit(5, "mm"),
+      # panel.spacing.x = grid::unit(5, "mm"),
+      # legend.spacing = grid::unit(2, "mm"),
+      # panel.border = ggplot2::element_rect(colour = "black", fill = NA),
+      # plot.margin = grid::unit(rep(3, 4), "mm"),
+      # # legend.title = element_blank(),
+      # legend.box.spacing = grid::unit(3, "mm"),
+      # legend.margin = ggplot2::margin(t = -5),
+      panel.grid = ggplot2::element_blank(),
+      # panel.grid = ggplot2::element_line(color = "gray10", size = 0.4, linetype = 3),
+      panel.ontop = TRUE)
+}

### 2021-01-05 R/zzz.R

@@ -1,0 +1,3 @@
+globalVariables(c(":=", ".", ".NATURAL"))
+
+.datatable.aware <- TRUE

### 2021-01-05 analysis/scripts/00-download_era.R

@@ -1,0 +1,47 @@
+library(lubridate)
+library(ecmwfr)
+
+# Todo esto no sirve! Baja ERA interim y hay que bajar ERA5
+
+levs <- c(10, 20, 30, 50, 70, 100, 150, 200, 250, 300, 400, 500, 600, 700, 850,
+          925, 1000)
+
+dates <- seq.Date(as.Date("1979-01-01"), as.Date("2018-12-31"), "1 month")
+dates <- paste0(year(dates),
+                formatC(month(dates), width = 2, flag = "0"),
+                formatC(day(dates), width = 2, flag = "0"),
+                collapse = "/")
+
+hgt_request <- list(
+  class    = "ei",
+  dataset  = "interim",
+  date     = dates,
+  expver   = "1",
+  grid     = "2.5/2.5",
+  levelist = levs,
+  levtype  = "pl",
+  param    = "129.128",
+  stream   = "moda",
+  type     = "an",
+  format   = "netcdf",
+  target   = "hgt.mon.mean.nc"
+)
+
+u_request <- list(
+  class    = "ei",
+  dataset  = "interim",
+  date     = dates,
+  expver   = "1",
+  grid     = "2.5/2.5",
+  levelist = levs,
+  levtype  = "pl",
+  param    = "131.128",
+  stream   = "moda",
+  type     = "an",
+  format   = "netcdf",
+  target   = "uwnd.mon.mean.nc"
+)
+
+path <- here::here("datos", "ERA-Interim")
+wf_request(hgt_request, job_name = "hgt_era", path = path)
+wf_request(u_request, job_name = "u_era", path = path)

### 2021-01-05 analysis/scripts/01-compute_uv.R

@@ -1,0 +1,25 @@
+subset = list(lat = -90:10)
+levels = c(50, 200, 500)
+
+years <- seq(1979, 2018)
+
+uv <- data.table::rbindlist(lapply(years, function(y) {
+  cat("Processing year ", y, "\r")
+  ufile <- paste0("datos/NCEP Reanalysis/daily/uwnd.", y, ".nc")
+  vfile <- paste0("datos/NCEP Reanalysis/daily/vwnd.", y, ".nc")
+  
+  wnd <- metR::ReadNetCDF(ufile, "uwnd", subset = subset)
+  wnd[, vwnd := metR::ReadNetCDF(vfile, "vwnd", subset = subset, out = "vector")[[1]]]
+  wnd <- wnd[level %in% levels]
+  
+  wnd[, year := year(time[1]), by = time]
+  wnd[, season := metR::season(time[1]), by = time]
+  
+  wnd[, .(uv = cov(uwnd, vwnd)), by = .(level, lon, lat, year, season)]
+}))
+
+data.table::setnames(uv, "level", "lev")
+uv[, dataset := "ncep"]
+cat("Saving data.")
+saveRDS(uv, "datos/uv.Rds")
+

### 2021-01-05 analysis/scripts/01-get-data.R

@@ -1,0 +1,24 @@
+library(magrittr)
+library(shceof)
+
+force <- FALSE
+
+# Download CMAP -----------------------------------------------------------
+CMAP(force)
+
+# Download HadSST -----------------------------------------------------------
+HADSST(force)
+
+# Download Streamfunction NCEP ----------------------------------------------------
+NCEP_PSI(force)
+
+# Download Vorticity NCEP ----------------------------------------------------
+NCEP_VOR(force)
+
+# Download ERA5 -----------------------------------------------------------
+ERA5(force)
+ERA5_BE(force)
+ERA5_SST(force)
+
+# Download ERA20C -----------------------------------------------------------
+ERA20C(force)

### 2021-01-05 analysis/scripts/01-preprocess_data.R

@@ -1,0 +1,54 @@
+library(metR)
+library(data.table)
+library(magrittr)
+library(here)
+
+
+setnames <- function(x, ...) {
+  names <- c(...)
+  # print(names)
+  data.table::setnames(x, names(names), unname(names))
+}
+
+
+subset_ncep <- list(lat = -90:0,
+                    time = c("1979-01-01", "2018-12-31"))
+subset_era <- list(latitude = -90:0,
+                   time = c("1979-01-01", "2018-12-31"))
+
+ncep <- ReadNetCDF(here("datos", "NCEP Reanalysis", "hgt.mon.mean.nc"),
+                   subset = subset_ncep) %>%
+  .[, dataset := "ncep"] %>% 
+  setnames(., level = "lev") %>% 
+  .[lev %in% c(200, 50)]
+
+era <- ReadNetCDF(here("datos", "ERA-Interim", "hgt.mon.mean.nc"),
+                  c(hgt = "z"),
+                  subset = subset_era) %>%
+  .[, hgt := hgt/9.8] %>%
+  .[, dataset := "era"] %>%
+  setnames(longitude = "lon", latitude = "lat", level = "lev") %>% 
+  .[lev %in% c(200, 50)]
+
+era20 <- rbind(ReadNetCDF(here("datos", "ERA-20C", "hgt.mon.mean.nc"),
+                          c(hgt = "z"),
+                          subset = list(latitude = -90:0, 
+                                        level = 50)),
+               ReadNetCDF(here("datos", "ERA-20C", "hgt.mon.mean.nc"),
+                          c(hgt = "z"),
+                          subset = list(latitude = -90:0, 
+                                        level = 200))) %>% 
+  .[, hgt := hgt/9.8] %>%
+  .[, dataset := "era20"] %>%
+  setnames(longitude = "lon", latitude = "lat", level = "lev")
+
+datos <- rbind(era20, era, ncep)
+# datos <- ncep
+# remove(ncep)
+remove(era, ncep, era20)
+
+datos <- datos[, .(hgt = mean(hgt)), 
+               by = .(lev, lon, lat, year(time), season(time), dataset)]
+
+
+saveRDS(datos, here("datos", "datos.Rds"))

### 2021-01-05 analysis/scripts/helperfun.R

@@ -1,0 +1,1689 @@
+## Archivo con funciones
+
+#Librerías
+# library(ggplot2)
+# library(ggforce)
+# library(stringi)
+# # library(ggthemes)
+# library(magrittr)
+# library(data.table)
+# library(lubridate)
+# library(akima)
+
+# library(RColorBrewer)
+library(compiler)
+enableJIT(0)
+
+# source(here::here("scripts/eof_methods.R"))
+
+`%b%` <- function(x, y) data.table::`%between%`(x, y) 
+
+# Mapa
+BuildMap <- function(res = 1, smooth = 1, pm = 180,
+                     countries = FALSE, ...) {
+   # Caraga datos de mapas y cambia el meridiano principal, suaviza y cambia
+   # resolución para plotear más rápido.
+   # Entra:
+   #   res: resolución (cuantos puntos se eliminan entre los que quedan)
+   #   smooth: factor de suavifzado (ventana del promedio corrido)
+   #   pm: longitud del meridiano central
+   #   countries: ¿datos a nivel país?
+   # Sale:
+   #   un data.table con las coordenadas de cada polígono y su grupo
+   library(data.table)
+   if (countries) {
+      library(rworldxtra)
+      data("countriesLow")
+      m <- countriesLow
+   } else {
+      library(rworldmap)
+      data(coastsCoarse)
+      m <- coastsCoarse
+   }
+   
+   m <- as.data.table(fortify(m))
+   m[, group := as.numeric(group)]
+   m[, id := as.numeric(id)]
+   
+   # Cambio el prime meridian.
+   m2 <- copy(m)
+   m2[, long := long + 360]
+   m2[, group := group + max(group) + 1]
+   m <- rbind(m, m2)
+   m <- m[long >= pm - 180 & long <= 180 + pm]
+   
+   m[, MO := max(order), by = group]
+   
+   # Suavizo.
+   if (smooth > 1) {
+      cut <- max(smooth, res)
+      notsmooth <- m[, .N, by = group][N < cut + 4, group]
+      m <- m[!(group %in% notsmooth)]    # saco los grupos muy chicos
+      
+      m[order != 1 & order != MO,
+        `:=`(long = zoo::rollmean(long, smooth, fill = "extend"),
+             lat = zoo::rollmean(lat, smooth, fill = "extend")),
+        by = group]
+   }
+   # Bajo la resolución.
[…]

### 2021-01-05 analysis/scripts/preprocess_data.R

@@ -1,0 +1,47 @@
+library(metR)
+library(data.table)
+library(magrittr)
+library(here)
+
+
+setnames <- function(x, ...) {
+  names <- c(...)
+  # print(names)
+  data.table::setnames(x, names(names), unname(names))
+}
+
+
+subset_ncep <- list(lat = -90:0,
+                    time = c("1979-01-01", "2018-12-31"))
+subset_era <- list(latitude = -90:0,
+                   time = c("1979-01-01", "2018-12-31"))
+
+ncep <- ReadNetCDF(here("datos", "NCEP Reanalysis", "hgt.mon.mean.nc"),
+                   subset = subset_ncep) %>% 
+  .[, u := ReadNetCDF(here("datos", "NCEP Reanalysis", "uwnd.mon.mean.nc"),
+                      out = "vector", 
+                      subset = subset_ncep)] %>% 
+  .[, dataset := "ncep"] %>% 
+  setnames(., level = "lev")
+
+era <- ReadNetCDF(here("datos", "ERA-Interim", "hgt.mon.mean.nc"),
+                  c(hgt = "z"),
+                  subset = subset_era) %>%
+  .[, hgt := hgt/9.8] %>%
+  .[, u := ReadNetCDF(here("datos", "ERA-Interim", "uwnd.mon.mean.nc"),
+                      c(u = "u"),
+                      out = "vector",
+                      subset = subset_era)] %>%
+  .[, dataset := "era"] %>%
+  setnames(longitude = "lon", latitude = "lat", level = "lev")
+
+datos <- rbind(era, ncep)
+# datos <- ncep
+# remove(ncep)
+remove(era, ncep)
+
+datos <- datos[, .(u = mean(u), hgt = mean(hgt)), 
+               by = .(lev, lon, lat, year(time), season(time), dataset)]
+
+
+saveRDS(datos, "datos/datos.Rda")

### 2021-01-05 src/comf.f

@@ -1,0 +1,186 @@
+C
+C     file comf.f
+C
+C     * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * *
+C     *                                                               *
+C     *                  copyright (c) 1999 by UCAR                   *
+C     *                                                               *
+C     *       UNIVERSITY CORPORATION for ATMOSPHERIC RESEARCH         *
+C     *                                                               *
+C     *                      all rights reserved                      *
+C     *                                                               *
+C     *                      FISHPACK version 4.1                     *
+C     *                                                               *
+C     *     A PACKAGE OF FORTRAN SUBPROGRAMS FOR THE SOLUTION OF      *
+C     *                                                               *
+C     *      SEPARABLE ELLIPTIC PARTIAL DIFFERENTIAL EQUATIONS        *
+C     *                                                               *
+C     *                             BY                                *
+C     *                                                               *
+C     *        JOHN ADAMS, PAUL SWARZTRAUBER AND ROLAND SWEET         *
+C     *                                                               *
+C     *                             OF                                *
+C     *                                                               *
+C     *         THE NATIONAL CENTER FOR ATMOSPHERIC RESEARCH          *
+C     *                                                               *
+C     *                BOULDER, COLORADO  (80307)  U.S.A.             *
+C     *                                                               *
+C     *                   WHICH IS SPONSORED BY                       *
+C     *                                                               *
+C     *              THE NATIONAL SCIENCE FOUNDATION                  *
+C     *                                                               *
+C     * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * *
+C
+C
+C PACKAGE COMF           THE ENTRIES IN THIS PACKAGE ARE LOWLEVEL       
+C                        ENTRIES, SUPPORTING FISHPACK ENTRIES BLKTRI
+C                        AND CBLKTRI. THAT IS, THESE ROUTINES ARE       
+C                        NOT CALLED DIRECTLY BY USERS, BUT RATHER       
+C                        BY ENTRIES WITHIN BLKTRI AND CBLKTRI.          
+C                        DESCRIPTION OF ENTRIES EPMACH AND PIMACH       
+C                        FOLLOW BELOW.                                  
+C                                                                       
+C LATEST REVISION        NOVEMBER 1988
+C                                                                       
+C SPECIAL CONDITIONS     NONE                                           
+C                                                                       
+C I/O                    NONE                                           
+C                                                                       
+C PRECISION              SINGLE                                         
[…]

### 2021-01-05 src/gnbnaux.f

@@ -1,0 +1,314 @@
+C
+C     file gnbnaux.f
+C
+C     * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * *
+C     *                                                               *
+C     *                  copyright (c) 1999 by UCAR                   *
+C     *                                                               *
+C     *       UNIVERSITY CORPORATION for ATMOSPHERIC RESEARCH         *
+C     *                                                               *
+C     *                      all rights reserved                      *
+C     *                                                               *
+C     *                      FISHPACK version 4.1                     *
+C     *                                                               *
+C     *     A PACKAGE OF FORTRAN SUBPROGRAMS FOR THE SOLUTION OF      *
+C     *                                                               *
+C     *      SEPARABLE ELLIPTIC PARTIAL DIFFERENTIAL EQUATIONS        *
+C     *                                                               *
+C     *                             BY                                *
+C     *                                                               *
+C     *        JOHN ADAMS, PAUL SWARZTRAUBER AND ROLAND SWEET         *
+C     *                                                               *
+C     *                             OF                                *
+C     *                                                               *
+C     *         THE NATIONAL CENTER FOR ATMOSPHERIC RESEARCH          *
+C     *                                                               *
+C     *                BOULDER, COLORADO  (80307)  U.S.A.             *
[…]

### 2021-01-05 src/hwsssp.f

@@ -1,0 +1,780 @@
+C
+C     file hwsssp.f
+C
+      SUBROUTINE HWSSSP_(TS,TF,M,MBDCND,BDTS,BDTF,PS,PF,N,NBDCND,BDPS,
+     1                   BDPF,ELMBDA,F,IDIMF,PERTRB,IERROR,W)
+C
+C     * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * *
+C     *                                                               *
+C     *                  copyright (c) 1999 by UCAR                   *
+C     *                                                               *
+C     *       UNIVERSITY CORPORATION for ATMOSPHERIC RESEARCH         *
+C     *                                                               *
+C     *                      all rights reserved                      *
+C     *                                                               *
+C     *                      FISHPACK version 4.1                     *
+C     *                                                               *
+C     *     A PACKAGE OF FORTRAN SUBPROGRAMS FOR THE SOLUTION OF      *
+C     *                                                               *
+C     *      SEPARABLE ELLIPTIC PARTIAL DIFFERENTIAL EQUATIONS        *
+C     *                                                               *
+C     *                             BY                                *
+C     *                                                               *
+C     *        JOHN ADAMS, PAUL SWARZTRAUBER AND ROLAND SWEET         *
+C     *                                                               *
+C     *                             OF                                *
+C     *                                                               *
+C     *         THE NATIONAL CENTER FOR ATMOSPHERIC RESEARCH          *
+C     *                                                               *
+C     *                BOULDER, COLORADO  (80307)  U.S.A.             *
+C     *                                                               *
+C     *                   WHICH IS SPONSORED BY                       *
+C     *                                                               *
+C     *              THE NATIONAL SCIENCE FOUNDATION                  *
+C     *                                                               *
[…]

### 2021-01-05 analysis/paper/paper.Rmd
Muchos cambios

@@
 guide_colorsteps_bottom <- function(width = 25, height = 0.5, ...) {
-  guide_colorsteps(title.position = "top", title.hjust = 0.5,
+  guide_colorsteps(show.limits = TRUE, even.steps = FALSE,
+                   title.position = "top", title.hjust = 0.5,
                    barheight = height,
@@
 
+```{r test-irlba}
+has_irlba <- requireNamespace("rilba", quietly = TRUE)
+if (has_irlba) {
+  stop("If installed, metR::EOF uses irlba, but that package has a bug with complex matrices\n(see: https://github.com/bwlewis/irlba/issues/55).
+Uninstall irlba with renv::remove(\"irlba\") and try again.")
+}
+
+```
+
 ```{r read-data}
@@
 
-```{r}
+```{r compute-eof-naive}
 eof_naive <- era5[lev %in% c(50, 200) & lat %between% c(-90, -20)] %>% 
@@
 
-```{r}
+```{r define-plot_ceof}
 plot_ceof <- function(data, n = 1:2, formula = lev  ~  PC) {
@@
 The real part of the first EOF is, therefore, capturing the mean zonally anomalous field *and the variability that projects onto it*.
-
-```{r, fig.cap = "Correlación entre el las anomalías zonales de geopotential de cada mes y el campo medio de anomalías zonales de geopotencial (negro) y correlación (inversa) entre anomalías zonales de altura geopotencial y la parte real del EOF1 en (rojo) entre 80S y 20S."}
-# cor_mean <- era5 %>% 
-#   .[lev %in% c(50, 200)] %>% 
-#   .[lat <= -20] %>% 
-#   # .[lat >= -80] %>% 
-#   .[, hgt := Anomaly(hgt), by = .(time, lat, lev)] %>% 
-#   .[, hgt_mean := mean(hgt), by = .(lat, lon, lev)] %>% 
-#   .[, .(mean = cor(hgt_mean, hgt)), by = .(lev, time)] %>%
-#   identity()
-# 
-# 
-# ceof$eof[[1]] %>% 
-#   .$right %>% 
-# 
-#   .[PC == "PC1"] %>% 
-#   sep_ReIm(format = "wide") %>% 
-#   .[, hgt := NULL] %>% 
-#   .[, PC := NULL] %>% 
-#   era5[., on = .NATURAL] %>% 
-#   .[, hgt := Anomaly(hgt), by = .(time, lat, lev)] %>% 
-#   .[, .(hgt = -cor(hgt, R)), by = .(lev, time)] %>% 
-# 
-#   ggplot(aes(time, hgt))  +
-#   geom_line(color = "red") +
-#   geom_line(data = cor_mean, aes(y = mean)) +
-#   facet_grid(lev~.)
-
-```
 
@@
 
-```{r}
+```{r compute-regression}
 regr <- ceof[, eof[[1]]$left, by = season] %>%
@@
   .[, hgt := NULL] %>%
-  .[, `:=`(R = R/sd(R), I = I/sd(I)), by = .(season, PC)] %>%
+  .[, `:=`(R = as.numeric(scale(R)), I = as.numeric(scale(I))), by = .(season, PC)] %>%
   dcast(time ~ PC, value.var = c("R", "I")) %>%
@@
   .[, c("term", "PC") := tstrsplit(term, "_")] %>%
-  rm_intercept()
+  rm_intercept() %>% 
+  .[, term := factor(term, levels = c("R", "I"), labels = c("Real", "Imaginary"))]
 ```
 
-```{r}
+```{r define-plot_regr}
 plot_regr <- function(data, which_PC, which_levs, which_var) {
@@
 
+  var_title <- switch(which_var, 
+                      hgt = "Geopotential height", 
+                      t = "Temperature")
+  lab_pc <- c(PC1 = "EOF1", PC2 = "EOF2")
+
   if (length(which_PC) > 1) {
-    facet <- ggh4x::facet_nested(PC ~  term, labeller = labeller(term = lab_cplx))
+    facet <- ggh4x::facet_nested(PC ~  term)
   } else {
-    facet <- ggh4x::facet_nested(lev ~  term, labeller = labeller(term = lab_cplx,
-                                                                  lev = lab_lev))
+    facet <- ggh4x::facet_nested(lev ~ PC +  term, labeller = labeller(lev = lab_lev, PC = lab_pc))
   }
@@
         geom_contour_tanaka() +
-        scale_fill_divergent(NULL, guide = guide_colorstrip_bottom())
+        scale_fill_divergent(var_title, guide = guide_colorstrip_bottom())
     ) +
@@
 
-```{r, fig.cap = "Regresiones con altura geopotencial PC1"}
+(ref:eof1-regr-gh-cap) Regression coefficients of the real and imaginary part of the first complex EOF on SON geopotential height for the 1979 -- 2019 period.
+These coefficients come from multiple linear regression involving the real and imaginary part of both EFO2.
+
+```{r eof1-regr-gh, fig.cap = "(ref:eof1-regr-gh-cap)"}
 plot_regr(regr, "PC1", c(200, 50), "hgt")
@@
 
-```{r, fig.cap = "Regresiones con temperatura y PC1"}
+(ref:eof1-regr-t-cap) Same as Figure \@ref(fig:eof1-regr-gh) but for air temperature.
+
+```{r eof1-regr-t, fig.cap = "(ref:eof1-regr-t-cap)"}
 plot_regr(regr, "PC1", c(200, 50), "t") 
@@
 
-```{r, fig.cap = "Regresiones con altura geopotencial y PC2"}
+(ref:eof2-regr-gh-cap) Same as Figure \@ref(fig:eof1-regr-gh) but for the second EOF.
+
+```{r eof2-regr-gh, fig.cap = "(ref:eof2-regr-gh-cap)"}
 plot_regr(regr, "PC2", c(200, 50), "hgt")
@@
 
-```{r, fig.cap = "Regresiones con temperatura y PC2"}
+(ref:eof2-regr-t-cap) Same as Figure \@ref(fig:eof1-regr-gh) but for the second EOF and temperature.
+
+```{r eof2-regr-t, fig.cap = "(ref:eof2-regr-t-cap)"}
 plot_regr(regr, "PC2", c(200, 50), "t")
@@
 
-```{r}
-regr_simpler <- ceof[, eof[[1]]$left, by = season] %>%
-  .[season == "SON"] %>%
-  sep_ReIm(format = "wide") %>%
-  .[, hgt := NULL] %>%
-  .[, `:=`(R = R/sd(R), I = I/sd(I)), by = .(season, PC)] %>%
-  .[era5, on = "time", allow.cartesian = TRUE] %>%
-  .[lev %in% c(200, 50)] %>%
-  .[, FitLm(hgt, I, R, se = TRUE), by = .(lev, lon, lat, PC, season)] %>%
-  rm_intercept()
-```
-
-```{r, fig.cap = "Varianza temporal de altura geopotencial explicada por la variación temporal de cada componente principal."}
-regr_simpler %>%
-  # .[term == "R"] %>%
-  .[lat <= -20] %>%
-  periodic(lon = c(0, 360)) %>%
-  ggplot(aes(lon, lat)) +
-
-  geom_contour_fill(aes(z = r.squared, fill = ..level..),
-                    breaks = MakeBreaks(0.1, exclude = 0)) +
-  scale_fill_divergent_discretised(NULL, guide = guide_colorsteps_bottom())  +
-
-  geom_qmap(~.x[lat <= -20]) +
-  scale_x_longitude() +
-  scale_y_latitude() +
-  ggh4x::facet_nested(lev ~ PC, labeller = labeller(lev = lab_lev)) +
-  coord_polar() +
-  axis_labs_smol
-```
-
-```{r, fig.cap = "Amplitud de ondas zonales de la regresión con altura geopotencial (figuras anteriores)"}
+```{r zonal-waves, fig.cap = "Amplitud de ondas zonales de la regresión con altura geopotencial (figuras anteriores)"}
 regr %>%
@@
   # .[PC == "PC1"] %>%
-  .[, FitWave(estimate, 1:3), by = .(lev, PC, term, lat)] %>%
+  .[, FitWave(estimate, 0:3), by = .(lev, PC, term, lat)] %>%
   ggplot(aes(lat, amplitude)) +
@@
 
-```{r}
+```{r read-indices}
 sam <- "/home/elio/Documents/CONICET/asymsam/analysis/data/derived_data/indexes.Rds" %>%
@@
     scale_color_brewer("Part", palette = "Dark2") +
-    facet_grid(part~PC)
+    facet_grid(part ~ PC, labeller = labeller(PC = c(PC1 = "EOF1", PC2 = "EOF2")))
 }
@@
 
-```{r, fig.cap = "Relación del EOF (estandarizado) con SAM"}
+```{r sam-eof, fig.cap = "Standarised temporal EOFs and SAM index computed as the leading EOF of geopotential height anomalies at 700 hPa."}
 plot_with_index(ceof, sam[lev == 700 & symmetry == "full"][, .(time, sam)])
@@
 
-```{r}
+(ref:sam-eof-vertical-cap) Coefficient of determination between the real and imaginary part of each EOF and the SAM, Asymmetric SAM and Symmetric SAM indices computed at each level according to Campitelli on press??.
+
+```{r sam-eof-vertical, fig.cap = "(ref:sam-eof-vertical-cap)"}
 
@@
                                                          sym = "Symmetric SAM")) +
-  facet_grid(part ~ PC) +
+  facet_grid(part ~ PC, labeller = labeller(PC = c(PC1 = "EOF1", PC2 = "EOF2"))) +
   coord_flip()
@@
 
-```{r, fig.cap = "Idem anterior pero con ENSO"}
+(ref:oni-eof-cap) Same as Figure \@ref(fig:sam-eof) but for the Oceanic Enso Index (REF??).
+
+```{r oni-eof, fig.cap = "(ref:oni-eof-cap)"}
 plot_with_index(ceof, enso)
@@
 
-```{r}
+```{r compute-psi}
 psi <- ReadNetCDF(ERA5(), vars = "vo",
@@
 
-```{r}
+```{r regr_psi}
 regr_psi <- ceof[, eof[[1]]$left, by = season] %>%
@@
 
-```{r}
-
-plot_psi <- function(data, which_pc, which_lev) {
-  data %>%
-    .[PC == which_pc] %>%
-    .[lev == which_lev] %>%
-    # .[abs(estimate) > 1, estimate := NA] %>%
-    .[lat %between% c(-90, 10)] %>%
-    .[, p_val := Pvaluate(estimate, std.error, df, "fdr"), by = .(PC, term)] %>%
-    periodic(lon = c(0, 360)) %>%
-    ggplot(aes(lon, lat)) +
-    geom_contour_fill(aes(z = estimate), breaks = AnchorBreaks(exclude = 0)) +
-
-    geom_contour2(aes(z = p_val), breaks = 0.05, size = 0.1) +
-    stat_subset(aes(subset = p_val <= 0.05 & is.cross(lon, lat, 2)),
-                geom = "point", size = 0.2, alpha = 0.5) +
-    geom_streamline(aes(dx = fx, dy = fy), min.L = 3, L = 10, n = 30,
-                    size = 0.2, res = 2) +
-    geom_qmap(keep = .015) +
-    scale_fill_divergent(guide = "none") +
-    scale_x_longitude() +
-    scale_y_latitude() +
-    scale_mag(guide = "none") +
-    facet_grid(term ~ PC, labeller = labeller(term = lab_cplx)) +
-    coord_quickmap(ylim = c(NA, 10)) +
-    # ggh4x::facet_nested(term ~ PC) +
-    axis_labs_smol
-}
-
-```
-
-```{r}
+```{r sst}
 sst <- ERA5_SST() %>%
@@
 
-```{r}
+```{r regr_sst}
 regr_sst <- ceof[, eof[[1]]$left, by = season] %>%
@@
 
-```{r}
-plot_sst <- function(data, which_pc) {
-  data %>%
+```{r plot_sst_psi}
+plot_sst_psi <- function(which_pc, which_lev) {
+  sst_gdata <- regr_sst[PC == which_pc] %>%
+    .[lat %between% c(-90, 10)] %>%
+    .[, p_val := Pvaluate(estimate, std.error, df, "fdr"), by = .(PC, term)] %>%
+    .[, var := "SST"]
+  regr_psi %>% 
     .[PC == which_pc] %>%
+    .[lev == which_lev] %>%
     # .[abs(estimate) > 1, estimate := NA] %>%
@@
     .[, p_val := Pvaluate(estimate, std.error, df, "fdr"), by = .(PC, term)] %>%
+    .[, var := "Streamfunction"] %>% 
     periodic(lon = c(0, 360)) %>%
     ggplot(aes(lon, lat)) +
-    ditto(
-      geom_contour_fill(aes(z = estimate), breaks = AnchorBreaks(exclude = 0)) +
-        geom_contour_tanaka(aes(z = estimate), breaks = AnchorBreaks(exclude = 0)) +
-        scale_fill_divergent(NULL, guide = guide_colorstrip_bottom(width = 15))
-    ) +
-
+    geom_contour_fill(aes(z = estimate*1e8, fill = ..level..),  breaks = AnchorBreaks(exclude = 0)) +
     geom_contour2(aes(z = p_val), breaks = 0.05, size = 0.1) +
@@
                 geom = "point", size = 0.2, alpha = 0.5) +
+    geom_streamline(aes(dx = fx, dy = fy), min.L = 3, L = 10, n = 30,
+                    size = 0.2, res = 2) +
     geom_qmap(keep = .015) +
+    scale_fill_divergent_discretised("Streamfunction", guide = "none") +
+    ggnewscale::new_scale_fill() +
 
+    geom_contour_fill(data = sst_gdata,
+                      aes(z = estimate, fill = ..level..), breaks = AnchorBreaks(exclude = 0)) +
+    geom_contour_tanaka(data = sst_gdata, 
+                        aes(z = estimate), breaks = AnchorBreaks(exclude = 0)) +
+    geom_contour2(data = sst_gdata, 
+                  aes(z = p_val), breaks = 0.05, size = 0.1) +
+    stat_subset(data = sst_gdata,
+                aes(subset = p_val <= 0.05 & is.cross(lon, lat, 2)),
+                geom = "point", size = 0.2, alpha = 0.5) +
+    scale_fill_divergent_discretised("SST", guide = guide_colorsteps_bottom(order = 2)) +
     scale_x_longitude() +
     scale_y_latitude() +
-    facet_grid(term ~ PC, labeller = labeller(term = lab_cplx)) +
-    axis_labs_smol +
-    coord_quickmap(ylim = c(-90, 10))
+    scale_mag(guide = "none") +
+    facet_grid(term ~ var, labeller = labeller(term = lab_cplx, 
+                                               PC = c(PC1 = "EOF1", PC2 = "EOF2"))) +
+    coord_quickmap(ylim = c(NA, 10)) +
+    theme(legend.box = "horizontal") +
+    # ggh4x::facet_nested(term ~ PC) +
+    axis_labs_smol
 }
-
 ```
 
-```{r, fig.cap = "Regresión con PSI en 200 hPa y sst, PC1"}
-plot_psi(regr_psi, "PC1", 200) +
-  plot_sst(regr_sst, "PC1")
+```{r sst-psi-1,  fig.cap = "Regresión con PSI en 200 hPa y sst, PC1"}
+plot_sst_psi(which_pc = "PC1", which_lev = 200)
 ```
 
-```{r, fig.cap = "Regresión con PSI en 200 hPa y sst, PC2"}
-plot_psi(regr_psi, "PC2", 200) +
-  plot_sst(regr_sst, "PC2")
+```{r sst-psi-2,  fig.cap = "Regresión con PSI en 200 hPa y sst, PC1"}
+plot_sst_psi(which_pc = "PC2", which_lev = 200)
+```
+
+```{r mean-psi, fig.cap = "mean streamfunction"}
+psi %>% 
+  .[lev == 200, .(psi.z = -mean(psi.z)), by = .(lon, lat)] %>% 
+  .[, c("fx", "fy") := WaveFlux(.SD)] %>% 
+  .[lat <= 10] %>% 
+  ggplot(aes(lon, lat)) +
+  geom_contour_fill(aes(z = psi.z)) +
+  geom_streamline(aes(dx = fx, dy = fy), min.L = 3, L = 10, n = 30,
+                  size = 0.2, res = 2) +
+  geom_qmap(keep = .015) +
+  scale_fill_divergent(NULL, guide = guide_colorstrip_bottom(width = 15)) +
+  scale_x_longitude() +
+  scale_y_latitude()  +
+  coord_quickmap(ylim = c(-90, 10))
 ```
@@
 
-```{r}
+```{r cmap}
 cmap <- ReadNetCDF(CMAP(), vars = c(pp = "precip"),
@@
 
-```{r}
+```{r regr_pp}
 regr_pp <- ceof[, eof[[1]]$left, by = season] %>%
@@
 
-```{r}
-
+```{r add_continents}
 continents <- list(africa  = list(lon = c(0, 60),
@@
                    oceania = list(lon = c(100, 180),
-                                  lat = c(-50, -5)),
+                                  lat = c(-50, -13)),
                    america = list(lon = c(265, 325),
@@
 }
-
 ```
 
-```{r}
-plot_pp <- function(data, which_pc, which_term = c("R", "I"), which_continent) {
+```{r plot_pp}
+plot_pp <- function(data, which_pc, which_term = c("R", "I"), which_continent = NULL) {
+  filter_continent <- function(data) {
+    if (!is.null(which_continent)) {
+    data[ continent == which_continent]  
+    } else {
+      data
+    }
+
+  }
   data %>%
@@
     .[, lev := "sfc"] %>%
+    .[lat >= -60] %>% 
     add_continent() %>%
-    .[continent == which_continent] %>%
-    .[, p_val := Pvaluate(estimate, std.error, df, "fdr"), by = .(lev, PC, term)] %>%
+    filter_continent() %>% 
+    # .[abs(estimate) <= 2] %>% 
+    .[, p_val := Pvaluate(estimate, std.error, df, "none"), by = .(lev, PC, term)] %>%
     periodic(lon = c(0, 360)) %>%
     ggplot(aes(lon, lat)) +
-    ditto(
-      geom_contour_fill(aes(z = estimate), breaks = AnchorBreaks(exclude = 0)) +
-        geom_contour_tanaka() +
-        scale_fill_divergent(NULL, guide = guide_colorstrip_bottom(15))
-    ) +
+    # geom_raster(aes(fill = estimate)) +
+    geom_contour_fill(aes(z = estimate, fill = ..level..), breaks = AnchorBreaks(exclude = 0)) +
+    geom_contour_tanaka(aes(z = estimate), breaks = AnchorBreaks(exclude = 0)) +
 
+    # scale_fill_divergent() +
+    scale_fill_divergent_discretised(NULL, guide = guide_colorsteps_bottom(15)) +
     geom_contour2(aes(z = p_val), breaks = 0.05, size = 0.1) +
     stat_subset(aes(subset = p_val <= 0.05),
-                geom = "point", size = 1) +
-    geom_qmap(~.x %>% add_continent() %>% .[continent == which_continent], keep = 1) +
+                geom = "point", size = 0.2) +
+    geom_qmap(~.x %>% .[lat %between% c(-60, max(data$lat))] %>% add_continent() %>% filter_continent(),
+              keep = 1) +
 
@@
 
-```{r, fig.cap = "Regresión de cada PC con la precipitación"}
-plot_pp(regr_pp, "PC1", which_continent =  "america")  +
-  plot_pp(regr_pp, "PC2", which_continent =  "america")
+```{r pp-america, fig.cap = "Regresión de cada PC con la precipitación"}
+plot_pp(data = regr_pp, which_pc= c("PC1", "PC2"), which_continent =  "america")  
 ```
 
-```{r, fig.cap = "Regresión de cada PC con la precipitación"}
-plot_pp(regr_pp, "PC1", which_continent = "africa") +
-  plot_pp(regr_pp, "PC2", which_continent = "africa")
+```{r pp-africa, fig.cap = "Regresión de cada PC con la precipitación"}
+plot_pp(regr_pp, c("PC1", "PC2"), which_continent = "africa") 
 ```
 
-```{r, fig.cap = "Regresión de cada PC con la precipitación"}
-plot_pp(regr_pp, "PC1", which_continent =  "oceania") +
-  plot_pp(regr_pp, "PC2", which_continent =  "oceania")
+```{r pp-oceania, fig.cap = "Regresión de cada PC con la precipitación"}
+plot_pp(regr_pp, c("PC1", "PC2"), which_continent =  "oceania") 
 ```

### 2021-01-08 analysis/paper/paper.Rmd
Agrega SAM datos y más (🥐️ para charles)

@@
               I = "Imaginary")
+lab_PC <- c(PC1 = "EOF1", 
+            PC2 = "EOF2")
+
 combine_lists <- function(...) {
@@
 Note that in Figure \@ref(fig:ceofs-2), the real part of EOF1 has non zero mean.
-This is due to th fact that the geopotential fields that enter into the algorithm are anomalies with respect to the zonal mean, not the time mean.
+This is due to th fact that the geopotential fields that enter into the algorithm are anomalies with respect to the zonal mean (shown in Figure \@ref(fig:mean-anom)), not the time mean.
 The real part of the first EOF is, therefore, capturing the mean zonally anomalous field *and the variability that projects onto it*.
 
-(ref:extended-series-cap) Temporal series extended using ERA5 Back extended preliminary edition (period 1950 -- 1978), ERA5 (period 1979 -- 2019) and ERA20C (period 1900 -- 2010).
+```{r mean-anom, fig.cap = "Mean SON zonal geopotential height anomalies at (a) 50 hPa and (b) 200 hPa"}
+era5 %>% 
+  .[, mean(-hgt), by = .(lon, lat, lev)] %>% 
+  .[lev %in% c(50, 200)] %>% 
+  .[lat <= -20] %>% 
+  .[, V2 := V1 - mean(V1), by = .(lat, lev)] %>% 
+  periodic(lon = c(0, 360)) %>% 
+  ggplot(aes(lon, lat)) +
+  geom_contour_fill(aes(z = V2), global.breaks = FALSE, breaks = AnchorBreaks(exclude = 0)) +
+  geom_contour_tanaka(aes(z = V2), global.breaks = FALSE, breaks = AnchorBreaks(exclude = 0)) +
+  geom_qmap(~ .x[lat <= -20]) +
+  scale_fill_divergent(NULL, breaks = AnchorBreaks(exclude = 0), 
+                       guide = guide_colorstrip_bottom()) +
+  scale_x_longitude() +
+  scale_y_latitude() +
+  facet_grid(~lev, labeller = labeller(lev = lab_lev)) +
+  coord_polar() +
+  axis_labs_smol +
+  tag_facets()
+```
+
+(ref:extended-series-cap) Temporal series extended using ERA5 back extended preliminary edition (period 1950 -- 1978) and ERA5 (period 1979 -- 2019) and ERA20C (period 1900 -- 2010).
 Each series is computed by projecting monthly geopotential height zonal anomalies standardised by level south of 20ºS onto the corresponding spatial pattern.
-In numbers, the correlation between the ERA20C-based series and the two ERA5-based series.
+In numbers, the coefficient of determination between the ERA20C-based series and the two ERA5-based series.
 The ERA5 + ERABC and the ERA20C series are centred and scaled using the mean and standard deviation for their period of overlap (1950 -- 2010).
@@
 extended <- rbindlist(list(era20c = era20c,
-                           era5 = era_more),  use.names = TRUE, idcol = "dataset") %>%
-  .[, hgt := (hgt - mean(hgt[dataset == "era5"], na.rm = TRUE))/sd(hgt[dataset == "era5"], na.rm = TRUE),
-    by = .(part, PC)]
+                           era5 = era_more),  use.names = TRUE, idcol = "dataset")
+
+trends <- extended %>% 
+  .[time %between% common_range(time, dataset)] %>% 
+  .[, FitLm(hgt, time, se = TRUE), by = .(dataset, part, PC)] %>% 
+  rm_intercept() %>% 
+  .[, p_val := Pvaluate(estimate, std.error, df)] %>% 
+  .[, p_report := report_p(p_val)]
+
+trends_all <- extended %>% 
+  .[, FitLm(hgt, time, se = TRUE), by = .(dataset, part, PC)] %>% 
+  rm_intercept() %>% 
+  .[, p_val := Pvaluate(estimate, std.error, df)] %>% 
+  .[, p_report := report_p(p_val)]
+
+periods <- range(extended[time %between% common_range(time, dataset)]$time)
+periods <- c(periods[1], lubridate::as_datetime("1979-01-01"), periods[2])
 
@@
   dcast(time + PC + part ~ dataset, value.var = "hgt") %>% 
-  na.omit() %>% 
-  .[, cor(era20c, era5), 
-    by = .(PC, part, time = ifelse(year(time) >= 1979, 2005, 1960))] %>% 
-  .[, time := lubridate::as_datetime(paste0(time,  "-01-01"))] %>% 
-  .[, dataset := ifelse(year(time) >= 1979, "era5", "era5be")]
+  na.omit() %>% {
+    d <- .
+    rbind(
+  d[, .(correlation = cor(era20c, era5)), 
+    by = .(PC, part, time = ifelse(year(time) >= 1979, 
+                                   as.character(mean(periods[2:3])),
+                                   as.character(mean(periods[1:2]))))],
+
+    d[, .(correlation = cor(era20c, era5)), 
+    by = .(PC, part)] %>% 
+    .[, time := as.character(mean(periods[c(1, 3)]))]
+    )
+  } %>%
+
+  .[, time := lubridate::as_datetime(time)]
 
@@
   copy() %>% 
-  .[dataset == "era5", dataset := ifelse(year(time) < 1979, "era5be", "era5")] %>% 
+  .[, ref := time %between% common_range(time, dataset)] %>%
+  .[, hgt := (hgt - mean(hgt[ref]))/sd(hgt[ref]), by = .(PC, part, dataset)] %>% 
+  .[, dataset_o := dataset] %>% 
+  # .[dataset == "era5", dataset := ifelse(year(time) < 1979, "era5be", "era5")] %>% 
   ggplot(aes(time, hgt)) +
+  # Zero line + rangeframe
   geom_hline(yintercept = 0, size = 0.1, color = "gray30") +
+  ggthemes::geom_rangeframe(sides = "lr", size = 0.1, color = "gray30") +
+
   geom_line(aes(color = dataset)) +
-  geom_text(data = correlations, aes(color = dataset, y = 2.3, label = signif(V1, 2))) +
+  scale_alpha_manual(values = c(1, 0.2)) +
+  # Correlations
+  annotate("errorbarh", xmin = periods[1:2], xmax = periods[2:3], y = -2.7, size = 0.3) +
+  annotate("errorbarh", xmin = periods[1], xmax = periods[3], y = -4, size = 0.3) +
+
+  geom_label(data = correlations[time %in% range(time)], aes(y = -2.7, label = signif(correlation^2, 2)), 
+             label.r = grid::unit(0, "lines"), label.size = 0) +
+  geom_label(data = correlations[!(time %in% range(time))], aes(y = -4, label = signif(correlation^2, 2)), 
+             label.r = grid::unit(0, "lines"), label.size = 0) +
+
+  geom_smooth(data = ~.x[dataset == "era20c"], 
+              method = "lm", se = FALSE,  aes(color = dataset)) +
+  geom_smooth(data = ~copy(.x)[time %between% common_range(time, dataset)],
+              aes(color = dataset), method = "lm", se = FALSE) +
+
   scale_x_datetime(NULL) +
   scale_y_continuous("Standarised EOF") +
-  scale_color_brewer("Data", palette = "Dark2", labels = c(era5 = "ERA5",
-                                                           era20c = "ERA20C",
-                                                           era5be = "ERA5 BE"),
+  scale_color_brewer(NULL, palette = "Dark2", labels = c(era20c = "ERA20C",
+                                                           era5 = "ERA5 BE"),
                      guide = guide_legend(override.aes = list(size = 1), order = 1)) +
-  ggnewscale::new_scale_colour() +
-  geom_smooth(data = ~.x[dataset == "era20c"][, dataset := dataset == "era20c"], 
-              method = "lm", se = FALSE,  aes(color = dataset)) +
-  geom_smooth(data = ~copy(.x)[, dataset := dataset == "era20c"][time %between% common_range(time, dataset)],
-              aes(color = dataset), method = "lm", se = FALSE) +
+  facet_grid(part ~ PC, labeller = labeller(PC = lab_PC)) +
+  theme(panel.spacing = grid::unit(1.5, "lines"))  +
 
-  scale_color_brewer("Trend", palette = "Dark2", direction = -1,
-                     guide = guide_legend(override.aes = list(size = 1), order = 99),
-                     labels = c("FALSE" = "ERA5+ERA5 BE",
-                                "TRUE" = "ERA2CO")) +
-  facet_grid(part ~ PC) +
-  theme(panel.spacing = grid::unit(1.5, "lines"))  +
-  ggthemes::geom_rangeframe(sides = "lr", size = 0.1, color = "gray30") +
   tag_facets("cr")
@@
 
-Figure \@ref(fig:extended-series) shows temporal series of the two complex EOFs extended beyond the satellite era using the preliminary ERA5 back extension going back to 1950 and using ERA20C for the period 1900 -- 2010.
-The hybrid ERA5 + ERA5 BE series and the ERE20C series agree relatively well during the period of overlap, although correlations are higher in the satellite period.
-The ERA20C time-series has a significant upward linear trend in the imaginary part of EOF1.
-However, there is no trend in the corresponding hybrid modern dataset during the period of overlap and the correlations between the old and modern series is much poorer.
-This suggests that the observed trends is not robust.
+Figure \@ref(fig:extended-series) shows temporal series of the two complex EOFs extended beyond the satellite era using the preliminary ERA5 back extension going back to 1950 (which we call the "hybrid ERA5" reanalysis) and using ERA20C for the period 1900 -- 2010.
+The hybrid ERA5 series and the ERE20C series agree relatively well during the period of overlap, although correlations are higher in the satellite period.
+There is a slight downwards trend in the real part of EOF1 in the hybrid reanalysis (Figure \@ref(fig:extended-series)a.1, p-value `r trends[part == "Real" & PC == "PC1" & dataset != "era20c"]$p_report` for the period 1950 -- 2010) that appears to be part of a long-term signal, judging by the statistically significant trend observed in the ERA20C reanalysis from 1900 to 2010 (p-value `r trends_all[part == "Real" & PC == "PC1" & dataset == "era20c"]$p_report`) with similar magnitude during the period of overlap.
+
+The ERA20C time-series has a also significant upward linear trend in the imaginary part of EOF1 (Figure \@(ref:fig-extended-series)a.2, p-value `r trends[part == "Imaginary" & PC == "PC1" & dataset == "era20c"]$p_report`.
+However, there is no trend in the hybrid dataset during the period of overlap (p-value `r trends[part == "Imaginary" & PC == "PC1" & dataset != "era20c"]$p_report`) and the correlations between the old and modern series are much poorer for this variable.
+This suggests that the observed trend is not robust.
+
+There is no significant trend in any of the complex parts of EOF2.
+The correlations between the hybrid and ERA20C datasets are much higher than for the EOF1, which suggests that ERA20C could be used as a valid dataset to study this index in the past.
+
+<!-- (ref:extended-magnitude-cap) Same as Figure \@ref(fig:extended-series) but for EOF magnitude defined as $\sqrt{\mathrm{Real}^2 + \mathrm{Imaginary}^2}$.  -->
+
+<!-- ```{r extended-magnitude, fig.cap = "(ref:extended-magnitude-cap)"} -->
+<!-- magnitudes <- extended %>%  -->
+<!--   copy() %>%  -->
+<!--   dcast(time + PC + dataset ~ part, value.var = "hgt") %>%  -->
+<!--   .[, ref := time %between% common_range(time, dataset)] %>% -->
+<!--   .[, hgt := sqrt(Real^2 + Imaginary^2)] %>%  -->
+<!--   .[, hgt := (hgt - mean(hgt[ref]))/sd(hgt[ref]), by = .(PC, dataset)] -->
+
+<!-- correlations_mag <- magnitudes %>%  -->
+<!--   dcast(time + PC  ~ dataset, value.var = "hgt") %>%  -->
+<!--   na.omit() %>% { -->
+<!--     d <- . -->
+<!--     rbind( -->
+<!--   d[, .(correlation = cor(era20c, era5)),  -->
+<!--     by = .(PC, time = ifelse(year(time) >= 1979,  -->
+<!--                                    as.character(mean(periods[2:3])), -->
+<!--                                    as.character(mean(periods[1:2]))))], -->
+
+<!--     d[, .(correlation = cor(era20c, era5)),  -->
+<!--     by = .(PC)] %>%  -->
+<!--     .[, time := as.character(mean(periods[c(1, 3)]))] -->
+<!--     ) -->
+<!--   } %>%  -->
+
+<!--   .[, time := lubridate::as_datetime(time)] -->
+
+<!-- trends_mag <- magnitudes %>%  -->
+<!--   .[time %between% common_range(time, dataset)] %>%  -->
+<!--   .[, FitLm(hgt, time, se = TRUE), by = .(dataset, PC)] %>%  -->
+<!--   rm_intercept() %>%  -->
+<!--   .[, p_val := Pvaluate(estimate, std.error, df)] %>%  -->
+<!--   .[, p_report := report_p(p_val)] -->
+
+<!-- trends_all_mag <- magnitudes %>%  -->
+<!--   .[, FitLm(hgt, time, se = TRUE), by = .(dataset, PC)] %>%  -->
+<!--   rm_intercept() %>%  -->
+<!--   .[, p_val := Pvaluate(estimate, std.error, df)] %>%  -->
+<!--   .[, p_report := report_p(p_val)] -->
+
+<!-- magnitudes %>%  -->
+<!--   ggplot(aes(time, hgt)) + -->
+<!--   # Zero line + rangeframe -->
+<!--   geom_hline(yintercept = 0, size = 0.1, color = "gray30") + -->
+<!--   ggthemes::geom_rangeframe(sides = "lr", size = 0.1, color = "gray30") + -->
+
+<!--   geom_line(aes(color = dataset)) + -->
+
+<!--   # Correlations -->
+<!--   annotate("errorbarh", xmin = periods[1:2], xmax = periods[2:3], y = -2.7, size = 0.3) + -->
+<!--   annotate("errorbarh", xmin = periods[1], xmax = periods[3], y = -4, size = 0.3) + -->
+
+<!--   geom_label(data = correlations_mag[time %in% range(time)], aes(y = -2.7, label = signif(correlation^2, 2)),  -->
+<!--              label.r = grid::unit(0, "lines"), label.size = 0) + -->
+<!--   geom_label(data = correlations_mag[!(time %in% range(time))], aes(y = -4, label = signif(correlation^2, 2)),  -->
+<!--              label.r = grid::unit(0, "lines"), label.size = 0) + -->
+
+<!--     geom_smooth(data = ~.x[dataset == "era20c"],  -->
+<!--               method = "lm", se = FALSE,  aes(color = dataset)) + -->
+<!--   geom_smooth(data = ~copy(.x)[time %between% common_range(time, dataset)], -->
+<!--               aes(color = dataset), method = "lm", se = FALSE) + -->
+
+<!--   scale_x_datetime(NULL) + -->
+<!--   scale_y_continuous("Standarised EOF magnitude") + -->
+<!--   scale_color_brewer(NULL, palette = "Dark2", labels = c(era20c = "ERA20C", -->
+<!--                                                            era5 = "ERA5 BE"), -->
+<!--                      guide = guide_legend(override.aes = list(size = 1), order = 1)) + -->
+<!--   facet_wrap(~PC, labeller = labeller(PC = lab_PC), ncol = 2) + -->
+<!--   theme(panel.spacing = grid::unit(1.5, "lines"))  + -->
+
+<!--   tag_facets() -->
+<!-- ``` -->
 
@@
     coord_polar() +
-    axis_labs_smol 
+    axis_labs_smol +
+    tag_facets("rc")
 }
@@
 ```
+Figure \@ref(fig:ref:eof1-regr-gh) shows regression patterns of SON geopotential height on the real and imaginary part of the first EOF at 50 hPa and 200 hPa.
 
@@
 ```{r sam-eof-vertical, fig.cap = "(ref:sam-eof-vertical-cap)"}
-
 time <- ceof[, eof[[1]]$left] %>%
@@
 
-"/home/elio/Documents/CONICET/asymsam/analysis/data/derived_data/indexes.Rds" %>%
-  readRDS() %>%
-  # .[lev %in% c(200, 50)] %>%
-  # .[term != "full"] %>%
-  rm_singleton() %>%
-  .[season(time) == "SON"] %>%
-  .[, .(estimate_norm = mean(estimate_norm)), by = .(lev, term, time = seasonally(time))] %>%
+sams <- readRDS(data_path("derived", "sam.Rds"))
+
+sams %>% 
   .[time, on =.NATURAL, allow.cartesian = TRUE] %>%
   na.omit() %>%
-  .[, cor(hgt, estimate_norm)^2, by = .(lev, term, part, PC)] %>%
+  .[, cor(hgt, estimate)^2, by = .(lev, term, part, PC)] %>%
+  ggplot(aes(lev, V1)) +
+  geom_line(aes(color = term, size = term)) +
+  scale_x_level() +
+  scale_y_continuous("r²") +
+  scale_size_manual(guide = "none",
+                    values = c(full = 1,
+                               asym = 0.5,
+                               sym = 0.5)) +
+  scale_color_brewer(NULL, palette = "Dark2", labels = c(full = "SAM",
+                                                         asym = "Asymmetric SAM",
+                                                         sym = "Symmetric SAM")) +
+  facet_grid(part ~ PC, labeller = labeller(PC = c(PC1 = "EOF1", PC2 = "EOF2"))) +
+  coord_flip()
+
+```
+
+```{r}
+sams <- ERA5() %>% 
+  ReadNetCDF(vars = c("hgt" = "z"), 
+             subset = list(latitude = c(-90, -20))) %>% 
+  normalise_coords() %>% 
+  .[season(time) == "SON"] %>% 
+  .[, .(hgt = mean(hgt)), by = .(lon, lat, lev, time = seasonally(time))] %>% 
+  .[, hgt := Anomaly(hgt), by = .(lon, lat, lev)] %>% 
+  .[, asymsam::eof_asym(hgt, lon, lat, time), by = lev]
+```
+
+```{r}
+sams %>% 
+  .[, PC := NULL] %>% 
+ .[time, on =.NATURAL, allow.cartesian = TRUE] %>%
+  na.omit() %>%
+  .[, cor(hgt, estimate)^2, by = .(lev, term, part, PC)] %>%
   ggplot(aes(lev, V1)) +

### 2021-01-08 R/data_locations.R

@@ -67,7 +67,7 @@ ERA20C <- data_nc_fun(file = data_path("raw", "era20c.mon.mean.nc"),
 #' @export
 #' @rdname data_locations
 NCEP_PSI <- data_nc_fun(file = data_path("raw", "psi.mon.mean.nc"),
-                        download = "ftp://ftp2.psl.noaa.gov/Datasets/ncep.reanalysis.derived/sigma/vor.mon.mean.nc")
+                        download = "ftp://ftp2.psl.noaa.gov/Datasets/ncep.reanalysis.derived/sigma/psi.mon.mean.nc")
 
 #' @export
 #' @rdname data_locations

### 2021-01-08 R/lm_lite.R

@@ -44,9 +44,9 @@ Detrend <- function(y, x) {
    return(y + m)
 }
 
-#' Compute p.values
+#' Compute p-values
 #'
-#' Computes p.values from the result of [metR::FitLm()] (or any assumed student-t-distributed
+#' Computes p-values from the result of [metR::FitLm()] (or any assumed student-t-distributed
 #' statistic) with optional adjustments.
 #'
 #' @param estimate estimate of the regression (or other fit)
@@ -59,6 +59,23 @@ Pvaluate <- function(estimate, std.error, df, adjustment = "none") {
   stats::p.adjust(2*stats::pt(abs(estimate)/std.error, df, lower.tail = FALSE), method = adjustment)
 }
 
+#' Report p-values
+#'
+#' Creates a "pretty" string from p-values. Reports p-values (adding an equal sign)
+#' rounded to two  significant if they are greater than 0.001, in which case,
+#' it reports " < 0.001".
+#'
+#' @param p numeric vector of p-values
+#'
+#'
+#' @return
+#' A character vector of the same length as `p` with `" < 0.001"` for values
+#' lower than 0.001 and `" = signif(p, 2)"` otherwise.
+#'
+#' @export
+report_p <- function(p) {
+  ifelse(p < 0.001, " < 0.001", paste0(" = ", signif(p, 2)))
+}
 
 #' Remove intercept
 #'

### 2021-01-08 analysis/scripts/02-compute-sam.R

@@ -1,0 +1,28 @@
+sam_file <- data_path("derived", "sam.Rds")
+
+if (file.exists(sam_file)) {
+  message("Sam file already exists. Exiting.")
+} else {
+  check_asymsam <- requireNamespace("asymsam", quietly = TRUE)
+  if (!check_asymsam) {
+    stop("Computing the Asymmetric and Symmetric SAM requires package asymsam.\n",
+         "Install it with remotes::install_github(\"eliocamp/asymsam\").")
+  }
+
+  library(shceof)
+  library(metR)
+
+  sams <- ERA5() %>%
+    ReadNetCDF(vars = c("hgt" = "z"),
+               subset = list(latitude = c(-90, -20))) %>%
+    normalise_coords() %>%
+    .[, hgt := Anomaly(hgt), by = .(lon, lat, lev, month(time))] %>%
+    .[, asymsam::eof_asym(hgt, lon, lat, time), by = lev] %>%
+    rm_singleton() %>%
+    .[season(time) == "SON"] %>%
+    .[, .(estimate = mean(estimate)), by = .(lev, term, time = seasonally(time))]
+
+  saveRDS(sams, sam_file)
+}
+
+

### 2021-02-08 analysis/paper/paper.Rmd
Rotación y ENSO

@@
   fig.path = "../figures/",
-  dpi = 300
+  dpi = 300,
+  dev = c("pdf", "png")
 )
@@
 
+```{r compute-psi}
+psi <- ReadNetCDF(ERA5(), vars = "vo",
+                  subset = list(level = c(50, 200),
+                                time = c("1979-01-01", "2019-12-01"))) %>%
+  normalise_coords() %>%
+  .[season(time) == "SON"] %>%
+  .[, .(vo = mean(vo)), by = .(lon, lat, lev,time = seasonally(time))] %>%
+  .[, solve_poisson(vo, lon, lat), by = .(lev, time)] %>%
+  .[, psi.z := Anomaly(value), by = .(time, lev, lat)] %>%
+  .[, value := NULL]
+```
+
 # Introduction
@@
 # Data & Methods
+
+Whereas for traditional "real" PCA, each principal component is defined up to it's sign. In complex PCA, principal components are defined up to their phase [@horel1984].
 
@@
 That is, each group of two EOFs represents the same zonal wave structure that changes in magnitude and location.
-Similarly, at 200 hPa EOF2 and EOF3 represent the same zonal wave structure with wave number ~3.
+Similarly, at 200 hPa EOF2 and EOF3 represent the same zonal wave structure with wave number \~3.
 
@@
 
+```{r rotate_eofs}
+rotate <- function(z, angle = 0) {
+  complex(real = cos(angle), imaginary = sin(angle)) * z
+}
+
+box <- list(lon = c(150, 195),
+           lat = c(-25, -5))
+
+psi_box <- psi %>% 
+  .[lev == 200] %>% 
+  .[lat %between% box$lat & lon %between% box$lon] %>% 
+  .[, .(psi.z = weighted.mean(psi.z, cos(lat*pi/180))), by = .(time, lev)]
+
+psi_box <- cut(ceof$eof[[1]], 2)$left %>% 
+  copy() %>% 
+  .[psi_box, on = "time"]
+
+angles <- seq(-pi, pi, by = 1*pi/180)
+
+rotations_pc1 <- lapply(angles, function(a) {
+  ceof$eof[[1]]$left[PC == "PC1"] %>% 
+    copy() %>% 
+    .[, hgt := rotate(hgt, a)] %>% 
+    sep_ReIm() %>% 
+    .[, .(mean = mean(hgt)), by = part] %>%
+    .[, rotation := a]
+}) %>% 
+  rbindlist()
+
+best_rotation_pc1 <- rotations_pc1[part == "Imaginary"][which.min(abs(mean))]$rotation
+
+rotations_pc2 <- lapply(angles, function(a) {
+  psi_box %>% 
+    .[, hgt2 := rotate(hgt, a)] %>% 
+    .[, .(R = cor(Re(hgt2), psi.z),
+          I = cor(Im(hgt2), psi.z))] %>% 
+    .[, rotation := a]
+}) %>% 
+  rbindlist()
+
+best_rotation_pc2 <- rotations_pc2[which.min(abs(R^2))]$rotation
+
+eof_rotated_left <- ceof$eof[[1]]$left %>% 
+  copy() %>% 
+  .[PC == "PC2", hgt := rotate(hgt, best_rotation_pc2)] %>% 
+  .[PC == "PC1", hgt := rotate(hgt, best_rotation_pc1)]
+
+eof_rotated_right <-  ceof$eof[[1]]$right %>% 
+  copy() %>% 
+  .[PC == "PC2", hgt := rotate(hgt, best_rotation_pc2)] %>% 
+  .[PC == "PC1", hgt := rotate(hgt, best_rotation_pc1)]
+
+ceof_unrotated <- copy(ceof)
+
+ceof$eof[[1]]$left <- eof_rotated_left
+ceof$eof[[1]]$right <- eof_rotated_right
+
+ceof <- rbindlist(list(rotated = ceof,
+               unrotated = ceof_unrotated), 
+          idcol = "rotation")
+
+rotated <- function(x) x[rotation == "rotated"] 
+unrotated <- function(x) x[rotation == "unrotated"] 
+```
+
+```{r check-rotation}
+# Check that rotated and unrotated eofs 
+# predict the same fields (up to floating point errors)
+bad_cells <- ceof[, predict(eof[[1]]), by = rotation] %>% 
+  .[, hgt := as.numeric(Re(hgt))] %>% 
+  dcast(time + lon + lat + lev ~ rotation, value.var = "hgt") %>% 
+  .[, dif := rotated - unrotated] %>% 
+  .[abs(rotated - unrotated)  > 1e-10]
+
+if (nrow(bad_cells) != 0) {
+  stop("Rotated and unrotated eofs are not equivalent!")
+}
+
+ceof <- ceof[rotation == "rotated"][, rotation := NULL][]
+```
+
 ```{r define-plot_ceof}
@@
 ```
+
 Both observations motivate the decision of performing complex EOF jointly between levels.
@@
 
-The wave patterns described by these complex EOFs match the patterns seen in the *navie* EOFs of Figure \@ref(fig: cor-eof-naive-1): The first is a wave 1, while the second is a wave 3.
+The wave patterns described by these complex EOFs match the patterns seen in the navie EOFs of Figure \@ref(fig: cor-eof-naive-1): The first is a wave 1, while the second is a wave 3.
 Note that in Figure \@ref(fig:ceofs-2), the real part of EOF1 has non zero mean.
 This is due to th fact that the geopotential fields that enter into the algorithm are anomalies with respect to the zonal mean (shown in Figure \@ref(fig:mean-anom)), not the time mean.
-The real part of the first EOF is, therefore, capturing the mean zonally anomalous field *and the variability that projects onto it*.
+The real part of the first EOF is, therefore, capturing the mean zonally anomalous field and the variability that projects onto it.
 
@@
 ```{r extended-series, fig.cap = "(ref:extended-series-cap)"}
-
 pattern <- ceof$eof[[1]]$right %>%
@@
 
+# Check that the extended index coincides with the 
+# original index.
+cors <- era_more %>% 
+  .[sep_ReIm(ceof$eof[[1]]$left), on = c("time", "PC", "part")] %>% 
+  na.omit() %>% 
+  .[, cor(hgt, i.hgt), by = .(PC, part)]
+
+if (any(cors$V1 < 0.98)) {
+  stop("Extended indices didn't work!")
+}
+
 extended <- rbindlist(list(era20c = era20c,
@@
 
-  geom_label(data = correlations[time %in% range(time)], aes(y = -2.7, label = signif(correlation^2, 2)), 
+  geom_label(data = correlations[time %in% range(time)], 
+             aes(y = -2.7, label = paste0("r²: ", signif(correlation^2, 2))), 
              label.r = grid::unit(0, "lines"), label.size = 0) +
-  geom_label(data = correlations[!(time %in% range(time))], aes(y = -4, label = signif(correlation^2, 2)), 
+  geom_label(data = correlations[!(time %in% range(time))], 
+             aes(y = -4, label = paste0("r²: ", signif(correlation^2, 2))), 
              label.r = grid::unit(0, "lines"), label.size = 0) +
@@
 
-  scale_x_datetime(NULL) +
+  scale_x_datetime(NULL, date_labels = "%Y",
+                   breaks = function(x) sort(unique(c(scales::pretty_breaks()(x), 
+                                                          lubridate::as_datetime("1979-10-01"))))) +
   scale_y_continuous("Standarised EOF") +
@@
   theme(panel.spacing = grid::unit(1.5, "lines"))  +
-
   tag_facets("cr")
@@
 The hybrid ERA5 series and the ERE20C series agree relatively well during the period of overlap, although correlations are higher in the satellite period.
-There is a slight downwards trend in the real part of EOF1 in the hybrid reanalysis (Figure \@ref(fig:extended-series)a.1, p-value `r trends[part == "Real" & PC == "PC1" & dataset != "era20c"]$p_report` for the period 1950 -- 2010) that appears to be part of a long-term signal, judging by the statistically significant trend observed in the ERA20C reanalysis from 1900 to 2010 (p-value `r trends_all[part == "Real" & PC == "PC1" & dataset == "era20c"]$p_report`) with similar magnitude during the period of overlap.
+There is a downward trend in the real part of EOF1 in the hybrid reanalysis (Figure \@ref(fig:extended-series)a.1, p-value `r trends[part == "Real" & PC == "PC1" & dataset != "era20c"]$p_report` for the period 1950 -- 2010) that appears to be part of a long-term signal, judging by the statistically significant trend observed in the ERA20C reanalysis from 1900 to 2010 (p-value `r trends_all[part == "Real" & PC == "PC1" & dataset == "era20c"]$p_report`) with similar magnitude during the period of overlap.
 
@@
 
-<!-- (ref:extended-magnitude-cap) Same as Figure \@ref(fig:extended-series) but for EOF magnitude defined as $\sqrt{\mathrm{Real}^2 + \mathrm{Imaginary}^2}$.  -->
-
-<!-- ```{r extended-magnitude, fig.cap = "(ref:extended-magnitude-cap)"} -->
-<!-- magnitudes <- extended %>%  -->
-<!--   copy() %>%  -->
-<!--   dcast(time + PC + dataset ~ part, value.var = "hgt") %>%  -->
-<!--   .[, ref := time %between% common_range(time, dataset)] %>% -->
-<!--   .[, hgt := sqrt(Real^2 + Imaginary^2)] %>%  -->
-<!--   .[, hgt := (hgt - mean(hgt[ref]))/sd(hgt[ref]), by = .(PC, dataset)] -->
-
-<!-- correlations_mag <- magnitudes %>%  -->
-<!--   dcast(time + PC  ~ dataset, value.var = "hgt") %>%  -->
-<!--   na.omit() %>% { -->
-<!--     d <- . -->
-<!--     rbind( -->
-<!--   d[, .(correlation = cor(era20c, era5)),  -->
-<!--     by = .(PC, time = ifelse(year(time) >= 1979,  -->
-<!--                                    as.character(mean(periods[2:3])), -->
-<!--                                    as.character(mean(periods[1:2]))))], -->
-
-<!--     d[, .(correlation = cor(era20c, era5)),  -->
-<!--     by = .(PC)] %>%  -->
-<!--     .[, time := as.character(mean(periods[c(1, 3)]))] -->
-<!--     ) -->
-<!--   } %>%  -->
-
-<!--   .[, time := lubridate::as_datetime(time)] -->
-
-<!-- trends_mag <- magnitudes %>%  -->
-<!--   .[time %between% common_range(time, dataset)] %>%  -->
-<!--   .[, FitLm(hgt, time, se = TRUE), by = .(dataset, PC)] %>%  -->
-<!--   rm_intercept() %>%  -->
-<!--   .[, p_val := Pvaluate(estimate, std.error, df)] %>%  -->
-<!--   .[, p_report := report_p(p_val)] -->
-
-<!-- trends_all_mag <- magnitudes %>%  -->
-<!--   .[, FitLm(hgt, time, se = TRUE), by = .(dataset, PC)] %>%  -->
-<!--   rm_intercept() %>%  -->
-<!--   .[, p_val := Pvaluate(estimate, std.error, df)] %>%  -->
-<!--   .[, p_report := report_p(p_val)] -->
-
-<!-- magnitudes %>%  -->
-<!--   ggplot(aes(time, hgt)) + -->
-<!--   # Zero line + rangeframe -->
-<!--   geom_hline(yintercept = 0, size = 0.1, color = "gray30") + -->
-<!--   ggthemes::geom_rangeframe(sides = "lr", size = 0.1, color = "gray30") + -->
-
-<!--   geom_line(aes(color = dataset)) + -->
-
-<!--   # Correlations -->
-<!--   annotate("errorbarh", xmin = periods[1:2], xmax = periods[2:3], y = -2.7, size = 0.3) + -->
-<!--   annotate("errorbarh", xmin = periods[1], xmax = periods[3], y = -4, size = 0.3) + -->
-
-<!--   geom_label(data = correlations_mag[time %in% range(time)], aes(y = -2.7, label = signif(correlation^2, 2)),  -->
-<!--              label.r = grid::unit(0, "lines"), label.size = 0) + -->
-<!--   geom_label(data = correlations_mag[!(time %in% range(time))], aes(y = -4, label = signif(correlation^2, 2)),  -->
-<!--              label.r = grid::unit(0, "lines"), label.size = 0) + -->
-
-<!--     geom_smooth(data = ~.x[dataset == "era20c"],  -->
-<!--               method = "lm", se = FALSE,  aes(color = dataset)) + -->
-<!--   geom_smooth(data = ~copy(.x)[time %between% common_range(time, dataset)], -->
-<!--               aes(color = dataset), method = "lm", se = FALSE) + -->
-
-<!--   scale_x_datetime(NULL) + -->
-<!--   scale_y_continuous("Standarised EOF magnitude") + -->
-<!--   scale_color_brewer(NULL, palette = "Dark2", labels = c(era20c = "ERA20C", -->
-<!--                                                            era5 = "ERA5 BE"), -->
-<!--                      guide = guide_legend(override.aes = list(size = 1), order = 1)) + -->
-<!--   facet_wrap(~PC, labeller = labeller(PC = lab_PC), ncol = 2) + -->
-<!--   theme(panel.spacing = grid::unit(1.5, "lines"))  + -->
-
-<!--   tag_facets() -->
-<!-- ``` -->
-
 ## Regression
@@
 ```{r compute-regression}
-regr <- ceof[, eof[[1]]$left, by = season] %>%
+regr <- ceof[, eof[[1]]$left, by = .(season)] %>%
   .[season == "SON"] %>%
@@
   .[, hgt := NULL] %>%
-  .[, `:=`(R = as.numeric(scale(R)), I = as.numeric(scale(I))), by = .(season, PC)] %>%
+  .[, `:=`(R = as.numeric(scale(R)), I = as.numeric(scale(I))), 
+    by = .(season, PC)] %>%
   dcast(time ~ PC, value.var = c("R", "I")) %>%
@@
   # .[variable == "hgt"] %>%
-  .[, FitLm(value, I_PC1, R_PC1, I_PC2, R_PC2, se = TRUE), by = .(lev, lon, lat, variable)] %>%
+  .[, FitLm(value, I_PC1, R_PC1, I_PC2, R_PC2, se = TRUE), 
+    by = .(lev, lon, lat, variable)] %>%
   .[, c("term", "PC") := tstrsplit(term, "_")] %>%
@@
 ```
-Figure \@ref(fig:ref:eof1-regr-gh) shows regression patterns of SON geopotential height on the real and imaginary part of the first EOF at 50 hPa and 200 hPa.
+
+Figure \@ref(fig:eof1-regr-gh) shows regression patterns of SON geopotential height on the real and imaginary part of the first EOF at 50 hPa and 200 hPa.
 
@@
 
-* Parte real de EOF1: variabilidady que se proyecta en la dirección del campo medio
-
-* Parte imaginaria de EOF1: variabilidad asociada a la parte simétrica del SAM
-
-* EOF2: PSA1 y PSA2.  
+-   Parte real de EOF1: variabilidady que se proyecta en la dirección del campo medio
+
+-   Parte imaginaria de EOF1: variabilidad asociada a la parte simétrica del SAM
+
+-   EOF2: PSA1 y PSA2.\
+
 -   Parte imaginaria: proyecta sobre la parte asimétrica del SAM (PSA2)
+
 -   Parte real: relativamente independiente del SAM
@@
 ## Relationship with other indices
+
+### SAM
 
@@
                                 bg.colour = "white", bg.r = 0.2,
-                                size = 5,
+                                size = 5, show.legend = FALSE,
                                 aes(x, y, color = part, label = paste0("r² = ", round(cor^2, 2)))) +
@@
 
-```{r sam-eof, fig.cap = "Standarised temporal EOFs and SAM index computed as the leading EOF of geopotential height anomalies at 700 hPa."}
-plot_with_index(ceof, sam[lev == 700 & symmetry == "full"][, .(time, sam)])
-```
-
-(ref:sam-eof-vertical-cap) Coefficient of determination between the real and imaginary part of each EOF and the SAM, Asymmetric SAM and Symmetric SAM indices computed at each level according to Campitelli on press??.
+(ref:sam-eof-vertical-cap) Coefficient of determination between the real and imaginary part of each EOF and the SAM, Asymmetric SAM (A-SAM) and Symmetric SAM (S-SAM) indices computed at each level according to @campitelli2020a.
+Points mark estimates with p-value \< 0.01 corrected for False Detection Rate [@benjamini1995].
 
 ```{r sam-eof-vertical, fig.cap = "(ref:sam-eof-vertical-cap)"}
-time <- ceof[, eof[[1]]$left] %>%
+time <- ceof[, eof[[1]]$left, by = .(season)] %>%
   sep_ReIm() %>%
@@
 
-sams <- readRDS(data_path("derived", "sam.Rds"))
-
-sams %>% 
-  .[time, on =.NATURAL, allow.cartesian = TRUE] %>%
-  na.omit() %>%
-  .[, cor(hgt, estimate)^2, by = .(lev, term, part, PC)] %>%
-  ggplot(aes(lev, V1)) +
+sams <- readRDS(data_path("derived", "sam.Rds")) %>% 
+  .[season(time) == "SON"] %>%
+  .[, dim :=  lubridate::days_in_month(time)] %>% 
+  .[, .(estimate = weighted.mean(estimate, dim)),
+    by = .(lev, term, time = seasonally(time))]
+
+sam_r2 <- sams %>% 
+  .[time, on = .NATURAL, allow.cartesian = TRUE] %>%
+  .[, with(cor.test(hgt, estimate), 
+           list(estimate = estimate, 
+                low = conf.int[1],
+                high = conf.int[2],
+                p_value = p.value)), by = .(lev, term, part, PC)] %>%
+
+  .[, low2 := pmin(low^2, high^2, ifelse(sign(high) != sign(low), 0, 1))] %>% 
+  # .[low2 <= 0, low2 := 0] %>% 
+  .[, high2 := pmax(low^2, high^2)] %>% 
+  .[PC != "PC3"]
+
+max_r2 <- sam_r2[part == "Imaginary" & term == "asym"][, .SD[which.max(estimate^2)],
+                                                       by = .(PC)]
+
+sam_r2 %>% 
+  ggplot(aes(lev, estimate^2)) +
+  # geom_ribbon(data =  ~.x[p.adjust(p_value, "fdr") < 0.001], 
+  # aes(ymin = low2, ymax = high2, color = term), alpha= 0.1) +
+  geom_point(data = ~.x[p.adjust(p_value, "fdr") < 0.01], aes(color = term)) +
   geom_line(aes(color = term, size = term)) +
   scale_x_level() +
-  scale_y_continuous("r²") +
+  scale_y_continuous("r²", limits = c(0, 1)) +
   scale_size_manual(guide = "none",
-                    values = c(full = 1,
+                    values = c(full = 1.2,
                                asym = 0.5,
@@
   scale_color_brewer(NULL, palette = "Dark2", labels = c(full = "SAM",
-                                                         asym = "Asymmetric SAM",
-                                                         sym = "Symmetric SAM")) +
+                                                         asym = "A-SAM",
+                                                         sym = "S-SAM")) +
   facet_grid(part ~ PC, labeller = labeller(PC = c(PC1 = "EOF1", PC2 = "EOF2"))) +
-  coord_flip()
-
-```
-
-```{r}
-sams <- ERA5() %>% 
-  ReadNetCDF(vars = c("hgt" = "z"), 
-             subset = list(latitude = c(-90, -20))) %>% 
-  normalise_coords() %>% 
-  .[season(time) == "SON"] %>% 
-  .[, .(hgt = mean(hgt)), by = .(lon, lat, lev, time = seasonally(time))] %>% 
-  .[, hgt := Anomaly(hgt), by = .(lon, lat, lev)] %>% 
-  .[, asymsam::eof_asym(hgt, lon, lat, time), by = lev]
-```
-
-```{r}
-sams %>% 
-  .[, PC := NULL] %>% 
- .[time, on =.NATURAL, allow.cartesian = TRUE] %>%
-  na.omit() %>%
-  .[, cor(hgt, estimate)^2, by = .(lev, term, part, PC)] %>%
-  ggplot(aes(lev, V1)) +
-  geom_line(aes(color = term, size = term)) +
-  scale_x_level() +
-  scale_y_continuous("r²") +
-  scale_size_manual(guide = "none",
-                    values = c(full = 1,
-                               asym = 0.5,
-                               sym = 0.5)) +
-  scale_color_brewer(NULL, palette = "Dark2", labels = c(full = "SAM",
-                                                         asym = "Asymmetric SAM",
-                                                         sym = "Symmetric SAM")) +
-  facet_grid(part ~ PC, labeller = labeller(PC = c(PC1 = "EOF1", PC2 = "EOF2"))) +
-  coord_flip()
-
-```
-
-(ref:oni-eof-cap) Same as Figure \@ref(fig:sam-eof) but for the Oceanic Enso Index (REF??).
-
-```{r oni-eof, fig.cap = "(ref:oni-eof-cap)"}
-plot_with_index(ceof, enso)
-```
-
-::: notes
-
-En la tropósfera, la pate imaginaria del PC2 captura la parte asimétrica del SAM.
-La parte real es muy parecida en estructura, pero como está desfasada, no coincide con el SAM y entonces no proyecta fuerte en él.
-
-La Parte imaginaria está muy influenciada por ENSO, como se ve en la relación con el íncide y en la regresión con la SST.
-La fase Real parece no tener influencia del océano tropical.
-Sin embargo, los flujos de acción de onda.
-
-A su vez, la parte imaginaria del PC1 captura también parte del SAM, pero principalmente la parte simétrica.
-
-:::
-
-```{r compute-psi}
-psi <- ReadNetCDF(ERA5(), vars = "vo",
-                  subset = list(level = c(50, 200),
-                                time = c("1979-01-01", "2019-12-01"))) %>%
-  normalise_coords() %>%
-  .[season(time) == "SON"] %>%
-  .[, .(vo = mean(vo)), by = .(lon, lat, lev,time = seasonally(time))] %>%
-  .[, solve_poisson(vo, lon, lat), by = .(lev, time)] %>%
-  .[, psi.z := Anomaly(value), by = .(time, lev, lat)] %>%
-  .[, value := NULL]
-```
+  coord_flip() +
+  tag_facets("rc")
+```
+
+Figure \@ref(fig:sam-eof-vertical) shows the coefficient of determination between the EOFs and three SAM indices.
+The SAM index represents the leading EOF of monthly geopotential height fields south of 20ºS at each level.
+The A-SAM and S-SAM indices represent, respectively, the zonally asymmetric and symmetric component of the SAM and are obtained by projecting the zonally asymmetric and zonally symmetric part of the SAM spatial pattern onto monthly geopotential height fields (@campitelli2020a).
+Since the analysis here is only for the SON trimester, monthly values were averaged across semesters (weighted by the number of days in each month).
+
+Both EOFs bear some modest relationship with the SAM index (thick green line in Figure \@ref(fig:sam-eof-vertical)), although not statistically significant at every level.
+The split between A-SAM and S-SAM gives more insight into the nature of the relationship.
+The relationship between the SAM and the imaginary EOF1 (Figure \@ref(fig:sam-eof-vertical).b1) is mediated by S-SAM in the troposphere, but by the A-SAM in the stratosphere.
+
+The Imaginary EOF2 is related with the SAM through the A-SAM in the troposphere, with up to `r scales::percent(max_r2[PC == "PC2"]$estimate, 2)` of shared variance, reached at `r max_r2[PC == "PC2"]$lev` hPa (Figure \@ref(fig:sam-eof-vertical).b2).
+The Imaginary EOF2 is, then, an almost a pure representation of the tropospheric asymmetric part of the SAM.
+Since the Real EOF2 represents a quarter-wavelength rotation of the same zonal wave 3 structure, the Imaginary part is the phase that coincides with the SAM.
+As a corollary, this means that the asymmetric component of the SAM represents only one phase of a zonally varying zonal wave.
 
 ```{r regr_psi}
-regr_psi <- ceof[, eof[[1]]$left, by = season] %>%
+regr_psi <- ceof[, eof[[1]]$left, by = .(season)] %>%
   sep_ReIm(hgt, "wide") %>%
@@
   .[, psi.z := estimate] %>%
-  .[, c("fx", "fy") := shceof::WaveFlux(.SD, p = 200), by = .(lev, PC, season, term)]
+  .[, c("fx", "fy") := shceof::WaveFlux(.SD, p = 200), by = .(lev, PC, season, term)] %>% 
+  .[, term := factor(term, levels = c("R", "I"), labels = c("Real", "Imaginary"),
+                          ordered = TRUE)]
 ```
@@
 ```{r regr_sst}
-regr_sst <- ceof[, eof[[1]]$left, by = season] %>%
+regr_sst <- ceof[, eof[[1]]$left, by = .(season)] %>%
   sep_ReIm(hgt, "wide") %>%
@@
   .[, FitLm(t, I, R, se = TRUE), by = .(lon, lat, PC, season)] %>%
-  rm_intercept()
+  rm_intercept() %>% 
+  .[, term := factor(term, levels = c("R", "I"), labels = c("Real", "Imaginary"),
+                          ordered = TRUE)]
 ```
@@
     .[, var := "SST"]
+
   regr_psi %>% 
@@
     ggplot(aes(lon, lat)) +
-    geom_contour_fill(aes(z = estimate*1e8, fill = ..level..),  breaks = AnchorBreaks(exclude = 0)) +
+    geom_contour_fill(aes(z = estimate*1e8, fill = ..level..),
+                      breaks = AnchorBreaks(exclude = 0)) +
     geom_contour2(aes(z = p_val), breaks = 0.05, size = 0.1) +
-    stat_subset(aes(subset = p_val <= 0.05 & is.cross(lon, lat, 2)),
+    stat_subset(aes(subset = p_val <= 0.05 & is.cross(lon, lat, 1)),
                 geom = "point", size = 0.2, alpha = 0.5) +
@@
     stat_subset(data = sst_gdata,
-                aes(subset = p_val <= 0.05 & is.cross(lon, lat, 2)),
+                aes(subset = p_val <= 0.05 & is.cross(lon, lat, 3)),
                 geom = "point", size = 0.2, alpha = 0.5) +
@@
     scale_mag(guide = "none") +
-    facet_grid(term ~ var, labeller = labeller(term = lab_cplx, 
-                                               PC = c(PC1 = "EOF1", PC2 = "EOF2"))) +
+    facet_grid(term ~ var, labeller = labeller(PC = c(PC1 = "EOF1", PC2 = "EOF2"))) +
     coord_quickmap(ylim = c(NA, 10)) +
@@
     # ggh4x::facet_nested(term ~ PC) +
-    axis_labs_smol
-}
-```
-
-```{r sst-psi-1,  fig.cap = "Regresión con PSI en 200 hPa y sst, PC1"}
-plot_sst_psi(which_pc = "PC1", which_lev = 200)
+    axis_labs_smol +
+    tag_facets("cr")
+}
 ```
@@
 plot_sst_psi(which_pc = "PC2", which_lev = 200) 
-```
-
-```{r mean-psi, fig.cap = "mean streamfunction"}
-psi %>% 
-  .[lev == 200, .(psi.z = -mean(psi.z)), by = .(lon, lat)] %>% 
-  .[, c("fx", "fy") := WaveFlux(.SD)] %>% 
-  .[lat <= 10] %>% 
-  ggplot(aes(lon, lat)) +
-  geom_contour_fill(aes(z = psi.z)) +
-  geom_streamline(aes(dx = fx, dy = fy), min.L = 3, L = 10, n = 30,
-                  size = 0.2, res = 2) +
-  geom_qmap(keep = .015) +
-  scale_fill_divergent(NULL, guide = guide_colorstrip_bottom(width = 15)) +
-  scale_x_longitude() +
-  scale_y_latitude()  +
-  coord_quickmap(ylim = c(-90, 10))
-```
-
-### Precipitation
+# +   annotate("rect", xmin = 170, xmax = 195, ymin = -25, ymax = -5, color = "black", fill = NA)
+```
+
+???
+-\> esta intro hay que cambiarla.
+To help identify those sources Figure \@ref(fig:sst-psi-2) shows regression maps between EOF2 and STT and Streamfunction at 200 hPa.
+As expected from the correlations between each component of EOF2 and ENSO, Imaginary EOF2 is associated with strong ENSO-like SST anomalies and streamfunction (Figure \@ref(fig:sst-psi-2).a2).
+Streamfunction anomalies (Figure \@ref(fig:sst-psi-2).b2) are indicative of tropical forcing, which is compatible with SST anomalies.
+Real EOF2, on the other hand, is not associated with significant SST anomalies (Figure \@ref(fig:sst-psi-2).a1), as expected by its lack of relationship with ENSO, and streamfunction anomalies in the tropics are much weaker and non significant.
+
+```{r enso_cor}
+enso_cor <-  ceof[, eof[[1]]$left, by = .(season)] %>%
+    sep_ReIm() %>%
+    .[, hgt := scale(hgt), by = .(PC, part)] %>%
+    enso[., on = "time"] %>%
+    na.omit() %>% 
+   .[, with(cor.test(hgt, oni), list(estimate = estimate,
+                                     p.value = p.value,
+                                     low = conf.int[1],
+                                     hig = conf.int[2])), 
+     by = .(PC, part)] %>% 
+  .[, cor_text := paste0(signif(estimate, 2), " (CI: ", 
+                         signif(low, 2), " -- ", signif(hig, 2), ")")]
+
+```
+
+```{r enso-phase, fig.cap = "Relationship between ENSO and phase of EOF2 for the period 1979 -- 2019.
+Colours denote years with magnitude of EOF2 greater or smaller than the 50th percentile.
+Black line is the fit ONI ~ cos(phase) + sin(phase) computed by OLS weighted by the magnitude of EOF2."}
+ceof[, eof[[1]]$left, by = .(season)] %>% 
+  .[, arg := Arg(hgt)] %>% 
+  .[, mag := Mod(hgt)] %>% 
+  .[enso, on = "time"] %>% 
+  na.omit() %>% 
+  .[PC == "PC2"] %>% 
+  .[year(time) >= 1979] %>%
+  qwrap(arg = c(-pi, pi) ~ c(-2*pi, 2*pi)) %>% 
+  ggplot(aes(arg, oni)) +
+  geom_hline(yintercept = c(-0.5, 0, 0.5), size = 0.2, color = "gray50") +
+  geom_vline(xintercept = 0, size = 0.2, color = "gray50") +
+  geom_smooth(method = "lm", aes(weight = mag),
+              formula = y ~ I(cos(x)) + I(sin(x)), 
+              se = FALSE, color = "black", size = 0.5) +
+  geom_point(aes(colour = mag > quantile(mag, 0.5)), size = 3) +
+  scale_color_brewer(NULL, palette = "Dark2", 
+                     labels = c("TRUE" = "|EOF2| > 50%",
+                                "FALSE" = "|EOF2| < 50%")) +
+  scale_y_continuous("ONI") +
+  scale_x_continuous("Phase", 
+                     breaks = seq(-pi, pi, by = 30*pi/180), 
+                     labels = function(x) paste0(round(x*180/pi, 5), "º")) +
+  coord_cartesian(xlim = c(-pi, pi))
+```
+
+The zonal wave represented by the EOF2 pattern is very similar to the wave train associated with ENSO.
+Indeed, the correlation between the Imaginary EOF2 and the Oceanic ENSO Index ([@bamston1997]) is `r enso_cor[PC == "PC2" & part == "Imaginary"]$cor_text`.
+The correlation between the Real EOF2 and ENSO is not significant (`r enso_cor[PC == "PC2" & part == "Real"]$cor_text`).
+This suggests that, even though both EOF2 patterns represent the same wave train with a different phase, these two phases are exited by different sources.
+
+Figure \@ref(fig:sst-psi-2) suggests, then, that the wave train represented by EOF2 can be forced both by the tropics and internally, and that it has preferred phases in each case.
+When it is forced by the tropics, it aligns roughly with the Imaginary phase, and when it's the Real phase is closer to the internally-varying wave.
+However, Figure \@ref(fig:sst-psi-2).b1 also indicates that this separation is not completely clean.
+It's not expected..
+??
+
+## Precipitation
 
@@
 
+# Appendix
+
+## Chosen rotations of the EOFs
+
+```{r rotations-pc1, fig.cap = "Rotations of PC1"}
+rotations_pc1 %>% 
+  ggplot(aes(rotation, mean)) +
+  geom_hline(yintercept = 0, color = "gray") +
+  geom_line(aes(color = part), size = 1) +
+  scale_x_continuous("Rotation", breaks = seq(-pi, pi, by = 30*pi/180), 
+                     labels = function(x) paste0(round(x*180/pi, 5), "º")) +
+  scale_y_continuous("Mean value") +
+  scale_color_brewer(NULL, palette = "Dark2", labels = lab_cplx)
+```
+
+```{r regr-rotation, fig.cap = "Regression with streamfunction PC2"}
+regr_psi %>% 
+  .[PC == "PC2"] %>%
+  .[lev == 200] %>%
+  .[lat %between% c(-90, 10)] %>%
+  periodic(lon = c(0, 360)) %>%
+  ggplot(aes(lon, lat)) +
+  geom_contour_fill(aes(z = estimate*1e8, fill = ..level..),
+                    breaks = AnchorBreaks(exclude = 0)) +
+  geom_contour_tanaka(aes(z = estimate*1e8),
+                    breaks = AnchorBreaks(exclude = 0)) +
+
+  geom_qmap(keep = .015) +
+  scale_fill_divergent_discretised("Streamfunction", guide = "none") +
+
+  scale_x_longitude() +
+  scale_y_latitude() +
+  scale_mag(guide = "none") +
+  facet_grid(term ~ ., labeller = labeller(PC = c(PC1 = "EOF1", PC2 = "EOF2"))) +
+  coord_quickmap(ylim = c(NA, 10)) +
+  theme(legend.box = "horizontal") +
+  axis_labs_smol +
+  annotate("rect", 
+           xmin = box$lon[1], xmax = box$lon[2], 
+           ymin = box$lat[1], ymax = box$lat[2], color = "black", fill = NA)
+```
+
+```{r rotations-pc2, fig.cap = "R2 between EOF2 Real and Imaginary parts and zonal anomalies of streamfunction at BOX?? for different rotation parameters of the real and imaginary parts."}
+rotations_pc2 %>% 
+  melt(id.vars = "rotation") %>% 
+  ggplot(aes(rotation, value^2)) +
+  geom_hline(yintercept = 0, color = "gray") +
+  geom_line(aes(color = variable), size = 1) +
+  scale_x_continuous("Rotation", breaks = seq(-pi, pi, by = 30*pi/180), 
+                     labels = function(x) paste0(round(x*180/pi, 5), "º")) +
+  scale_y_continuous("r^2") +
+  scale_color_brewer(NULL, palette = "Dark2", labels = lab_cplx)
+```
+
 # References

### 2021-02-08 R/0-download.R

@@ -66,7 +66,7 @@ request_era5be <- list(
 request_era5 <- list(
   format = "netcdf",
   product_type = "monthly_averaged_reanalysis",
-  variable = c("geopotential", "temperature", "vorticity"),
+  variable = c("geopotential", "temperature", "vorticity", "divergence"),
   pressure_level = c("1", "2", "3", "5", "7", "10", "20", "30", "50",
                      "70", "100", "125", "150", "175", "200", "225",
                      "250", "300", "350", "400", "450", "500", "550",

### 2021-02-08 R/ReIm.R

@@ -39,7 +39,8 @@ sep_ReIm <- function(data, column, format = c("longer", "wider")) {
   if (format[1] == "longer") {
     data[, c(col) := NULL]
     data <- data.table::setDT(tidyr::pivot_longer(data, R:I, names_to = "part", values_to = col))
-    data[, part := factor(part, levels = c("R", "I"), labels = c("Real", "Imaginary"))]
+    data[, part := factor(part, levels = c("R", "I"), labels = c("Real", "Imaginary"),
+                          ordered = TRUE)]
   }
 
   return(data[])

### 2021-02-08 R/theme.R

@@ -13,7 +13,7 @@ theme_shceof <- function(base_size = 16) {
       # text = element_text(family = font_rc),
       legend.position = "bottom", legend.box = "vertical",
       tagger.panel.tag.background = ggplot2::element_rect(color = NA),
-      strip.text = ggplot2::element_text(size = 12),
+      strip.text = ggplot2::element_text(size = ggplot2::rel(12/16)),
       # panel.spacing.y = grid::unit(5, "mm"),
       # panel.spacing.x = grid::unit(5, "mm"),
       # legend.spacing = grid::unit(2, "mm"),

### 2021-02-08 analysis/scripts/02-compute-sam.R

@@ -18,9 +18,7 @@ if (file.exists(sam_file)) {
     normalise_coords() %>%
     .[, hgt := Anomaly(hgt), by = .(lon, lat, lev, month(time))] %>%
     .[, asymsam::eof_asym(hgt, lon, lat, time), by = lev] %>%
-    rm_singleton() %>%
-    .[season(time) == "SON"] %>%
-    .[, .(estimate = mean(estimate)), by = .(lev, term, time = seasonally(time))]
+    rm_singleton()
 
   saveRDS(sams, sam_file)
 }

### 2021-02-10 analysis/paper/paper.Rmd
Tabla

@@
 library(shceof)
+library(kableExtra)
 
@@
 
-If
-
 # Results
+
+No mostrar los EOFs no complejos. No tiene sentido mostralos si no se analizas.
+
+Las correlaciones entre PCs Podrían ir en una tabla. Quizás suplementaria.
+
+Relacionar 50hPa con vórtice polar? Temperatura también suma. 
+En 50 el negativo es más intenso que el negativo. ¿Por qué?
+
+Ver qué pasa con U con el EOF1.
+
+En 200 no se ve la señal de jet de la temperatura ¿qué pasa más abajo? 
+en 700 la anomalía de temp coincide más o menos con el geopotentical
+
+ondas antes e la tempreatura
+
+Mostrar EOF espaciales.
+
+Tirar ERA20C a la basura. ERA20c suplementaria?
+
+Pp, ¿cuál es la regresión con SAM asim en 200 hPa?
 
@@
 
-(ref:cap-eof-naive) Leading 4 EOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2009.
-
-(ref:subcap-eof-naive) Spatial patterns (arbitrary units). The numbers at the top-left of each panel is the variance explained by each EOF.
-
-(ref:subcap-cor-eof-naive) Coefficient of determination of the temporal index of each EOF between levels.
-
-```{r cor-eof-naive, fig.cap = "(ref:cap-eof-naive)", fig.subcap = c("(ref:subcap-eof-naive)", "(ref:subcap-cor-eof-naive)"), fig.ncol = 1}
-eof_naive[, eof[[1]]$right, by = lev] %>% 
-  periodic(lon = c(0, 360)) %>% 
-  ggplot(aes(lon, lat)) +
-
-  geom_contour_fill(aes(z = hgt, fill = ..level..), 
-                    breaks = AnchorBreaks(0, 0.01, exclude = 0)) +
-  geom_contour_tanaka(aes(z = hgt), breaks = AnchorBreaks(0, 0.01, exclude = 0)) +
-
-  geom_qmap(~.x[lat <= -20]) +
-  scale_fill_divergent_discretised(guide = "none")  +
-  scale_y_latitude() +
-  scale_x_longitude() +
-  coord_polar() +
-  facet_grid(lev~EOF, labeller = labeller(lev = lab_lev))  +
-  axis_labs_smol +
-  tagger::tag_facets(tag = "panel", tag_pool = explained, tag_suffix = "")
+```{r}
+
+background_palette <- grDevices::colorRampPalette(c("white", scales::muted("red")), 
+                                                  space = "Lab")
 
@@
   .[, item2 := tstrsplit(item2, "_")[[2]]] %>% 
-  ggplot(aes(item1, item2)) +
-  geom_raster(aes(fill = correlation^2)) +
-  shadowtext::geom_shadowtext(aes(label = round(correlation^2, 2)), 
-                              color = "black", bg.colour = "white", bg.r = 0.2) +
-  scale_fill_gradient(guide = "none", low = "white", high = scales::muted("red"), limits = c(0, 1)) +
-  scale_x_discrete("200 hPa") +
-  scale_y_discrete("50 hPa") +
-  plot_layout(ncol = 1)
+  .[, r2 := round(correlation^2, 2)] %>% 
+  dcast(item1 ~ item2, value.var = "r2") %>% 
+  setnames("item1", "200 hPa") %>% 
+  kbl2(booktabs = TRUE, caption = "$R^2$ between pairs of EOFs of each level rounded to two decimals.") %>% 
+  # add_header_above(c("", "50 hPa" = 4)) %>% 
+  columns_spec(2:5,  background = spec_color(.table[[.col]], option = background_palette, 
+                                              scale_from = c(0, 1)),
+               color = ifelse(.table[[.col]] > .5, "white", "black")) %>% 
+  add_header_above(c("", "50 hPa" = 4)) %>%
+  kable_classic()
 ```
@@
 
-(ref:cap-splitted) Coefficient of determination of the absolute magnitude of complex EOFs between 200 hPa and 50 hPa computing EOF separatedly for each level.
-The high correlation between levels for the two leading EOFs justifies treating the pattern as a mode of covariability between the stratosphere and the troposphere and, thus, computing the EOFs using both levels at the same time.
-(probably not to be shown.)
-
-```{r corr-ceof-splitted, fig.cap = "(ref:cap-splitted)"}
+```{r corr-ceof-splitted}
 ceof_splitted[, eof[[1]]$left, by = .(lev = levs)] %>%
@@
   .[, item2 := tstrsplit(item2, "_")[[2]]] %>%
-  ggplot(aes(item1, item2)) +
-  geom_raster(aes(fill = correlation^2)) +
-  shadowtext::geom_shadowtext(aes(label = round(correlation^2, 2)),
-                              color = "black", bg.colour = "white", bg.r = 0.2) +
-  scale_fill_gradient(guide = "none", low = "white", high = scales::muted("red")) +
-  scale_x_discrete("200 hPa") +
-  scale_y_discrete("50 hPa")
+  .[, r2 := round(correlation^2, 2)] %>%
+  dcast(item1 ~ item2, value.var = "r2") %>%
+  setnames("item1", "200 hPa") %>%
+  kbl2(booktabs = TRUE, caption = "$R^2$ of the absolute magnitude of complex EOFs between 200 hPa and 50 hPa computing EOF separatedly for each level.
+The high correlation between levels for the two leading EOFs justifies treating the pattern as a mode of covariability between the stratosphere and the troposphere and, thus, computing the EOFs using both levels at the same time.") %>%
+  columns_spec(2:4, background = spec_color(.table[[.col]], option = background_palette,
+                                            scale_from = c(0, 1)),
+               color = ifelse(.table[[.col]] > .5, "white", "black")) %>% 
+  add_header_above(c("", "50 hPa" = 3)) %>%
+  kable_classic() 
 ```
@@
 ```{r mean-anom, fig.cap = "Mean SON zonal geopotential height anomalies at (a) 50 hPa and (b) 200 hPa"}
-era5 %>% 
-  .[, mean(-hgt), by = .(lon, lat, lev)] %>% 
-  .[lev %in% c(50, 200)] %>% 
-  .[lat <= -20] %>% 
-  .[, V2 := V1 - mean(V1), by = .(lat, lev)] %>% 
-  periodic(lon = c(0, 360)) %>% 
-  ggplot(aes(lon, lat)) +
-  geom_contour_fill(aes(z = V2), global.breaks = FALSE, breaks = AnchorBreaks(exclude = 0)) +
-  geom_contour_tanaka(aes(z = V2), global.breaks = FALSE, breaks = AnchorBreaks(exclude = 0)) +
-  geom_qmap(~ .x[lat <= -20]) +
-  scale_fill_divergent(NULL, breaks = AnchorBreaks(exclude = 0), 
-                       guide = guide_colorstrip_bottom()) +
-  scale_x_longitude() +
-  scale_y_latitude() +
-  facet_grid(~lev, labeller = labeller(lev = lab_lev)) +
-  coord_polar() +
-  axis_labs_smol +
-  tag_facets()
+# era5 %>%
+#   .[, mean(-hgt), by = .(lon, lat, lev)] %>%
+#   .[lev %in% c(50, 200)] %>%
+#   .[lat <= -20] %>%
+#   .[, V2 := V1 - mean(V1), by = .(lat, lev)] %>%
+#   periodic(lon = c(0, 360)) %>%
+#   ggplot(aes(lon, lat)) +
+#   geom_contour_fill(aes(z = V2), global.breaks = FALSE, breaks = AnchorBreaks(exclude = 0)) +
+#   geom_contour_tanaka(aes(z = V2), global.breaks = FALSE, breaks = AnchorBreaks(exclude = 0)) +
+#   geom_qmap(~ .x[lat <= -20]) +
+#   scale_fill_divergent(NULL, breaks = AnchorBreaks(exclude = 0),
+#                        guide = guide_colorstrip_bottom()) +
+#   scale_x_longitude() +
+#   scale_y_latitude() +
+#   facet_grid(~lev, labeller = labeller(lev = lab_lev)) +
+#   coord_polar() +
+#   axis_labs_smol +
+#   tag_facets()
 ```
@@
 
+```{r}
+
+```
+
 Figure \@ref(fig:extended-series) shows temporal series of the two complex EOFs extended beyond the satellite era using the preliminary ERA5 back extension going back to 1950 (which we call the "hybrid ERA5" reanalysis) and using ERA20C for the period 1900 -- 2010.
@@
 
-The ERA20C time-series has a also significant upward linear trend in the imaginary part of EOF1 (Figure \@(ref:fig-extended-series)a.2, p-value `r trends[part == "Imaginary" & PC == "PC1" & dataset == "era20c"]$p_report`.
+The ERA20C time-series has a also significant upward linear trend in the imaginary part of EOF1 (Figure \@ref(fig:extended-series)a.2, p-value `r trends[part == "Imaginary" & PC == "PC1" & dataset == "era20c"]$p_report`.
 However, there is no trend in the hybrid dataset during the period of overlap (p-value `r trends[part == "Imaginary" & PC == "PC1" & dataset != "era20c"]$p_report`) and the correlations between the old and modern series are much poorer for this variable.
@@
 
-(ref:sam-eof-vertical-cap) Coefficient of determination between the real and imaginary part of each EOF and the SAM, Asymmetric SAM (A-SAM) and Symmetric SAM (S-SAM) indices computed at each level according to @campitelli2020a.
+(ref:sam-eof-vertical-cap) Coefficient of determination between the real and imaginary part of each EOF and the SAM, Asymmetric SAM (A-SAM) and Symmetric SAM (S-SAM) indices computed at each level according to @campitelli2021.
 Points mark estimates with p-value \< 0.01 corrected for False Detection Rate [@benjamini1995].
@@
   .[time, on = .NATURAL, allow.cartesian = TRUE] %>%
-  .[, with(cor.test(hgt, estimate), 
-           list(estimate = estimate, 
-                low = conf.int[1],
-                high = conf.int[2],
-                p_value = p.value)), by = .(lev, term, part, PC)] %>%
-
-  .[, low2 := pmin(low^2, high^2, ifelse(sign(high) != sign(low), 0, 1))] %>% 
-  # .[low2 <= 0, low2 := 0] %>% 
-  .[, high2 := pmax(low^2, high^2)] %>% 
-  .[PC != "PC3"]
+  .[, correlate(hgt, estimate), by = .(lev, term, part, PC)]
 
@@
   # aes(ymin = low2, ymax = high2, color = term), alpha= 0.1) +
-  geom_point(data = ~.x[p.adjust(p_value, "fdr") < 0.01], aes(color = term)) +
+  geom_point(data = ~.x[p.adjust(p.value, "fdr") < 0.01], aes(color = term)) +
   geom_line(aes(color = term, size = term)) +
@@
 The SAM index represents the leading EOF of monthly geopotential height fields south of 20ºS at each level.
-The A-SAM and S-SAM indices represent, respectively, the zonally asymmetric and symmetric component of the SAM and are obtained by projecting the zonally asymmetric and zonally symmetric part of the SAM spatial pattern onto monthly geopotential height fields (@campitelli2020a).
+The A-SAM and S-SAM indices represent, respectively, the zonally asymmetric and symmetric component of the SAM and are obtained by projecting the zonally asymmetric and zonally symmetric part of the SAM spatial pattern onto monthly geopotential height fields (@campitelli2021).
 Since the analysis here is only for the SON trimester, monthly values were averaged across semesters (weighted by the number of days in each month).
@@
     axis_labs_smol +
-    tag_facets("cr")
-}
-```
-
-```{r sst-psi-2,  fig.cap = "Regresión con PSI en 200 hPa y sst, PC1"}
-plot_sst_psi(which_pc = "PC2", which_lev = 200) 
-# +   annotate("rect", xmin = 170, xmax = 195, ymin = -25, ymax = -5, color = "black", fill = NA)
-```
-
-???
--\> esta intro hay que cambiarla.
-To help identify those sources Figure \@ref(fig:sst-psi-2) shows regression maps between EOF2 and STT and Streamfunction at 200 hPa.
-As expected from the correlations between each component of EOF2 and ENSO, Imaginary EOF2 is associated with strong ENSO-like SST anomalies and streamfunction (Figure \@ref(fig:sst-psi-2).a2).
-Streamfunction anomalies (Figure \@ref(fig:sst-psi-2).b2) are indicative of tropical forcing, which is compatible with SST anomalies.
-Real EOF2, on the other hand, is not associated with significant SST anomalies (Figure \@ref(fig:sst-psi-2).a1), as expected by its lack of relationship with ENSO, and streamfunction anomalies in the tropics are much weaker and non significant.
+    tag_facets("cr", position = "bl")
+}
+```
 
@@
   na.omit() %>%
-   .[, with(cor.test(hgt, oni), list(estimate = estimate,
-                                     p.value = p.value,
-                                     low = conf.int[1],
-                                     hig = conf.int[2])), 
-     by = .(PC, part)] %>% 
-  .[, cor_text := paste0(signif(estimate, 2), " (CI: ", 
-                         signif(low, 2), " -- ", signif(hig, 2), ")")]
-
-```
+  .[, correlate(hgt, oni), by = .(PC, part)]
+```
+
+```{r sst-psi-2,  fig.cap = "Regresión con PSI en 200 hPa y sst, PC1"}
+plot_sst_psi(which_pc = "PC2", which_lev = 200)
+```
+
+???
+-\> esta intro hay que cambiarla.
+Figure \@ref(fig:sst-psi-2) shows regression maps between EOF2 and Sea Surface Temperatures (SST) Streamfunction at 200 hPa.
+The Imaginary EOF2 is associated with strong negative SST anomalies on the Central Pacific and positive anomalies on an area across the North of Australia and New Zealand, the South Pacific Convergence Zone (SPCZ) (Figure \@ref(fig:sst-psi-2).a2).
+This pattern is almost canonically negative ENSO and indeed, the correlation between the Imaginary EOF2 and the Oceanic ENSO Index ([@bamston1997]) is `r enso_cor[PC == "PC2" & part == "Imaginary"]$cor_text`.
+Streamfunction anomalies show a coherent picture.
+The Imaginary EOF2 is associated with strong wave-like streamfunction anomalies emanating from the tropics (Figure \@ref(fig:sst-psi-2).b2).
+.
+This is consistent with what we know of the effect of ENSO on the extratropics: SST anomalies initiate anomalous convection that excites meridionally-moving Rossby waves.
+
+One could expect that the Real EOF2, being just a different phase of the same wave train would show a similar pattern but, perhaps with a slight translation of the main tropical sources.
+However, the Real EOF2 is associated neither with any significant SST (Figure \@ref(fig:sst-psi-2).a1) nor streamfunction anomalies in the tropics.
+The correlation between the Real EOF2 and ENSO is also not significant (`r enso_cor[PC == "PC2" & part == "Real"]$cor_text`).
 
@@
 Black line is the fit ONI ~ cos(phase) + sin(phase) computed by OLS weighted by the magnitude of EOF2."}
+cor_enso_mag <- ceof[, eof[[1]]$left, by = .(season)] %>%
+  .[, arg := Arg(hgt)] %>%
+  .[, mag := Mod(hgt)] %>%
+  .[enso, on = "time"] %>%
+  na.omit() %>%
+  .[PC == "PC2"] %>%
+  .[, correlate(mag, abs(oni))]
+
 ceof[, eof[[1]]$left, by = .(season)] %>%
@@
   .[PC == "PC2"] %>%
-  .[year(time) >= 1979] %>%
-  qwrap(arg = c(-pi, pi) ~ c(-2*pi, 2*pi)) %>% 
   ggplot(aes(arg, oni)) +
@@
   geom_vline(xintercept = 0, size = 0.2, color = "gray50") +
-  geom_smooth(method = "lm", aes(weight = mag),
+  annotate(shadowtext:::GeomShadowText,
+           label = c("- Real", "- Imaginary", "+ Real", "+ Imaginary"),
+           color = "gray60", bg.colour = "white", bg.r = 0.2,
+           size = 6,
+           y = -2, x = c(-pi, -pi/2, 0, pi/2), angle = 90, hjust = 0) +
+
+  geom_smooth(data = ~qwrap(.x, arg = c(-pi, pi) ~ c(-2*pi, 2*pi)),
+              method = "lm", aes(weight = mag),
               formula = y ~  I(cos(x)) + I(sin(x)),
               se = FALSE, color = "black", size = 0.5) +
-  geom_point(aes(colour = mag > quantile(mag, 0.5)), size = 3) +
+  geom_point(aes(colour = mag >= quantile(mag, 0.5),
+                 shape  = mag >= quantile(mag, 0.5)),
+             size = 4) +
+
   scale_color_brewer(NULL, palette = "Dark2",
+                     labels = c("TRUE"  = "|EOF2| > 50%",
+                                "FALSE" = "|EOF2| < 50%")) +
+  scale_shape_manual(NULL, values = c("TRUE" = 20,
+                                      "FALSE" = 18),
                      labels = c("TRUE"  = "|EOF2| > 50%",
@@
 
-The zonal wave represented by the EOF2 pattern is very similar to the wave train associated with ENSO.
-Indeed, the correlation between the Imaginary EOF2 and the Oceanic ENSO Index ([@bamston1997]) is `r enso_cor[PC == "PC2" & part == "Imaginary"]$cor_text`.
-The correlation between the Real EOF2 and ENSO is not significant (`r enso_cor[PC == "PC2" & part == "Real"]$cor_text`).
-This suggests that, even though both EOF2 patterns represent the same wave train with a different phase, these two phases are exited by different sources.
-
-Figure \@ref(fig:sst-psi-2) suggests, then, that the wave train represented by EOF2 can be forced both by the tropics and internally, and that it has preferred phases in each case.
-When it is forced by the tropics, it aligns roughly with the Imaginary phase, and when it's the Real phase is closer to the internally-varying wave.
-However, Figure \@ref(fig:sst-psi-2).b1 also indicates that this separation is not completely clean.
-It's not expected..
-??
+We show a more complete picture of the relationship between ENSO and EOF2 in Figure \@ref(fig:enso-phase), which plots the ONI index value and the phase of the EOF2 for each year between 1979 and 2019 highlighting years in which the magnitude of EOF2 was higher than 50% of the years.
+In years with positive ENSO, the phase of the EOF2 is always around -90º (corresponding with negative imaginary part) and vice versa.
+In years with near neutral ENSO, the phase of the EOF2 is much more variable between 0º and -180º.
+There is also a relationship between the magnitude of the EOF2 and the magnitude of the ENSO.
+The strong EOF2 years are almost invariably associated with strong ENSO years.
+The correlation between the absolute magnitude of the ONI and the magnitude of the EOF2 is `r cor_enso_mag$cor_text`.
+
+All this suggests that the wave train represented by EOF2 can be both forced by tropical SSTs and also a part of the internal variability of the extratropical atmosphere.
+When internally forced, there is little phase preference and amplitudes are relatively low.
+However, this same wave train, when excited by tropical SST anomalies will prefer the imaginary phase, with the sign of the geopotential anomalies depending on the sign of the tropical SSTs anomalies, and can reach higher magnitudes.
+
+```{r  sst-psi-1, fig.cap = "Same as figure befoe but for EOF1"}
+plot_sst_psi(which_pc = "PC1", which_lev = 200)
+```
+
+```{r}
+psi %>%
+  .[lev == 200] %>%
+  .[, .(psi.z = mean(psi.z)), by = .(lon, lat)] %>%
+  .[, c("fx", "fy") := WaveFlux(.SD)] %>%
+  .[lat <= 0] %>%
+  ggplot(aes(lon, lat)) +
+  geom_contour_fill(aes(z = psi.z)) +
+  geom_streamline(aes(dx = fx, dy = fy)) +
+  geom_qmap(~.x[lat<0], keep = .015) +
+  scale_fill_divergent()
+```
 
@@
 
+## Naive EOFs
+
+(ref:cap-eof-naive) Leading 4 EOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2009.
+
+(ref:subcap-eof-naive) Spatial patterns (arbitrary units). The numbers at the top-left of each panel is the variance explained by each EOF.
+
+(ref:subcap-cor-eof-naive) Coefficient of determination of the temporal index of each EOF between levels.
+
+```{r cor-eof-naive, fig.cap = "(ref:cap-eof-naive)", fig.subcap = c("(ref:subcap-eof-naive)", "(ref:subcap-cor-eof-naive)"), fig.ncol = 1}
+eof_naive[, eof[[1]]$right, by = lev] %>%
+  periodic(lon = c(0, 360)) %>%
+  ggplot(aes(lon, lat)) +
+
+  geom_contour_fill(aes(z = hgt, fill = ..level..),
+                    breaks = AnchorBreaks(0, 0.01, exclude = 0)) +
+  geom_contour_tanaka(aes(z = hgt), breaks = AnchorBreaks(0, 0.01, exclude = 0)) +
+
+  geom_qmap(~.x[lat <= -20]) +
+  scale_fill_divergent_discretised(guide = "none")  +
+  scale_y_latitude() +
+  scale_x_longitude() +
+  coord_polar() +
+  facet_grid(lev~EOF, labeller = labeller(lev = lab_lev))  +
+  axis_labs_smol +
+  tagger::tag_facets(tag = "panel", tag_pool = explained, tag_suffix = "")
+
+eof_naive[, eof[[1]]$left, by = lev] %>%
+  .[, item := paste(lev, EOF, sep = "_")] %>%
+  widyr::pairwise_cor(item, time, hgt) %>%
+  setDT() %>%
+  # .[]
+  .[tstrsplit(item1, "_")[[1]] != tstrsplit(item2, "_")[[1]], ] %>%
+  .[tstrsplit(item1, "_")[[1]] != 50] %>%
+  .[, item1 := tstrsplit(item1, "_")[[2]]] %>%
+  .[, item2 := tstrsplit(item2, "_")[[2]]] %>%
+  ggplot(aes(item1, item2)) +
+  geom_raster(aes(fill = correlation^2)) +
+  shadowtext::geom_shadowtext(aes(label = round(correlation^2, 2)),
+                              color = "black", bg.colour = "white", bg.r = 0.2) +
+  scale_fill_gradient(guide = "none", low = "white", high = scales::muted("red"), limits = c(0, 1)) +
+  scale_x_discrete("200 hPa") +
+  scale_y_discrete("50 hPa") +
+  plot_layout(ncol = 1)
+```
+
 ## Chosen rotations of the EOFs

### 2021-02-10 R/correlate.R

@@ -1,0 +1,26 @@
+#' Compute correlations
+#'
+#' Thin wrapper around [cor.test] which returns estimate, p.value, low and high
+#' 95% confidence interval and a text ready for reporting.
+#'
+#' @param x,y numeric variables to correlate.
+#' @param signif number of significant figures to keep in the reporting text.
+#'
+#' @export
+correlate <- function(x, y, signif = 2) {
+  correlation <- cor.test(x, y)
+  out <- with(correlation,
+              list(estimate = estimate,
+                   p.value = p.value,
+                   low = conf.int[1],
+                   hig = conf.int[2]))
+
+
+
+  out$text <- with(out,
+                   paste0(signif(estimate, signif), " (CI: ",
+                          signif(low, signif), " -- ", signif(hig, signif), ")"))
+
+  out
+
+}

### 2021-02-11 analysis/paper/paper.Rmd
Fija correlación con ENSO y signo de PC1

@@
 library(patchwork)
+library(kableExtra)
 library(shceof)
-library(kableExtra)
 
@@
   })
+}
+```
+
+```{r read-indices}
+enso <- rsoi::download_oni(TRUE, data_path("raw", "oni.csv")) %>%
+  as.data.table() %>%
+  .[, .(time = lubridate::as_datetime(Date), oni = ONI)] %>%
+  na.omit() %>%
+  .[season(time) == "SON"] %>%
+  .[, .(oni = mean(oni)), by = .(time = seasonally(time))]
+
+sams <- readRDS(data_path("derived", "sam.Rds")) %>%
+  .[season(time) == "SON"] %>%
+  .[, dim :=  lubridate::days_in_month(time)] %>%
+  .[, .(estimate = weighted.mean(estimate, dim)),
+    by = .(lev, term, time = seasonally(time))]
+
+levs <- unique(sams$lev)
+
+for (l in seq_along(levs)[-1]) {
+  cor <- sams[lev %in% levs[seq(l - 1, l)]] %>% 
+    .[term == "full"] %>%
+    dcast(time ~ lev, valuev.ar = "estimate") %>% 
+    setnames(., colnames(.), c("time", "old", "new")) %>% 
+    .[, cor(old, new)]
+
+  sams[lev == levs[l], estimate := sign(cor)*estimate]
 }
@@
 
-Las correlaciones entre PCs Podrían ir en una tabla. Quizás suplementaria.
-
 Relacionar 50hPa con vórtice polar? Temperatura también suma. 
@@
 
-ondas antes e la tempreatura
-
 Mostrar EOF espaciales.
-
-Tirar ERA20C a la basura. ERA20c suplementaria?
 
@@
   # add_header_above(c("", "50 hPa" = 4)) %>% 
-  columns_spec(2:5,  background = spec_color(.table[[.col]], option = background_palette, 
+  column_spec2(2:5,  background = spec_color2(.table[[.col]], option = background_palette, 
                                               scale_from = c(0, 1)),
@@
 The high correlation between levels for the two leading EOFs justifies treating the pattern as a mode of covariability between the stratosphere and the troposphere and, thus, computing the EOFs using both levels at the same time.") %>%
-  columns_spec(2:4, background = spec_color(.table[[.col]], option = background_palette,
+  column_spec2(2:4, background = spec_color2(.table[[.col]], option = background_palette,
                                             scale_from = c(0, 1)),
@@
 
-best_rotation_pc1 <- rotations_pc1[part == "Imaginary"][which.min(abs(mean))]$rotation
+best_rotation_pc1 <- rotations_pc1[part == "Real"][which.min(mean)]$rotation
 
@@
 ceof <- ceof[rotation == "rotated"][, rotation := NULL][]
+```
+
+```{r enso_cor}
+enso_cor <-  ceof[, eof[[1]]$left, by = .(season)] %>%
+  sep_ReIm() %>%
+  .[, hgt := scale(hgt), by = .(PC, part)] %>%
+  enso[., on = "time"] %>%
+  na.omit() %>%
+  .[, correlate(hgt, oni), by = .(PC, part)]
+
+# Defines the sign of the 2nd EOF to have a positive correlation 
+# with ENSO
+sign <- sign(enso_cor[PC == "PC2" & part == "Imaginary"]$estimate)
+
+if (sign != 1) {
+  ceof$eof[[1]] <- flip_ceof(ceof$eof[[1]], n = 2)
+}
+
+# Re-run the correlation in case it changed. 
+enso_cor <-  ceof[, eof[[1]]$left, by = .(season)] %>%
+  sep_ReIm() %>%
+  .[, hgt := scale(hgt), by = .(PC, part)] %>%
+  enso[., on = "time"] %>%
+  na.omit() %>%
+  .[, correlate(hgt, oni), by = .(PC, part)]
 ```
@@
 
-```{r mean-anom, fig.cap = "Mean SON zonal geopotential height anomalies at (a) 50 hPa and (b) 200 hPa"}
-# era5 %>%
-#   .[, mean(-hgt), by = .(lon, lat, lev)] %>%
-#   .[lev %in% c(50, 200)] %>%
-#   .[lat <= -20] %>%
-#   .[, V2 := V1 - mean(V1), by = .(lat, lev)] %>%
-#   periodic(lon = c(0, 360)) %>%
-#   ggplot(aes(lon, lat)) +
-#   geom_contour_fill(aes(z = V2), global.breaks = FALSE, breaks = AnchorBreaks(exclude = 0)) +
-#   geom_contour_tanaka(aes(z = V2), global.breaks = FALSE, breaks = AnchorBreaks(exclude = 0)) +
-#   geom_qmap(~ .x[lat <= -20]) +
-#   scale_fill_divergent(NULL, breaks = AnchorBreaks(exclude = 0),
-#                        guide = guide_colorstrip_bottom()) +
-#   scale_x_longitude() +
-#   scale_y_latitude() +
-#   facet_grid(~lev, labeller = labeller(lev = lab_lev)) +
-#   coord_polar() +
-#   axis_labs_smol +
-#   tag_facets()
-```
-
-(ref:extended-series-cap) Temporal series extended using ERA5 back extended preliminary edition (period 1950 -- 1978) and ERA5 (period 1979 -- 2019) and ERA20C (period 1900 -- 2010).
+(ref:extended-series-cap) Temporal series extended using ERA5 back extended preliminary edition (period 1950 -- 1978) and ERA5 (period 1979 -- 2019).
 Each series is computed by projecting monthly geopotential height zonal anomalies standardised by level south of 20ºS onto the corresponding spatial pattern.
-In numbers, the coefficient of determination between the ERA20C-based series and the two ERA5-based series.
-The ERA5 + ERABC and the ERA20C series are centred and scaled using the mean and standard deviation for their period of overlap (1950 -- 2010).
 
@@
 
-era20c <- ERA20C() %>%
-  ReadNetCDF(vars = c(hgt = "z"),
+era_more <- rbind( 
+  ReadNetCDF(ERA5_BE(), vars = c(hgt = "z"),
              subset = list(latitude = range(pattern$lat),
-                           level = list(200, 50))) %>%
-  normalise_coords() %>%
-  .[season(time) == "SON"] %>%
-  .[, .(hgt = mean(hgt)), by = .(lev, lon, lat, time = seasonally(time))] %>%
-  .[, season := season(time)] %>%
-  .[, hgt := hgt/sd(hgt), by = .(lev)] %>%
-  .[pattern, on = .NATURAL, allow.cartesian = TRUE] %>%
-  .[, .(hgt = weighted.mean(hgt*EOF, w = cos(lat*pi/180))),  by = .(part, PC, time)]
-
-era_more <- ERA5_BE() %>%
-  ReadNetCDF(vars = c(hgt = "z"),
-             subset = list(latitude = range(pattern$lat),
-                           level = list(200, 50))) %>%
-  rbind(.,
+                           level = list(200, 50))),
   ReadNetCDF(ERA5(), vars = c(hgt = "z"),
@@
   .[pattern, on = .NATURAL, allow.cartesian = TRUE] %>%
-  .[, .(hgt = weighted.mean(hgt*EOF, w = cos(lat*pi/180))),  by = .(part, PC, time)] %>%
-  era20c[., on = c("part", "PC", "time")] %>%
-  .[, hgt := (i.hgt - mean(i.hgt))/sd(i.hgt)*sd(hgt, na.rm = TRUE) + mean(hgt, na.rm = TRUE),
-    by = .(part, PC)] %>%
-  .[, i.hgt := NULL]
+  .[, .(hgt = weighted.mean(hgt*EOF, w = cos(lat*pi/180))),  by = .(part, PC, time)]
 
@@
 
-extended <- rbindlist(list(era20c = era20c,
-                           era5 = era_more),  use.names = TRUE, idcol = "dataset")
-
-trends <- extended %>%
-  .[time %between% common_range(time, dataset)] %>%
-  .[, FitLm(hgt, time, se = TRUE), by = .(dataset, part, PC)] %>%
+trends <- era_more %>%
+  .[, FitLm(hgt, time, se = TRUE), by = .(part, PC)] %>%
   rm_intercept() %>%
@@
 
-trends_all <- extended %>%
-  .[, FitLm(hgt, time, se = TRUE), by = .(dataset, part, PC)] %>%
-  rm_intercept() %>%
-  .[, p_val := Pvaluate(estimate, std.error, df)] %>%
-  .[, p_report := report_p(p_val)]
-
-periods <- range(extended[time %between% common_range(time, dataset)]$time)
-periods <- c(periods[1], lubridate::as_datetime("1979-01-01"), periods[2])
-
-correlations <- extended %>%
-  dcast(time + PC + part ~ dataset, value.var = "hgt") %>%
-  na.omit() %>% {
-    d <- .
-    rbind(
-      d[, .(correlation = cor(era20c, era5)),
-        by = .(PC, part, time = ifelse(year(time) >= 1979,
-                                       as.character(mean(periods[2:3])),
-                                       as.character(mean(periods[1:2]))))],
-
-      d[, .(correlation = cor(era20c, era5)),
-        by = .(PC, part)] %>%
-        .[, time := as.character(mean(periods[c(1, 3)]))]
-    )
-  } %>%
-
-  .[, time := lubridate::as_datetime(time)]
-
-extended %>%
+era_more %>%
   copy() %>% 
-  .[, ref := time %between% common_range(time, dataset)] %>%
-  .[, hgt := (hgt - mean(hgt[ref]))/sd(hgt[ref]), by = .(PC, part, dataset)] %>%
-  .[, dataset_o := dataset] %>%
-  # .[dataset == "era5", dataset := ifelse(year(time) < 1979, "era5be", "era5")] %>%
+  # .[, hgt := as.numeric(scale(hgt)), by = .(part, PC)] %>%
   ggplot(aes(time, hgt)) +
   # Zero line + rangeframe
-  geom_hline(yintercept = 0, size = 0.1, color = "gray30") +
+  geom_hline(data = ~.x[, mean(hgt), by  = .(PC, part)], 
+             aes(yintercept = V1), size = 0.1, color = "gray30") +
   ggthemes::geom_rangeframe(sides = "lr", size = 0.1, color = "gray30") +
-
-  geom_line(aes(color = dataset)) +
+  geom_line() +
   scale_alpha_manual(values = c(1, 0.2)) +
-  # Correlations
-  annotate("errorbarh", xmin = periods[1:2], xmax = periods[2:3], y = -2.7, size = 0.3) +
-  annotate("errorbarh", xmin = periods[1], xmax = periods[3], y = -4, size = 0.3) +
-
-  geom_label(data = correlations[time %in% range(time)],
-             aes(y = -2.7, label = paste0("r²: ", signif(correlation^2, 2))),
-             label.r = grid::unit(0, "lines"), label.size = 0) +
-  geom_label(data = correlations[!(time %in% range(time))],
-             aes(y = -4, label = paste0("r²: ", signif(correlation^2, 2))),
-             label.r = grid::unit(0, "lines"), label.size = 0) +
-
-  geom_smooth(data = ~.x[dataset == "era20c"],
-              method = "lm", se = FALSE,  aes(color = dataset)) +
-  geom_smooth(data = ~copy(.x)[time %between% common_range(time, dataset)],
-              aes(color = dataset), method = "lm", se = FALSE) +
-
-  scale_x_datetime(NULL, date_labels = "%Y",
-                   breaks = function(x) sort(unique(c(scales::pretty_breaks()(x),
-                                                      lubridate::as_datetime("1979-10-01"))))) +
-  scale_y_continuous("Standarised EOF") +
-  scale_color_brewer(NULL, palette = "Dark2", labels = c(era20c = "ERA20C",
-                                                         era5 = "ERA5 BE"),
-                     guide = guide_legend(override.aes = list(size = 1), order = 1)) +
-  facet_grid(part ~ PC, labeller = labeller(PC = lab_PC)) +
+
+  geom_smooth(method = "lm", color = "black") +
+
+  scale_x_datetime(NULL, date_labels = "%Y") +
+  scale_y_continuous("EOF") +
+
+  facet_grid(PC ~ part, labeller = labeller(PC = lab_PC), scales = "free_y") +
   theme(panel.spacing = grid::unit(1.5, "lines"))  +
@@
 
-```{r}
-
-```
-
-Figure \@ref(fig:extended-series) shows temporal series of the two complex EOFs extended beyond the satellite era using the preliminary ERA5 back extension going back to 1950 (which we call the "hybrid ERA5" reanalysis) and using ERA20C for the period 1900 -- 2010.
-The hybrid ERA5 series and the ERE20C series agree relatively well during the period of overlap, although correlations are higher in the satellite period.
-There is a downward trend in the real part of EOF1 in the hybrid reanalysis (Figure \@ref(fig:extended-series)a.1, p-value `r trends[part == "Real" & PC == "PC1" & dataset != "era20c"]$p_report` for the period 1950 -- 2010) that appears to be part of a long-term signal, judging by the statistically significant trend observed in the ERA20C reanalysis from 1900 to 2010 (p-value `r trends_all[part == "Real" & PC == "PC1" & dataset == "era20c"]$p_report`) with similar magnitude during the period of overlap.
-
-The ERA20C time-series has a also significant upward linear trend in the imaginary part of EOF1 (Figure \@ref(fig:extended-series)a.2, p-value `r trends[part == "Imaginary" & PC == "PC1" & dataset == "era20c"]$p_report`.
-However, there is no trend in the hybrid dataset during the period of overlap (p-value `r trends[part == "Imaginary" & PC == "PC1" & dataset != "era20c"]$p_report`) and the correlations between the old and modern series are much poorer for this variable.
-This suggests that the observed trend is not robust.
-
+Figure \@ref(fig:extended-series) shows temporal series of the two complex EOFs extended beyond the satellite era using the preliminary ERA5 back extension going back to 1950 (which we call the "hybrid ERA5" reanalysis).
+There is a downward trend in the real part of EOF1 in the hybrid reanalysis (Figure \@ref(fig:extended-series)a.1, p-value `r trends[part == "Real" & PC == "PC1"]$p_report`).
 There is no significant trend in any of the complex parts of EOF2.
-The correlations between the hybrid and ERA20C datasets are much higher than for the EOF1, which suggests that ERA20C could be used as a valid dataset to study this index in the past.
 
@@
 
-(ref:eof1-regr-t-cap) Same as Figure \@ref(fig:eof1-regr-gh) but for air temperature.
-
-```{r eof1-regr-t, fig.cap = "(ref:eof1-regr-t-cap)"}
-plot_regr(regr, "PC1", c(200, 50), "t")
-```
-
 (ref:eof2-regr-gh-cap) Same as Figure \@ref(fig:eof1-regr-gh) but for the second EOF.
@@
 plot_regr(regr, "PC2", c(200, 50), "hgt")
-```
-
-(ref:eof2-regr-t-cap) Same as Figure \@ref(fig:eof1-regr-gh) but for the second EOF and temperature.
-
-```{r eof2-regr-t, fig.cap = "(ref:eof2-regr-t-cap)"}
-plot_regr(regr, "PC2", c(200, 50), "t")
 ```
@@
 
+(ref:eof1-regr-t-cap) Same as Figure \@ref(fig:eof1-regr-gh) but for air temperature.
+
+```{r eof1-regr-t, fig.cap = "(ref:eof1-regr-t-cap)"}
+plot_regr(regr, "PC1", c(200, 50), "t")
+```
+
+(ref:eof2-regr-t-cap) Same as Figure \@ref(fig:eof1-regr-gh) but for the second EOF and temperature.
+
+```{r eof2-regr-t, fig.cap = "(ref:eof2-regr-t-cap)"}
+plot_regr(regr, "PC2", c(200, 50), "t")
+```
+
 ## Relationship with other indices
@@
 ### SAM
-
-```{r read-indices}
-sam <- "/home/elio/Documents/CONICET/asymsam/analysis/data/derived_data/indexes.Rds" %>%
-  readRDS() %>%
-  .[, .(time, sam = -estimate_norm, lev, symmetry = term)] %>%
-  .[, sam := sam/sd(sam), by = .(lev, symmetry)] %>%
-  .[season(time) == "SON"] %>%
-  .[, .(sam = mean(sam)), by = .(symmetry, lev, time = seasonally(time))]
-
-enso <- rsoi::download_oni(TRUE, data_path("raw", "oni.csv")) %>%
-  as.data.table() %>%
-  .[, .(time = lubridate::as_datetime(Date), oni = ONI)] %>%
-  na.omit() %>%
-  .[season(time) == "SON"] %>%
-  .[, .(oni = mean(oni)), by = .(time = seasonally(time))]
-
-plot_with_index <- function(ceof, index) {
-  index_col <- colnames(index)[2]
-
-  ceof[, eof[[1]]$left] %>%
-    sep_ReIm() %>%
-    .[, hgt := scale(hgt), by = .(PC, part)] %>%
-    index[., on = "time"] %>%
-    na.omit() %>%
-    .[, index := get(index_col)] %>%
-    ggplot(aes(hgt, index)) +
-    geom_point(aes(color = part)) +
-    geom_smooth(aes(color = part), method = "lm") +
-    shadowtext::geom_shadowtext(data = ~.x[, .(cor = cor(index, hgt),
-                                               x = -1.2,
-                                               y = sign(-cor(index, hgt))*.1 + predict(lm(index ~ hgt),
-                                                                                       newdata = data.frame(hgt = -1.2))),
-                                           by = .(PC, part)],
-                                bg.colour = "white", bg.r = 0.2,
-                                size = 5, show.legend = FALSE,
-                                aes(x, y, color = part, label = paste0("r² = ", round(cor^2, 2)))) +
-    scale_x_continuous("EOF") +
-    scale_y_continuous(toupper(index_col)) +
-    scale_color_brewer("Part", palette = "Dark2") +
-    facet_grid(part ~ PC, labeller = labeller(PC = c(PC1 = "EOF1", PC2 = "EOF2")))
-}
-```
 
@@
   sep_ReIm() %>%
-  .[, hgt := scale(hgt), by = .(PC, part)] %>%
-  .[]
-
-sams <- readRDS(data_path("derived", "sam.Rds")) %>%
-  .[season(time) == "SON"] %>%
-  .[, dim :=  lubridate::days_in_month(time)] %>%
-  .[, .(estimate = weighted.mean(estimate, dim)),
-    by = .(lev, term, time = seasonally(time))]
+  .[, hgt := scale(hgt), by = .(PC, part)]
 
@@
 
-max_r2 <- sam_r2[part == "Imaginary" & term == "asym"][, .SD[which.max(estimate^2)],
-                                                       by = .(PC)]
+max_r2 <- sam_r2[part == "Imaginary" & term == "asym"] %>% 
+  .[, .SD[which.max(estimate^2)], by = .(PC)]
 
@@
   ggplot(aes(lev, estimate^2)) +
-  # geom_ribbon(data =  ~.x[p.adjust(p_value, "fdr") < 0.001],
-  # aes(ymin = low2, ymax = high2, color = term), alpha= 0.1) +
   geom_point(data = ~.x[p.adjust(p.value, "fdr") < 0.01], aes(color = term)) +
@@
 
-```{r enso_cor}
-enso_cor <-  ceof[, eof[[1]]$left, by = .(season)] %>%
-  sep_ReIm() %>%
-  .[, hgt := scale(hgt), by = .(PC, part)] %>%
-  enso[., on = "time"] %>%
-  na.omit() %>%
-  .[, correlate(hgt, oni), by = .(PC, part)]
-```
-
 ```{r sst-psi-2,  fig.cap = "Regresión con PSI en 200 hPa y sst, PC1"}
@@
 Figure \@ref(fig:sst-psi-2) shows regression maps between EOF2 and Sea Surface Temperatures (SST) Streamfunction at 200 hPa.
-The Imaginary EOF2 is associated with strong negative SST anomalies on the Central Pacific and positive anomalies on an area across the North of Australia and New Zealand, the South Pacific Convergence Zone (SPCZ) (Figure \@ref(fig:sst-psi-2).a2).
-This pattern is almost canonically negative ENSO and indeed, the correlation between the Imaginary EOF2 and the Oceanic ENSO Index ([@bamston1997]) is `r enso_cor[PC == "PC2" & part == "Imaginary"]$cor_text`.
+The Imaginary EOF2 is associated with strong positive SST anomalies on the Central Pacific and negative anomalies on an area across the North of Australia and New Zealand, the South Pacific Convergence Zone (SPCZ) (Figure \@ref(fig:sst-psi-2).a2).
+This pattern is almost canonically positive ENSO and indeed, the correlation between the Imaginary EOF2 and the Oceanic ENSO Index ([@bamston1997]) is very high `r enso_cor[PC == "PC2" & part == "Imaginary"]$text`.
 Streamfunction anomalies show a coherent picture.
@@
 
-One could expect that the Real EOF2, being just a different phase of the same wave train would show a similar pattern but, perhaps with a slight translation of the main tropical sources.
-However, the Real EOF2 is associated neither with any significant SST (Figure \@ref(fig:sst-psi-2).a1) nor streamfunction anomalies in the tropics.
+Since the Real EOF2 represents just a different phase of the same wave train, one would expect that t would show a similar forcing pattern to the Imaginary EOF with a slight translation of its location.
+However, Figure \@ref(fig:sst-psi-2).a1 and b1 show that the Real EOF2 is not associated either with any significant SST nor streamfunction anomalies in the tropics.
 The correlation between the Real EOF2 and ENSO is also not significant (`r enso_cor[PC == "PC2" & part == "Real"]$cor_text`).
+This lack of tropical signal suggests a radically different nature of the different phases of the EOF2 wave train.
 
@@
 Black line is the fit ONI ~ cos(phase) + sin(phase) computed by OLS weighted by the magnitude of EOF2."}
-cor_enso_mag <- ceof[, eof[[1]]$left, by = .(season)] %>%
+gdata <- ceof[, eof[[1]]$left, by = .(season)] %>%
   .[, arg := Arg(hgt)] %>%
@@
   na.omit() %>%
-  .[PC == "PC2"] %>%
-  .[, correlate(mag, abs(oni))]
-
-ceof[, eof[[1]]$left, by = .(season)] %>%
-  .[, arg := Arg(hgt)] %>%
-  .[, mag := Mod(hgt)] %>%
-  .[enso, on = "time"] %>%
-  na.omit() %>%
-  .[PC == "PC2"] %>%
+  .[PC == "PC2"]
+
+cor_enso_mag <- gdata[, correlate(mag, abs(oni))]
+
+cos_model <- gdata %>% 
+  qwrap(arg = c(-pi, pi) ~ c(-2*pi, 2*pi)) %>% 
+  .[, lm(oni ~ I(sin(arg)) - 1, weights = mag )]
+
+gdata %>% 
   ggplot(aes(arg, oni)) +
@@
               method = "lm", aes(weight = mag),
-              formula = y ~  I(cos(x)) + I(sin(x)),
+              formula = y ~ I(sin(x)) -1,
               se = FALSE, color = "black", size = 0.5) +
@@
 
-We show a more complete picture of the relationship between ENSO and EOF2 in Figure \@ref(fig:enso-phase), which plots the ONI index value and the phase of the EOF2 for each year between 1979 and 2019 highlighting years in which the magnitude of EOF2 was higher than 50% of the years.
-In years with positive ENSO, the phase of the EOF2 is always around -90º (corresponding with negative imaginary part) and vice versa.
-In years with near neutral ENSO, the phase of the EOF2 is much more variable between 0º and -180º.
+To better explore the relationship between tropical forcing and phase of the EOF2, Figure \@ref(fig:enso-phase) plots the ONI index and the phase of the EOF2 for each year between 1979 and 2019, highlighting years in which the magnitude of EOF2 was higher than 50% of the years.
+In years with positive ENSO, the phase of the EOF2 is always around +90º (corresponding with positive imaginary part) and vice versa.
+In years with near neutral ENSO, the phase of the EOF2 is much more variable.
+The black line in Fiure \@ref(fig:enso-phase) represents the equation $ONI = `r coef(cos_model)[1]`\sin{phase}$, with coefficient fitted by ordinary least squares and $r^2=`r summary(cos_model)$r.squared`$.
+
 There is also a relationship between the magnitude of the EOF2 and the magnitude of the ENSO.
-The strong EOF2 years are almost invariably associated with strong ENSO years.
-The correlation between the absolute magnitude of the ONI and the magnitude of the EOF2 is `r cor_enso_mag$cor_text`.
+The strong EOF2 years tend to coincide with strong ENSO years.
+The correlation between the absolute magnitude of the ONI and the magnitude of the EOF2 is `r cor_enso_mag$text`.
 
 All this suggests that the wave train represented by EOF2 can be both forced by tropical SSTs and also a part of the internal variability of the extratropical atmosphere.
-When internally forced, there is little phase preference and amplitudes are relatively low.
-However, this same wave train, when excited by tropical SST anomalies will prefer the imaginary phase, with the sign of the geopotential anomalies depending on the sign of the tropical SSTs anomalies, and can reach higher magnitudes.
+When internally forced, the wave train has little phase preference and its amplitude is relatively low.
+When excited by tropical SST anomalies, it tends to remain locked to the imaginary phase, with the sign of the geopotential anomalies depending on the sign of the tropical SSTs anomalies, and with larger amplitude.
 
@@
 plot_sst_psi(which_pc = "PC1", which_lev = 200)
-```
-
-```{r}
-psi %>%
-  .[lev == 200] %>%
-  .[, .(psi.z = mean(psi.z)), by = .(lon, lat)] %>%
-  .[, c("fx", "fy") := WaveFlux(.SD)] %>%
-  .[lat <= 0] %>%
-  ggplot(aes(lon, lat)) +
-  geom_contour_fill(aes(z = psi.z)) +
-  geom_streamline(aes(dx = fx, dy = fy)) +
-  geom_qmap(~.x[lat<0], keep = .015) +
-  scale_fill_divergent()
 ```
@@
     geom_contour2(aes(z = p_val), breaks = 0.05, size = 0.1) +
-    stat_subset(aes(subset = p_val <= 0.05),
+    stat_subset(aes(subset = p_val <= 0.05 & is.cross(lon, lat)),
                 geom = "point", size = 0.2) +
@@
 ```{r pp-america, fig.cap = "Regresión de cada PC con la precipitación"}
-plot_pp(data = regr_pp, which_pc= c("PC1", "PC2"), which_continent =  "america")
+plot_pp(data = regr_pp, which_pc = c("PC1", "PC2"), which_continent = "america")
 ```

### 2021-02-11 R/eof.R

@@ -1,0 +1,24 @@
+#' Flip sign of eofs
+#'
+#' @param  eof eof object returned from [metR::EOF()]
+#' @param n vector of eofs to flip. If `NULL`, defaults to all.
+#'
+#'
+#' @export
+flip_ceof <- function(eof, n = NULL) {
+  eof <- copy(eof)
+
+  pc_col <- attr(eof, "suffix", TRUE)
+  val_col <- attr(eof, "value.var", TRUE)
+
+  if (is.null(n)) {
+    n <- eof$sdev[[pc_col]]
+  } else {
+    n <- paste0(pc_col, n)
+  }
+
+  eof$left[get(pc_col) %in% n, c(val_col) := -get(val_col)]
+  eof$right[get(pc_col) %in% n, c(val_col) := -get(val_col)]
+
+  return(eof)
+}

### 2021-02-17 analysis/paper/paper.Rmd
PSA y SAM filtrado

@@
   comment = "#>",
-  cache.extra = 43,
+  cache.extra = 42,
   cache.path = "../cache/paper/",
@@
   .[, solve_poisson(vo, lon, lat), by = .(lev, time)] %>%
-  .[, psi.z := Anomaly(value), by = .(time, lev, lat)] %>%
-  .[, value := NULL]
+  setnames("value", "psi") %>% 
+  .[, psi.z := Anomaly(psi), by = .(time, lev, lat)] 
+```
+
+```{r psa}
+psa <- ReadNetCDF(ERA5(), vars = c(hgt = "z"),
+                   subset = list(level = list(500),
+                                 latitude = -90:0)) %>% 
+  normalise_coords() %>% 
+  .[, .(hgt = mean(hgt)), by = .(lon, lat, time = seasonally(time))] %>% 
+  .[, value := Anomaly(hgt)*sqrt(cos(lat*pi/180)), by = .(lon, lat, season(time))] %>%
+  EOF(value ~ time | lon + lat, n = 1:3, data = .)
+
+psa <- lapply(psa, function(x) {
+  copy(x)[, PC := fcase(PC == "PC1", "SAM", 
+                        PC == "PC2", "PSA1", 
+                        PC == "PC3", "PSA2")] 
+})
+
+psa_time <- psa$left[season(time) == "SON"]
 ```
@@
 
-Whereas for traditional "real" PCA, each principal component is defined up to it's sign. In complex PCA, principal components are defined up to their phase [@horel1984].
-
-ERA5, ERA5 BE, ERA 20C
+The phase of principal components is defined up to an additive constant.
+For real principal components, this constant can be either 0 or $\pi$, corresponding to a change in sign.
+For complex principal components, it can be any number between 0 and $2\pi$ [@horel1984].
+In this paper we chose the phase of each of the two complex principal components so that the real and imaginary part are easier to interpret.
+For the first complex principal component, the phase was chosen so that the time series corresponding to the imaginary part had a time mean of zero.
+
+For the second complex principal component, the phase was chosen so that the coefficent of determination between the Oceanic Niño Index [@bamston1997] and the real part was minimised.
+It's important to reassure the reader that this procedure does not create a spurious correlation, it only takes whatever relationship that already exist and aligns it with a specific phase.
+If the relationship is very weak, the rotation cannot generate a strong relationship.
+
+ERA5, ERA5 BE,
 
@@
 
-Relacionar 50hPa con vórtice polar? Temperatura también suma. 
-En 50 el negativo es más intenso que el negativo. ¿Por qué?
+Relacionar 50hPa con vórtice polar? Temperatura también suma. En 50 el negativo es más intenso que el negativo. ¿Por qué?
 
@@
 
-En 200 no se ve la señal de jet de la temperatura ¿qué pasa más abajo? 
-en 700 la anomalía de temp coincide más o menos con el geopotentical
+En 200 no se ve la señal de jet de la temperatura ¿qué pasa más abajo? en 700 la anomalía de temp coincide más o menos con el geopotentical
 
@@
 
-```{r}
+```{r corr-eofs}
 
@@
 
-box <- list(lon = c(150, 195),
-            lat = c(-25, -5))
-
-psi_box <- psi %>%
-  .[lev == 200] %>%
-  .[lat %between% box$lat & lon %between% box$lon] %>%
-  .[, .(psi.z = weighted.mean(psi.z, cos(lat*pi/180))), by = .(time, lev)]
-
-psi_box <- cut(ceof$eof[[1]], 2)$left %>%
-  copy() %>%
-  .[psi_box, on = "time"]
-
-angles <- seq(-pi, pi, by = 1*pi/180)
+angles <- seq(-pi, pi, by = .5*pi/180)
 
@@
 
+with_enso <-  cut(ceof$eof[[1]], 2)$left %>%
+  copy() %>%
+  .[enso, on = "time"] %>% 
+  na.omit()
+
 rotations_pc2 <- lapply(angles, function(a) {
-  psi_box %>%
+  with_enso %>%
     .[, hgt2 := rotate(hgt, a)] %>%
-    .[, .(R = cor(Re(hgt2), psi.z),
-          I = cor(Im(hgt2), psi.z))] %>%
+    .[, .(R = cor(Re(hgt2), oni),
+          I = cor(Im(hgt2), oni))] %>%
     .[, rotation := a]
@@
 
-best_rotation_pc2 <- rotations_pc2[which.min(abs(R^2))]$rotation
+best_rotation_pc2 <- rotations_pc2[I > 0][which.min(abs(R))]$rotation
 
@@
 ceof <- ceof[rotation == "rotated"][, rotation := NULL][]
-```
-
-```{r enso_cor}
-enso_cor <-  ceof[, eof[[1]]$left, by = .(season)] %>%
-  sep_ReIm() %>%
-  .[, hgt := scale(hgt), by = .(PC, part)] %>%
-  enso[., on = "time"] %>%
-  na.omit() %>%
-  .[, correlate(hgt, oni), by = .(PC, part)]
-
-# Defines the sign of the 2nd EOF to have a positive correlation 
-# with ENSO
-sign <- sign(enso_cor[PC == "PC2" & part == "Imaginary"]$estimate)
-
-if (sign != 1) {
-  ceof$eof[[1]] <- flip_ceof(ceof$eof[[1]], n = 2)
-}
-
-# Re-run the correlation in case it changed. 
-enso_cor <-  ceof[, eof[[1]]$left, by = .(season)] %>%
-  sep_ReIm() %>%
-  .[, hgt := scale(hgt), by = .(PC, part)] %>%
-  enso[., on = "time"] %>%
-  na.omit() %>%
-  .[, correlate(hgt, oni), by = .(PC, part)]
 ```
@@
 
+### PSA1 and PSA2
+
+```{r}
+pc2 <- ceof$eof[[1]]$left[PC == "PC2"] %>% 
+  copy() %>% 
+  .[, PC := NULL] %>% 
+  .[]
+
+psa2 <- psa$left[PC == "PSA2"][season(time) == "SON"]
+
+rotations_psa <- lapply(angles, function(a) {
+  pc2 %>% 
+    copy() %>% 
+    .[, hgt := rotate(hgt, a)] %>% 
+    .[psa2, on = "time"] %>% 
+    .[, cor(value, Re(hgt))] 
+})
+
+psa_angle <- angles[which.min(unlist(rotations_psa))]
+
+pc2[psa_time, on = "time"] %>% 
+  .[PC != "SAM"] %>% 
+  copy() %>% 
+  .[, hgt := rotate(hgt, psa_angle)] %>%
+  sep_ReIm() %>% 
+  .[] %>% 
+  ggplot(aes(value, hgt)) +
+  geom_point() +
+  facet_grid(part ~ PC)
+```
+
 ### SAM
@@
 The Imaginary EOF2 is related with the SAM through the A-SAM in the troposphere, with up to `r scales::percent(max_r2[PC == "PC2"]$estimate, 2)` of shared variance, reached at `r max_r2[PC == "PC2"]$lev` hPa (Figure \@ref(fig:sam-eof-vertical).b2).
-The Imaginary EOF2 is, then, an almost a pure representation of the tropospheric asymmetric part of the SAM.
-Since the Real EOF2 represents a quarter-wavelength rotation of the same zonal wave 3 structure, the Imaginary part is the phase that coincides with the SAM.
-As a corollary, this means that the asymmetric component of the SAM represents only one phase of a zonally varying zonal wave.
+
+```{r eof-filtered, fig.cap = "Spatial pattern of the leading EOF of 200 hPa geopotential height with the variability of EOF2 filtered out South of 20º. Arbitrary scale."}
+eof_filtered <- era5 %>% 
+  .[sep_ReIm(pc2, format = "wide")[, hgt := NULL], on = .NATURAL] %>% 
+  .[lev == 200] %>% 
+  .[lat <= -20] %>% 
+  .[season(time) == "SON"] %>% 
+  .[, hgt := resid(.lm.fit(cbind(1, R, I), hgt)), .(lon, lat, lev, season(time))] %>%
+  .[, eof := Anomaly(hgt)*sqrt(cos(lat*pi/180)), by = .(lon, lat, lev, season(time))] %>% 
+  EOF(eof ~ lon + lat | time, n = 1, data = .)
+
+eof_filtered_cor <- eof_filtered$right[sams, on = "time"] %>% 
+  .[lev == 200] %>% 
+  .[, correlate(estimate, eof), by = term]
+
+eof_filtered %>% 
+  .$left %>%
+  periodic(lon = c(0, 360)) %>% 
+  ggplot(aes(lon, lat)) +
+  geom_contour_fill(aes(z = eof), breaks = AnchorBreaks(0, exclude = 0)) +
+  geom_contour_tanaka(aes(z = eof), breaks = AnchorBreaks(0, exclude = 0)) +
+  scale_fill_divergent(guide = "none") +
+  geom_qmap(~.x[lat <= -20]) +  
+  scale_x_longitude() +
+  scale_y_latitude() +
+  coord_polar() +
+  axis_labs_smol 
+```
+
+As a further illustration, Figure \@ref(fig:eof-filtered) shows the spatial pattern of the leading EOF of 200 hPa geopotential height when the variability associated with EOF2 is removed.
+The resulting pattern is an nearly zonally symmetrical annular mode.
+This mode is highly correlated with SAM (`r eof_filtered_cor[term == "full"]$text`), but only with the symmetrical part (`r eof_filtered_cor[term == "sym"]$text`).
+Note that the usual definition of the PSA modes as the second and third EOFs creates modes orthogonal to the SAM (defined as the first EOF) and thus impedes this kind of filtering.
+
+This suggests that the separation between the asymmetric and symmetric part of the SAM is not only statistically possible, but physically plausible.
+The asymmetric part of the SAM might be statistical contamination from one phase of the PSA mode.
 
@@
   rm_intercept() %>%
-  .[, psi.z := estimate] %>%
+  .[, psi.z := Anomaly(estimate), by = .(PC, season, lev, lat)] %>%
   .[, c("fx", "fy") := shceof::WaveFlux(.SD, p = 200), by = .(lev, PC, season, term)] %>%
@@
                      ordered = TRUE)]
+```
+
+```{r enso_cor}
+enso_cor <-  ceof[, eof[[1]]$left, by = .(season)] %>%
+  sep_ReIm() %>%
+  .[, hgt := scale(hgt), by = .(PC, part)] %>%
+  enso[., on = "time"] %>%
+  na.omit() %>%
+  .[, correlate(hgt, oni), by = .(PC, part)]
 ```
@@
 
-```{r sst-psi-2,  fig.cap = "Regresión con PSI en 200 hPa y sst, PC1"}
+(ref:sst-psi-2-cap) Regression maps of EOF2 with SST (column a) and streamfunction zonal anomalies with their corresponding activity wave flux (column b).
+Areas marked with dots have p-values smaller than 0.05 adjusted for FDR.
+
+```{r sst-psi-2,  fig.cap = "(ref:sst-psi-2-cap)"}
 plot_sst_psi(which_pc = "PC2", which_lev = 200)
@@
 
-???
--\> esta intro hay que cambiarla.
 Figure \@ref(fig:sst-psi-2) shows regression maps between EOF2 and Sea Surface Temperatures (SST) Streamfunction at 200 hPa.
 The Imaginary EOF2 is associated with strong positive SST anomalies on the Central Pacific and negative anomalies on an area across the North of Australia and New Zealand, the South Pacific Convergence Zone (SPCZ) (Figure \@ref(fig:sst-psi-2).a2).
-This pattern is almost canonically positive ENSO and indeed, the correlation between the Imaginary EOF2 and the Oceanic ENSO Index ([@bamston1997]) is very high `r enso_cor[PC == "PC2" & part == "Imaginary"]$text`.
+This pattern is almost canonically positive ENSO and indeed, the correlation between the Imaginary EOF2 and the Oceanic Niño Index ([@bamston1997]) is very high `r enso_cor[PC == "PC2" & part == "Imaginary"]$text`.
 Streamfunction anomalies show a coherent picture.
 The Imaginary EOF2 is associated with strong wave-like streamfunction anomalies emanating from the tropics (Figure \@ref(fig:sst-psi-2).b2).
-.
-This is consistent with what we know of the effect of ENSO on the extratropics: SST anomalies initiate anomalous convection that excites meridionally-moving Rossby waves.
-
-Since the Real EOF2 represents just a different phase of the same wave train, one would expect that t would show a similar forcing pattern to the Imaginary EOF with a slight translation of its location.
+This is consistent with what we know of the effect of ENSO on the extratropics: SST anomalies initiate anomalous convection that excites Rossby waves that propagate meridionally towards higher latitudes.
+
+?? Compare with Figura 1 b and c (Streamfunction) and Figure 2 d and FIgure 3 d (SST) from [@mo2001] and Figure 7 [@irving2016].
+
+Imaginary more global than real -> is that why it appears mixed with SAM?
+
+Since the Real EOF2 represents just a different phase of the same wave train, one would expect that it would show a similar forcing pattern to the Imaginary EOF with a slight translation of its location.
 However, Figure \@ref(fig:sst-psi-2).a1 and b1 show that the Real EOF2 is not associated either with any significant SST nor streamfunction anomalies in the tropics.
@@
 gdata <- ceof[, eof[[1]]$left, by = .(season)] %>%
+  .[PC == "PC2"] %>% 
+  # .[, hgt := Anomaly(hgt)] %>% 
   .[, arg := Arg(hgt)] %>%
@@
   .[enso, on = "time"] %>%
-  na.omit() %>%
-  .[PC == "PC2"]
+  na.omit()
 
@@
 
+cor_enso_spearman <- gdata[, cor.test(mag, abs(oni), method = "spearman")]
+
+cor_enso_mag_outlier <- gdata[oni < 1.5][, correlate(mag, abs(oni))]
+
 cos_model <- gdata %>% 
-  qwrap(arg = c(-pi, pi) ~ c(-2*pi, 2*pi)) %>% 
   .[, lm(oni ~ I(sin(arg)) - 1, weights = mag )]
@@
 In years with near neutral ENSO, the phase of the EOF2 is much more variable.
-The black line in Fiure \@ref(fig:enso-phase) represents the equation $ONI = `r coef(cos_model)[1]`\sin{phase}$, with coefficient fitted by ordinary least squares and $r^2=`r summary(cos_model)$r.squared`$.
-
-There is also a relationship between the magnitude of the EOF2 and the magnitude of the ENSO.
-The strong EOF2 years tend to coincide with strong ENSO years.
+The black line in Figure \@ref(fig:enso-phase) tries to quantify this relationship.
+Is the equation $ONI = `r signif(coef(cos_model)[1], 2)`\sin{phase}$, whose coefficient we fitted by weighted least squares using the magnitude of the EFO2 as weight.
+The $r^2$ corresponding to the fit is `r signif(summary(cos_model)$r.squared, 2)`, with p-value `r report_p(coef(summary(cos_model))[4])`.
+
+Figure \@ref(fig:enso-phase) suggest that strong EOF2 years tend to coincide with strong ENSO years.
 The correlation between the absolute magnitude of the ONI and the magnitude of the EOF2 is `r cor_enso_mag$text`.
+This relationship appears to be driven only by the three years with strongest ENSO events in the period (`r knitr::combine_words(year(gdata[order(-oni)][1:3]$time))`) which also coindice with the three years with strongest EOF2 magnitude.
+If those years are removed, the correlation becomes non-significant (`r cor_enso_mag_outlier$text`).
+Furthermore, the Spearman correlation, which is robust to outliers is also non-significant (`r signif(cor_enso_spearman$estimate, 2)`, p-value = `r report_p(cor_enso_spearman$p.value)`).
+Therefore, the relationship between the magnitude of the EOF2 train wave and ENSO remains uncertain.
 
 All this suggests that the wave train represented by EOF2 can be both forced by tropical SSTs and also a part of the internal variability of the extratropical atmosphere.
-When internally forced, the wave train has little phase preference and its amplitude is relatively low.
-When excited by tropical SST anomalies, it tends to remain locked to the imaginary phase, with the sign of the geopotential anomalies depending on the sign of the tropical SSTs anomalies, and with larger amplitude.
-
-```{r  sst-psi-1, fig.cap = "Same as figure befoe but for EOF1"}
+When internally forced, the wave train has little phase preference.
+When excited by tropical SST anomalies, it tends to remain locked to the imaginary phase, with the sign of the geopotential anomalies depending on the sign of the tropical SSTs anomalies.
+
+(ref:sst-psi-1-cap) Same as Figure \@ref(fig:sst-psi-2) but for EOF1.
+
+```{r sst-psi-1, fig.cap = "(ref:sst-psi-1-cap)"}
 plot_sst_psi(which_pc = "PC1", which_lev = 200)
 ```
+
+Figure \@ref(fig:sst-psi-1) shows SST and streamfunction regression maps for EOF1.
+There is no significant pattern of SST anomalies associated with either the Real or Imaginary EOF1.
+Consistently, streamfunction anomalies don't show any tropical influence.
+On the contrary, the Real EOF1 is associated with wave activity fluxes that flow equatorward from the coast of Antarctica around 150ºE along with wave activity fluxes that move between the positive and negative of Streamfunction anomalies along Antarctica.
 
@@
     # geom_raster(aes(fill = estimate)) +
-    geom_contour_fill(aes(z = estimate, fill = ..level..), breaks = AnchorBreaks(exclude = 0)) +
+    geom_contour_fill(aes(z = estimate, fill = ..level..), 
+                      breaks = AnchorBreaks(exclude = 0)) +
     geom_contour_tanaka(aes(z = estimate), breaks = AnchorBreaks(exclude = 0)) +
@@
     # scale_fill_divergent() +
-    scale_fill_divergent_discretised(NULL, guide = guide_colorsteps_bottom(15)) +
+    scale_fill_divergent_discretised(NULL, 
+                         # limits = c(-.8, .8),
+                         # oob = scales::squish,
+                         low = "#8E521C",
+                         high = "#00665E",
+                         guide = guide_colorsteps_bottom(15)) +
+    # scale_fill_divergent_discretised(NULL, guide = guide_colorsteps_bottom(15)) +
     geom_contour2(aes(z = p_val), breaks = 0.05, size = 0.1) +
@@
     axis_labs_smol +
+    tag_facets("cr") +
     coord_quickmap()
@@
 
-```{r pp-america, fig.cap = "Regresión de cada PC con la precipitación"}
+```{r sesa}
+sesa <- list(lon = c(300, 315),
+             lat = c(-35, -22))
+```
+
+(ref:pp-america-cap) Regression of SON mean precipitation anomalies in South America (mm per day, shaded) and (column a) EOF1 the (row 1) Imaginary and (column 1) Real phase.
+For the 1979 -- 2018 period.
+Black contours with dots indicate areas with p-value smaller than 0.05 controlling for False Detection Rate.
+
+```{r pp-america, fig.cap = "(ref:pp-america-cap)"}
 plot_pp(data = regr_pp, which_pc = c("PC1", "PC2"), which_continent = "america") 
-```
-
-```{r pp-africa, fig.cap = "Regresión de cada PC con la precipitación"}
+  # annotate("rect", xmin = min(sesa$lon), xmax = max(sesa$lon), 
+  #          ymin = min(sesa$lat), ymax = max(sesa$lat),
+  #          fill = NA, color = "gray50", size = 0.3)
+```
+
+Figure \@ref(fig:pp-america) shows regression maps of seasonal precipitation with each EOF in South America.
+EOF1 (Figure \@ref(fig:pp-america) column a) is associated with increased precipitation over Paraguay and and some parts of Brazil.
+Although these positive anomalies are not very strong.
+The strongest precipitation anomalies are the ones associated with the Imaginary EOF2.
+The positive anomalies on Southeaster South America (SESA) and Chile, and negative anomalies over Southern Brazil is a well known springtime precipitation signature of ENSO [@cai2020a] and it's also virtually identical to the precipitation anomalies associated with the Asymmetric SAM [@campitelli2021].
+This is not surprising considering the close relationship between the ONI, the Asymmetric SAM index and the Imaginary EOF2 shown previously.
+The Real EOF2, on the other hand, is associated with negative precipitation anomalies in a smaller area of SESA.
+The relationship between precipitation anomalies in SESA and the phase of EOF2 follows a curve similar to that of Figure \@ref(fig:enso-phase) (not shown).
+
+(ref:pp-africa-cap) Same as Figure \@ref(fig:pp-america) but for South of Africa.
+
+```{r pp-africa, fig.cap = "(ref:pp-africa-cap)"}
 plot_pp(regr_pp, c("PC1", "PC2"), which_continent = "africa")
@@
 
-```{r pp-oceania, fig.cap = "Regresión de cada PC con la precipitación"}
+(ref:pp-oceania-cap) Same as Figure \@ref(fig:pp-america) but for New Zealand and neighbouring islands.
+
+```{r pp-oceania, fig.cap = "(ref:pp-oceania-cap)"}
 plot_pp(regr_pp, c("PC1", "PC2"), which_continent =  "oceania")
@@
 ```{r rotations-pc1, fig.cap = "Rotations of PC1"}
+
 rotations_pc1 %>%
   ggplot(aes(rotation, mean)) +
+  geom_vline(xintercept = best_rotation_pc1) +
   geom_hline(yintercept = 0, color = "gray") +
@@
 
-```{r regr-rotation, fig.cap = "Regression with streamfunction PC2"}
-regr_psi %>%
-  .[PC == "PC2"] %>%
-  .[lev == 200] %>%
-  .[lat %between% c(-90, 10)] %>%
-  periodic(lon = c(0, 360)) %>%
-  ggplot(aes(lon, lat)) +
-  geom_contour_fill(aes(z = estimate*1e8, fill = ..level..),
-                    breaks = AnchorBreaks(exclude = 0)) +
-  geom_contour_tanaka(aes(z = estimate*1e8),
-                      breaks = AnchorBreaks(exclude = 0)) +
-
-  geom_qmap(keep = .015) +
-  scale_fill_divergent_discretised("Streamfunction", guide = "none") +
-
-  scale_x_longitude() +
-  scale_y_latitude() +
-  scale_mag(guide = "none") +
-  facet_grid(term ~ ., labeller = labeller(PC = c(PC1 = "EOF1", PC2 = "EOF2"))) +
-  coord_quickmap(ylim = c(NA, 10)) +
-  theme(legend.box = "horizontal") +
-  axis_labs_smol +
-  annotate("rect",
-           xmin = box$lon[1], xmax = box$lon[2],
-           ymin = box$lat[1], ymax = box$lat[2], color = "black", fill = NA)
-```
-
-```{r rotations-pc2, fig.cap = "R2 between EOF2 Real and Imaginary parts and zonal anomalies of streamfunction at BOX?? for different rotation parameters of the real and imaginary parts."}
+```{r rotations-pc2, fig.cap = "R2 between EOF2 Real and Imaginary parts and ONI index for different rotation parameters of the real and imaginary parts."}
 rotations_pc2 %>%
@@
   geom_hline(yintercept = 0, color = "gray") +
+  geom_vline(xintercept = best_rotation_pc2) +
   geom_line(aes(color = variable), size = 1) +

### 2021-02-19 analysis/paper/paper.Rmd
O3

@@
   comment = "#>",
-  cache.extra = 42,
+  cache.extra = 40,
   cache.path = "../cache/paper/",
@@
 
-pattern_crosshatch <- ggplot2::ggproto("GeomCrosshatch", ggpattern::GeomPolygonPattern)
-
-ggplot2::update_geom_defaults(pattern_crosshatch, 
-                              list(pattern = "crosshatch",
+pattern_dots <- ggplot2::ggproto("GeomDots", ggpattern::GeomPolygonPattern)
+ggplot2::update_geom_defaults(pattern_dots, 
+                              list(pattern = "circle",
+                                   colour = NA,
                                    pattern_colour = "black",
                                    pattern_fill = "black",
-                                   pattern_density = 0.01,
-                                   pattern_alpha = 0.2,
+                                   pattern_density = 0.1,
+                                   pattern_alpha = 1,
+                                   fill = NA,
                                    pattern_spacing = 0.02))
+
+geom_contour_pval <- function(mapping, p.value = 0.01, size = 0.1) {
+  list(
+    stat_contour2(mapping, breaks = p.value, geom = pattern_dots),
+    geom_contour2(mapping, breaks = p.value, size = size)
+  )
+}
 
@@
     .[term == "full"] %>%
-    dcast(time ~ lev, valuev.ar = "estimate") %>% 
+    dcast(time ~ lev, value.var = "estimate") %>% 
     setnames(., colnames(.), c("time", "old", "new")) %>% 
@@
   .[, psi.z := Anomaly(psi), by = .(time, lev, lat)] 
+```
+
+```{r o3}
+o3 <- O3() %>% 
+  readRDS() %>% 
+  .[lat <= -20] %>% 
+  .[, .(toc = mean(total_ozone_column, na.rm = TRUE)), by = .(lon, lat, time = seasonally(time))] 
+o3[, time := year(time)]
 ```
@@
 
-rotations_pc1 <- lapply(angles, function(a) {
-  ceof$eof[[1]]$left[PC == "PC1"] %>%
+# rotations_pc1 <- lapply(angles, function(a) {
+#   ceof$eof[[1]]$left[PC == "PC1"] %>%
+#     copy() %>%
+#     .[, hgt := rotate(hgt, a)] %>%
+#     sep_ReIm() %>%
+#     .[, .(mean = mean(hgt)), by = part] %>%
+#     .[, rotation := a]
+# }) %>%
+#   rbindlist()
+# 
+# 
+# best_rotation_pc1 <- rotations_pc1[part == "Real"][which.min(mean)]$rotation
+
+wave1_o3 <- o3 %>% 
+  .[lat %between% c(-70, -45)] %>% 
+  .[, .(toc = weighted.mean(toc, w = cos(lat*pi/180), na.rm = TRUE)), by = .(time, lon)] %>% 
+  .[, FitWave(toc, 1), by = .(time)]
+
+with_o3 <- ceof$eof[[1]]$left %>% 
+  .[PC == "PC1"] %>% 
+  .[, time := year(time)] %>% 
+  .[wave1_o3, on = "time"]
+
+rotations_o3 <- lapply(angles, function(a) {
+  with_o3 %>% 
     copy() %>% 
     .[, hgt := rotate(hgt, a)] %>% 
+    na.omit() %>% 
     sep_ReIm() %>% 
-    .[, .(mean = mean(hgt)), by = part] %>%
+    .[, .(cor = cor(amplitude, hgt)), by = part] %>% 
     .[, rotation := a]
@@
 
-best_rotation_pc1 <- rotations_pc1[part == "Real"][which.min(mean)]$rotation
+best_rotation_pc1 <- rotations_o3[part == "Real"][which.max(cor)]$rotation
 
@@
 (ref:extended-series-cap) Temporal series extended using ERA5 back extended preliminary edition (period 1950 -- 1978) and ERA5 (period 1979 -- 2019).
-Each series is computed by projecting monthly geopotential height zonal anomalies standardised by level south of 20ºS onto the corresponding spatial pattern.
+Each series is computed by projecting monthly geopotential height zonal anomalies standardised by level south of 20ºS onto the corresponding spatial pattern and scaled to zero mean and unit standard deviation.
 
@@
 
+# era_less <-   ceof %>%
+#   .[, eof[[1]]$left, by = season] %>%
+#   sep_ReIm(hgt) %>%
+#   .[season == "SON"] %>% 
+#   .[, hgt := as.numeric(scale(hgt)), by = .(part, PC)]
+
 era_more %>%
   copy() %>% 
-  # .[, hgt := as.numeric(scale(hgt)), by = .(part, PC)] %>%
+  .[, ref := year(time) > 1979] %>% 
+  .[, hgt := scale(hgt), by = .(part, PC)] %>%
   ggplot(aes(time, hgt)) +
@@
   geom_line() +
+  # geom_line(data = era_less) +
   scale_alpha_manual(values = c(1, 0.2)) +
@@
   .[, hgt := NULL] %>%
-  .[, `:=`(R = as.numeric(scale(R)), I = as.numeric(scale(I))),
+  .[, `:=`(Real = as.numeric(scale(Real)), Imaginary = as.numeric(scale(Imaginary))),
     by = .(season, PC)] %>%
-  dcast(time ~ PC, value.var = c("R", "I")) %>%
+  dcast(time ~ PC, value.var = c("Real", "Imaginary")) %>%
   .[melt(era5, id.vars = c("time", "lon", "lat", "lev")), on = "time", allow.cartesian = TRUE] %>%
@@
   # .[variable == "hgt"] %>%
-  .[, FitLm(value, I_PC1, R_PC1, I_PC2, R_PC2, se = TRUE),
+  .[, FitLm(value, Imaginary_PC1, Real_PC1, Imaginary_PC2, Real_PC2, se = TRUE),
     by = .(lev, lon, lat, variable)] %>%
@@
   rm_intercept() %>%
-  .[, term := factor(term, levels = c("R", "I"), labels = c("Real", "Imaginary"))]
+  .[, term := factor_ReIm(term)]
 ```
@@
     # .[term == "I" & lev == 50] %>%
-    .[, p_val := Pvaluate(estimate, std.error, df, "fdr"), by = .(lev, PC, term)] %>%
+    .[, p.value := Pvaluate(estimate, std.error, df, "fdr"), by = .(lev, PC, term)] %>%
     periodic(lon = c(0, 360)) %>%
@@
     ) +
-
-    geom_contour2(aes(z = p_val), breaks = 0.05, size = 0.1) +
-    stat_subset(aes(subset = p_val <= 0.05 & is.cross(lon, lat, 1)),
-                geom = "point", size = 0.2, alpha = 0.1) +
+    geom_contour_pval(aes(z = p.value)) +
+
     geom_qmap(~ .x[lat <= -20]) +
@@
 
+```{r t}
+t <- ERA5() %>% 
+  ReadNetCDF(vars = "t", 
+             subset = list(latitude = c(-90, -20))) %>% 
+  normalise_coords()
+```
+
+```{r t-vertical, fig.cap = "Regression of EOF1 with mean air temperature between 80ºS and 25ºS."}
+# t_PC1_vertical <- t %>% 
+#   .[cut(ceof$eof[[1]], 1)$left, on = "time"] %>% 
+#   sep_ReIm(format = "wider") %>% 
+#   .[, FitLm(t, Real, Imaginary, se = TRUE), by = .(lev, lon, lat)] %>% 
+#   rm_intercept()
+# t_PC1_vertical %>% 
+#   .[lev %in% c(1, 5, 10, 50)] %>% 
+#   .[, p.value := Pvaluate(estimate, std.error, df, "fdr"), by = .(term, lev)] %>% 
+#     ggplot(aes(lon, lat)) +
+#   geom_contour_fill(aes(z = estimate)) +
+#   stat_subset(aes(subset = p.value <= 0.01 & is.cross(lon, lat)),
+#               size = .5, alpha = 0.5) +
+#   geom_hline(yintercept = c(-80, -35)) +
+#   geom_qmap(~.x[lat <= -20]) +
+#   scale_fill_divergent() + 
+#   coord_polar() +
+#   facet_grid(lev ~ term)
+
+t %>% 
+  .[lat %between% c(-80, -35)] %>% 
+  .[, .(t = weighted.mean(t, cos(lat*pi/180))), by = .(lev, time, lon)] %>%  
+  .[cut(ceof$eof[[1]], 1)$left, on = "time"] %>% 
+  sep_ReIm(format = "wider") %>% 
+  .[, FitLm(t, Real, Imaginary, se = TRUE), by = .(lev, lon)] %>% 
+  rm_intercept() %>% 
+  .[, term := factor_ReIm(term)] %>% 
+  .[, p.value := Pvaluate(estimate, std.error, df, "fdr"), by = .(term)] %>%
+  ggplot(aes(lon, lev)) +
+  geom_contour_fill(aes(z = estimate, fill = ..level..), breaks = AnchorBreaks(0, exclude = 0)) +
+  geom_contour_tanaka(aes(z = estimate), 
+                      breaks = AnchorBreaks(0, exclude = 0)) +
+
+  geom_contour_pval(aes(z = p.value)) +
+  scale_fill_divergent_discretised("Air temperature", guide = guide_colorsteps_bottom()) + 
+  scale_y_level() +
+  scale_x_longitude() +
+  facet_grid(~term) +
+  axis_labs_smol
+```
+
+```{r o3_regr} 
+temp <- ceof$eof[[1]]$left %>% 
+  .[PC == "PC1"] %>% 
+    copy() %>% 
+  sep_ReIm(format = "wider") %>% 
+  .[, `:=`(Real = as.numeric(scale(Real)), 
+           Imaginary = as.numeric(scale(Imaginary))),
+    by = .(PC)] %>%
+  .[, time := year(time)]
+
+o3_regr <- temp %>% 
+  o3[., on = "time", allow.cartesian = TRUE] %>% 
+  .[, FitLm(toc, Real, Imaginary, se = TRUE), by = .(lon, lat, PC)] %>% 
+  rm_intercept() %>% 
+  .[, term := factor_ReIm(term)]
+```
+
+```{r o3-regr, fig.cap = "Regression of standarised EOF1 and Total Ozone column."}
+o3_regr %>% 
+  .[, p.value := Pvaluate(estimate, std.error, df, "fdr"), by = .(PC, term)] %>% 
+  ggplot(aes(lon, lat)) +
+  geom_contour_fill(aes(z = estimate, fill = ..level..),
+                    breaks = AnchorBreaks(0, exclude = 0)) +
+  geom_contour_tanaka(aes(z = estimate), breaks = AnchorBreaks(0, exclude = 0)) +
+  geom_contour_pval(aes(z = p.value)) +
+  geom_qmap(~.x[lat <= -20]) +
+  scale_fill_divergent_discretised("Total Ozone Column", 
+                                   guide = guide_colorsteps_bottom()) + 
+  scale_x_longitude() +
+  scale_y_latitude() +
+  coord_polar() +
+  facet_grid(PC~term) +
+  axis_labs_smol
+```
+
+The ozone hole, as measured by Total Ozone Column (TOC) is climatologically not centered over the South Pole, but shifted towards the Weddell Sea, which translates to a large amplitude zonal wave 1 with minimum over the Weddell Sea and maximum over the Davis Sea [@wirth1993].
+The regression
+
+```{r wave1-o3, fig.cap = "EOF1 and wave1 o3 amplitude."}
+wave1_o3 <- o3 %>% 
+  .[lat %between% c(-70, -45)] %>% 
+  .[, .(toc = mean(toc, na.rm = TRUE)), by = .(time, lon)] %>% 
+  .[, FitWave(toc, 1), by = .(time)]
+
+temp <- ceof$eof[[1]]$left %>% 
+  .[PC == "PC1"] %>% 
+  .[, time := year(time)] %>% 
+  .[wave1_o3, on = "time"]
+
+temp %>% 
+  copy() %>% 
+  na.omit() %>% 
+  sep_ReIm() %>% 
+  .[, hgt := scale(hgt), by  = .(part)] %>% 
+  ggplot(aes(amplitude, hgt)) +
+  geom_point() +
+  geom_smooth(method = "lm") +
+  facet_grid(PC ~ part)
+```
+
 ```{r eof2-regr-t, fig.cap = "(ref:eof2-regr-t-cap)"}
@@
 
-```{r}
+```{r psa-eof2, fig.cap = "EFO2 and PSA"}
 pc2 <- ceof$eof[[1]]$left[PC == "PC2"] %>% 
@@
   sep_ReIm() %>% 
-  .[] %>% 
   ggplot(aes(value, hgt)) +
   geom_point() +
+  scale_y_continuous("EOF2") +
+  scale_x_continuous("PSA") +
   facet_grid(part ~ PC)
@@
   .[season(time) == "SON"] %>% 
-  .[, hgt := resid(.lm.fit(cbind(1, R, I), hgt)), .(lon, lat, lev, season(time))] %>%
+  .[, hgt := resid(.lm.fit(cbind(1, Real, Imaginary), hgt)), .(lon, lat, lev, season(time))] %>%
   .[, eof := Anomaly(hgt)*sqrt(cos(lat*pi/180)), by = .(lon, lat, lev, season(time))] %>% 
@@
   .[, hgt := NULL] %>%
-  .[, `:=`(R = R/sd(R), I = I/sd(I)), by = .(season, PC)] %>%
+  .[, `:=`(Real = Real/sd(Real), Imaginary = Imaginary/sd(Imaginary)), by = .(season, PC)] %>%
   .[psi, on = "time", allow.cartesian = TRUE] %>%
   .[season(time) == "SON"] %>%
-  .[, FitLm(psi.z, R, I, se = TRUE),by = .(lon, lat, lev, PC, season)] %>%
+  .[, FitLm(psi.z, Real, Imaginary, se = TRUE),by = .(lon, lat, lev, PC, season)] %>%
   rm_intercept() %>%
@@
   .[, c("fx", "fy") := shceof::WaveFlux(.SD, p = 200), by = .(lev, PC, season, term)] %>%
-  .[, term := factor(term, levels = c("R", "I"), labels = c("Real", "Imaginary"),
-                     ordered = TRUE)]
+  .[, term := factor_ReIm(term)]
 ```
@@
   .[, hgt := NULL] %>%
-  .[, `:=`(R = R/sd(R), I = I/sd(I)), by = .(season, PC)] %>%
+  .[, `:=`(Real = Real/sd(Real), Imaginary = Imaginary/sd(Imaginary)), by = .(season, PC)] %>%
   .[sst, on = "time", allow.cartesian = TRUE] %>%
   .[season(time) == "SON"] %>%
-  .[, FitLm(t, I, R, se = TRUE), by = .(lon, lat, PC, season)] %>%
+  .[, FitLm(t, Imaginary, Real, se = TRUE), by = .(lon, lat, PC, season)] %>%
   rm_intercept() %>%
-  .[, term := factor(term, levels = c("R", "I"), labels = c("Real", "Imaginary"),
-                     ordered = TRUE)]
+  .[, term := factor_ReIm(term)]
 ```
@@
     .[lat %between% c(-90, 10)] %>%
-    .[, p_val := Pvaluate(estimate, std.error, df, "fdr"), by = .(PC, term)] %>%
+    .[, p.value := Pvaluate(estimate, std.error, df, "fdr"), by = .(PC, term)] %>%
     .[, var := "Streamfunction"] %>%
@@
                       breaks = AnchorBreaks(exclude = 0)) +
-    geom_contour2(aes(z = p_val), breaks = 0.05, size = 0.1) +
-    stat_subset(aes(subset = p_val <= 0.05 & is.cross(lon, lat, 1)),
-                geom = "point", size = 0.2, alpha = 0.5) +
+    geom_contour_pval(aes(z = p.value)) +
     geom_streamline(aes(dx = fx, dy = fy), min.L = 3, L = 10, n = 30,
@@
   .[, hgt := NULL] %>%
-  .[, `:=`(R = R/sd(R), I = I/sd(I)), by = .(season, PC)] %>%
+  .[, `:=`(Real = Real/sd(Real), Imaginary = Imaginary/sd(Imaginary)), by = .(season, PC)] %>%
   .[melt(cmap, id.vars = c("lon", "lat", "time")), on = "time", allow.cartesian = TRUE] %>%
   .[season == "SON"] %>%
-  .[, FitLm(value, I, R, se = TRUE), by = .(lon, lat, PC, variable, season)] %>%
-  rm_intercept()
+  .[, FitLm(value, Imaginary, Real, se = TRUE), by = .(lon, lat, PC, variable, season)] %>%
+  rm_intercept() %>% 
+  .[, term := factor_ReIm(term)]
 ```
@@
 ```{r plot_pp}
-plot_pp <- function(data, which_pc, which_term = c("R", "I"), which_continent = NULL) {
+plot_pp <- function(data, which_pc, which_term = c("Real", "Imaginary"), which_continent = NULL) {
   filter_continent <- function(data) {
     if (!is.null(which_continent)) {
-      data[ continent == which_continent]
+      data[continent == which_continent]
     } else {
@@
     # .[abs(estimate) <= 2] %>%
-    .[, p_val := Pvaluate(estimate, std.error, df, "none"), by = .(lev, PC, term)] %>%
+    .[, p.value := Pvaluate(estimate, std.error, df, "none"), by = .(lev, PC, term)] %>%
     periodic(lon = c(0, 360)) %>%
@@
     # scale_fill_divergent_discretised(NULL, guide = guide_colorsteps_bottom(15)) +
-    geom_contour2(aes(z = p_val), breaks = 0.05, size = 0.1) +
-    stat_subset(aes(subset = p_val <= 0.05 & is.cross(lon, lat)),
-                geom = "point", size = 0.2) +
+    geom_contour_pval(aes(z = p.value)) +
     geom_qmap(~.x %>% .[lat %between% c(-60, max(data$lat))] %>% add_continent() %>% filter_continent(),
@@
     scale_y_latitude(ticks = 15) +
-    ggh4x::facet_nested(term ~ PC, labeller = labeller(term = lab_cplx)) +
+    ggh4x::facet_nested(term ~ PC) +
     axis_labs_smol +
@@
 
-rotations_pc1 %>%
-  ggplot(aes(rotation, mean)) +
+rotations_o3 %>%
+  ggplot(aes(rotation, cor)) +
   geom_vline(xintercept = best_rotation_pc1) +
@@
                      labels = function(x) paste0(round(x*180/pi, 5), "º")) +
-  scale_y_continuous("Mean value") +
+  scale_y_continuous("Correlation with O3 zonal wave 1") +
   scale_color_brewer(NULL, palette = "Dark2", labels = lab_cplx)

### 2021-02-19 R/0-download.R

@@ -24,6 +24,39 @@ download_cds <- function(request) {
 }
 
 
+download_ozone <- function(request) {
+  force(request)
+  function(file) {
+    user <- Sys.getenv("CDSUSER", NA)
+    key <- Sys.getenv("CDSKEY", NA)
+    filename <- basename(file)
+
+    if (is.na(user) | is.na(key)) {
+      stop("Please set the environmental variables CDSUSER and CDSKEY")
+    }
+
+
+    ecmwfr::wf_set_key(user = user,
+                       key = key,
+                       service = "cds")
+
+    file_zip <- ecmwfr::wf_request(request = request,
+                                user = user,
+                                time_out = 5*3600,
+                                path = tempdir())
+
+
+    files <- unzip(file_zip, exdir = tempdir())
+
+    data <- do.call(rbind, lapply(files, metR::ReadNetCDF))
+    data <- normalise_coords(data)
+    data[, lon := metR::ConvertLongitude(lon)]
+
+    saveRDS(data, file)
+    file
+  }
+}
+
 download_webapi <- function(request) {
   force(request)
   function(file) {
@@ -56,7 +89,7 @@ request_era5be <- list(
   variable = c("geopotential", "temperature", "vorticity"),
   pressure_level = c("1", "2", "3", "5", "7", "10", "20", "30", "50", "70", "100", "125", "150", "175", "200", "225", "250", "300", "350", "400", "450", "500", "550", "600", "650", "700", "750", "775", "800", "825", "850", "875", "900", "925", "950", "975", "1000"),
   year = as.character(1950:1978),
-  month = c("01", "02", "03", "04", "05", "06", "07", "08", "09", "10", "11", "12"),
+  month = c("09", "10", "11"),
   time = "00:00",
   grid = c("2.5", "2.5"),
   dataset_short_name = "reanalysis-era5-pressure-levels-monthly-means-preliminary-back-extension"
@@ -66,14 +99,14 @@ request_era5be <- list(
 request_era5 <- list(
   format = "netcdf",
   product_type = "monthly_averaged_reanalysis",
-  variable = c("geopotential", "temperature", "vorticity", "divergence"),
+  variable = c("geopotential", "temperature", "vorticity", "divergence", "total_column_ozone"),
   pressure_level = c("1", "2", "3", "5", "7", "10", "20", "30", "50",
[…]

### 2021-02-19 R/ReIm.R

@@ -14,7 +14,7 @@ ReIm <- function(complex) {
 #' @rdname ReIm
 sep_ReIm <- function(data, column, format = c("longer", "wider")) {
   R <- part <- I <- NULL
-  names <- c("R", "I")
+  names <- c("Real", "Imaginary")
 
 
   if (missing(column)) {
@@ -38,10 +38,16 @@ sep_ReIm <- function(data, column, format = c("longer", "wider")) {
 
   if (format[1] == "longer") {
     data[, c(col) := NULL]
-    data <- data.table::setDT(tidyr::pivot_longer(data, R:I, names_to = "part", values_to = col))
-    data[, part := factor(part, levels = c("R", "I"), labels = c("Real", "Imaginary"),
-                          ordered = TRUE)]
+    data <- data.table::setDT(tidyr::pivot_longer(data, Real:Imaginary, names_to = "part", values_to = col))
+    data[, part := factor(part, levels = names, ordered = TRUE)]
   }
 
   return(data[])
+}
+
+#' @param part Character vector with elements "Real" and "Imaginary"
+#' @export
+#' @rdname ReIm
+factor_ReIm <- function(part) {
+  factor(part, levels = c("Real", "Imaginary"), ordered = TRUE)
 }

### 2021-02-19 R/data_locations.R

@@ -39,6 +39,23 @@ data_nc_fun <- function(file, download) {
 }
 
 
+data_rds_fun <- function(file, download) {
+   force(file)
+
+   if (is.character(download)) {
+      download <- simple_download(download)
+   }
+
+   function(force_download = FALSE) {
+      if (force_download || !file.exists(file)) {
+         file <- download(file)
+      }
+      return(file)
+   }
+}
+
+
+
 #' @param force_download Logical indicating wether to force download of data.
 #' @export
 #' @rdname data_locations
@@ -85,3 +102,8 @@ HADSST <- data_nc_fun(file = data_path("raw", "hadsst.mon.mean.nc"),
 CMAP <- data_nc_fun(file = data_path("raw", "cmap.mon.mean.nc"),
                     download = "ftp://ftp.cdc.noaa.gov/Datasets/cmap/std/precip.mon.mean.nc" )
 
+#' @param force_download Logical indicating wether to force download of data.
+#' @export
+#' @rdname data_locations
+O3 <- data_rds_fun(file = data_path("raw", "o3.mon.mean.Rds"),
+                    download = download_ozone(request_o3))

### 2021-02-19 analysis/scripts/00-download_era.R

@@ -1,47 +1,0 @@
-library(lubridate)
-library(ecmwfr)
-
-# Todo esto no sirve! Baja ERA interim y hay que bajar ERA5
-
-levs <- c(10, 20, 30, 50, 70, 100, 150, 200, 250, 300, 400, 500, 600, 700, 850,
-          925, 1000)
-
-dates <- seq.Date(as.Date("1979-01-01"), as.Date("2018-12-31"), "1 month")
-dates <- paste0(year(dates),
-                formatC(month(dates), width = 2, flag = "0"),
-                formatC(day(dates), width = 2, flag = "0"),
-                collapse = "/")
-
-hgt_request <- list(
-  class    = "ei",
-  dataset  = "interim",
-  date     = dates,
-  expver   = "1",
-  grid     = "2.5/2.5",
-  levelist = levs,
-  levtype  = "pl",
-  param    = "129.128",
-  stream   = "moda",
-  type     = "an",
-  format   = "netcdf",
-  target   = "hgt.mon.mean.nc"
-)
-
-u_request <- list(
-  class    = "ei",
-  dataset  = "interim",
-  date     = dates,
-  expver   = "1",
-  grid     = "2.5/2.5",
-  levelist = levs,
-  levtype  = "pl",
-  param    = "131.128",
-  stream   = "moda",
-  type     = "an",
-  format   = "netcdf",
-  target   = "uwnd.mon.mean.nc"
-)
-
-path <- here::here("datos", "ERA-Interim")
-wf_request(hgt_request, job_name = "hgt_era", path = path)
-wf_request(u_request, job_name = "u_era", path = path)

### 2021-02-19 analysis/scripts/00-download_era_daily.R

@@ -1,18 +1,0 @@
-request <- list(
-  class    = "ei",
-  dataset  = "interim",
-  date     = "1979-01-01/to/2018-12-31",
-  expver   = "1",
-  grid     = "2.5/2.5",
-  levelist = "500",
-  levtype  = "pl",
-  param    = "131.128/132.128",
-  step     = "0",
-  stream   = "oper",
-  time     = "00:00:00/06:00:00/12:00:00/18:00:00",
-  type     = "an",
-  target   = "u_v_daily.nc",
-  format   = "netcdf"
-)
-
-ecmwfr::wf_request(request, job_name = "era_daily")

### 2021-02-19 analysis/scripts/01-get-data.R

@@ -1,4 +1,3 @@
-library(magrittr)
 library(shceof)
 
 force <- FALSE
@@ -22,3 +21,6 @@ ERA5_SST(force)
 
 # Download ERA20C -----------------------------------------------------------
 ERA20C(force)
+
+# Download O3 --------------------------------
+O3(force)

### 2021-02-19 analysis/paper/paper.Rmd
Readme and data download

@@
   list(
-    stat_contour2(mapping, breaks = p.value, geom = pattern_dots),
+    stat_contour_fill(mapping, breaks = c(0, p.value), fill = NA, geom = pattern_dots),
     geom_contour2(mapping, breaks = p.value, size = size)
@@
 
-For the second complex principal component, the phase was chosen so that the coefficent of determination between the Oceanic Niño Index [@bamston1997] and the real part was minimised.
+For the second complex principal component, the phase was chosen so that the coefficient of determination between the Oceanic Niño Index [@bamston1997] and the real part was minimised.
 It's important to reassure the reader that this procedure does not create a spurious correlation, it only takes whatever relationship that already exist and aligns it with a specific phase.
 If the relationship is very weak, the rotation cannot generate a strong relationship.
+
+We calculated the PSA1 and PSA2 modes as the third and fourth leading EOF of seasonal mean 500 hPa geopotential height with all seasons together following @mo2001.
 
@@
   copy() %>% 
-  .[, hgt := rotate(hgt, psa_angle)] %>%
+  # .[, hgt := rotate(hgt, psa_angle)] %>%
   sep_ReIm() %>% 
@@
   ggplot(aes(lev, estimate^2)) +
+  geom_vline(xintercept = c(50, 200),  size = 0.2, color = "gray50") +
   geom_point(data = ~.x[p.adjust(p.value, "fdr") < 0.01], aes(color = term)) +
@@
   geom_vline(xintercept = best_rotation_pc1) +
-  geom_hline(yintercept = 0, color = "gray") +
+  geom_hline(yintercept = 0,  size = 0.2, color = "gray50") +
   geom_line(aes(color = part), size = 1) +
@@
   ggplot(aes(rotation, value^2)) +
-  geom_hline(yintercept = 0, color = "gray") +
+  geom_hline(yintercept = 0,  size = 0.2, color = "gray50") +
   geom_vline(xintercept = best_rotation_pc2) +

### 2021-02-19 R/0-download.R

@@ -106,7 +106,7 @@ request_era5 <- list(
                      "600", "650", "700", "750", "775", "800", "825",
                      "850", "875", "900", "925", "950", "975", "1000"),
   year = as.character(1979:2019),
-  month =  c("09", "10", "11"),
+  month = formatC(1:12, width = 2, flag = 0),   # Need all months to compute PSA
   time = "00:00",
   grid = c("2.5", "2.5"),
   dataset_short_name = "reanalysis-era5-pressure-levels-monthly-means"

### 2021-02-19 R/correlate.R

@@ -15,12 +15,9 @@ correlate <- function(x, y, signif = 2) {
                    low = conf.int[1],
                    hig = conf.int[2]))
 
-
-
   out$text <- with(out,
                    paste0(signif(estimate, signif), " (CI: ",
                           signif(low, signif), " -- ", signif(hig, signif), ")"))
 
   out
-
 }

### 2021-02-19 R/lm_lite.R

@@ -28,7 +28,8 @@ lm_lite <- function(y, x, weights = rep(1, length(y)), r2 = FALSE) {
 #' @keywords internal
 var.wt <- function(x, w, na.rm = FALSE) {
   if (na.rm) {
-    w <- w[i <- !is.na(x)]
+    i <- !is.na(x)
+    w <- w[i]
     x <- x[i]
   }
   sum.w <- sum(w)

### 2021-02-26 analysis/paper/paper.Rmd
Más PSA SAM y texto

@@
     reference_doc: ../templates/template.pptx
-bibliography: references.bib
+bibliography: [references.bib, data.bib]
 csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
@@
   comment = "#>",
-  cache.extra = 40,
+  cache.extra = 41,
   cache.path = "../cache/paper/",
@@
 
+panel_background <- theme(panel.background = element_rect(fill = "#fbfbfb", color = NA), 
+                          panel.ontop = FALSE,
+                          tagger.panel.tag.background = ggplot2::element_rect(color = NA, 
+                                                                              fill = "#fbfbfb"))
+
 guide_colorstrip_bottom <- function(width = 25, height = 0.5, ...) {
@@
 
-guide_colorsteps_bottom <- function(width = 25, height = 0.5, ...) {
-  guide_colorsteps(show.limits = TRUE, even.steps = FALSE,
+guide_colorsteps_bottom <- function(width = 25, height = 0.5, even.steps = FALSE, ...) {
+  guide_colorsteps(show.limits = TRUE, even.steps = even.steps,
                    title.position = "top", title.hjust = 0.5,
@@
 geom_contour_pval <- function(mapping, p.value = 0.01, size = 0.1) {
+  mapping2 <- mapping
+  mapping2$fill <- aes(fill = NA)$fill
   list(
-    stat_contour_fill(mapping, breaks = c(0, p.value), fill = NA, geom = pattern_dots),
+    stat_contour_filled(mapping2, breaks = c(0, p.value), fill = NA, geom = pattern_dots),
     geom_contour2(mapping, breaks = p.value, size = size)
@@
   .[lat <= -20] %>% 
-  .[, .(toc = mean(total_ozone_column, na.rm = TRUE)), by = .(lon, lat, time = seasonally(time))] 
+  .[, .(toc = mean(total_ozone_column, na.rm = TRUE)), by = .(lon, lat, time = seasonally(time))] %>% 
+  unique(by = c("lon", "lat", "time"))
 o3[, time := year(time)]
@@
 
+@era5 and @era5be
+
 The phase of principal components is defined up to an additive constant.
@@
 
-# Results
-
 No mostrar los EOFs no complejos. No tiene sentido mostralos si no se analizas.
@@
 Pp, ¿cuál es la regresión con SAM asim en 200 hPa?
+
+## Motivation to use complex EOF
 
@@
 
+Both observations motivate the decision of performing complex EOF jointly between levels.
+The computation of the EOFs was carried out using data from both levels at the same time, therefore, each complex EOF has a spatial part that depends on longitude, latitude and level, an a temporal part that depends only on time.
+
+# Results
+
+## Description of EOFs
+
 ```{r rotate_eofs}
@@
   theme(panel.spacing = grid::unit(1.5, "lines"))  +
-  ggthemes::geom_rangeframe(sides = "lr", size = 0.1, color = "gray30")
-```
-
-Both observations motivate the decision of performing complex EOF jointly between levels.
-The computation of the EOFs was carried out using data from both levels at the same time, therefore, each complex EOF has a spatial part that depends on longitude, latitude and level, an a temporal part that depends only on time.
+  ggthemes::geom_rangeframe(sides = "lr", size = 0.1, color = "gray30") +
+  panel_background
+```
+
 Figures \@ref(fig:ceofs-1) and \@ref(fig:ceofs-2) show, respectively, the spatial and temporal parts of the first two leading complex EOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa.
 In the spatial patterns in Figure \@ref(fig:ceofs-1), the real (in shading) and the imaginary parts (in contour) are in quadrature by construction, so that each EOF describe a single wave-like pattern whose amplitude and position (i.e.
-phase) is controlled by the magnitude and argument of the complex temporal EOF.
+phase) is controlled by the magnitude and phase of the complex temporal EOF.
 
 The wave patterns described by these complex EOFs match the patterns seen in the navie EOFs of Figure \@ref(fig: cor-eof-naive-1): The first is a wave 1, while the second is a wave 3.
-Note that in Figure \@ref(fig:ceofs-2), the real part of EOF1 has non zero mean.
+Note that in Figure \@ref(fig:ceofs-2), both parts of the EOF1 have non zero mean.
 This is due to th fact that the geopotential fields that enter into the algorithm are anomalies with respect to the zonal mean (shown in Figure \@ref(fig:mean-anom)), not the time mean.
-The real part of the first EOF is, therefore, capturing the mean zonally anomalous field and the variability that projects onto it.
+The variability associated with the first EOF is variability that projects onto the mean zonally anomalous field.
 
@@
   theme(panel.spacing = grid::unit(1.5, "lines"))  +
-  tag_facets("cr")
+  tag_facets("cr") +
+  panel_background
 ```
@@
 
-## Regression
+## Regressions
+
+The spatial patterns shown in Figure \@ref(fig:ceofs-1) are "pure" EOF patterns.
+But as they are derived by removing the zonally symmetric circulation, they might not include all the variability that is actually associated with the EOF time series.
+They are also idealised in that they are forced by construction, to be Hilber transforms of each other.
+To get a more realistic view of the real-world variability described by these EOFS, we computed regression patterns of each EOF with the complete geopotential fields.
 
@@
 ```{r eof1-regr-gh, fig.cap = "(ref:eof1-regr-gh-cap)"}
-plot_regr(regr, "PC1", c(200, 50), "hgt")
-```
-
-Figure \@ref(fig:eof1-regr-gh) shows regression patterns of SON geopotential height on the real and imaginary part of the first EOF at 50 hPa and 200 hPa.
+plot_regr(regr, "PC1", c(200, 50), "hgt") +
+  # geom_contour2(data = ~copy(.x)[, estimate := Anomaly(estimate), by = .(term, lev, lat)],
+  #              aes(z = estimate, linetype = factor(-sign(..level..))), 
+  #              global.breaks = FALSE, breaks = AnchorBreaks(0, bins = 2, exclude = 0)) +
+  # scale_linetype(guide = "none") +
+  NULL
+```
+
+Figure \@ref(fig:eof1-regr-gh) shows regression patterns of EOF1 and geopotential height.
+At 50 hPa (Figure \@ref(fig:eof1-regr-gh) row a), both the Real and Imaginary EOF1 are associated with planetary wave 1 patterns, that are 90º off phase.
+Their phases coincide with the ones shown in Figure \@ref(fig:ceofs-1)a.1, with the positive centre of the Real EOF1 located towards the dateline, and the one of the Imaginary EOF1 located over Eastern Antarctica.
+However, the Real EOF1 pattern is substantially altered by the zonally symmetric circulation.
+Instead of a clear wave 1 pattern, the regression pattern can be better describes as a monopole with its centre displaced from the South Pole.
+Similarly affected are the regression patterns at 200 hPa (Figure \@ref(fig:eof1-regr-gh) row b).
+It is only possible to distinguish some of the wave 1 pattern in relation with the Real EOF1 (Figure @\ref(fig:eof1-regr-gh)b.1).
+The Imaginary EOF1 shows a much more zonally symmetrical pattern resembling a negative Southern Annular Mode.
+
+With the exception of the Imaginary EFO1, it's clear that these patterns are very different than the idealised versions shown in Figure \@ref(fig:ceofs-1), particularly at 200 hPa.
+Moreover, only in the stratosphere these patterns actually show a distinguishable wave 1 pattern shifted in phase by 90º.
+This should not be utterly surprising, since the spatial patterns of the *naive* EOFs only show this feature at 50 hPa, suggesting that using the complex EOF method is "forcing" a wave structure where there is none.
+
+Therefore, the magnitude and phase of the EOF1 are associated with the magnitude and phase of a zonal wave only in the stratosphere.
+While in the troposphere they are associated with slightly off-centre monopoles.
 
@@
 
+Figure \@ref(fig: eof2-regr-gh) shows the regression pattern of geopotential height and the EOF2.
+Unlike in the case of EOF1, in this case the regression patterns represent relatively well the *idealised* patterns from Figure \@ref(fig:ceofs-1).
+Although there are some differences (particularly in 50 hPa), the wave trains identified before are well characterised and patterns associated with the Real EOF2 are 90º out of phase with those associated with the Imaginary EOF2.
+Zonal wave 3 dominates all fields, but only in the western hemisphere.
+
+EOF2 then represents a equivalent barotropic wave train that is very similar to the the Pacific South American Patterns [@mo2001].
+Comparing the location of the positive anomaly near 90ºW in column b of Figure \@ref(fig: eof2-regr-gh) with Figure 1 a and b from @mo2001, the Real EOF2 can roughly be identified with the PSA2, while the Imaginary EOF2 resembles PSA1.
+This separation is not completely meaningful in this case, since by treating this pattern as a single wave train any one pair of orthogonal waves can characterise this variability.
+
 ```{r zonal-waves, fig.cap = "Amplitud de ondas zonales de la regresión con altura geopotencial (figuras anteriores)"}
-regr %>%
-  .[lat <= -20] %>%
-  .[lev %in% c(200, 50)] %>%
-  # .[PC == "PC1"] %>%
-  .[, FitWave(estimate, 0:3), by = .(lev, PC, term, lat)] %>%
-  ggplot(aes(lat, amplitude)) +
-  geom_line(aes(color = factor(k))) +
-  ggh4x::facet_nested(lev + PC ~ term, scales = "free") +
-  scale_color_brewer("Wavenumber", palette = "Dark2") +
-  scale_y_continuous("Amplitude") +
-  scale_x_latitude() +
-  coord_flip()
-```
-
-Interpretaciones:
-
--   Parte real de EOF1: variabilidady que se proyecta en la dirección del campo medio
-
--   Parte imaginaria de EOF1: variabilidad asociada a la parte simétrica del SAM
-
--   EOF2: PSA1 y PSA2.\
-
--   Parte imaginaria: proyecta sobre la parte asimétrica del SAM (PSA2)
-
--   Parte real: relativamente independiente del SAM
-
-El primer EOF es esencialmente una medida de la variabilidad proyectada en la dirección el campo medio. La parte importante, de todas formas, es la Real.
+# regr %>%
+#   .[lat <= -20] %>%
+#   .[lev %in% c(200, 50)] %>%
+#   # .[PC == "PC1"] %>%
+#   .[, FitWave(estimate, 0:3), by = .(lev, PC, term, lat)] %>%
+#   # .[, pseudor2 := amplitudepseudor2^2/sum(amplitude^2), by = .(lev, PC, term, lat)] %>% 
+#   ggplot(aes(lat, amplitude)) +
+#   geom_line(aes(color = factor(k))) +
+#   ggh4x::facet_nested(PC + lev ~ term, scales = "free", labeller = labeller(lev = lab_lev)) +
+#   scale_color_brewer("Wavenumber", palette = "Dark2") +
+#   scale_y_continuous("Amplitude") +
+#   scale_x_latitude() +
+#   coord_flip() +
+#   panel_background
+```
+
+## Relationship with other variables
+
+### Ozone
 
@@
 ```
-
-(ref:eof2-regr-t-cap) Same as Figure \@ref(fig:eof1-regr-gh) but for the second EOF and temperature.
 
@@
              subset = list(latitude = c(-90, -20))) %>% 
-  normalise_coords()
+  normalise_coords() %>% 
+    .[season(time) == "SON"] %>%
+  .[, .(t = mean(t)), by = .(lev, lon, lat, time = seasonally(time))]
 ```
@@
 ```{r t-vertical, fig.cap = "Regression of EOF1 with mean air temperature between 80ºS and 25ºS."}
-# t_PC1_vertical <- t %>% 
-#   .[cut(ceof$eof[[1]], 1)$left, on = "time"] %>% 
-#   sep_ReIm(format = "wider") %>% 
-#   .[, FitLm(t, Real, Imaginary, se = TRUE), by = .(lev, lon, lat)] %>% 
-#   rm_intercept()
-# t_PC1_vertical %>% 
-#   .[lev %in% c(1, 5, 10, 50)] %>% 
-#   .[, p.value := Pvaluate(estimate, std.error, df, "fdr"), by = .(term, lev)] %>% 
-#     ggplot(aes(lon, lat)) +
-#   geom_contour_fill(aes(z = estimate)) +
-#   stat_subset(aes(subset = p.value <= 0.01 & is.cross(lon, lat)),
-#               size = .5, alpha = 0.5) +
-#   geom_hline(yintercept = c(-80, -35)) +
-#   geom_qmap(~.x[lat <= -20]) +
-#   scale_fill_divergent() + 
-#   coord_polar() +
-#   facet_grid(lev ~ term)
-
 t %>% 
@@
   .[, .(t = weighted.mean(t, cos(lat*pi/180))), by = .(lev, time, lon)] %>%  
+  .[, t := Anomaly(t), by = .(time, lev)] %>% 
   .[cut(ceof$eof[[1]], 1)$left, on = "time"] %>% 
   sep_ReIm(format = "wider") %>% 
+  .[, ":="(Real = as.numeric(scale(Real)), Imaginary = as.numeric(scale(Imaginary))), by = .(PC)] %>% 
   .[, FitLm(t, Real, Imaginary, se = TRUE), by = .(lev, lon)] %>% 
@@
   ggplot(aes(lon, lev)) +
-  geom_contour_fill(aes(z = estimate, fill = ..level..), breaks = AnchorBreaks(0, exclude = 0)) +
+  geom_contour_fill(aes(z = estimate, fill = ..level..), breaks = AnchorBreaks(0, 0.25, exclude = 0)) +
   geom_contour_tanaka(aes(z = estimate), 
-                      breaks = AnchorBreaks(0, exclude = 0)) +
-
+                      breaks = AnchorBreaks(0, 0.25, exclude = 0)) +
   geom_contour_pval(aes(z = p.value)) +
-  scale_fill_divergent_discretised("Air temperature", guide = guide_colorsteps_bottom()) + 
+  scale_fill_divergent_discretised("Air temperature", guide = guide_colorsteps_bottom(25)) + 
   scale_y_level() +
@@
 ```
+
+Figure \@ref(fig:eof1-regr-t) shows regression patterns of EOF1 with air temperature.
+In both levels, the distribution of temperature anomalies mirror the geopotential height anomalies in 50 hPa.
+To understand the vertical distribution of these anomalies, Figure @\ref(fig:t-vertical)) shows the regression pattern of EOF1 and zonal anomalies of air temperature averaged between 80ºS and 35ºS.
+Both pars of EOF1 are associated with a temperature wave 1 pattern which changes phase around 10 hPa.
+This switch in the sign of the correlation in temperature at 10 hPa suggest a relationship with Ozone, since its indicative of the effects of planetary waves in the advection of Ozone [@hartmann1979;@rood1985;@smith1995].
 
@@
   facet_grid(PC~term) +
-  axis_labs_smol
-```
-
-The ozone hole, as measured by Total Ozone Column (TOC) is climatologically not centered over the South Pole, but shifted towards the Weddell Sea, which translates to a large amplitude zonal wave 1 with minimum over the Weddell Sea and maximum over the Davis Sea [@wirth1993].
-The regression
+  axis_labs_smol +
+  tag_facets("panel")
+```
+
+To confirm this possibility, Figure \@ref(fig:o3-regr) shows regression maps of EOF1 with fields of Total Ozone Column (TOC).
+It shows zonal wave 1 pattens in TOC associated with both phases of EOF1.
+Climatologically, the springtime Ozone minimum is located off the South Pole and towards the Weddell Sea [@wirth1993].
+Thus, the Real EOF1 regression pattern (Figure \@ref(fig:o3-regr)a) coincides with the climatological position of the ozone whole while the one for the Imaginary EOF1 is shifted by 90º.
+This suggest that the EOF1 can be a complete description of the TOC zonal wave 1.
 
@@
 
+# temp %>% 
+#   copy() %>% 
+#   na.omit() %>% 
+#   sep_ReIm() %>% 
+#   .[, hgt := scale(hgt), by  = .(part)] %>% 
+#   ggplot(aes(amplitude, hgt)) +
+#   geom_point() +
+#   geom_smooth(method = "lm", color = "black") +
+#   facet_grid(PC ~ part) +
+#   panel_background
+# temp %>% 
+#   .[, correlate(amplitude, Mod(hgt))]
+# 
+# temp %>% 
+#   .[, correlate(phase, Arg(hgt))]
+
+correlation_statistics <- function(x, n, signif = 2) {
+  z <- atanh(x)
+  se <- 1/sqrt(n - 3)
+
+  low <- qnorm(.025, mean = z, sd = se)
+  low <- tanh(low)
+  hig <- qnorm(.975, mean = z, sd = se)
+  hig <- tanh(hig)
+
+  p.value <- pnorm(abs(x)/se, lower.tail = FALSE)
+
+  text <- paste0(signif(x, signif), " (CI: ",
+                 signif(low, signif), " -- ", signif(hig, signif), ")")
+
+  list(estimate = x, 
+       p.value = p.value, 
+       low = low,
+       hig = hig,
+       text = text
+  )
+}
+
+wave_correlation <- temp %>% 
+  na.omit() %>% 
+  .[, .(cor = mean(cos(phase - Arg(hgt))),
+        n = .N)] %>% 
+  .[, correlation_statistics(cor, n)]
+
+temp %>% 
+  ggplot(aes(Mod(hgt), amplitude)) +
+  scale_x_continuous("EOF1 Amplitude") +
+  scale_y_continuous("O3 Wave-1 Amplitude") +
+
 temp %>% 
@@
   na.omit() %>% 
-  sep_ReIm() %>% 
-  .[, hgt := scale(hgt), by  = .(part)] %>% 
-  ggplot(aes(amplitude, hgt)) +
-  geom_point() +
-  geom_smooth(method = "lm") +
-  facet_grid(PC ~ part)
-```
-
-```{r eof2-regr-t, fig.cap = "(ref:eof2-regr-t-cap)"}
-plot_regr(regr, "PC2", c(200, 50), "t")
-```
-
-## Relationship with other indices
-
-### PSA1 and PSA2
+  qwrap(phase = c(0, 2*pi) ~ c(-pi, pi)) %>% 
+  ggplot(aes(Arg(hgt), phase)) +
+  scale_x_continuous("EOF1 Phase",
+                     breaks = seq(-pi, pi, by = 15*pi/180),
+                     labels = function(x) paste0(round(x*180/pi, 5), "º")) +
+  scale_y_continuous("O3 Wave-1 Phase",
+                     breaks = seq(-pi, pi, by = 15*pi/180),
+                     labels = function(x) paste0(round(x*180/pi, 5), "º"))  +
+  plot_layout(ncol  = 1) &
+  geom_point()  &
+  geom_smooth(method = "lm") 
+```
+
+Indeed, Figure \@ref(fig:o3-regr) shows the relationship between amplitude and phase of the planetary wave 1 in Total Ozone Column between 70ºS and 45ºS and amplitude and phase of the EOF1.
+The correlation between the zonal wave 1 of ozone and EOF1 (computed as the mean cosine of the difference in phase) is `r wave_correlation$text`.
+As expected from the location of anomalies in Figure \@ref(fig: o3-regr), the Real EOF1 drives the relationship with amplitude and the Imaginary EOF1 drives the relationship with phase (not shown).
+
+### PSA
+
+We will show that the EOF2 offers an alternative way of representing the PSA which has several advantages over using the second and third principal components .
 
@@
   scale_x_continuous("PSA") +
-  facet_grid(part ~ PC)
-```
+  facet_grid(part ~ PC) +
+  panel_background +
+  tag_facets("rc")
+```
+
+Figure @\ref(fig:psa-eof2) shows the relationship between the two PSA indices and the Real and Imaginary phase of EOF2.
+As anticipated by Figure @\ref(fig: eof2-regr-gh), there is a strong correlation between PSA1 and Imaginary EOF2 (Figure @\ref(fig:psa-eof2)b.1), and between PSA2 and Real EOF2 (Figure @\ref(fig:psa-eof2)a.2).
+Conversely, there is no relationship between PSA1 and Real EOF2, and between PSA2 and Imaginary EOF2 (Figure @\ref(fig:psa-eof2) panels a.1 and b.2).
+So not only this EOF2 represents well both the spatial structure and temporal evolution of the PSA modes, but it's also possible to make a rather clean association between its two phases and the two PSA modes.
+
+This particular rotation of EOF2 is the one which maximises the relationship between ENSO and Imaginary EOF2.
+It is also the same that maximises this clean association between EFO2 parts and PSA modes.
+The reason that the "naive" principal component analysis arrives naturally to this optimal separation is that the imaginary phase is the most common phase (Figure \@ref(fig:phase-histogram)), so it naturally appears first when performing a Principal Component Analysis.
+
+```{r phase-histogram, fig.cap = "Histogram of phase distributions of EOF2."}
+cut(ceof$eof[[1]], 2)$left %>% 
+  .[, arg := Arg(hgt)] %>% 
+  .[, w := Mod(hgt)/sum(Mod(hgt))] %>% 
+  qwrap(arg = c(-pi, pi) ~  c(-pi - 1/4*pi, pi - 1/4*pi)) %>% 
+  ggplot(aes(arg)) +
+  geom_histogram(binwidth = 90*pi/180, center = 0, color = "black", fill = "gray70") +
+  annotate("text",
+           label = c("- Real", "- Imaginary", "+ Real", "+ Imaginary"),
+           color = "black", size = 6,
+           y = 0.5, x = c(-pi, -pi/2, 0, pi/2), angle = 90, hjust = 0) +
+  geom_rug() +
+  scale_y_continuous("Number of years") +
+  scale_x_continuous("Phase",
+                     breaks = seq(-pi - 1/4*pi, pi - 1/4*pi, by = 45*pi/180),
+                     labels = function(x) paste0(round(x*180/pi, 5), "º")) 
+```
+
+By aligning he imaginary phase with the direction with maximum relationship with ENSO, we also aligned it with the phase of maximum occurrence.
+This is analogous to the method in @irving2016, who used reprojection and Fourier filtering to detect "PSA-like" variability and defined as the proper PSA using the peaks of the phase distribution (Compare our Figure \@ref:(phase-histogram) with their Figure 6).
+
+The advantage of our method is that it is much simpler to implement, it provides magnitude and phase naturally, and it facilitates the description of this mode as a propagating wave instead of as standing oscillation.
 
@@
 (ref:sam-eof-vertical-cap) Coefficient of determination between the real and imaginary part of each EOF and the SAM, Asymmetric SAM (A-SAM) and Symmetric SAM (S-SAM) indices computed at each level according to @campitelli2021.
-Points mark estimates with p-value \< 0.01 corrected for False Detection Rate [@benjamini1995].
+Points mark estimates with p-value < 0.01 corrected for False Detection Rate [@benjamini1995].
 
@@
   coord_flip() +
-  tag_facets("rc")
+  tag_facets("rc") +
+  panel_background
 ```
@@
 
-This suggests that the separation between the asymmetric and symmetric part of the SAM is not only statistically possible, but physically plausible.
-The asymmetric part of the SAM might be statistical contamination from one phase of the PSA mode.
+```{r fake-eof, fig.cap = "Principal compomnents of a sinthetic data set of geopotential height as the sum of the fields predicted by the EOF2 plus perfectly zonally symmetric SAM-like annomalies with random amplitude."}
+pattern <- eof_filtered$right %>% 
+  .[era5[lev == 200], on = "time"] %>% 
+  .[, FitLm(hgt, eof), by = .(lon, lat)] %>% 
+  .[, estimate := mean(estimate), by = .(lat, term)] %>%
+  dcast(lon + lat ~ term, value.var = "estimate")
+
+set.seed(42)
+fake_sam <- eof_filtered$right %>% 
+  copy() %>% 
+  .[, eof := sample(eof, replace = TRUE)]  %>%
+  setnames("eof", "value") %>% 
+  .[, pattern[, .(sam = value*eof), by = .(lon, lat)], by = time]
+
+# fake_sam %>% 
+#   .[time %in% unique(time)[1:4]] %>% 
+#   ggplot(aes(lon, lat)) +
+#   geom_contour_fill(aes(z = sam)) +
+#   facet_wrap(~factor(time)) +
+#   coord_polar()
+
+fake_eof <- era5 %>% 
+  .[sep_ReIm(pc2, format = "wide")[, hgt := NULL], on = .NATURAL] %>% 
+  .[lev == 200] %>% 
+  .[lat <= -20] %>% 
+  .[season(time) == "SON"] %>% 
+  .[, hgt := hgt - .lm.fit(cbind(1, Imaginary, Real), hgt)$residuals, .(lon, lat, lev, season(time))] %>% 
+  .[, hgt := Anomaly(hgt), by = .(lat)] %>% 
+  fake_sam[., on = .NATURAL] %>% 
+  .[, hgt := hgt + sam] %>% 
+  .[, eof := Anomaly(hgt)*sqrt(cos(lat*pi/180)), by = .(lon, lat, season(time))] %>% 
+  EOF(eof ~ lon + lat | time, n = 1:3, data = .)
+
+# era5 %>% 
+#   .[sep_ReIm(pc2, format = "wide")[, hgt := NULL], on = .NATURAL] %>% 
+#   .[lev == 200] %>% 
+#   .[lat <= -20] %>% 
+#   .[season(time) == "SON"] %>% 
+#   .[, FitLm(hgt, Real, Imaginary), .(lon, lat)] %>% 
+#   rm_intercept() %>% 
+#   .[, .(lon, lat, eof = estimate, PC = term)] %>% 
+#   rbind(pattern[, .(lon, lat, eof, PC = "SAM")]) %>% 
+#   .[, PC := factor(PC, levels = c("SAM", "Imaginary", "Real"), ordered = TRUE)] %>% 
+#   periodic(lon = c(0, 360)) %>% 
+#   ggplot(aes(lon, lat)) +
+#   geom_contour_fill(aes(z = eof), global.breaks = FALSE, breaks = AnchorBreaks(0, exclude = 0)) +
+#   geom_contour_tanaka(aes(z = eof), global.breaks = FALSE,  breaks = AnchorBreaks(0, exclude = 0)) +
+#   scale_fill_divergent(guide = "none") +
+#   geom_qmap(~.x[lat <= -20]) +  
+#   scale_x_longitude() +
+#   scale_y_latitude() +
+#   coord_polar() +
+#   facet_wrap(~PC) +
+#   axis_labs_smol +
+
+fake_eof %>% 
+  .$left %>% 
+  copy() %>% 
+  .[lat <= -20] %>% 
+  # .[, eof := Anomaly(eof), by = .(lat, PC)] %>%
+  periodic(lon = c(0, 360)) %>% 
+  ggplot(aes(lon, lat)) +
+  geom_contour_fill(aes(z = eof), global.breaks = FALSE, breaks = AnchorBreaks(0, exclude = 0)) +
+  geom_contour_tanaka(aes(z = eof), global.breaks = FALSE,  breaks = AnchorBreaks(0, exclude = 0)) +
+  scale_fill_divergent(guide = "none") +
+  geom_qmap(~.x[lat <= -20]) +  
+  scale_x_longitude() +
+  scale_y_latitude() +
+  coord_polar() +
+  facet_wrap(~PC) +
+  axis_labs_smol
+
+```
+
+This suggests that the asymmetric part of the SAM might be statistical contamination from one phase of the PSA.
+Figure \@ref(fig:fake-eof) serves as an illustration of this kind of statistical mixing of independent modes.
+If shows the leading (and only) EOFs of a synthetic dataset if `r uniqueN(fake_sam$time)` years created by adding a perfectly zonally symmetrical SAM-like pattern with random amplitudes taken from the real SAM, and geopotential height anomalies predicted by EFO2.
+Even though by construction this dataset has a perfectly symmetric SAM, the EOF decomposition mixes this mode with parts of the the zonally asymmetric EOF2 pattern, resulting in a pattern eerily similar to the real SAM pattern.
+
+This demonstrates that the zonal asymmetries of the SAM could very well be mostly a statistical artifact.
+Since many surface impacts are mediated by the asymmetric component, as well as the relationship between SAM and ENSO [@campitelli2021], this potential issue could affect the interpretability of many results involving the SAM.
+Studies on the relationship between the SAM and the PSA pattern would be particularly difficult to interpret.
+
+## Tropical sources
 
@@
 
-?? Compare with Figura 1 b and c (Streamfunction) and Figure 2 d and FIgure 3 d (SST) from [@mo2001] and Figure 7 [@irving2016].
-
-Imaginary more global than real -> is that why it appears mixed with SAM?
-
 Since the Real EOF2 represents just a different phase of the same wave train, one would expect that it would show a similar forcing pattern to the Imaginary EOF with a slight translation of its location.
@@
    annotate(shadowtext:::GeomShadowText,
-           label = c("- Real", "- Imaginary", "+ Real", "+ Imaginary"),
+           label = c("- Real", "- Imaginary", "+ Real", "+ Imaginary", "+ Real"),
            color = "gray60", bg.colour = "white", bg.r = 0.2,
            size = 6,
-           y = -2, x = c(-pi, -pi/2, 0, pi/2), angle = 90, hjust = 0) +
+           y = -Inf, x = c(-pi, -pi/2, 0, pi/2, pi), angle = 90, hjust = 0) +
 
@@
                      labels = function(x) paste0(round(x*180/pi, 5), "º")) +
-  coord_cartesian(xlim = c(-pi, pi))
+  coord_cartesian(xlim = c(-pi, pi)) +
+  panel_background
 ```
@@
 ```{r rotations-pc1, fig.cap = "Rotations of PC1"}
-
 rotations_o3 %>%

### 2021-02-26 R/data_locations.R

@@ -20,7 +20,7 @@ data_path <- function(type = c("raw", "derived"), ...) {
 }
 
 
-data_nc_fun <- function(file, download) {
+data_nc_fun <- function(file, download, citation = NULL) {
    force(file)
 
    if (is.character(download)) {
@@ -30,6 +30,10 @@ data_nc_fun <- function(file, download) {
    function(force_download = FALSE) {
       if (force_download || !file.exists(file)) {
          file <- download(file)
+         if (!is.null(citation)) {
+            write_citation(citation)
+         }
+
       }
 
       checkmate::assert_access(file, access = "r")
@@ -39,11 +43,14 @@ data_nc_fun <- function(file, download) {
 }
 
 
-data_rds_fun <- function(file, download) {
+data_rds_fun <- function(file, download, citation = NULL) {
    force(file)
 
    if (is.character(download)) {
       download <- simple_download(download)
+      if (!is.null(citation)) {
+         write_citation(citation)
+      }
    }
 
    function(force_download = FALSE) {
@@ -60,14 +67,16 @@ data_rds_fun <- function(file, download) {
 #' @export
 #' @rdname data_locations
 ERA5 <- data_nc_fun(file = data_path("raw", "era5.mon.mean.nc"),
-                    download = download_cds(request_era5))
+                    download = download_cds(request_era5),
+                    citation = citation_era5)
 
 
 #' @param force_download Logical indicating wether to force download of data.
 #' @export
 #' @rdname data_locations
 ERA5_BE <- data_nc_fun(file = data_path("raw", "era5be.mon.mean.nc"),
-                       download = download_cds(request_era5be))
+                       download = download_cds(request_era5be),
+                       citation = citation_era5be)
 
 
 #' @export
@@ -100,7 +109,8 @@ HADSST <- data_nc_fun(file = data_path("raw", "hadsst.mon.mean.nc"),
 #' @export
 #' @rdname data_locations
 CMAP <- data_nc_fun(file = data_path("raw", "cmap.mon.mean.nc"),
-                    download = "ftp://ftp.cdc.noaa.gov/Datasets/cmap/std/precip.mon.mean.nc" )
+                    download = "ftp://ftp.cdc.noaa.gov/Datasets/cmap/std/precip.mon.mean.nc",
+                    citation = citation_cmap)
 
 #' @param force_download Logical indicating wether to force download of data.
 #' @export

### 2021-02-26 R/write_data_citation.R

@@ -1,0 +1,87 @@
+write_citation <- function(citation) {
+  date <- format(Sys.Date(), "%d-%m-%Y")
+  text <- citation(date)
+
+  lines <- strsplit(text, "\n")[[1]]
+  line_start <- lines[1]
+  line_end <- rev(lines)[1]
+
+  file <- here::here("analysis", "paper", "data.bib")
+
+  if (!file.exists(file)) {
+    file.create(file)
+  }
+
+  lines <- readLines(file)
+  start <- which(lines == line_start)
+  end <- which(lines == line_end)
+
+  if (length(end) != 0 & length(start) != 0) {
+    lines <- lines[-seq(start, end)]
+  }
+
+  lines <- c(lines, text)
+  lines <- lines[nchar(lines) != 0]
+
+  unlink(file)
+  file.create(file)
+  writeLines(lines, file)
+  return(invisible(file))
+}
+
+
+citation_era5be <- function(date) {
+  paste0(
+r"(--start ERA5BE
+@article{era5be,
+  title = {{{ERA5}} Monthly Averaged Data on Pressure Levels from 1950 to 1978 (Preliminary Version)},
+  author = {Bell, B. and Hersbach, H. and  Berrisford, P. and  Dahlgren, P. and Horányi, A. and Muñoz Sabater, J. and Nicolas, J. and Radu, R. and Schepers, D. and Simmons, A. and Soci, C. and Thépaut, J-N.},
+  year = {2020},
+  journal = {{{Copernicus Climate Change Service}} ({{C3S}}) {{Climate Data Store}} ({{CDS}})},
+  volume = {(Accessed on {$<$})", date, r"({$>$}), https://cds.climate.copernicus.eu/cdsapp\#!/dataset/reanalysis-era5-pressure-levels-monthly-means-preliminary-back-extension?tab=overview}
+}
+--end ERA5BE
+)"
+  )
+}
+
+citation_era5 <- function(date) {
+  paste0(
+r"(--start ERA5
+@article{era5,
+    title = {{{ERA5}} monthly averaged data on pressure levels from 1979 to present},
+    author = {Hersbach, H. and Bell, B. and Berrisford, P. and Biavati, G. and Horányi, A. and Muñoz Sabater, J. and Nicolas, J. and Peubey, C. and Radu, R. and Rozum, I. and Schepers, D. and Simmons, A. and Soci, C. and Dee, D. and Thépaut, J-N.},
[…]

### 2021-03-01 analysis/paper/paper.Rmd
Métodos, datos y formato de paper

@@
 title: "Title Goes Here"
-author:
-  - Myfirstname Mylastname:
-      email: fl@oneplace.edu
-      institute: [UofO]
-      correspondence: true
-  - Yourfirstname Yourlastname:
-      email: fl@another.edu
-      institute: [UofA]
-      correspondence: false
-institute:
-  - UofO: University of One Place
-  - UofA: University of Another Place
+titlerunning:  Title
+authors: 
+- name: Elio Campitelli
+  address: |
+    Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos, Buenos Aires, Argentina
+    CONICET – Universidad de Buenos Aires, Centro de Investigaciones del Mar y la Atmósfera (CIMA), Buenos Aires, Argentina
+    CNRS – IRD – CONICET – UBA, Instituto Franco‐Argentino para el Estudio del Clima y sus Impactos (UMI 3351 IFAECI), Buenos Aires, Argentina
+  email: elio.campitelli@cima.fcen.uba.ar
+
+- name:  Leandro B. Díaz
+  address: |
+    Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos, Buenos Aires, Argentina
+    CONICET – Universidad de Buenos Aires, Centro de Investigaciones del Mar y la Atmósfera (CIMA), Buenos Aires, Argentina
+    CNRS – IRD – CONICET – UBA, Instituto Franco‐Argentino para el Estudio del Clima y sus Impactos (UMI 3351 IFAECI), Buenos Aires, Argentina
+
+- name:  Carolina Vera
+  address: |
+    Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos, Buenos Aires, Argentina
+    CONICET – Universidad de Buenos Aires, Centro de Investigaciones del Mar y la Atmósfera (CIMA), Buenos Aires, Argentina
+    CNRS – IRD – CONICET – UBA, Instituto Franco‐Argentino para el Estudio del Clima y sus Impactos (UMI 3351 IFAECI), Buenos Aires, Argentina
+
+thanks: | 
+  The research was supported by UBACyT20020170100428BA and the CLIMAX Project funded by Belmont Forum/ANR-15-JCL/-0002-01. Elio Campitelli was supported by a PhD grant from CONICET, Argentina.
+
+keywords:
+
+abstract: |
+  abstract
+
 date: "`r format(Sys.time(), '%d %B, %Y')`"
+
+bibliography: 
+  - references.bib
+  - data.bib
+  - packages.bib
+
+bibstyle: spbasic
+# bibstyle options spbasic(default), spphys, spmpsci
 output: 
-  bookdown::pdf_document2:
-    fig_height: 7.6
-    fig_width: 12.8
-  powerpoint_presentation:
-    fig_height: 7.6
-    fig_width: 12.8
-    reference_doc: ../templates/template.pptx
-bibliography: [references.bib, data.bib]
-csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
+  bookdown::pdf_book: 
+      base_format: rticles::springer_article
+      latex_engine: "xelatex"
+      citation_package: "natbib"
+      # pandoc_args: ["--lua-filter=abstract-to-meta.lua"]
 header-includes: 
+ - \usepackage{gensymb}
  - \usepackage{subfig}
-# abstract: |
-#   Text of abstract
-# keywords: |
-#   keyword 1; keyword 2; keyword 3
-# highlights: |
-#   These are the highlights. 
+ - \usepackage[inline]{showlabels}
+ - \usepackage{chngcntr}
+ - \usepackage{natbib}
+ - \usepackage{lineno}
+ - \linenumbers 
 ---
@@
 
+## Figure classes
+DefineClass <- function(class.opts) {
+    class.name <- deparse(substitute(class.opts))
+    class.fun <- function(options) {
+        class <- options[[class.name]]
+        class <- class.opts[[class]]
+        options[names(class)] <- class
+        options
+    }
+    invisible(knitr::opts_hooks$set(
+        setNames(list(class.fun), class.name)))
+}
+
+page.width <- 210 - 30 - 25
+text.width <- page.width - 20
+text.height <- page.height <- 297 - 25 - 25
+
+fig.class <- list(
+    pagewidth = list(
+        fig.width = page.width/25.4,
+        out.width = paste0(page.width, "mm"),
+        # fig.align = "center",
+        fig.env = "figure*",
+        out.extra = " "
+    )
+)
+
+DefineClass(fig.class)
+
 library(metR)
@@
 grid <- theme(panel.grid = element_line(color = "gray10", size = 0.4, linetype = 3))
+no_grid <- theme(panel.grid = element_blank())
 
@@
 o3 <- O3() %>% 
-  readRDS() %>% 
+  ReadNetCDF(vars = c(toc = "tco3")) %>% 
+  normalise_coords() %>% 
+  .[, lon := ConvertLongitude(lon)] %>% 
   .[lat <= -20] %>% 
-  .[, .(toc = mean(total_ozone_column, na.rm = TRUE)), by = .(lon, lat, time = seasonally(time))] %>% 
-  unique(by = c("lon", "lat", "time"))
-o3[, time := year(time)]
+  .[, toc := toc/2.1415e-5] %>%    # transform to Dobson Units (https://sacs.aeronomie.be/info/dobson.php)
+  .[, .(toc = mean(toc, na.rm = TRUE)), by = .(lon, lat, time = seasonally(time))]
+
+o3wave_lats <- c(-70, -45)
 ```
@@
 
-# Data & Methods
-
-@era5 and @era5be
-
-The phase of principal components is defined up to an additive constant.
-For real principal components, this constant can be either 0 or $\pi$, corresponding to a change in sign.
-For complex principal components, it can be any number between 0 and $2\pi$ [@horel1984].
-In this paper we chose the phase of each of the two complex principal components so that the real and imaginary part are easier to interpret.
-For the first complex principal component, the phase was chosen so that the time series corresponding to the imaginary part had a time mean of zero.
-
-For the second complex principal component, the phase was chosen so that the coefficient of determination between the Oceanic Niño Index [@bamston1997] and the real part was minimised.
-It's important to reassure the reader that this procedure does not create a spurious correlation, it only takes whatever relationship that already exist and aligns it with a specific phase.
-If the relationship is very weak, the rotation cannot generate a strong relationship.
-
-We calculated the PSA1 and PSA2 modes as the third and fourth leading EOF of seasonal mean 500 hPa geopotential height with all seasons together following @mo2001.
-
-ERA5, ERA5 BE,
-
-No mostrar los EOFs no complejos. No tiene sentido mostralos si no se analizas.
-
-Relacionar 50hPa con vórtice polar? Temperatura también suma. En 50 el negativo es más intenso que el negativo. ¿Por qué?
-
-Ver qué pasa con U con el EOF1.
-
-En 200 no se ve la señal de jet de la temperatura ¿qué pasa más abajo? en 700 la anomalía de temp coincide más o menos con el geopotentical
-
-Mostrar EOF espaciales.
-
-Pp, ¿cuál es la regresión con SAM asim en 200 hPa?
-
-## Motivation to use complex EOF
+# Methods
+
+## Data
+
+We used monthly geopotential height, air temperature and Total Ozone COlumn (TOC) at 2.5\degree\ longitude by 2.5\degree\ latitude of horizontal resolution and 37 vertical isobaric levels from ERA5 [@era5] for the period 1979 to 2019.
+We restrict most of our analysis to the post-satellite era to avoid any confounding factors arising from the incorporation of satellite observations, but we also use the preliminary back extension of ERA5 from 1950 to 1978 [@era5be] to look at long-term trends.
+We derived streamfunction at 200 hPa from ERA5 vorticity using the FORTRAN subroutine FISHPACK [@fishpack] and computed horizontal wave activity fluxes following @plumb1985 (Equation 5.7).
+
+Precipitation data we used monthly data from the CPC Merged Analysis of Precipitation [@cmap].
+
+We compute the amplitude and phase of the zonal wave 1 by averaging SON TOC between `r knitr::combine_words(LatLabel(o3wave_lats))` for each spring  and computing the Fourier spectrum.
+
+We explore the relationship with the Southern Annular Mode (SAM).
+We compute the level-dependent SAM index as the leading EOF of year-round monthly geopotential height anomalies south of 20ºS at each level for the whole period [@baldwin2009].
+We further split the SAM into its zonally symmetric and zonally asymmetric components (S-SAM and A-SAM indices respectively).
+These are obtained by projecting the zonally asymmetric and zonally symmetric part of the SAM spatial pattern onto monthly geopotential height fields [@campitelli2021].
+Since the analysis here is only for the SON trimester, monthly values were averaged across trimesters.
+
+We also calculated the Pacific South American patterns (PSA1 and PSA2) as the third and fourth leading EOF of seasonal mean 500 hPa geopotential height with all seasons together following @mo2001.
+
+## Complex Empirical Orthogonal Functions
+
+(ref:eof-naive-cap) Spatial patterns of the leading 4 EOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2009.
+(arbitrary units).
+The numbers at the top-left of each panel is the variance explained by each EOF.
 
@@
 
+```{r eof-naive, fig.cap = "(ref:eof-naive-cap)"}
+eof_naive[, eof[[1]]$right, by = lev] %>%
+  periodic(lon = c(0, 360)) %>%
+  ggplot(aes(lon, lat)) +
+  geom_contour_fill(aes(z = hgt, fill = ..level..),
+                    breaks = AnchorBreaks(0, 0.01, exclude = 0)) +
+  geom_contour_tanaka(aes(z = hgt), breaks = AnchorBreaks(0, 0.01, exclude = 0)) +
+
+  geom_qmap(~.x[lat <= -20]) +
+  scale_fill_divergent_discretised(guide = "none")  +
+  scale_y_latitude() +
+  scale_x_longitude() +
+  coord_polar() +
+  theme(axis.text = element_blank()) +
+  facet_grid(lev~EOF, labeller = labeller(lev = lab_lev))  +
+  tagger::tag_facets(tag = "panel", tag_pool = explained, tag_suffix = "")
+```
+
+In traditional EOF analysis zonal waves appear as pairs of EOFs, usually degenerate, that represent similar patterns but shifted in phase.
+For instance, Figure \@ref(fig:eof-naive) shows the leading 4 EOFs of zonally anomalous September through October geopotential height at 50 hPa and 100 hPa.
+It is clear that at 50 hPa the first two EFS represent the same zonal wave 1 pattern and the last 2 represent the same zonal wave 3 pattern shifted by 1/4 wavelength.
+Since these pairs of EOFs seem to represent the same phase-varying structure, it would be desirable to combine them into a single index with amplitude and phase.
+
+Complex Empirical Orthogonal Functions (cEOF) is a useful method to characterise these waves [@horel1984].
+This methods involves augmenting the original fields with the Hilbert transform of the data and then computing the Singular Value Decomposition of it.
+In this work, instead of applying the Hilbert transform to the time series at each point, we apply it to each latitude circle at each moment in time.
+Since the latitude circle is a periodic domain, this procedure does not suffer from edge effects.
+
+The result of the cEOF is a set of complex spatial patterns and complex time series.
+The real and imaginary part of each spatial pattern represent two phases wave-like spatial pattern that are in quadrature.
+The magnitude and argument of each complex time series represent the amplitude and phase of each zonal wave.
+
 ```{r corr-eofs}
@@
                                                   space = "Lab")
-
-eof_naive[, eof[[1]]$left, by = lev] %>% 
-  .[, item := paste(lev, EOF, sep = "_")] %>% 
-  widyr::pairwise_cor(item, time, hgt) %>% 
-  setDT() %>%
-  # .[]
-  .[tstrsplit(item1, "_")[[1]] != tstrsplit(item2, "_")[[1]], ] %>% 
-  .[tstrsplit(item1, "_")[[1]] != 50] %>% 
-  .[, item1 := tstrsplit(item1, "_")[[2]]] %>%
-  .[, item2 := tstrsplit(item2, "_")[[2]]] %>% 
-  .[, r2 := round(correlation^2, 2)] %>% 
-  dcast(item1 ~ item2, value.var = "r2") %>% 
-  setnames("item1", "200 hPa") %>% 
-  kbl2(booktabs = TRUE, caption = "$R^2$ between pairs of EOFs of each level rounded to two decimals.") %>% 
+# 
+# eof_naive[, eof[[1]]$left, by = lev] %>% 
+#   .[, item := paste(lev, EOF, sep = "_")] %>% 
+#   widyr::pairwise_cor(item, time, hgt) %>% 
+#   setDT() %>%
+#   # .[]
+#   .[tstrsplit(item1, "_")[[1]] != tstrsplit(item2, "_")[[1]], ] %>% 
+#   .[tstrsplit(item1, "_")[[1]] != 50] %>% 
+#   .[, item1 := tstrsplit(item1, "_")[[2]]] %>%
+#   .[, item2 := tstrsplit(item2, "_")[[2]]] %>% 
+#   .[, r2 := round(correlation^2, 2)] %>% 
+#   dcast(item1 ~ item2, value.var = "r2") %>% 
+#   setnames("item1", "200 hPa") %>% 
+#   kbl2(booktabs = TRUE, caption = "$R^2$ between pairs of EOFs of each level rounded to two decimals.") %>% 
+#   # add_header_above(c("", "50 hPa" = 4)) %>% 
+#   column_spec2(2:5,  background = spec_color2(.table[[.col]], option = background_palette, 
+#                                               scale_from = c(0, 1)),
+#                color = ifelse(.table[[.col]] > .5, "white", "black")) %>% 
 #   add_header_above(c("", "50 hPa" = 4)) %>%
-  column_spec2(2:5,  background = spec_color2(.table[[.col]], option = background_palette, 
-                                              scale_from = c(0, 1)),
-               color = ifelse(.table[[.col]] > .5, "white", "black")) %>% 
-  add_header_above(c("", "50 hPa" = 4)) %>%
-  kable_classic()
-```
-
-Figure \@ref(fig:cor-eof-naive-1) show the spatial pattern of the 4 leading EOF of zonal anomalies of geopotential height for 50 hPa and 200 hPa computed separately.
-The percentages XXX .
-At 50 hPa, EOF1 and EOF2 on one hand, and EOF3 and EOF4 on the other are clearly pairs of zonal waves with wave numbers 1 and 2-3 respectively shifted by 1/4 wavelength.
-That is, each group of two EOFs represents the same zonal wave structure that changes in magnitude and location.
-Similarly, at 200 hPa EOF2 and EOF3 represent the same zonal wave structure with wave number \~3.
-
-Furthermore, the wave 1 pattern represented by the first two EOFs at 50 hPa is similar to the wave 1 pattern shown as the leading EOF at 200 hPa, and the wave 2-3 pattern represented by the third and fourth EOF at 50 hPa is similar to the wave 3 pattern present in EOF2 and EOF3 at 200 hPa.
-These similarities suggest some level of joint variability across levels.
-Figure \@ref(fig:cor-eof-naive-2) shows the coefficient of determination between temporal series of each EOF at each level.
-EOF3 and EOF4 at 50 hPa are highly correlated with EOF2 and EOF3 at 200 hPa.
-Not only each pair of EOFs represent a single structure within each level, but they represent a coherent pattern of variability between levels.
-Similarly, the leading EOF of each level are relatively highly correlated, again suggesting an unique mode of joint variability.
+#   kable_classic()
+```
 
@@
   setnames("item1", "200 hPa") %>%
-  kbl2(booktabs = TRUE, caption = "$R^2$ of the absolute magnitude of complex EOFs between 200 hPa and 50 hPa computing EOF separatedly for each level.
-The high correlation between levels for the two leading EOFs justifies treating the pattern as a mode of covariability between the stratosphere and the troposphere and, thus, computing the EOFs using both levels at the same time.") %>%
+  kbl2(booktabs = TRUE, caption = "$R^2$ of the absolute magnitude of complex EOFs between 200 hPa and 50 hPa computing EOF separatedly for each level.") %>%
   column_spec2(2:4, background = spec_color2(.table[[.col]], option = background_palette,
@@
 
+Another characteristic seen in Figure \@ref(fig:eof-naive) is that EOF1 at 200 hPa is very similar to EFO1 at 50 hPa.
+The zonal wave 3 structure in EOF2 and EOF3 at 200 hPa are also very similar to the one in EOF3 and EOF4 at 50 hPa.
+Furthermore, Table \@ref(tab:corr-ceof-splitted) shows the coefficient of determination between temporal series of the amplitude of each complex EOF across levels.
+There's a high degree of correlation between the respective EOF1 and EOF2 at each level.
+Both the spatial similarities and temporal correlations suggest some level of joint variability between levels.
+
 Both observations motivate the decision of performing complex EOF jointly between levels.
 The computation of the EOFs was carried out using data from both levels at the same time, therefore, each complex EOF has a spatial part that depends on longitude, latitude and level, an a temporal part that depends only on time.
+
+The phase of principal components is defined up to an additive constant.
+For real principal components, this constant can be either 0 or $\pi$, corresponding to a change in sign.
+For complex principal components, it can be any number between 0 and $2\pi$ [@horel1984].
+Since any choice is arbitrary and equally valid, we chose the phase of each EOF so that the real and imaginary parts are aligned with meaningful phases in our analysis.
+This was done after some exploratory analysis.
+
+For the first complex principal component, the phase was chosen so that the time series corresponding to the real part has the maximum correlation with the zonal wave 1 of Total Ozone Column between `r knitr::combine_words(LatLabel(o3wave_lats))`.
+This also nearly minimises the correlation with the imaginary part (\@ref(fig:rotations-pc1)).
+
+For the second complex principal component, the phase was chosen so that the coefficient of determination between the Oceanic Niño Index [@bamston1997] and the real part was minimised, which also nearly maximises the correlation with the imaginary part (\@ref(fig:rotations-pc2)).
+It's important to reassure the reader that this procedure does not create a spurious correlation, it only takes whatever relationship that already exist and aligns it with a specific phase.
+If the relationship is very weak, the rotation cannot generate a strong relationship.
+
+## Computation procedures
+
+We performed all analysis in this paper using the R programming language [@rcoreteam2020], using the data.table package [@dowle2020] and the metR package [@campitelli2020].
+All graphics are made using ggplot2 [@wickham2009].
+We downloaded data from reanalysis using the ecmwfr package [@hufkens2020] and indices of the ENSO with the rsoi package [@albers2020].
+The paper was rendered using knitr and rmarkdown [@xie2015;@allaire2019].
 
@@
 wave1_o3 <- o3 %>% 
-  .[lat %between% c(-70, -45)] %>% 
+  .[lat %between% o3wave_lats] %>% 
   .[, .(toc = weighted.mean(toc, w = cos(lat*pi/180), na.rm = TRUE)), by = .(time, lon)] %>% 
@@
   .[PC == "PC1"] %>% 
-  .[, time := year(time)] %>% 
   .[wave1_o3, on = "time"]
@@
 
-```{r ceofs, fig.cap = "Leading 2 EOFs of zonal anomalies of geopotential height in 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2009.", fig.subcap = c("Spatial patterns.
-Real part in shading, imaginary part in contours.", "Temproal series."), fig.ncol = 1}
+```{r ceofs-1, fig.cap = "Leading 2 EOFs of zonal anomalies of geopotential height in 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2009.
+Spatial patterns.
+Real part in shading, imaginary part in contours"}
 ceof[season == "SON"] %>%
@@
   tag_facets(tag = "cr")
-
+```
+
+```{r ceofs-2, fig.cap = "Temproal series."}
 ceof %>%
@@
 
-The wave patterns described by these complex EOFs match the patterns seen in the navie EOFs of Figure \@ref(fig: cor-eof-naive-1): The first is a wave 1, while the second is a wave 3.
+The wave patterns described by these complex EOFs match the patterns seen in the traditional EOFs of Figure \@ref(fig:eof-naive): The first is a wave 1, while the second is a wave 3.
 Note that in Figure \@ref(fig:ceofs-2), both parts of the EOF1 have non zero mean.
-This is due to th fact that the geopotential fields that enter into the algorithm are anomalies with respect to the zonal mean (shown in Figure \@ref(fig:mean-anom)), not the time mean.
-The variability associated with the first EOF is variability that projects onto the mean zonally anomalous field.
+This is due to th fact that the geopotential fields that enter into the algorithm are anomalies with respect to the zonal mean, not the time mean.
+The variability associated with the first EOF includes variability that projects onto the mean zonally anomalous field.
 
@@
 Similarly affected are the regression patterns at 200 hPa (Figure \@ref(fig:eof1-regr-gh) row b).
-It is only possible to distinguish some of the wave 1 pattern in relation with the Real EOF1 (Figure @\ref(fig:eof1-regr-gh)b.1).
+It is only possible to distinguish some of the wave 1 pattern in relation with the Real EOF1 (Figure \@ref(fig:eof1-regr-gh)b.1).
 The Imaginary EOF1 shows a much more zonally symmetrical pattern resembling a negative Southern Annular Mode.
@@
 In both levels, the distribution of temperature anomalies mirror the geopotential height anomalies in 50 hPa.
-To understand the vertical distribution of these anomalies, Figure @\ref(fig:t-vertical)) shows the regression pattern of EOF1 and zonal anomalies of air temperature averaged between 80ºS and 35ºS.
+To understand the vertical distribution of these anomalies, Figure \@ref(fig:t-vertical)) shows the regression pattern of EOF1 and zonal anomalies of air temperature averaged between 80ºS and 35ºS.
 Both pars of EOF1 are associated with a temperature wave 1 pattern which changes phase around 10 hPa.
@@
            Imaginary = as.numeric(scale(Imaginary))),
-    by = .(PC)] %>%
-  .[, time := year(time)]
+    by = .(PC)]
 
@@
 o3_regr %>% 
+  periodic(lon = c(0, 360)) %>% 
   .[, p.value := Pvaluate(estimate, std.error, df, "fdr"), by = .(PC, term)] %>% 
@@
 wave1_o3 <- o3 %>% 
-  .[lat %between% c(-70, -45)] %>% 
+  .[lat %between% o3wave_lats] %>% 
   .[, .(toc = mean(toc, na.rm = TRUE)), by = .(time, lon)] %>% 
@@
 
-Figure @\ref(fig:psa-eof2) shows the relationship between the two PSA indices and the Real and Imaginary phase of EOF2.
-As anticipated by Figure @\ref(fig: eof2-regr-gh), there is a strong correlation between PSA1 and Imaginary EOF2 (Figure @\ref(fig:psa-eof2)b.1), and between PSA2 and Real EOF2 (Figure @\ref(fig:psa-eof2)a.2).
-Conversely, there is no relationship between PSA1 and Real EOF2, and between PSA2 and Imaginary EOF2 (Figure @\ref(fig:psa-eof2) panels a.1 and b.2).
+Figure \@ref(fig:psa-eof2) shows the relationship between the two PSA indices and the Real and Imaginary phase of EOF2.
+As anticipated by Figure \@ref(fig: eof2-regr-gh), there is a strong correlation between PSA1 and Imaginary EOF2 (Figure \@ref(fig:psa-eof2)b.1), and between PSA2 and Real EOF2 (Figure \@ref(fig:psa-eof2)a.2).
+Conversely, there is no relationship between PSA1 and Real EOF2, and between PSA2 and Imaginary EOF2 (Figure \@ref(fig:psa-eof2) panels a.1 and b.2).
 So not only this EOF2 represents well both the spatial structure and temporal evolution of the PSA modes, but it's also possible to make a rather clean association between its two phases and the two PSA modes.
@@
 Figure \@ref(fig:sam-eof-vertical) shows the coefficient of determination between the EOFs and three SAM indices.
-The SAM index represents the leading EOF of monthly geopotential height fields south of 20ºS at each level.
-The A-SAM and S-SAM indices represent, respectively, the zonally asymmetric and symmetric component of the SAM and are obtained by projecting the zonally asymmetric and zonally symmetric part of the SAM spatial pattern onto monthly geopotential height fields (@campitelli2021).
-Since the analysis here is only for the SON trimester, monthly values were averaged across semesters (weighted by the number of days in each month).
 
@@
 The correlation between the absolute magnitude of the ONI and the magnitude of the EOF2 is `r cor_enso_mag$text`.
-This relationship appears to be driven only by the three years with strongest ENSO events in the period (`r knitr::combine_words(year(gdata[order(-oni)][1:3]$time))`) which also coindice with the three years with strongest EOF2 magnitude.
+This relationship, however, appears to be driven only by the three years with strongest ENSO events in the period (`r knitr::combine_words(year(gdata[order(-oni)][1:3]$time))`) which also coincide with the three years with strongest EOF2 magnitude.
 If those years are removed, the correlation becomes non-significant (`r cor_enso_mag_outlier$text`).
-Furthermore, the Spearman correlation, which is robust to outliers is also non-significant (`r signif(cor_enso_spearman$estimate, 2)`, p-value = `r report_p(cor_enso_spearman$p.value)`).
+Furthermore, even when using all the datapoints, the Spearman correlation --which is robust to outliers-- is also non-significant (`r signif(cor_enso_spearman$estimate, 2)`, p-value = `r report_p(cor_enso_spearman$p.value)`).
 Therefore, the relationship between the magnitude of the EOF2 train wave and ENSO remains uncertain.
@@
 When excited by tropical SST anomalies, it tends to remain locked to the imaginary phase, with the sign of the geopotential anomalies depending on the sign of the tropical SSTs anomalies.
+This explains the relative overabundance of years with EOF2 near positive and negative imaginary phase in Figure \@ref(fig:phase-histogram).
 
@@
 
-## Naive EOFs
-
-(ref:cap-eof-naive) Leading 4 EOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2009.
-
-(ref:subcap-eof-naive) Spatial patterns (arbitrary units). The numbers at the top-left of each panel is the variance explained by each EOF.
-
-(ref:subcap-cor-eof-naive) Coefficient of determination of the temporal index of each EOF between levels.
-
-```{r cor-eof-naive, fig.cap = "(ref:cap-eof-naive)", fig.subcap = c("(ref:subcap-eof-naive)", "(ref:subcap-cor-eof-naive)"), fig.ncol = 1}
-eof_naive[, eof[[1]]$right, by = lev] %>%
-  periodic(lon = c(0, 360)) %>%
-  ggplot(aes(lon, lat)) +
-
-  geom_contour_fill(aes(z = hgt, fill = ..level..),
-                    breaks = AnchorBreaks(0, 0.01, exclude = 0)) +
-  geom_contour_tanaka(aes(z = hgt), breaks = AnchorBreaks(0, 0.01, exclude = 0)) +
-
-  geom_qmap(~.x[lat <= -20]) +
-  scale_fill_divergent_discretised(guide = "none")  +
-  scale_y_latitude() +
-  scale_x_longitude() +
-  coord_polar() +
-  facet_grid(lev~EOF, labeller = labeller(lev = lab_lev))  +
-  axis_labs_smol +
-  tagger::tag_facets(tag = "panel", tag_pool = explained, tag_suffix = "")
-
-eof_naive[, eof[[1]]$left, by = lev] %>%
-  .[, item := paste(lev, EOF, sep = "_")] %>%
-  widyr::pairwise_cor(item, time, hgt) %>%
-  setDT() %>%
-  # .[]
-  .[tstrsplit(item1, "_")[[1]] != tstrsplit(item2, "_")[[1]], ] %>%
-  .[tstrsplit(item1, "_")[[1]] != 50] %>%
-  .[, item1 := tstrsplit(item1, "_")[[2]]] %>%
-  .[, item2 := tstrsplit(item2, "_")[[2]]] %>%
-  ggplot(aes(item1, item2)) +
-  geom_raster(aes(fill = correlation^2)) +
-  shadowtext::geom_shadowtext(aes(label = round(correlation^2, 2)),
-                              color = "black", bg.colour = "white", bg.r = 0.2) +
-  scale_fill_gradient(guide = "none", low = "white", high = scales::muted("red"), limits = c(0, 1)) +
-  scale_x_discrete("200 hPa") +
-  scale_y_discrete("50 hPa") +
-  plot_layout(ncol = 1)
-```
-
 ## Chosen rotations of the EOFs
@@
 
+## Code availability {-}
+
+A version-controlled repository of the code used to create this analysis, including the code used to download the data can be found at https://github.com/eliocamp/shceof.
+
 # References

### 2021-03-01 analysis/paper/paper.tex
Métodos, datos y formato de paper

% !TeX program = pdfLaTeX
\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage[utf8]{inputenc}

\usepackage[hyphens]{url} % not crucial - just used below for the URL
\usepackage{hyperref}
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

%
% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
%
% insert here the call for the packages your document requires
%\usepackage{latexsym}
% etc.
%
% please place your own definitions here and don't use \def but
% \newcommand{}{}
%
% Insert the name of "your journal" with
% \journalname{myjournal}
%

%% load any required packages here

% Pandoc citation processing

\usepackage{gensymb}
\usepackage{subfig}
\usepackage[inline]{showlabels}
\usepackage{chngcntr}
\usepackage{natbib}
\usepackage{lineno}
\linenumbers
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\begin{document}

\title{Title Goes Here \thanks{The research was supported by UBACyT20020170100428BA and the CLIMAX Project funded by Belmont Forum/ANR-15-JCL/-0002-01. Elio Campitelli was supported by a PhD grant from CONICET, Argentina.} }

    \titlerunning{Title}

\author{  Elio Campitelli \and  Leandro B. Díaz \and  Carolina Vera \and  }

\institute{
        Elio Campitelli \at
     Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos, Buenos Aires, Argentina
CONICET -- Universidad de Buenos Aires, Centro de Investigaciones del Mar y la Atmósfera (CIMA), Buenos Aires, Argentina
CNRS -- IRD -- CONICET -- UBA, Instituto Franco‐Argentino para el Estudio del Clima y sus Impactos (UMI 3351 IFAECI), Buenos Aires, Argentina \\
     \email{\href{mailto:elio.campitelli@cima.fcen.uba.ar}{\nolinkurl{elio.campitelli@cima.fcen.uba.ar}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        Leandro B. Díaz \at
     Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos, Buenos Aires, Argentina
CONICET -- Universidad de Buenos Aires, Centro de Investigaciones del Mar y la Atmósfera (CIMA), Buenos Aires, Argentina
CNRS -- IRD -- CONICET -- UBA, Instituto Franco‐Argentino para el Estudio del Clima y sus Impactos (UMI 3351 IFAECI), Buenos Aires, Argentina \\
     %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        Carolina Vera \at
     Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos, Buenos Aires, Argentina
CONICET -- Universidad de Buenos Aires, Centro de Investigaciones del Mar y la Atmósfera (CIMA), Buenos Aires, Argentina
CNRS -- IRD -- CONICET -- UBA, Instituto Franco‐Argentino para el Estudio del Clima y sus Impactos (UMI 3351 IFAECI), Buenos Aires, Argentina \\
     %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
    }

\date{Received: date / Accepted: date}
% The correct dates will be entered by the editor

\maketitle

\begin{abstract}
abstract
\\
\keywords{
         \and
    }

\end{abstract}

\def\spacingset#1{\renewcommand{\baselinestretch}%
{#1}\small\normalsize} \spacingset{1}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

\hypertarget{methods}{%
\section{Methods}\label{methods}}

\hypertarget{data}{%
\subsection{Data}\label{data}}

We used monthly geopotential height, air temperature and Total Ozone COlumn (TOC) at 2.5\degree~longitude by 2.5\degree~latitude of horizontal resolution and 37 vertical isobaric levels from ERA5 \citep{era5} for the period 1979 to 2019. We restrict most of our analysis to the post-satellite era to avoid any confounding factors arising from the incorporation of satellite observations, but we also use the preliminary back extension of ERA5 from 1950 to 1978 \citep{era5be} to look at long-term trends. We derived streamfunction at 200 hPa from ERA5 vorticity using the FORTRAN subroutine FISHPACK \citep{fishpack} and computed horizontal wave activity fluxes following \citet{plumb1985} (Equation 5.7).

Precipitation data we used monthly data from the CPC Merged Analysis of Precipitation \citep{cmap}.

We compute the amplitude and phase of the zonal wave 1 by averaging SON TOC between 70°S and 45°S for each spring and computing the Fourier spectrum.

We explore the relationship with the Southern Annular Mode (SAM). We compute the level-dependent SAM index as the leading EOF of year-round monthly geopotential height anomalies south of 20ºS at each level for the whole period \citep{baldwin2009}. We further split the SAM into its zonally symmetric and zonally asymmetric components (S-SAM and A-SAM indices respectively). These are obtained by projecting the zonally asymmetric and zonally symmetric part of the SAM spatial pattern onto monthly geopotential height fields \citep{campitelli2021}. Since the analysis here is only for the SON trimester, monthly values were averaged across trimesters.

We also calculated the Pacific South American patterns (PSA1 and PSA2) as the third and fourth leading EOF of seasonal mean 500 hPa geopotential height with all seasons together following \citet{mo2001}.

\hypertarget{complex-empirical-orthogonal-functions}{%
\subsection{Complex Empirical Orthogonal Functions}\label{complex-empirical-orthogonal-functions}}

\begin{figure}
\centering
\includegraphics{../figures/eof-naive-1.pdf}
\caption{\label{fig:eof-naive}Spatial patterns of the leading 4 EOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2009. (arbitrary units). The numbers at the top-left of each panel is the variance explained by each EOF.}
\end{figure}

In traditional EOF analysis zonal waves appear as pairs of EOFs, usually degenerate, that represent similar patterns but shifted in phase. For instance, Figure \ref{fig:eof-naive} shows the leading 4 EOFs of zonally anomalous September through October geopotential height at 50 hPa and 100 hPa. It is clear that at 50 hPa the first two EFS represent the same zonal wave 1 pattern and the last 2 represent the same zonal wave 3 pattern shifted by 1/4 wavelength. Since these pairs of EOFs seem to represent the same phase-varying structure, it would be desirable to combine them into a single index with amplitude and phase.

Complex Empirical Orthogonal Functions (cEOF) is a useful method to characterise these waves \citep{horel1984}. This methods involves augmenting the original fields with the Hilbert transform of the data and then computing the Singular Value Decomposition of it. In this work, instead of applying the Hilbert transform to the time series at each point, we apply it to each latitude circle at each moment in time. Since the latitude circle is a periodic domain, this procedure does not suffer from edge effects.

The result of the cEOF is a set of complex spatial patterns and complex time series. The real and imaginary part of each spatial pattern represent two phases wave-like spatial pattern that are in quadrature. The magnitude and argument of each complex time series represent the amplitude and phase of each zonal wave.

\begin{table}

\caption{\label{tab:corr-ceof-splitted}$R^2$ of the absolute magnitude of complex EOFs between 200 hPa and 50 hPa computing EOF separatedly for each level.}
\centering
\begin{tabular}[t]{l>{}r>{}r>{}r}
\toprule
\multicolumn{1}{c}{} & \multicolumn{3}{c}{50 hPa} \\
\cmidrule(l{3pt}r{3pt}){2-4}
200 hPa & PC1 & PC2 & PC3\\
\midrule
PC1 & \cellcolor[HTML]{E1C0BB}{\textcolor{black}{0.28}} & \cellcolor[HTML]{FDFAFA}{\textcolor{black}{0.02}} & \cellcolor[HTML]{FDFAFA}{\textcolor{black}{0.02}}\\
PC2 & \cellcolor[HTML]{FFFFFF}{\textcolor{black}{0.00}} & \cellcolor[HTML]{BB7A73}{\textcolor{white}{0.60}} & \cellcolor[HTML]{FDFAFA}{\textcolor{black}{0.02}}\\
PC3 & \cellcolor[HTML]{FFFFFF}{\textcolor{black}{0.00}} & \cellcolor[HTML]{FFFFFF}{\textcolor{black}{0.00}} & \cellcolor[HTML]{FDFCFC}{\textcolor{black}{0.01}}\\
\bottomrule
\end{tabular}
\end{table}

Another characteristic seen in Figure \ref{fig:eof-naive} is that EOF1 at 200 hPa is very similar to EFO1 at 50 hPa. The zonal wave 3 structure in EOF2 and EOF3 at 200 hPa are also very similar to the one in EOF3 and EOF4 at 50 hPa. Furthermore, Table \ref{tab:corr-ceof-splitted} shows the coefficient of determination between temporal series of the amplitude of each complex EOF across levels. There's a high degree of correlation between the respective EOF1 and EOF2 at each level. Both the spatial similarities and temporal correlations suggest some level of joint variability between levels.

Both observations motivate the decision of performing complex EOF jointly between levels. The computation of the EOFs was carried out using data from both levels at the same time, therefore, each complex EOF has a spatial part that depends on longitude, latitude and level, an a temporal part that depends only on time.

The phase of principal components is defined up to an additive constant. For real principal components, this constant can be either 0 or \(\pi\), corresponding to a change in sign. For complex principal components, it can be any number between 0 and \(2\pi\) \citep{horel1984}. Since any choice is arbitrary and equally valid, we chose the phase of each EOF so that the real and imaginary parts are aligned with meaningful phases in our analysis. This was done after some exploratory analysis.

For the first complex principal component, the phase was chosen so that the time series corresponding to the real part has the maximum correlation with the zonal wave 1 of Total Ozone Column between 70°S and 45°S. This also nearly minimises the correlation with the imaginary part (\ref{fig:rotations-pc1}).

For the second complex principal component, the phase was chosen so that the coefficient of determination between the Oceanic Niño Index \citep{bamston1997} and the real part was minimised, which also nearly maximises the correlation with the imaginary part (\ref{fig:rotations-pc2}). It's important to reassure the reader that this procedure does not create a spurious correlation, it only takes whatever relationship that already exist and aligns it with a specific phase. If the relationship is very weak, the rotation cannot generate a strong relationship.

\hypertarget{computation-procedures}{%
\subsection{Computation procedures}\label{computation-procedures}}

We performed all analysis in this paper using the R programming language \citep{rcoreteam2020}, using the data.table package \citep{dowle2020} and the metR package \citep{campitelli2020}. All graphics are made using ggplot2 \citep{wickham2009}. We downloaded data from reanalysis using the ecmwfr package \citep{hufkens2020} and indices of the ENSO with the rsoi package \citep{albers2020}. The paper was rendered using knitr and rmarkdown \citep{xie2015, allaire2019}.

\hypertarget{results}{%
\section{Results}\label{results}}

\hypertarget{description-of-eofs}{%
\subsection{Description of EOFs}\label{description-of-eofs}}

\begin{figure}
\centering
\includegraphics{../figures/ceofs-1-1.pdf}
\caption{\label{fig:ceofs-1}Leading 2 EOFs of zonal anomalies of geopotential height in 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2009. Spatial patterns. Real part in shading, imaginary part in contours}
\end{figure}

\begin{figure}
\centering
\includegraphics{../figures/ceofs-2-1.pdf}
\caption{\label{fig:ceofs-2}Temproal series.}
\end{figure}

Figures \ref{fig:ceofs-1} and \ref{fig:ceofs-2} show, respectively, the spatial and temporal parts of the first two leading complex EOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa. In the spatial patterns in Figure \ref{fig:ceofs-1}, the real (in shading) and the imaginary parts (in contour) are in quadrature by construction, so that each EOF describe a single wave-like pattern whose amplitude and position (i.e.~phase) is controlled by the magnitude and phase of the complex temporal EOF.

The wave patterns described by these complex EOFs match the patterns seen in the traditional EOFs of Figure \ref{fig:eof-naive}: The first is a wave 1, while the second is a wave 3. Note that in Figure \ref{fig:ceofs-2}, both parts of the EOF1 have non zero mean. This is due to th fact that the geopotential fields that enter into the algorithm are anomalies with respect to the zonal mean, not the time mean. The variability associated with the first EOF includes variability that projects onto the mean zonally anomalous field.

\begin{figure}
\centering
\includegraphics{../figures/extended-series-1.pdf}
\caption{\label{fig:extended-series}Temporal series extended using ERA5 back extended preliminary edition (period 1950 -- 1978) and ERA5 (period 1979 -- 2019). Each series is computed by projecting monthly geopotential height zonal anomalies standardised by level south of 20ºS onto the corresponding spatial pattern and scaled to zero mean and unit standard deviation.}
\end{figure}

Figure \ref{fig:extended-series} shows temporal series of the two complex EOFs extended beyond the satellite era using the preliminary ERA5 back extension going back to 1950 (which we call the ``hybrid ERA5'' reanalysis). There is a downward trend in the real part of EOF1 in the hybrid reanalysis (Figure \ref{fig:extended-series}a.1, p-value = 0.0011). There is no significant trend in any of the complex parts of EOF2.

\hypertarget{regressions}{%
\subsection{Regressions}\label{regressions}}

The spatial patterns shown in Figure \ref{fig:ceofs-1} are ``pure'' EOF patterns. But as they are derived by removing the zonally symmetric circulation, they might not include all the variability that is actually associated with the EOF time series. They are also idealised in that they are forced by construction, to be Hilber transforms of each other. To get a more realistic view of the real-world variability described by these EOFS, we computed regression patterns of each EOF with the complete geopotential fields.

\begin{figure}
\centering
\includegraphics{../figures/eof1-regr-gh-1.pdf}
\caption{\label{fig:eof1-regr-gh}Regression coefficients of the real and imaginary part of the first complex EOF on SON geopotential height for the 1979 -- 2019 period. These coefficients come from multiple linear regression involving the real and imaginary part of both EFO2.}
\end{figure}

Figure \ref{fig:eof1-regr-gh} shows regression patterns of EOF1 and geopotential height. At 50 hPa (Figure \ref{fig:eof1-regr-gh} row a), both the Real and Imaginary EOF1 are associated with planetary wave 1 patterns, that are 90º off phase. Their phases coincide with the ones shown in Figure \ref{fig:ceofs-1}a.1, with the positive centre of the Real EOF1 located towards the dateline, and the one of the Imaginary EOF1 located over Eastern Antarctica. However, the Real EOF1 pattern is substantially altered by the zonally symmetric circulation. Instead of a clear wave 1 pattern, the regression pattern can be better describes as a monopole with its centre displaced from the South Pole.
Similarly affected are the regression patterns at 200 hPa (Figure \ref{fig:eof1-regr-gh} row b). It is only possible to distinguish some of the wave 1 pattern in relation with the Real EOF1 (Figure \ref{fig:eof1-regr-gh}b.1). The Imaginary EOF1 shows a much more zonally symmetrical pattern resembling a negative Southern Annular Mode.

With the exception of the Imaginary EFO1, it's clear that these patterns are very different than the idealised versions shown in Figure \ref{fig:ceofs-1}, particularly at 200 hPa. Moreover, only in the stratosphere these patterns actually show a distinguishable wave 1 pattern shifted in phase by 90º. This should not be utterly surprising, since the spatial patterns of the \emph{naive} EOFs only show this feature at 50 hPa, suggesting that using the complex EOF method is ``forcing'' a wave structure where there is none.

Therefore, the magnitude and phase of the EOF1 are associated with the magnitude and phase of a zonal wave only in the stratosphere. While in the troposphere they are associated with slightly off-centre monopoles.

\begin{figure}
\centering
\includegraphics{../figures/eof2-regr-gh-1.pdf}
\caption{\label{fig:eof2-regr-gh}Same as Figure \ref{fig:eof1-regr-gh} but for the second EOF.}
\end{figure}

Figure @ref(fig: eof2-regr-gh) shows the regression pattern of geopotential height and the EOF2. Unlike in the case of EOF1, in this case the regression patterns represent relatively well the \emph{idealised} patterns from Figure \ref{fig:ceofs-1}. Although there are some differences (particularly in 50 hPa), the wave trains identified before are well characterised and patterns associated with the Real EOF2 are 90º out of phase with those associated with the Imaginary EOF2. Zonal wave 3 dominates all fields, but only in the western hemisphere.

EOF2 then represents a equivalent barotropic wave train that is very similar to the the Pacific South American Patterns \citep{mo2001}. Comparing the location of the positive anomaly near 90ºW in column b of Figure @ref(fig: eof2-regr-gh) with Figure 1 a and b from \citet{mo2001}, the Real EOF2 can roughly be identified with the PSA2, while the Imaginary EOF2 resembles PSA1. This separation is not completely meaningful in this case, since by treating this pattern as a single wave train any one pair of orthogonal waves can characterise this variability.

\hypertarget{relationship-with-other-variables}{%
\subsection{Relationship with other variables}\label{relationship-with-other-variables}}

\hypertarget{ozone}{%
\subsubsection{Ozone}\label{ozone}}

\begin{figure}
\centering
\includegraphics{../figures/eof1-regr-t-1.pdf}
\caption{\label{fig:eof1-regr-t}Same as Figure \ref{fig:eof1-regr-gh} but for air temperature.}
\end{figure}

\begin{figure}
\centering
\includegraphics{../figures/t-vertical-1.pdf}
\caption{\label{fig:t-vertical}Regression of EOF1 with mean air temperature between 80ºS and 25ºS.}
\end{figure}

Figure \ref{fig:eof1-regr-t} shows regression patterns of EOF1 with air temperature. In both levels, the distribution of temperature anomalies mirror the geopotential height anomalies in 50 hPa. To understand the vertical distribution of these anomalies, Figure \ref{fig:t-vertical}) shows the regression pattern of EOF1 and zonal anomalies of air temperature averaged between 80ºS and 35ºS. Both pars of EOF1 are associated with a temperature wave 1 pattern which changes phase around 10 hPa. This switch in the sign of the correlation in temperature at 10 hPa suggest a relationship with Ozone, since its indicative of the effects of planetary waves in the advection of Ozone \citep{hartmann1979, rood1985, smith1995}.

\begin{figure}
\centering
\includegraphics{../figures/o3-regr-1.pdf}
\caption{\label{fig:o3-regr}Regression of standarised EOF1 and Total Ozone column.}
\end{figure}

To confirm this possibility, Figure \ref{fig:o3-regr} shows regression maps of EOF1 with fields of Total Ozone Column (TOC). It shows zonal wave 1 pattens in TOC associated with both phases of EOF1. Climatologically, the springtime Ozone minimum is located off the South Pole and towards the Weddell Sea \citep{wirth1993}. Thus, the Real EOF1 regression pattern (Figure \ref{fig:o3-regr}a) coincides with the climatological position of the ozone whole while the one for the Imaginary EOF1 is shifted by 90º. This suggest that the EOF1 can be a complete description of the TOC zonal wave 1.

\begin{figure}
\centering
\includegraphics{../figures/wave1-o3-1.pdf}
\caption{\label{fig:wave1-o3}EOF1 and wave1 o3 amplitude.}
\end{figure}

Indeed, Figure \ref{fig:o3-regr} shows the relationship between amplitude and phase of the planetary wave 1 in Total Ozone Column between 70ºS and 45ºS and amplitude and phase of the EOF1. The correlation between the zonal wave 1 of ozone and EOF1 (computed as the mean cosine of the difference in phase) is 0.87 (CI: 0.76 -- 0.93). As expected from the location of anomalies in Figure @ref(fig: o3-regr), the Real EOF1 drives the relationship with amplitude and the Imaginary EOF1 drives the relationship with phase (not shown).

\hypertarget{psa}{%
\subsubsection{PSA}\label{psa}}

We will show that the EOF2 offers an alternative way of representing the PSA which has several advantages over using the second and third principal components .

\begin{figure}
\centering
\includegraphics{../figures/psa-eof2-1.pdf}
\caption{\label{fig:psa-eof2}EFO2 and PSA}
\end{figure}

Figure \ref{fig:psa-eof2} shows the relationship between the two PSA indices and the Real and Imaginary phase of EOF2. As anticipated by Figure @ref(fig: eof2-regr-gh), there is a strong correlation between PSA1 and Imaginary EOF2 (Figure \ref{fig:psa-eof2}b.1), and between PSA2 and Real EOF2 (Figure \ref{fig:psa-eof2}a.2). Conversely, there is no relationship between PSA1 and Real EOF2, and between PSA2 and Imaginary EOF2 (Figure \ref{fig:psa-eof2} panels a.1 and b.2). So not only this EOF2 represents well both the spatial structure and temporal evolution of the PSA modes, but it's also possible to make a rather clean association between its two phases and the two PSA modes.

This particular rotation of EOF2 is the one which maximises the relationship between ENSO and Imaginary EOF2. It is also the same that maximises this clean association between EFO2 parts and PSA modes. The reason that the ``naive'' principal component analysis arrives naturally to this optimal separation is that the imaginary phase is the most common phase (Figure \ref{fig:phase-histogram}), so it naturally appears first when performing a Principal Component Analysis.

\begin{figure}
\centering
\includegraphics{../figures/phase-histogram-1.pdf}
\caption{\label{fig:phase-histogram}Histogram of phase distributions of EOF2.}
\end{figure}

By aligning he imaginary phase with the direction with maximum relationship with ENSO, we also aligned it with the phase of maximum occurrence. This is analogous to the method in \citet{irving2016}, who used reprojection and Fourier filtering to detect ``PSA-like'' variability and defined as the proper PSA using the peaks of the phase distribution (Compare our Figure @ref:(phase-histogram) with their Figure 6).

The advantage of our method is that it is much simpler to implement, it provides magnitude and phase naturally, and it facilitates the description of this mode as a propagating wave instead of as standing oscillation.

\hypertarget{sam}{%
\subsubsection{SAM}\label{sam}}

\begin{figure}
\centering
\includegraphics{../figures/sam-eof-vertical-1.pdf}
\caption{\label{fig:sam-eof-vertical}Coefficient of determination between the real and imaginary part of each EOF and the SAM, Asymmetric SAM (A-SAM) and Symmetric SAM (S-SAM) indices computed at each level according to \citet{campitelli2021}. Points mark estimates with p-value \textless{} 0.01 corrected for False Detection Rate \citep{benjamini1995}.}
\end{figure}

Figure \ref{fig:sam-eof-vertical} shows the coefficient of determination between the EOFs and three SAM indices.

Both EOFs bear some modest relationship with the SAM index (thick green line in Figure \ref{fig:sam-eof-vertical}), although not statistically significant at every level. The split between A-SAM and S-SAM gives more insight into the nature of the relationship. The relationship between the SAM and the imaginary EOF1 (Figure \ref{fig:sam-eof-vertical}.b1) is mediated by S-SAM in the troposphere, but by the A-SAM in the stratosphere.

The Imaginary EOF2 is related with the SAM through the A-SAM in the troposphere, with up to -96\% of shared variance, reached at 225 hPa (Figure \ref{fig:sam-eof-vertical}.b2).

\begin{figure}
\centering
\includegraphics{../figures/eof-filtered-1.pdf}
\caption{\label{fig:eof-filtered}Spatial pattern of the leading EOF of 200 hPa geopotential height with the variability of EOF2 filtered out South of 20º. Arbitrary scale.}
\end{figure}

As a further illustration, Figure \ref{fig:eof-filtered} shows the spatial pattern of the leading EOF of 200 hPa geopotential height when the variability associated with EOF2 is removed. The resulting pattern is an nearly zonally symmetrical annular mode. This mode is highly correlated with SAM (0.75 (CI: 0.58 -- 0.86)), but only with the symmetrical part (0.88 (CI: 0.79 -- 0.94)). Note that the usual definition of the PSA modes as the second and third EOFs creates modes orthogonal to the SAM (defined as the first EOF) and thus impedes this kind of filtering.

\begin{figure}
\centering
\includegraphics{../figures/fake-eof-1.pdf}
\caption{\label{fig:fake-eof}Principal compomnents of a sinthetic data set of geopotential height as the sum of the fields predicted by the EOF2 plus perfectly zonally symmetric SAM-like annomalies with random amplitude.}
\end{figure}

This suggests that the asymmetric part of the SAM might be statistical contamination from one phase of the PSA. Figure \ref{fig:fake-eof} serves as an illustration of this kind of statistical mixing of independent modes. If shows the leading (and only) EOFs of a synthetic dataset if 41 years created by adding a perfectly zonally symmetrical SAM-like pattern with random amplitudes taken from the real SAM, and geopotential height anomalies predicted by EFO2. Even though by construction this dataset has a perfectly symmetric SAM, the EOF decomposition mixes this mode with parts of the the zonally asymmetric EOF2 pattern, resulting in a pattern eerily similar to the real SAM pattern.

This demonstrates that the zonal asymmetries of the SAM could very well be mostly a statistical artifact. Since many surface impacts are mediated by the asymmetric component, as well as the relationship between SAM and ENSO \citep{campitelli2021}, this potential issue could affect the interpretability of many results involving the SAM. Studies on the relationship between the SAM and the PSA pattern would be particularly difficult to interpret.

\hypertarget{tropical-sources}{%
\subsection{Tropical sources}\label{tropical-sources}}

\begin{figure}
\centering
\includegraphics{../figures/sst-psi-2-1.pdf}
\caption{\label{fig:sst-psi-2}Regression maps of EOF2 with SST (column a) and streamfunction zonal anomalies with their corresponding activity wave flux (column b). Areas marked with dots have p-values smaller than 0.05 adjusted for FDR.}
\end{figure}

Figure \ref{fig:sst-psi-2} shows regression maps between EOF2 and Sea Surface Temperatures (SST) Streamfunction at 200 hPa. The Imaginary EOF2 is associated with strong positive SST anomalies on the Central Pacific and negative anomalies on an area across the North of Australia and New Zealand, the South Pacific Convergence Zone (SPCZ) (Figure \ref{fig:sst-psi-2}.a2). This pattern is almost canonically positive ENSO and indeed, the correlation between the Imaginary EOF2 and the Oceanic Niño Index (\citep{bamston1997}) is very high 0.76 (CI: 0.59 -- 0.86). Streamfunction anomalies show a coherent picture. The Imaginary EOF2 is associated with strong wave-like streamfunction anomalies emanating from the tropics (Figure \ref{fig:sst-psi-2}.b2). This is consistent with what we know of the effect of ENSO on the extratropics: SST anomalies initiate anomalous convection that excites Rossby waves that propagate meridionally towards higher latitudes.

Since the Real EOF2 represents just a different phase of the same wave train, one would expect that it would show a similar forcing pattern to the Imaginary EOF with a slight translation of its location. However, Figure \ref{fig:sst-psi-2}.a1 and b1 show that the Real EOF2 is not associated either with any significant SST nor streamfunction anomalies in the tropics. The correlation between the Real EOF2 and ENSO is also not significant (). This lack of tropical signal suggests a radically different nature of the different phases of the EOF2 wave train.

\begin{figure}
\centering
\includegraphics{../figures/enso-phase-1.pdf}
\caption{\label{fig:enso-phase}Relationship between ENSO and phase of EOF2 for the period 1979 -- 2019. Colours denote years with magnitude of EOF2 greater or smaller than the 50th percentile. Black line is the fit ONI \textasciitilde{} cos(phase) + sin(phase) computed by OLS weighted by the magnitude of EOF2.}
\end{figure}

To better explore the relationship between tropical forcing and phase of the EOF2, Figure \ref{fig:enso-phase} plots the ONI index and the phase of the EOF2 for each year between 1979 and 2019, highlighting years in which the magnitude of EOF2 was higher than 50\% of the years. In years with positive ENSO, the phase of the EOF2 is always around +90º (corresponding with positive imaginary part) and vice versa. In years with near neutral ENSO, the phase of the EOF2 is much more variable. The black line in Figure \ref{fig:enso-phase} tries to quantify this relationship. Is the equation \(ONI = 1.1\sin{phase}\), whose coefficient we fitted by weighted least squares using the magnitude of the EFO2 as weight. The \(r^2\) corresponding to the fit is 0.56, with p-value \textless{} 0.001.

Figure \ref{fig:enso-phase} suggest that strong EOF2 years tend to coincide with strong ENSO years. The correlation between the absolute magnitude of the ONI and the magnitude of the EOF2 is 0.44 (CI: 0.16 -- 0.66). This relationship, however, appears to be driven only by the three years with strongest ENSO events in the period (2015, 1997, and 1982) which also coincide with the three years with strongest EOF2 magnitude. If those years are removed, the correlation becomes non-significant (0.044 (CI: -0.28 -- 0.36)). Furthermore, even when using all the datapoints, the Spearman correlation --which is robust to outliers-- is also non-significant (0.2, p-value = = 0.2). Therefore, the relationship between the magnitude of the EOF2 train wave and ENSO remains uncertain.

All this suggests that the wave train represented by EOF2 can be both forced by tropical SSTs and also a part of the internal variability of the extratropical atmosphere. When internally forced, the wave train has little phase preference. When excited by tropical SST anomalies, it tends to remain locked to the imaginary phase, with the sign of the geopotential anomalies depending on the sign of the tropical SSTs anomalies. This explains the relative overabundance of years with EOF2 near positive and negative imaginary phase in Figure \ref{fig:phase-histogram}.

\begin{figure}
\centering
\includegraphics{../figures/sst-psi-1-1.pdf}
\caption{\label{fig:sst-psi-1}Same as Figure \ref{fig:sst-psi-2} but for EOF1.}
\end{figure}

Figure \ref{fig:sst-psi-1} shows SST and streamfunction regression maps for EOF1. There is no significant pattern of SST anomalies associated with either the Real or Imaginary EOF1. Consistently, streamfunction anomalies don't show any tropical influence. On the contrary, the Real EOF1 is associated with wave activity fluxes that flow equatorward from the coast of Antarctica around 150ºE along with wave activity fluxes that move between the positive and negative of Streamfunction anomalies along Antarctica.

\hypertarget{precipitation}{%
\subsection{Precipitation}\label{precipitation}}

\begin{figure}
\centering
\includegraphics{../figures/pp-america-1.pdf}
\caption{\label{fig:pp-america}Regression of SON mean precipitation anomalies in South America (mm per day, shaded) and (column a) EOF1 the (row 1) Imaginary and (column 1) Real phase. For the 1979 -- 2018 period. Black contours with dots indicate areas with p-value smaller than 0.05 controlling for False Detection Rate.}
\end{figure}

Figure \ref{fig:pp-america} shows regression maps of seasonal precipitation with each EOF in South America. EOF1 (Figure \ref{fig:pp-america} column a) is associated with increased precipitation over Paraguay and and some parts of Brazil. Although these positive anomalies are not very strong. The strongest precipitation anomalies are the ones associated with the Imaginary EOF2. The positive anomalies on Southeaster South America (SESA) and Chile, and negative anomalies over Southern Brazil is a well known springtime precipitation signature of ENSO \citep{cai2020a} and it's also virtually identical to the precipitation anomalies associated with the Asymmetric SAM \citep{campitelli2021}. This is not surprising considering the close relationship between the ONI, the Asymmetric SAM index and the Imaginary EOF2 shown previously. The Real EOF2, on the other hand, is associated with negative precipitation anomalies in a smaller area of SESA. The relationship between precipitation anomalies in SESA and the phase of EOF2 follows a curve similar to that of Figure \ref{fig:enso-phase} (not shown).

\begin{figure}
\centering
\includegraphics{../figures/pp-africa-1.pdf}
\caption{\label{fig:pp-africa}Same as Figure \ref{fig:pp-america} but for South of Africa.}
\end{figure}

\begin{figure}
\centering
\includegraphics{../figures/pp-oceania-1.pdf}
\caption{\label{fig:pp-oceania}Same as Figure \ref{fig:pp-america} but for New Zealand and neighbouring islands.}
\end{figure}

\hypertarget{conclusions}{%
\section{Conclusions}\label{conclusions}}

\hypertarget{appendix}{%
\section{Appendix}\label{appendix}}

\hypertarget{chosen-rotations-of-the-eofs}{%
\subsection{Chosen rotations of the EOFs}\label{chosen-rotations-of-the-eofs}}

\begin{figure}
\centering
\includegraphics{../figures/rotations-pc1-1.pdf}
\caption{\label{fig:rotations-pc1}Rotations of PC1}
\end{figure}

\begin{figure}
\centering
\includegraphics{../figures/rotations-pc2-1.pdf}
\caption{\label{fig:rotations-pc2}R2 between EOF2 Real and Imaginary parts and ONI index for different rotation parameters of the real and imaginary parts.}
\end{figure}

\hypertarget{code-availability}{%
\subsection*{Code availability}\label{code-availability}}
\addcontentsline{toc}{subsection}{Code availability}

A version-controlled repository of the code used to create this analysis, including the code used to download the data can be found at \url{https://github.com/eliocamp/shceof}.

\hypertarget{references}{%
\section{References}\label{references}}

\bibliographystyle{spbasic}
\bibliography{references.bib,data.bib,packages.bib}

\end{document}

### 2021-03-01 R/0-download.R

@@ -99,7 +99,7 @@ request_era5be <- list(
 request_era5 <- list(
   format = "netcdf",
   product_type = "monthly_averaged_reanalysis",
-  variable = c("geopotential", "temperature", "vorticity", "divergence", "total_column_ozone"),
+  variable = c("geopotential", "temperature", "vorticity", "divergence"),
   pressure_level = c("1", "2", "3", "5", "7", "10", "20", "30", "50",
                      "70", "100", "125", "150", "175", "200", "225",
                      "250", "300", "350", "400", "450", "500", "550",
@@ -111,6 +111,20 @@ request_era5 <- list(
   grid = c("2.5", "2.5"),
   dataset_short_name = "reanalysis-era5-pressure-levels-monthly-means"
 )
+
+request_o3_era5 <- list(
+  format = "netcdf",
+  variable = "total_column_ozone",
+  product_type = "monthly_averaged_reanalysis",
+  year = as.character(1979:2019),
+  month = formatC(9:11, width = 2, flag = 0),
+  time = "00:00",
+  area = c(0, -180, -90, 180),
+  grid = c("2.5", "2.5"),
+  dataset_short_name = "reanalysis-era5-single-levels-monthly-means",
+  target = "ozone.era5.nc"
+)
+
 
 request_era5_sst <- list(
   format = "netcdf",
@@ -158,6 +172,8 @@ request_o3 <- list(
   dataset_short_name = "satellite-ozone-v1",
   target = "ozone.zip"
 )
+
+
 
 simple_download <- function(url) {
   force(url)

### 2021-03-01 R/write_data_citation.R

@@ -1,6 +1,5 @@
 write_citation <- function(citation) {
-  date <- format(Sys.Date(), "%d-%m-%Y")
-  text <- citation(date)
+  text <- citation()
 
   lines <- strsplit(text, "\n")[[1]]
   line_start <- lines[1]
@@ -30,7 +29,8 @@ write_citation <- function(citation) {
 }
 
 
-citation_era5be <- function(date) {
+citation_era5be <- function() {
+  date <- format(Sys.Date(), "%d-%m-%Y")
   paste0(
 r"(--start ERA5BE
 @article{era5be,
@@ -45,7 +45,8 @@ r"(--start ERA5BE
   )
 }
 
-citation_era5 <- function(date) {
+citation_era5 <- function() {
+  date <- format(Sys.Date(), "%d-%m-%Y")
   paste0(
 r"(--start ERA5
 @article{era5,
@@ -61,7 +62,28 @@ r"(--start ERA5
   )
 }
 
+
+citation_o3_era5 <- function() {
+  date <- format(Sys.Date(), "%d-%m-%Y")
+  paste0(
+r"(--start o3-ERA5
+@article{o3-era5,
+    title = {{{ERA5}} monthly averaged data on single levelss from 1979 to present},
+    author = {Hersbach, H. and Bell, B. and Berrisford, P. and Biavati, G. and Horányi, A. and Muñoz Sabater, J. and Nicolas, J. and Peubey, C. and Radu, R. and Rozum, I. and Schepers, D. and Simmons, A. and Soci, C. and Dee, D. and Thépaut, J-N.},
+    year = {2019},
+    journal = {{{Copernicus Climate Change Service}} ({{C3S}}) {{Climate Data Store}} ({{CDS}})},
+    volume = {(Accessed on {$<$})", date, r"({$>$})},
+    doi = {10.24381/cds.f17050d7}
+    }
+--end o3-ERA5
+)"
+  )
+}
+
+
+
 citation_cmap <- function() {
+  date <- format(Sys.Date(), "%d-%m-%Y")
   paste0(
 r"(--start CMAP
 @article{cmap,
@@ -70,7 +92,7 @@ r"(--start CMAP
     author = {Xie, Pingping and Arkin, Phillip A.},
     year = {1997},
     month = nov,
-    volume = {78},
+    volume = {78 (Accessed on {$<$})", date, r"({$>$})},
     pages = {2539--2558},
     publisher = {{American Meteorological Society}},
     issn = {0003-0007},

### 2021-03-04 analysis/paper/paper.Rmd
Draft resultados

@@
   comment = "#>",
-  cache.extra = 41,
+  cache.extra = 42,
   cache.path = "../cache/paper/",
@@
 fig.class <- list(
+  colwidth = list(
+    fig.width = 5
+  ),
     pagewidth = list(
@@
 
-theme_set(theme_shceof(base_size = 16))
+theme_set(theme_shceof(base_size = 12))
 
@@
 
-guide_colorsteps_bottom <- function(width = 25, height = 0.5, even.steps = FALSE, ...) {
+guide_colorsteps_bottom <- function(width = 20, height = 0.5, even.steps = FALSE, ...) {
   guide_colorsteps(show.limits = TRUE, even.steps = even.steps,
@@
                                    pattern_fill = "black",
-                                   pattern_density = 0.1,
+                                   pattern_density = 0.05,
                                    pattern_alpha = 1,
                                    fill = NA,
-                                   pattern_spacing = 0.02))
+                                   pattern_spacing = 0.025))
+
+update_geom_defaults(GeomSmooth, list(colour = "#0d52bf"))
 
@@
 
-We compute the amplitude and phase of the zonal wave 1 by averaging SON TOC between `r knitr::combine_words(LatLabel(o3wave_lats))` for each spring  and computing the Fourier spectrum.
+We compute the amplitude and phase of the zonal wave 1 by averaging SON TOC between `r knitr::combine_words(LatLabel(o3wave_lats))` for each Spring and computing the Fourier spectrum.
 
@@
 
-(ref:eof-naive-cap) Spatial patterns of the leading 4 EOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2009.
-(arbitrary units).
-The numbers at the top-left of each panel is the variance explained by each EOF.
+(ref:eof-naive-cap) Spatial patterns of the leading 4 EOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2009 (arbitrary units).
+The numbers at the top-left of each panel ate the variance explained by each EOF.
 
@@
 The zonal wave 3 structure in EOF2 and EOF3 at 200 hPa are also very similar to the one in EOF3 and EOF4 at 50 hPa.
-Furthermore, Table \@ref(tab:corr-ceof-splitted) shows the coefficient of determination between temporal series of the amplitude of each complex EOF across levels.
+Furthermore, Table \@ref(tab:corr-ceof-splitted) shows the coefficient of determination between time series of the amplitude of each complex EOF across levels.
 There's a high degree of correlation between the respective EOF1 and EOF2 at each level.
@@
 This was done after some exploratory analysis.
-
-For the first complex principal component, the phase was chosen so that the time series corresponding to the real part has the maximum correlation with the zonal wave 1 of Total Ozone Column between `r knitr::combine_words(LatLabel(o3wave_lats))`.
-This also nearly minimises the correlation with the imaginary part (\@ref(fig:rotations-pc1)).
-
-For the second complex principal component, the phase was chosen so that the coefficient of determination between the Oceanic Niño Index [@bamston1997] and the real part was minimised, which also nearly maximises the correlation with the imaginary part (\@ref(fig:rotations-pc2)).
-It's important to reassure the reader that this procedure does not create a spurious correlation, it only takes whatever relationship that already exist and aligns it with a specific phase.
-If the relationship is very weak, the rotation cannot generate a strong relationship.
-
-## Computation procedures
-
-We performed all analysis in this paper using the R programming language [@rcoreteam2020], using the data.table package [@dowle2020] and the metR package [@campitelli2020].
-All graphics are made using ggplot2 [@wickham2009].
-We downloaded data from reanalysis using the ecmwfr package [@hufkens2020] and indices of the ENSO with the rsoi package [@albers2020].
-The paper was rendered using knitr and rmarkdown [@xie2015;@allaire2019].
-
-# Results
-
-## Description of EOFs
 
@@
 
+(ref:rotations-cap) a) Correlation of the Real and Imaginary EOF1 with wave 1 amplitude of  Total Ozone Column averaged between `r knitr::combine_words(LatLabel(o3wave_lats))` and b) $R^2$ between the Real and Imaginary EOF2 and the ONI index for different values of rotation.
+The vertical line marks the selected rotation, which maximises de correlation.
+
+```{r rotations, fig.cap = "(ref:rotations-cap)", fig.width=5, fig.height=2.5}
+r2 <- expression(`$r^2$` = paste("", "r", phantom()^{
+    paste("2")
+}, "", ""))
+
+rotations_o3 %>%
+  ggplot(aes(rotation, cor)) +
+  geom_vline(xintercept = best_rotation_pc1) +
+  geom_hline(yintercept = 0,  size = 0.2, color = "gray50") +
+  geom_line(aes(color = part)) +
+  scale_x_continuous("Rotation", breaks = seq(-pi, pi, by = 60*pi/180),
+                     labels = function(x) paste0(round(x*180/pi, 5), "º")) +
+  scale_y_continuous("Correlation") +
+  scale_color_brewer(NULL, palette = "Dark2", labels = lab_cplx) +
+
+rotations_pc2 %>%
+  melt(id.vars = "rotation") %>%
+  ggplot(aes(rotation, value^2)) +
+  geom_hline(yintercept = 0,  size = 0.2, color = "gray50") +
+  geom_vline(xintercept = best_rotation_pc2) +
+  geom_line(aes(color = variable)) +
+  scale_x_continuous("Rotation", breaks = seq(-pi, pi, by = 60*pi/180),
+                     labels = function(x) paste0(round(x*180/pi, 5), "º")) +
+  scale_y_continuous(r2) +
+  scale_color_brewer(NULL, palette = "Dark2", labels = lab_cplx) +
+
+  plot_layout(guides = "collect") +
+  plot_annotation(tag_levels = "a", tag_suffix = ")")
+```
+
+For the first complex principal component, the phase was chosen so that the time series corresponding to the real part has the maximum correlation with the zonal wave 1 of Total Ozone Column between `r knitr::combine_words(LatLabel(o3wave_lats))`.
+This also nearly minimises the correlation with the imaginary part (Figure \@ref(fig:rotations)a).
+
+For the second complex principal component, the phase was chosen so that the coefficient of determination between the Oceanic Niño Index [@bamston1997] and the real part was minimised, which also nearly maximises the correlation with the imaginary part (Figure \@ref(fig:rotations)b).
+It's important to reassure the reader that this procedure does not create a spurious correlation, it only takes whatever relationship that already exist and aligns it with a specific phase.
+If the relationship is very weak, the rotation cannot generate a strong relationship.
+
+## Computation procedures
+
+We performed all analysis in this paper using the R programming language [@rcoreteam2020], using the data.table package [@dowle2020] and the metR package [@campitelli2020].
+All graphics are made using ggplot2 [@wickham2009].
+We downloaded data from reanalysis using the ecmwfr package [@hufkens2020] and indices of the ENSO with the rsoi package [@albers2020].
+The paper was rendered using knitr and rmarkdown [@xie2015;@allaire2019].
+
+# Results
+
+## Description of EOFs
+
 ```{r define-plot_ceof}
@@
 
-```{r ceofs-1, fig.cap = "Leading 2 EOFs of zonal anomalies of geopotential height in 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2009.
-Spatial patterns.
-Real part in shading, imaginary part in contours"}
+```{r ceofs-1, fig.cap = "Spatial patterns of the leading 2 complex EOFs of zonal anomalies of geopotential height in 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2009.
+Real part in shading, imaginary part in contours", fig.width=5, fig.height=5}
 ceof[season == "SON"] %>%
@@
 
-```{r ceofs-2, fig.cap = "Temproal series."}
+```{r ceofs-2, fig.cap = "Time series of the leading 2 complex EOFs of zonal anomalies of geopotential height in 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2009.", fig.height=4, fig.width = 5}
 ceof %>%
@@
 
-(ref:extended-series-cap) Temporal series extended using ERA5 back extended preliminary edition (period 1950 -- 1978) and ERA5 (period 1979 -- 2019).
-Each series is computed by projecting monthly geopotential height zonal anomalies standardised by level south of 20ºS onto the corresponding spatial pattern and scaled to zero mean and unit standard deviation.
-
-```{r extended-series, fig.cap = "(ref:extended-series-cap)"}
+```{r era_more}
 pattern <- ceof$eof[[1]]$right %>%
@@
 
+```
+
+(ref:extended-series-cap) Time series extended using ERA5 back extended preliminary edition (period 1950 -- 1978) and ERA5 (period 1979 -- 2019).
+Each series is computed by projecting monthly geopotential height zonal anomalies standardised by level south of 20ºS onto the corresponding spatial pattern and scaled to zero mean and unit standard deviation.
+
+```{r extended-series, fig.cap = "(ref:extended-series-cap)", fig.height=4, fig.width = 5}
 # Check that the extended index coincides with the
@@
 
-Figure \@ref(fig:extended-series) shows temporal series of the two complex EOFs extended beyond the satellite era using the preliminary ERA5 back extension going back to 1950 (which we call the "hybrid ERA5" reanalysis).
+Figure \@ref(fig:extended-series) shows time series of the two complex EOFs extended beyond the satellite era using the preliminary ERA5 back extension going back to 1950 (which we call the "hybrid ERA5" reanalysis).
 There is a downward trend in the real part of EOF1 in the hybrid reanalysis (Figure \@ref(fig:extended-series)a.1, p-value `r trends[part == "Real" & PC == "PC1"]$p_report`).
@@
   }
+
   data[lat <= -20] %>%
@@
     ggplot(aes(lon, lat)) +
-    ditto(
     geom_contour_fill(aes(z = estimate), breaks = AnchorBreaks(exclude = 0), global.breaks = FALSE) +
-        geom_contour_tanaka() +
-        scale_fill_divergent(var_title, guide = guide_colorstrip_bottom())
-    ) +
+    geom_contour_tanaka(aes(z = estimate), breaks = AnchorBreaks(exclude = 0), global.breaks = FALSE) +
+    scale_fill_divergent(var_title, breaks = AnchorBreaks(0, exclude = 0),
+                         guide = guide_colorstrip_bottom(20)) +
     geom_contour_pval(aes(z = p.value)) +
@@
 }
-
 ```
@@
 
-```{r eof1-regr-gh, fig.cap = "(ref:eof1-regr-gh-cap)"}
+```{r eof1-regr-gh, fig.cap = "(ref:eof1-regr-gh-cap)", fig.height=5.4, fig.width=5}
 plot_regr(regr, "PC1", c(200, 50), "hgt") +
@@
   NULL
+
 ```
@@
 
-```{r eof2-regr-gh, fig.cap = "(ref:eof2-regr-gh-cap)"}
+```{r eof2-regr-gh, fig.cap = "(ref:eof2-regr-gh-cap)", fig.height=5.4, fig.width=5}
 plot_regr(regr, "PC2", c(200, 50), "hgt")
+
 ```
@@
 
-```{r eof1-regr-t, fig.cap = "(ref:eof1-regr-t-cap)"}
+```{r eof1-regr-t, fig.cap = "(ref:eof1-regr-t-cap)",  fig.height=5.4, fig.width=5}
 plot_regr(regr, "PC1", c(200, 50), "t")
+
 ```
@@
 
-```{r t-vertical, fig.cap = "Regression of EOF1 with mean air temperature between 80ºS and 25ºS."}
+(ref:t-vertical-cap) Regression of EOF1 with mean air temperature averaged between 80ºS and 25ºS for the period 1979 -- 2019.
+
+```{r t-vertical, fig.cap = "(ref:t-vertical-cap)", fig.height=3.5, fig.width=5}
 t %>% 
@@
   ggplot(aes(lon, lev)) +
-  geom_contour_fill(aes(z = estimate, fill = ..level..), breaks = AnchorBreaks(0, 0.25, exclude = 0)) +
+  geom_contour_fill(aes(z = estimate, fill = ..level..),
+                    breaks = AnchorBreaks(0, 0.25, exclude = 0)) +
   geom_contour_tanaka(aes(z = estimate), 
@@
   geom_contour_pval(aes(z = p.value)) +
-  scale_fill_divergent_discretised("Air temperature", guide = guide_colorsteps_bottom(25)) + 
+  scale_fill_divergent_discretised("Air temperature", guide = guide_colorsteps_bottom(20)) + 
   scale_y_level() +
-  scale_x_longitude() +
+  scale_x_longitude(ticks = 60) +
   facet_grid(~term) +
@@
 
-```{r o3-regr, fig.cap = "Regression of standarised EOF1 and Total Ozone column."}
+(raf:o3-regr-cap) Regression coefficients of the real and imaginary part of the first complex EOF on SON mean Total Ozone Column for the 1979 -- 2019 period.
+
+```{r o3-regr, fig.cap = "(raf:o3-regr-cap)", fig.height=4, fig.width = 5}
 o3_regr %>% 
@@
   scale_fill_divergent_discretised("Total Ozone Column", 
-                                   guide = guide_colorsteps_bottom()) + 
+                                   guide = guide_colorsteps_bottom(20)) + 
   scale_x_longitude() +
@@
   tag_facets("panel")
+
 ```
@@
 
-```{r wave1-o3, fig.cap = "EOF1 and wave1 o3 amplitude."}
+(ref:wave1-o3-cap) Relationship between EOF1 amplitude and phase with amplitude and phase of the zonal wave 1 in Total Ozone Column averaged between `r knitr::combine_words(LatLabel(o3wave_lats))` for the period 1979 -- 2019.
+
+```{r wave1-o3, fig.cap = "(ref:wave1-o3-cap)", fig.width=5, fig.height=3}
 wave1_o3 <- o3 %>% 
@@
   scale_x_continuous("EOF1 Amplitude") +
-  scale_y_continuous("O3 Wave-1 Amplitude") +
+  scale_y_continuous("O3 Wave-1\nAmplitude") +
 
@@
                      labels = function(x) paste0(round(x*180/pi, 5), "º")) +
-  scale_y_continuous("O3 Wave-1 Phase",
-                     breaks = seq(-pi, pi, by = 15*pi/180),
+  scale_y_continuous("O3 Wave-1\nPhase",
+                     breaks = seq(-pi, pi, by = 30*pi/180),
                      labels = function(x) paste0(round(x*180/pi, 5), "º"))  +
@@
   geom_smooth(method = "lm")
+
 ```
@@
 
-```{r psa-eof2, fig.cap = "EFO2 and PSA"}
+(ref:psa-eof2-cap) Relationship between the Real and Imaginary parts of EOF2 and the PSA1 and PSA2 modes computed as the second and third EOFs of seasonal geopotential height anomalies [@mo2001].
+For the period 1979 -- 2019.
+
+```{r psa-eof2, fig.cap = "(ref:psa-eof2-cap)", fig.width=5, fig.height=4}
 pc2 <- ceof$eof[[1]]$left[PC == "PC2"] %>% 
@@
   geom_point() +
+  geom_smooth(method = "lm") +
+  geom_smooth(method = "lm") + 
   scale_y_continuous("EOF2") +
@@
   tag_facets("rc")
+
 ```
@@
 
-```{r phase-histogram, fig.cap = "Histogram of phase distributions of EOF2."}
+(ref:phase-histogram-cap) Histogram showing phase distribution of EOF2 for the period 1979 -- 2019.
+
+```{r phase-histogram, fig.cap = "(ref:phase-histogram-cap)", fig.width=2.5, fig.height=2.5}
 cut(ceof$eof[[1]], 2)$left %>% 
@@
            label = c("- Real", "- Imaginary", "+ Real", "+ Imaginary"),
-           color = "black", size = 6,
+           color = "black", size = 3,
            y = 0.5, x = c(-pi, -pi/2, 0, pi/2), angle = 90, hjust = 0) +
@@
   scale_x_continuous("Phase",
-                     breaks = seq(-pi - 1/4*pi, pi - 1/4*pi, by = 45*pi/180),
+                     breaks = seq(-pi - 1/4*pi, pi - 1/4*pi, by = 90*pi/180),
                      labels = function(x) paste0(round(x*180/pi, 5), "º")) 
@@
 
-```{r sam-eof-vertical, fig.cap = "(ref:sam-eof-vertical-cap)"}
+```{r sam-eof-vertical, fig.cap = "(ref:sam-eof-vertical-cap)", fig.height=4, fig.width=5}
 time <- ceof[, eof[[1]]$left, by = .(season)] %>%
@@
   scale_x_level() +
-  scale_y_continuous("r²", limits = c(0, 1)) +
+  scale_y_continuous(r2, limits = c(0, 1)) +
   scale_size_manual(guide = "none",
@@
 
-```{r eof-filtered, fig.cap = "Spatial pattern of the leading EOF of 200 hPa geopotential height with the variability of EOF2 filtered out South of 20º. Arbitrary scale."}
+```{r eof-filtered, fig.cap = "Spatial pattern of the leading EOF of 200 hPa geopotential height with the variability of EOF2 filtered out South of 20º.
+Arbitrary scale.", fig.width=2.5, fig.height=2.5}
 eof_filtered <- era5 %>% 
@@
   coord_polar() +
-  axis_labs_smol 
+  theme(axis.text = element_blank())
 ```
@@
 
-```{r fake-eof, fig.cap = "Principal compomnents of a sinthetic data set of geopotential height as the sum of the fields predicted by the EOF2 plus perfectly zonally symmetric SAM-like annomalies with random amplitude."}
+(ref:fake-eof-cap) EOFs of a synthetic data set of 41 times steps of geopotential height fields as the sum of the fields predicted by the EOF2 plus perfectly zonally symmetric SAM-like anomalies with random amplitude.
+
+```{r fake-eof, fig.cap = "(ref:fake-eof-cap)", fig.width=5, fig.height=2.5}
 pattern <- eof_filtered$right %>% 
@@
   facet_wrap(~PC) +
-  axis_labs_smol
-
+  theme(axis.text = element_blank())
 ```
@@
     geom_streamline(aes(dx = fx, dy = fy), min.L = 3, L = 10, n = 30,
-                    size = 0.2, res = 2) +
+                    size = 0.2, res = 2, nx = 25, ny = 15) +
     geom_qmap(keep = .015) +
@@
 
-```{r sst-psi-2,  fig.cap = "(ref:sst-psi-2-cap)"}
+```{r sst-psi-2,  fig.cap = "(ref:sst-psi-2-cap)", fig.with = 6, fig.height=4}
 plot_sst_psi(which_pc = "PC2", which_lev = 200)
@@
 Colours denote years with magnitude of EOF2 greater or smaller than the 50th percentile.
-Black line is the fit ONI ~ cos(phase) + sin(phase) computed by OLS weighted by the magnitude of EOF2."}
+Black line is the fit ONI ~ cos(phase) + sin(phase) computed by OLS weighted by the magnitude of EOF2.", fig.width = 5, fig.height=3}
 gdata <- ceof[, eof[[1]]$left, by = .(season)] %>%
@@
            color = "gray60", bg.colour = "white", bg.r = 0.2,
-           size = 6,
+           size = 3,
            y = -Inf, x = c(-pi, -pi/2, 0, pi/2, pi), angle = 90, hjust = 0) +
@@
                  shape  = mag >= quantile(mag, 0.5)),
-             size = 4) +
+             size = 3) +
 
@@
   scale_x_continuous("Phase",
-                     breaks = seq(-pi, pi, by = 30*pi/180),
+                     breaks = seq(-pi, pi, by = 45*pi/180),
                      labels = function(x) paste0(round(x*180/pi, 5), "º")) +
@@
 The black line in Figure \@ref(fig:enso-phase) tries to quantify this relationship.
-Is the equation $ONI = `r signif(coef(cos_model)[1], 2)`\sin{phase}$, whose coefficient we fitted by weighted least squares using the magnitude of the EFO2 as weight.
+Is the equation $ONI = `r signif(coef(cos_model)[1], 2)`\sin{(phase)}$, whose coefficient we fitted by weighted least squares using the magnitude of the EFO2 as weight.
 The $r^2$ corresponding to the fit is `r signif(summary(cos_model)$r.squared, 2)`, with p-value `r report_p(coef(summary(cos_model))[4])`.
@@
 If those years are removed, the correlation becomes non-significant (`r cor_enso_mag_outlier$text`).
-Furthermore, even when using all the datapoints, the Spearman correlation --which is robust to outliers-- is also non-significant (`r signif(cor_enso_spearman$estimate, 2)`, p-value = `r report_p(cor_enso_spearman$p.value)`).
+Furthermore, even when using all the datapoints, the Spearman correlation --which is robust to outliers-- is also non-significant (`r signif(cor_enso_spearman$estimate, 2)`, p-value`r report_p(cor_enso_spearman$p.value)`).
 Therefore, the relationship between the magnitude of the EOF2 train wave and ENSO remains uncertain.
@@
 
-```{r sst-psi-1, fig.cap = "(ref:sst-psi-1-cap)"}
+```{r sst-psi-1, fig.cap = "(ref:sst-psi-1-cap)", fig.with = 6, fig.height=4}
 plot_sst_psi(which_pc = "PC1", which_lev = 200)
@@
 
-```{r pp-america, fig.cap = "(ref:pp-america-cap)"}
+```{r pp-america, fig.cap = "(ref:pp-america-cap)", fig.width=5, fig.height=6}
 plot_pp(data = regr_pp, which_pc = c("PC1", "PC2"), which_continent = "america") 
@@
   #          fill = NA, color = "gray50", size = 0.3)
+  #          
 ```
@@
 Figure \@ref(fig:pp-america) shows regression maps of seasonal precipitation with each EOF in South America.
-EOF1 (Figure \@ref(fig:pp-america) column a) is associated with increased precipitation over Paraguay and and some parts of Brazil.
-Although these positive anomalies are not very strong.
+EOF1 is associated with slight decreases in precipitation in Southern Brazil and Paraguay, although these are not statistically significant \@ref(fig:pp-america) column a.
 The strongest precipitation anomalies are the ones associated with the Imaginary EOF2.
 The positive anomalies on Southeaster South America (SESA) and Chile, and negative anomalies over Southern Brazil is a well known springtime precipitation signature of ENSO [@cai2020a] and it's also virtually identical to the precipitation anomalies associated with the Asymmetric SAM [@campitelli2021].
-This is not surprising considering the close relationship between the ONI, the Asymmetric SAM index and the Imaginary EOF2 shown previously.
+This is not surprising considering the close relationship between the ONI, the Asymmetric SAM index and the Imaginary EOF2 shown previously, but further consolidates the identification of this mode with the PSA pattern.
 The Real EOF2, on the other hand, is associated with negative precipitation anomalies in a smaller area of SESA.
@@
 
+(ref:pp-oceania-cap) Same as Figure \@ref(fig:pp-america) but for New Zealand and neighbouring islands.
+
+```{r box_australia}
+box_australia <- data.frame(PC = "PC2", 
+                            lon = c(130, 155),
+                            lat = c(-40, -15))
+```
+
+```{r pp-oceania, fig.cap = "(ref:pp-oceania-cap)", fig.width=5, fig.height=4}
+plot_pp(regr_pp, c("PC1", "PC2"), which_continent =  "oceania") +
+  geom_rect(data = box_australia, aes(xmin = min(lon), xmax = max(lon),
+                                      ymin = min(lat), ymax = max(lat)),
+            fill = NA, colour = "gray20") 
+```
+
+On the antipodes, precipitation anomalies in New Zealand and neighbouring islands associated with EOF1 are also weak and not statistically significant (Figure \@ref(fig:pp-oceania) column a) whereas EFO2 is associated with large and expansive anomalies.
+The Real EOF2 is associated with positive seasonal anomalies in Northern and Eastern Australia, and negative ones over New Zealand, while precipitation anomalies associated with Imaginary EOF2 are negative over all Eastern Australia.
+Again, these negative anomalies are similar to the springtime precipitation anomalies observed in relation with the Asymmetric SAM [@campitelli2021].
+
+(ref:australia-pp-phase-cap) Same as Figure \@ref(fig:enso-phase) but for precipitacion anomalies averaged between  `r knitr::combine_words(LonLabel(box_australia$lon))` and `r knitr::combine_words(LatLabel(box_australia$lat))` (box shown in Figure \@ref(fig:pp-oceania) column b).
+
+```{r australia-pp-phase, fig.cap = "(ref:australia-pp-phase-cap)", fig.width = 5, fig.height=3}
+gdata <- cmap %>% 
+  .[lon %between% box_australia$lon  & lat %between% box_australia$lat] %>% 
+  .[, .(pp = weighted.mean(pp, cos(lat*pi/180))), by = time] %>% 
+  .[, pp := Anomaly(pp), by = season(time)] %>% 
+  .[ceof$eof[[1]]$left, on  = "time"] %>% 
+  .[PC == "PC2"] %>% 
+  .[, arg := Arg(hgt)] %>% 
+  .[, mag := Mod(hgt)]
+
+model <- lm(pp ~ I(sin(arg)) + I(cos(arg)) - 1, data = gdata)
+fstatistic <- summary(model)$fstatistic
+p_value <- pf(fstatistic["value"], fstatistic["numdf"], fstatistic["dendf"], 
+        lower.tail = FALSE)
+
+gdata %>% 
+  ggplot(aes(Arg(hgt), pp)) +
+  geom_hline(yintercept = 0, size = 0.2, color = "gray50") +
+  geom_vline(xintercept = 0, size = 0.2, color = "gray50") +
+  annotate(shadowtext:::GeomShadowText,
+           label = c("- Real", "- Imaginary", "+ Real", "+ Imaginary", "+ Real"),
+           color = "gray60", bg.colour = "white", bg.r = 0.2,
+           size = 3,
+           y = -Inf, x = c(-pi, -pi/2, 0, pi/2, pi), angle = 90, hjust = 0) +
+
+  geom_smooth(data = ~qwrap(.x, arg = c(-pi, pi) ~ c(-2*pi, 2*pi)),
+              method = "lm", aes(weight = mag),
+              formula = y ~ I(sin(x)) + I(cos(x)) -1,
+              se = FALSE, color = "black", size = 0.5) +
+    geom_point(aes(colour = mag >= quantile(mag, 0.5),
+                 shape  = mag >= quantile(mag, 0.5)),
+             size = 3) +
+
+  scale_color_brewer(NULL, palette = "Dark2",
+                     labels = c("TRUE"  = "|EOF2| > 50%",
+                                "FALSE" = "|EOF2| < 50%")) +
+  scale_shape_manual(NULL, values = c("TRUE" = 20,
+                                      "FALSE" = 18),
+                     labels = c("TRUE"  = "|EOF2| > 50%",
+                                "FALSE" = "|EOF2| < 50%")) +
+  scale_y_continuous("Precipitation anomaly") +
+  scale_x_continuous("Phase",
+                     breaks = seq(-pi, pi, by = 30*pi/180),
+                     labels = function(x) paste0(round(x*180/pi, 5), "º")) +
+  coord_cartesian(xlim = c(-pi, pi)) +
+  panel_background
+```
+
+To better understand the relationship between precipitation anomalies over Western Australia and the EOF2, Figure \@ref(fig:australia-pp-phase) plots precipitation anomalies averaged over Eastern Australia (box in Figure \@ref(fig:pp-oceania)) as a function of EOF2 phase.
+The black line is the equation $\mathrm{Precipitation} = `r signif(coef(model)[1], 2)`\sin{(\mathrm{phase})} + `r signif(coef(model)[2], 2)`\cos{(\mathrm{phase})}$, whose coefficient we fitted by weighted least squares using the magnitude of the EFO2 as weight.
+The $r^2$ corresponding to the fit is `r signif(summary(model)$r.squared, 2)`, with p-value `r report_p(p_value)`.
+There is a strong relationship between the phase of the EOF2 and precipitation anomalies in Eastern Australia, but Real and Imaginary phases we chose are not aligned with the direction that maximises this relationship.
+If one were to perform a more in-depth analysis of the relationship between this pattern and precipitation in this region, one would conceivably align one of the axis (either the Real or the Imaginary) to the one that maximises it.
+
 (ref:pp-africa-cap) Same as Figure \@ref(fig:pp-america) but for South of Africa.
@@
 ```{r pp-africa, fig.cap = "(ref:pp-africa-cap)"}
-plot_pp(regr_pp, c("PC1", "PC2"), which_continent = "africa")
-```
-
-(ref:pp-oceania-cap) Same as Figure \@ref(fig:pp-america) but for New Zealand and neighbouring islands.
-
-```{r pp-oceania, fig.cap = "(ref:pp-oceania-cap)"}
-plot_pp(regr_pp, c("PC1", "PC2"), which_continent =  "oceania")
+# plot_pp(regr_pp, c("PC1", "PC2"), which_continent = "africa")
 ```
@@
 
-# Appendix
-
-## Chosen rotations of the EOFs
-
-```{r rotations-pc1, fig.cap = "Rotations of PC1"}
-rotations_o3 %>%
-  ggplot(aes(rotation, cor)) +
-  geom_vline(xintercept = best_rotation_pc1) +
-  geom_hline(yintercept = 0,  size = 0.2, color = "gray50") +
-  geom_line(aes(color = part), size = 1) +
-  scale_x_continuous("Rotation", breaks = seq(-pi, pi, by = 30*pi/180),
-                     labels = function(x) paste0(round(x*180/pi, 5), "º")) +
-  scale_y_continuous("Correlation with O3 zonal wave 1") +
-  scale_color_brewer(NULL, palette = "Dark2", labels = lab_cplx)
-```
-
-```{r rotations-pc2, fig.cap = "R2 between EOF2 Real and Imaginary parts and ONI index for different rotation parameters of the real and imaginary parts."}
-rotations_pc2 %>%
-  melt(id.vars = "rotation") %>%
-  ggplot(aes(rotation, value^2)) +
-  geom_hline(yintercept = 0,  size = 0.2, color = "gray50") +
-  geom_vline(xintercept = best_rotation_pc2) +
-  geom_line(aes(color = variable), size = 1) +
-  scale_x_continuous("Rotation", breaks = seq(-pi, pi, by = 30*pi/180),
-                     labels = function(x) paste0(round(x*180/pi, 5), "º")) +
-  scale_y_continuous("r^2") +
-  scale_color_brewer(NULL, palette = "Dark2", labels = lab_cplx)
-```
-
 ## Code availability {-}

### 2021-03-04 analysis/paper/paper.tex
Draft resultados

@@
 
-We compute the amplitude and phase of the zonal wave 1 by averaging SON TOC between 70°S and 45°S for each spring and computing the Fourier spectrum.
+We compute the amplitude and phase of the zonal wave 1 by averaging SON TOC between 70°S and 45°S for each Spring and computing the Fourier spectrum.
 
@@
 \includegraphics{../figures/eof-naive-1.pdf}
-\caption{\label{fig:eof-naive}Spatial patterns of the leading 4 EOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2009.
-(arbitrary units).
-The numbers at the top-left of each panel is the variance explained by each EOF.}
+\caption{\label{fig:eof-naive}Spatial patterns of the leading 4 EOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2009 (arbitrary units).
+The numbers at the top-left of each panel ate the variance explained by each EOF.}
 \end{figure}
@@
 The zonal wave 3 structure in EOF2 and EOF3 at 200 hPa are also very similar to the one in EOF3 and EOF4 at 50 hPa.
-Furthermore, Table \ref{tab:corr-ceof-splitted} shows the coefficient of determination between temporal series of the amplitude of each complex EOF across levels.
+Furthermore, Table \ref{tab:corr-ceof-splitted} shows the coefficient of determination between time series of the amplitude of each complex EOF across levels.
 There's a high degree of correlation between the respective EOF1 and EOF2 at each level.
@@
 
+\begin{figure}
+\centering
+\includegraphics{../figures/rotations-1.pdf}
+\caption{\label{fig:rotations}a) Correlation of the Real and Imaginary EOF1 with wave 1 amplitude of Total Ozone Column averaged between 70°S and 45°S and b) \(R^2\) between the Real and Imaginary EOF2 and the ONI index for different values of rotation.
+The vertical line marks the selected rotation, which maximises de correlation.}
+\end{figure}
+
 For the first complex principal component, the phase was chosen so that the time series corresponding to the real part has the maximum correlation with the zonal wave 1 of Total Ozone Column between 70°S and 45°S.
-This also nearly minimises the correlation with the imaginary part (\ref{fig:rotations-pc1}).
+This also nearly minimises the correlation with the imaginary part (Figure \ref{fig:rotations}a).
 
-For the second complex principal component, the phase was chosen so that the coefficient of determination between the Oceanic Niño Index \citep{bamston1997} and the real part was minimised, which also nearly maximises the correlation with the imaginary part (\ref{fig:rotations-pc2}).
+For the second complex principal component, the phase was chosen so that the coefficient of determination between the Oceanic Niño Index \citep{bamston1997} and the real part was minimised, which also nearly maximises the correlation with the imaginary part (Figure \ref{fig:rotations}b).
 It's important to reassure the reader that this procedure does not create a spurious correlation, it only takes whatever relationship that already exist and aligns it with a specific phase.
@@
 \includegraphics{../figures/ceofs-1-1.pdf}
-\caption{\label{fig:ceofs-1}Leading 2 EOFs of zonal anomalies of geopotential height in 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2009.
-Spatial patterns.
+\caption{\label{fig:ceofs-1}Spatial patterns of the leading 2 complex EOFs of zonal anomalies of geopotential height in 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2009.
 Real part in shading, imaginary part in contours}
@@
 \includegraphics{../figures/ceofs-2-1.pdf}
-\caption{\label{fig:ceofs-2}Temproal series.}
+\caption{\label{fig:ceofs-2}Time series of the leading 2 complex EOFs of zonal anomalies of geopotential height in 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2009.}
 \end{figure}
@@
 \includegraphics{../figures/extended-series-1.pdf}
-\caption{\label{fig:extended-series}Temporal series extended using ERA5 back extended preliminary edition (period 1950 -- 1978) and ERA5 (period 1979 -- 2019).
+\caption{\label{fig:extended-series}Time series extended using ERA5 back extended preliminary edition (period 1950 -- 1978) and ERA5 (period 1979 -- 2019).
 Each series is computed by projecting monthly geopotential height zonal anomalies standardised by level south of 20ºS onto the corresponding spatial pattern and scaled to zero mean and unit standard deviation.}
@@
 
-Figure \ref{fig:extended-series} shows temporal series of the two complex EOFs extended beyond the satellite era using the preliminary ERA5 back extension going back to 1950 (which we call the ``hybrid ERA5'' reanalysis).
-There is a downward trend in the real part of EOF1 in the hybrid reanalysis (Figure \ref{fig:extended-series}a.1, p-value = 0.0011).
+Figure \ref{fig:extended-series} shows time series of the two complex EOFs extended beyond the satellite era using the preliminary ERA5 back extension going back to 1950 (which we call the ``hybrid ERA5'' reanalysis).
+There is a downward trend in the real part of EOF1 in the hybrid reanalysis (Figure \ref{fig:extended-series}a.1, p-value \textless{} 0.001).
 There is no significant trend in any of the complex parts of EOF2.
@@
 \includegraphics{../figures/t-vertical-1.pdf}
-\caption{\label{fig:t-vertical}Regression of EOF1 with mean air temperature between 80ºS and 25ºS.}
+\caption{\label{fig:t-vertical}Regression of EOF1 with mean air temperature averaged between 80ºS and 25ºS for the period 1979 -- 2019.}
 \end{figure}
@@
 
+(raf:o3-regr-cap) Regression coefficients of the real and imaginary part of the first complex EOF on SON mean Total Ozone Column for the 1979 -- 2019 period.
+
 \begin{figure}
@@
 \includegraphics{../figures/o3-regr-1.pdf}
-\caption{\label{fig:o3-regr}Regression of standarised EOF1 and Total Ozone column.}
+\caption{\label{fig:o3-regr}(raf:o3-regr-cap)}
 \end{figure}
@@
 \includegraphics{../figures/wave1-o3-1.pdf}
-\caption{\label{fig:wave1-o3}EOF1 and wave1 o3 amplitude.}
+\caption{\label{fig:wave1-o3}Relationship between EOF1 amplitude and phase with amplitude and phase of the zonal wave 1 in Total Ozone Column averaged between 70°S and 45°S for the period 1979 -- 2019.}
 \end{figure}
@@
 Indeed, Figure \ref{fig:o3-regr} shows the relationship between amplitude and phase of the planetary wave 1 in Total Ozone Column between 70ºS and 45ºS and amplitude and phase of the EOF1.
-The correlation between the zonal wave 1 of ozone and EOF1 (computed as the mean cosine of the difference in phase) is 0.87 (CI: 0.76 -- 0.93).
+The correlation between the zonal wave 1 of ozone and EOF1 (computed as the mean cosine of the difference in phase) is 0.86 (CI: 0.76 -- 0.93).
 As expected from the location of anomalies in Figure @ref(fig: o3-regr), the Real EOF1 drives the relationship with amplitude and the Imaginary EOF1 drives the relationship with phase (not shown).
@@
 \includegraphics{../figures/psa-eof2-1.pdf}
-\caption{\label{fig:psa-eof2}EFO2 and PSA}
+\caption{\label{fig:psa-eof2}Relationship between the Real and Imaginary parts of EOF2 and the PSA1 and PSA2 modes computed as the second and third EOFs of seasonal geopotential height anomalies \citep{mo2001}.
+For the period 1979 -- 2019.}
 \end{figure}
@@
 \includegraphics{../figures/phase-histogram-1.pdf}
-\caption{\label{fig:phase-histogram}Histogram of phase distributions of EOF2.}
+\caption{\label{fig:phase-histogram}Histogram showing phase distribution of EOF2 for the period 1979 -- 2019.}
 \end{figure}
@@
 \includegraphics{../figures/fake-eof-1.pdf}
-\caption{\label{fig:fake-eof}Principal compomnents of a sinthetic data set of geopotential height as the sum of the fields predicted by the EOF2 plus perfectly zonally symmetric SAM-like annomalies with random amplitude.}
+\caption{\label{fig:fake-eof}EOFs of a synthetic data set of 41 times steps of geopotential height fields as the sum of the fields predicted by the EOF2 plus perfectly zonally symmetric SAM-like anomalies with random amplitude.}
 \end{figure}
@@
 The black line in Figure \ref{fig:enso-phase} tries to quantify this relationship.
-Is the equation \(ONI = 1.1\sin{phase}\), whose coefficient we fitted by weighted least squares using the magnitude of the EFO2 as weight.
+Is the equation \(ONI = 1.1\sin{(phase)}\), whose coefficient we fitted by weighted least squares using the magnitude of the EFO2 as weight.
 The \(r^2\) corresponding to the fit is 0.56, with p-value \textless{} 0.001.
@@
 If those years are removed, the correlation becomes non-significant (0.044 (CI: -0.28 -- 0.36)).
-Furthermore, even when using all the datapoints, the Spearman correlation --which is robust to outliers-- is also non-significant (0.2, p-value = = 0.2).
+Furthermore, even when using all the datapoints, the Spearman correlation --which is robust to outliers-- is also non-significant (0.2, p-value = 0.2).
 Therefore, the relationship between the magnitude of the EOF2 train wave and ENSO remains uncertain.
@@
 Figure \ref{fig:pp-america} shows regression maps of seasonal precipitation with each EOF in South America.
-EOF1 (Figure \ref{fig:pp-america} column a) is associated with increased precipitation over Paraguay and and some parts of Brazil.
-Although these positive anomalies are not very strong.
+EOF1 is associated with slight decreases in precipitation in Southern Brazil and Paraguay, although these are not statistically significant \ref{fig:pp-america} column a.
 The strongest precipitation anomalies are the ones associated with the Imaginary EOF2.
 The positive anomalies on Southeaster South America (SESA) and Chile, and negative anomalies over Southern Brazil is a well known springtime precipitation signature of ENSO \citep{cai2020a} and it's also virtually identical to the precipitation anomalies associated with the Asymmetric SAM \citep{campitelli2021}.
-This is not surprising considering the close relationship between the ONI, the Asymmetric SAM index and the Imaginary EOF2 shown previously.
+This is not surprising considering the close relationship between the ONI, the Asymmetric SAM index and the Imaginary EOF2 shown previously, but further consolidates the identification of this mode with the PSA pattern.
 The Real EOF2, on the other hand, is associated with negative precipitation anomalies in a smaller area of SESA.
 The relationship between precipitation anomalies in SESA and the phase of EOF2 follows a curve similar to that of Figure \ref{fig:enso-phase} (not shown).
-
-\begin{figure}
-\centering
-\includegraphics{../figures/pp-africa-1.pdf}
-\caption{\label{fig:pp-africa}Same as Figure \ref{fig:pp-america} but for South of Africa.}
-\end{figure}
 
@@
 
-\hypertarget{conclusions}{%
-\section{Conclusions}\label{conclusions}}
-
-\hypertarget{appendix}{%
-\section{Appendix}\label{appendix}}
-
-\hypertarget{chosen-rotations-of-the-eofs}{%
-\subsection{Chosen rotations of the EOFs}\label{chosen-rotations-of-the-eofs}}
+On the antipodes, precipitation anomalies in New Zealand and neighbouring islands associated with EOF1 are also weak and not statistically significant (Figure \ref{fig:pp-oceania} column a) whereas EFO2 is associated with large and expansive anomalies.
+The Real EOF2 is associated with positive seasonal anomalies in Northern and Eastern Australia, and negative ones over New Zealand, while precipitation anomalies associated with Imaginary EOF2 are negative over all Eastern Australia.
+Again, these negative anomalies are similar to the springtime precipitation anomalies observed in relation with the Asymmetric SAM \citep{campitelli2021}.
 
@@
 \centering
-\includegraphics{../figures/rotations-pc1-1.pdf}
-\caption{\label{fig:rotations-pc1}Rotations of PC1}
+\includegraphics{../figures/australia-pp-phase-1.pdf}
+\caption{\label{fig:australia-pp-phase}Same as Figure \ref{fig:enso-phase} but for precipitacion anomalies averaged between 130°E and 155°E and 40°S and 15°S (box shown in Figure \ref{fig:pp-oceania} column b).}
 \end{figure}
 
-\begin{figure}
-\centering
-\includegraphics{../figures/rotations-pc2-1.pdf}
-\caption{\label{fig:rotations-pc2}R2 between EOF2 Real and Imaginary parts and ONI index for different rotation parameters of the real and imaginary parts.}
-\end{figure}
+To better understand the relationship between precipitation anomalies over Western Australia and the EOF2, Figure \ref{fig:australia-pp-phase} plots precipitation anomalies averaged over Eastern Australia (box in Figure \ref{fig:pp-oceania}) as a function of EOF2 phase.
+The black line is the equation \(\mathrm{Precipitation} = -0.26\sin{(\mathrm{phase})} + 0.22\cos{(\mathrm{phase})}\), whose coefficient we fitted by weighted least squares using the magnitude of the EFO2 as weight.
+The \(r^2\) corresponding to the fit is 0.37, with p-value \textless{} 0.001.
+There is a strong relationship between the phase of the EOF2 and precipitation anomalies in Eastern Australia, but Real and Imaginary phases we chose are not aligned with the direction that maximises this relationship.
+If one were to perform a more in-depth analysis of the relationship between this pattern and precipitation in this region, one would conceivably align one of the axis (either the Real or the Imaginary) to the one that maximises it.
+
+\hypertarget{conclusions}{%
+\section{Conclusions}\label{conclusions}}
 

### 2021-03-04 R/correlate.R

@@ -16,8 +16,8 @@ correlate <- function(x, y, signif = 2) {
                    hig = conf.int[2]))
 
   out$text <- with(out,
-                   paste0(signif(estimate, signif), " (CI: ",
-                          signif(low, signif), " -- ", signif(hig, signif), ")"))
+                   paste0(signif(estimate, signif), " (CI:\ ",
+                          signif(low, signif), "\ --\ ", signif(hig, signif), ")"))
 
   out
 }

### 2021-03-04 R/lm_lite.R

@@ -75,7 +75,7 @@ Pvaluate <- function(estimate, std.error, df, adjustment = "none") {
 #'
 #' @export
 report_p <- function(p) {
-  ifelse(p < 0.001, " < 0.001", paste0(" = ", signif(p, 2)))
+  ifelse(p < 0.001, "\ <\ 0.001", paste0("\ =\ ", signif(p, 2)))
 }
 
 #' Remove intercept

### 2021-03-04 analysis/paper/zome_ozone.R

@@ -1,0 +1,264 @@
+# https://cliflo.niwa.co.nz/pls/niwp/wgenf.genform1
+
+
+file <- data_path("raw", "wgenf.genform1_proc.txt")
+
+
+stations <- fread(file, skip = 1, nrows = 17, na.strings = "N/A",
+                  col.names = c("name", "agent_number", "network_number", "lat", "lon", "height", "pos", "auth")) %>%
+  .[, lon := ConvertLongitude(lon)]
+
+variable_codes <-  fread(file, skip = 24, nrow = 4)
+
+
+data <- fread(file, skip = 34, na.strings = "-", nrow = 1251 - 36) %>%
+  .[!(Stats_Code %in%  c("83", "83_date"))] %>%
+  .[, -"Annual"] %>%
+  melt(id.vars = c("Station", "Stats_Code", "Year")) %>%
+  .[, month := setNames(1:12, month.abb)[variable]] %>%
+  .[, time := lubridate::make_datetime(Year, month)] %>%
+  .[, .(agent_number = Station, code = Stats_Code, time, value)]
+
+
+uv2 <- fread(data_path("raw", "NILU-UV_2000-01-01_2010-12-31_020321203210.txt"), skip = 2,
+             col.names = c("time", "uv_marambio", "uvmax_marambio", "uv_ushuaia", "uvmax_ushuaia")) %>%
+  melt(id.vars = "time") %>%
+  tidyr::separate(variable, c("variable", "station")) %>%
+  tidyr::complete(variable, station, time = seq(min(time), max(time), "1 day")) %>%
+  as.data.table() %>%
+  .[!is.finite(value), value := NA]
+
+
+
+
+
+
+data[code == "84"] %>%
+  ggplot(aes(time, value)) +
+  geom_line() +
+  facet_wrap(~agent_number)
+
+
+uv <- data[code == "84"] %>%
+  tidyr::complete(agent_number, code, time = seq(min(time), max(time), "1 month")) %>%
+  as.data.table() %>%
+  .[, .(value = mean(value, na.rm = TRUE),
+        n = sum(!is.na(value))), by = .(agent_number, code, time = seasonally(time))] %>%
+  .[n > 2] %>%
+  .[season(time) == "SON"] %>%
+  .[, n := .N, by = .(code, agent_number)] %>%
+  .[n > 20]
+
[…]

### 2021-03-07 analysis/paper/paper.Rmd
Agrega regresión hgt vertical

@@
 plot_regr(regr, "PC1", c(200, 50), "t")
-
 ```
@@
 t <- ERA5() %>% 
-  ReadNetCDF(vars = "t", 
+  ReadNetCDF(vars = c("t", hgt = "z"), 
              subset = list(latitude = c(-90, -20))) %>% 
@@
   .[season(time) == "SON"] %>%
-  .[, .(t = mean(t)), by = .(lev, lon, lat, time = seasonally(time))]
+  .[, .(t = mean(t),
+        hgt = mean(hgt)), by = .(lev, lon, lat, time = seasonally(time))]
 ```
@@
   .[lat %between% c(-80, -35)] %>% 
-  .[, .(t = weighted.mean(t, cos(lat*pi/180))), by = .(lev, time, lon)] %>%  
-  .[, t := Anomaly(t), by = .(time, lev)] %>% 
+  .[, .(hgt = weighted.mean(hgt, cos(lat*pi/180)),
+        t = weighted.mean(t, cos(lat*pi/180))), by = .(lev, time, lon)] %>%  
+  melt(id.vars = c("lev", "time", "lon")) %>% 
+  .[, value := Anomaly(value), by = .(time, lev, variable)] %>% 
   .[cut(ceof$eof[[1]], 1)$left, on = "time"] %>% 
@@
   .[, ":="(Real = as.numeric(scale(Real)), Imaginary = as.numeric(scale(Imaginary))), by = .(PC)] %>% 
-  .[, FitLm(t, Real, Imaginary, se = TRUE), by = .(lev, lon)] %>% 
+  .[, FitLm(value, Real, Imaginary, se = TRUE), by = .(lev, lon, variable)] %>% 
   rm_intercept() %>% 
   .[, term := factor_ReIm(term)] %>% 
-  .[, p.value := Pvaluate(estimate, std.error, df, "fdr"), by = .(term)] %>%
+  # .[, z := metR:::standard_atmosphere$altitude(lev*100)] %>% 
+  # .[, dz1 := c(NA, diff(estimate)/diff(z)) , by = .(term, lon, variable)] %>% 
+  # .[, dz2 := c(diff(estimate)/diff(z), NA) , by = .(term, lon, variable)] %>% 
+  # .[, dz := (dz1 + dz2)/2] %>% 
+  # .[, p.value := Pvaluate(estimate, std.error, df, "fdr"), by = .(term, variable)] %>%
+  # .[lev < 900] %>% 
   ggplot(aes(lon, lev)) +
-  geom_contour_fill(aes(z = estimate, fill = ..level..),
-                    breaks = AnchorBreaks(0, 0.25, exclude = 0)) +
-  geom_contour_tanaka(aes(z = estimate), 
-                      breaks = AnchorBreaks(0, 0.25, exclude = 0)) +
-  geom_contour_pval(aes(z = p.value)) +
+  geom_contour_fill(data = ~.x[variable == "t"],
+                    aes(z = estimate, fill = ..level..),
+                    breaks = AnchorBreaks(0, binwidth = 0.25, exclude = 0)) +
+  geom_contour_tanaka(data = ~.x[variable == "t"],
+                      aes(z = estimate),
+                      breaks = AnchorBreaks(0, binwidth = 0.25, exclude = 0)) +
+  geom_contour2(data = ~.x[variable == "hgt"], 
+                aes(z = estimate, linetype = factor(-sign(..level..))), 
+                breaks = AnchorBreaks(0, exclude = 0)) +
+  geom_text_contour(data = ~.x[variable == "hgt"], 
+                aes(z = estimate), skip = 1, stroke = 0.15, stroke.colour = "white",
+                breaks = AnchorBreaks(0, exclude = 0)) +
+
   scale_fill_divergent_discretised("Air temperature", guide = guide_colorsteps_bottom(20)) + 
   scale_y_level() +
+  scale_linetype(guide = "none") +
   scale_x_longitude(ticks = 60) +
   facet_grid(~term) +
-  axis_labs_smol
+  axis_labs_smol +
+  coord_cartesian(clip = "off")
 ```
@@
 In both levels, the distribution of temperature anomalies mirror the geopotential height anomalies in 50 hPa.
-To understand the vertical distribution of these anomalies, Figure \@ref(fig:t-vertical)) shows the regression pattern of EOF1 and zonal anomalies of air temperature averaged between 80ºS and 35ºS.
-Both pars of EOF1 are associated with a temperature wave 1 pattern which changes phase around 10 hPa.
-This switch in the sign of the correlation in temperature at 10 hPa suggest a relationship with Ozone, since its indicative of the effects of planetary waves in the advection of Ozone [@hartmann1979;@rood1985;@smith1995].
+To understand the vertical distribution of these anomalies, Figure \@ref(fig:t-vertical)) shows the regression pattern of EOF1 with zonal anomalies of air temperature and zonal anomalies of geopotential height averaged between 80ºS and 35ºS.
+The zonal wave 1 in geopotential height reach its maximum at around 10 hPa with some westerly tilt.
+Temperature anomalies are positive below geopotential height maximum, negative above, and vice versa for geopotential height minimums.
+This reflects the thermodynamic relationship between temperature and layer width.
 
@@
 
-To confirm this possibility, Figure \@ref(fig:o3-regr) shows regression maps of EOF1 with fields of Total Ozone Column (TOC).
+Planetary waves that penetrate into the stratosphere are known to affect ozone distribution [@hartmann1979].
+Figure \@ref(fig:o3-regr) shows regression maps of EOF1 with fields of Total Ozone Column (TOC).
 It shows zonal wave 1 pattens in TOC associated with both phases of EOF1.
@@
 Thus, the Real EOF1 regression pattern (Figure \@ref(fig:o3-regr)a) coincides with the climatological position of the ozone whole while the one for the Imaginary EOF1 is shifted by 90º.
+THe maximum TOC anomalies are co-located with the minimum temperature anomalies above 10 hPa and mix the maximum temperature anomalies below 10 hPa; that is, the TOC zonal wave 1 is anticorrelated with the temperature zonal wave 1 in the upper stratosphere, and directly correlated in the upper stratosphere.
+These phase differences are due to the change between the photochemically-dominated upper stratosphere, where cold temperatures inhibit the destruction of ozone, and the advectively-dominated lower stratosphere [@hartmann1979;@smith1995].
+
+This is explained by the inverse temperature dependence of the photochemical desctruction of ozone.
+
 This suggest that the EOF1 can be a complete description of the TOC zonal wave 1.
+
+Temperature anomalies in
+
+Both pars of EOF1 are associated with a temperature wave 1 pattern which changes phase around 10 hPa.
+This switch in the sign of the correlation in temperature at 10 hPa suggest a relationship with Ozone, since its indicative of the effects of planetary waves in the advection of Ozone [@hartmann1979;@rood1985;@smith1995].
+
+To confirm this possibility,
 

### 2021-03-08 analysis/paper/paper.Rmd
O3 mixing ratio

--- 2021-03-04 analysis/paper/paper.Rmd
@@
 
-geom_contour_pval <- function(mapping, p.value = 0.01, size = 0.1) {
+geom_contour_pval <- function(mapping, p.value = 0.01, size = 0.1, ...) {
   mapping2 <- mapping
@@
   list(
-    stat_contour_filled(mapping2, breaks = c(0, p.value), fill = NA, geom = pattern_dots),
-    geom_contour2(mapping, breaks = p.value, size = size)
+    stat_contour_filled(mapping2, breaks = c(0, p.value), fill = NA, geom = pattern_dots, ...),
+    geom_contour2(mapping, breaks = p.value, size = size, ...)
   )
@@
 
-We used monthly geopotential height, air temperature and Total Ozone COlumn (TOC) at 2.5\degree\ longitude by 2.5\degree\ latitude of horizontal resolution and 37 vertical isobaric levels from ERA5 [@era5] for the period 1979 to 2019.
+We used monthly geopotential height, air temperature, ozone mixing ratio, and Total Ozone COlumn (TOC) at 2.5\degree\ longitude by 2.5\degree\ latitude of horizontal resolution and 37 vertical isobaric levels from ERA5 [@era5] for the period 1979 to 2019.
 We restrict most of our analysis to the post-satellite era to avoid any confounding factors arising from the incorporation of satellite observations, but we also use the preliminary back extension of ERA5 from 1950 to 1978 [@era5be] to look at long-term trends.
@@
 In traditional EOF analysis zonal waves appear as pairs of EOFs, usually degenerate, that represent similar patterns but shifted in phase.
-For instance, Figure \@ref(fig:eof-naive) shows the leading 4 EOFs of zonally anomalous September through October geopotential height at 50 hPa and 100 hPa.
+For instance, Figure\ \@ref(fig:eof-naive) shows the leading 4 EOFs of zonally anomalous September through October geopotential height at 50 hPa and 100 hPa.
 It is clear that at 50 hPa the first two EFS represent the same zonal wave 1 pattern and the last 2 represent the same zonal wave 3 pattern shifted by 1/4 wavelength.
@@
 
-Another characteristic seen in Figure \@ref(fig:eof-naive) is that EOF1 at 200 hPa is very similar to EFO1 at 50 hPa.
+Another characteristic seen in Figure\ \@ref(fig:eof-naive) is that EOF1 at 200 hPa is very similar to EFO1 at 50 hPa.
 The zonal wave 3 structure in EOF2 and EOF3 at 200 hPa are also very similar to the one in EOF3 and EOF4 at 50 hPa.
@@
   geom_line(aes(color = part)) +
-  scale_x_continuous("Rotation", breaks = seq(-pi, pi, by = 60*pi/180),
+  scale_x_continuous("Rotation", breaks = seq(-pi, pi, by = 90*pi/180),
                      labels = function(x) paste0(round(x*180/pi, 5), "º")) +
@@
   geom_line(aes(color = variable)) +
-  scale_x_continuous("Rotation", breaks = seq(-pi, pi, by = 60*pi/180),
+  scale_x_continuous("Rotation", breaks = seq(-pi, pi, by = 90*pi/180),
                      labels = function(x) paste0(round(x*180/pi, 5), "º")) +
@@
 For the first complex principal component, the phase was chosen so that the time series corresponding to the real part has the maximum correlation with the zonal wave 1 of Total Ozone Column between `r knitr::combine_words(LatLabel(o3wave_lats))`.
-This also nearly minimises the correlation with the imaginary part (Figure \@ref(fig:rotations)a).
-
-For the second complex principal component, the phase was chosen so that the coefficient of determination between the Oceanic Niño Index [@bamston1997] and the real part was minimised, which also nearly maximises the correlation with the imaginary part (Figure \@ref(fig:rotations)b).
+This also nearly minimises the correlation with the imaginary part (Figure\ \@ref(fig:rotations)a).
+
+For the second complex principal component, the phase was chosen so that the coefficient of determination between the Oceanic Niño Index [@bamston1997] and the real part was minimised, which also nearly maximises the correlation with the imaginary part (Figure\ \@ref(fig:rotations)b).
 It's important to reassure the reader that this procedure does not create a spurious correlation, it only takes whatever relationship that already exist and aligns it with a specific phase.
@@
 Figures \@ref(fig:ceofs-1) and \@ref(fig:ceofs-2) show, respectively, the spatial and temporal parts of the first two leading complex EOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa.
-In the spatial patterns in Figure \@ref(fig:ceofs-1), the real (in shading) and the imaginary parts (in contour) are in quadrature by construction, so that each EOF describe a single wave-like pattern whose amplitude and position (i.e.
+In the spatial patterns in Figure\ \@ref(fig:ceofs-1), the real (in shading) and the imaginary parts (in contour) are in quadrature by construction, so that each EOF describe a single wave-like pattern whose amplitude and position (i.e.
 phase) is controlled by the magnitude and phase of the complex temporal EOF.
 
-The wave patterns described by these complex EOFs match the patterns seen in the traditional EOFs of Figure \@ref(fig:eof-naive): The first is a wave 1, while the second is a wave 3.
-Note that in Figure \@ref(fig:ceofs-2), both parts of the EOF1 have non zero mean.
+The wave patterns described by these complex EOFs match the patterns seen in the traditional EOFs of Figure\ \@ref(fig:eof-naive): The first is a wave 1, while the second is a wave 3.
+Note that in Figure\ \@ref(fig:ceofs-2), both parts of the EOF1 have non zero mean.
 This is due to th fact that the geopotential fields that enter into the algorithm are anomalies with respect to the zonal mean, not the time mean.
@@
 
-Figure \@ref(fig:extended-series) shows time series of the two complex EOFs extended beyond the satellite era using the preliminary ERA5 back extension going back to 1950 (which we call the "hybrid ERA5" reanalysis).
-There is a downward trend in the real part of EOF1 in the hybrid reanalysis (Figure \@ref(fig:extended-series)a.1, p-value `r trends[part == "Real" & PC == "PC1"]$p_report`).
+Figure\ \@ref(fig:extended-series) shows time series of the two complex EOFs extended beyond the satellite era using the preliminary ERA5 back extension going back to 1950 (which we call the "hybrid ERA5" reanalysis).
+There is a downward trend in the real part of EOF1 in the hybrid reanalysis (Figure\ \@ref(fig:extended-series)a.1, p-value `r trends[part == "Real" & PC == "PC1"]$p_report`).
 There is no significant trend in any of the complex parts of EOF2.
@@
 
-The spatial patterns shown in Figure \@ref(fig:ceofs-1) are "pure" EOF patterns.
+The spatial patterns shown in Figure\ \@ref(fig:ceofs-1) are "pure" EOF patterns.
 But as they are derived by removing the zonally symmetric circulation, they might not include all the variability that is actually associated with the EOF time series.
@@
 
-Figure \@ref(fig:eof1-regr-gh) shows regression patterns of EOF1 and geopotential height.
-At 50 hPa (Figure \@ref(fig:eof1-regr-gh) row a), both the Real and Imaginary EOF1 are associated with planetary wave 1 patterns, that are 90º off phase.
-Their phases coincide with the ones shown in Figure \@ref(fig:ceofs-1)a.1, with the positive centre of the Real EOF1 located towards the dateline, and the one of the Imaginary EOF1 located over Eastern Antarctica.
+Figure\ \@ref(fig:eof1-regr-gh) shows regression patterns of EOF1 and geopotential height.
+At 50 hPa (Figure\ \@ref(fig:eof1-regr-gh) row a), both the Real and Imaginary EOF1 are associated with planetary wave 1 patterns, that are 90º off phase.
+Their phases coincide with the ones shown in Figure\ \@ref(fig:ceofs-1)a.1, with the positive centre of the Real EOF1 located towards the dateline, and the one of the Imaginary EOF1 located over Eastern Antarctica.
 However, the Real EOF1 pattern is substantially altered by the zonally symmetric circulation.
 Instead of a clear wave 1 pattern, the regression pattern can be better describes as a monopole with its centre displaced from the South Pole.
-Similarly affected are the regression patterns at 200 hPa (Figure \@ref(fig:eof1-regr-gh) row b).
-It is only possible to distinguish some of the wave 1 pattern in relation with the Real EOF1 (Figure \@ref(fig:eof1-regr-gh)b.1).
+
+Similarly affected are the regression patterns at 200 hPa (Figure\ \@ref(fig:eof1-regr-gh) row b).
+It is only possible to distinguish some of the wave 1 pattern in relation with the Real EOF1 (Figure\ \@ref(fig:eof1-regr-gh)b.1).
 The Imaginary EOF1 shows a much more zonally symmetrical pattern resembling a negative Southern Annular Mode.
 
-With the exception of the Imaginary EFO1, it's clear that these patterns are very different than the idealised versions shown in Figure \@ref(fig:ceofs-1), particularly at 200 hPa.
+With the exception of the Imaginary EFO1, it's clear that these patterns are very different than the idealised versions shown in Figure\ \@ref(fig:ceofs-1), particularly at 200 hPa.
 Moreover, only in the stratosphere these patterns actually show a distinguishable wave 1 pattern shifted in phase by 90º.
@@
 
-(ref:eof2-regr-gh-cap) Same as Figure \@ref(fig:eof1-regr-gh) but for the second EOF.
+(ref:eof2-regr-gh-cap) Same as Figure\ \@ref(fig:eof1-regr-gh) but for the second EOF.
 
@@
 
-Figure \@ref(fig: eof2-regr-gh) shows the regression pattern of geopotential height and the EOF2.
-Unlike in the case of EOF1, in this case the regression patterns represent relatively well the *idealised* patterns from Figure \@ref(fig:ceofs-1).
+Figure\ \@ref(fig:eof2-regr-gh) shows the regression pattern of geopotential height and the EOF2.
+Unlike in the case of EOF1, in this case the regression patterns represent relatively well the *idealised* patterns from Figure\ \@ref(fig:ceofs-1).
 Although there are some differences (particularly in 50 hPa), the wave trains identified before are well characterised and patterns associated with the Real EOF2 are 90º out of phase with those associated with the Imaginary EOF2.
@@
 EOF2 then represents a equivalent barotropic wave train that is very similar to the the Pacific South American Patterns [@mo2001].
-Comparing the location of the positive anomaly near 90ºW in column b of Figure \@ref(fig: eof2-regr-gh) with Figure 1 a and b from @mo2001, the Real EOF2 can roughly be identified with the PSA2, while the Imaginary EOF2 resembles PSA1.
+Comparing the location of the positive anomaly near 90ºW in column b of Figure\ \@ref(fig:eof2-regr-gh) with Figure 1 a and b from @mo2001, the Real EOF2 can roughly be identified with the PSA2, while the Imaginary EOF2 resembles PSA1.
 This separation is not completely meaningful in this case, since by treating this pattern as a single wave train any one pair of orthogonal waves can characterise this variability.
@@
 
-(ref:eof1-regr-t-cap) Same as Figure \@ref(fig:eof1-regr-gh) but for air temperature.
+(ref:eof1-regr-t-cap) Same as Figure\ \@ref(fig:eof1-regr-gh) but for air temperature.
 
@@
 t <- ERA5() %>% 
-  ReadNetCDF(vars = "t", 
+  ReadNetCDF(vars = c("t", hgt = "z", "o3"), 
              subset = list(latitude = c(-90, -20))) %>% 
@@
   .[season(time) == "SON"] %>%
-  .[, .(t = mean(t)), by = .(lev, lon, lat, time = seasonally(time))]
-```
-
-(ref:t-vertical-cap) Regression of EOF1 with mean air temperature averaged between 80ºS and 25ºS for the period 1979 -- 2019.
+  .[, .(t = mean(t),
+        o3 = mean(o3),
+        hgt = mean(hgt)), by = .(lev, lon, lat, time = seasonally(time))]
+```
+
+(ref:t-vertical-cap) Regression of EOF1 with zonal anomalies of mean air temperature (shaded) and ozone mixing ratio (contours, negative contours with dashed lines, labels in parts per billion by mass) averaged between 80ºS and 25ºS for the period 1979 -- 2019.
+
+```{r lable_placement_min}
+lable_placement_min <- function(direction = c("vertical", "horizontal"))  {
+  direction <- direction[1]
+  function(x, y) {
+    if (direction == "vertical") {
+      return(which.min(y))
+    }
+    else if (direction == "horizontal") {
+      return(which.min(x))
+    }
+  }
+}
+```
 
@@
   .[lat %between% c(-80, -35)] %>% 
-  .[, .(t = weighted.mean(t, cos(lat*pi/180))), by = .(lev, time, lon)] %>%  
-  .[, t := Anomaly(t), by = .(time, lev)] %>% 
+  .[, .(hgt = weighted.mean(hgt, cos(lat*pi/180)),
+        o3 = weighted.mean(o3, cos(lat*pi/180)),
+        t = weighted.mean(t, cos(lat*pi/180))), by = .(lev, time, lon)] %>%  
+  melt(id.vars = c("lev", "time", "lon")) %>% 
+  .[, value := Anomaly(value), by = .(time, lev, variable)] %>%
   .[cut(ceof$eof[[1]], 1)$left, on = "time"] %>% 
@@
   .[, ":="(Real = as.numeric(scale(Real)), Imaginary = as.numeric(scale(Imaginary))), by = .(PC)] %>% 
-  .[, FitLm(t, Real, Imaginary, se = TRUE), by = .(lev, lon)] %>% 
+  .[, FitLm(value, Real, Imaginary, se = TRUE), by = .(lev, lon, variable)] %>% 
   rm_intercept() %>% 
   .[, term := factor_ReIm(term)] %>% 
-  .[, p.value := Pvaluate(estimate, std.error, df, "fdr"), by = .(term)] %>%
+  # .[, z := metR:::standard_atmosphere$altitude(lev*100)] %>% 
+  # .[, dz1 := c(NA, diff(estimate)/diff(z)) , by = .(term, lon, variable)] %>% 
+  # .[, dz2 := c(diff(estimate)/diff(z), NA) , by = .(term, lon, variable)] %>% 
+  # .[, dz := (dz1 + dz2)/2] %>% 
+  # .[, p.value := Pvaluate(estimate, std.error, df, "fdr"), by = .(term, variable)] %>%
+  # .[lev < 900] %>% 
   ggplot(aes(lon, lev)) +
-  geom_contour_fill(aes(z = estimate, fill = ..level..),
-                    breaks = AnchorBreaks(0, 0.25, exclude = 0)) +
-  geom_contour_tanaka(aes(z = estimate), 
-                      breaks = AnchorBreaks(0, 0.25, exclude = 0)) +
-  geom_contour_pval(aes(z = p.value)) +
+  geom_contour_fill(data = ~.x[variable == "t"],
+                    aes(z = estimate, fill = ..level..),
+                    breaks = AnchorBreaks(0, binwidth = 0.25, exclude = 0)) +
+  geom_contour_tanaka(data = ~.x[variable == "t"],
+                      aes(z = estimate),
+                      breaks = AnchorBreaks(0, binwidth = 0.25, exclude = 0)) +
+  geom_contour2(data = ~.x[variable == "o3"], 
+                aes(z = estimate, linetype = factor(-sign(..level..))), 
+                breaks = AnchorBreaks(0, exclude = 0)) +
+  geom_text_contour(data = ~.x[variable == "o3"], 
+                aes(z = estimate*1e9, label = signif(..level.., 2)),
+                size = 3,
+                skip = 1, stroke = 0.15, stroke.colour = "white", 
+                label.placement = lable_placement_min(),
+                breaks = AnchorBreaks(0, exclude = 0)) +
+
   scale_fill_divergent_discretised("Air temperature", guide = guide_colorsteps_bottom(20)) + 
   scale_y_level() +
+  scale_linetype(guide = "none") +
   scale_x_longitude(ticks = 60) +
   facet_grid(~term) +
-  axis_labs_smol
-```
-
-Figure \@ref(fig:eof1-regr-t) shows regression patterns of EOF1 with air temperature.
+  axis_labs_smol +
+  coord_cartesian(clip = "off")
+```
+
+Figure\ \@ref(fig:eof1-regr-t) shows regression patterns of EOF1 with air temperature.
 In both levels, the distribution of temperature anomalies mirror the geopotential height anomalies in 50 hPa.
-To understand the vertical distribution of these anomalies, Figure \@ref(fig:t-vertical)) shows the regression pattern of EOF1 and zonal anomalies of air temperature averaged between 80ºS and 35ºS.
-Both pars of EOF1 are associated with a temperature wave 1 pattern which changes phase around 10 hPa.
-This switch in the sign of the correlation in temperature at 10 hPa suggest a relationship with Ozone, since its indicative of the effects of planetary waves in the advection of Ozone [@hartmann1979;@rood1985;@smith1995].
+To understand the vertical distribution of these anomalies, Figure\ \@ref(fig:t-vertical)) shows the regression pattern of EOF1 with zonal anomalies of air temperature and zonal anomalies of ozone mixing ratio averaged between 80ºS and 35ºS.
+Temperature zonal anomalies associated with EOF1 show a clear wave 1 pattern throughout the atmosphere above 200 hPa with a change in sign above 10 hPa.
+Following hydrostatic balance, this is the level in which the geopotential anomaly have maximum amplitude (not shown).
+
+The maximum ozone anomalies are co-located with the minimum temperature anomalies above 10 hPa and with the maximum temperature anomalies below 10 hPa; that is, the ozone zonal wave 1 is anticorrelated with the temperature zonal wave 1 in the upper stratosphere, and directly correlated in the upper stratosphere.
+This change in phase is observed in ozone anomalies forced by planetary waves that reach the stratosphere.In the photochemically-dominated upper stratosphere, cold temperatures inhibit the destruction of ozone, and the advectively-dominated lower stratosphere, ozone anomalies are 90º out of phase with horizontal and vertical transport, which is 90º our of phase with temperature anomalies [@hartmann1979;@wirth1993;@smith1995].
+
+```{r o3-t-phase}
+# t %>% 
+#   .[lat %between% c(-80, -35)] %>% 
+#   .[, .(o3 = weighted.mean(o3, cos(lat*pi/180)),
+#         t = weighted.mean(t, cos(lat*pi/180))), by = .(lev, time, lon)] %>%  
+#   melt(id.vars = c("lev", "time", "lon")) %>% 
+#   # .[, value := Anomaly(value), by = .(time, lev, variable)] %>%
+#   .[cut(ceof$eof[[1]], 1)$left, on = "time"] %>% 
+#   sep_ReIm(format = "wider") %>% 
+#   .[, ":="(Real = as.numeric(scale(Real)), Imaginary = as.numeric(scale(Imaginary))), by = .(PC)] %>% 
+#   .[, FitLm(value, Real, Imaginary, se = TRUE), by = .(lev, lon, variable)] %>% 
+#   rm_intercept() %>% 
+#   .[, term := factor_ReIm(term)] %>% 
+#   .[, FitWave(estimate, 1), by = .(lev, variable, term)] %>% 
+#   dcast(lev + term ~ variable, value.var = c("phase", "amplitude")) %>%
+#   ggplot(aes(lev, acos(cos(phase_t - phase_o3)))) +
+#   geom_line(aes(color = term)) +
+#   scale_x_level() +
+#   scale_colour_brewer(NULL, palette = "Dark2") +
+#   scale_y_continuous("Wave-1 phase difference\nbetween Temperature and Ozone",
+#                      breaks = seq(-pi, pi, by = 30*pi/180),
+#                      labels = function(x) paste0(round(x*180/pi, 5), "º"))  +
+#   coord_flip()
+```
 
@@
     by = .(PC)]
-
+o3_mean <- o3 %>% 
+  .[, .(toc = mean(toc)), by = .(lon, lat)] %>% 
+  .[, toc_z := Anomaly(toc), by = .(lat)]
 o3_regr <- temp %>% 
@@
 
-(raf:o3-regr-cap) Regression coefficients of the real and imaginary part of the first complex EOF on SON mean Total Ozone Column for the 1979 -- 2019 period.
-
-```{r o3-regr, fig.cap = "(raf:o3-regr-cap)", fig.height=4, fig.width = 5}
+(ref:o3-regr-cap) Regression coefficients of the real and imaginary part of the first complex EOF on SON mean Total Ozone Column.
+On contours, the mean zonal anomaly of Total OZone Column (negative contours in dashed lines).
+For  the 1979 -- 2019 period.
+
+```{r o3-regr, fig.cap = "(ref:o3-regr-cap)", fig.height=4, fig.width = 5}
+
 o3_regr %>% 
@@
                     breaks = AnchorBreaks(0, exclude = 0)) +
+  geom_contour2(data = periodic(o3_mean, lon = c(0, 360)),
+                aes(z = toc_z, linetype = factor(-sign(..level..))), breaks = AnchorBreaks(0, exclude = 0), 
+                bins = 5) +
   geom_contour_tanaka(aes(z = estimate), breaks = AnchorBreaks(0, exclude = 0)) +
   geom_contour_pval(aes(z = p.value)) +
+
   geom_qmap(~.x[lat <= -20]) +
@@
   scale_y_latitude() +
+  scale_linetype(guide = "none") +
   coord_polar() +
@@
 
-To confirm this possibility, Figure \@ref(fig:o3-regr) shows regression maps of EOF1 with fields of Total Ozone Column (TOC).
+Figure\ \@ref(fig:o3-regr) shows regression maps of EOF1 with fields of Total Ozone Column (TOC).
 It shows zonal wave 1 pattens in TOC associated with both phases of EOF1.
 Climatologically, the springtime Ozone minimum is located off the South Pole and towards the Weddell Sea [@wirth1993].
-Thus, the Real EOF1 regression pattern (Figure \@ref(fig:o3-regr)a) coincides with the climatological position of the ozone whole while the one for the Imaginary EOF1 is shifted by 90º.
-This suggest that the EOF1 can be a complete description of the TOC zonal wave 1.
+Thus, the Real EOF1 regression pattern (Figure\ \@ref(fig:o3-regr)a) coincides with the climatological position of the ozone whole while the one for the Imaginary EOF1 is shifted by 90º.
 
@@
 
-Indeed, Figure \@ref(fig:o3-regr) shows the relationship between amplitude and phase of the planetary wave 1 in Total Ozone Column between 70ºS and 45ºS and amplitude and phase of the EOF1.
+Indeed, Figure\ \@ref(fig:wave1-o3) shows the relationship between amplitude and phase of the planetary wave 1 in Total Ozone Column between 70ºS and 45ºS and amplitude and phase of the EOF1.
 The correlation between the zonal wave 1 of ozone and EOF1 (computed as the mean cosine of the difference in phase) is `r wave_correlation$text`.
-As expected from the location of anomalies in Figure \@ref(fig: o3-regr), the Real EOF1 drives the relationship with amplitude and the Imaginary EOF1 drives the relationship with phase (not shown).
+As expected from the location of anomalies in Figure\ \@ref(fig:wave1-o3), the Real EOF1 drives the relationship with amplitude and the Imaginary EOF1 drives the relationship with phase (not shown).
 
@@
 
-Figure \@ref(fig:psa-eof2) shows the relationship between the two PSA indices and the Real and Imaginary phase of EOF2.
-As anticipated by Figure \@ref(fig: eof2-regr-gh), there is a strong correlation between PSA1 and Imaginary EOF2 (Figure \@ref(fig:psa-eof2)b.1), and between PSA2 and Real EOF2 (Figure \@ref(fig:psa-eof2)a.2).
-Conversely, there is no relationship between PSA1 and Real EOF2, and between PSA2 and Imaginary EOF2 (Figure \@ref(fig:psa-eof2) panels a.1 and b.2).
+Figure\ \@ref(fig:psa-eof2) shows the relationship between the two PSA indices and the Real and Imaginary phase of EOF2.
+As anticipated by Figure\ \@ref(fig:eof2-regr-gh), there is a strong correlation between PSA1 and Imaginary EOF2 (Figure\ \@ref(fig:psa-eof2)b.1), and between PSA2 and Real EOF2 (Figure\ \@ref(fig:psa-eof2)a.2).
+Conversely, there is no relationship between PSA1 and Real EOF2, and between PSA2 and Imaginary EOF2 (Figure\ \@ref(fig:psa-eof2) panels a.1 and b.2).
 So not only this EOF2 represents well both the spatial structure and temporal evolution of the PSA modes, but it's also possible to make a rather clean association between its two phases and the two PSA modes.
@@
 It is also the same that maximises this clean association between EFO2 parts and PSA modes.
-The reason that the "naive" principal component analysis arrives naturally to this optimal separation is that the imaginary phase is the most common phase (Figure \@ref(fig:phase-histogram)), so it naturally appears first when performing a Principal Component Analysis.
+The reason that the "naive" principal component analysis arrives naturally to this optimal separation is that the imaginary phase is the most common phase (Figure\ \@ref(fig:phase-histogram)), so it naturally appears first when performing a Principal Component Analysis.
 
@@
 By aligning he imaginary phase with the direction with maximum relationship with ENSO, we also aligned it with the phase of maximum occurrence.
-This is analogous to the method in @irving2016, who used reprojection and Fourier filtering to detect "PSA-like" variability and defined as the proper PSA using the peaks of the phase distribution (Compare our Figure \@ref:(phase-histogram) with their Figure 6).
+This is analogous to the method in @irving2016, who used reprojection and Fourier filtering to detect "PSA-like" variability and defined as the proper PSA using the peaks of the phase distribution (Compare our Figure\ \@ref:(phase-histogram) with their Figure 6).
 
@@
 
-Figure \@ref(fig:sam-eof-vertical) shows the coefficient of determination between the EOFs and three SAM indices.
-
-Both EOFs bear some modest relationship with the SAM index (thick green line in Figure \@ref(fig:sam-eof-vertical)), although not statistically significant at every level.
+Figure\ \@ref(fig:sam-eof-vertical) shows the coefficient of determination between the EOFs and three SAM indices.
+
+Both EOFs bear some modest relationship with the SAM index (thick green line in Figure\ \@ref(fig:sam-eof-vertical)), although not statistically significant at every level.
 The split between A-SAM and S-SAM gives more insight into the nature of the relationship.
-The relationship between the SAM and the imaginary EOF1 (Figure \@ref(fig:sam-eof-vertical).b1) is mediated by S-SAM in the troposphere, but by the A-SAM in the stratosphere.
-
-The Imaginary EOF2 is related with the SAM through the A-SAM in the troposphere, with up to `r scales::percent(max_r2[PC == "PC2"]$estimate, 2)` of shared variance, reached at `r max_r2[PC == "PC2"]$lev` hPa (Figure \@ref(fig:sam-eof-vertical).b2).
+The relationship between the SAM and the imaginary EOF1 (Figure\ \@ref(fig:sam-eof-vertical).b1) is mediated by S-SAM in the troposphere, but by the A-SAM in the stratosphere.
+
+The Imaginary EOF2 is related with the SAM through the A-SAM in the troposphere, with up to `r scales::percent(max_r2[PC == "PC2"]$estimate, 2)` of shared variance, reached at `r max_r2[PC == "PC2"]$lev` hPa (Figure\ \@ref(fig:sam-eof-vertical).b2).
 
@@
 
-As a further illustration, Figure \@ref(fig:eof-filtered) shows the spatial pattern of the leading EOF of 200 hPa geopotential height when the variability associated with EOF2 is removed.
+As a further illustration, Figure\ \@ref(fig:eof-filtered) shows the spatial pattern of the leading EOF of 200 hPa geopotential height when the variability associated with EOF2 is removed.
 The resulting pattern is an nearly zonally symmetrical annular mode.
@@
 This suggests that the asymmetric part of the SAM might be statistical contamination from one phase of the PSA.
-Figure \@ref(fig:fake-eof) serves as an illustration of this kind of statistical mixing of independent modes.
+Figure\ \@ref(fig:fake-eof) serves as an illustration of this kind of statistical mixing of independent modes.
 If shows the leading (and only) EOFs of a synthetic dataset if `r uniqueN(fake_sam$time)` years created by adding a perfectly zonally symmetrical SAM-like pattern with random amplitudes taken from the real SAM, and geopotential height anomalies predicted by EFO2.
@@
     .[lat %between% c(-90, 10)] %>%
-    .[, p.value := Pvaluate(estimate, std.error, df, "fdr"), by = .(PC, term)] %>%
+    .[, p_val := Pvaluate(estimate, std.error, df, "fdr"), by = .(PC, term)] %>%
     .[, var := "Streamfunction"] %>%
@@
                       breaks = AnchorBreaks(exclude = 0)) +
-    geom_contour_pval(aes(z = p.value)) +
+    geom_contour_pval(aes(z = p_val)) +
     geom_streamline(aes(dx = fx, dy = fy), min.L = 3, L = 10, n = 30,
-                    size = 0.2, res = 2, nx = 25, ny = 15) +
+                    size = 0.2, res = 2, nx = 25, ny = 15, 
+                    arrow.angle = 14, arrow.length = 0.2) +
     geom_qmap(keep = .015) +
@@
                       aes(z = estimate, fill = ..level..), breaks = AnchorBreaks(exclude = 0)) +
-    geom_contour_tanaka(data = sst_gdata,
-                        aes(z = estimate), breaks = AnchorBreaks(exclude = 0)) +
-    geom_contour2(data = sst_gdata,
-                  aes(z = p_val), breaks = 0.05, size = 0.1) +
-    stat_subset(data = sst_gdata,
-                aes(subset = p_val <= 0.05 & is.cross(lon, lat, 3)),
-                geom = "point", size = 0.2, alpha = 0.5) +
+    geom_contour_pval(data = sst_gdata, aes(z = p_val)) +
     scale_fill_divergent_discretised("SST", guide = guide_colorsteps_bottom(order = 2)) +
@@
 
-```{r sst-psi-2,  fig.cap = "(ref:sst-psi-2-cap)", fig.with = 6, fig.height=4}
+```{r sst-psi-2,  fig.cap = "(ref:sst-psi-2-cap)", fig.width = 6, fig.height=4, dependson="plot_sst_psi"}
 plot_sst_psi(which_pc = "PC2", which_lev = 200)
@@
 
-Figure \@ref(fig:sst-psi-2) shows regression maps between EOF2 and Sea Surface Temperatures (SST) Streamfunction at 200 hPa.
-The Imaginary EOF2 is associated with strong positive SST anomalies on the Central Pacific and negative anomalies on an area across the North of Australia and New Zealand, the South Pacific Convergence Zone (SPCZ) (Figure \@ref(fig:sst-psi-2).a2).
+Figure\ \@ref(fig:sst-psi-2) shows regression maps between EOF2 and Sea Surface Temperatures (SST) Streamfunction at 200 hPa.
+The Imaginary EOF2 is associated with strong positive SST anomalies on the Central Pacific and negative anomalies on an area across the North of Australia and New Zealand, the South Pacific Convergence Zone (SPCZ) (Figure\ \@ref(fig:sst-psi-2).a2).
 This pattern is almost canonically positive ENSO and indeed, the correlation between the Imaginary EOF2 and the Oceanic Niño Index ([@bamston1997]) is very high `r enso_cor[PC == "PC2" & part == "Imaginary"]$text`.
 Streamfunction anomalies show a coherent picture.
-The Imaginary EOF2 is associated with strong wave-like streamfunction anomalies emanating from the tropics (Figure \@ref(fig:sst-psi-2).b2).
+The Imaginary EOF2 is associated with strong wave-like streamfunction anomalies emanating from the tropics (Figure\ \@ref(fig:sst-psi-2).b2).
 This is consistent with what we know of the effect of ENSO on the extratropics: SST anomalies initiate anomalous convection that excites Rossby waves that propagate meridionally towards higher latitudes.
@@
 Since the Real EOF2 represents just a different phase of the same wave train, one would expect that it would show a similar forcing pattern to the Imaginary EOF with a slight translation of its location.
-However, Figure \@ref(fig:sst-psi-2).a1 and b1 show that the Real EOF2 is not associated either with any significant SST nor streamfunction anomalies in the tropics.
+However, Figure\ \@ref(fig:sst-psi-2).a1 and b1 show that the Real EOF2 is not associated either with any significant SST nor streamfunction anomalies in the tropics.
 The correlation between the Real EOF2 and ENSO is also not significant (`r enso_cor[PC == "PC2" & part == "Real"]$cor_text`).
@@
 
-To better explore the relationship between tropical forcing and phase of the EOF2, Figure \@ref(fig:enso-phase) plots the ONI index and the phase of the EOF2 for each year between 1979 and 2019, highlighting years in which the magnitude of EOF2 was higher than 50% of the years.
+To better explore the relationship between tropical forcing and phase of the EOF2, Figure\ \@ref(fig:enso-phase) plots the ONI index and the phase of the EOF2 for each year between 1979 and 2019, highlighting years in which the magnitude of EOF2 was higher than 50% of the years.
 In years with positive ENSO, the phase of the EOF2 is always around +90º (corresponding with positive imaginary part) and vice versa.
 In years with near neutral ENSO, the phase of the EOF2 is much more variable.
-The black line in Figure \@ref(fig:enso-phase) tries to quantify this relationship.
+The black line in Figure\ \@ref(fig:enso-phase) tries to quantify this relationship.
 Is the equation $ONI = `r signif(coef(cos_model)[1], 2)`\sin{(phase)}$, whose coefficient we fitted by weighted least squares using the magnitude of the EFO2 as weight.
@@
 
-Figure \@ref(fig:enso-phase) suggest that strong EOF2 years tend to coincide with strong ENSO years.
+Figure\ \@ref(fig:enso-phase) suggest that strong EOF2 years tend to coincide with strong ENSO years.
 The correlation between the absolute magnitude of the ONI and the magnitude of the EOF2 is `r cor_enso_mag$text`.
@@
 When excited by tropical SST anomalies, it tends to remain locked to the imaginary phase, with the sign of the geopotential anomalies depending on the sign of the tropical SSTs anomalies.
-This explains the relative overabundance of years with EOF2 near positive and negative imaginary phase in Figure \@ref(fig:phase-histogram).
-
-(ref:sst-psi-1-cap) Same as Figure \@ref(fig:sst-psi-2) but for EOF1.
-
-```{r sst-psi-1, fig.cap = "(ref:sst-psi-1-cap)", fig.with = 6, fig.height=4}
+This explains the relative overabundance of years with EOF2 near positive and negative imaginary phase in Figure\ \@ref(fig:phase-histogram).
+
+(ref:sst-psi-1-cap) Same as Figure\ \@ref(fig:sst-psi-2) but for EOF1.
+
+```{r sst-psi-1, fig.cap = "(ref:sst-psi-1-cap)", fig.width = 6, fig.height=4, dependson="plot_sst_psi"}
 plot_sst_psi(which_pc = "PC1", which_lev = 200)
@@
 
-Figure \@ref(fig:sst-psi-1) shows SST and streamfunction regression maps for EOF1.
+Figure\ \@ref(fig:sst-psi-1) shows SST and streamfunction regression maps for EOF1.
 There is no significant pattern of SST anomalies associated with either the Real or Imaginary EOF1.
@@
 
-Figure \@ref(fig:pp-america) shows regression maps of seasonal precipitation with each EOF in South America.
+Figure\ \@ref(fig:pp-america) shows regression maps of seasonal precipitation with each EOF in South America.
 EOF1 is associated with slight decreases in precipitation in Southern Brazil and Paraguay, although these are not statistically significant \@ref(fig:pp-america) column a.
@@
 The Real EOF2, on the other hand, is associated with negative precipitation anomalies in a smaller area of SESA.
-The relationship between precipitation anomalies in SESA and the phase of EOF2 follows a curve similar to that of Figure \@ref(fig:enso-phase) (not shown).
-
-(ref:pp-oceania-cap) Same as Figure \@ref(fig:pp-america) but for New Zealand and neighbouring islands.
+The relationship between precipitation anomalies in SESA and the phase of EOF2 follows a curve similar to that of Figure\ \@ref(fig:enso-phase) (not shown).
+
+(ref:pp-oceania-cap) Same as Figure\ \@ref(fig:pp-america) but for New Zealand and neighbouring islands.
 
@@
 
-On the antipodes, precipitation anomalies in New Zealand and neighbouring islands associated with EOF1 are also weak and not statistically significant (Figure \@ref(fig:pp-oceania) column a) whereas EFO2 is associated with large and expansive anomalies.
+On the antipodes, precipitation anomalies in New Zealand and neighbouring islands associated with EOF1 are also weak and not statistically significant (Figure\ \@ref(fig:pp-oceania) column a) whereas EFO2 is associated with large and expansive anomalies.
 The Real EOF2 is associated with positive seasonal anomalies in Northern and Eastern Australia, and negative ones over New Zealand, while precipitation anomalies associated with Imaginary EOF2 are negative over all Eastern Australia.
@@
 
-(ref:australia-pp-phase-cap) Same as Figure \@ref(fig:enso-phase) but for precipitacion anomalies averaged between  `r knitr::combine_words(LonLabel(box_australia$lon))` and `r knitr::combine_words(LatLabel(box_australia$lat))` (box shown in Figure \@ref(fig:pp-oceania) column b).
+(ref:australia-pp-phase-cap) Same as Figure\ \@ref(fig:enso-phase) but for precipitacion anomalies averaged between  `r knitr::combine_words(LonLabel(box_australia$lon))` and `r knitr::combine_words(LatLabel(box_australia$lat))` (box shown in Figure\ \@ref(fig:pp-oceania) column b).
 
@@
 
-To better understand the relationship between precipitation anomalies over Western Australia and the EOF2, Figure \@ref(fig:australia-pp-phase) plots precipitation anomalies averaged over Eastern Australia (box in Figure \@ref(fig:pp-oceania)) as a function of EOF2 phase.
+To better understand the relationship between precipitation anomalies over Western Australia and the EOF2, Figure\ \@ref(fig:australia-pp-phase) plots precipitation anomalies averaged over Eastern Australia (box in Figure\ \@ref(fig:pp-oceania)) as a function of EOF2 phase.
 The black line is the equation $\mathrm{Precipitation} = `r signif(coef(model)[1], 2)`\sin{(\mathrm{phase})} + `r signif(coef(model)[2], 2)`\cos{(\mathrm{phase})}$, whose coefficient we fitted by weighted least squares using the magnitude of the EFO2 as weight.
@@
 
-(ref:pp-africa-cap) Same as Figure \@ref(fig:pp-america) but for South of Africa.
+(ref:pp-africa-cap) Same as Figure\ \@ref(fig:pp-america) but for South of Africa.
 

### 2021-03-08 analysis/paper/paper.tex
O3 mixing ratio

@@
 
-We used monthly geopotential height, air temperature and Total Ozone COlumn (TOC) at 2.5\degree~longitude by 2.5\degree~latitude of horizontal resolution and 37 vertical isobaric levels from ERA5 \citep{era5} for the period 1979 to 2019.
+We used monthly geopotential height, air temperature, ozone mixing ratio, and Total Ozone COlumn (TOC) at 2.5\degree~longitude by 2.5\degree~latitude of horizontal resolution and 37 vertical isobaric levels from ERA5 \citep{era5} for the period 1979 to 2019.
 We restrict most of our analysis to the post-satellite era to avoid any confounding factors arising from the incorporation of satellite observations, but we also use the preliminary back extension of ERA5 from 1950 to 1978 \citep{era5be} to look at long-term trends.
@@
 In traditional EOF analysis zonal waves appear as pairs of EOFs, usually degenerate, that represent similar patterns but shifted in phase.
-For instance, Figure \ref{fig:eof-naive} shows the leading 4 EOFs of zonally anomalous September through October geopotential height at 50 hPa and 100 hPa.
+For instance, Figure~\ref{fig:eof-naive} shows the leading 4 EOFs of zonally anomalous September through October geopotential height at 50 hPa and 100 hPa.
 It is clear that at 50 hPa the first two EFS represent the same zonal wave 1 pattern and the last 2 represent the same zonal wave 3 pattern shifted by 1/4 wavelength.
@@
 
-Another characteristic seen in Figure \ref{fig:eof-naive} is that EOF1 at 200 hPa is very similar to EFO1 at 50 hPa.
+Another characteristic seen in Figure~\ref{fig:eof-naive} is that EOF1 at 200 hPa is very similar to EFO1 at 50 hPa.
 The zonal wave 3 structure in EOF2 and EOF3 at 200 hPa are also very similar to the one in EOF3 and EOF4 at 50 hPa.
@@
 For the first complex principal component, the phase was chosen so that the time series corresponding to the real part has the maximum correlation with the zonal wave 1 of Total Ozone Column between 70°S and 45°S.
-This also nearly minimises the correlation with the imaginary part (Figure \ref{fig:rotations}a).
+This also nearly minimises the correlation with the imaginary part (Figure~\ref{fig:rotations}a).
 
-For the second complex principal component, the phase was chosen so that the coefficient of determination between the Oceanic Niño Index \citep{bamston1997} and the real part was minimised, which also nearly maximises the correlation with the imaginary part (Figure \ref{fig:rotations}b).
+For the second complex principal component, the phase was chosen so that the coefficient of determination between the Oceanic Niño Index \citep{bamston1997} and the real part was minimised, which also nearly maximises the correlation with the imaginary part (Figure~\ref{fig:rotations}b).
 It's important to reassure the reader that this procedure does not create a spurious correlation, it only takes whatever relationship that already exist and aligns it with a specific phase.
@@
 Figures \ref{fig:ceofs-1} and \ref{fig:ceofs-2} show, respectively, the spatial and temporal parts of the first two leading complex EOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa.
-In the spatial patterns in Figure \ref{fig:ceofs-1}, the real (in shading) and the imaginary parts (in contour) are in quadrature by construction, so that each EOF describe a single wave-like pattern whose amplitude and position (i.e.~phase) is controlled by the magnitude and phase of the complex temporal EOF.
+In the spatial patterns in Figure~\ref{fig:ceofs-1}, the real (in shading) and the imaginary parts (in contour) are in quadrature by construction, so that each EOF describe a single wave-like pattern whose amplitude and position (i.e.~phase) is controlled by the magnitude and phase of the complex temporal EOF.
 
-The wave patterns described by these complex EOFs match the patterns seen in the traditional EOFs of Figure \ref{fig:eof-naive}: The first is a wave 1, while the second is a wave 3.
-Note that in Figure \ref{fig:ceofs-2}, both parts of the EOF1 have non zero mean.
+The wave patterns described by these complex EOFs match the patterns seen in the traditional EOFs of Figure~\ref{fig:eof-naive}: The first is a wave 1, while the second is a wave 3.
+Note that in Figure~\ref{fig:ceofs-2}, both parts of the EOF1 have non zero mean.
 This is due to th fact that the geopotential fields that enter into the algorithm are anomalies with respect to the zonal mean, not the time mean.
@@
 
-Figure \ref{fig:extended-series} shows time series of the two complex EOFs extended beyond the satellite era using the preliminary ERA5 back extension going back to 1950 (which we call the ``hybrid ERA5'' reanalysis).
-There is a downward trend in the real part of EOF1 in the hybrid reanalysis (Figure \ref{fig:extended-series}a.1, p-value \textless{} 0.001).
+Figure~\ref{fig:extended-series} shows time series of the two complex EOFs extended beyond the satellite era using the preliminary ERA5 back extension going back to 1950 (which we call the ``hybrid ERA5'' reanalysis).
+There is a downward trend in the real part of EOF1 in the hybrid reanalysis (Figure~\ref{fig:extended-series}a.1, p-value \textless{} 0.001).
 There is no significant trend in any of the complex parts of EOF2.
@@
 
-The spatial patterns shown in Figure \ref{fig:ceofs-1} are ``pure'' EOF patterns.
+The spatial patterns shown in Figure~\ref{fig:ceofs-1} are ``pure'' EOF patterns.
 But as they are derived by removing the zonally symmetric circulation, they might not include all the variability that is actually associated with the EOF time series.
@@
 
-Figure \ref{fig:eof1-regr-gh} shows regression patterns of EOF1 and geopotential height.
-At 50 hPa (Figure \ref{fig:eof1-regr-gh} row a), both the Real and Imaginary EOF1 are associated with planetary wave 1 patterns, that are 90º off phase.
-Their phases coincide with the ones shown in Figure \ref{fig:ceofs-1}a.1, with the positive centre of the Real EOF1 located towards the dateline, and the one of the Imaginary EOF1 located over Eastern Antarctica.
+Figure~\ref{fig:eof1-regr-gh} shows regression patterns of EOF1 and geopotential height.
+At 50 hPa (Figure~\ref{fig:eof1-regr-gh} row a), both the Real and Imaginary EOF1 are associated with planetary wave 1 patterns, that are 90º off phase.
+Their phases coincide with the ones shown in Figure~\ref{fig:ceofs-1}a.1, with the positive centre of the Real EOF1 located towards the dateline, and the one of the Imaginary EOF1 located over Eastern Antarctica.
 However, the Real EOF1 pattern is substantially altered by the zonally symmetric circulation.
 Instead of a clear wave 1 pattern, the regression pattern can be better describes as a monopole with its centre displaced from the South Pole.
-Similarly affected are the regression patterns at 200 hPa (Figure \ref{fig:eof1-regr-gh} row b).
-It is only possible to distinguish some of the wave 1 pattern in relation with the Real EOF1 (Figure \ref{fig:eof1-regr-gh}b.1).
+
+Similarly affected are the regression patterns at 200 hPa (Figure~\ref{fig:eof1-regr-gh} row b).
+It is only possible to distinguish some of the wave 1 pattern in relation with the Real EOF1 (Figure~\ref{fig:eof1-regr-gh}b.1).
 The Imaginary EOF1 shows a much more zonally symmetrical pattern resembling a negative Southern Annular Mode.
 
-With the exception of the Imaginary EFO1, it's clear that these patterns are very different than the idealised versions shown in Figure \ref{fig:ceofs-1}, particularly at 200 hPa.
+With the exception of the Imaginary EFO1, it's clear that these patterns are very different than the idealised versions shown in Figure~\ref{fig:ceofs-1}, particularly at 200 hPa.
 Moreover, only in the stratosphere these patterns actually show a distinguishable wave 1 pattern shifted in phase by 90º.
@@
 \includegraphics{../figures/eof2-regr-gh-1.pdf}
-\caption{\label{fig:eof2-regr-gh}Same as Figure \ref{fig:eof1-regr-gh} but for the second EOF.}
+\caption{\label{fig:eof2-regr-gh}Same as Figure~\ref{fig:eof1-regr-gh} but for the second EOF.}
 \end{figure}
 
-Figure @ref(fig: eof2-regr-gh) shows the regression pattern of geopotential height and the EOF2.
-Unlike in the case of EOF1, in this case the regression patterns represent relatively well the \emph{idealised} patterns from Figure \ref{fig:ceofs-1}.
+Figure~\ref{fig:eof2-regr-gh} shows the regression pattern of geopotential height and the EOF2.
+Unlike in the case of EOF1, in this case the regression patterns represent relatively well the \emph{idealised} patterns from Figure~\ref{fig:ceofs-1}.
 Although there are some differences (particularly in 50 hPa), the wave trains identified before are well characterised and patterns associated with the Real EOF2 are 90º out of phase with those associated with the Imaginary EOF2.
@@
 EOF2 then represents a equivalent barotropic wave train that is very similar to the the Pacific South American Patterns \citep{mo2001}.
-Comparing the location of the positive anomaly near 90ºW in column b of Figure @ref(fig: eof2-regr-gh) with Figure 1 a and b from \citet{mo2001}, the Real EOF2 can roughly be identified with the PSA2, while the Imaginary EOF2 resembles PSA1.
+Comparing the location of the positive anomaly near 90ºW in column b of Figure~\ref{fig:eof2-regr-gh} with Figure 1 a and b from \citet{mo2001}, the Real EOF2 can roughly be identified with the PSA2, while the Imaginary EOF2 resembles PSA1.
 This separation is not completely meaningful in this case, since by treating this pattern as a single wave train any one pair of orthogonal waves can characterise this variability.
@@
 \includegraphics{../figures/eof1-regr-t-1.pdf}
-\caption{\label{fig:eof1-regr-t}Same as Figure \ref{fig:eof1-regr-gh} but for air temperature.}
+\caption{\label{fig:eof1-regr-t}Same as Figure~\ref{fig:eof1-regr-gh} but for air temperature.}
 \end{figure}
@@
 \includegraphics{../figures/t-vertical-1.pdf}
-\caption{\label{fig:t-vertical}Regression of EOF1 with mean air temperature averaged between 80ºS and 25ºS for the period 1979 -- 2019.}
+\caption{\label{fig:t-vertical}Regression of EOF1 with zonal anomalies of mean air temperature (shaded) and ozone mixing ratio (contours, negative contours with dashed lines, labels in parts per billion by mass) averaged between 80ºS and 25ºS for the period 1979 -- 2019.}
 \end{figure}
 
-Figure \ref{fig:eof1-regr-t} shows regression patterns of EOF1 with air temperature.
+Figure~\ref{fig:eof1-regr-t} shows regression patterns of EOF1 with air temperature.
 In both levels, the distribution of temperature anomalies mirror the geopotential height anomalies in 50 hPa.
-To understand the vertical distribution of these anomalies, Figure \ref{fig:t-vertical}) shows the regression pattern of EOF1 and zonal anomalies of air temperature averaged between 80ºS and 35ºS.
-Both pars of EOF1 are associated with a temperature wave 1 pattern which changes phase around 10 hPa.
-This switch in the sign of the correlation in temperature at 10 hPa suggest a relationship with Ozone, since its indicative of the effects of planetary waves in the advection of Ozone \citep{hartmann1979, rood1985, smith1995}.
+To understand the vertical distribution of these anomalies, Figure~\ref{fig:t-vertical}) shows the regression pattern of EOF1 with zonal anomalies of air temperature and zonal anomalies of ozone mixing ratio averaged between 80ºS and 35ºS.
+Temperature zonal anomalies associated with EOF1 show a clear wave 1 pattern throughout the atmosphere above 200 hPa with a change in sign above 10 hPa.
+Following hydrostatic balance, this is the level in which the geopotential anomaly have maximum amplitude (not shown).
 
-(raf:o3-regr-cap) Regression coefficients of the real and imaginary part of the first complex EOF on SON mean Total Ozone Column for the 1979 -- 2019 period.
+The maximum ozone anomalies are co-located with the minimum temperature anomalies above 10 hPa and with the maximum temperature anomalies below 10 hPa; that is, the ozone zonal wave 1 is anticorrelated with the temperature zonal wave 1 in the upper stratosphere, and directly correlated in the upper stratosphere.
+This change in phase is observed in ozone anomalies forced by planetary waves that reach the stratosphere.In the photochemically-dominated upper stratosphere, cold temperatures inhibit the destruction of ozone, and the advectively-dominated lower stratosphere, ozone anomalies are 90º out of phase with horizontal and vertical transport, which is 90º our of phase with temperature anomalies \citep{hartmann1979, wirth1993, smith1995}.
 
@@
 \includegraphics{../figures/o3-regr-1.pdf}
-\caption{\label{fig:o3-regr}(raf:o3-regr-cap)}
+\caption{\label{fig:o3-regr}Regression coefficients of the real and imaginary part of the first complex EOF on SON mean Total Ozone Column.
+On contours, the mean zonal anomaly of Total OZone Column (negative contours in dashed lines).
+For the 1979 -- 2019 period.}
 \end{figure}
 
-To confirm this possibility, Figure \ref{fig:o3-regr} shows regression maps of EOF1 with fields of Total Ozone Column (TOC).
+Figure~\ref{fig:o3-regr} shows regression maps of EOF1 with fields of Total Ozone Column (TOC).
 It shows zonal wave 1 pattens in TOC associated with both phases of EOF1.
 Climatologically, the springtime Ozone minimum is located off the South Pole and towards the Weddell Sea \citep{wirth1993}.
-Thus, the Real EOF1 regression pattern (Figure \ref{fig:o3-regr}a) coincides with the climatological position of the ozone whole while the one for the Imaginary EOF1 is shifted by 90º.
-This suggest that the EOF1 can be a complete description of the TOC zonal wave 1.
+Thus, the Real EOF1 regression pattern (Figure~\ref{fig:o3-regr}a) coincides with the climatological position of the ozone whole while the one for the Imaginary EOF1 is shifted by 90º.
 
@@
 
-Indeed, Figure \ref{fig:o3-regr} shows the relationship between amplitude and phase of the planetary wave 1 in Total Ozone Column between 70ºS and 45ºS and amplitude and phase of the EOF1.
+Indeed, Figure~\ref{fig:wave1-o3} shows the relationship between amplitude and phase of the planetary wave 1 in Total Ozone Column between 70ºS and 45ºS and amplitude and phase of the EOF1.
 The correlation between the zonal wave 1 of ozone and EOF1 (computed as the mean cosine of the difference in phase) is 0.86 (CI: 0.76 -- 0.93).
-As expected from the location of anomalies in Figure @ref(fig: o3-regr), the Real EOF1 drives the relationship with amplitude and the Imaginary EOF1 drives the relationship with phase (not shown).
+As expected from the location of anomalies in Figure~\ref{fig:wave1-o3}, the Real EOF1 drives the relationship with amplitude and the Imaginary EOF1 drives the relationship with phase (not shown).
 
@@
 
-Figure \ref{fig:psa-eof2} shows the relationship between the two PSA indices and the Real and Imaginary phase of EOF2.
-As anticipated by Figure @ref(fig: eof2-regr-gh), there is a strong correlation between PSA1 and Imaginary EOF2 (Figure \ref{fig:psa-eof2}b.1), and between PSA2 and Real EOF2 (Figure \ref{fig:psa-eof2}a.2).
-Conversely, there is no relationship between PSA1 and Real EOF2, and between PSA2 and Imaginary EOF2 (Figure \ref{fig:psa-eof2} panels a.1 and b.2).
+Figure~\ref{fig:psa-eof2} shows the relationship between the two PSA indices and the Real and Imaginary phase of EOF2.
+As anticipated by Figure~\ref{fig:eof2-regr-gh}, there is a strong correlation between PSA1 and Imaginary EOF2 (Figure~\ref{fig:psa-eof2}b.1), and between PSA2 and Real EOF2 (Figure~\ref{fig:psa-eof2}a.2).
+Conversely, there is no relationship between PSA1 and Real EOF2, and between PSA2 and Imaginary EOF2 (Figure~\ref{fig:psa-eof2} panels a.1 and b.2).
 So not only this EOF2 represents well both the spatial structure and temporal evolution of the PSA modes, but it's also possible to make a rather clean association between its two phases and the two PSA modes.
@@
 It is also the same that maximises this clean association between EFO2 parts and PSA modes.
-The reason that the ``naive'' principal component analysis arrives naturally to this optimal separation is that the imaginary phase is the most common phase (Figure \ref{fig:phase-histogram}), so it naturally appears first when performing a Principal Component Analysis.
+The reason that the ``naive'' principal component analysis arrives naturally to this optimal separation is that the imaginary phase is the most common phase (Figure~\ref{fig:phase-histogram}), so it naturally appears first when performing a Principal Component Analysis.
 
@@
 By aligning he imaginary phase with the direction with maximum relationship with ENSO, we also aligned it with the phase of maximum occurrence.
-This is analogous to the method in \citet{irving2016}, who used reprojection and Fourier filtering to detect ``PSA-like'' variability and defined as the proper PSA using the peaks of the phase distribution (Compare our Figure @ref:(phase-histogram) with their Figure 6).
+This is analogous to the method in \citet{irving2016}, who used reprojection and Fourier filtering to detect ``PSA-like'' variability and defined as the proper PSA using the peaks of the phase distribution (Compare our Figure~@ref:(phase-histogram) with their Figure 6).
 
@@
 
-Figure \ref{fig:sam-eof-vertical} shows the coefficient of determination between the EOFs and three SAM indices.
+Figure~\ref{fig:sam-eof-vertical} shows the coefficient of determination between the EOFs and three SAM indices.
 
-Both EOFs bear some modest relationship with the SAM index (thick green line in Figure \ref{fig:sam-eof-vertical}), although not statistically significant at every level.
+Both EOFs bear some modest relationship with the SAM index (thick green line in Figure~\ref{fig:sam-eof-vertical}), although not statistically significant at every level.
 The split between A-SAM and S-SAM gives more insight into the nature of the relationship.
-The relationship between the SAM and the imaginary EOF1 (Figure \ref{fig:sam-eof-vertical}.b1) is mediated by S-SAM in the troposphere, but by the A-SAM in the stratosphere.
+The relationship between the SAM and the imaginary EOF1 (Figure~\ref{fig:sam-eof-vertical}.b1) is mediated by S-SAM in the troposphere, but by the A-SAM in the stratosphere.
 
-The Imaginary EOF2 is related with the SAM through the A-SAM in the troposphere, with up to -96\% of shared variance, reached at 225 hPa (Figure \ref{fig:sam-eof-vertical}.b2).
+The Imaginary EOF2 is related with the SAM through the A-SAM in the troposphere, with up to -96\% of shared variance, reached at 225 hPa (Figure~\ref{fig:sam-eof-vertical}.b2).
 
@@
 
-As a further illustration, Figure \ref{fig:eof-filtered} shows the spatial pattern of the leading EOF of 200 hPa geopotential height when the variability associated with EOF2 is removed.
+As a further illustration, Figure~\ref{fig:eof-filtered} shows the spatial pattern of the leading EOF of 200 hPa geopotential height when the variability associated with EOF2 is removed.
 The resulting pattern is an nearly zonally symmetrical annular mode.
@@
 This suggests that the asymmetric part of the SAM might be statistical contamination from one phase of the PSA.
-Figure \ref{fig:fake-eof} serves as an illustration of this kind of statistical mixing of independent modes.
+Figure~\ref{fig:fake-eof} serves as an illustration of this kind of statistical mixing of independent modes.
 If shows the leading (and only) EOFs of a synthetic dataset if 41 years created by adding a perfectly zonally symmetrical SAM-like pattern with random amplitudes taken from the real SAM, and geopotential height anomalies predicted by EFO2.
@@
 
-Figure \ref{fig:sst-psi-2} shows regression maps between EOF2 and Sea Surface Temperatures (SST) Streamfunction at 200 hPa.
-The Imaginary EOF2 is associated with strong positive SST anomalies on the Central Pacific and negative anomalies on an area across the North of Australia and New Zealand, the South Pacific Convergence Zone (SPCZ) (Figure \ref{fig:sst-psi-2}.a2).
+Figure~\ref{fig:sst-psi-2} shows regression maps between EOF2 and Sea Surface Temperatures (SST) Streamfunction at 200 hPa.
+The Imaginary EOF2 is associated with strong positive SST anomalies on the Central Pacific and negative anomalies on an area across the North of Australia and New Zealand, the South Pacific Convergence Zone (SPCZ) (Figure~\ref{fig:sst-psi-2}.a2).
 This pattern is almost canonically positive ENSO and indeed, the correlation between the Imaginary EOF2 and the Oceanic Niño Index (\citep{bamston1997}) is very high 0.76 (CI: 0.59 -- 0.86).
 Streamfunction anomalies show a coherent picture.
-The Imaginary EOF2 is associated with strong wave-like streamfunction anomalies emanating from the tropics (Figure \ref{fig:sst-psi-2}.b2).
+The Imaginary EOF2 is associated with strong wave-like streamfunction anomalies emanating from the tropics (Figure~\ref{fig:sst-psi-2}.b2).
 This is consistent with what we know of the effect of ENSO on the extratropics: SST anomalies initiate anomalous convection that excites Rossby waves that propagate meridionally towards higher latitudes.
@@
 Since the Real EOF2 represents just a different phase of the same wave train, one would expect that it would show a similar forcing pattern to the Imaginary EOF with a slight translation of its location.
-However, Figure \ref{fig:sst-psi-2}.a1 and b1 show that the Real EOF2 is not associated either with any significant SST nor streamfunction anomalies in the tropics.
+However, Figure~\ref{fig:sst-psi-2}.a1 and b1 show that the Real EOF2 is not associated either with any significant SST nor streamfunction anomalies in the tropics.
 The correlation between the Real EOF2 and ENSO is also not significant ().
@@
 
-To better explore the relationship between tropical forcing and phase of the EOF2, Figure \ref{fig:enso-phase} plots the ONI index and the phase of the EOF2 for each year between 1979 and 2019, highlighting years in which the magnitude of EOF2 was higher than 50\% of the years.
+To better explore the relationship between tropical forcing and phase of the EOF2, Figure~\ref{fig:enso-phase} plots the ONI index and the phase of the EOF2 for each year between 1979 and 2019, highlighting years in which the magnitude of EOF2 was higher than 50\% of the years.
 In years with positive ENSO, the phase of the EOF2 is always around +90º (corresponding with positive imaginary part) and vice versa.
 In years with near neutral ENSO, the phase of the EOF2 is much more variable.
-The black line in Figure \ref{fig:enso-phase} tries to quantify this relationship.
+The black line in Figure~\ref{fig:enso-phase} tries to quantify this relationship.
 Is the equation \(ONI = 1.1\sin{(phase)}\), whose coefficient we fitted by weighted least squares using the magnitude of the EFO2 as weight.
@@
 
-Figure \ref{fig:enso-phase} suggest that strong EOF2 years tend to coincide with strong ENSO years.
+Figure~\ref{fig:enso-phase} suggest that strong EOF2 years tend to coincide with strong ENSO years.
 The correlation between the absolute magnitude of the ONI and the magnitude of the EOF2 is 0.44 (CI: 0.16 -- 0.66).
@@
 When excited by tropical SST anomalies, it tends to remain locked to the imaginary phase, with the sign of the geopotential anomalies depending on the sign of the tropical SSTs anomalies.
-This explains the relative overabundance of years with EOF2 near positive and negative imaginary phase in Figure \ref{fig:phase-histogram}.
+This explains the relative overabundance of years with EOF2 near positive and negative imaginary phase in Figure~\ref{fig:phase-histogram}.
 
@@
 \includegraphics{../figures/sst-psi-1-1.pdf}
-\caption{\label{fig:sst-psi-1}Same as Figure \ref{fig:sst-psi-2} but for EOF1.}
+\caption{\label{fig:sst-psi-1}Same as Figure~\ref{fig:sst-psi-2} but for EOF1.}
 \end{figure}
 
-Figure \ref{fig:sst-psi-1} shows SST and streamfunction regression maps for EOF1.
+Figure~\ref{fig:sst-psi-1} shows SST and streamfunction regression maps for EOF1.
 There is no significant pattern of SST anomalies associated with either the Real or Imaginary EOF1.
@@
 
-Figure \ref{fig:pp-america} shows regression maps of seasonal precipitation with each EOF in South America.
+Figure~\ref{fig:pp-america} shows regression maps of seasonal precipitation with each EOF in South America.
 EOF1 is associated with slight decreases in precipitation in Southern Brazil and Paraguay, although these are not statistically significant \ref{fig:pp-america} column a.
@@
 The Real EOF2, on the other hand, is associated with negative precipitation anomalies in a smaller area of SESA.
-The relationship between precipitation anomalies in SESA and the phase of EOF2 follows a curve similar to that of Figure \ref{fig:enso-phase} (not shown).
+The relationship between precipitation anomalies in SESA and the phase of EOF2 follows a curve similar to that of Figure~\ref{fig:enso-phase} (not shown).
 
@@
 \includegraphics{../figures/pp-oceania-1.pdf}
-\caption{\label{fig:pp-oceania}Same as Figure \ref{fig:pp-america} but for New Zealand and neighbouring islands.}
+\caption{\label{fig:pp-oceania}Same as Figure~\ref{fig:pp-america} but for New Zealand and neighbouring islands.}
 \end{figure}
 
-On the antipodes, precipitation anomalies in New Zealand and neighbouring islands associated with EOF1 are also weak and not statistically significant (Figure \ref{fig:pp-oceania} column a) whereas EFO2 is associated with large and expansive anomalies.
+On the antipodes, precipitation anomalies in New Zealand and neighbouring islands associated with EOF1 are also weak and not statistically significant (Figure~\ref{fig:pp-oceania} column a) whereas EFO2 is associated with large and expansive anomalies.
 The Real EOF2 is associated with positive seasonal anomalies in Northern and Eastern Australia, and negative ones over New Zealand, while precipitation anomalies associated with Imaginary EOF2 are negative over all Eastern Australia.
@@
 \includegraphics{../figures/australia-pp-phase-1.pdf}
-\caption{\label{fig:australia-pp-phase}Same as Figure \ref{fig:enso-phase} but for precipitacion anomalies averaged between 130°E and 155°E and 40°S and 15°S (box shown in Figure \ref{fig:pp-oceania} column b).}
+\caption{\label{fig:australia-pp-phase}Same as Figure~\ref{fig:enso-phase} but for precipitacion anomalies averaged between 130°E and 155°E and 40°S and 15°S (box shown in Figure~\ref{fig:pp-oceania} column b).}
 \end{figure}
 
-To better understand the relationship between precipitation anomalies over Western Australia and the EOF2, Figure \ref{fig:australia-pp-phase} plots precipitation anomalies averaged over Eastern Australia (box in Figure \ref{fig:pp-oceania}) as a function of EOF2 phase.
+To better understand the relationship between precipitation anomalies over Western Australia and the EOF2, Figure~\ref{fig:australia-pp-phase} plots precipitation anomalies averaged over Eastern Australia (box in Figure~\ref{fig:pp-oceania}) as a function of EOF2 phase.
 The black line is the equation \(\mathrm{Precipitation} = -0.26\sin{(\mathrm{phase})} + 0.22\cos{(\mathrm{phase})}\), whose coefficient we fitted by weighted least squares using the magnitude of the EFO2 as weight.

### 2021-03-08 R/0-download.R

@@ -99,13 +99,13 @@ request_era5be <- list(
 request_era5 <- list(
   format = "netcdf",
   product_type = "monthly_averaged_reanalysis",
-  variable = c("geopotential", "temperature", "vorticity", "divergence"),
+  variable = c("geopotential", "temperature", "vorticity", "ozone_mass_mixing_ratio"),
   pressure_level = c("1", "2", "3", "5", "7", "10", "20", "30", "50",
                      "70", "100", "125", "150", "175", "200", "225",
                      "250", "300", "350", "400", "450", "500", "550",
                      "600", "650", "700", "750", "775", "800", "825",
                      "850", "875", "900", "925", "950", "975", "1000"),
-  year = as.character(1979:2019),
+  year = as.character(1979:2020),
   month = formatC(1:12, width = 2, flag = 0),   # Need all months to compute PSA
   time = "00:00",
   grid = c("2.5", "2.5"),
@@ -116,7 +116,7 @@ request_o3_era5 <- list(
   format = "netcdf",
   variable = "total_column_ozone",
   product_type = "monthly_averaged_reanalysis",
-  year = as.character(1979:2019),
+  year = as.character(1979:2020),
   month = formatC(9:11, width = 2, flag = 0),
   time = "00:00",
   area = c(0, -180, -90, 180),
@@ -130,7 +130,7 @@ request_era5_sst <- list(
   format = "netcdf",
   variable = "sea_surface_temperature",
   product_type = "monthly_averaged_reanalysis",
-  year = as.character(1979:2019),
+  year = as.character(1979:2020),
   month =  c("09", "10", "11"),
   time = "00:00",
   grid = c("2.5", "2.5"),

### 2021-03-13 analysis/paper/paper.Rmd
Corrige tamaño de las figuras

@@
 
-## Figure classes
-DefineClass <- function(class.opts) {
-    class.name <- deparse(substitute(class.opts))
-    class.fun <- function(options) {
-        class <- options[[class.name]]
-        class <- class.opts[[class]]
-        options[names(class)] <- class
-        options
-    }
-    invisible(knitr::opts_hooks$set(
-        setNames(list(class.fun), class.name)))
-}
-
-page.width <- 210 - 30 - 25
-text.width <- page.width - 20
-text.height <- page.height <- 297 - 25 - 25
-
-fig.class <- list(
-  colwidth = list(
-    fig.width = 5
-  ),
-    pagewidth = list(
-        fig.width = page.width/25.4,
-        out.width = paste0(page.width, "mm"),
-        # fig.align = "center",
-        fig.env = "figure*",
-        out.extra = " "
-    )
-)
-
-DefineClass(fig.class)
+# ## Figure classes
+# DefineClass <- function(class.opts) {
+#     class.name <- deparse(substitute(class.opts))
+#     class.fun <- function(options) {
+#         class <- options[[class.name]]
+#         class <- class.opts[[class]]
+#         options[names(class)] <- class
+#         options
+#     }
+#     invisible(knitr::opts_hooks$set(
+#         setNames(list(class.fun), class.name)))
+# }
+# 
+# page.width <- 210 - 30 - 25
+# text.width <- page.width - 20
+# text.height <- page.height <- 297 - 25 - 25
+# 
+# fig.class <- list(
+#   colwidth = list(
+#     fig.width = 5
+#   ),
+#     pagewidth = list(
+#         fig.width = page.width/25.4,
+#         out.width = paste0(page.width, "mm"),
+#         # fig.align = "center",
+#         fig.env = "figure*",
+#         out.extra = " "
+#     )
+# )
+# 
+# DefineClass(fig.class)
+
+width_column <- 3.6
 
@@
 
-theme_set(theme_shceof(base_size = 12))
+theme_set(theme_shceof(base_size = 10))
 
@@
 
-guide_colorstrip_bottom <- function(width = 25, height = 0.5, ...) {
+guide_colorstrip_bottom <- function(width = 15, height = 0.5, ...) {
   guide_colorstrip(title.position = "top", title.hjust = 0.5,
@@
 
-guide_colorsteps_bottom <- function(width = 20, height = 0.5, even.steps = FALSE, ...) {
+guide_colorsteps_bottom <- function(width = 15, height = 0.5, even.steps = FALSE, ...) {
   guide_colorsteps(show.limits = TRUE, even.steps = even.steps,
@@
 We used monthly geopotential height, air temperature, ozone mixing ratio, and Total Ozone COlumn (TOC) at 2.5\degree\ longitude by 2.5\degree\ latitude of horizontal resolution and 37 vertical isobaric levels from ERA5 [@era5] for the period 1979 to 2019.
-We restrict most of our analysis to the post-satellite era to avoid any confounding factors arising from the incorporation of satellite observations, but we also use the preliminary back extension of ERA5 from 1950 to 1978 [@era5be] to look at long-term trends.
+We restricted most of our analysis to the post-satellite era to avoid any confounding factors arising from the incorporation of satellite observations, but we also used the preliminary back extension of ERA5 from 1950 to 1978 [@era5be] to look at long-term trends.
 We derived streamfunction at 200 hPa from ERA5 vorticity using the FORTRAN subroutine FISHPACK [@fishpack] and computed horizontal wave activity fluxes following @plumb1985 (Equation 5.7).
-
-Precipitation data we used monthly data from the CPC Merged Analysis of Precipitation [@cmap].
-
-We compute the amplitude and phase of the zonal wave 1 by averaging SON TOC between `r knitr::combine_words(LatLabel(o3wave_lats))` for each Spring and computing the Fourier spectrum.
-
-We explore the relationship with the Southern Annular Mode (SAM).
-We compute the level-dependent SAM index as the leading EOF of year-round monthly geopotential height anomalies south of 20ºS at each level for the whole period [@baldwin2009].
+For precipitation data we used monthly data from the CPC Merged Analysis of Precipitation [@cmap].
+
+We computed the amplitude and phase of the zonal wave 1 by averaging variables (temperature, geopotential height, ozone mixing ratio and Total Ozone Column) `r knitr::combine_words(LatLabel(o3wave_lats))` for each Spring and computing the Fourier spectrum.
+
+We computed the level-dependent Southern Annular Mode index as the leading EOF of year-round monthly geopotential height anomalies south of 20ºS at each level for the whole period [@baldwin2009].
 We further split the SAM into its zonally symmetric and zonally asymmetric components (S-SAM and A-SAM indices respectively).
@@
 (ref:eof-naive-cap) Spatial patterns of the leading 4 EOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2009 (arbitrary units).
-The numbers at the top-left of each panel ate the variance explained by each EOF.
+The numbers at the top-left of each panel ate the zonal variance explained by each EOF.
 
@@
 
-```{r eof-naive, fig.cap = "(ref:eof-naive-cap)"}
+```{r eof-naive, fig.cap = "(ref:eof-naive-cap)", fig.width=3.7, fig.height=2.2}
 eof_naive[, eof[[1]]$right, by = lev] %>%
@@
                     breaks = AnchorBreaks(0, 0.01, exclude = 0)) +
-  geom_contour_tanaka(aes(z = hgt), breaks = AnchorBreaks(0, 0.01, exclude = 0)) +
-
-  geom_qmap(~.x[lat <= -20]) +
+  geom_contour_tanaka(aes(z = hgt), breaks = AnchorBreaks(0, 0.01, exclude = 0), 
+                    range = c(0.01, 0.2)) +
+
+  geom_qmap(~.x[lat <= -20], size = 0.1) +
   scale_fill_divergent_discretised(guide = "none")  +
@@
   facet_grid(lev~EOF, labeller = labeller(lev = lab_lev))  +
-  tagger::tag_facets(tag = "panel", tag_pool = explained, tag_suffix = "")
+  tagger::tag_facets("rc")  +
+  theme(panel.spacing = grid::unit(-.7, "lines"))
 ```
@@
 
-```{r rotations, fig.cap = "(ref:rotations-cap)", fig.width=5, fig.height=2.5}
+```{r rotations, fig.cap = "(ref:rotations-cap)", fig.width=width_column, fig.height=2.5}
 r2 <- expression(`$r^2$` = paste("", "r", phantom()^{
@@
 ```{r ceofs-1, fig.cap = "Spatial patterns of the leading 2 complex EOFs of zonal anomalies of geopotential height in 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2009.
-Real part in shading, imaginary part in contours", fig.width=5, fig.height=5}
+Real part in shading, imaginary part in contours", fig.width=width_column, fig.height=width_column}
 ceof[season == "SON"] %>%
@@
 
-```{r ceofs-2, fig.cap = "Time series of the leading 2 complex EOFs of zonal anomalies of geopotential height in 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2009.", fig.height=4, fig.width = 5}
+```{r ceofs-2, fig.cap = "Time series of the leading 2 complex EOFs of zonal anomalies of geopotential height in 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2009.", fig.height=width_column, fig.width = width_column}
 ceof %>%
@@
 
-```{r extended-series, fig.cap = "(ref:extended-series-cap)", fig.height=4, fig.width = 5}
+```{r extended-series, fig.cap = "(ref:extended-series-cap)", fig.height=3, fig.width = 7, fig.env = "figure*"}
 # Check that the extended index coincides with the
@@
   .[, ref := year(time) > 1979] %>% 
-  .[, hgt := scale(hgt), by = .(part, PC)] %>%
-  ggplot(aes(time, hgt)) +
+   # .[, hgt := scale(hgt, center = FALSE, scale = sd(hgt[ref])), by = .(part, PC)] %>%
+  ggplot(aes(time, hgt*1e3)) +
   # Zero line + rangeframe
-  geom_hline(data = ~.x[, mean(hgt), by  = .(PC, part)], 
+  geom_hline(data = ~.x[, mean(hgt)*1e3, by  = .(PC, part)], 
              aes(yintercept = V1), size = 0.1, color = "gray30") +
@@
   scale_alpha_manual(values = c(1, 0.2)) +
-
   geom_smooth(method = "lm", color = "black") +
-
   scale_x_datetime(NULL, date_labels = "%Y") +
@@
 
-Figure\ \@ref(fig:extended-series) shows time series of the two complex EOFs extended beyond the satellite era using the preliminary ERA5 back extension going back to 1950 (which we call the "hybrid ERA5" reanalysis).
-There is a downward trend in the real part of EOF1 in the hybrid reanalysis (Figure\ \@ref(fig:extended-series)a.1, p-value `r trends[part == "Real" & PC == "PC1"]$p_report`).
-There is no significant trend in any of the complex parts of EOF2.
+Figure\ \@ref(fig:extended-series) shows time series of the two complex EOFs extended beyond the satellite era using the preliminary ERA5 back extension going back to 1950.There is an upward trend in the real part of EOF1 (Figure\ \@ref(fig:extended-series)a.1, p-value `r trends[part == "Real" & PC == "PC1"]$p_report`) and no significant trend in any of the complex parts of EOF2.
+The positive trend in the Real EOF1 translates into a positive trend in EOF1 magnitude, but not in phase (not shown).
 
@@
 
-```{r define-plot_regr}
+```{r plot_regr}
 plot_regr <- function(data, which_PC, which_levs, which_var) {
@@
     scale_fill_divergent(var_title, breaks = AnchorBreaks(0, exclude = 0),
-                         guide = guide_colorstrip_bottom(20)) +
+                         guide = guide_colorstrip_bottom(15)) +
     geom_contour_pval(aes(z = p.value)) +
@@
 
-```{r eof1-regr-gh, fig.cap = "(ref:eof1-regr-gh-cap)", fig.height=5.4, fig.width=5}
-plot_regr(regr, "PC1", c(200, 50), "hgt") +
+```{r eof1-regr-gh, fig.cap = "(ref:eof1-regr-gh-cap)", fig.height=1.2*width_column, fig.width=width_column, dependson="plot_regr"}
+plot_regr(regr , "PC1", c(200, 50), "hgt") +
   # geom_contour2(data = ~copy(.x)[, estimate := Anomaly(estimate), by = .(term, lev, lat)],
@@
 
-```{r eof2-regr-gh, fig.cap = "(ref:eof2-regr-gh-cap)", fig.height=5.4, fig.width=5}
+```{r eof2-regr-gh, fig.cap = "(ref:eof2-regr-gh-cap)", fig.height=1.3*width_column, fig.width=width_column, dependson="plot_regr"}
 plot_regr(regr, "PC2", c(200, 50), "hgt")
@@
 
-```{r eof1-regr-t, fig.cap = "(ref:eof1-regr-t-cap)",  fig.height=5.4, fig.width=5}
+```{r eof1-regr-t, fig.cap = "(ref:eof1-regr-t-cap)",  dependson="plot_regr", fig.height=1.3*width_column, fig.width=width_column}
 plot_regr(regr, "PC1", c(200, 50), "t")
@@
 
-```{r t-vertical, fig.cap = "(ref:t-vertical-cap)", fig.height=3.5, fig.width=5}
+```{r t-vertical, fig.cap = "(ref:t-vertical-cap)", fig.height=2.5, fig.width=width_column}
 t %>% 
@@
                 aes(z = estimate*1e9, label = signif(..level.., 2)),
-                size = 3,
+                size = 2,
                 skip = 1, stroke = 0.15, stroke.colour = "white", 
@@
 
-  scale_fill_divergent_discretised("Air temperature", guide = guide_colorsteps_bottom(20)) + 
+  scale_fill_divergent_discretised("Air temperature", guide = guide_colorsteps_bottom(15)) + 
   scale_y_level() +
@@
 
-```{r o3-regr, fig.cap = "(ref:o3-regr-cap)", fig.height=4, fig.width = 5}
-
+```{r o3-regr, fig.cap = "(ref:o3-regr-cap)", fig.height=3, fig.width = width_column}
 o3_regr %>% 
@@
   geom_contour2(data = periodic(o3_mean, lon = c(0, 360)),
-                aes(z = toc_z, linetype = factor(-sign(..level..))), breaks = AnchorBreaks(0, exclude = 0), 
+                aes(z = toc_z, linetype = factor(-sign(..level..))), 
+                breaks = AnchorBreaks(0, exclude = 0), 
                 bins = 5) +
@@
   scale_fill_divergent_discretised("Total Ozone Column", 
-                                   guide = guide_colorsteps_bottom(20)) + 
+                                   guide = guide_colorsteps_bottom(15)) + 
   scale_x_longitude() +
@@
 
-```{r wave1-o3, fig.cap = "(ref:wave1-o3-cap)", fig.width=5, fig.height=3}
+```{r wave1-o3, fig.cap = "(ref:wave1-o3-cap)", fig.width=width_column, fig.height=width_column*.7}
 wave1_o3 <- o3 %>% 
@@
 
-```{r psa-eof2, fig.cap = "(ref:psa-eof2-cap)", fig.width=5, fig.height=4}
+```{r psa-eof2, fig.cap = "(ref:psa-eof2-cap)", fig.width=7, fig.height=3, fig.env="figure*"}
 pc2 <- ceof$eof[[1]]$left[PC == "PC2"] %>% 
@@
 
-```{r phase-histogram, fig.cap = "(ref:phase-histogram-cap)", fig.width=2.5, fig.height=2.5}
+```{r phase-histogram, fig.cap = "(ref:phase-histogram-cap)", fig.width=width_column*.8, fig.height=2.5}
 cut(ceof$eof[[1]], 2)$left %>% 
@@
 
-```{r sam-eof-vertical, fig.cap = "(ref:sam-eof-vertical-cap)", fig.height=4, fig.width=5}
+```{r sam-eof-vertical, fig.cap = "(ref:sam-eof-vertical-cap)", fig.height=4, fig.width=width_column}
 time <- ceof[, eof[[1]]$left, by = .(season)] %>%
@@
 ```{r eof-filtered, fig.cap = "Spatial pattern of the leading EOF of 200 hPa geopotential height with the variability of EOF2 filtered out South of 20º.
-Arbitrary scale.", fig.width=2.5, fig.height=2.5}
+Arbitrary units.", fig.width=2.5, fig.height=2.5}
 eof_filtered <- era5 %>% 
@@
 
-```{r fake-eof, fig.cap = "(ref:fake-eof-cap)", fig.width=5, fig.height=2.5}
+```{r fake_eof}
 pattern <- eof_filtered$right %>% 
@@
   EOF(eof ~ lon + lat | time, n = 1:3, data = .)  
-
+```
+
+```{r fake-eof, fig.cap = "(ref:fake-eof-cap)", fig.width=3.7, fig.height=1.7}
 # era5 %>% 
@@
 
-```{r sst-psi-2,  fig.cap = "(ref:sst-psi-2-cap)", fig.width = 6, fig.height=4, dependson="plot_sst_psi"}
+```{r sst-psi-2,  fig.cap = "(ref:sst-psi-2-cap)", fig.width = 7, fig.height=4, dependson="plot_sst_psi", fig.env="figure*"}
 plot_sst_psi(which_pc = "PC2", which_lev = 200)
@@
 Colours denote years with magnitude of EOF2 greater or smaller than the 50th percentile.
-Black line is the fit ONI ~ cos(phase) + sin(phase) computed by OLS weighted by the magnitude of EOF2.", fig.width = 5, fig.height=3}
+Black line is the fit ONI ~ cos(phase) + sin(phase) computed by OLS weighted by the magnitude of EOF2.", fig.width = width_column, fig.height=3}
 gdata <- ceof[, eof[[1]]$left, by = .(season)] %>%
@@
 
-```{r sst-psi-1, fig.cap = "(ref:sst-psi-1-cap)", fig.width = 6, fig.height=4, dependson="plot_sst_psi"}
+```{r sst-psi-1, fig.cap = "(ref:sst-psi-1-cap)", fig.width = 7, fig.height=4, dependson="plot_sst_psi", fig.env = "figure*"}
 plot_sst_psi(which_pc = "PC1", which_lev = 200)
@@
 
-```{r pp-america, fig.cap = "(ref:pp-america-cap)", fig.width=5, fig.height=6}
+```{r pp-america, fig.cap = "(ref:pp-america-cap)", fig.width=width_column, fig.height=width_column*1.1}
 plot_pp(data = regr_pp, which_pc = c("PC1", "PC2"), which_continent = "america") 
@@
 
-```{r pp-oceania, fig.cap = "(ref:pp-oceania-cap)", fig.width=5, fig.height=4}
+```{r pp-oceania, fig.cap = "(ref:pp-oceania-cap)", fig.width=width_column, fig.height=width_column*.8}
 plot_pp(regr_pp, c("PC1", "PC2"), which_continent =  "oceania") +
@@
 
-```{r australia-pp-phase, fig.cap = "(ref:australia-pp-phase-cap)", fig.width = 5, fig.height=3}
+```{r australia-pp-phase, fig.cap = "(ref:australia-pp-phase-cap)", fig.width = width_column, fig.height=3}
 gdata <- cmap %>% 

### 2021-03-13 analysis/paper/paper.tex
Corrige tamaño de las figuras

@@
 We used monthly geopotential height, air temperature, ozone mixing ratio, and Total Ozone COlumn (TOC) at 2.5\degree~longitude by 2.5\degree~latitude of horizontal resolution and 37 vertical isobaric levels from ERA5 \citep{era5} for the period 1979 to 2019.
-We restrict most of our analysis to the post-satellite era to avoid any confounding factors arising from the incorporation of satellite observations, but we also use the preliminary back extension of ERA5 from 1950 to 1978 \citep{era5be} to look at long-term trends.
+We restricted most of our analysis to the post-satellite era to avoid any confounding factors arising from the incorporation of satellite observations, but we also used the preliminary back extension of ERA5 from 1950 to 1978 \citep{era5be} to look at long-term trends.
 We derived streamfunction at 200 hPa from ERA5 vorticity using the FORTRAN subroutine FISHPACK \citep{fishpack} and computed horizontal wave activity fluxes following \citet{plumb1985} (Equation 5.7).
+For precipitation data we used monthly data from the CPC Merged Analysis of Precipitation \citep{cmap}.
 
-Precipitation data we used monthly data from the CPC Merged Analysis of Precipitation \citep{cmap}.
+We computed the amplitude and phase of the zonal wave 1 by averaging variables (temperature, geopotential height, ozone mixing ratio and Total Ozone Column) 70°S and 45°S for each Spring and computing the Fourier spectrum.
 
-We compute the amplitude and phase of the zonal wave 1 by averaging SON TOC between 70°S and 45°S for each Spring and computing the Fourier spectrum.
-
-We explore the relationship with the Southern Annular Mode (SAM).
-We compute the level-dependent SAM index as the leading EOF of year-round monthly geopotential height anomalies south of 20ºS at each level for the whole period \citep{baldwin2009}.
+We computed the level-dependent Southern Annular Mode index as the leading EOF of year-round monthly geopotential height anomalies south of 20ºS at each level for the whole period \citep{baldwin2009}.
 We further split the SAM into its zonally symmetric and zonally asymmetric components (S-SAM and A-SAM indices respectively).
@@
 \caption{\label{fig:eof-naive}Spatial patterns of the leading 4 EOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2009 (arbitrary units).
-The numbers at the top-left of each panel ate the variance explained by each EOF.}
+The numbers at the top-left of each panel ate the zonal variance explained by each EOF.}
 \end{figure}
@@
 
-\begin{figure}
-\centering
-\includegraphics{../figures/extended-series-1.pdf}
-\caption{\label{fig:extended-series}Time series extended using ERA5 back extended preliminary edition (period 1950 -- 1978) and ERA5 (period 1979 -- 2019).
-Each series is computed by projecting monthly geopotential height zonal anomalies standardised by level south of 20ºS onto the corresponding spatial pattern and scaled to zero mean and unit standard deviation.}
-\end{figure}
+\begin{figure*}
+\includegraphics{../figures/extended-series-1} \caption{Time series extended using ERA5 back extended preliminary edition (period 1950 -- 1978) and ERA5 (period 1979 -- 2019).
+Each series is computed by projecting monthly geopotential height zonal anomalies standardised by level south of 20ºS onto the corresponding spatial pattern and scaled to zero mean and unit standard deviation.}\label{fig:extended-series}
+\end{figure*}
 
-Figure~\ref{fig:extended-series} shows time series of the two complex EOFs extended beyond the satellite era using the preliminary ERA5 back extension going back to 1950 (which we call the ``hybrid ERA5'' reanalysis).
-There is a downward trend in the real part of EOF1 in the hybrid reanalysis (Figure~\ref{fig:extended-series}a.1, p-value \textless{} 0.001).
-There is no significant trend in any of the complex parts of EOF2.
+Figure~\ref{fig:extended-series} shows time series of the two complex EOFs extended beyond the satellite era using the preliminary ERA5 back extension going back to 1950.There is an upward trend in the real part of EOF1 (Figure~\ref{fig:extended-series}a.1, p-value \textless{} 0.001) and no significant trend in any of the complex parts of EOF2.
+The positive trend in the Real EOF1 translates into a positive trend in EOF1 magnitude, but not in phase (not shown).
 
@@
 
-\begin{figure}
-\centering
-\includegraphics{../figures/psa-eof2-1.pdf}
-\caption{\label{fig:psa-eof2}Relationship between the Real and Imaginary parts of EOF2 and the PSA1 and PSA2 modes computed as the second and third EOFs of seasonal geopotential height anomalies \citep{mo2001}.
-For the period 1979 -- 2019.}
-\end{figure}
+\begin{figure*}
+\includegraphics{../figures/psa-eof2-1} \caption{Relationship between the Real and Imaginary parts of EOF2 and the PSA1 and PSA2 modes computed as the second and third EOFs of seasonal geopotential height anomalies \citep{mo2001}.
+For the period 1979 -- 2019.}\label{fig:psa-eof2}
+\end{figure*}
 
@@
 \includegraphics{../figures/eof-filtered-1.pdf}
-\caption{\label{fig:eof-filtered}Spatial pattern of the leading EOF of 200 hPa geopotential height with the variability of EOF2 filtered out South of 20º. Arbitrary scale.}
+\caption{\label{fig:eof-filtered}Spatial pattern of the leading EOF of 200 hPa geopotential height with the variability of EOF2 filtered out South of 20º. Arbitrary units.}
 \end{figure}
@@
 
-\begin{figure}
-\centering
-\includegraphics{../figures/sst-psi-2-1.pdf}
-\caption{\label{fig:sst-psi-2}Regression maps of EOF2 with SST (column a) and streamfunction zonal anomalies with their corresponding activity wave flux (column b).
-Areas marked with dots have p-values smaller than 0.05 adjusted for FDR.}
-\end{figure}
+\begin{figure*}
+\includegraphics{../figures/sst-psi-2-1} \caption{Regression maps of EOF2 with SST (column a) and streamfunction zonal anomalies with their corresponding activity wave flux (column b).
+Areas marked with dots have p-values smaller than 0.05 adjusted for FDR.}\label{fig:sst-psi-2}
+\end{figure*}
 
@@
 
-\begin{figure}
-\centering
-\includegraphics{../figures/sst-psi-1-1.pdf}
-\caption{\label{fig:sst-psi-1}Same as Figure~\ref{fig:sst-psi-2} but for EOF1.}
-\end{figure}
+\begin{figure*}
+\includegraphics{../figures/sst-psi-1-1} \caption{Same as Figure~\ref{fig:sst-psi-2} but for EOF1.}\label{fig:sst-psi-1}
+\end{figure*}
 

### 2021-03-22 analysis/paper/paper.Rmd
Conclusion e intro

@@
  - \linenumbers 
+ - \usepackage{pdfcomment}
 ---
@@
 
+One typical way of describing the zonally asymmetric circulation is by the amplitude and phase of zonal waves obtained by Fourier decomposition of geopotential height at each latitude [e.g.
+@vanloon1972;@trenberth1980].
+These methodologies rely on the assumption that the circulation can be meaningfully understood in terms of zonal waves of constant amplitude along a latitude circle.
+This is not valid for meridionally propagating waves or zonal waves with localised amplitudes.
+In the case of the wave 3, for example, @trenberth1985 observed that wave 3 played a role in blocking events, but that in many cases, this was due to increased amplitude of a longitudinally localised wave train and not a hemispheric-scale zonal wave 3.
+
+Another approach to characterising the Southern Hemisphere circulation is by using Empirical Orthogonal Functions (EOF, also known as Principal Component Analysis).
+Within the EOF framework, the Southern Annular Mode appears as the leading mode of variability of the Southern Hemisphere (SAM) [@fogt2020] followed by the two Pacific–South American Patterns (PSA) [@mo2001].
+The SAM represents a relatively zonally symmetric pattern of alternating low pressures in polar latitude and a ring of high pressures in high latitudes with an embedded wave 3 pattern that is more prominent in the Pacific sector.
+The PSA1 and PSA2 describe the two out-of-phase sides of a meridionally propagating wave train that originates in the Eastern equatorial Pacific and travels towards the South Atlantic following a great-circle arch along the Antarctic Peninsula.
+These patterns are derived by applying EOF to temporal anomalies, but @raphael2003 applied EOF methods specifically to zonal anomalies.
+
+EOFs are more flexible in the sense that they can characterise a richer set of spatial patterns, but they are restricted to standing oscillation modes and have difficulty dealing with propagating or phase-varying modes such as zonal waves.
+A single EOF can also represent a mixture of two or more physical modes.
+
+The inability of a single EOF to describe travelling waves is what forces the PSA pattern to be described by two EOFs.
+@irving2016 characterises the Pacific–South American Pattern in a Fourier framework.
+Being a meridionally propagating mode, the PSA cannot be correctly characterised by Fourier decomposition at each latitude circle, so they reprojected meridional wind fields so that the path of the PSA laid on the equator.
+On this new projection, they could identify the PSA using Fourier decomposition.
+
+Finally, another, hybrid, methodology consist on identifying particular features of interest and creating indices using simple methods such as averages and differences.
+The Marshal SAM Index [@marshall2003], for example, tries to characterise the SAM by the average difference of sea level pressure between 40ºS and 65ºS, following @gong1999.
+@raphael2004 used a similar methodology to describe the wave 3 in the Southern Hemisphere.
+Instead of using Fourier to compute the amplitude and phase of the zonal wave, they averaged standardised geopotential height anomalies in three points representing the location of the ridges of the zonal wave 3 of the climatological mean.
+This is equivalent to an index of the amplitude of the zonal wave 3 that projects into the phase of the zonal wave 3 of the mean field.
+Similarly, @hobbs2010 posit that the zonally asymmetric circulation in the Southern Hemisphere can be described by the strength and location of two anticyclones located in the sub-Antarctic western hemisphere.
+
+These "feature-based" methods are grounded on other methods to identify the interesting features and can be useful to characterise features that are not readily apparent with Fourier or EOF methods but that nonetheless interesting or important.
+The use of data of single points, which makes them suitable for extending into the data-poor past also makes them not very robust.
+
+In this paper we propose using complex Empirical Orthogonal Functions [@horel1984] as a robust extension of EOF analysis that is capable of describing rich structures with meridional and zonal propagation, and amplitude modulation such as the PSA.
+We apply this method to characterise the Springtime zonally asymmetric circulation of the extratopical Southern Hemisphere.
+
+\pdfcomment[]{¿Agrego algo sobre impactos? Por ejemplo, impactos de PSA /SAM.}
+
 # Methods
@@
 Both observations motivate the decision of performing complex EOF jointly between levels.
+Since the variability of the stratosphere is far greater in magnitude than the variability of the troposphere, we standardised the values by the standard deviation of each level.
 The computation of the EOFs was carried out using data from both levels at the same time, therefore, each complex EOF has a spatial part that depends on longitude, latitude and level, an a temporal part that depends only on time.
@@
 
-Figure\ \@ref(fig:extended-series) shows time series of the two complex EOFs extended beyond the satellite era using the preliminary ERA5 back extension going back to 1950.There is an upward trend in the real part of EOF1 (Figure\ \@ref(fig:extended-series)a.1, p-value `r trends[part == "Real" & PC == "PC1"]$p_report`) and no significant trend in any of the complex parts of EOF2.
+Figure\ \@ref(fig:extended-series) shows time series of the two complex EOFs extended beyond the satellite era using the preliminary ERA5 back extension going back to 1950.There is an upward trend in the real part of EOF1 (Figure\ \@ref(fig:extended-series)a.1, p-value\ `r trends[part == "Real" & PC == "PC1"]$p_report`) and no significant trend in any of the complex parts of EOF2.
 The positive trend in the Real EOF1 translates into a positive trend in EOF1 magnitude, but not in phase (not shown).
@@
 (ref:o3-regr-cap) Regression coefficients of the real and imaginary part of the first complex EOF on SON mean Total Ozone Column.
-On contours, the mean zonal anomaly of Total OZone Column (negative contours in dashed lines).
+On contours, the mean zonal anomaly of Total Ozone Column (negative contours in dashed lines).
 For  the 1979 -- 2019 period.
@@
 By aligning he imaginary phase with the direction with maximum relationship with ENSO, we also aligned it with the phase of maximum occurrence.
-This is analogous to the method in @irving2016, who used reprojection and Fourier filtering to detect "PSA-like" variability and defined as the proper PSA using the peaks of the phase distribution (Compare our Figure\ \@ref:(phase-histogram) with their Figure 6).
+This is analogous to the method in @irving2016, who used reprojection and Fourier filtering to detect "PSA-like" variability and defined as the proper PSA using the peaks of the phase distribution (Compare our Figure\ \@ref(fig:phase-histogram) with their Figure 6).
 
@@
 (ref:sam-eof-vertical-cap) Coefficient of determination between the real and imaginary part of each EOF and the SAM, Asymmetric SAM (A-SAM) and Symmetric SAM (S-SAM) indices computed at each level according to @campitelli2021.
-Points mark estimates with p-value < 0.01 corrected for False Detection Rate [@benjamini1995].
+Points mark estimates with p-value\ <\ 0.01 corrected for False Detection Rate [@benjamini1995].
 
@@
                     arrow.angle = 14, arrow.length = 0.2) +
+    # geom_vector(aes(dx = fx*1e20, dy = fy*1e20), min.mag = 1e-10,
+    #             skip = 3,
+    #                 arrow.angle = 14, arrow.length = 0.2) +
+    # scale_mag() +
+
     geom_qmap(keep = .015) +
@@
     scale_y_latitude() +
-    scale_mag(guide = "none") +
+
     facet_grid(term ~ var, labeller = labeller(PC = c(PC1 = "EOF1", PC2 = "EOF2"))) +
@@
 
+We studied the two leading modes of co-variability of September-October-November tropospheric and stratospheric zonal anomaliesby applying complex Empirical Orthogonal Functions.
+
+The first complex EOF represents the variability of the zonal wave 1 in the stratosphere, but represents a more zonally symmetric monopole in the troposphere.
+There is a statistically positive trend in the magnitude of this EOF, which is consistent with previous studies which showed secular changes in springtime wave-1-like patterns [e.g.
+@raphael2003].
+This mode is closely related to anomalies in Total Ozone Column.
+
+THe second complex EOF represents a wave-3 pattern with maximum magnitude in the Pacific sector.
+Essentially, it is an alternate representation of the PSA1 and PSA2 patterns [@mo2001].
+We show that the Imaginary EOF2 can be identified with the PSA1 and the Real EOF2 with the PSA2.
+
+We show that this EOF prefers a phase aligned with the Imaginary EOF2.
+This is because the when ENSO is active, it forces the EOF2 to be in the Imaginary phase; positive for El Niño and negative for La Niña.
+When ENSO is neutral, the EOF2 is still active, but with no particular phase.
+This mirrors the results of @cai2002, who showed that the CSIRO Model can create PSA-like variability even in the absence of ENSO forcing (with a climatological run), but the variability of one of the PSA modes was enhanced when adding the ENSO signal.
+
+We further show that the Imaginary EOF2 is closely related to the Southern Annular Mode in the troposphere.
+In fact, it is a virtually perfect indicator of the zonally asymmetric portion of the SAM.
+To the extreme that the leading EOF of geopotential height with the EOF2 removes results in an annular mode that has none of the typical zonal asymmetries of the observed SAM.
+
+This raises the possibility that the asymmetric component of the SAM is actually a statistical contamination of the PSA mode.
+
+Precipitation anomalies in South America associated with the Imaginary EOF2 show a clear ENSO-like impact, with positive anomalies in South-Eastern Southamerica, negative anomalies in Southern Brazil and positive anomalies in central Chile.
+Precipitation anomalies associated with the Real EOF2 are low and not statistically significant, showing that the Imaginary phase is optimally aligned with the direction of maximum precipitation impacts.
+
+On the other hand, over Australia, both the Real and Imaginary phase are associated with significant precipitation anomalies, and we further show that the direction of maximum impact is not aligned with our chosen rotation of EOF2.
+However, this underscore the benefit of using complex EOF, since it would trivial to rotate it.
+
 ## Code availability {-}

### 2021-03-22 analysis/paper/paper.tex
Conclusion e intro

@@
 \linenumbers
+\usepackage{pdfcomment}
 \usepackage{booktabs}
@@
 \section{Introduction}\label{introduction}}
+
+One typical way of describing the zonally asymmetric circulation is by the amplitude and phase of zonal waves obtained by Fourier decomposition of geopotential height at each latitude \citep[e.g.][]{vanloon1972, trenberth1980}.
+These methodologies rely on the assumption that the circulation can be meaningfully understood in terms of zonal waves of constant amplitude along a latitude circle.
+This is not valid for meridionally propagating waves or zonal waves with localised amplitudes.
+In the case of the wave 3, for example, \citet{trenberth1985} observed that wave 3 played a role in blocking events, but that in many cases, this was due to increased amplitude of a longitudinally localised wave train and not a hemispheric-scale zonal wave 3.
+
+Another approach to characterising the Southern Hemisphere circulation is by using Empirical Orthogonal Functions (EOF, also known as Principal Component Analysis).
+Within the EOF framework, the Southern Annular Mode appears as the leading mode of variability of the Southern Hemisphere (SAM) \citep{fogt2020} followed by the two Pacific--South American Patterns (PSA) \citep{mo2001}.
+The SAM represents a relatively zonally symmetric pattern of alternating low pressures in polar latitude and a ring of high pressures in high latitudes with an embedded wave 3 pattern that is more prominent in the Pacific sector.
+The PSA1 and PSA2 describe the two out-of-phase sides of a meridionally propagating wave train that originates in the Eastern equatorial Pacific and travels towards the South Atlantic following a great-circle arch along the Antarctic Peninsula.
+These patterns are derived by applying EOF to temporal anomalies, but \citet{raphael2003} applied EOF methods specifically to zonal anomalies.
+
+EOFs are more flexible in the sense that they can characterise a richer set of spatial patterns, but they are restricted to standing oscillation modes and have difficulty dealing with propagating or phase-varying modes such as zonal waves.
+A single EOF can also represent a mixture of two or more physical modes.
+
+The inability of a single EOF to describe travelling waves is what forces the PSA pattern to be described by two EOFs.
+\citet{irving2016} characterises the Pacific--South American Pattern in a Fourier framework.
+Being a meridionally propagating mode, the PSA cannot be correctly characterised by Fourier decomposition at each latitude circle, so they reprojected meridional wind fields so that the path of the PSA laid on the equator.
+On this new projection, they could identify the PSA using Fourier decomposition.
+
+Finally, another, hybrid, methodology consist on identifying particular features of interest and creating indices using simple methods such as averages and differences.
+The Marshal SAM Index \citep{marshall2003}, for example, tries to characterise the SAM by the average difference of sea level pressure between 40ºS and 65ºS, following \citet{gong1999}.
+\citet{raphael2004} used a similar methodology to describe the wave 3 in the Southern Hemisphere.
+Instead of using Fourier to compute the amplitude and phase of the zonal wave, they averaged standardised geopotential height anomalies in three points representing the location of the ridges of the zonal wave 3 of the climatological mean.
+This is equivalent to an index of the amplitude of the zonal wave 3 that projects into the phase of the zonal wave 3 of the mean field.
+Similarly, \citet{hobbs2010} posit that the zonally asymmetric circulation in the Southern Hemisphere can be described by the strength and location of two anticyclones located in the sub-Antarctic western hemisphere.
+
+These ``feature-based'' methods are grounded on other methods to identify the interesting features and can be useful to characterise features that are not readily apparent with Fourier or EOF methods but that nonetheless interesting or important.
+The use of data of single points, which makes them suitable for extending into the data-poor past also makes them not very robust.
+
+In this paper we propose using complex Empirical Orthogonal Functions \citep{horel1984} as a robust extension of EOF analysis that is capable of describing rich structures with meridional and zonal propagation, and amplitude modulation such as the PSA.
+We apply this method to characterise the Springtime zonally asymmetric circulation of the extratopical Southern Hemisphere.
+
+\pdfcomment[]{¿Agrego algo sobre impactos? Por ejemplo, impactos de PSA /SAM.}
 
@@
 
-Figure~\ref{fig:extended-series} shows time series of the two complex EOFs extended beyond the satellite era using the preliminary ERA5 back extension going back to 1950.There is an upward trend in the real part of EOF1 (Figure~\ref{fig:extended-series}a.1, p-value \textless{} 0.001) and no significant trend in any of the complex parts of EOF2.
+Figure~\ref{fig:extended-series} shows time series of the two complex EOFs extended beyond the satellite era using the preliminary ERA5 back extension going back to 1950.There is an upward trend in the real part of EOF1 (Figure~\ref{fig:extended-series}a.1, p-value~ \textless{} 0.001) and no significant trend in any of the complex parts of EOF2.
 The positive trend in the Real EOF1 translates into a positive trend in EOF1 magnitude, but not in phase (not shown).
@@
 \caption{\label{fig:o3-regr}Regression coefficients of the real and imaginary part of the first complex EOF on SON mean Total Ozone Column.
-On contours, the mean zonal anomaly of Total OZone Column (negative contours in dashed lines).
+On contours, the mean zonal anomaly of Total Ozone Column (negative contours in dashed lines).
 For the 1979 -- 2019 period.}
@@
 By aligning he imaginary phase with the direction with maximum relationship with ENSO, we also aligned it with the phase of maximum occurrence.
-This is analogous to the method in \citet{irving2016}, who used reprojection and Fourier filtering to detect ``PSA-like'' variability and defined as the proper PSA using the peaks of the phase distribution (Compare our Figure~@ref:(phase-histogram) with their Figure 6).
+This is analogous to the method in \citet{irving2016}, who used reprojection and Fourier filtering to detect ``PSA-like'' variability and defined as the proper PSA using the peaks of the phase distribution (Compare our Figure~\ref{fig:phase-histogram} with their Figure 6).
 
@@
 \caption{\label{fig:sam-eof-vertical}Coefficient of determination between the real and imaginary part of each EOF and the SAM, Asymmetric SAM (A-SAM) and Symmetric SAM (S-SAM) indices computed at each level according to \citet{campitelli2021}.
-Points mark estimates with p-value \textless{} 0.01 corrected for False Detection Rate \citep{benjamini1995}.}
+Points mark estimates with p-value~\textless~0.01 corrected for False Detection Rate \citep{benjamini1995}.}
 \end{figure}
@@
 If those years are removed, the correlation becomes non-significant (0.044 (CI: -0.28 -- 0.36)).
-Furthermore, even when using all the datapoints, the Spearman correlation --which is robust to outliers-- is also non-significant (0.2, p-value = 0.2).
+Furthermore, even when using all the datapoints, the Spearman correlation --which is robust to outliers-- is also non-significant (0.2, p-value= 0.2).
 Therefore, the relationship between the magnitude of the EOF2 train wave and ENSO remains uncertain.
@@
 
+We studied the two leading modes of co-variability of September-October-November tropospheric and stratospheric zonal anomaliesby applying complex Empirical Orthogonal Functions.
+
+The first complex EOF represents the variability of the zonal wave 1 in the stratosphere, but represents a more zonally symmetric monopole in the troposphere.
+There is a statistically positive trend in the magnitude of this EOF, which is consistent with previous studies which showed secular changes in springtime wave-1-like patterns \citep[e.g.][]{raphael2003}.
+This mode is closely related to anomalies in Total Ozone Column.
+
+THe second complex EOF represents a wave-3 pattern with maximum magnitude in the Pacific sector.
+Essentially, it is an alternate representation of the PSA1 and PSA2 patterns \citep{mo2001}.
+We show that the Imaginary EOF2 can be identified with the PSA1 and the Real EOF2 with the PSA2.
+
+We show that this EOF prefers a phase aligned with the Imaginary EOF2.
+This is because the when ENSO is active, it forces the EOF2 to be in the Imaginary phase; positive for El Niño and negative for La Niña.
+When ENSO is neutral, the EOF2 is still active, but with no particular phase.
+This mirrors the results of \citet{cai2002}, who showed that the CSIRO Model can create PSA-like variability even in the absence of ENSO forcing (with a climatological run), but the variability of one of the PSA modes was enhanced when adding the ENSO signal.
+
+We further show that the Imaginary EOF2 is closely related to the Southern Annular Mode in the troposphere.
+In fact, it is a virtually perfect indicator of the zonally asymmetric portion of the SAM.
+To the extreme that the leading EOF of geopotential height with the EOF2 removes results in an annular mode that has none of the typical zonal asymmetries of the observed SAM.
+
+This raises the possibility that the asymmetric component of the SAM is actually a statistical contamination of the PSA mode.
+
+Precipitation anomalies in South America associated with the Imaginary EOF2 show a clear ENSO-like impact, with positive anomalies in South-Eastern Southamerica, negative anomalies in Southern Brazil and positive anomalies in central Chile.
+Precipitation anomalies associated with the Real EOF2 are low and not statistically significant, showing that the Imaginary phase is optimally aligned with the direction of maximum precipitation impacts.
+
+On the other hand, over Australia, both the Real and Imaginary phase are associated with significant precipitation anomalies, and we further show that the direction of maximum impact is not aligned with our chosen rotation of EOF2.
+However, this underscore the benefit of using complex EOF, since it would trivial to rotate it.
+
 \hypertarget{code-availability}{%

### 2021-03-22 R/lm_lite.R

@@ -75,7 +75,7 @@ Pvaluate <- function(estimate, std.error, df, adjustment = "none") {
 #'
 #' @export
 report_p <- function(p) {
-  ifelse(p < 0.001, "\ <\ 0.001", paste0("\ =\ ", signif(p, 2)))
+  ifelse(p < 0.001, "<\ 0.001", paste0("=\ ", signif(p, 2)))
 }
 
 #' Remove intercept

### 2021-03-23 analysis/paper/paper.Rmd
Comentarios Lean 1ra version

--- 2021-03-13 analysis/paper/paper.Rmd
@@
 
-# Methods
+# Data and Methods
 
@@
 We restricted most of our analysis to the post-satellite era to avoid any confounding factors arising from the incorporation of satellite observations, but we also used the preliminary back extension of ERA5 from 1950 to 1978 [@era5be] to look at long-term trends.
-We derived streamfunction at 200 hPa from ERA5 vorticity using the FORTRAN subroutine FISHPACK [@fishpack] and computed horizontal wave activity fluxes following @plumb1985 (Equation 5.7).
-For precipitation data we used monthly data from the CPC Merged Analysis of Precipitation [@cmap].
+We derived streamfunction at 200 hPa from ERA5 vorticity using the FORTRAN subroutine FISHPACK [@fishpack] and computed horizontal wave activity fluxes following @plumb1985.
+For precipitation data we used monthly data from the CPC Merged Analysis of Precipitation [@cmap], with a 2.5\degree\ resolution in latitude and longitude.
+This rainfall gridded dataset is based on information from different sources such as rain gauge observations, satellite
+inferred estimations and the NCEP-NCAR reanalysis, and it is available since 1979 to present.
+
+## Methods
+
+\textcolor{red}{Aclarar qué a lo largo del trabajo se utilizan los datos de primavera y cómo se calcula eso.
+Tambień creo que acá se podría aclarar que el análisis se concentra en dos niveles para representar la estratósfera y la troposféra alta}
 
@@
 
-We computed the level-dependent Southern Annular Mode index as the leading EOF of year-round monthly geopotential height anomalies south of 20ºS at each level for the whole period [@baldwin2009].
+We computed the level-dependent Southern Annular Mode (SAM) index as the leading EOF of year-round monthly geopotential height anomalies south of 20ºS at each level for the whole period [@baldwin2009].
 We further split the SAM into its zonally symmetric and zonally asymmetric components (S-SAM and A-SAM indices respectively).
@@
 
-## Complex Empirical Orthogonal Functions
-
-(ref:eof-naive-cap) Spatial patterns of the leading 4 EOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2009 (arbitrary units).
+\textcolor{red}{Creo que habría que expandir un poco esta parte e ir comentando todoas las metodologías que vas usando: regresiones, tendencias, cómo evlauas significancia en cada caso, etc. }
+
+## Complex Empirical Orthogonal Functions (cEOF)
+
+(ref:eof-naive-cap) Spatial patterns of the leading 4 EOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2019 (arbitrary units).
 The numbers at the top-left of each panel ate the zonal variance explained by each EOF.
@@
 
-In traditional EOF analysis zonal waves appear as pairs of EOFs, usually degenerate, that represent similar patterns but shifted in phase.
-For instance, Figure\ \@ref(fig:eof-naive) shows the leading 4 EOFs of zonally anomalous September through October geopotential height at 50 hPa and 100 hPa.
-It is clear that at 50 hPa the first two EFS represent the same zonal wave 1 pattern and the last 2 represent the same zonal wave 3 pattern shifted by 1/4 wavelength.
+In traditional EOF analysis zonal waves appear as pairs of EOFs, usually degenerate, that represent similar patterns but shifted in phase \textcolor{red}{(Agregaría algunas referencias)}.
+For instance, Figure\ \@ref(fig:eof-naive) shows the leading 4 EOFs of zonally anomalous September through October \textcolor{red}{(November supongo.
+Sería bueno que ya arriba indiques lo de SON y ya después lo menciones directamente así)} geopotential height at 50 hPa and 200 hPa.
+It is clear that at 50 hPa the first two EOFs represent the same zonal wave 1 pattern and the last 2 represent the same zonal wave 3  \textcolor{red}{(No es visualmente obvia una estructura de onda 3.
+Solo se ven 4 centros)} pattern shifted by 1/4 wavelength.
+Similarly, EOFs 2 and 3 at 200 hPa represent the a zonal wave 3 pattern shifted by 1/4 wavelength.
 Since these pairs of EOFs seem to represent the same phase-varying structure, it would be desirable to combine them into a single index with amplitude and phase.
@@
 Complex Empirical Orthogonal Functions (cEOF) is a useful method to characterise these waves [@horel1984].
-This methods involves augmenting the original fields with the Hilbert transform of the data and then computing the Singular Value Decomposition of it.
+This method involves augmenting the original fields with the Hilbert transform of the data and then computing the Singular Value Decomposition of it.
 In this work, instead of applying the Hilbert transform to the time series at each point, we apply it to each latitude circle at each moment in time.
 Since the latitude circle is a periodic domain, this procedure does not suffer from edge effects.
-
 The result of the cEOF is a set of complex spatial patterns and complex time series.
@@
   setnames("item1", "200 hPa") %>%
-  kbl2(booktabs = TRUE, caption = "$R^2$ of the absolute magnitude of complex EOFs between 200 hPa and 50 hPa computing EOF separatedly for each level.") %>%
+  kbl2(booktabs = TRUE, caption = "Coefficient of determination ($R^2$) of the absolute magnitude of complex EOFs between 200 hPa and 50 hPa computing EOF separatedly for each level.") %>%
   column_spec2(2:4, background = spec_color2(.table[[.col]], option = background_palette,
@@
 
-Another characteristic seen in Figure\ \@ref(fig:eof-naive) is that EOF1 at 200 hPa is very similar to EFO1 at 50 hPa.
-The zonal wave 3 structure in EOF2 and EOF3 at 200 hPa are also very similar to the one in EOF3 and EOF4 at 50 hPa.
+Another characteristic seen in Figure\ \@ref(fig:eof-naive) is that EOF1 at 200 hPa is very similar to EOF1 at 50 hPa  \textcolor{red}{(En la figura que yo veo, no diría que no similares.
+Aclara mejor a qué te referis)}.
+The zonal wave 3 structure in EOF2 and EOF3 at 200 hPa are also very similar to the one in EOF3 and EOF4 at 50 hPa, excecpt for the behaviour above the Indian Ocean.
 Furthermore, Table \@ref(tab:corr-ceof-splitted) shows the coefficient of determination between time series of the amplitude of each complex EOF across levels.
@@
 Both the spatial similarities and temporal correlations suggest some level of joint variability between levels.
-
-Both observations motivate the decision of performing complex EOF jointly between levels.
-The computation of the EOFs was carried out using data from both levels at the same time, therefore, each complex EOF has a spatial part that depends on longitude, latitude and level, an a temporal part that depends only on time.
+\textcolor{red}{(Revisar el parráfo considerando el comentario)}
+
+Both observations \textcolor{red}{(¿Qué serían las dos osbervaciones?)}  motivate the decision of performing complex EOF jointly between levels.
+The computation of the EOFs was carried out using data from both levels at the same time, therefore, each complex EOF has a spatial component that depends on longitude, latitude and level, and a temporal component that depends only on time.
 
@@
 Since any choice is arbitrary and equally valid, we chose the phase of each EOF so that the real and imaginary parts are aligned with meaningful phases in our analysis.
-This was done after some exploratory analysis.
+This procedure does not create a spurious correlation, it only takes whatever relationship that already exist and aligns it with a specific phase.
 
@@
 
+\textcolor{red}{(No mostraría la figura 2. Me parece que agrega una complejidad extra y no sé entiende demasiado, sino se la explica muy bien)}
+
 For the first complex principal component, the phase was chosen so that the time series corresponding to the real part has the maximum correlation with the zonal wave 1 of Total Ozone Column between `r knitr::combine_words(LatLabel(o3wave_lats))`.
@@
 
-For the second complex principal component, the phase was chosen so that the coefficient of determination between the Oceanic Niño Index [@bamston1997] and the real part was minimised, which also nearly maximises the correlation with the imaginary part (Figure\ \@ref(fig:rotations)b).
-It's important to reassure the reader that this procedure does not create a spurious correlation, it only takes whatever relationship that already exist and aligns it with a specific phase.
-If the relationship is very weak, the rotation cannot generate a strong relationship.
+For the second complex principal component, the phase was chosen so that the coefficient of determination between the Oceanic Niño Index [@bamston1997] and the real part was minimized, which also nearly maximizes the correlation with the imaginary part (Figure\ \@ref(fig:rotations)b).
 
@@
 
-## Description of EOFs
+## Zonal anomalies circulation modes (no me convence del todo, pero debería ser algo así)
 
@@
 
-```{r ceofs-1, fig.cap = "Spatial patterns of the leading 2 complex EOFs of zonal anomalies of geopotential height in 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2009.
+```{r ceofs-1, fig.cap = "Spatial patterns of the leading 2 complex EOFs of zonal anomalies of geopotential height in 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2019.
 Real part in shading, imaginary part in contours", fig.width=width_column, fig.height=width_column}
@@
 
-```{r ceofs-2, fig.cap = "Time series of the leading 2 complex EOFs of zonal anomalies of geopotential height in 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2009.", fig.height=width_column, fig.width = width_column}
+```{r ceofs-2, fig.cap = "Time series of the leading 2 complex EOFs of zonal anomalies of geopotential height in 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2019.", fig.height=width_column, fig.width = width_column}
 ceof %>%
@@
 
-Figures \@ref(fig:ceofs-1) and \@ref(fig:ceofs-2) show, respectively, the spatial and temporal parts of the first two leading complex EOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa.
-In the spatial patterns in Figure\ \@ref(fig:ceofs-1), the real (in shading) and the imaginary parts (in contour) are in quadrature by construction, so that each EOF describe a single wave-like pattern whose amplitude and position (i.e.
+To describe the spatial and temporal variability of the circulation zonal anomalies, the spatial and temporal parts of the first two leading cEOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa are shown in Figures \@ref(fig:ceofs-1) and \@ref(fig:ceofs-2).
+The first mode (CEOF1) explains 82% of the variance, while the second mode (CEOF1) explains a smaller fraction (7%).
+In the spatial patterns (Fig.\ \@ref(fig:ceofs-1)), the real (in shading) and the imaginary components (in contour) are in quadrature by construction, so that each cEOF describe a single wave-like pattern whose amplitude and position (i.e.
 phase) is controlled by the magnitude and phase of the complex temporal EOF.
-
-The wave patterns described by these complex EOFs match the patterns seen in the traditional EOFs of Figure\ \@ref(fig:eof-naive): The first is a wave 1, while the second is a wave 3.
-Note that in Figure\ \@ref(fig:ceofs-2), both parts of the EOF1 have non zero mean.
-This is due to th fact that the geopotential fields that enter into the algorithm are anomalies with respect to the zonal mean, not the time mean.
-The variability associated with the first EOF includes variability that projects onto the mean zonally anomalous field.
+The wave patterns described by these cEOFs match the patterns seen in the traditional EOFs of Figure\ \@ref(fig:eof-naive): The first is mainly a wave 1, while the second is a wave 3.
+
+The temporal variability for both components of the cEOF1 have non zero mean (Fig.\ \@ref(fig:ceofs-2)).
+This is due to the fact that the geopotential fields that enter into the cEOFs algorithm are anomalies with respect to the zonal mean, not the time mean.
+As a consequence, the variability associated with the first cEOF includes variability that projects onto the mean zonally anomalous field.
+
+\textcolor{red}{(Acá creo que habría que ampliar un poco más la discusión.
+diciendo en más detalle lo que se ve en la parte espacial (donde se ven los centros más importantes, qué diferencias hay entre estratósfera y troposféra) y temporal, por ej.
+qué tipo de variabilidad se observa, si están correlacionadas)}
 
@@
 
-Figure\ \@ref(fig:extended-series) shows time series of the two complex EOFs extended beyond the satellite era using the preliminary ERA5 back extension going back to 1950.There is an upward trend in the real part of EOF1 (Figure\ \@ref(fig:extended-series)a.1, p-value `r trends[part == "Real" & PC == "PC1"]$p_report`) and no significant trend in any of the complex parts of EOF2.
-The positive trend in the Real EOF1 translates into a positive trend in EOF1 magnitude, but not in phase (not shown).
-
-## Regressions
-
-The spatial patterns shown in Figure\ \@ref(fig:ceofs-1) are "pure" EOF patterns.
-But as they are derived by removing the zonally symmetric circulation, they might not include all the variability that is actually associated with the EOF time series.
-They are also idealised in that they are forced by construction, to be Hilber transforms of each other.
-To get a more realistic view of the real-world variability described by these EOFS, we computed regression patterns of each EOF with the complete geopotential fields.
+Figure\ \@ref(fig:extended-series) shows time series of the two complex EOFs extended beyond the satellite era using the preliminary ERA5 back extension going back to 1950.
+\textcolor{red}{(Me viene una duda metodológica que vendría bien aclarar.
+Cuando decis que lo extendes, ¿es que lo calculaste para todo ese período o que proyectas el modo en los otros tiempos?)} There is a significant upward trend in the real component of cEOF1 (Figure\ \@ref(fig:extended-series)a.1, p-value `r trends[part == "Real" & PC == "PC1"]$p_report`) and no significant trend in any of the complex components of cEOF2.
+The positive trend in the real cEOF1 translates into a positive trend in cEOF1 magnitude, but not in the phase (not shown).
+\textcolor{red}{(¿Qué interpretación se le podría dar a esto?
+¿Que las amplitudes de las anomalías asociadas al modo son cada vez mayores?)}
+
+## Regressions  (me parece que habría que decirlo de una forma más precisa o juntarlo a la sección anterior)
+
+The spatial patterns shown in Figure\ \@ref(fig:ceofs-1) are derived by removing the zonally symmetric circulation, so they might not include all the variability that is actually associated with the cEOF time series.
+They are also idealised in that they are forced by construction, to be Hilbert transforms of each other \textcolor{red}{(¿Cómo se interpretaría esto?, la oración no es muy clara)}.
+To get a more realistic view of the real-world variability described by these cEOFs, we computed regression patterns of each cEOF with the complete geopotential fields.
+
+\textcolor{red}{(El parráfo anterior quizás hay que trabajarlo un poco más, porque tiene algunos términos que pueden no entenderse bien o que no son los más correctos ("real-world").
+También quizás esto conviene directamente contarlo cuando se describe la metodología.)}
 
@@
 
-Figure\ \@ref(fig:eof1-regr-gh) shows regression patterns of EOF1 and geopotential height.
+Figure\ \@ref(fig:eof1-regr-gh) shows regression patterns between geopotential height anomalies and cEOF1.
 At 50 hPa (Figure\ \@ref(fig:eof1-regr-gh) row a), both the Real and Imaginary EOF1 are associated with planetary wave 1 patterns, that are 90º off phase.
-Their phases coincide with the ones shown in Figure\ \@ref(fig:ceofs-1)a.1, with the positive centre of the Real EOF1 located towards the dateline, and the one of the Imaginary EOF1 located over Eastern Antarctica.
-However, the Real EOF1 pattern is substantially altered by the zonally symmetric circulation.
+Their phases coincide with the ones shown in Figure\ \@ref(fig:ceofs-1)a.1, with the positive centre of the real cEOF1 located towards the dateline, and the one of the Imaginary cEOF1 located over Eastern Antarctica.
+However, the cReal EOF1 pattern is substantially altered by the zonally symmetric circulation.
 Instead of a clear wave 1 pattern, the regression pattern can be better describes as a monopole with its centre displaced from the South Pole.
 
-Similarly affected are the regression patterns at 200 hPa (Figure\ \@ref(fig:eof1-regr-gh) row b).
-It is only possible to distinguish some of the wave 1 pattern in relation with the Real EOF1 (Figure\ \@ref(fig:eof1-regr-gh)b.1).
-The Imaginary EOF1 shows a much more zonally symmetrical pattern resembling a negative Southern Annular Mode.
-
-With the exception of the Imaginary EFO1, it's clear that these patterns are very different than the idealised versions shown in Figure\ \@ref(fig:ceofs-1), particularly at 200 hPa.
+The regression patterns at 200 hPa (Figure\ \@ref(fig:eof1-regr-gh) row b) are similarly influenced by the zonally symmetric circulation.
+It is only possible to distinguish partially the wave 1 pattern in relation with the real cEOF1 (Figure\ \@ref(fig:eof1-regr-gh)b.1).
+The imaginary cEOF1 shows a much more zonally symmetrical pattern resembling a negative Southern Annular Mode.
+
+With the exception of the imaginary cEOF1, it is clear that these patterns are very different than the fully zonally asymmetric versions (Fig.\ \@ref(fig:ceofs-1)), particularly at 200 hPa.
 Moreover, only in the stratosphere these patterns actually show a distinguishable wave 1 pattern shifted in phase by 90º.
-This should not be utterly surprising, since the spatial patterns of the *naive* EOFs only show this feature at 50 hPa, suggesting that using the complex EOF method is "forcing" a wave structure where there is none.
-
+This should not be utterly surprising, since the spatial patterns of the *naive* EOFs only show this feature at 50 hPa (Fig.\ \@ref(fig:eof-naive)), suggesting that using the cEOF method is artificially generating a wave structure at 200hPa.
 Therefore, the magnitude and phase of the EOF1 are associated with the magnitude and phase of a zonal wave only in the stratosphere.
-While in the troposphere they are associated with slightly off-centre monopoles.
+While in the troposphere, they are associated with slightly off-centre monopoles.
 
@@
 
-Figure\ \@ref(fig:eof2-regr-gh) shows the regression pattern of geopotential height and the EOF2.
-Unlike in the case of EOF1, in this case the regression patterns represent relatively well the *idealised* patterns from Figure\ \@ref(fig:ceofs-1).
-Although there are some differences (particularly in 50 hPa), the wave trains identified before are well characterised and patterns associated with the Real EOF2 are 90º out of phase with those associated with the Imaginary EOF2.
-Zonal wave 3 dominates all fields, but only in the western hemisphere.
-
-EOF2 then represents a equivalent barotropic wave train that is very similar to the the Pacific South American Patterns [@mo2001].
-Comparing the location of the positive anomaly near 90ºW in column b of Figure\ \@ref(fig:eof2-regr-gh) with Figure 1 a and b from @mo2001, the Real EOF2 can roughly be identified with the PSA2, while the Imaginary EOF2 resembles PSA1.
-This separation is not completely meaningful in this case, since by treating this pattern as a single wave train any one pair of orthogonal waves can characterise this variability.
+Figure\ \@ref(fig:eof2-regr-gh) shows the regression pattern of geopotential height and the cEOF2.
+Unlike for cEOF1, in this case the regression patterns are similar to the fully zonally assymetric patterns from Figure\ \@ref(fig:ceofs-1).
+Although there are some differences (particularly in 50 hPa), the wave trains identified before are well characterised and patterns associated with the real cEOF2 are 90º out of phase with those associated with the imaginary cEOF2.
+Zonal wave 3 dominates all fields, but only in the western hemisphere, over the Pacific and Atlantic Oceans.
+cEOF2 then represents a equivalent barotropic wave train that is very similar to the the Pacific South American Patterns [@mo2001].
+Comparing the location of the positive anomaly near 90ºW in column b of Figure\ \@ref(fig:eof2-regr-gh) with Figures 1.a and b from @mo2001, the real cEOF2 can roughly be identified with PSA2, while the imaginary cEOF2 resembles PSA1.
+This separation is not completely meaningful in this case, since by treating this pattern as a single wave train any one pair of orthogonal waves can characterise this variability \textcolor{red}{(No me quedo del todo claro el sentido de esta última oración.
+Si podes ampliarla o reescribirla de otra manera, mejor)}.
 
@@
 
-## Relationship with other variables
-
-### Ozone
+## cEOFs relationship with other variables
+
+### Temperature and Ozone
 
@@
 
-(ref:t-vertical-cap) Regression of EOF1 with zonal anomalies of mean air temperature (shaded) and ozone mixing ratio (contours, negative contours with dashed lines, labels in parts per billion by mass) averaged between 80ºS and 25ºS for the period 1979 -- 2019.
+(ref:t-vertical-cap) Regression coefficients between cEOF1 and zonal anomalies of mean air temperature (shaded) and ozone mixing ratio (contours, negative contours with dashed lines, labels in parts per billion by mass) averaged between 80ºS and 35ºS for the period 1979 -- 2019.
 
@@
 
-Figure\ \@ref(fig:eof1-regr-t) shows regression patterns of EOF1 with air temperature.
-In both levels, the distribution of temperature anomalies mirror the geopotential height anomalies in 50 hPa.
-To understand the vertical distribution of these anomalies, Figure\ \@ref(fig:t-vertical)) shows the regression pattern of EOF1 with zonal anomalies of air temperature and zonal anomalies of ozone mixing ratio averaged between 80ºS and 35ºS.
-Temperature zonal anomalies associated with EOF1 show a clear wave 1 pattern throughout the atmosphere above 200 hPa with a change in sign above 10 hPa.
+Figure\ \@ref(fig:eof1-regr-t) shows regression patterns of air temperature at 50hPa and 200hPa onto cEOF1.
+In both levels, the distribution of temperature anomalies \textcolor{red}{(Ojo que cuando se habla de regresiones uno no muestra anomalías sino coeficientes de regresión)} mirror the geopotential height anomalies in 50 hPa (Fig.
+XX) \textcolor{red}{(Dar un poco más de detalle de lo que se ve, aunque sea lo mismo de antes)}.
+The vertical distribution of these patterns is shown in Figure\ \@ref(fig:t-vertical)) that shows the regression coefficientes between cEOF1 and zonal anomalies of air temperature and zonal anomalies of ozone mixing ratio averaged between 80ºS and 35ºS.
+Temperature zonal anomalies associated with cEOF1 show a clear wave 1 pattern for both real and imaginary components throughout the atmosphere above 200 hPa  \textcolor{red}{(¿Empiezan en 200 hPa los contornos?)}  with a change in sign above 10 hPa.
 Following hydrostatic balance, this is the level in which the geopotential anomaly have maximum amplitude (not shown).
 
-The maximum ozone anomalies are co-located with the minimum temperature anomalies above 10 hPa and with the maximum temperature anomalies below 10 hPa; that is, the ozone zonal wave 1 is anticorrelated with the temperature zonal wave 1 in the upper stratosphere, and directly correlated in the upper stratosphere.
-This change in phase is observed in ozone anomalies forced by planetary waves that reach the stratosphere.In the photochemically-dominated upper stratosphere, cold temperatures inhibit the destruction of ozone, and the advectively-dominated lower stratosphere, ozone anomalies are 90º out of phase with horizontal and vertical transport, which is 90º our of phase with temperature anomalies [@hartmann1979;@wirth1993;@smith1995].
+The maximum ozone anomalies are co-located with the minimum temperature anomalies above 10 hPa and with the maximum temperature anomalies below 10 hPa (Fig.\ \@ref(fig:t-vertical))); that is, the ozone zonal wave 1 is anticorrelated with the temperature zonal wave 1 in the upper stratosphere, and directly correlated in the upper stratosphere  \textcolor{red}{(¿Sabes si se le puede creer mucho a los datos allá arriba o agregamos alguna nota precautoria?)} .
+This change in phase is observed in ozone anomalies forced by planetary waves that reach the stratosphere.
+In the photochemically-dominated upper stratosphere, cold temperatures inhibit the destruction of ozone, and the advectively-dominated lower stratosphere, ozone anomalies are 90º out of phase with horizontal and vertical transport, which is 90º our of phase with temperature anomalies [@hartmann1979;@wirth1993;@smith1995].
 
@@
 
-(ref:o3-regr-cap) Regression coefficients of the real and imaginary part of the first complex EOF on SON mean Total Ozone Column.
+(ref:o3-regr-cap) Regression coefficients of SON mean Total Ozone Column anomalies onto the real (a) and imaginary (b) components of the cEOF1.
 On contours, the mean zonal anomaly of Total OZone Column (negative contours in dashed lines).
+¿Units?
 For  the 1979 -- 2019 period.
@@
 Climatologically, the springtime Ozone minimum is located off the South Pole and towards the Weddell Sea [@wirth1993].
-Thus, the Real EOF1 regression pattern (Figure\ \@ref(fig:o3-regr)a) coincides with the climatological position of the ozone whole while the one for the Imaginary EOF1 is shifted by 90º.
+Thus, the Real EOF1 regression pattern (Figure\ \@ref(fig:o3-regr)a) coincides with the climatological position of the ozone hole while the one for the Imaginary EOF1 is shifted by 90º.
 
@@
 
-Indeed, Figure\ \@ref(fig:wave1-o3) shows the relationship between amplitude and phase of the planetary wave 1 in Total Ozone Column between 70ºS and 45ºS and amplitude and phase of the EOF1.
+Indeed, Figure\ \@ref(fig:wave1-o3) shows the relationship between amplitude and phase of the planetary wave 1 in Total Ozone Column between 70ºS and 45ºS \textcolor{red}{(Habría que justifica mejor estas elecciones de bandas latitudinales a paritr del gráfico anterior)} and amplitude and phase of the EOF1.
 The correlation between the zonal wave 1 of ozone and EOF1 (computed as the mean cosine of the difference in phase) is `r wave_correlation$text`.
@@
 
-We will show that the EOF2 offers an alternative way of representing the PSA which has several advantages over using the second and third principal components .
+We will show that the EOF2 offers an alternative way of representing the PSA which has several advantages over using the second and third principal components.
+\textcolor{red}{(Esto queda raro porque estas adelantando la conclusión)}
 
@@
 
+The relationship between PSAs and the modes described from the cEOFs is also studied.
 Figure\ \@ref(fig:psa-eof2) shows the relationship between the two PSA indices and the Real and Imaginary phase of EOF2.
@@
 It is also the same that maximises this clean association between EFO2 parts and PSA modes.
-The reason that the "naive" principal component analysis arrives naturally to this optimal separation is that the imaginary phase is the most common phase (Figure\ \@ref(fig:phase-histogram)), so it naturally appears first when performing a Principal Component Analysis.
-
-(ref:phase-histogram-cap) Histogram showing phase distribution of EOF2 for the period 1979 -- 2019.
+
+\textcolor{red}{(Lo del histograma habría que explicarlo con un pocoo más detalle porque no es tan trivial.
+Primero decir para qué se hace, qué se hace exactamente y después recien decir cuál es el resultado y qué información aporta)}
+
+The reason that the EOF analysis arrives to this optimal separation is that the imaginary cEOF is the most common phase (Figure\ \@ref(fig:phase-histogram)), so it naturally appears first when performing a Principal Component Analysis.
+
+(ref:phase-histogram-cap) Histogram showing phase distribution of EOF2 for the period 1979 -- 2019. \textcolor{red}{(También ampliar)}
 
@@
 
-By aligning he imaginary phase with the direction with maximum relationship with ENSO, we also aligned it with the phase of maximum occurrence.
-This is analogous to the method in @irving2016, who used reprojection and Fourier filtering to detect "PSA-like" variability and defined as the proper PSA using the peaks of the phase distribution (Compare our Figure\ \@ref:(phase-histogram) with their Figure 6).
-
-The advantage of our method is that it is much simpler to implement, it provides magnitude and phase naturally, and it facilitates the description of this mode as a propagating wave instead of as standing oscillation.
+By aligning he imaginary phase with the direction with maximum relationship with ENSO, we also aligned it with the phase of maximum occurrence of XXX.
+This is analogous to the method of @irving2016, who used reprojection and Fourier filtering to detect "PSA-like" variability and defined as the proper PSA using the peaks of the phase distribution (Compare our Figure\ \@ref:(phase-histogram) with their Figure 6).
+\textcolor{red}{(Amplia un poco la oración anterior que no queda del todo clara)} The advantage of our method is that it is much simpler to implement, it provides magnitude and phase naturally, and it facilitates the description of this mode as a propagating wave instead of as standing oscillation.
+As a consequence, the cEOF2 offers an alternative way of representing the PSA which has several advantages over using the second and third EOFs.
 
@@
 
-Figure\ \@ref(fig:sam-eof-vertical) shows the coefficient of determination between the EOFs and three SAM indices.
-
-Both EOFs bear some modest relationship with the SAM index (thick green line in Figure\ \@ref(fig:sam-eof-vertical)), although not statistically significant at every level.
+To explore the relationship between SAM and the modes described from the cEOF, it is computed the coefficient of determination between the cEOFs timeseries and the three SAM indices (SAM, A-SAM and S-SAM) at each level (Fig.\ \@ref(fig:sam-eof-vertical)).
+Both cEOFs bear some modest relationship with the full SAM index (thick green line in Figure\ \@ref(fig:sam-eof-vertical)), although not statistically significant at every level.
+
+ \textcolor{red}{(Esta explicación habría que hacerla más detallada. Explicar más sobre lo que se ve en cada panel y después ir interpretando.  )}
 The split between A-SAM and S-SAM gives more insight into the nature of the relationship.
-The relationship between the SAM and the imaginary EOF1 (Figure\ \@ref(fig:sam-eof-vertical).b1) is mediated by S-SAM in the troposphere, but by the A-SAM in the stratosphere.
-
-The Imaginary EOF2 is related with the SAM through the A-SAM in the troposphere, with up to `r scales::percent(max_r2[PC == "PC2"]$estimate, 2)` of shared variance, reached at `r max_r2[PC == "PC2"]$lev` hPa (Figure\ \@ref(fig:sam-eof-vertical).b2).
+The relationship between the SAM and ¿the imaginary EOF1?
+(Figure\ \@ref(fig:sam-eof-vertical).b1) is mediated by S-SAM in the troposphere, but by the A-SAM in the stratosphere.
+
+The Imaginary EOF2 is related with the SAM through the A-SAM in the troposphere, with up to \textcolor{red}{(Acá aparece un -)} `r scales::percent(max_r2[PC == "PC2"]$estimate, 2)` of shared variance, reached at `r max_r2[PC == "PC2"]$lev` hPa (Figure\ \@ref(fig:sam-eof-vertical).b2).
 
@@
 
-As a further illustration, Figure\ \@ref(fig:eof-filtered) shows the spatial pattern of the leading EOF of 200 hPa geopotential height when the variability associated with EOF2 is removed.
+\textcolor{red}{(La discusión de los próximos parráfos me resulta confusa.
+Deberías aclarar mejor por qué mostras cada cosa que mostras y cómo se relaciona con lo que mostraste antes.
+También me resulta como un resultado que se va un poco de tema, por eso creo que debería estar mejor contada esta parte para poder incluirla)}
+
+As a further illustration \textcolor{red}{(¿De qué?)}, Figure\ \@ref(fig:eof-filtered) shows the spatial pattern of the leading EOF of 200 hPa geopotential height when the variability associated with EOF2 is removed.
 The resulting pattern is an nearly zonally symmetrical annular mode.
@@
 
-This demonstrates that the zonal asymmetries of the SAM could very well be mostly a statistical artifact.
+This suggests that the zonal asymmetries of the SAM could very well be mostly a statistical artifact.
 Since many surface impacts are mediated by the asymmetric component, as well as the relationship between SAM and ENSO [@campitelli2021], this potential issue could affect the interpretability of many results involving the SAM.
@@
 The Imaginary EOF2 is associated with strong positive SST anomalies on the Central Pacific and negative anomalies on an area across the North of Australia and New Zealand, the South Pacific Convergence Zone (SPCZ) (Figure\ \@ref(fig:sst-psi-2).a2).
-This pattern is almost canonically positive ENSO and indeed, the correlation between the Imaginary EOF2 and the Oceanic Niño Index ([@bamston1997]) is very high `r enso_cor[PC == "PC2" & part == "Imaginary"]$text`.
+This pattern is almost canonically positive ENSO and indeed, the correlation between the Imaginary EOF2 and the Oceanic Niño Index [@bamston1997] is very high `r enso_cor[PC == "PC2" & part == "Imaginary"]$text`.
 Streamfunction anomalies show a coherent picture.
 The Imaginary EOF2 is associated with strong wave-like streamfunction anomalies emanating from the tropics (Figure\ \@ref(fig:sst-psi-2).b2).
-This is consistent with what we know of the effect of ENSO on the extratropics: SST anomalies initiate anomalous convection that excites Rossby waves that propagate meridionally towards higher latitudes.
+This is consistent with what we know of the effect of ENSO on the extratropics: SST anomalies initiate anomalous convection that excites Rossby waves that propagate meridionally towards higher latitudes \textcolor{red}{(REFS??)}.
 
@@
 However, Figure\ \@ref(fig:sst-psi-2).a1 and b1 show that the Real EOF2 is not associated either with any significant SST nor streamfunction anomalies in the tropics.
-The correlation between the Real EOF2 and ENSO is also not significant (`r enso_cor[PC == "PC2" & part == "Real"]$cor_text`).
-This lack of tropical signal suggests a radically different nature of the different phases of the EOF2 wave train.
+The correlation between the Real EOF2 and ENSO is also not significant (`r enso_cor[PC == "PC2" & part == "Real"]$cor_text`) \textcolor{red}{(No aparece el valor en el PDF)}.
+This lack of tropical signal suggests different natures of the different phases of the EOF2 wave train.
 
@@
 
-To better explore the relationship between tropical forcing and phase of the EOF2, Figure\ \@ref(fig:enso-phase) plots the ONI index and the phase of the EOF2 for each year between 1979 and 2019, highlighting years in which the magnitude of EOF2 was higher than 50% of the years.
-In years with positive ENSO, the phase of the EOF2 is always around +90º (corresponding with positive imaginary part) and vice versa.
+To better explore the relationship between tropical forcing and phase of the cEOF2, Figure\ \@ref(fig:enso-phase) plots the ONI index and the phase of the EOF2 for each SON trimester between 1979 and 2019, highlighting years in which the magnitude of EOF2 is above the median.
+In years with positive ONI, the phase of the EOF2 is mostly around +90º (corresponding with positive imaginary part) and vice versa.
 In years with near neutral ENSO, the phase of the EOF2 is much more variable.
-The black line in Figure\ \@ref(fig:enso-phase) tries to quantify this relationship.
-Is the equation $ONI = `r signif(coef(cos_model)[1], 2)`\sin{(phase)}$, whose coefficient we fitted by weighted least squares using the magnitude of the EFO2 as weight.
-The $r^2$ corresponding to the fit is `r signif(summary(cos_model)$r.squared, 2)`, with p-value `r report_p(coef(summary(cos_model))[4])`.
-
-Figure\ \@ref(fig:enso-phase) suggest that strong EOF2 years tend to coincide with strong ENSO years.
+The black line in Figure\ \@ref(fig:enso-phase) is an harmonical fit of the relationship between ONI and cEOF2 phase.
+The $r^2$ corresponding to the fit is `r signif(summary(cos_model)$r.squared, 2)`, with p-value `r report_p(coef(summary(cos_model))[4])`, indicating a quasi-sinusoidal relation between these two variables.
+
+Furthemore, Figure\ \@ref(fig:enso-phase) suggest that strong EOF2 years tend to coincide with strong ENSO years.
 The correlation between the absolute magnitude of the ONI and the magnitude of the EOF2 is `r cor_enso_mag$text`.
@@
 Furthermore, even when using all the datapoints, the Spearman correlation --which is robust to outliers-- is also non-significant (`r signif(cor_enso_spearman$estimate, 2)`, p-value`r report_p(cor_enso_spearman$p.value)`).
-Therefore, the relationship between the magnitude of the EOF2 train wave and ENSO remains uncertain.
-
-All this suggests that the wave train represented by EOF2 can be both forced by tropical SSTs and also a part of the internal variability of the extratropical atmosphere.
-When internally forced, the wave train has little phase preference.
-When excited by tropical SST anomalies, it tends to remain locked to the imaginary phase, with the sign of the geopotential anomalies depending on the sign of the tropical SSTs anomalies.
+Therefore, the relationship between the magnitude of the cEOF2 train wave and ONI remains uncertain.
+
+It could be concluded that the wave train represented by cEOF2 can be both part of the internal variability of the extratropical atmosphere or forced by tropical SSTs.
+In the former case, the wave train has little phase preference.
+However, when cEOF2 is excited by tropical SST variability, it tends to remain locked to the imaginary phase.
 This explains the relative overabundance of years with EOF2 near positive and negative imaginary phase in Figure\ \@ref(fig:phase-histogram).
@@
 There is no significant pattern of SST anomalies associated with either the Real or Imaginary EOF1.
-Consistently, streamfunction anomalies don't show any tropical influence.
-On the contrary, the Real EOF1 is associated with wave activity fluxes that flow equatorward from the coast of Antarctica around 150ºE along with wave activity fluxes that move between the positive and negative of Streamfunction anomalies along Antarctica.
+Consistently, streamfunction anomalies do not show any tropical influence.
+Instead of that, the real and imaginary CEOF1 are associated with zonally wave activity fluxes in the extra-tropics, except for an equatorward flow from the coast of Antarctica around 150ºE in the real component.
 
@@
 
-Figure\ \@ref(fig:pp-america) shows regression maps of seasonal precipitation with each EOF in South America.
-EOF1 is associated with slight decreases in precipitation in Southern Brazil and Paraguay, although these are not statistically significant \@ref(fig:pp-america) column a.
+The influence of cEOFs variability in continental rainfall of extra-tropical Southern Hemisphere is also explored.
+We focus on South America and Oceania sectors, because there are not relevant signals in Southern Africa (not shown).
+
+Figure\ \@ref(fig:pp-america) shows regression maps of seasonal precipitation with each cEOF in South America.
+cEOF1 is associated with slight decreases in precipitation in Southern Brazil and Paraguay, although these are not statistically significant \@ref(fig:pp-america) column a.
 The strongest precipitation anomalies are the ones associated with the Imaginary EOF2.
-The positive anomalies on Southeaster South America (SESA) and Chile, and negative anomalies over Southern Brazil is a well known springtime precipitation signature of ENSO [@cai2020a] and it's also virtually identical to the precipitation anomalies associated with the Asymmetric SAM [@campitelli2021].
-This is not surprising considering the close relationship between the ONI, the Asymmetric SAM index and the Imaginary EOF2 shown previously, but further consolidates the identification of this mode with the PSA pattern.
-The Real EOF2, on the other hand, is associated with negative precipitation anomalies in a smaller area of SESA.
-The relationship between precipitation anomalies in SESA and the phase of EOF2 follows a curve similar to that of Figure\ \@ref(fig:enso-phase) (not shown).
-
-(ref:pp-oceania-cap) Same as Figure\ \@ref(fig:pp-america) but for New Zealand and neighbouring islands.
+The positive anomalies on Southeastern South America (SESA) and Central Chile, and negative anomalies over South Atlantic Convergence Zone is a well known springtime precipitation signature of ENSO [@cai2020a] and it is also similar to the precipitation anomalies associated with the A-SAM [@campitelli2021].
+This is not surprising considering the close relationship between the ONI, the Asymmetric SAM index and the Imaginary cEOF2 shown previously, but further consolidates the identification of this mode with the PSA pattern.
+The Real cEOF2, on the other hand, is associated with negative precipitation anomalies in a smaller area of SESA.
+There is a phase dependence of the precipitation anomalies in SESA (not shown), which resembles the relationship between ONI and the phase of EOF2.
+\textcolor{red}{(Podrías agregar acá el coeficiente de determinación)}
+
+  (ref:pp-oceania-cap) As Figure\ \@ref(fig:pp-america) but for Oceania.
 
@@
 
-On the antipodes, precipitation anomalies in New Zealand and neighbouring islands associated with EOF1 are also weak and not statistically significant (Figure\ \@ref(fig:pp-oceania) column a) whereas EFO2 is associated with large and expansive anomalies.
-The Real EOF2 is associated with positive seasonal anomalies in Northern and Eastern Australia, and negative ones over New Zealand, while precipitation anomalies associated with Imaginary EOF2 are negative over all Eastern Australia.
+The precipitation anomalies in Oceania associated with EOF1 are also weak and not statistically significant (Figure\ \@ref(fig:pp-oceania) column a) whereas EFO2 is associated with large and expansive anomalies.
+The Real cEOF2 is associated with positive seasonal anomalies in Northern and Eastern Australia, and negative ones over New Zealand, while precipitation anomalies associated with Imaginary EOF2 are negative over all Eastern Australia.
 Again, these negative anomalies are similar to the springtime precipitation anomalies observed in relation with the Asymmetric SAM [@campitelli2021].
+\textcolor{red}{(Habría que agregar alguna referencia del ENSO que es más conocido)}
 
@@
 The black line is the equation $\mathrm{Precipitation} = `r signif(coef(model)[1], 2)`\sin{(\mathrm{phase})} + `r signif(coef(model)[2], 2)`\cos{(\mathrm{phase})}$, whose coefficient we fitted by weighted least squares using the magnitude of the EFO2 as weight.
+\textcolor{red}{(Como te puse antes la ecuación como mucho la pondría sobre la figura)}
 The $r^2$ corresponding to the fit is `r signif(summary(model)$r.squared, 2)`, with p-value `r report_p(p_value)`.
-There is a strong relationship between the phase of the EOF2 and precipitation anomalies in Eastern Australia, but Real and Imaginary phases we chose are not aligned with the direction that maximises this relationship.
+There is a strong relationship between the phase of the EOF2 and precipitation anomalies in Eastern Australia, but Real and Imaginary phases we chose are not exactly aligned with the direction that maximises this relationship.
 If one were to perform a more in-depth analysis of the relationship between this pattern and precipitation in this region, one would conceivably align one of the axis (either the Real or the Imaginary) to the one that maximises it.

### 2021-03-30 analysis/paper/paper.Rmd
Re-agrega introducción y conclusión

@@
 
+One typical way of describing the zonally asymmetric circulation is by the amplitude and phase of zonal waves obtained by Fourier decomposition of geopotential height at each latitude [e.g.
+@vanloon1972;@trenberth1980].
+These methodologies rely on the assumption that the circulation can be meaningfully understood in terms of zonal waves of constant amplitude along a latitude circle.
+This is not valid for meridionally propagating waves or zonal waves with localised amplitudes.
+In the case of the wave 3, for example, @trenberth1985 observed that wave 3 played a role in blocking events, but that in many cases, this was due to increased amplitude of a longitudinally localised wave train and not a hemispheric-scale zonal wave 3.
+
+Another approach to characterising the Southern Hemisphere circulation is by using Empirical Orthogonal Functions (EOF, also known as Principal Component Analysis).
+Within the EOF framework, the Southern Annular Mode appears as the leading mode of variability of the Southern Hemisphere (SAM) [@fogt2020] followed by the two Pacific–South American Patterns (PSA) [@mo2001].
+The SAM represents a relatively zonally symmetric pattern of alternating low pressures in polar latitude and a ring of high pressures in high latitudes with an embedded wave 3 pattern that is more prominent in the Pacific sector.
+The PSA1 and PSA2 describe the two out-of-phase sides of a meridionally propagating wave train that originates in the Eastern equatorial Pacific and travels towards the South Atlantic following a great-circle arch along the Antarctic Peninsula.
+These patterns are derived by applying EOF to temporal anomalies, but @raphael2003 applied EOF methods specifically to zonal anomalies.
+
+EOFs are more flexible in the sense that they can characterise a richer set of spatial patterns, but they are restricted to standing oscillation modes and have difficulty dealing with propagating or phase-varying modes such as zonal waves.
+A single EOF can also represent a mixture of two or more physical modes.
+
+The inability of a single EOF to describe travelling waves is what forces the PSA pattern to be described by two EOFs.
+@irving2016 characterises the Pacific–South American Pattern in a Fourier framework.
+Being a meridionally propagating mode, the PSA cannot be correctly characterised by Fourier decomposition at each latitude circle, so they reprojected meridional wind fields so that the path of the PSA laid on the equator.
+On this new projection, they could identify the PSA using Fourier decomposition.
+
+Finally, another, hybrid, methodology consist on identifying particular features of interest and creating indices using simple methods such as averages and differences.
+The Marshal SAM Index [@marshall2003], for example, tries to characterise the SAM by the average difference of sea level pressure between 40ºS and 65ºS, following @gong1999.
+@raphael2004 used a similar methodology to describe the wave 3 in the Southern Hemisphere.
+Instead of using Fourier to compute the amplitude and phase of the zonal wave, they averaged standardised geopotential height anomalies in three points representing the location of the ridges of the zonal wave 3 of the climatological mean.
+This is equivalent to an index of the amplitude of the zonal wave 3 that projects into the phase of the zonal wave 3 of the mean field.
+Similarly, @hobbs2010 posit that the zonally asymmetric circulation in the Southern Hemisphere can be described by the strength and location of two anticyclones located in the sub-Antarctic western hemisphere.
+
+These "feature-based" methods are grounded on other methods to identify the interesting features and can be useful to characterise features that are not readily apparent with Fourier or EOF methods but that nonetheless interesting or important.
+The use of data of single points, which makes them suitable for extending into the data-poor past also makes them not very robust.
+
+In this paper we propose using complex Empirical Orthogonal Functions [@horel1984] as a robust extension of EOF analysis that is capable of describing rich structures with meridional and zonal propagation, and amplitude modulation such as the PSA.
+We apply this method to characterise the Springtime zonally asymmetric circulation of the extratopical Southern Hemisphere.
+
 # Data and Methods
@@
 For precipitation data we used monthly data from the CPC Merged Analysis of Precipitation [@cmap], with a 2.5\degree\ resolution in latitude and longitude.
-This rainfall gridded dataset is based on information from different sources such as rain gauge observations, satellite
-inferred estimations and the NCEP-NCAR reanalysis, and it is available since 1979 to present.
+This rainfall gridded dataset is based on information from different sources such as rain gauge observations, satellite inferred estimations and the NCEP-NCAR reanalysis, and it is available since 1979 to present.
 

### 2021-03-30 analysis/paper/paper.Rmd
Actualiza afiliación.
#22

@@
   address: |
-    Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos, Buenos Aires, Argentina
-    CONICET – Universidad de Buenos Aires, Centro de Investigaciones del Mar y la Atmósfera (CIMA), Buenos Aires, Argentina
-    CNRS – IRD – CONICET – UBA, Instituto Franco‐Argentino para el Estudio del Clima y sus Impactos (UMI 3351 IFAECI), Buenos Aires, Argentina
+    (1) Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos. Buenos Aires, Argentina.
+    (2) CONICET – Universidad de Buenos Aires. Centro de Investigaciones del Mar y la Atmósfera (CIMA). Buenos Aires, Argentina.
+    (3) CNRS – IRD – CONICET – UBA. Instituto Franco-Argentino para el Estudio del Clima y sus Impactos (IRL 3351 IFAECI). Buenos Aires, Argentina.
   email: elio.campitelli@cima.fcen.uba.ar
@@
   address: |
-    Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos, Buenos Aires, Argentina
-    CONICET – Universidad de Buenos Aires, Centro de Investigaciones del Mar y la Atmósfera (CIMA), Buenos Aires, Argentina
-    CNRS – IRD – CONICET – UBA, Instituto Franco‐Argentino para el Estudio del Clima y sus Impactos (UMI 3351 IFAECI), Buenos Aires, Argentina
+    (1) Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos. Buenos Aires, Argentina.
+    (2) CONICET – Universidad de Buenos Aires. Centro de Investigaciones del Mar y la Atmósfera (CIMA). Buenos Aires, Argentina.
+    (3) CNRS – IRD – CONICET – UBA. Instituto Franco-Argentino para el Estudio del Clima y sus Impactos (IRL 3351 IFAECI). Buenos Aires, Argentina.
 

### 2021-03-30 analysis/paper/paper.Rmd
Agrega justificación de ondas zonales
Closes #7

@@
 ## Methods
+
+We computed the amplitude and phase of the zonal wave 1 by averaging (area-weighted) variables (temperature, geopotential height, ozone mixing ratio and Total Ozone Column) between `r knitr::combine_words(LatLabel(o3wave_lats))` for each SON and extracting the wave-1 component of the Fourier spectrum.
+We chose this latitude band because it is wide enough to capture most of the relevant anomalies.
 

### 2021-03-30 analysis/paper/paper.Rmd
Agrega unidades en o3-regr-cap
Closes #6
Agrega justificación de ondas zonales (ups anterior)

@@
 
-We used monthly geopotential height, air temperature, ozone mixing ratio, and Total Ozone COlumn (TOC) at 2.5\degree\ longitude by 2.5\degree\ latitude of horizontal resolution and 37 vertical isobaric levels from ERA5 [@era5] for the period 1979 to 2019.
+We used monthly geopotential height, air temperature, ozone mixing ratio, and Total Ozone Column (TOC) at 2.5\degree\ longitude by 2.5\degree\ latitude of horizontal resolution and 37 vertical isobaric levels from ERA5 [@era5] for the period 1979 to 2019.
 We restricted most of our analysis to the post-satellite era to avoid any confounding factors arising from the incorporation of satellite observations, but we also used the preliminary back extension of ERA5 from 1950 to 1978 [@era5be] to look at long-term trends.
@@
 
+We restrict our study to  the September-October-November (SON) trimester.
+For that, we use seasonal means, averaging monthly values weighted by the number of days in each  month.
+We also use 200 hPa to represent the high troposphere and 50 hPa to represent the lower stratosphere.
+
 We computed the amplitude and phase of the zonal wave 1 by averaging (area-weighted) variables (temperature, geopotential height, ozone mixing ratio and Total Ozone Column) between `r knitr::combine_words(LatLabel(o3wave_lats))` for each SON and extracting the wave-1 component of the Fourier spectrum.
 We chose this latitude band because it is wide enough to capture most of the relevant anomalies.
-
-\textcolor{red}{Aclarar qué a lo largo del trabajo se utilizan los datos de primavera y cómo se calcula eso.
-Tambień creo que acá se podría aclarar que el análisis se concentra en dos niveles para representar la estratósfera y la troposféra alta}
-
-We computed the amplitude and phase of the zonal wave 1 by averaging variables (temperature, geopotential height, ozone mixing ratio and Total Ozone Column) `r knitr::combine_words(LatLabel(o3wave_lats))` for each Spring and computing the Fourier spectrum.
 
@@
 (ref:o3-regr-cap) Regression coefficients of SON mean Total Ozone Column anomalies onto the real (a) and imaginary (b) components of the cEOF1.
-On contours, the mean zonal anomaly of Total OZone Column (negative contours in dashed lines).
-¿Units?
+On contours, the mean zonal anomaly of Total OZone Column (negative contours in dashed lines, Dobson Units).
 For  the 1979 -- 2019 period.

### 2021-03-30 analysis/paper/paper.Rmd
Quita intro de PSA
Closes #21
Boludeces de estilo

@@
   .[season(time) == "SON"] %>%
-  .[, .(vo = mean(vo)), by = .(lon, lat, lev,time = seasonally(time))] %>%
+  .[, .(vo = mean(vo)), by = .(lon, lat, lev, time = seasonally(time))] %>%
   .[, solve_poisson(vo, lon, lat), by = .(lev, time)] %>%
@@
   .[, hgt  := Anomaly(hgt)*sqrt(cos(lat*pi/180)), by = .(time, lev, lat)] %>% 
-  .[, .(eof = list(EOF(hgt ~ time | lon + lat, n = 1:4, data = .SD, suffix = "EOF"))), by = lev]
+  .[, .(eof = list(EOF(hgt ~ time | lon + lat, n = 1:4, data = .SD, suffix = "EOF"))), 
+    by = lev]
 
@@
 
-We will show that the EOF2 offers an alternative way of representing the PSA which has several advantages over using the second and third principal components.
-\textcolor{red}{(Esto queda raro porque estas adelantando la conclusión)}
-
 (ref:psa-eof2-cap) Relationship between the Real and Imaginary parts of EOF2 and the PSA1 and PSA2 modes computed as the second and third EOFs of seasonal geopotential height anomalies [@mo2001].

### 2021-04-05 analysis/paper/paper.Rmd
Introducción regresiones
Closes #5

@@
 
+## Regressions
+
 The spatial patterns shown in Figure\ \@ref(fig:ceofs-1) are derived by removing the zonally symmetric circulation, so they might not include all the variability that is actually associated with the cEOF time series.
-They are also idealised in that they are forced by construction, to be Hilbert transforms of each other \textcolor{red}{(¿Cómo se interpretaría esto?, la oración no es muy clara)}.
-To get a more realistic view of the real-world variability described by these cEOFs, we computed regression patterns of each cEOF with the complete geopotential fields.
-
-\textcolor{red}{(El parráfo anterior quizás hay que trabajarlo un poco más, porque tiene algunos términos que pueden no entenderse bien o que no son los más correctos ("real-world").
-También quizás esto conviene directamente contarlo cuando se describe la metodología.)}
-
+To understand the full geopotential height field changes associated with each cEOF, we computed regression patterns of each cEOF with the complete geopotential fields.
 ```{r compute-regression}

### 2021-04-05 analysis/paper/paper.Rmd
Actualiza afiliaciones
(#22)

@@
   address: |
-    (1) Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos. Buenos Aires, Argentina.
-    (2) CONICET – Universidad de Buenos Aires. Centro de Investigaciones del Mar y la Atmósfera (CIMA). Buenos Aires, Argentina.
-    (3) CNRS – IRD – CONICET – UBA. Instituto Franco-Argentino para el Estudio del Clima y sus Impactos (IRL 3351 IFAECI). Buenos Aires, Argentina.
+    Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos. Buenos Aires, Argentina.
+    CONICET – Universidad de Buenos Aires. Centro de Investigaciones del Mar y la Atmósfera (CIMA). Buenos Aires, Argentina.
+    CNRS – IRD – CONICET – UBA. Instituto Franco-Argentino para el Estudio del Clima y sus Impactos (IRL 3351 IFAECI). Buenos Aires, Argentina.
   email: elio.campitelli@cima.fcen.uba.ar
@@
   address: |
-    (1) Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos. Buenos Aires, Argentina.
-    (2) CONICET – Universidad de Buenos Aires. Centro de Investigaciones del Mar y la Atmósfera (CIMA). Buenos Aires, Argentina.
-    (3) CNRS – IRD – CONICET – UBA. Instituto Franco-Argentino para el Estudio del Clima y sus Impactos (IRL 3351 IFAECI). Buenos Aires, Argentina.
+    Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos. Buenos Aires, Argentina.
+    CONICET – Universidad de Buenos Aires. Centro de Investigaciones del Mar y la Atmósfera (CIMA). Buenos Aires, Argentina.
+    CNRS – IRD – CONICET – UBA. Instituto Franco-Argentino para el Estudio del Clima y sus Impactos (IRL 3351 IFAECI). Buenos Aires, Argentina.
 

### 2021-04-06 analysis/paper/paper.Rmd
Muchos cambios

@@
   comment = "#>",
-  cache.extra = 42,
+  cache.extra = 41,
   cache.path = "../cache/paper/",
@@
               I = "Imaginary")
-lab_PC <- c(PC1 = "EOF1", 
-            PC2 = "EOF2")
 
@@
   .[, hgt  := Anomaly(hgt)*sqrt(cos(lat*pi/180)), by = .(time, lev, lat)] %>% 
-  .[, .(eof = list(EOF(hgt ~ time | lon + lat, n = 1:4, data = .SD, suffix = "EOF"))), 
+  .[, .(eof = list(EOF(hgt ~ time | lon + lat, n = 1:4, data = .SD, suffix = "cEOF"))), 
     by = lev]
@@
 explained <- eof_naive[, eof[[1]]$sdev, by = lev] %>% 
-  .[order(lev, EOF)] %>% 
+  .[order(lev, cEOF)] %>% 
   .[, scales::percent(r2, accuracy = .1)]
@@
 
-```{r eof-naive, fig.cap = "(ref:eof-naive-cap)", fig.width=3.7, fig.height=2.2}
+```{r eof-naive, fig.cap = "(ref:eof-naive-cap)", fig.width=2.5, fig.height=3}
 eof_naive[, eof[[1]]$right, by = lev] %>%
+  .[lev == 50] %>% 
   periodic(lon = c(0, 360)) %>%
@@
   theme(axis.text = element_blank()) +
-  facet_grid(lev~EOF, labeller = labeller(lev = lab_lev))  +
+  facet_wrap(.~cEOF, ncol = 2, labeller = labeller(lev = lab_lev))  +
   tagger::tag_facets("rc")  +
@@
 
-In traditional EOF analysis zonal waves appear as pairs of EOFs, usually degenerate, that represent similar patterns but shifted in phase \textcolor{red}{(Agregaría algunas referencias)}.
-For instance, Figure\ \@ref(fig:eof-naive) shows the leading 4 EOFs of zonally anomalous September through October \textcolor{red}{(November supongo.
-Sería bueno que ya arriba indiques lo de SON y ya después lo menciones directamente así)} geopotential height at 50 hPa and 200 hPa.
-It is clear that at 50 hPa the first two EOFs represent the same zonal wave 1 pattern and the last 2 represent the same zonal wave 3  \textcolor{red}{(No es visualmente obvia una estructura de onda 3.
-Solo se ven 4 centros)} pattern shifted by 1/4 wavelength.
-Similarly, EOFs 2 and 3 at 200 hPa represent the a zonal wave 3 pattern shifted by 1/4 wavelength.
-Since these pairs of EOFs seem to represent the same phase-varying structure, it would be desirable to combine them into a single index with amplitude and phase.
+In traditional EOF analysis zonal waves appear as pairs of EOFs, usually degenerate, that represent similar patterns but shifted in phase [@horel1984].
+For instance, Figure\ \@ref(fig:eof-naive) shows the leading 4 EOFs of zonally anomalous SON geopotential height at 50 hPa and 200 hPa.
+It is clear that at 50 hPa the first two EOFs represent the same zonal wave 1 pattern and the last 2 represent the same zonal wave pattern with shorter wavenumber shifted by 1/4 wavelength.
+A similar EOF structure can be seen in 200 hPa (not shown).
+Since each pair of EOFs seem to represent the same phase-varying structure, it would be desirable to combine them into a single index with amplitude and phase.
 
@@
   .[, mag  := abs(hgt)] %>%
-  .[, item := paste(lev, PC, sep = "_")] %>%
+  .[, item := paste(lev, cEOF, sep = "_")] %>%
   widyr::pairwise_cor(item, time, mag) %>%
   setDT() %>%
-  # .[]
   .[tstrsplit(item1, "_")[[1]] != tstrsplit(item2, "_")[[1]], ] %>%
@@
 
-Another characteristic seen in Figure\ \@ref(fig:eof-naive) is that EOF1 at 200 hPa is very similar to EOF1 at 50 hPa  \textcolor{red}{(En la figura que yo veo, no diría que no similares.
-Aclara mejor a qué te referis)}.
-The zonal wave 3 structure in EOF2 and EOF3 at 200 hPa are also very similar to the one in EOF3 and EOF4 at 50 hPa, excecpt for the behaviour above the Indian Ocean.
-Furthermore, Table \@ref(tab:corr-ceof-splitted) shows the coefficient of determination between time series of the amplitude of each complex EOF across levels.
-There's a high degree of correlation between the respective EOF1 and EOF2 at each level.
-Both the spatial similarities and temporal correlations suggest some level of joint variability between levels.
-\textcolor{red}{(Revisar el parráfo considerando el comentario)}
-
-Both observations \textcolor{red}{(¿Qué serían las dos osbervaciones?)}  motivate the decision of performing complex EOF jointly between levels.
+Table \@ref(tab:corr-ceof-splitted) shows the coefficient of determination between time series of the amplitude of each complex EOF (cEOF) across levels.
+There's a high degree of correlation between the respective cEOF1 and cEOF2 at each level.
+The spatial patterns of the 50 hPa and 200 hPa cEOFs are also similar (not shown)
+
+Both the spatial pattern similarity and the high temporal correlation of cOEFs computed at 50 hPa and 200 hPa suggest that these are, to a large extent, modes of joint variability.
+This motivates the decision of performing complex EOF jointly between levels.
 The computation of the EOFs was carried out using data from both levels at the same time, therefore, each complex EOF has a spatial component that depends on longitude, latitude and level, and a temporal component that depends only on time.
@@
 
-# rotations_pc1 <- lapply(angles, function(a) {
-#   ceof$eof[[1]]$left[PC == "PC1"] %>%
+# rotations_cEOF1 <- lapply(angles, function(a) {
+#   ceof$eof[[1]]$left[cEOF == "cEOF1"] %>%
 #     copy() %>%
@@
 # 
-# best_rotation_pc1 <- rotations_pc1[part == "Real"][which.min(mean)]$rotation
+# best_rotation_cEOF1 <- rotations_cEOF1[part == "Real"][which.min(mean)]$rotation
 
@@
 with_o3 <- ceof$eof[[1]]$left %>% 
-  .[PC == "PC1"] %>% 
+  .[cEOF == "cEOF1"] %>% 
   .[wave1_o3, on = "time"]
@@
 
-best_rotation_pc1 <- rotations_o3[part == "Real"][which.max(cor)]$rotation
+best_rotation_cEOF1 <- rotations_o3[part == "Real"][which.max(cor)]$rotation
 
@@
 
-rotations_pc2 <- lapply(angles, function(a) {
+rotations_cEOF2 <- lapply(angles, function(a) {
   with_enso %>%
@@
 
-best_rotation_pc2 <- rotations_pc2[I > 0][which.min(abs(R))]$rotation
+best_rotation_cEOF2 <- rotations_cEOF2[I > 0][which.min(abs(R))]$rotation
 
@@
   copy() %>%
-  .[PC == "PC2", hgt := rotate(hgt, best_rotation_pc2)] %>%
-  .[PC == "PC1", hgt := rotate(hgt, best_rotation_pc1)]
+  .[cEOF == "cEOF2", hgt := rotate(hgt, best_rotation_cEOF2)] %>%
+  .[cEOF == "cEOF1", hgt := rotate(hgt, best_rotation_cEOF1)]
 
@@
 
-```{r rotations, fig.cap = "(ref:rotations-cap)", fig.width=width_column, fig.height=2.5}
+```{r rotations, fig.cap="(ref:rotations-cap)", fig.height=2.5, fig.width=width_column, include=FALSE}
 r2 <- expression(`$r^2$` = paste("", "r", phantom()^{
@@
   ggplot(aes(rotation, cor)) +
-  geom_vline(xintercept = best_rotation_pc1) +
+  geom_vline(xintercept = best_rotation_cEOF1) +
   geom_hline(yintercept = 0,  size = 0.2, color = "gray50") +
@@
 
-rotations_pc2 %>%
+rotations_cEOF2 %>%
   melt(id.vars = "rotation") %>%
@@
   geom_hline(yintercept = 0,  size = 0.2, color = "gray50") +
-  geom_vline(xintercept = best_rotation_pc2) +
+  geom_vline(xintercept = best_rotation_cEOF2) +
   geom_line(aes(color = variable)) +
@@
 ```
-
-\textcolor{red}{(No mostraría la figura 2. Me parece que agrega una complejidad extra y no sé entiende demasiado, sino se la explica muy bien)}
-
 For the first complex principal component, the phase was chosen so that the time series corresponding to the real part has the maximum correlation with the zonal wave 1 of Total Ozone Column between `r knitr::combine_words(LatLabel(o3wave_lats))`.
-This also nearly minimises the correlation with the imaginary part (Figure\ \@ref(fig:rotations)a).
-
-For the second complex principal component, the phase was chosen so that the coefficient of determination between the Oceanic Niño Index [@bamston1997] and the real part was minimized, which also nearly maximizes the correlation with the imaginary part (Figure\ \@ref(fig:rotations)b).
+This also nearly minimises the correlation with the imaginary part.
+
+For the second complex principal component, the phase was chosen so that the coefficient of determination between the Oceanic Niño Index [@bamston1997] and the real part was minimized, which also nearly maximizes the correlation with the imaginary part.
 
@@
 
-## Zonal anomalies circulation modes (no me convence del todo, pero debería ser algo así)
+## cEOF patterns
 
 ```{r define-plot_ceof}
-plot_ceof <- function(data, n = 1:2, formula = lev  ~  PC) {
+plot_ceof <- function(data, n = 1:2, formula = lev  ~  cEOF) {
 
@@
   var <- data[, cut(eof[[1]], n)$sdev, by = .(season)] %>%
-    .[, setNames(paste0(PC, " (", scales::percent(r2), ")"),
-                 PC)]
+    .[, setNames(paste0(cEOF, " (", scales::percent(r2), ")"),
+                 cEOF)]
 
@@
                labeller = labeller(lev = lab_lev,
-                                   PC = var)) +
+                                   cEOF = var)) +
     coord_quickmap(ylim = c(NA, -20)) +
@@
 
-```{r ceofs-1, fig.cap = "Spatial patterns of the leading 2 complex EOFs of zonal anomalies of geopotential height in 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2019.
-Real part in shading, imaginary part in contours", fig.width=width_column, fig.height=width_column}
+```{r ceofs-1, fig.cap = "Spatial patterns of the leading 2 complex EOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2019.
+Real part in shading, imaginary part in contours.", fig.width=width_column, fig.height=width_column}
 ceof[season == "SON"] %>%
@@
 
-```{r ceofs-2, fig.cap = "Time series of the leading 2 complex EOFs of zonal anomalies of geopotential height in 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2019.", fig.height=width_column, fig.width = width_column}
+```{r ceofs-2, fig.cap = "Time series of the leading 2 complex EOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2019.", fig.height=width_column, fig.width = width_column}
 ceof %>%
@@
   .[season == "SON"] %>%
-  # .[, hgt := scale(hgt), by = .(part, PC)] %>%
+  # .[, hgt := scale(hgt), by = .(part, cEOF)] %>%
   ggplot(aes(time, hgt)) +
@@
   scale_x_datetime(NULL) +
-  scale_y_continuous("EOF") +
+  scale_y_continuous(NULL) +
   scale_color_brewer(NULL, palette = "Dark2") +
-  facet_grid(PC ~ .) +
-  theme(panel.spacing = grid::unit(1.5, "lines"))  +
+  facet_grid(cEOF ~ ., switch = "y") +
+  theme(panel.spacing = grid::unit(1.5, "lines"), strip.placement = "outside")  +
   ggthemes::geom_rangeframe(sides = "lr", size = 0.1, color = "gray30") +
@@
 phase) is controlled by the magnitude and phase of the complex temporal EOF.
-The wave patterns described by these cEOFs match the patterns seen in the traditional EOFs of Figure\ \@ref(fig:eof-naive): The first is mainly a wave 1, while the second is a wave 3.
+The wave patterns described by these cEOFs match the patterns seen in the traditional EOFs of Figure\ \@ref(fig:eof-naive).
+
+The first cEOF (Fig.\ \@ref(fig:eof-naive) column\ a) is a hemispheric wave 1 pattern with maximum amplitude at high latitudes.
+At 50 hPa the Real cEOF1 has the maximum of the wave 1 at 150ºE and at 200 hPa, the maximum is locates at around 175ºE indicating a westerly shift in phase.
+
+The second cEOF (Fig.\ \@ref(fig:eof-naive) column\ b) show also a zonal wave-like structure with maximum amplitude at high latitudes, but with shorter spatial scales.
+In particular, it's dominant structure at both levels is a wave 3 but with larger amplitude in the pacific sector.
+This modulated amplitude is especially apparent at 200 hPa.
+There is no apparent phase shift, suggesting a rather barotropic structure.
 
@@
 As a consequence, the variability associated with the first cEOF includes variability that projects onto the mean zonally anomalous field.
-
-\textcolor{red}{(Acá creo que habría que ampliar un poco más la discusión.
-diciendo en más detalle lo que se ve en la parte espacial (donde se ven los centros más importantes, qué diferencias hay entre estratósfera y troposféra) y temporal, por ej.
-qué tipo de variabilidad se observa, si están correlacionadas)}
+There are no significant simultanous correlation between the time series.
 
@@
   .[pattern, on = .NATURAL, allow.cartesian = TRUE] %>%
-  .[, .(hgt = weighted.mean(hgt*EOF, w = cos(lat*pi/180))),  by = .(part, PC, time)]
+  .[, .(hgt = weighted.mean(hgt*EOF, w = cos(lat*pi/180))),  by = .(part, cEOF, time)]
 
@@
 cors <- era_more %>%
-  .[sep_ReIm(ceof$eof[[1]]$left), on = c("time", "PC", "part")] %>%
+  .[sep_ReIm(ceof$eof[[1]]$left), on = c("time", "cEOF", "part")] %>%
   na.omit() %>%
-  .[, cor(hgt, i.hgt), by = .(PC, part)]
+  .[, cor(hgt, i.hgt), by = .(cEOF, part)]
 
@@
 trends <- era_more %>%
-  .[, FitLm(hgt, time, se = TRUE), by = .(part, PC)] %>%
+  .[, FitLm(hgt, time, se = TRUE), by = .(part, cEOF)] %>%
   rm_intercept() %>%
@@
 #   .[season == "SON"] %>% 
-#   .[, hgt := as.numeric(scale(hgt)), by = .(part, PC)]
+#   .[, hgt := as.numeric(scale(hgt)), by = .(part, cEOF)]
 
@@
   .[, ref := year(time) > 1979] %>% 
-   # .[, hgt := scale(hgt, center = FALSE, scale = sd(hgt[ref])), by = .(part, PC)] %>%
+   # .[, hgt := scale(hgt, center = FALSE, scale = sd(hgt[ref])), by = .(part, cEOF)] %>%
   ggplot(aes(time, hgt*1e3)) +
   # Zero line + rangeframe
-  geom_hline(data = ~.x[, mean(hgt)*1e3, by  = .(PC, part)], 
+  geom_hline(data = ~.x[, mean(hgt)*1e3, by  = .(cEOF, part)], 
              aes(yintercept = V1), size = 0.1, color = "gray30") +
@@
 
-  facet_grid(PC ~ part, labeller = labeller(PC = lab_PC), scales = "free_y") +
+  facet_grid(cEOF ~ part, scales = "free_y") +
   theme(panel.spacing = grid::unit(1.5, "lines"))  +
@@
 
-Figure\ \@ref(fig:extended-series) shows time series of the two complex EOFs extended beyond the satellite era using the preliminary ERA5 back extension going back to 1950.
-\textcolor{red}{(Me viene una duda metodológica que vendría bien aclarar.
-Cuando decis que lo extendes, ¿es que lo calculaste para todo ese período o que proyectas el modo en los otros tiempos?)} There is a significant upward trend in the real component of cEOF1 (Figure\ \@ref(fig:extended-series)a.1, p-value `r trends[part == "Real" & PC == "PC1"]$p_report`) and no significant trend in any of the complex components of cEOF2.
+Both cEOF shot year-to-year variability but show no evidence of decadal variabilty.
+The relatively short timespan (`r diff(range(unique(year(era5$time))))` years) hampers our habilty to detect longer trends.
+Therefore, we extend the cEOF time series using the preliminary ERA5 back extension going back to 1950 by projecting the older geopogential height zonal anomaly fields onto the spatial patterns shown in Figure\ \@ref(fig:ceofs-1).
+
+The extended cEOF timeseires are shown in Figure\ \@ref(fig:extended-series).
+There is a significant upward trend in the real component of cEOF1 (Fig.\ \@ref(fig:extended-series)a.1, p-value\ `r trends[part == "Real" & cEOF == "cEOF1"]$p_report`) and no significant trend in any of the complex components of cEOF2.
 The positive trend in the real cEOF1 translates into a positive trend in cEOF1 magnitude, but not in the phase (not shown).
-\textcolor{red}{(¿Qué interpretación se le podría dar a esto?
-¿Que las amplitudes de las anomalías asociadas al modo son cada vez mayores?)}
-
-## Regressions  (me parece que habría que decirlo de una forma más precisa o juntarlo a la sección anterior)
+
+This long-term change indicates an increase in the magnitude of the high latitude zonal wave 1.
+A similar observation was made by @raphael2003, who detected a step after around 1975 in the leading EOF of August-September-October 500 hPa zonal geopotential height anomalies.
 
@@
 To understand the full geopotential height field changes associated with each cEOF, we computed regression patterns of each cEOF with the complete geopotential fields.
+
 ```{r compute-regression}
@@
   .[, `:=`(Real = as.numeric(scale(Real)), Imaginary = as.numeric(scale(Imaginary))),
-    by = .(season, PC)] %>%
-  dcast(time ~ PC, value.var = c("Real", "Imaginary")) %>%
+    by = .(season, cEOF)] %>%
+  dcast(time ~ cEOF, value.var = c("Real", "Imaginary")) %>%
   .[melt(era5, id.vars = c("time", "lon", "lat", "lev")), on = "time", allow.cartesian = TRUE] %>%
@@
   # .[variable == "hgt"] %>%
-  .[, FitLm(value, Imaginary_PC1, Real_PC1, Imaginary_PC2, Real_PC2, se = TRUE),
+  .[, FitLm(value, Imaginary_cEOF1, Real_cEOF1, Imaginary_cEOF2, Real_cEOF2, se = TRUE),
     by = .(lev, lon, lat, variable)] %>%
-  .[, c("term", "PC") := tstrsplit(term, "_")] %>%
+  .[, c("term", "cEOF") := tstrsplit(term, "_")] %>%
   rm_intercept() %>%
@@
 ```{r plot_regr}
-plot_regr <- function(data, which_PC, which_levs, which_var) {
+plot_regr <- function(data, which_cEOF, which_levs, which_var) {
   data <- data %>%
-    .[ PC %in% which_PC] %>%
+    .[ cEOF %in% which_cEOF] %>%
     .[variable == which_var]
@@
                       t = "Temperature")
-  lab_pc <- c(PC1 = "EOF1", PC2 = "EOF2")
-
-  if (length(which_PC) > 1) {
-    facet <- ggh4x::facet_nested(PC ~  term)
+
+  if (length(which_cEOF) > 1) {
+    facet <- ggh4x::facet_nested(cEOF ~  term)
   } else {
-    facet <- ggh4x::facet_nested(lev ~ PC +  term, labeller = labeller(lev = lab_lev, PC = lab_pc))
+    facet <- ggh4x::facet_nested(lev ~ cEOF +  term, 
+                                 labeller = labeller(lev = lab_lev))
   }
@@
     # .[term == "I" & lev == 50] %>%
-    .[, p.value := Pvaluate(estimate, std.error, df, "fdr"), by = .(lev, PC, term)] %>%
+    .[, p.value := Pvaluate(estimate, std.error, df, "fdr"), by = .(lev, cEOF, term)] %>%
     periodic(lon = c(0, 360)) %>%
@@
 ```{r eof1-regr-gh, fig.cap = "(ref:eof1-regr-gh-cap)", fig.height=1.2*width_column, fig.width=width_column, dependson="plot_regr"}
-plot_regr(regr , "PC1", c(200, 50), "hgt") +
+plot_regr(regr , "cEOF1", c(200, 50), "hgt") +
   # geom_contour2(data = ~copy(.x)[, estimate := Anomaly(estimate), by = .(term, lev, lat)],
@@
 ```{r eof2-regr-gh, fig.cap = "(ref:eof2-regr-gh-cap)", fig.height=1.3*width_column, fig.width=width_column, dependson="plot_regr"}
-plot_regr(regr, "PC2", c(200, 50), "hgt")
+plot_regr(regr, "cEOF2", c(200, 50), "hgt")
 
@@
 Figure\ \@ref(fig:eof2-regr-gh) shows the regression pattern of geopotential height and the cEOF2.
-Unlike for cEOF1, in this case the regression patterns are similar to the fully zonally assymetric patterns from Figure\ \@ref(fig:ceofs-1).
+Unlike for cEOF1, in this case the regression patterns are similar to the fully zonally asymmetric patterns from Figure\ \@ref(fig:ceofs-1).
 Although there are some differences (particularly in 50 hPa), the wave trains identified before are well characterised and patterns associated with the real cEOF2 are 90º out of phase with those associated with the imaginary cEOF2.
@@
 Comparing the location of the positive anomaly near 90ºW in column b of Figure\ \@ref(fig:eof2-regr-gh) with Figures 1.a and b from @mo2001, the real cEOF2 can roughly be identified with PSA2, while the imaginary cEOF2 resembles PSA1.
-This separation is not completely meaningful in this case, since by treating this pattern as a single wave train any one pair of orthogonal waves can characterise this variability \textcolor{red}{(No me quedo del todo claro el sentido de esta última oración.
-Si podes ampliarla o reescribirla de otra manera, mejor)}.
 
@@
 #   .[lev %in% c(200, 50)] %>%
-#   # .[PC == "PC1"] %>%
-#   .[, FitWave(estimate, 0:3), by = .(lev, PC, term, lat)] %>%
-#   # .[, pseudor2 := amplitudepseudor2^2/sum(amplitude^2), by = .(lev, PC, term, lat)] %>% 
+#   # .[cEOF == "cEOF1"] %>%
+#   .[, FitWave(estimate, 0:3), by = .(lev, cEOF, term, lat)] %>%
+#   # .[, pseudor2 := amplitudepseudor2^2/sum(amplitude^2), by = .(lev, cEOF, term, lat)] %>% 
 #   ggplot(aes(lat, amplitude)) +
 #   geom_line(aes(color = factor(k))) +
-#   ggh4x::facet_nested(PC + lev ~ term, scales = "free", labeller = labeller(lev = lab_lev)) +
+#   ggh4x::facet_nested(cEOF + lev ~ term, scales = "free", labeller = labeller(lev = lab_lev)) +
 #   scale_color_brewer("Wavenumber", palette = "Dark2") +
@@
 ```{r eof1-regr-t, fig.cap = "(ref:eof1-regr-t-cap)",  dependson="plot_regr", fig.height=1.3*width_column, fig.width=width_column}
-plot_regr(regr, "PC1", c(200, 50), "t")
+plot_regr(regr, "cEOF1", c(200, 50), "t")
 ```
@@
   sep_ReIm(format = "wider") %>% 
-  .[, ":="(Real = as.numeric(scale(Real)), Imaginary = as.numeric(scale(Imaginary))), by = .(PC)] %>% 
+  .[, ":="(Real = as.numeric(scale(Real)), Imaginary = as.numeric(scale(Imaginary))), by = .(cEOF)] %>% 
   .[, FitLm(value, Real, Imaginary, se = TRUE), by = .(lev, lon, variable)] %>% 
@@
 Figure\ \@ref(fig:eof1-regr-t) shows regression patterns of air temperature at 50hPa and 200hPa onto cEOF1.
-In both levels, the distribution of temperature anomalies \textcolor{red}{(Ojo que cuando se habla de regresiones uno no muestra anomalías sino coeficientes de regresión)} mirror the geopotential height anomalies in 50 hPa (Fig.
-XX) \textcolor{red}{(Dar un poco más de detalle de lo que se ve, aunque sea lo mismo de antes)}.
+In both levels, the Real cEOF1 is associated with positive monopole over the South Pole with its centre moved slightly towards 150ºE (Fig.\ \@ref(fig:eof1-regr-t) column 1).
+Regression maps on the Imaginary cEOF1, on the other hand, show a more clear wae 1 pattern with its maximum around 60ºE.
+The distribution of temperature regression coefficients at 50 hPa and at 200 hPa mirror the geopotential height regression maps at 50 hPa (Fig.\ \@ref(fig:eof1-regr-gh)).
+
 The vertical distribution of these patterns is shown in Figure\ \@ref(fig:t-vertical)) that shows the regression coefficientes between cEOF1 and zonal anomalies of air temperature and zonal anomalies of ozone mixing ratio averaged between 80ºS and 35ºS.
-Temperature zonal anomalies associated with cEOF1 show a clear wave 1 pattern for both real and imaginary components throughout the atmosphere above 200 hPa  \textcolor{red}{(¿Empiezan en 200 hPa los contornos?)}  with a change in sign above 10 hPa.
+Temperature zonal anomalies associated with cEOF1 show a clear wave 1 pattern for both real and imaginary components throughout the atmosphere above 250 hPa  with a change in sign above 10 hPa.
 Following hydrostatic balance, this is the level in which the geopotential anomaly have maximum amplitude (not shown).
 
-The maximum ozone anomalies are co-located with the minimum temperature anomalies above 10 hPa and with the maximum temperature anomalies below 10 hPa (Fig.\ \@ref(fig:t-vertical))); that is, the ozone zonal wave 1 is anticorrelated with the temperature zonal wave 1 in the upper stratosphere, and directly correlated in the upper stratosphere  \textcolor{red}{(¿Sabes si se le puede creer mucho a los datos allá arriba o agregamos alguna nota precautoria?)} .
+The maximum ozone anomalies are co-located with the minimum temperature anomalies above 10 hPa and with the maximum temperature anomalies below 10 hPa (Fig.\ \@ref(fig:t-vertical))); that is, the ozone zonal wave 1 is anticorrelated with the temperature zonal wave 1 in the upper stratosphere, and directly correlated in the upper stratosphere .
 This change in phase is observed in ozone anomalies forced by planetary waves that reach the stratosphere.
@@
 #   sep_ReIm(format = "wider") %>% 
-#   .[, ":="(Real = as.numeric(scale(Real)), Imaginary = as.numeric(scale(Imaginary))), by = .(PC)] %>% 
+#   .[, ":="(Real = as.numeric(scale(Real)), Imaginary = as.numeric(scale(Imaginary))), by = .(cEOF)] %>% 
 #   .[, FitLm(value, Real, Imaginary, se = TRUE), by = .(lev, lon, variable)] %>% 
@@
 temp <- ceof$eof[[1]]$left %>% 
-  .[PC == "PC1"] %>% 
+  .[cEOF == "cEOF1"] %>% 
   copy() %>% 
@@
            Imaginary = as.numeric(scale(Imaginary))),
-    by = .(PC)]
+    by = .(cEOF)]
 o3_mean <- o3 %>% 
@@
   o3[., on = "time", allow.cartesian = TRUE] %>% 
-  .[, FitLm(toc, Real, Imaginary, se = TRUE), by = .(lon, lat, PC)] %>% 
+  .[, FitLm(toc, Real, Imaginary, se = TRUE), by = .(lon, lat, cEOF)] %>% 
   rm_intercept() %>% 
@@
   periodic(lon = c(0, 360)) %>% 
-  .[, p.value := Pvaluate(estimate, std.error, df, "fdr"), by = .(PC, term)] %>% 
+  .[, p.value := Pvaluate(estimate, std.error, df, "fdr"), by = .(cEOF, term)] %>% 
   ggplot(aes(lon, lat)) +
@@
   coord_polar() +
-  facet_grid(PC~term) +
+  facet_grid(cEOF~term) +
   axis_labs_smol +
@@
 temp <- ceof$eof[[1]]$left %>% 
-  .[PC == "PC1"] %>% 
+  .[cEOF == "cEOF1"] %>% 
   .[wave1_o3, on = "time"]
@@
 #   geom_smooth(method = "lm", color = "black") +
-#   facet_grid(PC ~ part) +
+#   facet_grid(cEOF ~ part) +
 #   panel_background
@@
 ```{r psa-eof2, fig.cap = "(ref:psa-eof2-cap)", fig.width=7, fig.height=3, fig.env="figure*"}
-pc2 <- ceof$eof[[1]]$left[PC == "PC2"] %>% 
+cEOF2 <- ceof$eof[[1]]$left[cEOF == "cEOF2"] %>% 
   copy() %>% 
-  .[, PC := NULL] %>% 
+  .[, cEOF := NULL] %>% 
   .[]
@@
 rotations_psa <- lapply(angles, function(a) {
-  pc2 %>% 
+  cEOF2 %>% 
     copy() %>% 
@@
 
-pc2[psa_time, on = "time"] %>% 
+cEOF2[psa_time, on = "time"] %>% 
   .[PC != "SAM"] %>% 
@@
   sep_ReIm() %>%
-  .[, hgt := scale(hgt), by = .(PC, part)]
+  .[, hgt := scale(hgt), by = .(cEOF, part)]
 
@@
   .[time, on = .NATURAL, allow.cartesian = TRUE] %>%
-  .[, correlate(hgt, estimate), by = .(lev, term, part, PC)]
+  .[, correlate(hgt, estimate), by = .(lev, term, part, cEOF)]
 
 max_r2 <- sam_r2[part == "Imaginary" & term == "asym"] %>% 
-  .[, .SD[which.max(estimate^2)], by = .(PC)]
+  .[, .SD[which.max(estimate^2)], by = .(cEOF)]
 
@@
                                                          sym = "S-SAM")) +
-  facet_grid(part ~ PC, labeller = labeller(PC = c(PC1 = "EOF1", PC2 = "EOF2"))) +
+  facet_grid(part ~ cEOF) +
   coord_flip() +
@@
 
-The Imaginary EOF2 is related with the SAM through the A-SAM in the troposphere, with up to \textcolor{red}{(Acá aparece un -)} `r scales::percent(max_r2[PC == "PC2"]$estimate, 2)` of shared variance, reached at `r max_r2[PC == "PC2"]$lev` hPa (Figure\ \@ref(fig:sam-eof-vertical).b2).
+The Imaginary EOF2 is related with the SAM through the A-SAM in the troposphere, with up to \textcolor{red}{(Acá aparece un -)} `r scales::percent(max_r2[cEOF == "cEOF2"]$estimate, 2)` of shared variance, reached at `r max_r2[cEOF == "cEOF2"]$lev` hPa (Figure\ \@ref(fig:sam-eof-vertical).b2).
 
@@
 eof_filtered <- era5 %>% 
-  .[sep_ReIm(pc2, format = "wide")[, hgt := NULL], on = .NATURAL] %>% 
+  .[sep_ReIm(cEOF2, format = "wide")[, hgt := NULL], on = .NATURAL] %>% 
   .[lev == 200] %>% 
@@
 fake_eof <- era5 %>% 
-  .[sep_ReIm(pc2, format = "wide")[, hgt := NULL], on = .NATURAL] %>% 
+  .[sep_ReIm(cEOF2, format = "wide")[, hgt := NULL], on = .NATURAL] %>% 
   .[lev == 200] %>% 
@@
   .[season(time) == "SON"] %>% 
-  .[, hgt := hgt - .lm.fit(cbind(1, Imaginary, Real), hgt)$residuals, .(lon, lat, lev, season(time))] %>% 
+  .[, hgt := hgt - .lm.fit(cbind(1, Imaginary, Real), hgt)$residuals,
+    .(lon, lat, lev, season(time))] %>% 
   .[, hgt := Anomaly(hgt), by = .(lat)] %>% 
@@
 # era5 %>% 
-#   .[sep_ReIm(pc2, format = "wide")[, hgt := NULL], on = .NATURAL] %>% 
+#   .[sep_ReIm(cEOF2, format = "wide")[, hgt := NULL], on = .NATURAL] %>% 
 #   .[lev == 200] %>% 
@@
   .[lat <= -20] %>% 
-  # .[, eof := Anomaly(eof), by = .(lat, PC)] %>%
+  # .[, eof := Anomaly(eof), by = .(lat, cEOF)] %>%
   periodic(lon = c(0, 360)) %>% 
@@
   .[, hgt := NULL] %>%
-  .[, `:=`(Real = Real/sd(Real), Imaginary = Imaginary/sd(Imaginary)), by = .(season, PC)] %>%
+  .[, `:=`(Real = Real/sd(Real), Imaginary = Imaginary/sd(Imaginary)), 
+    by = .(season, cEOF)] %>%
   .[psi, on = "time", allow.cartesian = TRUE] %>%
   .[season(time) == "SON"] %>%
-  .[, FitLm(psi.z, Real, Imaginary, se = TRUE),by = .(lon, lat, lev, PC, season)] %>%
+  .[, FitLm(psi.z, Real, Imaginary, se = TRUE),by = .(lon, lat, lev, cEOF, season)] %>%
   rm_intercept() %>%
-  .[, psi.z := Anomaly(estimate), by = .(PC, season, lev, lat)] %>%
-  .[, c("fx", "fy") := shceof::WaveFlux(.SD, p = 200), by = .(lev, PC, season, term)] %>%
+  .[, psi.z := Anomaly(estimate), by = .(cEOF, season, lev, lat)] %>%
+  .[, c("fx", "fy") := shceof::WaveFlux(.SD, p = 200), by = .(lev, cEOF, season, term)] %>%
   .[, term := factor_ReIm(term)]
@@
   .[, hgt := NULL] %>%
-  .[, `:=`(Real = Real/sd(Real), Imaginary = Imaginary/sd(Imaginary)), by = .(season, PC)] %>%
+  .[, `:=`(Real = Real/sd(Real), Imaginary = Imaginary/sd(Imaginary)), 
+    by = .(season, cEOF)] %>%
   .[sst, on = "time", allow.cartesian = TRUE] %>%
   .[season(time) == "SON"] %>%
-  .[, FitLm(t, Imaginary, Real, se = TRUE), by = .(lon, lat, PC, season)] %>%
+  .[, FitLm(t, Imaginary, Real, se = TRUE), by = .(lon, lat, cEOF, season)] %>%
   rm_intercept() %>%
@@
   sep_ReIm() %>%
-  .[, hgt := scale(hgt), by = .(PC, part)] %>%
+  .[, hgt := scale(hgt), by = .(cEOF, part)] %>%
   enso[., on = "time"] %>%
   na.omit() %>%
-  .[, correlate(hgt, oni), by = .(PC, part)]
+  .[, correlate(hgt, oni), by = .(cEOF, part)]
 ```
@@
 ```{r plot_sst_psi}
-plot_sst_psi <- function(which_pc, which_lev) {
-  sst_gdata <- regr_sst[PC == which_pc] %>%
+plot_sst_psi <- function(which_cEOF, which_lev) {
+  sst_gdata <- regr_sst[cEOF == which_cEOF] %>%
     .[lat %between% c(-90, 10)] %>%
-    .[, p_val := Pvaluate(estimate, std.error, df, "fdr"), by = .(PC, term)] %>%
+    .[, p_val := Pvaluate(estimate, std.error, df, "fdr"), by = .(cEOF, term)] %>%
     .[, var := "SST"]
@@
   regr_psi %>%
-    .[PC == which_pc] %>%
+    .[cEOF == which_cEOF] %>%
     .[lev == which_lev] %>%
@@
     .[lat %between% c(-90, 10)] %>%
-    .[, p_val := Pvaluate(estimate, std.error, df, "fdr"), by = .(PC, term)] %>%
+    .[, p_val := Pvaluate(estimate, std.error, df, "fdr"), by = .(cEOF, term)] %>%
     .[, var := "Streamfunction"] %>%
@@
     scale_mag(guide = "none") +
-    facet_grid(term ~ var, labeller = labeller(PC = c(PC1 = "EOF1", PC2 = "EOF2"))) +
+    facet_grid(term ~ var) +
     coord_quickmap(ylim = c(NA, 10)) +
     theme(legend.box = "horizontal") +
-    # ggh4x::facet_nested(term ~ PC) +
     axis_labs_smol +
@@
 ```{r sst-psi-2,  fig.cap = "(ref:sst-psi-2-cap)", fig.width = 7, fig.height=4, dependson="plot_sst_psi", fig.env="figure*"}
-plot_sst_psi(which_pc = "PC2", which_lev = 200)
+plot_sst_psi(which_cEOF = "cEOF2", which_lev = 200)
 ```
@@
 The Imaginary EOF2 is associated with strong positive SST anomalies on the Central Pacific and negative anomalies on an area across the North of Australia and New Zealand, the South Pacific Convergence Zone (SPCZ) (Figure\ \@ref(fig:sst-psi-2).a2).
-This pattern is almost canonically positive ENSO and indeed, the correlation between the Imaginary EOF2 and the Oceanic Niño Index [@bamston1997] is very high `r enso_cor[PC == "PC2" & part == "Imaginary"]$text`.
+This pattern is almost canonically positive ENSO and indeed, the correlation between the Imaginary EOF2 and the Oceanic Niño Index [@bamston1997] is very high `r enso_cor[cEOF == "cEOF2" & part == "Imaginary"]$text`.
 Streamfunction anomalies show a coherent picture.
@@
 However, Figure\ \@ref(fig:sst-psi-2).a1 and b1 show that the Real EOF2 is not associated either with any significant SST nor streamfunction anomalies in the tropics.
-The correlation between the Real EOF2 and ENSO is also not significant (`r enso_cor[PC == "PC2" & part == "Real"]$cor_text`) \textcolor{red}{(No aparece el valor en el PDF)}.
+The correlation between the Real EOF2 and ENSO is also not significant (`r enso_cor[cEOF == "cEOF2" & part == "Real"]$cor_text`) \textcolor{red}{(No aparece el valor en el PDF)}.
 This lack of tropical signal suggests different natures of the different phases of the EOF2 wave train.
@@
 gdata <- ceof[, eof[[1]]$left, by = .(season)] %>%
-  .[PC == "PC2"] %>% 
+  .[cEOF == "cEOF2"] %>% 
   # .[, hgt := Anomaly(hgt)] %>% 
@@
 ```{r sst-psi-1, fig.cap = "(ref:sst-psi-1-cap)", fig.width = 7, fig.height=4, dependson="plot_sst_psi", fig.env = "figure*"}
-plot_sst_psi(which_pc = "PC1", which_lev = 200)
+plot_sst_psi(which_cEOF = "cEOF1", which_lev = 200)
 ```
@@
   .[, hgt := NULL] %>%
-  .[, `:=`(Real = Real/sd(Real), Imaginary = Imaginary/sd(Imaginary)), by = .(season, PC)] %>%
+  .[, `:=`(Real = Real/sd(Real), Imaginary = Imaginary/sd(Imaginary)), 
+    by = .(season, cEOF)] %>%
   .[melt(cmap, id.vars = c("lon", "lat", "time")), on = "time", allow.cartesian = TRUE] %>%
   .[season == "SON"] %>%
-  .[, FitLm(value, Imaginary, Real, se = TRUE), by = .(lon, lat, PC, variable, season)] %>%
+  .[, FitLm(value, Imaginary, Real, se = TRUE), by = .(lon, lat, cEOF, variable, season)] %>%
   rm_intercept() %>% 
@@
 ```{r plot_pp}
-plot_pp <- function(data, which_pc, which_term = c("Real", "Imaginary"), which_continent = NULL) {
+plot_pp <- function(data, which_cEOF, which_term = c("Real", "Imaginary"), which_continent = NULL) {
   filter_continent <- function(data) {
@@
   data %>%
-    .[PC %in% which_pc] %>%
+    .[cEOF %in% which_cEOF] %>%
     .[term %in% which_term] %>%
@@
     # .[abs(estimate) <= 2] %>%
-    .[, p.value := Pvaluate(estimate, std.error, df, "none"), by = .(lev, PC, term)] %>%
+    .[, p.value := Pvaluate(estimate, std.error, df, "none"), by = .(lev, cEOF, term)] %>%
     periodic(lon = c(0, 360)) %>%
@@
     scale_y_latitude(ticks = 15) +
-    ggh4x::facet_nested(term ~ PC) +
+    ggh4x::facet_nested(term ~ cEOF) +
     axis_labs_smol +
@@
 ```{r pp-america, fig.cap = "(ref:pp-america-cap)", fig.width=width_column, fig.height=width_column*1.1}
-plot_pp(data = regr_pp, which_pc = c("PC1", "PC2"), which_continent = "america") 
+plot_pp(data = regr_pp, which_cEOF = c("cEOF1", "cEOF2"), which_continent = "america") 
   # annotate("rect", xmin = min(sesa$lon), xmax = max(sesa$lon), 
@@
 ```{r box_australia}
-box_australia <- data.frame(PC = "PC2", 
+box_australia <- data.frame(cEOF = "cEOF2", 
                             lon = c(130, 155),
@@
 ```{r pp-oceania, fig.cap = "(ref:pp-oceania-cap)", fig.width=width_column, fig.height=width_column*.8}
-plot_pp(regr_pp, c("PC1", "PC2"), which_continent =  "oceania") +
+plot_pp(regr_pp, c("cEOF1", "cEOF2"), which_continent =  "oceania") +
   geom_rect(data = box_australia, aes(xmin = min(lon), xmax = max(lon),
@@
   .[ceof$eof[[1]]$left, on  = "time"] %>% 
-  .[PC == "PC2"] %>% 
+  .[cEOF == "cEOF2"] %>% 
   .[, arg := Arg(hgt)] %>% 
@@
 ```{r pp-africa, fig.cap = "(ref:pp-africa-cap)"}
-# plot_pp(regr_pp, c("PC1", "PC2"), which_continent = "africa")
+# plot_pp(regr_pp, c("cEOF1", "cEOF2"), which_continent = "africa")
 ```

### 2021-04-06 analysis/paper/paper.tex
Muchos cambios

@@
 \linenumbers
-\usepackage{pdfcomment}
 \usepackage{booktabs}
@@
         Elio Campitelli \at
-     Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos, Buenos Aires, Argentina
-CONICET -- Universidad de Buenos Aires, Centro de Investigaciones del Mar y la Atmósfera (CIMA), Buenos Aires, Argentina
-CNRS -- IRD -- CONICET -- UBA, Instituto Franco‐Argentino para el Estudio del Clima y sus Impactos (UMI 3351 IFAECI), Buenos Aires, Argentina \\
+     Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos. Buenos Aires, Argentina.
+CONICET -- Universidad de Buenos Aires. Centro de Investigaciones del Mar y la Atmósfera (CIMA). Buenos Aires, Argentina.
+CNRS -- IRD -- CONICET -- UBA. Instituto Franco-Argentino para el Estudio del Clima y sus Impactos (IRL 3351 IFAECI). Buenos Aires, Argentina. \\
      \email{\href{mailto:elio.campitelli@cima.fcen.uba.ar}{\nolinkurl{elio.campitelli@cima.fcen.uba.ar}}}  %  \\
@@
         Leandro B. Díaz \at
-     Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos, Buenos Aires, Argentina
-CONICET -- Universidad de Buenos Aires, Centro de Investigaciones del Mar y la Atmósfera (CIMA), Buenos Aires, Argentina
-CNRS -- IRD -- CONICET -- UBA, Instituto Franco‐Argentino para el Estudio del Clima y sus Impactos (UMI 3351 IFAECI), Buenos Aires, Argentina \\
+     Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos. Buenos Aires, Argentina.
+CONICET -- Universidad de Buenos Aires. Centro de Investigaciones del Mar y la Atmósfera (CIMA). Buenos Aires, Argentina.
+CNRS -- IRD -- CONICET -- UBA. Instituto Franco-Argentino para el Estudio del Clima y sus Impactos (IRL 3351 IFAECI). Buenos Aires, Argentina. \\
      %  \\
@@
         Carolina Vera \at
-     Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos, Buenos Aires, Argentina
-CONICET -- Universidad de Buenos Aires, Centro de Investigaciones del Mar y la Atmósfera (CIMA), Buenos Aires, Argentina
-CNRS -- IRD -- CONICET -- UBA, Instituto Franco‐Argentino para el Estudio del Clima y sus Impactos (UMI 3351 IFAECI), Buenos Aires, Argentina \\
+     Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos. Buenos Aires, Argentina.
+CONICET -- Universidad de Buenos Aires. Centro de Investigaciones del Mar y la Atmósfera (CIMA). Buenos Aires, Argentina.
+CNRS -- IRD -- CONICET -- UBA. Instituto Franco-Argentino para el Estudio del Clima y sus Impactos (IRL 3351 IFAECI). Buenos Aires, Argentina. \\
      %  \\
@@
 
-\pdfcomment[]{¿Agrego algo sobre impactos? Por ejemplo, impactos de PSA /SAM.}
-
-\hypertarget{methods}{%
-\section{Methods}\label{methods}}
+\hypertarget{data-and-methods}{%
+\section{Data and Methods}\label{data-and-methods}}
 
@@
 
-We used monthly geopotential height, air temperature, ozone mixing ratio, and Total Ozone COlumn (TOC) at 2.5\degree~longitude by 2.5\degree~latitude of horizontal resolution and 37 vertical isobaric levels from ERA5 \citep{era5} for the period 1979 to 2019.
+We used monthly geopotential height, air temperature, ozone mixing ratio, and Total Ozone Column (TOC) at 2.5\degree~longitude by 2.5\degree~latitude of horizontal resolution and 37 vertical isobaric levels from ERA5 \citep{era5} for the period 1979 to 2019.
 We restricted most of our analysis to the post-satellite era to avoid any confounding factors arising from the incorporation of satellite observations, but we also used the preliminary back extension of ERA5 from 1950 to 1978 \citep{era5be} to look at long-term trends.
-We derived streamfunction at 200 hPa from ERA5 vorticity using the FORTRAN subroutine FISHPACK \citep{fishpack} and computed horizontal wave activity fluxes following \citet{plumb1985} (Equation 5.7).
-For precipitation data we used monthly data from the CPC Merged Analysis of Precipitation \citep{cmap}.
+We derived streamfunction at 200 hPa from ERA5 vorticity using the FORTRAN subroutine FISHPACK \citep{fishpack} and computed horizontal wave activity fluxes following \citet{plumb1985}.
+For precipitation data we used monthly data from the CPC Merged Analysis of Precipitation \citep{cmap}, with a 2.5\degree~resolution in latitude and longitude.
+This rainfall gridded dataset is based on information from different sources such as rain gauge observations, satellite inferred estimations and the NCEP-NCAR reanalysis, and it is available since 1979 to present.
 
-We computed the amplitude and phase of the zonal wave 1 by averaging variables (temperature, geopotential height, ozone mixing ratio and Total Ozone Column) 70°S and 45°S for each Spring and computing the Fourier spectrum.
+\hypertarget{methods}{%
+\subsection{Methods}\label{methods}}
 
-We computed the level-dependent Southern Annular Mode index as the leading EOF of year-round monthly geopotential height anomalies south of 20ºS at each level for the whole period \citep{baldwin2009}.
+We restrict our study to the September-October-November (SON) trimester.
+For that, we use seasonal means, averaging monthly values weighted by the number of days in each month.
+We also use 200 hPa to represent the high troposphere and 50 hPa to represent the lower stratosphere.
+
+We computed the amplitude and phase of the zonal wave 1 by averaging (area-weighted) variables (temperature, geopotential height, ozone mixing ratio and Total Ozone Column) between 70°S and 45°S for each SON and extracting the wave-1 component of the Fourier spectrum.
+We chose this latitude band because it is wide enough to capture most of the relevant anomalies.
+
+We computed the level-dependent Southern Annular Mode (SAM) index as the leading EOF of year-round monthly geopotential height anomalies south of 20ºS at each level for the whole period \citep{baldwin2009}.
 We further split the SAM into its zonally symmetric and zonally asymmetric components (S-SAM and A-SAM indices respectively).
@@
 
-\hypertarget{complex-empirical-orthogonal-functions}{%
-\subsection{Complex Empirical Orthogonal Functions}\label{complex-empirical-orthogonal-functions}}
+\textcolor{red}{Creo que habría que expandir un poco esta parte e ir comentando todoas las metodologías que vas usando: regresiones, tendencias, cómo evlauas significancia en cada caso, etc. }
+
+\hypertarget{complex-empirical-orthogonal-functions-ceof}{%
+\subsection{Complex Empirical Orthogonal Functions (cEOF)}\label{complex-empirical-orthogonal-functions-ceof}}
 
@@
 \includegraphics{../figures/eof-naive-1.pdf}
-\caption{\label{fig:eof-naive}Spatial patterns of the leading 4 EOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2009 (arbitrary units).
+\caption{\label{fig:eof-naive}Spatial patterns of the leading 4 EOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2019 (arbitrary units).
 The numbers at the top-left of each panel ate the zonal variance explained by each EOF.}
@@
 
-In traditional EOF analysis zonal waves appear as pairs of EOFs, usually degenerate, that represent similar patterns but shifted in phase.
-For instance, Figure~\ref{fig:eof-naive} shows the leading 4 EOFs of zonally anomalous September through October geopotential height at 50 hPa and 100 hPa.
-It is clear that at 50 hPa the first two EFS represent the same zonal wave 1 pattern and the last 2 represent the same zonal wave 3 pattern shifted by 1/4 wavelength.
-Since these pairs of EOFs seem to represent the same phase-varying structure, it would be desirable to combine them into a single index with amplitude and phase.
+In traditional EOF analysis zonal waves appear as pairs of EOFs, usually degenerate, that represent similar patterns but shifted in phase \citep{horel1984}.
+For instance, Figure~\ref{fig:eof-naive} shows the leading 4 EOFs of zonally anomalous SON geopotential height at 50 hPa and 200 hPa.
+It is clear that at 50 hPa the first two EOFs represent the same zonal wave 1 pattern and the last 2 represent the same zonal wave pattern with shorter wavenumber shifted by 1/4 wavelength.
+A similar EOF structure can be seen in 200 hPa (not shown).
+Since each pair of EOFs seem to represent the same phase-varying structure, it would be desirable to combine them into a single index with amplitude and phase.
 
 Complex Empirical Orthogonal Functions (cEOF) is a useful method to characterise these waves \citep{horel1984}.
-This methods involves augmenting the original fields with the Hilbert transform of the data and then computing the Singular Value Decomposition of it.
+This method involves augmenting the original fields with the Hilbert transform of the data and then computing the Singular Value Decomposition of it.
 In this work, instead of applying the Hilbert transform to the time series at each point, we apply it to each latitude circle at each moment in time.
 Since the latitude circle is a periodic domain, this procedure does not suffer from edge effects.
-
 The result of the cEOF is a set of complex spatial patterns and complex time series.
@@
 
-\caption{\label{tab:corr-ceof-splitted}$R^2$ of the absolute magnitude of complex EOFs between 200 hPa and 50 hPa computing EOF separatedly for each level.}
+\caption{\label{tab:corr-ceof-splitted}Coefficient of determination ($R^2$) of the absolute magnitude of complex EOFs between 200 hPa and 50 hPa computing EOF separatedly for each level.}
 \centering
@@
 \cmidrule(l{3pt}r{3pt}){2-4}
-200 hPa & PC1 & PC2 & PC3\\
+200 hPa & cEOF1 & cEOF2 & cEOF3\\
 \midrule
-PC1 & \cellcolor[HTML]{E1C0BB}{\textcolor{black}{0.28}} & \cellcolor[HTML]{FDFAFA}{\textcolor{black}{0.02}} & \cellcolor[HTML]{FDFAFA}{\textcolor{black}{0.02}}\\
-PC2 & \cellcolor[HTML]{FFFFFF}{\textcolor{black}{0.00}} & \cellcolor[HTML]{BB7A73}{\textcolor{white}{0.60}} & \cellcolor[HTML]{FDFAFA}{\textcolor{black}{0.02}}\\
-PC3 & \cellcolor[HTML]{FFFFFF}{\textcolor{black}{0.00}} & \cellcolor[HTML]{FFFFFF}{\textcolor{black}{0.00}} & \cellcolor[HTML]{FDFCFC}{\textcolor{black}{0.01}}\\
+cEOF1 & \cellcolor[HTML]{E1C0BB}{\textcolor{black}{0.28}} & \cellcolor[HTML]{FDFAFA}{\textcolor{black}{0.02}} & \cellcolor[HTML]{FDFAFA}{\textcolor{black}{0.02}}\\
+cEOF2 & \cellcolor[HTML]{FFFFFF}{\textcolor{black}{0.00}} & \cellcolor[HTML]{BB7A73}{\textcolor{white}{0.60}} & \cellcolor[HTML]{FDFAFA}{\textcolor{black}{0.02}}\\
+cEOF3 & \cellcolor[HTML]{FFFFFF}{\textcolor{black}{0.00}} & \cellcolor[HTML]{FFFFFF}{\textcolor{black}{0.00}} & \cellcolor[HTML]{FDFCFC}{\textcolor{black}{0.01}}\\
 \bottomrule
@@
 
-Another characteristic seen in Figure~\ref{fig:eof-naive} is that EOF1 at 200 hPa is very similar to EFO1 at 50 hPa.
-The zonal wave 3 structure in EOF2 and EOF3 at 200 hPa are also very similar to the one in EOF3 and EOF4 at 50 hPa.
-Furthermore, Table \ref{tab:corr-ceof-splitted} shows the coefficient of determination between time series of the amplitude of each complex EOF across levels.
-There's a high degree of correlation between the respective EOF1 and EOF2 at each level.
-Both the spatial similarities and temporal correlations suggest some level of joint variability between levels.
+Table \ref{tab:corr-ceof-splitted} shows the coefficient of determination between time series of the amplitude of each complex EOF (cEOF) across levels.
+There's a high degree of correlation between the respective cEOF1 and cEOF2 at each level.
+The spatial patterns of the 50 hPa and 200 hPa cEOFs are also similar (not shown)
 
-Both observations motivate the decision of performing complex EOF jointly between levels.
-Since the variability of the stratosphere is far greater in magnitude than the variability of the troposphere, we standardised the values by the standard deviation of each level.
-The computation of the EOFs was carried out using data from both levels at the same time, therefore, each complex EOF has a spatial part that depends on longitude, latitude and level, an a temporal part that depends only on time.
+Both the spatial pattern similarity and the high temporal correlation of cOEFs computed at 50 hPa and 200 hPa suggest that these are, to a large extent, modes of joint variability.
+This motivates the decision of performing complex EOF jointly between levels.
+The computation of the EOFs was carried out using data from both levels at the same time, therefore, each complex EOF has a spatial component that depends on longitude, latitude and level, and a temporal component that depends only on time.
 
@@
 Since any choice is arbitrary and equally valid, we chose the phase of each EOF so that the real and imaginary parts are aligned with meaningful phases in our analysis.
-This was done after some exploratory analysis.
-
-\begin{figure}
-\centering
-\includegraphics{../figures/rotations-1.pdf}
-\caption{\label{fig:rotations}a) Correlation of the Real and Imaginary EOF1 with wave 1 amplitude of Total Ozone Column averaged between 70°S and 45°S and b) \(R^2\) between the Real and Imaginary EOF2 and the ONI index for different values of rotation.
-The vertical line marks the selected rotation, which maximises de correlation.}
-\end{figure}
+This procedure does not create a spurious correlation, it only takes whatever relationship that already exist and aligns it with a specific phase.
 
 For the first complex principal component, the phase was chosen so that the time series corresponding to the real part has the maximum correlation with the zonal wave 1 of Total Ozone Column between 70°S and 45°S.
-This also nearly minimises the correlation with the imaginary part (Figure~\ref{fig:rotations}a).
+This also nearly minimises the correlation with the imaginary part.
 
-For the second complex principal component, the phase was chosen so that the coefficient of determination between the Oceanic Niño Index \citep{bamston1997} and the real part was minimised, which also nearly maximises the correlation with the imaginary part (Figure~\ref{fig:rotations}b).
-It's important to reassure the reader that this procedure does not create a spurious correlation, it only takes whatever relationship that already exist and aligns it with a specific phase.
-If the relationship is very weak, the rotation cannot generate a strong relationship.
+For the second complex principal component, the phase was chosen so that the coefficient of determination between the Oceanic Niño Index \citep{bamston1997} and the real part was minimized, which also nearly maximizes the correlation with the imaginary part.
 
@@
 
-\hypertarget{description-of-eofs}{%
-\subsection{Description of EOFs}\label{description-of-eofs}}
+\hypertarget{ceof-patterns}{%
+\subsection{cEOF patterns}\label{ceof-patterns}}
 
@@
 \includegraphics{../figures/ceofs-1-1.pdf}
-\caption{\label{fig:ceofs-1}Spatial patterns of the leading 2 complex EOFs of zonal anomalies of geopotential height in 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2009.
-Real part in shading, imaginary part in contours}
+\caption{\label{fig:ceofs-1}Spatial patterns of the leading 2 complex EOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2019.
+Real part in shading, imaginary part in contours.}
 \end{figure}
@@
 \includegraphics{../figures/ceofs-2-1.pdf}
-\caption{\label{fig:ceofs-2}Time series of the leading 2 complex EOFs of zonal anomalies of geopotential height in 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2009.}
+\caption{\label{fig:ceofs-2}Time series of the leading 2 complex EOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2019.}
 \end{figure}
 
-Figures \ref{fig:ceofs-1} and \ref{fig:ceofs-2} show, respectively, the spatial and temporal parts of the first two leading complex EOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa.
-In the spatial patterns in Figure~\ref{fig:ceofs-1}, the real (in shading) and the imaginary parts (in contour) are in quadrature by construction, so that each EOF describe a single wave-like pattern whose amplitude and position (i.e.~phase) is controlled by the magnitude and phase of the complex temporal EOF.
+To describe the spatial and temporal variability of the circulation zonal anomalies, the spatial and temporal parts of the first two leading cEOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa are shown in Figures \ref{fig:ceofs-1} and \ref{fig:ceofs-2}.
+The first mode (CEOF1) explains 82\% of the variance, while the second mode (CEOF1) explains a smaller fraction (7\%).
+In the spatial patterns (Fig.~\ref{fig:ceofs-1}), the real (in shading) and the imaginary components (in contour) are in quadrature by construction, so that each cEOF describe a single wave-like pattern whose amplitude and position (i.e.~phase) is controlled by the magnitude and phase of the complex temporal EOF.
+The wave patterns described by these cEOFs match the patterns seen in the traditional EOFs of Figure~\ref{fig:eof-naive}.
 
-The wave patterns described by these complex EOFs match the patterns seen in the traditional EOFs of Figure~\ref{fig:eof-naive}: The first is a wave 1, while the second is a wave 3.
-Note that in Figure~\ref{fig:ceofs-2}, both parts of the EOF1 have non zero mean.
-This is due to th fact that the geopotential fields that enter into the algorithm are anomalies with respect to the zonal mean, not the time mean.
-The variability associated with the first EOF includes variability that projects onto the mean zonally anomalous field.
+The first cEOF (Fig.~\ref{fig:eof-naive} column~a) is a hemispheric wave 1 pattern with maximum amplitude at high latitudes.
+At 50 hPa the Real cEOF1 has the maximum of the wave 1 at 150ºE and at 200 hPa, the maximum is locates at around 175ºE indicating a westerly shift in phase.
+
+The second cEOF (Fig.~\ref{fig:eof-naive} column~b) show also a zonal wave-like structure with maximum amplitude at high latitudes, but with shorter spatial scales.
+In particular, it's dominant structure at both levels is a wave 3 but with larger amplitude in the pacific sector.
+This modulated amplitude is especially apparent at 200 hPa.
+There is no apparent phase shift, suggesting a rather barotropic structure.
+
+The temporal variability for both components of the cEOF1 have non zero mean (Fig.~\ref{fig:ceofs-2}).
+This is due to the fact that the geopotential fields that enter into the cEOFs algorithm are anomalies with respect to the zonal mean, not the time mean.
+As a consequence, the variability associated with the first cEOF includes variability that projects onto the mean zonally anomalous field.
+There are no significant simultanous correlation between the time series.
 
@@
 
-Figure~\ref{fig:extended-series} shows time series of the two complex EOFs extended beyond the satellite era using the preliminary ERA5 back extension going back to 1950.There is an upward trend in the real part of EOF1 (Figure~\ref{fig:extended-series}a.1, p-value~ \textless{} 0.001) and no significant trend in any of the complex parts of EOF2.
-The positive trend in the Real EOF1 translates into a positive trend in EOF1 magnitude, but not in phase (not shown).
+Both cEOF shot year-to-year variability but show no evidence of decadal variabilty.
+The relatively short timespan (40 years) hampers our habilty to detect longer trends.
+Therefore, we extend the cEOF time series using the preliminary ERA5 back extension going back to 1950 by projecting the older geopogential height zonal anomaly fields onto the spatial patterns shown in Figure~\ref{fig:ceofs-1}.
+
+The extended cEOF timeseires are shown in Figure~\ref{fig:extended-series}.
+There is a significant upward trend in the real component of cEOF1 (Fig.~\ref{fig:extended-series}a.1, p-value~\textless{} 0.001) and no significant trend in any of the complex components of cEOF2.
+The positive trend in the real cEOF1 translates into a positive trend in cEOF1 magnitude, but not in the phase (not shown).
+
+This long-term change indicates an increase in the magnitude of the high latitude zonal wave 1.
+A similar observation was made by \citet{raphael2003}, who detected a step after around 1975 in the leading EOF of August-September-October 500 hPa zonal geopotential height anomalies.
 
@@
 
-The spatial patterns shown in Figure~\ref{fig:ceofs-1} are ``pure'' EOF patterns.
-But as they are derived by removing the zonally symmetric circulation, they might not include all the variability that is actually associated with the EOF time series.
-They are also idealised in that they are forced by construction, to be Hilber transforms of each other.
-To get a more realistic view of the real-world variability described by these EOFS, we computed regression patterns of each EOF with the complete geopotential fields.
+The spatial patterns shown in Figure~\ref{fig:ceofs-1} are derived by removing the zonally symmetric circulation, so they might not include all the variability that is actually associated with the cEOF time series.
+To understand the full geopotential height field changes associated with each cEOF, we computed regression patterns of each cEOF with the complete geopotential fields.
 
@@
 
-Figure~\ref{fig:eof1-regr-gh} shows regression patterns of EOF1 and geopotential height.
+Figure~\ref{fig:eof1-regr-gh} shows regression patterns between geopotential height anomalies and cEOF1.
 At 50 hPa (Figure~\ref{fig:eof1-regr-gh} row a), both the Real and Imaginary EOF1 are associated with planetary wave 1 patterns, that are 90º off phase.
-Their phases coincide with the ones shown in Figure~\ref{fig:ceofs-1}a.1, with the positive centre of the Real EOF1 located towards the dateline, and the one of the Imaginary EOF1 located over Eastern Antarctica.
-However, the Real EOF1 pattern is substantially altered by the zonally symmetric circulation.
+Their phases coincide with the ones shown in Figure~\ref{fig:ceofs-1}a.1, with the positive centre of the real cEOF1 located towards the dateline, and the one of the Imaginary cEOF1 located over Eastern Antarctica.
+However, the cReal EOF1 pattern is substantially altered by the zonally symmetric circulation.
 Instead of a clear wave 1 pattern, the regression pattern can be better describes as a monopole with its centre displaced from the South Pole.
 
-Similarly affected are the regression patterns at 200 hPa (Figure~\ref{fig:eof1-regr-gh} row b).
-It is only possible to distinguish some of the wave 1 pattern in relation with the Real EOF1 (Figure~\ref{fig:eof1-regr-gh}b.1).
-The Imaginary EOF1 shows a much more zonally symmetrical pattern resembling a negative Southern Annular Mode.
+The regression patterns at 200 hPa (Figure~\ref{fig:eof1-regr-gh} row b) are similarly influenced by the zonally symmetric circulation.
+It is only possible to distinguish partially the wave 1 pattern in relation with the real cEOF1 (Figure~\ref{fig:eof1-regr-gh}b.1).
+The imaginary cEOF1 shows a much more zonally symmetrical pattern resembling a negative Southern Annular Mode.
 
-With the exception of the Imaginary EFO1, it's clear that these patterns are very different than the idealised versions shown in Figure~\ref{fig:ceofs-1}, particularly at 200 hPa.
+With the exception of the imaginary cEOF1, it is clear that these patterns are very different than the fully zonally asymmetric versions (Fig.~\ref{fig:ceofs-1}), particularly at 200 hPa.
 Moreover, only in the stratosphere these patterns actually show a distinguishable wave 1 pattern shifted in phase by 90º.
-This should not be utterly surprising, since the spatial patterns of the \emph{naive} EOFs only show this feature at 50 hPa, suggesting that using the complex EOF method is ``forcing'' a wave structure where there is none.
-
+This should not be utterly surprising, since the spatial patterns of the \emph{naive} EOFs only show this feature at 50 hPa (Fig.~\ref{fig:eof-naive}), suggesting that using the cEOF method is artificially generating a wave structure at 200hPa.
 Therefore, the magnitude and phase of the EOF1 are associated with the magnitude and phase of a zonal wave only in the stratosphere.
-While in the troposphere they are associated with slightly off-centre monopoles.
+While in the troposphere, they are associated with slightly off-centre monopoles.
 
@@
 
-Figure~\ref{fig:eof2-regr-gh} shows the regression pattern of geopotential height and the EOF2.
-Unlike in the case of EOF1, in this case the regression patterns represent relatively well the \emph{idealised} patterns from Figure~\ref{fig:ceofs-1}.
-Although there are some differences (particularly in 50 hPa), the wave trains identified before are well characterised and patterns associated with the Real EOF2 are 90º out of phase with those associated with the Imaginary EOF2.
-Zonal wave 3 dominates all fields, but only in the western hemisphere.
+Figure~\ref{fig:eof2-regr-gh} shows the regression pattern of geopotential height and the cEOF2.
+Unlike for cEOF1, in this case the regression patterns are similar to the fully zonally asymmetric patterns from Figure~\ref{fig:ceofs-1}.
+Although there are some differences (particularly in 50 hPa), the wave trains identified before are well characterised and patterns associated with the real cEOF2 are 90º out of phase with those associated with the imaginary cEOF2.
+Zonal wave 3 dominates all fields, but only in the western hemisphere, over the Pacific and Atlantic Oceans.
+cEOF2 then represents a equivalent barotropic wave train that is very similar to the the Pacific South American Patterns \citep{mo2001}.
+Comparing the location of the positive anomaly near 90ºW in column b of Figure~\ref{fig:eof2-regr-gh} with Figures 1.a and b from \citet{mo2001}, the real cEOF2 can roughly be identified with PSA2, while the imaginary cEOF2 resembles PSA1.
 
-EOF2 then represents a equivalent barotropic wave train that is very similar to the the Pacific South American Patterns \citep{mo2001}.
-Comparing the location of the positive anomaly near 90ºW in column b of Figure~\ref{fig:eof2-regr-gh} with Figure 1 a and b from \citet{mo2001}, the Real EOF2 can roughly be identified with the PSA2, while the Imaginary EOF2 resembles PSA1.
-This separation is not completely meaningful in this case, since by treating this pattern as a single wave train any one pair of orthogonal waves can characterise this variability.
+\hypertarget{ceofs-relationship-with-other-variables}{%
+\subsection{cEOFs relationship with other variables}\label{ceofs-relationship-with-other-variables}}
 
-\hypertarget{relationship-with-other-variables}{%
-\subsection{Relationship with other variables}\label{relationship-with-other-variables}}
-
-\hypertarget{ozone}{%
-\subsubsection{Ozone}\label{ozone}}
+\hypertarget{temperature-and-ozone}{%
+\subsubsection{Temperature and Ozone}\label{temperature-and-ozone}}
 
@@
 \includegraphics{../figures/t-vertical-1.pdf}
-\caption{\label{fig:t-vertical}Regression of EOF1 with zonal anomalies of mean air temperature (shaded) and ozone mixing ratio (contours, negative contours with dashed lines, labels in parts per billion by mass) averaged between 80ºS and 25ºS for the period 1979 -- 2019.}
+\caption{\label{fig:t-vertical}Regression coefficients between cEOF1 and zonal anomalies of mean air temperature (shaded) and ozone mixing ratio (contours, negative contours with dashed lines, labels in parts per billion by mass) averaged between 80ºS and 35ºS for the period 1979 -- 2019.}
 \end{figure}
 
-Figure~\ref{fig:eof1-regr-t} shows regression patterns of EOF1 with air temperature.
-In both levels, the distribution of temperature anomalies mirror the geopotential height anomalies in 50 hPa.
-To understand the vertical distribution of these anomalies, Figure~\ref{fig:t-vertical}) shows the regression pattern of EOF1 with zonal anomalies of air temperature and zonal anomalies of ozone mixing ratio averaged between 80ºS and 35ºS.
-Temperature zonal anomalies associated with EOF1 show a clear wave 1 pattern throughout the atmosphere above 200 hPa with a change in sign above 10 hPa.
+Figure~\ref{fig:eof1-regr-t} shows regression patterns of air temperature at 50hPa and 200hPa onto cEOF1.
+In both levels, the Real cEOF1 is associated with positive monopole over the South Pole with its centre moved slightly towards 150ºE (Fig.~\ref{fig:eof1-regr-t} column 1).
+Regression maps on the Imaginary cEOF1, on the other hand, show a more clear wae 1 pattern with its maximum around 60ºE.
+The distribution of temperature regression coefficients at 50 hPa and at 200 hPa mirror the geopotential height regression maps at 50 hPa (Fig.~\ref{fig:eof1-regr-gh}).
+
+The vertical distribution of these patterns is shown in Figure~\ref{fig:t-vertical}) that shows the regression coefficientes between cEOF1 and zonal anomalies of air temperature and zonal anomalies of ozone mixing ratio averaged between 80ºS and 35ºS.
+Temperature zonal anomalies associated with cEOF1 show a clear wave 1 pattern for both real and imaginary components throughout the atmosphere above 250 hPa with a change in sign above 10 hPa.
 Following hydrostatic balance, this is the level in which the geopotential anomaly have maximum amplitude (not shown).
 
-The maximum ozone anomalies are co-located with the minimum temperature anomalies above 10 hPa and with the maximum temperature anomalies below 10 hPa; that is, the ozone zonal wave 1 is anticorrelated with the temperature zonal wave 1 in the upper stratosphere, and directly correlated in the upper stratosphere.
-This change in phase is observed in ozone anomalies forced by planetary waves that reach the stratosphere.In the photochemically-dominated upper stratosphere, cold temperatures inhibit the destruction of ozone, and the advectively-dominated lower stratosphere, ozone anomalies are 90º out of phase with horizontal and vertical transport, which is 90º our of phase with temperature anomalies \citep{hartmann1979, wirth1993, smith1995}.
+The maximum ozone anomalies are co-located with the minimum temperature anomalies above 10 hPa and with the maximum temperature anomalies below 10 hPa (Fig.~\ref{fig:t-vertical})); that is, the ozone zonal wave 1 is anticorrelated with the temperature zonal wave 1 in the upper stratosphere, and directly correlated in the upper stratosphere .
+This change in phase is observed in ozone anomalies forced by planetary waves that reach the stratosphere.
+In the photochemically-dominated upper stratosphere, cold temperatures inhibit the destruction of ozone, and the advectively-dominated lower stratosphere, ozone anomalies are 90º out of phase with horizontal and vertical transport, which is 90º our of phase with temperature anomalies \citep{hartmann1979, wirth1993, smith1995}.
 
@@
 \includegraphics{../figures/o3-regr-1.pdf}
-\caption{\label{fig:o3-regr}Regression coefficients of the real and imaginary part of the first complex EOF on SON mean Total Ozone Column.
-On contours, the mean zonal anomaly of Total Ozone Column (negative contours in dashed lines).
+\caption{\label{fig:o3-regr}Regression coefficients of SON mean Total Ozone Column anomalies onto the real (a) and imaginary (b) components of the cEOF1.
+On contours, the mean zonal anomaly of Total OZone Column (negative contours in dashed lines, Dobson Units).
 For the 1979 -- 2019 period.}
@@
 Climatologically, the springtime Ozone minimum is located off the South Pole and towards the Weddell Sea \citep{wirth1993}.
-Thus, the Real EOF1 regression pattern (Figure~\ref{fig:o3-regr}a) coincides with the climatological position of the ozone whole while the one for the Imaginary EOF1 is shifted by 90º.
+Thus, the Real EOF1 regression pattern (Figure~\ref{fig:o3-regr}a) coincides with the climatological position of the ozone hole while the one for the Imaginary EOF1 is shifted by 90º.
 
@@
 
-Indeed, Figure~\ref{fig:wave1-o3} shows the relationship between amplitude and phase of the planetary wave 1 in Total Ozone Column between 70ºS and 45ºS and amplitude and phase of the EOF1.
+Indeed, Figure~\ref{fig:wave1-o3} shows the relationship between amplitude and phase of the planetary wave 1 in Total Ozone Column between 70ºS and 45ºS \textcolor{red}{(Habría que justifica mejor estas elecciones de bandas latitudinales a paritr del gráfico anterior)} and amplitude and phase of the EOF1.
 The correlation between the zonal wave 1 of ozone and EOF1 (computed as the mean cosine of the difference in phase) is 0.86 (CI: 0.76 -- 0.93).
@@
 
-We will show that the EOF2 offers an alternative way of representing the PSA which has several advantages over using the second and third principal components .
-
 \begin{figure*}
@@
 
+The relationship between PSAs and the modes described from the cEOFs is also studied.
 Figure~\ref{fig:psa-eof2} shows the relationship between the two PSA indices and the Real and Imaginary phase of EOF2.
@@
 It is also the same that maximises this clean association between EFO2 parts and PSA modes.
-The reason that the ``naive'' principal component analysis arrives naturally to this optimal separation is that the imaginary phase is the most common phase (Figure~\ref{fig:phase-histogram}), so it naturally appears first when performing a Principal Component Analysis.
+
+\textcolor{red}{(Lo del histograma habría que explicarlo con un pocoo más detalle porque no es tan trivial.
+Primero decir para qué se hace, qué se hace exactamente y después recien decir cuál es el resultado y qué información aporta)}
+
+The reason that the EOF analysis arrives to this optimal separation is that the imaginary cEOF is the most common phase (Figure~\ref{fig:phase-histogram}), so it naturally appears first when performing a Principal Component Analysis.
 
@@
 \includegraphics{../figures/phase-histogram-1.pdf}
-\caption{\label{fig:phase-histogram}Histogram showing phase distribution of EOF2 for the period 1979 -- 2019.}
+\caption{\label{fig:phase-histogram}Histogram showing phase distribution of EOF2 for the period 1979 -- 2019. \textcolor{red}{(También ampliar)}}
 \end{figure}
 
-By aligning he imaginary phase with the direction with maximum relationship with ENSO, we also aligned it with the phase of maximum occurrence.
-This is analogous to the method in \citet{irving2016}, who used reprojection and Fourier filtering to detect ``PSA-like'' variability and defined as the proper PSA using the peaks of the phase distribution (Compare our Figure~\ref{fig:phase-histogram} with their Figure 6).
-
-The advantage of our method is that it is much simpler to implement, it provides magnitude and phase naturally, and it facilitates the description of this mode as a propagating wave instead of as standing oscillation.
+By aligning he imaginary phase with the direction with maximum relationship with ENSO, we also aligned it with the phase of maximum occurrence of XXX.
+This is analogous to the method of \citet{irving2016}, who used reprojection and Fourier filtering to detect ``PSA-like'' variability and defined as the proper PSA using the peaks of the phase distribution (Compare our Figure~@ref:(phase-histogram) with their Figure 6).
+\textcolor{red}{(Amplia un poco la oración anterior que no queda del todo clara)} The advantage of our method is that it is much simpler to implement, it provides magnitude and phase naturally, and it facilitates the description of this mode as a propagating wave instead of as standing oscillation.
+As a consequence, the cEOF2 offers an alternative way of representing the PSA which has several advantages over using the second and third EOFs.
 
@@
 \caption{\label{fig:sam-eof-vertical}Coefficient of determination between the real and imaginary part of each EOF and the SAM, Asymmetric SAM (A-SAM) and Symmetric SAM (S-SAM) indices computed at each level according to \citet{campitelli2021}.
-Points mark estimates with p-value~\textless~0.01 corrected for False Detection Rate \citep{benjamini1995}.}
+Points mark estimates with p-value \textless{} 0.01 corrected for False Detection Rate \citep{benjamini1995}.}
 \end{figure}
 
-Figure~\ref{fig:sam-eof-vertical} shows the coefficient of determination between the EOFs and three SAM indices.
+To explore the relationship between SAM and the modes described from the cEOF, it is computed the coefficient of determination between the cEOFs timeseries and the three SAM indices (SAM, A-SAM and S-SAM) at each level (Fig.~\ref{fig:sam-eof-vertical}).
+Both cEOFs bear some modest relationship with the full SAM index (thick green line in Figure~\ref{fig:sam-eof-vertical}), although not statistically significant at every level.
 
-Both EOFs bear some modest relationship with the SAM index (thick green line in Figure~\ref{fig:sam-eof-vertical}), although not statistically significant at every level.
+\textcolor{red}{(Esta explicación habría que hacerla más detallada. Explicar más sobre lo que se ve en cada panel y después ir interpretando.  )}
 The split between A-SAM and S-SAM gives more insight into the nature of the relationship.
-The relationship between the SAM and the imaginary EOF1 (Figure~\ref{fig:sam-eof-vertical}.b1) is mediated by S-SAM in the troposphere, but by the A-SAM in the stratosphere.
+The relationship between the SAM and ¿the imaginary EOF1?
+(Figure~\ref{fig:sam-eof-vertical}.b1) is mediated by S-SAM in the troposphere, but by the A-SAM in the stratosphere.
 
-The Imaginary EOF2 is related with the SAM through the A-SAM in the troposphere, with up to -96\% of shared variance, reached at 225 hPa (Figure~\ref{fig:sam-eof-vertical}.b2).
+The Imaginary EOF2 is related with the SAM through the A-SAM in the troposphere, with up to \textcolor{red}{(Acá aparece un -)} -96\% of shared variance, reached at 225 hPa (Figure~\ref{fig:sam-eof-vertical}.b2).
 
@@
 
-As a further illustration, Figure~\ref{fig:eof-filtered} shows the spatial pattern of the leading EOF of 200 hPa geopotential height when the variability associated with EOF2 is removed.
+\textcolor{red}{(La discusión de los próximos parráfos me resulta confusa.
+Deberías aclarar mejor por qué mostras cada cosa que mostras y cómo se relaciona con lo que mostraste antes.
+También me resulta como un resultado que se va un poco de tema, por eso creo que debería estar mejor contada esta parte para poder incluirla)}
+
+As a further illustration \textcolor{red}{(¿De qué?)}, Figure~\ref{fig:eof-filtered} shows the spatial pattern of the leading EOF of 200 hPa geopotential height when the variability associated with EOF2 is removed.
 The resulting pattern is an nearly zonally symmetrical annular mode.
@@
 
-This demonstrates that the zonal asymmetries of the SAM could very well be mostly a statistical artifact.
+This suggests that the zonal asymmetries of the SAM could very well be mostly a statistical artifact.
 Since many surface impacts are mediated by the asymmetric component, as well as the relationship between SAM and ENSO \citep{campitelli2021}, this potential issue could affect the interpretability of many results involving the SAM.
@@
 The Imaginary EOF2 is associated with strong positive SST anomalies on the Central Pacific and negative anomalies on an area across the North of Australia and New Zealand, the South Pacific Convergence Zone (SPCZ) (Figure~\ref{fig:sst-psi-2}.a2).
-This pattern is almost canonically positive ENSO and indeed, the correlation between the Imaginary EOF2 and the Oceanic Niño Index (\citep{bamston1997}) is very high 0.76 (CI: 0.59 -- 0.86).
+This pattern is almost canonically positive ENSO and indeed, the correlation between the Imaginary EOF2 and the Oceanic Niño Index \citep{bamston1997} is very high 0.76 (CI: 0.59 -- 0.86).
 Streamfunction anomalies show a coherent picture.
 The Imaginary EOF2 is associated with strong wave-like streamfunction anomalies emanating from the tropics (Figure~\ref{fig:sst-psi-2}.b2).
-This is consistent with what we know of the effect of ENSO on the extratropics: SST anomalies initiate anomalous convection that excites Rossby waves that propagate meridionally towards higher latitudes.
+This is consistent with what we know of the effect of ENSO on the extratropics: SST anomalies initiate anomalous convection that excites Rossby waves that propagate meridionally towards higher latitudes \textcolor{red}{(REFS??)}.
 
@@
 However, Figure~\ref{fig:sst-psi-2}.a1 and b1 show that the Real EOF2 is not associated either with any significant SST nor streamfunction anomalies in the tropics.
-The correlation between the Real EOF2 and ENSO is also not significant ().
-This lack of tropical signal suggests a radically different nature of the different phases of the EOF2 wave train.
+The correlation between the Real EOF2 and ENSO is also not significant () \textcolor{red}{(No aparece el valor en el PDF)}.
+This lack of tropical signal suggests different natures of the different phases of the EOF2 wave train.
 
@@
 
-To better explore the relationship between tropical forcing and phase of the EOF2, Figure~\ref{fig:enso-phase} plots the ONI index and the phase of the EOF2 for each year between 1979 and 2019, highlighting years in which the magnitude of EOF2 was higher than 50\% of the years.
-In years with positive ENSO, the phase of the EOF2 is always around +90º (corresponding with positive imaginary part) and vice versa.
+To better explore the relationship between tropical forcing and phase of the cEOF2, Figure~\ref{fig:enso-phase} plots the ONI index and the phase of the EOF2 for each SON trimester between 1979 and 2019, highlighting years in which the magnitude of EOF2 is above the median.
+In years with positive ONI, the phase of the EOF2 is mostly around +90º (corresponding with positive imaginary part) and vice versa.
 In years with near neutral ENSO, the phase of the EOF2 is much more variable.
-The black line in Figure~\ref{fig:enso-phase} tries to quantify this relationship.
-Is the equation \(ONI = 1.1\sin{(phase)}\), whose coefficient we fitted by weighted least squares using the magnitude of the EFO2 as weight.
-The \(r^2\) corresponding to the fit is 0.56, with p-value \textless{} 0.001.
+The black line in Figure~\ref{fig:enso-phase} is an harmonical fit of the relationship between ONI and cEOF2 phase.
+The \(r^2\) corresponding to the fit is 0.56, with p-value \textless{} 0.001, indicating a quasi-sinusoidal relation between these two variables.
 
-Figure~\ref{fig:enso-phase} suggest that strong EOF2 years tend to coincide with strong ENSO years.
+Furthemore, Figure~\ref{fig:enso-phase} suggest that strong EOF2 years tend to coincide with strong ENSO years.
 The correlation between the absolute magnitude of the ONI and the magnitude of the EOF2 is 0.44 (CI: 0.16 -- 0.66).
@@
 Furthermore, even when using all the datapoints, the Spearman correlation --which is robust to outliers-- is also non-significant (0.2, p-value= 0.2).
-Therefore, the relationship between the magnitude of the EOF2 train wave and ENSO remains uncertain.
+Therefore, the relationship between the magnitude of the cEOF2 train wave and ONI remains uncertain.
 
-All this suggests that the wave train represented by EOF2 can be both forced by tropical SSTs and also a part of the internal variability of the extratropical atmosphere.
-When internally forced, the wave train has little phase preference.
-When excited by tropical SST anomalies, it tends to remain locked to the imaginary phase, with the sign of the geopotential anomalies depending on the sign of the tropical SSTs anomalies.
+It could be concluded that the wave train represented by cEOF2 can be both part of the internal variability of the extratropical atmosphere or forced by tropical SSTs.
+In the former case, the wave train has little phase preference.
+However, when cEOF2 is excited by tropical SST variability, it tends to remain locked to the imaginary phase.
 This explains the relative overabundance of years with EOF2 near positive and negative imaginary phase in Figure~\ref{fig:phase-histogram}.
@@
 
-Figure~\ref{fig:pp-america} shows regression maps of seasonal precipitation with each EOF in South America.
-EOF1 is associated with slight decreases in precipitation in Southern Brazil and Paraguay, although these are not statistically significant \ref{fig:pp-america} column a.
+The influence of cEOFs variability in continental rainfall of extra-tropical Southern Hemisphere is also explored.
+We focus on South America and Oceania sectors, because there are not relevant signals in Southern Africa (not shown).
+
+Figure~\ref{fig:pp-america} shows regression maps of seasonal precipitation with each cEOF in South America.
+cEOF1 is associated with slight decreases in precipitation in Southern Brazil and Paraguay, although these are not statistically significant \ref{fig:pp-america} column a.
 The strongest precipitation anomalies are the ones associated with the Imaginary EOF2.
-The positive anomalies on Southeaster South America (SESA) and Chile, and negative anomalies over Southern Brazil is a well known springtime precipitation signature of ENSO \citep{cai2020a} and it's also virtually identical to the precipitation anomalies associated with the Asymmetric SAM \citep{campitelli2021}.
-This is not surprising considering the close relationship between the ONI, the Asymmetric SAM index and the Imaginary EOF2 shown previously, but further consolidates the identification of this mode with the PSA pattern.
-The Real EOF2, on the other hand, is associated with negative precipitation anomalies in a smaller area of SESA.
-The relationship between precipitation anomalies in SESA and the phase of EOF2 follows a curve similar to that of Figure~\ref{fig:enso-phase} (not shown).
+The positive anomalies on Southeastern South America (SESA) and Central Chile, and negative anomalies over South Atlantic Convergence Zone is a well known springtime precipitation signature of ENSO \citep{cai2020a} and it is also similar to the precipitation anomalies associated with the A-SAM \citep{campitelli2021}.
+This is not surprising considering the close relationship between the ONI, the Asymmetric SAM index and the Imaginary cEOF2 shown previously, but further consolidates the identification of this mode with the PSA pattern.
+The Real cEOF2, on the other hand, is associated with negative precipitation anomalies in a smaller area of SESA.
+There is a phase dependence of the precipitation anomalies in SESA (not shown), which resembles the relationship between ONI and the phase of EOF2.
+\textcolor{red}{(Podrías agregar acá el coeficiente de determinación)}
 
@@
 \includegraphics{../figures/pp-oceania-1.pdf}
-\caption{\label{fig:pp-oceania}Same as Figure~\ref{fig:pp-america} but for New Zealand and neighbouring islands.}
+\caption{\label{fig:pp-oceania}As Figure~\ref{fig:pp-america} but for Oceania.}
 \end{figure}
 
-On the antipodes, precipitation anomalies in New Zealand and neighbouring islands associated with EOF1 are also weak and not statistically significant (Figure~\ref{fig:pp-oceania} column a) whereas EFO2 is associated with large and expansive anomalies.
-The Real EOF2 is associated with positive seasonal anomalies in Northern and Eastern Australia, and negative ones over New Zealand, while precipitation anomalies associated with Imaginary EOF2 are negative over all Eastern Australia.
+The precipitation anomalies in Oceania associated with EOF1 are also weak and not statistically significant (Figure~\ref{fig:pp-oceania} column a) whereas EFO2 is associated with large and expansive anomalies.
+The Real cEOF2 is associated with positive seasonal anomalies in Northern and Eastern Australia, and negative ones over New Zealand, while precipitation anomalies associated with Imaginary EOF2 are negative over all Eastern Australia.
 Again, these negative anomalies are similar to the springtime precipitation anomalies observed in relation with the Asymmetric SAM \citep{campitelli2021}.
+\textcolor{red}{(Habría que agregar alguna referencia del ENSO que es más conocido)}
 
@@
 The black line is the equation \(\mathrm{Precipitation} = -0.26\sin{(\mathrm{phase})} + 0.22\cos{(\mathrm{phase})}\), whose coefficient we fitted by weighted least squares using the magnitude of the EFO2 as weight.
+\textcolor{red}{(Como te puse antes la ecuación como mucho la pondría sobre la figura)}
 The \(r^2\) corresponding to the fit is 0.37, with p-value \textless{} 0.001.
-There is a strong relationship between the phase of the EOF2 and precipitation anomalies in Eastern Australia, but Real and Imaginary phases we chose are not aligned with the direction that maximises this relationship.
+There is a strong relationship between the phase of the EOF2 and precipitation anomalies in Eastern Australia, but Real and Imaginary phases we chose are not exactly aligned with the direction that maximises this relationship.
 If one were to perform a more in-depth analysis of the relationship between this pattern and precipitation in this region, one would conceivably align one of the axis (either the Real or the Imaginary) to the one that maximises it.

### 2021-04-06 R/compute-ceof.R

@@ -27,7 +27,7 @@ compute_ceof <- function(hgt, lon, lat, lev, temporal = FALSE, lats.eof = c(-90,
     .[, hgt.cpx := spectral::analyticFunction(hgt),
       by = .(lat, time, lev)] %>%
     .[, hgt := hgt.cpx] %>%
-    EOF(hgt ~ time | lon + lat + lev, n = n, suffix = "PC", data = .)
+    EOF(hgt ~ time | lon + lat + lev, n = n, suffix = "cEOF", data = .)
 }
 
 # compute_ceof <- function(dt, temporal = FALSE, lats = c(-80, -20), n = 1:2) {

### 2021-04-06 analysis/paper/shceof.Rmd
Justifica latitudes de ozono

--- 2021-04-06 analysis/paper/paper.Rmd
@@
 
-o3wave_lats <- c(-70, -45)
+o3wave_lats <- c(-75, -45)
 ```
@@
 
-(ref:t-vertical-cap) Regression coefficients between cEOF1 and zonal anomalies of mean air temperature (shaded) and ozone mixing ratio (contours, negative contours with dashed lines, labels in parts per billion by mass) averaged between 80ºS and 35ºS for the period 1979 -- 2019.
+(ref:t-vertical-cap) Regression coefficients between cEOF1 and zonal anomalies of mean air temperature (shaded) and ozone mixing ratio (contours, negative contours with dashed lines, labels in parts per billion by mass) averaged between `r knitr::combine_words(LatLabel(o3wave_lats))` for the period 1979 -- 2019.
 
@@
 t %>% 
-  .[lat %between% c(-80, -35)] %>% 
+  .[lat %between% o3wave_lats] %>% 
   .[, .(hgt = weighted.mean(hgt, cos(lat*pi/180)),
@@
 
-The vertical distribution of these patterns is shown in Figure\ \@ref(fig:t-vertical)) that shows the regression coefficientes between cEOF1 and zonal anomalies of air temperature and zonal anomalies of ozone mixing ratio averaged between 80ºS and 35ºS.
+The vertical distribution of these patterns is shown in Figure\ \@ref(fig:t-vertical)) that shows the regression coefficientes between cEOF1 and zonal anomalies of air temperature and zonal anomalies of ozone mixing ratio averaged between `r knitr::combine_words(LatLabel(o3wave_lats))`.
 Temperature zonal anomalies associated with cEOF1 show a clear wave 1 pattern for both real and imaginary components throughout the atmosphere above 250 hPa  with a change in sign above 10 hPa.
@@
                     breaks = AnchorBreaks(0, exclude = 0)) +
+  geom_contour_tanaka(aes(z = estimate), breaks = AnchorBreaks(0, exclude = 0)) +
   geom_contour2(data = periodic(o3_mean, lon = c(0, 360)),
@@
                 bins = 5) +
-  geom_contour_tanaka(aes(z = estimate), breaks = AnchorBreaks(0, exclude = 0)) +
+
   geom_contour_pval(aes(z = p.value)) +
-
+  geom_hline(yintercept = o3wave_lats, size = 0.3, alpha = 0.5) +
   geom_qmap(~.x[lat <= -20]) +
@@
   coord_polar() +
+
   facet_grid(cEOF~term) +
@@
 
-Indeed, Figure\ \@ref(fig:wave1-o3) shows the relationship between amplitude and phase of the planetary wave 1 in Total Ozone Column between 70ºS and 45ºS \textcolor{red}{(Habría que justifica mejor estas elecciones de bandas latitudinales a paritr del gráfico anterior)} and amplitude and phase of the EOF1.
+Indeed, Figure\ \@ref(fig:wave1-o3) shows the relationship between amplitude and phase of the cEOF1 and the amplitude and phase of the planetary wave 1 in Total Ozone Column between `r knitr::combine_words(LatLabel(o3wave_lats))`; as shown in Figure\ \@ref(fig:o3-regr), this is a wide latitude band that captures most of the area with significant regression coefficients.
 The correlation between the zonal wave 1 of ozone and EOF1 (computed as the mean cosine of the difference in phase) is `r wave_correlation$text`.

### 2021-04-06 analysis/paper/shceof.Rmd
Explica más el histograma.
Closes #8

@@
 
-This particular rotation of EOF2 is the one which maximises the relationship between ENSO and Imaginary EOF2.
-It is also the same that maximises this clean association between EFO2 parts and PSA modes.
+This particular rotation of cEOF2 is the one which maximises the relationship between ENSO and Imaginary cEOF2.
+It is also the same that maximises this clean association between cEOF2 parts and PSA modes.
 
@@
 
-The reason that the EOF analysis arrives to this optimal separation is that the imaginary cEOF is the most common phase (Figure\ \@ref(fig:phase-histogram)), so it naturally appears first when performing a Principal Component Analysis.
+The reason that the EOF analysis arrives to this optimal separation is that the Imaginary cEOF is the most common phase (either negative or positive)
+
+The reason the conventional EOF analysis arrives at the same separation than our particular cEOF rotation is probably the fact that not all phases are equal.
+Figure\ \@ref(fig:phase-histogram) is an histogram that counts the number of SON years in which the cEOF2 was near a particular phase with the observations marked as rugs on the horizontal axis.
+For instance, years with cEOF2 phase within 45º of 0º are nearest the "positive Real" phase.
+About two thirds of time cEOF2 has a phase similar to either the negative or positive Imaginary phase.
 
@@
 ```{r phase-histogram, fig.cap = "(ref:phase-histogram-cap)", fig.width=width_column*.8, fig.height=2.5}
+N <- nrow(cut(ceof$eof[[1]], 2)$left)
 cut(ceof$eof[[1]], 2)$left %>% 
   .[, arg := Arg(hgt)] %>% 
-  .[, w := Mod(hgt)/sum(Mod(hgt))] %>% 
+  .[, w := Mod(hgt)/mean(Mod(hgt))] %>% 
   qwrap(arg = c(-pi, pi) ~  c(-pi - 1/4*pi, pi - 1/4*pi)) %>% 
@@
   geom_rug() +
-  scale_y_continuous("Number of years") +
+  scale_y_continuous("Number of years", 
+                     sec.axis = sec_axis(~.x/N, 
+                                         name = "Proportion of years", 
+                                         labels = scales::percent)) +
   scale_x_continuous("Phase",
@@
 
-By aligning he imaginary phase with the direction with maximum relationship with ENSO, we also aligned it with the phase of maximum occurrence of XXX.
-This is analogous to the method of @irving2016, who used reprojection and Fourier filtering to detect "PSA-like" variability and defined as the proper PSA using the peaks of the phase distribution (Compare our Figure\ \@ref:(phase-histogram) with their Figure 6).
-\textcolor{red}{(Amplia un poco la oración anterior que no queda del todo clara)} The advantage of our method is that it is much simpler to implement, it provides magnitude and phase naturally, and it facilitates the description of this mode as a propagating wave instead of as standing oscillation.
+This phase-locking behaviour indicates that the Imaginary phase both is the direction that has the maximum relationship with ENSO and is also the most common phase.
+Therefore, the Imaginary cEOF2 explains more variance than the Real cEOF2 and conventional EOF analysis will tend to separate the two.
+
+This is analogous to the method of @irving2016, who used reprojection and Fourier filtering to detect "PSA-like" variability and defined as the proper PSA using the peaks of the phase distribution (Compare our Figure\ \@ref(fig:phase-histogram) with their Figure 6).
+
+\textcolor{red}{(Amplia un poco la oración anterior que no queda del todo clara)}
+
+The advantage of our method is that it is much simpler to implement, it provides magnitude and phase naturally, and it facilitates the description of this mode as a propagating wave instead of as standing oscillation.
 As a consequence, the cEOF2 offers an alternative way of representing the PSA which has several advantages over using the second and third EOFs.

### 2021-04-06 analysis/paper/shceof.Rmd
Amplía el epígrafe del histograma.
#8
Más histograma
(#8)

@@
 
-\textcolor{red}{(Lo del histograma habría que explicarlo con un pocoo más detalle porque no es tan trivial.
-Primero decir para qué se hace, qué se hace exactamente y después recien decir cuál es el resultado y qué información aporta)}
-
-The reason that the EOF analysis arrives to this optimal separation is that the Imaginary cEOF is the most common phase (either negative or positive)
-
 The reason the conventional EOF analysis arrives at the same separation than our particular cEOF rotation is probably the fact that not all phases are equal.
@@
 
-(ref:phase-histogram-cap) Histogram showing phase distribution of EOF2 for the period 1979 -- 2019. \textcolor{red}{(También ampliar)}
+(ref:phase-histogram-cap) Histogram of phase distribution of cEOF2 for the period 1979 -- 2019. Bins are centred at 90º, 0º, -90º, -180º with a binwidth of 90º.
 
@@
   ggplot(aes(arg)) +
-  geom_histogram(binwidth = 90*pi/180, center = 0, color = "black", fill = "gray70") +
+  geom_histogram(binwidth = 90*pi/180, center = 0, size = 0.2,
+                 color = "black", fill = "gray70") +
   annotate("text",
@@
                                          labels = scales::percent)) +
-  scale_x_continuous("Phase",
-                     breaks = seq(-pi - 1/4*pi, pi - 1/4*pi, by = 90*pi/180),
-                     labels = function(x) paste0(round(x*180/pi, 5), "º")) 
+  scale_x_continuous("Phase", trans = scales::reverse_trans(),
+                     breaks = seq(-pi - 1/4*pi, pi - 1/4*pi, by = 45*pi/180),
+                     labels = function(x) paste0(round(x*180/pi, 5), "º")) +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
 ```

### 2021-04-06 analysis/paper/shceof.Rmd
Cambia todo a cEOF
Closes #1 y 
Menciona #22

@@
 This motivates the decision of performing complex EOF jointly between levels.
-The computation of the EOFs was carried out using data from both levels at the same time, therefore, each complex EOF has a spatial component that depends on longitude, latitude and level, and a temporal component that depends only on time.
+The computation of the cEOFs was carried out using data from both levels at the same time, therefore, each cEOF has a spatial component that depends on longitude, latitude and level, and a temporal component that depends only on time.
 
@@
 For complex principal components, it can be any number between 0 and $2\pi$ [@horel1984].
-Since any choice is arbitrary and equally valid, we chose the phase of each EOF so that the real and imaginary parts are aligned with meaningful phases in our analysis.
+Since any choice is arbitrary and equally valid, we chose the phase of each cEOF so that the real and imaginary parts are aligned with meaningful phases in our analysis.
 This procedure does not create a spurious correlation, it only takes whatever relationship that already exist and aligns it with a specific phase.
@@
 
-(ref:rotations-cap) a) Correlation of the Real and Imaginary EOF1 with wave 1 amplitude of  Total Ozone Column averaged between `r knitr::combine_words(LatLabel(o3wave_lats))` and b) $R^2$ between the Real and Imaginary EOF2 and the ONI index for different values of rotation.
+(ref:rotations-cap) a) Correlation of the Real and Imaginary cEOF1 with wave 1 amplitude of  Total Ozone Column averaged between `r knitr::combine_words(LatLabel(o3wave_lats))` and b) $R^2$ between the Real and Imaginary cEOF2 and the ONI index for different values of rotation.
 The vertical line marks the selected rotation, which maximises de correlation.
@@
 
-```{r ceofs-1, fig.cap = "Spatial patterns of the leading 2 complex EOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2019.
+```{r ceofs-1, fig.cap = "Spatial patterns of the leading 2 cEOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2019.
 Real part in shading, imaginary part in contours.", fig.width=width_column, fig.height=width_column}
@@
 
-```{r ceofs-2, fig.cap = "Time series of the leading 2 complex EOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2019.", fig.height=width_column, fig.width = width_column}
+```{r ceofs-2, fig.cap = "Time series of the leading 2 cEOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2019.", fig.height=width_column, fig.width = width_column}
 ceof %>%
@@
 In the spatial patterns (Fig.\ \@ref(fig:ceofs-1)), the real (in shading) and the imaginary components (in contour) are in quadrature by construction, so that each cEOF describe a single wave-like pattern whose amplitude and position (i.e.
-phase) is controlled by the magnitude and phase of the complex temporal EOF.
+phase) is controlled by the magnitude and phase of the temporal cEOF.
 The wave patterns described by these cEOFs match the patterns seen in the traditional EOFs of Figure\ \@ref(fig:eof-naive).
@@
 
-(ref:eof1-regr-gh-cap) Regression coefficients of the real and imaginary part of the first complex EOF on SON geopotential height for the 1979 -- 2019 period.
-These coefficients come from multiple linear regression involving the real and imaginary part of both EFO2.
+(ref:eof1-regr-gh-cap) Regression coefficients of the real and imaginary part of the first cEOF on SON geopotential height for the 1979 -- 2019 period.
+These coefficients come from multiple linear regression involving the real and imaginary part of both cEFO2.
 
@@
 Figure\ \@ref(fig:eof1-regr-gh) shows regression patterns between geopotential height anomalies and cEOF1.
-At 50 hPa (Figure\ \@ref(fig:eof1-regr-gh) row a), both the Real and Imaginary EOF1 are associated with planetary wave 1 patterns, that are 90º off phase.
+At 50 hPa (Figure\ \@ref(fig:eof1-regr-gh) row a), both the Real and Imaginary cEOF1 are associated with planetary wave 1 patterns, that are 90º off phase.
 Their phases coincide with the ones shown in Figure\ \@ref(fig:ceofs-1)a.1, with the positive centre of the real cEOF1 located towards the dateline, and the one of the Imaginary cEOF1 located over Eastern Antarctica.
-However, the cReal EOF1 pattern is substantially altered by the zonally symmetric circulation.
+However, the Real cEOF1 pattern is substantially altered by the zonally symmetric circulation.
 Instead of a clear wave 1 pattern, the regression pattern can be better describes as a monopole with its centre displaced from the South Pole.
@@
 This should not be utterly surprising, since the spatial patterns of the *naive* EOFs only show this feature at 50 hPa (Fig.\ \@ref(fig:eof-naive)), suggesting that using the cEOF method is artificially generating a wave structure at 200hPa.
-Therefore, the magnitude and phase of the EOF1 are associated with the magnitude and phase of a zonal wave only in the stratosphere.
+Therefore, the magnitude and phase of the cEOF1 are associated with the magnitude and phase of a zonal wave only in the stratosphere.
 While in the troposphere, they are associated with slightly off-centre monopoles.
 
-(ref:eof2-regr-gh-cap) Same as Figure\ \@ref(fig:eof1-regr-gh) but for the second EOF.
+(ref:eof2-regr-gh-cap) Same as Figure\ \@ref(fig:eof1-regr-gh) but for the second cEOF.
 
@@
 
-Figure\ \@ref(fig:o3-regr) shows regression maps of EOF1 with fields of Total Ozone Column (TOC).
-It shows zonal wave 1 pattens in TOC associated with both phases of EOF1.
+Figure\ \@ref(fig:o3-regr) shows regression maps of cEOF1 with fields of Total Ozone Column (TOC).
+It shows zonal wave 1 pattens in TOC associated with both phases of cEOF1.
 Climatologically, the springtime Ozone minimum is located off the South Pole and towards the Weddell Sea [@wirth1993].
-Thus, the Real EOF1 regression pattern (Figure\ \@ref(fig:o3-regr)a) coincides with the climatological position of the ozone hole while the one for the Imaginary EOF1 is shifted by 90º.
-
-(ref:wave1-o3-cap) Relationship between EOF1 amplitude and phase with amplitude and phase of the zonal wave 1 in Total Ozone Column averaged between `r knitr::combine_words(LatLabel(o3wave_lats))` for the period 1979 -- 2019.
+Thus, the Real cEOF1 regression pattern (Figure\ \@ref(fig:o3-regr)a) coincides with the climatological position of the ozone hole while the one for the Imaginary cEOF1 is shifted by 90º.
+
+(ref:wave1-o3-cap) Relationship between cEOF1 amplitude and phase with amplitude and phase of the zonal wave 1 in Total Ozone Column averaged between `r knitr::combine_words(LatLabel(o3wave_lats))` for the period 1979 -- 2019.
 
@@
 Indeed, Figure\ \@ref(fig:wave1-o3) shows the relationship between amplitude and phase of the cEOF1 and the amplitude and phase of the planetary wave 1 in Total Ozone Column between `r knitr::combine_words(LatLabel(o3wave_lats))`; as shown in Figure\ \@ref(fig:o3-regr), this is a wide latitude band that captures most of the area with significant regression coefficients.
-The correlation between the zonal wave 1 of ozone and EOF1 (computed as the mean cosine of the difference in phase) is `r wave_correlation$text`.
-As expected from the location of anomalies in Figure\ \@ref(fig:wave1-o3), the Real EOF1 drives the relationship with amplitude and the Imaginary EOF1 drives the relationship with phase (not shown).
+The correlation between the zonal wave 1 of ozone and cEOF1 (computed as the mean cosine of the difference in phase) is `r wave_correlation$text`.
+As expected from the location of anomalies in Figure\ \@ref(fig:wave1-o3), the Real cEOF1 drives the relationship with amplitude and the Imaginary cEOF1 drives the relationship with phase (not shown).
 
@@
 
-(ref:psa-eof2-cap) Relationship between the Real and Imaginary parts of EOF2 and the PSA1 and PSA2 modes computed as the second and third EOFs of seasonal geopotential height anomalies [@mo2001].
+(ref:psa-eof2-cap) Relationship between the Real and Imaginary parts of cEOF2 and the PSA1 and PSA2 modes computed as the second and third EOFs of seasonal geopotential height anomalies [@mo2001].
 For the period 1979 -- 2019.
@@
 The relationship between PSAs and the modes described from the cEOFs is also studied.
-Figure\ \@ref(fig:psa-eof2) shows the relationship between the two PSA indices and the Real and Imaginary phase of EOF2.
-As anticipated by Figure\ \@ref(fig:eof2-regr-gh), there is a strong correlation between PSA1 and Imaginary EOF2 (Figure\ \@ref(fig:psa-eof2)b.1), and between PSA2 and Real EOF2 (Figure\ \@ref(fig:psa-eof2)a.2).
-Conversely, there is no relationship between PSA1 and Real EOF2, and between PSA2 and Imaginary EOF2 (Figure\ \@ref(fig:psa-eof2) panels a.1 and b.2).
-So not only this EOF2 represents well both the spatial structure and temporal evolution of the PSA modes, but it's also possible to make a rather clean association between its two phases and the two PSA modes.
+Figure\ \@ref(fig:psa-eof2) shows the relationship between the two PSA indices and the Real and Imaginary phase of cEOF2.
+As anticipated by Figure\ \@ref(fig:eof2-regr-gh), there is a strong correlation between PSA1 and Imaginary cEOF2 (Figure\ \@ref(fig:psa-eof2)b.1), and between PSA2 and Real cEOF2 (Figure\ \@ref(fig:psa-eof2)a.2).
+Conversely, there is no relationship between PSA1 and Real cEOF2, and between PSA2 and Imaginary cEOF2 (Figure\ \@ref(fig:psa-eof2) panels a.1 and b.2).
+So not only this cEOF2 represents well both the spatial structure and temporal evolution of the PSA modes, but it's also possible to make a rather clean association between its two phases and the two PSA modes.
 
@@
 This phase-locking behaviour indicates that the Imaginary phase both is the direction that has the maximum relationship with ENSO and is also the most common phase.
-Therefore, the Imaginary cEOF2 explains more variance than the Real cEOF2 and conventional EOF analysis will tend to separate the two.
+Therefore, the Imaginary cEOF2 explains more variance than the Real cEOF2 and conventional cEOF analysis will tend to separate the two.
 
@@
 
-(ref:sam-eof-vertical-cap) Coefficient of determination between the real and imaginary part of each EOF and the SAM, Asymmetric SAM (A-SAM) and Symmetric SAM (S-SAM) indices computed at each level according to @campitelli2021.
+(ref:sam-eof-vertical-cap) Coefficient of determination between the real and imaginary part of each cEOF and the SAM, Asymmetric SAM (A-SAM) and Symmetric SAM (S-SAM) indices computed at each level according to @campitelli2021.
 Points mark estimates with p-value < 0.01 corrected for False Detection Rate [@benjamini1995].
@@
 The split between A-SAM and S-SAM gives more insight into the nature of the relationship.
-The relationship between the SAM and ¿the imaginary EOF1?
+The relationship between the SAM and ¿the imaginary cEOF1?
 (Figure\ \@ref(fig:sam-eof-vertical).b1) is mediated by S-SAM in the troposphere, but by the A-SAM in the stratosphere.
 
-The Imaginary EOF2 is related with the SAM through the A-SAM in the troposphere, with up to \textcolor{red}{(Acá aparece un -)} `r scales::percent(max_r2[cEOF == "cEOF2"]$estimate, 2)` of shared variance, reached at `r max_r2[cEOF == "cEOF2"]$lev` hPa (Figure\ \@ref(fig:sam-eof-vertical).b2).
-
-```{r eof-filtered, fig.cap = "Spatial pattern of the leading EOF of 200 hPa geopotential height with the variability of EOF2 filtered out South of 20º.
+The Imaginary cEOF2 is related with the SAM through the A-SAM in the troposphere, with up to \textcolor{red}{(Acá aparece un -)} `r scales::percent(max_r2[cEOF == "cEOF2"]$estimate, 2)` of shared variance, reached at `r max_r2[cEOF == "cEOF2"]$lev` hPa (Figure\ \@ref(fig:sam-eof-vertical).b2).
+
+```{r eof-filtered, fig.cap = "Spatial pattern of the leading EOF of 200 hPa geopotential height with the variability of cEOF2 filtered out South of 20º.
 Arbitrary units.", fig.width=2.5, fig.height=2.5}
@@
 
-As a further illustration \textcolor{red}{(¿De qué?)}, Figure\ \@ref(fig:eof-filtered) shows the spatial pattern of the leading EOF of 200 hPa geopotential height when the variability associated with EOF2 is removed.
+As a further illustration \textcolor{red}{(¿De qué?)}, Figure\ \@ref(fig:eof-filtered) shows the spatial pattern of the leading EOF of 200 hPa geopotential height when the variability associated with cEOF2 is removed.
 The resulting pattern is an nearly zonally symmetrical annular mode.
@@
 
-(ref:fake-eof-cap) EOFs of a synthetic data set of 41 times steps of geopotential height fields as the sum of the fields predicted by the EOF2 plus perfectly zonally symmetric SAM-like anomalies with random amplitude.
+(ref:fake-eof-cap) EOFs of a synthetic data set of 41 times steps of geopotential height fields as the sum of the fields predicted by the cEOF2 plus perfectly zonally symmetric SAM-like anomalies with random amplitude.
 
@@
 If shows the leading (and only) EOFs of a synthetic dataset if `r uniqueN(fake_sam$time)` years created by adding a perfectly zonally symmetrical SAM-like pattern with random amplitudes taken from the real SAM, and geopotential height anomalies predicted by EFO2.
-Even though by construction this dataset has a perfectly symmetric SAM, the EOF decomposition mixes this mode with parts of the the zonally asymmetric EOF2 pattern, resulting in a pattern eerily similar to the real SAM pattern.
+Even though by construction this dataset has a perfectly symmetric SAM, the EOF decomposition mixes this mode with parts of the the zonally asymmetric cEOF2 pattern, resulting in a pattern eerily similar to the real SAM pattern.
 
@@
 
-(ref:sst-psi-2-cap) Regression maps of EOF2 with SST (column a) and streamfunction zonal anomalies with their corresponding activity wave flux (column b).
+(ref:sst-psi-2-cap) Regression maps of cEOF2 with SST (column a) and streamfunction zonal anomalies with their corresponding activity wave flux (column b).
 Areas marked with dots have p-values smaller than 0.05 adjusted for FDR.
@@
 
-Figure\ \@ref(fig:sst-psi-2) shows regression maps between EOF2 and Sea Surface Temperatures (SST) Streamfunction at 200 hPa.
-The Imaginary EOF2 is associated with strong positive SST anomalies on the Central Pacific and negative anomalies on an area across the North of Australia and New Zealand, the South Pacific Convergence Zone (SPCZ) (Figure\ \@ref(fig:sst-psi-2).a2).
-This pattern is almost canonically positive ENSO and indeed, the correlation between the Imaginary EOF2 and the Oceanic Niño Index [@bamston1997] is very high `r enso_cor[cEOF == "cEOF2" & part == "Imaginary"]$text`.
+Figure\ \@ref(fig:sst-psi-2) shows regression maps between cEOF2 and Sea Surface Temperatures (SST) Streamfunction at 200 hPa.
+The Imaginary cEOF2 is associated with strong positive SST anomalies on the Central Pacific and negative anomalies on an area across the North of Australia and New Zealand, the South Pacific Convergence Zone (SPCZ) (Figure\ \@ref(fig:sst-psi-2).a2).
+This pattern is almost canonically positive ENSO and indeed, the correlation between the Imaginary cEOF2 and the Oceanic Niño Index [@bamston1997] is very high `r enso_cor[cEOF == "cEOF2" & part == "Imaginary"]$text`.
 Streamfunction anomalies show a coherent picture.
-The Imaginary EOF2 is associated with strong wave-like streamfunction anomalies emanating from the tropics (Figure\ \@ref(fig:sst-psi-2).b2).
+The Imaginary cEOF2 is associated with strong wave-like streamfunction anomalies emanating from the tropics (Figure\ \@ref(fig:sst-psi-2).b2).
 This is consistent with what we know of the effect of ENSO on the extratropics: SST anomalies initiate anomalous convection that excites Rossby waves that propagate meridionally towards higher latitudes \textcolor{red}{(REFS??)}.
 
-Since the Real EOF2 represents just a different phase of the same wave train, one would expect that it would show a similar forcing pattern to the Imaginary EOF with a slight translation of its location.
-However, Figure\ \@ref(fig:sst-psi-2).a1 and b1 show that the Real EOF2 is not associated either with any significant SST nor streamfunction anomalies in the tropics.
-The correlation between the Real EOF2 and ENSO is also not significant (`r enso_cor[cEOF == "cEOF2" & part == "Real"]$cor_text`) \textcolor{red}{(No aparece el valor en el PDF)}.
-This lack of tropical signal suggests different natures of the different phases of the EOF2 wave train.
-
-```{r enso-phase, fig.cap = "Relationship between ENSO and phase of EOF2 for the period 1979 -- 2019.
-Colours denote years with magnitude of EOF2 greater or smaller than the 50th percentile.
-Black line is the fit ONI ~ cos(phase) + sin(phase) computed by OLS weighted by the magnitude of EOF2.", fig.width = width_column, fig.height=3}
+Since the Real cEOF2 represents just a different phase of the same wave train, one would expect that it would show a similar forcing pattern to the Imaginary cEOF with a slight translation of its location.
+However, Figure\ \@ref(fig:sst-psi-2).a1 and b1 show that the Real cEOF2 is not associated either with any significant SST nor streamfunction anomalies in the tropics.
+The correlation between the Real cEOF2 and ENSO is also not significant (`r enso_cor[cEOF == "cEOF2" & part == "Real"]$cor_text`) \textcolor{red}{(No aparece el valor en el PDF)}.
+This lack of tropical signal suggests different natures of the different phases of the cEOF2 wave train.
+
+```{r enso-phase, fig.cap = "Relationship between ENSO and phase of cEOF2 for the period 1979 -- 2019.
+Colours denote years with magnitude of cEOF2 greater or smaller than the 50th percentile.
+Black line is the fit ONI ~ cos(phase) + sin(phase) computed by OLS weighted by the magnitude of cEOF2.", fig.width = width_column, fig.height=3}
 gdata <- ceof[, eof[[1]]$left, by = .(season)] %>%
@@
 
-To better explore the relationship between tropical forcing and phase of the cEOF2, Figure\ \@ref(fig:enso-phase) plots the ONI index and the phase of the EOF2 for each SON trimester between 1979 and 2019, highlighting years in which the magnitude of EOF2 is above the median.
-In years with positive ONI, the phase of the EOF2 is mostly around +90º (corresponding with positive imaginary part) and vice versa.
-In years with near neutral ENSO, the phase of the EOF2 is much more variable.
+To better explore the relationship between tropical forcing and phase of the cEOF2, Figure\ \@ref(fig:enso-phase) plots the ONI index and the phase of the cEOF2 for each SON trimester between 1979 and 2019, highlighting years in which the magnitude of cEOF2 is above the median.
+In years with positive ONI, the phase of the cEOF2 is mostly around +90º (corresponding with positive imaginary part) and vice versa.
+In years with near neutral ENSO, the phase of the cEOF2 is much more variable.
 The black line in Figure\ \@ref(fig:enso-phase) is an harmonical fit of the relationship between ONI and cEOF2 phase.
@@
 
-Furthemore, Figure\ \@ref(fig:enso-phase) suggest that strong EOF2 years tend to coincide with strong ENSO years.
-The correlation between the absolute magnitude of the ONI and the magnitude of the EOF2 is `r cor_enso_mag$text`.
-This relationship, however, appears to be driven only by the three years with strongest ENSO events in the period (`r knitr::combine_words(year(gdata[order(-oni)][1:3]$time))`) which also coincide with the three years with strongest EOF2 magnitude.
+Furthemore, Figure\ \@ref(fig:enso-phase) suggest that strong cEOF2 years tend to coincide with strong ENSO years.
+The correlation between the absolute magnitude of the ONI and the magnitude of the cEOF2 is `r cor_enso_mag$text`.
+This relationship, however, appears to be driven only by the three years with strongest ENSO events in the period (`r knitr::combine_words(year(gdata[order(-oni)][1:3]$time))`) which also coincide with the three years with strongest cEOF2 magnitude.
 If those years are removed, the correlation becomes non-significant (`r cor_enso_mag_outlier$text`).
@@
 However, when cEOF2 is excited by tropical SST variability, it tends to remain locked to the imaginary phase.
-This explains the relative overabundance of years with EOF2 near positive and negative imaginary phase in Figure\ \@ref(fig:phase-histogram).
-
-(ref:sst-psi-1-cap) Same as Figure\ \@ref(fig:sst-psi-2) but for EOF1.
+This explains the relative overabundance of years with cEOF2 near positive and negative imaginary phase in Figure\ \@ref(fig:phase-histogram).
+
+(ref:sst-psi-1-cap) Same as Figure\ \@ref(fig:sst-psi-2) but for cEOF1.
 
@@
 
-Figure\ \@ref(fig:sst-psi-1) shows SST and streamfunction regression maps for EOF1.
-There is no significant pattern of SST anomalies associated with either the Real or Imaginary EOF1.
+Figure\ \@ref(fig:sst-psi-1) shows SST and streamfunction regression maps for cEOF1.
+There is no significant pattern of SST anomalies associated with either the Real or Imaginary cEOF1.
 Consistently, streamfunction anomalies do not show any tropical influence.
@@
 
-(ref:pp-america-cap) Regression of SON mean precipitation anomalies in South America (mm per day, shaded) and (column a) EOF1 the (row 1) Imaginary and (column 1) Real phase.
+(ref:pp-america-cap) Regression of SON mean precipitation anomalies in South America (mm per day, shaded) and (column a) cEOF1 the (row 1) Imaginary and (column 1) Real phase.
 For the 1979 -- 2018 period.
@@
 cEOF1 is associated with slight decreases in precipitation in Southern Brazil and Paraguay, although these are not statistically significant \@ref(fig:pp-america) column a.
-The strongest precipitation anomalies are the ones associated with the Imaginary EOF2.
+The strongest precipitation anomalies are the ones associated with the Imaginary cEOF2.
 The positive anomalies on Southeastern South America (SESA) and Central Chile, and negative anomalies over South Atlantic Convergence Zone is a well known springtime precipitation signature of ENSO [@cai2020a] and it is also similar to the precipitation anomalies associated with the A-SAM [@campitelli2021].
@@
 The Real cEOF2, on the other hand, is associated with negative precipitation anomalies in a smaller area of SESA.
-There is a phase dependence of the precipitation anomalies in SESA (not shown), which resembles the relationship between ONI and the phase of EOF2.
+There is a phase dependence of the precipitation anomalies in SESA (not shown), which resembles the relationship between ONI and the phase of cEOF2.
 \textcolor{red}{(Podrías agregar acá el coeficiente de determinación)}
@@
 
-The precipitation anomalies in Oceania associated with EOF1 are also weak and not statistically significant (Figure\ \@ref(fig:pp-oceania) column a) whereas EFO2 is associated with large and expansive anomalies.
-The Real cEOF2 is associated with positive seasonal anomalies in Northern and Eastern Australia, and negative ones over New Zealand, while precipitation anomalies associated with Imaginary EOF2 are negative over all Eastern Australia.
+The precipitation anomalies in Oceania associated with cEOF1 are also weak and not statistically significant (Figure\ \@ref(fig:pp-oceania) column a) whereas EFO2 is associated with large and expansive anomalies.
+The Real cEOF2 is associated with positive seasonal anomalies in Northern and Eastern Australia, and negative ones over New Zealand, while precipitation anomalies associated with Imaginary cEOF2 are negative over all Eastern Australia.
 Again, these negative anomalies are similar to the springtime precipitation anomalies observed in relation with the Asymmetric SAM [@campitelli2021].
@@
 
-To better understand the relationship between precipitation anomalies over Western Australia and the EOF2, Figure\ \@ref(fig:australia-pp-phase) plots precipitation anomalies averaged over Eastern Australia (box in Figure\ \@ref(fig:pp-oceania)) as a function of EOF2 phase.
+To better understand the relationship between precipitation anomalies over Western Australia and the cEOF2, Figure\ \@ref(fig:australia-pp-phase) plots precipitation anomalies averaged over Eastern Australia (box in Figure\ \@ref(fig:pp-oceania)) as a function of cEOF2 phase.
 The black line is the equation $\mathrm{Precipitation} = `r signif(coef(model)[1], 2)`\sin{(\mathrm{phase})} + `r signif(coef(model)[2], 2)`\cos{(\mathrm{phase})}$, whose coefficient we fitted by weighted least squares using the magnitude of the EFO2 as weight.
@@
 The $r^2$ corresponding to the fit is `r signif(summary(model)$r.squared, 2)`, with p-value `r report_p(p_value)`.
-There is a strong relationship between the phase of the EOF2 and precipitation anomalies in Eastern Australia, but Real and Imaginary phases we chose are not exactly aligned with the direction that maximises this relationship.
+There is a strong relationship between the phase of the cEOF2 and precipitation anomalies in Eastern Australia, but Real and Imaginary phases we chose are not exactly aligned with the direction that maximises this relationship.
 If one were to perform a more in-depth analysis of the relationship between this pattern and precipitation in this region, one would conceivably align one of the axis (either the Real or the Imaginary) to the one that maximises it.
@@
 The first complex EOF represents the variability of the zonal wave 1 in the stratosphere, but represents a more zonally symmetric monopole in the troposphere.
-There is a statistically positive trend in the magnitude of this EOF, which is consistent with previous studies which showed secular changes in springtime wave-1-like patterns [e.g.
+There is a statistically positive trend in the magnitude of this cEOF, which is consistent with previous studies which showed secular changes in springtime wave-1-like patterns [e.g.
 @raphael2003].
@@
 Essentially, it is an alternate representation of the PSA1 and PSA2 patterns [@mo2001].
-We show that the Imaginary EOF2 can be identified with the PSA1 and the Real EOF2 with the PSA2.
-
-We show that this EOF prefers a phase aligned with the Imaginary EOF2.
-This is because the when ENSO is active, it forces the EOF2 to be in the Imaginary phase; positive for El Niño and negative for La Niña.
-When ENSO is neutral, the EOF2 is still active, but with no particular phase.
+We show that the Imaginary cEOF2 can be identified with the PSA1 and the Real cEOF2 with the PSA2.
+
+We show that this cEOF prefers a phase aligned with the Imaginary cEOF2.
+This is because the when ENSO is active, it forces the cEOF2 to be in the Imaginary phase; positive for El Niño and negative for La Niña.
+When ENSO is neutral, the cEOF2 is still active, but with no particular phase.
 This mirrors the results of @cai2002, who showed that the CSIRO Model can create PSA-like variability even in the absence of ENSO forcing (with a climatological run), but the variability of one of the PSA modes was enhanced when adding the ENSO signal.
 
-We further show that the Imaginary EOF2 is closely related to the Southern Annular Mode in the troposphere.
+We further show that the Imaginary cEOF2 is closely related to the Southern Annular Mode in the troposphere.
 In fact, it is a virtually perfect indicator of the zonally asymmetric portion of the SAM.
-To the extreme that the leading EOF of geopotential height with the EOF2 removes results in an annular mode that has none of the typical zonal asymmetries of the observed SAM.
+To the extreme that the leading EOF of geopotential height with the cEOF2 removes results in an annular mode that has none of the typical zonal asymmetries of the observed SAM.
 
@@
 
-Precipitation anomalies in South America associated with the Imaginary EOF2 show a clear ENSO-like impact, with positive anomalies in South-Eastern Southamerica, negative anomalies in Southern Brazil and positive anomalies in central Chile.
-Precipitation anomalies associated with the Real EOF2 are low and not statistically significant, showing that the Imaginary phase is optimally aligned with the direction of maximum precipitation impacts.
-
-On the other hand, over Australia, both the Real and Imaginary phase are associated with significant precipitation anomalies, and we further show that the direction of maximum impact is not aligned with our chosen rotation of EOF2.
+Precipitation anomalies in South America associated with the Imaginary cEOF2 show a clear ENSO-like impact, with positive anomalies in South-Eastern Southamerica, negative anomalies in Southern Brazil and positive anomalies in central Chile.
+Precipitation anomalies associated with the Real cEOF2 are low and not statistically significant, showing that the Imaginary phase is optimally aligned with the direction of maximum precipitation impacts.
+
+On the other hand, over Australia, both the Real and Imaginary phase are associated with significant precipitation anomalies, and we further show that the direction of maximum impact is not aligned with our chosen rotation of cEOF2.
 However, this underscore the benefit of using complex EOF, since it would trivial to rotate it.

### 2021-04-06 analysis/paper/shceof.tex
Renderiza

--- 2021-04-06 analysis/paper/paper.tex
@@
 
-We computed the amplitude and phase of the zonal wave 1 by averaging (area-weighted) variables (temperature, geopotential height, ozone mixing ratio and Total Ozone Column) between 70°S and 45°S for each SON and extracting the wave-1 component of the Fourier spectrum.
+We computed the amplitude and phase of the zonal wave 1 by averaging (area-weighted) variables (temperature, geopotential height, ozone mixing ratio and Total Ozone Column) between 75°S and 45°S for each SON and extracting the wave-1 component of the Fourier spectrum.
 We chose this latitude band because it is wide enough to capture most of the relevant anomalies.
@@
 For complex principal components, it can be any number between 0 and \(2\pi\) \citep{horel1984}.
-Since any choice is arbitrary and equally valid, we chose the phase of each EOF so that the real and imaginary parts are aligned with meaningful phases in our analysis.
+Since any choice is arbitrary and equally valid, we chose the phase of each cEOF so that the real and imaginary parts are aligned with meaningful phases in our analysis.
 This procedure does not create a spurious correlation, it only takes whatever relationship that already exist and aligns it with a specific phase.
 
-For the first complex principal component, the phase was chosen so that the time series corresponding to the real part has the maximum correlation with the zonal wave 1 of Total Ozone Column between 70°S and 45°S.
+For the first complex principal component, the phase was chosen so that the time series corresponding to the real part has the maximum correlation with the zonal wave 1 of Total Ozone Column between 75°S and 45°S.
 This also nearly minimises the correlation with the imaginary part.
@@
 \includegraphics{../figures/ceofs-1-1.pdf}
-\caption{\label{fig:ceofs-1}Spatial patterns of the leading 2 complex EOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2019.
+\caption{\label{fig:ceofs-1}Spatial patterns of the leading 2 cEOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2019.
 Real part in shading, imaginary part in contours.}
@@
 \includegraphics{../figures/ceofs-2-1.pdf}
-\caption{\label{fig:ceofs-2}Time series of the leading 2 complex EOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2019.}
+\caption{\label{fig:ceofs-2}Time series of the leading 2 cEOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2019.}
 \end{figure}
@@
 The first mode (CEOF1) explains 82\% of the variance, while the second mode (CEOF1) explains a smaller fraction (7\%).
-In the spatial patterns (Fig.~\ref{fig:ceofs-1}), the real (in shading) and the imaginary components (in contour) are in quadrature by construction, so that each cEOF describe a single wave-like pattern whose amplitude and position (i.e.~phase) is controlled by the magnitude and phase of the complex temporal EOF.
+In the spatial patterns (Fig.~\ref{fig:ceofs-1}), the real (in shading) and the imaginary components (in contour) are in quadrature by construction, so that each cEOF describe a single wave-like pattern whose amplitude and position (i.e.~phase) is controlled by the magnitude and phase of the temporal cEOF.
 The wave patterns described by these cEOFs match the patterns seen in the traditional EOFs of Figure~\ref{fig:eof-naive}.
@@
 \includegraphics{../figures/eof1-regr-gh-1.pdf}
-\caption{\label{fig:eof1-regr-gh}Regression coefficients of the real and imaginary part of the first complex EOF on SON geopotential height for the 1979 -- 2019 period.
-These coefficients come from multiple linear regression involving the real and imaginary part of both EFO2.}
+\caption{\label{fig:eof1-regr-gh}Regression coefficients of the real and imaginary part of the first cEOF on SON geopotential height for the 1979 -- 2019 period.
+These coefficients come from multiple linear regression involving the real and imaginary part of both cEFO2.}
 \end{figure}
@@
 Figure~\ref{fig:eof1-regr-gh} shows regression patterns between geopotential height anomalies and cEOF1.
-At 50 hPa (Figure~\ref{fig:eof1-regr-gh} row a), both the Real and Imaginary EOF1 are associated with planetary wave 1 patterns, that are 90º off phase.
+At 50 hPa (Figure~\ref{fig:eof1-regr-gh} row a), both the Real and Imaginary cEOF1 are associated with planetary wave 1 patterns, that are 90º off phase.
 Their phases coincide with the ones shown in Figure~\ref{fig:ceofs-1}a.1, with the positive centre of the real cEOF1 located towards the dateline, and the one of the Imaginary cEOF1 located over Eastern Antarctica.
-However, the cReal EOF1 pattern is substantially altered by the zonally symmetric circulation.
+However, the Real cEOF1 pattern is substantially altered by the zonally symmetric circulation.
 Instead of a clear wave 1 pattern, the regression pattern can be better describes as a monopole with its centre displaced from the South Pole.
@@
 This should not be utterly surprising, since the spatial patterns of the \emph{naive} EOFs only show this feature at 50 hPa (Fig.~\ref{fig:eof-naive}), suggesting that using the cEOF method is artificially generating a wave structure at 200hPa.
-Therefore, the magnitude and phase of the EOF1 are associated with the magnitude and phase of a zonal wave only in the stratosphere.
+Therefore, the magnitude and phase of the cEOF1 are associated with the magnitude and phase of a zonal wave only in the stratosphere.
 While in the troposphere, they are associated with slightly off-centre monopoles.
@@
 \includegraphics{../figures/eof2-regr-gh-1.pdf}
-\caption{\label{fig:eof2-regr-gh}Same as Figure~\ref{fig:eof1-regr-gh} but for the second EOF.}
+\caption{\label{fig:eof2-regr-gh}Same as Figure~\ref{fig:eof1-regr-gh} but for the second cEOF.}
 \end{figure}
@@
 \includegraphics{../figures/t-vertical-1.pdf}
-\caption{\label{fig:t-vertical}Regression coefficients between cEOF1 and zonal anomalies of mean air temperature (shaded) and ozone mixing ratio (contours, negative contours with dashed lines, labels in parts per billion by mass) averaged between 80ºS and 35ºS for the period 1979 -- 2019.}
+\caption{\label{fig:t-vertical}Regression coefficients between cEOF1 and zonal anomalies of mean air temperature (shaded) and ozone mixing ratio (contours, negative contours with dashed lines, labels in parts per billion by mass) averaged between 75°S and 45°S for the period 1979 -- 2019.}
 \end{figure}
@@
 
-The vertical distribution of these patterns is shown in Figure~\ref{fig:t-vertical}) that shows the regression coefficientes between cEOF1 and zonal anomalies of air temperature and zonal anomalies of ozone mixing ratio averaged between 80ºS and 35ºS.
+The vertical distribution of these patterns is shown in Figure~\ref{fig:t-vertical}) that shows the regression coefficientes between cEOF1 and zonal anomalies of air temperature and zonal anomalies of ozone mixing ratio averaged between 75°S and 45°S.
 Temperature zonal anomalies associated with cEOF1 show a clear wave 1 pattern for both real and imaginary components throughout the atmosphere above 250 hPa with a change in sign above 10 hPa.
@@
 
-Figure~\ref{fig:o3-regr} shows regression maps of EOF1 with fields of Total Ozone Column (TOC).
-It shows zonal wave 1 pattens in TOC associated with both phases of EOF1.
+Figure~\ref{fig:o3-regr} shows regression maps of cEOF1 with fields of Total Ozone Column (TOC).
+It shows zonal wave 1 pattens in TOC associated with both phases of cEOF1.
 Climatologically, the springtime Ozone minimum is located off the South Pole and towards the Weddell Sea \citep{wirth1993}.
-Thus, the Real EOF1 regression pattern (Figure~\ref{fig:o3-regr}a) coincides with the climatological position of the ozone hole while the one for the Imaginary EOF1 is shifted by 90º.
+Thus, the Real cEOF1 regression pattern (Figure~\ref{fig:o3-regr}a) coincides with the climatological position of the ozone hole while the one for the Imaginary cEOF1 is shifted by 90º.
 
@@
 \includegraphics{../figures/wave1-o3-1.pdf}
-\caption{\label{fig:wave1-o3}Relationship between EOF1 amplitude and phase with amplitude and phase of the zonal wave 1 in Total Ozone Column averaged between 70°S and 45°S for the period 1979 -- 2019.}
+\caption{\label{fig:wave1-o3}Relationship between cEOF1 amplitude and phase with amplitude and phase of the zonal wave 1 in Total Ozone Column averaged between 75°S and 45°S for the period 1979 -- 2019.}
 \end{figure}
 
-Indeed, Figure~\ref{fig:wave1-o3} shows the relationship between amplitude and phase of the planetary wave 1 in Total Ozone Column between 70ºS and 45ºS \textcolor{red}{(Habría que justifica mejor estas elecciones de bandas latitudinales a paritr del gráfico anterior)} and amplitude and phase of the EOF1.
-The correlation between the zonal wave 1 of ozone and EOF1 (computed as the mean cosine of the difference in phase) is 0.86 (CI: 0.76 -- 0.93).
-As expected from the location of anomalies in Figure~\ref{fig:wave1-o3}, the Real EOF1 drives the relationship with amplitude and the Imaginary EOF1 drives the relationship with phase (not shown).
+Indeed, Figure~\ref{fig:wave1-o3} shows the relationship between amplitude and phase of the cEOF1 and the amplitude and phase of the planetary wave 1 in Total Ozone Column between 75°S and 45°S; as shown in Figure~\ref{fig:o3-regr}, this is a wide latitude band that captures most of the area with significant regression coefficients.
+The correlation between the zonal wave 1 of ozone and cEOF1 (computed as the mean cosine of the difference in phase) is 0.86 (CI: 0.76 -- 0.93).
+As expected from the location of anomalies in Figure~\ref{fig:wave1-o3}, the Real cEOF1 drives the relationship with amplitude and the Imaginary cEOF1 drives the relationship with phase (not shown).
 
@@
 \begin{figure*}
-\includegraphics{../figures/psa-eof2-1} \caption{Relationship between the Real and Imaginary parts of EOF2 and the PSA1 and PSA2 modes computed as the second and third EOFs of seasonal geopotential height anomalies \citep{mo2001}.
+\includegraphics{../figures/psa-eof2-1} \caption{Relationship between the Real and Imaginary parts of cEOF2 and the PSA1 and PSA2 modes computed as the second and third EOFs of seasonal geopotential height anomalies \citep{mo2001}.
 For the period 1979 -- 2019.}\label{fig:psa-eof2}
@@
 The relationship between PSAs and the modes described from the cEOFs is also studied.
-Figure~\ref{fig:psa-eof2} shows the relationship between the two PSA indices and the Real and Imaginary phase of EOF2.
-As anticipated by Figure~\ref{fig:eof2-regr-gh}, there is a strong correlation between PSA1 and Imaginary EOF2 (Figure~\ref{fig:psa-eof2}b.1), and between PSA2 and Real EOF2 (Figure~\ref{fig:psa-eof2}a.2).
-Conversely, there is no relationship between PSA1 and Real EOF2, and between PSA2 and Imaginary EOF2 (Figure~\ref{fig:psa-eof2} panels a.1 and b.2).
-So not only this EOF2 represents well both the spatial structure and temporal evolution of the PSA modes, but it's also possible to make a rather clean association between its two phases and the two PSA modes.
+Figure~\ref{fig:psa-eof2} shows the relationship between the two PSA indices and the Real and Imaginary phase of cEOF2.
+As anticipated by Figure~\ref{fig:eof2-regr-gh}, there is a strong correlation between PSA1 and Imaginary cEOF2 (Figure~\ref{fig:psa-eof2}b.1), and between PSA2 and Real cEOF2 (Figure~\ref{fig:psa-eof2}a.2).
+Conversely, there is no relationship between PSA1 and Real cEOF2, and between PSA2 and Imaginary cEOF2 (Figure~\ref{fig:psa-eof2} panels a.1 and b.2).
+So not only this cEOF2 represents well both the spatial structure and temporal evolution of the PSA modes, but it's also possible to make a rather clean association between its two phases and the two PSA modes.
 
-This particular rotation of EOF2 is the one which maximises the relationship between ENSO and Imaginary EOF2.
-It is also the same that maximises this clean association between EFO2 parts and PSA modes.
+This particular rotation of cEOF2 is the one which maximises the relationship between ENSO and Imaginary cEOF2.
+It is also the same that maximises this clean association between cEOF2 parts and PSA modes.
 
-\textcolor{red}{(Lo del histograma habría que explicarlo con un pocoo más detalle porque no es tan trivial.
-Primero decir para qué se hace, qué se hace exactamente y después recien decir cuál es el resultado y qué información aporta)}
-
-The reason that the EOF analysis arrives to this optimal separation is that the imaginary cEOF is the most common phase (Figure~\ref{fig:phase-histogram}), so it naturally appears first when performing a Principal Component Analysis.
+The reason the conventional EOF analysis arrives at the same separation than our particular cEOF rotation is probably the fact that not all phases are equal.
+Figure~\ref{fig:phase-histogram} is an histogram that counts the number of SON years in which the cEOF2 was near a particular phase with the observations marked as rugs on the horizontal axis.
+For instance, years with cEOF2 phase within 45º of 0º are nearest the ``positive Real'' phase.
+About two thirds of time cEOF2 has a phase similar to either the negative or positive Imaginary phase.
 
@@
 \includegraphics{../figures/phase-histogram-1.pdf}
-\caption{\label{fig:phase-histogram}Histogram showing phase distribution of EOF2 for the period 1979 -- 2019. \textcolor{red}{(También ampliar)}}
+\caption{\label{fig:phase-histogram}Histogram of phase distribution of cEOF2 for the period 1979 -- 2019. Bins are centred at 90º, 0º, -90º, -180º with a binwidth of 90º.}
 \end{figure}
 
-By aligning he imaginary phase with the direction with maximum relationship with ENSO, we also aligned it with the phase of maximum occurrence of XXX.
-This is analogous to the method of \citet{irving2016}, who used reprojection and Fourier filtering to detect ``PSA-like'' variability and defined as the proper PSA using the peaks of the phase distribution (Compare our Figure~@ref:(phase-histogram) with their Figure 6).
-\textcolor{red}{(Amplia un poco la oración anterior que no queda del todo clara)} The advantage of our method is that it is much simpler to implement, it provides magnitude and phase naturally, and it facilitates the description of this mode as a propagating wave instead of as standing oscillation.
+This phase-locking behaviour indicates that the Imaginary phase both is the direction that has the maximum relationship with ENSO and is also the most common phase.
+Therefore, the Imaginary cEOF2 explains more variance than the Real cEOF2 and conventional cEOF analysis will tend to separate the two.
+
+This is analogous to the method of \citet{irving2016}, who used reprojection and Fourier filtering to detect ``PSA-like'' variability and defined as the proper PSA using the peaks of the phase distribution (Compare our Figure~\ref{fig:phase-histogram} with their Figure 6).
+
+\textcolor{red}{(Amplia un poco la oración anterior que no queda del todo clara)}
+
+The advantage of our method is that it is much simpler to implement, it provides magnitude and phase naturally, and it facilitates the description of this mode as a propagating wave instead of as standing oscillation.
 As a consequence, the cEOF2 offers an alternative way of representing the PSA which has several advantages over using the second and third EOFs.
@@
 \includegraphics{../figures/sam-eof-vertical-1.pdf}
-\caption{\label{fig:sam-eof-vertical}Coefficient of determination between the real and imaginary part of each EOF and the SAM, Asymmetric SAM (A-SAM) and Symmetric SAM (S-SAM) indices computed at each level according to \citet{campitelli2021}.
+\caption{\label{fig:sam-eof-vertical}Coefficient of determination between the real and imaginary part of each cEOF and the SAM, Asymmetric SAM (A-SAM) and Symmetric SAM (S-SAM) indices computed at each level according to \citet{campitelli2021}.
 Points mark estimates with p-value \textless{} 0.01 corrected for False Detection Rate \citep{benjamini1995}.}
@@
 The split between A-SAM and S-SAM gives more insight into the nature of the relationship.
-The relationship between the SAM and ¿the imaginary EOF1?
+The relationship between the SAM and ¿the imaginary cEOF1?
 (Figure~\ref{fig:sam-eof-vertical}.b1) is mediated by S-SAM in the troposphere, but by the A-SAM in the stratosphere.
 
-The Imaginary EOF2 is related with the SAM through the A-SAM in the troposphere, with up to \textcolor{red}{(Acá aparece un -)} -96\% of shared variance, reached at 225 hPa (Figure~\ref{fig:sam-eof-vertical}.b2).
+The Imaginary cEOF2 is related with the SAM through the A-SAM in the troposphere, with up to \textcolor{red}{(Acá aparece un -)} -96\% of shared variance, reached at 225 hPa (Figure~\ref{fig:sam-eof-vertical}.b2).
 
@@
 \includegraphics{../figures/eof-filtered-1.pdf}
-\caption{\label{fig:eof-filtered}Spatial pattern of the leading EOF of 200 hPa geopotential height with the variability of EOF2 filtered out South of 20º. Arbitrary units.}
+\caption{\label{fig:eof-filtered}Spatial pattern of the leading EOF of 200 hPa geopotential height with the variability of cEOF2 filtered out South of 20º. Arbitrary units.}
 \end{figure}
@@
 
-As a further illustration \textcolor{red}{(¿De qué?)}, Figure~\ref{fig:eof-filtered} shows the spatial pattern of the leading EOF of 200 hPa geopotential height when the variability associated with EOF2 is removed.
+As a further illustration \textcolor{red}{(¿De qué?)}, Figure~\ref{fig:eof-filtered} shows the spatial pattern of the leading EOF of 200 hPa geopotential height when the variability associated with cEOF2 is removed.
 The resulting pattern is an nearly zonally symmetrical annular mode.
@@
 \includegraphics{../figures/fake-eof-1.pdf}
-\caption{\label{fig:fake-eof}EOFs of a synthetic data set of 41 times steps of geopotential height fields as the sum of the fields predicted by the EOF2 plus perfectly zonally symmetric SAM-like anomalies with random amplitude.}
+\caption{\label{fig:fake-eof}EOFs of a synthetic data set of 41 times steps of geopotential height fields as the sum of the fields predicted by the cEOF2 plus perfectly zonally symmetric SAM-like anomalies with random amplitude.}
 \end{figure}
@@
 If shows the leading (and only) EOFs of a synthetic dataset if 41 years created by adding a perfectly zonally symmetrical SAM-like pattern with random amplitudes taken from the real SAM, and geopotential height anomalies predicted by EFO2.
-Even though by construction this dataset has a perfectly symmetric SAM, the EOF decomposition mixes this mode with parts of the the zonally asymmetric EOF2 pattern, resulting in a pattern eerily similar to the real SAM pattern.
+Even though by construction this dataset has a perfectly symmetric SAM, the EOF decomposition mixes this mode with parts of the the zonally asymmetric cEOF2 pattern, resulting in a pattern eerily similar to the real SAM pattern.
 
@@
 \begin{figure*}
-\includegraphics{../figures/sst-psi-2-1} \caption{Regression maps of EOF2 with SST (column a) and streamfunction zonal anomalies with their corresponding activity wave flux (column b).
+\includegraphics{../figures/sst-psi-2-1} \caption{Regression maps of cEOF2 with SST (column a) and streamfunction zonal anomalies with their corresponding activity wave flux (column b).
 Areas marked with dots have p-values smaller than 0.05 adjusted for FDR.}\label{fig:sst-psi-2}
@@
 
-Figure~\ref{fig:sst-psi-2} shows regression maps between EOF2 and Sea Surface Temperatures (SST) Streamfunction at 200 hPa.
-The Imaginary EOF2 is associated with strong positive SST anomalies on the Central Pacific and negative anomalies on an area across the North of Australia and New Zealand, the South Pacific Convergence Zone (SPCZ) (Figure~\ref{fig:sst-psi-2}.a2).
-This pattern is almost canonically positive ENSO and indeed, the correlation between the Imaginary EOF2 and the Oceanic Niño Index \citep{bamston1997} is very high 0.76 (CI: 0.59 -- 0.86).
+Figure~\ref{fig:sst-psi-2} shows regression maps between cEOF2 and Sea Surface Temperatures (SST) Streamfunction at 200 hPa.
+The Imaginary cEOF2 is associated with strong positive SST anomalies on the Central Pacific and negative anomalies on an area across the North of Australia and New Zealand, the South Pacific Convergence Zone (SPCZ) (Figure~\ref{fig:sst-psi-2}.a2).
+This pattern is almost canonically positive ENSO and indeed, the correlation between the Imaginary cEOF2 and the Oceanic Niño Index \citep{bamston1997} is very high 0.76 (CI: 0.59 -- 0.86).
 Streamfunction anomalies show a coherent picture.
-The Imaginary EOF2 is associated with strong wave-like streamfunction anomalies emanating from the tropics (Figure~\ref{fig:sst-psi-2}.b2).
+The Imaginary cEOF2 is associated with strong wave-like streamfunction anomalies emanating from the tropics (Figure~\ref{fig:sst-psi-2}.b2).
 This is consistent with what we know of the effect of ENSO on the extratropics: SST anomalies initiate anomalous convection that excites Rossby waves that propagate meridionally towards higher latitudes \textcolor{red}{(REFS??)}.
 
-Since the Real EOF2 represents just a different phase of the same wave train, one would expect that it would show a similar forcing pattern to the Imaginary EOF with a slight translation of its location.
-However, Figure~\ref{fig:sst-psi-2}.a1 and b1 show that the Real EOF2 is not associated either with any significant SST nor streamfunction anomalies in the tropics.
-The correlation between the Real EOF2 and ENSO is also not significant () \textcolor{red}{(No aparece el valor en el PDF)}.
-This lack of tropical signal suggests different natures of the different phases of the EOF2 wave train.
+Since the Real cEOF2 represents just a different phase of the same wave train, one would expect that it would show a similar forcing pattern to the Imaginary cEOF with a slight translation of its location.
+However, Figure~\ref{fig:sst-psi-2}.a1 and b1 show that the Real cEOF2 is not associated either with any significant SST nor streamfunction anomalies in the tropics.
+The correlation between the Real cEOF2 and ENSO is also not significant () \textcolor{red}{(No aparece el valor en el PDF)}.
+This lack of tropical signal suggests different natures of the different phases of the cEOF2 wave train.
 
@@
 \includegraphics{../figures/enso-phase-1.pdf}
-\caption{\label{fig:enso-phase}Relationship between ENSO and phase of EOF2 for the period 1979 -- 2019.
-Colours denote years with magnitude of EOF2 greater or smaller than the 50th percentile.
-Black line is the fit ONI \textasciitilde{} cos(phase) + sin(phase) computed by OLS weighted by the magnitude of EOF2.}
+\caption{\label{fig:enso-phase}Relationship between ENSO and phase of cEOF2 for the period 1979 -- 2019.
+Colours denote years with magnitude of cEOF2 greater or smaller than the 50th percentile.
+Black line is the fit ONI \textasciitilde{} cos(phase) + sin(phase) computed by OLS weighted by the magnitude of cEOF2.}
 \end{figure}
 
-To better explore the relationship between tropical forcing and phase of the cEOF2, Figure~\ref{fig:enso-phase} plots the ONI index and the phase of the EOF2 for each SON trimester between 1979 and 2019, highlighting years in which the magnitude of EOF2 is above the median.
-In years with positive ONI, the phase of the EOF2 is mostly around +90º (corresponding with positive imaginary part) and vice versa.
-In years with near neutral ENSO, the phase of the EOF2 is much more variable.
+To better explore the relationship between tropical forcing and phase of the cEOF2, Figure~\ref{fig:enso-phase} plots the ONI index and the phase of the cEOF2 for each SON trimester between 1979 and 2019, highlighting years in which the magnitude of cEOF2 is above the median.
+In years with positive ONI, the phase of the cEOF2 is mostly around +90º (corresponding with positive imaginary part) and vice versa.
+In years with near neutral ENSO, the phase of the cEOF2 is much more variable.
 The black line in Figure~\ref{fig:enso-phase} is an harmonical fit of the relationship between ONI and cEOF2 phase.
@@
 
-Furthemore, Figure~\ref{fig:enso-phase} suggest that strong EOF2 years tend to coincide with strong ENSO years.
-The correlation between the absolute magnitude of the ONI and the magnitude of the EOF2 is 0.44 (CI: 0.16 -- 0.66).
-This relationship, however, appears to be driven only by the three years with strongest ENSO events in the period (2015, 1997, and 1982) which also coincide with the three years with strongest EOF2 magnitude.
+Furthemore, Figure~\ref{fig:enso-phase} suggest that strong cEOF2 years tend to coincide with strong ENSO years.
+The correlation between the absolute magnitude of the ONI and the magnitude of the cEOF2 is 0.44 (CI: 0.16 -- 0.66).
+This relationship, however, appears to be driven only by the three years with strongest ENSO events in the period (2015, 1997, and 1982) which also coincide with the three years with strongest cEOF2 magnitude.
 If those years are removed, the correlation becomes non-significant (0.044 (CI: -0.28 -- 0.36)).
@@
 However, when cEOF2 is excited by tropical SST variability, it tends to remain locked to the imaginary phase.
-This explains the relative overabundance of years with EOF2 near positive and negative imaginary phase in Figure~\ref{fig:phase-histogram}.
+This explains the relative overabundance of years with cEOF2 near positive and negative imaginary phase in Figure~\ref{fig:phase-histogram}.
 
 \begin{figure*}
-\includegraphics{../figures/sst-psi-1-1} \caption{Same as Figure~\ref{fig:sst-psi-2} but for EOF1.}\label{fig:sst-psi-1}
+\includegraphics{../figures/sst-psi-1-1} \caption{Same as Figure~\ref{fig:sst-psi-2} but for cEOF1.}\label{fig:sst-psi-1}
 \end{figure*}
 
-Figure~\ref{fig:sst-psi-1} shows SST and streamfunction regression maps for EOF1.
-There is no significant pattern of SST anomalies associated with either the Real or Imaginary EOF1.
+Figure~\ref{fig:sst-psi-1} shows SST and streamfunction regression maps for cEOF1.
+There is no significant pattern of SST anomalies associated with either the Real or Imaginary cEOF1.
 Consistently, streamfunction anomalies do not show any tropical influence.
@@
 \includegraphics{../figures/pp-america-1.pdf}
-\caption{\label{fig:pp-america}Regression of SON mean precipitation anomalies in South America (mm per day, shaded) and (column a) EOF1 the (row 1) Imaginary and (column 1) Real phase.
+\caption{\label{fig:pp-america}Regression of SON mean precipitation anomalies in South America (mm per day, shaded) and (column a) cEOF1 the (row 1) Imaginary and (column 1) Real phase.
 For the 1979 -- 2018 period.
@@
 cEOF1 is associated with slight decreases in precipitation in Southern Brazil and Paraguay, although these are not statistically significant \ref{fig:pp-america} column a.
-The strongest precipitation anomalies are the ones associated with the Imaginary EOF2.
+The strongest precipitation anomalies are the ones associated with the Imaginary cEOF2.
 The positive anomalies on Southeastern South America (SESA) and Central Chile, and negative anomalies over South Atlantic Convergence Zone is a well known springtime precipitation signature of ENSO \citep{cai2020a} and it is also similar to the precipitation anomalies associated with the A-SAM \citep{campitelli2021}.
@@
 
-The precipitation anomalies in Oceania associated with EOF1 are also weak and not statistically significant (Figure~\ref{fig:pp-oceania} column a) whereas EFO2 is associated with large and expansive anomalies.
-The Real cEOF2 is associated with positive seasonal anomalies in Northern and Eastern Australia, and negative ones over New Zealand, while precipitation anomalies associated with Imaginary EOF2 are negative over all Eastern Australia.
+The precipitation anomalies in Oceania associated with cEOF1 are also weak and not statistically significant (Figure~\ref{fig:pp-oceania} column a) whereas EFO2 is associated with large and expansive anomalies.
+The Real cEOF2 is associated with positive seasonal anomalies in Northern and Eastern Australia, and negative ones over New Zealand, while precipitation anomalies associated with Imaginary cEOF2 are negative over all Eastern Australia.
 Again, these negative anomalies are similar to the springtime precipitation anomalies observed in relation with the Asymmetric SAM \citep{campitelli2021}.
@@
 
-To better understand the relationship between precipitation anomalies over Western Australia and the EOF2, Figure~\ref{fig:australia-pp-phase} plots precipitation anomalies averaged over Eastern Australia (box in Figure~\ref{fig:pp-oceania}) as a function of EOF2 phase.
+To better understand the relationship between precipitation anomalies over Western Australia and the cEOF2, Figure~\ref{fig:australia-pp-phase} plots precipitation anomalies averaged over Eastern Australia (box in Figure~\ref{fig:pp-oceania}) as a function of cEOF2 phase.
 The black line is the equation \(\mathrm{Precipitation} = -0.26\sin{(\mathrm{phase})} + 0.22\cos{(\mathrm{phase})}\), whose coefficient we fitted by weighted least squares using the magnitude of the EFO2 as weight.
@@
 The \(r^2\) corresponding to the fit is 0.37, with p-value \textless{} 0.001.
-There is a strong relationship between the phase of the EOF2 and precipitation anomalies in Eastern Australia, but Real and Imaginary phases we chose are not exactly aligned with the direction that maximises this relationship.
+There is a strong relationship between the phase of the cEOF2 and precipitation anomalies in Eastern Australia, but Real and Imaginary phases we chose are not exactly aligned with the direction that maximises this relationship.
 If one were to perform a more in-depth analysis of the relationship between this pattern and precipitation in this region, one would conceivably align one of the axis (either the Real or the Imaginary) to the one that maximises it.
@@
 The first complex EOF represents the variability of the zonal wave 1 in the stratosphere, but represents a more zonally symmetric monopole in the troposphere.
-There is a statistically positive trend in the magnitude of this EOF, which is consistent with previous studies which showed secular changes in springtime wave-1-like patterns \citep[e.g.][]{raphael2003}.
+There is a statistically positive trend in the magnitude of this cEOF, which is consistent with previous studies which showed secular changes in springtime wave-1-like patterns \citep[e.g.][]{raphael2003}.
 This mode is closely related to anomalies in Total Ozone Column.
@@
 Essentially, it is an alternate representation of the PSA1 and PSA2 patterns \citep{mo2001}.
-We show that the Imaginary EOF2 can be identified with the PSA1 and the Real EOF2 with the PSA2.
+We show that the Imaginary cEOF2 can be identified with the PSA1 and the Real cEOF2 with the PSA2.
 
-We show that this EOF prefers a phase aligned with the Imaginary EOF2.
-This is because the when ENSO is active, it forces the EOF2 to be in the Imaginary phase; positive for El Niño and negative for La Niña.
-When ENSO is neutral, the EOF2 is still active, but with no particular phase.
+We show that this cEOF prefers a phase aligned with the Imaginary cEOF2.
+This is because the when ENSO is active, it forces the cEOF2 to be in the Imaginary phase; positive for El Niño and negative for La Niña.
+When ENSO is neutral, the cEOF2 is still active, but with no particular phase.
 This mirrors the results of \citet{cai2002}, who showed that the CSIRO Model can create PSA-like variability even in the absence of ENSO forcing (with a climatological run), but the variability of one of the PSA modes was enhanced when adding the ENSO signal.
 
-We further show that the Imaginary EOF2 is closely related to the Southern Annular Mode in the troposphere.
+We further show that the Imaginary cEOF2 is closely related to the Southern Annular Mode in the troposphere.
 In fact, it is a virtually perfect indicator of the zonally asymmetric portion of the SAM.
-To the extreme that the leading EOF of geopotential height with the EOF2 removes results in an annular mode that has none of the typical zonal asymmetries of the observed SAM.
+To the extreme that the leading EOF of geopotential height with the cEOF2 removes results in an annular mode that has none of the typical zonal asymmetries of the observed SAM.
 
@@
 
-Precipitation anomalies in South America associated with the Imaginary EOF2 show a clear ENSO-like impact, with positive anomalies in South-Eastern Southamerica, negative anomalies in Southern Brazil and positive anomalies in central Chile.
-Precipitation anomalies associated with the Real EOF2 are low and not statistically significant, showing that the Imaginary phase is optimally aligned with the direction of maximum precipitation impacts.
+Precipitation anomalies in South America associated with the Imaginary cEOF2 show a clear ENSO-like impact, with positive anomalies in South-Eastern Southamerica, negative anomalies in Southern Brazil and positive anomalies in central Chile.
+Precipitation anomalies associated with the Real cEOF2 are low and not statistically significant, showing that the Imaginary phase is optimally aligned with the direction of maximum precipitation impacts.
 
-On the other hand, over Australia, both the Real and Imaginary phase are associated with significant precipitation anomalies, and we further show that the direction of maximum impact is not aligned with our chosen rotation of EOF2.
+On the other hand, over Australia, both the Real and Imaginary phase are associated with significant precipitation anomalies, and we further show that the direction of maximum impact is not aligned with our chosen rotation of cEOF2.
 However, this underscore the benefit of using complex EOF, since it would trivial to rotate it.

### 2021-04-07 analysis/paper/shceof.Rmd
Explica mejor irving2016
(parte de la explicación está en la introducción)
Closes #9

@@
 
-This is analogous to the method of @irving2016, who used reprojection and Fourier filtering to detect "PSA-like" variability and defined as the proper PSA using the peaks of the phase distribution (Compare our Figure\ \@ref(fig:phase-histogram) with their Figure 6).
-
-\textcolor{red}{(Amplia un poco la oración anterior que no queda del todo clara)}
+This is similar to the method used by @irving2016.
+From their Fourier decomposition of reprojected meridional wind fields, they show that the phase distribution is bimodal and isolate the PSA pattern from the rest of the PSA-like variability by selecting events that are near the peacks of the distribution (compare our Figure\ \@ref(fig:phase-histogram) with their Figure 6).
 

### 2021-04-07 analysis/paper/shceof.Rmd
Arregla r2 negativo (estaba mostrando r)
Closes #11

@@
 
-The Imaginary cEOF2 is related with the SAM through the A-SAM in the troposphere, with up to \textcolor{red}{(Acá aparece un -)} `r scales::percent(max_r2[cEOF == "cEOF2"]$estimate, 2)` of shared variance, reached at `r max_r2[cEOF == "cEOF2"]$lev` hPa (Figure\ \@ref(fig:sam-eof-vertical).b2).
+The Imaginary cEOF2 is related with the SAM through the A-SAM in the troposphere, with up to `r scales::percent(max_r2[cEOF == "cEOF2"]$estimate^2, 2)` of shared variance, reached at `r max_r2[cEOF == "cEOF2"]$lev` hPa (Figure\ \@ref(fig:sam-eof-vertical).b2).
 

### 2021-04-07 analysis/paper/shceof.Rmd
Mejora descripción de SAM / cEOF

@@
 (ref:sam-eof-vertical-cap) Coefficient of determination between the real and imaginary part of each cEOF and the SAM, Asymmetric SAM (A-SAM) and Symmetric SAM (S-SAM) indices computed at each level according to @campitelli2021.
-Points mark estimates with p-value < 0.01 corrected for False Detection Rate [@benjamini1995].
+Thick lines represent estimates with p-value < 0.01 corrected for False Detection Rate [@benjamini1995].
 
@@
 sam_r2 %>%
+  .[, p.value_a := p.adjust(p.value, "fdr")] %>% 
   ggplot(aes(lev, estimate^2)) +
   geom_vline(xintercept = c(50, 200),  size = 0.2, color = "gray50") +
-  geom_point(data = ~.x[p.adjust(p.value, "fdr") < 0.01], aes(color = term)) +
-  geom_line(aes(color = term, size = term)) +
+  # geom_point(data = ~.x[p.adjust(p.value, "fdr") < 0.01], aes(color = term)) +
+  geom_line(aes(color = term, group = term, size = p.value_a <= 0.01)) +
   scale_x_level() +
   scale_y_continuous(r2, limits = c(0, 1)) +
-  scale_size_manual(guide = "none",
-                    values = c(full = 1.2,
-                               asym = 0.5,
-                               sym = 0.5)) +
+  scale_size_manual("p-value", values = c("TRUE" = 1, 
+                                      "FALSE" = 0.5), 
+                    guide = "none") +
   scale_color_brewer(NULL, palette = "Dark2", labels = c(full = "SAM",
@@
 
-To explore the relationship between SAM and the modes described from the cEOF, it is computed the coefficient of determination between the cEOFs timeseries and the three SAM indices (SAM, A-SAM and S-SAM) at each level (Fig.\ \@ref(fig:sam-eof-vertical)).
-Both cEOFs bear some modest relationship with the full SAM index (thick green line in Figure\ \@ref(fig:sam-eof-vertical)), although not statistically significant at every level.
-
- \textcolor{red}{(Esta explicación habría que hacerla más detallada. Explicar más sobre lo que se ve en cada panel y después ir interpretando.  )}
-The split between A-SAM and S-SAM gives more insight into the nature of the relationship.
-The relationship between the SAM and ¿the imaginary cEOF1?
-(Figure\ \@ref(fig:sam-eof-vertical).b1) is mediated by S-SAM in the troposphere, but by the A-SAM in the stratosphere.
-
-The Imaginary cEOF2 is related with the SAM through the A-SAM in the troposphere, with up to `r scales::percent(max_r2[cEOF == "cEOF2"]$estimate^2, 2)` of shared variance, reached at `r max_r2[cEOF == "cEOF2"]$lev` hPa (Figure\ \@ref(fig:sam-eof-vertical).b2).
+To explore the relationship between SAM and the modes described from the cEOF, we compute.
+the coefficient of determination between the cEOFs timeseries and the three SAM indices (SAM, A-SAM and S-SAM) at each level (Fig.\ \@ref(fig:sam-eof-vertical)).
+The SAM index is moderately correlated with the Real cEOF1 in all levels, and with the Imaginary cEOF1 and Imaginary cEOF2 in the troposphere.
+It is not correlated with the Real cEOF2 at any level.
+
+The relationship between the tropospheric SAM and cEOF1 is explained entirely by the zonally symmetric component of the SAM as shown by the low and statistically non-significant correlations between the A-SAM and either the Real or Imaginary cEOF1 and the high correlation with the S-SAM below 100 hPa.
+In the stratosphere, the Real cEOF1 is correlated with both A-SAM and S-SAM, while the Imaginary cEOF1 is highly correlated only with the A-SAM.
+These correlations are consistent with the regression patterns of geopotential height in Figure\ \@ref(fig:eof1-regr-gh).
+
+In the case of cEOF2, the moderate correlation between SAM and Imaginary cEOF2 becomes extremely high when only the zonally asymmetric variability of the SAM is considered.
+The Imaginary cEOF2 and the A-SAM share up to `r scales::percent(max_r2[cEOF == "cEOF2"]$estimate^2, 2)` of their variance, reached at `r max_r2[cEOF == "cEOF2"]$lev` hPa (Figure\ \@ref(fig:sam-eof-vertical).b2).
 

### 2021-04-07 analysis/paper/shceof.Rmd
Corrije la correlación que no aparecía
Closes #14

@@
 However, Figure\ \@ref(fig:sst-psi-2).a1 and b1 show that the Real cEOF2 is not associated either with any significant SST nor streamfunction anomalies in the tropics.
-The correlation between the Real cEOF2 and ENSO is also not significant (`r enso_cor[cEOF == "cEOF2" & part == "Real"]$cor_text`) \textcolor{red}{(No aparece el valor en el PDF)}.
+The correlation between the Real cEOF2 and ENSO is also not significant (`r enso_cor[cEOF == "cEOF2" & part == "Real"]$text`) \textcolor{red}{(No aparece el valor en el PDF)}.
 This lack of tropical signal suggests different natures of the different phases of the cEOF2 wave train.

### 2021-04-07 analysis/paper/shceof.Rmd
Explica más la cuestión de SAM y PSA
Closes #13

@@
 (ref:eof-naive-cap) Spatial patterns of the leading 4 EOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2019 (arbitrary units).
-The numbers at the top-left of each panel ate the zonal variance explained by each EOF.
 
@@
 
-In the case of cEOF2, the moderate correlation between SAM and Imaginary cEOF2 becomes extremely high when only the zonally asymmetric variability of the SAM is considered.
-The Imaginary cEOF2 and the A-SAM share up to `r scales::percent(max_r2[cEOF == "cEOF2"]$estimate^2, 2)` of their variance, reached at `r max_r2[cEOF == "cEOF2"]$lev` hPa (Figure\ \@ref(fig:sam-eof-vertical).b2).
-
 ```{r eof-filtered, fig.cap = "Spatial pattern of the leading EOF of 200 hPa geopotential height with the variability of cEOF2 filtered out South of 20º.
@@
 
-eof_filtered %>% 
-  .$left %>% 
-  periodic(lon = c(0, 360)) %>% 
-  ggplot(aes(lon, lat)) +
-  geom_contour_fill(aes(z = eof), breaks = AnchorBreaks(0, exclude = 0)) +
-  geom_contour_tanaka(aes(z = eof), breaks = AnchorBreaks(0, exclude = 0)) +
-  scale_fill_divergent(guide = "none") +
-  geom_qmap(~.x[lat <= -20]) +  
-  scale_x_longitude() +
-  scale_y_latitude() +
-  coord_polar() +
-  theme(axis.text = element_blank())
+# eof_filtered %>% 
+#   .$left %>% 
+#   periodic(lon = c(0, 360)) %>% 
+#   ggplot(aes(lon, lat)) +
+#   geom_contour_fill(aes(z = eof), breaks = AnchorBreaks(0, exclude = 0)) +
+#   geom_contour_tanaka(aes(z = eof), breaks = AnchorBreaks(0, exclude = 0)) +
+#   scale_fill_divergent(guide = "none") +
+#   geom_qmap(~.x[lat <= -20]) +  
+#   scale_x_longitude() +
+#   scale_y_latitude() +
+#   coord_polar() +
+#   theme(axis.text = element_blank())
 ```
@@
 
-As a further illustration \textcolor{red}{(¿De qué?)}, Figure\ \@ref(fig:eof-filtered) shows the spatial pattern of the leading EOF of 200 hPa geopotential height when the variability associated with cEOF2 is removed.
+In the case of cEOF2, the moderate correlation between SAM and Imaginary cEOF2 becomes extremely high when only the zonally asymmetric variability of the SAM is considered.
+The Imaginary cEOF2 and the A-SAM share up to `r scales::percent(max_r2[cEOF == "cEOF2"]$estimate^2, 2)` of their variance, reached at `r max_r2[cEOF == "cEOF2"]$lev` hPa (Figure\ \@ref(fig:sam-eof-vertical).b2).
+Such extremely high correlation between the asymmetric SAM and the Imaginary cEOF2 suggests that these are not different features that share a causal link, but that they might be different ways of characterising the same phenomenon.
+And, in fact, suggests that the asymmetric part of the SAM might be statistical contamination from one phase of the PSA.
+
+<!-- As a further illustration \textcolor{red}{(¿De qué?)}, Figure\ \@ref(fig:eof-filtered) shows the spatial pattern of the leading EOF of 200 hPa geopotential height when the variability associated with cEOF2 is removed.
 The resulting pattern is an nearly zonally symmetrical annular mode.
@@
 Note that the usual definition of the PSA modes as the second and third EOFs creates modes orthogonal to the SAM (defined as the first EOF) and thus impedes this kind of filtering.
-
-(ref:fake-eof-cap) EOFs of a synthetic data set of 41 times steps of geopotential height fields as the sum of the fields predicted by the cEOF2 plus perfectly zonally symmetric SAM-like anomalies with random amplitude.
+-->
+
+(ref:fake-eof-cap) Row a: The three components used to create a synthetic dataset of 41 years of geopotential height at 200 hPa.
+Row b: The three EOF computed from the synthetic dataset with the variance explained by each EOF in parenthesis.
+Units are arbitrary.
+Negative values in blue and positive values in red.
 
@@
   .[, eof := Anomaly(hgt)*sqrt(cos(lat*pi/180)), by = .(lon, lat, season(time))] %>% 
-  EOF(eof ~ lon + lat | time, n = 1:3, data = .)  
-```
-
-```{r fake-eof, fig.cap = "(ref:fake-eof-cap)", fig.width=3.7, fig.height=1.7}
-# era5 %>% 
-#   .[sep_ReIm(cEOF2, format = "wide")[, hgt := NULL], on = .NATURAL] %>% 
-#   .[lev == 200] %>% 
-#   .[lat <= -20] %>% 
-#   .[season(time) == "SON"] %>% 
-#   .[, FitLm(hgt, Real, Imaginary), .(lon, lat)] %>% 
-#   rm_intercept() %>% 
-#   .[, .(lon, lat, eof = estimate, PC = term)] %>% 
-#   rbind(pattern[, .(lon, lat, eof, PC = "SAM")]) %>% 
-#   .[, PC := factor(PC, levels = c("SAM", "Imaginary", "Real"), ordered = TRUE)] %>% 
-#   periodic(lon = c(0, 360)) %>% 
-#   ggplot(aes(lon, lat)) +
-#   geom_contour_fill(aes(z = eof), global.breaks = FALSE, breaks = AnchorBreaks(0, exclude = 0)) +
-#   geom_contour_tanaka(aes(z = eof), global.breaks = FALSE,  breaks = AnchorBreaks(0, exclude = 0)) +
-#   scale_fill_divergent(guide = "none") +
-#   geom_qmap(~.x[lat <= -20]) +  
-#   scale_x_longitude() +
-#   scale_y_latitude() +
-#   coord_polar() +
-#   facet_wrap(~PC) +
-#   axis_labs_smol +
-
-fake_eof %>% 
-  .$left %>% 
+  EOF(eof ~ lon + lat | time, n = 1:3, data = ., suffix = "EOF")  
+```
+
+```{r fake-eof, fig.cap = "(ref:fake-eof-cap)", fig.width=3.7, fig.height=3.2}
+north_thumb <- function(sdev, n){
+  sdev <- sdev^2
+  delta <- sqrt(2 / n) * sdev
+
+  # Separation between each pair of eofs must be 
+  # greater than the sum of their errors.
+  separation <- abs(diff(sdev))
+  separation >  delta[-length(delta)] + delta[-1]
+}
+
+separated <- north_thumb(fake_eof$sdev$sd^2, 41)
+stopifnot(all(separated))
+
+var <- fake_eof$sdev %>%
   copy() %>% 
+    .[, setNames(paste0(EOF, " (", scales::percent(r2), ")"),
+                 EOF)]
+var <- c(var, setNames(c("SAM", "Imaginary", "Real"), c("SAM", "Imaginary", "Real")))
+
+era5 %>%
+  .[sep_ReIm(cEOF2, format = "wide")[, hgt := NULL], on = .NATURAL] %>%
+  .[lev == 200] %>%
+
+  .[season(time) == "SON"] %>%
+  .[, FitLm(hgt, Real, Imaginary), .(lon, lat)] %>%
+  rm_intercept() %>%
+  .[, .(lon, lat, eof = estimate, EOF = term)] %>%
+  rbind(pattern[, .(lon, lat, eof, EOF = "SAM")]) %>%
+  rbind(fake_eof$left, use.names = TRUE) %>% 
    .[lat <= -20] %>%
-  # .[, eof := Anomaly(eof), by = .(lat, cEOF)] %>%
+  .[, EOF := factor(EOF, c("SAM", "Imaginary", "Real", paste0("EOF", 1:3)),
+                    ordered = TRUE)] %>% 
+  .[EOF %in% c("EOF1", "EOF3"), eof := -eof] %>%
+
   periodic(lon = c(0, 360)) %>%
   ggplot(aes(lon, lat)) +
-  geom_contour_fill(aes(z = eof), global.breaks = FALSE, breaks = AnchorBreaks(0, exclude = 0)) +
+
+  geom_contour_fill(data = ~.x[EOF %in% c("SAM", "Imaginary", "Real")],
+    aes(z = eof), global.breaks = FALSE, breaks = AnchorBreaks(0, exclude = 0)) +
+  scale_fill_divergent(guide = "none") +
+  ggnewscale::new_scale_fill() +
+  geom_contour_fill(data = ~.x[!(EOF %in% c("SAM", "Imaginary", "Real"))],
+    aes(z = eof), global.breaks = FALSE, breaks = AnchorBreaks(0, exclude = 0)) +
+
   geom_contour_tanaka(aes(z = eof), global.breaks = FALSE,  breaks = AnchorBreaks(0, exclude = 0)) +
@@
   coord_polar() +
-  facet_wrap(~PC) +
-  theme(axis.text = element_blank())
-```
-
-This suggests that the asymmetric part of the SAM might be statistical contamination from one phase of the PSA.
-Figure\ \@ref(fig:fake-eof) serves as an illustration of this kind of statistical mixing of independent modes.
-If shows the leading (and only) EOFs of a synthetic dataset if `r uniqueN(fake_sam$time)` years created by adding a perfectly zonally symmetrical SAM-like pattern with random amplitudes taken from the real SAM, and geopotential height anomalies predicted by EFO2.
-Even though by construction this dataset has a perfectly symmetric SAM, the EOF decomposition mixes this mode with parts of the the zonally asymmetric cEOF2 pattern, resulting in a pattern eerily similar to the real SAM pattern.
+  facet_wrap(~EOF, ncol = 3, labeller = labeller(EOF = var)) +
+  axis_labs_smol +
+  theme(axis.text = element_blank()) +
+  tag_facets("rc")
+```
+
+To test the hypothesis that a PSA-like feature can appear as statistic contamination of an otherwise zonally symmetrical and uncorrelated SAM, we create a synthetic dataset of `r uniqueN(fake_sam$time)` fields by reconstructing the 200 hPa geopotential field explained by the cEOF2 and adding a perfectly zonally symmetric SAM-like structure with random sign and value.
+To this synthetic dataset we then apply traditional EOF.
+By construction this dataset contains two independent components: a zonally symmetric standing oscillation and a PSA of varying amplitude and phase (Figure\ \@ref(fig:fake-eof) row a).
+The former one is ar standing oscillation which could be represented by a single EOF, while the latter, being a wave which can change phase, needs two EOFs to be represented correctly.
+Thus, the full variance of this synthetic dataset can be perfectly explained by three EOFs.
+
+The three EOFs of the synthetic dataset are shown in Figure\ \@ref(fig:fake-eof) row b.
+The synthetic SAM can be seen in the first EOF while the PSA is explained by the second and third.
+However, even though the synthetic SAM is perfectly zonally symmetric by construction, the leading EOF is not.
+The EOF decomposition mixes this mode with parts of the the zonally asymmetric cEOF2 pattern, resulting in a pattern eerily similar to the real SAM pattern.
+Of note, the three EOFs are well separated according to North's rule of thumb [@north1982].
 
@@
 
-Furthemore, Figure\ \@ref(fig:enso-phase) suggest that strong cEOF2 years tend to coincide with strong ENSO years.
+Furthermore, Figure\ \@ref(fig:enso-phase) suggest that strong cEOF2 years tend to coincide with strong ENSO years.
 The correlation between the absolute magnitude of the ONI and the magnitude of the cEOF2 is `r cor_enso_mag$text`.

### 2021-04-07 analysis/paper/shceof.tex
Renderiza todo

@@
 \includegraphics{../figures/eof-naive-1.pdf}
-\caption{\label{fig:eof-naive}Spatial patterns of the leading 4 EOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2019 (arbitrary units).
-The numbers at the top-left of each panel ate the zonal variance explained by each EOF.}
+\caption{\label{fig:eof-naive}Spatial patterns of the leading 4 EOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2019 (arbitrary units).}
 \end{figure}
@@
 
-This is analogous to the method of \citet{irving2016}, who used reprojection and Fourier filtering to detect ``PSA-like'' variability and defined as the proper PSA using the peaks of the phase distribution (Compare our Figure~\ref{fig:phase-histogram} with their Figure 6).
-
-\textcolor{red}{(Amplia un poco la oración anterior que no queda del todo clara)}
+This is similar to the method used by \citet{irving2016}.
+From their Fourier decomposition of reprojected meridional wind fields, they show that the phase distribution is bimodal and isolate the PSA pattern from the rest of the PSA-like variability by selecting events that are near the peacks of the distribution (compare our Figure~\ref{fig:phase-histogram} with their Figure 6).
 
@@
 \caption{\label{fig:sam-eof-vertical}Coefficient of determination between the real and imaginary part of each cEOF and the SAM, Asymmetric SAM (A-SAM) and Symmetric SAM (S-SAM) indices computed at each level according to \citet{campitelli2021}.
-Points mark estimates with p-value \textless{} 0.01 corrected for False Detection Rate \citep{benjamini1995}.}
+Thick lines represent estimates with p-value \textless{} 0.01 corrected for False Detection Rate \citep{benjamini1995}.}
 \end{figure}
 
-To explore the relationship between SAM and the modes described from the cEOF, it is computed the coefficient of determination between the cEOFs timeseries and the three SAM indices (SAM, A-SAM and S-SAM) at each level (Fig.~\ref{fig:sam-eof-vertical}).
-Both cEOFs bear some modest relationship with the full SAM index (thick green line in Figure~\ref{fig:sam-eof-vertical}), although not statistically significant at every level.
+To explore the relationship between SAM and the modes described from the cEOF, we compute.
+the coefficient of determination between the cEOFs timeseries and the three SAM indices (SAM, A-SAM and S-SAM) at each level (Fig.~\ref{fig:sam-eof-vertical}).
+The SAM index is moderately correlated with the Real cEOF1 in all levels, and with the Imaginary cEOF1 and Imaginary cEOF2 in the troposphere.
+It is not correlated with the Real cEOF2 at any level.
 
-\textcolor{red}{(Esta explicación habría que hacerla más detallada. Explicar más sobre lo que se ve en cada panel y después ir interpretando.  )}
-The split between A-SAM and S-SAM gives more insight into the nature of the relationship.
-The relationship between the SAM and ¿the imaginary cEOF1?
-(Figure~\ref{fig:sam-eof-vertical}.b1) is mediated by S-SAM in the troposphere, but by the A-SAM in the stratosphere.
-
-The Imaginary cEOF2 is related with the SAM through the A-SAM in the troposphere, with up to \textcolor{red}{(Acá aparece un -)} -96\% of shared variance, reached at 225 hPa (Figure~\ref{fig:sam-eof-vertical}.b2).
-
-\begin{figure}
-\centering
-\includegraphics{../figures/eof-filtered-1.pdf}
-\caption{\label{fig:eof-filtered}Spatial pattern of the leading EOF of 200 hPa geopotential height with the variability of cEOF2 filtered out South of 20º. Arbitrary units.}
-\end{figure}
+The relationship between the tropospheric SAM and cEOF1 is explained entirely by the zonally symmetric component of the SAM as shown by the low and statistically non-significant correlations between the A-SAM and either the Real or Imaginary cEOF1 and the high correlation with the S-SAM below 100 hPa.
+In the stratosphere, the Real cEOF1 is correlated with both A-SAM and S-SAM, while the Imaginary cEOF1 is highly correlated only with the A-SAM.
+These correlations are consistent with the regression patterns of geopotential height in Figure~\ref{fig:eof1-regr-gh}.
 
@@
 
-As a further illustration \textcolor{red}{(¿De qué?)}, Figure~\ref{fig:eof-filtered} shows the spatial pattern of the leading EOF of 200 hPa geopotential height when the variability associated with cEOF2 is removed.
-The resulting pattern is an nearly zonally symmetrical annular mode.
-This mode is highly correlated with SAM (0.75 (CI: 0.58 -- 0.86)), but only with the symmetrical part (0.88 (CI: 0.79 -- 0.94)).
-Note that the usual definition of the PSA modes as the second and third EOFs creates modes orthogonal to the SAM (defined as the first EOF) and thus impedes this kind of filtering.
+In the case of cEOF2, the moderate correlation between SAM and Imaginary cEOF2 becomes extremely high when only the zonally asymmetric variability of the SAM is considered.
+The Imaginary cEOF2 and the A-SAM share up to 92\% of their variance, reached at 225 hPa (Figure~\ref{fig:sam-eof-vertical}.b2).
+Such extremely high correlation between the asymmetric SAM and the Imaginary cEOF2 suggests that these are not different features that share a causal link, but that they might be different ways of characterising the same phenomenon.
+And, in fact, suggests that the asymmetric part of the SAM might be statistical contamination from one phase of the PSA.
 
@@
 \includegraphics{../figures/fake-eof-1.pdf}
-\caption{\label{fig:fake-eof}EOFs of a synthetic data set of 41 times steps of geopotential height fields as the sum of the fields predicted by the cEOF2 plus perfectly zonally symmetric SAM-like anomalies with random amplitude.}
+\caption{\label{fig:fake-eof}Row a: The three components used to create a synthetic dataset of 41 years of geopotential height at 200 hPa.
+Row b: The three EOF computed from the synthetic dataset with the variance explained by each EOF in parenthesis.
+Units are arbitrary.
+Negative values in blue and positive values in red.}
 \end{figure}
 
-This suggests that the asymmetric part of the SAM might be statistical contamination from one phase of the PSA.
-Figure~\ref{fig:fake-eof} serves as an illustration of this kind of statistical mixing of independent modes.
-If shows the leading (and only) EOFs of a synthetic dataset if 41 years created by adding a perfectly zonally symmetrical SAM-like pattern with random amplitudes taken from the real SAM, and geopotential height anomalies predicted by EFO2.
-Even though by construction this dataset has a perfectly symmetric SAM, the EOF decomposition mixes this mode with parts of the the zonally asymmetric cEOF2 pattern, resulting in a pattern eerily similar to the real SAM pattern.
+To test the hypothesis that a PSA-like feature can appear as statistic contamination of an otherwise zonally symmetrical and uncorrelated SAM, we create a synthetic dataset of 41 fields by reconstructing the 200 hPa geopotential field explained by the cEOF2 and adding a perfectly zonally symmetric SAM-like structure with random sign and value.
+To this synthetic dataset we then apply traditional EOF.
+By construction this dataset contains two independent components: a zonally symmetric standing oscillation and a PSA of varying amplitude and phase (Figure~\ref{fig:fake-eof} row a).
+The former one is ar standing oscillation which could be represented by a single EOF, while the latter, being a wave which can change phase, needs two EOFs to be represented correctly.
+Thus, the full variance of this synthetic dataset can be perfectly explained by three EOFs.
+
+The three EOFs of the synthetic dataset are shown in Figure~\ref{fig:fake-eof} row b.
+The synthetic SAM can be seen in the first EOF while the PSA is explained by the second and third.
+However, even though the synthetic SAM is perfectly zonally symmetric by construction, the leading EOF is not.
+The EOF decomposition mixes this mode with parts of the the zonally asymmetric cEOF2 pattern, resulting in a pattern eerily similar to the real SAM pattern.
+Of note, the three EOFs are well separated according to North's rule of thumb \citep{north1982}.
 
@@
 However, Figure~\ref{fig:sst-psi-2}.a1 and b1 show that the Real cEOF2 is not associated either with any significant SST nor streamfunction anomalies in the tropics.
-The correlation between the Real cEOF2 and ENSO is also not significant () \textcolor{red}{(No aparece el valor en el PDF)}.
+The correlation between the Real cEOF2 and ENSO is also not significant (-0.0017 (CI: -0.31 -- 0.31)) \textcolor{red}{(No aparece el valor en el PDF)}.
 This lack of tropical signal suggests different natures of the different phases of the cEOF2 wave train.
@@
 
-Furthemore, Figure~\ref{fig:enso-phase} suggest that strong cEOF2 years tend to coincide with strong ENSO years.
+Furthermore, Figure~\ref{fig:enso-phase} suggest that strong cEOF2 years tend to coincide with strong ENSO years.
 The correlation between the absolute magnitude of the ONI and the magnitude of the cEOF2 is 0.44 (CI: 0.16 -- 0.66).

### 2021-04-08 analysis/paper/shceof.Rmd
Elimina comentario de #13

@@
 
-\textcolor{red}{(La discusión de los próximos parráfos me resulta confusa.
-Deberías aclarar mejor por qué mostras cada cosa que mostras y cómo se relaciona con lo que mostraste antes.
-También me resulta como un resultado que se va un poco de tema, por eso creo que debería estar mejor contada esta parte para poder incluirla)}
-
 In the case of cEOF2, the moderate correlation between SAM and Imaginary cEOF2 becomes extremely high when only the zonally asymmetric variability of the SAM is considered.

### 2021-04-08 analysis/paper/shceof.Rmd
Agrega referencia a teleconexión

@@
 The Imaginary cEOF2 is associated with strong wave-like streamfunction anomalies emanating from the tropics (Figure\ \@ref(fig:sst-psi-2).b2).
-This is consistent with what we know of the effect of ENSO on the extratropics: SST anomalies initiate anomalous convection that excites Rossby waves that propagate meridionally towards higher latitudes \textcolor{red}{(REFS??)}.
+This is consistent with what we know of the effect of ENSO on the extratropics: SST anomalies initiate anomalous convection that excites Rossby waves that propagate meridionally towards higher latitudes [@mo2000].
 
@@
 However, Figure\ \@ref(fig:sst-psi-2).a1 and b1 show that the Real cEOF2 is not associated either with any significant SST nor streamfunction anomalies in the tropics.
-The correlation between the Real cEOF2 and ENSO is also not significant (`r enso_cor[cEOF == "cEOF2" & part == "Real"]$text`) \textcolor{red}{(No aparece el valor en el PDF)}.
+The correlation between the Real cEOF2 and ENSO is also not significant (`r enso_cor[cEOF == "cEOF2" & part == "Real"]$text`).
 This lack of tropical signal suggests different natures of the different phases of the cEOF2 wave train.

### 2021-04-08 analysis/paper/shceof.Rmd
Agrega fit de modelo sinusoidal en SESA
Closes #15

@@
 
+```{r sinusoidal-sesa}
+gdata <- cmap %>% 
+  .[lon %between% sesa$lon  & lat %between% sesa$lat] %>% 
+  .[, .(pp = weighted.mean(pp, cos(lat*pi/180))), by = time] %>% 
+  .[, pp := Anomaly(pp), by = season(time)] %>% 
+  .[ceof$eof[[1]]$left, on  = "time"] %>% 
+  .[cEOF == "cEOF2"] %>% 
+  .[, arg := Arg(hgt)] %>% 
+  .[, mag := Mod(hgt)]
+
+model <- lm(pp ~ I(sin(arg)) + I(cos(arg)) - 1, data = gdata)
+
+fstatistic <- summary(model)$fstatistic
+p_value <- pf(fstatistic["value"], fstatistic["numdf"], fstatistic["dendf"], 
+        lower.tail = FALSE)
+```
+
 Figure\ \@ref(fig:pp-america) shows regression maps of seasonal precipitation with each cEOF in South America.
@@
 There is a phase dependence of the precipitation anomalies in SESA (not shown), which resembles the relationship between ONI and the phase of cEOF2.
-\textcolor{red}{(Podrías agregar acá el coeficiente de determinación)}
+A sinusoidal fit has an $r^2$\ of\ `r round(summary(model)$r.squared, 2)` with p-value\ `r report_p(p_value)`.
 

### 2021-04-08 analysis/paper/shceof.Rmd
Agerga referencia de ENSO en Australia

@@
 The Real cEOF2 is associated with positive seasonal anomalies in Northern and Eastern Australia, and negative ones over New Zealand, while precipitation anomalies associated with Imaginary cEOF2 are negative over all Eastern Australia.
-Again, these negative anomalies are similar to the springtime precipitation anomalies observed in relation with the Asymmetric SAM [@campitelli2021].
-\textcolor{red}{(Habría que agregar alguna referencia del ENSO que es más conocido)}
+Again, these negative anomalies are similar to the springtime precipitation anomalies observed in relation with the Asymmetric SAM [@campitelli2021] and ENSO [@cai2011].
 

### 2021-04-08 analysis/paper/shceof.Rmd
Pone ecuación en el plot y agrega medidas del fit en el txto
Closes #16

@@
 Colours denote years with magnitude of cEOF2 greater or smaller than the 50th percentile.
-Black line is the fit ONI ~ cos(phase) + sin(phase) computed by OLS weighted by the magnitude of cEOF2.", fig.width = width_column, fig.height=3}
+Black line is the fit ONI ~ sin(phase) computed by OLS weighted by the magnitude of cEOF2.", fig.width = width_column, fig.height=3}
 gdata <- ceof[, eof[[1]]$left, by = .(season)] %>%
@@
   .[, lm(oni ~ I(sin(arg)) - 1, weights = mag )]
+
+equation <- latex2exp::TeX(paste0("$\\mathrm{ONI} = ", 
+        signif(coef(cos_model)[1], 2), "\\sin{(\\mathrm{phase})}$"),
+        output = "text")
 
@@
   coord_cartesian(xlim = c(-pi, pi)) +
+  labs(subtitle = equation) +
   panel_background
@@
 In years with near neutral ENSO, the phase of the cEOF2 is much more variable.
-The black line in Figure\ \@ref(fig:enso-phase) is an harmonical fit of the relationship between ONI and cEOF2 phase.
+The black line in Figure\ \@ref(fig:enso-phase) is a sinusoidal fit of the relationship between ONI and cEOF2 phase.
 The $r^2$ corresponding to the fit is `r signif(summary(cos_model)$r.squared, 2)`, with p-value `r report_p(coef(summary(cos_model))[4])`, indicating a quasi-sinusoidal relation between these two variables.
@@
 
-(ref:australia-pp-phase-cap) Same as Figure\ \@ref(fig:enso-phase) but for precipitacion anomalies averaged between  `r knitr::combine_words(LonLabel(box_australia$lon))` and `r knitr::combine_words(LatLabel(box_australia$lat))` (box shown in Figure\ \@ref(fig:pp-oceania) column b).
+(ref:australia-pp-phase-cap) Same as Figure\ \@ref(fig:enso-phase) but for precipitation anomalies averaged between  `r knitr::combine_words(LonLabel(box_australia$lon))` and `r knitr::combine_words(LatLabel(box_australia$lat))` (box shown in Figure\ \@ref(fig:pp-oceania) column b).
 
@@
 
+equation <- latex2exp::TeX(paste0("$\\mathrm{Precipitation} = ", 
+        signif(coef(model)[1], 2), "\\sin{(\\mathrm{phase})} + ", 
+        signif(coef(model)[2], 2), "\\cos{(\\mathrm{phase})}$"), output = "text")
+
 gdata %>% 
@@
              size = 3) +
+
+  # annotate("text", x = 90*pi/180, y = 1, label = equation, parse = TRUE) +
 
@@
 To better understand the relationship between precipitation anomalies over Western Australia and the cEOF2, Figure\ \@ref(fig:australia-pp-phase) plots precipitation anomalies averaged over Eastern Australia (box in Figure\ \@ref(fig:pp-oceania)) as a function of cEOF2 phase.
-The black line is the equation $\mathrm{Precipitation} = `r signif(coef(model)[1], 2)`\sin{(\mathrm{phase})} + `r signif(coef(model)[2], 2)`\cos{(\mathrm{phase})}$, whose coefficient we fitted by weighted least squares using the magnitude of the EFO2 as weight.
-\textcolor{red}{(Como te puse antes la ecuación como mucho la pondría sobre la figura)}
+The black line shows a sinusoidal fit, whose coefficients we fitted by weighted least squares using the magnitude of the EFO2 as weight.
 The $r^2$ corresponding to the fit is `r signif(summary(model)$r.squared, 2)`, with p-value `r report_p(p_value)`.

### 2021-04-08 analysis/paper/shceof.Rmd
Muestra mejor significancia en SAm-vertical

@@
   # geom_point(data = ~.x[p.adjust(p.value, "fdr") < 0.01], aes(color = term)) +
-  geom_line(aes(color = term, group = term, size = p.value_a <= 0.01)) +
+  geom_line(data = ~copy(.x)[p.value_a > 0.01, estimate := NA], 
+            aes(colour = stage(term, after_scale = scales::alpha(colour, 0.5))), 
+            size = 2) +
+  geom_line(aes(color = term, group = term)) +
+
   scale_x_level() +

### 2021-04-08 analysis/paper/shceof.tex
Renderiza todo

@@
 The Imaginary cEOF2 is associated with strong wave-like streamfunction anomalies emanating from the tropics (Figure~\ref{fig:sst-psi-2}.b2).
-This is consistent with what we know of the effect of ENSO on the extratropics: SST anomalies initiate anomalous convection that excites Rossby waves that propagate meridionally towards higher latitudes \textcolor{red}{(REFS??)}.
+This is consistent with what we know of the effect of ENSO on the extratropics: SST anomalies initiate anomalous convection that excites Rossby waves that propagate meridionally towards higher latitudes \citep{mo2000}.
 
@@
 However, Figure~\ref{fig:sst-psi-2}.a1 and b1 show that the Real cEOF2 is not associated either with any significant SST nor streamfunction anomalies in the tropics.
-The correlation between the Real cEOF2 and ENSO is also not significant (-0.0017 (CI: -0.31 -- 0.31)) \textcolor{red}{(No aparece el valor en el PDF)}.
+The correlation between the Real cEOF2 and ENSO is also not significant (-0.0017 (CI: -0.31 -- 0.31)).
 This lack of tropical signal suggests different natures of the different phases of the cEOF2 wave train.
@@
 Colours denote years with magnitude of cEOF2 greater or smaller than the 50th percentile.
-Black line is the fit ONI \textasciitilde{} cos(phase) + sin(phase) computed by OLS weighted by the magnitude of cEOF2.}
+Black line is the fit ONI \textasciitilde{} sin(phase) computed by OLS weighted by the magnitude of cEOF2.}
 \end{figure}
@@
 In years with near neutral ENSO, the phase of the cEOF2 is much more variable.
-The black line in Figure~\ref{fig:enso-phase} is an harmonical fit of the relationship between ONI and cEOF2 phase.
+The black line in Figure~\ref{fig:enso-phase} is a sinusoidal fit of the relationship between ONI and cEOF2 phase.
 The \(r^2\) corresponding to the fit is 0.56, with p-value \textless{} 0.001, indicating a quasi-sinusoidal relation between these two variables.
@@
 There is a phase dependence of the precipitation anomalies in SESA (not shown), which resembles the relationship between ONI and the phase of cEOF2.
-\textcolor{red}{(Podrías agregar acá el coeficiente de determinación)}
+A sinusoidal fit has an \(r^2\)~of~0.27 with p-value~= 0.002.
 
@@
 The Real cEOF2 is associated with positive seasonal anomalies in Northern and Eastern Australia, and negative ones over New Zealand, while precipitation anomalies associated with Imaginary cEOF2 are negative over all Eastern Australia.
-Again, these negative anomalies are similar to the springtime precipitation anomalies observed in relation with the Asymmetric SAM \citep{campitelli2021}.
-\textcolor{red}{(Habría que agregar alguna referencia del ENSO que es más conocido)}
+Again, these negative anomalies are similar to the springtime precipitation anomalies observed in relation with the Asymmetric SAM \citep{campitelli2021} and ENSO \citep{cai2011}.
 
@@
 \includegraphics{../figures/australia-pp-phase-1.pdf}
-\caption{\label{fig:australia-pp-phase}Same as Figure~\ref{fig:enso-phase} but for precipitacion anomalies averaged between 130°E and 155°E and 40°S and 15°S (box shown in Figure~\ref{fig:pp-oceania} column b).}
+\caption{\label{fig:australia-pp-phase}Same as Figure~\ref{fig:enso-phase} but for precipitation anomalies averaged between 130°E and 155°E and 40°S and 15°S (box shown in Figure~\ref{fig:pp-oceania} column b).}
 \end{figure}
@@
 To better understand the relationship between precipitation anomalies over Western Australia and the cEOF2, Figure~\ref{fig:australia-pp-phase} plots precipitation anomalies averaged over Eastern Australia (box in Figure~\ref{fig:pp-oceania}) as a function of cEOF2 phase.
-The black line is the equation \(\mathrm{Precipitation} = -0.26\sin{(\mathrm{phase})} + 0.22\cos{(\mathrm{phase})}\), whose coefficient we fitted by weighted least squares using the magnitude of the EFO2 as weight.
-\textcolor{red}{(Como te puse antes la ecuación como mucho la pondría sobre la figura)}
+The black line shows a sinusoidal fit, whose coefficients we fitted by weighted least squares using the magnitude of the EFO2 as weight.
 The \(r^2\) corresponding to the fit is 0.37, with p-value \textless{} 0.001.

### 2021-04-09 analysis/paper/shceof.Rmd
Promedia ponderando cantidad de días
Closes #23
Un montón de laburo y bastante al pedo porque (naturalmente) no cambió casi nada.

@@
   .[season(time) == "SON"] %>%
+  .[, oni := oni*season_weights(time)] %>% 
   .[, .(oni = mean(oni)), by = .(time = seasonally(time))]
@@
   .[season(time) == "SON"] %>%
-  .[, dim :=  lubridate::days_in_month(time)] %>%
-  .[, .(estimate = weighted.mean(estimate, dim)),
+  .[, estimate := estimate*season_weights(time)] %>% 
+  .[, .(estimate = mean(estimate)),
     by = .(lev, term, time = seasonally(time))]
@@
   .[season(time) == "SON"] %>%
-  .[, .(hgt = mean(hgt), 
-        t = mean(t)), by = .(lev, lon, lat, time = seasonally(time))]
+  .[, w := season_weights(time)] %>% 
+  .[, .(hgt = mean(hgt*w), 
+        t = mean(t*w)), by = .(lev, lon, lat, time = seasonally(time))]
 ```
@@
   .[season(time) == "SON"] %>%
-  .[, .(vo = mean(vo)), by = .(lon, lat, lev, time = seasonally(time))] %>%
+  .[, w := season_weights(time)] %>% 
+  .[, .(vo = mean(vo*w)), by = .(lon, lat, lev, time = seasonally(time))] %>%
   .[, solve_poisson(vo, lon, lat), by = .(lev, time)] %>%
@@
   .[, toc := toc/2.1415e-5] %>%    # transform to Dobson Units (https://sacs.aeronomie.be/info/dobson.php)
-  .[, .(toc = mean(toc, na.rm = TRUE)), by = .(lon, lat, time = seasonally(time))]
+  .[, w := season_weights(time)] %>% 
+  .[, .(toc = mean(toc*w)), by = .(lon, lat, time = seasonally(time))]
 
@@
   normalise_coords() %>% 
-  .[, .(hgt = mean(hgt)), by = .(lon, lat, time = seasonally(time))] %>% 
+  .[, w := season_weights(time)] %>% 
+  .[, .(hgt = mean(hgt*w)), by = .(lon, lat, time = seasonally(time))] %>% 
   .[, value := Anomaly(hgt)*sqrt(cos(lat*pi/180)), by = .(lon, lat, season(time))] %>%
@@
     .[season(time) == "SON"] %>%
-    .[, .(hgt = mean(hgt)), by = .(lev, lon, lat, time = seasonally(time))] %>%
+    .[, w := season_weights(time)] %>% 
+    .[, .(hgt = mean(hgt*w)), by = .(lev, lon, lat, time = seasonally(time))] %>%
     .[, season := season(time)] %>%
@@
   .[season(time) == "SON"] %>%
-  .[, .(hgt = mean(hgt)), by = .(lev, lon, lat, time = seasonally(time))] %>%
+  .[, w := season_weights(time)] %>% 
+  .[, .(hgt = mean(hgt*w)), by = .(lev, lon, lat, time = seasonally(time))] %>%
   .[, season := season(time)] %>%
@@
   .[season(time) == "SON"] %>% 
-  .[, .(t = mean(t),
-        o3 = mean(o3),
-        hgt = mean(hgt)), by = .(lev, lon, lat, time = seasonally(time))]
+  .[, w := season_weights(time)] %>% 
+  .[, .(t   = mean(t*w),
+        o3  = mean(o3*w),
+        hgt = mean(hgt*w)), by = .(lev, lon, lat, time = seasonally(time))]
 ```
@@
 To this synthetic dataset we then apply traditional EOF.
-By construction this dataset contains two independent components: a zonally symmetric standing oscillation and a PSA of varying amplitude and phase (Figure\ \@ref(fig:fake-eof) row a).
+By construction this dataset contains two independent components: a zonally symmetric standing oscillation and a PSA-like of varying amplitude and phase (Figure\ \@ref(fig:fake-eof) row a).
 The former one is ar standing oscillation which could be represented by a single EOF, while the latter, being a wave which can change phase, needs two EOFs to be represented correctly.
@@
   .[season(time) == "SON"] %>%
-  .[, .(t = mean(t)), by = .(lon, lat, time = seasonally(time))]
+  .[, w := season_weights(time)] %>% 
+  .[, .(t = mean(t*w)), by = .(lon, lat, time = seasonally(time))]
 ```
@@
    annotate(shadowtext:::GeomShadowText,
-           label = c("- Real", "- Imaginary", "+ Real", "+ Imaginary", "+ Real"),
+           label = c("- Real", "- Imaginary", "+ Real", "+ Imaginary", "- Real"),
            color = "gray60", bg.colour = "white", bg.r = 0.2,
@@
   normalise_coords() %>%
-  .[, .(pp = mean(pp, na.rm = TRUE)), by = .(lon, lat, time = seasonally(time))]
+  .[, w := season_weights(time)] %>% 
+  .[, .(pp = mean(pp*w)), by = .(lon, lat, time = seasonally(time))]
 ```
@@
   annotate(shadowtext:::GeomShadowText,
-           label = c("- Real", "- Imaginary", "+ Real", "+ Imaginary", "+ Real"),
+           label = c("- Real", "- Imaginary", "+ Real", "+ Imaginary", "- Real"),
            color = "gray60", bg.colour = "white", bg.r = 0.2,

### 2021-04-09 analysis/paper/shceof.Rmd
Cambia a ERSST v5
Closes #28

@@
 We derived streamfunction at 200 hPa from ERA5 vorticity using the FORTRAN subroutine FISHPACK [@fishpack] and computed horizontal wave activity fluxes following @plumb1985.
+Sea Surface Temperature (SST) fields were taken from Extended Reconstructed Sea Surface Temperature (ERSST) v5 [@huang2017].
 For precipitation data we used monthly data from the CPC Merged Analysis of Precipitation [@cmap], with a 2.5\degree\ resolution in latitude and longitude.
@@
 ```{r sst}
-sst <- ERA5_SST() %>%
-  ReadNetCDF(vars = c("t" = "sst")) %>%
+sst <- ERSST() %>%
+  readRDS() %>% 
+  .[season(time) == "SON"] %>%
   normalise_coords() %>%
-  .[season(time) == "SON"] %>%
   .[, w := season_weights(time)] %>% 
@@
   .[, term := factor_ReIm(term)]
+
 ```
@@
 plot_sst_psi(which_cEOF = "cEOF2", which_lev = 200)
+
 ```
@@
 plot_sst_psi(which_cEOF = "cEOF1", which_lev = 200)
+
 ```

### 2021-04-09 R/0-download.R

@@ -184,3 +184,24 @@ simple_download <- function(url) {
 }
 
 
+download_ersst <- function(file) {
+  base_url <- "https://www.ncei.noaa.gov/pub/data/cmb/ersst/v5/netcdf/ersst.v5."
+
+  years <- 1970:2019
+  months <- formatC(9:11, width = 2, flag = "0")
+  dates <- data.table::CJ(years, months)[, paste0(years, months)]
+
+  urls <- paste0(base_url, dates, ".nc")
+
+  files <- vapply(urls, function(url) {
+    file <- tempfile(fileext = ".nc")
+    download.file(url, file)
+    file
+  }, character(1))
+
+  data <- lapply(files, function(file) metR::ReadNetCDF(file, vars = c(t = "sst")))
+  data <- data.table::rbindlist(data)
+
+  saveRDS(data, file)
+  file
+}

### 2021-04-09 R/data_locations.R

@@ -118,3 +118,8 @@ CMAP <- data_nc_fun(file = data_path("raw", "cmap.mon.mean.nc"),
 O3 <- data_nc_fun(file = data_path("raw", "o3.mon.mean.nc"),
                     download = download_cds(request_o3_era5),
                     citation = citation_o3_era5)
+
+#' @export
+#' @rdname data_locations
+ERSST <- data_rds_fun(file = data_path("raw", "ersst.Rds"),
+                      download = download_ersst)

### 2021-04-09 analysis/scripts/01-get-data.R

@@ -17,7 +17,7 @@ NCEP_VOR(force)
 # Download ERA5 -----------------------------------------------------------
 ERA5(force)
 ERA5_BE(force)
-ERA5_SST(force)
+ERSST(force)
 
 # Download ERA20C -----------------------------------------------------------
 ERA20C(force)

### 2021-04-09 analysis/paper/shceof.Rmd
Agrega la barra de colores a todas las figuras
Closes #30

@@
 
-```{r define-plot_ceof}
-plot_ceof <- function(data, n = 1:2, formula = lev  ~  cEOF) {
-
-  gdata <- data[, cut(eof[[1]], n)$right, by = .(season)]
-
-  var <- data[, cut(eof[[1]], n)$sdev, by = .(season)] %>%
+```{r ceofs-1, fig.cap = "Spatial patterns of the leading 2 cEOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2019.
+Real part in shading, imaginary part in contours.
+Arbitrary units.", fig.width=width_column, fig.height=width_column*1.2}
+
+gdata <- ceof[, eof[[1]]$right, by = .(season)] %>% 
+  copy() %>% 
+  .[, hgt := hgt/sd(abs(hgt))]
+
+var <- ceof[, eof[[1]]$sdev, by = .(season)] %>%
   .[, setNames(paste0(cEOF, " (", scales::percent(r2), ")"),
@@
   geom_contour_fill(aes(z = Re(hgt), fill = ..level..),
-                      breaks = AnchorBreaks(0, 0.005, exclude = 0)) +
+                    breaks = AnchorBreaks(0, 1, exclude = 0)) +
   geom_contour2(aes(z = Im(hgt),  linetype = factor(sign(..level..))),
-                  breaks = AnchorBreaks(0, 0.005, exclude = 0)) +
-
-    scale_fill_divergent_discretised(guide = "none") +
+                breaks = AnchorBreaks(0, 1, exclude = 0)) +
+
+  scale_fill_divergent_discretised(guide = guide_colorsteps_bottom()) +
 
@@
                         guide = "none") +
-    facet_grid(formula,
+  facet_grid( lev  ~  cEOF,
              labeller = labeller(lev = lab_lev,
@@
   coord_quickmap(ylim = c(NA, -20)) +
-    theme(legend.title = element_blank())
-}
-```
-
-```{r ceofs-1, fig.cap = "Spatial patterns of the leading 2 cEOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2019.
-Real part in shading, imaginary part in contours.", fig.width=width_column, fig.height=width_column}
-ceof[season == "SON"] %>%
-  plot_ceof() +
+  theme(legend.title = element_blank()) +
   coord_polar() +
@@
   tag_facets(tag = "cr")
+
 ```
@@
 Units are arbitrary.
-Negative values in blue and positive values in red.
 
@@
 
-```{r fake-eof, fig.cap = "(ref:fake-eof-cap)", fig.width=3.7, fig.height=3.2}
+```{r fake-eof, fig.cap = "(ref:fake-eof-cap)", fig.width=3.7, fig.height=3.3}
 north_thumb <- function(sdev, n){
@@
   .[EOF %in% c("EOF1", "EOF3"), eof := -eof] %>% 
+  .[!(EOF %in% c("SAM", "Imaginary", "Real")), eof := eof/sd(eof)] %>% 
+  .[EOF %in% c("SAM", "Imaginary", "Real"), eof := eof/sd(eof)] %>%
+
+  # geom_contour_tanaka has problem with the perfectly symemtric SAM
+  # adding just a tiny bit of noise solves it. 
+  .[EOF == "SAM", eof := eof + rnorm(.N, sd = 1e-5)] %>%
 
@@
   ggplot(aes(lon, lat)) +
-
-  geom_contour_fill(data = ~.x[EOF %in% c("SAM", "Imaginary", "Real")],
-    aes(z = eof), global.breaks = FALSE, breaks = AnchorBreaks(0, exclude = 0)) +
-  scale_fill_divergent(guide = "none") +
-  ggnewscale::new_scale_fill() +
-  geom_contour_fill(data = ~.x[!(EOF %in% c("SAM", "Imaginary", "Real"))],
-    aes(z = eof), global.breaks = FALSE, breaks = AnchorBreaks(0, exclude = 0)) +
-
-  geom_contour_tanaka(aes(z = eof), global.breaks = FALSE,  breaks = AnchorBreaks(0, exclude = 0)) +
-  scale_fill_divergent(guide = "none") +
+  geom_contour_fill(aes(z = eof, fill = ..level..), global.breaks = TRUE, 
+                    breaks = AnchorBreaks(0, 0.5, exclude = 0)) +
+
+  geom_contour_tanaka(aes(z = eof), global.breaks = TRUE, 
+                      breaks = AnchorBreaks(0,0.5,  exclude = 0)) +
+  scale_fill_divergent_discretised(NULL, guide = guide_colorsteps_bottom()) +
   geom_qmap(~.x[lat <= -20]) +

### 2021-04-09 analysis/paper/shceof.Rmd
Agrega más barras de colores
(#30)

@@
 
-```{r eof-naive, fig.cap = "(ref:eof-naive-cap)", fig.width=2.5, fig.height=3}
+```{r eof-naive, fig.cap = "(ref:eof-naive-cap)", fig.width=2.5, fig.height=3.2}
 eof_naive[, eof[[1]]$right, by = lev] %>%
   .[lev == 50] %>% 
+  .[ , hgt := hgt/sd(hgt)] %>% 
   periodic(lon = c(0, 360)) %>%
@@
   geom_contour_fill(aes(z = hgt, fill = ..level..),
-                    breaks = AnchorBreaks(0, 0.01, exclude = 0)) +
-  geom_contour_tanaka(aes(z = hgt), breaks = AnchorBreaks(0, 0.01, exclude = 0), 
+                    breaks = AnchorBreaks(0, 1, exclude = 0)) +
+  geom_contour_tanaka(aes(z = hgt), breaks = AnchorBreaks(0, 1, exclude = 0), 
                     range = c(0.01, 0.2)) +
@@
   geom_qmap(~.x[lat <= -20], size = 0.1) +
-  scale_fill_divergent_discretised(guide = "none")  +
+  scale_fill_divergent_discretised(NULL, guide = guide_colorsteps_bottom(10))  +
   scale_y_latitude() +

### 2021-04-09 analysis/paper/shceof.Rmd
Cambia orden (SAM -> PSA)
Closes #36

@@
 
-### PSA
-
-(ref:psa-eof2-cap) Relationship between the Real and Imaginary parts of cEOF2 and the PSA1 and PSA2 modes computed as the second and third EOFs of seasonal geopotential height anomalies [@mo2001].
-For the period 1979 -- 2019.
-
-```{r psa-eof2, fig.cap = "(ref:psa-eof2-cap)", fig.width=7, fig.height=3, fig.env="figure*"}
+### SAM
+
+(ref:sam-eof-vertical-cap) Coefficient of determination between the real and imaginary part of each cEOF and the SAM, Asymmetric SAM (A-SAM) and Symmetric SAM (S-SAM) indices computed at each level according to @campitelli2021.
+Thick lines represent estimates with p-value < 0.01 corrected for False Detection Rate [@benjamini1995].
+
+```{r sam-eof-vertical, fig.cap = "(ref:sam-eof-vertical-cap)", fig.height=4, fig.width=width_column}
+time <- ceof[, eof[[1]]$left, by = .(season)] %>%
+  sep_ReIm() %>%
+  .[, hgt := scale(hgt), by = .(cEOF, part)]
+
+sam_r2 <- sams %>%
+  .[time, on = .NATURAL, allow.cartesian = TRUE] %>%
+  .[, correlate(hgt, estimate), by = .(lev, term, part, cEOF)]
+
+max_r2 <- sam_r2[part == "Imaginary" & term == "asym"] %>% 
+  .[, .SD[which.max(estimate^2)], by = .(cEOF)]
+
+sam_r2 %>%
+  .[, p.value_a := p.adjust(p.value, "fdr")] %>% 
+  ggplot(aes(lev, estimate^2)) +
+  geom_vline(xintercept = c(50, 200),  size = 0.2, color = "gray50") +
+  # geom_point(data = ~.x[p.adjust(p.value, "fdr") < 0.01], aes(color = term)) +
+  geom_line(data = ~copy(.x)[p.value_a > 0.01, estimate := NA], 
+            aes(colour = stage(term, after_scale = scales::alpha(colour, 0.5))), 
+            size = 2) +
+  geom_line(aes(color = term, group = term)) +
+
+  scale_x_level() +
+  scale_y_continuous(r2, limits = c(0, 1)) +
+  scale_size_manual("p-value", values = c("TRUE" = 1, 
+                                      "FALSE" = 0.5), 
+                    guide = "none") +
+  scale_color_brewer(NULL, palette = "Dark2", labels = c(full = "SAM",
+                                                         asym = "A-SAM",
+                                                         sym = "S-SAM")) +
+  facet_grid(part ~ cEOF) +
+  coord_flip() +
+  tag_facets("rc") +
+  panel_background
+```
+
+To explore the relationship between SAM and the modes described from the cEOF, we compute.
+the coefficient of determination between the cEOFs timeseries and the three SAM indices (SAM, A-SAM and S-SAM) at each level (Fig.\ \@ref(fig:sam-eof-vertical)).
+The SAM index is moderately correlated with the Real cEOF1 in all levels, and with the Imaginary cEOF1 and Imaginary cEOF2 in the troposphere.
+It is not correlated with the Real cEOF2 at any level.
+
+The relationship between the tropospheric SAM and cEOF1 is explained entirely by the zonally symmetric component of the SAM as shown by the low and statistically non-significant correlations between the A-SAM and either the Real or Imaginary cEOF1 and the high correlation with the S-SAM below 100 hPa.
+In the stratosphere, the Real cEOF1 is correlated with both A-SAM and S-SAM, while the Imaginary cEOF1 is highly correlated only with the A-SAM.
+These correlations are consistent with the regression patterns of geopotential height in Figure\ \@ref(fig:eof1-regr-gh).
+
+```{r eof-filtered, fig.cap = "Spatial pattern of the leading EOF of 200 hPa geopotential height with the variability of cEOF2 filtered out South of 20º.
+Arbitrary units.", fig.width=2.5, fig.height=2.5}
+eof_filtered <- era5 %>% 
+  .[sep_ReIm(cEOF2, format = "wide")[, hgt := NULL], on = .NATURAL] %>% 
+  .[lev == 200] %>% 
+  .[lat <= -20] %>% 
+  .[season(time) == "SON"] %>% 
+  .[, hgt := resid(.lm.fit(cbind(1, Real, Imaginary), hgt)), .(lon, lat, lev, season(time))] %>%
+  .[, eof := Anomaly(hgt)*sqrt(cos(lat*pi/180)), by = .(lon, lat, lev, season(time))] %>% 
+  EOF(eof ~ lon + lat | time, n = 1, data = .)
+
+eof_filtered_cor <- eof_filtered$right[sams, on = "time"] %>% 
+  .[lev == 200] %>% 
+  .[, correlate(estimate, eof), by = term]
+
+# eof_filtered %>% 
+#   .$left %>% 
+#   periodic(lon = c(0, 360)) %>% 
+#   ggplot(aes(lon, lat)) +
+#   geom_contour_fill(aes(z = eof), breaks = AnchorBreaks(0, exclude = 0)) +
+#   geom_contour_tanaka(aes(z = eof), breaks = AnchorBreaks(0, exclude = 0)) +
+#   scale_fill_divergent(guide = "none") +
+#   geom_qmap(~.x[lat <= -20]) +  
+#   scale_x_longitude() +
+#   scale_y_latitude() +
+#   coord_polar() +
+#   theme(axis.text = element_blank())
+```
+
+In the case of cEOF2, the moderate correlation between SAM and Imaginary cEOF2 becomes extremely high when only the zonally asymmetric variability of the SAM is considered.
+The Imaginary cEOF2 and the A-SAM share up to `r scales::percent(max_r2[cEOF == "cEOF2"]$estimate^2, 2)` of their variance, reached at `r max_r2[cEOF == "cEOF2"]$lev` hPa (Figure\ \@ref(fig:sam-eof-vertical).b2).
+Such extremely high correlation between the asymmetric SAM and the Imaginary cEOF2 suggests that these are not different features that share a causal link, but that they might be different ways of characterising the same phenomenon.
+
+<!-- As a further illustration \textcolor{red}{(¿De qué?)}, Figure\ \@ref(fig:eof-filtered) shows the spatial pattern of the leading EOF of 200 hPa geopotential height when the variability associated with cEOF2 is removed.
+The resulting pattern is an nearly zonally symmetrical annular mode.
+This mode is highly correlated with SAM (`r eof_filtered_cor[term == "full"]$text`), but only with the symmetrical part (`r eof_filtered_cor[term == "sym"]$text`).
+Note that the usual definition of the PSA modes as the second and third EOFs creates modes orthogonal to the SAM (defined as the first EOF) and thus impedes this kind of filtering.
+-->
+
+(ref:fake-eof-cap) Row a: The three components used to create a synthetic dataset of 41 years of geopotential height at 200 hPa.
+Row b: The three EOF computed from the synthetic dataset with the variance explained by each EOF in parenthesis.
+Units are arbitrary.
+
+```{r fake_eof}
 cEOF2 <- ceof$eof[[1]]$left[cEOF == "cEOF2"] %>% 
@@
 
+pattern <- eof_filtered$right %>% 
+  .[era5[lev == 200], on = "time"] %>% 
+  .[, FitLm(hgt, eof), by = .(lon, lat)] %>% 
+  .[, estimate := mean(estimate), by = .(lat, term)] %>%
+  dcast(lon + lat ~ term, value.var = "estimate")
+
+set.seed(42)
+fake_sam <- eof_filtered$right %>% 
+  copy() %>% 
+  .[, eof := sample(eof, replace = TRUE)]
+
+cor_with_eof2 <- fake_sam[cEOF2, on = "time"] %>% 
+  sep_ReIm() %>% 
+  .[, correlate(eof, hgt), by = part]
+
+fake_sam <- fake_sam %>% 
+  setnames("eof", "value") %>% 
+  .[, pattern[, .(sam = value*eof), by = .(lon, lat)], by = time]
+
+# fake_sam %>% 
+#   .[time %in% unique(time)[1:4]] %>% 
+#   ggplot(aes(lon, lat)) +
+#   geom_contour_fill(aes(z = sam)) +
+#   facet_wrap(~factor(time)) +
+#   coord_polar()
+
+fake_eof <- era5 %>% 
+  .[sep_ReIm(cEOF2, format = "wide")[, hgt := NULL], on = .NATURAL] %>% 
+  .[lev == 200] %>% 
+  .[lat <= -20] %>% 
+  .[season(time) == "SON"] %>% 
+  .[, hgt := hgt - .lm.fit(cbind(1, Imaginary, Real), hgt)$residuals,
+    .(lon, lat, lev, season(time))] %>% 
+  .[, hgt := Anomaly(hgt), by = .(lat)] %>% 
+  fake_sam[., on = .NATURAL] %>% 
+  .[, hgt := hgt + sam] %>% 
+  .[, eof := Anomaly(hgt)*sqrt(cos(lat*pi/180)), by = .(lon, lat, season(time))] %>% 
+  EOF(eof ~ lon + lat | time, n = 1:3, data = ., suffix = "EOF")  
+```
+
+```{r fake-eof, fig.cap = "(ref:fake-eof-cap)", fig.width=3.7, fig.height=3.3}
+north_thumb <- function(sdev, n){
+  sdev <- sdev^2
+  delta <- sqrt(2 / n) * sdev
+
+  # Separation between each pair of eofs must be 
+  # greater than the sum of their errors.
+  separation <- abs(diff(sdev))
+  separation >  delta[-length(delta)] + delta[-1]
+}
+
+separated <- north_thumb(fake_eof$sdev$sd^2, 41)
+stopifnot(all(separated))
+
+var <- fake_eof$sdev %>%
+  copy() %>% 
+    .[, setNames(paste0(EOF, " (", scales::percent(r2), ")"),
+                 EOF)]
+var <- c(var, setNames(c("SAM", "Imaginary", "Real"), c("SAM", "Imaginary", "Real")))
+
+era5 %>%
+  .[sep_ReIm(cEOF2, format = "wide")[, hgt := NULL], on = .NATURAL] %>%
+  .[lev == 200] %>%
+
+  .[season(time) == "SON"] %>%
+  .[, FitLm(hgt, Real, Imaginary), .(lon, lat)] %>%
+  rm_intercept() %>%
+  .[, .(lon, lat, eof = estimate, EOF = term)] %>%
+  rbind(pattern[, .(lon, lat, eof, EOF = "SAM")]) %>%
+  rbind(fake_eof$left, use.names = TRUE) %>% 
+   .[lat <= -20] %>%
+  .[, EOF := factor(EOF, c("SAM", "Imaginary", "Real", paste0("EOF", 1:3)),
+                    ordered = TRUE)] %>% 
+  .[EOF %in% c("EOF1", "EOF3"), eof := -eof] %>% 
+  .[!(EOF %in% c("SAM", "Imaginary", "Real")), eof := eof/sd(eof)] %>% 
+  .[EOF %in% c("SAM", "Imaginary", "Real"), eof := eof/sd(eof)] %>%
+
+  # geom_contour_tanaka has problem with the perfectly symemtric SAM
+  # adding just a tiny bit of noise solves it. 
+  .[EOF == "SAM", eof := eof + rnorm(.N, sd = 1e-5)] %>%
+
+  periodic(lon = c(0, 360)) %>%
+  ggplot(aes(lon, lat)) +
+  geom_contour_fill(aes(z = eof, fill = ..level..), global.breaks = TRUE, 
+                    breaks = AnchorBreaks(0, 0.5, exclude = 0)) +
+
+  geom_contour_tanaka(aes(z = eof), global.breaks = TRUE, 
+                      breaks = AnchorBreaks(0,0.5,  exclude = 0)) +
+  scale_fill_divergent_discretised(NULL, guide = guide_colorsteps_bottom()) +
+  geom_qmap(~.x[lat <= -20]) +
+  scale_x_longitude() +
+  scale_y_latitude() +
+  coord_polar() +
+  facet_wrap(~EOF, ncol = 3, labeller = labeller(EOF = var)) +
+  axis_labs_smol +
+  theme(axis.text = element_blank()) +
+  tag_facets("rc")
+```
+
+To test the hypothesis that a PSA-like feature can appear as statistic contamination of an otherwise zonally symmetrical and uncorrelated SAM, we create a synthetic dataset of `r uniqueN(fake_sam$time)` fields by reconstructing the 200 hPa geopotential field explained by the cEOF2 and adding a perfectly zonally symmetric SAM-like structure with random sign and value sampled with replacement from the real SAM index.
+To this synthetic dataset we then apply traditional EOF.
+By construction this dataset contains two independent components: a zonally symmetric standing oscillation and a PSA-like of varying amplitude and phase (Figure\ \@ref(fig:fake-eof) row a).
+The former one is a standing oscillation which could be represented by a single EOF, while the latter, being a wave which can change phase, needs two EOFs to be represented correctly.
+Thus, the full variance of this synthetic dataset can be perfectly explained by three EOFs.
+
+The three EOFs of the synthetic dataset are shown in Figure\ \@ref(fig:fake-eof) row b.
+The synthetic SAM can be seen in the first EOF while the cEOF2 is explained by the second and third.
+However, even though the synthetic SAM is perfectly zonally symmetric by construction, the leading EOF is not.
+The EOF decomposition mixes this mode with parts of the the zonally asymmetric cEOF2 pattern, resulting in a pattern eerily similar to the real SAM pattern.
+This happens because, although in this synthetic dataset the zonally symmetrical SAM and the cEOF2 are uncorrelated by construction, the sample correlation is not null.
+In this particular example, it is `r cor_with_eof2[part == "Real"]$text` with the Real part and `r cor_with_eof2[part == "Imaginary"]$text` with the Imaginary part.
+Of note, the three EOFs are well separated according to North's rule of thumb [@north1982].
+
+This suggests that the zonal asymmetries of the SAM could very well be mostly a statistical artefact.
+Since many surface impacts are mediated by the asymmetric component, as well as the relationship between SAM and ENSO [@campitelli2021], this potential issue could affect the interpretability of many results involving the SAM.
+Studies on the relationship between the SAM and PSA-like patterns would be particularly difficult to interpret.
+
+### PSA
+
+(ref:psa-eof2-cap) Relationship between the Real and Imaginary parts of cEOF2 and the PSA1 and PSA2 modes computed as the second and third EOFs of seasonal geopotential height anomalies [@mo2001].
+For the period 1979 -- 2019.
+
+```{r psa-eof2, fig.cap = "(ref:psa-eof2-cap)", fig.width=7, fig.height=3, fig.env="figure*"}
 psa2 <- psa$left[PC == "PSA2"][season(time) == "SON"]
@@
   tag_facets("rc")
-
 ```
@@
 
-### SAM
-
-(ref:sam-eof-vertical-cap) Coefficient of determination between the real and imaginary part of each cEOF and the SAM, Asymmetric SAM (A-SAM) and Symmetric SAM (S-SAM) indices computed at each level according to @campitelli2021.
-Thick lines represent estimates with p-value < 0.01 corrected for False Detection Rate [@benjamini1995].
-
-```{r sam-eof-vertical, fig.cap = "(ref:sam-eof-vertical-cap)", fig.height=4, fig.width=width_column}
-time <- ceof[, eof[[1]]$left, by = .(season)] %>%
-  sep_ReIm() %>%
-  .[, hgt := scale(hgt), by = .(cEOF, part)]
-
-sam_r2 <- sams %>%
-  .[time, on = .NATURAL, allow.cartesian = TRUE] %>%
-  .[, correlate(hgt, estimate), by = .(lev, term, part, cEOF)]
-
-max_r2 <- sam_r2[part == "Imaginary" & term == "asym"] %>% 
-  .[, .SD[which.max(estimate^2)], by = .(cEOF)]
-
-sam_r2 %>%
-  .[, p.value_a := p.adjust(p.value, "fdr")] %>% 
-  ggplot(aes(lev, estimate^2)) +
-  geom_vline(xintercept = c(50, 200),  size = 0.2, color = "gray50") +
-  # geom_point(data = ~.x[p.adjust(p.value, "fdr") < 0.01], aes(color = term)) +
-  geom_line(data = ~copy(.x)[p.value_a > 0.01, estimate := NA], 
-            aes(colour = stage(term, after_scale = scales::alpha(colour, 0.5))), 
-            size = 2) +
-  geom_line(aes(color = term, group = term)) +
-
-  scale_x_level() +
-  scale_y_continuous(r2, limits = c(0, 1)) +
-  scale_size_manual("p-value", values = c("TRUE" = 1, 
-                                      "FALSE" = 0.5), 
-                    guide = "none") +
-  scale_color_brewer(NULL, palette = "Dark2", labels = c(full = "SAM",
-                                                         asym = "A-SAM",
-                                                         sym = "S-SAM")) +
-  facet_grid(part ~ cEOF) +
-  coord_flip() +
-  tag_facets("rc") +
-  panel_background
-```
-
-To explore the relationship between SAM and the modes described from the cEOF, we compute.
-the coefficient of determination between the cEOFs timeseries and the three SAM indices (SAM, A-SAM and S-SAM) at each level (Fig.\ \@ref(fig:sam-eof-vertical)).
-The SAM index is moderately correlated with the Real cEOF1 in all levels, and with the Imaginary cEOF1 and Imaginary cEOF2 in the troposphere.
-It is not correlated with the Real cEOF2 at any level.
-
-The relationship between the tropospheric SAM and cEOF1 is explained entirely by the zonally symmetric component of the SAM as shown by the low and statistically non-significant correlations between the A-SAM and either the Real or Imaginary cEOF1 and the high correlation with the S-SAM below 100 hPa.
-In the stratosphere, the Real cEOF1 is correlated with both A-SAM and S-SAM, while the Imaginary cEOF1 is highly correlated only with the A-SAM.
-These correlations are consistent with the regression patterns of geopotential height in Figure\ \@ref(fig:eof1-regr-gh).
-
-```{r eof-filtered, fig.cap = "Spatial pattern of the leading EOF of 200 hPa geopotential height with the variability of cEOF2 filtered out South of 20º.
-Arbitrary units.", fig.width=2.5, fig.height=2.5}
-eof_filtered <- era5 %>% 
-  .[sep_ReIm(cEOF2, format = "wide")[, hgt := NULL], on = .NATURAL] %>% 
-  .[lev == 200] %>% 
-  .[lat <= -20] %>% 
-  .[season(time) == "SON"] %>% 
-  .[, hgt := resid(.lm.fit(cbind(1, Real, Imaginary), hgt)), .(lon, lat, lev, season(time))] %>%
-  .[, eof := Anomaly(hgt)*sqrt(cos(lat*pi/180)), by = .(lon, lat, lev, season(time))] %>% 
-  EOF(eof ~ lon + lat | time, n = 1, data = .)
-
-eof_filtered_cor <- eof_filtered$right[sams, on = "time"] %>% 
-  .[lev == 200] %>% 
-  .[, correlate(estimate, eof), by = term]
-
-# eof_filtered %>% 
-#   .$left %>% 
-#   periodic(lon = c(0, 360)) %>% 
-#   ggplot(aes(lon, lat)) +
-#   geom_contour_fill(aes(z = eof), breaks = AnchorBreaks(0, exclude = 0)) +
-#   geom_contour_tanaka(aes(z = eof), breaks = AnchorBreaks(0, exclude = 0)) +
-#   scale_fill_divergent(guide = "none") +
-#   geom_qmap(~.x[lat <= -20]) +  
-#   scale_x_longitude() +
-#   scale_y_latitude() +
-#   coord_polar() +
-#   theme(axis.text = element_blank())
-```
-
-In the case of cEOF2, the moderate correlation between SAM and Imaginary cEOF2 becomes extremely high when only the zonally asymmetric variability of the SAM is considered.
-The Imaginary cEOF2 and the A-SAM share up to `r scales::percent(max_r2[cEOF == "cEOF2"]$estimate^2, 2)` of their variance, reached at `r max_r2[cEOF == "cEOF2"]$lev` hPa (Figure\ \@ref(fig:sam-eof-vertical).b2).
-Such extremely high correlation between the asymmetric SAM and the Imaginary cEOF2 suggests that these are not different features that share a causal link, but that they might be different ways of characterising the same phenomenon.
-And, in fact, suggests that the asymmetric part of the SAM might be statistical contamination from one phase of the PSA.
-
-<!-- As a further illustration \textcolor{red}{(¿De qué?)}, Figure\ \@ref(fig:eof-filtered) shows the spatial pattern of the leading EOF of 200 hPa geopotential height when the variability associated with cEOF2 is removed.
-The resulting pattern is an nearly zonally symmetrical annular mode.
-This mode is highly correlated with SAM (`r eof_filtered_cor[term == "full"]$text`), but only with the symmetrical part (`r eof_filtered_cor[term == "sym"]$text`).
-Note that the usual definition of the PSA modes as the second and third EOFs creates modes orthogonal to the SAM (defined as the first EOF) and thus impedes this kind of filtering.
--->
-
-(ref:fake-eof-cap) Row a: The three components used to create a synthetic dataset of 41 years of geopotential height at 200 hPa.
-Row b: The three EOF computed from the synthetic dataset with the variance explained by each EOF in parenthesis.
-Units are arbitrary.
-
-```{r fake_eof}
-pattern <- eof_filtered$right %>% 
-  .[era5[lev == 200], on = "time"] %>% 
-  .[, FitLm(hgt, eof), by = .(lon, lat)] %>% 
-  .[, estimate := mean(estimate), by = .(lat, term)] %>%
-  dcast(lon + lat ~ term, value.var = "estimate")
-
-set.seed(42)
-fake_sam <- eof_filtered$right %>% 
-  copy() %>% 
-  .[, eof := sample(eof, replace = TRUE)]  %>%
-  setnames("eof", "value") %>% 
-  .[, pattern[, .(sam = value*eof), by = .(lon, lat)], by = time]
-
-# fake_sam %>% 
-#   .[time %in% unique(time)[1:4]] %>% 
-#   ggplot(aes(lon, lat)) +
-#   geom_contour_fill(aes(z = sam)) +
-#   facet_wrap(~factor(time)) +
-#   coord_polar()
-
-fake_eof <- era5 %>% 
-  .[sep_ReIm(cEOF2, format = "wide")[, hgt := NULL], on = .NATURAL] %>% 
-  .[lev == 200] %>% 
-  .[lat <= -20] %>% 
-  .[season(time) == "SON"] %>% 
-  .[, hgt := hgt - .lm.fit(cbind(1, Imaginary, Real), hgt)$residuals,
-    .(lon, lat, lev, season(time))] %>% 
-  .[, hgt := Anomaly(hgt), by = .(lat)] %>% 
-  fake_sam[., on = .NATURAL] %>% 
-  .[, hgt := hgt + sam] %>% 
-  .[, eof := Anomaly(hgt)*sqrt(cos(lat*pi/180)), by = .(lon, lat, season(time))] %>% 
-  EOF(eof ~ lon + lat | time, n = 1:3, data = ., suffix = "EOF")  
-```
-
-```{r fake-eof, fig.cap = "(ref:fake-eof-cap)", fig.width=3.7, fig.height=3.3}
-north_thumb <- function(sdev, n){
-  sdev <- sdev^2
-  delta <- sqrt(2 / n) * sdev
-
-  # Separation between each pair of eofs must be 
-  # greater than the sum of their errors.
-  separation <- abs(diff(sdev))
-  separation >  delta[-length(delta)] + delta[-1]
-}
-
-separated <- north_thumb(fake_eof$sdev$sd^2, 41)
-stopifnot(all(separated))
-
-var <- fake_eof$sdev %>%
-  copy() %>% 
-    .[, setNames(paste0(EOF, " (", scales::percent(r2), ")"),
-                 EOF)]
-var <- c(var, setNames(c("SAM", "Imaginary", "Real"), c("SAM", "Imaginary", "Real")))
-
-era5 %>%
-  .[sep_ReIm(cEOF2, format = "wide")[, hgt := NULL], on = .NATURAL] %>%
-  .[lev == 200] %>%
-
-  .[season(time) == "SON"] %>%
-  .[, FitLm(hgt, Real, Imaginary), .(lon, lat)] %>%
-  rm_intercept() %>%
-  .[, .(lon, lat, eof = estimate, EOF = term)] %>%
-  rbind(pattern[, .(lon, lat, eof, EOF = "SAM")]) %>%
-  rbind(fake_eof$left, use.names = TRUE) %>% 
-   .[lat <= -20] %>%
-  .[, EOF := factor(EOF, c("SAM", "Imaginary", "Real", paste0("EOF", 1:3)),
-                    ordered = TRUE)] %>% 
-  .[EOF %in% c("EOF1", "EOF3"), eof := -eof] %>% 
-  .[!(EOF %in% c("SAM", "Imaginary", "Real")), eof := eof/sd(eof)] %>% 
-  .[EOF %in% c("SAM", "Imaginary", "Real"), eof := eof/sd(eof)] %>%
-
-  # geom_contour_tanaka has problem with the perfectly symemtric SAM
-  # adding just a tiny bit of noise solves it. 
-  .[EOF == "SAM", eof := eof + rnorm(.N, sd = 1e-5)] %>%
-
-  periodic(lon = c(0, 360)) %>%
-  ggplot(aes(lon, lat)) +
-  geom_contour_fill(aes(z = eof, fill = ..level..), global.breaks = TRUE, 
-                    breaks = AnchorBreaks(0, 0.5, exclude = 0)) +
-
-  geom_contour_tanaka(aes(z = eof), global.breaks = TRUE, 
-                      breaks = AnchorBreaks(0,0.5,  exclude = 0)) +
-  scale_fill_divergent_discretised(NULL, guide = guide_colorsteps_bottom()) +
-  geom_qmap(~.x[lat <= -20]) +
-  scale_x_longitude() +
-  scale_y_latitude() +
-  coord_polar() +
-  facet_wrap(~EOF, ncol = 3, labeller = labeller(EOF = var)) +
-  axis_labs_smol +
-  theme(axis.text = element_blank()) +
-  tag_facets("rc")
-```
-
-To test the hypothesis that a PSA-like feature can appear as statistic contamination of an otherwise zonally symmetrical and uncorrelated SAM, we create a synthetic dataset of `r uniqueN(fake_sam$time)` fields by reconstructing the 200 hPa geopotential field explained by the cEOF2 and adding a perfectly zonally symmetric SAM-like structure with random sign and value.
-To this synthetic dataset we then apply traditional EOF.
-By construction this dataset contains two independent components: a zonally symmetric standing oscillation and a PSA-like of varying amplitude and phase (Figure\ \@ref(fig:fake-eof) row a).
-The former one is ar standing oscillation which could be represented by a single EOF, while the latter, being a wave which can change phase, needs two EOFs to be represented correctly.
-Thus, the full variance of this synthetic dataset can be perfectly explained by three EOFs.
-
-The three EOFs of the synthetic dataset are shown in Figure\ \@ref(fig:fake-eof) row b.
-The synthetic SAM can be seen in the first EOF while the PSA is explained by the second and third.
-However, even though the synthetic SAM is perfectly zonally symmetric by construction, the leading EOF is not.
-The EOF decomposition mixes this mode with parts of the the zonally asymmetric cEOF2 pattern, resulting in a pattern eerily similar to the real SAM pattern.
-Of note, the three EOFs are well separated according to North's rule of thumb [@north1982].
-
-This suggests that the zonal asymmetries of the SAM could very well be mostly a statistical artifact.
-Since many surface impacts are mediated by the asymmetric component, as well as the relationship between SAM and ENSO [@campitelli2021], this potential issue could affect the interpretability of many results involving the SAM.
-Studies on the relationship between the SAM and the PSA pattern would be particularly difficult to interpret.
-
 ## Tropical sources

### 2021-04-09 analysis/paper/shceof.Rmd
Agrega líneas de cero
Closes #38

@@
   ggplot(aes(value, hgt)) +
+  geom_hline(yintercept = 0, size = 0.2, color = "gray50") +
+  geom_vline(xintercept = 0, size = 0.2, color = "gray50") +
   geom_point() +

### 2021-04-09 analysis/paper/shceof.tex
Renderiza todo

@@
         Elio Campitelli \at
-     Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos. Buenos Aires, Argentina.
-CONICET -- Universidad de Buenos Aires. Centro de Investigaciones del Mar y la Atmósfera (CIMA). Buenos Aires, Argentina.
+     Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos. Buenos Aires, Argentina.\\
+CONICET -- Universidad de Buenos Aires. Centro de Investigaciones del Mar y la Atmósfera (CIMA). Buenos Aires, Argentina.\\
 CNRS -- IRD -- CONICET -- UBA. Instituto Franco-Argentino para el Estudio del Clima y sus Impactos (IRL 3351 IFAECI). Buenos Aires, Argentina. \\
@@
 We derived streamfunction at 200 hPa from ERA5 vorticity using the FORTRAN subroutine FISHPACK \citep{fishpack} and computed horizontal wave activity fluxes following \citet{plumb1985}.
+Sea Surface Temperature (SST) fields were taken from Extended Reconstructed Sea Surface Temperature (ERSST) v5 \citep{huang2017}.
 For precipitation data we used monthly data from the CPC Merged Analysis of Precipitation \citep{cmap}, with a 2.5\degree~resolution in latitude and longitude.
@@
 \midrule
-cEOF1 & \cellcolor[HTML]{E1C0BB}{\textcolor{black}{0.28}} & \cellcolor[HTML]{FDFAFA}{\textcolor{black}{0.02}} & \cellcolor[HTML]{FDFAFA}{\textcolor{black}{0.02}}\\
+cEOF1 & \cellcolor[HTML]{E1C0BB}{\textcolor{black}{0.28}} & \cellcolor[HTML]{FDFCFC}{\textcolor{black}{0.01}} & \cellcolor[HTML]{FDFAFA}{\textcolor{black}{0.02}}\\
 cEOF2 & \cellcolor[HTML]{FFFFFF}{\textcolor{black}{0.00}} & \cellcolor[HTML]{BB7A73}{\textcolor{white}{0.60}} & \cellcolor[HTML]{FDFAFA}{\textcolor{black}{0.02}}\\
-cEOF3 & \cellcolor[HTML]{FFFFFF}{\textcolor{black}{0.00}} & \cellcolor[HTML]{FFFFFF}{\textcolor{black}{0.00}} & \cellcolor[HTML]{FDFCFC}{\textcolor{black}{0.01}}\\
+cEOF3 & \cellcolor[HTML]{FFFFFF}{\textcolor{black}{0.00}} & \cellcolor[HTML]{FFFFFF}{\textcolor{black}{0.00}} & \cellcolor[HTML]{FDFAFA}{\textcolor{black}{0.02}}\\
 \bottomrule
@@
 \caption{\label{fig:ceofs-1}Spatial patterns of the leading 2 cEOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2019.
-Real part in shading, imaginary part in contours.}
+Real part in shading, imaginary part in contours.
+Arbitrary units.}
 \end{figure}
@@
 Indeed, Figure~\ref{fig:wave1-o3} shows the relationship between amplitude and phase of the cEOF1 and the amplitude and phase of the planetary wave 1 in Total Ozone Column between 75°S and 45°S; as shown in Figure~\ref{fig:o3-regr}, this is a wide latitude band that captures most of the area with significant regression coefficients.
-The correlation between the zonal wave 1 of ozone and cEOF1 (computed as the mean cosine of the difference in phase) is 0.86 (CI: 0.76 -- 0.93).
+The correlation between the zonal wave 1 of ozone and cEOF1 (computed as the mean cosine of the difference in phase) is 0.87 (CI: 0.77 -- 0.93).
 As expected from the location of anomalies in Figure~\ref{fig:wave1-o3}, the Real cEOF1 drives the relationship with amplitude and the Imaginary cEOF1 drives the relationship with phase (not shown).
+
+\hypertarget{sam}{%
+\subsubsection{SAM}\label{sam}}
+
+\begin{figure}
+\centering
+\includegraphics{../figures/sam-eof-vertical-1.pdf}
+\caption{\label{fig:sam-eof-vertical}Coefficient of determination between the real and imaginary part of each cEOF and the SAM, Asymmetric SAM (A-SAM) and Symmetric SAM (S-SAM) indices computed at each level according to \citet{campitelli2021}.
+Thick lines represent estimates with p-value \textless{} 0.01 corrected for False Detection Rate \citep{benjamini1995}.}
+\end{figure}
+
+To explore the relationship between SAM and the modes described from the cEOF, we compute.
+the coefficient of determination between the cEOFs timeseries and the three SAM indices (SAM, A-SAM and S-SAM) at each level (Fig.~\ref{fig:sam-eof-vertical}).
+The SAM index is moderately correlated with the Real cEOF1 in all levels, and with the Imaginary cEOF1 and Imaginary cEOF2 in the troposphere.
+It is not correlated with the Real cEOF2 at any level.
+
+The relationship between the tropospheric SAM and cEOF1 is explained entirely by the zonally symmetric component of the SAM as shown by the low and statistically non-significant correlations between the A-SAM and either the Real or Imaginary cEOF1 and the high correlation with the S-SAM below 100 hPa.
+In the stratosphere, the Real cEOF1 is correlated with both A-SAM and S-SAM, while the Imaginary cEOF1 is highly correlated only with the A-SAM.
+These correlations are consistent with the regression patterns of geopotential height in Figure~\ref{fig:eof1-regr-gh}.
+
+In the case of cEOF2, the moderate correlation between SAM and Imaginary cEOF2 becomes extremely high when only the zonally asymmetric variability of the SAM is considered.
+The Imaginary cEOF2 and the A-SAM share up to 92\% of their variance, reached at 225 hPa (Figure~\ref{fig:sam-eof-vertical}.b2).
+Such extremely high correlation between the asymmetric SAM and the Imaginary cEOF2 suggests that these are not different features that share a causal link, but that they might be different ways of characterising the same phenomenon.
+
+\begin{figure}
+\centering
+\includegraphics{../figures/fake-eof-1.pdf}
+\caption{\label{fig:fake-eof}Row a: The three components used to create a synthetic dataset of 41 years of geopotential height at 200 hPa.
+Row b: The three EOF computed from the synthetic dataset with the variance explained by each EOF in parenthesis.
+Units are arbitrary.}
+\end{figure}
+
+To test the hypothesis that a PSA-like feature can appear as statistic contamination of an otherwise zonally symmetrical and uncorrelated SAM, we create a synthetic dataset of 41 fields by reconstructing the 200 hPa geopotential field explained by the cEOF2 and adding a perfectly zonally symmetric SAM-like structure with random sign and value sampled with replacement from the real SAM index.
+To this synthetic dataset we then apply traditional EOF.
+By construction this dataset contains two independent components: a zonally symmetric standing oscillation and a PSA-like of varying amplitude and phase (Figure~\ref{fig:fake-eof} row a).
+The former one is a standing oscillation which could be represented by a single EOF, while the latter, being a wave which can change phase, needs two EOFs to be represented correctly.
+Thus, the full variance of this synthetic dataset can be perfectly explained by three EOFs.
+
+The three EOFs of the synthetic dataset are shown in Figure~\ref{fig:fake-eof} row b.
+The synthetic SAM can be seen in the first EOF while the cEOF2 is explained by the second and third.
+However, even though the synthetic SAM is perfectly zonally symmetric by construction, the leading EOF is not.
+The EOF decomposition mixes this mode with parts of the the zonally asymmetric cEOF2 pattern, resulting in a pattern eerily similar to the real SAM pattern.
+This happens because, although in this synthetic dataset the zonally symmetrical SAM and the cEOF2 are uncorrelated by construction, the sample correlation is not null.
+In this particular example, it is -0.064 (CI: -0.36 -- 0.25) with the Real part and -0.13 (CI: -0.42 -- 0.19) with the Imaginary part.
+Of note, the three EOFs are well separated according to North's rule of thumb \citep{north1982}.
+
+This suggests that the zonal asymmetries of the SAM could very well be mostly a statistical artefact.
+Since many surface impacts are mediated by the asymmetric component, as well as the relationship between SAM and ENSO \citep{campitelli2021}, this potential issue could affect the interpretability of many results involving the SAM.
+Studies on the relationship between the SAM and PSA-like patterns would be particularly difficult to interpret.
 
@@
 
-\hypertarget{sam}{%
-\subsubsection{SAM}\label{sam}}
-
-\begin{figure}
-\centering
-\includegraphics{../figures/sam-eof-vertical-1.pdf}
-\caption{\label{fig:sam-eof-vertical}Coefficient of determination between the real and imaginary part of each cEOF and the SAM, Asymmetric SAM (A-SAM) and Symmetric SAM (S-SAM) indices computed at each level according to \citet{campitelli2021}.
-Thick lines represent estimates with p-value \textless{} 0.01 corrected for False Detection Rate \citep{benjamini1995}.}
-\end{figure}
-
-To explore the relationship between SAM and the modes described from the cEOF, we compute.
-the coefficient of determination between the cEOFs timeseries and the three SAM indices (SAM, A-SAM and S-SAM) at each level (Fig.~\ref{fig:sam-eof-vertical}).
-The SAM index is moderately correlated with the Real cEOF1 in all levels, and with the Imaginary cEOF1 and Imaginary cEOF2 in the troposphere.
-It is not correlated with the Real cEOF2 at any level.
-
-The relationship between the tropospheric SAM and cEOF1 is explained entirely by the zonally symmetric component of the SAM as shown by the low and statistically non-significant correlations between the A-SAM and either the Real or Imaginary cEOF1 and the high correlation with the S-SAM below 100 hPa.
-In the stratosphere, the Real cEOF1 is correlated with both A-SAM and S-SAM, while the Imaginary cEOF1 is highly correlated only with the A-SAM.
-These correlations are consistent with the regression patterns of geopotential height in Figure~\ref{fig:eof1-regr-gh}.
-
-In the case of cEOF2, the moderate correlation between SAM and Imaginary cEOF2 becomes extremely high when only the zonally asymmetric variability of the SAM is considered.
-The Imaginary cEOF2 and the A-SAM share up to 92\% of their variance, reached at 225 hPa (Figure~\ref{fig:sam-eof-vertical}.b2).
-Such extremely high correlation between the asymmetric SAM and the Imaginary cEOF2 suggests that these are not different features that share a causal link, but that they might be different ways of characterising the same phenomenon.
-And, in fact, suggests that the asymmetric part of the SAM might be statistical contamination from one phase of the PSA.
-
-\begin{figure}
-\centering
-\includegraphics{../figures/fake-eof-1.pdf}
-\caption{\label{fig:fake-eof}Row a: The three components used to create a synthetic dataset of 41 years of geopotential height at 200 hPa.
-Row b: The three EOF computed from the synthetic dataset with the variance explained by each EOF in parenthesis.
-Units are arbitrary.
-Negative values in blue and positive values in red.}
-\end{figure}
-
-To test the hypothesis that a PSA-like feature can appear as statistic contamination of an otherwise zonally symmetrical and uncorrelated SAM, we create a synthetic dataset of 41 fields by reconstructing the 200 hPa geopotential field explained by the cEOF2 and adding a perfectly zonally symmetric SAM-like structure with random sign and value.
-To this synthetic dataset we then apply traditional EOF.
-By construction this dataset contains two independent components: a zonally symmetric standing oscillation and a PSA of varying amplitude and phase (Figure~\ref{fig:fake-eof} row a).
-The former one is ar standing oscillation which could be represented by a single EOF, while the latter, being a wave which can change phase, needs two EOFs to be represented correctly.
-Thus, the full variance of this synthetic dataset can be perfectly explained by three EOFs.
-
-The three EOFs of the synthetic dataset are shown in Figure~\ref{fig:fake-eof} row b.
-The synthetic SAM can be seen in the first EOF while the PSA is explained by the second and third.
-However, even though the synthetic SAM is perfectly zonally symmetric by construction, the leading EOF is not.
-The EOF decomposition mixes this mode with parts of the the zonally asymmetric cEOF2 pattern, resulting in a pattern eerily similar to the real SAM pattern.
-Of note, the three EOFs are well separated according to North's rule of thumb \citep{north1982}.
-
-This suggests that the zonal asymmetries of the SAM could very well be mostly a statistical artifact.
-Since many surface impacts are mediated by the asymmetric component, as well as the relationship between SAM and ENSO \citep{campitelli2021}, this potential issue could affect the interpretability of many results involving the SAM.
-Studies on the relationship between the SAM and the PSA pattern would be particularly difficult to interpret.
-
 \hypertarget{tropical-sources}{%
@@
 The Imaginary cEOF2 is associated with strong positive SST anomalies on the Central Pacific and negative anomalies on an area across the North of Australia and New Zealand, the South Pacific Convergence Zone (SPCZ) (Figure~\ref{fig:sst-psi-2}.a2).
-This pattern is almost canonically positive ENSO and indeed, the correlation between the Imaginary cEOF2 and the Oceanic Niño Index \citep{bamston1997} is very high 0.76 (CI: 0.59 -- 0.86).
+This pattern is almost canonically positive ENSO and indeed, the correlation between the Imaginary cEOF2 and the Oceanic Niño Index \citep{bamston1997} is very high 0.76 (CI: 0.59 -- 0.87).
 Streamfunction anomalies show a coherent picture.
@@
 However, Figure~\ref{fig:sst-psi-2}.a1 and b1 show that the Real cEOF2 is not associated either with any significant SST nor streamfunction anomalies in the tropics.
-The correlation between the Real cEOF2 and ENSO is also not significant (-0.0017 (CI: -0.31 -- 0.31)).
+The correlation between the Real cEOF2 and ENSO is also not significant (-0.0042 (CI: -0.31 -- 0.3)).
 This lack of tropical signal suggests different natures of the different phases of the cEOF2 wave train.
@@
 Furthermore, Figure~\ref{fig:enso-phase} suggest that strong cEOF2 years tend to coincide with strong ENSO years.
-The correlation between the absolute magnitude of the ONI and the magnitude of the cEOF2 is 0.44 (CI: 0.16 -- 0.66).
+The correlation between the absolute magnitude of the ONI and the magnitude of the cEOF2 is 0.45 (CI: 0.17 -- 0.67).
 This relationship, however, appears to be driven only by the three years with strongest ENSO events in the period (2015, 1997, and 1982) which also coincide with the three years with strongest cEOF2 magnitude.
-If those years are removed, the correlation becomes non-significant (0.044 (CI: -0.28 -- 0.36)).
-Furthermore, even when using all the datapoints, the Spearman correlation --which is robust to outliers-- is also non-significant (0.2, p-value= 0.2).
+If those years are removed, the correlation becomes non-significant (0.047 (CI: -0.28 -- 0.36)).
+Furthermore, even when using all the datapoints, the Spearman correlation --which is robust to outliers-- is also non-significant (0.2, p-value= 0.21).
 Therefore, the relationship between the magnitude of the cEOF2 train wave and ONI remains uncertain.
@@
 There is a phase dependence of the precipitation anomalies in SESA (not shown), which resembles the relationship between ONI and the phase of cEOF2.
-A sinusoidal fit has an \(r^2\)~of~0.27 with p-value~= 0.002.
+A sinusoidal fit has an \(r^2\)~of~0.27 with p-value~= 0.0022.
 

### 2021-04-09 analysis/paper/shceof.Rmd
Reemplaza gráfico PSA vs EOF2 por tabla de correlación
#34

--- 2021-04-09 analysis/paper/shceof.Rmd
@@
 
-(ref:psa-eof2-cap) Relationship between the Real and Imaginary parts of cEOF2 and the PSA1 and PSA2 modes computed as the second and third EOFs of seasonal geopotential height anomalies [@mo2001].
+(ref:psa-eof2-cap) Correlation between the Real and Imaginary parts of cEOF2 and the PSA1 and PSA2 modes computed as the second and third EOFs of seasonal geopotential height anomalies [@mo2001].
 For the period 1979 -- 2019.
@@
 
+make_numeric <- function(x) {
+  x <- strsplit(x, "\ ")
+  vapply(x, function(x) as.numeric(x[[1]]), numeric(1))
+}
+
+background_palette <- grDevices::colorRampPalette(c(scales::muted("blue"),
+                                                    "white",
+                                                    scales::muted("red")),
+                                                  space = "Lab")
+
 cEOF2[psa_time, on = "time"] %>% 
@@
   copy() %>% 
-  # .[, hgt := rotate(hgt, psa_angle)] %>%
   sep_ReIm() %>% 
-  ggplot(aes(value, hgt)) +
-  geom_point() +
-  geom_smooth(method = "lm") +
-  geom_smooth(method = "lm") + 
-  scale_y_continuous("EOF2") +
-  scale_x_continuous("PSA") +
-  facet_grid(part ~ PC) +
-  panel_background +
-  tag_facets("rc")
+  .[, correlate(value, hgt), by = .(PC, part)] %>% 
+  dcast(PC ~ part, value.var = "text") %>% 
+  kbl2(booktabs = TRUE, caption = "(ref:psa-eof2-cap)") %>%
+  column_spec2(2:3, background = spec_color2(make_numeric(.table[[.col]]),
+                                             option = background_palette,
+                                             scale_from = c(-1, 1)), 
+               color = ifelse(abs(make_numeric(.table[[.col]])) > .5, "white", "black")) %>%
+  add_header_above(c("", "cEOF2" = 2)) %>%
+  kable_classic()
+
+# cEOF2[psa_time, on = "time"] %>% 
+#   .[PC != "SAM"] %>% 
+#   copy() %>% 
+#   # .[, hgt := rotate(hgt, psa_angle)] %>%
+#   sep_ReIm() %>% 
+#   ggplot(aes(value, hgt)) +
+#   geom_hline(yintercept = 0, size = 0.2, color = "gray50") +
+#   geom_vline(xintercept = 0, size = 0.2, color = "gray50") +
+#   geom_point() +
+#   geom_smooth(method = "lm") +
+#   geom_smooth(method = "lm") + 
+#   scale_y_continuous("EOF2") +
+#   scale_x_continuous("PSA") +
+#   facet_grid(part ~ PC) +
+#   panel_background +
+#   tag_facets("rc")
 ```
@@
 The relationship between PSAs and the modes described from the cEOFs is also studied.
-Figure\ \@ref(fig:psa-eof2) shows the relationship between the two PSA indices and the Real and Imaginary phase of cEOF2.
-As anticipated by Figure\ \@ref(fig:eof2-regr-gh), there is a strong correlation between PSA1 and Imaginary cEOF2 (Figure\ \@ref(fig:psa-eof2)b.1), and between PSA2 and Real cEOF2 (Figure\ \@ref(fig:psa-eof2)a.2).
-Conversely, there is no relationship between PSA1 and Real cEOF2, and between PSA2 and Imaginary cEOF2 (Figure\ \@ref(fig:psa-eof2) panels a.1 and b.2).
+Figure\ \@ref(tab:psa-eof2) shows the relationship between the two PSA indices and the Real and Imaginary phase of cEOF2.
+As anticipated by Figure\ \@ref(fig:eof2-regr-gh), there is a strong correlation between PSA1 and Imaginary cEOF2, and between PSA2 and Real cEOF2.
+Conversely, there is no relationship between PSA1 and Real cEOF2, and between PSA2 and Imaginary cEOF2.
 So not only this cEOF2 represents well both the spatial structure and temporal evolution of the PSA modes, but it's also possible to make a rather clean association between its two phases and the two PSA modes.

### 2021-04-09 analysis/paper/shceof.tex
Renderiza todo

@@
 
-\begin{figure*}
-\includegraphics{../figures/psa-eof2-1} \caption{Relationship between the Real and Imaginary parts of cEOF2 and the PSA1 and PSA2 modes computed as the second and third EOFs of seasonal geopotential height anomalies \citep{mo2001}.
-For the period 1979 -- 2019.}\label{fig:psa-eof2}
-\end{figure*}
+\begin{table}
+
+\caption{\label{tab:psa-eof2}Correlation between the Real and Imaginary parts of cEOF2 and the PSA1 and PSA2 modes computed as the second and third EOFs of seasonal geopotential height anomalies \citep{mo2001}.
+For the period 1979 -- 2019.}
+\centering
+\begin{tabular}[t]{l>{}l>{}l}
+\toprule
+\multicolumn{1}{c}{} & \multicolumn{2}{c}{cEOF2} \\
+\cmidrule(l{3pt}r{3pt}){2-3}
+PC & Real & Imaginary\\
+\midrule
+PSA1 & \cellcolor[HTML]{C3BBDE}{\textcolor{black}{-0.32 (CI: -0.57 -- -0.01)}} & \cellcolor[HTML]{6058A9}{\textcolor{white}{-0.83 (CI: -0.9 -- -0.69)}}\\
+PSA2 & \cellcolor[HTML]{645BAB}{\textcolor{white}{-0.81 (CI: -0.9 -- -0.67)}} & \cellcolor[HTML]{F9F1F0}{\textcolor{black}{0.059 (CI: -0.25 -- 0.36)}}\\
+\bottomrule
+\end{tabular}
+\end{table}
 
 The relationship between PSAs and the modes described from the cEOFs is also studied.
-Figure~\ref{fig:psa-eof2} shows the relationship between the two PSA indices and the Real and Imaginary phase of cEOF2.
-As anticipated by Figure~\ref{fig:eof2-regr-gh}, there is a strong correlation between PSA1 and Imaginary cEOF2 (Figure~\ref{fig:psa-eof2}b.1), and between PSA2 and Real cEOF2 (Figure~\ref{fig:psa-eof2}a.2).
-Conversely, there is no relationship between PSA1 and Real cEOF2, and between PSA2 and Imaginary cEOF2 (Figure~\ref{fig:psa-eof2} panels a.1 and b.2).
+Figure~\ref{tab:psa-eof2} shows the relationship between the two PSA indices and the Real and Imaginary phase of cEOF2.
+As anticipated by Figure~\ref{fig:eof2-regr-gh}, there is a strong correlation between PSA1 and Imaginary cEOF2, and between PSA2 and Real cEOF2.
+Conversely, there is no relationship between PSA1 and Real cEOF2, and between PSA2 and Imaginary cEOF2.
 So not only this cEOF2 represents well both the spatial structure and temporal evolution of the PSA modes, but it's also possible to make a rather clean association between its two phases and the two PSA modes.

### 2021-04-09 R/correlate.R

@@ -5,9 +5,11 @@
 #'
 #' @param x,y numeric variables to correlate.
 #' @param signif number of significant figures to keep in the reporting text.
+#' @param sep separator to use between the estimate and the confidence interval.
+#' The default uses a protected space.
 #'
 #' @export
-correlate <- function(x, y, signif = 2) {
+correlate <- function(x, y, signif = 2, sep = "\ ") {
   correlation <- cor.test(x, y)
   out <- with(correlation,
               list(estimate = estimate,
@@ -15,8 +17,9 @@ correlate <- function(x, y, signif = 2) {
                    low = conf.int[1],
                    hig = conf.int[2]))
 
+
   out$text <- with(out,
-                   paste0(signif(estimate, signif), " (CI:\ ",
+                   paste0(signif(estimate, signif), sep, "(CI:\ ",
                           signif(low, signif), "\ --\ ", signif(hig, signif), ")"))
 
   out

### 2021-04-13 analysis/paper/shceof.Rmd
Agrega círculos de latitud
Closes #29

@@
 no_grid <- theme(panel.grid = element_blank())
+grid_y <- theme(panel.grid.major.y = element_line(color = "gray60", size = 0.1))
 
@@
   scale_fill_divergent_discretised(NULL, guide = guide_colorsteps_bottom(10))  +
-  scale_y_latitude() +
+  scale_y_latitude(labels = NULL,
+                   minor_breaks = NULL) +
   scale_x_longitude() +
@@
   tagger::tag_facets("rc")  +
-  theme(panel.spacing = grid::unit(-.7, "lines"))
+  theme(panel.spacing = grid::unit(-.7, "lines")) +
+  grid_y
 ```
@@
   scale_x_longitude() +
-  scale_y_latitude() +
+    scale_y_latitude(labels = NULL,
+                   minor_breaks = NULL) +
   scale_linetype_manual(values = c("1" = 1, "-1" = 2), labels = c("1" = "+", "-1" = "-"),
@@
   axis_labs_smol +
-  tag_facets(tag = "cr")
-
+  tag_facets(tag = "cr") +
+  grid_y
 ```
@@
     scale_x_longitude() +
-    scale_y_latitude() +
+      scale_y_latitude(labels = NULL,
+                   minor_breaks = NULL) +
+    grid_y +
     facet +
@@
   geom_contour_pval(aes(z = p.value)) +
-  geom_hline(yintercept = o3wave_lats, size = 0.3, alpha = 0.5) +
+  geom_hline(yintercept = o3wave_lats, size = 0.3, alpha = 0.8) +
   geom_qmap(~.x[lat <= -20]) +
@@
   scale_x_longitude() +
-  scale_y_latitude() +
+    scale_y_latitude(labels = NULL,
+                   minor_breaks = NULL) +
+  grid_y +
   scale_linetype(guide = "none") +
@@
 Arbitrary units.", fig.width=2.5, fig.height=2.5}
+cEOF2 <- ceof$eof[[1]]$left[cEOF == "cEOF2"] %>% 
+  copy() %>% 
+  .[, cEOF := NULL] %>% 
+  .[]
+
 eof_filtered <- era5 %>% 
@@
   scale_x_longitude() +
-  scale_y_latitude() +
+    scale_y_latitude(labels = NULL,
+                   minor_breaks = NULL) +
+  grid_y +
   coord_polar() +

### 2021-04-13 analysis/paper/shceof.Rmd
Hace las barras de colores simétricas
Closes #33

@@
     ggplot(aes(lon, lat)) +
-    geom_contour_fill(aes(z = estimate), breaks = AnchorBreaks(exclude = 0), global.breaks = FALSE) +
-    geom_contour_tanaka(aes(z = estimate), breaks = AnchorBreaks(exclude = 0), global.breaks = FALSE) +
+    geom_contour_fill(aes(z = estimate), breaks = AnchorBreaks(exclude = 0),
+                      global.breaks = FALSE) +
+    geom_contour_tanaka(aes(z = estimate), 
+                        breaks = AnchorBreaks(exclude = 0), global.breaks = FALSE) +
     scale_fill_divergent(var_title, breaks = AnchorBreaks(0, exclude = 0),
@@
 plot_regr(regr , "cEOF1", c(200, 50), "hgt") +
-  # geom_contour2(data = ~copy(.x)[, estimate := Anomaly(estimate), by = .(term, lev, lat)],
-  #              aes(z = estimate, linetype = factor(-sign(..level..))), 
-  #              global.breaks = FALSE, breaks = AnchorBreaks(0, bins = 2, exclude = 0)) +
-  # scale_linetype(guide = "none") +
-  NULL
-
+  scale_fill_divergent("Geopotential height", breaks = AnchorBreaks(0, exclude = 0),
+                       guide = guide_colorstrip_bottom(15), 
+                       limits = c(-1800, 1800))
 ```
@@
 ```{r eof2-regr-gh, fig.cap = "(ref:eof2-regr-gh-cap)", fig.height=1.3*width_column, fig.width=width_column, dependson="plot_regr"}
-plot_regr(regr, "cEOF2", c(200, 50), "hgt")
-
+plot_regr(regr, "cEOF2", c(200, 50), "hgt") +
+    scale_fill_divergent("Geopotential height", 
+                         breaks = AnchorBreaks(0, 200, exclude = 0),
+                       guide = guide_colorstrip_bottom(15), 
+                       limits = c(-1000+1, 1000-1))
 ```
@@
 ```{r eof1-regr-t, fig.cap = "(ref:eof1-regr-t-cap)",  dependson="plot_regr", fig.height=1.3*width_column, fig.width=width_column}
-plot_regr(regr, "cEOF1", c(200, 50), "t")
+plot_regr(regr, "cEOF1", c(200, 50), "t") +
+    scale_fill_divergent("Temperature", breaks = AnchorBreaks(0, exclude = 0),
+                       guide = guide_colorstrip_bottom(15), 
+                       limits = c(-4.5, 4.5))
 ```
@@
                     aes(z = estimate, fill = ..level..),
-                    breaks = AnchorBreaks(0, binwidth = 0.25, exclude = 0)) +
+                    breaks = AnchorBreaks(0, binwidth = 0.3, exclude = 0)) +
   geom_contour_tanaka(data = ~.x[variable == "t"],
                       aes(z = estimate),
-                      breaks = AnchorBreaks(0, binwidth = 0.25, exclude = 0)) +
+                      breaks = AnchorBreaks(0, binwidth = 0.3, exclude = 0)) +
   geom_contour2(data = ~.x[variable == "o3"], 
@@
 
-  scale_fill_divergent_discretised("Air temperature", guide = guide_colorsteps_bottom(15)) + 
+  scale_fill_divergent_discretised("Air temperature", 
+                                   # limits = c(-2.25, 2.25),
+                                   guide = guide_colorsteps_bottom(15)) + 
   scale_y_level() +
@@
   scale_fill_divergent_discretised("Total Ozone Column", 
+                                   limits = c(-30, 30),  
                                    guide = guide_colorsteps_bottom(15)) + 
@@
                       breaks = AnchorBreaks(0,0.5,  exclude = 0)) +
-  scale_fill_divergent_discretised(NULL, guide = guide_colorsteps_bottom()) +
+  scale_fill_divergent_discretised(NULL, guide = guide_colorsteps_bottom(), 
+                                   limits = c(-4, 4)) +
   geom_qmap(~.x[lat <= -20]) +
@@
 ```{r sst-psi-2,  fig.cap = "(ref:sst-psi-2-cap)", fig.width = 7, fig.height=4, dependson="plot_sst_psi", fig.env="figure*"}
-plot_sst_psi(which_cEOF = "cEOF2", which_lev = 200)
-
+plot_sst_psi(which_cEOF = "cEOF2", which_lev = 200) +
+  scale_fill_divergent_discretised("SST",
+                                   limits = c(-1, 1),
+                                   guide = guide_colorsteps_bottom(20, order = 2)) 
 ```
@@
 ```{r sst-psi-1, fig.cap = "(ref:sst-psi-1-cap)", fig.width = 7, fig.height=4, dependson="plot_sst_psi", fig.env = "figure*"}
-plot_sst_psi(which_cEOF = "cEOF1", which_lev = 200)
-
+plot_sst_psi(which_cEOF = "cEOF1", which_lev = 200) +
+  scale_fill_divergent_discretised("SST",
+                                   limits = c(-.3, .3),
+                                   guide = guide_colorsteps_bottom(20, order = 2)) 
 ```
@@
 ```{r pp-america, fig.cap = "(ref:pp-america-cap)", fig.width=width_column, fig.height=width_column*1.1}
-plot_pp(data = regr_pp, which_cEOF = c("cEOF1", "cEOF2"), which_continent = "america") 
+plot_pp(data = regr_pp, which_cEOF = c("cEOF1", "cEOF2"), which_continent = "america")  +
+  scale_fill_divergent_discretised(NULL, 
+                         limits = c(-1, 1),
+                         # oob = scales::squish,
+                         low = "#8E521C",
+                         high = "#00665E",
+                         guide = guide_colorsteps_bottom(15)) 
   # annotate("rect", xmin = min(sesa$lon), xmax = max(sesa$lon), 

### 2021-04-13 analysis/paper/shceof.Rmd
Quita figura de ZW1 de O3 vs cEOF1
#34

@@
 
-```{r wave1-o3, fig.cap = "(ref:wave1-o3-cap)", fig.width=width_column, fig.height=width_column*.7}
+```{r wave1-o3, fig.cap = "(ref:wave1-o3-cap)", fig.width=width_column, fig.height=width_column*.7, include = FALSE}
 wave1_o3 <- o3 %>% 
@@
 #   panel_background
-# temp %>% 
-#   .[, correlate(amplitude, Mod(hgt))]
-# 
-# temp %>% 
-#   .[, correlate(phase, Arg(hgt))]
+wave1_cor <- rbind(temp %>%
+                     .[, correlate(amplitude, Mod(hgt))] %>% 
+                     .[, with := "amplitude"],
+
+                   temp %>%
+                     .[, correlate(phase, Arg(hgt))] %>% 
+                     .[, with := "phase"])
 
@@
 
-Indeed, Figure\ \@ref(fig:wave1-o3) shows the relationship between amplitude and phase of the cEOF1 and the amplitude and phase of the planetary wave 1 in Total Ozone Column between `r knitr::combine_words(LatLabel(o3wave_lats))`; as shown in Figure\ \@ref(fig:o3-regr), this is a wide latitude band that captures most of the area with significant regression coefficients.
-The correlation between the zonal wave 1 of ozone and cEOF1 (computed as the mean cosine of the difference in phase) is `r wave_correlation$text`.
-As expected from the location of anomalies in Figure\ \@ref(fig:wave1-o3), the Real cEOF1 drives the relationship with amplitude and the Imaginary cEOF1 drives the relationship with phase (not shown).
+There is a close relationship between amplitude and phase of the cEOF1 and the amplitude and phase of the planetary wave 1 in Total Ozone Column between `r knitr::combine_words(LatLabel(o3wave_lats))`; as shown in Figure\ \@ref(fig:o3-regr), this is a wide latitude band that captures most of the area with significant regression coefficients.
+The correlation between the amplitude of each index is `r wave1_cor[with == "amplitude"]$text` the overall correlation between the zonal wave 1 of ozone and cEOF1 (computed as the mean cosine of the difference in phase) is `r wave_correlation$text`.
+As expected from the location of anomalies in Figure\ \@ref(fig:o3-regr), the Real cEOF1 drives the relationship with amplitude and the Imaginary cEOF1 drives the relationship with phase (not shown).
 

### 2021-04-13 analysis/paper/shceof.tex
Renderiza todo

@@
 
-\begin{figure}
-\centering
-\includegraphics{../figures/wave1-o3-1.pdf}
-\caption{\label{fig:wave1-o3}Relationship between cEOF1 amplitude and phase with amplitude and phase of the zonal wave 1 in Total Ozone Column averaged between 75°S and 45°S for the period 1979 -- 2019.}
-\end{figure}
-
-Indeed, Figure~\ref{fig:wave1-o3} shows the relationship between amplitude and phase of the cEOF1 and the amplitude and phase of the planetary wave 1 in Total Ozone Column between 75°S and 45°S; as shown in Figure~\ref{fig:o3-regr}, this is a wide latitude band that captures most of the area with significant regression coefficients.
-The correlation between the zonal wave 1 of ozone and cEOF1 (computed as the mean cosine of the difference in phase) is 0.87 (CI: 0.77 -- 0.93).
-As expected from the location of anomalies in Figure~\ref{fig:wave1-o3}, the Real cEOF1 drives the relationship with amplitude and the Imaginary cEOF1 drives the relationship with phase (not shown).
+There is a close relationship between amplitude and phase of the cEOF1 and the amplitude and phase of the planetary wave 1 in Total Ozone Column between 75°S and 45°S; as shown in Figure~\ref{fig:o3-regr}, this is a wide latitude band that captures most of the area with significant regression coefficients.
+The correlation between the amplitude of each index is 0.77 (CI: 0.61 -- 0.87) the overall correlation between the zonal wave 1 of ozone and cEOF1 (computed as the mean cosine of the difference in phase) is 0.87 (CI: 0.77 -- 0.93).
+As expected from the location of anomalies in Figure~\ref{fig:o3-regr}, the Real cEOF1 drives the relationship with amplitude and the Imaginary cEOF1 drives the relationship with phase (not shown).
 

### 2021-04-13 analysis/paper/shceof.Rmd
Agrega detalles de regersiones y significancia en la metodología
Closes #24

@@
 
-\textcolor{red}{Creo que habría que expandir un poco esta parte e ir comentando todoas las metodologías que vas usando: regresiones, tendencias, cómo evlauas significancia en cada caso, etc. }
+We performed linear regressions to quantify the association between the cEOFs and other variables.
+For each cEOF, we create regression maps by computing fitting a multiple linear model involving both the real and the imaginary part.
+We evaluated statistical significance using a two-sided t-test and, in the case of regression maps, p-values were adjusted by controlling for the False Discovery Rate [@benjamini1995;@wilks2016] to avoid misleading results from the high number of regressions [@walker1914;@katz1991].
+
+Linear trends were computed by Ordinary Least Squares and the 95% confidence interval  was computed assuming a t-distribution with the appropriate residual degrees of freedom.
+To the amplitude of the zonal waves is defined through computing the Fourier transform of the spatial field at each latitude circle.
 

### 2021-04-13 analysis/paper/shceof.Rmd
Renderiza todo

@@
               formula = y ~ I(sin(x)) + I(cos(x)) -1,
-              se = FALSE, color = "black", size = 0.5) +
+              se = FALSE, color = "black", size = 0.5, 
+              fullrange = TRUE) +
     geom_point(aes(colour = mag >= quantile(mag, 0.5),

### 2021-04-13 analysis/paper/shceof.tex
Renderiza todo

@@
 
-\textcolor{red}{Creo que habría que expandir un poco esta parte e ir comentando todoas las metodologías que vas usando: regresiones, tendencias, cómo evlauas significancia en cada caso, etc. }
+We performed linear regressions to quantify the association between the cEOFs and other variables.
+For each cEOF, we create regression maps by computing fitting a multiple linear model involving both the real and the imaginary part.
+We evaluated statistical significance using a two-sided t-test and, in the case of regression maps, p-values were adjusted by controlling for the False Discovery Rate \citep{benjamini1995, wilks2016} to avoid misleading results from the high number of regressions \citep{walker1914, katz1991}.
+
+Linear trends were computed by Ordinary Least Squares and the 95\% confidence interval was computed assuming a t-distribution with the appropriate residual degrees of freedom.
+To the amplitude of the zonal waves is defined through computing the Fourier transform of the spatial field at each latitude circle.
 

### 2021-04-20 analysis/paper/shceof.Rmd
2da Revision

@@
 ---
-title: "Title Goes Here"
+title: " Austral Spring Extratropical Southern Hemisphere zonally assymetric circulation"
 titlerunning:  Title
@@
 
+\textcolor{red}{Agregar parráfo que introduzca el problema del estudio de la ciriculación extra-tropical en SH, su variaiblidad y sus cambios.
+Quizás algo también de los impactos.
+También algo habría que decir de la primavera, ya que solo nos concentrmaos en esa estación}
+
+The large-scale extratropical circulation in the Southern Hemisphere is strongly zonally symmetric, but its zonal departures are highly relevant for surface impacts and could be related to high-impact climate extremes.
+Zonal asymmetries of extratropical circulation in the Southern Hemisphere strongly modulate weather systems and regional cli8 mate through latitudinal transport of heat, humidity, and momentum [@trenberth1980].
+
 One typical way of describing the zonally asymmetric circulation is by the amplitude and phase of zonal waves obtained by Fourier decomposition of geopotential height at each latitude [e.g.
-@vanloon1972;@trenberth1980].
-These methodologies rely on the assumption that the circulation can be meaningfully understood in terms of zonal waves of constant amplitude along a latitude circle.
+@vanloon1972;@trenberth1980;\textcolor{red}{Agregar alguna referencia más actual}].
+This approach suggests that zonal waves 1 and 3 explained most of the variance of the tropospheric extratropical Southern Hemisphere circulation (@vanloon1972).
+However, this methodology rely on the assumption that the circulation can be meaningfully understood in terms of zonal waves of constant amplitude along a latitude circle.
 This is not valid for meridionally propagating waves or zonal waves with localised amplitudes.
-In the case of the wave 3, for example, @trenberth1985 observed that wave 3 played a role in blocking events, but that in many cases, this was due to increased amplitude of a longitudinally localised wave train and not a hemispheric-scale zonal wave 3.
+In the case of the wave 3, for example, @trenberth1985 observed that it played a role in blocking events, but this was mostly due to increased amplitude of a longitudinally localised wave train instead of an hemispheric-scale zonal wave 3.
 
 Another approach to characterising the Southern Hemisphere circulation is by using Empirical Orthogonal Functions (EOF, also known as Principal Component Analysis).
-Within the EOF framework, the Southern Annular Mode appears as the leading mode of variability of the Southern Hemisphere (SAM) [@fogt2020] followed by the two Pacific–South American Patterns (PSA) [@mo2001].
+Within the EOF framework, the Southern Annular Mode (SAM) appears as the leading mode of variability of the tropospheric Southern Hemisphere circulation [@fogt2020] followed by the two Pacific–South American Patterns (PSA) [@mo2001].
 The SAM represents a relatively zonally symmetric pattern of alternating low pressures in polar latitude and a ring of high pressures in high latitudes with an embedded wave 3 pattern that is more prominent in the Pacific sector.
@@
 
-EOFs are more flexible in the sense that they can characterise a richer set of spatial patterns, but they are restricted to standing oscillation modes and have difficulty dealing with propagating or phase-varying modes such as zonal waves.
-A single EOF can also represent a mixture of two or more physical modes.
-
+EOFs are more flexible than Fourier descomposition modes in the sense that they can characterise \textcolor{red}{Esto hay que exresarlo más claramente}.
+Nonetheless, they are restricted to standing oscillation modes and could not represent properly propagating or phase-varying modes such as zonal waves.
+Besides, a single EOF can also represent a mixture of two or more physical modes.
 The inability of a single EOF to describe travelling waves is what forces the PSA pattern to be described by two EOFs.
@@
 On this new projection, they could identify the PSA using Fourier decomposition.
-
-Finally, another, hybrid, methodology consist on identifying particular features of interest and creating indices using simple methods such as averages and differences.
+Finally, another, hybrid, methodology commonly used consist on identifying particular features of interest and creating indices using simple methods such as averages and differences.
 The Marshal SAM Index [@marshall2003], for example, tries to characterise the SAM by the average difference of sea level pressure between 40ºS and 65ºS, following @gong1999.
@@
 This is equivalent to an index of the amplitude of the zonal wave 3 that projects into the phase of the zonal wave 3 of the mean field.
-Similarly, @hobbs2010 posit that the zonally asymmetric circulation in the Southern Hemisphere can be described by the strength and location of two anticyclones located in the sub-Antarctic western hemisphere.
-
-These "feature-based" methods are grounded on other methods to identify the interesting features and can be useful to characterise features that are not readily apparent with Fourier or EOF methods but that nonetheless interesting or important.
-The use of data of single points, which makes them suitable for extending into the data-poor past also makes them not very robust.
+Similarly, @hobbs2010 propose that the zonally asymmetric circulation in the Southern Hemisphere can be described by the strength and location of two anticyclones located in the sub-Antarctic western hemisphere.
+These methods are grounded on other methods to identify the centers of action for the described phenomena and can be useful to characterise features that are not readily apparent with Fourier or EOF methods.
+The use of data of only specific regions or locations makes them suitable for easy computation and for extending into less quality data from the past, but also makes them not very robust and could not capture non-stationary behaviours of the patterns.
+
+\textcolor{red}{Falta un parráfo que deje más claro por qué es necesario considerar un enfoque alternativo y cuál va a ser el beneficio de ello.
+También agregar qué se estudiara conjuntamente la troposfera y la estratósfera }
+
+\textcolor{red}{Falta el objetivo que justifica el uso de la metodología}
 
@@
 
+\textcolor{red}{The data and methods are described in Section 2 :P ...}
+
 # Data and Methods
@@
 We used monthly geopotential height, air temperature, ozone mixing ratio, and Total Ozone Column (TOC) at 2.5\degree\ longitude by 2.5\degree\ latitude of horizontal resolution and 37 vertical isobaric levels from ERA5 [@era5] for the period 1979 to 2019.
-We restricted most of our analysis to the post-satellite era to avoid any confounding factors arising from the incorporation of satellite observations, but we also used the preliminary back extension of ERA5 from 1950 to 1978 [@era5be] to look at long-term trends.
-We derived streamfunction at 200 hPa from ERA5 vorticity using the FORTRAN subroutine FISHPACK [@fishpack] and computed horizontal wave activity fluxes following @plumb1985.
-Sea Surface Temperature (SST) fields were taken from Extended Reconstructed Sea Surface Temperature (ERSST) v5 [@huang2017].
-For precipitation data we used monthly data from the CPC Merged Analysis of Precipitation [@cmap], with a 2.5\degree\ resolution in latitude and longitude.
+Most of our analysis was restricted to the post-satellite era to avoid any confounding factors arising from the incorporation of satellite observations, but we also used the preliminary back extension of ERA5 from 1950 to 1978 [@era5be] to look at long-term trends.
+Streamfunction was derived from ERA5 vorticity at 200 hPa using the FORTRAN subroutine FISHPACK [@fishpack] and horizontal wave activity fluxes was computed following @plumb1985.
+We usded Sea Surface Temperature (SST) monthly fields from Extended Reconstructed Sea Surface Temperature (ERSST) v5 [@huang2017] and precipitation monthly data from the CPC Merged Analysis of Precipitation [@cmap], with a 2.5\degree\ resolution in latitude and longitude.
 This rainfall gridded dataset is based on information from different sources such as rain gauge observations, satellite inferred estimations and the NCEP-NCAR reanalysis, and it is available since 1979 to present.
@@
 
-We restrict our study to  the September-October-November (SON) trimester.
-For that, we use seasonal means, averaging monthly values weighted by the number of days in each  month.
-We also use 200 hPa to represent the high troposphere and 50 hPa to represent the lower stratosphere.
+The study is restricted to the spring season, defined as the September-October-November (SON) trimester.
+We compute seasonal means for the different variables, averaging monthly values weighted by the number of days in each month.
+We use 200 hPa level to represent the high troposphere and 50 hPa to represent the lower stratosphere.
 
@@
 We further split the SAM into its zonally symmetric and zonally asymmetric components (S-SAM and A-SAM indices respectively).
-These are obtained by projecting the zonally asymmetric and zonally symmetric part of the SAM spatial pattern onto monthly geopotential height fields [@campitelli2021].
-Since the analysis here is only for the SON trimester, monthly values were averaged across trimesters.
-
-We also calculated the Pacific South American patterns (PSA1 and PSA2) as the third and fourth leading EOF of seasonal mean 500 hPa geopotential height with all seasons together following @mo2001.
-
-We performed linear regressions to quantify the association between the cEOFs and other variables.
-For each cEOF, we create regression maps by computing fitting a multiple linear model involving both the real and the imaginary part.
-We evaluated statistical significance using a two-sided t-test and, in the case of regression maps, p-values were adjusted by controlling for the False Discovery Rate [@benjamini1995;@wilks2016] to avoid misleading results from the high number of regressions [@walker1914;@katz1991].
-
-Linear trends were computed by Ordinary Least Squares and the 95% confidence interval  was computed assuming a t-distribution with the appropriate residual degrees of freedom.
-To the amplitude of the zonal waves is defined through computing the Fourier transform of the spatial field at each latitude circle.
+These indices were obtained by projecting the zonally asymmetric and zonally symmetric part of the SAM spatial pattern onto monthly geopotential height fields, as proposed by @campitelli2021.
+The Pacific South American patterns (PSA1 and PSA2) were calculated as the third and fourth leading EOF of seasonal mean for 500 hPa geopotential height with all seasons together \textcolor{red}{No me quedo claro lo de "all seasons together"} following @mo2001.
+
+Linear trends were computed by Ordinary Least Squares and the 95% confidence interval was computed assuming a t-distribution with the appropriate residual degrees of freedom \textcolor{red}{Referencia?}.
+The amplitude of the zonal waves was defined through computing the Fourier transform of the spatial field at each latitude circle.
+\textcolor{red}{Esto está repetido con lo de arriba o querías referirte a otra cosa}
 
@@
 
-(ref:eof-naive-cap) Spatial patterns of the leading 4 EOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2019 (arbitrary units).
+(ref:eof-naive-cap) Spatial patterns of the leading 4 EOFs of SON zonal anomalies of geopotential height at 50 hPa for the 1979 -- 2019 period (arbitrary units).
 
@@
 
-In traditional EOF analysis zonal waves appear as pairs of EOFs, usually degenerate, that represent similar patterns but shifted in phase [@horel1984].
-For instance, Figure\ \@ref(fig:eof-naive) shows the leading 4 EOFs of zonally anomalous SON geopotential height at 50 hPa and 200 hPa.
-It is clear that at 50 hPa the first two EOFs represent the same zonal wave 1 pattern and the last 2 represent the same zonal wave pattern with shorter wavenumber shifted by 1/4 wavelength.
+In traditional EOF analysis zonal waves appear as pairs of EOFs, that could be degenerated representing similar patterns but shifted in phase [@horel1984].
+For instance, Figure\ \@ref(fig:eof-naive) shows the leading 4 EOFs of SON geopotential height zonal anomalies at 50 hPa \textcolor{red}{Creo que no aclaras ni acá, ni para los complejos, cuál sería el dominio para el cuál estás calculando el EOF) La barra de colores quedaría mejor si le agregas un contorno que la encierre.}.
+It is clear that the first two EOFs represent the same zonal wave 1 pattern and the last two represent a same zonal wave pattern with shorter wavenumber and four centers of action shifted by 1/4 wavelength.
 A similar EOF structure can be seen in 200 hPa (not shown).
@@
 
-Complex Empirical Orthogonal Functions (cEOF) is a useful method to characterise these waves [@horel1984].
-This method involves augmenting the original fields with the Hilbert transform of the data and then computing the Singular Value Decomposition of it.
-In this work, instead of applying the Hilbert transform to the time series at each point, we apply it to each latitude circle at each moment in time.
-Since the latitude circle is a periodic domain, this procedure does not suffer from edge effects.
-The result of the cEOF is a set of complex spatial patterns and complex time series.
-The real and imaginary part of each spatial pattern represent two phases wave-like spatial pattern that are in quadrature.
+ \textcolor{red}{Detalles para la figura 1: a) Les pusiste cEOF, cuando debería ser EOF; b) te quedaron muy superpuestos los distintos subplots; c) La barra de colores quedaría mejor si le agregas un contorno que la encierre.}
+
+Complex Empirical Orthogonal Functions (cEOF) is a useful method to characterise zonal waves, considering phase-varying structures [@horel1984].
+This method involves augmenting \textcolor{red}{No sé si es un vocabulario específico de usar este método, pero no me queda claro que se entiende por "augmenting" exactamente} the original fields with the Hilbert transform of the data and then computing the Singular Value Decomposition of it.
+In this work, instead of applying the Hilbert transform to the time series at each point, we apply it to each latitude circle at each moment in time \textcolor{red}{Esta aclaración la haces por qué es la manera usual de hacer la metodología, si es asi aclaralo.}.
+Since each latitude circle is a periodic domain, this procedure does not suffer from edge effects.
+The result of the cEOF methodology is a set of complex spatial patterns and complex time series.
+The real and imaginary part of each spatial pattern represent two phases wave-like spatial patterns that are in quadrature.
 The magnitude and argument of each complex time series represent the amplitude and phase of each zonal wave.
@@
 
+The cEOF methodology is applied to SON geopotential height zonal anomalies at 50 and 200 hPa, representing a stratsopheric and upper troposhperic level respectively.
 Table \@ref(tab:corr-ceof-splitted) shows the coefficient of determination between time series of the amplitude of each complex EOF (cEOF) across levels.
 There's a high degree of correlation between the respective cEOF1 and cEOF2 at each level.
-The spatial patterns of the 50 hPa and 200 hPa cEOFs are also similar (not shown)
+The spatial patterns of the 50 hPa and 200 hPa cEOFs are also similar (not shown).
 
@@
 
+We performed linear regressions to quantify the association between the cEOFs and other variables.
+For each cEOF, we create regression maps by computing fitting a multiple linear model involving both the real and the imaginary part.
+We evaluated statistical significance using a two-sided t-test and, in the case of regression maps, p-values were adjusted by controlling for the False Discovery Rate [@benjamini1995;@wilks2016] to avoid misleading results from the high number of regressions [@walker1914;@katz1991].
+
 ## Computation procedures
 
-We performed all analysis in this paper using the R programming language [@rcoreteam2020], using the data.table package [@dowle2020] and the metR package [@campitelli2020].
+We performed all analysis in this paper using the R programming language [@rcoreteam2020], using data.table [@dowle2020] and metR [@campitelli2020] packages.
 All graphics are made using ggplot2 [@wickham2009].
@@
 
-## cEOF patterns
-
-```{r ceofs-1, fig.cap = "Spatial patterns of the leading 2 cEOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2019.
-Real part in shading, imaginary part in contours.
+## cEOF spatial patterns
+
+```{r ceofs-1, fig.cap = "Spatial patterns for the two leading cEOFs of SON zonal anomalies of geopotential height at 50 hPa and 200 hPa for the 1979 -- 2019 period.
+The shading (contours) corresponds to real (imaginary) part.
 Arbitrary units.", fig.width=width_column, fig.height=width_column*1.2}
@@
 
-```{r ceofs-2, fig.cap = "Time series of the leading 2 cEOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2019.", fig.height=width_column, fig.width = width_column}
+```{r ceofs-2, fig.cap = "Time series of the leading two cEOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2019.", fig.height=width_column, fig.width = width_column}
 ceof %>%
@@
 
-To describe the spatial and temporal variability of the circulation zonal anomalies, the spatial and temporal parts of the first two leading cEOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa are shown in Figures \@ref(fig:ceofs-1) and \@ref(fig:ceofs-2).
-The first mode (CEOF1) explains 82% of the variance, while the second mode (CEOF1) explains a smaller fraction (7%).
-In the spatial patterns (Fig.\ \@ref(fig:ceofs-1)), the real (in shading) and the imaginary components (in contour) are in quadrature by construction, so that each cEOF describe a single wave-like pattern whose amplitude and position (i.e.
+To describe the variability of the circulation zonal anomalies, the spatial and temporal parts of the first two leading cEOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa, compueted jointly at both levels, are shown in Figures \@ref(fig:ceofs-1) and \@ref(fig:ceofs-2).
+The first mode (cEOF1) explains 82% of the variance, while the second mode (cEOF2) explains a smaller fraction (7%).
+In the spatial patterns (Fig.\ \@ref(fig:ceofs-1)), the real and the imaginary components are in quadrature by construction, so that each cEOF describe a single wave-like pattern whose amplitude and position (i.e.
 phase) is controlled by the magnitude and phase of the temporal cEOF.
@@
 The first cEOF (Fig.\ \@ref(fig:eof-naive) column\ a) is a hemispheric wave 1 pattern with maximum amplitude at high latitudes.
-At 50 hPa the Real cEOF1 has the maximum of the wave 1 at 150ºE and at 200 hPa, the maximum is locates at around 175ºE indicating a westerly shift in phase.
-
-The second cEOF (Fig.\ \@ref(fig:eof-naive) column\ b) show also a zonal wave-like structure with maximum amplitude at high latitudes, but with shorter spatial scales.
-In particular, it's dominant structure at both levels is a wave 3 but with larger amplitude in the pacific sector.
+At 50 hPa the Real cEOF1 has the maximum of the wave 1 at 150ºE and at 200 hPa, the maximum is located at around 175ºE indicating a westerly shift in phase.
+The second cEOF (Fig.\ \@ref(fig:eof-naive) column\ b) shows also a zonal wave-like structure with maximum amplitude at high latitudes, but with shorter spatial scales.
+In particular, its dominant structure at both levels is a wave 3 but with larger amplitude in the pacific sector.
 This modulated amplitude is especially apparent at 200 hPa.
@@
 This is due to the fact that the geopotential fields that enter into the cEOFs algorithm are anomalies with respect to the zonal mean, not the time mean.
-As a consequence, the variability associated with the first cEOF includes variability that projects onto the mean zonally anomalous field.
+As a consequence, the variability associated with the cEOF1 includes variability that projects onto the mean zonally anomalous field.
 There are no significant simultanous correlation between the time series.
+Both cEOFs show year-to-year variability but show no evidence of decadal variabilty.
+The relatively short timespan (`r diff(range(unique(year(era5$time))))` years) hampers our habilty to detect longer trends.
 
@@
 Each series is computed by projecting monthly geopotential height zonal anomalies standardised by level south of 20ºS onto the corresponding spatial pattern and scaled to zero mean and unit standard deviation.
+\textcolor{red}{Esta aclaración metodológica debería estar en la secció de métodos y revisa como expresas este caption para que sea consistente con el resto.}
 
@@
 
-Both cEOF shot year-to-year variability but show no evidence of decadal variabilty.
-The relatively short timespan (`r diff(range(unique(year(era5$time))))` years) hampers our habilty to detect longer trends.
-Therefore, we extend the cEOF time series using the preliminary ERA5 back extension going back to 1950 by projecting the older geopogential height zonal anomaly fields onto the spatial patterns shown in Figure\ \@ref(fig:ceofs-1).
-
-The extended cEOF timeseires are shown in Figure\ \@ref(fig:extended-series).
+Therefore, we extend the cEOFs time series using the preliminary ERA5 back extension going back to 1950 by projecting the older SON geopogential height zonal anomaly fields onto the spatial patterns shown in Figure\ \@ref(fig:ceofs-1).
 There is a significant upward trend in the real component of cEOF1 (Fig.\ \@ref(fig:extended-series)a.1, p-value\ `r trends[part == "Real" & cEOF == "cEOF1"]$p_report`) and no significant trend in any of the complex components of cEOF2.
@@
 
+\textcolor{red}{Revisa la figura 4 que te están quedando bastante desparejos los subplots.
+También sugieron mantener la convención de colores de la figura anterior (real/imaginario) para hacer más directa la interpretación.}
+
 This long-term change indicates an increase in the magnitude of the high latitude zonal wave 1.
@@
 
-## Regressions
+## cEOFs Regressions
 
@@
 
-(ref:eof1-regr-gh-cap) Regression coefficients of the real and imaginary part of the first cEOF on SON geopotential height for the 1979 -- 2019 period.
+(ref:eof1-regr-gh-cap) Regression coefficients of the real and imaginary part of the first cEOF on SON geopotential height anomalies for the 1979 -- 2019 period.
 These coefficients come from multiple linear regression involving the real and imaginary part of both cEFO2.
@@
 
-Figure\ \@ref(fig:eof1-regr-gh) shows regression patterns between geopotential height anomalies and cEOF1.
+\textcolor{red}{Creo que te había hecho el mismo comentario en la versión anterior.
+Tenés que ser más preciso cuando hablas/mostras regresiones indicando en qué dirección va la regresión "regressed upon".
+Revisalo a lo largo del texto.
+En la figura 5, ¿no está al revés?}
+
+\textcolor{red}{En el caption de la Figura 5 falta aclarar qué son los puntos.}
+
+Figure\ \@ref(fig:eof1-regr-gh) shows regression maps of SON geopotential height anomalies upon cEOF1.
 At 50 hPa (Figure\ \@ref(fig:eof1-regr-gh) row a), both the Real and Imaginary cEOF1 are associated with planetary wave 1 patterns, that are 90º off phase.
@@
 It is only possible to distinguish partially the wave 1 pattern in relation with the real cEOF1 (Figure\ \@ref(fig:eof1-regr-gh)b.1).
-The imaginary cEOF1 shows a much more zonally symmetrical pattern resembling a negative Southern Annular Mode.
+The imaginary cEOF1 shows a much more zonally symmetrical pattern resembling the negative SAM phase.
 
 With the exception of the imaginary cEOF1, it is clear that these patterns are very different than the fully zonally asymmetric versions (Fig.\ \@ref(fig:ceofs-1)), particularly at 200 hPa.
-Moreover, only in the stratosphere these patterns actually show a distinguishable wave 1 pattern shifted in phase by 90º.
-This should not be utterly surprising, since the spatial patterns of the *naive* EOFs only show this feature at 50 hPa (Fig.\ \@ref(fig:eof-naive)), suggesting that using the cEOF method is artificially generating a wave structure at 200hPa.
+Moreover, only in the stratosphere these patterns actually show a distinguishable wave 1 pattern shifted in phase by 90º, suggesting that using the cEOF method is artificially generating a wave structure at 200hPa.
 Therefore, the magnitude and phase of the cEOF1 are associated with the magnitude and phase of a zonal wave only in the stratosphere.
@@
 
-Figure\ \@ref(fig:eof2-regr-gh) shows the regression pattern of geopotential height and the cEOF2.
+Figure\ \@ref(fig:eof2-regr-gh) shows the regression pattern of geopotential height anomalies upon the cEOF2.
 Unlike for cEOF1, in this case the regression patterns are similar to the fully zonally asymmetric patterns from Figure\ \@ref(fig:ceofs-1).
@@
 
-## cEOFs relationship with other variables
+## cEOFs relationship with other variables (Quedo medio rara la divsión con los títulos.
+Me parece que convendrá discutir primero la relación con otros modos de circulación conocidos (SAM, Ozono) y después con las otras variables(Temp, TOC, SST, precip) )
 
@@
 
+The relation between cEOFs and temperature was also evaluated.
 Figure\ \@ref(fig:eof1-regr-t) shows regression patterns of air temperature at 50hPa and 200hPa onto cEOF1.
 In both levels, the Real cEOF1 is associated with positive monopole over the South Pole with its centre moved slightly towards 150ºE (Fig.\ \@ref(fig:eof1-regr-t) column 1).
-Regression maps on the Imaginary cEOF1, on the other hand, show a more clear wae 1 pattern with its maximum around 60ºE.
+On the other hand, the regression maps on the Imaginary cEOF1 show a more clear wave 1 pattern with its maximum around 60ºE.
 The distribution of temperature regression coefficients at 50 hPa and at 200 hPa mirror the geopotential height regression maps at 50 hPa (Fig.\ \@ref(fig:eof1-regr-gh)).
 
-The vertical distribution of these patterns is shown in Figure\ \@ref(fig:t-vertical)) that shows the regression coefficientes between cEOF1 and zonal anomalies of air temperature and zonal anomalies of ozone mixing ratio averaged between `r knitr::combine_words(LatLabel(o3wave_lats))`.
+Figure \ \@ref(fig:t-vertical) shows the vertical distribution of the regression coefficientes on cEOF1 from zonal anomalies of air temperature and zonal anomalies of ozone mixing ratio averaged between `r knitr::combine_words(LatLabel(o3wave_lats))`.
+\textcolor{red}{Deberias aclarar qué significa la terminología que estas utilizando.
+Lo podes hacer en los métodos y ya queda para todo el texto}.
 Temperature zonal anomalies associated with cEOF1 show a clear wave 1 pattern for both real and imaginary components throughout the atmosphere above 250 hPa  with a change in sign above 10 hPa.
@@
 
-The maximum ozone anomalies are co-located with the minimum temperature anomalies above 10 hPa and with the maximum temperature anomalies below 10 hPa (Fig.\ \@ref(fig:t-vertical))); that is, the ozone zonal wave 1 is anticorrelated with the temperature zonal wave 1 in the upper stratosphere, and directly correlated in the upper stratosphere .
+The maximum ozone anomalies are co-located with the minimum temperature anomalies above 10 hPa and with the maximum temperature anomalies below 10 hPa (Fig.\ \@ref(fig:t-vertical))).
+Therefore, the ozone zonal wave 1 is anticorrelated with the temperature zonal wave 1 in the upper stratosphere, and directly correlated in the upper stratosphere .
 This change in phase is observed in ozone anomalies forced by planetary waves that reach the stratosphere.
 In the photochemically-dominated upper stratosphere, cold temperatures inhibit the destruction of ozone, and the advectively-dominated lower stratosphere, ozone anomalies are 90º out of phase with horizontal and vertical transport, which is 90º our of phase with temperature anomalies [@hartmann1979;@wirth1993;@smith1995].
+
+\textcolor{red}{Revisa que todos los captions tengan las unidades correspondientes}
 
@@
 
-(ref:o3-regr-cap) Regression coefficients of SON mean Total Ozone Column anomalies onto the real (a) and imaginary (b) components of the cEOF1.
+(ref:o3-regr-cap) Regression coefficients (shaded) of SON mean Total Ozone Column anomalies onto the real (a) and imaginary (b) components of the cEOF1 for the 1979 -- 2019 period.
 On contours, the mean zonal anomaly of Total OZone Column (negative contours in dashed lines, Dobson Units).
-For  the 1979 -- 2019 period.
 
@@
 
-Figure\ \@ref(fig:o3-regr) shows regression maps of cEOF1 with fields of Total Ozone Column (TOC).
-It shows zonal wave 1 pattens in TOC associated with both phases of cEOF1.
-Climatologically, the springtime Ozone minimum is located off the South Pole and towards the Weddell Sea [@wirth1993].
+The regression maps of cEOF1 with fields of Total Ozone Column (TOC) (Fig.
+\ \@ref(fig:o3-regr)) show zonal wave 1 pattens in TOC associated with both phases of cEOF1.
+Climatologically, the springtime Ozone minimum is located off the South Pole and towards the Weddell Sea [@wirth1993, \textcolor{red}{¿Alguna referencia más nueva?}].
 Thus, the Real cEOF1 regression pattern (Figure\ \@ref(fig:o3-regr)a) coincides with the climatological position of the ozone hole while the one for the Imaginary cEOF1 is shifted by 90º.
@@
 
-There is a close relationship between amplitude and phase of the cEOF1 and the amplitude and phase of the planetary wave 1 in Total Ozone Column between `r knitr::combine_words(LatLabel(o3wave_lats))`; as shown in Figure\ \@ref(fig:o3-regr), this is a wide latitude band that captures most of the area with significant regression coefficients.
-The correlation between the amplitude of each index is `r wave1_cor[with == "amplitude"]$text` the overall correlation between the zonal wave 1 of ozone and cEOF1 (computed as the mean cosine of the difference in phase) is `r wave_correlation$text`.
-As expected from the location of anomalies in Figure\ \@ref(fig:o3-regr), the Real cEOF1 drives the relationship with amplitude and the Imaginary cEOF1 drives the relationship with phase (not shown).
+There is a close spatial relationship between amplitudes and phases of the cEOF1 and TOC planetary wave 1 between `r knitr::combine_words(LatLabel(o3wave_lats))`.
+The correlation between the amplitude of both indices is `r wave1_cor[with == "amplitude"]$text` the overall correlation between the zonal wave 1 of ozone and cEOF1 (computed as the mean cosine of the difference in phase) is `r wave_correlation$text`.
+As expected from the location of regressions in Figure\ \@ref(fig:o3-regr), the Real cEOF1 drives the relationship of amplitudes and the Imaginary cEOF1 drives the relationship of phases between the timeseries (not shown).
 
@@
 
-To explore the relationship between SAM and the modes described from the cEOF, we compute.
-the coefficient of determination between the cEOFs timeseries and the three SAM indices (SAM, A-SAM and S-SAM) at each level (Fig.\ \@ref(fig:sam-eof-vertical)).
+To explore the relationship between SAM and the modes described from the cEOF, we compute the coefficient of determination between the cEOFs timeseries and the three SAM indices (SAM, A-SAM and S-SAM) at each vertical level (Fig.\ \@ref(fig:sam-eof-vertical)).
 The SAM index is moderately correlated with the Real cEOF1 in all levels, and with the Imaginary cEOF1 and Imaginary cEOF2 in the troposphere.
@@
 
-In the case of cEOF2, the moderate correlation between SAM and Imaginary cEOF2 becomes extremely high when only the zonally asymmetric variability of the SAM is considered.
+In the case of cEOF2, the moderate correlation between SAM and Imaginary cEOF2 becomes extremely high only when the zonally asymmetric variability of the SAM is considered.
 The Imaginary cEOF2 and the A-SAM share up to `r scales::percent(max_r2[cEOF == "cEOF2"]$estimate^2, 2)` of their variance, reached at `r max_r2[cEOF == "cEOF2"]$lev` hPa (Figure\ \@ref(fig:sam-eof-vertical).b2).
-Such extremely high correlation between the asymmetric SAM and the Imaginary cEOF2 suggests that these are not different features that share a causal link, but that they might be different ways of characterising the same phenomenon.
+Such extremely high correlation between the asymmetric SAM and the Imaginary cEOF2 suggests that these might be different ways of characterising the same phenomenon.
 
@@
 
-To test the hypothesis that a PSA-like feature can appear as statistic contamination of an otherwise zonally symmetrical and uncorrelated SAM, we create a synthetic dataset of `r uniqueN(fake_sam$time)` fields by reconstructing the 200 hPa geopotential field explained by the cEOF2 and adding a perfectly zonally symmetric SAM-like structure with random sign and value sampled with replacement from the real SAM index.
-To this synthetic dataset we then apply traditional EOF.
+\textcolor{red}{Esta parte creo que quedaría mejor como una subsección, aunque no se me viene rápido a la cabeza el título y la pondría después de la parte de PSA}
+
+To test the hypothesis that a PSA-like feature can appear as statistic contamination of an otherwise zonally symmetrical and uncorrelated SAM, we create a synthetic dataset of `r uniqueN(fake_sam$time)` fields by reconstructing the SON 200 hPa geopotential height field explained by the cEOF2 and adding a perfectly zonally symmetric SAM-like structure with random sign and value sampled with replacement from the real SAM index.
+To this synthetic dataset (\textcolor{red}{No me suena mucho la palabra "synthetic", ¿se usa en este contexto?}) we then apply traditional EOF.
 By construction this dataset contains two independent components: a zonally symmetric standing oscillation and a PSA-like of varying amplitude and phase (Figure\ \@ref(fig:fake-eof) row a).
-The former one is a standing oscillation which could be represented by a single EOF, while the latter, being a wave which can change phase, needs two EOFs to be represented correctly.
+While the former is a standing oscillation which could be represented by a single EOF, the latter is a wave which can change phase, so it needs two EOFs to be represented correctly.
 Thus, the full variance of this synthetic dataset can be perfectly explained by three EOFs.
+
+\textcolor{red}{Los titulos de la figura 11 superior se prestan a la confusión, ya que no es exactamente el SAM y no se entiende qué sería "Imaginary" y "Real"}
 
@@
 However, even though the synthetic SAM is perfectly zonally symmetric by construction, the leading EOF is not.
-The EOF decomposition mixes this mode with parts of the the zonally asymmetric cEOF2 pattern, resulting in a pattern eerily similar to the real SAM pattern.
-This happens because, although in this synthetic dataset the zonally symmetrical SAM and the cEOF2 are uncorrelated by construction, the sample correlation is not null.
+The EOF decomposition mixes this mode with parts of the the zonally asymmetric cEOF2 pattern, resulting in a pattern pretty similar to the real SAM pattern (\textcolor{red}{REF}).
+This happens because, although in this synthetic dataset the zonally symmetrical SAM and the cEOF2 are uncorrelated by construction, the sample correlation is not null \textcolor{red}{No me quedo 100% clara esta parte, ¿podrías expresarla un poco distinto?}.
 In this particular example, it is `r cor_with_eof2[part == "Real"]$text` with the Real part and `r cor_with_eof2[part == "Imaginary"]$text` with the Imaginary part.
@@
 
-This suggests that the zonal asymmetries of the SAM could very well be mostly a statistical artefact.
+This suggests that the zonal asymmetries of the SON SAM could very well be mostly a statistical artefact.
 Since many surface impacts are mediated by the asymmetric component, as well as the relationship between SAM and ENSO [@campitelli2021], this potential issue could affect the interpretability of many results involving the SAM.
@@
 
-(ref:psa-eof2-cap) Correlation between the Real and Imaginary parts of cEOF2 and the PSA1 and PSA2 modes computed as the second and third EOFs of seasonal geopotential height anomalies [@mo2001].
-For the period 1979 -- 2019.
+(ref:psa-eof2-cap) Correlations between the Real and Imaginary parts of cEOF2 and the PSA1 and PSA2 modes computed as the second and third EOFs of seasonal geopotential height anomalies [following @mo2001] for the  1979 -- 2019 period.
+\textcolor{red}{Acclarar lo que está entre parentésis y que sería la coloración.
+Me parece también que hay un error en el del PSA2 imaginario.}
 
@@
 The relationship between PSAs and the modes described from the cEOFs is also studied.
-Figure\ \@ref(tab:psa-eof2) shows the relationship between the two PSA indices and the Real and Imaginary phase of cEOF2.
+Table\ \@ref(tab:psa-eof2) shows the correlations between the two PSA indices and the timeseries for Real and Imaginary phase of cEOF2.
 As anticipated by Figure\ \@ref(fig:eof2-regr-gh), there is a strong correlation between PSA1 and Imaginary cEOF2, and between PSA2 and Real cEOF2.
 Conversely, there is no relationship between PSA1 and Real cEOF2, and between PSA2 and Imaginary cEOF2.
-So not only this cEOF2 represents well both the spatial structure and temporal evolution of the PSA modes, but it's also possible to make a rather clean association between its two phases and the two PSA modes.
-
-This particular rotation of cEOF2 is the one which maximises the relationship between ENSO and Imaginary cEOF2.
-It is also the same that maximises this clean association between cEOF2 parts and PSA modes.
+As a result, cEOF2 represents well both the spatial structure and temporal evolution of the PSA modes, as well as it is possible to make a rather clean association between its two phases and the two PSA modes.
+So, It could be concluded that the same particular rotation of cEOF2 that maximises the association between cEOF2 parts and PSA modes, is the one which maximises the relationship between ENSO and Imaginary cEOF2.
 
 The reason the conventional EOF analysis arrives at the same separation than our particular cEOF rotation is probably the fact that not all phases are equal.
-Figure\ \@ref(fig:phase-histogram) is an histogram that counts the number of SON years in which the cEOF2 was near a particular phase with the observations marked as rugs on the horizontal axis.
+\textcolor{red}{No me quedo claro ese "equal", ¿igual de probables/posibles/factibles?}.
+To visualize that,  Figure\ \@ref(fig:phase-histogram) shows an histogram that counts the number of SON years in which the cEOF2 was close to each of the four particular phases (positive/negative Real/Imaginary), with the observations for each season marked as rugs on the horizontal axis.
 For instance, years with cEOF2 phase within 45º of 0º are nearest the "positive Real" phase.
 About two thirds of time cEOF2 has a phase similar to either the negative or positive Imaginary phase.
-
-(ref:phase-histogram-cap) Histogram of phase distribution of cEOF2 for the period 1979 -- 2019. Bins are centred at 90º, 0º, -90º, -180º with a binwidth of 90º.
+It could be seen that Imaginary phase is the most common phase and It is also the direction that has the maximum relationship with ENSO.
+Therefore, the Imaginary cEOF2 explains more variance than the Real cEOF2 and conventional EOF analysis will tend to separate the two.
+
+(ref:phase-histogram-cap) Histogram of phase distribution of cEOF2 for the 1979 -- 2019 period. Bins are centred at 90º, 0º, -90º, -180º with a binwidth of 90º. \textcolor{red}{Rugs...}
 
@@
 
-This phase-locking behaviour indicates that the Imaginary phase both is the direction that has the maximum relationship with ENSO and is also the most common phase.
-Therefore, the Imaginary cEOF2 explains more variance than the Real cEOF2 and conventional cEOF analysis will tend to separate the two.
-
-This is similar to the method used by @irving2016.
-From their Fourier decomposition of reprojected meridional wind fields, they show that the phase distribution is bimodal and isolate the PSA pattern from the rest of the PSA-like variability by selecting events that are near the peacks of the distribution (compare our Figure\ \@ref(fig:phase-histogram) with their Figure 6).
+The result obtaneid here is similar to the one obtained by @irving2016.
+From their Fourier decomposition of reprojected meridional wind fields \textcolor{red}{¿En qué región?}, they show that the phase distribution is bimodal and isolate the PSA pattern from the rest of the PSA-like variability by selecting events that are near the peaks of the distribution (compare our Figure\ \@ref(fig:phase-histogram) with their Figure 6).
 
@@
 
-(ref:sst-psi-2-cap) Regression maps of cEOF2 with SST (column a) and streamfunction zonal anomalies with their corresponding activity wave flux (column b).
+(ref:sst-psi-2-cap) Regression maps of cEOF2 with SST (column a) and streamfunction zonal anomalies (shaded) with their corresponding activity wave flux (vectors) (column b).
 Areas marked with dots have p-values smaller than 0.05 adjusted for FDR.
@@
 
-Figure\ \@ref(fig:sst-psi-2) shows regression maps between cEOF2 and Sea Surface Temperatures (SST) Streamfunction at 200 hPa.
-The Imaginary cEOF2 is associated with strong positive SST anomalies on the Central Pacific and negative anomalies on an area across the North of Australia and New Zealand, the South Pacific Convergence Zone (SPCZ) (Figure\ \@ref(fig:sst-psi-2).a2).
-This pattern is almost canonically positive ENSO and indeed, the correlation between the Imaginary cEOF2 and the Oceanic Niño Index [@bamston1997] is very high `r enso_cor[cEOF == "cEOF2" & part == "Imaginary"]$text`.
-Streamfunction anomalies show a coherent picture.
-The Imaginary cEOF2 is associated with strong wave-like streamfunction anomalies emanating from the tropics (Figure\ \@ref(fig:sst-psi-2).b2).
-This is consistent with what we know of the effect of ENSO on the extratropics: SST anomalies initiate anomalous convection that excites Rossby waves that propagate meridionally towards higher latitudes [@mo2000].
-
-Since the Real cEOF2 represents just a different phase of the same wave train, one would expect that it would show a similar forcing pattern to the Imaginary cEOF with a slight translation of its location.
+The connections between cEOFs and Tropical sources is also assesed.
+Figure\ \@ref(fig:sst-psi-2) shows the regression maps of Sea Surface Temperatures (SST) and Streamfunction at 200 hPa respectively upon cEOF2.
+The Imaginary cEOF2 is associated with strong positive SST anomalies on the Central Pacific and negative anomalies over an area across the North of Australia and New Zealand, the South Pacific Convergence Zone (SPCZ) (Figure\ \@ref(fig:sst-psi-2).a2).
+This pattern is almost canonically positive ENSO (\textcolor{red}{REF ENSO}) and indeed, the correlation between the Imaginary cEOF2 and the Oceanic Niño Index [@bamston1997] is very high `r enso_cor[cEOF == "cEOF2" & part == "Imaginary"]$text`.
+Besides, the Imaginary cEOF2 is associated with strong wave-like streamfunction anomalies emanating from the tropics (Figure\ \@ref(fig:sst-psi-2).b2).
+This is consistent with the effect of ENSO on the extratropics: SST anomalies initiate anomalous convection that excites Rossby waves that propagate meridionally towards higher latitudes [@mo2000].
+
+\textcolor{red}{Revisa los labels de la colorbar para la figura 13.
+Además ojo que solamente sirven para la columna a, faltan los correspondientes a la streamfunction y el vector de referencia para wave flux}
+
+Since the Real cEOF2 represents just a different phase of the same wave train, it would be expected to show a similar forcing pattern to the Imaginary cEOF with a slight translation of its location.
 However, Figure\ \@ref(fig:sst-psi-2).a1 and b1 show that the Real cEOF2 is not associated either with any significant SST nor streamfunction anomalies in the tropics.
 The correlation between the Real cEOF2 and ENSO is also not significant (`r enso_cor[cEOF == "cEOF2" & part == "Real"]$text`).
-This lack of tropical signal suggests different natures of the different phases of the cEOF2 wave train.
-
-```{r enso-phase, fig.cap = "Relationship between ENSO and phase of cEOF2 for the period 1979 -- 2019.
-Colours denote years with magnitude of cEOF2 greater or smaller than the 50th percentile.
+This lack of tropical signal suggests different natures of the different phases of the Real cEOF2 wave train, which has the maximum amplitudes in the surroundings of southern South America.
+
+```{r enso-phase, fig.cap = " ONI plotted against cEOF2 phase for th 1979 -- 2019 period.
+Years with magnitude of cEOF2 greater or smaller than the 50th percentile are shown as orange circles and green diamomnds respectively.
 Black line is the fit ONI ~ sin(phase) computed by OLS weighted by the magnitude of cEOF2.", fig.width = width_column, fig.height=3}
@@
 
-To better explore the relationship between tropical forcing and phase of the cEOF2, Figure\ \@ref(fig:enso-phase) plots the ONI index and the phase of the cEOF2 for each SON trimester between 1979 and 2019, highlighting years in which the magnitude of cEOF2 is above the median.
+To better explore the relationship between tropical forcing and phase of the cEOF2, Figure\ \@ref(fig:enso-phase) shows the ONI index against the cEOF2 phase for each SON trimester between 1979 and 2019, highlighting years in which the magnitude of cEOF2 is above the median.
 In years with positive ONI, the phase of the cEOF2 is mostly around +90º (corresponding with positive imaginary part) and vice versa.
@@
 
-Furthermore, Figure\ \@ref(fig:enso-phase) suggest that strong cEOF2 years tend to coincide with strong ENSO years.
+\textcolor{red}{¿Qué significa para el fit "weighted by the magnitude of cEOF2"?}
+
+Furthermore, Figure\ \@ref(fig:enso-phase) suggest that higher amplitude cEOF2 years tend to coincide with stronger ENSO years.
 The correlation between the absolute magnitude of the ONI and the magnitude of the cEOF2 is `r cor_enso_mag$text`.
@@
 
+\textcolor{red}{Se que la cantidad de años hace bastante limitada la muestra, pero creo que es un resultado a tener en cuenta que en esos eventos muy intensos se alineen.
+¿Probaste hacer la correlación con todos los puntos naranjas?}
+
 It could be concluded that the wave train represented by cEOF2 can be both part of the internal variability of the extratropical atmosphere or forced by tropical SSTs.
@@
 Figure\ \@ref(fig:sst-psi-1) shows SST and streamfunction regression maps for cEOF1.
-There is no significant pattern of SST anomalies associated with either the Real or Imaginary cEOF1.
+Unlike the cEOF2 case, there is no significant pattern of SST anomalies associated with either the Real or Imaginary cEOF1.
 Consistently, streamfunction anomalies do not show any tropical influence.
-Instead of that, the real and imaginary CEOF1 are associated with zonally wave activity fluxes in the extra-tropics, except for an equatorward flow from the coast of Antarctica around 150ºE in the real component.
+Instead of that, the real and imaginary CEOF1 are associated with zonally wave activity fluxes in the extra-tropics around 60ºS, except for an equatorward flow from the coast of Antarctica around 150ºE in the real component.
 
@@
 (ref:pp-america-cap) Regression of SON mean precipitation anomalies in South America (mm per day, shaded) and (column a) cEOF1 the (row 1) Imaginary and (column 1) Real phase.
-For the 1979 -- 2018 period.
+For the 1979 -- 2018 \textcolor{red}{¿Cambiaste el año o es un typo?}  period.
 Black contours with dots indicate areas with p-value smaller than 0.05 controlling for False Detection Rate.
@@
 Figure\ \@ref(fig:pp-america) shows regression maps of seasonal precipitation with each cEOF in South America.
-cEOF1 is associated with slight decreases in precipitation in Southern Brazil and Paraguay, although these are not statistically significant \@ref(fig:pp-america) column a.
-The strongest precipitation anomalies are the ones associated with the Imaginary cEOF2.
+cEOF1 is associated with slight decreases \textcolor{red}{Revisa como estás expresando toda esta parte (y en el resto del texto), recorda que son regresiones así que no corresponde decir postive/negative sino haces alguna referencia a la variable con las que los estás regresionando.}
+ in precipitation in Southern Brazil and Paraguay, although these are not statistically significant (Fig.\ \@ref(fig:pp-america) column a).
+The strongest precipitation regression coefccients are the ones associated with the Imaginary cEOF2.
 The positive anomalies on Southeastern South America (SESA) and Central Chile, and negative anomalies over South Atlantic Convergence Zone is a well known springtime precipitation signature of ENSO [@cai2020a] and it is also similar to the precipitation anomalies associated with the A-SAM [@campitelli2021].
-This is not surprising considering the close relationship between the ONI, the Asymmetric SAM index and the Imaginary cEOF2 shown previously, but further consolidates the identification of this mode with the PSA pattern.
-The Real cEOF2, on the other hand, is associated with negative precipitation anomalies in a smaller area of SESA.
-There is a phase dependence of the precipitation anomalies in SESA (not shown), which resembles the relationship between ONI and the phase of cEOF2.
-A sinusoidal fit has an $r^2$\ of\ `r round(summary(model)$r.squared, 2)` with p-value\ `r report_p(p_value)`.
+This is not surprising considering the close relationship between the ONI, the A-SAM index and the Imaginary cEOF2 shown previously, but further consolidates the identification of this mode with the PSA pattern.
+On the other hand, the Real cEOF2 is associated with negative precipitation anomalies in a smaller area of SESA.
+Resembling the relationship between ONI and the phase of cEOF2 (Fig.\ \@ref(fig:enso-phase)), there is a cEOF2 phase dependence of the precipitation anomalies in SESA (not shown).
+These variable could be related wth a signficant sinusoidal fit that has a coefficient of determination of\ `r round(summary(model)$r.squared, 2)` (p-value\ `r report_p(p_value)`).
 
@@
 
-The precipitation anomalies in Oceania associated with cEOF1 are also weak and not statistically significant (Figure\ \@ref(fig:pp-oceania) column a) whereas EFO2 is associated with large and expansive anomalies.
+The precipitation anomalies in Oceania associated with cEOF1 are also weak and not statistically significant (Figure\ \@ref(fig:pp-oceania) column a) whereas EFO2 is associated with large and widespread anomalies.
 The Real cEOF2 is associated with positive seasonal anomalies in Northern and Eastern Australia, and negative ones over New Zealand, while precipitation anomalies associated with Imaginary cEOF2 are negative over all Eastern Australia.
@@
 To better understand the relationship between precipitation anomalies over Western Australia and the cEOF2, Figure\ \@ref(fig:australia-pp-phase) plots precipitation anomalies averaged over Eastern Australia (box in Figure\ \@ref(fig:pp-oceania)) as a function of cEOF2 phase.
-The black line shows a sinusoidal fit, whose coefficients we fitted by weighted least squares using the magnitude of the EFO2 as weight.
-The $r^2$ corresponding to the fit is `r signif(summary(model)$r.squared, 2)`, with p-value `r report_p(p_value)`.
-There is a strong relationship between the phase of the cEOF2 and precipitation anomalies in Eastern Australia, but Real and Imaginary phases we chose are not exactly aligned with the direction that maximises this relationship.
-If one were to perform a more in-depth analysis of the relationship between this pattern and precipitation in this region, one would conceivably align one of the axis (either the Real or the Imaginary) to the one that maximises it.
+The black line shows a sinusoidal fit, whose coefficients are fitted by weighted least squares using the magnitude of the EFO2 as weight.
+The $r^2$ corresponding to the fit is `r signif(summary(model)$r.squared, 2)` (p-value `r report_p(p_value)`).
+There is a strong relationship between the phase of the cEOF2 and precipitation anomalies in Eastern Australia, but the Real and Imaginary phases we chose are not exactly aligned with the direction that maximises this relationship.
+To perform a more in-depth analysis of the relationship between this pattern and precipitation in this region,  it would be recommended to align one of the axis (either the Real or the Imaginary) to the phase that maximises the described relationship.
+
+\textcolor{red}{¿Había alguna razón para poner SA antes que Oceanía?
+Sino invertiría el orden, porque creo que es más obvio lo que decis en Sudamérica sobre el sinusoidal fit, si mostras antes de lo de Oceanía.}
+
+\textcolor{red}{La formula en la figura trata de escribirla en una forma más precisa "precipitación"/"phase" son muy generales, y también teniendo en cuenta las unidades.
+También incluilo en el caption.}
 
@@
 
-We studied the two leading modes of co-variability of September-October-November tropospheric and stratospheric zonal anomaliesby applying complex Empirical Orthogonal Functions.
+\textcolor{red}{Al igual que en la introducción, te falta establecer más claro el objetivo, después iría lo que escribiste.}
+
+We studied the two leading modes of co-variability of September-October-November tropospheric and stratospheric zonal anomalies by applying complex Empirical Orthogonal Functions.
 
@@
 We show that the Imaginary cEOF2 can be identified with the PSA1 and the Real cEOF2 with the PSA2.
-
-We show that this cEOF prefers a phase aligned with the Imaginary cEOF2.
-This is because the when ENSO is active, it forces the cEOF2 to be in the Imaginary phase; positive for El Niño and negative for La Niña.
+This cEOF prefers a phase aligned with the Imaginary cEOF2.
+This is because the when ENSO is active, it forces the cEOF2 to be in the Imaginary phase.
 When ENSO is neutral, the cEOF2 is still active, but with no particular phase.
@@
 
-We further show that the Imaginary cEOF2 is closely related to the Southern Annular Mode in the troposphere.
-In fact, it is a virtually perfect indicator of the zonally asymmetric portion of the SAM.
-To the extreme that the leading EOF of geopotential height with the cEOF2 removes results in an annular mode that has none of the typical zonal asymmetries of the observed SAM.
+We further show that the Imaginary cEOF2 is closely related to the Southern Annular Mode in the troposphere. In fact, it has a closer resemblance of the zonally asymmetric portion of the SAM.
+
+To the extreme that the leading EOF of geopotential height with the cEOF2 removes results in an annular mode that has none of the typical zonal asymmetries of the observed SAM.\textcolor{red}{No me pareció del todo claro lo que estás queriendo expresar acá}
 
@@
 
-Precipitation anomalies in South America associated with the Imaginary cEOF2 show a clear ENSO-like impact, with positive anomalies in South-Eastern Southamerica, negative anomalies in Southern Brazil and positive anomalies in central Chile.
+Precipitation anomalies in South America associated with the Imaginary cEOF2 show a clear ENSO-like impact, with positive anomalies in South-Eastern South America, negative anomalies in Southern Brazil and positive anomalies in central Chile.
 Precipitation anomalies associated with the Real cEOF2 are low and not statistically significant, showing that the Imaginary phase is optimally aligned with the direction of maximum precipitation impacts.
@@
 
+\textcolor{red}{Debería cerrarse con algún parráfo que diga qué tiene de bueno trabajar con esta metodología y qué permitiría hacer a futuro (se puede poner que sería deseable extenderlo a otras estaciones, etc.)}
+
 ## Code availability {-}

### 2021-05-31 analysis/paper/shceof.Rmd
Mueve todo a wrap por oración.

@@
 Another approach to characterising the Southern Hemisphere circulation is by using Empirical Orthogonal Functions (EOF, also known as Principal Component Analysis).
-Within the EOF framework, the Southern Annular Mode (SAM) appears as the leading mode of variability of the tropospheric Southern Hemisphere circulation [@fogt2020] followed by the two Pacific–South American Patterns (PSA) [@mo2001].
+Within the EOF framework, the Southern Annular Mode (SAM) appears as the leading mode of variability of the tropospheric Southern Hemisphere circulation [@fogt2020] followed by the two Pacific--South American Patterns (PSA) [@mo2001].
 The SAM represents a relatively zonally symmetric pattern of alternating low pressures in polar latitude and a ring of high pressures in high latitudes with an embedded wave 3 pattern that is more prominent in the Pacific sector.
@@
 The inability of a single EOF to describe travelling waves is what forces the PSA pattern to be described by two EOFs.
-@irving2016 characterises the Pacific–South American Pattern in a Fourier framework.
+@irving2016 characterises the Pacific--South American Pattern in a Fourier framework.
 Being a meridionally propagating mode, the PSA cannot be correctly characterised by Fourier decomposition at each latitude circle, so they reprojected meridional wind fields so that the path of the PSA laid on the equator.
 On this new projection, they could identify the PSA using Fourier decomposition.
+
 Finally, another, hybrid, methodology commonly used consist on identifying particular features of interest and creating indices using simple methods such as averages and differences.
@@
 
-We used monthly geopotential height, air temperature, ozone mixing ratio, and Total Ozone Column (TOC) at 2.5\degree\ longitude by 2.5\degree\ latitude of horizontal resolution and 37 vertical isobaric levels from ERA5 [@era5] for the period 1979 to 2019.
+We used monthly geopotential height, air temperature, ozone mixing ratio, and Total Ozone Column (TOC) at 2.5\degree longitude by 2.5\degree latitude of horizontal resolution and 37 vertical isobaric levels from ERA5 [@era5] for the period 1979 to 2019.
 Most of our analysis was restricted to the post-satellite era to avoid any confounding factors arising from the incorporation of satellite observations, but we also used the preliminary back extension of ERA5 from 1950 to 1978 [@era5be] to look at long-term trends.
 Streamfunction was derived from ERA5 vorticity at 200 hPa using the FORTRAN subroutine FISHPACK [@fishpack] and horizontal wave activity fluxes was computed following @plumb1985.
-We usded Sea Surface Temperature (SST) monthly fields from Extended Reconstructed Sea Surface Temperature (ERSST) v5 [@huang2017] and precipitation monthly data from the CPC Merged Analysis of Precipitation [@cmap], with a 2.5\degree\ resolution in latitude and longitude.
+We usded Sea Surface Temperature (SST) monthly fields from Extended Reconstructed Sea Surface Temperature (ERSST) v5 [@huang2017] and precipitation monthly data from the CPC Merged Analysis of Precipitation [@cmap], with a 2.5\degree resolution in latitude and longitude.
 This rainfall gridded dataset is based on information from different sources such as rain gauge observations, satellite inferred estimations and the NCEP-NCAR reanalysis, and it is available since 1979 to present.
@@
 In traditional EOF analysis zonal waves appear as pairs of EOFs, that could be degenerated representing similar patterns but shifted in phase [@horel1984].
-For instance, Figure\ \@ref(fig:eof-naive) shows the leading 4 EOFs of SON geopotential height zonal anomalies at 50 hPa \textcolor{red}{Creo que no aclaras ni acá, ni para los complejos, cuál sería el dominio para el cuál estás calculando el EOF) La barra de colores quedaría mejor si le agregas un contorno que la encierre.}.
+For instance, Figure \@ref(fig:eof-naive) shows the leading 4 EOFs of SON geopotential height zonal anomalies at 50 hPa \textcolor{red}{Creo que no aclaras ni acá, ni para los complejos, cuál sería el dominio para el cuál estás calculando el EOF) La barra de colores quedaría mejor si le agregas un contorno que la encierre.}.
 It is clear that the first two EOFs represent the same zonal wave 1 pattern and the last two represent a same zonal wave pattern with shorter wavenumber and four centers of action shifted by 1/4 wavelength.
@@
 ```
+
 For the first complex principal component, the phase was chosen so that the time series corresponding to the real part has the maximum correlation with the zonal wave 1 of Total Ozone Column between `r knitr::combine_words(LatLabel(o3wave_lats))`.
@@
 For each cEOF, we create regression maps by computing fitting a multiple linear model involving both the real and the imaginary part.
-We evaluated statistical significance using a two-sided t-test and, in the case of regression maps, p-values were adjusted by controlling for the False Discovery Rate [@benjamini1995;@wilks2016] to avoid misleading results from the high number of regressions [@walker1914;@katz1991].
+We evaluated statistical significance using a two-sided t-test and, in the case of regression maps, p-values were adjusted by controlling for the False Discovery Rate [@benjamini1995; @wilks2016] to avoid misleading results from the high number of regressions [@walker1914; @katz1991].
 
@@
 We downloaded data from reanalysis using the ecmwfr package [@hufkens2020] and indices of the ENSO with the rsoi package [@albers2020].
-The paper was rendered using knitr and rmarkdown [@xie2015;@allaire2019].
+The paper was rendered using knitr and rmarkdown [@xie2015; @allaire2019].
 
@@
 The first mode (cEOF1) explains 82% of the variance, while the second mode (cEOF2) explains a smaller fraction (7%).
-In the spatial patterns (Fig.\ \@ref(fig:ceofs-1)), the real and the imaginary components are in quadrature by construction, so that each cEOF describe a single wave-like pattern whose amplitude and position (i.e.
+In the spatial patterns (Fig. \@ref(fig:ceofs-1)), the real and the imaginary components are in quadrature by construction, so that each cEOF describe a single wave-like pattern whose amplitude and position (i.e.
 phase) is controlled by the magnitude and phase of the temporal cEOF.
-The wave patterns described by these cEOFs match the patterns seen in the traditional EOFs of Figure\ \@ref(fig:eof-naive).
-
-The first cEOF (Fig.\ \@ref(fig:eof-naive) column\ a) is a hemispheric wave 1 pattern with maximum amplitude at high latitudes.
+The wave patterns described by these cEOFs match the patterns seen in the traditional EOFs of Figure \@ref(fig:eof-naive).
+
+The first cEOF (Fig. \@ref(fig:eof-naive) column a) is a hemispheric wave 1 pattern with maximum amplitude at high latitudes.
 At 50 hPa the Real cEOF1 has the maximum of the wave 1 at 150ºE and at 200 hPa, the maximum is located at around 175ºE indicating a westerly shift in phase.
-The second cEOF (Fig.\ \@ref(fig:eof-naive) column\ b) shows also a zonal wave-like structure with maximum amplitude at high latitudes, but with shorter spatial scales.
+The second cEOF (Fig. \@ref(fig:eof-naive) column b) shows also a zonal wave-like structure with maximum amplitude at high latitudes, but with shorter spatial scales.
 In particular, its dominant structure at both levels is a wave 3 but with larger amplitude in the pacific sector.
@@
 
-The temporal variability for both components of the cEOF1 have non zero mean (Fig.\ \@ref(fig:ceofs-2)).
+The temporal variability for both components of the cEOF1 have non zero mean (Fig. \@ref(fig:ceofs-2)).
 This is due to the fact that the geopotential fields that enter into the cEOFs algorithm are anomalies with respect to the zonal mean, not the time mean.
@@
 
-Therefore, we extend the cEOFs time series using the preliminary ERA5 back extension going back to 1950 by projecting the older SON geopogential height zonal anomaly fields onto the spatial patterns shown in Figure\ \@ref(fig:ceofs-1).
-There is a significant upward trend in the real component of cEOF1 (Fig.\ \@ref(fig:extended-series)a.1, p-value\ `r trends[part == "Real" & cEOF == "cEOF1"]$p_report`) and no significant trend in any of the complex components of cEOF2.
+Therefore, we extend the cEOFs time series using the preliminary ERA5 back extension going back to 1950 by projecting the older SON geopogential height zonal anomaly fields onto the spatial patterns shown in Figure \@ref(fig:ceofs-1).
+There is a significant upward trend in the real component of cEOF1 (Fig. \@ref(fig:extended-series)a.1, p-value `r trends[part == "Real" & cEOF == "cEOF1"]$p_report`) and no significant trend in any of the complex components of cEOF2.
 The positive trend in the real cEOF1 translates into a positive trend in cEOF1 magnitude, but not in the phase (not shown).
@@
 
-The spatial patterns shown in Figure\ \@ref(fig:ceofs-1) are derived by removing the zonally symmetric circulation, so they might not include all the variability that is actually associated with the cEOF time series.
+The spatial patterns shown in Figure \@ref(fig:ceofs-1) are derived by removing the zonally symmetric circulation, so they might not include all the variability that is actually associated with the cEOF time series.
 To understand the full geopotential height field changes associated with each cEOF, we computed regression patterns of each cEOF with the complete geopotential fields.
@@
 
-Figure\ \@ref(fig:eof1-regr-gh) shows regression maps of SON geopotential height anomalies upon cEOF1.
-At 50 hPa (Figure\ \@ref(fig:eof1-regr-gh) row a), both the Real and Imaginary cEOF1 are associated with planetary wave 1 patterns, that are 90º off phase.
-Their phases coincide with the ones shown in Figure\ \@ref(fig:ceofs-1)a.1, with the positive centre of the real cEOF1 located towards the dateline, and the one of the Imaginary cEOF1 located over Eastern Antarctica.
+Figure \@ref(fig:eof1-regr-gh) shows regression maps of SON geopotential height anomalies upon cEOF1.
+At 50 hPa (Figure \@ref(fig:eof1-regr-gh) row a), both the Real and Imaginary cEOF1 are associated with planetary wave 1 patterns, that are 90º off phase.
+Their phases coincide with the ones shown in Figure \@ref(fig:ceofs-1)a.1, with the positive centre of the real cEOF1 located towards the dateline, and the one of the Imaginary cEOF1 located over Eastern Antarctica.
 However, the Real cEOF1 pattern is substantially altered by the zonally symmetric circulation.
@@
 
-The regression patterns at 200 hPa (Figure\ \@ref(fig:eof1-regr-gh) row b) are similarly influenced by the zonally symmetric circulation.
-It is only possible to distinguish partially the wave 1 pattern in relation with the real cEOF1 (Figure\ \@ref(fig:eof1-regr-gh)b.1).
+The regression patterns at 200 hPa (Figure \@ref(fig:eof1-regr-gh) row b) are similarly influenced by the zonally symmetric circulation.
+It is only possible to distinguish partially the wave 1 pattern in relation with the real cEOF1 (Figure \@ref(fig:eof1-regr-gh)b.1).
 The imaginary cEOF1 shows a much more zonally symmetrical pattern resembling the negative SAM phase.
 
-With the exception of the imaginary cEOF1, it is clear that these patterns are very different than the fully zonally asymmetric versions (Fig.\ \@ref(fig:ceofs-1)), particularly at 200 hPa.
+With the exception of the imaginary cEOF1, it is clear that these patterns are very different than the fully zonally asymmetric versions (Fig. \@ref(fig:ceofs-1)), particularly at 200 hPa.
 Moreover, only in the stratosphere these patterns actually show a distinguishable wave 1 pattern shifted in phase by 90º, suggesting that using the cEOF method is artificially generating a wave structure at 200hPa.
@@
 
-(ref:eof2-regr-gh-cap) Same as Figure\ \@ref(fig:eof1-regr-gh) but for the second cEOF.
+(ref:eof2-regr-gh-cap) Same as Figure \@ref(fig:eof1-regr-gh) but for the second cEOF.
 
@@
 
-Figure\ \@ref(fig:eof2-regr-gh) shows the regression pattern of geopotential height anomalies upon the cEOF2.
-Unlike for cEOF1, in this case the regression patterns are similar to the fully zonally asymmetric patterns from Figure\ \@ref(fig:ceofs-1).
+Figure \@ref(fig:eof2-regr-gh) shows the regression pattern of geopotential height anomalies upon the cEOF2.
+Unlike for cEOF1, in this case the regression patterns are similar to the fully zonally asymmetric patterns from Figure \@ref(fig:ceofs-1).
 Although there are some differences (particularly in 50 hPa), the wave trains identified before are well characterised and patterns associated with the real cEOF2 are 90º out of phase with those associated with the imaginary cEOF2.
@@
 cEOF2 then represents a equivalent barotropic wave train that is very similar to the the Pacific South American Patterns [@mo2001].
-Comparing the location of the positive anomaly near 90ºW in column b of Figure\ \@ref(fig:eof2-regr-gh) with Figures 1.a and b from @mo2001, the real cEOF2 can roughly be identified with PSA2, while the imaginary cEOF2 resembles PSA1.
+Comparing the location of the positive anomaly near 90ºW in column b of Figure \@ref(fig:eof2-regr-gh) with Figures 1.a and b from @mo2001, the real cEOF2 can roughly be identified with PSA2, while the imaginary cEOF2 resembles PSA1.
 
@@
 
-(ref:eof1-regr-t-cap) Same as Figure\ \@ref(fig:eof1-regr-gh) but for air temperature.
+(ref:eof1-regr-t-cap) Same as Figure \@ref(fig:eof1-regr-gh) but for air temperature.
 
@@
 The relation between cEOFs and temperature was also evaluated.
-Figure\ \@ref(fig:eof1-regr-t) shows regression patterns of air temperature at 50hPa and 200hPa onto cEOF1.
-In both levels, the Real cEOF1 is associated with positive monopole over the South Pole with its centre moved slightly towards 150ºE (Fig.\ \@ref(fig:eof1-regr-t) column 1).
+Figure \@ref(fig:eof1-regr-t) shows regression patterns of air temperature at 50hPa and 200hPa onto cEOF1.
+In both levels, the Real cEOF1 is associated with positive monopole over the South Pole with its centre moved slightly towards 150ºE (Fig. \@ref(fig:eof1-regr-t) column 1).
 On the other hand, the regression maps on the Imaginary cEOF1 show a more clear wave 1 pattern with its maximum around 60ºE.
-The distribution of temperature regression coefficients at 50 hPa and at 200 hPa mirror the geopotential height regression maps at 50 hPa (Fig.\ \@ref(fig:eof1-regr-gh)).
-
-Figure \ \@ref(fig:t-vertical) shows the vertical distribution of the regression coefficientes on cEOF1 from zonal anomalies of air temperature and zonal anomalies of ozone mixing ratio averaged between `r knitr::combine_words(LatLabel(o3wave_lats))`.
-\textcolor{red}{Deberias aclarar qué significa la terminología que estas utilizando.
-Lo podes hacer en los métodos y ya queda para todo el texto}.
+The distribution of temperature regression coefficients at 50 hPa and at 200 hPa mirror the geopotential height regression maps at 50 hPa (Fig. \@ref(fig:eof1-regr-gh)).
+
+Figure  \@ref(fig:t-vertical) shows the vertical distribution of the regression coefficientes on cEOF1 from zonal anomalies of air temperature and zonal anomalies of ozone mixing ratio averaged between `r knitr::combine_words(LatLabel(o3wave_lats))`.
+\textcolor{red}{Deberias aclarar qué significa la terminología que estas utilizando. Lo podes hacer en los métodos y ya queda para todo el texto}.
 Temperature zonal anomalies associated with cEOF1 show a clear wave 1 pattern for both real and imaginary components throughout the atmosphere above 250 hPa with a change in sign above 10 hPa.
@@
 
-The maximum ozone anomalies are co-located with the minimum temperature anomalies above 10 hPa and with the maximum temperature anomalies below 10 hPa (Fig.\ \@ref(fig:t-vertical))).
+The maximum ozone anomalies are co-located with the minimum temperature anomalies above 10 hPa and with the maximum temperature anomalies below 10 hPa (Fig. \@ref(fig:t-vertical))).
 Therefore, the ozone zonal wave 1 is anticorrelated with the temperature zonal wave 1 in the upper stratosphere, and directly correlated in the upper stratosphere .
 This change in phase is observed in ozone anomalies forced by planetary waves that reach the stratosphere.
-In the photochemically-dominated upper stratosphere, cold temperatures inhibit the destruction of ozone, and the advectively-dominated lower stratosphere, ozone anomalies are 90º out of phase with horizontal and vertical transport, which is 90º our of phase with temperature anomalies [@hartmann1979;@wirth1993;@smith1995].
+In the photochemically-dominated upper stratosphere, cold temperatures inhibit the destruction of ozone, and the advectively-dominated lower stratosphere, ozone anomalies are 90º out of phase with horizontal and vertical transport, which is 90º our of phase with temperature anomalies [@hartmann1979; @wirth1993; @smith1995].
 
@@
 The regression maps of cEOF1 with fields of Total Ozone Column (TOC) (Fig.
-\ \@ref(fig:o3-regr)) show zonal wave 1 pattens in TOC associated with both phases of cEOF1.
+ \@ref(fig:o3-regr)) show zonal wave 1 pattens in TOC associated with both phases of cEOF1.
 Climatologically, the springtime Ozone minimum is located off the South Pole and towards the Weddell Sea [@wirth1993, \textcolor{red}{¿Alguna referencia más nueva?}].
-Thus, the Real cEOF1 regression pattern (Figure\ \@ref(fig:o3-regr)a) coincides with the climatological position of the ozone hole while the one for the Imaginary cEOF1 is shifted by 90º.
+Thus, the Real cEOF1 regression pattern (Figure \@ref(fig:o3-regr)a) coincides with the climatological position of the ozone hole while the one for the Imaginary cEOF1 is shifted by 90º.
 
@@
 The correlation between the amplitude of both indices is `r wave1_cor[with == "amplitude"]$text` the overall correlation between the zonal wave 1 of ozone and cEOF1 (computed as the mean cosine of the difference in phase) is `r wave_correlation$text`.
-As expected from the location of regressions in Figure\ \@ref(fig:o3-regr), the Real cEOF1 drives the relationship of amplitudes and the Imaginary cEOF1 drives the relationship of phases between the timeseries (not shown).
+As expected from the location of regressions in Figure \@ref(fig:o3-regr), the Real cEOF1 drives the relationship of amplitudes and the Imaginary cEOF1 drives the relationship of phases between the timeseries (not shown).
 
@@
 (ref:sam-eof-vertical-cap) Coefficient of determination between the real and imaginary part of each cEOF and the SAM, Asymmetric SAM (A-SAM) and Symmetric SAM (S-SAM) indices computed at each level according to @campitelli2021.
-Thick lines represent estimates with p-value < 0.01 corrected for False Detection Rate [@benjamini1995].
+Thick lines represent estimates with p-value \< 0.01 corrected for False Detection Rate [@benjamini1995].
 
@@
 
-To explore the relationship between SAM and the modes described from the cEOF, we compute the coefficient of determination between the cEOFs timeseries and the three SAM indices (SAM, A-SAM and S-SAM) at each vertical level (Fig.\ \@ref(fig:sam-eof-vertical)).
+To explore the relationship between SAM and the modes described from the cEOF, we compute the coefficient of determination between the cEOFs timeseries and the three SAM indices (SAM, A-SAM and S-SAM) at each vertical level (Fig. \@ref(fig:sam-eof-vertical)).
 The SAM index is moderately correlated with the Real cEOF1 in all levels, and with the Imaginary cEOF1 and Imaginary cEOF2 in the troposphere.
@@
 In the stratosphere, the Real cEOF1 is correlated with both A-SAM and S-SAM, while the Imaginary cEOF1 is highly correlated only with the A-SAM.
-These correlations are consistent with the regression patterns of geopotential height in Figure\ \@ref(fig:eof1-regr-gh).
+These correlations are consistent with the regression patterns of geopotential height in Figure \@ref(fig:eof1-regr-gh).
 
@@
 In the case of cEOF2, the moderate correlation between SAM and Imaginary cEOF2 becomes extremely high only when the zonally asymmetric variability of the SAM is considered.
-The Imaginary cEOF2 and the A-SAM share up to `r scales::percent(max_r2[cEOF == "cEOF2"]$estimate^2, 2)` of their variance, reached at `r max_r2[cEOF == "cEOF2"]$lev` hPa (Figure\ \@ref(fig:sam-eof-vertical).b2).
+The Imaginary cEOF2 and the A-SAM share up to `r scales::percent(max_r2[cEOF == "cEOF2"]$estimate^2, 2)` of their variance, reached at `r max_r2[cEOF == "cEOF2"]$lev` hPa (Figure \@ref(fig:sam-eof-vertical).b2).
 Such extremely high correlation between the asymmetric SAM and the Imaginary cEOF2 suggests that these might be different ways of characterising the same phenomenon.
 
-<!-- As a further illustration \textcolor{red}{(¿De qué?)}, Figure\ \@ref(fig:eof-filtered) shows the spatial pattern of the leading EOF of 200 hPa geopotential height when the variability associated with cEOF2 is removed.
+<!-- As a further illustration \textcolor{red}{(¿De qué?)}, Figure\ @ref(fig:eof-filtered) shows the spatial pattern of the leading EOF of 200 hPa geopotential height when the variability associated with cEOF2 is removed.
 The resulting pattern is an nearly zonally symmetrical annular mode.
@@
 To this synthetic dataset (\textcolor{red}{No me suena mucho la palabra "synthetic", ¿se usa en este contexto?}) we then apply traditional EOF.
-By construction this dataset contains two independent components: a zonally symmetric standing oscillation and a PSA-like of varying amplitude and phase (Figure\ \@ref(fig:fake-eof) row a).
+By construction this dataset contains two independent components: a zonally symmetric standing oscillation and a PSA-like of varying amplitude and phase (Figure \@ref(fig:fake-eof) row a).
 While the former is a standing oscillation which could be represented by a single EOF, the latter is a wave which can change phase, so it needs two EOFs to be represented correctly.
@@
 
-The three EOFs of the synthetic dataset are shown in Figure\ \@ref(fig:fake-eof) row b.
+The three EOFs of the synthetic dataset are shown in Figure \@ref(fig:fake-eof) row b.
 The synthetic SAM can be seen in the first EOF while the cEOF2 is explained by the second and third.
@@
 The EOF decomposition mixes this mode with parts of the the zonally asymmetric cEOF2 pattern, resulting in a pattern pretty similar to the real SAM pattern (\textcolor{red}{REF}).
-This happens because, although in this synthetic dataset the zonally symmetrical SAM and the cEOF2 are uncorrelated by construction, the sample correlation is not null \textcolor{red}{No me quedo 100% clara esta parte, ¿podrías expresarla un poco distinto?}.
+This happens because, although in this synthetic dataset the zonally symmetrical SAM and the cEOF2 are uncorrelated by construction, the sample correlation is not null \\textcolor{red}{No me quedo 100% clara esta parte, ¿podrías expresarla un poco distinto?}
+.
 In this particular example, it is `r cor_with_eof2[part == "Real"]$text` with the Real part and `r cor_with_eof2[part == "Imaginary"]$text` with the Imaginary part.
@@
 The relationship between PSAs and the modes described from the cEOFs is also studied.
-Table\ \@ref(tab:psa-eof2) shows the correlations between the two PSA indices and the timeseries for Real and Imaginary phase of cEOF2.
-As anticipated by Figure\ \@ref(fig:eof2-regr-gh), there is a strong correlation between PSA1 and Imaginary cEOF2, and between PSA2 and Real cEOF2.
+Table \@ref(tab:psa-eof2) shows the correlations between the two PSA indices and the timeseries for Real and Imaginary phase of cEOF2.
+As anticipated by Figure \@ref(fig:eof2-regr-gh), there is a strong correlation between PSA1 and Imaginary cEOF2, and between PSA2 and Real cEOF2.
 Conversely, there is no relationship between PSA1 and Real cEOF2, and between PSA2 and Imaginary cEOF2.
@@
 \textcolor{red}{No me quedo claro ese "equal", ¿igual de probables/posibles/factibles?}.
-To visualize that,  Figure\ \@ref(fig:phase-histogram) shows an histogram that counts the number of SON years in which the cEOF2 was close to each of the four particular phases (positive/negative Real/Imaginary), with the observations for each season marked as rugs on the horizontal axis.
+To visualize that, Figure \@ref(fig:phase-histogram) shows an histogram that counts the number of SON years in which the cEOF2 was close to each of the four particular phases (positive/negative Real/Imaginary), with the observations for each season marked as rugs on the horizontal axis.
 For instance, years with cEOF2 phase within 45º of 0º are nearest the "positive Real" phase.
@@
 The result obtaneid here is similar to the one obtained by @irving2016.
-From their Fourier decomposition of reprojected meridional wind fields \textcolor{red}{¿En qué región?}, they show that the phase distribution is bimodal and isolate the PSA pattern from the rest of the PSA-like variability by selecting events that are near the peaks of the distribution (compare our Figure\ \@ref(fig:phase-histogram) with their Figure 6).
+From their Fourier decomposition of reprojected meridional wind fields \textcolor{red}{¿En qué región?}, they show that the phase distribution is bimodal and isolate the PSA pattern from the rest of the PSA-like variability by selecting events that are near the peaks of the distribution (compare our Figure \@ref(fig:phase-histogram) with their Figure 6).
 
@@
 The connections between cEOFs and Tropical sources is also assesed.
-Figure\ \@ref(fig:sst-psi-2) shows the regression maps of Sea Surface Temperatures (SST) and Streamfunction at 200 hPa respectively upon cEOF2.
-The Imaginary cEOF2 is associated with strong positive SST anomalies on the Central Pacific and negative anomalies over an area across the North of Australia and New Zealand, the South Pacific Convergence Zone (SPCZ) (Figure\ \@ref(fig:sst-psi-2).a2).
+Figure \@ref(fig:sst-psi-2) shows the regression maps of Sea Surface Temperatures (SST) and Streamfunction at 200 hPa respectively upon cEOF2.
+The Imaginary cEOF2 is associated with strong positive SST anomalies on the Central Pacific and negative anomalies over an area across the North of Australia and New Zealand, the South Pacific Convergence Zone (SPCZ) (Figure \@ref(fig:sst-psi-2).a2).
 This pattern is almost canonically positive ENSO (\textcolor{red}{REF ENSO}) and indeed, the correlation between the Imaginary cEOF2 and the Oceanic Niño Index [@bamston1997] is very high `r enso_cor[cEOF == "cEOF2" & part == "Imaginary"]$text`.
-Besides, the Imaginary cEOF2 is associated with strong wave-like streamfunction anomalies emanating from the tropics (Figure\ \@ref(fig:sst-psi-2).b2).
+Besides, the Imaginary cEOF2 is associated with strong wave-like streamfunction anomalies emanating from the tropics (Figure \@ref(fig:sst-psi-2).b2).
 This is consistent with the effect of ENSO on the extratropics: SST anomalies initiate anomalous convection that excites Rossby waves that propagate meridionally towards higher latitudes [@mo2000].
@@
 Since the Real cEOF2 represents just a different phase of the same wave train, it would be expected to show a similar forcing pattern to the Imaginary cEOF with a slight translation of its location.
-However, Figure\ \@ref(fig:sst-psi-2).a1 and b1 show that the Real cEOF2 is not associated either with any significant SST nor streamfunction anomalies in the tropics.
+However, Figure \@ref(fig:sst-psi-2).a1 and b1 show that the Real cEOF2 is not associated either with any significant SST nor streamfunction anomalies in the tropics.
 The correlation between the Real cEOF2 and ENSO is also not significant (`r enso_cor[cEOF == "cEOF2" & part == "Real"]$text`).
@@
 
-To better explore the relationship between tropical forcing and phase of the cEOF2, Figure\ \@ref(fig:enso-phase) shows the ONI index against the cEOF2 phase for each SON trimester between 1979 and 2019, highlighting years in which the magnitude of cEOF2 is above the median.
+To better explore the relationship between tropical forcing and phase of the cEOF2, Figure \@ref(fig:enso-phase) shows the ONI index against the cEOF2 phase for each SON trimester between 1979 and 2019, highlighting years in which the magnitude of cEOF2 is above the median.
 In years with positive ONI, the phase of the cEOF2 is mostly around +90º (corresponding with positive imaginary part) and vice versa.
 In years with near neutral ENSO, the phase of the cEOF2 is much more variable.
-The black line in Figure\ \@ref(fig:enso-phase) is a sinusoidal fit of the relationship between ONI and cEOF2 phase.
+The black line in Figure \@ref(fig:enso-phase) is a sinusoidal fit of the relationship between ONI and cEOF2 phase.
 The $r^2$ corresponding to the fit is `r signif(summary(cos_model)$r.squared, 2)`, with p-value `r report_p(coef(summary(cos_model))[4])`, indicating a quasi-sinusoidal relation between these two variables.
@@
 
-Furthermore, Figure\ \@ref(fig:enso-phase) suggest that higher amplitude cEOF2 years tend to coincide with stronger ENSO years.
+Furthermore, Figure \@ref(fig:enso-phase) suggest that higher amplitude cEOF2 years tend to coincide with stronger ENSO years.
 The correlation between the absolute magnitude of the ONI and the magnitude of the cEOF2 is `r cor_enso_mag$text`.
@@
 However, when cEOF2 is excited by tropical SST variability, it tends to remain locked to the imaginary phase.
-This explains the relative overabundance of years with cEOF2 near positive and negative imaginary phase in Figure\ \@ref(fig:phase-histogram).
-
-(ref:sst-psi-1-cap) Same as Figure\ \@ref(fig:sst-psi-2) but for cEOF1.
+This explains the relative overabundance of years with cEOF2 near positive and negative imaginary phase in Figure \@ref(fig:phase-histogram).
+
+(ref:sst-psi-1-cap) Same as Figure \@ref(fig:sst-psi-2) but for cEOF1.
 
@@
 
-Figure\ \@ref(fig:sst-psi-1) shows SST and streamfunction regression maps for cEOF1.
+Figure \@ref(fig:sst-psi-1) shows SST and streamfunction regression maps for cEOF1.
 Unlike the cEOF2 case, there is no significant pattern of SST anomalies associated with either the Real or Imaginary cEOF1.
@@
 
-Figure\ \@ref(fig:pp-america) shows regression maps of seasonal precipitation with each cEOF in South America.
-cEOF1 is associated with slight decreases \textcolor{red}{Revisa como estás expresando toda esta parte (y en el resto del texto), recorda que son regresiones así que no corresponde decir postive/negative sino haces alguna referencia a la variable con las que los estás regresionando.}
- in precipitation in Southern Brazil and Paraguay, although these are not statistically significant (Fig.\ \@ref(fig:pp-america) column a).
+Figure \@ref(fig:pp-america) shows regression maps of seasonal precipitation with each cEOF in South America.
+cEOF1 is associated with slight decreases \textcolor{red}{Revisa como estás expresando toda esta parte (y en el resto del texto), recorda que son regresiones así que no corresponde decir postive/negative sino haces alguna referencia a la variable con las que los estás regresionando.} in precipitation in Southern Brazil and Paraguay, although these are not statistically significant (Fig. \@ref(fig:pp-america) column a).
 The strongest precipitation regression coefccients are the ones associated with the Imaginary cEOF2.
@@
 On the other hand, the Real cEOF2 is associated with negative precipitation anomalies in a smaller area of SESA.
-Resembling the relationship between ONI and the phase of cEOF2 (Fig.\ \@ref(fig:enso-phase)), there is a cEOF2 phase dependence of the precipitation anomalies in SESA (not shown).
-These variable could be related wth a signficant sinusoidal fit that has a coefficient of determination of\ `r round(summary(model)$r.squared, 2)` (p-value\ `r report_p(p_value)`).
-
-(ref:pp-oceania-cap) As Figure\ \@ref(fig:pp-america) but for Oceania.
+Resembling the relationship between ONI and the phase of cEOF2 (Fig. \@ref(fig:enso-phase)), there is a cEOF2 phase dependence of the precipitation anomalies in SESA (not shown).
+These variable could be related wth a signficant sinusoidal fit that has a coefficient of determination of `r round(summary(model)$r.squared, 2)` (p-value `r report_p(p_value)`).
+
+(ref:pp-oceania-cap) As Figure \@ref(fig:pp-america) but for Oceania.
 
@@
 
-The precipitation anomalies in Oceania associated with cEOF1 are also weak and not statistically significant (Figure\ \@ref(fig:pp-oceania) column a) whereas EFO2 is associated with large and widespread anomalies.
+The precipitation anomalies in Oceania associated with cEOF1 are also weak and not statistically significant (Figure \@ref(fig:pp-oceania) column a) whereas EFO2 is associated with large and widespread anomalies.
 The Real cEOF2 is associated with positive seasonal anomalies in Northern and Eastern Australia, and negative ones over New Zealand, while precipitation anomalies associated with Imaginary cEOF2 are negative over all Eastern Australia.
@@
 
-(ref:australia-pp-phase-cap) Same as Figure\ \@ref(fig:enso-phase) but for precipitation anomalies averaged between  `r knitr::combine_words(LonLabel(box_australia$lon))` and `r knitr::combine_words(LatLabel(box_australia$lat))` (box shown in Figure\ \@ref(fig:pp-oceania) column b).
+(ref:australia-pp-phase-cap) Same as Figure \@ref(fig:enso-phase) but for precipitation anomalies averaged between `r knitr::combine_words(LonLabel(box_australia$lon))` and `r knitr::combine_words(LatLabel(box_australia$lat))` (box shown in Figure \@ref(fig:pp-oceania) column b).
 
@@
 
-To better understand the relationship between precipitation anomalies over Western Australia and the cEOF2, Figure\ \@ref(fig:australia-pp-phase) plots precipitation anomalies averaged over Eastern Australia (box in Figure\ \@ref(fig:pp-oceania)) as a function of cEOF2 phase.
+To better understand the relationship between precipitation anomalies over Western Australia and the cEOF2, Figure \@ref(fig:australia-pp-phase) plots precipitation anomalies averaged over Eastern Australia (box in Figure \@ref(fig:pp-oceania)) as a function of cEOF2 phase.
 The black line shows a sinusoidal fit, whose coefficients are fitted by weighted least squares using the magnitude of the EFO2 as weight.
@@
 
-(ref:pp-africa-cap) Same as Figure\ \@ref(fig:pp-america) but for South of Africa.
+(ref:pp-africa-cap) Same as Figure \@ref(fig:pp-america) but for South of Africa.
 
@@
 The first complex EOF represents the variability of the zonal wave 1 in the stratosphere, but represents a more zonally symmetric monopole in the troposphere.
-There is a statistically positive trend in the magnitude of this cEOF, which is consistent with previous studies which showed secular changes in springtime wave-1-like patterns [e.g.
-@raphael2003].
+There is a statistically positive trend in the magnitude of this cEOF, which is consistent with previous studies which showed secular changes in springtime wave-1-like patterns [e.g. @raphael2003].
 This mode is closely related to anomalies in Total Ozone Column.
@@
 
-We further show that the Imaginary cEOF2 is closely related to the Southern Annular Mode in the troposphere. In fact, it has a closer resemblance of the zonally asymmetric portion of the SAM.
+We further show that the Imaginary cEOF2 is closely related to the Southern Annular Mode in the troposphere.
+In fact, it has a closer resemblance of the zonally asymmetric portion of the SAM.
 
@@
 
-## Code availability {-}
-
-A version-controlled repository of the code used to create this analysis, including the code used to download the data can be found at https://github.com/eliocamp/shceof.
+## Code availability {.unnumbered}
+
+A version-controlled repository of the code used to create this analysis, including the code used to download the data can be found at <https://github.com/eliocamp/shceof>.
 

### 2021-05-31 analysis/paper/shceof.Rmd
Corrige formato de un comentario

@@
 The large-scale extratropical circulation in the Southern Hemisphere is strongly zonally symmetric, but its zonal departures are highly relevant for surface impacts and could be related to high-impact climate extremes.
-Zonal asymmetries of extratropical circulation in the Southern Hemisphere strongly modulate weather systems and regional cli8 mate through latitudinal transport of heat, humidity, and momentum [@trenberth1980].
+Zonal asymmetries of extratropical circulation in the Southern Hemisphere strongly modulate weather systems and regional climate through latitudinal transport of heat, humidity, and momentum [@trenberth1980].
 
@@
 
-## cEOFs relationship with other variables (Quedo medio rara la divsión con los títulos.
-Me parece que convendrá discutir primero la relación con otros modos de circulación conocidos (SAM, Ozono) y después con las otras variables(Temp, TOC, SST, precip) )
+## cEOFs relationship with other variables
+
+\textcolor{red}{Quedo medio rara la divsión con los títulos.
+Me parece que convendrá discutir primero la relación con otros modos de circulación conocidos (SAM, Ozono) y después con las otras variables(Temp, TOC, SST, precip)}
 

### 2021-05-31 analysis/paper/shceof.Rmd
Agrega referencia más actual
#42
Cambia PSA
(#43)

@@
 One typical way of describing the zonally asymmetric circulation is by the amplitude and phase of zonal waves obtained by Fourier decomposition of geopotential height at each latitude [e.g.
-@vanloon1972;@trenberth1980;\textcolor{red}{Agregar alguna referencia más actual}].
+@vanloon1972;@trenberth1980;@turner2017].
 This approach suggests that zonal waves 1 and 3 explained most of the variance of the tropospheric extratropical Southern Hemisphere circulation (@vanloon1972).
@@
 These indices were obtained by projecting the zonally asymmetric and zonally symmetric part of the SAM spatial pattern onto monthly geopotential height fields, as proposed by @campitelli2021.
-The Pacific South American patterns (PSA1 and PSA2) were calculated as the third and fourth leading EOF of seasonal mean for 500 hPa geopotential height with all seasons together \textcolor{red}{No me quedo claro lo de "all seasons together"} following @mo2001.
+The Pacific South American patterns (PSA1 and PSA2) were calculated as the third and fourth leading EOF of seasonal mean for 500 hPa geopotential height combining all seasons following @mo2001.
 

### 2021-06-01 analysis/paper/shceof.Rmd
Cambia método PSA
Closes #43

@@
 These indices were obtained by projecting the zonally asymmetric and zonally symmetric part of the SAM spatial pattern onto monthly geopotential height fields, as proposed by @campitelli2021.
-The Pacific South American patterns (PSA1 and PSA2) were calculated as the third and fourth leading EOF of seasonal mean for 500 hPa geopotential height combining all seasons following @mo2001.
+The Pacific South American patterns (PSA1 and PSA2) were calculated following @mo2001 as the third and fourth leading EOF of seasonal mean anomalies for SH 500 hPa geopotential height combining all seasons.
 

### 2021-06-01 analysis/paper/shceof.Rmd
Agrega referencia de OLS
Closes #44

@@
 
-Linear trends were computed by Ordinary Least Squares and the 95% confidence interval was computed assuming a t-distribution with the appropriate residual degrees of freedom \textcolor{red}{Referencia?}.
+Linear trends were computed by Ordinary Least Squares and the 95% confidence interval was computed assuming a t-distribution with the appropriate residual degrees of freedom [@wilks2011r].
 The amplitude of the zonal waves was defined through computing the Fourier transform of the spatial field at each latitude circle.

### 2021-06-01 analysis/paper/shceof.Rmd
cEOF -> EOF en figura 1

@@
 
+The amplitude of the zonal waves was defined through computing the Fourier transform of the spatial field at each latitude circle.
 We computed the amplitude and phase of the zonal wave 1 by averaging (area-weighted) variables (temperature, geopotential height, ozone mixing ratio and Total Ozone Column) between `r knitr::combine_words(LatLabel(o3wave_lats))` for each SON and extracting the wave-1 component of the Fourier spectrum.
@@
 Linear trends were computed by Ordinary Least Squares and the 95% confidence interval was computed assuming a t-distribution with the appropriate residual degrees of freedom [@wilks2011r].
-The amplitude of the zonal waves was defined through computing the Fourier transform of the spatial field at each latitude circle.
-\textcolor{red}{Esto está repetido con lo de arriba o querías referirte a otra cosa}
 
@@
 
-(ref:eof-naive-cap) Spatial patterns of the leading 4 EOFs of SON zonal anomalies of geopotential height at 50 hPa for the 1979 -- 2019 period (arbitrary units).
+(ref:eof-naive-cap) Spatial patterns of the leading 4 EOFs of SON zonal anomalies of geopotential height at 50 hPa south of 20\degree S for the 1979 -- 2019 period (arbitrary units).
 
@@
   .[, hgt  := Anomaly(hgt)*sqrt(cos(lat*pi/180)), by = .(time, lev, lat)] %>% 
-  .[, .(eof = list(EOF(hgt ~ time | lon + lat, n = 1:4, data = .SD, suffix = "cEOF"))), 
+  .[, .(eof = list(EOF(hgt ~ time | lon + lat, n = 1:4, data = .SD, suffix = "EOF"))), 
     by = lev]
@@
 explained <- eof_naive[, eof[[1]]$sdev, by = lev] %>% 
-  .[order(lev, cEOF)] %>% 
+  .[order(lev, EOF)] %>% 
   .[, scales::percent(r2, accuracy = .1)]
@@
   theme(axis.text = element_blank()) +
-  facet_wrap(.~cEOF, ncol = 2, labeller = labeller(lev = lab_lev))  +
+  facet_wrap(.~EOF, ncol = 2)  +
   tagger::tag_facets("rc")  +
@@
 In traditional EOF analysis zonal waves appear as pairs of EOFs, that could be degenerated representing similar patterns but shifted in phase [@horel1984].
-For instance, Figure \@ref(fig:eof-naive) shows the leading 4 EOFs of SON geopotential height zonal anomalies at 50 hPa \textcolor{red}{Creo que no aclaras ni acá, ni para los complejos, cuál sería el dominio para el cuál estás calculando el EOF) La barra de colores quedaría mejor si le agregas un contorno que la encierre.}.
+For instance, Figure \@ref(fig:eof-naive) shows the leading 4 EOFs of SON geopotential height zonal anomalies at 50 hPa south of 20\degree S. 
 It is clear that the first two EOFs represent the same zonal wave 1 pattern and the last two represent a same zonal wave pattern with shorter wavenumber and four centers of action shifted by 1/4 wavelength.
@@
 Since each pair of EOFs seem to represent the same phase-varying structure, it would be desirable to combine them into a single index with amplitude and phase.
-
-\textcolor{red}{Detalles para la figura 1: a) Les pusiste cEOF, cuando debería ser EOF; b) te quedaron muy superpuestos los distintos subplots; c) La barra de colores quedaría mejor si le agregas un contorno que la encierre.}
 

### 2021-06-03 analysis/paper/shceof.Rmd
Describe color en tablas

@@
 
+(ref:corr-ceof-splitted-cap) Coefficient of determination ($R^2$) of the absolute magnitude of complex EOFs between 200 hPa and 50 hPa computing EOF separatedly for each level. 
+Background colour indicates the value of $R^2$.
+
 ```{r corr-ceof-splitted}
@@
   setnames("item1", "200 hPa") %>%
-  kbl2(booktabs = TRUE, caption = "Coefficient of determination ($R^2$) of the absolute magnitude of complex EOFs between 200 hPa and 50 hPa computing EOF separatedly for each level.") %>%
+  kbl2(booktabs = TRUE, caption = "(ref:corr-ceof-splitted-cap)") %>%
   column_spec2(2:4, background = spec_color2(.table[[.col]], option = background_palette,
@@
 (ref:psa-eof2-cap) Correlations between the Real and Imaginary parts of cEOF2 and the PSA1 and PSA2 modes computed as the second and third EOFs of seasonal geopotential height anomalies [following @mo2001] for the 1979 -- 2019 period.
-\textcolor{red}{Acclarar lo que está entre parentésis y que sería la coloración.
-Me parece también que hay un error en el del PSA2 imaginario.}
+95% confidence intervals in parenthesis.
+Background colour indicates the value of correlation.
 

### 2021-06-03 analysis/paper/shceof.Rmd
Agrega extensión de series en métodos

@@
 
+While we compute these complex pricipal components using data from 1979 to 2019, we extended the complex time series back to the 1950 -- 1978 period by projecting monthly geopotential height zonal anomalies standardised by level south of 20ºS onto the corresponding spatial patterns.
+
+Each series is computed by projecting monthly geopotential height zonal anomalies standardised by level south of 20ºS onto the corresponding spatial pattern and scaled to zero mean and unit standard deviation.
+
 We performed linear regressions to quantify the association between the cEOFs and other variables.
@@
 (ref:extended-series-cap) Time series extended using ERA5 back extended preliminary edition (period 1950 -- 1978) and ERA5 (period 1979 -- 2019).
-Each series is computed by projecting monthly geopotential height zonal anomalies standardised by level south of 20ºS onto the corresponding spatial pattern and scaled to zero mean and unit standard deviation.
-\textcolor{red}{Esta aclaración metodológica debería estar en la secció de métodos y revisa como expresas este caption para que sea consistente con el resto.}
+Each series is scaled to zero mean and unit standard deviation.
 

### 2021-06-03 analysis/paper/shceof.Rmd
Pequeños detalles

@@
 
-The cEOF methodology is applied to SON geopotential height zonal anomalies at 50 and 200 hPa, representing a stratsopheric and upper troposhperic level respectively.
+The cEOF methodology is applied to SON geopotential height zonal anomalies south of 20\degree S at 50 and 200 hPa, representing a stratsopheric and upper troposhperic level respectively.
 Table \@ref(tab:corr-ceof-splitted) shows the coefficient of determination between time series of the amplitude of each complex EOF (cEOF) across levels.
@@
   ggthemes::geom_rangeframe(sides = "lr", size = 0.1, color = "gray30") +
-  geom_line() +
+  geom_line(aes(color = part)) +
   # geom_line(data = era_less) +
   scale_alpha_manual(values = c(1, 0.2)) +
-  geom_smooth(method = "lm", color = "black") +
+  geom_smooth(method = "lm", aes(color = part)) +
   scale_x_datetime(NULL, date_labels = "%Y") +
   scale_y_continuous("EOF") +
-
+  scale_color_brewer(NULL, palette = "Dark2") +
   facet_grid(cEOF ~ part, scales = "free_y") +
@@
 (ref:eof1-regr-gh-cap) Regression coefficients of the real and imaginary part of the first cEOF on SON geopotential height anomalies for the 1979 -- 2019 period. 
-These coefficients come from multiple linear regression involving the real and imaginary part of both cEFO2.
+These coefficients come from multiple linear regression involving the real and imaginary part of both cEFOs.
 

### 2021-06-03 analysis/paper/shceof.Rmd
Mejora la descripción del método cEOF
Closes  #45

@@
 Complex Empirical Orthogonal Functions (cEOF) is a useful method to characterise zonal waves, considering phase-varying structures [@horel1984]. 
-This method involves augmenting \textcolor{red}{No sé si es un vocabulario específico de usar este método, pero no me queda claro que se entiende por "augmenting" exactamente} the original fields with the Hilbert transform of the data and then computing the Singular Value Decomposition of it.
-In this work, instead of applying the Hilbert transform to the time series at each point, we apply it to each latitude circle at each moment in time \textcolor{red}{Esta aclaración la haces por qué es la manera usual de hacer la metodología, si es asi aclaralo.}.
+This method involves computing EOF on the analytic representation of the original field. 
+This representation is a complex field in which the real part is the original variable and the imaginary part is the original data shifted by 90\degree at each spectral frequency -- i.e.
+its Hilbert transform.
+The Hilbert transform is usually understood in terms of time-varying signal. 
+However, in this work we apply the Hilbert transform to each latitude circle at each moment in time.
 Since each latitude circle is a periodic domain, this procedure does not suffer from edge effects.
+
 The result of the cEOF methodology is a set of complex spatial patterns and complex time series.

### 2021-06-09 analysis/paper/shceof.Rmd
Agrega más detalle de metodología de regresión múltipl
Closes #48

@@
 
-While we compute these complex pricipal components using data from 1979 to 2019, we extended the complex time series back to the 1950 -- 1978 period by projecting monthly geopotential height zonal anomalies standardised by level south of 20ºS onto the corresponding spatial patterns.
-
-Each series is computed by projecting monthly geopotential height zonal anomalies standardised by level south of 20ºS onto the corresponding spatial pattern and scaled to zero mean and unit standard deviation.
+While we compute these complex principal components using data from 1979 to 2019, we extended the complex time series back to the 1950 -- 1978 period by projecting monthly geopotential height zonal anomalies standardised by level south of 20ºS onto the corresponding spatial patterns.
 
@@
 For each cEOF, we create regression maps by computing fitting a multiple linear model involving both the real and the imaginary part.
+To obtain the linear coefficients of a variable $X$ (geopotential, temperature, precipitation, etc...) with the Imaginary and Real parts of each cEOF we fit the equation
+
+$$
+X(\lambda, \phi, t) = \alpha(\lambda, \phi) \operatorname{Im}(\operatorname{cEOF}) + \beta(\lambda, \phi) \operatorname{Re}(\operatorname{cEOF}) + X_0(\lambda, \phi) +  \epsilon(\lambda, \phi, t)
+$$
+
+where $\lambda$ and $\phi$ are the longitude and latitude, $t$ is the time, $\alpha$ and $\beta$ are the linear regression coefficients, $X_0$ and $\epsilon$ are the constant and error terms respectively.
+
 We evaluated statistical significance using a two-sided t-test and, in the case of regression maps, p-values were adjusted by controlling for the False Discovery Rate [@benjamini1995; @wilks2016] to avoid misleading results from the high number of regressions [@walker1914; @katz1991].
@@
     by = .(season, cEOF)] %>%
-  dcast(time ~ cEOF, value.var = c("Real", "Imaginary")) %>%
+  # dcast(time ~ cEOF, value.var = c("Real", "Imaginary")) %>%
   .[melt(era5, id.vars = c("time", "lon", "lat", "lev")), on = "time", allow.cartesian = TRUE] %>%
@@
   # .[variable == "hgt"] %>%
-  .[, FitLm(value, Imaginary_cEOF1, Real_cEOF1, Imaginary_cEOF2, Real_cEOF2, se = TRUE),
-    by = .(lev, lon, lat, variable)] %>%
-  .[, c("term", "cEOF") := tstrsplit(term, "_")] %>%
+  .[, FitLm(value, Imaginary, Real, se = TRUE),
+    by = .(lev, lon, lat, variable, cEOF)] %>%
+  # .[, c("term", "cEOF") := tstrsplit(term, "_")] %>%
   rm_intercept() %>%

### 2021-06-09 analysis/paper/shceof.Rmd
Arregla figura extended-series
#46

@@
                           panel.ontop = FALSE,
+                          panel.grid = element_line(size = 0.5, color = "gray90"),
                           tagger.panel.tag.background = ggplot2::element_rect(color = NA, 
@@
   ggplot(aes(time, hgt)) +
-  geom_hline(yintercept = 0, size = 0.1, color = "gray30") +
   geom_line(aes(color = part)) +
@@
   theme(panel.spacing = grid::unit(1.5, "lines"), strip.placement = "outside")  +
-  ggthemes::geom_rangeframe(sides = "lr", size = 0.1, color = "gray30") +
+
   panel_background
@@
 (ref:extended-series-cap) Time series extended using ERA5 back extended preliminary edition (period 1950 -- 1978) and ERA5 (period 1979 -- 2019).
-Each series is scaled to zero mean and unit standard deviation.
-
-```{r extended-series, fig.cap = "(ref:extended-series-cap)", fig.height=3, fig.width = 7, fig.env = "figure*"}
+Dark straight line is the OLS trend.
+Black horizontal and vertical line mark the mean value and range of each time series respectively.
+
+```{r extended-series, fig.cap = "(ref:extended-series-cap)", fig.height=4, fig.width = 7, fig.env = "figure*"}
 # Check that the extended index coincides with the
@@
   geom_hline(data = ~.x[, mean(hgt)*1e3, by  = .(cEOF, part)], 
-             aes(yintercept = V1), size = 0.1, color = "gray30") +
-  ggthemes::geom_rangeframe(sides = "lr", size = 0.1, color = "gray30") +
+             aes(yintercept = V1), size = 0.2, color = "gray30") +
+  ggthemes::geom_rangeframe(sides = "r", size = 0.2, color = "gray30") +
   geom_line(aes(color = part)) +
@@
   scale_alpha_manual(values = c(1, 0.2)) +
-  geom_smooth(method = "lm", aes(color = part)) +
+  geom_smooth(method = "lm", aes(color = stage(part, after_scale = scales::muted(color))),
+              show.legend = FALSE, size = 0.8) +
   scale_x_datetime(NULL, date_labels = "%Y") +
-  scale_y_continuous("EOF") +
-  scale_color_brewer(NULL, palette = "Dark2") +
-  facet_grid(cEOF ~ part, scales = "free_y") +
+  scale_y_continuous("EOF", minor_breaks = NULL) +
+  scale_color_brewer(NULL, palette = "Dark2", 
+                     guide = guide_legend(override.aes = list(fill = NA))) +
+  facet_grid(cEOF ~ part) +
   theme(panel.spacing = grid::unit(1.5, "lines"))  +
@@
 
-\textcolor{red}{Revisa la figura 4 que te están quedando bastante desparejos los subplots.
-También sugieron mantener la convención de colores de la figura anterior (real/imaginario) para hacer más directa la interpretación.}
-
 This long-term change indicates an increase in the magnitude of the high latitude zonal wave 1.
@@
 The spatial patterns shown in Figure \@ref(fig:ceofs-1) are derived by removing the zonally symmetric circulation, so they might not include all the variability that is actually associated with the cEOF time series.
-To understand the full geopotential height field changes associated with each cEOF, we computed regression patterns of each cEOF with the complete geopotential fields.
+To understand the full geopotential height field changes associated with each cEOF, we computed regression patterns of the full geopotential fields onto the 
+of each cEOF with the complete geopotential fields.
 
@@
 
-  scale_x_level() +
+  scale_x_level(minor_breaks = NULL) +
   scale_y_continuous(r2, limits = c(0, 1)) +

### 2021-06-09 analysis/paper/shceof.Rmd
Elimina la Figura ceof-2
Closes #46

@@
 
-```{r ceofs-2, fig.cap = "Time series of the leading two cEOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2019.", fig.height=width_column, fig.width = width_column}
-ceof %>%
-  .[, eof[[1]]$left, by = season] %>%
-  sep_ReIm(hgt) %>%
-  .[season == "SON"] %>%
-  # .[, hgt := scale(hgt), by = .(part, cEOF)] %>%
-  ggplot(aes(time, hgt)) +
-  geom_line(aes(color = part)) +
-  scale_x_datetime(NULL) +
-  scale_y_continuous(NULL) +
-  scale_color_brewer(NULL, palette = "Dark2") +
-  facet_grid(cEOF ~ ., switch = "y") +
-  theme(panel.spacing = grid::unit(1.5, "lines"), strip.placement = "outside")  +
-
-  panel_background
-```
-
-To describe the variability of the circulation zonal anomalies, the spatial and temporal parts of the first two leading cEOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa, compueted jointly at both levels, are shown in Figures \@ref(fig:ceofs-1) and \@ref(fig:ceofs-2).
-The first mode (cEOF1) explains 82% of the variance, while the second mode (cEOF2) explains a smaller fraction (7%).
-In the spatial patterns (Fig. \@ref(fig:ceofs-1)), the real and the imaginary components are in quadrature by construction, so that each cEOF describe a single wave-like pattern whose amplitude and position (i.e.
-phase) is controlled by the magnitude and phase of the temporal cEOF.
-The wave patterns described by these cEOFs match the patterns seen in the traditional EOFs of Figure \@ref(fig:eof-naive).
-
-The first cEOF (Fig. \@ref(fig:eof-naive) column a) is a hemispheric wave 1 pattern with maximum amplitude at high latitudes.
-At 50 hPa the Real cEOF1 has the maximum of the wave 1 at 150ºE and at 200 hPa, the maximum is located at around 175ºE indicating a westerly shift in phase.
-The second cEOF (Fig. \@ref(fig:eof-naive) column b) shows also a zonal wave-like structure with maximum amplitude at high latitudes, but with shorter spatial scales.
-In particular, its dominant structure at both levels is a wave 3 but with larger amplitude in the pacific sector.
-This modulated amplitude is especially apparent at 200 hPa.
-There is no apparent phase shift, suggesting a rather barotropic structure.
-
-The temporal variability for both components of the cEOF1 have non zero mean (Fig. \@ref(fig:ceofs-2)).
-This is due to the fact that the geopotential fields that enter into the cEOFs algorithm are anomalies with respect to the zonal mean, not the time mean.
-As a consequence, the variability associated with the cEOF1 includes variability that projects onto the mean zonally anomalous field.
-There are no significant simultanous correlation between the time series.
-Both cEOFs show year-to-year variability but show no evidence of decadal variabilty.
-The relatively short timespan (`r diff(range(unique(year(era5$time))))` years) hampers our habilty to detect longer trends.
-
 ```{r era_more}
@@
 
-Therefore, we extend the cEOFs time series using the preliminary ERA5 back extension going back to 1950 by projecting the older SON geopogential height zonal anomaly fields onto the spatial patterns shown in Figure \@ref(fig:ceofs-1).
+To describe the variability of the circulation zonal anomalies, the spatial and temporal parts of the first two leading cEOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa, computed jointly at both levels, are shown in Figures \@ref(fig:ceofs-1) and \@ref(fig:extended-series,).
+The first mode (cEOF1) explains 82% of the variance, while the second mode (cEOF2) explains a smaller fraction (7%).
+In the spatial patterns (Fig. \@ref(fig:ceofs-1)), the real and the imaginary components are in quadrature by construction, so that each cEOF describe a single wave-like pattern whose amplitude and position (i.e.
+phase) is controlled by the magnitude and phase of the temporal cEOF.
+The wave patterns described by these cEOFs match the patterns seen in the traditional EOFs of Figure \@ref(fig:eof-naive).
+
+The first cEOF (Fig. \@ref(fig:eof-naive) column a) is a hemispheric wave 1 pattern with maximum amplitude at high latitudes.
+At 50 hPa the Real cEOF1 has the maximum of the wave 1 at 150ºE and at 200 hPa, the maximum is located at around 175ºE indicating a westerly shift in phase.
+The second cEOF (Fig. \@ref(fig:eof-naive) column b) shows also a zonal wave-like structure with maximum amplitude at high latitudes, but with shorter spatial scales.
+In particular, its dominant structure at both levels is a wave 3 but with larger amplitude in the pacific sector.
+This modulated amplitude is especially apparent at 200 hPa.
+There is no apparent phase shift, suggesting a rather barotropic structure.
+
+The temporal variability for both components of the cEOF1 have non zero mean (Fig. \@ref(fig:extended-series,)).
+This is due to the fact that the geopotential fields that enter into the cEOFs algorithm are anomalies with respect to the zonal mean, not the time mean.
+As a consequence, the variability associated with the cEOF1 includes variability that projects onto the mean zonally anomalous field.
+There are no significant simultaneous correlation between the time series.
+Both cEOFs show year-to-year variability but show no evidence of decadal variability.
+
 There is a significant upward trend in the real component of cEOF1 (Fig. \@ref(fig:extended-series)a.1, p-value `r trends[part == "Real" & cEOF == "cEOF1"]$p_report`) and no significant trend in any of the complex components of cEOF2.

### 2021-06-09 analysis/paper/shceof.Rmd
Mejora lenguaje de regresiones y tags en los epígrafes
#32 y #26

@@
 The spatial patterns shown in Figure \@ref(fig:ceofs-1) are derived by removing the zonally symmetric circulation, so they might not include all the variability that is actually associated with the cEOF time series.
-To understand the full geopotential height field changes associated with each cEOF, we computed regression patterns of the full geopotential fields onto the 
-of each cEOF with the complete geopotential fields.
+To understand the full geopotential height field changes associated with each cEOF, we computed regression patterns of the full geopotential fields onto each cEOF.
 
@@
 
-(ref:eof1-regr-gh-cap) Regression coefficients of the real and imaginary part of the first cEOF on SON geopotential height anomalies for the 1979 -- 2019 period. 
-These coefficients come from multiple linear regression involving the real and imaginary part of both cEFOs.
+(ref:eof1-regr-gh-cap) Regression of SON geopotential height anomalies (meters) with the (column 1) real and (column 2) imaginary parts of the first cEOF for the 1979 -- 2019 period at (row a) 50 hPa and (row b) 200 hPa.
+These coefficients come from multiple linear regression involving the real and imaginary parts. 
+Areas marked with dots have p-values smaller than 0.05 adjusted for False Detection Rate.
 
@@
 
-(ref:t-vertical-cap) Regression coefficients between cEOF1 and zonal anomalies of mean air temperature (shaded) and ozone mixing ratio (contours, negative contours with dashed lines, labels in parts per billion by mass) averaged between `r knitr::combine_words(LatLabel(o3wave_lats))` for the period 1979 -- 2019.
+(ref:t-vertical-cap) Regression of SON anomalies of mean air temperature (shaded, Kelvin) and ozone mixing ratio (contours, negative contours with dashed lines, labels in parts per billion by mass) averaged between `r knitr::combine_words(LatLabel(o3wave_lats))` with the (a) real and (b) imaginary parts of the first cEOF for the 1979 -- 2019 period.
 
@@
   axis_labs_smol +
-  coord_cartesian(clip = "off") 
+  coord_cartesian(clip = "off") +
+  tag_facets()
 ```
@@
 
-(ref:o3-regr-cap) Regression coefficients (shaded) of SON mean Total Ozone Column anomalies onto the real (a) and imaginary (b) components of the cEOF1 for the 1979 -- 2019 period.
-On contours, the mean zonal anomaly of Total OZone Column (negative contours in dashed lines, Dobson Units).
+(ref:o3-regr-cap) Regression of SON mean Total Ozone Column anomalies (shaded, Dobson Units) with the (a) real and (b) imaginary parts of the first cEOF for the 1979 -- 2019 period.
+On contours, the mean zonal anomaly of Total Ozone Column (negative contours in dashed lines, Dobson Units).
+Areas marked with dots have p-values smaller than 0.05 adjusted for False Detection Rate.
 
@@
 For the 1979 -- 2018 \textcolor{red}{¿Cambiaste el año o es un typo?} period.
-Black contours with dots indicate areas with p-value smaller than 0.05 controlling for False Detection Rate.
+Areas marked with dots have p-values smaller than 0.05 adjusted for False Detection Rate.
 

### 2021-06-14 analysis/paper/shceof.Rmd
Mueve sección de si existe el SAM

@@
 ```
-
-\textcolor{red}{Creo que te había hecho el mismo comentario en la versión anterior.
-Tenés que ser más preciso cuando hablas/mostras regresiones indicando en qué dirección va la regresión "regressed upon".
-Revisalo a lo largo del texto.
-En la figura 5, ¿no está al revés?}
-
-\textcolor{red}{En el caption de la Figura 5 falta aclarar qué son los puntos.}
 
@@
 
-Figure  \@ref(fig:t-vertical) shows the vertical distribution of the regression coefficientes on cEOF1 from zonal anomalies of air temperature and zonal anomalies of ozone mixing ratio averaged between `r knitr::combine_words(LatLabel(o3wave_lats))`.
+Figure \@ref(fig:t-vertical) shows the vertical distribution of the regression coefficients on cEOF1 from zonal anomalies of air temperature and zonal anomalies of ozone mixing ratio averaged between `r knitr::combine_words(LatLabel(o3wave_lats))`.
 \textcolor{red}{Deberias aclarar qué significa la terminología que estas utilizando. Lo podes hacer en los métodos y ya queda para todo el texto}.
@@
 In the photochemically-dominated upper stratosphere, cold temperatures inhibit the destruction of ozone, and the advectively-dominated lower stratosphere, ozone anomalies are 90º out of phase with horizontal and vertical transport, which is 90º our of phase with temperature anomalies [@hartmann1979; @wirth1993; @smith1995].
-
-\textcolor{red}{Revisa que todos los captions tengan las unidades correspondientes}
 
@@
 
-<!-- As a further illustration \textcolor{red}{(¿De qué?)}, Figure\ @ref(fig:eof-filtered) shows the spatial pattern of the leading EOF of 200 hPa geopotential height when the variability associated with cEOF2 is removed.
-The resulting pattern is an nearly zonally symmetrical annular mode.
-This mode is highly correlated with SAM (`r eof_filtered_cor[term == "full"]$text`), but only with the symmetrical part (`r eof_filtered_cor[term == "sym"]$text`).
-Note that the usual definition of the PSA modes as the second and third EOFs creates modes orthogonal to the SAM (defined as the first EOF) and thus impedes this kind of filtering.
--->
-
-(ref:fake-eof-cap) Row a: The three components used to create a synthetic dataset of 41 years of geopotential height at 200 hPa.
-Row b: The three EOF computed from the synthetic dataset with the variance explained by each EOF in parenthesis.
-Units are arbitrary.
-
-```{r fake_eof}
-cEOF2 <- ceof$eof[[1]]$left[cEOF == "cEOF2"] %>% 
-  copy() %>% 
-  .[, cEOF := NULL] %>% 
-  .[]
-
-pattern <- eof_filtered$right %>% 
-  .[era5[lev == 200], on = "time"] %>% 
-  .[, FitLm(hgt, eof), by = .(lon, lat)] %>% 
-  .[, estimate := mean(estimate), by = .(lat, term)] %>%
-  dcast(lon + lat ~ term, value.var = "estimate")
-
-set.seed(42)
-fake_sam <- eof_filtered$right %>% 
-  copy() %>% 
-  .[, eof := sample(eof, replace = TRUE)]
-
-cor_with_eof2 <- fake_sam[cEOF2, on = "time"] %>% 
-  sep_ReIm() %>% 
-  .[, correlate(eof, hgt), by = part]
-
-fake_sam <- fake_sam %>% 
-  setnames("eof", "value") %>% 
-  .[, pattern[, .(sam = value*eof), by = .(lon, lat)], by = time]
-
-# fake_sam %>% 
-#   .[time %in% unique(time)[1:4]] %>% 
-#   ggplot(aes(lon, lat)) +
-#   geom_contour_fill(aes(z = sam)) +
-#   facet_wrap(~factor(time)) +
-#   coord_polar()
-
-fake_eof <- era5 %>% 
-  .[sep_ReIm(cEOF2, format = "wide")[, hgt := NULL], on = .NATURAL] %>% 
-  .[lev == 200] %>% 
-  .[lat <= -20] %>% 
-  .[season(time) == "SON"] %>% 
-  .[, hgt := hgt - .lm.fit(cbind(1, Imaginary, Real), hgt)$residuals,
-    .(lon, lat, lev, season(time))] %>% 
-  .[, hgt := Anomaly(hgt), by = .(lat)] %>% 
-  fake_sam[., on = .NATURAL] %>% 
-  .[, hgt := hgt + sam] %>% 
-  .[, eof := Anomaly(hgt)*sqrt(cos(lat*pi/180)), by = .(lon, lat, season(time))] %>% 
-  EOF(eof ~ lon + lat | time, n = 1:3, data = ., suffix = "EOF")  
-```
-
-```{r fake-eof, fig.cap = "(ref:fake-eof-cap)", fig.width=3.7, fig.height=3.3}
-north_thumb <- function(sdev, n){
-  sdev <- sdev^2
-  delta <- sqrt(2 / n) * sdev
-
-  # Separation between each pair of eofs must be 
-  # greater than the sum of their errors.
-  separation <- abs(diff(sdev))
-  separation >  delta[-length(delta)] + delta[-1]
-}
-
-separated <- north_thumb(fake_eof$sdev$sd^2, 41)
-stopifnot(all(separated))
-
-var <- fake_eof$sdev %>%
-  copy() %>% 
-    .[, setNames(paste0(EOF, " (", scales::percent(r2), ")"),
-                 EOF)]
-var <- c(var, setNames(c("SAM", "Imaginary", "Real"), c("SAM", "Imaginary", "Real")))
-
-era5 %>%
-  .[sep_ReIm(cEOF2, format = "wide")[, hgt := NULL], on = .NATURAL] %>%
-  .[lev == 200] %>%
-
-  .[season(time) == "SON"] %>%
-  .[, FitLm(hgt, Real, Imaginary), .(lon, lat)] %>%
-  rm_intercept() %>%
-  .[, .(lon, lat, eof = estimate, EOF = term)] %>%
-  rbind(pattern[, .(lon, lat, eof, EOF = "SAM")]) %>%
-  rbind(fake_eof$left, use.names = TRUE) %>% 
-   .[lat <= -20] %>%
-  .[, EOF := factor(EOF, c("SAM", "Imaginary", "Real", paste0("EOF", 1:3)),
-                    ordered = TRUE)] %>% 
-  .[EOF %in% c("EOF1", "EOF3"), eof := -eof] %>% 
-  .[!(EOF %in% c("SAM", "Imaginary", "Real")), eof := eof/sd(eof)] %>% 
-  .[EOF %in% c("SAM", "Imaginary", "Real"), eof := eof/sd(eof)] %>%
-
-  # geom_contour_tanaka has problem with the perfectly symemtric SAM
-  # adding just a tiny bit of noise solves it. 
-  .[EOF == "SAM", eof := eof + rnorm(.N, sd = 1e-5)] %>%
-
-  periodic(lon = c(0, 360)) %>%
-  ggplot(aes(lon, lat)) +
-  geom_contour_fill(aes(z = eof, fill = ..level..), global.breaks = TRUE, 
-                    breaks = AnchorBreaks(0, 0.5, exclude = 0)) +
-
-  geom_contour_tanaka(aes(z = eof), global.breaks = TRUE, 
-                      breaks = AnchorBreaks(0,0.5,  exclude = 0)) +
-  scale_fill_divergent_discretised(NULL, guide = guide_colorsteps_bottom(), 
-                                   limits = c(-4, 4)) +
-  geom_qmap(~.x[lat <= -20]) +
-  scale_x_longitude() +
-    scale_y_latitude(labels = NULL,
-                   minor_breaks = NULL) +
-  grid_y +
-  coord_polar() +
-  facet_wrap(~EOF, ncol = 3, labeller = labeller(EOF = var)) +
-  axis_labs_smol +
-  theme(axis.text = element_blank()) +
-  tag_facets("rc")
-```
-
-\textcolor{red}{Esta parte creo que quedaría mejor como una subsección, aunque no se me viene rápido a la cabeza el título y la pondría después de la parte de PSA}
-
-To test the hypothesis that a PSA-like feature can appear as statistic contamination of an otherwise zonally symmetrical and uncorrelated SAM, we create a synthetic dataset of `r uniqueN(fake_sam$time)` fields by reconstructing the SON 200 hPa geopotential height field explained by the cEOF2 and adding a perfectly zonally symmetric SAM-like structure with random sign and value sampled with replacement from the real SAM index.
-To this synthetic dataset (\textcolor{red}{No me suena mucho la palabra "synthetic", ¿se usa en este contexto?}) we then apply traditional EOF.
-By construction this dataset contains two independent components: a zonally symmetric standing oscillation and a PSA-like of varying amplitude and phase (Figure \@ref(fig:fake-eof) row a).
-While the former is a standing oscillation which could be represented by a single EOF, the latter is a wave which can change phase, so it needs two EOFs to be represented correctly.
-Thus, the full variance of this synthetic dataset can be perfectly explained by three EOFs.
-
-\textcolor{red}{Los titulos de la figura 11 superior se prestan a la confusión, ya que no es exactamente el SAM y no se entiende qué sería "Imaginary" y "Real"}
-
-The three EOFs of the synthetic dataset are shown in Figure \@ref(fig:fake-eof) row b.
-The synthetic SAM can be seen in the first EOF while the cEOF2 is explained by the second and third.
-However, even though the synthetic SAM is perfectly zonally symmetric by construction, the leading EOF is not.
-The EOF decomposition mixes this mode with parts of the the zonally asymmetric cEOF2 pattern, resulting in a pattern pretty similar to the real SAM pattern (\textcolor{red}{REF}).
-This happens because, although in this synthetic dataset the zonally symmetrical SAM and the cEOF2 are uncorrelated by construction, the sample correlation is not null \\textcolor{red}{No me quedo 100% clara esta parte, ¿podrías expresarla un poco distinto?}
-.
-In this particular example, it is `r cor_with_eof2[part == "Real"]$text` with the Real part and `r cor_with_eof2[part == "Imaginary"]$text` with the Imaginary part.
-Of note, the three EOFs are well separated according to North's rule of thumb [@north1982].
-
-This suggests that the zonal asymmetries of the SON SAM could very well be mostly a statistical artefact.
-Since many surface impacts are mediated by the asymmetric component, as well as the relationship between SAM and ENSO [@campitelli2021], this potential issue could affect the interpretability of many results involving the SAM.
-Studies on the relationship between the SAM and PSA-like patterns would be particularly difficult to interpret.
-
 ### PSA
@@
 
+### Is the Asymmetric SAM real?
+
+(ref:fake-eof-cap) Row a: The three components used to create a synthetic dataset of 41 years of geopotential height at 200 hPa.
+Row b: The three EOF computed from the synthetic dataset with the variance explained by each EOF in parenthesis.
+Units are arbitrary.
+
+```{r make_fake_eof}
+cEOF2 <- ceof$eof[[1]]$left[cEOF == "cEOF2"] %>% 
+  copy() %>% 
+  .[, cEOF := NULL] %>% 
+  .[]
+
+pattern <- eof_filtered$right %>% 
+  .[era5[lev == 200], on = "time"] %>% 
+  .[, FitLm(hgt, eof), by = .(lon, lat)] %>% 
+  .[, estimate := mean(estimate), by = .(lat, term)] %>%
+  dcast(lon + lat ~ term, value.var = "estimate")
+
+set.seed(42)
+fake_sam <- eof_filtered$right %>% 
+  copy() %>% 
+  .[, eof := sample(eof, replace = TRUE)]
+
+cor_with_eof2 <- fake_sam[cEOF2, on = "time"] %>% 
+  sep_ReIm() %>% 
+  .[, correlate(eof, hgt), by = part]
+
+fake_sam <- fake_sam %>% 
+  setnames("eof", "value") %>% 
+  .[, pattern[, .(sam = value*eof), by = .(lon, lat)], by = time]
+
+# fake_sam %>% 
+#   .[time %in% unique(time)[1:4]] %>% 
+#   ggplot(aes(lon, lat)) +
+#   geom_contour_fill(aes(z = sam)) +
+#   facet_wrap(~factor(time)) +
+#   coord_polar()
+
+fake_eof <- era5 %>% 
+  .[sep_ReIm(cEOF2, format = "wide")[, hgt := NULL], on = .NATURAL] %>% 
+  .[lev == 200] %>% 
+  .[lat <= -20] %>% 
+  .[season(time) == "SON"] %>% 
+  .[, hgt := hgt - .lm.fit(cbind(1, Imaginary, Real), hgt)$residuals,
+    .(lon, lat, lev, season(time))] %>% 
+  .[, hgt := Anomaly(hgt), by = .(lat)] %>% 
+  fake_sam[., on = .NATURAL] %>% 
+  .[, hgt := hgt + sam] %>% 
+  .[, eof := Anomaly(hgt)*sqrt(cos(lat*pi/180)), by = .(lon, lat, season(time))] %>% 
+  EOF(eof ~ lon + lat | time, n = 1:3, data = ., suffix = "EOF")  
+```
+
+```{r fake-eof, fig.cap = "(ref:fake-eof-cap)", fig.width=3.7, fig.height=3.3}
+north_thumb <- function(sdev, n){
+  sdev <- sdev^2
+  delta <- sqrt(2 / n) * sdev
+
+  # Separation between each pair of eofs must be 
+  # greater than the sum of their errors.
+  separation <- abs(diff(sdev))
+  separation >  delta[-length(delta)] + delta[-1]
+}
+
+separated <- north_thumb(fake_eof$sdev$sd^2, 41)
+stopifnot(all(separated))
+
+var <- fake_eof$sdev %>%
+  copy() %>% 
+    .[, setNames(paste0(EOF, " (", scales::percent(r2), ")"),
+                 EOF)]
+var <- c(var, setNames(c("SAM", "Imaginary", "Real"), c("SAM", "Imaginary", "Real")))
+
+era5 %>%
+  .[sep_ReIm(cEOF2, format = "wide")[, hgt := NULL], on = .NATURAL] %>%
+  .[lev == 200] %>%
+
+  .[season(time) == "SON"] %>%
+  .[, FitLm(hgt, Real, Imaginary), .(lon, lat)] %>%
+  rm_intercept() %>%
+  .[, .(lon, lat, eof = estimate, EOF = term)] %>%
+  rbind(pattern[, .(lon, lat, eof, EOF = "SAM")]) %>%
+  rbind(fake_eof$left, use.names = TRUE) %>% 
+   .[lat <= -20] %>%
+  .[, EOF := factor(EOF, c("SAM", "Imaginary", "Real", paste0("EOF", 1:3)),
+                    ordered = TRUE)] %>% 
+  .[EOF %in% c("EOF1", "EOF3"), eof := -eof] %>% 
+  .[!(EOF %in% c("SAM", "Imaginary", "Real")), eof := eof/sd(eof)] %>% 
+  .[EOF %in% c("SAM", "Imaginary", "Real"), eof := eof/sd(eof)] %>%
+
+  # geom_contour_tanaka has problem with the perfectly symemtric SAM
+  # adding just a tiny bit of noise solves it. 
+  .[EOF == "SAM", eof := eof + rnorm(.N, sd = 1e-5)] %>%
+
+  periodic(lon = c(0, 360)) %>%
+  ggplot(aes(lon, lat)) +
+  geom_contour_fill(aes(z = eof, fill = ..level..), global.breaks = TRUE, 
+                    breaks = AnchorBreaks(0, 0.5, exclude = 0)) +
+
+  geom_contour_tanaka(aes(z = eof), global.breaks = TRUE, 
+                      breaks = AnchorBreaks(0,0.5,  exclude = 0)) +
+  scale_fill_divergent_discretised(NULL, guide = guide_colorsteps_bottom(), 
+                                   limits = c(-4, 4)) +
+  geom_qmap(~.x[lat <= -20]) +
+  scale_x_longitude() +
+    scale_y_latitude(labels = NULL,
+                   minor_breaks = NULL) +
+  grid_y +
+  coord_polar() +
+  facet_wrap(~EOF, ncol = 3, labeller = labeller(EOF = var)) +
+  axis_labs_smol +
+  theme(axis.text = element_blank()) +
+  tag_facets("rc")
+```
+
+To test the hypothesis that a PSA-like feature can appear as statistic contamination of an otherwise zonally symmetrical and uncorrelated SAM, we create a synthetic dataset of `r uniqueN(fake_sam$time)` fields by reconstructing the SON 200 hPa geopotential height field explained by the cEOF2 and adding a perfectly zonally symmetric SAM-like structure with random sign and value sampled with replacement from the real SAM index.
+To this synthetic dataset (\textcolor{red}{No me suena mucho la palabra "synthetic", ¿se usa en este contexto?}) we then apply traditional EOF.
+By construction this dataset contains two independent components: a zonally symmetric standing oscillation and a PSA-like of varying amplitude and phase (Figure \@ref(fig:fake-eof) row a).
+While the former is a standing oscillation which could be represented by a single EOF, the latter is a wave which can change phase, so it needs two EOFs to be represented correctly.
+Thus, the full variance of this synthetic dataset can be perfectly explained by three EOFs.
+
+\textcolor{red}{Los titulos de la figura 11 superior se prestan a la confusión, ya que no es exactamente el SAM y no se entiende qué sería "Imaginary" y "Real"}
+
+The three EOFs of the synthetic dataset are shown in Figure \@ref(fig:fake-eof) row b.
+The synthetic SAM can be seen in the first EOF while the cEOF2 is explained by the second and third.
+However, even though the synthetic SAM is perfectly zonally symmetric by construction, the leading EOF is not.
+The EOF decomposition mixes this mode with parts of the the zonally asymmetric cEOF2 pattern, resulting in a pattern pretty similar to the real SAM pattern (\textcolor{red}{REF}).
+This happens because, although in this synthetic dataset the zonally symmetrical SAM and the cEOF2 are uncorrelated by construction, the sample correlation is not null \\textcolor{red}{No me quedo 100% clara esta parte, ¿podrías expresarla un poco distinto?}
+.
+In this particular example, it is `r cor_with_eof2[part == "Real"]$text` with the Real part and `r cor_with_eof2[part == "Imaginary"]$text` with the Imaginary part.
+Of note, the three EOFs are well separated according to North's rule of thumb [@north1982].
+
+This suggests that the zonal asymmetries of the SON SAM could very well be mostly a statistical artefact.
+Since many surface impacts are mediated by the asymmetric component, as well as the relationship between SAM and ENSO [@campitelli2021], this potential issue could affect the interpretability of many results involving the SAM.
+Studies on the relationship between the SAM and PSA-like patterns would be particularly difficult to interpret.
 ## Tropical sources
@@
 (ref:sst-psi-2-cap) Regression maps of cEOF2 with SST (column a) and streamfunction zonal anomalies (shaded) with their corresponding activity wave flux (vectors) (column b).
-Areas marked with dots have p-values smaller than 0.05 adjusted for FDR.
+Areas marked with dots have p-values smaller than 0.01 adjusted for FDR.
 

### 2021-06-15 analysis/paper/shceof.Rmd
Varios camios

@@
  \@ref(fig:o3-regr)) show zonal wave 1 pattens in TOC associated with both phases of cEOF1.
-Climatologically, the springtime Ozone minimum is located off the South Pole and towards the Weddell Sea [@wirth1993, \textcolor{red}{¿Alguna referencia más nueva?}].
+Climatologically, the springtime Ozone minimum is located off the South Pole and towards the Weddell Sea [@grytsai2011].
 Thus, the Real cEOF1 regression pattern (Figure \@ref(fig:o3-regr)a) coincides with the climatological position of the ozone hole while the one for the Imaginary cEOF1 is shifted by 90º.
@@
 
-The reason the conventional EOF analysis arrives at the same separation than our particular cEOF rotation is probably the fact that not all phases are equal.
-\textcolor{red}{No me quedo claro ese "equal", ¿igual de probables/posibles/factibles?}.
+The reason the conventional EOF analysis arrives at the same separation than our particular cEOF rotation is probably the fact that not all phases are equally frequent.
 To visualize that, Figure \@ref(fig:phase-histogram) shows an histogram that counts the number of SON years in which the cEOF2 was close to each of the four particular phases (positive/negative Real/Imaginary), with the observations for each season marked as rugs on the horizontal axis.
@@
 
-(ref:phase-histogram-cap) Histogram of phase distribution of cEOF2 for the 1979 -- 2019 period. Bins are centred at 90º, 0º, -90º, -180º with a binwidth of 90º. \textcolor{red}{Rugs...}
+(ref:phase-histogram-cap) Histogram of phase distribution of cEOF2 for the 1979 -- 2019 period.
+Bins are centred at 90º, 0º, -90º, -180º with a binwidth of 90º.
+The small vertical lines near the horizontal axis mark the observations.
 
@@
 
-pattern <- eof_filtered$right %>% 
-  .[era5[lev == 200], on = "time"] %>% 
-  .[, FitLm(hgt, eof), by = .(lon, lat)] %>% 
-  .[, estimate := mean(estimate), by = .(lat, term)] %>%
-  dcast(lon + lat ~ term, value.var = "estimate")
+pattern <- sams[term == "full" & lev == 200] %>% 
+  .[era5[lev == 200], on = c("lev", "time")] %>% 
+  .[, FitLm(hgt, estimate), by = .(lon, lat, lev)] %>% 
+  rm_intercept() %>% 
+  .[, estimate := mean(estimate), by = .(lat)] %>% 
+  .[, term := NULL] %>% 
+  setnames("estimate", "eof")
 
 set.seed(42)
-fake_sam <- eof_filtered$right %>% 
+
+fake_sam <-  sams[term == "full" & lev == 200] %>% 
   copy() %>% 
-  .[, eof := sample(eof, replace = TRUE)]
+  .[, value := sample(estimate, replace = TRUE)] %>% 
+  .[, estimate := NULL] %>% 
+  .[, term := NULL]
+
+# complement <- function(x, y) {
+#   
+#   e <- residuals(lm(x ~ y))
+#   
+#   e*sd(x) + mean(x)
+#   
+# }
+# 
+# fake_sam <-  sams[term == "full" & lev == 200] %>% 
+#   copy() %>% 
+#   .[, value := complement(estimate, y)] %>% 
+#   .[, estimate := NULL] %>% 
+#   .[, term := NULL]
 
@@
   sep_ReIm() %>% 
-  .[, correlate(eof, hgt), by = part]
+  .[, correlate(value, hgt), by = part]
 
 fake_sam <- fake_sam %>% 
-  setnames("eof", "value") %>% 
-  .[, pattern[, .(sam = value*eof), by = .(lon, lat)], by = time]
+  .[pattern, on = "lev", allow.cartesian = TRUE] %>% 
+  .[, sam := value*eof]
 
@@
                  EOF)]
-var <- c(var, setNames(c("SAM", "Imaginary", "Real"), c("SAM", "Imaginary", "Real")))
+var <- c(var, setNames(c("Annular Pattern", "Imaginary cEOF2", "Real cEOF2"), c("SAM", "Imaginary", "Real")))
 
@@
   # adding just a tiny bit of noise solves it.
+
+  .[, eof := scale(eof), by = EOF] %>% 
   .[EOF == "SAM", eof := eof + rnorm(.N, sd = 1e-5)] %>%
-
+  .[EOF %in% c("EOF1", "EOF2"), eof := -eof] %>% 
   periodic(lon = c(0, 360)) %>%
@@
 
-To test the hypothesis that a PSA-like feature can appear as statistic contamination of an otherwise zonally symmetrical and uncorrelated SAM, we create a synthetic dataset of `r uniqueN(fake_sam$time)` fields by reconstructing the SON 200 hPa geopotential height field explained by the cEOF2 and adding a perfectly zonally symmetric SAM-like structure with random sign and value sampled with replacement from the real SAM index.
-To this synthetic dataset (\textcolor{red}{No me suena mucho la palabra "synthetic", ¿se usa en este contexto?}) we then apply traditional EOF.
+To test the hypothesis that a PSA-like feature can appear as statistic contamination of an otherwise zonally symmetrical and uncorrelated annular pattern, we create a synthetic dataset of `r uniqueN(fake_sam$time)` fields by reconstructing the SON 200 hPa geopotential height field explained by the cEOF2 and adding a perfectly zonally symmetric annylar pattern with random sign and value sampled with replacement from the real SAM index.
+To this synthetic dataset we then apply traditional EOF.
 By construction this dataset contains two independent components: a zonally symmetric standing oscillation and a PSA-like of varying amplitude and phase (Figure \@ref(fig:fake-eof) row a).
-While the former is a standing oscillation which could be represented by a single EOF, the latter is a wave which can change phase, so it needs two EOFs to be represented correctly.
+While the former is a standing oscillation which could be represented by a single EOF, the latter is a wave which can change phase, so traditional EOF analysis needs two EOFs to represented it correctly.
 Thus, the full variance of this synthetic dataset can be perfectly explained by three EOFs.
 
-\textcolor{red}{Los titulos de la figura 11 superior se prestan a la confusión, ya que no es exactamente el SAM y no se entiende qué sería "Imaginary" y "Real"}
-
 The three EOFs of the synthetic dataset are shown in Figure \@ref(fig:fake-eof) row b.
-The synthetic SAM can be seen in the first EOF while the cEOF2 is explained by the second and third.
-However, even though the synthetic SAM is perfectly zonally symmetric by construction, the leading EOF is not.
-The EOF decomposition mixes this mode with parts of the the zonally asymmetric cEOF2 pattern, resulting in a pattern pretty similar to the real SAM pattern (\textcolor{red}{REF}).
-This happens because, although in this synthetic dataset the zonally symmetrical SAM and the cEOF2 are uncorrelated by construction, the sample correlation is not null \\textcolor{red}{No me quedo 100% clara esta parte, ¿podrías expresarla un poco distinto?}
-.
+The annular pattern is captured mainly by the first EOF while the cEOF2 is explained by the second and third.
+However, even though the annular pattern used to construct these data is perfectly zonally symmetric, the leading EOF is not.
+The EOF decomposition mixes the true annular pattern with parts of the the zonally asymmetric cEOF2 pattern, resulting in a pattern pretty similar to the real SAM pattern [@fogt2020].
+
+\textcolor{red}{Todo esto está mal.}
+
+Rotating makes it even worse.
+
+The reason this happens is that the spatial correlation between the annular pattern and the wave patterns is not zero; i.e. they are not orthogonal. And since EOFs
+
 In this particular example, it is `r cor_with_eof2[part == "Real"]$text` with the Real part and `r cor_with_eof2[part == "Imaginary"]$text` with the Imaginary part.
@@
 Studies on the relationship between the SAM and PSA-like patterns would be particularly difficult to interpret.
+
 ## Tropical sources
@@
                                    limits = c(-1, 1),
-                                   guide = guide_colorsteps_bottom(20, order = 2)) 
+                                   guide = guide_colorstrip_bottom(20, order = 2)) 
 ```
@@
 The Imaginary cEOF2 is associated with strong positive SST anomalies on the Central Pacific and negative anomalies over an area across the North of Australia and New Zealand, the South Pacific Convergence Zone (SPCZ) (Figure \@ref(fig:sst-psi-2).a2).
-This pattern is almost canonically positive ENSO (\textcolor{red}{REF ENSO}) and indeed, the correlation between the Imaginary cEOF2 and the Oceanic Niño Index [@bamston1997] is very high `r enso_cor[cEOF == "cEOF2" & part == "Imaginary"]$text`.
+This pattern is almost canonically positive ENSO [@bamston1997] and indeed, the correlation between the Imaginary cEOF2 and the Oceanic Niño Index [@bamston1997] is very high `r enso_cor[cEOF == "cEOF2" & part == "Imaginary"]$text`.
 Besides, the Imaginary cEOF2 is associated with strong wave-like streamfunction anomalies emanating from the tropics (Figure \@ref(fig:sst-psi-2).b2).
@@
 Years with magnitude of cEOF2 greater or smaller than the 50th percentile are shown as orange circles and green diamomnds respectively.
-Black line is the fit ONI ~ sin(phase) computed by OLS weighted by the magnitude of cEOF2.", fig.width = width_column, fig.height=3}
+Black line is the fit ONI ~ sin(phase) computed by weighted OLS using the magnitude of the cEOF2 as weights.", fig.width = width_column, fig.height=3}
 gdata <- ceof[, eof[[1]]$left, by = .(season)] %>%
@@
 cor_enso_mag_outlier <- gdata[oni < 1.5][, correlate(mag, abs(oni))]
+
+cor_enso_mag_quantile <- gdata[mag >= quantile(mag, 0.5)] %>% 
+  .[oni < 1.5] %>% 
+  .[, correlate(mag, abs(oni))]
 
@@
   scale_color_brewer(NULL, palette = "Dark2",
-                     labels = c("TRUE"  = "|EOF2| > 50%",
-                                "FALSE" = "|EOF2| < 50%")) +
+                     labels = c("TRUE"  = "|cEOF2| > 50%",
+                                "FALSE" = "|cEOF2| < 50%")) +
   scale_shape_manual(NULL, values = c("TRUE" = 20,
                                       "FALSE" = 18),
-                     labels = c("TRUE"  = "|EOF2| > 50%",
-                                "FALSE" = "|EOF2| < 50%")) +
+                     labels = c("TRUE"  = "|cEOF2| > 50%",
+                                "FALSE" = "|cEOF2| < 50%")) +
   scale_y_continuous("ONI") +
@@
 ¿Probaste hacer la correlación con todos los puntos naranjas?}
+
+```{r}
+# gdata %>% 
+#   ggplot(aes(oni, mag)) +
+#   geom_point(aes(color = mag >= quantile(mag, 0.5)))  +
+#    scale_color_brewer(NULL, palette = "Dark2",
+#                      labels = c("TRUE"  = "|cEOF2| > 50%",
+#                                 "FALSE" = "|cEOF2| < 50%")) +
+#   panel_background
+```
 
@@
 
-(ref:pp-america-cap) Regression of SON mean precipitation anomalies in South America (mm per day, shaded) and (column a) cEOF1 the (row 1) Imaginary and (column 1) Real phase.
-For the 1979 -- 2018 \textcolor{red}{¿Cambiaste el año o es un typo?} period.
-Areas marked with dots have p-values smaller than 0.05 adjusted for False Detection Rate.
-
-```{r pp-america, fig.cap = "(ref:pp-america-cap)", fig.width=width_column, fig.height=width_column*1.1}
-plot_pp(data = regr_pp, which_cEOF = c("cEOF1", "cEOF2"), which_continent = "america")  +
-  scale_fill_divergent_discretised(NULL, 
-                         limits = c(-1, 1),
-                         # oob = scales::squish,
-                         low = "#8E521C",
-                         high = "#00665E",
-                         guide = guide_colorsteps_bottom(15)) 
-  # annotate("rect", xmin = min(sesa$lon), xmax = max(sesa$lon), 
-  #          ymin = min(sesa$lat), ymax = max(sesa$lat),
-  #          fill = NA, color = "gray50", size = 0.3)
-  #          
-```
-
 The influence of cEOFs variability in continental rainfall of extra-tropical Southern Hemisphere is also explored.
@@
 
-```{r sinusoidal-sesa}
-gdata <- cmap %>% 
-  .[lon %between% sesa$lon  & lat %between% sesa$lat] %>% 
-  .[, .(pp = weighted.mean(pp, cos(lat*pi/180))), by = time] %>% 
-  .[, pp := Anomaly(pp), by = season(time)] %>% 
-  .[ceof$eof[[1]]$left, on  = "time"] %>% 
-  .[cEOF == "cEOF2"] %>% 
-  .[, arg := Arg(hgt)] %>% 
-  .[, mag := Mod(hgt)]
-
-model <- lm(pp ~ I(sin(arg)) + I(cos(arg)) - 1, data = gdata)
-
-fstatistic <- summary(model)$fstatistic
-p_value <- pf(fstatistic["value"], fstatistic["numdf"], fstatistic["dendf"], 
-        lower.tail = FALSE)
-```
-
-Figure \@ref(fig:pp-america) shows regression maps of seasonal precipitation with each cEOF in South America.
-cEOF1 is associated with slight decreases \textcolor{red}{Revisa como estás expresando toda esta parte (y en el resto del texto), recorda que son regresiones así que no corresponde decir postive/negative sino haces alguna referencia a la variable con las que los estás regresionando.} in precipitation in Southern Brazil and Paraguay, although these are not statistically significant (Fig. \@ref(fig:pp-america) column a).
-The strongest precipitation regression coefccients are the ones associated with the Imaginary cEOF2.
-The positive anomalies on Southeastern South America (SESA) and Central Chile, and negative anomalies over South Atlantic Convergence Zone is a well known springtime precipitation signature of ENSO [@cai2020a] and it is also similar to the precipitation anomalies associated with the A-SAM [@campitelli2021].
-This is not surprising considering the close relationship between the ONI, the A-SAM index and the Imaginary cEOF2 shown previously, but further consolidates the identification of this mode with the PSA pattern.
-On the other hand, the Real cEOF2 is associated with negative precipitation anomalies in a smaller area of SESA.
-Resembling the relationship between ONI and the phase of cEOF2 (Fig. \@ref(fig:enso-phase)), there is a cEOF2 phase dependence of the precipitation anomalies in SESA (not shown).
-These variable could be related wth a signficant sinusoidal fit that has a coefficient of determination of `r round(summary(model)$r.squared, 2)` (p-value `r report_p(p_value)`).
-
-(ref:pp-oceania-cap) As Figure \@ref(fig:pp-america) but for Oceania.
+(ref:pp-oceania-cap) Regression of SON mean precipitation anomalies in Oceania (mm per day, shaded) and (column a) cEOF1 the (row 1) Imaginary and (column 1) Real phase. For the 1979 -- 2019.
+Areas marked with dots have p-values smaller than 0.05 adjusted for False Detection Rate.
 
@@
 (ref:australia-pp-phase-cap) Same as Figure \@ref(fig:enso-phase) but for precipitation anomalies averaged between `r knitr::combine_words(LonLabel(box_australia$lon))` and `r knitr::combine_words(LatLabel(box_australia$lat))` (box shown in Figure \@ref(fig:pp-oceania) column b).
+Black line is the fit precipitation ~ sin(phase) + cos(phase) computed by weighted OLS using the magnitude of the cEOF2 as weights.
 
@@
 
-model <- lm(pp ~ I(sin(arg)) + I(cos(arg)) - 1, data = gdata)
+model <- lm(pp ~ I(sin(arg)) + I(cos(arg)) - 1, weights = mag, data = gdata)
 fstatistic <- summary(model)$fstatistic
@@
 equation <- latex2exp::TeX(paste0("$\\mathrm{Precipitation} = ", 
-        signif(coef(model)[1], 2), "\\sin{(\\mathrm{phase})} + ", 
-        signif(coef(model)[2], 2), "\\cos{(\\mathrm{phase})}$"), output = "text")
+        signif(coef(model)[1], 2), "\\frac{mm}{day} \\sin{(\\mathrm{phase})} + ", 
+        signif(coef(model)[2], 2), "\\frac{mm}{day} \\cos{(\\mathrm{phase})}$"), output = "text")
 
@@
   labs(subtitle = equation) +
+  theme(plot.subtitle = element_text(size = 8)) +
   panel_background
@@
 
-To better understand the relationship between precipitation anomalies over Western Australia and the cEOF2, Figure \@ref(fig:australia-pp-phase) plots precipitation anomalies averaged over Eastern Australia (box in Figure \@ref(fig:pp-oceania)) as a function of cEOF2 phase.
+To better understand the relationship between precipitation anomalies over Eastern Australia and the cEOF2, Figure \@ref(fig:australia-pp-phase) plots precipitation anomalies averaged over Eastern Australia (box in Figure \@ref(fig:pp-oceania)) as a function of cEOF2 phase.
 The black line shows a sinusoidal fit, whose coefficients are fitted by weighted least squares using the magnitude of the EFO2 as weight.
@@
 
-\textcolor{red}{¿Había alguna razón para poner SA antes que Oceanía?
-Sino invertiría el orden, porque creo que es más obvio lo que decis en Sudamérica sobre el sinusoidal fit, si mostras antes de lo de Oceanía.}
+(ref:pp-america-cap) As Figure \@ref(fig:pp-america) but for South America
+
+```{r pp-america, fig.cap = "(ref:pp-america-cap)", fig.width=width_column, fig.height=width_column*1.1}
+plot_pp(data = regr_pp, which_cEOF = c("cEOF1", "cEOF2"), which_continent = "america")  +
+  scale_fill_divergent_discretised(NULL, 
+                         limits = c(-1, 1),
+                         # oob = scales::squish,
+                         low = "#8E521C",
+                         high = "#00665E",
+                         guide = guide_colorsteps_bottom(15)) 
+  # annotate("rect", xmin = min(sesa$lon), xmax = max(sesa$lon), 
+  #          ymin = min(sesa$lat), ymax = max(sesa$lat),
+  #          fill = NA, color = "gray50", size = 0.3)
+  #          
+```
+
+```{r sinusoidal-sesa}
+gdata <- cmap %>% 
+  .[lon %between% sesa$lon  & lat %between% sesa$lat] %>% 
+  .[, .(pp = weighted.mean(pp, cos(lat*pi/180))), by = time] %>% 
+  .[, pp := Anomaly(pp), by = season(time)] %>% 
+  .[ceof$eof[[1]]$left, on  = "time"] %>% 
+  .[cEOF == "cEOF2"] %>% 
+  .[, arg := Arg(hgt)] %>% 
+  .[, mag := Mod(hgt)]
+
+model <- lm(pp ~ I(sin(arg)) + I(cos(arg)) - 1, data = gdata)
+
+fstatistic <- summary(model)$fstatistic
+p_value <- pf(fstatistic["value"], fstatistic["numdf"], fstatistic["dendf"], 
+        lower.tail = FALSE)
+```
+
+Figure \@ref(fig:pp-america) shows regression maps of seasonal precipitation with each cEOF in South America.
+cEOF1 is associated with slight decreases \textcolor{red}{Revisa como estás expresando toda esta parte (y en el resto del texto), recorda que son regresiones así que no corresponde decir postive/negative sino haces alguna referencia a la variable con las que los estás regresionando.} in precipitation in Southern Brazil and Paraguay, although these are not statistically significant (Fig. \@ref(fig:pp-america) column a).
+The strongest precipitation regression coefccients are the ones associated with the Imaginary cEOF2.
+The positive anomalies on Southeastern South America (SESA) and Central Chile, and negative anomalies over South Atlantic Convergence Zone is a well known springtime precipitation signature of ENSO [@cai2020a] and it is also similar to the precipitation anomalies associated with the A-SAM [@campitelli2021].
+This is not surprising considering the close relationship between the ONI, the A-SAM index and the Imaginary cEOF2 shown previously, but further consolidates the identification of this mode with the PSA pattern.
+On the other hand, the Real cEOF2 is associated with negative precipitation anomalies in a smaller area of SESA.
+Resembling the relationship between ONI and the phase of cEOF2 (Fig. \@ref(fig:enso-phase)), there is a cEOF2 phase dependence of the precipitation anomalies in SESA (not shown).
+These variable could be related wth a signficant sinusoidal fit that has a coefficient of determination of `r round(summary(model)$r.squared, 2)` (p-value `r report_p(p_value)`).
 
@@
 
-To the extreme that the leading EOF of geopotential height with the cEOF2 removes results in an annular mode that has none of the typical zonal asymmetries of the observed SAM.\textcolor{red}{No me pareció del todo claro lo que estás queriendo expresar acá}
-
 This raises the possibility that the asymmetric component of the SAM is actually a statistical contamination of the PSA mode.

### 2021-06-15 analysis/paper/shceof.tex
Render

@@
 
-\title{Title Goes Here \thanks{The research was supported by UBACyT20020170100428BA and the CLIMAX Project funded by Belmont Forum/ANR-15-JCL/-0002-01.
+\title{Austral Spring Extratropical Southern Hemisphere zonally assymetric circulation \thanks{The research was supported by UBACyT20020170100428BA and the CLIMAX Project funded by Belmont Forum/ANR-15-JCL/-0002-01.
 Elio Campitelli was supported by a PhD grant from CONICET, Argentina.} }
@@
 
-One typical way of describing the zonally asymmetric circulation is by the amplitude and phase of zonal waves obtained by Fourier decomposition of geopotential height at each latitude \citep[e.g.][]{vanloon1972, trenberth1980}.
-These methodologies rely on the assumption that the circulation can be meaningfully understood in terms of zonal waves of constant amplitude along a latitude circle.
+\textcolor{red}{Agregar parráfo que introduzca el problema del estudio de la ciriculación extra-tropical en SH, su variaiblidad y sus cambios.
+Quizás algo también de los impactos.
+También algo habría que decir de la primavera, ya que solo nos concentrmaos en esa estación}
+
+The large-scale extratropical circulation in the Southern Hemisphere is strongly zonally symmetric, but its zonal departures are highly relevant for surface impacts and could be related to high-impact climate extremes.
+Zonal asymmetries of extratropical circulation in the Southern Hemisphere strongly modulate weather systems and regional climate through latitudinal transport of heat, humidity, and momentum \citep{trenberth1980}.
+
+One typical way of describing the zonally asymmetric circulation is by the amplitude and phase of zonal waves obtained by Fourier decomposition of geopotential height at each latitude \citep[e.g.][]{vanloon1972, trenberth1980, turner2017}.
+This approach suggests that zonal waves 1 and 3 explained most of the variance of the tropospheric extratropical Southern Hemisphere circulation (\citet{vanloon1972}).
+However, this methodology rely on the assumption that the circulation can be meaningfully understood in terms of zonal waves of constant amplitude along a latitude circle.
 This is not valid for meridionally propagating waves or zonal waves with localised amplitudes.
-In the case of the wave 3, for example, \citet{trenberth1985} observed that wave 3 played a role in blocking events, but that in many cases, this was due to increased amplitude of a longitudinally localised wave train and not a hemispheric-scale zonal wave 3.
+In the case of the wave 3, for example, \citet{trenberth1985} observed that it played a role in blocking events, but this was mostly due to increased amplitude of a longitudinally localised wave train instead of an hemispheric-scale zonal wave 3.
 
 Another approach to characterising the Southern Hemisphere circulation is by using Empirical Orthogonal Functions (EOF, also known as Principal Component Analysis).
-Within the EOF framework, the Southern Annular Mode appears as the leading mode of variability of the Southern Hemisphere (SAM) \citep{fogt2020} followed by the two Pacific--South American Patterns (PSA) \citep{mo2001}.
+Within the EOF framework, the Southern Annular Mode (SAM) appears as the leading mode of variability of the tropospheric Southern Hemisphere circulation \citep{fogt2020} followed by the two Pacific--South American Patterns (PSA) \citep{mo2001}.
 The SAM represents a relatively zonally symmetric pattern of alternating low pressures in polar latitude and a ring of high pressures in high latitudes with an embedded wave 3 pattern that is more prominent in the Pacific sector.
@@
 
-Finally, another, hybrid, methodology consist on identifying particular features of interest and creating indices using simple methods such as averages and differences.
+Finally, another, hybrid, methodology commonly used consist on identifying particular features of interest and creating indices using simple methods such as averages and differences.
 The Marshal SAM Index \citep{marshall2003}, for example, tries to characterise the SAM by the average difference of sea level pressure between 40ºS and 65ºS, following \citet{gong1999}.
@@
 This is equivalent to an index of the amplitude of the zonal wave 3 that projects into the phase of the zonal wave 3 of the mean field.
-Similarly, \citet{hobbs2010} posit that the zonally asymmetric circulation in the Southern Hemisphere can be described by the strength and location of two anticyclones located in the sub-Antarctic western hemisphere.
+Similarly, \citet{hobbs2010} propose that the zonally asymmetric circulation in the Southern Hemisphere can be described by the strength and location of two anticyclones located in the sub-Antarctic western hemisphere.
+These methods are grounded on other methods to identify the centers of action for the described phenomena and can be useful to characterise features that are not readily apparent with Fourier or EOF methods.
+The use of data of only specific regions or locations makes them suitable for easy computation and for extending into less quality data from the past, but also makes them not very robust and could not capture non-stationary behaviours of the patterns.
 
-These ``feature-based'' methods are grounded on other methods to identify the interesting features and can be useful to characterise features that are not readily apparent with Fourier or EOF methods but that nonetheless interesting or important.
-The use of data of single points, which makes them suitable for extending into the data-poor past also makes them not very robust.
+\textcolor{red}{Falta un parráfo que deje más claro por qué es necesario considerar un enfoque alternativo y cuál va a ser el beneficio de ello.
+También agregar qué se estudiara conjuntamente la troposfera y la estratósfera }
+
+\textcolor{red}{Falta el objetivo que justifica el uso de la metodología}
 
@@
 We apply this method to characterise the Springtime zonally asymmetric circulation of the extratopical Southern Hemisphere.
+
+\textcolor{red}{The data and methods are described in Section 2 :P ...}
 
@@
 We used monthly geopotential height, air temperature, ozone mixing ratio, and Total Ozone Column (TOC) at 2.5\degree~longitude by 2.5\degree~latitude of horizontal resolution and 37 vertical isobaric levels from ERA5 \citep{era5} for the period 1979 to 2019.
-We restricted most of our analysis to the post-satellite era to avoid any confounding factors arising from the incorporation of satellite observations, but we also used the preliminary back extension of ERA5 from 1950 to 1978 \citep{era5be} to look at long-term trends.
-We derived streamfunction at 200 hPa from ERA5 vorticity using the FORTRAN subroutine FISHPACK \citep{fishpack} and computed horizontal wave activity fluxes following \citet{plumb1985}.
-Sea Surface Temperature (SST) fields were taken from Extended Reconstructed Sea Surface Temperature (ERSST) v5 \citep{huang2017}.
-For precipitation data we used monthly data from the CPC Merged Analysis of Precipitation \citep{cmap}, with a 2.5\degree~resolution in latitude and longitude.
+Most of our analysis was restricted to the post-satellite era to avoid any confounding factors arising from the incorporation of satellite observations, but we also used the preliminary back extension of ERA5 from 1950 to 1978 \citep{era5be} to look at long-term trends.
+Streamfunction was derived from ERA5 vorticity at 200 hPa using the FORTRAN subroutine FISHPACK \citep{fishpack} and horizontal wave activity fluxes was computed following \citet{plumb1985}.
+We usded Sea Surface Temperature (SST) monthly fields from Extended Reconstructed Sea Surface Temperature (ERSST) v5 \citep{huang2017} and precipitation monthly data from the CPC Merged Analysis of Precipitation \citep{cmap}, with a 2.5\degree~resolution in latitude and longitude.
 This rainfall gridded dataset is based on information from different sources such as rain gauge observations, satellite inferred estimations and the NCEP-NCAR reanalysis, and it is available since 1979 to present.
@@
 
-We restrict our study to the September-October-November (SON) trimester.
-For that, we use seasonal means, averaging monthly values weighted by the number of days in each month.
-We also use 200 hPa to represent the high troposphere and 50 hPa to represent the lower stratosphere.
+The study is restricted to the spring season, defined as the September-October-November (SON) trimester.
+We compute seasonal means for the different variables, averaging monthly values weighted by the number of days in each month.
+We use 200 hPa level to represent the high troposphere and 50 hPa to represent the lower stratosphere.
 
+The amplitude of the zonal waves was defined through computing the Fourier transform of the spatial field at each latitude circle.
 We computed the amplitude and phase of the zonal wave 1 by averaging (area-weighted) variables (temperature, geopotential height, ozone mixing ratio and Total Ozone Column) between 75°S and 45°S for each SON and extracting the wave-1 component of the Fourier spectrum.
@@
 We further split the SAM into its zonally symmetric and zonally asymmetric components (S-SAM and A-SAM indices respectively).
-These are obtained by projecting the zonally asymmetric and zonally symmetric part of the SAM spatial pattern onto monthly geopotential height fields \citep{campitelli2021}.
-Since the analysis here is only for the SON trimester, monthly values were averaged across trimesters.
+These indices were obtained by projecting the zonally asymmetric and zonally symmetric part of the SAM spatial pattern onto monthly geopotential height fields, as proposed by \citet{campitelli2021}.
+The Pacific South American patterns (PSA1 and PSA2) were calculated following \citet{mo2001} as the third and fourth leading EOF of seasonal mean anomalies for SH 500 hPa geopotential height combining all seasons.
 
-We also calculated the Pacific South American patterns (PSA1 and PSA2) as the third and fourth leading EOF of seasonal mean 500 hPa geopotential height with all seasons together following \citet{mo2001}.
-
-We performed linear regressions to quantify the association between the cEOFs and other variables.
-For each cEOF, we create regression maps by computing fitting a multiple linear model involving both the real and the imaginary part.
-We evaluated statistical significance using a two-sided t-test and, in the case of regression maps, p-values were adjusted by controlling for the False Discovery Rate \citep{benjamini1995, wilks2016} to avoid misleading results from the high number of regressions \citep{walker1914, katz1991}.
-
-Linear trends were computed by Ordinary Least Squares and the 95\% confidence interval was computed assuming a t-distribution with the appropriate residual degrees of freedom.
-To the amplitude of the zonal waves is defined through computing the Fourier transform of the spatial field at each latitude circle.
+Linear trends were computed by Ordinary Least Squares and the 95\% confidence interval was computed assuming a t-distribution with the appropriate residual degrees of freedom \citep{wilks2011r}.
 
@@
 \includegraphics{../figures/eof-naive-1.pdf}
-\caption{\label{fig:eof-naive}Spatial patterns of the leading 4 EOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2019 (arbitrary units).}
+\caption{\label{fig:eof-naive}Spatial patterns of the leading 4 EOFs of SON zonal anomalies of geopotential height at 50 hPa south of 20\degree S for the 1979 -- 2019 period (arbitrary units).}
 \end{figure}
 
-In traditional EOF analysis zonal waves appear as pairs of EOFs, usually degenerate, that represent similar patterns but shifted in phase \citep{horel1984}.
-For instance, Figure~\ref{fig:eof-naive} shows the leading 4 EOFs of zonally anomalous SON geopotential height at 50 hPa and 200 hPa.
-It is clear that at 50 hPa the first two EOFs represent the same zonal wave 1 pattern and the last 2 represent the same zonal wave pattern with shorter wavenumber shifted by 1/4 wavelength.
+In traditional EOF analysis zonal waves appear as pairs of EOFs, that could be degenerated representing similar patterns but shifted in phase \citep{horel1984}.
+For instance, Figure~\ref{fig:eof-naive} shows the leading 4 EOFs of SON geopotential height zonal anomalies at 50 hPa south of 20\degree S.
+It is clear that the first two EOFs represent the same zonal wave 1 pattern and the last two represent a same zonal wave pattern with shorter wavenumber and four centers of action shifted by 1/4 wavelength.
 A similar EOF structure can be seen in 200 hPa (not shown).
@@
 
-Complex Empirical Orthogonal Functions (cEOF) is a useful method to characterise these waves \citep{horel1984}.
-This method involves augmenting the original fields with the Hilbert transform of the data and then computing the Singular Value Decomposition of it.
-In this work, instead of applying the Hilbert transform to the time series at each point, we apply it to each latitude circle at each moment in time.
-Since the latitude circle is a periodic domain, this procedure does not suffer from edge effects.
-The result of the cEOF is a set of complex spatial patterns and complex time series.
-The real and imaginary part of each spatial pattern represent two phases wave-like spatial pattern that are in quadrature.
+Complex Empirical Orthogonal Functions (cEOF) is a useful method to characterise zonal waves, considering phase-varying structures \citep{horel1984}.
+This method involves computing EOF on the analytic representation of the original field.
+This representation is a complex field in which the real part is the original variable and the imaginary part is the original data shifted by 90\degree at each spectral frequency -- i.e.~its Hilbert transform.
+The Hilbert transform is usually understood in terms of time-varying signal.
+However, in this work we apply the Hilbert transform to each latitude circle at each moment in time.
+Since each latitude circle is a periodic domain, this procedure does not suffer from edge effects.
+
+The result of the cEOF methodology is a set of complex spatial patterns and complex time series.
+The real and imaginary part of each spatial pattern represent two phases wave-like spatial patterns that are in quadrature.
 The magnitude and argument of each complex time series represent the amplitude and phase of each zonal wave.
@@
 
-\caption{\label{tab:corr-ceof-splitted}Coefficient of determination ($R^2$) of the absolute magnitude of complex EOFs between 200 hPa and 50 hPa computing EOF separatedly for each level.}
+\caption{\label{tab:corr-ceof-splitted}Coefficient of determination (\(R^2\)) of the absolute magnitude of complex EOFs between 200 hPa and 50 hPa computing EOF separatedly for each level.
+Background colour indicates the value of \(R^2\).}
 \centering
@@
 
+The cEOF methodology is applied to SON geopotential height zonal anomalies south of 20\degree S at 50 and 200 hPa, representing a stratsopheric and upper troposhperic level respectively.
 Table \ref{tab:corr-ceof-splitted} shows the coefficient of determination between time series of the amplitude of each complex EOF (cEOF) across levels.
 There's a high degree of correlation between the respective cEOF1 and cEOF2 at each level.
-The spatial patterns of the 50 hPa and 200 hPa cEOFs are also similar (not shown)
+The spatial patterns of the 50 hPa and 200 hPa cEOFs are also similar (not shown).
 
@@
 
+While we compute these complex principal components using data from 1979 to 2019, we extended the complex time series back to the 1950 -- 1978 period by projecting monthly geopotential height zonal anomalies standardised by level south of 20ºS onto the corresponding spatial patterns.
+
+We performed linear regressions to quantify the association between the cEOFs and other variables.
+For each cEOF, we create regression maps by computing fitting a multiple linear model involving both the real and the imaginary part.
+To obtain the linear coefficients of a variable \(X\) (geopotential, temperature, precipitation, etc\ldots) with the Imaginary and Real parts of each cEOF we fit the equation
+
+\[
+X(\lambda, \phi, t) = \alpha(\lambda, \phi) \operatorname{Im}(\operatorname{cEOF}) + \beta(\lambda, \phi) \operatorname{Re}(\operatorname{cEOF}) + X_0(\lambda, \phi) +  \epsilon(\lambda, \phi, t)
+\]
+
+where \(\lambda\) and \(\phi\) are the longitude and latitude, \(t\) is the time, \(\alpha\) and \(\beta\) are the linear regression coefficients, \(X_0\) and \(\epsilon\) are the constant and error terms respectively.
+
+We evaluated statistical significance using a two-sided t-test and, in the case of regression maps, p-values were adjusted by controlling for the False Discovery Rate \citep{benjamini1995, wilks2016} to avoid misleading results from the high number of regressions \citep{walker1914, katz1991}.
+
 \hypertarget{computation-procedures}{%
@@
 
-We performed all analysis in this paper using the R programming language \citep{rcoreteam2020}, using the data.table package \citep{dowle2020} and the metR package \citep{campitelli2020}.
+We performed all analysis in this paper using the R programming language \citep{rcoreteam2020}, using data.table \citep{dowle2020} and metR \citep{campitelli2020} packages.
 All graphics are made using ggplot2 \citep{wickham2009}.
@@
 
-\hypertarget{ceof-patterns}{%
-\subsection{cEOF patterns}\label{ceof-patterns}}
+\hypertarget{ceof-spatial-patterns}{%
+\subsection{cEOF spatial patterns}\label{ceof-spatial-patterns}}
 
@@
 \includegraphics{../figures/ceofs-1-1.pdf}
-\caption{\label{fig:ceofs-1}Spatial patterns of the leading 2 cEOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2019.
-Real part in shading, imaginary part in contours.
+\caption{\label{fig:ceofs-1}Spatial patterns for the two leading cEOFs of SON zonal anomalies of geopotential height at 50 hPa and 200 hPa for the 1979 -- 2019 period.
+The shading (contours) corresponds to real (imaginary) part.
 Arbitrary units.}
@@
 
-\begin{figure}
-\centering
-\includegraphics{../figures/ceofs-2-1.pdf}
-\caption{\label{fig:ceofs-2}Time series of the leading 2 cEOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2019.}
-\end{figure}
+\begin{figure*}
+\includegraphics{../figures/extended-series-1} \caption{Time series extended using ERA5 back extended preliminary edition (period 1950 -- 1978) and ERA5 (period 1979 -- 2019).
+Dark straight line is the OLS trend.
+Black horizontal and vertical line mark the mean value and range of each time series respectively.}\label{fig:extended-series}
+\end{figure*}
 
-To describe the spatial and temporal variability of the circulation zonal anomalies, the spatial and temporal parts of the first two leading cEOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa are shown in Figures \ref{fig:ceofs-1} and \ref{fig:ceofs-2}.
-The first mode (CEOF1) explains 82\% of the variance, while the second mode (CEOF1) explains a smaller fraction (7\%).
-In the spatial patterns (Fig.~\ref{fig:ceofs-1}), the real (in shading) and the imaginary components (in contour) are in quadrature by construction, so that each cEOF describe a single wave-like pattern whose amplitude and position (i.e.~phase) is controlled by the magnitude and phase of the temporal cEOF.
+To describe the variability of the circulation zonal anomalies, the spatial and temporal parts of the first two leading cEOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa, computed jointly at both levels, are shown in Figures \ref{fig:ceofs-1} and @ref(fig:extended-series,).
+The first mode (cEOF1) explains 82\% of the variance, while the second mode (cEOF2) explains a smaller fraction (7\%).
+In the spatial patterns (Fig.~\ref{fig:ceofs-1}), the real and the imaginary components are in quadrature by construction, so that each cEOF describe a single wave-like pattern whose amplitude and position (i.e.~phase) is controlled by the magnitude and phase of the temporal cEOF.
 The wave patterns described by these cEOFs match the patterns seen in the traditional EOFs of Figure~\ref{fig:eof-naive}.
@@
 The first cEOF (Fig.~\ref{fig:eof-naive} column~a) is a hemispheric wave 1 pattern with maximum amplitude at high latitudes.
-At 50 hPa the Real cEOF1 has the maximum of the wave 1 at 150ºE and at 200 hPa, the maximum is locates at around 175ºE indicating a westerly shift in phase.
-
-The second cEOF (Fig.~\ref{fig:eof-naive} column~b) show also a zonal wave-like structure with maximum amplitude at high latitudes, but with shorter spatial scales.
-In particular, it's dominant structure at both levels is a wave 3 but with larger amplitude in the pacific sector.
+At 50 hPa the Real cEOF1 has the maximum of the wave 1 at 150ºE and at 200 hPa, the maximum is located at around 175ºE indicating a westerly shift in phase.
+The second cEOF (Fig.~\ref{fig:eof-naive} column~b) shows also a zonal wave-like structure with maximum amplitude at high latitudes, but with shorter spatial scales.
+In particular, its dominant structure at both levels is a wave 3 but with larger amplitude in the pacific sector.
 This modulated amplitude is especially apparent at 200 hPa.
@@
 
-The temporal variability for both components of the cEOF1 have non zero mean (Fig.~\ref{fig:ceofs-2}).
+The temporal variability for both components of the cEOF1 have non zero mean (Fig.~@ref(fig:extended-series,)).
 This is due to the fact that the geopotential fields that enter into the cEOFs algorithm are anomalies with respect to the zonal mean, not the time mean.
-As a consequence, the variability associated with the first cEOF includes variability that projects onto the mean zonally anomalous field.
-There are no significant simultanous correlation between the time series.
+As a consequence, the variability associated with the cEOF1 includes variability that projects onto the mean zonally anomalous field.
+There are no significant simultaneous correlation between the time series.
+Both cEOFs show year-to-year variability but show no evidence of decadal variability.
 
-\begin{figure*}
-\includegraphics{../figures/extended-series-1} \caption{Time series extended using ERA5 back extended preliminary edition (period 1950 -- 1978) and ERA5 (period 1979 -- 2019).
-Each series is computed by projecting monthly geopotential height zonal anomalies standardised by level south of 20ºS onto the corresponding spatial pattern and scaled to zero mean and unit standard deviation.}\label{fig:extended-series}
-\end{figure*}
-
-Both cEOF shot year-to-year variability but show no evidence of decadal variabilty.
-The relatively short timespan (40 years) hampers our habilty to detect longer trends.
-Therefore, we extend the cEOF time series using the preliminary ERA5 back extension going back to 1950 by projecting the older geopogential height zonal anomaly fields onto the spatial patterns shown in Figure~\ref{fig:ceofs-1}.
-
-The extended cEOF timeseires are shown in Figure~\ref{fig:extended-series}.
 There is a significant upward trend in the real component of cEOF1 (Fig.~\ref{fig:extended-series}a.1, p-value~\textless{} 0.001) and no significant trend in any of the complex components of cEOF2.
@@
 
-\hypertarget{regressions}{%
-\subsection{Regressions}\label{regressions}}
+\hypertarget{ceofs-regressions}{%
+\subsection{cEOFs Regressions}\label{ceofs-regressions}}
 
 The spatial patterns shown in Figure~\ref{fig:ceofs-1} are derived by removing the zonally symmetric circulation, so they might not include all the variability that is actually associated with the cEOF time series.
-To understand the full geopotential height field changes associated with each cEOF, we computed regression patterns of each cEOF with the complete geopotential fields.
+To understand the full geopotential height field changes associated with each cEOF, we computed regression patterns of the full geopotential fields onto each cEOF.
 
@@
 \includegraphics{../figures/eof1-regr-gh-1.pdf}
-\caption{\label{fig:eof1-regr-gh}Regression coefficients of the real and imaginary part of the first cEOF on SON geopotential height for the 1979 -- 2019 period.
-These coefficients come from multiple linear regression involving the real and imaginary part of both cEFO2.}
+\caption{\label{fig:eof1-regr-gh}Regression of SON geopotential height anomalies (meters) with the (column 1) real and (column 2) imaginary parts of the first cEOF for the 1979 -- 2019 period at (row a) 50 hPa and (row b) 200 hPa.
+These coefficients come from multiple linear regression involving the real and imaginary parts.
+Areas marked with dots have p-values smaller than 0.05 adjusted for False Detection Rate.}
 \end{figure}
 
-Figure~\ref{fig:eof1-regr-gh} shows regression patterns between geopotential height anomalies and cEOF1.
+Figure~\ref{fig:eof1-regr-gh} shows regression maps of SON geopotential height anomalies upon cEOF1.
 At 50 hPa (Figure~\ref{fig:eof1-regr-gh} row a), both the Real and Imaginary cEOF1 are associated with planetary wave 1 patterns, that are 90º off phase.
@@
 It is only possible to distinguish partially the wave 1 pattern in relation with the real cEOF1 (Figure~\ref{fig:eof1-regr-gh}b.1).
-The imaginary cEOF1 shows a much more zonally symmetrical pattern resembling a negative Southern Annular Mode.
+The imaginary cEOF1 shows a much more zonally symmetrical pattern resembling the negative SAM phase.
 
 With the exception of the imaginary cEOF1, it is clear that these patterns are very different than the fully zonally asymmetric versions (Fig.~\ref{fig:ceofs-1}), particularly at 200 hPa.
-Moreover, only in the stratosphere these patterns actually show a distinguishable wave 1 pattern shifted in phase by 90º.
-This should not be utterly surprising, since the spatial patterns of the \emph{naive} EOFs only show this feature at 50 hPa (Fig.~\ref{fig:eof-naive}), suggesting that using the cEOF method is artificially generating a wave structure at 200hPa.
+Moreover, only in the stratosphere these patterns actually show a distinguishable wave 1 pattern shifted in phase by 90º, suggesting that using the cEOF method is artificially generating a wave structure at 200hPa.
 Therefore, the magnitude and phase of the cEOF1 are associated with the magnitude and phase of a zonal wave only in the stratosphere.
@@
 
-Figure~\ref{fig:eof2-regr-gh} shows the regression pattern of geopotential height and the cEOF2.
+Figure~\ref{fig:eof2-regr-gh} shows the regression pattern of geopotential height anomalies upon the cEOF2.
 Unlike for cEOF1, in this case the regression patterns are similar to the fully zonally asymmetric patterns from Figure~\ref{fig:ceofs-1}.
@@
 \subsection{cEOFs relationship with other variables}\label{ceofs-relationship-with-other-variables}}
+
+\textcolor{red}{Quedo medio rara la divsión con los títulos.
+Me parece que convendrá discutir primero la relación con otros modos de circulación conocidos (SAM, Ozono) y después con las otras variables(Temp, TOC, SST, precip)}
 
@@
 \includegraphics{../figures/t-vertical-1.pdf}
-\caption{\label{fig:t-vertical}Regression coefficients between cEOF1 and zonal anomalies of mean air temperature (shaded) and ozone mixing ratio (contours, negative contours with dashed lines, labels in parts per billion by mass) averaged between 75°S and 45°S for the period 1979 -- 2019.}
+\caption{\label{fig:t-vertical}Regression of SON anomalies of mean air temperature (shaded, Kelvin) and ozone mixing ratio (contours, negative contours with dashed lines, labels in parts per billion by mass) averaged between 75°S and 45°S with the (a) real and (b) imaginary parts of the first cEOF for the 1979 -- 2019 period.}
 \end{figure}
 
+The relation between cEOFs and temperature was also evaluated.
 Figure~\ref{fig:eof1-regr-t} shows regression patterns of air temperature at 50hPa and 200hPa onto cEOF1.
 In both levels, the Real cEOF1 is associated with positive monopole over the South Pole with its centre moved slightly towards 150ºE (Fig.~\ref{fig:eof1-regr-t} column 1).
-Regression maps on the Imaginary cEOF1, on the other hand, show a more clear wae 1 pattern with its maximum around 60ºE.
+On the other hand, the regression maps on the Imaginary cEOF1 show a more clear wave 1 pattern with its maximum around 60ºE.
 The distribution of temperature regression coefficients at 50 hPa and at 200 hPa mirror the geopotential height regression maps at 50 hPa (Fig.~\ref{fig:eof1-regr-gh}).
 
-The vertical distribution of these patterns is shown in Figure~\ref{fig:t-vertical}) that shows the regression coefficientes between cEOF1 and zonal anomalies of air temperature and zonal anomalies of ozone mixing ratio averaged between 75°S and 45°S.
+Figure \ref{fig:t-vertical} shows the vertical distribution of the regression coefficients on cEOF1 from zonal anomalies of air temperature and zonal anomalies of ozone mixing ratio averaged between 75°S and 45°S.
+\textcolor{red}{Deberias aclarar qué significa la terminología que estas utilizando. Lo podes hacer en los métodos y ya queda para todo el texto}.
 Temperature zonal anomalies associated with cEOF1 show a clear wave 1 pattern for both real and imaginary components throughout the atmosphere above 250 hPa with a change in sign above 10 hPa.
@@
 
-The maximum ozone anomalies are co-located with the minimum temperature anomalies above 10 hPa and with the maximum temperature anomalies below 10 hPa (Fig.~\ref{fig:t-vertical})); that is, the ozone zonal wave 1 is anticorrelated with the temperature zonal wave 1 in the upper stratosphere, and directly correlated in the upper stratosphere .
+The maximum ozone anomalies are co-located with the minimum temperature anomalies above 10 hPa and with the maximum temperature anomalies below 10 hPa (Fig.~\ref{fig:t-vertical})).
+Therefore, the ozone zonal wave 1 is anticorrelated with the temperature zonal wave 1 in the upper stratosphere, and directly correlated in the upper stratosphere .
 This change in phase is observed in ozone anomalies forced by planetary waves that reach the stratosphere.
@@
 \includegraphics{../figures/o3-regr-1.pdf}
-\caption{\label{fig:o3-regr}Regression coefficients of SON mean Total Ozone Column anomalies onto the real (a) and imaginary (b) components of the cEOF1.
-On contours, the mean zonal anomaly of Total OZone Column (negative contours in dashed lines, Dobson Units).
-For the 1979 -- 2019 period.}
+\caption{\label{fig:o3-regr}Regression of SON mean Total Ozone Column anomalies (shaded, Dobson Units) with the (a) real and (b) imaginary parts of the first cEOF for the 1979 -- 2019 period.
+On contours, the mean zonal anomaly of Total Ozone Column (negative contours in dashed lines, Dobson Units).
+Areas marked with dots have p-values smaller than 0.05 adjusted for False Detection Rate.}
 \end{figure}
 
-Figure~\ref{fig:o3-regr} shows regression maps of cEOF1 with fields of Total Ozone Column (TOC).
-It shows zonal wave 1 pattens in TOC associated with both phases of cEOF1.
-Climatologically, the springtime Ozone minimum is located off the South Pole and towards the Weddell Sea \citep{wirth1993}.
+The regression maps of cEOF1 with fields of Total Ozone Column (TOC) (Fig.
+~\ref{fig:o3-regr}) show zonal wave 1 pattens in TOC associated with both phases of cEOF1.
+Climatologically, the springtime Ozone minimum is located off the South Pole and towards the Weddell Sea \citep{grytsai2011}.
 Thus, the Real cEOF1 regression pattern (Figure~\ref{fig:o3-regr}a) coincides with the climatological position of the ozone hole while the one for the Imaginary cEOF1 is shifted by 90º.
 
-There is a close relationship between amplitude and phase of the cEOF1 and the amplitude and phase of the planetary wave 1 in Total Ozone Column between 75°S and 45°S; as shown in Figure~\ref{fig:o3-regr}, this is a wide latitude band that captures most of the area with significant regression coefficients.
-The correlation between the amplitude of each index is 0.77 (CI: 0.61 -- 0.87) the overall correlation between the zonal wave 1 of ozone and cEOF1 (computed as the mean cosine of the difference in phase) is 0.87 (CI: 0.77 -- 0.93).
-As expected from the location of anomalies in Figure~\ref{fig:o3-regr}, the Real cEOF1 drives the relationship with amplitude and the Imaginary cEOF1 drives the relationship with phase (not shown).
+There is a close spatial relationship between amplitudes and phases of the cEOF1 and TOC planetary wave 1 between 75°S and 45°S.
+The correlation between the amplitude of both indices is 0.77 (CI: 0.61 -- 0.87) the overall correlation between the zonal wave 1 of ozone and cEOF1 (computed as the mean cosine of the difference in phase) is 0.87 (CI: 0.77 -- 0.93).
+As expected from the location of regressions in Figure~\ref{fig:o3-regr}, the Real cEOF1 drives the relationship of amplitudes and the Imaginary cEOF1 drives the relationship of phases between the timeseries (not shown).
 
@@
 
-To explore the relationship between SAM and the modes described from the cEOF, we compute.
-the coefficient of determination between the cEOFs timeseries and the three SAM indices (SAM, A-SAM and S-SAM) at each level (Fig.~\ref{fig:sam-eof-vertical}).
+To explore the relationship between SAM and the modes described from the cEOF, we compute the coefficient of determination between the cEOFs timeseries and the three SAM indices (SAM, A-SAM and S-SAM) at each vertical level (Fig.~\ref{fig:sam-eof-vertical}).
 The SAM index is moderately correlated with the Real cEOF1 in all levels, and with the Imaginary cEOF1 and Imaginary cEOF2 in the troposphere.
@@
 
-In the case of cEOF2, the moderate correlation between SAM and Imaginary cEOF2 becomes extremely high when only the zonally asymmetric variability of the SAM is considered.
+In the case of cEOF2, the moderate correlation between SAM and Imaginary cEOF2 becomes extremely high only when the zonally asymmetric variability of the SAM is considered.
 The Imaginary cEOF2 and the A-SAM share up to 92\% of their variance, reached at 225 hPa (Figure~\ref{fig:sam-eof-vertical}.b2).
-Such extremely high correlation between the asymmetric SAM and the Imaginary cEOF2 suggests that these are not different features that share a causal link, but that they might be different ways of characterising the same phenomenon.
-
-\begin{figure}
-\centering
-\includegraphics{../figures/fake-eof-1.pdf}
-\caption{\label{fig:fake-eof}Row a: The three components used to create a synthetic dataset of 41 years of geopotential height at 200 hPa.
-Row b: The three EOF computed from the synthetic dataset with the variance explained by each EOF in parenthesis.
-Units are arbitrary.}
-\end{figure}
-
-To test the hypothesis that a PSA-like feature can appear as statistic contamination of an otherwise zonally symmetrical and uncorrelated SAM, we create a synthetic dataset of 41 fields by reconstructing the 200 hPa geopotential field explained by the cEOF2 and adding a perfectly zonally symmetric SAM-like structure with random sign and value sampled with replacement from the real SAM index.
-To this synthetic dataset we then apply traditional EOF.
-By construction this dataset contains two independent components: a zonally symmetric standing oscillation and a PSA-like of varying amplitude and phase (Figure~\ref{fig:fake-eof} row a).
-The former one is a standing oscillation which could be represented by a single EOF, while the latter, being a wave which can change phase, needs two EOFs to be represented correctly.
-Thus, the full variance of this synthetic dataset can be perfectly explained by three EOFs.
-
-The three EOFs of the synthetic dataset are shown in Figure~\ref{fig:fake-eof} row b.
-The synthetic SAM can be seen in the first EOF while the cEOF2 is explained by the second and third.
-However, even though the synthetic SAM is perfectly zonally symmetric by construction, the leading EOF is not.
-The EOF decomposition mixes this mode with parts of the the zonally asymmetric cEOF2 pattern, resulting in a pattern eerily similar to the real SAM pattern.
-This happens because, although in this synthetic dataset the zonally symmetrical SAM and the cEOF2 are uncorrelated by construction, the sample correlation is not null.
-In this particular example, it is -0.064 (CI: -0.36 -- 0.25) with the Real part and -0.13 (CI: -0.42 -- 0.19) with the Imaginary part.
-Of note, the three EOFs are well separated according to North's rule of thumb \citep{north1982}.
-
-This suggests that the zonal asymmetries of the SAM could very well be mostly a statistical artefact.
-Since many surface impacts are mediated by the asymmetric component, as well as the relationship between SAM and ENSO \citep{campitelli2021}, this potential issue could affect the interpretability of many results involving the SAM.
-Studies on the relationship between the SAM and PSA-like patterns would be particularly difficult to interpret.
+Such extremely high correlation between the asymmetric SAM and the Imaginary cEOF2 suggests that these might be different ways of characterising the same phenomenon.
 
@@
 
-\caption{\label{tab:psa-eof2}Correlation between the Real and Imaginary parts of cEOF2 and the PSA1 and PSA2 modes computed as the second and third EOFs of seasonal geopotential height anomalies \citep{mo2001}.
-For the period 1979 -- 2019.}
+\caption{\label{tab:psa-eof2}Correlations between the Real and Imaginary parts of cEOF2 and the PSA1 and PSA2 modes computed as the second and third EOFs of seasonal geopotential height anomalies \citep[following][]{mo2001} for the 1979 -- 2019 period.
+95\% confidence intervals in parenthesis.
+Background colour indicates the value of correlation.}
 \centering
@@
 The relationship between PSAs and the modes described from the cEOFs is also studied.
-Figure~\ref{tab:psa-eof2} shows the relationship between the two PSA indices and the Real and Imaginary phase of cEOF2.
+Table~\ref{tab:psa-eof2} shows the correlations between the two PSA indices and the timeseries for Real and Imaginary phase of cEOF2.
 As anticipated by Figure~\ref{fig:eof2-regr-gh}, there is a strong correlation between PSA1 and Imaginary cEOF2, and between PSA2 and Real cEOF2.
 Conversely, there is no relationship between PSA1 and Real cEOF2, and between PSA2 and Imaginary cEOF2.
-So not only this cEOF2 represents well both the spatial structure and temporal evolution of the PSA modes, but it's also possible to make a rather clean association between its two phases and the two PSA modes.
+As a result, cEOF2 represents well both the spatial structure and temporal evolution of the PSA modes, as well as it is possible to make a rather clean association between its two phases and the two PSA modes.
+So, It could be concluded that the same particular rotation of cEOF2 that maximises the association between cEOF2 parts and PSA modes, is the one which maximises the relationship between ENSO and Imaginary cEOF2.
 
-This particular rotation of cEOF2 is the one which maximises the relationship between ENSO and Imaginary cEOF2.
-It is also the same that maximises this clean association between cEOF2 parts and PSA modes.
-
-The reason the conventional EOF analysis arrives at the same separation than our particular cEOF rotation is probably the fact that not all phases are equal.
-Figure~\ref{fig:phase-histogram} is an histogram that counts the number of SON years in which the cEOF2 was near a particular phase with the observations marked as rugs on the horizontal axis.
+The reason the conventional EOF analysis arrives at the same separation than our particular cEOF rotation is probably the fact that not all phases are equally frequent.
+To visualize that, Figure~\ref{fig:phase-histogram} shows an histogram that counts the number of SON years in which the cEOF2 was close to each of the four particular phases (positive/negative Real/Imaginary), with the observations for each season marked as rugs on the horizontal axis.
 For instance, years with cEOF2 phase within 45º of 0º are nearest the ``positive Real'' phase.
 About two thirds of time cEOF2 has a phase similar to either the negative or positive Imaginary phase.
+It could be seen that Imaginary phase is the most common phase and It is also the direction that has the maximum relationship with ENSO.
+Therefore, the Imaginary cEOF2 explains more variance than the Real cEOF2 and conventional EOF analysis will tend to separate the two.
 
@@
 \includegraphics{../figures/phase-histogram-1.pdf}
-\caption{\label{fig:phase-histogram}Histogram of phase distribution of cEOF2 for the period 1979 -- 2019. Bins are centred at 90º, 0º, -90º, -180º with a binwidth of 90º.}
+\caption{\label{fig:phase-histogram}Histogram of phase distribution of cEOF2 for the 1979 -- 2019 period.
+Bins are centred at 90º, 0º, -90º, -180º with a binwidth of 90º.
+The small vertical lines near the horizontal axis mark the observations.}
 \end{figure}
 
-This phase-locking behaviour indicates that the Imaginary phase both is the direction that has the maximum relationship with ENSO and is also the most common phase.
-Therefore, the Imaginary cEOF2 explains more variance than the Real cEOF2 and conventional cEOF analysis will tend to separate the two.
-
-This is similar to the method used by \citet{irving2016}.
-From their Fourier decomposition of reprojected meridional wind fields, they show that the phase distribution is bimodal and isolate the PSA pattern from the rest of the PSA-like variability by selecting events that are near the peacks of the distribution (compare our Figure~\ref{fig:phase-histogram} with their Figure 6).
+The result obtaneid here is similar to the one obtained by \citet{irving2016}.
+From their Fourier decomposition of reprojected meridional wind fields \textcolor{red}{¿En qué región?}, they show that the phase distribution is bimodal and isolate the PSA pattern from the rest of the PSA-like variability by selecting events that are near the peaks of the distribution (compare our Figure~\ref{fig:phase-histogram} with their Figure 6).
 
@@
 As a consequence, the cEOF2 offers an alternative way of representing the PSA which has several advantages over using the second and third EOFs.
+
+\hypertarget{is-the-asymmetric-sam-real}{%
+\subsubsection{Is the Asymmetric SAM real?}\label{is-the-asymmetric-sam-real}}
+
+\begin{figure}
+\centering
+\includegraphics{../figures/fake-eof-1.pdf}
+\caption{\label{fig:fake-eof}Row a: The three components used to create a synthetic dataset of 41 years of geopotential height at 200 hPa.
+Row b: The three EOF computed from the synthetic dataset with the variance explained by each EOF in parenthesis.
+Units are arbitrary.}
+\end{figure}
+
+To test the hypothesis that a PSA-like feature can appear as statistic contamination of an otherwise zonally symmetrical and uncorrelated annular pattern, we create a synthetic dataset of 41 fields by reconstructing the SON 200 hPa geopotential height field explained by the cEOF2 and adding a perfectly zonally symmetric annylar pattern with random sign and value sampled with replacement from the real SAM index.
+To this synthetic dataset we then apply traditional EOF.
+By construction this dataset contains two independent components: a zonally symmetric standing oscillation and a PSA-like of varying amplitude and phase (Figure~\ref{fig:fake-eof} row a).
+While the former is a standing oscillation which could be represented by a single EOF, the latter is a wave which can change phase, so traditional EOF analysis needs two EOFs to represented it correctly.
+Thus, the full variance of this synthetic dataset can be perfectly explained by three EOFs.
+
+The three EOFs of the synthetic dataset are shown in Figure~\ref{fig:fake-eof} row b.
+The annular pattern is captured mainly by the first EOF while the cEOF2 is explained by the second and third.
+However, even though the annular pattern used to construct these data is perfectly zonally symmetric, the leading EOF is not.
+The EOF decomposition mixes the true annular pattern with parts of the the zonally asymmetric cEOF2 pattern, resulting in a pattern pretty similar to the real SAM pattern \citep{fogt2020}.
+
+\textcolor{red}{Todo esto está mal.}
+
+Rotating makes it even worse.
+
+The reason this happens is that the spatial correlation between the annular pattern and the wave patterns is not zero; i.e.~they are not orthogonal. And since EOFs
+
+In this particular example, it is 0.022 (CI: -0.29 -- 0.33) with the Real part and 0.016 (CI: -0.29 -- 0.32) with the Imaginary part.
+Of note, the three EOFs are well separated according to North's rule of thumb \citep{north1982}.
+
+This suggests that the zonal asymmetries of the SON SAM could very well be mostly a statistical artefact.
+Since many surface impacts are mediated by the asymmetric component, as well as the relationship between SAM and ENSO \citep{campitelli2021}, this potential issue could affect the interpretability of many results involving the SAM.
+Studies on the relationship between the SAM and PSA-like patterns would be particularly difficult to interpret.
 
@@
 \begin{figure*}
-\includegraphics{../figures/sst-psi-2-1} \caption{Regression maps of cEOF2 with SST (column a) and streamfunction zonal anomalies with their corresponding activity wave flux (column b).
-Areas marked with dots have p-values smaller than 0.05 adjusted for FDR.}\label{fig:sst-psi-2}
+\includegraphics{../figures/sst-psi-2-1} \caption{Regression maps of cEOF2 with SST (column a) and streamfunction zonal anomalies (shaded) with their corresponding activity wave flux (vectors) (column b).
+Areas marked with dots have p-values smaller than 0.01 adjusted for FDR.}\label{fig:sst-psi-2}
 \end{figure*}
 
-Figure~\ref{fig:sst-psi-2} shows regression maps between cEOF2 and Sea Surface Temperatures (SST) Streamfunction at 200 hPa.
-The Imaginary cEOF2 is associated with strong positive SST anomalies on the Central Pacific and negative anomalies on an area across the North of Australia and New Zealand, the South Pacific Convergence Zone (SPCZ) (Figure~\ref{fig:sst-psi-2}.a2).
-This pattern is almost canonically positive ENSO and indeed, the correlation between the Imaginary cEOF2 and the Oceanic Niño Index \citep{bamston1997} is very high 0.76 (CI: 0.59 -- 0.87).
-Streamfunction anomalies show a coherent picture.
-The Imaginary cEOF2 is associated with strong wave-like streamfunction anomalies emanating from the tropics (Figure~\ref{fig:sst-psi-2}.b2).
-This is consistent with what we know of the effect of ENSO on the extratropics: SST anomalies initiate anomalous convection that excites Rossby waves that propagate meridionally towards higher latitudes \citep{mo2000}.
+The connections between cEOFs and Tropical sources is also assesed.
+Figure~\ref{fig:sst-psi-2} shows the regression maps of Sea Surface Temperatures (SST) and Streamfunction at 200 hPa respectively upon cEOF2.
+The Imaginary cEOF2 is associated with strong positive SST anomalies on the Central Pacific and negative anomalies over an area across the North of Australia and New Zealand, the South Pacific Convergence Zone (SPCZ) (Figure~\ref{fig:sst-psi-2}.a2).
+This pattern is almost canonically positive ENSO \citep{bamston1997} and indeed, the correlation between the Imaginary cEOF2 and the Oceanic Niño Index \citep{bamston1997} is very high 0.76 (CI: 0.59 -- 0.87).
+Besides, the Imaginary cEOF2 is associated with strong wave-like streamfunction anomalies emanating from the tropics (Figure~\ref{fig:sst-psi-2}.b2).
+This is consistent with the effect of ENSO on the extratropics: SST anomalies initiate anomalous convection that excites Rossby waves that propagate meridionally towards higher latitudes \citep{mo2000}.
 
-Since the Real cEOF2 represents just a different phase of the same wave train, one would expect that it would show a similar forcing pattern to the Imaginary cEOF with a slight translation of its location.
+\textcolor{red}{Revisa los labels de la colorbar para la figura 13.
+Además ojo que solamente sirven para la columna a, faltan los correspondientes a la streamfunction y el vector de referencia para wave flux}
+
+Since the Real cEOF2 represents just a different phase of the same wave train, it would be expected to show a similar forcing pattern to the Imaginary cEOF with a slight translation of its location.
 However, Figure~\ref{fig:sst-psi-2}.a1 and b1 show that the Real cEOF2 is not associated either with any significant SST nor streamfunction anomalies in the tropics.
 The correlation between the Real cEOF2 and ENSO is also not significant (-0.0042 (CI: -0.31 -- 0.3)).
-This lack of tropical signal suggests different natures of the different phases of the cEOF2 wave train.
+This lack of tropical signal suggests different natures of the different phases of the Real cEOF2 wave train, which has the maximum amplitudes in the surroundings of southern South America.
 
@@
 \includegraphics{../figures/enso-phase-1.pdf}
-\caption{\label{fig:enso-phase}Relationship between ENSO and phase of cEOF2 for the period 1979 -- 2019.
-Colours denote years with magnitude of cEOF2 greater or smaller than the 50th percentile.
-Black line is the fit ONI \textasciitilde{} sin(phase) computed by OLS weighted by the magnitude of cEOF2.}
+\caption{\label{fig:enso-phase} ONI plotted against cEOF2 phase for th 1979 -- 2019 period.
+Years with magnitude of cEOF2 greater or smaller than the 50th percentile are shown as orange circles and green diamomnds respectively.
+Black line is the fit ONI \textasciitilde{} sin(phase) computed by weighted OLS using the magnitude of the cEOF2 as weights.}
 \end{figure}
 
-To better explore the relationship between tropical forcing and phase of the cEOF2, Figure~\ref{fig:enso-phase} plots the ONI index and the phase of the cEOF2 for each SON trimester between 1979 and 2019, highlighting years in which the magnitude of cEOF2 is above the median.
+To better explore the relationship between tropical forcing and phase of the cEOF2, Figure~\ref{fig:enso-phase} shows the ONI index against the cEOF2 phase for each SON trimester between 1979 and 2019, highlighting years in which the magnitude of cEOF2 is above the median.
 In years with positive ONI, the phase of the cEOF2 is mostly around +90º (corresponding with positive imaginary part) and vice versa.
@@
 
-Furthermore, Figure~\ref{fig:enso-phase} suggest that strong cEOF2 years tend to coincide with strong ENSO years.
+\textcolor{red}{¿Qué significa para el fit "weighted by the magnitude of cEOF2"?}
+
+Furthermore, Figure~\ref{fig:enso-phase} suggest that higher amplitude cEOF2 years tend to coincide with stronger ENSO years.
 The correlation between the absolute magnitude of the ONI and the magnitude of the cEOF2 is 0.45 (CI: 0.17 -- 0.67).
@@
 Therefore, the relationship between the magnitude of the cEOF2 train wave and ONI remains uncertain.
+
+\textcolor{red}{Se que la cantidad de años hace bastante limitada la muestra, pero creo que es un resultado a tener en cuenta que en esos eventos muy intensos se alineen.
+¿Probaste hacer la correlación con todos los puntos naranjas?}
 
@@
 Figure~\ref{fig:sst-psi-1} shows SST and streamfunction regression maps for cEOF1.
-There is no significant pattern of SST anomalies associated with either the Real or Imaginary cEOF1.
+Unlike the cEOF2 case, there is no significant pattern of SST anomalies associated with either the Real or Imaginary cEOF1.
 Consistently, streamfunction anomalies do not show any tropical influence.
-Instead of that, the real and imaginary CEOF1 are associated with zonally wave activity fluxes in the extra-tropics, except for an equatorward flow from the coast of Antarctica around 150ºE in the real component.
+Instead of that, the real and imaginary CEOF1 are associated with zonally wave activity fluxes in the extra-tropics around 60ºS, except for an equatorward flow from the coast of Antarctica around 150ºE in the real component.
 
@@
 
-\begin{figure}
-\centering
-\includegraphics{../figures/pp-america-1.pdf}
-\caption{\label{fig:pp-america}Regression of SON mean precipitation anomalies in South America (mm per day, shaded) and (column a) cEOF1 the (row 1) Imaginary and (column 1) Real phase.
-For the 1979 -- 2018 period.
-Black contours with dots indicate areas with p-value smaller than 0.05 controlling for False Detection Rate.}
-\end{figure}
-
 The influence of cEOFs variability in continental rainfall of extra-tropical Southern Hemisphere is also explored.
 We focus on South America and Oceania sectors, because there are not relevant signals in Southern Africa (not shown).
-
-Figure~\ref{fig:pp-america} shows regression maps of seasonal precipitation with each cEOF in South America.
-cEOF1 is associated with slight decreases in precipitation in Southern Brazil and Paraguay, although these are not statistically significant \ref{fig:pp-america} column a.
-The strongest precipitation anomalies are the ones associated with the Imaginary cEOF2.
-The positive anomalies on Southeastern South America (SESA) and Central Chile, and negative anomalies over South Atlantic Convergence Zone is a well known springtime precipitation signature of ENSO \citep{cai2020a} and it is also similar to the precipitation anomalies associated with the A-SAM \citep{campitelli2021}.
-This is not surprising considering the close relationship between the ONI, the Asymmetric SAM index and the Imaginary cEOF2 shown previously, but further consolidates the identification of this mode with the PSA pattern.
-The Real cEOF2, on the other hand, is associated with negative precipitation anomalies in a smaller area of SESA.
-There is a phase dependence of the precipitation anomalies in SESA (not shown), which resembles the relationship between ONI and the phase of cEOF2.
-A sinusoidal fit has an \(r^2\)~of~0.27 with p-value~= 0.0022.
 
@@
 \includegraphics{../figures/pp-oceania-1.pdf}
-\caption{\label{fig:pp-oceania}As Figure~\ref{fig:pp-america} but for Oceania.}
+\caption{\label{fig:pp-oceania}Regression of SON mean precipitation anomalies in Oceania (mm per day, shaded) and (column a) cEOF1 the (row 1) Imaginary and (column 1) Real phase.
+For the 1979 -- 2019.
+Areas marked with dots have p-values smaller than 0.05 adjusted for False Detection Rate.}
 \end{figure}
 
-The precipitation anomalies in Oceania associated with cEOF1 are also weak and not statistically significant (Figure~\ref{fig:pp-oceania} column a) whereas EFO2 is associated with large and expansive anomalies.
+The precipitation anomalies in Oceania associated with cEOF1 are also weak and not statistically significant (Figure~\ref{fig:pp-oceania} column a) whereas EFO2 is associated with large and widespread anomalies.
 The Real cEOF2 is associated with positive seasonal anomalies in Northern and Eastern Australia, and negative ones over New Zealand, while precipitation anomalies associated with Imaginary cEOF2 are negative over all Eastern Australia.
@@
 \includegraphics{../figures/australia-pp-phase-1.pdf}
-\caption{\label{fig:australia-pp-phase}Same as Figure~\ref{fig:enso-phase} but for precipitation anomalies averaged between 130°E and 155°E and 40°S and 15°S (box shown in Figure~\ref{fig:pp-oceania} column b).}
+\caption{\label{fig:australia-pp-phase}Same as Figure~\ref{fig:enso-phase} but for precipitation anomalies averaged between 130°E and 155°E and 40°S and 15°S (box shown in Figure~\ref{fig:pp-oceania} column b).
+Black line is the fit precipitation \textasciitilde{} sin(phase) + cos(phase) computed by weighted OLS using the magnitude of the cEOF2 as weights.}
 \end{figure}
 
-To better understand the relationship between precipitation anomalies over Western Australia and the cEOF2, Figure~\ref{fig:australia-pp-phase} plots precipitation anomalies averaged over Eastern Australia (box in Figure~\ref{fig:pp-oceania}) as a function of cEOF2 phase.
-The black line shows a sinusoidal fit, whose coefficients we fitted by weighted least squares using the magnitude of the EFO2 as weight.
-The \(r^2\) corresponding to the fit is 0.37, with p-value \textless{} 0.001.
-There is a strong relationship between the phase of the cEOF2 and precipitation anomalies in Eastern Australia, but Real and Imaginary phases we chose are not exactly aligned with the direction that maximises this relationship.
-If one were to perform a more in-depth analysis of the relationship between this pattern and precipitation in this region, one would conceivably align one of the axis (either the Real or the Imaginary) to the one that maximises it.
+To better understand the relationship between precipitation anomalies over Eastern Australia and the cEOF2, Figure~\ref{fig:australia-pp-phase} plots precipitation anomalies averaged over Eastern Australia (box in Figure~\ref{fig:pp-oceania}) as a function of cEOF2 phase.
+The black line shows a sinusoidal fit, whose coefficients are fitted by weighted least squares using the magnitude of the EFO2 as weight.
+The \(r^2\) corresponding to the fit is 0.41 (p-value \textless{} 0.001).
+There is a strong relationship between the phase of the cEOF2 and precipitation anomalies in Eastern Australia, but the Real and Imaginary phases we chose are not exactly aligned with the direction that maximises this relationship.
+To perform a more in-depth analysis of the relationship between this pattern and precipitation in this region, it would be recommended to align one of the axis (either the Real or the Imaginary) to the phase that maximises the described relationship.
+
+\begin{figure}
+\centering
+\includegraphics{../figures/pp-america-1.pdf}
+\caption{\label{fig:pp-america}As Figure~\ref{fig:pp-america} but for South America}
+\end{figure}
+
+Figure~\ref{fig:pp-america} shows regression maps of seasonal precipitation with each cEOF in South America.
+cEOF1 is associated with slight decreases \textcolor{red}{Revisa como estás expresando toda esta parte (y en el resto del texto), recorda que son regresiones así que no corresponde decir postive/negative sino haces alguna referencia a la variable con las que los estás regresionando.} in precipitation in Southern Brazil and Paraguay, although these are not statistically significant (Fig.~\ref{fig:pp-america} column a).
+The strongest precipitation regression coefccients are the ones associated with the Imaginary cEOF2.
+The positive anomalies on Southeastern South America (SESA) and Central Chile, and negative anomalies over South Atlantic Convergence Zone is a well known springtime precipitation signature of ENSO \citep{cai2020a} and it is also similar to the precipitation anomalies associated with the A-SAM \citep{campitelli2021}.
+This is not surprising considering the close relationship between the ONI, the A-SAM index and the Imaginary cEOF2 shown previously, but further consolidates the identification of this mode with the PSA pattern.
+On the other hand, the Real cEOF2 is associated with negative precipitation anomalies in a smaller area of SESA.
+Resembling the relationship between ONI and the phase of cEOF2 (Fig.~\ref{fig:enso-phase}), there is a cEOF2 phase dependence of the precipitation anomalies in SESA (not shown).
+These variable could be related wth a signficant sinusoidal fit that has a coefficient of determination of~0.27 (p-value~= 0.0022).
+
+\textcolor{red}{La formula en la figura trata de escribirla en una forma más precisa "precipitación"/"phase" son muy generales, y también teniendo en cuenta las unidades.
+También incluilo en el caption.}
 
@@
 We further show that the Imaginary cEOF2 is closely related to the Southern Annular Mode in the troposphere.
-In fact, it is a virtually perfect indicator of the zonally asymmetric portion of the SAM.
-To the extreme that the leading EOF of geopotential height with the cEOF2 removes results in an annular mode that has none of the typical zonal asymmetries of the observed SAM.
+In fact, it has a closer resemblance of the zonally asymmetric portion of the SAM.
 
@@
 However, this underscore the benefit of using complex EOF, since it would trivial to rotate it.
+
+\textcolor{red}{Debería cerrarse con algún parráfo que diga qué tiene de bueno trabajar con esta metodología y qué permitiría hacer a futuro (se puede poner que sería deseable extenderlo a otras estaciones, etc.)}
 

### 2021-07-26 analysis/paper/shceof.Rmd
Quita lo de si existe A-SAM
Closes #50

@@
 
-guide_colorstrip_bottom <- function(width = 15, height = 0.5, ...) {
-  guide_colorstrip(title.position = "top", title.hjust = 0.5,
+guide_colorstrip_bottom <- function(width = 15, height = 0.5, 
+                                    title.position = "top",...) {
+  guide_colorstrip(title.position = title.position, title.hjust = 0.5,
                    barheight = height,
@@
 
-guide_colorsteps_bottom <- function(width = 15, height = 0.5, even.steps = FALSE, ...) {
+guide_colorsteps_bottom <- function(width = 15, height = 0.5, even.steps = FALSE, 
+                                    title.position = "top", ...) {
   guide_colorsteps(show.limits = TRUE, even.steps = even.steps,
-                   title.position = "top", title.hjust = 0.5,
+                   title.position = title.position, title.hjust = 0.5,
                    barheight = height,
@@
   .[, .(hgt = mean(hgt*w)), by = .(lon, lat, time = seasonally(time))] %>% 
-  .[, value := Anomaly(hgt)*sqrt(cos(lat*pi/180)), by = .(lon, lat, season(time))] %>%
+  .[, value := Anomaly(hgt)*sqrt(cos(lat*pi/180)), 
+    by = .(lon, lat, season(time))] %>%
   EOF(value ~ time | lon + lat, n = 1:3, data = .)
@@
 
-```{r eof-filtered, fig.cap = "Spatial pattern of the leading EOF of 200 hPa geopotential height with the variability of cEOF2 filtered out South of 20º.
-Arbitrary units.", fig.width=2.5, fig.height=2.5}
+```{r eof-filtered, eval=FALSE, fig.cap="Spatial pattern of the leading EOF of 200 hPa geopotential height with the variability of cEOF2 filtered out South of 20º.
+Arbitrary units.", fig.height=2.5, fig.width=2.5, include=FALSE}
 cEOF2 <- ceof$eof[[1]]$left[cEOF == "cEOF2"] %>% 
@@
 
-# eof_filtered %>% 
-#   .$left %>% 
-#   periodic(lon = c(0, 360)) %>% 
-#   ggplot(aes(lon, lat)) +
-#   geom_contour_fill(aes(z = eof), breaks = AnchorBreaks(0, exclude = 0)) +
-#   geom_contour_tanaka(aes(z = eof), breaks = AnchorBreaks(0, exclude = 0)) +
-#   scale_fill_divergent(guide = "none") +
-#   geom_qmap(~.x[lat <= -20]) +  
-#   scale_x_longitude() +
-#   scale_y_latitude() +
-#   coord_polar() +
-#   theme(axis.text = element_blank())
+eof_filtered %>%
+  .$left %>%
+  periodic(lon = c(0, 360)) %>%
+  ggplot(aes(lon, lat)) +
+  geom_contour_fill(aes(z = eof), breaks = AnchorBreaks(0, exclude = 0)) +
+  geom_contour_tanaka(aes(z = eof), breaks = AnchorBreaks(0, exclude = 0)) +
+  scale_fill_divergent(guide = "none") +
+  geom_qmap(~.x[lat <= -20]) +
+  scale_x_longitude() +
+  scale_y_latitude() +
+  coord_polar() +
+  theme(axis.text = element_blank())
+
+eof_filtered_cor %>% 
+  .[, .(term, text)] %>%
+  .[, term := factor(term, labels = c("SAM", "A-SAM", "S-SAM"))] %>%
+  knitr::kable(col.names = c("Índice SAM", "Correlación"))
 ```
@@
 psa2 <- psa$left[PC == "PSA2"][season(time) == "SON"]
+cEOF2 <- ceof$eof[[1]]$left[cEOF == "cEOF2"] %>% 
+  copy() %>% 
+  .[, cEOF := NULL] %>% 
+  .[]
 
@@
 
-The result obtaneid here is similar to the one obtained by @irving2016.
+The result obtained here is similar to the one obtained by @irving2016.
 From their Fourier decomposition of reprojected meridional wind fields \textcolor{red}{¿En qué región?}, they show that the phase distribution is bimodal and isolate the PSA pattern from the rest of the PSA-like variability by selecting events that are near the peaks of the distribution (compare our Figure \@ref(fig:phase-histogram) with their Figure 6).
@@
 As a consequence, the cEOF2 offers an alternative way of representing the PSA which has several advantages over using the second and third EOFs.
-
-### Is the Asymmetric SAM real?
-
-(ref:fake-eof-cap) Row a: The three components used to create a synthetic dataset of 41 years of geopotential height at 200 hPa.
-Row b: The three EOF computed from the synthetic dataset with the variance explained by each EOF in parenthesis.
-Units are arbitrary.
-
-```{r make_fake_eof}
-cEOF2 <- ceof$eof[[1]]$left[cEOF == "cEOF2"] %>% 
-  copy() %>% 
-  .[, cEOF := NULL] %>% 
-  .[]
-
-pattern <- sams[term == "full" & lev == 200] %>% 
-  .[era5[lev == 200], on = c("lev", "time")] %>% 
-  .[, FitLm(hgt, estimate), by = .(lon, lat, lev)] %>% 
-  rm_intercept() %>% 
-  .[, estimate := mean(estimate), by = .(lat)] %>% 
-  .[, term := NULL] %>% 
-  setnames("estimate", "eof")
-
-set.seed(42)
-
-fake_sam <-  sams[term == "full" & lev == 200] %>% 
-  copy() %>% 
-  .[, value := sample(estimate, replace = TRUE)] %>% 
-  .[, estimate := NULL] %>% 
-  .[, term := NULL]
-
-# complement <- function(x, y) {
-#   
-#   e <- residuals(lm(x ~ y))
-#   
-#   e*sd(x) + mean(x)
-#   
-# }
-# 
-# fake_sam <-  sams[term == "full" & lev == 200] %>% 
-#   copy() %>% 
-#   .[, value := complement(estimate, y)] %>% 
-#   .[, estimate := NULL] %>% 
-#   .[, term := NULL]
-
-cor_with_eof2 <- fake_sam[cEOF2, on = "time"] %>% 
-  sep_ReIm() %>% 
-  .[, correlate(value, hgt), by = part]
-
-fake_sam <- fake_sam %>% 
-  .[pattern, on = "lev", allow.cartesian = TRUE] %>% 
-  .[, sam := value*eof]
-
-# fake_sam %>% 
-#   .[time %in% unique(time)[1:4]] %>% 
-#   ggplot(aes(lon, lat)) +
-#   geom_contour_fill(aes(z = sam)) +
-#   facet_wrap(~factor(time)) +
-#   coord_polar()
-
-fake_eof <- era5 %>% 
-  .[sep_ReIm(cEOF2, format = "wide")[, hgt := NULL], on = .NATURAL] %>% 
-  .[lev == 200] %>% 
-  .[lat <= -20] %>% 
-  .[season(time) == "SON"] %>% 
-  .[, hgt := hgt - .lm.fit(cbind(1, Imaginary, Real), hgt)$residuals,
-    .(lon, lat, lev, season(time))] %>% 
-  .[, hgt := Anomaly(hgt), by = .(lat)] %>% 
-  fake_sam[., on = .NATURAL] %>% 
-  .[, hgt := hgt + sam] %>% 
-  .[, eof := Anomaly(hgt)*sqrt(cos(lat*pi/180)), by = .(lon, lat, season(time))] %>% 
-  EOF(eof ~ lon + lat | time, n = 1:3, data = ., suffix = "EOF")  
-```
-
-```{r fake-eof, fig.cap = "(ref:fake-eof-cap)", fig.width=3.7, fig.height=3.3}
-north_thumb <- function(sdev, n){
-  sdev <- sdev^2
-  delta <- sqrt(2 / n) * sdev
-
-  # Separation between each pair of eofs must be 
-  # greater than the sum of their errors.
-  separation <- abs(diff(sdev))
-  separation >  delta[-length(delta)] + delta[-1]
-}
-
-separated <- north_thumb(fake_eof$sdev$sd^2, 41)
-stopifnot(all(separated))
-
-var <- fake_eof$sdev %>%
-  copy() %>% 
-    .[, setNames(paste0(EOF, " (", scales::percent(r2), ")"),
-                 EOF)]
-var <- c(var, setNames(c("Annular Pattern", "Imaginary cEOF2", "Real cEOF2"), c("SAM", "Imaginary", "Real")))
-
-era5 %>%
-  .[sep_ReIm(cEOF2, format = "wide")[, hgt := NULL], on = .NATURAL] %>%
-  .[lev == 200] %>%
-
-  .[season(time) == "SON"] %>%
-  .[, FitLm(hgt, Real, Imaginary), .(lon, lat)] %>%
-  rm_intercept() %>%
-  .[, .(lon, lat, eof = estimate, EOF = term)] %>%
-  rbind(pattern[, .(lon, lat, eof, EOF = "SAM")]) %>%
-  rbind(fake_eof$left, use.names = TRUE) %>% 
-   .[lat <= -20] %>%
-  .[, EOF := factor(EOF, c("SAM", "Imaginary", "Real", paste0("EOF", 1:3)),
-                    ordered = TRUE)] %>% 
-  .[EOF %in% c("EOF1", "EOF3"), eof := -eof] %>% 
-  .[!(EOF %in% c("SAM", "Imaginary", "Real")), eof := eof/sd(eof)] %>% 
-  .[EOF %in% c("SAM", "Imaginary", "Real"), eof := eof/sd(eof)] %>%
-
-  # geom_contour_tanaka has problem with the perfectly symemtric SAM
-  # adding just a tiny bit of noise solves it.
-
-  .[, eof := scale(eof), by = EOF] %>% 
-  .[EOF == "SAM", eof := eof + rnorm(.N, sd = 1e-5)] %>%
-  .[EOF %in% c("EOF1", "EOF2"), eof := -eof] %>% 
-  periodic(lon = c(0, 360)) %>%
-  ggplot(aes(lon, lat)) +
-  geom_contour_fill(aes(z = eof, fill = ..level..), global.breaks = TRUE, 
-                    breaks = AnchorBreaks(0, 0.5, exclude = 0)) +
-
-  geom_contour_tanaka(aes(z = eof), global.breaks = TRUE,
-                      breaks = AnchorBreaks(0,0.5,  exclude = 0)) +
-  scale_fill_divergent_discretised(NULL, guide = guide_colorsteps_bottom(), 
-                                   limits = c(-4, 4)) +
-  geom_qmap(~.x[lat <= -20]) +
-  scale_x_longitude() +
-    scale_y_latitude(labels = NULL,
-                   minor_breaks = NULL) +
-  grid_y +
-  coord_polar() +
-  facet_wrap(~EOF, ncol = 3, labeller = labeller(EOF = var)) +
-  axis_labs_smol +
-  theme(axis.text = element_blank()) +
-  tag_facets("rc")
-```
-
-To test the hypothesis that a PSA-like feature can appear as statistic contamination of an otherwise zonally symmetrical and uncorrelated annular pattern, we create a synthetic dataset of `r uniqueN(fake_sam$time)` fields by reconstructing the SON 200 hPa geopotential height field explained by the cEOF2 and adding a perfectly zonally symmetric annylar pattern with random sign and value sampled with replacement from the real SAM index.
-To this synthetic dataset we then apply traditional EOF.
-By construction this dataset contains two independent components: a zonally symmetric standing oscillation and a PSA-like of varying amplitude and phase (Figure \@ref(fig:fake-eof) row a).
-While the former is a standing oscillation which could be represented by a single EOF, the latter is a wave which can change phase, so traditional EOF analysis needs two EOFs to represented it correctly.
-Thus, the full variance of this synthetic dataset can be perfectly explained by three EOFs.
-
-The three EOFs of the synthetic dataset are shown in Figure \@ref(fig:fake-eof) row b.
-The annular pattern is captured mainly by the first EOF while the cEOF2 is explained by the second and third.
-However, even though the annular pattern used to construct these data is perfectly zonally symmetric, the leading EOF is not.
-The EOF decomposition mixes the true annular pattern with parts of the the zonally asymmetric cEOF2 pattern, resulting in a pattern pretty similar to the real SAM pattern [@fogt2020].
-
-\textcolor{red}{Todo esto está mal.}
-
-Rotating makes it even worse.
-
-The reason this happens is that the spatial correlation between the annular pattern and the wave patterns is not zero; i.e. they are not orthogonal. And since EOFs
-
-In this particular example, it is `r cor_with_eof2[part == "Real"]$text` with the Real part and `r cor_with_eof2[part == "Imaginary"]$text` with the Imaginary part.
-Of note, the three EOFs are well separated according to North's rule of thumb [@north1982].
-
-This suggests that the zonal asymmetries of the SON SAM could very well be mostly a statistical artefact.
-Since many surface impacts are mediated by the asymmetric component, as well as the relationship between SAM and ENSO [@campitelli2021], this potential issue could affect the interpretability of many results involving the SAM.
-Studies on the relationship between the SAM and PSA-like patterns would be particularly difficult to interpret.
 
@@
     geom_contour_pval(aes(z = p_val)) +
-    geom_streamline(aes(dx = fx, dy = fy), min.L = 3, L = 10, n = 30,
-                    size = 0.2, res = 2, nx = 25, ny = 15, 
+    geom_streamline(aes(dx = fx, dy = fy), min.L = 3, L = 10,
+                    size = 0.1, res = 2, nx = 22, ny = 12, 
                     arrow.angle = 14, arrow.length = 0.2) +
     geom_qmap(keep = .015) +
-    scale_fill_divergent_discretised("Streamfunction", guide = "none") +
+    scale_fill_divergent_discretised("Streamfunction", 
+                                     guide = guide_colorstrip_bottom(20, order = 1)) +
     ggnewscale::new_scale_fill() +
@@
     geom_contour_fill(data = sst_gdata,
-                      aes(z = estimate, fill = ..level..), breaks = AnchorBreaks(exclude = 0)) +
+                      aes(z = estimate, fill = ..level..), 
+                      breaks = AnchorBreaks(exclude = 0)) +
     geom_contour_pval(data = sst_gdata, aes(z = p_val)) +
-    scale_fill_divergent_discretised("SST", guide = guide_colorsteps_bottom(order = 2)) +
+    scale_fill_divergent_discretised("SST", 
+                                     guide = guide_colorsteps_bottom(order = 2)) +
     scale_x_longitude() +
@@
     coord_quickmap(ylim = c(NA, 10)) +
-    theme(legend.box = "horizontal") +
+    theme(legend.box = "vertical") +
     axis_labs_smol +

### 2021-07-26 analysis/paper/shceof.Rmd
Agregrega amplitud de ENSO en conclusiones
Closes #53

@@
 
-```{r enso-phase, fig.cap = " ONI plotted against cEOF2 phase for th 1979 -- 2019 period.
+```{r enso-phase, fig.cap = "ONI plotted against cEOF2 phase for th 1979 -- 2019 period.
 Years with magnitude of cEOF2 greater or smaller than the 50th percentile are shown as orange circles and green diamomnds respectively.
@@
 
-\textcolor{red}{¿Qué significa para el fit "weighted by the magnitude of cEOF2"?}
-
 Furthermore, Figure \@ref(fig:enso-phase) suggest that higher amplitude cEOF2 years tend to coincide with stronger ENSO years.
@@
 Therefore, the relationship between the magnitude of the cEOF2 train wave and ONI remains uncertain.
-
-\textcolor{red}{Se que la cantidad de años hace bastante limitada la muestra, pero creo que es un resultado a tener en cuenta que en esos eventos muy intensos se alineen.
-¿Probaste hacer la correlación con todos los puntos naranjas?}
 
@@
   scale_fill_divergent_discretised("SST",
-                                   limits = c(-.3, .3),
+                                   limits = c(-1, 1),
                                    guide = guide_colorsteps_bottom(20, order = 2)) 
@@
 
-THe second complex EOF represents a wave-3 pattern with maximum magnitude in the Pacific sector.
+The second complex EOF represents a wave-3 pattern with maximum magnitude in the Pacific sector.
 Essentially, it is an alternate representation of the PSA1 and PSA2 patterns [@mo2001].
@@
 This is because the when ENSO is active, it forces the cEOF2 to be in the Imaginary phase.
-When ENSO is neutral, the cEOF2 is still active, but with no particular phase.
+When ENSO is neutral, the cEOF2 is still active, but with no particular preferred phase.
 This mirrors the results of @cai2002, who showed that the CSIRO Model can create PSA-like variability even in the absence of ENSO forcing (with a climatological run), but the variability of one of the PSA modes was enhanced when adding the ENSO signal.
+The magnitude of the cEOF2 appears not to be related to the magnitude of ENSO, except frot the three strongest ENSO events in the period (`r knitr::combine_words(year(gdata[order(-oni)][1:3]$time))`) which coincide with the three strongest cEOF2 years.
 

### 2021-07-26 analysis/paper/shceof.Rmd
Agrega escala correcta de función corriente
Closes #51

@@
 ```{r compute-psi}
+# Scaling magical numbers. (come from psi-test.Rmd)
+# Not really important, though.
+scaling <- c(add = -2.364646e+06, mult = 3.954828e+13)
+
 psi <- ReadNetCDF(ERA5(), vars = "vo",
@@
   setnames("value", "psi") %>% 
+  .[, psi := psi*scaling[2] + scaling[1]] %>%
   .[, psi.z := Anomaly(psi), by = .(time, lev, lat)] 
@@
     ggplot(aes(lon, lat)) +
-    geom_contour_fill(aes(z = estimate*1e8, fill = ..level..),
+    geom_contour_fill(aes(z = estimate*1e-7, fill = ..level..),
                       breaks = AnchorBreaks(exclude = 0)) +
@@
     geom_qmap(keep = .015) +
-    scale_fill_divergent_discretised("Streamfunction", 
+    scale_fill_divergent_discretised("Streamfunction (*1e-7)", 
                                      guide = guide_colorstrip_bottom(20, order = 1)) +

### 2021-07-26 analysis/paper/shceof.Rmd
🙄 "En la sección métodos están los médotos..."

@@
 
-\textcolor{red}{The data and methods are described in Section 2 :P ...}
-
-# Data and Methods
+In Section \@ref(methods) we describe the methods.
+In Section \@ref(temporal) we describe the temporal variability and vertical coherence of the indices.
+In Section \@ref(spatial), we analyse the spatial patterns of each complex EOF.
+In Section \@ref(regressions) we study the spatial regressions with geopotential height and temperatuer. 
+In Section \@ref(other-variables), we analyise the relationship between cEOF1 and Ozone, and between EOF2 and the Southern Annular Mode and the PSA. 
+In Section \@ref(tropical) we look for tropical forcings that explain the variability of each cEOF. 
+Finally, in Section \@ref(precipitation) we show the relationship between these modes of variabiltiy and precipitation in South America and Oceania.
+
+# Data and Methods {#methods}
 
@@
 
-## cEOF spatial patterns
+## cEOF spatial patterns {#spatial}
 
@@
 
-## cEOFs Regressions
+## cEOFs Regressions {#regressions}
 
@@
 
-## cEOFs relationship with other variables
+## cEOFs relationship with other variables {#other-variables}
 
@@
 
-## Tropical sources
+## Tropical sources  {#tropical}
 
@@
 
-## Precipitation
+## Precipitation {#precipitation}
 

### 2021-07-26 analysis/paper/shceof.Rmd
Arregla ecuación en precipitación
(#54)

@@
 
-EOFs are more flexible than Fourier descomposition modes in the sense that they can characterise \textcolor{red}{Esto hay que exresarlo más claramente}.
+EOFs are more flexible than Fourier descomposition modes in the sense that they can capture oscillation patterns that cannot be characterised by purely sinusoidal waves with constant amplitude.
 Nonetheless, they are restricted to standing oscillation modes and could not represent properly propagating or phase-varying modes such as zonal waves.
-Besides, a single EOF can also represent a mixture of two or more physical modes.
+A single EOF can also represent a mixture of two or more physical modes.
 The inability of a single EOF to describe travelling waves is what forces the PSA pattern to be described by two EOFs.
@@
     geom_qmap(keep = .015) +
-    scale_fill_divergent_discretised("Streamfunction (*1e-7)", 
+    scale_fill_divergent_discretised("Streamfunction (×1e-7)", 
                                      guide = guide_colorstrip_bottom(20, order = 1)) +
@@
 
-(ref:australia-pp-phase-cap) Same as Figure \@ref(fig:enso-phase) but for precipitation anomalies averaged between `r knitr::combine_words(LonLabel(box_australia$lon))` and `r knitr::combine_words(LatLabel(box_australia$lat))` (box shown in Figure \@ref(fig:pp-oceania) column b).
+(ref:australia-pp-phase-cap) Same as Figure \@ref(fig:enso-phase) but for precipitation anomalies averaged between `r knitr::combine_words(LonLabel(box_australia$lon))` and `r knitr::combine_words(LatLabel(box_australia$lat))` (box shown in Figure \@ref(fig:pp-oceania) column b) in $\frac{\mathrm{mm}}{\mathrm{day}}$.
 Black line is the fit precipitation ~ sin(phase) + cos(phase) computed by weighted OLS using the magnitude of the cEOF2 as weights.
@@
 
-equation <- latex2exp::TeX(paste0("$\\mathrm{Precipitation} = ", 
-                                  signif(coef(model)[1], 2), "\\frac{mm}{day} \\sin{(\\mathrm{phase})} + ", 
-                                  signif(coef(model)[2], 2), "\\frac{mm}{day} \\cos{(\\mathrm{phase})}$"), output = "text")
+equation <- latex2exp::TeX(paste0("$\\mathrm{Precipitation}_{\\mathrm{West Australia}} = ", 
+                                  signif(coef(model)[1], 2), "\\sin{(\\mathrm{phase})} + ", 
+                                  signif(coef(model)[2], 2), "\\cos{(\\mathrm{phase})}$"), 
+                           output = "text")
 
@@
 
-\textcolor{red}{La formula en la figura trata de escribirla en una forma más precisa "precipitación"/"phase" son muy generales, y también teniendo en cuenta las unidades.
-También incluilo en el caption.}
-
 (ref:pp-africa-cap) Same as Figure \@ref(fig:pp-america) but for South of Africa.
@@
 This mirrors the results of @cai2002, who showed that the CSIRO Model can create PSA-like variability even in the absence of ENSO forcing (with a climatological run), but the variability of one of the PSA modes was enhanced when adding the ENSO signal.
-The magnitude of the cEOF2 appears not to be related to the magnitude of ENSO, except frot the three strongest ENSO events in the period (`r knitr::combine_words(year(gdata[order(-oni)][1:3]$time))`) which coincide with the three strongest cEOF2 years.
+The magnitude of the cEOF2 appears not to be related to the magnitude of ENSO, except frot the three strongest ENSO events in the period which coincide with the three strongest cEOF2 years.
 

### 2021-07-27 analysis/paper/shceof.Rmd
Cambia orden de secciones
Muchos cambios

@@
 ---
-title: " Austral Spring Extratropical Southern Hemisphere zonally assymetric circulation"
+title: "Austral Spring Extratropical Southern Hemisphere zonally assymetric circulation"
 titlerunning:  Title
@@
 )
-
-# ## Figure classes
-# DefineClass <- function(class.opts) {
-#     class.name <- deparse(substitute(class.opts))
-#     class.fun <- function(options) {
-#         class <- options[[class.name]]
-#         class <- class.opts[[class]]
-#         options[names(class)] <- class
-#         options
-#     }
-#     invisible(knitr::opts_hooks$set(
-#         setNames(list(class.fun), class.name)))
-# }
-# 
-# page.width <- 210 - 30 - 25
-# text.width <- page.width - 20
-# text.height <- page.height <- 297 - 25 - 25
-# 
-# fig.class <- list(
-#   colwidth = list(
-#     fig.width = 5
-#   ),
-#     pagewidth = list(
-#         fig.width = page.width/25.4,
-#         out.width = paste0(page.width, "mm"),
-#         # fig.align = "center",
-#         fig.env = "figure*",
-#         out.extra = " "
-#     )
-# )
-# 
-# DefineClass(fig.class)
 
@@
 
+panel_background <- theme(panel.background = element_rect(fill = "#fbfbfb", color = NA), 
+                          panel.ontop = FALSE,
+                          tagger.panel.tag.background = ggplot2::element_rect(color = NA, 
+                                                                              fill = "#fbfbfb"))
+
 grid <- theme(panel.grid = element_line(color = "gray10", size = 0.4, linetype = 3))
@@
 grid_y <- theme(panel.grid.major.y = element_line(color = "gray60", size = 0.1))
-
-panel_background <- theme(panel.background = element_rect(fill = "#fbfbfb", color = NA), 
-                          panel.ontop = FALSE,
-                          panel.grid = element_line(size = 0.5, color = "gray90"),
-                          tagger.panel.tag.background = ggplot2::element_rect(color = NA, 
-                                                                              fill = "#fbfbfb"))
 
@@
                    barheight = height,
-                   barwidth = width, ...)
+                   barwidth = width, frame.colour = "black", ...)
 }
@@
     geom_contour2(mapping, breaks = p.value, size = size, ...)
+  )
+}
+
+geom_coords <- function() {
+  list(
+    no_grid,
+    annotate("segment", 
+             x = seq(0, 360 - 30, by = 30),
+             xend =  seq(0, 360 - 30, by = 30),
+             y = -90 + 15,
+             yend = Inf,
+             size = 0.1, alpha = 0.5),
+    geom_hline(yintercept = seq(-90, 0, by = 15), size = 0.1, alpha = 0.5),
+    shadowtext::geom_shadowtext(data = data.frame(x = 0, y = seq(-90 + 15, 0, 
+                                                                 by = 15)),
+                                aes(x, y, label = LatLabel(y)), size = 1.5, 
+                                alpha = 0.7,
+                                colour = "black",
+                                bg.colour = "white")
+
   )
@@
 This approach suggests that zonal waves 1 and 3 explained most of the variance of the tropospheric extratropical Southern Hemisphere circulation (@vanloon1972).
-However, this methodology rely on the assumption that the circulation can be meaningfully understood in terms of zonal waves of constant amplitude along a latitude circle.
+However, this methodology relies on the assumption that the circulation can be meaningfully understood in terms of zonal waves of constant amplitude along a latitude circle.
 This is not valid for meridionally propagating waves or zonal waves with localised amplitudes.
@@
 
-EOFs are more flexible than Fourier descomposition modes in the sense that they can capture oscillation patterns that cannot be characterised by purely sinusoidal waves with constant amplitude.
+EOFs are more flexible than Fourier decomposition modes in the sense that they can capture oscillation patterns that cannot be characterised by purely sinusoidal waves with constant amplitude.
 Nonetheless, they are restricted to standing oscillation modes and could not represent properly propagating or phase-varying modes such as zonal waves.
@@
 Similarly, @hobbs2010 propose that the zonally asymmetric circulation in the Southern Hemisphere can be described by the strength and location of two anticyclones located in the sub-Antarctic western hemisphere.
-These methods are grounded on other methods to identify the centers of action for the described phenomena and can be useful to characterise features that are not readily apparent with Fourier or EOF methods.
-The use of data of only specific regions or locations makes them suitable for easy computation and for extending into less quality data from the past, but also makes them not very robust and could not capture non-stationary behaviours of the patterns.
-
-\textcolor{red}{Falta un parráfo que deje más claro por qué es necesario considerar un enfoque alternativo y cuál va a ser el beneficio de ello.
-También agregar qué se estudiara conjuntamente la troposfera y la estratósfera }
-
-\textcolor{red}{Falta el objetivo que justifica el uso de la metodología}
-
-In this paper we propose using complex Empirical Orthogonal Functions [@horel1984] as a robust extension of EOF analysis that is capable of describing rich structures with meridional and zonal propagation, and amplitude modulation such as the PSA.
-We apply this method to characterise the Springtime zonally asymmetric circulation of the extratopical Southern Hemisphere.
+These methods are grounded on other methods to identify the centres of action for the described phenomena and can be useful to characterise features that are not readily apparent with Fourier or EOF methods.
+The use of data of only specific regions or locations makes them suitable for easy computation and for extending back into the data-poor past, but also makes them not very robust and unable to capture non-stationary patterns.
+
+Our objective is to create an index that is able to capture the phase-varying nature of planetary waves and is flexible enough to describe planetary waves with non-constant amplitude in each latitude circle.
+We use complex Empirical Orthogonal Functions [@horel1984] as a robust extension of EOF analysis that is capable of describing rich structures with meridional and zonal propagation, and amplitude modulation such as the PSA.
+
+We apply this method to characterise the Springtime zonally asymmetric circulation of the extratropical Southern Hemisphere.
+To take vertical distribution into account, we analyse both the troposphere and the stratosphere. 
+We focus the analysis to the September-October-November because that's the trimester in which tropical forcing of zonal waves is maximised (cita).
 
@@
   scale_y_latitude(labels = NULL,
-                   minor_breaks = NULL) +
+                   minor_breaks = NULL, 
+                   limits = c(NA, -20)) +
   scale_x_longitude() +
+  geom_coords() +
   coord_polar() +
@@
   facet_wrap(.~EOF, ncol = 2)  +
-  tagger::tag_facets("rc")  +
-  theme(panel.spacing = grid::unit(-.7, "lines")) +
-  grid_y
+  tagger::tag_facets("rc")
+
 ```
@@
 Table \@ref(tab:corr-ceof-splitted) shows the coefficient of determination between time series of the amplitude of each complex EOF (cEOF) across levels.
-There's a high degree of correlation between the respective cEOF1 and cEOF2 at each level.
+There's a high degree of correlation between the magnitude of the respective cEOF1 and cEOF2 at each level.
 The spatial patterns of the 50 hPa and 200 hPa cEOFs are also similar (not shown).
@@
   plot_annotation(tag_levels = "a", tag_suffix = ")")
+
 ```
@@
   scale_fill_divergent_discretised(guide = guide_colorsteps_bottom()) +
-
   geom_qmap(~ .x[lat <= -20]) +
-  scale_x_longitude() +
+  scale_x_longitude(labels = NULL) +
   scale_y_latitude(labels = NULL,
+                   limits = c(NA, -20),
                    minor_breaks = NULL) +
@@
   coord_polar() +
-  axis_labs_smol +
-  tag_facets(tag = "cr") +
-  grid_y
+  geom_coords() +
+  tag_facets(tag = "cr")
+
 ```
@@
 
-    scale_x_longitude() +
+    scale_x_longitude(labels = NULL) +
     scale_y_latitude(labels = NULL,
+                     limits = c(NA, -20),
                      minor_breaks = NULL) +
@@
     coord_polar() +
-    axis_labs_smol +
+    geom_coords() +
     tag_facets("rc")
@@
 plot_regr(regr , "cEOF1", c(200, 50), "hgt") +
-  scale_fill_divergent("Geopotential height", breaks = AnchorBreaks(0, exclude = 0),
-                       guide = guide_colorstrip_bottom(15), 
+  scale_fill_divergent("Geopotential height", 
+                       breaks = AnchorBreaks(0, exclude = 0),
+                       guide = guide_colorsteps_bottom(15, even.steps = TRUE), 
                        limits = c(-1800, 1800))
@@
                        breaks = AnchorBreaks(0, 200, exclude = 0),
-                       guide = guide_colorstrip_bottom(15), 
+                       guide = guide_colorsteps_bottom(15, even.steps = TRUE), 
                        limits = c(-1000+1, 1000-1))
@@
 
-\textcolor{red}{Quedo medio rara la divsión con los títulos.
-Me parece que convendrá discutir primero la relación con otros modos de circulación conocidos (SAM, Ozono) y después con las otras variables(Temp, TOC, SST, precip)}
+### SAM
+
+(ref:sam-eof-vertical-cap) Coefficient of determination between the real and imaginary part of each cEOF and the SAM, Asymmetric SAM (A-SAM) and Symmetric SAM (S-SAM) indices computed at each level according to @campitelli2021.
+Thick lines represent estimates with p-value \< 0.01 corrected for False Detection Rate [@benjamini1995].
+
+```{r sam-eof-vertical, fig.cap = "(ref:sam-eof-vertical-cap)", fig.height=4, fig.width=width_column}
+time <- ceof[, eof[[1]]$left, by = .(season)] %>%
+  sep_ReIm() %>%
+  .[, hgt := scale(hgt), by = .(cEOF, part)]
+
+sam_r2 <- sams %>%
+  .[time, on = .NATURAL, allow.cartesian = TRUE] %>%
+  .[, correlate(hgt, estimate), by = .(lev, term, part, cEOF)]
+
+max_r2 <- sam_r2[part == "Imaginary" & term == "asym"] %>% 
+  .[, .SD[which.max(estimate^2)], by = .(cEOF)]
+
+sam_r2 %>%
+  .[, p.value_a := p.adjust(p.value, "fdr")] %>% 
+  ggplot(aes(lev, estimate^2)) +
+  geom_vline(xintercept = c(50, 200),  size = 0.2, color = "gray50") +
+  # geom_point(data = ~.x[p.adjust(p.value, "fdr") < 0.01], aes(color = term)) +
+  geom_line(data = ~copy(.x)[p.value_a > 0.01, estimate := NA], 
+            aes(colour = stage(term, after_scale = scales::alpha(colour, 0.5))), 
+            size = 2) +
+  geom_line(aes(color = term, group = term)) +
+
+  scale_x_level(minor_breaks = NULL) +
+  scale_y_continuous(r2, limits = c(0, 1)) +
+  scale_size_manual("p-value", values = c("TRUE" = 1, 
+                                          "FALSE" = 0.5), 
+                    guide = "none") +
+  scale_color_brewer(NULL, palette = "Dark2", labels = c(full = "SAM",
+                                                         asym = "A-SAM",
+                                                         sym = "S-SAM")) +
+  facet_grid(part ~ cEOF) +
+  coord_flip() +
+  tag_facets("rc") +
+  panel_background
+```
+
+To explore the relationship between SAM and the modes described from the cEOF, we compute the coefficient of determination between the cEOFs timeseries and the three SAM indices (SAM, A-SAM and S-SAM) at each vertical level (Fig. \@ref(fig:sam-eof-vertical)).
+The SAM index is moderately correlated with the Real cEOF1 in all levels, and with the Imaginary cEOF1 and Imaginary cEOF2 in the troposphere.
+It is not correlated with the Real cEOF2 at any level.
+
+The relationship between the tropospheric SAM and cEOF1 is explained entirely by the zonally symmetric component of the SAM as shown by the low and statistically non-significant correlations between the A-SAM and either the Real or Imaginary cEOF1 and the high correlation with the S-SAM below 100 hPa.
+In the stratosphere, the Real cEOF1 is correlated with both A-SAM and S-SAM, while the Imaginary cEOF1 is highly correlated only with the A-SAM.
+These correlations are consistent with the regression patterns of geopotential height in Figure \@ref(fig:eof1-regr-gh).
+
+```{r eof-filtered, eval=FALSE, fig.cap="Spatial pattern of the leading EOF of 200 hPa geopotential height with the variability of cEOF2 filtered out South of 20º.
+Arbitrary units.", fig.height=2.5, fig.width=2.5, include=FALSE}
+cEOF2 <- ceof$eof[[1]]$left[cEOF == "cEOF2"] %>% 
+  copy() %>% 
+  .[, cEOF := NULL] %>% 
+  .[]
+
+eof_filtered <- era5 %>% 
+  .[sep_ReIm(cEOF2, format = "wide")[, hgt := NULL], on = .NATURAL] %>% 
+  .[lev == 200] %>% 
+  .[lat <= -20] %>% 
+  .[season(time) == "SON"] %>% 
+  .[, hgt := resid(.lm.fit(cbind(1, Real, Imaginary), hgt)), .(lon, lat, lev, season(time))] %>%
+  .[, eof := Anomaly(hgt)*sqrt(cos(lat*pi/180)), by = .(lon, lat, lev, season(time))] %>% 
+  EOF(eof ~ lon + lat | time, n = 1, data = .)
+
+eof_filtered_cor <- eof_filtered$right[sams, on = "time"] %>% 
+  .[lev == 200] %>% 
+  .[, correlate(estimate, eof), by = term]
+
+eof_filtered %>%
+  .$left %>%
+  periodic(lon = c(0, 360)) %>%
+  ggplot(aes(lon, lat)) +
+  geom_contour_fill(aes(z = eof), breaks = AnchorBreaks(0, exclude = 0)) +
+  geom_contour_tanaka(aes(z = eof), breaks = AnchorBreaks(0, exclude = 0)) +
+  scale_fill_divergent(guide = "none") +
+  geom_qmap(~.x[lat <= -20]) +
+  scale_x_longitude() +
+  scale_y_latitude() +
+  coord_polar() +
+  geom_coords() +
+  theme(axis.text = element_blank())
+
+eof_filtered_cor %>% 
+  .[, .(term, text)] %>%
+  .[, term := factor(term, labels = c("SAM", "A-SAM", "S-SAM"))] %>%
+  knitr::kable(col.names = c("Índice SAM", "Correlación"))
+```
+
+In the case of cEOF2, the moderate correlation between SAM and Imaginary cEOF2 becomes extremely high only when the zonally asymmetric variability of the SAM is considered.
+The Imaginary cEOF2 and the A-SAM share up to `r scales::percent(max_r2[cEOF == "cEOF2"]$estimate^2, 2)` of their variance, reached at `r max_r2[cEOF == "cEOF2"]$lev` hPa (Figure \@ref(fig:sam-eof-vertical).b2).
+Such extremely high correlation between the asymmetric SAM and the Imaginary cEOF2 suggests that these might be different ways of characterising the same phenomenon.
+
+### PSA
+
+(ref:psa-eof2-cap) Correlations between the Real and Imaginary parts of cEOF2 and the PSA1 and PSA2 modes computed as the second and third EOFs of seasonal geopotential height anomalies [following @mo2001] for the 1979 -- 2019 period.
+95% confidence intervals in parenthesis.
+Background colour indicates the value of correlation.
+
+```{r psa-eof2, fig.cap = "(ref:psa-eof2-cap)", fig.width=7, fig.height=3, fig.env="figure*"}
+psa2 <- psa$left[PC == "PSA2"][season(time) == "SON"]
+cEOF2 <- ceof$eof[[1]]$left[cEOF == "cEOF2"] %>% 
+  copy() %>% 
+  .[, cEOF := NULL] %>% 
+  .[]
+
+rotations_psa <- lapply(angles, function(a) {
+  cEOF2 %>% 
+    copy() %>% 
+    .[, hgt := rotate(hgt, a)] %>% 
+    .[psa2, on = "time"] %>% 
+    .[, cor(value, Re(hgt))] 
+})
+
+psa_angle <- angles[which.min(unlist(rotations_psa))]
+
+make_numeric <- function(x) {
+  x <- strsplit(x, "\ ")
+  vapply(x, function(x) as.numeric(x[[1]]), numeric(1))
+}
+
+background_palette <- grDevices::colorRampPalette(c(scales::muted("blue"),
+                                                    "white",
+                                                    scales::muted("red")),
+                                                  space = "Lab")
+
+cEOF2[psa_time, on = "time"] %>% 
+  .[PC != "SAM"] %>% 
+  copy() %>% 
+  sep_ReIm() %>% 
+  .[, correlate(value, hgt), by = .(PC, part)] %>% 
+  dcast(PC ~ part, value.var = "text") %>% 
+  kbl2(booktabs = TRUE, caption = "(ref:psa-eof2-cap)") %>%
+  column_spec2(2:3, background = spec_color2(make_numeric(.table[[.col]]),
+                                             option = background_palette,
+                                             scale_from = c(-1, 1)), 
+               color = ifelse(abs(make_numeric(.table[[.col]])) > .5, "white", "black")) %>%
+  add_header_above(c("", "cEOF2" = 2)) %>%
+  kable_classic()
+
+# cEOF2[psa_time, on = "time"] %>% 
+#   .[PC != "SAM"] %>% 
+#   copy() %>% 
+#   # .[, hgt := rotate(hgt, psa_angle)] %>%
+#   sep_ReIm() %>% 
+#   ggplot(aes(value, hgt)) +
+#   geom_hline(yintercept = 0, size = 0.2, color = "gray50") +
+#   geom_vline(xintercept = 0, size = 0.2, color = "gray50") +
+#   geom_point() +
+#   geom_smooth(method = "lm") +
+#   geom_smooth(method = "lm") + 
+#   scale_y_continuous("EOF2") +
+#   scale_x_continuous("PSA") +
+#   facet_grid(part ~ PC) +
+#   panel_background +
+#   tag_facets("rc")
+```
+
+The relationship between PSAs and the modes described from the cEOFs is also studied.
+Table \@ref(tab:psa-eof2) shows the correlations between the two PSA indices and the timeseries for Real and Imaginary phase of cEOF2.
+As anticipated by Figure \@ref(fig:eof2-regr-gh), there is a strong correlation between PSA1 and Imaginary cEOF2, and between PSA2 and Real cEOF2.
+Conversely, there is no relationship between PSA1 and Real cEOF2, and between PSA2 and Imaginary cEOF2.
+As a result, cEOF2 represents well both the spatial structure and temporal evolution of the PSA modes, as well as it is possible to make a rather clean association between its two phases and the two PSA modes.
+So, It could be concluded that the same particular rotation of cEOF2 that maximises the association between cEOF2 parts and PSA modes, is the one which maximises the relationship between ENSO and Imaginary cEOF2.
+
+The reason the conventional EOF analysis arrives at the same separation than our particular cEOF rotation is probably the fact that not all phases are equally frequent.
+To visualize that, Figure \@ref(fig:phase-histogram) shows an histogram that counts the number of SON years in which the cEOF2 was close to each of the four particular phases (positive/negative Real/Imaginary), with the observations for each season marked as rugs on the horizontal axis.
+For instance, years with cEOF2 phase within 45º of 0º are nearest the "positive Real" phase.
+About two thirds of time cEOF2 has a phase similar to either the negative or positive Imaginary phase.
+It could be seen that Imaginary phase is the most common phase and It is also the direction that has the maximum relationship with ENSO.
+Therefore, the Imaginary cEOF2 explains more variance than the Real cEOF2 and conventional EOF analysis will tend to separate the two.
+
+(ref:phase-histogram-cap) Histogram of phase distribution of cEOF2 for the 1979 -- 2019 period.
+Bins are centred at 90º, 0º, -90º, -180º with a binwidth of 90º.
+The small vertical lines near the horizontal axis mark the observations.
+
+```{r phase-histogram, fig.cap = "(ref:phase-histogram-cap)", fig.width=width_column*.8, fig.height=2.5}
+N <- nrow(cut(ceof$eof[[1]], 2)$left)
+cut(ceof$eof[[1]], 2)$left %>% 
+  .[, arg := Arg(hgt)] %>% 
+  .[, w := Mod(hgt)/mean(Mod(hgt))] %>% 
+  qwrap(arg = c(-pi, pi) ~  c(-pi - 1/4*pi, pi - 1/4*pi)) %>% 
+  ggplot(aes(arg)) +
+  geom_histogram(binwidth = 90*pi/180, center = 0, size = 0.2,
+                 color = "black", fill = "gray70") +
+  annotate("text",
+           label = c("- Real", "- Imaginary", "+ Real", "+ Imaginary"),
+           color = "black", size = 3,
+           y = 0.5, x = c(-pi, -pi/2, 0, pi/2), angle = 90, hjust = 0) +
+  geom_rug() +
+  scale_y_continuous("Number of years", 
+                     sec.axis = sec_axis(~.x/N, 
+                                         name = "Proportion of years", 
+                                         labels = scales::percent)) +
+  scale_x_continuous("Phase", trans = scales::reverse_trans(),
+                     breaks = seq(-pi - 1/4*pi, pi - 1/4*pi, by = 45*pi/180),
+                     labels = function(x) paste0(round(x*180/pi, 5), "º")) +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
+  no_grid
+```
+
+The result obtained here is similar to the one obtained by @irving2016.
+They show that the phase distribution of the PSA-like variability they detected is bimodal and isolate the PSA pattern from the rest of the PSA-like variability by selecting events that are near the peaks of the distribution (compare our Figure \@ref(fig:phase-histogram) with their Figure 6).
+
+The advantage of our method is that it is much simpler to implement, it provides magnitude and phase naturally, and it facilitates the description of this mode as a propagating wave instead of as standing oscillation.
+As a consequence, the cEOF2 offers an alternative way of representing the PSA which has several advantages over using the second and third EOFs.
 
@@
   scale_fill_divergent("Temperature", breaks = AnchorBreaks(0, exclude = 0),
-                       guide = guide_colorstrip_bottom(15), 
+                       guide = guide_colorsteps_bottom(15, even.steps = TRUE), 
                        limits = c(-4.5, 4.5))
@@
   .[, term := factor_ReIm(term)] %>% 
-  # .[, z := metR:::standard_atmosphere$altitude(lev*100)] %>% 
-  # .[, dz1 := c(NA, diff(estimate)/diff(z)) , by = .(term, lon, variable)] %>% 
-  # .[, dz2 := c(diff(estimate)/diff(z), NA) , by = .(term, lon, variable)] %>% 
-  # .[, dz := (dz1 + dz2)/2] %>% 
-  # .[, p.value := Pvaluate(estimate, std.error, df, "fdr"), by = .(term, variable)] %>%
-  # .[lev < 900] %>% 
   ggplot(aes(lon, lev)) +
@@
 Figure \@ref(fig:t-vertical) shows the vertical distribution of the regression coefficients on cEOF1 from zonal anomalies of air temperature and zonal anomalies of ozone mixing ratio averaged between `r knitr::combine_words(LatLabel(o3wave_lats))`.
-\textcolor{red}{Deberias aclarar qué significa la terminología que estas utilizando. Lo podes hacer en los métodos y ya queda para todo el texto}.
 Temperature zonal anomalies associated with cEOF1 show a clear wave 1 pattern for both real and imaginary components throughout the atmosphere above 250 hPa with a change in sign above 10 hPa.
@@
                                    guide = guide_colorsteps_bottom(15)) + 
-  scale_x_longitude() +
+  scale_x_longitude(labels = NULL) +
   scale_y_latitude(labels = NULL,
+                   limits = c(NA, -20),
                    minor_breaks = NULL) +
-  grid_y +
   scale_linetype(guide = "none") +
   coord_polar() +
-
+  geom_coords() +
   facet_grid(cEOF~term) +
-  axis_labs_smol +
   tag_facets("panel")
@@
   geom_point()  &
-  geom_smooth(method = "lm")
+  geom_smooth(method = "lm") & 
+  panel_background
 
@@
 As expected from the location of regressions in Figure \@ref(fig:o3-regr), the Real cEOF1 drives the relationship of amplitudes and the Imaginary cEOF1 drives the relationship of phases between the timeseries (not shown).
-
-### SAM
-
-(ref:sam-eof-vertical-cap) Coefficient of determination between the real and imaginary part of each cEOF and the SAM, Asymmetric SAM (A-SAM) and Symmetric SAM (S-SAM) indices computed at each level according to @campitelli2021.
-Thick lines represent estimates with p-value \< 0.01 corrected for False Detection Rate [@benjamini1995].
-
-```{r sam-eof-vertical, fig.cap = "(ref:sam-eof-vertical-cap)", fig.height=4, fig.width=width_column}
-time <- ceof[, eof[[1]]$left, by = .(season)] %>%
-  sep_ReIm() %>%
-  .[, hgt := scale(hgt), by = .(cEOF, part)]
-
-sam_r2 <- sams %>%
-  .[time, on = .NATURAL, allow.cartesian = TRUE] %>%
-  .[, correlate(hgt, estimate), by = .(lev, term, part, cEOF)]
-
-max_r2 <- sam_r2[part == "Imaginary" & term == "asym"] %>% 
-  .[, .SD[which.max(estimate^2)], by = .(cEOF)]
-
-sam_r2 %>%
-  .[, p.value_a := p.adjust(p.value, "fdr")] %>% 
-  ggplot(aes(lev, estimate^2)) +
-  geom_vline(xintercept = c(50, 200),  size = 0.2, color = "gray50") +
-  # geom_point(data = ~.x[p.adjust(p.value, "fdr") < 0.01], aes(color = term)) +
-  geom_line(data = ~copy(.x)[p.value_a > 0.01, estimate := NA], 
-            aes(colour = stage(term, after_scale = scales::alpha(colour, 0.5))), 
-            size = 2) +
-  geom_line(aes(color = term, group = term)) +
-
-  scale_x_level(minor_breaks = NULL) +
-  scale_y_continuous(r2, limits = c(0, 1)) +
-  scale_size_manual("p-value", values = c("TRUE" = 1, 
-                                          "FALSE" = 0.5), 
-                    guide = "none") +
-  scale_color_brewer(NULL, palette = "Dark2", labels = c(full = "SAM",
-                                                         asym = "A-SAM",
-                                                         sym = "S-SAM")) +
-  facet_grid(part ~ cEOF) +
-  coord_flip() +
-  tag_facets("rc") +
-  panel_background
-```
-
-To explore the relationship between SAM and the modes described from the cEOF, we compute the coefficient of determination between the cEOFs timeseries and the three SAM indices (SAM, A-SAM and S-SAM) at each vertical level (Fig. \@ref(fig:sam-eof-vertical)).
-The SAM index is moderately correlated with the Real cEOF1 in all levels, and with the Imaginary cEOF1 and Imaginary cEOF2 in the troposphere.
-It is not correlated with the Real cEOF2 at any level.
-
-The relationship between the tropospheric SAM and cEOF1 is explained entirely by the zonally symmetric component of the SAM as shown by the low and statistically non-significant correlations between the A-SAM and either the Real or Imaginary cEOF1 and the high correlation with the S-SAM below 100 hPa.
-In the stratosphere, the Real cEOF1 is correlated with both A-SAM and S-SAM, while the Imaginary cEOF1 is highly correlated only with the A-SAM.
-These correlations are consistent with the regression patterns of geopotential height in Figure \@ref(fig:eof1-regr-gh).
-
-```{r eof-filtered, eval=FALSE, fig.cap="Spatial pattern of the leading EOF of 200 hPa geopotential height with the variability of cEOF2 filtered out South of 20º.
-Arbitrary units.", fig.height=2.5, fig.width=2.5, include=FALSE}
-cEOF2 <- ceof$eof[[1]]$left[cEOF == "cEOF2"] %>% 
-  copy() %>% 
-  .[, cEOF := NULL] %>% 
-  .[]
-
-eof_filtered <- era5 %>% 
-  .[sep_ReIm(cEOF2, format = "wide")[, hgt := NULL], on = .NATURAL] %>% 
-  .[lev == 200] %>% 
-  .[lat <= -20] %>% 
-  .[season(time) == "SON"] %>% 
-  .[, hgt := resid(.lm.fit(cbind(1, Real, Imaginary), hgt)), .(lon, lat, lev, season(time))] %>%
-  .[, eof := Anomaly(hgt)*sqrt(cos(lat*pi/180)), by = .(lon, lat, lev, season(time))] %>% 
-  EOF(eof ~ lon + lat | time, n = 1, data = .)
-
-eof_filtered_cor <- eof_filtered$right[sams, on = "time"] %>% 
-  .[lev == 200] %>% 
-  .[, correlate(estimate, eof), by = term]
-
-eof_filtered %>%
-  .$left %>%
-  periodic(lon = c(0, 360)) %>%
-  ggplot(aes(lon, lat)) +
-  geom_contour_fill(aes(z = eof), breaks = AnchorBreaks(0, exclude = 0)) +
-  geom_contour_tanaka(aes(z = eof), breaks = AnchorBreaks(0, exclude = 0)) +
-  scale_fill_divergent(guide = "none") +
-  geom_qmap(~.x[lat <= -20]) +
-  scale_x_longitude() +
-  scale_y_latitude() +
-  coord_polar() +
-  theme(axis.text = element_blank())
-
-eof_filtered_cor %>% 
-  .[, .(term, text)] %>%
-  .[, term := factor(term, labels = c("SAM", "A-SAM", "S-SAM"))] %>%
-  knitr::kable(col.names = c("Índice SAM", "Correlación"))
-```
-
-In the case of cEOF2, the moderate correlation between SAM and Imaginary cEOF2 becomes extremely high only when the zonally asymmetric variability of the SAM is considered.
-The Imaginary cEOF2 and the A-SAM share up to `r scales::percent(max_r2[cEOF == "cEOF2"]$estimate^2, 2)` of their variance, reached at `r max_r2[cEOF == "cEOF2"]$lev` hPa (Figure \@ref(fig:sam-eof-vertical).b2).
-Such extremely high correlation between the asymmetric SAM and the Imaginary cEOF2 suggests that these might be different ways of characterising the same phenomenon.
-
-### PSA
-
-(ref:psa-eof2-cap) Correlations between the Real and Imaginary parts of cEOF2 and the PSA1 and PSA2 modes computed as the second and third EOFs of seasonal geopotential height anomalies [following @mo2001] for the 1979 -- 2019 period.
-95% confidence intervals in parenthesis.
-Background colour indicates the value of correlation.
-
-```{r psa-eof2, fig.cap = "(ref:psa-eof2-cap)", fig.width=7, fig.height=3, fig.env="figure*"}
-psa2 <- psa$left[PC == "PSA2"][season(time) == "SON"]
-cEOF2 <- ceof$eof[[1]]$left[cEOF == "cEOF2"] %>% 
-  copy() %>% 
-  .[, cEOF := NULL] %>% 
-  .[]
-
-rotations_psa <- lapply(angles, function(a) {
-  cEOF2 %>% 
-    copy() %>% 
-    .[, hgt := rotate(hgt, a)] %>% 
-    .[psa2, on = "time"] %>% 
-    .[, cor(value, Re(hgt))] 
-})
-
-psa_angle <- angles[which.min(unlist(rotations_psa))]
-
-make_numeric <- function(x) {
-  x <- strsplit(x, "\ ")
-  vapply(x, function(x) as.numeric(x[[1]]), numeric(1))
-}
-
-background_palette <- grDevices::colorRampPalette(c(scales::muted("blue"),
-                                                    "white",
-                                                    scales::muted("red")),
-                                                  space = "Lab")
-
-cEOF2[psa_time, on = "time"] %>% 
-  .[PC != "SAM"] %>% 
-  copy() %>% 
-  sep_ReIm() %>% 
-  .[, correlate(value, hgt), by = .(PC, part)] %>% 
-  dcast(PC ~ part, value.var = "text") %>% 
-  kbl2(booktabs = TRUE, caption = "(ref:psa-eof2-cap)") %>%
-  column_spec2(2:3, background = spec_color2(make_numeric(.table[[.col]]),
-                                             option = background_palette,
-                                             scale_from = c(-1, 1)), 
-               color = ifelse(abs(make_numeric(.table[[.col]])) > .5, "white", "black")) %>%
-  add_header_above(c("", "cEOF2" = 2)) %>%
-  kable_classic()
-
-# cEOF2[psa_time, on = "time"] %>% 
-#   .[PC != "SAM"] %>% 
-#   copy() %>% 
-#   # .[, hgt := rotate(hgt, psa_angle)] %>%
-#   sep_ReIm() %>% 
-#   ggplot(aes(value, hgt)) +
-#   geom_hline(yintercept = 0, size = 0.2, color = "gray50") +
-#   geom_vline(xintercept = 0, size = 0.2, color = "gray50") +
-#   geom_point() +
-#   geom_smooth(method = "lm") +
-#   geom_smooth(method = "lm") + 
-#   scale_y_continuous("EOF2") +
-#   scale_x_continuous("PSA") +
-#   facet_grid(part ~ PC) +
-#   panel_background +
-#   tag_facets("rc")
-```
-
-The relationship between PSAs and the modes described from the cEOFs is also studied.
-Table \@ref(tab:psa-eof2) shows the correlations between the two PSA indices and the timeseries for Real and Imaginary phase of cEOF2.
-As anticipated by Figure \@ref(fig:eof2-regr-gh), there is a strong correlation between PSA1 and Imaginary cEOF2, and between PSA2 and Real cEOF2.
-Conversely, there is no relationship between PSA1 and Real cEOF2, and between PSA2 and Imaginary cEOF2.
-As a result, cEOF2 represents well both the spatial structure and temporal evolution of the PSA modes, as well as it is possible to make a rather clean association between its two phases and the two PSA modes.
-So, It could be concluded that the same particular rotation of cEOF2 that maximises the association between cEOF2 parts and PSA modes, is the one which maximises the relationship between ENSO and Imaginary cEOF2.
-
-The reason the conventional EOF analysis arrives at the same separation than our particular cEOF rotation is probably the fact that not all phases are equally frequent.
-To visualize that, Figure \@ref(fig:phase-histogram) shows an histogram that counts the number of SON years in which the cEOF2 was close to each of the four particular phases (positive/negative Real/Imaginary), with the observations for each season marked as rugs on the horizontal axis.
-For instance, years with cEOF2 phase within 45º of 0º are nearest the "positive Real" phase.
-About two thirds of time cEOF2 has a phase similar to either the negative or positive Imaginary phase.
-It could be seen that Imaginary phase is the most common phase and It is also the direction that has the maximum relationship with ENSO.
-Therefore, the Imaginary cEOF2 explains more variance than the Real cEOF2 and conventional EOF analysis will tend to separate the two.
-
-(ref:phase-histogram-cap) Histogram of phase distribution of cEOF2 for the 1979 -- 2019 period.
-Bins are centred at 90º, 0º, -90º, -180º with a binwidth of 90º.
-The small vertical lines near the horizontal axis mark the observations.
-
-```{r phase-histogram, fig.cap = "(ref:phase-histogram-cap)", fig.width=width_column*.8, fig.height=2.5}
-N <- nrow(cut(ceof$eof[[1]], 2)$left)
-cut(ceof$eof[[1]], 2)$left %>% 
-  .[, arg := Arg(hgt)] %>% 
-  .[, w := Mod(hgt)/mean(Mod(hgt))] %>% 
-  qwrap(arg = c(-pi, pi) ~  c(-pi - 1/4*pi, pi - 1/4*pi)) %>% 
-  ggplot(aes(arg)) +
-  geom_histogram(binwidth = 90*pi/180, center = 0, size = 0.2,
-                 color = "black", fill = "gray70") +
-  annotate("text",
-           label = c("- Real", "- Imaginary", "+ Real", "+ Imaginary"),
-           color = "black", size = 3,
-           y = 0.5, x = c(-pi, -pi/2, 0, pi/2), angle = 90, hjust = 0) +
-  geom_rug() +
-  scale_y_continuous("Number of years", 
-                     sec.axis = sec_axis(~.x/N, 
-                                         name = "Proportion of years", 
-                                         labels = scales::percent)) +
-  scale_x_continuous("Phase", trans = scales::reverse_trans(),
-                     breaks = seq(-pi - 1/4*pi, pi - 1/4*pi, by = 45*pi/180),
-                     labels = function(x) paste0(round(x*180/pi, 5), "º")) +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))
-```
-
-The result obtained here is similar to the one obtained by @irving2016.
-From their Fourier decomposition of reprojected meridional wind fields \textcolor{red}{¿En qué región?}, they show that the phase distribution is bimodal and isolate the PSA pattern from the rest of the PSA-like variability by selecting events that are near the peaks of the distribution (compare our Figure \@ref(fig:phase-histogram) with their Figure 6).
-
-The advantage of our method is that it is much simpler to implement, it provides magnitude and phase naturally, and it facilitates the description of this mode as a propagating wave instead of as standing oscillation.
-As a consequence, the cEOF2 offers an alternative way of representing the PSA which has several advantages over using the second and third EOFs.
 
@@
     scale_fill_divergent_discretised("Streamfunction (×1e-7)", 
-                                     guide = guide_colorstrip_bottom(20, order = 1)) +
+                                     guide = guide_colorsteps_bottom(20, order = 1)) +
     ggnewscale::new_scale_fill() +
@@
     theme(legend.box = "vertical") +
-    axis_labs_smol +
     tag_facets("cr", position = "bl") 
@@
                                    limits = c(-1, 1),
-                                   guide = guide_colorstrip_bottom(20, order = 2)) 
-```
-
-The connections between cEOFs and Tropical sources is also assesed.
+                                   guide = guide_colorsteps_bottom(20, order = 2)) 
+```
+
+The connections between cEOFs and Tropical sources is also assessed.
 Figure \@ref(fig:sst-psi-2) shows the regression maps of Sea Surface Temperatures (SST) and Streamfunction at 200 hPa respectively upon cEOF2.
@@
 This is consistent with the effect of ENSO on the extratropics: SST anomalies initiate anomalous convection that excites Rossby waves that propagate meridionally towards higher latitudes [@mo2000].
-
-\textcolor{red}{Revisa los labels de la colorbar para la figura 13.
-Además ojo que solamente sirven para la columna a, faltan los correspondientes a la streamfunction y el vector de referencia para wave flux}
 
@@
     geom_contour_pval(aes(z = p.value)) +
-    geom_qmap(~.x %>% .[lat %between% c(-60, max(data$lat))] %>% add_continent() %>% filter_continent(),
+    geom_qmap(~.x %>% .[lat %between% c(-60, max(data$lat))] %>% 
+                add_continent() %>% filter_continent(),
               keep = 1) +
@@
 The influence of cEOFs variability in continental rainfall of extra-tropical Southern Hemisphere is also explored.
-We focus on South America and Oceania sectors, because there are not relevant signals in Southern Africa (not shown).
+We focus on South America and Oceania sectors because there are no relevant signals in southern Africa (not shown).
 
@@
 
-The precipitation anomalies in Oceania associated with cEOF1 are also weak and not statistically significant (Figure \@ref(fig:pp-oceania) column a) whereas EFO2 is associated with large and widespread anomalies.
+The precipitation anomalies in Oceania associated with cEOF1 are weak and not statistically significant (Figure \@ref(fig:pp-oceania) column a) whereas EFO2 is associated with large and widespread anomalies.
 The Real cEOF2 is associated with positive seasonal anomalies in Northern and Eastern Australia, and negative ones over New Zealand, while precipitation anomalies associated with Imaginary cEOF2 are negative over all Eastern Australia.
@@
 Figure \@ref(fig:pp-america) shows regression maps of seasonal precipitation with each cEOF in South America.
-cEOF1 is associated with slight decreases \textcolor{red}{Revisa como estás expresando toda esta parte (y en el resto del texto), recorda que son regresiones así que no corresponde decir postive/negative sino haces alguna referencia a la variable con las que los estás regresionando.} in precipitation in Southern Brazil and Paraguay, although these are not statistically significant (Fig. \@ref(fig:pp-america) column a).
+cEOF1 is associated with slight decreases in precipitation in Southern Brazil and Paraguay, although these are not statistically significant (Fig. \@ref(fig:pp-america) column a).
 The strongest precipitation regression coefccients are the ones associated with the Imaginary cEOF2.
-The positive anomalies on Southeastern South America (SESA) and Central Chile, and negative anomalies over South Atlantic Convergence Zone is a well known springtime precipitation signature of ENSO [@cai2020a] and it is also similar to the precipitation anomalies associated with the A-SAM [@campitelli2021].
+The positive regression coefficients on Southeastern South America (SESA) and Central Chile, and negative ones over South Atlantic Convergence Zone is a well-known springtime precipitation signature of ENSO [@cai2020a] and it is also similar to the precipitation anomalies associated with the A-SAM [@campitelli2021].
 This is not surprising considering the close relationship between the ONI, the A-SAM index and the Imaginary cEOF2 shown previously, but further consolidates the identification of this mode with the PSA pattern.
@@
 Resembling the relationship between ONI and the phase of cEOF2 (Fig. \@ref(fig:enso-phase)), there is a cEOF2 phase dependence of the precipitation anomalies in SESA (not shown).
-These variable could be related wth a signficant sinusoidal fit that has a coefficient of determination of `r round(summary(model)$r.squared, 2)` (p-value `r report_p(p_value)`).
+These variables could be related wth a significant sinusoidal fit that has a coefficient of determination of `r round(summary(model)$r.squared, 2)` (p-value `r report_p(p_value)`).
 
@@
 
-\textcolor{red}{Al igual que en la introducción, te falta establecer más claro el objetivo, después iría lo que escribiste.}
-
-We studied the two leading modes of co-variability of September-October-November tropospheric and stratospheric zonal anomalies by applying complex Empirical Orthogonal Functions.
+In this study we derived two complex indices of the zonally asymmetric circulation of the September-October-November southern extratropics using Complex Empirical Orthogonal Functions. 
+This allowed us to characterise both the amplitude and phase of planetary waves that are not perfectly organised as sinusoidal waves with constant amplitude at each latitude circle.
 
@@
 We further show that the Imaginary cEOF2 is closely related to the Southern Annular Mode in the troposphere.
-In fact, it has a closer resemblance of the zonally asymmetric portion of the SAM.
-
+In fact, it has a very close resemblance to the zonally asymmetric portion of the SAM.
 This raises the possibility that the asymmetric component of the SAM is actually a statistical contamination of the PSA mode.
@@
 On the other hand, over Australia, both the Real and Imaginary phase are associated with significant precipitation anomalies, and we further show that the direction of maximum impact is not aligned with our chosen rotation of cEOF2.
-However, this underscore the benefit of using complex EOF, since it would trivial to rotate it.
-
-\textcolor{red}{Debería cerrarse con algún parráfo que diga qué tiene de bueno trabajar con esta metodología y qué permitiría hacer a futuro (se puede poner que sería deseable extenderlo a otras estaciones, etc.)}
+However, this underscores the benefit of using complex EOF, since it would trivial to rotate it.
+
+To the extent that the cEOF2 can be identified with the PSA, what is the advantage of our technique?
+The spatial fields that the cEOF2 index describes are in quadrature by construction, which makes it possible to derive a proper amplitude and phase.
+Furthermore, because of the way they are constructed, they are not forced to be orthogonal to the SAM.
+This makes it possible to study the relationship between the two modes free of the statistical contamination that comes with the usual definition of the PSA as the second and third EOFs.
+
+Further work should extend this analysis to other seasons and further study the relationship between the cEOF2 and the SAM. 
+In particular, it is not clear whether the close identification between the Imaginary cEOF2 and the Asymmetric SAM arises from physical processes or statistical contamination.
 

### 2021-07-27 analysis/paper/shceof.tex
Muchos cambios

@@
 
-\textcolor{red}{The data and methods are described in Section 2 :P ...}
+In Section \ref{methods} we describe the methods.
+In Section \ref{temporal} we describe the temporal variability and vertical coherence of the indices.
+In Section \ref{spatial}, we analyse the spatial patterns of each complex EOF.
+In Section \ref{regressions} we study the spatial regressions with geopotential height and temperatuer.
+In Section \ref{other-variables}, we analyise the relationship between cEOF1 and Ozone, and between EOF2 and the Southern Annular Mode and the PSA.
+In Section \ref{tropical} we look for tropical forcings that explain the variability of each cEOF.
+Finally, in Section \ref{precipitation} we show the relationship between these modes of variabiltiy and precipitation in South America and Oceania.
 
-\hypertarget{data-and-methods}{%
-\section{Data and Methods}\label{data-and-methods}}
+\hypertarget{methods}{%
+\section{Data and Methods}\label{methods}}
 
@@
 
-\hypertarget{methods}{%
-\subsection{Methods}\label{methods}}
+\hypertarget{methods-1}{%
+\subsection{Methods}\label{methods-1}}
 
@@
 
-\hypertarget{ceof-spatial-patterns}{%
-\subsection{cEOF spatial patterns}\label{ceof-spatial-patterns}}
+\hypertarget{spatial}{%
+\subsection{cEOF spatial patterns}\label{spatial}}
 
@@
 
-\hypertarget{ceofs-regressions}{%
-\subsection{cEOFs Regressions}\label{ceofs-regressions}}
+\hypertarget{regressions}{%
+\subsection{cEOFs Regressions}\label{regressions}}
 
@@
 
-\hypertarget{ceofs-relationship-with-other-variables}{%
-\subsection{cEOFs relationship with other variables}\label{ceofs-relationship-with-other-variables}}
+\hypertarget{other-variables}{%
+\subsection{cEOFs relationship with other variables}\label{other-variables}}
 
-\textcolor{red}{Quedo medio rara la divsión con los títulos.
-Me parece que convendrá discutir primero la relación con otros modos de circulación conocidos (SAM, Ozono) y después con las otras variables(Temp, TOC, SST, precip)}
+\hypertarget{sam}{%
+\subsubsection{SAM}\label{sam}}
+
+\begin{figure}
+\centering
+\includegraphics{../figures/sam-eof-vertical-1.pdf}
+\caption{\label{fig:sam-eof-vertical}Coefficient of determination between the real and imaginary part of each cEOF and the SAM, Asymmetric SAM (A-SAM) and Symmetric SAM (S-SAM) indices computed at each level according to \citet{campitelli2021}.
+Thick lines represent estimates with p-value \textless{} 0.01 corrected for False Detection Rate \citep{benjamini1995}.}
+\end{figure}
+
+To explore the relationship between SAM and the modes described from the cEOF, we compute the coefficient of determination between the cEOFs timeseries and the three SAM indices (SAM, A-SAM and S-SAM) at each vertical level (Fig.~\ref{fig:sam-eof-vertical}).
+The SAM index is moderately correlated with the Real cEOF1 in all levels, and with the Imaginary cEOF1 and Imaginary cEOF2 in the troposphere.
+It is not correlated with the Real cEOF2 at any level.
+
+The relationship between the tropospheric SAM and cEOF1 is explained entirely by the zonally symmetric component of the SAM as shown by the low and statistically non-significant correlations between the A-SAM and either the Real or Imaginary cEOF1 and the high correlation with the S-SAM below 100 hPa.
+In the stratosphere, the Real cEOF1 is correlated with both A-SAM and S-SAM, while the Imaginary cEOF1 is highly correlated only with the A-SAM.
+These correlations are consistent with the regression patterns of geopotential height in Figure~\ref{fig:eof1-regr-gh}.
+
+In the case of cEOF2, the moderate correlation between SAM and Imaginary cEOF2 becomes extremely high only when the zonally asymmetric variability of the SAM is considered.
+The Imaginary cEOF2 and the A-SAM share up to 92\% of their variance, reached at 225 hPa (Figure~\ref{fig:sam-eof-vertical}.b2).
+Such extremely high correlation between the asymmetric SAM and the Imaginary cEOF2 suggests that these might be different ways of characterising the same phenomenon.
+
+\hypertarget{psa}{%
+\subsubsection{PSA}\label{psa}}
+
+\begin{table}
+
+\caption{\label{tab:psa-eof2}Correlations between the Real and Imaginary parts of cEOF2 and the PSA1 and PSA2 modes computed as the second and third EOFs of seasonal geopotential height anomalies \citep[following][]{mo2001} for the 1979 -- 2019 period.
+95\% confidence intervals in parenthesis.
+Background colour indicates the value of correlation.}
+\centering
+\begin{tabular}[t]{l>{}l>{}l}
+\toprule
+\multicolumn{1}{c}{} & \multicolumn{2}{c}{cEOF2} \\
+\cmidrule(l{3pt}r{3pt}){2-3}
+PC & Real & Imaginary\\
+\midrule
+PSA1 & \cellcolor[HTML]{C3BBDE}{\textcolor{black}{-0.32 (CI: -0.57 -- -0.015)}} & \cellcolor[HTML]{6259AA}{\textcolor{white}{-0.82 (CI: -0.9 -- -0.69)}}\\
+PSA2 & \cellcolor[HTML]{9E4E47}{\textcolor{white}{0.81 (CI: 0.67 -- 0.9)}} & \cellcolor[HTML]{F3F2F8}{\textcolor{black}{-0.062 (CI: -0.36 -- 0.25)}}\\
+\bottomrule
+\end{tabular}
+\end{table}
+
+The relationship between PSAs and the modes described from the cEOFs is also studied.
+Table~\ref{tab:psa-eof2} shows the correlations between the two PSA indices and the timeseries for Real and Imaginary phase of cEOF2.
+As anticipated by Figure~\ref{fig:eof2-regr-gh}, there is a strong correlation between PSA1 and Imaginary cEOF2, and between PSA2 and Real cEOF2.
+Conversely, there is no relationship between PSA1 and Real cEOF2, and between PSA2 and Imaginary cEOF2.
+As a result, cEOF2 represents well both the spatial structure and temporal evolution of the PSA modes, as well as it is possible to make a rather clean association between its two phases and the two PSA modes.
+So, It could be concluded that the same particular rotation of cEOF2 that maximises the association between cEOF2 parts and PSA modes, is the one which maximises the relationship between ENSO and Imaginary cEOF2.
+
+The reason the conventional EOF analysis arrives at the same separation than our particular cEOF rotation is probably the fact that not all phases are equally frequent.
+To visualize that, Figure~\ref{fig:phase-histogram} shows an histogram that counts the number of SON years in which the cEOF2 was close to each of the four particular phases (positive/negative Real/Imaginary), with the observations for each season marked as rugs on the horizontal axis.
+For instance, years with cEOF2 phase within 45º of 0º are nearest the ``positive Real'' phase.
+About two thirds of time cEOF2 has a phase similar to either the negative or positive Imaginary phase.
+It could be seen that Imaginary phase is the most common phase and It is also the direction that has the maximum relationship with ENSO.
+Therefore, the Imaginary cEOF2 explains more variance than the Real cEOF2 and conventional EOF analysis will tend to separate the two.
+
+\begin{figure}
+\centering
+\includegraphics{../figures/phase-histogram-1.pdf}
+\caption{\label{fig:phase-histogram}Histogram of phase distribution of cEOF2 for the 1979 -- 2019 period.
+Bins are centred at 90º, 0º, -90º, -180º with a binwidth of 90º.
+The small vertical lines near the horizontal axis mark the observations.}
+\end{figure}
+
+The result obtained here is similar to the one obtained by \citet{irving2016}.
+From their Fourier decomposition of reprojected meridional wind fields \textcolor{red}{¿En qué región?}, they show that the phase distribution is bimodal and isolate the PSA pattern from the rest of the PSA-like variability by selecting events that are near the peaks of the distribution (compare our Figure~\ref{fig:phase-histogram} with their Figure 6).
+
+The advantage of our method is that it is much simpler to implement, it provides magnitude and phase naturally, and it facilitates the description of this mode as a propagating wave instead of as standing oscillation.
+As a consequence, the cEOF2 offers an alternative way of representing the PSA which has several advantages over using the second and third EOFs.
 
@@
 
-\hypertarget{sam}{%
-\subsubsection{SAM}\label{sam}}
-
-\begin{figure}
-\centering
-\includegraphics{../figures/sam-eof-vertical-1.pdf}
-\caption{\label{fig:sam-eof-vertical}Coefficient of determination between the real and imaginary part of each cEOF and the SAM, Asymmetric SAM (A-SAM) and Symmetric SAM (S-SAM) indices computed at each level according to \citet{campitelli2021}.
-Thick lines represent estimates with p-value \textless{} 0.01 corrected for False Detection Rate \citep{benjamini1995}.}
-\end{figure}
-
-To explore the relationship between SAM and the modes described from the cEOF, we compute the coefficient of determination between the cEOFs timeseries and the three SAM indices (SAM, A-SAM and S-SAM) at each vertical level (Fig.~\ref{fig:sam-eof-vertical}).
-The SAM index is moderately correlated with the Real cEOF1 in all levels, and with the Imaginary cEOF1 and Imaginary cEOF2 in the troposphere.
-It is not correlated with the Real cEOF2 at any level.
-
-The relationship between the tropospheric SAM and cEOF1 is explained entirely by the zonally symmetric component of the SAM as shown by the low and statistically non-significant correlations between the A-SAM and either the Real or Imaginary cEOF1 and the high correlation with the S-SAM below 100 hPa.
-In the stratosphere, the Real cEOF1 is correlated with both A-SAM and S-SAM, while the Imaginary cEOF1 is highly correlated only with the A-SAM.
-These correlations are consistent with the regression patterns of geopotential height in Figure~\ref{fig:eof1-regr-gh}.
-
-In the case of cEOF2, the moderate correlation between SAM and Imaginary cEOF2 becomes extremely high only when the zonally asymmetric variability of the SAM is considered.
-The Imaginary cEOF2 and the A-SAM share up to 92\% of their variance, reached at 225 hPa (Figure~\ref{fig:sam-eof-vertical}.b2).
-Such extremely high correlation between the asymmetric SAM and the Imaginary cEOF2 suggests that these might be different ways of characterising the same phenomenon.
-
-\hypertarget{psa}{%
-\subsubsection{PSA}\label{psa}}
-
-\begin{table}
-
-\caption{\label{tab:psa-eof2}Correlations between the Real and Imaginary parts of cEOF2 and the PSA1 and PSA2 modes computed as the second and third EOFs of seasonal geopotential height anomalies \citep[following][]{mo2001} for the 1979 -- 2019 period.
-95\% confidence intervals in parenthesis.
-Background colour indicates the value of correlation.}
-\centering
-\begin{tabular}[t]{l>{}l>{}l}
-\toprule
-\multicolumn{1}{c}{} & \multicolumn{2}{c}{cEOF2} \\
-\cmidrule(l{3pt}r{3pt}){2-3}
-PC & Real & Imaginary\\
-\midrule
-PSA1 & \cellcolor[HTML]{C3BBDE}{\textcolor{black}{-0.32 (CI: -0.57 -- -0.01)}} & \cellcolor[HTML]{6058A9}{\textcolor{white}{-0.83 (CI: -0.9 -- -0.69)}}\\
-PSA2 & \cellcolor[HTML]{645BAB}{\textcolor{white}{-0.81 (CI: -0.9 -- -0.67)}} & \cellcolor[HTML]{F9F1F0}{\textcolor{black}{0.059 (CI: -0.25 -- 0.36)}}\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-The relationship between PSAs and the modes described from the cEOFs is also studied.
-Table~\ref{tab:psa-eof2} shows the correlations between the two PSA indices and the timeseries for Real and Imaginary phase of cEOF2.
-As anticipated by Figure~\ref{fig:eof2-regr-gh}, there is a strong correlation between PSA1 and Imaginary cEOF2, and between PSA2 and Real cEOF2.
-Conversely, there is no relationship between PSA1 and Real cEOF2, and between PSA2 and Imaginary cEOF2.
-As a result, cEOF2 represents well both the spatial structure and temporal evolution of the PSA modes, as well as it is possible to make a rather clean association between its two phases and the two PSA modes.
-So, It could be concluded that the same particular rotation of cEOF2 that maximises the association between cEOF2 parts and PSA modes, is the one which maximises the relationship between ENSO and Imaginary cEOF2.
-
-The reason the conventional EOF analysis arrives at the same separation than our particular cEOF rotation is probably the fact that not all phases are equally frequent.
-To visualize that, Figure~\ref{fig:phase-histogram} shows an histogram that counts the number of SON years in which the cEOF2 was close to each of the four particular phases (positive/negative Real/Imaginary), with the observations for each season marked as rugs on the horizontal axis.
-For instance, years with cEOF2 phase within 45º of 0º are nearest the ``positive Real'' phase.
-About two thirds of time cEOF2 has a phase similar to either the negative or positive Imaginary phase.
-It could be seen that Imaginary phase is the most common phase and It is also the direction that has the maximum relationship with ENSO.
-Therefore, the Imaginary cEOF2 explains more variance than the Real cEOF2 and conventional EOF analysis will tend to separate the two.
-
-\begin{figure}
-\centering
-\includegraphics{../figures/phase-histogram-1.pdf}
-\caption{\label{fig:phase-histogram}Histogram of phase distribution of cEOF2 for the 1979 -- 2019 period.
-Bins are centred at 90º, 0º, -90º, -180º with a binwidth of 90º.
-The small vertical lines near the horizontal axis mark the observations.}
-\end{figure}
-
-The result obtaneid here is similar to the one obtained by \citet{irving2016}.
-From their Fourier decomposition of reprojected meridional wind fields \textcolor{red}{¿En qué región?}, they show that the phase distribution is bimodal and isolate the PSA pattern from the rest of the PSA-like variability by selecting events that are near the peaks of the distribution (compare our Figure~\ref{fig:phase-histogram} with their Figure 6).
-
-The advantage of our method is that it is much simpler to implement, it provides magnitude and phase naturally, and it facilitates the description of this mode as a propagating wave instead of as standing oscillation.
-As a consequence, the cEOF2 offers an alternative way of representing the PSA which has several advantages over using the second and third EOFs.
-
-\hypertarget{is-the-asymmetric-sam-real}{%
-\subsubsection{Is the Asymmetric SAM real?}\label{is-the-asymmetric-sam-real}}
-
-\begin{figure}
-\centering
-\includegraphics{../figures/fake-eof-1.pdf}
-\caption{\label{fig:fake-eof}Row a: The three components used to create a synthetic dataset of 41 years of geopotential height at 200 hPa.
-Row b: The three EOF computed from the synthetic dataset with the variance explained by each EOF in parenthesis.
-Units are arbitrary.}
-\end{figure}
-
-To test the hypothesis that a PSA-like feature can appear as statistic contamination of an otherwise zonally symmetrical and uncorrelated annular pattern, we create a synthetic dataset of 41 fields by reconstructing the SON 200 hPa geopotential height field explained by the cEOF2 and adding a perfectly zonally symmetric annylar pattern with random sign and value sampled with replacement from the real SAM index.
-To this synthetic dataset we then apply traditional EOF.
-By construction this dataset contains two independent components: a zonally symmetric standing oscillation and a PSA-like of varying amplitude and phase (Figure~\ref{fig:fake-eof} row a).
-While the former is a standing oscillation which could be represented by a single EOF, the latter is a wave which can change phase, so traditional EOF analysis needs two EOFs to represented it correctly.
-Thus, the full variance of this synthetic dataset can be perfectly explained by three EOFs.
-
-The three EOFs of the synthetic dataset are shown in Figure~\ref{fig:fake-eof} row b.
-The annular pattern is captured mainly by the first EOF while the cEOF2 is explained by the second and third.
-However, even though the annular pattern used to construct these data is perfectly zonally symmetric, the leading EOF is not.
-The EOF decomposition mixes the true annular pattern with parts of the the zonally asymmetric cEOF2 pattern, resulting in a pattern pretty similar to the real SAM pattern \citep{fogt2020}.
-
-\textcolor{red}{Todo esto está mal.}
-
-Rotating makes it even worse.
-
-The reason this happens is that the spatial correlation between the annular pattern and the wave patterns is not zero; i.e.~they are not orthogonal. And since EOFs
-
-In this particular example, it is 0.022 (CI: -0.29 -- 0.33) with the Real part and 0.016 (CI: -0.29 -- 0.32) with the Imaginary part.
-Of note, the three EOFs are well separated according to North's rule of thumb \citep{north1982}.
-
-This suggests that the zonal asymmetries of the SON SAM could very well be mostly a statistical artefact.
-Since many surface impacts are mediated by the asymmetric component, as well as the relationship between SAM and ENSO \citep{campitelli2021}, this potential issue could affect the interpretability of many results involving the SAM.
-Studies on the relationship between the SAM and PSA-like patterns would be particularly difficult to interpret.
-
-\hypertarget{tropical-sources}{%
-\subsection{Tropical sources}\label{tropical-sources}}
+\hypertarget{tropical}{%
+\subsection{Tropical sources}\label{tropical}}
 
@@
 
-The connections between cEOFs and Tropical sources is also assesed.
+The connections between cEOFs and Tropical sources is also assessed.
 Figure~\ref{fig:sst-psi-2} shows the regression maps of Sea Surface Temperatures (SST) and Streamfunction at 200 hPa respectively upon cEOF2.
@@
 However, Figure~\ref{fig:sst-psi-2}.a1 and b1 show that the Real cEOF2 is not associated either with any significant SST nor streamfunction anomalies in the tropics.
-The correlation between the Real cEOF2 and ENSO is also not significant (-0.0042 (CI: -0.31 -- 0.3)).
+The correlation between the Real cEOF2 and ENSO is also not significant (0.00025 (CI: -0.31 -- 0.31)).
 This lack of tropical signal suggests different natures of the different phases of the Real cEOF2 wave train, which has the maximum amplitudes in the surroundings of southern South America.
@@
 \includegraphics{../figures/enso-phase-1.pdf}
-\caption{\label{fig:enso-phase} ONI plotted against cEOF2 phase for th 1979 -- 2019 period.
+\caption{\label{fig:enso-phase}ONI plotted against cEOF2 phase for th 1979 -- 2019 period.
 Years with magnitude of cEOF2 greater or smaller than the 50th percentile are shown as orange circles and green diamomnds respectively.
@@
 
-\textcolor{red}{¿Qué significa para el fit "weighted by the magnitude of cEOF2"?}
-
 Furthermore, Figure~\ref{fig:enso-phase} suggest that higher amplitude cEOF2 years tend to coincide with stronger ENSO years.
@@
 
-The precipitation anomalies in Oceania associated with cEOF1 are also weak and not statistically significant (Figure~\ref{fig:pp-oceania} column a) whereas EFO2 is associated with large and widespread anomalies.
+The precipitation anomalies in Oceania associated with cEOF1 are weak and not statistically significant (Figure~\ref{fig:pp-oceania} column a) whereas EFO2 is associated with large and widespread anomalies.
 The Real cEOF2 is associated with positive seasonal anomalies in Northern and Eastern Australia, and negative ones over New Zealand, while precipitation anomalies associated with Imaginary cEOF2 are negative over all Eastern Australia.
@@
 \includegraphics{../figures/australia-pp-phase-1.pdf}
-\caption{\label{fig:australia-pp-phase}Same as Figure~\ref{fig:enso-phase} but for precipitation anomalies averaged between 130°E and 155°E and 40°S and 15°S (box shown in Figure~\ref{fig:pp-oceania} column b).
+\caption{\label{fig:australia-pp-phase}Same as Figure~\ref{fig:enso-phase} but for precipitation anomalies averaged between 130°E and 155°E and 40°S and 15°S (box shown in Figure~\ref{fig:pp-oceania} column b) in \(\frac{\mathrm{mm}}{\mathrm{day}}\).
 Black line is the fit precipitation \textasciitilde{} sin(phase) + cos(phase) computed by weighted OLS using the magnitude of the cEOF2 as weights.}
@@
 Figure~\ref{fig:pp-america} shows regression maps of seasonal precipitation with each cEOF in South America.
-cEOF1 is associated with slight decreases \textcolor{red}{Revisa como estás expresando toda esta parte (y en el resto del texto), recorda que son regresiones así que no corresponde decir postive/negative sino haces alguna referencia a la variable con las que los estás regresionando.} in precipitation in Southern Brazil and Paraguay, although these are not statistically significant (Fig.~\ref{fig:pp-america} column a).
+cEOF1 is associated with slight decreases in precipitation in Southern Brazil and Paraguay, although these are not statistically significant (Fig.~\ref{fig:pp-america} column a).
 The strongest precipitation regression coefccients are the ones associated with the Imaginary cEOF2.
-The positive anomalies on Southeastern South America (SESA) and Central Chile, and negative anomalies over South Atlantic Convergence Zone is a well known springtime precipitation signature of ENSO \citep{cai2020a} and it is also similar to the precipitation anomalies associated with the A-SAM \citep{campitelli2021}.
+The positive regression coefficients on Southeastern South America (SESA) and Central Chile, and negative ones over South Atlantic Convergence Zone is a well-known springtime precipitation signature of ENSO \citep{cai2020a} and it is also similar to the precipitation anomalies associated with the A-SAM \citep{campitelli2021}.
 This is not surprising considering the close relationship between the ONI, the A-SAM index and the Imaginary cEOF2 shown previously, but further consolidates the identification of this mode with the PSA pattern.
@@
 Resembling the relationship between ONI and the phase of cEOF2 (Fig.~\ref{fig:enso-phase}), there is a cEOF2 phase dependence of the precipitation anomalies in SESA (not shown).
-These variable could be related wth a signficant sinusoidal fit that has a coefficient of determination of~0.27 (p-value~= 0.0022).
-
-\textcolor{red}{La formula en la figura trata de escribirla en una forma más precisa "precipitación"/"phase" son muy generales, y también teniendo en cuenta las unidades.
-También incluilo en el caption.}
+These variables could be related wth a significant sinusoidal fit that has a coefficient of determination of~0.27 (p-value~= 0.0022).
 
@@
 
-THe second complex EOF represents a wave-3 pattern with maximum magnitude in the Pacific sector.
+The second complex EOF represents a wave-3 pattern with maximum magnitude in the Pacific sector.
 Essentially, it is an alternate representation of the PSA1 and PSA2 patterns \citep{mo2001}.
@@
 This is because the when ENSO is active, it forces the cEOF2 to be in the Imaginary phase.
-When ENSO is neutral, the cEOF2 is still active, but with no particular phase.
+When ENSO is neutral, the cEOF2 is still active, but with no particular preferred phase.
 This mirrors the results of \citet{cai2002}, who showed that the CSIRO Model can create PSA-like variability even in the absence of ENSO forcing (with a climatological run), but the variability of one of the PSA modes was enhanced when adding the ENSO signal.
+The magnitude of the cEOF2 appears not to be related to the magnitude of ENSO, except frot the three strongest ENSO events in the period which coincide with the three strongest cEOF2 years.
 

### 2021-07-27 R/theme.R

@@ -13,7 +13,7 @@ theme_shceof <- function(base_size = 16) {
       # text = element_text(family = font_rc),
       legend.position = "bottom", legend.box = "vertical",
       tagger.panel.tag.background = ggplot2::element_rect(color = NA),
-      strip.text = ggplot2::element_text(size = ggplot2::rel(12/16)),
+      strip.text = ggplot2::element_text(size =  ggplot2::rel(11/16)),
       # panel.spacing.y = grid::unit(5, "mm"),
       # panel.spacing.x = grid::unit(5, "mm"),
       # legend.spacing = grid::unit(2, "mm"),
@@ -22,7 +22,6 @@ theme_shceof <- function(base_size = 16) {
       # # legend.title = element_blank(),
       # legend.box.spacing = grid::unit(3, "mm"),
       # legend.margin = ggplot2::margin(t = -5),
-      panel.grid = ggplot2::element_blank(),
-      # panel.grid = ggplot2::element_line(color = "gray10", size = 0.4, linetype = 3),
+      panel.grid = ggplot2::element_line(color = scales::alpha("gray60", 0.5), size = 0.1),
       panel.ontop = TRUE)
 }

### 2021-07-29 analysis/paper/shceof.Rmd
Agrega más intro

@@
 
-\textcolor{red}{Agregar parráfo que introduzca el problema del estudio de la ciriculación extra-tropical en SH, su variaiblidad y sus cambios.
-Quizás algo también de los impactos.
-También algo habría que decir de la primavera, ya que solo nos concentrmaos en esa estación}
-
 The large-scale extratropical circulation in the Southern Hemisphere is strongly zonally symmetric, but its zonal departures are highly relevant for surface impacts and could be related to high-impact climate extremes.
 Zonal asymmetries of extratropical circulation in the Southern Hemisphere strongly modulate weather systems and regional climate through latitudinal transport of heat, humidity, and momentum [@trenberth1980].
+
+For example, @raphael2007 suggests that variability in the planetary wave 3 projected onto its climatological location is associated with Antarctic sea ice concentration anomalies.
+@irving2015 showed that planetary wave amplitude in general is associated sea ice concentration, as well as temperature and precipitation anomalies.
+Another characteristic of the zonally asymmetric circulation are the Pacific--South American Patterns [@mo2001], which are associated with precipitation anomalies in South America.
+
+Some of these patterns of variability appear to have experienced secular changes.
+@raphael2003 found that the amplitude of the zonal wave 1 experienced a large increase and that the zonal wave 3 experienced changes in its annual cycle between 1958 and 1996.
 
@@
 
+The fourier technique can be generalised to integrate all planetary wave intensity irregardless of wave number by computing the wave envelope as done by @irving2015.
+This makes it possible to represent planetary waves with different amplitude at different longitudes, but it removes all information about phase and wave number.
+
 Another approach to characterising the Southern Hemisphere circulation is by using Empirical Orthogonal Functions (EOF, also known as Principal Component Analysis).
-Within the EOF framework, the Southern Annular Mode (SAM) appears as the leading mode of variability of the tropospheric Southern Hemisphere circulation [@fogt2020] followed by the two Pacific--South American Patterns (PSA) [@mo2001].
+Within the EOF framework, the Southern Annular Mode (SAM) appears as the leading mode of variability of the tropospheric Southern Hemisphere circulation [@fogt2020] followed by the two PSA [@mo2001].
 The SAM represents a relatively zonally symmetric pattern of alternating low pressures in polar latitude and a ring of high pressures in high latitudes with an embedded wave 3 pattern that is more prominent in the Pacific sector.

### 2021-07-29 analysis/paper/shceof.tex
Render

@@
 
-\textcolor{red}{Agregar parráfo que introduzca el problema del estudio de la ciriculación extra-tropical en SH, su variaiblidad y sus cambios.
-Quizás algo también de los impactos.
-También algo habría que decir de la primavera, ya que solo nos concentrmaos en esa estación}
-
 The large-scale extratropical circulation in the Southern Hemisphere is strongly zonally symmetric, but its zonal departures are highly relevant for surface impacts and could be related to high-impact climate extremes.
@@
 
+For example, \citet{raphael2007} suggests that variability in the planetary wave 3 projected onto its climatological location is associated with Antarctic sea ice concentration anomalies.
+\citet{irving2015} showed that planetary wave amplitude in general is associated sea ice concentration, as well as temperature and precipitation anomalies.
+Another characteristic of the zonally asymmetric circulation are the Pacific--South American Patterns \citep{mo2001}, which are associated with precipitation anomalies in South America.
+
+Some of these patterns of variability appear to have experienced secular changes.
+\citet{raphael2003} found that the amplitude of the zonal wave 1 experienced a large increase and that the zonal wave 3 experienced changes in its annual cycle between 1958 and 1996.
+
 One typical way of describing the zonally asymmetric circulation is by the amplitude and phase of zonal waves obtained by Fourier decomposition of geopotential height at each latitude \citep[e.g.][]{vanloon1972, trenberth1980, turner2017}.
 This approach suggests that zonal waves 1 and 3 explained most of the variance of the tropospheric extratropical Southern Hemisphere circulation (\citet{vanloon1972}).
-However, this methodology rely on the assumption that the circulation can be meaningfully understood in terms of zonal waves of constant amplitude along a latitude circle.
+However, this methodology relies on the assumption that the circulation can be meaningfully understood in terms of zonal waves of constant amplitude along a latitude circle.
 This is not valid for meridionally propagating waves or zonal waves with localised amplitudes.
@@
 
+The fourier technique can be generalised to integrate all planetary wave intensity irregardless of wave number by computing the wave envelope as done by \citet{irving2015}.
+This makes it possible to represent planetary waves with different amplitude at different longitudes, but it removes all information about phase and wave number.
+
 Another approach to characterising the Southern Hemisphere circulation is by using Empirical Orthogonal Functions (EOF, also known as Principal Component Analysis).
-Within the EOF framework, the Southern Annular Mode (SAM) appears as the leading mode of variability of the tropospheric Southern Hemisphere circulation \citep{fogt2020} followed by the two Pacific--South American Patterns (PSA) \citep{mo2001}.
+Within the EOF framework, the Southern Annular Mode (SAM) appears as the leading mode of variability of the tropospheric Southern Hemisphere circulation \citep{fogt2020} followed by the two PSA \citep{mo2001}.
 The SAM represents a relatively zonally symmetric pattern of alternating low pressures in polar latitude and a ring of high pressures in high latitudes with an embedded wave 3 pattern that is more prominent in the Pacific sector.
@@
 Similarly, \citet{hobbs2010} propose that the zonally asymmetric circulation in the Southern Hemisphere can be described by the strength and location of two anticyclones located in the sub-Antarctic western hemisphere.
-These methods are grounded on other methods to identify the centers of action for the described phenomena and can be useful to characterise features that are not readily apparent with Fourier or EOF methods.
-The use of data of only specific regions or locations makes them suitable for easy computation and for extending into less quality data from the past, but also makes them not very robust and could not capture non-stationary behaviours of the patterns.
+These methods are grounded on other methods to identify the centres of action for the described phenomena and can be useful to characterise features that are not readily apparent with Fourier or EOF methods.
+The use of data of only specific regions or locations makes them suitable for easy computation and for extending back into the data-poor past, but also makes them not very robust and unable to capture non-stationary patterns.
 
-\textcolor{red}{Falta un parráfo que deje más claro por qué es necesario considerar un enfoque alternativo y cuál va a ser el beneficio de ello.
-También agregar qué se estudiara conjuntamente la troposfera y la estratósfera }
+Our objective is to create an index that is able to capture the phase-varying nature of planetary waves and is flexible enough to describe planetary waves with non-constant amplitude in each latitude circle.
+We use complex Empirical Orthogonal Functions \citep{horel1984} as a robust extension of EOF analysis that is capable of describing rich structures with meridional and zonal propagation, and amplitude modulation such as the PSA.
 
-\textcolor{red}{Falta el objetivo que justifica el uso de la metodología}
-
-In this paper we propose using complex Empirical Orthogonal Functions \citep{horel1984} as a robust extension of EOF analysis that is capable of describing rich structures with meridional and zonal propagation, and amplitude modulation such as the PSA.
-We apply this method to characterise the Springtime zonally asymmetric circulation of the extratopical Southern Hemisphere.
+We apply this method to characterise the Springtime zonally asymmetric circulation of the extratropical Southern Hemisphere.
+To take vertical distribution into account, we analyse both the troposphere and the stratosphere.
+We focus the analysis to the September-October-November because that's the trimester in which tropical forcing of zonal waves is maximised (cita).
 
@@
 Table \ref{tab:corr-ceof-splitted} shows the coefficient of determination between time series of the amplitude of each complex EOF (cEOF) across levels.
-There's a high degree of correlation between the respective cEOF1 and cEOF2 at each level.
+There's a high degree of correlation between the magnitude of the respective cEOF1 and cEOF2 at each level.
 The spatial patterns of the 50 hPa and 200 hPa cEOFs are also similar (not shown).
@@
 The result obtained here is similar to the one obtained by \citet{irving2016}.
-From their Fourier decomposition of reprojected meridional wind fields \textcolor{red}{¿En qué región?}, they show that the phase distribution is bimodal and isolate the PSA pattern from the rest of the PSA-like variability by selecting events that are near the peaks of the distribution (compare our Figure~\ref{fig:phase-histogram} with their Figure 6).
+They show that the phase distribution of the PSA-like variability they detected is bimodal and isolate the PSA pattern from the rest of the PSA-like variability by selecting events that are near the peaks of the distribution (compare our Figure~\ref{fig:phase-histogram} with their Figure 6).
 
@@
 Figure \ref{fig:t-vertical} shows the vertical distribution of the regression coefficients on cEOF1 from zonal anomalies of air temperature and zonal anomalies of ozone mixing ratio averaged between 75°S and 45°S.
-\textcolor{red}{Deberias aclarar qué significa la terminología que estas utilizando. Lo podes hacer en los métodos y ya queda para todo el texto}.
 Temperature zonal anomalies associated with cEOF1 show a clear wave 1 pattern for both real and imaginary components throughout the atmosphere above 250 hPa with a change in sign above 10 hPa.
@@
 This is consistent with the effect of ENSO on the extratropics: SST anomalies initiate anomalous convection that excites Rossby waves that propagate meridionally towards higher latitudes \citep{mo2000}.
-
-\textcolor{red}{Revisa los labels de la colorbar para la figura 13.
-Además ojo que solamente sirven para la columna a, faltan los correspondientes a la streamfunction y el vector de referencia para wave flux}
 
@@
 
-\textcolor{red}{Al igual que en la introducción, te falta establecer más claro el objetivo, después iría lo que escribiste.}
-
-We studied the two leading modes of co-variability of September-October-November tropospheric and stratospheric zonal anomalies by applying complex Empirical Orthogonal Functions.
+In this study we derived two complex indices of the zonally asymmetric circulation of the September-October-November southern extratropics using Complex Empirical Orthogonal Functions.
+This allowed us to characterise both the amplitude and phase of planetary waves that are not perfectly organised as sinusoidal waves with constant amplitude at each latitude circle.
 
@@
 On the other hand, over Australia, both the Real and Imaginary phase are associated with significant precipitation anomalies, and we further show that the direction of maximum impact is not aligned with our chosen rotation of cEOF2.
-However, this underscore the benefit of using complex EOF, since it would trivial to rotate it.
+However, this underscores the benefit of using complex EOF, since it would trivial to rotate it.
 
-\textcolor{red}{Debería cerrarse con algún parráfo que diga qué tiene de bueno trabajar con esta metodología y qué permitiría hacer a futuro (se puede poner que sería deseable extenderlo a otras estaciones, etc.)}
+To the extent that the cEOF2 can be identified with the PSA, what is the advantage of our technique?
+The spatial fields that the cEOF2 index describes are in quadrature by construction, which makes it possible to derive a proper amplitude and phase.
+Furthermore, because of the way they are constructed, they are not forced to be orthogonal to the SAM.
+This makes it possible to study the relationship between the two modes free of the statistical contamination that comes with the usual definition of the PSA as the second and third EOFs.
+
+Further work should extend this analysis to other seasons and further study the relationship between the cEOF2 and the SAM.
+In particular, it is not clear whether the close identification between the Imaginary cEOF2 and the Asymmetric SAM arises from physical processes or statistical contamination.
 

### 2021-08-03 analysis/paper/shceof.Rmd
revision lean

@@
 ---
-title: "Austral Spring Extratropical Southern Hemisphere zonally assymetric circulation"
+title: "Revisiting the Austral Spring Extratropical Southern Hemisphere zonally assymetric circulation using complex Empirical Orthogonal Functions"
 titlerunning:  Title
@@
 
-For example, @raphael2007 suggests that variability in the planetary wave 3 projected onto its climatological location is associated with Antarctic sea ice concentration anomalies.
+One of the main features of the zonally asymmetric circulation are the Pacific--South American Patterns [@mo2001], which are associated with precipitation anomalies in South America.
+Besides, @raphael2007 suggests that variability in the planetary wave 3 projected onto its climatological location is associated with Antarctic sea ice concentration anomalies.
 @irving2015 showed that planetary wave amplitude in general is associated sea ice concentration, as well as temperature and precipitation anomalies.
-Another characteristic of the zonally asymmetric circulation are the Pacific--South American Patterns [@mo2001], which are associated with precipitation anomalies in South America.
-
-Some of these patterns of variability appear to have experienced secular changes.
+
+Some of these patterns of variability associated to the zonally asymmetric circulation appear to have experienced secular changes.
 @raphael2003 found that the amplitude of the zonal wave 1 experienced a large increase and that the zonal wave 3 experienced changes in its annual cycle between 1958 and 1996.
+\textcolor{red}{¿esto es lo único que encontraste asociado a esta afirmación?
+en caso de que si, diría de agregar que "ha sido poco estudiado hasta el momento"}
 
@@
 The inability of a single EOF to describe travelling waves is what forces the PSA pattern to be described by two EOFs.
-@irving2016 characterises the Pacific--South American Pattern in a Fourier framework.
-Being a meridionally propagating mode, the PSA cannot be correctly characterised by Fourier decomposition at each latitude circle, so they reprojected meridional wind fields so that the path of the PSA laid on the equator.
+@irving2016 characterises this pattern in a Fourier framework.
+Being a meridionally propagating mode, the PSA cannot be correctly characterised by Fourier decomposition at each latitude circle.
+As a result, they reprojected meridional wind fields so that the path of the PSA laid on the equator.
 On this new projection, they could identify the PSA using Fourier decomposition.
@@
 Our objective is to create an index that is able to capture the phase-varying nature of planetary waves and is flexible enough to describe planetary waves with non-constant amplitude in each latitude circle.
+\textcolor{red}{yo no pondría que el objetivo es crear el índice, sino porfundizar el conocimiento de la variabilidad...
+utilizando un índice...}
 We use complex Empirical Orthogonal Functions [@horel1984] as a robust extension of EOF analysis that is capable of describing rich structures with meridional and zonal propagation, and amplitude modulation such as the PSA.
@@
 We focus the analysis to the September-October-November because that's the trimester in which tropical forcing of zonal waves is maximised (cita).
+\textcolor{red}{Este parráfo me parece que quedo un poco colgado.
+Creo que sería importante que esté desde más arriba la idea de la primavera, ya que todo lo otro es muy general, pero al final el análisis solo se va a restringir a esta estación}
 
@@
 
-We used monthly geopotential height, air temperature, ozone mixing ratio, and Total Ozone Column (TOC) at 2.5\degree longitude by 2.5\degree latitude of horizontal resolution and 37 vertical isobaric levels from ERA5 [@era5] for the period 1979 to 2019.
+We used monthly geopotential height, air temperature, ozone mixing ratio, and total ozone column (TOC) at 2.5\degree longitude by 2.5\degree latitude of horizontal resolution and 37 vertical isobaric levels from ERA5 reanalysis [@era5] for the period 1979-2019.
 Most of our analysis was restricted to the post-satellite era to avoid any confounding factors arising from the incorporation of satellite observations, but we also used the preliminary back extension of ERA5 from 1950 to 1978 [@era5be] to look at long-term trends.
 Streamfunction was derived from ERA5 vorticity at 200 hPa using the FORTRAN subroutine FISHPACK [@fishpack] and horizontal wave activity fluxes was computed following @plumb1985.
-We usded Sea Surface Temperature (SST) monthly fields from Extended Reconstructed Sea Surface Temperature (ERSST) v5 [@huang2017] and precipitation monthly data from the CPC Merged Analysis of Precipitation [@cmap], with a 2.5\degree resolution in latitude and longitude.
+We used Sea Surface Temperature (SST) monthly fields from Extended Reconstructed Sea Surface Temperature (ERSST) v5 [@huang2017] and precipitation monthly data from the CPC Merged Analysis of Precipitation [@cmap], with a 2.5\degree resolution in latitude and longitude.
 This rainfall gridded dataset is based on information from different sources such as rain gauge observations, satellite inferred estimations and the NCEP-NCAR reanalysis, and it is available since 1979 to present.
@@
 
-The amplitude of the zonal waves was defined through computing the Fourier transform of the spatial field at each latitude circle.
-We computed the amplitude and phase of the zonal wave 1 by averaging (area-weighted) variables (temperature, geopotential height, ozone mixing ratio and Total Ozone Column) between `r knitr::combine_words(LatLabel(o3wave_lats))` for each SON and extracting the wave-1 component of the Fourier spectrum.
-We chose this latitude band because it is wide enough to capture most of the relevant anomalies.
-
-We computed the level-dependent Southern Annular Mode (SAM) index as the leading EOF of year-round monthly geopotential height anomalies south of 20ºS at each level for the whole period [@baldwin2009].
+The amplitude of the zonal waves was obtained through the Fourier transform of the spatial field at each latitude circle.
+We computed the amplitude and phase of the zonal wave 1 by averaging (area-weighted) each variable between `r knitr::combine_words(LatLabel(o3wave_lats))` for each SON and extracting the wave-1 component of the Fourier spectrum.
+We chose this latitude band because it is wide enough to capture most of the relevant anomalies for southern hemisphere mid-latitudes.
+
+We computed the level-dependent SAM index as the leading EOF of year-round monthly geopotential height anomalies south of 20ºS at each level for the whole period according to @baldwin2009.
 We further split the SAM into its zonally symmetric and zonally asymmetric components (S-SAM and A-SAM indices respectively).
 These indices were obtained by projecting the zonally asymmetric and zonally symmetric part of the SAM spatial pattern onto monthly geopotential height fields, as proposed by @campitelli2021.
-The Pacific South American patterns (PSA1 and PSA2) were calculated following @mo2001 as the third and fourth leading EOF of seasonal mean anomalies for SH 500 hPa geopotential height combining all seasons.
-
-Linear trends were computed by Ordinary Least Squares and the 95% confidence interval was computed assuming a t-distribution with the appropriate residual degrees of freedom [@wilks2011r].
+The Pacific South American patterns (PSA1 and PSA2) were calculated following @mo2001 as the third and fourth leading EOF of seasonal mean anomalies for SH 500 hPa geopotential height combining all seasons \textcolor{red}{leyendo esto me queda la duda.
+Vos calculaste esto usando todos los meses y luego tomaste el valor de SON, no?
+Creo que valdría la pena aclararlo mejor, por lo menos a mi me resulta confuso}.
+
+Linear trends were computed by Ordinary Least Squares (OLS) and the 95% confidence interval was computed assuming a t-distribution with the appropriate residual degrees of freedom [@wilks2011r].
 
@@
 (ref:eof-naive-cap) Spatial patterns of the leading 4 EOFs of SON zonal anomalies of geopotential height at 50 hPa south of 20\degree S for the 1979 -- 2019 period (arbitrary units).
+
+\textcolor{red}{En la figura 1 te quedaron muy chicos los subplots, amplialos un poco}.
 
@@
 
+\textcolor{red}{Los colores de la Tabla 1 no dicen mucho en la medida que no se conoce cuál sería la paleta. Quizás se pueden remarcar en negrita ambos aclarando que son significativos (lo son?)}.
+
 ```{r corr-ceof-splitted}
@@
 Since any choice is arbitrary and equally valid, we chose the phase of each cEOF so that the real and imaginary parts are aligned with meaningful phases in our analysis.
-This procedure does not create a spurious correlation, it only takes whatever relationship that already exist and aligns it with a specific phase.
+This procedure does not create a spurious correlation, it only takes a known relationship and aligns it with a specific phase.
+For the first complex principal component, the phase was chosen so that the time series corresponding to the real part has the maximum correlation with the zonal wave 1 of Total Ozone Column between `r knitr::combine_words(LatLabel(o3wave_lats))`.
+This also nearly minimises the correlation with the imaginary part.
+For the second complex principal component, the phase was chosen so that the coefficient of determination between the Oceanic Niño Index [@bamston1997] and the real part was minimized, which also nearly maximizes the correlation with the imaginary part.
 
@@
 
-For the first complex principal component, the phase was chosen so that the time series corresponding to the real part has the maximum correlation with the zonal wave 1 of Total Ozone Column between `r knitr::combine_words(LatLabel(o3wave_lats))`.
-This also nearly minimises the correlation with the imaginary part.
-
-For the second complex principal component, the phase was chosen so that the coefficient of determination between the Oceanic Niño Index [@bamston1997] and the real part was minimized, which also nearly maximizes the correlation with the imaginary part.
-
 While we compute these complex principal components using data from 1979 to 2019, we extended the complex time series back to the 1950 -- 1978 period by projecting monthly geopotential height zonal anomalies standardised by level south of 20ºS onto the corresponding spatial patterns.
@@
 We performed linear regressions to quantify the association between the cEOFs and other variables.
-For each cEOF, we create regression maps by computing fitting a multiple linear model involving both the real and the imaginary part.
+For each cEOF, we create regression maps by fitting a multiple linear model involving both the real and the imaginary part.
 To obtain the linear coefficients of a variable $X$ (geopotential, temperature, precipitation, etc...) with the Imaginary and Real parts of each cEOF we fit the equation
@@
 
-where $\lambda$ and $\phi$ are the longitude and latitude, $t$ is the time, $\alpha$ and $\beta$ are the linear regression coefficients, $X_0$ and $\epsilon$ are the constant and error terms respectively.
+where $\lambda$ and $\phi$ are the longitude and latitude, $t$ is the time, $\alpha$ and $\beta$ are the linear regression coefficients for Imaginary and Real parts respectively, $X_0$ and $\epsilon$ are the constant and error terms respectively.
 
@@
 
-To describe the variability of the circulation zonal anomalies, the spatial and temporal parts of the first two leading cEOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa, computed jointly at both levels, are shown in Figures \@ref(fig:ceofs-1) and \@ref(fig:extended-series,).
+To describe the variability of the circulation zonal anomalies, the spatial and temporal parts of the first two leading cEOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa, computed jointly at both levels, are shown in Figures \@ref(fig:ceofs-1) and \@ref(fig:extended-series).
 The first mode (cEOF1) explains 82% of the variance, while the second mode (cEOF2) explains a smaller fraction (7%).
@@
 
-The first cEOF (Fig. \@ref(fig:eof-naive) column a) is a hemispheric wave 1 pattern with maximum amplitude at high latitudes.
+\textcolor{red}{tengo la impresión de que en la figura 2 podría estar un poco más grandes los subplots para que se vean más fácilmente los detalles}
+
+The first cEOF (Fig. \@ref(fig:ceofs-1) column a) is a hemispheric wave 1 pattern with maximum amplitude at high latitudes.
 At 50 hPa the Real cEOF1 has the maximum of the wave 1 at 150ºE and at 200 hPa, the maximum is located at around 175ºE indicating a westerly shift in phase.
-The second cEOF (Fig. \@ref(fig:eof-naive) column b) shows also a zonal wave-like structure with maximum amplitude at high latitudes, but with shorter spatial scales.
+The cEOF2 (Fig. \@ref(fig:eof-naive) column b) shows also a zonal wave-like structure with maximum amplitude at high latitudes, but with shorter spatial scales.
 In particular, its dominant structure at both levels is a wave 3 but with larger amplitude in the pacific sector.
@@
 
-The temporal variability for both components of the cEOF1 have non zero mean (Fig. \@ref(fig:extended-series,)).
+\textcolor{red}{se me vino una duda, ¿se pueden comparar de algún modo las amplitudes que mostras en esta figura?
+en caso de que si, creo que habría que mencionar algo sobre lo que pasa con el cEOF2 en la estratósfera}
+
+The temporal variability for both components of the cEOF1 have non zero mean (Fig. \@ref(fig:extended-series)). \textcolor{red}{Esto vale también para el cEOF2, ¿no? en ese caso, lo diría más en gral}
 This is due to the fact that the geopotential fields that enter into the cEOFs algorithm are anomalies with respect to the zonal mean, not the time mean.
@@
 
-There is a significant upward trend in the real component of cEOF1 (Fig. \@ref(fig:extended-series)a.1, p-value `r trends[part == "Real" & cEOF == "cEOF1"]$p_report`) and no significant trend in any of the complex components of cEOF2.
+\textcolor{red}{En la figura 3 agregaría dos cosas: i)el eje y en la columna derecha (si la queres ver más grande perdes la referencia) ii) creo que serviría agregar una líneas verticales en la grilla para poder tener una idea un poco mejor de a qué años corresponde más o menos cada valor.
+pueden ser un poco más finas que las que ya están, pero creo que sería una referencia util para visualmente poder decir a qué año corresponde cada punto.
+Además, fijate que la figura te quedo de un ancho mayor que el de la página y te quedo un "EOF" en el eje-y a la izquierda que no creo que sea lo más correcto.}
+
+There is only a significant upward trend in the real component of cEOF1 (Fig. \@ref(fig:extended-series)a.1, p-value `r trends[part == "Real" & cEOF == "cEOF1"]$p_report`) and no significant trend in any of the complex components of cEOF2.
 The positive trend in the real cEOF1 translates into a positive trend in cEOF1 magnitude, but not in the phase (not shown).
-
 This long-term change indicates an increase in the magnitude of the high latitude zonal wave 1.
 A similar observation was made by @raphael2003, who detected a step after around 1975 in the leading EOF of August-September-October 500 hPa zonal geopotential height anomalies.
+\textcolor{red}{Creo que acá faltaría decir que el modo del trabajo de Raphael se parece al que estás mostrando vos, sino no tendria mucho sentido la comapración.}
 
@@
 Figure \@ref(fig:eof1-regr-gh) shows regression maps of SON geopotential height anomalies upon cEOF1.
-At 50 hPa (Figure \@ref(fig:eof1-regr-gh) row a), both the Real and Imaginary cEOF1 are associated with planetary wave 1 patterns, that are 90º off phase.
+At 50 hPa (Figure \@ref(fig:eof1-regr-gh) row a), both the Real and Imaginary cEOF1 are associated with planetary wave 1 patterns, that are 90º out of phase.
 Their phases coincide with the ones shown in Figure \@ref(fig:ceofs-1)a.1, with the positive centre of the real cEOF1 located towards the dateline, and the one of the Imaginary cEOF1 located over Eastern Antarctica.
@@
 
-The regression patterns at 200 hPa (Figure \@ref(fig:eof1-regr-gh) row b) are similarly influenced by the zonally symmetric circulation.
-It is only possible to distinguish partially the wave 1 pattern in relation with the real cEOF1 (Figure \@ref(fig:eof1-regr-gh)b.1).
+\textcolor{red}{La figura 4 la veo muy saturada con los puntos que indican la significancia.
+¿se podrán hacer con un tamaño menor?
+¿Las unidades son correctas?
+Me llama la atención que te puedan dar tantos cientos de metros de altura geopotencial.}
+
+The regression patterns for cEOF1 at 200 hPa (Figure \@ref(fig:eof1-regr-gh) row b) are similarly influenced by the zonally symmetric circulation.
+It is only possible to partially distinguish the wave 1 pattern in relation with the real cEOF1 (Figure \@ref(fig:eof1-regr-gh)b.1).
 The imaginary cEOF1 shows a much more zonally symmetrical pattern resembling the negative SAM phase.
@@
 Zonal wave 3 dominates all fields, but only in the western hemisphere, over the Pacific and Atlantic Oceans.
-cEOF2 then represents a equivalent barotropic wave train that is very similar to the the Pacific South American Patterns [@mo2001].
+cEOF2 then represents an equivalent barotropic wave train that is very similar to the the Pacific South American Patterns [@mo2001].
 Comparing the location of the positive anomaly near 90ºW in column b of Figure \@ref(fig:eof2-regr-gh) with Figures 1.a and b from @mo2001, the real cEOF2 can roughly be identified with PSA2, while the imaginary cEOF2 resembles PSA1.
+
+\textcolor{red}{Figura 5. Me resulto raro que vaya hasta 999 en vez de 1000 }
 
@@
 
+\textcolor{red}{Este título queda raro porque después hablas de modos de variabilidad climática y no de lo que llamaríamos variables}
+
 ### SAM
@@
 To explore the relationship between SAM and the modes described from the cEOF, we compute the coefficient of determination between the cEOFs timeseries and the three SAM indices (SAM, A-SAM and S-SAM) at each vertical level (Fig. \@ref(fig:sam-eof-vertical)).
-The SAM index is moderately correlated with the Real cEOF1 in all levels, and with the Imaginary cEOF1 and Imaginary cEOF2 in the troposphere.
-It is not correlated with the Real cEOF2 at any level.
+The SAM index is statistically signficant correlated  with the Real cEOF1 in all levels, and with the Imaginary cEOF1 and Imaginary cEOF2 in the troposphere.
+Correlations between SAM and the Real cEOF2 are non-significant.
+
+\textcolor{red}{Fig.
+6.
+Te quedan muy pegados los subplots de la columna izquierda con los de la derecha.
+Recomendaría aumentar el espacio intermedio y repetir el eje y en las figuras de la derecha.
+Además podrías agregar el y-label (Level) a la izquierda.}
 
@@
 In the stratosphere, the Real cEOF1 is correlated with both A-SAM and S-SAM, while the Imaginary cEOF1 is highly correlated only with the A-SAM.
-These correlations are consistent with the regression patterns of geopotential height in Figure \@ref(fig:eof1-regr-gh).
+These correlations are consistent with the regression patterns of geopotential height in Figure \@ref(fig:eof1-regr-gh) and their comparison with those obtained for SAM, A-SAM and S-SAM by \textcolor{red}{Campitelli et al 2021}.
 
@@
 In the case of cEOF2, the moderate correlation between SAM and Imaginary cEOF2 becomes extremely high only when the zonally asymmetric variability of the SAM is considered.
-The Imaginary cEOF2 and the A-SAM share up to `r scales::percent(max_r2[cEOF == "cEOF2"]$estimate^2, 2)` of their variance, reached at `r max_r2[cEOF == "cEOF2"]$lev` hPa (Figure \@ref(fig:sam-eof-vertical).b2).
+The Imaginary cEOF2 explains up to `r scales::percent(max_r2[cEOF == "cEOF2"]$estimate^2, 2)` of A-SAM variance, reached the maximum value at `r max_r2[cEOF == "cEOF2"]$lev` hPa (Figure \@ref(fig:sam-eof-vertical).b2).
 Such extremely high correlation between the asymmetric SAM and the Imaginary cEOF2 suggests that these might be different ways of characterising the same phenomenon.
@@
 ### PSA
+
+\textcolor{red}{Tabla 2.
+Mismo comentario que antes con lo del background color (no dice mucho si no se conoce la escala), quizás podes destacar en negritas según la significancia.
+¿Está bien el intervalo de confianza de PSA2 vs cEOF2 Im?
+Hay algo raro ahí}
 
@@
 Table \@ref(tab:psa-eof2) shows the correlations between the two PSA indices and the timeseries for Real and Imaginary phase of cEOF2.
+\textcolor{red}{Valdría la pena aclarar por qué no consideras el cEOF1 para esta discusión}
 As anticipated by Figure \@ref(fig:eof2-regr-gh), there is a strong correlation between PSA1 and Imaginary cEOF2, and between PSA2 and Real cEOF2.
@@
 It could be seen that Imaginary phase is the most common phase and It is also the direction that has the maximum relationship with ENSO.
+\textcolor{red}{Ojo que todavía no estuvo la discusión con lo del ENSO, así que esto lo deberías decir más adelante.}
 Therefore, the Imaginary cEOF2 explains more variance than the Real cEOF2 and conventional EOF analysis will tend to separate the two.
@@
 The result obtained here is similar to the one obtained by @irving2016.
-They show that the phase distribution of the PSA-like variability they detected is bimodal and isolate the PSA pattern from the rest of the PSA-like variability by selecting events that are near the peaks of the distribution (compare our Figure \@ref(fig:phase-histogram) with their Figure 6).
-
-The advantage of our method is that it is much simpler to implement, it provides magnitude and phase naturally, and it facilitates the description of this mode as a propagating wave instead of as standing oscillation.
+They show that the phase distribution of the PSA-like variability obtained by them is bimodal and isolate the PSA pattern from the rest of the PSA-like variability by selecting events that are near the peaks of the distribution (compare our Figure \@ref(fig:phase-histogram) with their Figure 6).
+
+The advantage of our method is that it is much simpler to implement, it provides magnitude and phase naturally, and it facilitates the description of this mode as a propagating wave instead of as standing oscillations.
 As a consequence, the cEOF2 offers an alternative way of representing the PSA which has several advantages over using the second and third EOFs.
@@
 
-(ref:eof1-regr-t-cap) Same as Figure \@ref(fig:eof1-regr-gh) but for air temperature.
+(ref:eof1-regr-t-cap) Same as Figure \@ref(fig:eof1-regr-gh) but for air temperature \textcolor{red}{Unidades}.
 
@@
 
-The relation between cEOFs and temperature was also evaluated.
+The relation between cEOFs and air temperature was also evaluated.
 Figure \@ref(fig:eof1-regr-t) shows regression patterns of air temperature at 50hPa and 200hPa onto cEOF1.
-In both levels, the Real cEOF1 is associated with positive monopole over the South Pole with its centre moved slightly towards 150ºE (Fig. \@ref(fig:eof1-regr-t) column 1).
+In both levels, the Real cEOF1 is associated with a positive monopole over the South Pole with its centre moved slightly towards 150ºE (Fig. \@ref(fig:eof1-regr-t) column 1).
 On the other hand, the regression maps on the Imaginary cEOF1 show a more clear wave 1 pattern with its maximum around 60ºE.
@@
 
-The maximum ozone anomalies are co-located with the minimum temperature anomalies above 10 hPa and with the maximum temperature anomalies below 10 hPa (Fig. \@ref(fig:t-vertical))).
-Therefore, the ozone zonal wave 1 is anticorrelated with the temperature zonal wave 1 in the upper stratosphere, and directly correlated in the upper stratosphere .
+The maximum ozone anomalies are co-located with the minimum temperature anomalies above 10 hPa and with the maximum temperature anomalies below 10 hPa (Fig. \@ref(fig:t-vertical)).
+Therefore, the ozone zonal wave 1 is anticorrelated with the temperature zonal wave 1 in the upper stratosphere, and directly correlated in the upper stratosphere.
 This change in phase is observed in ozone anomalies forced by planetary waves that reach the stratosphere.
-In the photochemically-dominated upper stratosphere, cold temperatures inhibit the destruction of ozone, and the advectively-dominated lower stratosphere, ozone anomalies are 90º out of phase with horizontal and vertical transport, which is 90º our of phase with temperature anomalies [@hartmann1979; @wirth1993; @smith1995].
+In the photochemically-dominated upper stratosphere, cold temperatures inhibit the destruction of ozone.
+On the other hand, in the advectively-dominated lower stratosphere, ozone anomalies are 90º out of phase with horizontal and vertical transport, which are in adittion 90º out of phase with temperature anomalies [@hartmann1979; @wirth1993; @smith1995].
 
@@
 
-(ref:o3-regr-cap) Regression of SON mean Total Ozone Column anomalies (shaded, Dobson Units) with the (a) real and (b) imaginary parts of the first cEOF for the 1979 -- 2019 period.
+(ref:o3-regr-cap) Regression of SON mean Total Ozone Column anomalies (shaded, Dobson Units) with the (a) real and (b) imaginary parts of the cEOF1 for the 1979 -- 2019 period.
 On contours, the mean zonal anomaly of Total Ozone Column (negative contours in dashed lines, Dobson Units).
 Areas marked with dots have p-values smaller than 0.05 adjusted for False Detection Rate.
+
+\textcolor{red}{Revisa que tenes una incosistencia en el uso de mayúsculas en Real/Imaginary porque a veces las usas y a veces no.}
 
@@
 There is a close spatial relationship between amplitudes and phases of the cEOF1 and TOC planetary wave 1 between `r knitr::combine_words(LatLabel(o3wave_lats))`.
-The correlation between the amplitude of both indices is `r wave1_cor[with == "amplitude"]$text` the overall correlation between the zonal wave 1 of ozone and cEOF1 (computed as the mean cosine of the difference in phase) is `r wave_correlation$text`.
+The correlation between the amplitude of both indices is `r wave1_cor[with == "amplitude"]$text`, while the overall correlation between the zonal wave 1 of ozone and cEOF1 (computed as the mean cosine of the difference in phase) is `r wave_correlation$text`.
+\textcolor{red}{¿Podrías clarificarlo?
+No estoy seguro de qué es lo que hiciste}
 As expected from the location of regressions in Figure \@ref(fig:o3-regr), the Real cEOF1 drives the relationship of amplitudes and the Imaginary cEOF1 drives the relationship of phases between the timeseries (not shown).
+\textcolor{red}{Tampoco me quedo claro qué estás diciendo con esto y cómo se explica.}
 
@@
 Areas marked with dots have p-values smaller than 0.01 adjusted for FDR.
+\textcolor{red}{Agregar unidades.
+Quedaría mucho mejor el gráfico y creo que se prestaría a menos confusión si colocas las barras de colores en las columnas corrrespondientes.
+Otro detalle es que en gral.
+para el gráfico de streamfunction la escala de colores se usa al revés para que los azules (positivos) sean ciclónicos y los rojos anticiclónicos}
 
@@
 
-The connections between cEOFs and Tropical sources is also assessed.
-Figure \@ref(fig:sst-psi-2) shows the regression maps of Sea Surface Temperatures (SST) and Streamfunction at 200 hPa respectively upon cEOF2.
+The connections between cEOFs and Tropical sources are also assessed.
+Figure \@ref(fig:sst-psi-2) shows the regression maps of Sea Surface Temperatures (SST) and Streamfunction anomalies at 200 hPa respectively upon cEOF2.
 The Imaginary cEOF2 is associated with strong positive SST anomalies on the Central Pacific and negative anomalies over an area across the North of Australia and New Zealand, the South Pacific Convergence Zone (SPCZ) (Figure \@ref(fig:sst-psi-2).a2).
-This pattern is almost canonically positive ENSO [@bamston1997] and indeed, the correlation between the Imaginary cEOF2 and the Oceanic Niño Index [@bamston1997] is very high `r enso_cor[cEOF == "cEOF2" & part == "Imaginary"]$text`.
+This pattern is almost canonically positive ENSO [@bamston1997] and indeed, the correlation between the Imaginary cEOF2 and the Oceanic Niño Index [@bamston1997] is significant and very high `r enso_cor[cEOF == "cEOF2" & part == "Imaginary"]$text`.
 Besides, the Imaginary cEOF2 is associated with strong wave-like streamfunction anomalies emanating from the tropics (Figure \@ref(fig:sst-psi-2).b2).
-This is consistent with the effect of ENSO on the extratropics: SST anomalies initiate anomalous convection that excites Rossby waves that propagate meridionally towards higher latitudes [@mo2000].
-
-Since the Real cEOF2 represents just a different phase of the same wave train, it would be expected to show a similar forcing pattern to the Imaginary cEOF with a slight translation of its location.
+This is consistent with the effect of ENSO on the extratropics: SST anomalies initiate anomalous tropical convection that excites Rossby waves that propagate meridionally towards higher latitudes [@mo2000].
+
+Since the Real cEOF2 represents just a different phase of the same wave train, it would be expected to show a similar forcing pattern to the Imaginary cEOF2 with a slight translation of its location.
 However, Figure \@ref(fig:sst-psi-2).a1 and b1 show that the Real cEOF2 is not associated either with any significant SST nor streamfunction anomalies in the tropics.
 The correlation between the Real cEOF2 and ENSO is also not significant (`r enso_cor[cEOF == "cEOF2" & part == "Real"]$text`).
-This lack of tropical signal suggests different natures of the different phases of the Real cEOF2 wave train, which has the maximum amplitudes in the surroundings of southern South America.
+\textcolor{red}{Usaría notación científica por la cantidad de 0 que tiene}.
+This lack of tropical signal suggests different natures of the different phases of the Real cEOF2 wave train \textcolor{red}{¿Las diferentes fases del cEOF2 real?
+Me suena raro como quedo escrito.}
+, which has \textcolor{red}{"have" según como está escrito}
+ the maximum amplitudes in the surroundings of southern South America.
 
@@
 The black line in Figure \@ref(fig:enso-phase) is a sinusoidal fit of the relationship between ONI and cEOF2 phase.
-The $r^2$ corresponding to the fit is `r signif(summary(cos_model)$r.squared, 2)`, with p-value `r report_p(coef(summary(cos_model))[4])`, indicating a quasi-sinusoidal relation between these two variables.
-
-Furthermore, Figure \@ref(fig:enso-phase) suggest that higher amplitude cEOF2 years tend to coincide with stronger ENSO years.
+The $r^2$ corresponding to the fit is `r signif(summary(cos_model)$r.squared, 2)`, statistically significant with p-value `r report_p(coef(summary(cos_model))[4])`, indicating a quasi-sinusoidal relation between these two variables.
+
+\textcolor{red}{Fig. 12 Revisa cómo te quedo abajo de la figura la indicación de qué son cada uno de los puntos (te quedaron dos circulos y una línea extra que no va)}
+
+Furthermore, Figure \@ref(fig:enso-phase) suggest that highest amplitude cEOF2 years tend to coincide with strong ENSO years.
 The correlation between the absolute magnitude of the ONI and the magnitude of the cEOF2 is `r cor_enso_mag$text`.
@@
 Consistently, streamfunction anomalies do not show any tropical influence.
-Instead of that, the real and imaginary CEOF1 are associated with zonally wave activity fluxes in the extra-tropics around 60ºS, except for an equatorward flow from the coast of Antarctica around 150ºE in the real component.
+Instead of that, the real and imaginary cEOF1 are associated with zonally wave activity fluxes in the extra-tropics around 60ºS, except for an equatorward flow from the coast of Antarctica around 150ºE in the real component.
+\textcolor{red}{¿Esto sería como decir que este modo está más asociado a la variabilidad interna?}
+
+\textcolor{red}{Fig. 13 Tene en cuenta los mismos comentarios que en el figura 11 y revisa también la barra de colores para SST que quedo rara}
 
@@
 
+\textcolor{red}{Revisa el caption de la figura 14 que equedo un estilo distinto que los anteriores.
+Te falto aclarar también qué es la caja (¿la inventaste vos o habías utilizado la de algún otro trabajo?).
+Fijate también de aumentar la separación entre la columna izquierda y derecha para que no se superpongan los labels.
+También fijate que esta barra de colores no te quedo con el borde externo como el resto.
+}
+
 ```{r box_australia}
@@
 
-The precipitation anomalies in Oceania associated with cEOF1 are weak and not statistically significant (Figure \@ref(fig:pp-oceania) column a) whereas EFO2 is associated with large and widespread anomalies.
+The precipitation anomalies in Oceania associated with cEOF1 are weak and not statistically significant (Figure \@ref(fig:pp-oceania) column a) whereas cEOF2 is associated with large and widespread anomalies.
 The Real cEOF2 is associated with positive seasonal anomalies in Northern and Eastern Australia, and negative ones over New Zealand, while precipitation anomalies associated with Imaginary cEOF2 are negative over all Eastern Australia.
-Again, these negative anomalies are similar to the springtime precipitation anomalies observed in relation with the Asymmetric SAM [@campitelli2021] and ENSO [@cai2011].
+As it could be expected, these negative anomalies are similar to the springtime precipitation anomalies observed in relation with the Asymmetric SAM [@campitelli2021] and ENSO [@cai2011].
+\textcolor{red}{Acá te faltaría  sacar un poco más el jugo asociando con la figura 11. Además, ¿revisaste si podes también usar algún resultado asociado a los PSA? }
 
@@
 Black line is the fit precipitation ~ sin(phase) + cos(phase) computed by weighted OLS using the magnitude of the cEOF2 as weights.
+
+\textcolor{red}{Revisa lo mismo que para la otra figura del tipo. }
 
@@
 
-(ref:pp-america-cap) As Figure \@ref(fig:pp-america) but for South America
+(ref:pp-america-cap) As Figure \@ref(fig:pp-oceania) but for South America
 
@@
 Figure \@ref(fig:pp-america) shows regression maps of seasonal precipitation with each cEOF in South America.
-cEOF1 is associated with slight decreases in precipitation in Southern Brazil and Paraguay, although these are not statistically significant (Fig. \@ref(fig:pp-america) column a).
+cEOF1 is associated with negative regressions coefficients for precipitation in Southern Brazil and Paraguay, although these are not statistically significant (Fig. \@ref(fig:pp-america) column a).
 The strongest precipitation regression coefccients are the ones associated with the Imaginary cEOF2.
 The positive regression coefficients on Southeastern South America (SESA) and Central Chile, and negative ones over South Atlantic Convergence Zone is a well-known springtime precipitation signature of ENSO [@cai2020a] and it is also similar to the precipitation anomalies associated with the A-SAM [@campitelli2021].
+\textcolor{red}{Lo mismo que te puse antes. Te faltaría  sacar un poco más el jugo asociando con la figura 11. Además, ¿revisaste si podes también usar algún resultado asociado a los PSA? }
+
 This is not surprising considering the close relationship between the ONI, the A-SAM index and the Imaginary cEOF2 shown previously, but further consolidates the identification of this mode with the PSA pattern.
@@
 Resembling the relationship between ONI and the phase of cEOF2 (Fig. \@ref(fig:enso-phase)), there is a cEOF2 phase dependence of the precipitation anomalies in SESA (not shown).
-These variables could be related wth a significant sinusoidal fit that has a coefficient of determination of `r round(summary(model)$r.squared, 2)` (p-value `r report_p(p_value)`).
+These variables could be related wth a significant sinusoidal fit that has a statistically significant coefficient of determination of `r round(summary(model)$r.squared, 2)`.
 
@@
 
-In this study we derived two complex indices of the zonally asymmetric circulation of the September-October-November southern extratropics using Complex Empirical Orthogonal Functions. 
+In this study we assessed the Austral Spring Extratropical Southern Hemisphere zonally assymetric circulation.
+For this purpose, two complex indices were derived using Complex Empirical Orthogonal Functions.
 This allowed us to characterise both the amplitude and phase of planetary waves that are not perfectly organised as sinusoidal waves with constant amplitude at each latitude circle.
@@
 The first complex EOF represents the variability of the zonal wave 1 in the stratosphere, but represents a more zonally symmetric monopole in the troposphere.
-There is a statistically positive trend in the magnitude of this cEOF, which is consistent with previous studies which showed secular changes in springtime wave-1-like patterns [e.g. @raphael2003].
-This mode is closely related to anomalies in Total Ozone Column.
+There is a statistically positive trend in the magnitude of this cEOF, which is consistent with previous studies that showed secular changes in springtime wave-1-like patterns [e.g.
+@raphael2003].
+This mode is closely related to stratospheric variability such as anomalies in Total Ozone Column.
+Otherwise, this complex EOF is not related with SST variability and continental precipitation in the Southern Hemisphere.
 
@@
 We show that the Imaginary cEOF2 can be identified with the PSA1 and the Real cEOF2 with the PSA2.
-This cEOF prefers a phase aligned with the Imaginary cEOF2.
-This is because the when ENSO is active, it forces the cEOF2 to be in the Imaginary phase.
+This cEOF prefers a phase aligned with the Imaginary cEOF2, because when ENSO is active, it forces the cEOF2 to be in the Imaginary phase.
 When ENSO is neutral, the cEOF2 is still active, but with no particular preferred phase.
@@
 On the other hand, over Australia, both the Real and Imaginary phase are associated with significant precipitation anomalies, and we further show that the direction of maximum impact is not aligned with our chosen rotation of cEOF2.
-However, this underscores the benefit of using complex EOF, since it would trivial to rotate it.
-
-To the extent that the cEOF2 can be identified with the PSA, what is the advantage of our technique?
+However, this underscores the benefit of using complex EOF, since it would be trivial to rotate it. \textcolor{red}{creo que esta idea convendría desarrollarla mejor para el que no lo tenga tan claro}
+
+To the extent that the cEOF2 can be identified with the PSA, what is the advantage of the technique proposed here?
 The spatial fields that the cEOF2 index describes are in quadrature by construction, which makes it possible to derive a proper amplitude and phase.
 Furthermore, because of the way they are constructed, they are not forced to be orthogonal to the SAM.
-This makes it possible to study the relationship between the two modes free of the statistical contamination that comes with the usual definition of the PSA as the second and third EOFs.
+This makes it possible to study the relationship between the two modes \textcolor{red}{¿A qué te referis con los dos modos?
+puede ser confuso}
+ free of the statistical contamination that comes with the usual definition of the PSA as the second and third EOFs.
 

### 2021-08-06 analysis/paper/shceof.Rmd
Aplica sugerencias 1

@@
 
-We used monthly geopotential height, air temperature, ozone mixing ratio, and total ozone column (TOC) at 2.5\degree longitude by 2.5\degree latitude of horizontal resolution and 37 vertical isobaric levels from ERA5 reanalysis [@era5] for the period 1979-2019.
+We used monthly geopotential height, air temperature, ozone mixing ratio, and total ozone column (TOC) at 2.5\degree longitude by 2.5\degree latitude of horizontal resolution and 37 vertical isobaric levels from the European Centre for Medium-Range Weather Forecasts Reanalysis version 5 [ERA; @era5] for the period 1979-2019.
 Most of our analysis was restricted to the post-satellite era to avoid any confounding factors arising from the incorporation of satellite observations, but we also used the preliminary back extension of ERA5 from 1950 to 1978 [@era5be] to look at long-term trends.
@@
 
-The first cEOF (Fig. \@ref(fig:ceofs-1) column a) is a hemispheric wave 1 pattern with maximum amplitude at high latitudes.
+The cEOF1 (Fig. \@ref(fig:ceofs-1) column a) is a hemispheric wave 1 pattern with maximum amplitude at high latitudes.
 At 50 hPa the Real cEOF1 has the maximum of the wave 1 at 150ºE and at 200 hPa, the maximum is located at around 175ºE indicating a westerly shift in phase.
@@
 en caso de que si, creo que habría que mencionar algo sobre lo que pasa con el cEOF2 en la estratósfera}
-
-The temporal variability for both components of the cEOF1 have non zero mean (Fig. \@ref(fig:extended-series)). \textcolor{red}{Esto vale también para el cEOF2, ¿no? en ese caso, lo diría más en gral}
-This is due to the fact that the geopotential fields that enter into the cEOFs algorithm are anomalies with respect to the zonal mean, not the time mean.
-As a consequence, the variability associated with the cEOF1 includes variability that projects onto the mean zonally anomalous field.
+All cEOFs have non zero mean (Fig.
+\@ref(fig:extended-series)) because the geopotential fields that enter into the cEOFs algorithm are anomalies with respect to the zonal mean instead of the time mean.
+However, cEOF2's mean is almost zero, which indicates that only cEOF1 includes variability that significantly projects onto the mean zonally anomalous field.
 There are no significant simultaneous correlation between the time series.
@@
 In the stratosphere, the Real cEOF1 is correlated with both A-SAM and S-SAM, while the Imaginary cEOF1 is highly correlated only with the A-SAM.
-These correlations are consistent with the regression patterns of geopotential height in Figure \@ref(fig:eof1-regr-gh) and their comparison with those obtained for SAM, A-SAM and S-SAM by \textcolor{red}{Campitelli et al 2021}.
+These correlations are consistent with the regression patterns of geopotential height in Figure \@ref(fig:eof1-regr-gh) and their comparison with those obtained for SAM, A-SAM and S-SAM by @campitelli2021.
 
@@
 
-(ref:eof1-regr-t-cap) Same as Figure \@ref(fig:eof1-regr-gh) but for air temperature \textcolor{red}{Unidades}.
+(ref:eof1-regr-t-cap) Same as Figure \@ref(fig:eof1-regr-gh) but for air temperature (K).
 
@@
 There is a close spatial relationship between amplitudes and phases of the cEOF1 and TOC planetary wave 1 between `r knitr::combine_words(LatLabel(o3wave_lats))`.
-The correlation between the amplitude of both indices is `r wave1_cor[with == "amplitude"]$text`, while the overall correlation between the zonal wave 1 of ozone and cEOF1 (computed as the mean cosine of the difference in phase) is `r wave_correlation$text`.
-\textcolor{red}{¿Podrías clarificarlo?
-No estoy seguro de qué es lo que hiciste}
-As expected from the location of regressions in Figure \@ref(fig:o3-regr), the Real cEOF1 drives the relationship of amplitudes and the Imaginary cEOF1 drives the relationship of phases between the timeseries (not shown).
-\textcolor{red}{Tampoco me quedo claro qué estás diciendo con esto y cómo se explica.}
+The correlation between the amplitude of both indices is `r wave1_cor[with == "amplitude"]$text` and `r wave1_cor[with == "phase"]$text` between their phases.
 
@@
 \textcolor{red}{Usaría notación científica por la cantidad de 0 que tiene}.
-This lack of tropical signal suggests different natures of the different phases of the Real cEOF2 wave train \textcolor{red}{¿Las diferentes fases del cEOF2 real?
-Me suena raro como quedo escrito.}
-, which has \textcolor{red}{"have" según como está escrito}
- the maximum amplitudes in the surroundings of southern South America.
+This lack of tropical signal suggest that the Real cEOF2 and the Imaginary cEOF2 have very different sources of variability.
 
@@
 Consistently, streamfunction anomalies do not show any tropical influence.
-Instead of that, the real and imaginary cEOF1 are associated with zonally wave activity fluxes in the extra-tropics around 60ºS, except for an equatorward flow from the coast of Antarctica around 150ºE in the real component.
-\textcolor{red}{¿Esto sería como decir que este modo está más asociado a la variabilidad interna?}
+Instead, the real and imaginary cEOF1 are associated with zonally wave activity fluxes in the extra-tropics around 60ºS, except for an equatorward flow from the coast of Antarctica around 150ºE in the real component.
+This suggests that it's variability is driven primary by the internal variability of the extra-tropics.
 
@@
 Furthermore, because of the way they are constructed, they are not forced to be orthogonal to the SAM.
-This makes it possible to study the relationship between the two modes \textcolor{red}{¿A qué te referis con los dos modos?
-puede ser confuso}
- free of the statistical contamination that comes with the usual definition of the PSA as the second and third EOFs.
+This makes it possible to study their relationship with the SAM free of the statistical contamination that comes with the usual definition of the PSA as the second and third EOFs.
 

### 2021-08-06 analysis/paper/shceof.Rmd
Aplica sugerencias 2

@@
 
-Some of these patterns of variability associated to the zonally asymmetric circulation appear to have experienced secular changes.
-@raphael2003 found that the amplitude of the zonal wave 1 experienced a large increase and that the zonal wave 3 experienced changes in its annual cycle between 1958 and 1996.
-\textcolor{red}{¿esto es lo único que encontraste asociado a esta afirmación?
-en caso de que si, diría de agregar que "ha sido poco estudiado hasta el momento"}
+Some of these patterns of variability associated to the zonally asymmetric circulation appear to have experienced secular changes, although to our knowledge this has not been extensively studied.
+For instance, @raphael2003 found that the amplitude of the zonal wave 1 experienced a large increase and that the zonal wave 3 experienced changes in its annual cycle between 1958 and 1996.
 
@@
 
-Our objective is to create an index that is able to capture the phase-varying nature of planetary waves and is flexible enough to describe planetary waves with non-constant amplitude in each latitude circle.
-\textcolor{red}{yo no pondría que el objetivo es crear el índice, sino porfundizar el conocimiento de la variabilidad...
-utilizando un índice...}
+Our objective is to improve our understanding of planetary waves in the Southern Hemisphere using an index that is able to capture the phase-varying nature of planetary waves and is flexible enough to describe planetary waves with non-constant amplitude in each latitude circle.
 We use complex Empirical Orthogonal Functions [@horel1984] as a robust extension of EOF analysis that is capable of describing rich structures with meridional and zonal propagation, and amplitude modulation such as the PSA.
@@
 These indices were obtained by projecting the zonally asymmetric and zonally symmetric part of the SAM spatial pattern onto monthly geopotential height fields, as proposed by @campitelli2021.
-The Pacific South American patterns (PSA1 and PSA2) were calculated following @mo2001 as the third and fourth leading EOF of seasonal mean anomalies for SH 500 hPa geopotential height combining all seasons \textcolor{red}{leyendo esto me queda la duda.
-Vos calculaste esto usando todos los meses y luego tomaste el valor de SON, no?
-Creo que valdría la pena aclararlo mejor, por lo menos a mi me resulta confuso}.
+Seasonal indices of the Pacific South American patterns (PSA1 and PSA2) were calculated following @mo2001 as the third and fourth leading EOF of seasonal mean anomalies for SH 500 hPa geopotential height combining all seasons.
+In our study, only the SON values were used.
 
@@
 This modulated amplitude is especially apparent at 200 hPa.
-There is no apparent phase shift, suggesting a rather barotropic structure.
+There is no apparent phase shift but the amplitude of the pattern is greatly reduced in the stratosphere, suggesting that this barotropic mode represents mainly tropospheric variability.
 
@@
 This long-term change indicates an increase in the magnitude of the high latitude zonal wave 1.
-A similar observation was made by @raphael2003, who detected a step after around 1975 in the leading EOF of August-September-October 500 hPa zonal geopotential height anomalies.
-\textcolor{red}{Creo que acá faltaría decir que el modo del trabajo de Raphael se parece al que estás mostrando vos, sino no tendria mucho sentido la comapración.}
+A similar observation was made by @raphael2003, who detected a step after around 1975 in the leading EOF of August-September-October 500 hPa zonal geopotential height anomalies, which is similar to the cEOF1.
 
@@
 
-The relationship between PSAs and the modes described from the cEOFs is also studied.
+Due to the similarity between the PSA patterns and the cEOF2 spatial patterns (Fig. \@ref(fig:eof2-regr-gh)), we study the relationship between these modes. 
 Table \@ref(tab:psa-eof2) shows the correlations between the two PSA indices and the timeseries for Real and Imaginary phase of cEOF2.
-\textcolor{red}{Valdría la pena aclarar por qué no consideras el cEOF1 para esta discusión}
 As anticipated by Figure \@ref(fig:eof2-regr-gh), there is a strong correlation between PSA1 and Imaginary cEOF2, and between PSA2 and Real cEOF2.
@@
 About two thirds of time cEOF2 has a phase similar to either the negative or positive Imaginary phase.
-It could be seen that Imaginary phase is the most common phase and It is also the direction that has the maximum relationship with ENSO.
-\textcolor{red}{Ojo que todavía no estuvo la discusión con lo del ENSO, así que esto lo deberías decir más adelante.}
+It could be seen that Imaginary phase is the most common phase and It is also, by construction, the direction that has the maximum relationship with ENSO. 
 Therefore, the Imaginary cEOF2 explains more variance than the Real cEOF2 and conventional EOF analysis will tend to separate the two.

### 2021-08-06 analysis/paper/shceof.Rmd
Apply suggestions from code review

@@
 
-## cEOFs relationship with other variables {#other-variables}
-
-\textcolor{red}{Este título queda raro porque después hablas de modos de variabilidad climática y no de lo que llamaríamos variables}
-
+## cEOFs relationship with known modes of variability {#other-variables}
 ### SAM
@@
 Areas marked with dots have p-values smaller than 0.05 adjusted for False Detection Rate.
+In black is the West Australia region, defined as the box between `r knitr::combine_words(LonLabel(box_australia$lon))` and `r knitr::combine_words(LatLabel(box_australia$lat))`.
 

### 2021-08-11 analysis/paper/shceof.Rmd
Cambia coord_polar

@@
 geom_coords <- function() {
+
   list(
     no_grid,
+    annotate("segment",
+             y = seq(-90, 0, by = 15),
+             yend = seq(-90, 0, by = 15),
+             x = 0,
+             xend = 360,
+             size = 0.1, alpha = 0.5),
+
     annotate("segment",
@@
              size = 0.1, alpha = 0.5),
-    geom_hline(yintercept = seq(-90, 0, by = 15), size = 0.1, alpha = 0.5),
     shadowtext::geom_shadowtext(data = data.frame(x = 0, y = seq(-90 + 15, 0, 
@@
 
+  )
+}
+
+coord_polar <- function(ymax = -20) {
+
+  x <- c(seq(0, 360, length.out = 40), 
+         seq(360, 0, length.out = 40), 
+         0)
+  y <- c(rep(ymax, length.out = 40), 
+         rep(60, length.out = 40), 
+         ymax)
+
+  cbind(x, y) |> 
+    list() |> 
+    sf::st_polygon() |> 
+    sf::st_sfc(crs = "+proj=latlong") -> white
+
+  list(
+    geom_sf(data = white, inherit.aes = FALSE, 
+            fill = "white", 
+            colour = "white", size = 2),
+    coord_sf(ylim = c(-90, ymax), 
+             lims_method = "box",
+             crs = "+proj=laea +lat_0=-90",
+             default_crs = "+proj=longlat",
+             label_axes =  "----")
   )
@@
 In Section \@ref(methods) we describe the methods.
-In Section \@ref(temporal) we describe the temporal variability and vertical coherence of the indices.
 In Section \@ref(spatial), we analyse the spatial patterns of each complex EOF.
@@
 
-  geom_qmap(~.x[lat <= -20], size = 0.1) +
+  geom_qmap(size = 0.1) +
   scale_fill_divergent_discretised(NULL, guide = guide_colorsteps_bottom(10))  +
@@
 We downloaded data from reanalysis using the ecmwfr package [@hufkens2020] and indices of the ENSO with the rsoi package [@albers2020].
-The paper was rendered using knitr and rmarkdown [@xie2015; @allaire2019].
+The paper was rendered using knitr and rmarkdown [@xie2015; @allaire2020].
 
@@
   scale_fill_divergent_discretised(guide = guide_colorsteps_bottom()) +
-  geom_qmap(~ .x[lat <= -20]) +
+  geom_qmap() +
   scale_x_longitude(labels = NULL) +
   scale_y_latitude(labels = NULL,
-                   limits = c(NA, -20),
                    minor_breaks = NULL) +
-  scale_linetype_manual(values = c("1" = 1, "-1" = 2), labels = c("1" = "+", "-1" = "-"),
+  scale_linetype_manual(values = c("1" = 1, "-1" = 2),
+                        labels = c("1" = "+", "-1" = "-"),
                         guide = "none") +
@@
                                   cEOF = var)) +
-  coord_quickmap(ylim = c(NA, -20)) +
   theme(legend.title = element_blank()) +
+  geom_coords() +
   coord_polar() +
-  geom_coords() +
+
   tag_facets(tag = "cr")
@@
 
-    geom_qmap(~ .x[lat <= -20]) +
+    geom_qmap() +
 
@@
     scale_y_latitude(labels = NULL,
-                     limits = c(NA, -20),
                      minor_breaks = NULL) +
@@
     facet +
+    geom_coords() +
     coord_polar() +
-    geom_coords() +
     tag_facets("rc")
@@
   scale_fill_divergent(guide = "none") +
-  geom_qmap(~.x[lat <= -20]) +
+  geom_qmap() +
   scale_x_longitude() +
   scale_y_latitude() +
+  geom_coords() +
   coord_polar() +
-  geom_coords() +
+
   theme(axis.text = element_blank())
@@
                                    guide = guide_colorsteps_bottom(15)) + 
-  scale_y_level() +
+  scale_y_level(minor_breaks= NULL, 
+                sec.axis = sa_height_axis(breaks = seq(0, 50, by = 10),
+                                          labels = function(x) paste0("~", x, " km"))) +
   scale_linetype(guide = "none") +
@@
   geom_contour_pval(aes(z = p.value)) +
-  geom_hline(yintercept = o3wave_lats, size = 0.3, alpha = 0.8) +
-  geom_qmap(~.x[lat <= -20]) +
+  annotate("segment", 
+           y  = o3wave_lats,
+           yend = o3wave_lats,
+           x = 0,
+           xend = 360,
+           size = 0.3, alpha = 0.8
+  ) +
+  geom_qmap() +
   scale_fill_divergent_discretised("Total Ozone Column", 
@@
   scale_y_latitude(labels = NULL,
-                   limits = c(NA, -20),
                    minor_breaks = NULL) +
   scale_linetype(guide = "none") +
+  geom_coords() +
   coord_polar() +
-  geom_coords() +
   facet_grid(cEOF~term) +
@@
     facet_grid(term ~ var) +
-    coord_quickmap(ylim = c(NA, 10)) +
+    # coord_quickmap(ylim = c(NA, 10)) +
+    coord_sf(ylim = c(-90, 0),
+             crs = "+proj=longlat +pm=180 +over",
+             default_crs = "+proj=longlat +over") + 
   theme(legend.box = "vertical") +
@@
 }
+
 ```
@@
     # .[abs(estimate) <= 2] %>%
-    .[, p.value := Pvaluate(estimate, std.error, df, "none"), by = .(lev, cEOF, term)] %>%
+    .[, p.value := Pvaluate(estimate, std.error, df, "none"), 
+      by = .(lev, cEOF, term)] %>%
     periodic(lon = c(0, 360)) %>%
@@
     geom_contour_pval(aes(z = p.value)) +
-    geom_qmap(~.x %>% .[lat %between% c(-60, max(data$lat))] %>% 
-                add_continent() %>% filter_continent(),
-              keep = 1) +
+    geom_qmap(keep = 1) +
 
@@
     tag_facets("cr") +
-    coord_quickmap()
-}
-```
-
-```{r sesa}
+      coord_sf(ylim = continents[[which_continent]]$lat,
+               xlim = continents[[which_continent]]$lon,
+               crs = "+proj=longlat +pm=180 +over",
+               default_crs = "+proj=longlat +over")
+
+}
+```
+
+```{r boxes}
 sesa <- list(lon = c(300, 315),
              lat = c(-35, -22))
+box_australia <- data.frame(cEOF = "cEOF2", 
+                            lon = c(130, 155),
+                            lat = c(-40, -15))
 ```
@@
 }
-
-```{r box_australia}
-box_australia <- data.frame(cEOF = "cEOF2", 
-                            lon = c(130, 155),
-                            lat = c(-40, -15))
-```
 

### 2021-08-11 analysis/paper/shceof.tex
Cambia coord_polar

@@
 
-\title{Austral Spring Extratropical Southern Hemisphere zonally assymetric circulation \thanks{The research was supported by UBACyT20020170100428BA and the CLIMAX Project funded by Belmont Forum/ANR-15-JCL/-0002-01.
+\title{Revisiting the Austral Spring Extratropical Southern Hemisphere zonally assymetric circulation using complex Empirical Orthogonal Functions \thanks{The research was supported by UBACyT20020170100428BA and the CLIMAX Project funded by Belmont Forum/ANR-15-JCL/-0002-01.
 Elio Campitelli was supported by a PhD grant from CONICET, Argentina.} }
@@
 
-For example, \citet{raphael2007} suggests that variability in the planetary wave 3 projected onto its climatological location is associated with Antarctic sea ice concentration anomalies.
+One of the main features of the zonally asymmetric circulation are the Pacific--South American Patterns \citep{mo2001}, which are associated with precipitation anomalies in South America.
+Besides, \citet{raphael2007} suggests that variability in the planetary wave 3 projected onto its climatological location is associated with Antarctic sea ice concentration anomalies.
 \citet{irving2015} showed that planetary wave amplitude in general is associated sea ice concentration, as well as temperature and precipitation anomalies.
-Another characteristic of the zonally asymmetric circulation are the Pacific--South American Patterns \citep{mo2001}, which are associated with precipitation anomalies in South America.
 
-Some of these patterns of variability appear to have experienced secular changes.
-\citet{raphael2003} found that the amplitude of the zonal wave 1 experienced a large increase and that the zonal wave 3 experienced changes in its annual cycle between 1958 and 1996.
+Some of these patterns of variability associated to the zonally asymmetric circulation appear to have experienced secular changes, although to our knowledge this has not been extensively studied.
+For instance, \citet{raphael2003} found that the amplitude of the zonal wave 1 experienced a large increase and that the zonal wave 3 experienced changes in its annual cycle between 1958 and 1996.
 
@@
 The inability of a single EOF to describe travelling waves is what forces the PSA pattern to be described by two EOFs.
-\citet{irving2016} characterises the Pacific--South American Pattern in a Fourier framework.
-Being a meridionally propagating mode, the PSA cannot be correctly characterised by Fourier decomposition at each latitude circle, so they reprojected meridional wind fields so that the path of the PSA laid on the equator.
+\citet{irving2016} characterises this pattern in a Fourier framework.
+Being a meridionally propagating mode, the PSA cannot be correctly characterised by Fourier decomposition at each latitude circle.
+As a result, they reprojected meridional wind fields so that the path of the PSA laid on the equator.
 On this new projection, they could identify the PSA using Fourier decomposition.
@@
 
-Our objective is to create an index that is able to capture the phase-varying nature of planetary waves and is flexible enough to describe planetary waves with non-constant amplitude in each latitude circle.
+Our objective is to improve our understanding of planetary waves in the Southern Hemisphere using an index that is able to capture the phase-varying nature of planetary waves and is flexible enough to describe planetary waves with non-constant amplitude in each latitude circle.
 We use complex Empirical Orthogonal Functions \citep{horel1984} as a robust extension of EOF analysis that is capable of describing rich structures with meridional and zonal propagation, and amplitude modulation such as the PSA.
@@
 We focus the analysis to the September-October-November because that's the trimester in which tropical forcing of zonal waves is maximised (cita).
+\textcolor{red}{Este parráfo me parece que quedo un poco colgado.
+Creo que sería importante que esté desde más arriba la idea de la primavera, ya que todo lo otro es muy general, pero al final el análisis solo se va a restringir a esta estación}
 
 In Section \ref{methods} we describe the methods.
-In Section \ref{temporal} we describe the temporal variability and vertical coherence of the indices.
 In Section \ref{spatial}, we analyse the spatial patterns of each complex EOF.
@@
 
-We used monthly geopotential height, air temperature, ozone mixing ratio, and Total Ozone Column (TOC) at 2.5\degree~longitude by 2.5\degree~latitude of horizontal resolution and 37 vertical isobaric levels from ERA5 \citep{era5} for the period 1979 to 2019.
+We used monthly geopotential height, air temperature, ozone mixing ratio, and total ozone column (TOC) at 2.5\degree~longitude by 2.5\degree~latitude of horizontal resolution and 37 vertical isobaric levels from the European Centre for Medium-Range Weather Forecasts Reanalysis version 5 {[}ERA; \citet{era5}{]} for the period 1979-2019.
 Most of our analysis was restricted to the post-satellite era to avoid any confounding factors arising from the incorporation of satellite observations, but we also used the preliminary back extension of ERA5 from 1950 to 1978 \citep{era5be} to look at long-term trends.
 Streamfunction was derived from ERA5 vorticity at 200 hPa using the FORTRAN subroutine FISHPACK \citep{fishpack} and horizontal wave activity fluxes was computed following \citet{plumb1985}.
-We usded Sea Surface Temperature (SST) monthly fields from Extended Reconstructed Sea Surface Temperature (ERSST) v5 \citep{huang2017} and precipitation monthly data from the CPC Merged Analysis of Precipitation \citep{cmap}, with a 2.5\degree~resolution in latitude and longitude.
+We used Sea Surface Temperature (SST) monthly fields from Extended Reconstructed Sea Surface Temperature (ERSST) v5 \citep{huang2017} and precipitation monthly data from the CPC Merged Analysis of Precipitation \citep{cmap}, with a 2.5\degree~resolution in latitude and longitude.
 This rainfall gridded dataset is based on information from different sources such as rain gauge observations, satellite inferred estimations and the NCEP-NCAR reanalysis, and it is available since 1979 to present.
@@
 
-The amplitude of the zonal waves was defined through computing the Fourier transform of the spatial field at each latitude circle.
-We computed the amplitude and phase of the zonal wave 1 by averaging (area-weighted) variables (temperature, geopotential height, ozone mixing ratio and Total Ozone Column) between 75°S and 45°S for each SON and extracting the wave-1 component of the Fourier spectrum.
-We chose this latitude band because it is wide enough to capture most of the relevant anomalies.
+The amplitude of the zonal waves was obtained through the Fourier transform of the spatial field at each latitude circle.
+We computed the amplitude and phase of the zonal wave 1 by averaging (area-weighted) each variable between 75°S and 45°S for each SON and extracting the wave-1 component of the Fourier spectrum.
+We chose this latitude band because it is wide enough to capture most of the relevant anomalies for southern hemisphere mid-latitudes.
 
-We computed the level-dependent Southern Annular Mode (SAM) index as the leading EOF of year-round monthly geopotential height anomalies south of 20ºS at each level for the whole period \citep{baldwin2009}.
+We computed the level-dependent SAM index as the leading EOF of year-round monthly geopotential height anomalies south of 20ºS at each level for the whole period according to \citet{baldwin2009}.
 We further split the SAM into its zonally symmetric and zonally asymmetric components (S-SAM and A-SAM indices respectively).
 These indices were obtained by projecting the zonally asymmetric and zonally symmetric part of the SAM spatial pattern onto monthly geopotential height fields, as proposed by \citet{campitelli2021}.
-The Pacific South American patterns (PSA1 and PSA2) were calculated following \citet{mo2001} as the third and fourth leading EOF of seasonal mean anomalies for SH 500 hPa geopotential height combining all seasons.
+Seasonal indices of the Pacific South American patterns (PSA1 and PSA2) were calculated following \citet{mo2001} as the third and fourth leading EOF of seasonal mean anomalies for SH 500 hPa geopotential height combining all seasons.
+In our study, only the SON values were used.
 
-Linear trends were computed by Ordinary Least Squares and the 95\% confidence interval was computed assuming a t-distribution with the appropriate residual degrees of freedom \citep{wilks2011r}.
+Linear trends were computed by Ordinary Least Squares (OLS) and the 95\% confidence interval was computed assuming a t-distribution with the appropriate residual degrees of freedom \citep{wilks2011r}.
 
@@
 \subsection{Complex Empirical Orthogonal Functions (cEOF)}\label{complex-empirical-orthogonal-functions-ceof}}
+
+\textcolor{red}{En la figura 1 te quedaron muy chicos los subplots, amplialos un poco}.
 
@@
 The magnitude and argument of each complex time series represent the amplitude and phase of each zonal wave.
+
+\textcolor{red}{Los colores de la Tabla 1 no dicen mucho en la medida que no se conoce cuál sería la paleta. Quizás se pueden remarcar en negrita ambos aclarando que son significativos (lo son?)}.
 
@@
 Since any choice is arbitrary and equally valid, we chose the phase of each cEOF so that the real and imaginary parts are aligned with meaningful phases in our analysis.
-This procedure does not create a spurious correlation, it only takes whatever relationship that already exist and aligns it with a specific phase.
-
+This procedure does not create a spurious correlation, it only takes a known relationship and aligns it with a specific phase.
 For the first complex principal component, the phase was chosen so that the time series corresponding to the real part has the maximum correlation with the zonal wave 1 of Total Ozone Column between 75°S and 45°S.
 This also nearly minimises the correlation with the imaginary part.
-
 For the second complex principal component, the phase was chosen so that the coefficient of determination between the Oceanic Niño Index \citep{bamston1997} and the real part was minimized, which also nearly maximizes the correlation with the imaginary part.
@@
 We performed linear regressions to quantify the association between the cEOFs and other variables.
-For each cEOF, we create regression maps by computing fitting a multiple linear model involving both the real and the imaginary part.
+For each cEOF, we create regression maps by fitting a multiple linear model involving both the real and the imaginary part.
 To obtain the linear coefficients of a variable \(X\) (geopotential, temperature, precipitation, etc\ldots) with the Imaginary and Real parts of each cEOF we fit the equation
@@
 
-where \(\lambda\) and \(\phi\) are the longitude and latitude, \(t\) is the time, \(\alpha\) and \(\beta\) are the linear regression coefficients, \(X_0\) and \(\epsilon\) are the constant and error terms respectively.
+where \(\lambda\) and \(\phi\) are the longitude and latitude, \(t\) is the time, \(\alpha\) and \(\beta\) are the linear regression coefficients for Imaginary and Real parts respectively, \(X_0\) and \(\epsilon\) are the constant and error terms respectively.
 
@@
 We downloaded data from reanalysis using the ecmwfr package \citep{hufkens2020} and indices of the ENSO with the rsoi package \citep{albers2020}.
-The paper was rendered using knitr and rmarkdown \citep{xie2015, allaire2019}.
+The paper was rendered using knitr and rmarkdown \citep{xie2015, allaire2020}.
 
@@
 
-To describe the variability of the circulation zonal anomalies, the spatial and temporal parts of the first two leading cEOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa, computed jointly at both levels, are shown in Figures \ref{fig:ceofs-1} and @ref(fig:extended-series,).
+To describe the variability of the circulation zonal anomalies, the spatial and temporal parts of the first two leading cEOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa, computed jointly at both levels, are shown in Figures \ref{fig:ceofs-1} and \ref{fig:extended-series}.
 The first mode (cEOF1) explains 82\% of the variance, while the second mode (cEOF2) explains a smaller fraction (7\%).
@@
 
-The first cEOF (Fig.~\ref{fig:eof-naive} column~a) is a hemispheric wave 1 pattern with maximum amplitude at high latitudes.
+\textcolor{red}{tengo la impresión de que en la figura 2 podría estar un poco más grandes los subplots para que se vean más fácilmente los detalles}
+
+The cEOF1 (Fig.~\ref{fig:ceofs-1} column~a) is a hemispheric wave 1 pattern with maximum amplitude at high latitudes.
 At 50 hPa the Real cEOF1 has the maximum of the wave 1 at 150ºE and at 200 hPa, the maximum is located at around 175ºE indicating a westerly shift in phase.
-The second cEOF (Fig.~\ref{fig:eof-naive} column~b) shows also a zonal wave-like structure with maximum amplitude at high latitudes, but with shorter spatial scales.
+The cEOF2 (Fig.~\ref{fig:eof-naive} column~b) shows also a zonal wave-like structure with maximum amplitude at high latitudes, but with shorter spatial scales.
 In particular, its dominant structure at both levels is a wave 3 but with larger amplitude in the pacific sector.
 This modulated amplitude is especially apparent at 200 hPa.
-There is no apparent phase shift, suggesting a rather barotropic structure.
+There is no apparent phase shift but the amplitude of the pattern is greatly reduced in the stratosphere, suggesting that this barotropic mode represents mainly tropospheric variability.
 
-The temporal variability for both components of the cEOF1 have non zero mean (Fig.~@ref(fig:extended-series,)).
-This is due to the fact that the geopotential fields that enter into the cEOFs algorithm are anomalies with respect to the zonal mean, not the time mean.
-As a consequence, the variability associated with the cEOF1 includes variability that projects onto the mean zonally anomalous field.
+\textcolor{red}{se me vino una duda, ¿se pueden comparar de algún modo las amplitudes que mostras en esta figura?
+en caso de que si, creo que habría que mencionar algo sobre lo que pasa con el cEOF2 en la estratósfera}
+All cEOFs have non zero mean (Fig. \ref{fig:extended-series}) because the geopotential fields that enter into the cEOFs algorithm are anomalies with respect to the zonal mean instead of the time mean.
+However, cEOF2's mean is almost zero, which indicates that only cEOF1 includes variability that significantly projects onto the mean zonally anomalous field.
 There are no significant simultaneous correlation between the time series.
@@
 
-There is a significant upward trend in the real component of cEOF1 (Fig.~\ref{fig:extended-series}a.1, p-value~\textless{} 0.001) and no significant trend in any of the complex components of cEOF2.
+\textcolor{red}{En la figura 3 agregaría dos cosas: i)el eje y en la columna derecha (si la queres ver más grande perdes la referencia) ii) creo que serviría agregar una líneas verticales en la grilla para poder tener una idea un poco mejor de a qué años corresponde más o menos cada valor.
+pueden ser un poco más finas que las que ya están, pero creo que sería una referencia util para visualmente poder decir a qué año corresponde cada punto.
+Además, fijate que la figura te quedo de un ancho mayor que el de la página y te quedo un "EOF" en el eje-y a la izquierda que no creo que sea lo más correcto.}
+
+There is only a significant upward trend in the real component of cEOF1 (Fig.~\ref{fig:extended-series}a.1, p-value~\textless{} 0.001) and no significant trend in any of the complex components of cEOF2.
 The positive trend in the real cEOF1 translates into a positive trend in cEOF1 magnitude, but not in the phase (not shown).
-
 This long-term change indicates an increase in the magnitude of the high latitude zonal wave 1.
-A similar observation was made by \citet{raphael2003}, who detected a step after around 1975 in the leading EOF of August-September-October 500 hPa zonal geopotential height anomalies.
+A similar observation was made by \citet{raphael2003}, who detected a step after around 1975 in the leading EOF of August-September-October 500 hPa zonal geopotential height anomalies, which is similar to the cEOF1.
 
@@
 Figure~\ref{fig:eof1-regr-gh} shows regression maps of SON geopotential height anomalies upon cEOF1.
-At 50 hPa (Figure~\ref{fig:eof1-regr-gh} row a), both the Real and Imaginary cEOF1 are associated with planetary wave 1 patterns, that are 90º off phase.
+At 50 hPa (Figure~\ref{fig:eof1-regr-gh} row a), both the Real and Imaginary cEOF1 are associated with planetary wave 1 patterns, that are 90º out of phase.
 Their phases coincide with the ones shown in Figure~\ref{fig:ceofs-1}a.1, with the positive centre of the real cEOF1 located towards the dateline, and the one of the Imaginary cEOF1 located over Eastern Antarctica.
@@
 
-The regression patterns at 200 hPa (Figure~\ref{fig:eof1-regr-gh} row b) are similarly influenced by the zonally symmetric circulation.
-It is only possible to distinguish partially the wave 1 pattern in relation with the real cEOF1 (Figure~\ref{fig:eof1-regr-gh}b.1).
+\textcolor{red}{La figura 4 la veo muy saturada con los puntos que indican la significancia.
+¿se podrán hacer con un tamaño menor?
+¿Las unidades son correctas?
+Me llama la atención que te puedan dar tantos cientos de metros de altura geopotencial.}
+
+The regression patterns for cEOF1 at 200 hPa (Figure~\ref{fig:eof1-regr-gh} row b) are similarly influenced by the zonally symmetric circulation.
+It is only possible to partially distinguish the wave 1 pattern in relation with the real cEOF1 (Figure~\ref{fig:eof1-regr-gh}b.1).
 The imaginary cEOF1 shows a much more zonally symmetrical pattern resembling the negative SAM phase.
@@
 Zonal wave 3 dominates all fields, but only in the western hemisphere, over the Pacific and Atlantic Oceans.
-cEOF2 then represents a equivalent barotropic wave train that is very similar to the the Pacific South American Patterns \citep{mo2001}.
+cEOF2 then represents an equivalent barotropic wave train that is very similar to the the Pacific South American Patterns \citep{mo2001}.
 Comparing the location of the positive anomaly near 90ºW in column b of Figure~\ref{fig:eof2-regr-gh} with Figures 1.a and b from \citet{mo2001}, the real cEOF2 can roughly be identified with PSA2, while the imaginary cEOF2 resembles PSA1.
 
+\textcolor{red}{Figura 5. Me resulto raro que vaya hasta 999 en vez de 1000 }
+
 \hypertarget{other-variables}{%
-\subsection{cEOFs relationship with other variables}\label{other-variables}}
+\subsection{cEOFs relationship with known modes of variability}\label{other-variables}}
 
@@
 To explore the relationship between SAM and the modes described from the cEOF, we compute the coefficient of determination between the cEOFs timeseries and the three SAM indices (SAM, A-SAM and S-SAM) at each vertical level (Fig.~\ref{fig:sam-eof-vertical}).
-The SAM index is moderately correlated with the Real cEOF1 in all levels, and with the Imaginary cEOF1 and Imaginary cEOF2 in the troposphere.
-It is not correlated with the Real cEOF2 at any level.
+The SAM index is statistically signficant correlated with the Real cEOF1 in all levels, and with the Imaginary cEOF1 and Imaginary cEOF2 in the troposphere.
+Correlations between SAM and the Real cEOF2 are non-significant.
+
+\textcolor{red}{Fig.
+6.
+Te quedan muy pegados los subplots de la columna izquierda con los de la derecha.
+Recomendaría aumentar el espacio intermedio y repetir el eje y en las figuras de la derecha.
+Además podrías agregar el y-label (Level) a la izquierda.}
 
@@
 In the stratosphere, the Real cEOF1 is correlated with both A-SAM and S-SAM, while the Imaginary cEOF1 is highly correlated only with the A-SAM.
-These correlations are consistent with the regression patterns of geopotential height in Figure~\ref{fig:eof1-regr-gh}.
+These correlations are consistent with the regression patterns of geopotential height in Figure~\ref{fig:eof1-regr-gh} and their comparison with those obtained for SAM, A-SAM and S-SAM by \citet{campitelli2021}.
 
 In the case of cEOF2, the moderate correlation between SAM and Imaginary cEOF2 becomes extremely high only when the zonally asymmetric variability of the SAM is considered.
-The Imaginary cEOF2 and the A-SAM share up to 92\% of their variance, reached at 225 hPa (Figure~\ref{fig:sam-eof-vertical}.b2).
+The Imaginary cEOF2 explains up to 92\% of A-SAM variance, reached the maximum value at 225 hPa (Figure~\ref{fig:sam-eof-vertical}.b2).
 Such extremely high correlation between the asymmetric SAM and the Imaginary cEOF2 suggests that these might be different ways of characterising the same phenomenon.
@@
 \subsubsection{PSA}\label{psa}}
+
+\textcolor{red}{Tabla 2.
+Mismo comentario que antes con lo del background color (no dice mucho si no se conoce la escala), quizás podes destacar en negritas según la significancia.
+¿Está bien el intervalo de confianza de PSA2 vs cEOF2 Im?
+Hay algo raro ahí}
 
@@
 
-The relationship between PSAs and the modes described from the cEOFs is also studied.
-Table~\ref{tab:psa-eof2} shows the correlations between the two PSA indices and the timeseries for Real and Imaginary phase of cEOF2.
+Due to the similarity between the PSA patterns and the cEOF2 spatial patterns (Fig. \ref{fig:eof2-regr-gh}), we study the relationship between these modes.
+Table \ref{tab:psa-eof2} shows the correlations between the two PSA indices and the timeseries for Real and Imaginary phase of cEOF2.
 As anticipated by Figure~\ref{fig:eof2-regr-gh}, there is a strong correlation between PSA1 and Imaginary cEOF2, and between PSA2 and Real cEOF2.
@@
 About two thirds of time cEOF2 has a phase similar to either the negative or positive Imaginary phase.
-It could be seen that Imaginary phase is the most common phase and It is also the direction that has the maximum relationship with ENSO.
+It could be seen that Imaginary phase is the most common phase and It is also, by construction, the direction that has the maximum relationship with ENSO.
 Therefore, the Imaginary cEOF2 explains more variance than the Real cEOF2 and conventional EOF analysis will tend to separate the two.
@@
 The result obtained here is similar to the one obtained by \citet{irving2016}.
-They show that the phase distribution of the PSA-like variability they detected is bimodal and isolate the PSA pattern from the rest of the PSA-like variability by selecting events that are near the peaks of the distribution (compare our Figure~\ref{fig:phase-histogram} with their Figure 6).
+They show that the phase distribution of the PSA-like variability obtained by them is bimodal and isolate the PSA pattern from the rest of the PSA-like variability by selecting events that are near the peaks of the distribution (compare our Figure~\ref{fig:phase-histogram} with their Figure 6).
 
-The advantage of our method is that it is much simpler to implement, it provides magnitude and phase naturally, and it facilitates the description of this mode as a propagating wave instead of as standing oscillation.
+The advantage of our method is that it is much simpler to implement, it provides magnitude and phase naturally, and it facilitates the description of this mode as a propagating wave instead of as standing oscillations.
 As a consequence, the cEOF2 offers an alternative way of representing the PSA which has several advantages over using the second and third EOFs.
@@
 \includegraphics{../figures/eof1-regr-t-1.pdf}
-\caption{\label{fig:eof1-regr-t}Same as Figure~\ref{fig:eof1-regr-gh} but for air temperature.}
+\caption{\label{fig:eof1-regr-t}Same as Figure~\ref{fig:eof1-regr-gh} but for air temperature (K).}
 \end{figure}
@@
 
-The relation between cEOFs and temperature was also evaluated.
+The relation between cEOFs and air temperature was also evaluated.
 Figure~\ref{fig:eof1-regr-t} shows regression patterns of air temperature at 50hPa and 200hPa onto cEOF1.
-In both levels, the Real cEOF1 is associated with positive monopole over the South Pole with its centre moved slightly towards 150ºE (Fig.~\ref{fig:eof1-regr-t} column 1).
+In both levels, the Real cEOF1 is associated with a positive monopole over the South Pole with its centre moved slightly towards 150ºE (Fig.~\ref{fig:eof1-regr-t} column 1).
 On the other hand, the regression maps on the Imaginary cEOF1 show a more clear wave 1 pattern with its maximum around 60ºE.
@@
 
-The maximum ozone anomalies are co-located with the minimum temperature anomalies above 10 hPa and with the maximum temperature anomalies below 10 hPa (Fig.~\ref{fig:t-vertical})).
-Therefore, the ozone zonal wave 1 is anticorrelated with the temperature zonal wave 1 in the upper stratosphere, and directly correlated in the upper stratosphere .
+The maximum ozone anomalies are co-located with the minimum temperature anomalies above 10 hPa and with the maximum temperature anomalies below 10 hPa (Fig.~\ref{fig:t-vertical}).
+Therefore, the ozone zonal wave 1 is anticorrelated with the temperature zonal wave 1 in the upper stratosphere, and directly correlated in the upper stratosphere.
 This change in phase is observed in ozone anomalies forced by planetary waves that reach the stratosphere.
-In the photochemically-dominated upper stratosphere, cold temperatures inhibit the destruction of ozone, and the advectively-dominated lower stratosphere, ozone anomalies are 90º out of phase with horizontal and vertical transport, which is 90º our of phase with temperature anomalies \citep{hartmann1979, wirth1993, smith1995}.
+In the photochemically-dominated upper stratosphere, cold temperatures inhibit the destruction of ozone.
+On the other hand, in the advectively-dominated lower stratosphere, ozone anomalies are 90º out of phase with horizontal and vertical transport, which are in adittion 90º out of phase with temperature anomalies \citep{hartmann1979, wirth1993, smith1995}.
+
+\textcolor{red}{Revisa que tenes una incosistencia en el uso de mayúsculas en Real/Imaginary porque a veces las usas y a veces no.}
 
@@
 \includegraphics{../figures/o3-regr-1.pdf}
-\caption{\label{fig:o3-regr}Regression of SON mean Total Ozone Column anomalies (shaded, Dobson Units) with the (a) real and (b) imaginary parts of the first cEOF for the 1979 -- 2019 period.
+\caption{\label{fig:o3-regr}Regression of SON mean Total Ozone Column anomalies (shaded, Dobson Units) with the (a) real and (b) imaginary parts of the cEOF1 for the 1979 -- 2019 period.
 On contours, the mean zonal anomaly of Total Ozone Column (negative contours in dashed lines, Dobson Units).
@@
 There is a close spatial relationship between amplitudes and phases of the cEOF1 and TOC planetary wave 1 between 75°S and 45°S.
-The correlation between the amplitude of both indices is 0.77 (CI: 0.61 -- 0.87) the overall correlation between the zonal wave 1 of ozone and cEOF1 (computed as the mean cosine of the difference in phase) is 0.87 (CI: 0.77 -- 0.93).
-As expected from the location of regressions in Figure~\ref{fig:o3-regr}, the Real cEOF1 drives the relationship of amplitudes and the Imaginary cEOF1 drives the relationship of phases between the timeseries (not shown).
+The correlation between the amplitude of both indices is 0.77 (CI: 0.61 -- 0.87) and 0.49 (CI: 0.21 -- 0.69) between their phases.
 
@@
 \includegraphics{../figures/sst-psi-2-1} \caption{Regression maps of cEOF2 with SST (column a) and streamfunction zonal anomalies (shaded) with their corresponding activity wave flux (vectors) (column b).
-Areas marked with dots have p-values smaller than 0.01 adjusted for FDR.}\label{fig:sst-psi-2}
+Areas marked with dots have p-values smaller than 0.01 adjusted for FDR.
+\textcolor{red}{Agregar unidades.
+Quedaría mucho mejor el gráfico y creo que se prestaría a menos confusión si colocas las barras de colores en las columnas corrrespondientes.
+Otro detalle es que en gral.
+para el gráfico de streamfunction la escala de colores se usa al revés para que los azules (positivos) sean ciclónicos y los rojos anticiclónicos}}\label{fig:sst-psi-2}
 \end{figure*}
 
-The connections between cEOFs and Tropical sources is also assessed.
-Figure~\ref{fig:sst-psi-2} shows the regression maps of Sea Surface Temperatures (SST) and Streamfunction at 200 hPa respectively upon cEOF2.
+The connections between cEOFs and Tropical sources are also assessed.
+Figure~\ref{fig:sst-psi-2} shows the regression maps of Sea Surface Temperatures (SST) and Streamfunction anomalies at 200 hPa respectively upon cEOF2.
 The Imaginary cEOF2 is associated with strong positive SST anomalies on the Central Pacific and negative anomalies over an area across the North of Australia and New Zealand, the South Pacific Convergence Zone (SPCZ) (Figure~\ref{fig:sst-psi-2}.a2).
-This pattern is almost canonically positive ENSO \citep{bamston1997} and indeed, the correlation between the Imaginary cEOF2 and the Oceanic Niño Index \citep{bamston1997} is very high 0.76 (CI: 0.59 -- 0.87).
+This pattern is almost canonically positive ENSO \citep{bamston1997} and indeed, the correlation between the Imaginary cEOF2 and the Oceanic Niño Index \citep{bamston1997} is significant and very high 0.76 (CI: 0.59 -- 0.87).
 Besides, the Imaginary cEOF2 is associated with strong wave-like streamfunction anomalies emanating from the tropics (Figure~\ref{fig:sst-psi-2}.b2).
-This is consistent with the effect of ENSO on the extratropics: SST anomalies initiate anomalous convection that excites Rossby waves that propagate meridionally towards higher latitudes \citep{mo2000}.
+This is consistent with the effect of ENSO on the extratropics: SST anomalies initiate anomalous tropical convection that excites Rossby waves that propagate meridionally towards higher latitudes \citep{mo2000}.
 
-Since the Real cEOF2 represents just a different phase of the same wave train, it would be expected to show a similar forcing pattern to the Imaginary cEOF with a slight translation of its location.
+Since the Real cEOF2 represents just a different phase of the same wave train, it would be expected to show a similar forcing pattern to the Imaginary cEOF2 with a slight translation of its location.
 However, Figure~\ref{fig:sst-psi-2}.a1 and b1 show that the Real cEOF2 is not associated either with any significant SST nor streamfunction anomalies in the tropics.
 The correlation between the Real cEOF2 and ENSO is also not significant (0.00025 (CI: -0.31 -- 0.31)).
-This lack of tropical signal suggests different natures of the different phases of the Real cEOF2 wave train, which has the maximum amplitudes in the surroundings of southern South America.
+\textcolor{red}{Usaría notación científica por la cantidad de 0 que tiene}.
+This lack of tropical signal suggest that the Real cEOF2 and the Imaginary cEOF2 have very different sources of variability.
 
@@
 The black line in Figure~\ref{fig:enso-phase} is a sinusoidal fit of the relationship between ONI and cEOF2 phase.
-The \(r^2\) corresponding to the fit is 0.56, with p-value \textless{} 0.001, indicating a quasi-sinusoidal relation between these two variables.
+The \(r^2\) corresponding to the fit is 0.56, statistically significant with p-value \textless{} 0.001, indicating a quasi-sinusoidal relation between these two variables.
 
-Furthermore, Figure~\ref{fig:enso-phase} suggest that higher amplitude cEOF2 years tend to coincide with stronger ENSO years.
+\textcolor{red}{Fig. 12 Revisa cómo te quedo abajo de la figura la indicación de qué son cada uno de los puntos (te quedaron dos circulos y una línea extra que no va)}
+
+Furthermore, Figure~\ref{fig:enso-phase} suggest that highest amplitude cEOF2 years tend to coincide with strong ENSO years.
 The correlation between the absolute magnitude of the ONI and the magnitude of the cEOF2 is 0.45 (CI: 0.17 -- 0.67).
@@
 Consistently, streamfunction anomalies do not show any tropical influence.
-Instead of that, the real and imaginary CEOF1 are associated with zonally wave activity fluxes in the extra-tropics around 60ºS, except for an equatorward flow from the coast of Antarctica around 150ºE in the real component.
+Instead, the real and imaginary cEOF1 are associated with zonally wave activity fluxes in the extra-tropics around 60ºS, except for an equatorward flow from the coast of Antarctica around 150ºE in the real component.
+This suggests that it's variability is driven primary by the internal variability of the extra-tropics.
+
+\textcolor{red}{Fig. 13 Tene en cuenta los mismos comentarios que en el figura 11 y revisa también la barra de colores para SST que quedo rara}
 
@@
 
+\textcolor{red}{Revisa el caption de la figura 14 que equedo un estilo distinto que los anteriores.
+Te falto aclarar también qué es la caja (¿la inventaste vos o habías utilizado la de algún otro trabajo?).
+Fijate también de aumentar la separación entre la columna izquierda y derecha para que no se superpongan los labels.
+También fijate que esta barra de colores no te quedo con el borde externo como el resto.
+}
+
 \begin{figure}
@@
 For the 1979 -- 2019.
-Areas marked with dots have p-values smaller than 0.05 adjusted for False Detection Rate.}
+Areas marked with dots have p-values smaller than 0.05 adjusted for False Detection Rate.
+In black is the West Australia region, defined as the box between 130°E and 155°E and 40°S and 15°S.}
 \end{figure}
 
-The precipitation anomalies in Oceania associated with cEOF1 are weak and not statistically significant (Figure~\ref{fig:pp-oceania} column a) whereas EFO2 is associated with large and widespread anomalies.
+The precipitation anomalies in Oceania associated with cEOF1 are weak and not statistically significant (Figure~\ref{fig:pp-oceania} column a) whereas cEOF2 is associated with large and widespread anomalies.
 The Real cEOF2 is associated with positive seasonal anomalies in Northern and Eastern Australia, and negative ones over New Zealand, while precipitation anomalies associated with Imaginary cEOF2 are negative over all Eastern Australia.
-Again, these negative anomalies are similar to the springtime precipitation anomalies observed in relation with the Asymmetric SAM \citep{campitelli2021} and ENSO \citep{cai2011}.
+As it could be expected, these negative anomalies are similar to the springtime precipitation anomalies observed in relation with the Asymmetric SAM \citep{campitelli2021} and ENSO \citep{cai2011}.
+\textcolor{red}{Acá te faltaría  sacar un poco más el jugo asociando con la figura 11. Además, ¿revisaste si podes también usar algún resultado asociado a los PSA? }
+
+\textcolor{red}{Revisa lo mismo que para la otra figura del tipo. }
 
@@
 \includegraphics{../figures/pp-america-1.pdf}
-\caption{\label{fig:pp-america}As Figure~\ref{fig:pp-america} but for South America}
+\caption{\label{fig:pp-america}As Figure~\ref{fig:pp-oceania} but for South America}
 \end{figure}
@@
 Figure~\ref{fig:pp-america} shows regression maps of seasonal precipitation with each cEOF in South America.
-cEOF1 is associated with slight decreases in precipitation in Southern Brazil and Paraguay, although these are not statistically significant (Fig.~\ref{fig:pp-america} column a).
+cEOF1 is associated with negative regressions coefficients for precipitation in Southern Brazil and Paraguay, although these are not statistically significant (Fig.~\ref{fig:pp-america} column a).
 The strongest precipitation regression coefccients are the ones associated with the Imaginary cEOF2.
 The positive regression coefficients on Southeastern South America (SESA) and Central Chile, and negative ones over South Atlantic Convergence Zone is a well-known springtime precipitation signature of ENSO \citep{cai2020a} and it is also similar to the precipitation anomalies associated with the A-SAM \citep{campitelli2021}.
+\textcolor{red}{Lo mismo que te puse antes. Te faltaría  sacar un poco más el jugo asociando con la figura 11. Además, ¿revisaste si podes también usar algún resultado asociado a los PSA? }
+
 This is not surprising considering the close relationship between the ONI, the A-SAM index and the Imaginary cEOF2 shown previously, but further consolidates the identification of this mode with the PSA pattern.
@@
 Resembling the relationship between ONI and the phase of cEOF2 (Fig.~\ref{fig:enso-phase}), there is a cEOF2 phase dependence of the precipitation anomalies in SESA (not shown).
-These variables could be related wth a significant sinusoidal fit that has a coefficient of determination of~0.27 (p-value~= 0.0022).
+These variables could be related wth a significant sinusoidal fit that has a statistically significant coefficient of determination of~0.27.
 
@@
 
-In this study we derived two complex indices of the zonally asymmetric circulation of the September-October-November southern extratropics using Complex Empirical Orthogonal Functions.
+In this study we assessed the Austral Spring Extratropical Southern Hemisphere zonally assymetric circulation.
+For this purpose, two complex indices were derived using Complex Empirical Orthogonal Functions.
 This allowed us to characterise both the amplitude and phase of planetary waves that are not perfectly organised as sinusoidal waves with constant amplitude at each latitude circle.
@@
 The first complex EOF represents the variability of the zonal wave 1 in the stratosphere, but represents a more zonally symmetric monopole in the troposphere.
-There is a statistically positive trend in the magnitude of this cEOF, which is consistent with previous studies which showed secular changes in springtime wave-1-like patterns \citep[e.g.][]{raphael2003}.
-This mode is closely related to anomalies in Total Ozone Column.
+There is a statistically positive trend in the magnitude of this cEOF, which is consistent with previous studies that showed secular changes in springtime wave-1-like patterns \citep[e.g.][]{raphael2003}.
+This mode is closely related to stratospheric variability such as anomalies in Total Ozone Column.
+Otherwise, this complex EOF is not related with SST variability and continental precipitation in the Southern Hemisphere.
 
@@
 On the other hand, over Australia, both the Real and Imaginary phase are associated with significant precipitation anomalies, and we further show that the direction of maximum impact is not aligned with our chosen rotation of cEOF2.
-However, this underscores the benefit of using complex EOF, since it would trivial to rotate it.
+However, this underscores the benefit of using complex EOF, since it would be trivial to rotate it. \textcolor{red}{creo que esta idea convendría desarrollarla mejor para el que no lo tenga tan claro}
 
-To the extent that the cEOF2 can be identified with the PSA, what is the advantage of our technique?
+To the extent that the cEOF2 can be identified with the PSA, what is the advantage of the technique proposed here?
 The spatial fields that the cEOF2 index describes are in quadrature by construction, which makes it possible to derive a proper amplitude and phase.
 Furthermore, because of the way they are constructed, they are not forced to be orthogonal to the SAM.
-This makes it possible to study the relationship between the two modes free of the statistical contamination that comes with the usual definition of the PSA as the second and third EOFs.
+This makes it possible to study their relationship with the SAM free of the statistical contamination that comes with the usual definition of the PSA as the second and third EOFs.
 

### 2021-08-11 R/maps.R

@@ -8,40 +8,46 @@
 #' @param ... other arguments passed to [ggplot2::geom_polygon()]
 #'
 #' @export
-geom_qmap <- function(subset = identity, color = "gray50", size = 0.2,
+geom_qmap <- function(subset = identity,
+                      crop = NULL,
+                      color = "gray50", size = 0.2,
                       fill = NA, wrap = c(0, 360), weighting = 0.7,
                       keep = 0.015, ...) {
   lon <- lat <- group <- NULL
   data <- map_simple(wrap = wrap, keep  = keep, weighting = weighting)
 
-  data <- data %>%
-    fortify() %>%
-    data.table::as.data.table() %>%
-    .[, c("long", "lat", "group")] %>%
-    data.table::setnames("long", "lon")
+  if (!is.null(crop)) {
+    bbox <- sf::st_bbox(data)
+
+    for (n in names(crop)) {
+      bbox[[n]] <- crop[[n]]
+    }
+
+    data <- sf::st_crop(data, bbox)
+  }
+
+
   subset <- purrr::as_mapper(subset)
   data <- subset(data)
 
-  ggplot2::geom_polygon(data = data,
-                        ggplot2::aes(lon, lat, group = group),
-                        color = color,
-                        size = size,
-                        fill = fill,
-                        ...)
+  ggplot2::geom_sf(data = data,
+                   inherit.aes = FALSE,
+                   color = color,
+                   size = size,
+                   fill = fill,
+                   ...)
+
 }
 
 map_simple <- function(wrap = c(0, 360), keep = 0.015, weighting = 0.7) {
   map <- maps::map("world", fill = TRUE,
                    col = "transparent", plot = FALSE, wrap = wrap)
-  IDs <- vapply(strsplit(map$names, ":"), function(x) x[1],
-                "")
-  proj <- sp::CRS("+proj=longlat +datum=WGS84")
-  map <- maptools::map2SpatialPolygons(map, IDs = IDs,
-                                       proj4string = proj)
-
+  map <- sf::st_as_sf(map)
   if (keep != 1) {
     map <- rmapshaper::ms_simplify(map, keep = keep, weighting = weighting)
   }
+
+
   map
 }
 
@@ -54,3 +60,9 @@ fortify <- ggplot2::fortify
 }
 
 
+
+`%||%` <- function (x, y) {
+  if (is_null(x))
+    y
+  else x
+}

### 2021-08-11 analysis/paper/shceof.Rmd
Refina los puntos

@@
                                    colour = NA,
-                                   pattern_colour = "black",
+                                   pattern_colour = NA,
                                    pattern_fill = "black",
-                                   pattern_density = 0.05,
-                                   pattern_alpha = 1,
-                                   fill = NA,
-                                   pattern_spacing = 0.025))
-
+                                   pattern_density = 0.25,
+                                   pattern_alpha = 0.5,
+                                   pattern_spacing = 0.03,
+                                   fill = NA
+                                   ))
 update_geom_defaults(GeomSmooth, list(colour = "#0d52bf"))

### 2021-08-12 analysis/paper/shceof.Rmd
Arregla mayúsculuas

@@
 There is only a significant upward trend in the real component of cEOF1 (Fig. \@ref(fig:extended-series)a.1, p-value `r trends[part == "Real" & cEOF == "cEOF1"]$p_report`) and no significant trend in any of the complex components of cEOF2.
-The positive trend in the real cEOF1 translates into a positive trend in cEOF1 magnitude, but not in the phase (not shown).
+The positive trend in the Real cEOF1 translates into a positive trend in cEOF1 magnitude, but not in the phase (not shown).
 This long-term change indicates an increase in the magnitude of the high latitude zonal wave 1.
@@
 At 50 hPa (Figure \@ref(fig:eof1-regr-gh) row a), both the Real and Imaginary cEOF1 are associated with planetary wave 1 patterns, that are 90º out of phase.
-Their phases coincide with the ones shown in Figure \@ref(fig:ceofs-1)a.1, with the positive centre of the real cEOF1 located towards the dateline, and the one of the Imaginary cEOF1 located over Eastern Antarctica.
+Their phases coincide with the ones shown in Figure \@ref(fig:ceofs-1)a.1, with the positive centre of the Real cEOF1 located towards the dateline, and the one of the Imaginary cEOF1 located over Eastern Antarctica.
 However, the Real cEOF1 pattern is substantially altered by the zonally symmetric circulation.
@@
 The regression patterns for cEOF1 at 200 hPa (Figure \@ref(fig:eof1-regr-gh) row b) are similarly influenced by the zonally symmetric circulation.
-It is only possible to partially distinguish the wave 1 pattern in relation with the real cEOF1 (Figure \@ref(fig:eof1-regr-gh)b.1).
-The imaginary cEOF1 shows a much more zonally symmetrical pattern resembling the negative SAM phase.
-
-With the exception of the imaginary cEOF1, it is clear that these patterns are very different than the fully zonally asymmetric versions (Fig. \@ref(fig:ceofs-1)), particularly at 200 hPa.
+It is only possible to partially distinguish the wave 1 pattern in relation with the Real cEOF1 (Figure \@ref(fig:eof1-regr-gh)b.1).
+The Imaginary cEOF1 shows a much more zonally symmetrical pattern resembling the negative SAM phase.
+
+With the exception of the Imaginary cEOF1, it is clear that these patterns are very different than the fully zonally asymmetric versions (Fig. \@ref(fig:ceofs-1)), particularly at 200 hPa.
 Moreover, only in the stratosphere these patterns actually show a distinguishable wave 1 pattern shifted in phase by 90º, suggesting that using the cEOF method is artificially generating a wave structure at 200hPa.
@@
 Unlike for cEOF1, in this case the regression patterns are similar to the fully zonally asymmetric patterns from Figure \@ref(fig:ceofs-1).
-Although there are some differences (particularly in 50 hPa), the wave trains identified before are well characterised and patterns associated with the real cEOF2 are 90º out of phase with those associated with the imaginary cEOF2.
+Although there are some differences (particularly in 50 hPa), the wave trains identified before are well characterised and patterns associated with the Real cEOF2 are 90º out of phase with those associated with the Imaginary cEOF2.
 Zonal wave 3 dominates all fields, but only in the western hemisphere, over the Pacific and Atlantic Oceans.
 cEOF2 then represents an equivalent barotropic wave train that is very similar to the the Pacific South American Patterns [@mo2001].
-Comparing the location of the positive anomaly near 90ºW in column b of Figure \@ref(fig:eof2-regr-gh) with Figures 1.a and b from @mo2001, the real cEOF2 can roughly be identified with PSA2, while the imaginary cEOF2 resembles PSA1.
+Comparing the location of the positive anomaly near 90ºW in column b of Figure \@ref(fig:eof2-regr-gh) with Figures 1.a and b from @mo2001, the Real cEOF2 can roughly be identified with PSA2, while the Imaginary cEOF2 resembles PSA1.
 
@@
 
-\textcolor{red}{Revisa que tenes una incosistencia en el uso de mayúsculas en Real/Imaginary porque a veces las usas y a veces no.}
-
 ```{r o3-regr, fig.cap = "(ref:o3-regr-cap)", fig.height=3, fig.width = width_column}
@@
 Consistently, streamfunction anomalies do not show any tropical influence.
-Instead, the real and imaginary cEOF1 are associated with zonally wave activity fluxes in the extra-tropics around 60ºS, except for an equatorward flow from the coast of Antarctica around 150ºE in the real component.
+Instead, the real and Imaginary cEOF1 are associated with zonally wave activity fluxes in the extra-tropics around 60ºS, except for an equatorward flow from the coast of Antarctica around 150ºE in the real component.
 This suggests that it's variability is driven primary by the internal variability of the extra-tropics.

### 2021-08-12 analysis/paper/shceof.Rmd
Pasa a break a oración

@@
 One typical way of describing the zonally asymmetric circulation is by the amplitude and phase of zonal waves obtained by Fourier decomposition of geopotential height at each latitude [e.g.
-@vanloon1972;@trenberth1980;@turner2017].
+@vanloon1972; @trenberth1980; @turner2017].
 This approach suggests that zonal waves 1 and 3 explained most of the variance of the tropospheric extratropical Southern Hemisphere circulation (@vanloon1972).
@@
 \textcolor{red}{se me vino una duda, ¿se pueden comparar de algún modo las amplitudes que mostras en esta figura?
-en caso de que si, creo que habría que mencionar algo sobre lo que pasa con el cEOF2 en la estratósfera}
-All cEOFs have non zero mean (Fig.
+en caso de que si, creo que habría que mencionar algo sobre lo que pasa con el cEOF2 en la estratósfera} All cEOFs have non zero mean (Fig.
 \@ref(fig:extended-series)) because the geopotential fields that enter into the cEOFs algorithm are anomalies with respect to the zonal mean instead of the time mean.
@@
 ## cEOFs relationship with known modes of variability {#other-variables}
+
 ### SAM
@@
 
-(ref:pp-oceania-cap) Regression of SON mean precipitation anomalies in Oceania (mm per day, shaded) and (column a) cEOF1 the (row 1) Imaginary and (column 1) Real phase. For the 1979 -- 2019.
+(ref:pp-oceania-cap) Regression of SON mean precipitation anomalies in Oceania (mm per day, shaded) and (column a) cEOF1 the (row 1) Imaginary and (column 1) Real phase.
+For the 1979 -- 2019.
 Areas marked with dots have p-values smaller than 0.05 adjusted for False Detection Rate.
@@
 (ref:australia-pp-phase-cap) Same as Figure \@ref(fig:enso-phase) but for precipitation anomalies averaged between `r knitr::combine_words(LonLabel(box_australia$lon))` and `r knitr::combine_words(LatLabel(box_australia$lat))` (box shown in Figure \@ref(fig:pp-oceania) column b) in $\frac{\mathrm{mm}}{\mathrm{day}}$.
-Black line is the fit precipitation ~ sin(phase) + cos(phase) computed by weighted OLS using the magnitude of the cEOF2 as weights.
+Black line is the fit precipitation \~ sin(phase) + cos(phase) computed by weighted OLS using the magnitude of the cEOF2 as weights.
 
@@
 The first complex EOF represents the variability of the zonal wave 1 in the stratosphere, but represents a more zonally symmetric monopole in the troposphere.
-There is a statistically positive trend in the magnitude of this cEOF, which is consistent with previous studies that showed secular changes in springtime wave-1-like patterns [e.g.
-@raphael2003].
+There is a statistically positive trend in the magnitude of this cEOF, which is consistent with previous studies that showed secular changes in springtime wave-1-like patterns [e.g. @raphael2003].
 This mode is closely related to stratospheric variability such as anomalies in Total Ozone Column.
@@
 On the other hand, over Australia, both the Real and Imaginary phase are associated with significant precipitation anomalies, and we further show that the direction of maximum impact is not aligned with our chosen rotation of cEOF2.
-However, this underscores the benefit of using complex EOF, since it would be trivial to rotate it. \textcolor{red}{creo que esta idea convendría desarrollarla mejor para el que no lo tenga tan claro}
+However, this underscores the benefit of using complex EOF, since it would be trivial to rotate it.
+\textcolor{red}{creo que esta idea convendría desarrollarla mejor para el que no lo tenga tan claro}
 

### 2021-08-12 analysis/paper/shceof.Rmd
Arregla leyendas

@@
 
-theme_set(theme_shceof(base_size = 10))
+theme_set(theme_shceof(base_size = 10) +
+            theme(legend.box.spacing = grid::unit(-.5, "lines")))
 
@@
   coord_polar() +
-  theme(axis.text = element_blank()) +
+  theme(axis.text = element_blank(), 
+        legend.box.spacing = grid::unit(-1, "lines")) +
   facet_wrap(.~EOF, ncol = 2)  +
@@
   .[enso, on = "time"] %>%
-  na.omit()
+  na.omit() %>%
+  .[, q := ifelse(mag >= quantile(mag, 0.5), "|cEOF2| > 50%", "|cEOF2| < 50%")]
 
@@
               se = FALSE, color = "black", size = 0.5) +
-  geom_point(aes(colour = mag >= quantile(mag, 0.5),
-                 shape  = mag >= quantile(mag, 0.5)),
+  geom_point(aes(color = q,
+                 shape  = q),
              size = 3) +
 
-  scale_color_brewer(NULL, palette = "Dark2",
-                     labels = c("TRUE"  = "|cEOF2| > 50%",
-                                "FALSE" = "|cEOF2| < 50%")) +
-  scale_shape_manual(NULL, values = c("TRUE" = 20,
-                                      "FALSE" = 18),
-                     labels = c("TRUE"  = "|cEOF2| > 50%",
-                                "FALSE" = "|cEOF2| < 50%")) +
+  scale_color_brewer("", palette = "Dark2") +
+  scale_shape_manual("", values = c(20, 18)) +
   scale_y_continuous("ONI") +
@@
   .[, arg := Arg(hgt)] %>% 
-  .[, mag := Mod(hgt)]
+  .[, mag := Mod(hgt)] %>% 
+  .[, q := ifelse(mag >= quantile(mag, 0.5), "|cEOF2| > 50%", "|cEOF2| < 50%")]
 
@@
               fullrange = TRUE) +
-  geom_point(aes(colour = mag >= quantile(mag, 0.5),
-                 shape  = mag >= quantile(mag, 0.5)),
+  geom_point(aes(colour = q,
+                 shape  = q),
              size = 3) +
@@
 
-  scale_color_brewer(NULL, palette = "Dark2",
-                     labels = c("TRUE"  = "|EOF2| > 50%",
-                                "FALSE" = "|EOF2| < 50%")) +
-  scale_shape_manual(NULL, values = c("TRUE" = 20,
-                                      "FALSE" = 18),
-                     labels = c("TRUE"  = "|EOF2| > 50%",
-                                "FALSE" = "|EOF2| < 50%")) +
+  scale_color_brewer(NULL, palette = "Dark2") +
+  scale_shape_manual(NULL, values = c(20, 18)) +
   scale_y_continuous("Precipitation anomaly") +

### 2021-08-12 analysis/paper/shceof.Rmd
Mejora escala de colores stremfunction

@@
 guide_colorsteps_bottom <- function(width = 15, height = 0.5, even.steps = FALSE, 
-                                    title.position = "top", ...) {
-  guide_colorsteps(show.limits = TRUE, even.steps = even.steps,
+                                    title.position = "top", show.limits = TRUE,
+                                    ...) {
+  guide_colorsteps(show.limits = show.limits, even.steps = even.steps,
                    title.position = title.position, title.hjust = 0.5,
@@
 plot_sst_psi <- function(which_cEOF, which_lev) {
+  one_less <- function(x) {
+    n <- length(x)
+    if (n < 4) return(x)
+    x <- round(x, 5)
+
+    c( rev(JumpBy(rev(x[x < 0]), 2, fill = "")),
+       JumpBy(x[x > 0], 2, fill = ""))
+  }
+
   sst_gdata <- regr_sst[cEOF == which_cEOF] %>%
     .[lat %between% c(-90, 10)] %>%
-    .[, p_val := Pvaluate(estimate, std.error, df, "fdr"), by = .(cEOF, term)] %>%
+    .[, p_val := Pvaluate(estimate, std.error, df, "fdr"), 
+      by = .(cEOF, term)] %>%
     .[, var := "SST"]
 
-  regr_psi %>%
+  rsst <- max(range(abs(regr_sst[!is.na(cEOF), ]$estimate), na.rm = TRUE))
+  breaks_sst <- AnchorBreaks(0, bins = 20, exclude = 0)(c(-rsst, rsst))
+
+  stream_gdata <- regr_psi %>%
     .[cEOF == which_cEOF] %>%
@@
     .[, p_val := Pvaluate(estimate, std.error, df, "fdr"), by = .(cEOF, term)] %>%
-    .[, var := "Streamfunction"] %>%
+    .[, var := "Streamfunction"]
+
+  rstream <- max(range(abs(regr_psi[lev == which_lev]$estimate), 
+                       na.rm = TRUE))*1e-7
+  breaks_stream <- AnchorBreaks(0, bins = 20, exclude = 0)(c(-rstream, rstream))
+
+  # browser()
+  stream_gdata %>% 
     periodic(lon = c(0, 360)) %>%
@@
     geom_contour_fill(aes(z = estimate*1e-7, fill = ..level..),
-                      breaks = AnchorBreaks(exclude = 0)) +
+                      breaks = breaks_stream) +
     geom_contour_pval(aes(z = p_val)) +
@@
     geom_qmap(keep = .015) +
-    scale_fill_divergent_discretised("Streamfunction (×1e-7)", 
-                                     guide = guide_colorsteps_bottom(20, order = 1)) +
+    scale_fill_divergent_discretised("Streamfunction (m²/s×1e-7)", 
+                                     labels = one_less,
+                                      low = scales::muted("red"),
+                                     high = scales::muted("blue"),
+                                     guide = guide_colorsteps_bottom(15, show.limits = FALSE,
+                                                                     order = 99)) +
     ggnewscale::new_scale_fill() +
@@
                       aes(z = estimate, fill = ..level..), 
-                      breaks = AnchorBreaks(exclude = 0)) +
+                      breaks = breaks_sst) +
     geom_contour_pval(data = sst_gdata, aes(z = p_val)) +
-    scale_fill_divergent_discretised("SST", 
-                                     guide = guide_colorsteps_bottom(order = 2)) +
+    scale_fill_divergent_discretised("SST (K)", 
+                                     labels = one_less,
+                                     guide = guide_colorsteps_bottom(width = 15,
+                                                                     show.limits = FALSE,
+                                                                     order = 1)) +
     scale_x_longitude() +
@@
              default_crs = "+proj=longlat +over") + 
-  theme(legend.box = "vertical") +
+    theme(legend.box = "horizontal", legend.text = element_text(size = rel(0.7))) +
     tag_facets("cr", position = "bl") 
@@
 
-(ref:sst-psi-2-cap) Regression maps of cEOF2 with SST (column a) and streamfunction zonal anomalies (shaded) with their corresponding activity wave flux (vectors) (column b).
+(ref:sst-psi-2-cap) Regression maps of cEOF2 with SST (K, column a) and streamfunction zonal anomalies ($m^2/s\times10^-7$, shaded) with their corresponding activity wave flux (vectors) (column b).
 Areas marked with dots have p-values smaller than 0.01 adjusted for FDR.
-\textcolor{red}{Agregar unidades.
-Quedaría mucho mejor el gráfico y creo que se prestaría a menos confusión si colocas las barras de colores en las columnas corrrespondientes.
-Otro detalle es que en gral.
-para el gráfico de streamfunction la escala de colores se usa al revés para que los azules (positivos) sean ciclónicos y los rojos anticiclónicos}
 
 ```{r sst-psi-2,  fig.cap = "(ref:sst-psi-2-cap)", fig.width = 7, fig.height=4, dependson="plot_sst_psi", fig.env="figure*"}
-plot_sst_psi(which_cEOF = "cEOF2", which_lev = 200) +
-  scale_fill_divergent_discretised("SST",
-                                   limits = c(-1, 1),
-                                   guide = guide_colorsteps_bottom(20, order = 2)) 
+plot_sst_psi(which_cEOF = "cEOF2", which_lev = 200)
+
 ```
@@
 ```{r sst-psi-1, fig.cap = "(ref:sst-psi-1-cap)", fig.width = 7, fig.height=4, dependson="plot_sst_psi", fig.env = "figure*"}
-plot_sst_psi(which_cEOF = "cEOF1", which_lev = 200) +
-  scale_fill_divergent_discretised("SST",
-                                   limits = c(-1, 1),
-                                   guide = guide_colorsteps_bottom(20, order = 2)) 
+plot_sst_psi(which_cEOF = "cEOF1", which_lev = 200)
+
 ```

### 2021-08-12 analysis/paper/shceof.tex
Renderiza

@@
 \textcolor{red}{se me vino una duda, ¿se pueden comparar de algún modo las amplitudes que mostras en esta figura?
-en caso de que si, creo que habría que mencionar algo sobre lo que pasa con el cEOF2 en la estratósfera}
-All cEOFs have non zero mean (Fig. \ref{fig:extended-series}) because the geopotential fields that enter into the cEOFs algorithm are anomalies with respect to the zonal mean instead of the time mean.
+en caso de que si, creo que habría que mencionar algo sobre lo que pasa con el cEOF2 en la estratósfera} All cEOFs have non zero mean (Fig.
+\ref{fig:extended-series}) because the geopotential fields that enter into the cEOFs algorithm are anomalies with respect to the zonal mean instead of the time mean.
 However, cEOF2's mean is almost zero, which indicates that only cEOF1 includes variability that significantly projects onto the mean zonally anomalous field.
@@
 There is only a significant upward trend in the real component of cEOF1 (Fig.~\ref{fig:extended-series}a.1, p-value~\textless{} 0.001) and no significant trend in any of the complex components of cEOF2.
-The positive trend in the real cEOF1 translates into a positive trend in cEOF1 magnitude, but not in the phase (not shown).
+The positive trend in the Real cEOF1 translates into a positive trend in cEOF1 magnitude, but not in the phase (not shown).
 This long-term change indicates an increase in the magnitude of the high latitude zonal wave 1.
@@
 At 50 hPa (Figure~\ref{fig:eof1-regr-gh} row a), both the Real and Imaginary cEOF1 are associated with planetary wave 1 patterns, that are 90º out of phase.
-Their phases coincide with the ones shown in Figure~\ref{fig:ceofs-1}a.1, with the positive centre of the real cEOF1 located towards the dateline, and the one of the Imaginary cEOF1 located over Eastern Antarctica.
+Their phases coincide with the ones shown in Figure~\ref{fig:ceofs-1}a.1, with the positive centre of the Real cEOF1 located towards the dateline, and the one of the Imaginary cEOF1 located over Eastern Antarctica.
 However, the Real cEOF1 pattern is substantially altered by the zonally symmetric circulation.
@@
 The regression patterns for cEOF1 at 200 hPa (Figure~\ref{fig:eof1-regr-gh} row b) are similarly influenced by the zonally symmetric circulation.
-It is only possible to partially distinguish the wave 1 pattern in relation with the real cEOF1 (Figure~\ref{fig:eof1-regr-gh}b.1).
-The imaginary cEOF1 shows a much more zonally symmetrical pattern resembling the negative SAM phase.
+It is only possible to partially distinguish the wave 1 pattern in relation with the Real cEOF1 (Figure~\ref{fig:eof1-regr-gh}b.1).
+The Imaginary cEOF1 shows a much more zonally symmetrical pattern resembling the negative SAM phase.
 
-With the exception of the imaginary cEOF1, it is clear that these patterns are very different than the fully zonally asymmetric versions (Fig.~\ref{fig:ceofs-1}), particularly at 200 hPa.
+With the exception of the Imaginary cEOF1, it is clear that these patterns are very different than the fully zonally asymmetric versions (Fig.~\ref{fig:ceofs-1}), particularly at 200 hPa.
 Moreover, only in the stratosphere these patterns actually show a distinguishable wave 1 pattern shifted in phase by 90º, suggesting that using the cEOF method is artificially generating a wave structure at 200hPa.
@@
 Unlike for cEOF1, in this case the regression patterns are similar to the fully zonally asymmetric patterns from Figure~\ref{fig:ceofs-1}.
-Although there are some differences (particularly in 50 hPa), the wave trains identified before are well characterised and patterns associated with the real cEOF2 are 90º out of phase with those associated with the imaginary cEOF2.
+Although there are some differences (particularly in 50 hPa), the wave trains identified before are well characterised and patterns associated with the Real cEOF2 are 90º out of phase with those associated with the Imaginary cEOF2.
 Zonal wave 3 dominates all fields, but only in the western hemisphere, over the Pacific and Atlantic Oceans.
 cEOF2 then represents an equivalent barotropic wave train that is very similar to the the Pacific South American Patterns \citep{mo2001}.
-Comparing the location of the positive anomaly near 90ºW in column b of Figure~\ref{fig:eof2-regr-gh} with Figures 1.a and b from \citet{mo2001}, the real cEOF2 can roughly be identified with PSA2, while the imaginary cEOF2 resembles PSA1.
+Comparing the location of the positive anomaly near 90ºW in column b of Figure~\ref{fig:eof2-regr-gh} with Figures 1.a and b from \citet{mo2001}, the Real cEOF2 can roughly be identified with PSA2, while the Imaginary cEOF2 resembles PSA1.
 
@@
 
-\textcolor{red}{Revisa que tenes una incosistencia en el uso de mayúsculas en Real/Imaginary porque a veces las usas y a veces no.}
-
 \begin{figure}
@@
 \begin{figure*}
-\includegraphics{../figures/sst-psi-2-1} \caption{Regression maps of cEOF2 with SST (column a) and streamfunction zonal anomalies (shaded) with their corresponding activity wave flux (vectors) (column b).
+\includegraphics{../figures/sst-psi-2-1} \caption{Regression maps of cEOF2 with SST (K, column a) and streamfunction zonal anomalies (\(m^2/s\times10^-7\), shaded) with their corresponding activity wave flux (vectors) (column b).
 Areas marked with dots have p-values smaller than 0.01 adjusted for FDR.

### 2021-08-12 analysis/paper/shceof.Rmd
Agerga ticks

@@
 
+guide_colorsteps <- function (even.steps = TRUE, show.limits = NULL, ticks = FALSE, 
+  ...) {
+  guide <- guide_colourbar(raster = FALSE, ticks = ticks, 
+    nbin = 300, ...)
+  guide$even.steps <- even.steps
+  guide$show.limits <- show.limits
+  class(guide) <- c("colorsteps", class(guide))
+  guide
+}
 guide_colorsteps_bottom <- function(width = 15, height = 0.5, even.steps = FALSE, 
-                                    title.position = "top", show.limits = TRUE,
+                                    title.position = "top", show.limits = TRUE, ticks = TRUE,
                                     ...) {
@@
                    title.position = title.position, title.hjust = 0.5,
-                   barheight = height,
+                   barheight = height, ticks = ticks,
+                   ticks.colour = "black",
                    barwidth = width, frame.colour = "black", ...)
@@
                                      guide = guide_colorsteps_bottom(width = 15,
-                                                                     show.limits = FALSE,
+                                                                     show.limits = FALSE, ticks = TRUE,
                                                                      order = 1)) +

### 2021-08-12 analysis/paper/shceof.tex
Render

@@
 \includegraphics{../figures/sst-psi-2-1} \caption{Regression maps of cEOF2 with SST (K, column a) and streamfunction zonal anomalies (\(m^2/s\times10^-7\), shaded) with their corresponding activity wave flux (vectors) (column b).
-Areas marked with dots have p-values smaller than 0.01 adjusted for FDR.
-\textcolor{red}{Agregar unidades.
-Quedaría mucho mejor el gráfico y creo que se prestaría a menos confusión si colocas las barras de colores en las columnas corrrespondientes.
-Otro detalle es que en gral.
-para el gráfico de streamfunction la escala de colores se usa al revés para que los azules (positivos) sean ciclónicos y los rojos anticiclónicos}}\label{fig:sst-psi-2}
+Areas marked with dots have p-values smaller than 0.01 adjusted for FDR.}\label{fig:sst-psi-2}
 \end{figure*}

### 2021-08-13 analysis/paper/shceof.Rmd
Quita colores de tablas

@@
 }
+
+correlation_statistics <- function(x, n, signif = 2) {
+  z <- atanh(x)
+  se <- 1/sqrt(n - 3)
+
+  low <- qnorm(.025, mean = z, sd = se)
+  low <- tanh(low)
+  hig <- qnorm(.975, mean = z, sd = se)
+  hig <- tanh(hig)
+
+  p.value <- pnorm(abs(x)/se, lower.tail = FALSE)
+
+  text <- paste0(signif(x, signif), " (CI: ",
+                 signif(low, signif), " -- ", signif(hig, signif), ")")
+
+  list(estimate = x, 
+       p.value = p.value, 
+       low = low,
+       hig = hig,
+       text = text
+  )
+}
+
 ```
@@
 (ref:eof-naive-cap) Spatial patterns of the leading 4 EOFs of SON zonal anomalies of geopotential height at 50 hPa south of 20\degree S for the 1979 -- 2019 period (arbitrary units).
-
-\textcolor{red}{En la figura 1 te quedaron muy chicos los subplots, amplialos un poco}.
 
@@
 
-(ref:corr-ceof-splitted-cap) Coefficient of determination ($R^2$) of the absolute magnitude of complex EOFs between 200 hPa and 50 hPa computing EOF separatedly for each level.
-Background colour indicates the value of $R^2$.
-
-\textcolor{red}{Los colores de la Tabla 1 no dicen mucho en la medida que no se conoce cuál sería la paleta. Quizás se pueden remarcar en negrita ambos aclarando que son significativos (lo son?)}.
+(ref:corr-ceof-splitted-cap) Coefficient of determination ($R^2$) of the absolute magnitude of complex EOFs between 200 hPa and 50 hPa computing EOF separately for each level.
+p-values lower than 0.01 in bold.
 
 ```{r corr-ceof-splitted}
-ceof_splitted[, eof[[1]]$left, by = .(lev = levs)] %>%
+table <- ceof_splitted[, eof[[1]]$left, by = .(lev = levs)] %>%
   copy() %>%
   .[, mag  := abs(hgt)] %>%
-  .[, item := paste(lev, cEOF, sep = "_")] %>%
-  widyr::pairwise_cor(item, time, mag) %>%
+  .[, item := paste(lev, cEOF, sep = "_")]
+
+n <- uniqueN(table$time)
+
+table2 <- table %>%
+  widyr::pairwise_cor(item, time, mag) |> 
   setDT() %>%
+  .[, p.value := correlation_statistics(correlation, n)$p.value] %>% 
   .[tstrsplit(item1, "_")[[1]] != tstrsplit(item2, "_")[[1]], ] %>%
@@
   .[, item2 := tstrsplit(item2, "_")[[2]]] %>%
-  .[, r2 := round(correlation^2, 2)] %>%
-  dcast(item1 ~ item2, value.var = "r2") %>%
+  .[, r2 := round(correlation^2, 2)]
+
+pval <- table2 %>% 
+  dcast(item1 ~ item2, value.var = c("p.value"))
+
+table2 %>% 
+  dcast(item1 ~ item2, value.var = c("r2")) %>% 
   setnames("item1", "200 hPa") %>%
   kbl2(booktabs = TRUE, caption = "(ref:corr-ceof-splitted-cap)") %>%
-  column_spec2(2:4, background = spec_color2(.table[[.col]], option = background_palette,
-                                             scale_from = c(0, 1)),
-               color = ifelse(.table[[.col]] > .5, "white", "black")) %>% 
+  column_spec2(2:4, bold = pval[[.col]] < 0.01) %>%
   add_header_above(c("", "50 hPa" = 3)) %>%
@@
 95% confidence intervals in parenthesis.
-Background colour indicates the value of correlation.
+p-values lower than 0.01 in bold.
 
@@
 
-cEOF2[psa_time, on = "time"] %>% 
+table <- cEOF2[psa_time, on = "time"] %>% 
   .[PC != "SAM"] %>% 
@@
   sep_ReIm() %>% 
-  .[, correlate(value, hgt), by = .(PC, part)] %>% 
+  .[, correlate(value, hgt), by = .(PC, part)]
+
+p.val <- table %>% 
+  dcast(PC ~ part, value.var = "p.value")
+
+table %>% 
   dcast(PC ~ part, value.var = "text") %>% 
   kbl2(booktabs = TRUE, caption = "(ref:psa-eof2-cap)") %>%
-  column_spec2(2:3, background = spec_color2(make_numeric(.table[[.col]]),
-                                             option = background_palette,
-                                             scale_from = c(-1, 1)), 
-               color = ifelse(abs(make_numeric(.table[[.col]])) > .5, "white", "black")) %>%
+  column_spec2(2:3, bold = p.val[[.col]] < 0.01) %>%
   add_header_above(c("", "cEOF2" = 2)) %>%
@@
 
-correlation_statistics <- function(x, n, signif = 2) {
-  z <- atanh(x)
-  se <- 1/sqrt(n - 3)
-
-  low <- qnorm(.025, mean = z, sd = se)
-  low <- tanh(low)
-  hig <- qnorm(.975, mean = z, sd = se)
-  hig <- tanh(hig)
-
-  p.value <- pnorm(abs(x)/se, lower.tail = FALSE)
-
-  text <- paste0(signif(x, signif), " (CI: ",
-                 signif(low, signif), " -- ", signif(hig, signif), ")")
-
-  list(estimate = x, 
-       p.value = p.value, 
-       low = low,
-       hig = hig,
-       text = text
-  )
-}
-
 wave_correlation <- temp %>% 

### 2021-08-13 analysis/paper/shceof.Rmd
Aumenta densidad de puntitos significantes

@@
                                    pattern_fill = "black",
-                                   pattern_density = 0.25,
-                                   pattern_alpha = 0.5,
-                                   pattern_spacing = 0.03,
+                                   pattern_density = 0.3,
+                                   pattern_alpha = 1,
+                                   pattern_spacing = 0.02,
                                    fill = NA

### 2021-08-13 analysis/paper/shceof.Rmd
Arreglos varios a figuras

@@
 
-```{r extended-series, fig.cap = "(ref:extended-series-cap)", fig.height=4, fig.width = 7, fig.env = "figure*"}
+```{r extended-series, fig.cap = "(ref:extended-series-cap)", fig.height=3, fig.width = 6, fig.env = "figure*"}
 # Check that the extended index coincides with the
@@
   geom_line(aes(color = part)) +
+  geom_point(aes(color = part), size = 0.25) +
   # geom_line(data = era_less) +
@@
               show.legend = FALSE, size = 0.8) +
-  scale_x_datetime(NULL, date_labels = "%Y") +
-  scale_y_continuous("EOF", minor_breaks = NULL) +
+  scale_x_datetime(NULL, date_labels = "%Y", 
+                   breaks = seq(as.POSIXct("1950-01-01"), 
+                                as.POSIXct("2020-01-01"), by = "10 years"),
+                   minor_breaks = seq(as.POSIXct("1950-01-01"), 
+                                          as.POSIXct("2020-01-01"), by = "2 years")) +
+  scale_y_continuous(NULL, minor_breaks = NULL, sec.axis = dup_axis()) +
   scale_color_brewer(NULL, palette = "Dark2", 
@@
   facet_grid(cEOF ~ part) +
-  theme(panel.spacing = grid::unit(1.5, "lines"))  +
+  theme(panel.spacing = grid::unit(1.5, "lines"),
+        strip.placement = "outside", 
+        panel.grid.minor.x = element_line(linetype = 2))  +
   tag_facets("cr") +
@@
 The wave patterns described by these cEOFs match the patterns seen in the traditional EOFs of Figure \@ref(fig:eof-naive).
-
-\textcolor{red}{tengo la impresión de que en la figura 2 podría estar un poco más grandes los subplots para que se vean más fácilmente los detalles}
 
@@
 
-\textcolor{red}{se me vino una duda, ¿se pueden comparar de algún modo las amplitudes que mostras en esta figura?
-en caso de que si, creo que habría que mencionar algo sobre lo que pasa con el cEOF2 en la estratósfera} All cEOFs have non zero mean (Fig.
+All cEOFs have non zero mean (Fig.
 \@ref(fig:extended-series)) because the geopotential fields that enter into the cEOFs algorithm are anomalies with respect to the zonal mean instead of the time mean.
@@
 Both cEOFs show year-to-year variability but show no evidence of decadal variability.
-
-\textcolor{red}{En la figura 3 agregaría dos cosas: i)el eje y en la columna derecha (si la queres ver más grande perdes la referencia) ii) creo que serviría agregar una líneas verticales en la grilla para poder tener una idea un poco mejor de a qué años corresponde más o menos cada valor.
-pueden ser un poco más finas que las que ya están, pero creo que sería una referencia util para visualmente poder decir a qué año corresponde cada punto.
-Además, fijate que la figura te quedo de un ancho mayor que el de la página y te quedo un "EOF" en el eje-y a la izquierda que no creo que sea lo más correcto.}
 
@@
 
-\textcolor{red}{La figura 4 la veo muy saturada con los puntos que indican la significancia.
-¿se podrán hacer con un tamaño menor?
-¿Las unidades son correctas?
-Me llama la atención que te puedan dar tantos cientos de metros de altura geopotencial.}
-
 The regression patterns for cEOF1 at 200 hPa (Figure \@ref(fig:eof1-regr-gh) row b) are similarly influenced by the zonally symmetric circulation.
@@
                        guide = guide_colorsteps_bottom(15, even.steps = TRUE), 
-                       limits = c(-1000+1, 1000-1))
+                       limits = c(-1000, 1000))
 ```
@@
 Comparing the location of the positive anomaly near 90ºW in column b of Figure \@ref(fig:eof2-regr-gh) with Figures 1.a and b from @mo2001, the Real cEOF2 can roughly be identified with PSA2, while the Imaginary cEOF2 resembles PSA1.
-
-\textcolor{red}{Figura 5. Me resulto raro que vaya hasta 999 en vez de 1000 }
 
@@
 
-  scale_x_level(minor_breaks = NULL) +
+  scale_x_level("Level (hPa)", minor_breaks = NULL, 
+                sec.axis = sa_height_axis(breaks = seq(0, 50, by = 10),
+                                          labels = function(x) paste0("~", x, " km"))) +
   scale_y_continuous(r2, limits = c(0, 1)) +
@@
   tag_facets("rc") +
-  panel_background
+  panel_background +
+  theme(panel.spacing = grid::unit(1, "lines"), 
+        strip.placement = "outside")
 ```

### 2021-08-13 analysis/paper/shceof.Rmd
Cambios en figuras

@@
 
-width_column <- 3.6
+width_column <- 3.3
 
@@
 
-```{r eof-naive, fig.cap = "(ref:eof-naive-cap)", fig.width=2.5, fig.height=3.2}
+```{r eof-naive, fig.cap = "(ref:eof-naive-cap)", fig.width=width_column, fig.height=width_column*1.2}
 eof_naive[, eof[[1]]$right, by = lev] %>%
@@
 
-\textcolor{red}{Fig.
-6.
-Te quedan muy pegados los subplots de la columna izquierda con los de la derecha.
-Recomendaría aumentar el espacio intermedio y repetir el eje y en las figuras de la derecha.
-Además podrías agregar el y-label (Level) a la izquierda.}
-
 The relationship between the tropospheric SAM and cEOF1 is explained entirely by the zonally symmetric component of the SAM as shown by the low and statistically non-significant correlations between the A-SAM and either the Real or Imaginary cEOF1 and the high correlation with the S-SAM below 100 hPa.
@@
 ### PSA
-
-\textcolor{red}{Tabla 2.
-Mismo comentario que antes con lo del background color (no dice mucho si no se conoce la escala), quizás podes destacar en negritas según la significancia.
-¿Está bien el intervalo de confianza de PSA2 vs cEOF2 Im?
-Hay algo raro ahí}
 
@@
   .[, correlate(hgt, oni), by = .(cEOF, part)]
+
 ```
@@
                                                                      order = 1)) +
-    scale_x_longitude() +
-    scale_y_latitude() +
+    scale_x_continuous(NULL, expand = c(0, 0), 
+                       breaks = seq(-180, 180-60, by = 60),
+                       labels = LonLabel) +
+
+    scale_y_continuous(NULL, breaks  = seq(-90, 0, by = 30), expand = c(0, 0)) +
     scale_mag(guide = "none") +
@@
     # coord_quickmap(ylim = c(NA, 10)) +
-    coord_sf(ylim = c(-90, 0),
+    coord_sf(ylim = c(-89, 0), expand = FALSE,
              crs = "+proj=longlat +pm=180 +over",
-             default_crs = "+proj=longlat +over") + 
-    theme(legend.box = "horizontal", legend.text = element_text(size = rel(0.7))) +
+             default_crs = "+proj=longlat +lon_wrap=180 +over") + 
+    theme(legend.box = "horizontal",
+          legend.text = element_text(size = rel(0.7)), 
+          legend.box.spacing = grid::unit(0, "lines") ) +
     tag_facets("cr", position = "bl") 
@@
 plot_sst_psi(which_cEOF = "cEOF2", which_lev = 200) 
-
 ```
@@
 The correlation between the Real cEOF2 and ENSO is also not significant (`r enso_cor[cEOF == "cEOF2" & part == "Real"]$text`).
-\textcolor{red}{Usaría notación científica por la cantidad de 0 que tiene}.
 This lack of tropical signal suggest that the Real cEOF2 and the Imaginary cEOF2 have very different sources of variability.
@@
 
-\textcolor{red}{Fig. 12 Revisa cómo te quedo abajo de la figura la indicación de qué son cada uno de los puntos (te quedaron dos circulos y una línea extra que no va)}
-
 Furthermore, Figure \@ref(fig:enso-phase) suggest that highest amplitude cEOF2 years tend to coincide with strong ENSO years.
@@
 This suggests that it's variability is driven primary by the internal variability of the extra-tropics.
-
-\textcolor{red}{Fig. 13 Tene en cuenta los mismos comentarios que en el figura 11 y revisa también la barra de colores para SST que quedo rara}
 
@@
 
-    scale_x_longitude(ticks = 15) +
+    scale_x_continuous("", expand = c(0, 0), breaks = seq(0, 360, by = 15), 
+                       labels = LonLabel) +
     scale_y_latitude(ticks = 15) +
@@
     axis_labs_smol +
-    tag_facets("cr") +
+    tag_facets("rc") +
     coord_sf(ylim = continents[[which_continent]]$lat,
@@
              crs = "+proj=longlat +pm=180 +over",
-             default_crs = "+proj=longlat +over")
+             default_crs = "+proj=longlat +over") +
+    theme(panel.spacing = grid::unit(1, "lines"))
 
@@
 
-(ref:pp-oceania-cap) Regression of SON mean precipitation anomalies in Oceania (mm per day, shaded) and (column a) cEOF1 the (row 1) Imaginary and (column 1) Real phase.
+(ref:pp-oceania-cap) Regression of SON mean precipitation anomalies in Oceania (mm per day, shaded) with the (row a) real and (row b) imaginary pars of the (column 1) cEOF1 and (column 2) cEOF2.
 For the 1979 -- 2019.
@@
 In black is the West Australia region, defined as the box between `r knitr::combine_words(LonLabel(box_australia$lon))` and `r knitr::combine_words(LatLabel(box_australia$lat))`.
-
-\textcolor{red}{Revisa el caption de la figura 14 que equedo un estilo distinto que los anteriores.
-Te falto aclarar también qué es la caja (¿la inventaste vos o habías utilizado la de algún otro trabajo?).
-Fijate también de aumentar la separación entre la columna izquierda y derecha para que no se superpongan los labels.
-También fijate que esta barra de colores no te quedo con el borde externo como el resto.
-}
 
@@
             fill = NA, colour = "gray20")
-```
-
-The precipitation anomalies in Oceania associated with cEOF1 are weak and not statistically significant (Figure \@ref(fig:pp-oceania) column a) whereas cEOF2 is associated with large and widespread anomalies.
+
+```
+
+The precipitation anomalies in Oceania associated with cEOF1 are weak and not statistically significant (Figure \@ref(fig:pp-oceania) column 1) whereas cEOF2 is associated with large and widespread anomalies.
 The Real cEOF2 is associated with positive seasonal anomalies in Northern and Eastern Australia, and negative ones over New Zealand, while precipitation anomalies associated with Imaginary cEOF2 are negative over all Eastern Australia.
@@
 Black line is the fit precipitation \~ sin(phase) + cos(phase) computed by weighted OLS using the magnitude of the cEOF2 as weights.
-
-\textcolor{red}{Revisa lo mismo que para la otra figura del tipo. }
 
@@
 Figure \@ref(fig:pp-america) shows regression maps of seasonal precipitation with each cEOF in South America.
-cEOF1 is associated with negative regressions coefficients for precipitation in Southern Brazil and Paraguay, although these are not statistically significant (Fig. \@ref(fig:pp-america) column a).
-The strongest precipitation regression coefccients are the ones associated with the Imaginary cEOF2.
+cEOF1 is associated with negative regressions coefficients for precipitation in Southern Brazil and Paraguay, although these are not statistically significant (Fig. \@ref(fig:pp-america) column 1).
+The strongest precipitation regression coefficients are the ones associated with the Imaginary cEOF2.
 The positive regression coefficients on Southeastern South America (SESA) and Central Chile, and negative ones over South Atlantic Convergence Zone is a well-known springtime precipitation signature of ENSO [@cai2020a] and it is also similar to the precipitation anomalies associated with the A-SAM [@campitelli2021].
@@
 These variables could be related wth a significant sinusoidal fit that has a statistically significant coefficient of determination of `r round(summary(model)$r.squared, 2)`.
-
-(ref:pp-africa-cap) Same as Figure \@ref(fig:pp-america) but for South of Africa.
-
-```{r pp-africa, fig.cap = "(ref:pp-africa-cap)"}
-# plot_pp(regr_pp, c("cEOF1", "cEOF2"), which_continent = "africa")
-```
 

### 2021-08-17 analysis/paper/shceof.Rmd
Agrega unas palabras sobre primavera

@@
 
-<!-- cambiarn nomenclatura. Imaginario / real -> pasarlo a nombrar la fase -->
-
-<!-- Chequear con ncep -->
-
-<!-- The following code chunk defines some general settings how code chunks should behave. -->
-
 ```{r setup, echo = FALSE, warning=FALSE, message=FALSE}
@@
 
-One of the main features of the zonally asymmetric circulation are the Pacific--South American Patterns [@mo2001], which are associated with precipitation anomalies in South America.
+One of the main features of the zonally asymmetric circulation are the Pacific--South American Patterns [@mo2001], which are associated with precipitation anomalies in South America, particularly in the September-October-November trimester (cita).
 Besides, @raphael2007 suggests that variability in the planetary wave 3 projected onto its climatological location is associated with Antarctic sea ice concentration anomalies.

### 2021-08-17 analysis/paper/shceof.Rmd
Pequeños cambios

@@
   comment = "#>",
-  cache.extra = 42,
+  cache.extra = 45,
   cache.path = "../cache/paper/",
@@
 
+f <- formals(geom_contour_tanaka)
+f$range <-  c(0.01, 0.2)
+formals(geom_contour_tanaka) <- f
+
 geom_contour_pval <- function(mapping, p.value = 0.01, size = 0.1, ...) {
@@
 ```{r test-irlba}
-has_irlba <- requireNamespace("rilba", quietly = TRUE)
+has_irlba <- requireNamespace("irlba", quietly = TRUE)
 if (has_irlba) {
@@
                                   signif(coef(cos_model)[1], 2), "\\sin{(\\mathrm{phase})}$"),
-                           output = "text")
+                           output = "character")
 
@@
                                   signif(coef(model)[2], 2), "\\cos{(\\mathrm{phase})}$"), 
-                           output = "text")
+                           output = "character")
 

### 2021-08-17 analysis/paper/shceof.Rmd
Mejora gráficos

@@
 
+one_less <- function(x) {
+  n <- length(x)
+  if (n < 4) return(x)
+  x <- round(x, 5)
+
+  c( rev(JumpBy(rev(x[x < 0]), 2, fill = "")),
+     JumpBy(x[x > 0], 2, fill = ""))
+}
+
+AnchorBreaks <- function (anchor = 0, binwidth = NULL, exclude = NULL, bins = 10, sym = FALSE) {
+  force(anchor)
+  force(binwidth)
+  force(exclude)
+  force(bins)
+  force(sym)
+  function(x, binwidth2 = NULL) {
+    if (sym) {
+      length <- max(abs(x - anchor))
+      x <- c(anchor - length, anchor + length)
+    }
+    if (!is.null(binwidth)) 
+      binwidth2 <- binwidth
+    if (is.null(binwidth2)) {
+      binwidth2 <- diff(pretty(x, bins))[1]
+    }
+    mult <- ceiling((x[1] - anchor)/binwidth2) - 1L
+    start <- anchor + mult * binwidth2
+    b <- seq(start, x[2] + binwidth2, binwidth2)
+    b[!(b %in% exclude)]
+  }
+}
+
 lab_lev <- AddSuffix(" hPa")
@@
 ```{r test-irlba}
-has_irlba <- requireNamespace("irlba", quietly = TRUE)
+x <- "irlba"
+has_irlba <- requireNamespace(x, quietly = TRUE)
 if (has_irlba) {
@@
 To take vertical distribution into account, we analyse both the troposphere and the stratosphere.
-We focus the analysis to the September-October-November because that's the trimester in which tropical forcing of zonal waves is maximised (cita).
-\textcolor{red}{Este parráfo me parece que quedo un poco colgado.
-Creo que sería importante que esté desde más arriba la idea de la primavera, ya que todo lo otro es muy general, pero al final el análisis solo se va a restringir a esta estación}
 
@@
 }
-```
-
-(ref:eof1-regr-gh-cap) Regression of SON geopotential height anomalies (meters) with the (column 1) real and (column 2) imaginary parts of the first cEOF for the 1979 -- 2019 period at (row a) 50 hPa and (row b) 200 hPa.
+
+mgp <- latex2exp::TeX(r'(Geopotential height ($m^2s^{-1}$))')
+```
+
+(ref:eof1-regr-gh-cap) Regression of SON geopotential height anomalies ($m^2s^{-1}$) with the (column 1) real and (column 2) imaginary parts of the first cEOF for the 1979 -- 2019 period at (row a) 50 hPa and (row b) 200 hPa.
 These coefficients come from multiple linear regression involving the real and imaginary parts.
-Areas marked with dots have p-values smaller than 0.05 adjusted for False Detection Rate.
+Areas marked with dots have p-values smaller than 0.01 adjusted for False Detection Rate.
 
@@
 plot_regr(regr , "cEOF1", c(200, 50), "hgt") +
-  scale_fill_divergent("Geopotential height", 
+  scale_fill_divergent(mgp,
                        breaks = AnchorBreaks(0, exclude = 0),
@@
 plot_regr(regr, "cEOF2", c(200, 50), "hgt") +
-  scale_fill_divergent("Geopotential height", 
+  scale_fill_divergent(mgp, 
                        breaks = AnchorBreaks(0, 200, exclude = 0),
@@
 
-(ref:sam-eof-vertical-cap) Coefficient of determination between the real and imaginary part of each cEOF and the SAM, Asymmetric SAM (A-SAM) and Symmetric SAM (S-SAM) indices computed at each level according to @campitelli2021.
+(ref:sam-eof-vertical-cap) Coefficient of determination between the real and imaginary part of each cEOF and the SAM, Asymmetric SAM (A-SAM) and Symmetric SAM (S-SAM) indices computed at each level according to @campitelli2021 for the 1979 -- 2019 period.
 Thick lines represent estimates with p-value \< 0.01 corrected for False Detection Rate [@benjamini1995].
@@
                                           labels = function(x) paste0("~", x, " km"))) +
-  scale_y_continuous(r2, limits = c(0, 1)) +
+  scale_y_continuous(r2, limits = c(0, 1), 
+                     guide = guide_axis(check.overlap = TRUE)) +
   scale_size_manual("p-value", values = c("TRUE" = 1, 
@@
 plot_regr(regr, "cEOF1", c(200, 50), "t") +
-  scale_fill_divergent("Temperature", breaks = AnchorBreaks(0, exclude = 0),
+  scale_fill_divergent("Temperature (K)", breaks = AnchorBreaks(0, exclude = 0),
                        guide = guide_colorsteps_bottom(15, even.steps = TRUE), 
@@
                     aes(z = estimate, fill = ..level..),
-                    breaks = AnchorBreaks(0, binwidth = 0.3, exclude = 0)) +
+                    breaks = AnchorBreaks(0, binwidth = 0.25, exclude = 0, sym = TRUE)) +
   geom_contour_tanaka(data = ~.x[variable == "t"],
                       aes(z = estimate),
-                      breaks = AnchorBreaks(0, binwidth = 0.3, exclude = 0)) +
+                      breaks = AnchorBreaks(0, binwidth = 0.25, exclude = 0, sym = TRUE)) +
   geom_contour2(data = ~.x[variable == "o3"], 
@@
 
-  scale_fill_divergent_discretised("Air temperature", 
+  scale_fill_divergent_discretised("Air temperature (K)", 
                                    # limits = c(-2.25, 2.25),
-                                   guide = guide_colorsteps_bottom(15)) + 
+                                   labels = one_less,
+                                   guide = guide_colorsteps_bottom(15, show.limits = TRUE)) + 
   scale_y_level(minor_breaks= NULL, 
@@
 On contours, the mean zonal anomaly of Total Ozone Column (negative contours in dashed lines, Dobson Units).
-Areas marked with dots have p-values smaller than 0.05 adjusted for False Detection Rate.
+Areas marked with dots have p-values smaller than 0.01 adjusted for False Detection Rate.
 
@@
   geom_qmap() +
-  scale_fill_divergent_discretised("Total Ozone Column", 
+  scale_fill_divergent_discretised("Total Ozone Column (DU)", 
                                    limits = c(-30, 30), 
+                                   breaks = AnchorBreaks(exclude = 0, binwidth = 5)(c(-30, 30)),
                                    guide = guide_colorsteps_bottom(15)) + 
@@
 plot_sst_psi <- function(which_cEOF, which_lev) {
-  one_less <- function(x) {
-    n <- length(x)
-    if (n < 4) return(x)
-    x <- round(x, 5)
-
-    c( rev(JumpBy(rev(x[x < 0]), 2, fill = "")),
-       JumpBy(x[x > 0], 2, fill = ""))
-  }
-
   sst_gdata <- regr_sst[cEOF == which_cEOF] %>%
@@
     geom_qmap(keep = .015) +
-    scale_fill_divergent_discretised("Streamfunction (m²/s×1e-7)", 
+    scale_fill_divergent_discretised(latex2exp::TeX(r'(Streamfunction ($m^2s^{-1}\times 1e^{-7}$))'), 
                                      labels = one_less,
@@
 
-(ref:sst-psi-2-cap) Regression maps of cEOF2 with SST (K, column a) and streamfunction zonal anomalies ($m^2/s\times10^-7$, shaded) with their corresponding activity wave flux (vectors) (column b).
+(ref:sst-psi-2-cap) Regression maps of cEOF2 with SST (K, column a) and streamfunction zonal anomalies ($m^2/s\times10^-7$, shaded) with their corresponding activity wave flux (vectors) (column b) for the 1979 -- 2019 period.
 Areas marked with dots have p-values smaller than 0.01 adjusted for FDR.
@@
 
-```{r enso-phase, fig.cap = "ONI plotted against cEOF2 phase for th 1979 -- 2019 period.
+(ref:enso-phase-cap) ONI plotted against cEOF2 phase for th 1979 -- 2019 period.
 Years with magnitude of cEOF2 greater or smaller than the 50th percentile are shown as orange circles and green diamomnds respectively. 
-Black line is the fit ONI ~ sin(phase) computed by weighted OLS using the magnitude of the cEOF2 as weights.", fig.width = width_column, fig.height=3}
+Black line is the fit ONI ~ sin(phase) computed by weighted OLS using the magnitude of the cEOF2 as weights.
+
+```{r enso-phase, fig.cap = "(ref:enso-phase-cap)", fig.width = width_column, fig.height=3}
 gdata <- ceof[, eof[[1]]$left, by = .(season)] %>%
@@
 equation <- latex2exp::TeX(paste0("$\\mathrm{ONI} = ", 
-                                  signif(coef(cos_model)[1], 2), "\\sin{(\\mathrm{phase})}$"),
-                           output = "character")
+                                  signif(coef(cos_model)[1], 2), "\\sin{(\\mathrm{phase})}$"))
 
@@
 plot_sst_psi(which_cEOF = "cEOF1", which_lev = 200) 
-
 ```
@@
     # scale_fill_divergent() +
-    scale_fill_divergent_discretised(NULL, 
+    scale_fill_divergent_discretised("Precipitation (mm/day)", 
                                      # limits = c(-.8, .8),
@@
 For the 1979 -- 2019.
-Areas marked with dots have p-values smaller than 0.05 adjusted for False Detection Rate.
+Areas marked with dots have p-values smaller than 0.01 adjusted for False Detection Rate.
 In black is the West Australia region, defined as the box between `r knitr::combine_words(LonLabel(box_australia$lon))` and `r knitr::combine_words(LatLabel(box_australia$lat))`.
@@
                                   signif(coef(model)[1], 2), "\\sin{(\\mathrm{phase})} + ", 
-                                  signif(coef(model)[2], 2), "\\cos{(\\mathrm{phase})}$"), 
-                           output = "character")
+                                  signif(coef(model)[2], 2), "\\cos{(\\mathrm{phase})}$"))
 
@@
 plot_pp(data = regr_pp, which_cEOF = c("cEOF1", "cEOF2"), which_continent = "america")  +
-  scale_fill_divergent_discretised(NULL, 
+  scale_fill_divergent_discretised("Precipitation (mm/day)", 
                                    limits = c(-1, 1),
@@
 However, this underscores the benefit of using complex EOF, since it would be trivial to rotate it.
-\textcolor{red}{creo que esta idea convendría desarrollarla mejor para el que no lo tenga tan claro}
+That is, look for the rotation of complex EOF that maximises the relationship and use that as the relevant index for study or forecast.
 

### 2021-08-17 analysis/paper/shceof.Rmd
Render

@@
   comment = "#>",
-  cache.extra = 45,
+  cache.extra = 42,
   cache.path = "../cache/paper/",

### 2021-08-17 analysis/paper/shceof.tex
Render

@@
 
-One of the main features of the zonally asymmetric circulation are the Pacific--South American Patterns \citep{mo2001}, which are associated with precipitation anomalies in South America.
+One of the main features of the zonally asymmetric circulation are the Pacific--South American Patterns \citep{mo2001}, which are associated with precipitation anomalies in South America, particularly in the September-October-November trimester (cita).
 Besides, \citet{raphael2007} suggests that variability in the planetary wave 3 projected onto its climatological location is associated with Antarctic sea ice concentration anomalies.
@@
 
-\textcolor{red}{En la figura 1 te quedaron muy chicos los subplots, amplialos un poco}.
-
 \begin{figure}
@@
 
-\textcolor{red}{Los colores de la Tabla 1 no dicen mucho en la medida que no se conoce cuál sería la paleta. Quizás se pueden remarcar en negrita ambos aclarando que son significativos (lo son?)}.
-
 \begin{table}
 
-\caption{\label{tab:corr-ceof-splitted}Coefficient of determination (\(R^2\)) of the absolute magnitude of complex EOFs between 200 hPa and 50 hPa computing EOF separatedly for each level.
-Background colour indicates the value of \(R^2\).}
+\caption{\label{tab:corr-ceof-splitted}Coefficient of determination (\(R^2\)) of the absolute magnitude of complex EOFs between 200 hPa and 50 hPa computing EOF separately for each level.
+p-values lower than 0.01 in bold.}
 \centering
@@
 \midrule
-cEOF1 & \cellcolor[HTML]{E1C0BB}{\textcolor{black}{0.28}} & \cellcolor[HTML]{FDFCFC}{\textcolor{black}{0.01}} & \cellcolor[HTML]{FDFAFA}{\textcolor{black}{0.02}}\\
-cEOF2 & \cellcolor[HTML]{FFFFFF}{\textcolor{black}{0.00}} & \cellcolor[HTML]{BB7A73}{\textcolor{white}{0.60}} & \cellcolor[HTML]{FDFAFA}{\textcolor{black}{0.02}}\\
-cEOF3 & \cellcolor[HTML]{FFFFFF}{\textcolor{black}{0.00}} & \cellcolor[HTML]{FFFFFF}{\textcolor{black}{0.00}} & \cellcolor[HTML]{FDFAFA}{\textcolor{black}{0.02}}\\
+cEOF1 & \textbf{0.28} & 0.01 & 0.02\\
+cEOF2 & 0.00 & \textbf{0.60} & 0.02\\
+cEOF3 & 0.00 & 0.00 & 0.02\\
 \bottomrule
@@
 
-\textcolor{red}{tengo la impresión de que en la figura 2 podría estar un poco más grandes los subplots para que se vean más fácilmente los detalles}
-
 The cEOF1 (Fig.~\ref{fig:ceofs-1} column~a) is a hemispheric wave 1 pattern with maximum amplitude at high latitudes.
@@
 
-\textcolor{red}{se me vino una duda, ¿se pueden comparar de algún modo las amplitudes que mostras en esta figura?
-en caso de que si, creo que habría que mencionar algo sobre lo que pasa con el cEOF2 en la estratósfera} All cEOFs have non zero mean (Fig.
-\ref{fig:extended-series}) because the geopotential fields that enter into the cEOFs algorithm are anomalies with respect to the zonal mean instead of the time mean.
+All cEOFs have non zero mean (Fig. \ref{fig:extended-series}) because the geopotential fields that enter into the cEOFs algorithm are anomalies with respect to the zonal mean instead of the time mean.
 However, cEOF2's mean is almost zero, which indicates that only cEOF1 includes variability that significantly projects onto the mean zonally anomalous field.
@@
 \includegraphics{../figures/eof1-regr-gh-1.pdf}
-\caption{\label{fig:eof1-regr-gh}Regression of SON geopotential height anomalies (meters) with the (column 1) real and (column 2) imaginary parts of the first cEOF for the 1979 -- 2019 period at (row a) 50 hPa and (row b) 200 hPa.
+\caption{\label{fig:eof1-regr-gh}Regression of SON geopotential height anomalies (\(m^2s^{-1}\)) with the (column 1) real and (column 2) imaginary parts of the first cEOF for the 1979 -- 2019 period at (row a) 50 hPa and (row b) 200 hPa.
 These coefficients come from multiple linear regression involving the real and imaginary parts.
-Areas marked with dots have p-values smaller than 0.05 adjusted for False Detection Rate.}
+Areas marked with dots have p-values smaller than 0.01 adjusted for False Detection Rate.}
 \end{figure}
@@
 Instead of a clear wave 1 pattern, the regression pattern can be better describes as a monopole with its centre displaced from the South Pole.
-
-\textcolor{red}{La figura 4 la veo muy saturada con los puntos que indican la significancia.
-¿se podrán hacer con un tamaño menor?
-¿Las unidades son correctas?
-Me llama la atención que te puedan dar tantos cientos de metros de altura geopotencial.}
 
@@
 
-\textcolor{red}{Figura 5. Me resulto raro que vaya hasta 999 en vez de 1000 }
-
 \hypertarget{other-variables}{%
@@
 \includegraphics{../figures/sam-eof-vertical-1.pdf}
-\caption{\label{fig:sam-eof-vertical}Coefficient of determination between the real and imaginary part of each cEOF and the SAM, Asymmetric SAM (A-SAM) and Symmetric SAM (S-SAM) indices computed at each level according to \citet{campitelli2021}.
+\caption{\label{fig:sam-eof-vertical}Coefficient of determination between the real and imaginary part of each cEOF and the SAM, Asymmetric SAM (A-SAM) and Symmetric SAM (S-SAM) indices computed at each level according to \citet{campitelli2021} for the 1979 -- 2019 period.
 Thick lines represent estimates with p-value \textless{} 0.01 corrected for False Detection Rate \citep{benjamini1995}.}
@@
 Correlations between SAM and the Real cEOF2 are non-significant.
-
-\textcolor{red}{Fig.
-6.
-Te quedan muy pegados los subplots de la columna izquierda con los de la derecha.
-Recomendaría aumentar el espacio intermedio y repetir el eje y en las figuras de la derecha.
-Además podrías agregar el y-label (Level) a la izquierda.}
 
@@
 
-\textcolor{red}{Tabla 2.
-Mismo comentario que antes con lo del background color (no dice mucho si no se conoce la escala), quizás podes destacar en negritas según la significancia.
-¿Está bien el intervalo de confianza de PSA2 vs cEOF2 Im?
-Hay algo raro ahí}
-
 \begin{table}
@@
 95\% confidence intervals in parenthesis.
-Background colour indicates the value of correlation.}
+p-values lower than 0.01 in bold.}
 \centering
@@
 \midrule
-PSA1 & \cellcolor[HTML]{C3BBDE}{\textcolor{black}{-0.32 (CI: -0.57 -- -0.015)}} & \cellcolor[HTML]{6259AA}{\textcolor{white}{-0.82 (CI: -0.9 -- -0.69)}}\\
-PSA2 & \cellcolor[HTML]{9E4E47}{\textcolor{white}{0.81 (CI: 0.67 -- 0.9)}} & \cellcolor[HTML]{F3F2F8}{\textcolor{black}{-0.062 (CI: -0.36 -- 0.25)}}\\
+PSA1 & -0.32 (CI: -0.57 -- -0.02) & \textbf{-0.82 (CI: -0.9 -- -0.69)}\\
+PSA2 & \textbf{0.81 (CI: 0.67 -- 0.9)} & -0.06 (CI: -0.36 -- 0.25)\\
 \bottomrule
@@
 On contours, the mean zonal anomaly of Total Ozone Column (negative contours in dashed lines, Dobson Units).
-Areas marked with dots have p-values smaller than 0.05 adjusted for False Detection Rate.}
+Areas marked with dots have p-values smaller than 0.01 adjusted for False Detection Rate.}
 \end{figure}
@@
 \begin{figure*}
-\includegraphics{../figures/sst-psi-2-1} \caption{Regression maps of cEOF2 with SST (K, column a) and streamfunction zonal anomalies (\(m^2/s\times10^-7\), shaded) with their corresponding activity wave flux (vectors) (column b).
+\includegraphics{../figures/sst-psi-2-1} \caption{Regression maps of cEOF2 with SST (K, column a) and streamfunction zonal anomalies (\(m^2/s\times10^-7\), shaded) with their corresponding activity wave flux (vectors) (column b) for the 1979 -- 2019 period.
 Areas marked with dots have p-values smaller than 0.01 adjusted for FDR.}\label{fig:sst-psi-2}
@@
 However, Figure~\ref{fig:sst-psi-2}.a1 and b1 show that the Real cEOF2 is not associated either with any significant SST nor streamfunction anomalies in the tropics.
-The correlation between the Real cEOF2 and ENSO is also not significant (0.00025 (CI: -0.31 -- 0.31)).
-\textcolor{red}{Usaría notación científica por la cantidad de 0 que tiene}.
+The correlation between the Real cEOF2 and ENSO is also not significant (0 (CI: -0.31 -- 0.31)).
 This lack of tropical signal suggest that the Real cEOF2 and the Imaginary cEOF2 have very different sources of variability.
@@
 
-\textcolor{red}{Fig. 12 Revisa cómo te quedo abajo de la figura la indicación de qué son cada uno de los puntos (te quedaron dos circulos y una línea extra que no va)}
-
 Furthermore, Figure~\ref{fig:enso-phase} suggest that highest amplitude cEOF2 years tend to coincide with strong ENSO years.
@@
 This relationship, however, appears to be driven only by the three years with strongest ENSO events in the period (2015, 1997, and 1982) which also coincide with the three years with strongest cEOF2 magnitude.
-If those years are removed, the correlation becomes non-significant (0.047 (CI: -0.28 -- 0.36)).
+If those years are removed, the correlation becomes non-significant (0.05 (CI: -0.28 -- 0.36)).
 Furthermore, even when using all the datapoints, the Spearman correlation --which is robust to outliers-- is also non-significant (0.2, p-value= 0.21).
@@
 
-\textcolor{red}{Fig. 13 Tene en cuenta los mismos comentarios que en el figura 11 y revisa también la barra de colores para SST que quedo rara}
-
 \hypertarget{precipitation}{%
@@
 
-\textcolor{red}{Revisa el caption de la figura 14 que equedo un estilo distinto que los anteriores.
-Te falto aclarar también qué es la caja (¿la inventaste vos o habías utilizado la de algún otro trabajo?).
-Fijate también de aumentar la separación entre la columna izquierda y derecha para que no se superpongan los labels.
-También fijate que esta barra de colores no te quedo con el borde externo como el resto.
-}
-
 \begin{figure}
@@
 \includegraphics{../figures/pp-oceania-1.pdf}
-\caption{\label{fig:pp-oceania}Regression of SON mean precipitation anomalies in Oceania (mm per day, shaded) and (column a) cEOF1 the (row 1) Imaginary and (column 1) Real phase.
+\caption{\label{fig:pp-oceania}Regression of SON mean precipitation anomalies in Oceania (mm per day, shaded) with the (row a) real and (row b) imaginary pars of the (column 1) cEOF1 and (column 2) cEOF2.
 For the 1979 -- 2019.
-Areas marked with dots have p-values smaller than 0.05 adjusted for False Detection Rate.
+Areas marked with dots have p-values smaller than 0.01 adjusted for False Detection Rate.
 In black is the West Australia region, defined as the box between 130°E and 155°E and 40°S and 15°S.}
@@
 
-The precipitation anomalies in Oceania associated with cEOF1 are weak and not statistically significant (Figure~\ref{fig:pp-oceania} column a) whereas cEOF2 is associated with large and widespread anomalies.
+The precipitation anomalies in Oceania associated with cEOF1 are weak and not statistically significant (Figure~\ref{fig:pp-oceania} column 1) whereas cEOF2 is associated with large and widespread anomalies.
 The Real cEOF2 is associated with positive seasonal anomalies in Northern and Eastern Australia, and negative ones over New Zealand, while precipitation anomalies associated with Imaginary cEOF2 are negative over all Eastern Australia.
@@
 \textcolor{red}{Acá te faltaría  sacar un poco más el jugo asociando con la figura 11. Además, ¿revisaste si podes también usar algún resultado asociado a los PSA? }
-
-\textcolor{red}{Revisa lo mismo que para la otra figura del tipo. }
 
@@
 Figure~\ref{fig:pp-america} shows regression maps of seasonal precipitation with each cEOF in South America.
-cEOF1 is associated with negative regressions coefficients for precipitation in Southern Brazil and Paraguay, although these are not statistically significant (Fig.~\ref{fig:pp-america} column a).
-The strongest precipitation regression coefccients are the ones associated with the Imaginary cEOF2.
+cEOF1 is associated with negative regressions coefficients for precipitation in Southern Brazil and Paraguay, although these are not statistically significant (Fig.~\ref{fig:pp-america} column 1).
+The strongest precipitation regression coefficients are the ones associated with the Imaginary cEOF2.
 The positive regression coefficients on Southeastern South America (SESA) and Central Chile, and negative ones over South Atlantic Convergence Zone is a well-known springtime precipitation signature of ENSO \citep{cai2020a} and it is also similar to the precipitation anomalies associated with the A-SAM \citep{campitelli2021}.

### 2021-08-26 analysis/paper/shceof.Rmd
Muchos cambios

@@
 
-coord_polar <- function(ymax = -20) {
+coord_polar <- function(ymax = -20, ...) {
 
@@
 
-  cbind(x, y) |> 
-    list() |> 
-    sf::st_polygon() |> 
+  cbind(x, y) %>% 
+    list() %>% 
+    sf::st_polygon() %>% 
     sf::st_sfc(crs = "+proj=latlong") -> white
@@
              default_crs = "+proj=longlat",
-             label_axes =  "----")
+             label_axes =  "----", ...)
   )
@@
 ```{r read-data}
-era5 <- ReadNetCDF(ERA5(), vars = c(hgt = "z",
-                                    t = "t"),
-                   subset = list(level = list(850, 700, 200, 50),
+era5 <- ERA5_geopotential() %>% 
+  ReadNetCDF(vars = c(hgt = "z"),
+             subset = list(level = list(200, 50),
                            latitude = -90:10)) %>% 
@@
   .[, w := season_weights(time)] %>% 
-  .[, .(hgt = mean(hgt*w), 
-        t = mean(t*w)), by = .(lev, lon, lat, time = seasonally(time))]
+  .[, .(hgt = mean(hgt*w)), by = .(lev, lon, lat, time = seasonally(time))]
 ```
@@
 scaling <- c(add = -2.364646e+06, mult = 3.954828e+13)
-
-psi <- ReadNetCDF(ERA5(), vars = "vo",
-                  subset = list(level = c(50, 200),
-                                time = c("1979-01-01", "2019-12-01"))) %>%
+psi <- ERA5_vorticity() %>% 
+ ReadNetCDF(vars = "vo") %>%
   normalise_coords() %>%
+  .[, lev := 200] %>% 
   .[season(time) == "SON"] %>%
@@
 ```{r o3}
-o3 <- O3() %>% 
+o3 <- ERA5_TOC() %>% 
   ReadNetCDF(vars = c(toc = "tco3")) %>% 
   normalise_coords() %>% 
-  .[, lon := ConvertLongitude(lon)] %>% 
   .[lat <= -20] %>% 
@@
 ```{r psa}
-psa <- ReadNetCDF(ERA5(), vars = c(hgt = "z"),
+psa <- ERA5_geopotential() %>% 
+  ReadNetCDF(vars = c(hgt = "z"),
              subset = list(level = list(500),
@@
 The SAM represents a relatively zonally symmetric pattern of alternating low pressures in polar latitude and a ring of high pressures in high latitudes with an embedded wave 3 pattern that is more prominent in the Pacific sector.
-The PSA1 and PSA2 describe the two out-of-phase sides of a meridionally propagating wave train that originates in the Eastern equatorial Pacific and travels towards the South Atlantic following a great-circle arch along the Antarctic Peninsula.
+The PSA1 and PSA2 describe the two out-of-phase sides of a meridionally propagating wave train that originates in the eastern equatorial Pacific and travels towards the South Atlantic following a great-circle arch along the Antarctic Peninsula.
 These patterns are derived by applying EOF to temporal anomalies, but @raphael2003 applied EOF methods specifically to zonal anomalies.
@@
 invisible(
-  ReadNetCDF(ERA5(), vars = c(hgt = "z"),
+  ERA5_geopotential() %>% 
+    ReadNetCDF(vars = c(hgt = "z"),
                subset = list(latitude = -90:10,
@@
 table2 <- table %>%
-  widyr::pairwise_cor(item, time, mag) |> 
+  widyr::pairwise_cor(item, time, mag) %>% 
   setDT() %>%
@@
                            level = list(200, 50))),
-  ReadNetCDF(ERA5(), vars = c(hgt = "z"),
+  ReadNetCDF(ERA5_geopotential(), vars = c(hgt = "z"),
              subset = list(latitude = range(pattern$lat),
@@
 ```{r compute-regression}
+
+era5 <- ERA5_temperature() %>% 
+  ReadNetCDF(vars = c("t"), 
+             subset = list(level = list(50, 200))) %>% 
+  normalise_coords() %>% 
+  .[season(time) == "SON"] %>%
+  .[, w := season_weights(time)] %>% 
+  .[, .(t = mean(t*w)), by = .(lev, lon, lat, time = seasonally(time))] %>% 
+  .[era5, on = .NATURAL]
+
 regr <- ceof[, eof[[1]]$left, by = .(season)] %>%
@@
     by = .(season, cEOF)] %>%
-  # dcast(time ~ cEOF, value.var = c("Real", "Imaginary")) %>%
   .[melt(era5, id.vars = c("time", "lon", "lat", "lev")), on = "time", allow.cartesian = TRUE] %>%
-  .[lev %in% c(850, 700, 200, 50)] %>%
-  # .[variable == "hgt"] %>%
   .[, FitLm(value, Imaginary, Real, se = TRUE),
     by = .(lev, lon, lat, variable, cEOF)] %>%
-  # .[, c("term", "cEOF") := tstrsplit(term, "_")] %>%
   rm_intercept() %>%
@@
 
-mgp <- latex2exp::TeX(r'(Geopotential height ($m^2s^{-1}$))')
+mgp <- latex2exp::TeX('Geopotential height ($m^2s^{-1}$)')
 ```
@@
 At 50 hPa (Figure \@ref(fig:eof1-regr-gh) row a), both the Real and Imaginary cEOF1 are associated with planetary wave 1 patterns, that are 90º out of phase.
-Their phases coincide with the ones shown in Figure \@ref(fig:ceofs-1)a.1, with the positive centre of the Real cEOF1 located towards the dateline, and the one of the Imaginary cEOF1 located over Eastern Antarctica.
+Their phases coincide with the ones shown in Figure \@ref(fig:ceofs-1)a.1, with the positive centre of the Real cEOF1 located towards the dateline, and the one of the Imaginary cEOF1 located over East Antarctica.
 However, the Real cEOF1 pattern is substantially altered by the zonally symmetric circulation.
@@
   .[time, on = .NATURAL, allow.cartesian = TRUE] %>%
+  na.omit() %>% 
   .[, correlate(hgt, estimate), by = .(lev, term, part, cEOF)]
@@
 ```{r t}
-t <- ERA5() %>% 
-  ReadNetCDF(vars = c("t", hgt = "z", "o3"), 
-             subset = list(latitude = c(-90, -20))) %>% 
+t <- ERA5_temperature() %>% 
+  ReadNetCDF(vars = c("t"), 
+             subset = list(latitude = o3wave_lats)) %>% 
   normalise_coords() %>% 
   .[season(time) == "SON"] %>% 
+  .[, .(t = weighted.mean(t, cos(lat*pi/180))), by = .(lon, lev, time)]
+
+o3_temp <- ERA5_ozone() %>% 
+  ReadNetCDF(vars = c("o3"), 
+             subset = list(latitude = o3wave_lats)) %>% 
+  normalise_coords() %>% 
+  .[season(time) == "SON"] %>% 
+  .[, .(o3 = weighted.mean(o3, cos(lat*pi/180))), by = .(lon, lev, time)]
+
+t <- t[o3_temp, on = .NATURAL]
+rm(o3_temp)
+
+t <- t %>% 
   .[, w := season_weights(time)] %>% 
   .[, .(t   = mean(t*w),
-        o3  = mean(o3*w),
-        hgt = mean(hgt*w)), by = .(lev, lon, lat, time = seasonally(time))]
+        o3  = mean(o3*w)), by = .(lev, lon, time = seasonally(time))]
 ```
@@
 t %>% 
-  .[lat %between% o3wave_lats] %>% 
-  .[, .(hgt = weighted.mean(hgt, cos(lat*pi/180)),
-        o3 = weighted.mean(o3, cos(lat*pi/180)),
-        t = weighted.mean(t, cos(lat*pi/180))), by = .(lev, time, lon)] %>%  
   melt(id.vars = c("lev", "time", "lon")) %>% 
@@
   .[psi, on = "time", allow.cartesian = TRUE] %>%
+  .[lev == 200] %>% 
   .[season(time) == "SON"] %>%
@@
 ```{r plot_sst_psi}
-plot_sst_psi <- function(which_cEOF, which_lev) {
+plot_sst_psi <- function(regr_sst, regr_psi, which_cEOF, which_lev) {
   sst_gdata <- regr_sst[cEOF == which_cEOF] %>%
@@
 ```{r sst-psi-2,  fig.cap = "(ref:sst-psi-2-cap)", fig.width = 7, fig.height=4, dependson="plot_sst_psi", fig.env="figure*"}
-plot_sst_psi(which_cEOF = "cEOF2", which_lev = 200) 
+plot_sst_psi(regr_sst = regr_sst, regr_psi = regr_psi,
+             which_cEOF = "cEOF2", which_lev = 200) 
 ```
@@
 ```{r sst-psi-1, fig.cap = "(ref:sst-psi-1-cap)", fig.width = 7, fig.height=4, dependson="plot_sst_psi", fig.env = "figure*"}
-plot_sst_psi(which_cEOF = "cEOF1", which_lev = 200) 
+plot_sst_psi(regr_sst = regr_sst, regr_psi = regr_psi,
+             which_cEOF = "cEOF1", which_lev = 200) 
 ```
@@
 cmap <- ReadNetCDF(CMAP(), vars = c(pp = "precip"),
-                   subset = list(lat = -90:10)) %>%
+                   subset = list(lat = -60:10)) %>%
   normalise_coords() %>%
@@
   .[, .(pp = mean(pp*w)), by = .(lon, lat, time = seasonally(time))]
-```
-
-```{r regr_pp}
-regr_pp <- ceof[, eof[[1]]$left, by = season] %>%
-  sep_ReIm(hgt, "wide") %>%
-  .[, hgt := NULL] %>%
-  .[, `:=`(Real = Real/sd(Real), Imaginary = Imaginary/sd(Imaginary)), 
-    by = .(season, cEOF)] %>%
-  .[melt(cmap, id.vars = c("lon", "lat", "time")), on = "time", allow.cartesian = TRUE] %>%
-  .[season == "SON"] %>%
-  .[, FitLm(value, Imaginary, Real, se = TRUE), by = .(lon, lat, cEOF, variable, season)] %>%
-  rm_intercept() %>% 
-  .[, term := factor_ReIm(term)]
 ```
@@
 
+crop_continent <- function(continent) {
+  bbox <- c(xmin = min(continents[[continent]]$lon), 
+            xmax = max(continents[[continent]]$lon),
+            ymin = min(continents[[continent]]$lat),
+            ymax = max(continents[[continent]]$lat))
+
+  function(data) {
+    s2 <- sf::sf_use_s2(FALSE)
+    on.exit(sf::sf_use_s2(s2))
+    data <- sf::st_crop(data, bbox)
+    data$continent <- continent
+    data
+  }
+}
+```
+
+```{r}
+rotate_reim <- function(real, imaginary, angle = 0) {
+  z <- complex(real = real, imaginary = imaginary)
+  z <- rotate(z, angle)
+  list(real = Re(z),
+       imaginary = Im(z))
+}
+
+sst_for_regression <- ceof[, eof[[1]]$left, by = .(season)] %>%
+  .[cEOF == "cEOF2"] %>% 
+  copy() %>% 
+  sep_ReIm(hgt, "wide") %>%
+  .[, hgt := NULL] %>%
+  .[sst, on = "time", allow.cartesian = TRUE] %>%
+  .[season(time) == "SON"] %>% 
+  na.omit()
+
+angles_regr <- -c(0, 45, 90, 45+90)*pi/180
+
+sst_regr <- lapply(angles_regr, function(a) {
+  sst_for_regression %>% 
+    copy() %>% 
+    .[, c("Real", "Imaginary") := rotate_reim(Real, Imaginary, a)] %>% 
+    .[, `:=`(Real = Real/sd(Real), Imaginary = Imaginary/sd(Imaginary)), 
+      by = .(season, cEOF)] %>%
+    .[, FitLm(t, Imaginary, Real, se = TRUE), by = .(lon, lat, cEOF, season)] %>%
+    rm_intercept() %>%
+    .[term == "Real"] %>% 
+    .[, term := factor_ReIm(term)] %>% 
+    .[, angle := a]
+}) %>% 
+  rbindlist()
+
+psi_for_regression <-  ceof[, eof[[1]]$left, by = .(season)] %>%
+  .[cEOF == "cEOF2"] %>% 
+  copy() %>% 
+  sep_ReIm(hgt, "wide") %>%
+  .[, hgt := NULL] %>%
+  .[psi, on = "time", allow.cartesian = TRUE] %>%
+  .[season(time) == "SON"] %>%
+  .[lev == 200]
+
+psi_regr <- lapply(angles_regr, function(a) {
+  psi_for_regression %>% 
+    copy() %>% 
+    .[, c("Real", "Imaginary") := rotate_reim(Real, Imaginary, a)] %>% 
+    .[, `:=`(Real = Real/sd(Real), Imaginary = Imaginary/sd(Imaginary)), 
+      by = .(season, cEOF)] %>%
+    .[, FitLm(psi.z, Real, Imaginary, se = TRUE),by = .(lon, lat, lev, cEOF, season)] %>%
+    rm_intercept() %>%
+    .[term == "Real"] %>% 
+    .[, psi.z := Anomaly(estimate), by = .(cEOF, season, lev, lat)] %>%
+    .[, c("fx", "fy") := shceof::WaveFlux(.SD, p = 200), 
+      by = .(lev, cEOF, season, term)] %>%
+    .[, term := factor_ReIm(term)] %>% 
+    .[, angle := a]
+}) %>% 
+  rbindlist()
+
+pp_for_regression <- ceof[, eof[[1]]$left, by = season] %>%
+  .[cEOF == "cEOF2"] %>% 
+  copy() %>% 
+  sep_ReIm(hgt, "wide") %>%
+  .[, hgt := NULL] %>%
+  .[cmap, on = "time", allow.cartesian = TRUE] %>%
+  .[season(time) == "SON"] %>% 
+  na.omit() %>% 
+  .[, pp := pp/sd(pp), by = .(lon, lat)]
+
+pp_regr <- lapply(angles_regr, function(a) {
+  pp_for_regression %>% 
+    copy() %>% 
+    .[, c("Real", "Imaginary") := rotate_reim(Real, Imaginary, a)] %>% 
+    .[, `:=`(Real = Real/sd(Real), Imaginary = Imaginary/sd(Imaginary)), 
+      by = .(season, cEOF)] %>%
+    .[, FitLm(pp, Imaginary, Real, se = TRUE),
+      by = .(lon, lat, cEOF, season)] %>%
+    .[term == "Real"] %>% 
+    rm_intercept() %>% 
+    .[, term := factor_ReIm(term)] %>% 
+    .[, angle := a*180/pi]
+
+}) %>%
+  rbindlist()
+
+# ceof[, eof[[1]]$left, by = season] %>%
+#   .[cEOF == "cEOF2"] %>% 
+#   copy() %>% 
+#   sep_ReIm(hgt, "wide") %>%
+#   .[, hgt := NULL] %>%
+#   .[, on = "time", allow.cartesian = TRUE] %>%
+#   .[season(time) == "SON"] %>% 
+#   na.omit() %>% 
+#   .[, pp := pp/sd(pp), by = .(lon, lat)]
+
 ```
@@
 ```{r plot_pp}
-plot_pp <- function(data, which_cEOF, which_term = c("Real", "Imaginary"), which_continent = NULL) {
-  filter_continent <- function(data) {
-    if (!is.null(which_continent)) {
-      data[continent == which_continent]
-    } else {
-      data
-    }
-
-  }
+spoke <- function(data, coords) {
+  radius <-  grid::unit(.15, "snpc")
+
+  outer_circle <-  grid::circleGrob(x = 0, 
+                             y = 0,
+                             r = radius*1.2)
+
+  xcentre <- grid::unit(0, "npc") + grid:::grobHeight(outer_circle)/2
+  ycentre <-  grid::unit(0, "npc") + grid:::grobWidth(outer_circle)/2
+
+  circle <- grid::circleGrob(x = xcentre, 
+                             y = ycentre,
+                             r = radius, 
+                             gp = grid::gpar(fill = "white", 
+                                             alpha = 0.7))
+  # browser()
+  dx <- cos(coords$angle[1]*pi/180)*grid::grobHeight(circle)/2
+  dy <- sin(coords$angle[1]*pi/180)*grid::grobWidth(circle)/2
+
+  line <- grid::segmentsGrob(x0 = xcentre, y0 = ycentre, 
+                             x1 = xcentre + dx,
+                             y1 = ycentre + dy, 
+                             gp = grid::gpar(fill = "black"),
+                             arrow = grid::arrow(angle = 13, 
+                                                 length =  grid::unit(.15/2, "snpc"),
+                                                 type = "closed")
+  )
+
+  grid::gTree(children = grid::gList(circle, line))
+
+}
+
+plot_pp <- function(data, which_cEOF) {
+  lats <- c(-55, -10)
+  breaks <- AnchorBreaks(exclude = 0)(c(-.9, .9))
+
   data %>%
-    .[cEOF %in% which_cEOF] %>%
-    .[term %in% which_term] %>%
+    # .[angle == unique(angle)[2]] %>% 
+    .[lat %between% lats] %>% 
+    # .[cEOF %in% which_cEOF] %>%
+    # .[term %in% which_term] %>%
     .[, lev := "sfc"] %>%
-    .[lat >= -60] %>%
-    add_continent() %>%
-    filter_continent() %>%
-    # .[abs(estimate) <= 2] %>%
+
     .[, p.value := Pvaluate(estimate, std.error, df, "none"), 
@@
     ggplot(aes(lon, lat)) +
-    # geom_raster(aes(fill = estimate)) +
+
     geom_contour_fill(aes(z = estimate, fill = ..level..),
-                      breaks = AnchorBreaks(exclude = 0)) +
-    geom_contour_tanaka(aes(z = estimate), breaks = AnchorBreaks(exclude = 0)) +
-
-    # scale_fill_divergent() +
-    scale_fill_divergent_discretised("Precipitation (mm/day)", 
-                                     # limits = c(-.8, .8),
+                      breaks = breaks) +
+    geom_contour_tanaka(aes(z = estimate), breaks = breaks) +
+
+    scale_fill_divergent_discretised("Precipitation (correlation)", 
+                                     # limits = c(-1, 1),
                                      # oob = scales::squish,
@@
                                      guide = guide_colorsteps_bottom(15)) +
-    # scale_fill_divergent_discretised(NULL, guide = guide_colorsteps_bottom(15)) +
+
     geom_contour_pval(aes(z = p.value)) +
-    geom_qmap(keep = 1) +
+    geom_qmap() +
+
+    grid_panel(spoke, aes(angle = -angle)) +
 
@@
     scale_y_latitude(ticks = 15) +
-    ggh4x::facet_nested(term ~ cEOF) +
+    facet_grid(angle ~ cEOF) +
     axis_labs_smol +
     tag_facets("rc") +
-    coord_sf(ylim = continents[[which_continent]]$lat,
-             xlim = continents[[which_continent]]$lon,
+    coord_sf(ylim = lats, 
              crs = "+proj=longlat +pm=180 +over",
@@
 Areas marked with dots have p-values smaller than 0.01 adjusted for False Detection Rate.
-In black is the West Australia region, defined as the box between `r knitr::combine_words(LonLabel(box_australia$lon))` and `r knitr::combine_words(LatLabel(box_australia$lat))`.
+In black is the East Australia region, defined as the box between `r knitr::combine_words(LonLabel(box_australia$lon))` and `r knitr::combine_words(LatLabel(box_australia$lat))`.
+
+```{r, fig.cap="precipitation correlation", fig.width=6.6, fig.height=5.5}
+plot_pp(pp_regr, c("cEOF1", "cEOF2"))
+```
+
+```{r}
+# a <- angles_regr[2]
+# 
+# plot_pp(pp_regr, c("cEOF1", "cEOF2"))
+# 
+# (plot_sst_psi(regr_sst = sst_regr[angle == a], regr_psi = psi_regr[angle == a],
+#               which_cEOF = "cEOF2", which_lev = 200)) / 
+#   (
+#     plot_pp(pp_regr[angle == a], c("cEOF1", "cEOF2"))
+#   )
+```
 
 ```{r pp-oceania, fig.cap = "(ref:pp-oceania-cap)", fig.width=width_column, fig.height=width_column*.8}
-plot_pp(regr_pp, c("cEOF1", "cEOF2"), which_continent =  "oceania") +
-  geom_rect(data = box_australia, aes(xmin = min(lon), xmax = max(lon),
-                                      ymin = min(lat), ymax = max(lat)),
-            fill = NA, colour = "gray20")
+# plot_pp(regr_pp, c("cEOF1", "cEOF2"), which_continent =  "oceania") +
+#   geom_rect(data = box_australia, aes(xmin = min(lon), xmax = max(lon),
+#                                       ymin = min(lat), ymax = max(lat)),
+#             fill = NA, colour = "gray20")
 
@@
 The precipitation anomalies in Oceania associated with cEOF1 are weak and not statistically significant (Figure \@ref(fig:pp-oceania) column 1) whereas cEOF2 is associated with large and widespread anomalies.
-The Real cEOF2 is associated with positive seasonal anomalies in Northern and Eastern Australia, and negative ones over New Zealand, while precipitation anomalies associated with Imaginary cEOF2 are negative over all Eastern Australia.
-As it could be expected, these negative anomalies are similar to the springtime precipitation anomalies observed in relation with the Asymmetric SAM [@campitelli2021] and ENSO [@cai2011].
+The Real cEOF2 is associated with positive seasonal anomalies in northern and eastern Australia, and negative ones over New Zealand, while precipitation anomalies associated with Imaginary cEOF2 are negative over all eastern Australia.
+
+These precipitation anomalies associated with the cEOF2 are similar to the springtime precipitation anomalies observed in relation with the Asymmetric SAM [@campitelli2021] and ENSO [@cai2011],
+
 \textcolor{red}{Acá te faltaría  sacar un poco más el jugo asociando con la figura 11. Además, ¿revisaste si podes también usar algún resultado asociado a los PSA? }
@@
 
-equation <- latex2exp::TeX(paste0("$\\mathrm{Precipitation}_{\\mathrm{West Australia}} = ", 
+equation <- latex2exp::TeX(paste0("$\\mathrm{Precipitation}_{\\mathrm{East Australia}} = ", 
                                   signif(coef(model)[1], 2), "\\sin{(\\mathrm{phase})} + ", 
@@
 
-To better understand the relationship between precipitation anomalies over Eastern Australia and the cEOF2, Figure \@ref(fig:australia-pp-phase) plots precipitation anomalies averaged over Eastern Australia (box in Figure \@ref(fig:pp-oceania)) as a function of cEOF2 phase.
+To better understand the relationship between precipitation anomalies over eastern Australia and the cEOF2, Figure \@ref(fig:australia-pp-phase) plots precipitation anomalies averaged over eastern Australia (box in Figure \@ref(fig:pp-oceania)) as a function of cEOF2 phase.
 The black line shows a sinusoidal fit, whose coefficients are fitted by weighted least squares using the magnitude of the EFO2 as weight.
 The $r^2$ corresponding to the fit is `r signif(summary(model)$r.squared, 2)` (p-value `r report_p(p_value)`).
-There is a strong relationship between the phase of the cEOF2 and precipitation anomalies in Eastern Australia, but the Real and Imaginary phases we chose are not exactly aligned with the direction that maximises this relationship.
+There is a strong relationship between the phase of the cEOF2 and precipitation anomalies in eastern Australia, but the Real and Imaginary phases we chose are not exactly aligned with the direction that maximises this relationship.
 To perform a more in-depth analysis of the relationship between this pattern and precipitation in this region, it would be recommended to align one of the axis (either the Real or the Imaginary) to the phase that maximises the described relationship.
 
+@cai2011 -> precipitación ENSO
+
+ no vio relación con la pp. mezclan fases, pero la fase es crucial.
+
 (ref:pp-america-cap) As Figure \@ref(fig:pp-oceania) but for South America
@@
 ```{r pp-america, fig.cap = "(ref:pp-america-cap)", fig.width=width_column, fig.height=width_column*1.1}
-plot_pp(data = regr_pp, which_cEOF = c("cEOF1", "cEOF2"), which_continent = "america")  +
-  scale_fill_divergent_discretised("Precipitation (mm/day)", 
-                                   limits = c(-1, 1),
-                                   # oob = scales::squish,
-                                   low = "#8E521C",
-                                   high = "#00665E",
-                                   guide = guide_colorsteps_bottom(15)) 
+# plot_pp(data = regr_pp, which_cEOF = c("cEOF1", "cEOF2"), which_continent = "america")  +
+#   scale_fill_divergent_discretised("Precipitation (mm/day)", 
+#                                    limits = c(-1, 1),
+#                                    # oob = scales::squish,
+#                                    low = "#8E521C",
+#                                    high = "#00665E",
+#                                    guide = guide_colorsteps_bottom(15)) 
 # annotate("rect", xmin = min(sesa$lon), xmax = max(sesa$lon), 
@@
 This mirrors the results of @cai2002, who showed that the CSIRO Model can create PSA-like variability even in the absence of ENSO forcing (with a climatological run), but the variability of one of the PSA modes was enhanced when adding the ENSO signal.
-The magnitude of the cEOF2 appears not to be related to the magnitude of ENSO, except frot the three strongest ENSO events in the period which coincide with the three strongest cEOF2 years.
+The magnitude of the cEOF2 appears not to be related to the magnitude of ENSO, except for the three strongest ENSO events in the period which coincide with the three strongest cEOF2 years.
 
@@
 That is, look for the rotation of complex EOF that maximises the relationship and use that as the relevant index for study or forecast.
+
+Some aspects of our work are in direct contrast with @irving2016 analysis of the PSA. 
+They didn't find neither a strong association between PSA-like variability and ENSO, nor a relationship between it and Australian precipitation. 
+Part of the disagreement could be due to the fact that we focus data only for SON, which is when the teleconnection pattern is strongest, instead of the whole year. 
+@irving2016 use of ERA5 forectasted precipitation instead of the more observationally constrained CMAP dataset might also be an issue. 
+But in light of the previous discussion on the importance of the phase of the cEOF2 in its relationship with ENSO and with precipitation impacts, a big difference could be fact that @irving2016's methodology is insensitive to phase changes of the PSA-like variability.
 

### 2021-09-07 analysis/paper/shceof.Rmd
Cambia datos y hace muchos cambios al texto.

@@
   comment = "#>",
-  cache.extra = 41,
+  cache.extra = 43,
   cache.path = "../cache/paper/",
@@
         panel.grid.minor.x = element_line(linetype = 2))  +
-  tag_facets("cr") +
+  tag_facets("rc") +
   panel_background 
@@
 
-```{r regr_psi}
-regr_psi <- ceof[, eof[[1]]$left, by = .(season)] %>%
-  sep_ReIm(hgt, "wide") %>%
-  .[, hgt := NULL] %>%
-  .[, `:=`(Real = Real/sd(Real), Imaginary = Imaginary/sd(Imaginary)), 
-    by = .(season, cEOF)] %>%
-  .[psi, on = "time", allow.cartesian = TRUE] %>%
-  .[lev == 200] %>% 
-  .[season(time) == "SON"] %>%
-  .[, FitLm(psi.z, Real, Imaginary, se = TRUE),by = .(lon, lat, lev, cEOF, season)] %>%
-  rm_intercept() %>%
-  .[, psi.z := Anomaly(estimate), by = .(cEOF, season, lev, lat)] %>%
-  .[, c("fx", "fy") := shceof::WaveFlux(.SD, p = 200), by = .(lev, cEOF, season, term)] %>%
-  .[, term := factor_ReIm(term)]
-```
-
 ```{r sst}
@@
 
-```{r regr_sst}
-regr_sst <- ceof[, eof[[1]]$left, by = .(season)] %>%
-  sep_ReIm(hgt, "wide") %>%
-  .[, hgt := NULL] %>%
-  .[, `:=`(Real = Real/sd(Real), Imaginary = Imaginary/sd(Imaginary)), 
-    by = .(season, cEOF)] %>%
-  .[sst, on = "time", allow.cartesian = TRUE] %>%
-  .[season(time) == "SON"] %>%
-  .[, FitLm(t, Imaginary, Real, se = TRUE), by = .(lon, lat, cEOF, season)] %>%
-  rm_intercept() %>%
-  .[, term := factor_ReIm(term)]
-
-```
-
-```{r enso_cor}
-enso_cor <-  ceof[, eof[[1]]$left, by = .(season)] %>%
-  sep_ReIm() %>%
-  .[, hgt := scale(hgt), by = .(cEOF, part)] %>%
-  enso[., on = "time"] %>%
-  na.omit() %>%
-  .[, correlate(hgt, oni), by = .(cEOF, part)]
-
-```
-
-```{r plot_sst_psi}
-plot_sst_psi <- function(regr_sst, regr_psi, which_cEOF, which_lev) {
-  sst_gdata <- regr_sst[cEOF == which_cEOF] %>%
-    .[lat %between% c(-90, 10)] %>%
-    .[, p_val := Pvaluate(estimate, std.error, df, "fdr"), 
-      by = .(cEOF, term)] %>%
-    .[, var := "SST"]
-
-  rsst <- max(range(abs(regr_sst[!is.na(cEOF), ]$estimate), na.rm = TRUE))
-  breaks_sst <- AnchorBreaks(0, bins = 20, exclude = 0)(c(-rsst, rsst))
-
-  stream_gdata <- regr_psi %>%
-    .[cEOF == which_cEOF] %>%
-    .[lev == which_lev] %>%
-    # .[abs(estimate) > 1, estimate := NA] %>%
-    .[lat %between% c(-90, 10)] %>%
-    .[, p_val := Pvaluate(estimate, std.error, df, "fdr"), by = .(cEOF, term)] %>%
-    .[, var := "Streamfunction"]
-
-  rstream <- max(range(abs(regr_psi[lev == which_lev]$estimate), 
-                       na.rm = TRUE))*1e-7
-  breaks_stream <- AnchorBreaks(0, bins = 20, exclude = 0)(c(-rstream, rstream))
-
-  # browser()
-  stream_gdata %>% 
-    periodic(lon = c(0, 360)) %>%
-    ggplot(aes(lon, lat)) +
-    geom_contour_fill(aes(z = estimate*1e-7, fill = ..level..),
-                      breaks = breaks_stream) +
-    geom_contour_pval(aes(z = p_val)) +
-    geom_streamline(aes(dx = fx, dy = fy), min.L = 3, L = 10,
-                    size = 0.1, res = 2, nx = 22, ny = 12, 
-                    arrow.angle = 14, arrow.length = 0.2) +
-    geom_qmap(keep = .015) +
-    scale_fill_divergent_discretised(latex2exp::TeX(r'(Streamfunction ($m^2s^{-1}\times 1e^{-7}$))'), 
-                                     labels = one_less,
-                                     low = scales::muted("red"),
-                                     high = scales::muted("blue"),
-                                     guide = guide_colorsteps_bottom(15, show.limits = FALSE,
-                                                                     order = 99)) +
-    ggnewscale::new_scale_fill() +
-
-    geom_contour_fill(data = sst_gdata,
-                      aes(z = estimate, fill = ..level..), 
-                      breaks = breaks_sst) +
-    geom_contour_pval(data = sst_gdata, aes(z = p_val)) +
-    scale_fill_divergent_discretised("SST (K)", 
-                                     labels = one_less,
-                                     guide = guide_colorsteps_bottom(width = 15,
-                                                                     show.limits = FALSE, ticks = TRUE,
-                                                                     order = 1)) +
-    scale_x_continuous(NULL, expand = c(0, 0), 
-                       breaks = seq(-180, 180-60, by = 60),
-                       labels = LonLabel) +
-
-    scale_y_continuous(NULL, breaks  = seq(-90, 0, by = 30), expand = c(0, 0)) +
-    scale_mag(guide = "none") +
-    facet_grid(term ~ var) +
-    # coord_quickmap(ylim = c(NA, 10)) +
-    coord_sf(ylim = c(-89, 0), expand = FALSE,
-             crs = "+proj=longlat +pm=180 +over",
-             default_crs = "+proj=longlat +lon_wrap=180 +over") + 
-    theme(legend.box = "horizontal",
-          legend.text = element_text(size = rel(0.7)), 
-          legend.box.spacing = grid::unit(0, "lines") ) +
-    tag_facets("cr", position = "bl") 
-}
-
-```
-
-(ref:sst-psi-2-cap) Regression maps of cEOF2 with SST (K, column a) and streamfunction zonal anomalies ($m^2/s\times10^-7$, shaded) with their corresponding activity wave flux (vectors) (column b) for the 1979 -- 2019 period.
-Areas marked with dots have p-values smaller than 0.01 adjusted for FDR.
-
-```{r sst-psi-2,  fig.cap = "(ref:sst-psi-2-cap)", fig.width = 7, fig.height=4, dependson="plot_sst_psi", fig.env="figure*"}
-plot_sst_psi(regr_sst = regr_sst, regr_psi = regr_psi,
-             which_cEOF = "cEOF2", which_lev = 200) 
-```
-
-The connections between cEOFs and Tropical sources are also assessed.
-Figure \@ref(fig:sst-psi-2) shows the regression maps of Sea Surface Temperatures (SST) and Streamfunction anomalies at 200 hPa respectively upon cEOF2.
-The Imaginary cEOF2 is associated with strong positive SST anomalies on the Central Pacific and negative anomalies over an area across the North of Australia and New Zealand, the South Pacific Convergence Zone (SPCZ) (Figure \@ref(fig:sst-psi-2).a2).
-This pattern is almost canonically positive ENSO [@bamston1997] and indeed, the correlation between the Imaginary cEOF2 and the Oceanic Niño Index [@bamston1997] is significant and very high `r enso_cor[cEOF == "cEOF2" & part == "Imaginary"]$text`.
-Besides, the Imaginary cEOF2 is associated with strong wave-like streamfunction anomalies emanating from the tropics (Figure \@ref(fig:sst-psi-2).b2).
-This is consistent with the effect of ENSO on the extratropics: SST anomalies initiate anomalous tropical convection that excites Rossby waves that propagate meridionally towards higher latitudes [@mo2000].
-
-Since the Real cEOF2 represents just a different phase of the same wave train, it would be expected to show a similar forcing pattern to the Imaginary cEOF2 with a slight translation of its location.
-However, Figure \@ref(fig:sst-psi-2).a1 and b1 show that the Real cEOF2 is not associated either with any significant SST nor streamfunction anomalies in the tropics.
-The correlation between the Real cEOF2 and ENSO is also not significant (`r enso_cor[cEOF == "cEOF2" & part == "Real"]$text`).
-This lack of tropical signal suggest that the Real cEOF2 and the Imaginary cEOF2 have very different sources of variability.
-
-(ref:enso-phase-cap) ONI plotted against cEOF2 phase for th 1979 -- 2019 period.
-Years with magnitude of cEOF2 greater or smaller than the 50th percentile are shown as orange circles and green diamomnds respectively. 
-Black line is the fit ONI ~ sin(phase) computed by weighted OLS using the magnitude of the cEOF2 as weights.
-
-```{r enso-phase, fig.cap = "(ref:enso-phase-cap)", fig.width = width_column, fig.height=3}
-gdata <- ceof[, eof[[1]]$left, by = .(season)] %>%
-  .[cEOF == "cEOF2"] %>% 
-  # .[, hgt := Anomaly(hgt)] %>% 
-  .[, arg := Arg(hgt)] %>%
-  .[, mag := Mod(hgt)] %>%
-  .[enso, on = "time"] %>%
-  na.omit() %>%
-  .[, q := ifelse(mag >= quantile(mag, 0.5), "|cEOF2| > 50%", "|cEOF2| < 50%")]
-
-cor_enso_mag <- gdata[, correlate(mag, abs(oni))]
-
-cor_enso_spearman <- gdata[, cor.test(mag, abs(oni), method = "spearman")]
-
-cor_enso_mag_outlier <- gdata[oni < 1.5][, correlate(mag, abs(oni))]
-
-cor_enso_mag_quantile <- gdata[mag >= quantile(mag, 0.5)] %>% 
-  .[oni < 1.5] %>% 
-  .[, correlate(mag, abs(oni))]
-
-cos_model <- gdata %>% 
-  .[, lm(oni ~ I(sin(arg)) - 1, weights = mag )]
-
-equation <- latex2exp::TeX(paste0("$\\mathrm{ONI} = ", 
-                                  signif(coef(cos_model)[1], 2), "\\sin{(\\mathrm{phase})}$"))
-
-gdata %>% 
-  ggplot(aes(arg, oni)) +
-  geom_hline(yintercept = c(-0.5, 0, 0.5), size = 0.2, color = "gray50") +
-  geom_vline(xintercept = 0, size = 0.2, color = "gray50") +
-  annotate(shadowtext:::GeomShadowText,
-           label = c("- Real", "- Imaginary", "+ Real", "+ Imaginary", "- Real"),
-           color = "gray60", bg.colour = "white", bg.r = 0.2,
-           size = 3,
-           y = -Inf, x = c(-pi, -pi/2, 0, pi/2, pi), angle = 90, hjust = 0) +
-
-  geom_smooth(data = ~qwrap(.x, arg = c(-pi, pi) ~ c(-2*pi, 2*pi)),
-              method = "lm", aes(weight = mag),
-              formula = y ~ I(sin(x)) -1,
-              se = FALSE, color = "black", size = 0.5) +
-  geom_point(aes(color = q,
-                 shape  = q),
-             size = 3) +
-
-  scale_color_brewer("", palette = "Dark2") +
-  scale_shape_manual("", values = c(20, 18)) +
-  scale_y_continuous("ONI") +
-  scale_x_continuous("Phase",
-                     breaks = seq(-pi, pi, by = 45*pi/180),
-                     labels = function(x) paste0(round(x*180/pi, 5), "º")) +
-  coord_cartesian(xlim = c(-pi, pi)) +
-  labs(subtitle = equation) +
-  panel_background
-```
-
-To better explore the relationship between tropical forcing and phase of the cEOF2, Figure \@ref(fig:enso-phase) shows the ONI index against the cEOF2 phase for each SON trimester between 1979 and 2019, highlighting years in which the magnitude of cEOF2 is above the median.
-In years with positive ONI, the phase of the cEOF2 is mostly around +90º (corresponding with positive imaginary part) and vice versa.
-In years with near neutral ENSO, the phase of the cEOF2 is much more variable.
-The black line in Figure \@ref(fig:enso-phase) is a sinusoidal fit of the relationship between ONI and cEOF2 phase.
-The $r^2$ corresponding to the fit is `r signif(summary(cos_model)$r.squared, 2)`, statistically significant with p-value `r report_p(coef(summary(cos_model))[4])`, indicating a quasi-sinusoidal relation between these two variables.
-
-Furthermore, Figure \@ref(fig:enso-phase) suggest that highest amplitude cEOF2 years tend to coincide with strong ENSO years.
-The correlation between the absolute magnitude of the ONI and the magnitude of the cEOF2 is `r cor_enso_mag$text`.
-This relationship, however, appears to be driven only by the three years with strongest ENSO events in the period (`r knitr::combine_words(year(gdata[order(-oni)][1:3]$time))`) which also coincide with the three years with strongest cEOF2 magnitude.
-If those years are removed, the correlation becomes non-significant (`r cor_enso_mag_outlier$text`).
-Furthermore, even when using all the datapoints, the Spearman correlation --which is robust to outliers-- is also non-significant (`r signif(cor_enso_spearman$estimate, 2)`, p-value`r report_p(cor_enso_spearman$p.value)`).
-Therefore, the relationship between the magnitude of the cEOF2 train wave and ONI remains uncertain.
-
-```{r}
-# gdata %>% 
-#   ggplot(aes(oni, mag)) +
-#   geom_point(aes(color = mag >= quantile(mag, 0.5)))  +
-#    scale_color_brewer(NULL, palette = "Dark2",
-#                      labels = c("TRUE"  = "|cEOF2| > 50%",
-#                                 "FALSE" = "|cEOF2| < 50%")) +
-#   panel_background
-```
-
-It could be concluded that the wave train represented by cEOF2 can be both part of the internal variability of the extratropical atmosphere or forced by tropical SSTs.
-In the former case, the wave train has little phase preference.
-However, when cEOF2 is excited by tropical SST variability, it tends to remain locked to the imaginary phase.
-This explains the relative overabundance of years with cEOF2 near positive and negative imaginary phase in Figure \@ref(fig:phase-histogram).
-
-(ref:sst-psi-1-cap) Same as Figure \@ref(fig:sst-psi-2) but for cEOF1.
-
-```{r sst-psi-1, fig.cap = "(ref:sst-psi-1-cap)", fig.width = 7, fig.height=4, dependson="plot_sst_psi", fig.env = "figure*"}
-plot_sst_psi(regr_sst = regr_sst, regr_psi = regr_psi,
-             which_cEOF = "cEOF1", which_lev = 200) 
-```
-
-Figure \@ref(fig:sst-psi-1) shows SST and streamfunction regression maps for cEOF1.
-Unlike the cEOF2 case, there is no significant pattern of SST anomalies associated with either the Real or Imaginary cEOF1.
-Consistently, streamfunction anomalies do not show any tropical influence.
-Instead, the real and Imaginary cEOF1 are associated with zonally wave activity fluxes in the extra-tropics around 60ºS, except for an equatorward flow from the coast of Antarctica around 150ºE in the real component.
-This suggests that it's variability is driven primary by the internal variability of the extra-tropics.
-
-## Precipitation {#precipitation}
-
 ```{r cmap}
@@
 
-```{r add_continents}
-continents <- list(africa  = list(lon = c(0, 60),
-                                  lat = c(-40, 0)),
-                   oceania = list(lon = c(100, 180),
-                                  lat = c(-50, -13)),
-                   america = list(lon = c(265, 325),
-                                  lat = c(-60, -10)))
-
-add_continent <- function(dt) {
-  dt[, continent := fcase(lon %between% continents$africa$lon &
-                            lat %between% continents$africa$lat, "africa",
-                          lon %between% continents$oceania$lon &
-                            lat %between% continents$oceania$lat, "oceania",
-                          lon %between% continents$america$lon &
-                            lat %between% continents$america$lat, "america")] %>%
-    .[, continent := factor(continent, levels = c("africa", "oceania", "america"), ordered = TRUE)] %>%
-    .[]
-}
-
-crop_continent <- function(continent) {
-  bbox <- c(xmin = min(continents[[continent]]$lon), 
-            xmax = max(continents[[continent]]$lon),
-            ymin = min(continents[[continent]]$lat),
-            ymax = max(continents[[continent]]$lat))
-
-  function(data) {
-    s2 <- sf::sf_use_s2(FALSE)
-    on.exit(sf::sf_use_s2(s2))
-    data <- sf::st_crop(data, bbox)
-    data$continent <- continent
-    data
-  }
-}
-```
-
-```{r}
+```{r t2m}
+t2m <- ERA5_T2M() %>% 
+  ReadNetCDF() %>% 
+   normalise_coords() %>%
+  .[, w := season_weights(time)] %>% 
+  .[, .(t2m = mean(t2m*w)), by = .(lon, lat, time = seasonally(time))]
+```
+
+```{r rotated-regressions} 
 rotate_reim <- function(real, imaginary, angle = 0) {
@@
 
+angles_regr <- -c(0, 45, 90, 45+90)*pi/180
+angles_labs <- setNames(c("Real", "45º", "Imaginary", "135º"), angles_regr*180/pi)
+
+hgt850 <- ERA5_geopotential() %>% 
+  ReadNetCDF(vars = c(hgt = "z"), 
+             subset = list(level = 850)) %>% 
+   normalise_coords() %>%
+  .[season(time) == "SON"] %>% 
+  .[, w := season_weights(time)] %>% 
+  .[, .(hgt = mean(hgt*w)), by = .(lon, lat, time = seasonally(time))]
+
+hgt850 <- ceof[, eof[[1]]$left, by = .(season)] %>%
+  .[season == "SON"] %>%
+  sep_ReIm(format = "wide") %>%
+  .[, hgt := NULL] %>%
+  .[, `:=`(Real = as.numeric(scale(Real)), Imaginary = as.numeric(scale(Imaginary))),
+    by = .(season, cEOF)] %>%
+  .[hgt850, on = "time", allow.cartesian = TRUE]
+
+hgt850_regr <- lapply(angles_regr, function(a) {
+  hgt850 %>% 
+    copy() %>% 
+    .[, c("Real", "Imaginary") := rotate_reim(Real, Imaginary, a)] %>% 
+    .[, `:=`(Real = Real/sd(Real), Imaginary = Imaginary/sd(Imaginary)), 
+      by = .(season, cEOF)] %>%
+    .[, FitLm(hgt, Imaginary, Real, se = TRUE), by = .(lon, lat, cEOF, season)] %>%
+    rm_intercept() %>%
+    .[term == "Real"] %>% 
+    .[, term := factor_ReIm(term)] %>% 
+    .[, angle := a*180/pi]
+}) %>% 
+  rbindlist()
+
 sst_for_regression <- ceof[, eof[[1]]$left, by = .(season)] %>%
-  .[cEOF == "cEOF2"] %>% 
+  # .[cEOF == "cEOF2"] %>% 
   copy() %>% 
@@
   na.omit()
-
-angles_regr <- -c(0, 45, 90, 45+90)*pi/180
 
@@
     .[, term := factor_ReIm(term)] %>% 
-    .[, angle := a]
+    .[, angle := a*180/pi]
 }) %>% 
@@
 psi_for_regression <-  ceof[, eof[[1]]$left, by = .(season)] %>%
-  .[cEOF == "cEOF2"] %>% 
+  # .[cEOF == "cEOF2"] %>% 
   copy() %>% 
@@
 pp_for_regression <- ceof[, eof[[1]]$left, by = season] %>%
-  .[cEOF == "cEOF2"] %>% 
+  # .[cEOF == "cEOF2"] %>% 
   copy() %>% 
@@
 
-# ceof[, eof[[1]]$left, by = season] %>%
+t2m_for_regression <-  ceof[, eof[[1]]$left, by = season] %>%
   # .[cEOF == "cEOF2"] %>% 
-#   copy() %>% 
-#   sep_ReIm(hgt, "wide") %>%
-#   .[, hgt := NULL] %>%
-#   .[, on = "time", allow.cartesian = TRUE] %>%
-#   .[season(time) == "SON"] %>% 
-#   na.omit() %>% 
-#   .[, pp := pp/sd(pp), by = .(lon, lat)]
-
-```
-
-```{r plot_pp}
+  copy() %>% 
+  sep_ReIm(hgt, "wide") %>%
+  .[, hgt := NULL] %>%
+  .[t2m, on = "time", allow.cartesian = TRUE] %>%
+  .[season(time) == "SON"] %>% 
+  na.omit()
+
+t2m_regr <- lapply(angles_regr, function(a) {
+  t2m_for_regression %>% 
+    copy() %>% 
+    .[, c("Real", "Imaginary") := rotate_reim(Real, Imaginary, a)] %>% 
+    .[, `:=`(Real = Real/sd(Real), Imaginary = Imaginary/sd(Imaginary)), 
+      by = .(season, cEOF)] %>%
+    .[, FitLm(t2m, Imaginary, Real, se = TRUE),
+      by = .(lon, lat, cEOF, season)] %>%
+    .[term == "Real"] %>% 
+    rm_intercept() %>% 
+    .[, term := factor_ReIm(term)] %>% 
+    .[, angle := a*180/pi]
+
+}) %>%
+  rbindlist()
+```
+
+```{r enso_cor}
+enso_cor <-  ceof[, eof[[1]]$left, by = .(season)] %>%
+  sep_ReIm() %>%
+  .[, hgt := scale(hgt), by = .(cEOF, part)] %>%
+  enso[., on = "time"] %>%
+  na.omit() %>%
+  .[, correlate(hgt, oni), by = .(cEOF, part)]
+```
+
+```{r spoke}
 spoke <- function(data, coords) {
@@
 }
-
-plot_pp <- function(data, which_cEOF) {
-  lats <- c(-55, -10)
-  breaks <- AnchorBreaks(exclude = 0)(c(-.9, .9))
-
-  data %>%
-    # .[angle == unique(angle)[2]] %>% 
-    .[lat %between% lats] %>% 
-    # .[cEOF %in% which_cEOF] %>%
-    # .[term %in% which_term] %>%
-    .[, lev := "sfc"] %>%
-
-    .[, p.value := Pvaluate(estimate, std.error, df, "none"), 
-      by = .(lev, cEOF, term)] %>%
+```
+
+```{r plot_sst_psi}
+plot_sst_psi <- function(sst_regr, psi_regr) {
+  sst_gdata <- sst_regr %>%
+    # .[cEOF == "cEOF2"] %>% 
+    .[lat %between% c(-90, 10)] %>%
+    .[, p_val := Pvaluate(estimate, std.error, df, "fdr"), 
+      by = .(cEOF, term, angle)] %>%
+    .[, var := "SST"]
+
+  rsst <- max(range(abs(sst_regr[!is.na(cEOF), ]$estimate), na.rm = TRUE))
+  breaks_sst <- AnchorBreaks(0, bins = 20, exclude = 0)(c(-rsst, rsst))
+
+  stream_gdata <- psi_regr %>%
+    # .[cEOF == "cEOF2"] %>% 
+    # .[abs(estimate) > 1, estimate := NA] %>%
+    .[lat %between% c(-90, 10)] %>%
+    .[, p_val := Pvaluate(estimate, std.error, df, "fdr"), by = .(cEOF, angle, term)] %>%
+    .[, var := "Streamfunction"]
+
+  rstream <- max(range(abs(psi_regr$estimate), 
+                       na.rm = TRUE))*1e-7
+  breaks_stream <- AnchorBreaks(0, bins = 20, exclude = 0)(c(-rstream, rstream))
+
+  # browser()
+  stream_gdata %>% 
     periodic(lon = c(0, 360)) %>%
@@
 
+      geom_contour_fill(data = sst_gdata,
+                      aes(z = estimate, fill = ..level..),
+                      breaks = breaks_sst) +
+    geom_contour_pval(data = sst_gdata, aes(z = p_val)) +
+
+    scale_fill_divergent_discretised("SST (K)",
+                                     labels = one_less,
+                                     guide = guide_colorsteps_bottom(width = 15,
+                                                                     show.limits = FALSE, ticks = TRUE,
+                                                                     order = 1)) +
+    ggnewscale::new_scale_fill() +
+    geom_contour_fill(aes(z = estimate*1e-7, fill = ..level..),
+                      breaks = breaks_stream) +
+    geom_contour_pval(aes(z = p_val)) +
+    # geom_streamline(aes(dx = fx, dy = fy), min.L = 3, L = 10,
+    #                 size = 0.1, res = 2, nx = 22, ny = 12, 
+    #                 arrow.angle = 14, arrow.length = 0.2) +
+    geom_vector(aes(dx = fx, dy = fy), skip = 3, min.mag = 0.002, 
+                size = 0.1) +
+    scale_mag(max_size = .5, guide = "none") +
+    geom_qmap(keep = .015) +
+    scale_fill_divergent_discretised(latex2exp::TeX("Streamfunction ($m^2s^{-1}\\times 1e^{-7}$)"), 
+                                     labels = one_less,
+                                     low = scales::muted("red"),
+                                     high = scales::muted("blue"),
+                                     guide = guide_colorsteps_bottom(15, show.limits = FALSE,
+                                                                     order = 99)) +
+    grid_panel(spoke, aes(angle = -angle, var = "SST"), inherit.aes = FALSE) +
+    scale_x_continuous(NULL, expand = c(0, 0), 
+                       breaks = seq(-180, 180-60, by = 60),
+                       labels = LonLabel) +
+
+    scale_y_continuous(NULL, breaks  = seq(-90, 0, by = 30), expand = c(0, 0)) +
+
+    facet_grid(angle ~ var, labeller = labeller(angle = angles_labs)) +
+    # coord_quickmap(ylim = c(NA, 10)) +
+    coord_sf(ylim = c(-89, 10), expand = FALSE,
+             crs = "+proj=longlat +pm=180 +over",
+             default_crs = "+proj=longlat +lon_wrap=180 +over") + 
+    theme(legend.box = "horizontal",
+          legend.text = element_text(size = rel(0.7)), 
+          legend.box.spacing = grid::unit(0, "lines") ) +
+    tag_facets("rc") 
+}
+```
+
+```{r  sst-psi-2,  fig.cap = "(ref:sst-psi-2-cap)", fig.width = 7, fig.height=5.6, dependson="plot_sst_psi", fig.env="figure*"}
+plot_sst_psi(sst_regr[cEOF == "cEOF2"], psi_regr[cEOF == "cEOF2"]) 
+```
+
+(ref:sst-psi-2-cap) Regression maps of cEOF2 with SST (K, column 1) and streamfunction zonal anomalies ($m^2/s\times10^-7$, shaded) with their corresponding activity wave flux (vectors) (column 2) for the 1979 -- 2019 period and different phases of the cEOF (illustrated in the lower-left arrow).
+Areas marked with dots have p-values smaller than 0.01 adjusted for FDR.
+
+The connections between cEOFs and Tropical sources are also assessed.
+Figure \@ref(fig:sst-psi-2) shows the regression maps of Sea Surface Temperatures (SST) and streamfunction anomalies at 200 hPa respectively upon standardised cEOF2, which can be interpreted as the anomalies associated with a 1 standard deviation increase of each index.
+Besides showing the regression of the real and imaginary parts, we also show regressions in two intermediate directions. 
+The Imaginary cEOF2 is associated with strong positive SST anomalies on the Central Pacific and negative anomalies over an area across the North of Australia and New Zealand, the South Pacific Convergence Zone (SPCZ) (Figure \@ref(fig:sst-psi-2).b1).
+This pattern is almost canonically positive ENSO [@bamston1997] and indeed, the correlation between the Imaginary cEOF2 and the Oceanic Niño Index [@bamston1997] is significant and very high `r enso_cor[cEOF == "cEOF2" & part == "Imaginary"]$text`.
+On top of this ENSO-like pattern, there are positive anomalies in the western Indian Ocean and negative values in the eastern Indian Ocean, resembling a positive Indian Ocean Dipole.
+
+Fittingly, the Imaginary cEOF2 is also associated with strong wave-like streamfunction anomalies emanating from the tropics (Figure \@ref(fig:sst-psi-2).b2), both from the Central Pacific sector and the Incean Ocean.
+These are consistent with the effect of ENSO and the Indian Ocean Dipole on the extratropics: the SST anomalies initiate anomalous tropical convection that excites Rossby waves that propagate meridionally towards higher latitudes [@mo2000; @cai2011; @nuncio2015].
+
+Not all phases of the cEOF2 are associated with the same tropical SSTs, however.
+Figure \@ref(fig:sst-psi-2).d1 and d2 show that the Real cEOF2 is not associated either with any significant SST nor streamfunction anomalies in the tropics.
+The correlation between the Real cEOF2 and ENSO is also not significant (`r enso_cor[cEOF == "cEOF2" & part == "Real"]$text`). 
+Rows a and c in Fig. \@ref(fig:sst-psi-2) show that the intermediate phases are still associated with significant SST anomalies over the Pacific Ocean, but with with slightly different location.
+The 135º phase is associated with SST in the central Pacific, while the 45º phase is associated with SST in the eastern Pacific, which correspond roughly to the Eastern Pacific and Central Pacific "flavours" of ENSO (Citas).
+
+(ref:enso-phase-cap) ONI plotted against cEOF2 phase for th 1979 -- 2019 period.
+Years with magnitude of cEOF2 greater or smaller than the 50th percentile are shown as orange diamonds and green circles and respectively. 
+Black line is the fit ONI ~ sin(phase) computed by weighted OLS using the magnitude of the cEOF2 as weights.
+
+```{r enso-phase, fig.cap = "(ref:enso-phase-cap)", fig.width = width_column, fig.height=3, include = TRUE}
+gdata <- ceof[, eof[[1]]$left, by = .(season)] %>%
+  .[cEOF == "cEOF2"] %>% 
+  # .[, hgt := Anomaly(hgt)] %>% 
+  .[, arg := Arg(hgt)] %>%
+  .[, mag := Mod(hgt)] %>%
+  .[enso, on = "time"] %>%
+  na.omit() %>%
+  .[, q := ifelse(mag >= quantile(mag, 0.5), "|cEOF2| > 50%", "|cEOF2| < 50%")]
+
+cor_enso_mag <- gdata[, correlate(mag, abs(oni))]
+
+cor_enso_spearman <- gdata[, cor.test(mag, abs(oni), method = "spearman")]
+
+cor_enso_mag_outlier <- gdata[oni < 1.5][, correlate(mag, abs(oni))]
+
+cor_enso_mag_quantile <- gdata[mag >= quantile(mag, 0.5)] %>% 
+  .[oni < 1.5] %>% 
+  .[, correlate(mag, abs(oni))]
+
+cos_model <- gdata %>% 
+  .[, lm(oni ~ I(sin(arg)) - 1, weights = mag )]
+
+equation <- latex2exp::TeX(paste0("$\\mathrm{ONI} = ", 
+                                  signif(coef(cos_model)[1], 2), "\\sin{(\\mathrm{phase})}$"))
+
+gdata %>% 
+  ggplot(aes(arg, oni)) +
+  geom_hline(yintercept = c(-0.5, 0, 0.5), size = 0.2, color = "gray50") +
+  geom_vline(xintercept = 0, size = 0.2, color = "gray50") +
+  annotate(shadowtext:::GeomShadowText,
+           label = c("- Real", "- Imaginary", "+ Real", "+ Imaginary", "- Real"),
+           color = "gray60", bg.colour = "white", bg.r = 0.2,
+           size = 3,
+           y = -Inf, x = c(-pi, -pi/2, 0, pi/2, pi), angle = 90, hjust = 0) +
+
+  geom_smooth(data = ~qwrap(.x, arg = c(-pi, pi) ~ c(-2*pi, 2*pi)),
+              method = "lm", aes(weight = mag),
+              formula = y ~ I(sin(x)) -1,
+              se = FALSE, color = "black", size = 0.5) +
+  geom_point(aes(color = q,
+                 shape  = q),
+             size = 3) +
+  ggrepel::geom_text_repel(data = ~.x[order(-oni)][1:3],
+                           aes(label = year(time))) +
+
+  scale_color_brewer("", palette = "Dark2") +
+  scale_shape_manual("", values = c(20, 18)) +
+  scale_y_continuous("ONI") +
+  scale_x_continuous("Phase",
+                     breaks = seq(-pi, pi, by = 45*pi/180),
+                     labels = function(x) paste0(round(x*180/pi, 5), "º")) +
+  coord_cartesian(xlim = c(-pi, pi)) +
+  labs(subtitle = equation) +
+  panel_background
+```
+
+To better explore the relationship between tropical forcing and phase of the cEOF2, Figure \@ref(fig:enso-phase) shows the ONI index against the cEOF2 phase for each SON trimester between 1979 and 2019, highlighting years in which the magnitude of cEOF2 is above the median.
+In years with positive ONI, the phase of the cEOF2 is mostly around +90º (corresponding with positive imaginary part) and vice versa.
+In years with near neutral ENSO, the phase of the cEOF2 is much more variable.
+The black line in Figure \@ref(fig:enso-phase) is a sinusoidal fit of the relationship between ONI and cEOF2 phase.
+The $r^2$ corresponding to the fit is `r signif(summary(cos_model)$r.squared, 2)`, statistically significant with p-value `r report_p(coef(summary(cos_model))[4])`, indicating a quasi-sinusoidal relation between these two variables.
+
+The correlation between the absolute magnitude of the ONI and the magnitude of the cEOF2 is `r cor_enso_mag$text`.
+This relationship, however, appears to be driven only by the three years with strongest ENSO events in the period (`r knitr::combine_words(year(gdata[order(-oni)][1:3]$time))`) which also coincide with the three years with strongest cEOF2 magnitude.
+If those years are removed, the correlation becomes non-significant (`r cor_enso_mag_outlier$text`).
+Furthermore, even when using all years, the Spearman correlation --which is robust to outliers-- is also non-significant (`r signif(cor_enso_spearman$estimate, 2)`, p-value`r report_p(cor_enso_spearman$p.value)`).
+Therefore, although the location of tropical SST seem to have an effect in defining the phase of the cEOF2, the relationship between the magnitude of the cEOF2 train wave and ONI remains uncertain.
+
+It could be concluded that the wave train represented by cEOF2 can be both part of the internal variability of the extratropical atmosphere or forced by tropical SSTs.
+In the former case, the wave train has little phase preference.
+However, when cEOF2 is excited by tropical SST variability, it tends to remain locked to the imaginary phase.
+This explains the relative over-abundance of years with cEOF2 near positive and negative imaginary phase in Figure \@ref(fig:phase-histogram).
+
+Unlike the cEOF2 case, there is no significant pattern of SST anomalies associated with either the Real or Imaginary cEOF1 (Figure \@ref(fig:sst-psi-1)).
+Consistently, streamfunction anomalies do not show any tropical influence.
+Instead, the real and Imaginary cEOF1 are associated with zonally wave activity fluxes in the extra-tropics around 60ºS, except for an equatorward flow from the coast of Antarctica around 150ºE in the real component.
+This suggests that its variability is driven primary by the internal variability of the extra-tropics.
+
+## Precipitation and 2-metre temperature {#precipitation}
+
+```{r add_continents}
+continents <- list(africa  = list(lon = c(0, 60),
+                                  lat = c(-40, 0)),
+                   oceania = list(lon = c(100, 180),
+                                  lat = c(-50, -13)),
+                   america = list(lon = c(265, 325),
+                                  lat = c(-60, -10)))
+
+add_continent <- function(dt) {
+  dt[, continent := fcase(lon %between% continents$africa$lon &
+                            lat %between% continents$africa$lat, "africa",
+                          lon %between% continents$oceania$lon &
+                            lat %between% continents$oceania$lat, "oceania",
+                          lon %between% continents$america$lon &
+                            lat %between% continents$america$lat, "america")] %>%
+    .[, continent := factor(continent, levels = c("africa", "oceania", "america"), ordered = TRUE)] %>%
+    .[]
+}
+
+crop_continent <- function(continent) {
+  bbox <- c(xmin = min(continents[[continent]]$lon), 
+            xmax = max(continents[[continent]]$lon),
+            ymin = min(continents[[continent]]$lat),
+            ymax = max(continents[[continent]]$lat))
+
+  function(data) {
+    s2 <- sf::sf_use_s2(FALSE)
+    on.exit(sf::sf_use_s2(s2))
+    data <- sf::st_crop(data, bbox)
+    data$continent <- continent
+    data
+  }
+}
+```
+
+```{r plot_pp}
+plot_pp <- function(pp_regr, t2m_regr, hgt850_regr) {
+  lats <- c(-55, 10)
+  breaks_pp <- AnchorBreaks(exclude = 0)(c(-.9, .9))
+
+  pp_regr <- pp_regr %>% 
+    copy() %>% 
+    .[, var := "Precipitation"] %>% 
+        .[, p.value := Pvaluate(estimate, std.error, df, "none"), 
+      by = .(cEOF, angle)]
+
+  t2m_regr <- t2m_regr %>% 
+    copy() %>% 
+    .[, var := "2m Temperature"] %>% 
+    .[, p.value := Pvaluate(estimate, std.error, df, "none"), by = .(cEOF, angle)]
+
+  pp_regr %>%
+    # .[lat %between% lats] %>%
+
+    periodic(lon = c(0, 360)) %>%
+    ggplot(aes(lon, lat)) +
+
     geom_contour_fill(aes(z = estimate, fill = ..level..),
-                      breaks = breaks) +
-    geom_contour_tanaka(aes(z = estimate), breaks = breaks) +
+                      breaks = breaks_pp) +
+    geom_contour_tanaka(aes(z = estimate), breaks = breaks_pp) +
 
@@
                                      # limits = c(-1, 1),
-                                     # oob = scales::squish,
                                      low = "#8E521C",
                                      high = "#00665E",
-                                     guide = guide_colorsteps_bottom(15)) +
-
+                                     guide = guide_colorsteps_bottom(15, order = 99)) +
     geom_contour_pval(aes(z = p.value)) +
+
+    ggnewscale::new_scale_fill() +
+
+    geom_contour_fill(data = t2m_regr, 
+                      aes(z = estimate, fill = ..level..), 
+                      breaks = AnchorBreaks(exclude = 0)) +
+    geom_contour_tanaka(data = t2m_regr, 
+                        aes(z = estimate),
+                        breaks = AnchorBreaks(exclude = 0)) +
+
+    scale_fill_divergent_discretised("Temperature",
+                                     # limits = c(-1, 1),
+                                     guide = guide_colorsteps_bottom(15, order = 1)) +
+    geom_contour_pval(data = t2m_regr, aes(z = p.value)) +
+
+    geom_contour2(data = copy(hgt850_regr)[, var := "2m Temperature"], 
+                  aes(z = estimate, linetype = factor(-sign(..level..))), 
+                  size = 0.2, breaks = AnchorBreaks(0, exclude = 0)) +
+
     geom_qmap() +
 
-    grid_panel(spoke, aes(angle = -angle)) +
+    grid_panel(spoke, aes(angle = -angle, var = "SST"), inherit.aes = FALSE) +
 
@@
     scale_y_latitude(ticks = 15) +
-    facet_grid(angle ~ cEOF) +
+    scale_linetype_discrete(guide = "none") +
+    facet_grid(angle ~ var, labeller = labeller(angle = angles_labs)) +
     axis_labs_smol +
     tag_facets("rc") +
-    coord_sf(ylim = lats, 
+    coord_sf(ylim = c(-85, 10),
              crs = "+proj=longlat +pm=180 +over",
              default_crs = "+proj=longlat +over") +
-    theme(panel.spacing = grid::unit(1, "lines"))
+    theme(panel.spacing = grid::unit(1, "lines")) +
+    theme(legend.box = "horizontal",
+          legend.text = element_text(size = rel(0.7)), 
+          legend.box.spacing = grid::unit(0, "lines") )
 
@@
 
-The influence of cEOFs variability in continental rainfall of extra-tropical Southern Hemisphere is also explored.
-We focus on South America and Oceania sectors because there are no relevant signals in southern Africa (not shown).
-
-(ref:pp-oceania-cap) Regression of SON mean precipitation anomalies in Oceania (mm per day, shaded) with the (row a) real and (row b) imaginary pars of the (column 1) cEOF1 and (column 2) cEOF2.
+The influence of cEOFs variability in continental rainfall and 2-metre temperature in the Southern Hemisphere is also explored.
+Figure \@ref(fig:pp-temp-2) show regression maps of 2-meter temperature (column 1) and precipitation (column 2) upon standardised cEOF2 which, again, can be interpreted as the anomalies associates with a 1 standard deviation increase.
+
+(ref:pp-temp-cap) Regression of SON mean 2-meter temperature (K, shaded) and 850hPa geopotential height (m, contours) (column 1), and precipitation (correlation, column 2) with different phases of cEOF2.
 For the 1979 -- 2019.
 Areas marked with dots have p-values smaller than 0.01 adjusted for False Detection Rate.
-In black is the East Australia region, defined as the box between `r knitr::combine_words(LonLabel(box_australia$lon))` and `r knitr::combine_words(LatLabel(box_australia$lat))`.
-
-```{r, fig.cap="precipitation correlation", fig.width=6.6, fig.height=5.5}
-plot_pp(pp_regr, c("cEOF1", "cEOF2"))
-```
-
-```{r}
-# a <- angles_regr[2]
-# 
-# plot_pp(pp_regr, c("cEOF1", "cEOF2"))
-# 
-# (plot_sst_psi(regr_sst = sst_regr[angle == a], regr_psi = psi_regr[angle == a],
-#               which_cEOF = "cEOF2", which_lev = 200)) / 
-#   (
-#     plot_pp(pp_regr[angle == a], c("cEOF1", "cEOF2"))
-#   )
-```
-
-```{r pp-oceania, fig.cap = "(ref:pp-oceania-cap)", fig.width=width_column, fig.height=width_column*.8}
-# plot_pp(regr_pp, c("cEOF1", "cEOF2"), which_continent =  "oceania") +
-#   geom_rect(data = box_australia, aes(xmin = min(lon), xmax = max(lon),
-#                                       ymin = min(lat), ymax = max(lat)),
-#             fill = NA, colour = "gray20")
-
-```
-
-The precipitation anomalies in Oceania associated with cEOF1 are weak and not statistically significant (Figure \@ref(fig:pp-oceania) column 1) whereas cEOF2 is associated with large and widespread anomalies.
-The Real cEOF2 is associated with positive seasonal anomalies in northern and eastern Australia, and negative ones over New Zealand, while precipitation anomalies associated with Imaginary cEOF2 are negative over all eastern Australia.
-
-These precipitation anomalies associated with the cEOF2 are similar to the springtime precipitation anomalies observed in relation with the Asymmetric SAM [@campitelli2021] and ENSO [@cai2011],
-
-\textcolor{red}{Acá te faltaría  sacar un poco más el jugo asociando con la figura 11. Además, ¿revisaste si podes también usar algún resultado asociado a los PSA? }
+
+```{r pp-temp-2, fig.cap="(ref:pp-temp-cap)", fig.width=6.6, fig.height=5.5}
+plot_pp(pp_regr[cEOF == "cEOF2"], t2m_regr[cEOF == "cEOF2"], hgt850_regr[cEOF == "cEOF2"])
+```
+
+Temperature anomalies associated with the Imaginary cEOF (Fig.
+\@ref(fig:pp-temp-2).b1) show positive values in the tropical Pacific, consistent with SSTs anomalies associated with the same phase (Fig.
+\@ref(fig:sst-psi-2).b1).
+
+At higher latitudes there is a wave-like pattern of positive and negative values that coincide with the nodes of the 850hPa geopotential height regression patterns.
+This is consistent with temperature anomalies produced by meridional advection of temperature by the meridional winds arising from geostrophic balance.
+
+Over the continents, the Imaginary cEOF2 (Fig.
+\@ref(fig:pp-temp-2)b.1) is associated with positive temperatures in southern Australia and negative temperature in southern South America and the Antarctic Peninsula.
+
+The temperatures anomalies associated with the Real cEOF2 (Fig. \@ref(fig:pp-temp-2)d.1) are less extensive. 
+There is no significant signal over the continents except for positive anomalies near the Antarctic Peninsula.
+
+Tropical precipitation anomalies associated with the Imaginary cEOF2 are strong, with positive anomalies in the central Pacific and western Indian, an negative anomalies in the eastern Pacific. 
+This pattern is consistent with the SST anomalies shown before as the increased SST enhances tropical convection and vice versa.
+
+On the extra-tropics, the Imaginary cEOFs is correlated with drier conditions over eastern Australia; a similar signal as the one associated with ENSO @cai2011. 
+However, the Imaginary cEOF2 is not the direction most correlated with precipitation in that area. 
+The 135º direction (positive Imaginary + negative Real cEOF2) is associated with stronger and more extensive correlations with precipitation over Australia and New Zealand.
+This Australian signal is most likely there due to the more direct impacts of ENSO rather than on the interconnection pattern represented by the cEOF2.
 
@@
 
-```{r australia-pp-phase, fig.cap = "(ref:australia-pp-phase-cap)", fig.width = width_column, fig.height=3}
+```{r australia-pp-phase, fig.cap = "(ref:australia-pp-phase-cap)", fig.width = width_column, fig.height=3, include = FALSE}
 gdata <- cmap %>% 
@@
 
-To better understand the relationship between precipitation anomalies over eastern Australia and the cEOF2, Figure \@ref(fig:australia-pp-phase) plots precipitation anomalies averaged over eastern Australia (box in Figure \@ref(fig:pp-oceania)) as a function of cEOF2 phase.
-The black line shows a sinusoidal fit, whose coefficients are fitted by weighted least squares using the magnitude of the EFO2 as weight.
-The $r^2$ corresponding to the fit is `r signif(summary(model)$r.squared, 2)` (p-value `r report_p(p_value)`).
-There is a strong relationship between the phase of the cEOF2 and precipitation anomalies in eastern Australia, but the Real and Imaginary phases we chose are not exactly aligned with the direction that maximises this relationship.
-To perform a more in-depth analysis of the relationship between this pattern and precipitation in this region, it would be recommended to align one of the axis (either the Real or the Imaginary) to the phase that maximises the described relationship.
-
-@cai2011 -> precipitación ENSO
-
- no vio relación con la pp. mezclan fases, pero la fase es crucial.
+<!-- To better understand the relationship between precipitation anomalies over eastern Australia and the cEOF2, Figure \@ref(fig:australia-pp-phase) plots precipitation anomalies averaged over eastern Australia (box in Figure \@ref(fig:pp-oceania)) as a function of cEOF2 phase.
+-->
+<!-- The black line shows a sinusoidal fit, whose coefficients are fitted by weighted least squares using the magnitude of the EFO2 as weight. -->
+<!-- The $r^2$ corresponding to the fit is `r signif(summary(model)$r.squared, 2)` (p-value `r report_p(p_value)`). -->
+<!-- There is a strong relationship between the phase of the cEOF2 and precipitation anomalies in eastern Australia, but the Real and Imaginary phases we chose are not exactly aligned with the direction that maximises this relationship.
+-->
+
+Over South America, the Imaginary cEOF has positive correlations with precipitation in South Eastern South America (SESA) and central Chile, and negative correlations in eastern Brazil.
+This again is a well-known springtime precipitation signature of ENSO [@cai2020a] and it is also similar to the precipitation anomalies associated with the A-SAM [@campitelli2021].
 
@@
 
-Figure \@ref(fig:pp-america) shows regression maps of seasonal precipitation with each cEOF in South America.
-cEOF1 is associated with negative regressions coefficients for precipitation in Southern Brazil and Paraguay, although these are not statistically significant (Fig. \@ref(fig:pp-america) column 1).
-The strongest precipitation regression coefficients are the ones associated with the Imaginary cEOF2.
-The positive regression coefficients on Southeastern South America (SESA) and Central Chile, and negative ones over South Atlantic Convergence Zone is a well-known springtime precipitation signature of ENSO [@cai2020a] and it is also similar to the precipitation anomalies associated with the A-SAM [@campitelli2021].
-\textcolor{red}{Lo mismo que te puse antes. Te faltaría  sacar un poco más el jugo asociando con la figura 11. Además, ¿revisaste si podes también usar algún resultado asociado a los PSA? }
-
 This is not surprising considering the close relationship between the ONI, the A-SAM index and the Imaginary cEOF2 shown previously, but further consolidates the identification of this mode with the PSA pattern.
-On the other hand, the Real cEOF2 is associated with negative precipitation anomalies in a smaller area of SESA.
 Resembling the relationship between ONI and the phase of cEOF2 (Fig. \@ref(fig:enso-phase)), there is a cEOF2 phase dependence of the precipitation anomalies in SESA (not shown).
-These variables could be related wth a significant sinusoidal fit that has a statistically significant coefficient of determination of `r round(summary(model)$r.squared, 2)`.
+These variables could be related with a significant sinusoidal fit that has a statistically significant coefficient of determination of `r round(summary(model)$r.squared, 2)`.
+
+The precipitation signal associated with the Real cEOF2 is less strong. 
+There is a residual positive correlation in the equatorial eastern Pacific and small, not statistically significant positive correlations over eastern Australia and negative ones over New Zealand.
+
+Lastly, the cEOF1 doesn't have a significant signal associated with 2-metre temperature or precipitacion (Fig. \@ref(fig:pp-temp-cap-1)).
 
@@
 Essentially, it is an alternate representation of the PSA1 and PSA2 patterns [@mo2001].
-We show that the Imaginary cEOF2 can be identified with the PSA1 and the Real cEOF2 with the PSA2.
-This cEOF prefers a phase aligned with the Imaginary cEOF2, because when ENSO is active, it forces the cEOF2 to be in the Imaginary phase.
+We show that the Imaginary cEOF2 can be roughly identified with the PSA1 and the Real cEOF2 with the PSA2.
+
+There is some disagreement in the literature of whether the phase of the PSA pattern is affected by the location of tropical SST anomalies [@irving2016].
+We show not only that the cEOF2 tends to be in the imaginary phase (~PSA1) when the ENSO region is warm, but that central Pacific SST anomalies tend to move the cEOF towards the negative real phase and eastern Pacific SST anomalies tend to move it towards positive real phase.
 When ENSO is neutral, the cEOF2 is still active, but with no particular preferred phase.
@@
 The magnitude of the cEOF2 appears not to be related to the magnitude of ENSO, except for the three strongest ENSO events in the period which coincide with the three strongest cEOF2 years.
+
+The sensitivity of the phase of the cEOF2 to the location of the tropical SST anomalies was also seen by @ciasto2015, who detected similar Rossby wave patterns associated with central Pacific and eastern Pacific SST anomalies but with a change in phase.
 
@@
 
-# References
+\appendix
+
+\counterwithin{figure}{section}
+
+# Extra figures
+
+\newpage
 
 <!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->
+
+(ref:sst-psi-1-cap) Same as Figure \@ref(fig:sst-psi-2) but for cEOF1.
+
+```{r sst-psi-1, fig.cap = "(ref:sst-psi-1-cap)", fig.width = 7, fig.height=5.6, dependson="plot_sst_psi", fig.env = "figure*"}
+plot_sst_psi(sst_regr[cEOF == "cEOF1"], psi_regr[cEOF == "cEOF1"]) 
+```
+
+(ref:pp-temp-cap-1) Same as Figure \@ref(fig:pp-temp-2) but for cEOF1.
+
+```{r pp-temp-1, fig.cap="(ref:pp-temp-cap-1)", fig.width=6.6, fig.height=5.5}
+plot_pp(pp_regr[cEOF == "cEOF1"], t2m_regr[cEOF == "cEOF1"], hgt850_regr[cEOF == "cEOF1"])
+```

### 2021-09-07 R/0-download.R

@@ -103,7 +103,7 @@ request_geopotential <- list(
   format = "netcdf",
   product_type = "monthly_averaged_reanalysis",
   variable = c("geopotential"),
-  pressure_level = c("50", "200", "500"),
+  pressure_level = c("50", "200", "500", "850"),
   year = as.character(1979:2020),
   month = formatC(1:12, width = 2, flag = 0),   # Need all months to compute PSA
   time = "00:00",
@@ -189,9 +189,9 @@ request_toc <- list(
 )
 
 
-request_era5_sst <- list(
+request_era5_2mt <- list(
   format = "netcdf",
-  variable = c("sea_surface_temperature", "2m_temperature"),
+  variable = c("2m_temperature"),
   product_type = "monthly_averaged_reanalysis",
   year = as.character(1979:2020),
   month =  c("09", "10", "11"),

### 2021-09-07 R/data_locations.R

@@ -114,8 +114,8 @@ ERA5_BE <- data_nc_fun(file = data_path("raw", "era5be.mon.mean.nc"),
 
 #' @export
 #' @rdname data_locations
-ERA5_SST <- data_nc_fun(file = data_path("raw", "era5.sst.mon.mean.nc"),
-                        download = download_cds(request_era5_sst))
+ERA5_T2M <- data_nc_fun(file = data_path("raw", "era5.2mt.mon.mean.nc"),
+                        download = download_cds(request_era5_2mt))
 
 
 #' @export

### 2021-09-07 R/draw_panel.R

@@ -41,3 +41,5 @@ grid_panel <- function(grob=grid::nullGrob(),
         show.legend = FALSE,
         params = list(grob = grob, debug = debug, ...))
 }
+
+

### 2021-09-07 R/maps.R

@@ -10,7 +10,7 @@
 #' @export
 geom_qmap <- function(subset = identity,
                       crop = NULL,
-                      color = "gray50", size = 0.2,
+                      color = "gray50", size = 0.3,
                       fill = NA, wrap = c(0, 360), weighting = 0.7,
                       keep = 0.015, ...) {
   lon <- lat <- group <- NULL

### 2021-09-07 analysis/scripts/02-compute-sam.R

@@ -12,7 +12,7 @@ if (file.exists(sam_file)) {
   library(shceof)
   library(metR)
 
-  sams <- ERA5() %>%
+  sams <- ERA5_geopotential_all() %>%
     ReadNetCDF(vars = c("hgt" = "z"),
                subset = list(latitude = c(-90, -20))) %>%
     normalise_coords() %>%

### 2021-09-07 analysis/paper/shceof.tex
REnder

@@
 The SAM represents a relatively zonally symmetric pattern of alternating low pressures in polar latitude and a ring of high pressures in high latitudes with an embedded wave 3 pattern that is more prominent in the Pacific sector.
-The PSA1 and PSA2 describe the two out-of-phase sides of a meridionally propagating wave train that originates in the Eastern equatorial Pacific and travels towards the South Atlantic following a great-circle arch along the Antarctic Peninsula.
+The PSA1 and PSA2 describe the two out-of-phase sides of a meridionally propagating wave train that originates in the eastern equatorial Pacific and travels towards the South Atlantic following a great-circle arch along the Antarctic Peninsula.
 These patterns are derived by applying EOF to temporal anomalies, but \citet{raphael2003} applied EOF methods specifically to zonal anomalies.
@@
 \midrule
-cEOF1 & \textbf{0.28} & 0.01 & 0.02\\
-cEOF2 & 0.00 & \textbf{0.60} & 0.02\\
-cEOF3 & 0.00 & 0.00 & 0.02\\
+cEOF1 & \textbf{0.29} & 0.01 & 0.03\\
+cEOF2 & 0.00 & \textbf{0.59} & 0.02\\
+cEOF3 & 0.00 & 0.00 & 0.01\\
 \bottomrule
@@
 
-There is only a significant upward trend in the real component of cEOF1 (Fig.~\ref{fig:extended-series}a.1, p-value~\textless{} 0.001) and no significant trend in any of the complex components of cEOF2.
+There is only a significant upward trend in the real component of cEOF1 (Fig.~\ref{fig:extended-series}a.1, p-value~= 0.0037) and no significant trend in any of the complex components of cEOF2.
 The positive trend in the Real cEOF1 translates into a positive trend in cEOF1 magnitude, but not in the phase (not shown).
@@
 At 50 hPa (Figure~\ref{fig:eof1-regr-gh} row a), both the Real and Imaginary cEOF1 are associated with planetary wave 1 patterns, that are 90º out of phase.
-Their phases coincide with the ones shown in Figure~\ref{fig:ceofs-1}a.1, with the positive centre of the Real cEOF1 located towards the dateline, and the one of the Imaginary cEOF1 located over Eastern Antarctica.
+Their phases coincide with the ones shown in Figure~\ref{fig:ceofs-1}a.1, with the positive centre of the Real cEOF1 located towards the dateline, and the one of the Imaginary cEOF1 located over East Antarctica.
 However, the Real cEOF1 pattern is substantially altered by the zonally symmetric circulation.
@@
 \midrule
-PSA1 & -0.32 (CI: -0.57 -- -0.02) & \textbf{-0.82 (CI: -0.9 -- -0.69)}\\
-PSA2 & \textbf{0.81 (CI: 0.67 -- 0.9)} & -0.06 (CI: -0.36 -- 0.25)\\
+PSA1 & 0.26 (CI: -0.04 -- 0.52) & \textbf{0.82 (CI: 0.69 -- 0.9)}\\
+PSA2 & \textbf{0.79 (CI: 0.63 -- 0.88)} & -0.02 (CI: -0.32 -- 0.29)\\
 \bottomrule
@@
 There is a close spatial relationship between amplitudes and phases of the cEOF1 and TOC planetary wave 1 between 75°S and 45°S.
-The correlation between the amplitude of both indices is 0.77 (CI: 0.61 -- 0.87) and 0.49 (CI: 0.21 -- 0.69) between their phases.
+The correlation between the amplitude of both indices is 0.79 (CI: 0.63 -- 0.88) and -0.85 (CI: -0.92 -- -0.74) between their phases.
 
@@
 \begin{figure*}
-\includegraphics{../figures/sst-psi-2-1} \caption{Regression maps of cEOF2 with SST (K, column a) and streamfunction zonal anomalies (\(m^2/s\times10^-7\), shaded) with their corresponding activity wave flux (vectors) (column b) for the 1979 -- 2019 period.
+\includegraphics{../figures/sst-psi-2-1} \caption{Regression maps of cEOF2 with SST (K, column 1) and streamfunction zonal anomalies (\(m^2/s\times10^-7\), shaded) with their corresponding activity wave flux (vectors) (column 2) for the 1979 -- 2019 period and different phases of the cEOF (illustrated in the lower-left arrow).
 Areas marked with dots have p-values smaller than 0.01 adjusted for FDR.}\label{fig:sst-psi-2}
@@
 The connections between cEOFs and Tropical sources are also assessed.
-Figure~\ref{fig:sst-psi-2} shows the regression maps of Sea Surface Temperatures (SST) and Streamfunction anomalies at 200 hPa respectively upon cEOF2.
-The Imaginary cEOF2 is associated with strong positive SST anomalies on the Central Pacific and negative anomalies over an area across the North of Australia and New Zealand, the South Pacific Convergence Zone (SPCZ) (Figure~\ref{fig:sst-psi-2}.a2).
-This pattern is almost canonically positive ENSO \citep{bamston1997} and indeed, the correlation between the Imaginary cEOF2 and the Oceanic Niño Index \citep{bamston1997} is significant and very high 0.76 (CI: 0.59 -- 0.87).
-Besides, the Imaginary cEOF2 is associated with strong wave-like streamfunction anomalies emanating from the tropics (Figure~\ref{fig:sst-psi-2}.b2).
-This is consistent with the effect of ENSO on the extratropics: SST anomalies initiate anomalous tropical convection that excites Rossby waves that propagate meridionally towards higher latitudes \citep{mo2000}.
+Figure~\ref{fig:sst-psi-2} shows the regression maps of Sea Surface Temperatures (SST) and streamfunction anomalies at 200 hPa respectively upon standardised cEOF2, which can be interpreted as the anomalies associated with a 1 standard deviation increase of each index.
+Besides showing the regression of the real and imaginary parts, we also show regressions in two intermediate directions.
+The Imaginary cEOF2 is associated with strong positive SST anomalies on the Central Pacific and negative anomalies over an area across the North of Australia and New Zealand, the South Pacific Convergence Zone (SPCZ) (Figure~\ref{fig:sst-psi-2}.b1).
+This pattern is almost canonically positive ENSO \citep{bamston1997} and indeed, the correlation between the Imaginary cEOF2 and the Oceanic Niño Index \citep{bamston1997} is significant and very high 0.76 (CI: 0.6 -- 0.87).
+On top of this ENSO-like pattern, there are positive anomalies in the western Indian Ocean and negative values in the eastern Indian Ocean, resembling a positive Indian Ocean Dipole.
 
-Since the Real cEOF2 represents just a different phase of the same wave train, it would be expected to show a similar forcing pattern to the Imaginary cEOF2 with a slight translation of its location.
-However, Figure~\ref{fig:sst-psi-2}.a1 and b1 show that the Real cEOF2 is not associated either with any significant SST nor streamfunction anomalies in the tropics.
-The correlation between the Real cEOF2 and ENSO is also not significant (0 (CI: -0.31 -- 0.31)).
-This lack of tropical signal suggest that the Real cEOF2 and the Imaginary cEOF2 have very different sources of variability.
+Fittingly, the Imaginary cEOF2 is also associated with strong wave-like streamfunction anomalies emanating from the tropics (Figure~\ref{fig:sst-psi-2}.b2), both from the Central Pacific sector and the Incean Ocean.
+These are consistent with the effect of ENSO and the Indian Ocean Dipole on the extratropics: the SST anomalies initiate anomalous tropical convection that excites Rossby waves that propagate meridionally towards higher latitudes \citep{mo2000, cai2011, nuncio2015}.
+
+Not all phases of the cEOF2 are associated with the same tropical SSTs, however.
+Figure~\ref{fig:sst-psi-2}.d1 and d2 show that the Real cEOF2 is not associated either with any significant SST nor streamfunction anomalies in the tropics.
+The correlation between the Real cEOF2 and ENSO is also not significant (0 (CI: -0.3 -- 0.3)).
+Rows a and c in Fig.~\ref{fig:sst-psi-2} show that the intermediate phases are still associated with significant SST anomalies over the Pacific Ocean, but with with slightly different location.
+The 135º phase is associated with SST in the central Pacific, while the 45º phase is associated with SST in the eastern Pacific, which correspond roughly to the Eastern Pacific and Central Pacific ``flavours'' of ENSO (Citas).
 
@@
 \caption{\label{fig:enso-phase}ONI plotted against cEOF2 phase for th 1979 -- 2019 period.
-Years with magnitude of cEOF2 greater or smaller than the 50th percentile are shown as orange circles and green diamomnds respectively.
+Years with magnitude of cEOF2 greater or smaller than the 50th percentile are shown as orange diamonds and green circles and respectively.
 Black line is the fit ONI \textasciitilde{} sin(phase) computed by weighted OLS using the magnitude of the cEOF2 as weights.}
@@
 The black line in Figure~\ref{fig:enso-phase} is a sinusoidal fit of the relationship between ONI and cEOF2 phase.
-The \(r^2\) corresponding to the fit is 0.56, statistically significant with p-value \textless{} 0.001, indicating a quasi-sinusoidal relation between these two variables.
+The \(r^2\) corresponding to the fit is 0.57, statistically significant with p-value \textless{} 0.001, indicating a quasi-sinusoidal relation between these two variables.
 
-Furthermore, Figure~\ref{fig:enso-phase} suggest that highest amplitude cEOF2 years tend to coincide with strong ENSO years.
-The correlation between the absolute magnitude of the ONI and the magnitude of the cEOF2 is 0.45 (CI: 0.17 -- 0.67).
+The correlation between the absolute magnitude of the ONI and the magnitude of the cEOF2 is 0.45 (CI: 0.17 -- 0.66).
 This relationship, however, appears to be driven only by the three years with strongest ENSO events in the period (2015, 1997, and 1982) which also coincide with the three years with strongest cEOF2 magnitude.
-If those years are removed, the correlation becomes non-significant (0.05 (CI: -0.28 -- 0.36)).
-Furthermore, even when using all the datapoints, the Spearman correlation --which is robust to outliers-- is also non-significant (0.2, p-value= 0.21).
-Therefore, the relationship between the magnitude of the cEOF2 train wave and ONI remains uncertain.
+If those years are removed, the correlation becomes non-significant (0.04 (CI: -0.28 -- 0.35)).
+Furthermore, even when using all years, the Spearman correlation --which is robust to outliers-- is also non-significant (0.2, p-value= 0.21).
+Therefore, although the location of tropical SST seem to have an effect in defining the phase of the cEOF2, the relationship between the magnitude of the cEOF2 train wave and ONI remains uncertain.
 
@@
 However, when cEOF2 is excited by tropical SST variability, it tends to remain locked to the imaginary phase.
-This explains the relative overabundance of years with cEOF2 near positive and negative imaginary phase in Figure~\ref{fig:phase-histogram}.
+This explains the relative over-abundance of years with cEOF2 near positive and negative imaginary phase in Figure~\ref{fig:phase-histogram}.
 
-\begin{figure*}
-\includegraphics{../figures/sst-psi-1-1} \caption{Same as Figure~\ref{fig:sst-psi-2} but for cEOF1.}\label{fig:sst-psi-1}
-\end{figure*}
-
-Figure~\ref{fig:sst-psi-1} shows SST and streamfunction regression maps for cEOF1.
-Unlike the cEOF2 case, there is no significant pattern of SST anomalies associated with either the Real or Imaginary cEOF1.
+Unlike the cEOF2 case, there is no significant pattern of SST anomalies associated with either the Real or Imaginary cEOF1 (Figure~\ref{fig:sst-psi-1}).
 Consistently, streamfunction anomalies do not show any tropical influence.
 Instead, the real and Imaginary cEOF1 are associated with zonally wave activity fluxes in the extra-tropics around 60ºS, except for an equatorward flow from the coast of Antarctica around 150ºE in the real component.
-This suggests that it's variability is driven primary by the internal variability of the extra-tropics.
+This suggests that its variability is driven primary by the internal variability of the extra-tropics.
 
 \hypertarget{precipitation}{%
-\subsection{Precipitation}\label{precipitation}}
+\subsection{Precipitation and 2-metre temperature}\label{precipitation}}
 
-The influence of cEOFs variability in continental rainfall of extra-tropical Southern Hemisphere is also explored.
-We focus on South America and Oceania sectors because there are no relevant signals in southern Africa (not shown).
+The influence of cEOFs variability in continental rainfall and 2-metre temperature in the Southern Hemisphere is also explored.
+Figure \ref{fig:pp-temp-2} show regression maps of 2-meter temperature (column 1) and precipitation (column 2) upon standardised cEOF2 which, again, can be interpreted as the anomalies associates with a 1 standard deviation increase.
 
@@
 \centering
-\includegraphics{../figures/pp-oceania-1.pdf}
-\caption{\label{fig:pp-oceania}Regression of SON mean precipitation anomalies in Oceania (mm per day, shaded) with the (row a) real and (row b) imaginary pars of the (column 1) cEOF1 and (column 2) cEOF2.
+\includegraphics{../figures/pp-temp-2-1.pdf}
+\caption{\label{fig:pp-temp-2}Regression of SON mean 2-meter temperature (K, shaded) and 850hPa geopotential height (m, contours) (column 1), and precipitation (correlation, column 2) with different phases of cEOF2.
 For the 1979 -- 2019.
-Areas marked with dots have p-values smaller than 0.01 adjusted for False Detection Rate.
-In black is the West Australia region, defined as the box between 130°E and 155°E and 40°S and 15°S.}
+Areas marked with dots have p-values smaller than 0.01 adjusted for False Detection Rate.}
 \end{figure}
 
-The precipitation anomalies in Oceania associated with cEOF1 are weak and not statistically significant (Figure~\ref{fig:pp-oceania} column 1) whereas cEOF2 is associated with large and widespread anomalies.
-The Real cEOF2 is associated with positive seasonal anomalies in Northern and Eastern Australia, and negative ones over New Zealand, while precipitation anomalies associated with Imaginary cEOF2 are negative over all Eastern Australia.
-As it could be expected, these negative anomalies are similar to the springtime precipitation anomalies observed in relation with the Asymmetric SAM \citep{campitelli2021} and ENSO \citep{cai2011}.
-\textcolor{red}{Acá te faltaría  sacar un poco más el jugo asociando con la figura 11. Además, ¿revisaste si podes también usar algún resultado asociado a los PSA? }
+Temperature anomalies associated with the Imaginary cEOF (Fig.
+\ref{fig:pp-temp-2}.b1) show positive values in the tropical Pacific, consistent with SSTs anomalies associated with the same phase (Fig.
+\ref{fig:sst-psi-2}.b1).
 
-\begin{figure}
-\centering
-\includegraphics{../figures/australia-pp-phase-1.pdf}
-\caption{\label{fig:australia-pp-phase}Same as Figure~\ref{fig:enso-phase} but for precipitation anomalies averaged between 130°E and 155°E and 40°S and 15°S (box shown in Figure~\ref{fig:pp-oceania} column b) in \(\frac{\mathrm{mm}}{\mathrm{day}}\).
-Black line is the fit precipitation \textasciitilde{} sin(phase) + cos(phase) computed by weighted OLS using the magnitude of the cEOF2 as weights.}
-\end{figure}
+At higher latitudes there is a wave-like pattern of positive and negative values that coincide with the nodes of the 850hPa geopotential height regression patterns.
+This is consistent with temperature anomalies produced by meridional advection of temperature by the meridional winds arising from geostrophic balance.
 
-To better understand the relationship between precipitation anomalies over Eastern Australia and the cEOF2, Figure~\ref{fig:australia-pp-phase} plots precipitation anomalies averaged over Eastern Australia (box in Figure~\ref{fig:pp-oceania}) as a function of cEOF2 phase.
-The black line shows a sinusoidal fit, whose coefficients are fitted by weighted least squares using the magnitude of the EFO2 as weight.
-The \(r^2\) corresponding to the fit is 0.41 (p-value \textless{} 0.001).
-There is a strong relationship between the phase of the cEOF2 and precipitation anomalies in Eastern Australia, but the Real and Imaginary phases we chose are not exactly aligned with the direction that maximises this relationship.
-To perform a more in-depth analysis of the relationship between this pattern and precipitation in this region, it would be recommended to align one of the axis (either the Real or the Imaginary) to the phase that maximises the described relationship.
+Over the continents, the Imaginary cEOF2 (Fig.
+\ref{fig:pp-temp-2}b.1) is associated with positive temperatures in southern Australia and negative temperature in southern South America and the Antarctic Peninsula.
 
-\begin{figure}
-\centering
-\includegraphics{../figures/pp-america-1.pdf}
-\caption{\label{fig:pp-america}As Figure~\ref{fig:pp-oceania} but for South America}
-\end{figure}
+The temperatures anomalies associated with the Real cEOF2 (Fig. \ref{fig:pp-temp-2}d.1) are less extensive.
+There is no significant signal over the continents except for positive anomalies near the Antarctic Peninsula.
 
-Figure~\ref{fig:pp-america} shows regression maps of seasonal precipitation with each cEOF in South America.
-cEOF1 is associated with negative regressions coefficients for precipitation in Southern Brazil and Paraguay, although these are not statistically significant (Fig.~\ref{fig:pp-america} column 1).
-The strongest precipitation regression coefficients are the ones associated with the Imaginary cEOF2.
-The positive regression coefficients on Southeastern South America (SESA) and Central Chile, and negative ones over South Atlantic Convergence Zone is a well-known springtime precipitation signature of ENSO \citep{cai2020a} and it is also similar to the precipitation anomalies associated with the A-SAM \citep{campitelli2021}.
-\textcolor{red}{Lo mismo que te puse antes. Te faltaría  sacar un poco más el jugo asociando con la figura 11. Además, ¿revisaste si podes también usar algún resultado asociado a los PSA? }
+Tropical precipitation anomalies associated with the Imaginary cEOF2 are strong, with positive anomalies in the central Pacific and western Indian, an negative anomalies in the eastern Pacific.
+This pattern is consistent with the SST anomalies shown before as the increased SST enhances tropical convection and vice versa.
+
+On the extra-tropics, the Imaginary cEOFs is correlated with drier conditions over eastern Australia; a similar signal as the one associated with ENSO \citet{cai2011}.
+However, the Imaginary cEOF2 is not the direction most correlated with precipitation in that area.
+The 135º direction (positive Imaginary + negative Real cEOF2) is associated with stronger and more extensive correlations with precipitation over Australia and New Zealand.
+This Australian signal is most likely there due to the more direct impacts of ENSO rather than on the interconnection pattern represented by the cEOF2.
+
+Over South America, the Imaginary cEOF has positive correlations with precipitation in South Eastern South America (SESA) and central Chile, and negative correlations in eastern Brazil.
+This again is a well-known springtime precipitation signature of ENSO \citep{cai2020a} and it is also similar to the precipitation anomalies associated with the A-SAM \citep{campitelli2021}.
 
 This is not surprising considering the close relationship between the ONI, the A-SAM index and the Imaginary cEOF2 shown previously, but further consolidates the identification of this mode with the PSA pattern.
-On the other hand, the Real cEOF2 is associated with negative precipitation anomalies in a smaller area of SESA.
 Resembling the relationship between ONI and the phase of cEOF2 (Fig.~\ref{fig:enso-phase}), there is a cEOF2 phase dependence of the precipitation anomalies in SESA (not shown).
-These variables could be related wth a significant sinusoidal fit that has a statistically significant coefficient of determination of~0.27.
+These variables could be related with a significant sinusoidal fit that has a statistically significant coefficient of determination of~0.3.
+
+The precipitation signal associated with the Real cEOF2 is less strong.
+There is a residual positive correlation in the equatorial eastern Pacific and small, not statistically significant positive correlations over eastern Australia and negative ones over New Zealand.
+
+Lastly, the cEOF1 doesn't have a significant signal associated with 2-metre temperature or precipitacion (Fig. \ref{fig:pp-temp-cap-1}).
 
@@
 Essentially, it is an alternate representation of the PSA1 and PSA2 patterns \citep{mo2001}.
-We show that the Imaginary cEOF2 can be identified with the PSA1 and the Real cEOF2 with the PSA2.
-This cEOF prefers a phase aligned with the Imaginary cEOF2, because when ENSO is active, it forces the cEOF2 to be in the Imaginary phase.
+We show that the Imaginary cEOF2 can be roughly identified with the PSA1 and the Real cEOF2 with the PSA2.
+
+There is some disagreement in the literature of whether the phase of the PSA pattern is affected by the location of tropical SST anomalies \citep{irving2016}.
+We show not only that the cEOF2 tends to be in the imaginary phase (\textasciitilde PSA1) when the ENSO region is warm, but that central Pacific SST anomalies tend to move the cEOF towards the negative real phase and eastern Pacific SST anomalies tend to move it towards positive real phase.
 When ENSO is neutral, the cEOF2 is still active, but with no particular preferred phase.
 This mirrors the results of \citet{cai2002}, who showed that the CSIRO Model can create PSA-like variability even in the absence of ENSO forcing (with a climatological run), but the variability of one of the PSA modes was enhanced when adding the ENSO signal.
-The magnitude of the cEOF2 appears not to be related to the magnitude of ENSO, except frot the three strongest ENSO events in the period which coincide with the three strongest cEOF2 years.
+The magnitude of the cEOF2 appears not to be related to the magnitude of ENSO, except for the three strongest ENSO events in the period which coincide with the three strongest cEOF2 years.
+
+The sensitivity of the phase of the cEOF2 to the location of the tropical SST anomalies was also seen by \citet{ciasto2015}, who detected similar Rossby wave patterns associated with central Pacific and eastern Pacific SST anomalies but with a change in phase.
 
@@
 That is, look for the rotation of complex EOF that maximises the relationship and use that as the relevant index for study or forecast.
+
+Some aspects of our work are in direct contrast with \citet{irving2016} analysis of the PSA.
+They didn't find neither a strong association between PSA-like variability and ENSO, nor a relationship between it and Australian precipitation.
+Part of the disagreement could be due to the fact that we focus data only for SON, which is when the teleconnection pattern is strongest, instead of the whole year.
+\citet{irving2016} use of ERA5 forectasted precipitation instead of the more observationally constrained CMAP dataset might also be an issue.
+But in light of the previous discussion on the importance of the phase of the cEOF2 in its relationship with ENSO and with precipitation impacts, a big difference could be fact that \citet{irving2016}'s methodology is insensitive to phase changes of the PSA-like variability.
 
@@
 
-\hypertarget{references}{%
-\section{References}\label{references}}
+\appendix
+
+\counterwithin{figure}{section}
+
+\hypertarget{extra-figures}{%
+\section{Extra figures}\label{extra-figures}}
+
+\newpage
+
+\begin{figure*}
+\includegraphics{../figures/sst-psi-1-1} \caption{Same as Figure~\ref{fig:sst-psi-2} but for cEOF1.}\label{fig:sst-psi-1}
+\end{figure*}
+
+\begin{figure}
+\centering
+\includegraphics{../figures/pp-temp-1-1.pdf}
+\caption{\label{fig:pp-temp-1}Same as Figure~\ref{fig:pp-temp-2} but for cEOF1.}
+\end{figure}
 

### 2021-09-08 analysis/paper/shceof.Rmd
Agrega plot de varianza explicada

@@
 }
+
+# There's a bunch of issues with s2 right now
+sf::sf_use_s2(FALSE)
 
@@
     scale_x_continuous(NULL, expand = c(0, 0), 
-                       breaks = seq(-180, 180-60, by = 60),
+                       breaks = seq(0, 360, by = 60),
                        labels = LonLabel) +
@@
     coord_sf(ylim = c(-89, 10), expand = FALSE,
-             crs = "+proj=longlat +pm=180 +over",
+             # crs = "+proj=longlat +pm=180 +over",
              default_crs = "+proj=longlat +lon_wrap=180 +over") + 
@@
 
+```{r pp-t2m-r2, fig.cap = "(ref:cap-pp-t2m-r2)", fig.width=3.3*2, fig.height=3.3}
+breaks <- seq(.2, 1, by = .1)
+
+rbind(Precipitation = pp_regr,
+      `2-metre Temperature` = t2m_regr, idcol = "var") %>% 
+  .[angle == 0] %>% 
+  .[is.finite(r.squared)] %>% 
+  # .[cEOF == "cEOF2"] %>% 
+  periodic(lon = c(0, 360)) %>%
+  ggplot(aes(lon, lat)) +
+  geom_contour_fill(aes(z = r.squared, fill = ..level..), breaks = breaks) +
+  geom_contour_tanaka(aes(z = r.squared), breaks = breaks) +
+  # geom_text_contour(aes(z = r.squared, 
+  #                       label = ifelse(..level.. > 0.2, scales::percent(..level..), NA)), 
+  #                   breaks = breaks, size = 2,
+  #                   skip = 1, stroke = 0.15, stroke.colour = "white") +
+  geom_qmap(keep = .015) +
+  scale_fill_divergent_discretised(name = "Explained variance", midpoint = 0,
+                       labels = scales::percent_format(),
+                       guide = guide_colorsteps_bottom(15, show.limits = TRUE)) +
+
+  scale_x_continuous(NULL, expand = c(0, 0),
+                     breaks = seq(0, 360, by = 60),
+                     labels = LonLabel) +
+  scale_y_continuous(NULL, breaks  = seq(-90, 0, by = 30), expand = c(0, 0)) +
+
+  facet_grid(var ~ cEOF) +
+  # coord_quickmap(ylim = c(NA, 10)) +
+  coord_sf(ylim = c(NA, 10),
+           default_crs = "+proj=longlat +lon_wrap=180 +over") + 
+  theme(legend.box = "horizontal",
+        legend.text = element_text(size = rel(0.7)), 
+        legend.box.spacing = grid::unit(0, "lines") ) +
+  tag_facets("rc") 
+```
+
+(ref:cap-pp-t2m-r2) Explained variance ($r^2$) of 2-metre temperature (row a) and precipitation (row b) with cEOF1 (column 1) and cEOF2 (column 2).
+
+The influence of cEOFs variability in continental rainfall and 2-metre temperature in the Southern Hemisphere is also explored. 
+Figure \@ref(fig:pp-t2m-r2) shows the $r^2$ between the cEOF1 and cEOF2 and 2-metre temperature and precipitation. 
+For the cEOF1 there are no significant signals in precipitation and most of the temperature signal is low and restricted to the eastern coast of Antarctica.
+
+There is a strong signal between both variables and cEOF2, with explained variance greater than 50% in some regions. 
+For 2-metre temperature, there are high values in the tropical Pacific and the SPCZ, as well as the region following an arc between New Zealand and the South Atlantic. 
+Over the continents, moderate values of about 30% variance explained can be seen in southern Australia, some regions of South America and the Antarctic Peninsula. 
+For precipitation, there are high values over the tropics. 
+At higher latitudes, moderate values are observed over eastern Australia and some regions of southern South America.
+
+Since the cEOF1 has a relatively weak signal, we will focus on the cEOF2.
+
 ```{r add_continents}
@@
 
-    scale_x_continuous("", expand = c(0, 0), breaks = seq(0, 360, by = 15), 
+    scale_x_continuous("", expand = c(0, 0), breaks = seq(0, 360, by = 30), 
                        labels = LonLabel) +
-    scale_y_latitude(ticks = 15) +
+    scale_y_continuous("", expand = c(0, 0), breaks = seq(-90, 0, by = 15), 
+                       labels = LatLabel) +
+    # scale_y_latitude(ticks = 15, labels = LatLabel) +
     scale_linetype_discrete(guide = "none") +
@@
     tag_facets("rc") +
-    coord_sf(ylim = c(-85, 10),
-             crs = "+proj=longlat +pm=180 +over",
+    coord_sf(ylim = c(NA, 10),
              default_crs = "+proj=longlat +over") +
@@
 
-The influence of cEOFs variability in continental rainfall and 2-metre temperature in the Southern Hemisphere is also explored.
-Figure \@ref(fig:pp-temp-2) show regression maps of 2-meter temperature (column 1) and precipitation (column 2) upon standardised cEOF2 which, again, can be interpreted as the anomalies associates with a 1 standard deviation increase.
+Figure \@ref(fig:pp-temp-2) shows regression maps of 2-meter temperature (column 1) and precipitation (column 2) upon standardised cEOF2 which, again, can be interpreted as the anomalies associates with a 1 standard deviation increase.
 
@@
 
-```{r pp-temp-2, fig.cap="(ref:pp-temp-cap)", fig.width=6.6, fig.height=5.5}
+```{r pp-temp-2, fig.cap="(ref:pp-temp-cap)", fig.width=6.6, fig.height=5.5, dependson="plot_pp"}
 plot_pp(pp_regr[cEOF == "cEOF2"], t2m_regr[cEOF == "cEOF2"], hgt850_regr[cEOF == "cEOF2"])
@@
 
-```{r pp-temp-1, fig.cap="(ref:pp-temp-cap-1)", fig.width=6.6, fig.height=5.5}
+```{r pp-temp-1, fig.cap="(ref:pp-temp-cap-1)", fig.width=6.6, fig.height=5.5, dependson="plot_pp"}
 plot_pp(pp_regr[cEOF == "cEOF1"], t2m_regr[cEOF == "cEOF1"], hgt850_regr[cEOF == "cEOF1"])

### 2021-09-08 analysis/paper/shceof.tex
Render

@@
 {#1}\small\normalsize} \spacingset{1}
+
+\begin{verbatim}
+## Spherical geometry (s2) switched off
+\end{verbatim}
 
@@
 
+\begin{figure}
+\centering
+\includegraphics{../figures/pp-t2m-r2-1.pdf}
+\caption{\label{fig:pp-t2m-r2}Explained variance (\(r^2\)) of 2-metre temperature (row a) and precipitation (row b) with cEOF1 (column 1) and cEOF2 (column 2).}
+\end{figure}
+
 The influence of cEOFs variability in continental rainfall and 2-metre temperature in the Southern Hemisphere is also explored.
-Figure \ref{fig:pp-temp-2} show regression maps of 2-meter temperature (column 1) and precipitation (column 2) upon standardised cEOF2 which, again, can be interpreted as the anomalies associates with a 1 standard deviation increase.
+Figure \ref{fig:pp-t2m-r2} shows the \(r^2\) between the cEOF1 and cEOF2 and 2-metre temperature and precipitation.
+For the cEOF1 there are no significant signals in precipitation and most of the temperature signal is low and restricted to the eastern coast of Antarctica.
+
+There is a strong signal between both variables and cEOF2, with explained variance greater than 50\% in some regions.
+For 2-metre temperature, there are high values in the tropical Pacific and the SPCZ, as well as the region following an arc between New Zealand and the South Atlantic.
+Over the continents, moderate values of about 30\% variance explained can be seen in southern Australia, some regions of South America and the Antarctic Peninsula.
+For precipitation, there are high values over the tropics.
+At higher latitudes, moderate values are observed over eastern Australia and some regions of southern South America.
+
+Since the cEOF1 has a relatively weak signal, we will focus on the cEOF2.
+
+Figure \ref{fig:pp-temp-2} shows regression maps of 2-meter temperature (column 1) and precipitation (column 2) upon standardised cEOF2 which, again, can be interpreted as the anomalies associates with a 1 standard deviation increase.
 

### 2021-09-08 analysis/paper/shceof.Rmd
Saca mensaje de sf_s2
Render

@@
   comment = "#>",
-  cache.extra = 43,
+  cache.extra = 41,
   cache.path = "../cache/paper/",
@@
 # There's a bunch of issues with s2 right now
-sf::sf_use_s2(FALSE)
+sink <- capture.output(sf::sf_use_s2(FALSE))
 

### 2021-12-14 analysis/paper/shceof.Rmd
Actualiza cosas

@@
 output: 
+  bookdown::word_document2: default
   bookdown::pdf_book: 
@@
       # pandoc_args: ["--lua-filter=abstract-to-meta.lua"]
+always_allow_html: true
 header-includes: 

### 2022-01-24 analysis/paper/shceof.Rmd
Arregla un par de bugs.

@@
               show.legend = FALSE, size = 0.8) +
-  scale_x_datetime(NULL, date_labels = "%Y", 
-                   breaks = seq(as.POSIXct("1950-01-01"), 
-                                as.POSIXct("2020-01-01"), by = "10 years"),
-                   minor_breaks = seq(as.POSIXct("1950-01-01"), 
-                                      as.POSIXct("2020-01-01"), by = "2 years")) +
+  scale_x_date(NULL, date_labels = "%Y",
+               breaks = seq(as.Date("1950-01-01"),
+                            as.Date("2020-01-01"), by = "10 years"),
+               minor_breaks = seq(as.Date("1950-01-01"),
+                                  as.Date("2020-01-01"), by = "2 years")) +
   scale_y_continuous(NULL, minor_breaks = NULL, sec.axis = dup_axis()) +
@@
 ```{r lable_placement_min}
-lable_placement_min <- function(direction = c("vertical", "horizontal"))  {
+label_placement_min <- function(direction = c("vertical", "horizontal"))  {
   direction <- direction[1]
   function(x, y) {
+    browser()
+    y <- x[[1]]$y
+    x <- x[[1]]$x
+
     if (direction == "vertical") {
@@
                     skip = 1, stroke = 0.15, stroke.colour = "white",
-                    label.placement = lable_placement_min(),
+                    # label.placement = label_placement_min(),
                     breaks = AnchorBreaks(0, exclude = 0)) +
@@
   coord_cartesian(clip = "off") +
-  tag_facets()
+  tag_facets() +
+  NULL
 ```

### 2022-01-24 analysis/paper/shceof.Rmd
Agrega comentarios de Carolina

@@
 ---
-title: "Revisiting the Austral Spring Extratropical Southern Hemisphere zonally assymetric circulation using complex Empirical Orthogonal Functions"
+title: "Revisiting the Austral Spring Extratropical Southern Hemisphere zonally asymmetric circulation using complex Empirical Orthogonal Functions"
 titlerunning:  Title
@@
 
-The large-scale extratropical circulation in the Southern Hemisphere is strongly zonally symmetric, but its zonal departures are highly relevant for surface impacts and could be related to high-impact climate extremes.
-Zonal asymmetries of extratropical circulation in the Southern Hemisphere strongly modulate weather systems and regional climate through latitudinal transport of heat, humidity, and momentum [@trenberth1980].
-
-One of the main features of the zonally asymmetric circulation are the Pacific--South American Patterns [@mo2001], which are associated with precipitation anomalies in South America, particularly in the September-October-November trimester (cita).
-Besides, @raphael2007 suggests that variability in the planetary wave 3 projected onto its climatological location is associated with Antarctic sea ice concentration anomalies.
-@irving2015 showed that planetary wave amplitude in general is associated sea ice concentration, as well as temperature and precipitation anomalies.
-
-Some of these patterns of variability associated to the zonally asymmetric circulation appear to have experienced secular changes, although to our knowledge this has not been extensively studied.
+The large-scale extratropical circulation in the Southern Hemisphere (SH) is strongly zonally symmetric, but its zonal departures are highly relevant for regiona limpacts and could even be related to the occurrence of high-impact climate extremes (REFERENCE).
+Zonal asymmetries of extratropical circulation in the Southern Hemisphere strongly modulate weather systems and regional climate through promoting longitudinally different latitudinal transport of heat, humidity, and momentum [@trenberth1980].
+
+Well-known features of the zonally asymmetric circulation in the SH are the Pacific–South American Patterns [@mo2001], which are associated with Rossby wave trains influencing precipitation anomalies in regions like South America, particularly in the September-October-November trimester (cita).
+In addition, previous works have identified at extratropical and subpolar latitudes, wave-like patterns with dominant wavenumbers  3--4, also exerting distinctive regional impacts.  
+@raphael2007 shows that variability in the planetary wave 3 projected onto its climatological location is associated with anomalies in the Antarctic sea-ice concentration. 
+@irving2015 showed that planetary wave amplitude in general is associated to sea-ice concentration, as well as to temperature and precipitation anomalies {where?}.
+
+Some of these zonally asymmetric patterns of the SH circulation variability, appear to have experienced secular changes. 
 For instance, @raphael2003 found that the amplitude of the zonal wave 1 experienced a large increase and that the zonal wave 3 experienced changes in its annual cycle between 1958 and 1996. 
-
-One typical way of describing the zonally asymmetric circulation is by the amplitude and phase of zonal waves obtained by Fourier decomposition of geopotential height at each latitude [e.g.
+However, little is known yet about variability and trends of these patterns.
+
+Zonally asymmetric circulation is typically described by the amplitude and phase of zonal waves obtained by Fourier decomposition of geopotential heights or sea-level pressures at each latitude  [e.g.
 @vanloon1972; @trenberth1980; @turner2017].
-This approach suggests that zonal waves 1 and 3 explained most of the variance of the tropospheric extratropical Southern Hemisphere circulation (@vanloon1972).
-However, this methodology relies on the assumption that the circulation can be meaningfully understood in terms of zonal waves of constant amplitude along a latitude circle.
-This is not valid for meridionally propagating waves or zonal waves with localised amplitudes.
-In the case of the wave 3, for example, @trenberth1985 observed that it played a role in blocking events, but this was mostly due to increased amplitude of a longitudinally localised wave train instead of an hemispheric-scale zonal wave 3.
-
-The fourier technique can be generalised to integrate all planetary wave intensity irregardless of wave number by computing the wave envelope as done by @irving2015.
-This makes it possible to represent planetary waves with different amplitude at different longitudes, but it removes all information about phase and wave number.
-
-Another approach to characterising the Southern Hemisphere circulation is by using Empirical Orthogonal Functions (EOF, also known as Principal Component Analysis).
-Within the EOF framework, the Southern Annular Mode (SAM) appears as the leading mode of variability of the tropospheric Southern Hemisphere circulation [@fogt2020] followed by the two PSA [@mo2001].
-The SAM represents a relatively zonally symmetric pattern of alternating low pressures in polar latitude and a ring of high pressures in high latitudes with an embedded wave 3 pattern that is more prominent in the Pacific sector.
-The PSA1 and PSA2 describe the two out-of-phase sides of a meridionally propagating wave train that originates in the eastern equatorial Pacific and travels towards the South Atlantic following a great-circle arch along the Antarctic Peninsula.
-These patterns are derived by applying EOF to temporal anomalies, but @raphael2003 applied EOF methods specifically to zonal anomalies.
-
-EOFs are more flexible than Fourier decomposition modes in the sense that they can capture oscillation patterns that cannot be characterised by purely sinusoidal waves with constant amplitude.
-Nonetheless, they are restricted to standing oscillation modes and could not represent properly propagating or phase-varying modes such as zonal waves.
+This approach suggests that zonal waves 1 and 3 explain most of the variance of the SH tropospheric extratropical circulation (@vanloon1972). 
+However, this methodology relies on the assumption that the circulation can be meaningfully described in terms of zonal waves of constant amplitude along a latitude circle. 
+However, this is not valid for meridionally propagating waves or zonal waves with localised amplitudes.
+In the case of the wave 3, for example, @trenberth1985 concluded that it plays a role in the development of blocking events, but this was mostly due to increased amplitude of a longitudinally localised wave train instead of an hemispheric-scale zonal wave 3.
+
+The Fourier technique can be generalised to integrate all planetary wave amplitude? regardless of wave number, by computing the wave envelope @irving2015.
+The latter makes possible to represent planetary waves with different amplitude at different longitudes, but it removes all information about phase and wave number.
+
+Another extensively used approach to characterize the SH circulation anomalies, is by computing Empirical Orthogonal Functions (EOF, also known as Principal Component Analysis). 
+Within the EOF framework, the Southern Annular Mode (SAM) appears as the leading mode of variability of the SH circulation [@fogt2020] followed by two PSA-like patterns (Mo and Paegle 2001). 
+SAM represents a relatively zonally symmetric pattern of alternating low pressures in polar latitude and a ring of high pressures in high latitudes with an embedded wave 3 pattern that is more prominent in the Pacific sector.
+Moreover, the 2nd and 3rd EOFs, usually known as PSA1 and PSA2 patterns, respectively, describe meridionally propagating wave trains that originates in the eastern equatorial Pacific and Australian-Indian Ocean sector, and travel towards the South Atlantic following a great-circle arch along the Antarctic coast.
+These patterns are usually derived by applying EOF to temporal anomalies, but @raphael2003 also applied EOF methods specifically to zonal anomalies.
+
+Patterns resulting from EOF analyses are more flexible than Fourier decomposition derived modes in the sense that they can capture oscillation patterns that cannot be characterised by purely sinusoidal waves with constant amplitude.
+Nonetheless, they are restricted to standing oscillation modes and could not represent properly propagating or phase-varying features such as zonal waves. 
 A single EOF can also represent a mixture of two or more physical modes.
 The inability of a single EOF to describe travelling waves is what forces the PSA pattern to be described by two EOFs. 
-@irving2016 characterises this pattern in a Fourier framework.
-Being a meridionally propagating mode, the PSA cannot be correctly characterised by Fourier decomposition at each latitude circle.
-As a result, they reprojected meridional wind fields so that the path of the PSA laid on the equator.
-On this new projection, they could identify the PSA using Fourier decomposition.
-
-Finally, another, hybrid, methodology commonly used consist on identifying particular features of interest and creating indices using simple methods such as averages and differences.
-The Marshal SAM Index [@marshall2003], for example, tries to characterise the SAM by the average difference of sea level pressure between 40ºS and 65ºS, following @gong1999.
-@raphael2004 used a similar methodology to describe the wave 3 in the Southern Hemisphere.
-Instead of using Fourier to compute the amplitude and phase of the zonal wave, they averaged standardised geopotential height anomalies in three points representing the location of the ridges of the zonal wave 3 of the climatological mean.
-This is equivalent to an index of the amplitude of the zonal wave 3 that projects into the phase of the zonal wave 3 of the mean field.
-Similarly, @hobbs2010 propose that the zonally asymmetric circulation in the Southern Hemisphere can be described by the strength and location of two anticyclones located in the sub-Antarctic western hemisphere.
+@irving2016 characterise this pattern in a Fourier framework. Being a meridionally propagating mode, the PSA cannot be correctly characterised by Fourier decomposition at each latitude circle.
+As a result, they reprojected meridional wind fields so that the path of the PSA laid on the equator. On this new projection, they could identify the PSA using Fourier decomposition.
+
+Finally, another, hybrid methodology commonly used consists on identifying particular features of interest and creating indices using simple methods such as averages and differences.
+The SAM Index defined by [@marshall2003] following Gong and @gong1999, for example, describes SAM variability by the averaging the difference of sea level pressure between 40ºS and 65ºS.
+On the other hand, @raphael2004 used a similar methodology to describe the wave 3 activity in the SH circulation. 
+Averaging standardised geopotential-height anomalies at three locations of the ridges associated to the zonal wave 3 of the climatological mean. 
+By construction, the index is equivalent to projects the amplitude of the zonal wave 3 on the phase of the zonal wave 3 of the climatological mean field. 
+Similarly, @hobbs2010 propose that the zonally asymmetric circulation in the SH can be described by the strength and location of two anticyclones located in the sub-Antarctic western hemisphere. 
 These methods are grounded on other methods to identify the centres of action for the described phenomena and can be useful to characterise features that are not readily apparent with Fourier or EOF methods.
-The use of data of only specific regions or locations makes them suitable for easy computation and for extending back into the data-poor past, but also makes them not very robust and unable to capture non-stationary patterns.
-
-Our objective is to improve our understanding of planetary waves in the Southern Hemisphere using an index that is able to capture the phase-varying nature of planetary waves and is flexible enough to describe planetary waves with non-constant amplitude in each latitude circle.
+Using information at specific regions or locations makes easy the computation of indices like those, but they are not very robust or unable to capture non-stationary patterns.
+
+The general goal of this study is to improve the description and understanding of planetary waves in the SH using an index that is able to capture the phase-varying nature of planetary waves and is flexible enough to describe planetary waves with non-constant amplitude in each latitude circle.
 We use complex Empirical Orthogonal Functions [@horel1984] as a robust extension of EOF analysis that is capable of describing rich structures with meridional and zonal propagation, and amplitude modulation such as the PSA.
 
-We apply this method to characterise the Springtime zonally asymmetric circulation of the extratropical Southern Hemisphere.
-To take vertical distribution into account, we analyse both the troposphere and the stratosphere.
+We apply this method to characterise the springtime zonally asymmetric circulation of the extratropical SH circulation. 
+In addition to take vertical distribution into account, we analyse both the troposphere and the stratosphere.
 
@@
 We used monthly geopotential height, air temperature, ozone mixing ratio, and total ozone column (TOC) at 2.5\degree longitude by 2.5\degree latitude of horizontal resolution and 37 vertical isobaric levels from the European Centre for Medium-Range Weather Forecasts Reanalysis version 5 [ERA; @era5] for the period 1979-2019.
-Most of our analysis was restricted to the post-satellite era to avoid any confounding factors arising from the incorporation of satellite observations, but we also used the preliminary back extension of ERA5 from 1950 to 1978 [@era5be] to look at long-term trends.
-Streamfunction was derived from ERA5 vorticity at 200 hPa using the FORTRAN subroutine FISHPACK [@fishpack] and horizontal wave activity fluxes was computed following @plumb1985. 
-We used Sea Surface Temperature (SST) monthly fields from Extended Reconstructed Sea Surface Temperature (ERSST) v5 [@huang2017] and precipitation monthly data from the CPC Merged Analysis of Precipitation [@cmap], with a 2.5\degree resolution in latitude and longitude.
-This rainfall gridded dataset is based on information from different sources such as rain gauge observations, satellite inferred estimations and the NCEP-NCAR reanalysis, and it is available since 1979 to present.
+Most of our analysis is restricted to the post-satellite era to avoid any confounding factors arising from the incorporation of satellite observations, but we also used the preliminary back extension of ERA5 from 1950 to 1978 [@era5be] to describe long-term trends.
+We derived streamfunction at 200 hPa from ERA5 vorticity using the FORTRAN subroutine FISHPACK [@fishpack] and we computed horizontal wave activity fluxes following @plumb1985. 
+Sea Surface Temperature (SST) monthly fields are from Extended Reconstructed Sea Surface Temperature (ERSST) v5 [@huang2017] and precipitation monthly data from the CPC Merged Analysis of Precipitation [@cmap], with a 2.5\degree resolution in latitude and longitude.
+The rainfall gridded dataset is based on information from different sources such as rain gauge observations, satellite inferred estimations and the NCEP-NCAR reanalysis, and it is available since 1979 to present.
 
@@
 The amplitude of the zonal waves was obtained through the Fourier transform of the spatial field at each latitude circle. 
-We computed the amplitude and phase of the zonal wave 1 by averaging (area-weighted) each variable between `r knitr::combine_words(LatLabel(o3wave_lats))` for each SON and extracting the wave-1 component of the Fourier spectrum.
-We chose this latitude band because it is wide enough to capture most of the relevant anomalies for southern hemisphere mid-latitudes.
-
-We computed the level-dependent SAM index as the leading EOF of year-round monthly geopotential height anomalies south of 20ºS at each level for the whole period according to @baldwin2009.
+For the analysis of wave 1, we computed its amplitude and phase by averaging (area-weighted) the data for each variable and each SON between `r knitr::combine_words(LatLabel(o3wave_lats))`, and then extracting the wave-1 component of the Fourier spectrum.
+We chose this latitude band because it is wide enough to capture most of the relevant anomalies of SH mid-latitudes.
+
+We computed the level-dependent SAM index as the leading EOF of year-round monthly geopotential height anomalies south of 20ºS at each level for the whole period according [@baldwin2009].
 We further split the SAM into its zonally symmetric and zonally asymmetric components (S-SAM and A-SAM indices respectively).
 These indices were obtained by projecting the zonally asymmetric and zonally symmetric part of the SAM spatial pattern onto monthly geopotential height fields, as proposed by @campitelli2021.
-Seasonal indices of the Pacific South American patterns (PSA1 and PSA2) were calculated following @mo2001 as the third and fourth leading EOF of seasonal mean anomalies for SH 500 hPa geopotential height combining all seasons.
-In our study, only the SON values were used.
+Seasonal indices of the Pacific South American patterns (PSA1 and PSA2) were calculated -- in agreement with @mo2001 -- as the third and fourth leading EOF of seasonal mean anomalies for SH 500 hPa geopotential height.
 
@@
 
-(ref:eof-naive-cap) Spatial patterns of the leading 4 EOFs of SON zonal anomalies of geopotential height at 50 hPa south of 20\degree S for the 1979 -- 2019 period (arbitrary units).
+(ref:eof-naive-cap) Spatial patterns of the four leading EOFs of SON zonal anomalies of geopotential height at 50 hPa south of 20\degree S for the 1979 -- 2019 period (arbitrary units).
 
@@
 
-In traditional EOF analysis zonal waves appear as pairs of EOFs, that could be degenerated representing similar patterns but shifted in phase [@horel1984].
-For instance, Figure \@ref(fig:eof-naive) shows the leading 4 EOFs of SON geopotential height zonal anomalies at 50 hPa south of 20\degree S.
+In standard way of applying EOF analysis zonal waves may appear as pairs of EOFs, that could be degenerated representing similar patterns but shifted in phase [@horel1984].
+Figure \@ref(fig:eof-naive) shows the four eading EOFs of SON geopotential height zonal anomalies at 50 hPa south of 20\degree S.
 It is clear that the first two EOFs represent the same zonal wave 1 pattern and the last two represent a same zonal wave pattern with shorter wavenumber and four centers of action shifted by 1/4 wavelength.
@@
 
-Complex Empirical Orthogonal Functions (cEOF) is a useful method to characterise zonal waves, considering phase-varying structures [@horel1984].
+ON the other hand, complex Empirical Orthogonal Functions (cEOF) is a useful method to characterise zonal waves, associated with phase-varying structures [@horel1984].
 This method involves computing EOF on the analytic representation of the original field.
-This representation is a complex field in which the real part is the original variable and the imaginary part is the original data shifted by 90\degree at each spectral frequency -- i.e.
+That representation is a complex field in which the real part is the original data and the imaginary part is the original data shifted by 90\degree at each spectral frequency -- i.e.
 its Hilbert transform.
 The Hilbert transform is usually understood in terms of time-varying signal.
-However, in this work we apply the Hilbert transform to each latitude circle at each moment in time.
+However, in this work we apply the Hilbert transform at each latitude circle and at each considered time.
 Since each latitude circle is a periodic domain, this procedure does not suffer from edge effects.
@@
 The result of the cEOF methodology is a set of complex spatial patterns and complex time series.
-The real and imaginary part of each spatial pattern represent two phases wave-like spatial patterns that are in quadrature.
+The real and imaginary part of each spatial pattern represent the two phases of wave-like spatial patterns that are in quadrature.
 The magnitude and argument of each complex time series represent the amplitude and phase of each zonal wave.
@@
 
-(ref:corr-ceof-splitted-cap) Coefficient of determination ($R^2$) of the absolute magnitude of complex EOFs between 200 hPa and 50 hPa computing EOF separately for each level.
-p-values lower than 0.01 in bold.
+(ref:corr-ceof-splitted-cap) Coefficient of determination ($R^2$) between the absolute magnitude of complex EOFs computed separately at 200 hPa and 50 hPa (p-values lower than 0.01 in bold).
 
@@
 
-The cEOF methodology is applied to SON geopotential height zonal anomalies south of 20\degree S at 50 and 200 hPa, representing a stratsopheric and upper troposhperic level respectively.
-Table \@ref(tab:corr-ceof-splitted) shows the coefficient of determination between time series of the amplitude of each complex EOF (cEOF) across levels.
-There's a high degree of correlation between the magnitude of the respective cEOF1 and cEOF2 at each level.
+The cEOF methodology is applied to SON geopotential height zonal anomalies south of 20\degree S separately at 50 and 200 hPa.
+Table \@ref(tab:corr-ceof-splitted) shows the coefficient of determination between time series of the amplitude of each cEOF across levels.
+There is a high degree of correlation between the magnitude of the respective cEOF1 and cEOF2 at each level.
 The spatial patterns of the 50 hPa and 200 hPa cEOFs are also similar (not shown).
@@
 This motivates the decision of performing complex EOF jointly between levels.
-The computation of the cEOFs was carried out using data from both levels at the same time, therefore, each cEOF has a spatial component that depends on longitude, latitude and level, and a temporal component that depends only on time.
+Therefore cEOFs were computed using data from both levels at the same time.
+In that sense each cEOF has a spatial component that depends on longitude, latitude and level, and a temporal component that depends only on time.
 
@@
 
-We performed linear regressions to quantify the association between the cEOFs and other variables.
-For each cEOF, we create regression maps by fitting a multiple linear model involving both the real and the imaginary part.
-To obtain the linear coefficients of a variable $X$ (geopotential, temperature, precipitation, etc...) with the Imaginary and Real parts of each cEOF we fit the equation
+We performed linear regressions to quantify the association between the cEOFs and other variables (e.g. geopotential height, temperature, precipitation, and others).
+For each cEOF, we computed regression maps by fitting a multiple linear model involving both the real and the imaginary part.
+To obtain the linear coefficients of a variable $X$  with the imaginary and real parts of each cEOF we fit the equation
 
@@
 
-where $\lambda$ and $\phi$ are the longitude and latitude, $t$ is the time, $\alpha$ and $\beta$ are the linear regression coefficients for Imaginary and Real parts respectively, $X_0$ and $\epsilon$ are the constant and error terms respectively.
+where $\lambda$ and $\phi$ are the longitude and latitude, $t$ is the time, $\alpha$ and $\beta$ are the linear regression coefficients for imaginary and real parts respectively, $X_0$ and $\epsilon$ are the constant and error terms respectively.
 
@@
 (ref:extended-series-cap) Time series extended using ERA5 back extended preliminary edition (period 1950 -- 1978) and ERA5 (period 1979 -- 2019).
-Dark straight line is the OLS trend.
-Black horizontal and vertical line mark the mean value and range of each time series respectively.
+Dark straight line is the linear trend.
+Black horizontal and vertical line mark the mean value and range of each time series, respectively.
 
@@
 The cEOF2 (Fig. \@ref(fig:eof-naive) column b) shows also a zonal wave-like structure with maximum amplitude at high latitudes, but with shorter spatial scales.
-In particular, its dominant structure at both levels is a wave 3 but with larger amplitude in the pacific sector.
+In particular, the dominant structure at both levels is a wave 3 but with larger amplitude in the pacific sector.
 This modulated amplitude is especially apparent at 200 hPa.
@@
 However, cEOF2's mean is almost zero, which indicates that only cEOF1 includes variability that significantly projects onto the mean zonally anomalous field.
-There are no significant simultaneous correlation between the time series.
+There is no significant simultaneous correlation between the time series.
 Both cEOFs show year-to-year variability but show no evidence of decadal variability.
 
-There is only a significant upward trend in the real component of cEOF1 (Fig. \@ref(fig:extended-series)a.1, p-value `r trends[part == "Real" & cEOF == "cEOF1"]$p_report`) and no significant trend in any of the complex components of cEOF2.
+A significant upward trend in the real component of cEOF1 is evident (Fig. \@ref(fig:extended-series)a.1, p-value `r trends[part == "Real" & cEOF == "cEOF1"]$p_report`)  while there is  no significant trend in any of the complex components of cEOF2.
 The positive trend in the Real cEOF1 translates into a positive trend in cEOF1 magnitude, but not in the phase (not shown).
 This long-term change indicates an increase in the magnitude of the high latitude zonal wave 1.
-A similar observation was made by @raphael2003, who detected a step after around 1975 in the leading EOF of August-September-October 500 hPa zonal geopotential height anomalies, which is similar to the cEOF1.
+In agreement, @raphael2003 detected a step change after around 1975 in the temporal evolution of leading EOF computed from August-September-October 500 hPa zonal geopotential height anomalies, which is similar to the cEOF1.
 
@@
 
-The spatial patterns shown in Figure \@ref(fig:ceofs-1) are derived by removing the zonally symmetric circulation, so they might not include all the variability that is actually associated with the cEOF time series.
-To understand the full geopotential height field changes associated with each cEOF, we computed regression patterns of the full geopotential fields onto each cEOF.
+In the previous section, cEOFs were applied to zonal anomalies, that is anomalies derived by removing the zonally mean values in order to isolate the main characteristics of the main zonal waves characterizing the circulation in the SH.
+In this section, regression fields were computed using the full fields of the variables in order to describe the influence of the cEOFs on the temporal anomalies.
 
@@
 Their phases coincide with the ones shown in Figure \@ref(fig:ceofs-1)a.1, with the positive centre of the Real cEOF1 located towards the dateline, and the one of the Imaginary cEOF1 located over East Antarctica.
-However, the Real cEOF1 pattern is substantially altered by the zonally symmetric circulation.
-Instead of a clear wave 1 pattern, the regression pattern can be better describes as a monopole with its centre displaced from the South Pole.
-
-The regression patterns for cEOF1 at 200 hPa (Figure \@ref(fig:eof1-regr-gh) row b) are similarly influenced by the zonally symmetric circulation.
+Instead of a wave-1 pattern, regressed anomalies describe a monopole with its centre displaced from the South Pole. The latter indicates an influence of cEOF1 in modulating the polar vortex.
+
+The regression maps for cEOF1 at 200 hPa (Figure \@ref(fig:eof1-regr-gh) row b) are similarly influenced by the zonally symmetric circulation.
 It is only possible to partially distinguish the wave 1 pattern in relation with the Real cEOF1 (Figure \@ref(fig:eof1-regr-gh)b.1).

### 2022-03-15 analysis/paper/shceof.Rmd
Agrega comentarios de Leandro en la intro.

@@
 output: 
-  bookdown::word_document2: default
+  # bookdown::word_document2: default
   bookdown::pdf_book: 
@@
       latex_engine: "xelatex"
-      citation_package: "natbib"
+      # citation_package: "natbib"
       # pandoc_args: ["--lua-filter=abstract-to-meta.lua"]
@@
                                 bg.colour = "white")
-
   )
@@
 coord_polar <- function(ymax = -20, ...) {
-
   x <- c(seq(0, 360, length.out = 40), 
@@
 
-The large-scale extratropical circulation in the Southern Hemisphere (SH) is strongly zonally symmetric, but its zonal departures are highly relevant for regiona limpacts and could even be related to the occurrence of high-impact climate extremes (REFERENCE).
-Zonal asymmetries of extratropical circulation in the Southern Hemisphere strongly modulate weather systems and regional climate through promoting longitudinally different latitudinal transport of heat, humidity, and momentum [@trenberth1980].
-
-Well-known features of the zonally asymmetric circulation in the SH are the Pacific–South American Patterns [@mo2001], which are associated with Rossby wave trains influencing precipitation anomalies in regions like South America, particularly in the September-October-November trimester (cita).
-In addition, previous works have identified at extratropical and subpolar latitudes, wave-like patterns with dominant wavenumbers  3--4, also exerting distinctive regional impacts.  
+The large-scale extratropical circulation in the Southern Hemisphere (SH) is strongly zonally symmetric, but its zonal departures are highly relevant for regional impacts [e.g. @hoskins2005].
+Zonal asymmetries of extratropical circulation in the Southern Hemisphere strongly modulate weather systems and regional climate through promoting longitudinally different latitudinal transport of heat, humidity, and momentum [@trenberth1980a; @raphael2007] and could even be related to the occurrence of high-impact climate extremes [@pezza2012].
+
+Zonally asymmetric circulation is typically described by the amplitude and phase of zonal waves obtained by Fourier decomposition of geopotential heights or sea-level pressures at each latitude [e.g.
+@vanloon1972; @trenberth1980a; @turner2017].
+This approach suggests that zonal waves 1 and 3 explain almost 99% of the total variance in the annual mean 500-hPa pattern at 50ºS [@vanloon1972].
+@trenberth1985 concluded that wave 3 plays a role in the development of blocking events.
+In addition, previous works have identified at extratropical and subpolar latitudes, wave-like patterns with dominant wavenumbers 3-4, also exerting distinctive regional impacts.
 @raphael2007 shows that variability in the planetary wave 3 projected onto its climatological location is associated with anomalies in the Antarctic sea-ice concentration.
-@irving2015 showed that planetary wave amplitude in general is associated to sea-ice concentration, as well as to temperature and precipitation anomalies {where?}.
-
-Some of these zonally asymmetric patterns of the SH circulation variability, appear to have experienced secular changes. 
-For instance, @raphael2003 found that the amplitude of the zonal wave 1 experienced a large increase and that the zonal wave 3 experienced changes in its annual cycle between 1958 and 1996. 
-However, little is known yet about variability and trends of these patterns.
-
-Zonally asymmetric circulation is typically described by the amplitude and phase of zonal waves obtained by Fourier decomposition of geopotential heights or sea-level pressures at each latitude  [e.g.
-@vanloon1972; @trenberth1980; @turner2017].
-This approach suggests that zonal waves 1 and 3 explain most of the variance of the SH tropospheric extratropical circulation (@vanloon1972). 
-However, this methodology relies on the assumption that the circulation can be meaningfully described in terms of zonal waves of constant amplitude along a latitude circle. 
+
+The Fourier decomposition relies on the assumption that the circulation can be meaningfully described in terms of zonal waves of constant amplitude along a latitude circle.
 However, this is not valid for meridionally propagating waves or zonal waves with localised amplitudes.
-In the case of the wave 3, for example, @trenberth1985 concluded that it plays a role in the development of blocking events, but this was mostly due to increased amplitude of a longitudinally localised wave train instead of an hemispheric-scale zonal wave 3.
-
-The Fourier technique can be generalised to integrate all planetary wave amplitude? regardless of wave number, by computing the wave envelope @irving2015.
+Addressing this limitation, the Fourier technique can be generalized to integrate all planetary wave amplitude regardless of wave number, by computing the wave envelope [@irving2015].
 The latter makes possible to represent planetary waves with different amplitude at different longitudes, but it removes all information about phase and wave number.
-
-Another extensively used approach to characterize the SH circulation anomalies, is by computing Empirical Orthogonal Functions (EOF, also known as Principal Component Analysis). 
-Within the EOF framework, the Southern Annular Mode (SAM) appears as the leading mode of variability of the SH circulation [@fogt2020] followed by two PSA-like patterns (Mo and Paegle 2001). 
+With this approach, @irving2015 showed that planetary wave amplitude in general is associated to Antarctic sea-ice concentration and temperature, as well as to precipitation anomalies in regions of significant topography in SH mid-latitudes and Antarctica.
+
+Another extensively used approach to characterize the SH tropospheric circulation anomalies, is by computing Empirical Orthogonal Functions (EOF, also known as Principal Component Analysis).
+Within the EOF framework, the Southern Annular Mode (SAM) appears as the leading mode of variability of the SH circulation [@fogt2020].
 SAM represents a relatively zonally symmetric pattern of alternating low pressures in polar latitude and a ring of high pressures in high latitudes with an embedded wave 3 pattern that is more prominent in the Pacific sector.
-Moreover, the 2nd and 3rd EOFs, usually known as PSA1 and PSA2 patterns, respectively, describe meridionally propagating wave trains that originates in the eastern equatorial Pacific and Australian-Indian Ocean sector, and travel towards the South Atlantic following a great-circle arch along the Antarctic coast.
-These patterns are usually derived by applying EOF to temporal anomalies, but @raphael2003 also applied EOF methods specifically to zonal anomalies.
+Moreover, the 2nd and 3rd EOFs, usually known as Pacific--South American Patterns (PSA) 1 and PSA2 patterns, respectively, describe meridionally propagating wave trains that originates in the eastern equatorial Pacific and Australian-Indian Ocean sector, and travel towards the South Atlantic following a great-circle arch along the Antarctic coast [@mo2001].
+These patterns influence precipitation anomalies in regions like South America from autumn to spring (REF?).
+Although these patterns are usually derived by applying EOF to temporal anomalies, [@raphael2003] also applied EOF methods specifically to zonal anomalies.
+More recently, [@irving2016] proposed a novel methodology for objectively identifying the PSA pattern using Fourier decomposition.
 
@@
 A single EOF can also represent a mixture of two or more physical modes.
-The inability of a single EOF to describe travelling waves is what forces the PSA pattern to be described by two EOFs. 
-@irving2016 characterise this pattern in a Fourier framework. Being a meridionally propagating mode, the PSA cannot be correctly characterised by Fourier decomposition at each latitude circle.
-As a result, they reprojected meridional wind fields so that the path of the PSA laid on the equator. On this new projection, they could identify the PSA using Fourier decomposition.
-
-Finally, another, hybrid methodology commonly used consists on identifying particular features of interest and creating indices using simple methods such as averages and differences.
-The SAM Index defined by [@marshall2003] following Gong and @gong1999, for example, describes SAM variability by the averaging the difference of sea level pressure between 40ºS and 65ºS.
-On the other hand, @raphael2004 used a similar methodology to describe the wave 3 activity in the SH circulation. 
-Averaging standardised geopotential-height anomalies at three locations of the ridges associated to the zonal wave 3 of the climatological mean. 
-By construction, the index is equivalent to projects the amplitude of the zonal wave 3 on the phase of the zonal wave 3 of the climatological mean field. 
-Similarly, @hobbs2010 propose that the zonally asymmetric circulation in the SH can be described by the strength and location of two anticyclones located in the sub-Antarctic western hemisphere. 
-These methods are grounded on other methods to identify the centres of action for the described phenomena and can be useful to characterise features that are not readily apparent with Fourier or EOF methods.
-Using information at specific regions or locations makes easy the computation of indices like those, but they are not very robust or unable to capture non-stationary patterns.
-
-The general goal of this study is to improve the description and understanding of planetary waves in the SH using an index that is able to capture the phase-varying nature of planetary waves and is flexible enough to describe planetary waves with non-constant amplitude in each latitude circle.
-We use complex Empirical Orthogonal Functions [@horel1984] as a robust extension of EOF analysis that is capable of describing rich structures with meridional and zonal propagation, and amplitude modulation such as the PSA.
-
-We apply this method to characterise the springtime zonally asymmetric circulation of the extratropical SH circulation. 
-In addition to take vertical distribution into account, we analyse both the troposphere and the stratosphere.
-
-In Section \@ref(methods) we describe the methods.
-In Section \@ref(spatial), we analyse the spatial patterns of each complex EOF.
-In Section \@ref(regressions) we study the spatial regressions with geopotential height and temperatuer.
-In Section \@ref(other-variables), we analyise the relationship between cEOF1 and Ozone, and between EOF2 and the Southern Annular Mode and the PSA.
-In Section \@ref(tropical) we look for tropical forcings that explain the variability of each cEOF.
-Finally, in Section \@ref(precipitation) we show the relationship between these modes of variabiltiy and precipitation in South America and Oceania.
+
+A third methodology commonly used consists on identifying particular features of interest and creating indices using simple methods such as averages and differences.
+Examples of this methodology are the SAM Index of [@gong1999], the SH wave 3 activity index defined by @raphael2004 and the SH zonally asymmetric circulation index from @hobbs2010.
+These derived methods are grounded on other methods such as Fourier decomposition or EOF to identify the centres of action for the described phenomena and can be useful to characterise features that are not readily apparent with these methods.
+It is commonly easy the computation of these kind of indices, but they could be unable to capture non-stationary patterns.
+
+An alternative methodology that has been proposed to study travelling and standing waves is complex Empirical Orthogonal Functions [@horel1984].
+This method extends EOF analysis to capture oscillations with varying amplitude and phase and has been applied to the time domain on water level to study tide-river interactions [@laurel-castillo2020] and temperature and precipitation anomalies [e.g.
+@krokhin2007; @gelbrecht2018].
+To our knowledge, complex EOF has not been applied in the spatial domain to capture the phase-varying nature of planetary waves.
+
+Some of the zonally asymmetric patterns of the SH circulation variability described previously, appear to have experienced secular changes.
+For instance, @raphael2003 found that the amplitude of the zonal wave 1 experienced a large increase and that the zonal wave 3 experienced changes in its annual cycle between 1958 and 1996.
+However, little is known yet about variability and trends of these patterns.
+
+The general goal of this study is to improve the description and understanding of springtime zonally asymmetric extratropical SH circulation using complex Empirical Orthogonal Functions, which allow to describe phase varying planetary waves with variable amplitudes along a latitude circle.
+In addition, it is proposed to expand the knowledge of the simultaneous behaviour of SH asymmetric circulation in the troposphere and the stratosphere.
+
+In Section 2 we describe the methods.
+In Section 3.1, we analyse the spatial patterns of each complex EOF.
+In Section 3.2 we study the spatial regressions with geopotential height and temperature anomalies.
+In Section 3.3, we analyse the relationship between cEOF1 and Ozone, and between cEOF2 with the SAM and the PSA.
+In Section 3.4 we study tropical forcings that explain the variability of each cEOF.
+Finally, in Section 3.5 we show the relationship between these modes of variability and precipitation anomalies in South America and Oceania.
+
+<!-- The large-scale extratropical circulation in the Southern Hemisphere (SH) is strongly zonally symmetric, but its zonal departures are highly relevant for regional impacts and could even be related to the occurrence of high-impact climate extremes (REFERENCE).
+-->
+<!-- Zonal asymmetries of extratropical circulation in the Southern Hemisphere strongly modulate weather systems and regional climate through promoting longitudinally different latitudinal transport of heat, humidity, and momentum [@trenberth1980].
+-->
+
+<!-- Hablar simplemente de cómo estos fenómenos varían en longitud.  -->
+
+<!-- @hill2009 ->  -->
+
+<!-- Comparan eventos ENSO del 97 y de 2002 -->
+<!-- La posición de las anomalías de circulación (PSA) son importantes para determinar el impacto del ENSO en sudamérica.
+Sugiere que la localización de la anomalía de SST cambia la localización del PSA.
+-->
+
+<!-- > The meridional winds from the idealised experiments based on simplified and localised SST warming showed a significant difference between experiments forced by eastern versus central-west Pacific SST anomalies, with a clear phase shift of the PSA apparent.
+This phase shift in the idealised experiments is possibly the main factor responsible for the different responses in South American atmospheric circulation between the 1997/1998 and 2002/2003 El Niño events.
+The phase shift over southeastern South America was also seen in the low-level 850 hPa winds in the EN97 and EN02 experiments, suggesting that the PSA phase shift is responsible for the shift in the moisture transport over southeastern Brazil between the two El Niño events.
+In summary, the location of the maximum SST anomaly appears to be an important factor in controlling the amount of moisture transported into the SACZ region during El Niño events.
+-->
+
+<!-- @hobbs2010 -> muestra la importancia de la longitud de ciclones -->
+
+<!-- ciasto2015 -> no investiga impactos -->
+
+<!-- sun2013 -->
+
+<!-- wilson2014 -> -->
+
+<!-- Observaciones y modelos enso canonico y modoki. diferentes patrones de circulacion.  -->
+
+<!-- wang2021 ->  -->
+
+<!-- > However, the spatial pattern of the PSA wave train excited by ENSO seems to be -->
+<!-- > non-stationary. Fogt and Bromwich (2006) indicated that the PSA mode moved -->
+<!-- > eastwards by 20° longitude in the 1990s owing to the eastward shift of the tropical -->
+<!-- > convective position. In addition, Wilson et al. (2014, 2016) and Ciasto et al. (2015) -->
+<!-- > found that the different types of ENSO can contribute to the wave shift. Specifically, -->
+<!-- > Rossby waves excited by central Pacific (CP)–type El Niño shift westwards owing to -->
+<!-- > the westward movement of the tropical diabatic heating. -->
+
+<!-- Diferencia en la localización de los centros en LN y EN -> diferencias en los impactos en la península antártica.  -->
+
+<!-- Lo más importante es la localización de la salida del jet, no la perturbación de sst.  -->
+
+<!-- fogt2006 ->  -->
+
+<!-- cambios en la posición de las teleccionexiones. Pero es todo con composiciones y comparando décadas. No mira impactos. -->
+
+<!-- lou2021 ->  -->
+
+<!-- > In combination, they produce the single propagating PSA mode. -->
+
+<!-- Complicado y no entiendo mucho, pero cosas que puedo ver es que PSA forza el océano subtropical que luego es precursora de ENSO. No sé si eso es súper  -->
 
@@
 We further split the SAM into its zonally symmetric and zonally asymmetric components (S-SAM and A-SAM indices respectively).
-These indices were obtained by projecting the zonally asymmetric and zonally symmetric part of the SAM spatial pattern onto monthly geopotential height fields, as proposed by @campitelli2021.
+These indices were obtained by projecting the zonally asymmetric and zonally symmetric part of the SAM spatial pattern onto monthly geopotential height fields, as proposed by @campitelli2022.
 Seasonal indices of the Pacific South American patterns (PSA1 and PSA2) were calculated -- in agreement with @mo2001 -- as the third and fourth leading EOF of seasonal mean anomalies for SH 500 hPa geopotential height.
 
-Linear trends were computed by Ordinary Least Squares (OLS) and the 95% confidence interval was computed assuming a t-distribution with the appropriate residual degrees of freedom [@wilks2011r].
+Linear trends were computed by Ordinary Least Squares (OLS) and the 95% confidence interval was computed assuming a t-distribution with the appropriate residual degrees of freedom [@wilks2011].
 
@@
 
-(ref:sam-eof-vertical-cap) Coefficient of determination between the real and imaginary part of each cEOF and the SAM, Asymmetric SAM (A-SAM) and Symmetric SAM (S-SAM) indices computed at each level according to @campitelli2021 for the 1979 -- 2019 period.
+(ref:sam-eof-vertical-cap) Coefficient of determination between the real and imaginary part of each cEOF and the SAM, Asymmetric SAM (A-SAM) and Symmetric SAM (S-SAM) indices computed at each level according to @@campitelli2022 for the 1979 -- 2019 period.
 Thick lines represent estimates with p-value \< 0.01 corrected for False Detection Rate [@benjamini1995].
@@
 In the stratosphere, the Real cEOF1 is correlated with both A-SAM and S-SAM, while the Imaginary cEOF1 is highly correlated only with the A-SAM.
-These correlations are consistent with the regression patterns of geopotential height in Figure \@ref(fig:eof1-regr-gh) and their comparison with those obtained for SAM, A-SAM and S-SAM by @campitelli2021.
+These correlations are consistent with the regression patterns of geopotential height in Figure \@ref(fig:eof1-regr-gh) and their comparison with those obtained for SAM, A-SAM and S-SAM by @@campitelli2022.
 
@@
 Over South America, the Imaginary cEOF has positive correlations with precipitation in South Eastern South America (SESA) and central Chile, and negative correlations in eastern Brazil.
-This again is a well-known springtime precipitation signature of ENSO [@cai2020a] and it is also similar to the precipitation anomalies associated with the A-SAM [@campitelli2021].
+This again is a well-known springtime precipitation signature of ENSO [@cai2020a] and it is also similar to the precipitation anomalies associated with the A-SAM [@@campitelli2022].
 

### 2022-03-15 analysis/paper/shceof.Rmd
Agrega justificación de Primavera

@@
 
-The general goal of this study is to improve the description and understanding of springtime zonally asymmetric extratropical SH circulation using complex Empirical Orthogonal Functions, which allow to describe phase varying planetary waves with variable amplitudes along a latitude circle.
+The general goal of this study is to improve the description and understanding of the zonally asymmetric extratropical SH circulation using complex Empirical Orthogonal Functions, which allow to describe phase varying planetary waves with variable amplitudes along a latitude circle.
 In addition, it is proposed to expand the knowledge of the simultaneous behaviour of SH asymmetric circulation in the troposphere and the stratosphere.
+
+We restrict this paper to the September-October-November (SON) trimester.
+We chose this season because it showed the strongest and more clearly interpretable results in a preliminary analysis of the four season.
+This is is consistent with previous research that shows that this season experiences a maximum in tropical teleconnections [@cazes-boezio2003a] and stratosphere-troposphere interactions [@lim2018].
 

### 2022-03-21 analysis/paper/shceof.Rmd
Agrega cita de impactos de precipitación del PSA
Mo 2001 será muy vieja?

@@
 The large-scale extratropical circulation in the Southern Hemisphere (SH) is strongly zonally symmetric, but its zonal departures are highly relevant for regional impacts [e.g. @hoskins2005].
-Zonal asymmetries of extratropical circulation in the Southern Hemisphere strongly modulate weather systems and regional climate through promoting longitudinally different latitudinal transport of heat, humidity, and momentum [@trenberth1980a; @raphael2007] and could even be related to the occurrence of high-impact climate extremes [@pezza2012].
+They strongly modulate weather systems and regional climate through promoting longitudinally different latitudinal transport of heat, humidity, and momentum [@trenberth1980a; @raphael2007] and could even be related to the occurrence of high-impact climate extremes [@pezza2012].
 
@@
 SAM represents a relatively zonally symmetric pattern of alternating low pressures in polar latitude and a ring of high pressures in high latitudes with an embedded wave 3 pattern that is more prominent in the Pacific sector.
-Moreover, the 2nd and 3rd EOFs, usually known as Pacific--South American Patterns (PSA) 1 and PSA2 patterns, respectively, describe meridionally propagating wave trains that originates in the eastern equatorial Pacific and Australian-Indian Ocean sector, and travel towards the South Atlantic following a great-circle arch along the Antarctic coast [@mo2001].
-These patterns influence precipitation anomalies in regions like South America from autumn to spring (REF?).
+The 2nd and 3rd EOFs, usually known as Pacific--South American Patterns (PSA) 1 and PSA2 patterns, respectively, describe meridionally propagating wave trains that originate in the eastern equatorial Pacific and Australian-Indian Ocean sector, and travel towards the South Atlantic following a great-circle arch along the Antarctic coast [@mo2001].
+These patterns influence precipitation anomalies in South America [@mo2001].
 Although these patterns are usually derived by applying EOF to temporal anomalies, [@raphael2003] also applied EOF methods specifically to zonal anomalies.

### 2022-03-21 analysis/paper/shceof.tex
Render

@@
 \usepackage{hyperref}
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
 
@@
 
+% tightlist command for lists without linebreak
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+
+% From pandoc table feature
+\usepackage{longtable,booktabs,array}
+\usepackage{calc} % for calculating minipage widths
+% Correct order of tables after \paragraph or \subparagraph
+\usepackage{etoolbox}
+\makeatletter
+\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
+\makeatother
+% Allow footnotes in longtable head/foot
+\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
+\makesavenoteenv{longtable}
+
 % Pandoc citation processing
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+\newlength{\csllabelwidth}
+\setlength{\csllabelwidth}{3em}
+\newlength{\cslentryspacingunit} % times entry-spacing
+\setlength{\cslentryspacingunit}{\parskip}
+% for Pandoc 2.8 to 2.10.1
+\newenvironment{cslreferences}%
+  {}%
+  {\par}
+% For Pandoc 2.11+
+\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
+ {% don't indent paragraphs
+  \setlength{\parindent}{0pt}
+  % turn on hanging indent if param 1 is 1
+  \ifodd #1
+  \let\oldpar\par
+  \def\par{\hangindent=\cslhangindent\oldpar}
+  \fi
+  % set entry spacing
+  \setlength{\parskip}{#2\cslentryspacingunit}
+ }%
+ {}
+\usepackage{calc}
+\newcommand{\CSLBlock}[1]{#1\hfill\break}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
+\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 
@@
 \usepackage{xcolor}
-
 \begin{document}
 
-\title{Revisiting the Austral Spring Extratropical Southern Hemisphere zonally assymetric circulation using complex Empirical Orthogonal Functions \thanks{The research was supported by UBACyT20020170100428BA and the CLIMAX Project funded by Belmont Forum/ANR-15-JCL/-0002-01.
+\title{Revisiting the Austral Spring Extratropical Southern Hemisphere zonally asymmetric circulation using complex Empirical Orthogonal Functions \thanks{The research was supported by UBACyT20020170100428BA and the CLIMAX Project funded by Belmont Forum/ANR-15-JCL/-0002-01.
 Elio Campitelli was supported by a PhD grant from CONICET, Argentina.} }
@@
 
-\begin{verbatim}
-## Spherical geometry (s2) switched off
-\end{verbatim}
-
 \hypertarget{introduction}{%
@@
 
-The large-scale extratropical circulation in the Southern Hemisphere is strongly zonally symmetric, but its zonal departures are highly relevant for surface impacts and could be related to high-impact climate extremes.
-Zonal asymmetries of extratropical circulation in the Southern Hemisphere strongly modulate weather systems and regional climate through latitudinal transport of heat, humidity, and momentum \citep{trenberth1980}.
+The large-scale extratropical circulation in the Southern Hemisphere (SH) is strongly zonally symmetric, but its zonal departures are highly relevant for regional impacts (e.g.
+Hoskins and Hodges 2005).
+They strongly modulate weather systems and regional climate through promoting longitudinally different latitudinal transport of heat, humidity, and momentum (K.
+E.
+Trenberth 1980; M.
+N.
+Raphael 2007) and could even be related to the occurrence of high-impact climate extremes (Pezza, Rashid, and Simmonds 2012).
 
-One of the main features of the zonally asymmetric circulation are the Pacific--South American Patterns \citep{mo2001}, which are associated with precipitation anomalies in South America, particularly in the September-October-November trimester (cita).
-Besides, \citet{raphael2007} suggests that variability in the planetary wave 3 projected onto its climatological location is associated with Antarctic sea ice concentration anomalies.
-\citet{irving2015} showed that planetary wave amplitude in general is associated sea ice concentration, as well as temperature and precipitation anomalies.
+Zonally asymmetric circulation is typically described by the amplitude and phase of zonal waves obtained by Fourier decomposition of geopotential heights or sea-level pressures at each latitude (e.g.
+van Loon and Jenne 1972; K.
+E.
+Trenberth 1980; Turner et al.
+2017).
+This approach suggests that zonal waves 1 and 3 explain almost 99\% of the total variance in the annual mean 500-hPa pattern at 50ºS (van Loon and Jenne 1972).
+K. F. Trenberth and Mo (1985) concluded that wave 3 plays a role in the development of blocking events.
+In addition, previous works have identified at extratropical and subpolar latitudes, wave-like patterns with dominant wavenumbers 3-4, also exerting distinctive regional impacts.
+M. N. Raphael (2007) shows that variability in the planetary wave 3 projected onto its climatological location is associated with anomalies in the Antarctic sea-ice concentration.
 
-Some of these patterns of variability associated to the zonally asymmetric circulation appear to have experienced secular changes, although to our knowledge this has not been extensively studied.
-For instance, \citet{raphael2003} found that the amplitude of the zonal wave 1 experienced a large increase and that the zonal wave 3 experienced changes in its annual cycle between 1958 and 1996.
+The Fourier decomposition relies on the assumption that the circulation can be meaningfully described in terms of zonal waves of constant amplitude along a latitude circle.
+However, this is not valid for meridionally propagating waves or zonal waves with localised amplitudes.
+Addressing this limitation, the Fourier technique can be generalized to integrate all planetary wave amplitude regardless of wave number, by computing the wave envelope (Irving and Simmonds 2015).
+The latter makes possible to represent planetary waves with different amplitude at different longitudes, but it removes all information about phase and wave number.
+With this approach, Irving and Simmonds (2015) showed that planetary wave amplitude in general is associated to Antarctic sea-ice concentration and temperature, as well as to precipitation anomalies in regions of significant topography in SH mid-latitudes and Antarctica.
 
-One typical way of describing the zonally asymmetric circulation is by the amplitude and phase of zonal waves obtained by Fourier decomposition of geopotential height at each latitude \citep[e.g.][]{vanloon1972, trenberth1980, turner2017}.
-This approach suggests that zonal waves 1 and 3 explained most of the variance of the tropospheric extratropical Southern Hemisphere circulation (\citet{vanloon1972}).
-However, this methodology relies on the assumption that the circulation can be meaningfully understood in terms of zonal waves of constant amplitude along a latitude circle.
-This is not valid for meridionally propagating waves or zonal waves with localised amplitudes.
-In the case of the wave 3, for example, \citet{trenberth1985} observed that it played a role in blocking events, but this was mostly due to increased amplitude of a longitudinally localised wave train instead of an hemispheric-scale zonal wave 3.
+Another extensively used approach to characterize the SH tropospheric circulation anomalies, is by computing Empirical Orthogonal Functions (EOF, also known as Principal Component Analysis).
+Within the EOF framework, the Southern Annular Mode (SAM) appears as the leading mode of variability of the SH circulation (Fogt and Marshall 2020).
+SAM represents a relatively zonally symmetric pattern of alternating low pressures in polar latitude and a ring of high pressures in high latitudes with an embedded wave 3 pattern that is more prominent in the Pacific sector.
+The 2nd and 3rd EOFs, usually known as Pacific--South American Patterns (PSA) 1 and PSA2 patterns, respectively, describe meridionally propagating wave trains that originate in the eastern equatorial Pacific and Australian-Indian Ocean sector, and travel towards the South Atlantic following a great-circle arch along the Antarctic coast (Mo and Paegle 2001).
+These patterns influence precipitation anomalies in South America (Mo and Paegle 2001).
+Although these patterns are usually derived by applying EOF to temporal anomalies, (M. Raphael 2003) also applied EOF methods specifically to zonal anomalies.
+More recently, (Irving and Simmonds 2016) proposed a novel methodology for objectively identifying the PSA pattern using Fourier decomposition.
 
-The fourier technique can be generalised to integrate all planetary wave intensity irregardless of wave number by computing the wave envelope as done by \citet{irving2015}.
-This makes it possible to represent planetary waves with different amplitude at different longitudes, but it removes all information about phase and wave number.
+Patterns resulting from EOF analyses are more flexible than Fourier decomposition derived modes in the sense that they can capture oscillation patterns that cannot be characterised by purely sinusoidal waves with constant amplitude.
+Nonetheless, they are restricted to standing oscillation modes and could not represent properly propagating or phase-varying features such as zonal waves.
+A single EOF can also represent a mixture of two or more physical modes.
 
-Another approach to characterising the Southern Hemisphere circulation is by using Empirical Orthogonal Functions (EOF, also known as Principal Component Analysis).
-Within the EOF framework, the Southern Annular Mode (SAM) appears as the leading mode of variability of the tropospheric Southern Hemisphere circulation \citep{fogt2020} followed by the two PSA \citep{mo2001}.
-The SAM represents a relatively zonally symmetric pattern of alternating low pressures in polar latitude and a ring of high pressures in high latitudes with an embedded wave 3 pattern that is more prominent in the Pacific sector.
-The PSA1 and PSA2 describe the two out-of-phase sides of a meridionally propagating wave train that originates in the eastern equatorial Pacific and travels towards the South Atlantic following a great-circle arch along the Antarctic Peninsula.
-These patterns are derived by applying EOF to temporal anomalies, but \citet{raphael2003} applied EOF methods specifically to zonal anomalies.
+A third methodology commonly used consists on identifying particular features of interest and creating indices using simple methods such as averages and differences.
+Examples of this methodology are the SAM Index of (Gong and Wang 1999), the SH wave 3 activity index defined by M.
+N.
+Raphael (2004) and the SH zonally asymmetric circulation index from Hobbs and Raphael (2010).
+These derived methods are grounded on other methods such as Fourier decomposition or EOF to identify the centres of action for the described phenomena and can be useful to characterise features that are not readily apparent with these methods.
+It is commonly easy the computation of these kind of indices, but they could be unable to capture non-stationary patterns.
 
-EOFs are more flexible than Fourier decomposition modes in the sense that they can capture oscillation patterns that cannot be characterised by purely sinusoidal waves with constant amplitude.
-Nonetheless, they are restricted to standing oscillation modes and could not represent properly propagating or phase-varying modes such as zonal waves.
-A single EOF can also represent a mixture of two or more physical modes.
-The inability of a single EOF to describe travelling waves is what forces the PSA pattern to be described by two EOFs.
-\citet{irving2016} characterises this pattern in a Fourier framework.
-Being a meridionally propagating mode, the PSA cannot be correctly characterised by Fourier decomposition at each latitude circle.
-As a result, they reprojected meridional wind fields so that the path of the PSA laid on the equator.
-On this new projection, they could identify the PSA using Fourier decomposition.
+An alternative methodology that has been proposed to study travelling and standing waves is complex Empirical Orthogonal Functions (Horel 1984).
+This method extends EOF analysis to capture oscillations with varying amplitude and phase and has been applied to the time domain on water level to study tide-river interactions (Laurel-Castillo and Valle-Levinson 2020) and temperature and precipitation anomalies (e.g.
+Krokhin and Luxemburg 2007; Gelbrecht, Boers, and Jürgen Kurths 2018).
+To our knowledge, complex EOF has not been applied in the spatial domain to capture the phase-varying nature of planetary waves.
 
-Finally, another, hybrid, methodology commonly used consist on identifying particular features of interest and creating indices using simple methods such as averages and differences.
-The Marshal SAM Index \citep{marshall2003}, for example, tries to characterise the SAM by the average difference of sea level pressure between 40ºS and 65ºS, following \citet{gong1999}.
-\citet{raphael2004} used a similar methodology to describe the wave 3 in the Southern Hemisphere.
-Instead of using Fourier to compute the amplitude and phase of the zonal wave, they averaged standardised geopotential height anomalies in three points representing the location of the ridges of the zonal wave 3 of the climatological mean.
-This is equivalent to an index of the amplitude of the zonal wave 3 that projects into the phase of the zonal wave 3 of the mean field.
-Similarly, \citet{hobbs2010} propose that the zonally asymmetric circulation in the Southern Hemisphere can be described by the strength and location of two anticyclones located in the sub-Antarctic western hemisphere.
-These methods are grounded on other methods to identify the centres of action for the described phenomena and can be useful to characterise features that are not readily apparent with Fourier or EOF methods.
-The use of data of only specific regions or locations makes them suitable for easy computation and for extending back into the data-poor past, but also makes them not very robust and unable to capture non-stationary patterns.
+Some of the zonally asymmetric patterns of the SH circulation variability described previously, appear to have experienced secular changes.
+For instance, M. Raphael (2003) found that the amplitude of the zonal wave 1 experienced a large increase and that the zonal wave 3 experienced changes in its annual cycle between 1958 and 1996.
+However, little is known yet about variability and trends of these patterns.
 
-Our objective is to improve our understanding of planetary waves in the Southern Hemisphere using an index that is able to capture the phase-varying nature of planetary waves and is flexible enough to describe planetary waves with non-constant amplitude in each latitude circle.
-We use complex Empirical Orthogonal Functions \citep{horel1984} as a robust extension of EOF analysis that is capable of describing rich structures with meridional and zonal propagation, and amplitude modulation such as the PSA.
+The general goal of this study is to improve the description and understanding of the zonally asymmetric extratropical SH circulation using complex Empirical Orthogonal Functions, which allow to describe phase varying planetary waves with variable amplitudes along a latitude circle.
+In addition, it is proposed to expand the knowledge of the simultaneous behaviour of SH asymmetric circulation in the troposphere and the stratosphere.
 
-We apply this method to characterise the Springtime zonally asymmetric circulation of the extratropical Southern Hemisphere.
-To take vertical distribution into account, we analyse both the troposphere and the stratosphere.
+We restrict this paper to the September-October-November (SON) trimester.
+We chose this season because it showed the strongest and more clearly interpretable results in a preliminary analysis of the four season.
+This is is consistent with previous research that shows that this season experiences a maximum in tropical teleconnections (Cazes-Boezio, Robertson, and Mechoso 2003) and stratosphere-troposphere interactions (Lim, Hendon, and Thompson 2018).
 
-In Section \ref{methods} we describe the methods.
-In Section \ref{spatial}, we analyse the spatial patterns of each complex EOF.
-In Section \ref{regressions} we study the spatial regressions with geopotential height and temperatuer.
-In Section \ref{other-variables}, we analyise the relationship between cEOF1 and Ozone, and between EOF2 and the Southern Annular Mode and the PSA.
-In Section \ref{tropical} we look for tropical forcings that explain the variability of each cEOF.
-Finally, in Section \ref{precipitation} we show the relationship between these modes of variabiltiy and precipitation in South America and Oceania.
+In Section 2 we describe the methods.
+In Section 3.1, we analyse the spatial patterns of each complex EOF.
+In Section 3.2 we study the spatial regressions with geopotential height and temperature anomalies.
+In Section 3.3, we analyse the relationship between cEOF1 and Ozone, and between cEOF2 with the SAM and the PSA.
+In Section 3.4 we study tropical forcings that explain the variability of each cEOF.
+Finally, in Section 3.5 we show the relationship between these modes of variability and precipitation anomalies in South America and Oceania.
 
@@
 
-We used monthly geopotential height, air temperature, ozone mixing ratio, and total ozone column (TOC) at 2.5\degree~longitude by 2.5\degree~latitude of horizontal resolution and 37 vertical isobaric levels from the European Centre for Medium-Range Weather Forecasts Reanalysis version 5 {[}ERA; \citet{era5}{]} for the period 1979-2019.
-Most of our analysis was restricted to the post-satellite era to avoid any confounding factors arising from the incorporation of satellite observations, but we also used the preliminary back extension of ERA5 from 1950 to 1978 \citep{era5be} to look at long-term trends.
-Streamfunction was derived from ERA5 vorticity at 200 hPa using the FORTRAN subroutine FISHPACK \citep{fishpack} and horizontal wave activity fluxes was computed following \citet{plumb1985}.
-We used Sea Surface Temperature (SST) monthly fields from Extended Reconstructed Sea Surface Temperature (ERSST) v5 \citep{huang2017} and precipitation monthly data from the CPC Merged Analysis of Precipitation \citep{cmap}, with a 2.5\degree~resolution in latitude and longitude.
-This rainfall gridded dataset is based on information from different sources such as rain gauge observations, satellite inferred estimations and the NCEP-NCAR reanalysis, and it is available since 1979 to present.
+We used monthly geopotential height, air temperature, ozone mixing ratio, and total ozone column (TOC) at 2.5\degree~longitude by 2.5\degree~latitude of horizontal resolution and 37 vertical isobaric levels from the European Centre for Medium-Range Weather Forecasts Reanalysis version 5 {[}ERA; Hersbach et al.
+(2019){]} for the period 1979-2019.
+Most of our analysis is restricted to the post-satellite era to avoid any confounding factors arising from the incorporation of satellite observations, but we also used the preliminary back extension of ERA5 from 1950 to 1978 (Bell et al.
+2020) to describe long-term trends.
+We derived streamfunction at 200 hPa from ERA5 vorticity using the FORTRAN subroutine FISHPACK (Adams, Swartztrauber, and Sweet 1999) and we computed horizontal wave activity fluxes following Plumb (1985).
+Sea Surface Temperature (SST) monthly fields are from Extended Reconstructed Sea Surface Temperature (ERSST) v5 (Huang et al.
+2017) and precipitation monthly data from the CPC Merged Analysis of Precipitation (P.
+Xie and Arkin 1997), with a 2.5\degree~resolution in latitude and longitude.
+The rainfall gridded dataset is based on information from different sources such as rain gauge observations, satellite inferred estimations and the NCEP-NCAR reanalysis, and it is available since 1979 to present.
 
@@
 The amplitude of the zonal waves was obtained through the Fourier transform of the spatial field at each latitude circle.
-We computed the amplitude and phase of the zonal wave 1 by averaging (area-weighted) each variable between 75°S and 45°S for each SON and extracting the wave-1 component of the Fourier spectrum.
-We chose this latitude band because it is wide enough to capture most of the relevant anomalies for southern hemisphere mid-latitudes.
+For the analysis of wave 1, we computed its amplitude and phase by averaging (area-weighted) the data for each variable and each SON between 75°S and 45°S, and then extracting the wave-1 component of the Fourier spectrum.
+We chose this latitude band because it is wide enough to capture most of the relevant anomalies of SH mid-latitudes.
 
-We computed the level-dependent SAM index as the leading EOF of year-round monthly geopotential height anomalies south of 20ºS at each level for the whole period according to \citet{baldwin2009}.
+We computed the level-dependent SAM index as the leading EOF of year-round monthly geopotential height anomalies south of 20ºS at each level for the whole period according (Baldwin and Thompson 2009).
 We further split the SAM into its zonally symmetric and zonally asymmetric components (S-SAM and A-SAM indices respectively).
-These indices were obtained by projecting the zonally asymmetric and zonally symmetric part of the SAM spatial pattern onto monthly geopotential height fields, as proposed by \citet{campitelli2021}.
-Seasonal indices of the Pacific South American patterns (PSA1 and PSA2) were calculated following \citet{mo2001} as the third and fourth leading EOF of seasonal mean anomalies for SH 500 hPa geopotential height combining all seasons.
-In our study, only the SON values were used.
+These indices were obtained by projecting the zonally asymmetric and zonally symmetric part of the SAM spatial pattern onto monthly geopotential height fields, as proposed by Campitelli, Díaz, and Vera (2022).
+Seasonal indices of the Pacific South American patterns (PSA1 and PSA2) were calculated -- in agreement with Mo and Paegle (2001) -- as the third and fourth leading EOF of seasonal mean anomalies for SH 500 hPa geopotential height.
 
-Linear trends were computed by Ordinary Least Squares (OLS) and the 95\% confidence interval was computed assuming a t-distribution with the appropriate residual degrees of freedom \citep{wilks2011r}.
+Linear trends were computed by Ordinary Least Squares (OLS) and the 95\% confidence interval was computed assuming a t-distribution with the appropriate residual degrees of freedom (D. Wilks 2011).
 
@@
 \includegraphics{../figures/eof-naive-1.pdf}
-\caption{\label{fig:eof-naive}Spatial patterns of the leading 4 EOFs of SON zonal anomalies of geopotential height at 50 hPa south of 20\degree S for the 1979 -- 2019 period (arbitrary units).}
+\caption{\label{fig:eof-naive}Spatial patterns of the four leading EOFs of SON zonal anomalies of geopotential height at 50 hPa south of 20\degree S for the 1979 -- 2019 period (arbitrary units).}
 \end{figure}
 
-In traditional EOF analysis zonal waves appear as pairs of EOFs, that could be degenerated representing similar patterns but shifted in phase \citep{horel1984}.
-For instance, Figure~\ref{fig:eof-naive} shows the leading 4 EOFs of SON geopotential height zonal anomalies at 50 hPa south of 20\degree S.
+In standard way of applying EOF analysis zonal waves may appear as pairs of EOFs, that could be degenerated representing similar patterns but shifted in phase (Horel 1984).
+Figure~\ref{fig:eof-naive} shows the four eading EOFs of SON geopotential height zonal anomalies at 50 hPa south of 20\degree S.
 It is clear that the first two EOFs represent the same zonal wave 1 pattern and the last two represent a same zonal wave pattern with shorter wavenumber and four centers of action shifted by 1/4 wavelength.
@@
 
-Complex Empirical Orthogonal Functions (cEOF) is a useful method to characterise zonal waves, considering phase-varying structures \citep{horel1984}.
+ON the other hand, complex Empirical Orthogonal Functions (cEOF) is a useful method to characterise zonal waves, associated with phase-varying structures (Horel 1984).
 This method involves computing EOF on the analytic representation of the original field.
-This representation is a complex field in which the real part is the original variable and the imaginary part is the original data shifted by 90\degree at each spectral frequency -- i.e.~its Hilbert transform.
+That representation is a complex field in which the real part is the original data and the imaginary part is the original data shifted by 90\degree at each spectral frequency -- i.e.~its Hilbert transform.
 The Hilbert transform is usually understood in terms of time-varying signal.
-However, in this work we apply the Hilbert transform to each latitude circle at each moment in time.
+However, in this work we apply the Hilbert transform at each latitude circle and at each considered time.
 Since each latitude circle is a periodic domain, this procedure does not suffer from edge effects.
@@
 
-\caption{\label{tab:corr-ceof-splitted}Coefficient of determination (\(R^2\)) of the absolute magnitude of complex EOFs between 200 hPa and 50 hPa computing EOF separately for each level.
-p-values lower than 0.01 in bold.}
+\caption{\label{tab:corr-ceof-splitted}Coefficient of determination (\(R^2\)) between the absolute magnitude of complex EOFs computed separately at 200 hPa and 50 hPa (p-values lower than 0.01 in bold).}
 \centering
@@
 
-The cEOF methodology is applied to SON geopotential height zonal anomalies south of 20\degree S at 50 and 200 hPa, representing a stratsopheric and upper troposhperic level respectively.
-Table \ref{tab:corr-ceof-splitted} shows the coefficient of determination between time series of the amplitude of each complex EOF (cEOF) across levels.
-There's a high degree of correlation between the magnitude of the respective cEOF1 and cEOF2 at each level.
+The cEOF methodology is applied to SON geopotential height zonal anomalies south of 20\degree S separately at 50 and 200 hPa.
+Table \ref{tab:corr-ceof-splitted} shows the coefficient of determination between time series of the amplitude of each cEOF across levels.
+There is a high degree of correlation between the magnitude of the respective cEOF1 and cEOF2 at each level.
 The spatial patterns of the 50 hPa and 200 hPa cEOFs are also similar (not shown).
@@
 For real principal components, this constant can be either 0 or \(\pi\), corresponding to a change in sign.
-For complex principal components, it can be any number between 0 and \(2\pi\) \citep{horel1984}.
+For complex principal components, it can be any number between 0 and \(2\pi\) (Horel 1984).
 Since any choice is arbitrary and equally valid, we chose the phase of each cEOF so that the real and imaginary parts are aligned with meaningful phases in our analysis.
@@
 This also nearly minimises the correlation with the imaginary part.
-For the second complex principal component, the phase was chosen so that the coefficient of determination between the Oceanic Niño Index \citep{bamston1997} and the real part was minimized, which also nearly maximizes the correlation with the imaginary part.
+For the second complex principal component, the phase was chosen so that the coefficient of determination between the Oceanic Niño Index (Bamston, Chelliah, and Goldenberg 1997) and the real part was minimized, which also nearly maximizes the correlation with the imaginary part.
 
@@
 
-We performed linear regressions to quantify the association between the cEOFs and other variables.
-For each cEOF, we create regression maps by fitting a multiple linear model involving both the real and the imaginary part.
-To obtain the linear coefficients of a variable \(X\) (geopotential, temperature, precipitation, etc\ldots) with the Imaginary and Real parts of each cEOF we fit the equation
+We performed linear regressions to quantify the association between the cEOFs and other variables (e.g.~geopotential height, temperature, precipitation, and others).
+For each cEOF, we computed regression maps by fitting a multiple linear model involving both the real and the imaginary part.
+To obtain the linear coefficients of a variable \(X\) with the imaginary and real parts of each cEOF we fit the equation
 
@@
 
-where \(\lambda\) and \(\phi\) are the longitude and latitude, \(t\) is the time, \(\alpha\) and \(\beta\) are the linear regression coefficients for Imaginary and Real parts respectively, \(X_0\) and \(\epsilon\) are the constant and error terms respectively.
+where \(\lambda\) and \(\phi\) are the longitude and latitude, \(t\) is the time, \(\alpha\) and \(\beta\) are the linear regression coefficients for imaginary and real parts respectively, \(X_0\) and \(\epsilon\) are the constant and error terms respectively.
 
-We evaluated statistical significance using a two-sided t-test and, in the case of regression maps, p-values were adjusted by controlling for the False Discovery Rate \citep{benjamini1995, wilks2016} to avoid misleading results from the high number of regressions \citep{walker1914, katz1991}.
+We evaluated statistical significance using a two-sided t-test and, in the case of regression maps, p-values were adjusted by controlling for the False Discovery Rate (Benjamini and Hochberg 1995; D.
+S.
+Wilks 2016) to avoid misleading results from the high number of regressions (Walker 1914; Katz and Brown 1991).
 
@@
 
-We performed all analysis in this paper using the R programming language \citep{rcoreteam2020}, using data.table \citep{dowle2020} and metR \citep{campitelli2020} packages.
-All graphics are made using ggplot2 \citep{wickham2009}.
-We downloaded data from reanalysis using the ecmwfr package \citep{hufkens2020} and indices of the ENSO with the rsoi package \citep{albers2020}.
-The paper was rendered using knitr and rmarkdown \citep{xie2015, allaire2020}.
+We performed all analysis in this paper using the R programming language (R Core Team 2020), using data.table (Dowle and Srinivasan 2020) and metR (Campitelli 2020) packages.
+All graphics are made using ggplot2 (Wickham 2009).
+We downloaded data from reanalysis using the ecmwfr package (Hufkens 2020) and indices of the ENSO with the rsoi package (Albers and Campitelli 2020).
+The paper was rendered using knitr and rmarkdown (Y.
+Xie 2015; Allaire et al.
+2020).
 
@@
 \includegraphics{../figures/extended-series-1} \caption{Time series extended using ERA5 back extended preliminary edition (period 1950 -- 1978) and ERA5 (period 1979 -- 2019).
-Dark straight line is the OLS trend.
-Black horizontal and vertical line mark the mean value and range of each time series respectively.}\label{fig:extended-series}
+Dark straight line is the linear trend.
+Black horizontal and vertical line mark the mean value and range of each time series, respectively.}\label{fig:extended-series}
 \end{figure*}
@@
 The cEOF2 (Fig.~\ref{fig:eof-naive} column~b) shows also a zonal wave-like structure with maximum amplitude at high latitudes, but with shorter spatial scales.
-In particular, its dominant structure at both levels is a wave 3 but with larger amplitude in the pacific sector.
+In particular, the dominant structure at both levels is a wave 3 but with larger amplitude in the pacific sector.
 This modulated amplitude is especially apparent at 200 hPa.
@@
 However, cEOF2's mean is almost zero, which indicates that only cEOF1 includes variability that significantly projects onto the mean zonally anomalous field.
-There are no significant simultaneous correlation between the time series.
+There is no significant simultaneous correlation between the time series.
 Both cEOFs show year-to-year variability but show no evidence of decadal variability.
 
-There is only a significant upward trend in the real component of cEOF1 (Fig.~\ref{fig:extended-series}a.1, p-value~= 0.0037) and no significant trend in any of the complex components of cEOF2.
+A significant upward trend in the real component of cEOF1 is evident (Fig.~\ref{fig:extended-series}a.1, p-value~= 0.0037) while there is no significant trend in any of the complex components of cEOF2.
 The positive trend in the Real cEOF1 translates into a positive trend in cEOF1 magnitude, but not in the phase (not shown).
 This long-term change indicates an increase in the magnitude of the high latitude zonal wave 1.
-A similar observation was made by \citet{raphael2003}, who detected a step after around 1975 in the leading EOF of August-September-October 500 hPa zonal geopotential height anomalies, which is similar to the cEOF1.
+In agreement, M.
+Raphael (2003) detected a step change after around 1975 in the temporal evolution of leading EOF computed from August-September-October 500 hPa zonal geopotential height anomalies, which is similar to the cEOF1.
 
@@
 
-The spatial patterns shown in Figure~\ref{fig:ceofs-1} are derived by removing the zonally symmetric circulation, so they might not include all the variability that is actually associated with the cEOF time series.
-To understand the full geopotential height field changes associated with each cEOF, we computed regression patterns of the full geopotential fields onto each cEOF.
+In the previous section, cEOFs were applied to zonal anomalies, that is anomalies derived by removing the zonally mean values in order to isolate the main characteristics of the main zonal waves characterizing the circulation in the SH.
+In this section, regression fields were computed using the full fields of the variables in order to describe the influence of the cEOFs on the temporal anomalies.
 
@@
 Their phases coincide with the ones shown in Figure~\ref{fig:ceofs-1}a.1, with the positive centre of the Real cEOF1 located towards the dateline, and the one of the Imaginary cEOF1 located over East Antarctica.
-However, the Real cEOF1 pattern is substantially altered by the zonally symmetric circulation.
-Instead of a clear wave 1 pattern, the regression pattern can be better describes as a monopole with its centre displaced from the South Pole.
+Instead of a wave-1 pattern, regressed anomalies describe a monopole with its centre displaced from the South Pole. The latter indicates an influence of cEOF1 in modulating the polar vortex.
 
-The regression patterns for cEOF1 at 200 hPa (Figure~\ref{fig:eof1-regr-gh} row b) are similarly influenced by the zonally symmetric circulation.
+The regression maps for cEOF1 at 200 hPa (Figure~\ref{fig:eof1-regr-gh} row b) are similarly influenced by the zonally symmetric circulation.
 It is only possible to partially distinguish the wave 1 pattern in relation with the Real cEOF1 (Figure~\ref{fig:eof1-regr-gh}b.1).
@@
 Zonal wave 3 dominates all fields, but only in the western hemisphere, over the Pacific and Atlantic Oceans.
-cEOF2 then represents an equivalent barotropic wave train that is very similar to the the Pacific South American Patterns \citep{mo2001}.
-Comparing the location of the positive anomaly near 90ºW in column b of Figure~\ref{fig:eof2-regr-gh} with Figures 1.a and b from \citet{mo2001}, the Real cEOF2 can roughly be identified with PSA2, while the Imaginary cEOF2 resembles PSA1.
+cEOF2 then represents an equivalent barotropic wave train that is very similar to the the Pacific South American Patterns (Mo and Paegle 2001).
+Comparing the location of the positive anomaly near 90ºW in column b of Figure~\ref{fig:eof2-regr-gh} with Figures 1.a and b from Mo and Paegle (2001), the Real cEOF2 can roughly be identified with PSA2, while the Imaginary cEOF2 resembles PSA1.
 
@@
 \includegraphics{../figures/sam-eof-vertical-1.pdf}
-\caption{\label{fig:sam-eof-vertical}Coefficient of determination between the real and imaginary part of each cEOF and the SAM, Asymmetric SAM (A-SAM) and Symmetric SAM (S-SAM) indices computed at each level according to \citet{campitelli2021} for the 1979 -- 2019 period.
-Thick lines represent estimates with p-value \textless{} 0.01 corrected for False Detection Rate \citep{benjamini1995}.}
+\caption{\label{fig:sam-eof-vertical}Coefficient of determination between the real and imaginary part of each cEOF and the SAM, Asymmetric SAM (A-SAM) and Symmetric SAM (S-SAM) indices computed at each level according to @Campitelli, Díaz, and Vera (2022) for the 1979 -- 2019 period.
+Thick lines represent estimates with p-value \textless{} 0.01 corrected for False Detection Rate (Benjamini and Hochberg 1995).}
 \end{figure}
@@
 In the stratosphere, the Real cEOF1 is correlated with both A-SAM and S-SAM, while the Imaginary cEOF1 is highly correlated only with the A-SAM.
-These correlations are consistent with the regression patterns of geopotential height in Figure~\ref{fig:eof1-regr-gh} and their comparison with those obtained for SAM, A-SAM and S-SAM by \citet{campitelli2021}.
+These correlations are consistent with the regression patterns of geopotential height in Figure~\ref{fig:eof1-regr-gh} and their comparison with those obtained for SAM, A-SAM and S-SAM by @Campitelli, Díaz, and Vera (2022).
 
@@
 
-\caption{\label{tab:psa-eof2}Correlations between the Real and Imaginary parts of cEOF2 and the PSA1 and PSA2 modes computed as the second and third EOFs of seasonal geopotential height anomalies \citep[following][]{mo2001} for the 1979 -- 2019 period.
+\caption{\label{tab:psa-eof2}Correlations between the Real and Imaginary parts of cEOF2 and the PSA1 and PSA2 modes computed as the second and third EOFs of seasonal geopotential height anomalies (following Mo and Paegle 2001) for the 1979 -- 2019 period.
 95\% confidence intervals in parenthesis.
@@
 
-The result obtained here is similar to the one obtained by \citet{irving2016}.
+The result obtained here is similar to the one obtained by Irving and Simmonds (2016).
 They show that the phase distribution of the PSA-like variability obtained by them is bimodal and isolate the PSA pattern from the rest of the PSA-like variability by selecting events that are near the peaks of the distribution (compare our Figure~\ref{fig:phase-histogram} with their Figure 6).
@@
 In the photochemically-dominated upper stratosphere, cold temperatures inhibit the destruction of ozone.
-On the other hand, in the advectively-dominated lower stratosphere, ozone anomalies are 90º out of phase with horizontal and vertical transport, which are in adittion 90º out of phase with temperature anomalies \citep{hartmann1979, wirth1993, smith1995}.
+On the other hand, in the advectively-dominated lower stratosphere, ozone anomalies are 90º out of phase with horizontal and vertical transport, which are in adittion 90º out of phase with temperature anomalies (Hartmann and Garcia 1979; Wirth 1993; Smith 1995).
 
@@
 ~\ref{fig:o3-regr}) show zonal wave 1 pattens in TOC associated with both phases of cEOF1.
-Climatologically, the springtime Ozone minimum is located off the South Pole and towards the Weddell Sea \citep{grytsai2011}.
+Climatologically, the springtime Ozone minimum is located off the South Pole and towards the Weddell Sea (Grytsai 2011).
 Thus, the Real cEOF1 regression pattern (Figure~\ref{fig:o3-regr}a) coincides with the climatological position of the ozone hole while the one for the Imaginary cEOF1 is shifted by 90º.
@@
 The Imaginary cEOF2 is associated with strong positive SST anomalies on the Central Pacific and negative anomalies over an area across the North of Australia and New Zealand, the South Pacific Convergence Zone (SPCZ) (Figure~\ref{fig:sst-psi-2}.b1).
-This pattern is almost canonically positive ENSO \citep{bamston1997} and indeed, the correlation between the Imaginary cEOF2 and the Oceanic Niño Index \citep{bamston1997} is significant and very high 0.76 (CI: 0.6 -- 0.87).
+This pattern is almost canonically positive ENSO (Bamston, Chelliah, and Goldenberg 1997) and indeed, the correlation between the Imaginary cEOF2 and the Oceanic Niño Index (Bamston, Chelliah, and Goldenberg 1997) is significant and very high 0.76 (CI: 0.6 -- 0.87).
 On top of this ENSO-like pattern, there are positive anomalies in the western Indian Ocean and negative values in the eastern Indian Ocean, resembling a positive Indian Ocean Dipole.
@@
 Fittingly, the Imaginary cEOF2 is also associated with strong wave-like streamfunction anomalies emanating from the tropics (Figure~\ref{fig:sst-psi-2}.b2), both from the Central Pacific sector and the Incean Ocean.
-These are consistent with the effect of ENSO and the Indian Ocean Dipole on the extratropics: the SST anomalies initiate anomalous tropical convection that excites Rossby waves that propagate meridionally towards higher latitudes \citep{mo2000, cai2011, nuncio2015}.
+These are consistent with the effect of ENSO and the Indian Ocean Dipole on the extratropics: the SST anomalies initiate anomalous tropical convection that excites Rossby waves that propagate meridionally towards higher latitudes (Mo 2000; Cai et al.
+2011; Nuncio and Yuan 2015).
 
@@
 
-On the extra-tropics, the Imaginary cEOFs is correlated with drier conditions over eastern Australia; a similar signal as the one associated with ENSO \citet{cai2011}.
+On the extra-tropics, the Imaginary cEOFs is correlated with drier conditions over eastern Australia; a similar signal as the one associated with ENSO Cai et al. (2011).
 However, the Imaginary cEOF2 is not the direction most correlated with precipitation in that area.
@@
 Over South America, the Imaginary cEOF has positive correlations with precipitation in South Eastern South America (SESA) and central Chile, and negative correlations in eastern Brazil.
-This again is a well-known springtime precipitation signature of ENSO \citep{cai2020a} and it is also similar to the precipitation anomalies associated with the A-SAM \citep{campitelli2021}.
+This again is a well-known springtime precipitation signature of ENSO (Cai et al.
+2020) and it is also similar to the precipitation anomalies associated with the A-SAM (@ Campitelli, Díaz, and Vera 2022).
 
@@
 The first complex EOF represents the variability of the zonal wave 1 in the stratosphere, but represents a more zonally symmetric monopole in the troposphere.
-There is a statistically positive trend in the magnitude of this cEOF, which is consistent with previous studies that showed secular changes in springtime wave-1-like patterns \citep[e.g.][]{raphael2003}.
+There is a statistically positive trend in the magnitude of this cEOF, which is consistent with previous studies that showed secular changes in springtime wave-1-like patterns (e.g. M. Raphael 2003).
 This mode is closely related to stratospheric variability such as anomalies in Total Ozone Column.
@@
 The second complex EOF represents a wave-3 pattern with maximum magnitude in the Pacific sector.
-Essentially, it is an alternate representation of the PSA1 and PSA2 patterns \citep{mo2001}.
+Essentially, it is an alternate representation of the PSA1 and PSA2 patterns (Mo and Paegle 2001).
 We show that the Imaginary cEOF2 can be roughly identified with the PSA1 and the Real cEOF2 with the PSA2.
 
-There is some disagreement in the literature of whether the phase of the PSA pattern is affected by the location of tropical SST anomalies \citep{irving2016}.
+There is some disagreement in the literature of whether the phase of the PSA pattern is affected by the location of tropical SST anomalies (Irving and Simmonds 2016).
 We show not only that the cEOF2 tends to be in the imaginary phase (\textasciitilde PSA1) when the ENSO region is warm, but that central Pacific SST anomalies tend to move the cEOF towards the negative real phase and eastern Pacific SST anomalies tend to move it towards positive real phase.
 When ENSO is neutral, the cEOF2 is still active, but with no particular preferred phase.
-This mirrors the results of \citet{cai2002}, who showed that the CSIRO Model can create PSA-like variability even in the absence of ENSO forcing (with a climatological run), but the variability of one of the PSA modes was enhanced when adding the ENSO signal.
+This mirrors the results of Cai and Watterson (2002), who showed that the CSIRO Model can create PSA-like variability even in the absence of ENSO forcing (with a climatological run), but the variability of one of the PSA modes was enhanced when adding the ENSO signal.
 The magnitude of the cEOF2 appears not to be related to the magnitude of ENSO, except for the three strongest ENSO events in the period which coincide with the three strongest cEOF2 years.
 
-The sensitivity of the phase of the cEOF2 to the location of the tropical SST anomalies was also seen by \citet{ciasto2015}, who detected similar Rossby wave patterns associated with central Pacific and eastern Pacific SST anomalies but with a change in phase.
+The sensitivity of the phase of the cEOF2 to the location of the tropical SST anomalies was also seen by Ciasto, Simpkins, and England (2015), who detected similar Rossby wave patterns associated with central Pacific and eastern Pacific SST anomalies but with a change in phase.
 
@@
 
-Some aspects of our work are in direct contrast with \citet{irving2016} analysis of the PSA.
+Some aspects of our work are in direct contrast with Irving and Simmonds (2016) analysis of the PSA.
 They didn't find neither a strong association between PSA-like variability and ENSO, nor a relationship between it and Australian precipitation.
 Part of the disagreement could be due to the fact that we focus data only for SON, which is when the teleconnection pattern is strongest, instead of the whole year.
-\citet{irving2016} use of ERA5 forectasted precipitation instead of the more observationally constrained CMAP dataset might also be an issue.
-But in light of the previous discussion on the importance of the phase of the cEOF2 in its relationship with ENSO and with precipitation impacts, a big difference could be fact that \citet{irving2016}'s methodology is insensitive to phase changes of the PSA-like variability.
+Irving and Simmonds (2016) use of ERA5 forectasted precipitation instead of the more observationally constrained CMAP dataset might also be an issue.
+But in light of the previous discussion on the importance of the phase of the cEOF2 in its relationship with ENSO and with precipitation impacts, a big difference could be fact that Irving and Simmonds (2016)'s methodology is insensitive to phase changes of the PSA-like variability.
 
@@
 
+\hypertarget{refs}{}
+\begin{CSLReferences}{1}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-fishpack}{}}%
+Adams, John C., Paul N.
+Swartztrauber, and Roland Sweet.
+1999.
+{``{FISHPACK}, a Package of {Fortran} Subprograms for the Solution of Separable Elliptic Partial Differential Equations.''} 1999.
+\url{https://www2.cisl.ucar.edu/resources/legacy/fishpack}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-albers2020}{}}%
+Albers, Sam, and Elio Campitelli. 2020. {``Rsoi: {Import Various Northern} and {Southern Hemisphere Climate Indices}.''}
+
+\leavevmode\vadjust pre{\hypertarget{ref-allaire2020}{}}%
+Allaire, J. J., Yihui Xie {[}aut, cre, Jonathan McPherson, Javier Luraschi, Kevin Ushey, Aron Atkins, et al. 2020. {``Rmarkdown: {Dynamic Documents} for {R}.''}
+
+\leavevmode\vadjust pre{\hypertarget{ref-baldwin2009}{}}%
+Baldwin, Mark P., and David W.
+J.
+Thompson.
+2009.
+{``A Critical Comparison of Stratosphere--Troposphere Coupling Indices.''} \emph{Quarterly Journal of the Royal Meteorological Society} 135 (644): 1661--72.
+\url{https://doi.org/10.1002/qj.479}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-bamston1997}{}}%
+Bamston, Anthony G., Muthuvel Chelliah, and Stanley B.
+Goldenberg.
+1997.
+{``Documentation of a Highly {ENSO}‐related Sst Region in the Equatorial Pacific: {Research} Note.''} \emph{Atmosphere-Ocean} 35 (3): 367--83.
+\url{https://doi.org/10.1080/07055900.1997.9649597}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-era5be}{}}%
+Bell, B., H.
+Hersbach, P.
+Berrisford, P.
+Dahlgren, A.
+Horányi, J.
+Muñoz Sabater, J.
+Nicolas, et al.
+2020.
+{``{Era5} Monthly Averaged Data on Pressure Levels from 1950 to 1978 (Preliminary Version).''} \emph{{Copernicus Climate Change Service} ({C3s}) {Climate Data Store} ({CDS})} (Accessed on {\(<\)}26-08-2021{\(>\)}), https://cds.climate.copernicus.eu/cdsapp\#!/dataset/reanalysis-era5-pressure-levels-monthly-means-preliminary-back-extension?tab=overview.
+
+\leavevmode\vadjust pre{\hypertarget{ref-benjamini1995}{}}%
+Benjamini, Yoav, and Yosef Hochberg.
+1995.
+{``Controlling the {False Discovery Rate}: {A Practical} and {Powerful Approach} to {Multiple Testing}.''} \emph{Journal of the Royal Statistical Society: Series B (Methodological)} 57 (1): 289--300.
+\url{https://doi.org/10.1111/j.2517-6161.1995.tb02031.x}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-cai2020a}{}}%
+Cai, Wenju, Michael J.
+McPhaden, Alice M.
+Grimm, Regina R.
+Rodrigues, Andréa S.
+Taschetto, René D.
+Garreaud, Boris Dewitte, et al.
+2020.
+{``Climate Impacts of the {El Niño}--{Southern Oscillation} on {South America}.''} \emph{Nature Reviews Earth \& Environment} 1 (4, 4): 215--31.
+\url{https://doi.org/10.1038/s43017-020-0040-3}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-cai2011}{}}%
+Cai, Wenju, Peter van Rensch, Tim Cowan, and Harry H.
+Hendon.
+2011.
+{``Teleconnection {Pathways} of {ENSO} and the {IOD} and the {Mechanisms} for {Impacts} on {Australian Rainfall}.''} \emph{Journal of Climate} 24 (15): 3910--23.
+\url{https://doi.org/10.1175/2011JCLI4129.1}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-cai2002}{}}%
+Cai, Wenju, and Ian G.
+Watterson.
+2002.
+{``Modes of {Interannual Variability} of the {Southern Hemisphere Circulation Simulated} by the {CSIRO Climate Model}.''} \emph{Journal of Climate} 15 (10): 1159--74.
+\url{https://doi.org/10.1175/1520-0442(2002)015\%3C1159:MOIVOT\%3E2.0.CO;2}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-campitelli2020}{}}%
+Campitelli, Elio. 2020. {``{metR}: {Tools} for {Easier Analysis} of {Meteorological Fields}.''}
+
+\leavevmode\vadjust pre{\hypertarget{ref-campitelli2022}{}}%
+Campitelli, Elio, Leandro B.
+Díaz, and Carolina Vera.
+2022.
+{``Assessment of Zonally Symmetric and Asymmetric Components of the {Southern Annular Mode} Using a Novel Approach.''} \emph{Climate Dynamics} 58 (1): 161--78.
+\url{https://doi.org/10.1007/s00382-021-05896-5}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-cazes-boezio2003a}{}}%
+Cazes-Boezio, Gabriel, Andrew W.
+Robertson, and Carlos R.
+Mechoso.
+2003.
+{``Seasonal {Dependence} of {ENSO Teleconnections} over {South America} and {Relationships} with {Precipitation} in {Uruguay}.''} \emph{Journal of Climate} 16 (8): 1159--76.
+\url{https://doi.org/10.1175/1520-0442(2003)16\%3C1159:SDOETO\%3E2.0.CO;2}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-ciasto2015}{}}%
+Ciasto, Laura M., Graham R.
+Simpkins, and Matthew H.
+England.
+2015.
+{``Teleconnections Between {Tropical Pacific SST Anomalies} and {Extratropical Southern Hemisphere Climate}.''} \emph{Journal of Climate} 28 (1): 56--65.
+\url{https://doi.org/10.1175/JCLI-D-14-00438.1}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-dowle2020}{}}%
+Dowle, Matt, and Arun Srinivasan. 2020. {``Data.table: {Extension} of 'Data.frame'.''}
+
+\leavevmode\vadjust pre{\hypertarget{ref-fogt2020}{}}%
+Fogt, Ryan L., and Gareth J.
+Marshall.
+2020.
+{``The {Southern Annular Mode}: {Variability}, Trends, and Climate Impacts Across the {Southern Hemisphere}.''} \emph{WIREs Climate Change} 11 (4): e652.
+\url{https://doi.org/10.1002/wcc.652}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-gelbrecht2018}{}}%
+Gelbrecht, Maximilian, Niklas Boers, and Jürgen Kurths.
+2018.
+{``Phase Coherence Between Precipitation in {South America} and {Rossby} Waves.''} \emph{Science Advances} 4 (12): eaau3191.
+\url{https://doi.org/10.1126/sciadv.aau3191}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-gong1999}{}}%
+Gong, Daoyi, and Shaowu Wang. 1999. {``Definition of {Antarctic Oscillation} Index.''} \emph{Geophysical Research Letters} 26 (4): 459--62. \url{https://doi.org/10.1029/1999GL900003}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-grytsai2011}{}}%
+Grytsai, A.
+2011.
+{``Planetary Wave Peculiarities in {Antarctic} Ozone Distribution During 1979--2008.''} \emph{International Journal of Remote Sensing} 32 (11): 3139--51.
+\url{https://doi.org/10.1080/01431161.2010.541518}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-hartmann1979}{}}%
+Hartmann, Dennis L., and Rolando R.
+Garcia.
+1979.
+{``A {Mechanistic Model} of {Ozone Transport} by {Planetary Waves} in the {Stratosphere}.''} \emph{Journal of the Atmospheric Sciences} 36 (2): 350--64.
+\url{https://doi.org/10.1175/1520-0469(1979)036\%3C0350:AMMOOT\%3E2.0.CO;2}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-era5}{}}%
+Hersbach, H., B.
+Bell, P.
+Berrisford, G.
+Biavati, A.
+Horányi, J.
+Muñoz Sabater, J.
+Nicolas, et al.
+2019.
+{``{Era5} Monthly Averaged Data on Pressure Levels from 1979 to Present.''} \emph{{Copernicus Climate Change Service} ({C3s}) {Climate Data Store} ({CDS})} (Accessed on {\(<\)}07-09-2021{\(>\)}).
+\url{https://doi.org/10.24381/cds.6860a573}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-hobbs2010}{}}%
+Hobbs, William R., and Marilyn N.
+Raphael.
+2010.
+{``Characterizing the Zonally Asymmetric Component of the {SH} Circulation.''} \emph{Climate Dynamics} 35 (5): 859--73.
+\url{https://doi.org/10.1007/s00382-009-0663-z}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-horel1984}{}}%
+Horel, J.
+D.
+1984.
+{``Complex {Principal Component Analysis}: {Theory} and {Examples}.''} \emph{Journal of Applied Meteorology and Climatology} 23 (12): 1660--73.
+\url{https://doi.org/10.1175/1520-0450(1984)023\%3C1660:CPCATA\%3E2.0.CO;2}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-hoskins2005}{}}%
+Hoskins, B. J., and K. I. Hodges. 2005. {``A {New Perspective} on {Southern Hemisphere Storm Tracks}.''} \emph{Journal of Climate} 18 (20): 4108--29. \url{https://doi.org/10.1175/JCLI3570.1}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-huang2017}{}}%
+Huang, Boyin, Peter W.
+Thorne, Viva F.
+Banzon, Tim Boyer, Gennady Chepurin, Jay H.
+Lawrimore, Matthew J.
+Menne, Thomas M.
+Smith, Russell S.
+Vose, and Huai-Min Zhang.
+2017.
+{``Extended {Reconstructed Sea Surface Temperature}, {Version} 5 ({ERSSTv5}): {Upgrades}, {Validations}, and {Intercomparisons}.''} \emph{Journal of Climate} 30 (20): 8179--8205.
+\url{https://doi.org/10.1175/JCLI-D-16-0836.1}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-hufkens2020}{}}%
+Hufkens, Koen. 2020. {``Ecmwfr: {Programmatic} Interface to the Two {European Centre} for {Medium-Range Weather Forecasts API} Services.''}
+
+\leavevmode\vadjust pre{\hypertarget{ref-irving2015}{}}%
+Irving, Damien, and Ian Simmonds.
+2015.
+{``A {Novel Approach} to {Diagnosing Southern Hemisphere Planetary Wave Activity} and {Its Influence} on {Regional Climate Variability}.''} \emph{Journal of Climate} 28 (23): 9041--57.
+\url{https://doi.org/10.1175/JCLI-D-15-0287.1}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-irving2016}{}}%
+---------.
+2016.
+{``A {New Method} for {Identifying} the {Pacific}--{South American Pattern} and {Its Influence} on {Regional Climate Variability}.''} \emph{Journal of Climate} 29 (17): 6109--25.
+\url{https://doi.org/10.1175/JCLI-D-15-0843.1}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-katz1991}{}}%
+Katz, Richard W., and Barbara G.
+Brown.
+1991.
+{``The Problem of Multiplicity in Research on Teleconnections.''} \emph{International Journal of Climatology} 11 (5): 505--13.
+\url{https://doi.org/10.1002/joc.3370110504}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-krokhin2007}{}}%
+Krokhin, V.
+V., and W.
+M.
+J.
+Luxemburg.
+2007.
+{``Temperatures and Precipitation Totals over the {Russian Far East} and {Eastern Siberia}: Long-Term Variability and Its Links to Teleconnection Indices.''} \emph{Hydrology and Earth System Sciences} 11 (6): 1831--41.
+\url{https://doi.org/10.5194/hess-11-1831-2007}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-laurel-castillo2020}{}}%
+Laurel-Castillo, J.
+A., and A.
+Valle-Levinson.
+2020.
+{``Tidal and {Subtidal Variations} in {Water Level Produced} by {Ocean-River Interactions} in a {Subtropical Estuary}.''} \emph{Journal of Geophysical Research: Oceans} 125 (7): e2018JC014116.
+\url{https://doi.org/10.1029/2018JC014116}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-lim2018}{}}%
+Lim, E.-P., H.
+H.
+Hendon, and D.
+W.
+J.
+Thompson.
+2018.
+{``Seasonal {Evolution} of {Stratosphere-Troposphere Coupling} in the {Southern Hemisphere} and {Implications} for the {Predictability} of {Surface Climate}.''} \emph{Journal of Geophysical Research: Atmospheres} 123 (21): 12, 002--12, 016.
+\url{https://doi.org/10.1029/2018JD029321}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-vanloon1972}{}}%
+Loon, Harry van, and Roy L.
+Jenne.
+1972.
+{``The Zonal Harmonic Standing Waves in the Southern Hemisphere.''} \emph{Journal of Geophysical Research} 77 (6): 992--1003.
+\url{https://doi.org/10.1029/JC077i006p00992}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-mo2000}{}}%
+Mo, Kingtse C.
+2000.
+{``Relationships Between {Low-Frequency Variability} in the {Southern Hemisphere} and {Sea Surface Temperature Anomalies}.''} \emph{Journal of Climate} 13 (20): 3599--3610.
+\url{https://doi.org/10.1175/1520-0442(2000)013\%3C3599:RBLFVI\%3E2.0.CO;2}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-mo2001}{}}%
+Mo, Kingtse C., and Julia N.
+Paegle.
+2001.
+{``The {Pacific}--{South American} Modes and Their Downstream Effects.''} \emph{International Journal of Climatology} 21 (10): 1211--29.
+\url{https://doi.org/10.1002/joc.685}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-nuncio2015}{}}%
+Nuncio, M., and Xiaojun Yuan.
+2015.
+{``The {Influence} of the {Indian Ocean Dipole} on {Antarctic Sea Ice}*.''} \emph{Journal of Climate} 28 (7): 2682--90.
+\url{https://doi.org/10.1175/JCLI-D-14-00390.1}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-pezza2012}{}}%
+Pezza, Alexandre Bernardes, Harun A.
+Rashid, and Ian Simmonds.
+2012.
+{``Climate Links and Recent Extremes in Antarctic Sea Ice, High-Latitude Cyclones, {Southern Annular Mode} and {ENSO}.''} \emph{Climate Dynamics} 38 (1): 57--73.
+\url{https://doi.org/10.1007/s00382-011-1044-y}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-plumb1985}{}}%
+Plumb, R.
+Alan.
+1985.
+{``On the {Three-Dimensional Propagation} of {Stationary Waves}.''} \emph{Journal of the Atmospheric Sciences} 42 (3): 217--29.
+\url{https://doi.org/10.1175/1520-0469(1985)042\%3C0217:OTTDPO\%3E2.0.CO;2}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-rcoreteam2020}{}}%
+R Core Team. 2020. \emph{R: {A} Language and Environment for Statistical Computing}. Manual. {Vienna, Austria}: {R Foundation for Statistical Computing}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-raphael2004}{}}%
+Raphael, M. N. 2004. {``A Zonal Wave 3 Index for the {Southern Hemisphere}.''} \emph{Geophysical Research Letters} 31 (23). \url{https://doi.org/10.1029/2004GL020365}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-raphael2007}{}}%
+---------.
+2007.
+{``The Influence of Atmospheric Zonal Wave Three on {Antarctic} Sea Ice Variability.''} \emph{Journal of Geophysical Research: Atmospheres} 112 (D12).
+\url{https://doi.org/10.1029/2006JD007852}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-raphael2003}{}}%
+Raphael, Marilyn.
+2003.
+{``Recent, {Large-Scale Changes} in the {Extratropical Southern Hemisphere Atmospheric Circulation}.''} \emph{Journal of Climate} 16 (17): 2915--24.
+\url{https://doi.org/10.1175/1520-0442(2003)016\%3C2915:RLCITE\%3E2.0.CO;2}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-smith1995}{}}%
+Smith, Anne K.
+1995.
+{``Numerical Simulation of Global Variations of Temperature, Ozone, and Trace Species in the Stratosphere.''} \emph{Journal of Geophysical Research: Atmospheres} 100 (D1): 1253--69.
+\url{https://doi.org/10.1029/94JD02395}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-trenberth1980a}{}}%
+Trenberth, Kevin E.
+1980.
+{``Planetary {Waves} at 500 Mb in the {Southern Hemisphere}.''} \emph{Monthly Weather Review} 108 (9): 1378--89.
+\url{https://doi.org/10.1175/1520-0493(1980)108\%3C1378:PWAMIT\%3E2.0.CO;2}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-trenberth1985}{}}%
+Trenberth, Kevin F., and K.
+C.
+Mo.
+1985.
+{``Blocking in the {Southern Hemisphere}.''} \emph{Monthly Weather Review} 113 (1): 3--21.
+\url{https://doi.org/10.1175/1520-0493(1985)113\%3C0003:BITSH\%3E2.0.CO;2}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-turner2017}{}}%
+Turner, John, J.
+Scott Hosking, Thomas J.
+Bracegirdle, Tony Phillips, and Gareth J.
+Marshall.
+2017.
+{``Variability and Trends in the {Southern Hemisphere} High Latitude, Quasi-Stationary Planetary Waves.''} \emph{International Journal of Climatology} 37 (5): 2325--36.
+\url{https://doi.org/10.1002/joc.4848}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-walker1914}{}}%
+Walker, Sir Gilbert Thomas.
+1914.
+\emph{Correlation in {Seasonal Variations} of {Weather}, {III}: {On} the {Criterion} for the {Reality} of {Relationships Or Periodicities}}.
+{Meteorological Office}.
+\url{https://books.google.com?id=cuGtuAAACAAJ}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-wickham2009}{}}%
+Wickham, Hadley. 2009. \emph{Ggplot2: {Elegant Graphics} for {Data Analysis}}. Use {R}! {New York}: {Springer-Verlag}. \url{https://doi.org/10.1007/978-0-387-98141-3}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-wilks2016}{}}%
+Wilks, D.
+S.
+2016.
+{``{`{The Stippling Shows Statistically Significant Grid Points}'}: {How Research Results} Are {Routinely Overstated} and {Overinterpreted}, and {What} to {Do} about {It}.''} \emph{Bulletin of the American Meteorological Society} 97 (12): 2263--73.
+\url{https://doi.org/10.1175/BAMS-D-15-00267.1}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-wilks2011}{}}%
+Wilks, Daniel. 2011. \emph{Statistical {Methods} in the {Atmospheric Sciences}}. Vol. 100. {Elsevier}. \url{https://doi.org/10.1016/B978-0-12-385022-5.00022-1}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-wirth1993}{}}%
+Wirth, Volkmar.
+1993.
+{``Quasi-Stationary Planetary Waves in Total Ozone and Their Correlation with Lower Stratospheric Temperature.''} \emph{Journal of Geophysical Research: Atmospheres} 98 (D5): 8873--82.
+\url{https://doi.org/10.1029/92JD02820}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-cmap}{}}%
+Xie, Pingping, and Phillip A.
+Arkin.
+1997.
+{``Global {Precipitation}: {A} 17-{Year Monthly Analysis Based} on {Gauge Observations}, {Satellite Estimates}, and {Numerical Model Outputs}.''} \emph{Bull.
+Amer.
+Meteor.
+Soc.} 78 (Accessed on {\(<\)}26-08-2021{\(>\)}) (11): 2539--58.
+\url{https://doi.org/10.1175/1520-0477(1997)078\%3C2539:GPAYMA\%3E2.0.CO;2}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-xie2015}{}}%
+Xie, Yihui. 2015. \emph{Dynamic Documents with {R} and Knitr}. Second. {Boca Raton, Florida}: {Chapman and Hall/CRC}.
+
+\end{CSLReferences}
+
 \bibliographystyle{spbasic}

### 2022-03-21 analysis/paper/shceof.Rmd
Agrega DMI

@@
   .[, .(oni = mean(oni)), by = .(time = seasonally(time))]
+
+dmi <- download_dmi() %>%
+  as.data.table() %>%
+  .[, .(time = lubridate::as_datetime(Date), dmi = DMI)] %>%
+  na.omit() %>%
+  .[season(time) == "SON"] %>%
+  .[, dmi := dmi*season_weights(time)] %>% 
+  .[, .(dmi = mean(dmi)), by = .(time = seasonally(time))]
 
@@
 
+```{r dmi_cor}
+dmi_cor <-  ceof[, eof[[1]]$left, by = .(season)] %>%
+  sep_ReIm() %>%
+  .[, hgt := scale(hgt), by = .(cEOF, part)] %>%
+  dmi[., on = "time"] %>%
+  na.omit() %>%
+  .[, correlate(hgt, dmi), by = .(cEOF, part)]
+```
+
 ```{r spoke}
@@
 On top of this ENSO-like pattern, there are positive anomalies in the western Indian Ocean and negative values in the eastern Indian Ocean, resembling a positive Indian Ocean Dipole. 
+The correlation between the Imaginary cEOF2 and the DMI is `r dmi_cor[cEOF == "cEOF2" & part == "Imaginary"]$text`.
 

### 2022-03-21 analysis/paper/shceof.Rmd
Agrega datos de DMI / ENSO

@@
 The rainfall gridded dataset is based on information from different sources such as rain gauge observations, satellite inferred estimations and the NCEP-NCAR reanalysis, and it is available since 1979 to present.
+
+The Oceanic Niño Index [ONI, @bamston1997] comes from NOAA's Climate Prediction Center and and the Dipole Mode Index [DMI, @saji2003] from Global Climate Observing System Working Group on Surface Pressure.
+In both cases, the monthly time series averaged seasonally to produce seasonal time series.
 
@@
 All graphics are made using ggplot2 [@wickham2009].
-We downloaded data from reanalysis using the ecmwfr package [@hufkens2020] and indices of the ENSO with the rsoi package [@albers2020].
+We downloaded data from reanalysis using the ecmwfr package [@hufkens2020] and indices of ENSO and Indian Ocean Dipole with the rsoi package [@albers2020].
 The paper was rendered using knitr and rmarkdown [@xie2015; @allaire2020].
@@
 On top of this ENSO-like pattern, there are positive anomalies in the western Indian Ocean and negative values in the eastern Indian Ocean, resembling a positive Indian Ocean Dipole. 
-The correlation between the Imaginary cEOF2 and the DMI is `r dmi_cor[cEOF == "cEOF2" & part == "Imaginary"]$text`.
+Unsurprisingly, the correlation between the Imaginary cEOF2 and the DMI is `r dmi_cor[cEOF == "cEOF2" & part == "Imaginary"]$text`.
 
@@
 
+* The Oceanic Niño Index is available via NOAA's Climate Prediction Center: <https://www.cpc.ncep.noaa.gov/products/analysis_monitoring/ensostuff/detrend.nino34.ascii.txt>.
+
+* The Dipole Mode Index is available via Global Climate Observing System Working Group on Surface Pressure: <https://psl.noaa.gov/gcos_wgsp/Timeseries/Data/dmi.had.long.data>
+
 \appendix

### 2022-03-21 R/download-dmi.R

@@ -1,0 +1,39 @@
+#' Downloads DMI
+#'
+#' From https://github.com/boshek/rsoi/pull/43
+#'
+#' @export
+download_dmi <-  function() {
+  dmi_link = "https://psl.noaa.gov/gcos_wgsp/Timeseries/Data/dmi.had.long.data"
+
+  res = tryCatch(
+    rsoi:::check_response(dmi_link),
+    error = function(e) {
+      message(e)
+      return(invisible(NULL))
+    }
+  )
+
+  years = strsplit(trimws(readLines(res, n = 1)), " ")[[1]]
+  rows = diff(as.numeric(years)) + 1
+  months = month.abb
+
+  dmi =  read.csv(rsoi:::check_response(dmi_link),    # reset connection
+                  header = FALSE,
+                  col.names = c("Year", months),
+                  nrows = rows,
+                  skip = 1,
+                  na.strings = "-9999.000",
+                  sep = "",
+                  stringsAsFactors = FALSE)
+
+  grid = expand.grid(Year = dmi$Year, Month = months)
+  grid$DMI = c(as.matrix(dmi[, months]))
+  grid$Month = factor(grid$Month, levels = months)
+  grid$Date = as.Date(paste0(grid$Year, "-", as.numeric(grid$Month) ,"-01"), "%Y-%m-%d")
+
+  grid <- grid[, c("Year", "Month", "Date", "DMI")]
+  class(grid) = c("tbl_df", "tbl", "data.frame")
+
+  return(grid)
+}

### 2022-03-21 analysis/paper/shceof.Rmd
Pone referencias dinámicas a las "In Section 1..."

@@
 
-In Section 2 we describe the methods.
-In Section 3.1, we analyse the spatial patterns of each complex EOF.
-In Section 3.2 we study the spatial regressions with geopotential height and temperature anomalies.
-In Section 3.3, we analyse the relationship between cEOF1 and Ozone, and between cEOF2 with the SAM and the PSA.
-In Section 3.4 we study tropical forcings that explain the variability of each cEOF.
-Finally, in Section 3.5 we show the relationship between these modes of variability and precipitation anomalies in South America and Oceania.
+In Section \@ref(methods) we describe the methods.
+In Section \@ref(spatial) we analyse the spatial patterns of each complex EOF.
+In Section \@ref(regressions) we study the spatial regressions with geopotential height and temperature anomalies.
+In Section \@ref(other-variables) we analyse the relationship between cEOF1 and Ozone, and between cEOF2 with the SAM and the PSA.
+In Section \@ref(tropical) we study tropical forcings that explain the variability of each cEOF.
+Finally, in Section \@ref(precipitation) we show the relationship between these modes of variability and precipitation anomalies in South America and Oceania.
 

### 2022-03-21 analysis/paper/shceof.Rmd
Agrega detalle a los estudios de cEOF

@@
 An alternative methodology that has been proposed to study travelling and standing waves is complex Empirical Orthogonal Functions [@horel1984].
-This method extends EOF analysis to capture oscillations with varying amplitude and phase and has been applied to the time domain on water level to study tide-river interactions [@laurel-castillo2020] and temperature and precipitation anomalies [e.g.
-@krokhin2007; @gelbrecht2018].
+This method extends EOF analysis to capture oscillations with varying amplitude and phase and has been applied to the time domain. For instance, 
+@krokhin2007 applied cEOF to station-based monthly precipitation anomalies and monthly temperature anomalies in the Eastern Siberia and the Far East region to characterise the main modes of variability and their connection to teleconnection indices.
+Similarly, @gelbrecht2018 used cEOF applied to daily precipitation from reanalysis to study the propagating characteristics of the South American Monsoon. 
 To our knowledge, complex EOF has not been applied in the spatial domain to capture the phase-varying nature of planetary waves.

### 2022-03-23 analysis/paper/shceof.Rmd
Saca lineas al pedo
Elimina chunk innecesario

@@
 The magnitude and argument of each complex time series represent the amplitude and phase of each zonal wave.
-
-```{r corr-eofs}
-
-background_palette <- grDevices::colorRampPalette(c("white", scales::muted("red")),
-                                                  space = "Lab")
-# 
-# eof_naive[, eof[[1]]$left, by = lev] %>% 
-#   .[, item := paste(lev, EOF, sep = "_")] %>% 
-#   widyr::pairwise_cor(item, time, hgt) %>% 
-#   setDT() %>%
-#   # .[]
-#   .[tstrsplit(item1, "_")[[1]] != tstrsplit(item2, "_")[[1]], ] %>% 
-#   .[tstrsplit(item1, "_")[[1]] != 50] %>% 
-#   .[, item1 := tstrsplit(item1, "_")[[2]]] %>%
-#   .[, item2 := tstrsplit(item2, "_")[[2]]] %>% 
-#   .[, r2 := round(correlation^2, 2)] %>% 
-#   dcast(item1 ~ item2, value.var = "r2") %>% 
-#   setnames("item1", "200 hPa") %>% 
-#   kbl2(booktabs = TRUE, caption = "$R^2$ between pairs of EOFs of each level rounded to two decimals.") %>% 
-#   # add_header_above(c("", "50 hPa" = 4)) %>% 
-#   column_spec2(2:5,  background = spec_color2(.table[[.col]], option = background_palette, 
-#                                               scale_from = c(0, 1)),
-#                color = ifelse(.table[[.col]] > .5, "white", "black")) %>% 
-#   add_header_above(c("", "50 hPa" = 4)) %>%
-#   kable_classic()
-```
 
@@
 (ref:pp-temp-cap-1) Same as Figure \@ref(fig:pp-temp-2) but for cEOF1.
+(ref:pp-temp-cap-1) Same as Figure \@ref(fig:pp-temp-2) but for cEOF1.
 

### 2022-03-23 analysis/paper/shceof.Rmd
Revisión

@@
 Addressing this limitation, the Fourier technique can be generalized to integrate all planetary wave amplitude regardless of wave number, by computing the wave envelope [@irving2015].
-The latter makes possible to represent planetary waves with different amplitude at different longitudes, but it removes all information about phase and wave number.
+The latter makes it possible to represent planetary waves with different amplitude at different longitudes, but it removes all information about phase and wave number.
 With this approach, @irving2015 showed that planetary wave amplitude in general is associated to Antarctic sea-ice concentration and temperature, as well as to precipitation anomalies in regions of significant topography in SH mid-latitudes and Antarctica.
@@
 
-The Oceanic Niño Index [ONI, @bamston1997] comes from NOAA's Climate Prediction Center and and the Dipole Mode Index [DMI, @saji2003] from Global Climate Observing System Working Group on Surface Pressure.
-In both cases, the monthly time series averaged seasonally to produce seasonal time series.
+The Oceanic Niño Index [ONI, @bamston1997] comes from NOAA's Climate Prediction Center and the Dipole Mode Index [DMI, @saji2003] from Global Climate Observing System Working Group on Surface Pressure.
 
@@
 We compute seasonal means for the different variables, averaging monthly values weighted by the number of days in each month.
-We use 200 hPa level to represent the high troposphere and 50 hPa to represent the lower stratosphere.
+We use the 200 hPa level to represent the high troposphere and 50 hPa to represent the lower stratosphere.
 
@@
 
-We computed the level-dependent SAM index as the leading EOF of year-round monthly geopotential height anomalies south of 20ºS at each level for the whole period according [@baldwin2009].
+We computed the level-dependent SAM index as the leading EOF of year-round monthly geopotential height anomalies south of 20ºS at each level for the whole period [@baldwin2009].
 We further split the SAM into its zonally symmetric and zonally asymmetric components (S-SAM and A-SAM indices respectively).
 These indices were obtained by projecting the zonally asymmetric and zonally symmetric part of the SAM spatial pattern onto monthly geopotential height fields, as proposed by @campitelli2022.
-Seasonal indices of the Pacific South American patterns (PSA1 and PSA2) were calculated -- in agreement with @mo2001 -- as the third and fourth leading EOF of seasonal mean anomalies for SH 500 hPa geopotential height.
+Seasonal indices of the Pacific South American patterns (PSA1 and PSA2) were calculated, in agreement with @mo2001, as the third and fourth leading EOF of seasonal mean anomalies for 500-hPa geopotential heights at SH.
 
@@
 
-In standard way of applying EOF analysis zonal waves may appear as pairs of EOFs, that could be degenerated representing similar patterns but shifted in phase [@horel1984].
-Figure \@ref(fig:eof-naive) shows the four eading EOFs of SON geopotential height zonal anomalies at 50 hPa south of 20\degree S.
-It is clear that the first two EOFs represent the same zonal wave 1 pattern and the last two represent a same zonal wave pattern with shorter wavenumber and four centers of action shifted by 1/4 wavelength.
+In the standard EOF analysis, zonal waves may appear as pairs of (possibly degenerated) EOFs representing similar patterns but shifted in phase [@horel1984].
+Figure \@ref(fig:eof-naive) shows the four leading EOFs of SON geopotential height zonal anomalies at 50 hPa south of 20\degree S.
+It is clear that the first two EOFs represent a single phase-varying zonal wave 1 pattern and the last two represent a similarly phase-varying pattern with shorter wavenumber and four centres of action shifted by 1/4 wavelength.
 A similar EOF structure can be seen in 200 hPa (not shown).
-Since each pair of EOFs seem to represent the same phase-varying structure, it would be desirable to combine them into a single index with amplitude and phase.
-
-ON the other hand, complex Empirical Orthogonal Functions (cEOF) is a useful method to characterise zonal waves, associated with phase-varying structures [@horel1984].
+Since each pair of EOFs seems to represent the same phase-varying structure, it would be desirable to combine them into a single pattern described by amplitude and phase.
+
+Complex Empirical Orthogonal Functions (cEOF) are a useful method for characterising zonal waves, associated with phase-varying structures [@horel1984].
 This method involves computing EOF on the analytic representation of the original field.
@@
 
-The phase of principal components is defined up to an additive constant.
-For real principal components, this constant can be either 0 or $\pi$, corresponding to a change in sign.
-For complex principal components, it can be any number between 0 and $2\pi$ [@horel1984].
+The phase of cEOFs is defined up to an additive constant.
+For real EOFs, this constant can be either 0 or $\pi$, corresponding to a change in sign.
+For cEOFs, it can be any real number between 0 and $2\pi$ [@horel1984], corresponding to rotations in the complex plane.
 Since any choice is arbitrary and equally valid, we chose the phase of each cEOF so that the real and imaginary parts are aligned with meaningful phases in our analysis.
-This procedure does not create a spurious correlation, it only takes a known relationship and aligns it with a specific phase.
+This procedure does not create a spurious correlation, it only takes an existing relationship and aligns it with a specific phase.
 For the first complex principal component, the phase was chosen so that the time series corresponding to the real part has the maximum correlation with the zonal wave 1 of Total Ozone Column between `r knitr::combine_words(LatLabel(o3wave_lats))`.
@@
 
-Lastly, the cEOF1 doesn't have a significant signal associated with 2-metre temperature or precipitacion (Fig. \@ref(fig:pp-temp-cap-1)).
-
 # Conclusions
 
-In this study we assessed the Austral Spring Extratropical Southern Hemisphere zonally assymetric circulation.
+In this study we assessed the Austral Spring Extratropical Southern Hemisphere zonally asymmetric circulation.
 For this purpose, two complex indices were derived using Complex Empirical Orthogonal Functions.

### 2022-03-23 analysis/paper/shceof.Rmd
Justifica mejor la rotación de los cEOFs

@@
 For cEOFs, it can be any real number between 0 and $2\pi$ [@horel1984], corresponding to rotations in the complex plane.
-Since any choice is arbitrary and equally valid, we chose the phase of each cEOF so that the real and imaginary parts are aligned with meaningful phases in our analysis.
-This procedure does not create a spurious correlation, it only takes an existing relationship and aligns it with a specific phase.
-For the first complex principal component, the phase was chosen so that the time series corresponding to the real part has the maximum correlation with the zonal wave 1 of Total Ozone Column between `r knitr::combine_words(LatLabel(o3wave_lats))`.
-This also nearly minimises the correlation with the imaginary part.
-For the second complex principal component, the phase was chosen so that the coefficient of determination between the Oceanic Niño Index [@bamston1997] and the real part was minimized, which also nearly maximizes the correlation with the imaginary part.
+Since any choice is arbitrary and equally valid, we chose the phase of each cEOF so that the real or imaginary parts are aligned with meaningful phases in our analysis.
+
+Preliminary analysis showed that the first cEOF was closely related to the the zonal wave 1 of Total Ozone Column and the second cEOF was closely related to ENSO. 
+Therefore, to aid in the interpretation, we chose the phases of each cEOF so that these relationships are aligned with one of the complex components.
+
+We chose the phase of cEOF1 so that the time series corresponding to the real part has the maximum correlation with the zonal wave 1 of Total Ozone Column between `r knitr::combine_words(LatLabel(o3wave_lats))`.
+And we chose the phase of cEOF2 so that the coefficient of determination between the Oceanic Niño Index [@bamston1997] and the real part is minimized, which also nearly maximizes the correlation with the imaginary part.
+
+This procedure does not create spurious correlations, it only takes an existing relationship and aligns it with a specific phase.
 

### 2022-03-23 analysis/paper/shceof.Rmd
Uses rsoi::download_dmi()

@@
 
-dmi <- download_dmi() %>%
+dmi <- rsoi::download_dmi(TRUE, data_path("raw", "dmi.csv")) %>%
   as.data.table() %>%

### 2022-03-23 R/download-dmi.R

@@ -1,39 +1,0 @@
-#' Downloads DMI
-#'
-#' From https://github.com/boshek/rsoi/pull/43
-#'
-#' @export
-download_dmi <-  function() {
-  dmi_link = "https://psl.noaa.gov/gcos_wgsp/Timeseries/Data/dmi.had.long.data"
-
-  res = tryCatch(
-    rsoi:::check_response(dmi_link),
-    error = function(e) {
-      message(e)
-      return(invisible(NULL))
-    }
-  )
-
-  years = strsplit(trimws(readLines(res, n = 1)), " ")[[1]]
-  rows = diff(as.numeric(years)) + 1
-  months = month.abb
-
-  dmi =  read.csv(rsoi:::check_response(dmi_link),    # reset connection
-                  header = FALSE,
-                  col.names = c("Year", months),
-                  nrows = rows,
-                  skip = 1,
-                  na.strings = "-9999.000",
-                  sep = "",
-                  stringsAsFactors = FALSE)
-
-  grid = expand.grid(Year = dmi$Year, Month = months)
-  grid$DMI = c(as.matrix(dmi[, months]))
-  grid$Month = factor(grid$Month, levels = months)
-  grid$Date = as.Date(paste0(grid$Year, "-", as.numeric(grid$Month) ,"-01"), "%Y-%m-%d")
-
-  grid <- grid[, c("Year", "Month", "Date", "DMI")]
-  class(grid) = c("tbl_df", "tbl", "data.frame")
-
-  return(grid)
-}

### 2022-03-23 analysis/paper/shceof.Rmd
Revisión

@@
 
-We used monthly geopotential height, air temperature, ozone mixing ratio, and total ozone column (TOC) at 2.5\degree longitude by 2.5\degree latitude of horizontal resolution and 37 vertical isobaric levels from the European Centre for Medium-Range Weather Forecasts Reanalysis version 5 [ERA; @era5] for the period 1979-2019.
+We used monthly geopotential height, air temperature, ozone mixing ratio, and total ozone column (TOC) at 2.5\degree longitude by 2.5\degree latitude of horizontal resolution and 37 vertical isobaric levels from the European Centre for Medium-Range Weather Forecasts Reanalysis version 5 [ERA; @era5] for the period 1979 -- 2019.
 Most of our analysis is restricted to the post-satellite era to avoid any confounding factors arising from the incorporation of satellite observations, but we also used the preliminary back extension of ERA5 from 1950 to 1978 [@era5be] to describe long-term trends.
@@
 $$
-X(\lambda, \phi, t) = \alpha(\lambda, \phi) \operatorname{Im}(\operatorname{cEOF}) + \beta(\lambda, \phi) \operatorname{Re}(\operatorname{cEOF}) + X_0(\lambda, \phi) +  \epsilon(\lambda, \phi, t)
+X(\lambda, \phi, t) = \alpha(\lambda, \phi) \operatorname{Im}(\mathrm{cEOF}) + \beta(\lambda, \phi) \mathrm{Re}(\mathrm{cEOF}) + X_0(\lambda, \phi) +  \epsilon(\lambda, \phi, t)
 $$
@@
 
-(ref:extended-series-cap) Time series extended using ERA5 back extended preliminary edition (period 1950 -- 1978) and ERA5 (period 1979 -- 2019).
-Dark straight line is the linear trend.
-Black horizontal and vertical line mark the mean value and range of each time series, respectively.
+(ref:extended-series-cap) Time series of the two leading cEOFs of SON zonal anomalies of geopotential height at 50 hPa and 200 hPa.
+cEOF1 (row a) and cEOF2 (row b) separated in their real (column 1) and imaginary (column 2) components. 
+Dark straight line is the linear trend. Black horizontal and vertical line mark the mean value and range of each time series, respectively.
 
@@
 To describe the variability of the circulation zonal anomalies, the spatial and temporal parts of the first two leading cEOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa, computed jointly at both levels, are shown in Figures \@ref(fig:ceofs-1) and \@ref(fig:extended-series).
-The first mode (cEOF1) explains 82% of the variance, while the second mode (cEOF2) explains a smaller fraction (7%).
+The first mode (cEOF1) explains 82% of the variance of the zonally anomalous fields, while the second mode (cEOF2) explains a smaller fraction (7%).
 In the spatial patterns (Fig. \@ref(fig:ceofs-1)), the real and the imaginary components are in quadrature by construction, so that each cEOF describe a single wave-like pattern whose amplitude and position (i.e.
@@
 In particular, the dominant structure at both levels is a wave 3 but with larger amplitude in the pacific sector.
-This modulated amplitude is especially apparent at 200 hPa.
-There is no apparent phase shift but the amplitude of the pattern is greatly reduced in the stratosphere, suggesting that this barotropic mode represents mainly tropospheric variability.
-
-All cEOFs have non zero mean (Fig.
-\@ref(fig:extended-series)) because the geopotential fields that enter into the cEOFs algorithm are anomalies with respect to the zonal mean instead of the time mean.
-However, cEOF2's mean is almost zero, which indicates that only cEOF1 includes variability that significantly projects onto the mean zonally anomalous field.
-There is no significant simultaneous correlation between the time series.
-Both cEOFs show year-to-year variability but show no evidence of decadal variability.
+This modulated amplitude more evident at 200 hPa.
+There is no apparent phase shift with height but the amplitude of the pattern is greatly reduced in the stratosphere, suggesting that this barotropic mode represents mainly tropospheric variability.
+
+There is no significant simultaneous correlation between cEOFs time series.
+Both cEOFs show year-to-year variability but show no evidence of decadal variability (Fig. \@ref(fig:extended-series)).
+Because the geopotential fields that enter into the cEOFs algorithm are anomalies with respect to the zonal mean instead of the time mean, the cEOFs have non zero temporal mean. 
+However, cEOF2’s mean is almost zero, which indicates that only cEOF1 includes variability that significantly projects onto the mean zonally anomalous field.
 
 A significant upward trend in the real component of cEOF1 is evident (Fig. \@ref(fig:extended-series)a.1, p-value `r trends[part == "Real" & cEOF == "cEOF1"]$p_report`) while there is no significant trend in any of the complex components of cEOF2.
-The positive trend in the Real cEOF1 translates into a positive trend in cEOF1 magnitude, but not in the phase (not shown).
+The positive trend in the Real cEOF1 translates into a positive trend in cEOF1 magnitude, but not systematic change in phase (not shown).
 This long-term change indicates an increase in the magnitude of the high latitude zonal wave 1.
@@
 
-## cEOFs Regressions {#regressions}
+## cEOFs Regression maps {#regressions}
 

### 2022-03-23 analysis/paper/shceof.Rmd
Usa visual editor.

@@
 An alternative methodology that has been proposed to study travelling and standing waves is complex Empirical Orthogonal Functions [@horel1984].
-This method extends EOF analysis to capture oscillations with varying amplitude and phase and has been applied to the time domain. For instance, 
-@krokhin2007 applied cEOF to station-based monthly precipitation anomalies and monthly temperature anomalies in the Eastern Siberia and the Far East region to characterise the main modes of variability and their connection to teleconnection indices.
+This method extends EOF analysis to capture oscillations with varying amplitude and phase and has been applied to the time domain.
+For instance, @krokhin2007 applied cEOF to station-based monthly precipitation anomalies and monthly temperature anomalies in the Eastern Siberia and the Far East region to characterise the main modes of variability and their connection to teleconnection indices.
 Similarly, @gelbrecht2018 used cEOF applied to daily precipitation from reanalysis to study the propagating characteristics of the South American Monsoon.
@@
 -->
+
 <!-- Zonal asymmetries of extratropical circulation in the Southern Hemisphere strongly modulate weather systems and regional climate through promoting longitudinally different latitudinal transport of heat, humidity, and momentum [@trenberth1980].
@@
 <!-- Comparan eventos ENSO del 97 y de 2002 -->
+
 <!-- La posición de las anomalías de circulación (PSA) son importantes para determinar el impacto del ENSO en sudamérica.
@@
 <!-- > However, the spatial pattern of the PSA wave train excited by ENSO seems to be -->
+
 <!-- > non-stationary. Fogt and Bromwich (2006) indicated that the PSA mode moved -->
+
 <!-- > eastwards by 20° longitude in the 1990s owing to the eastward shift of the tropical -->
+
 <!-- > convective position. In addition, Wilson et al. (2014, 2016) and Ciasto et al. (2015) -->
+
 <!-- > found that the different types of ENSO can contribute to the wave shift. Specifically, -->
+
 <!-- > Rossby waves excited by central Pacific (CP)–type El Niño shift westwards owing to -->
+
 <!-- > the westward movement of the tropical diabatic heating. -->
@@
 cEOF1 (row a) and cEOF2 (row b) separated in their real (column 1) and imaginary (column 2) components.
-Dark straight line is the linear trend. Black horizontal and vertical line mark the mean value and range of each time series, respectively.
+Dark straight line is the linear trend.
+Black horizontal and vertical line mark the mean value and range of each time series, respectively.
 
@@
 Because the geopotential fields that enter into the cEOFs algorithm are anomalies with respect to the zonal mean instead of the time mean, the cEOFs have non zero temporal mean.
-However, cEOF2’s mean is almost zero, which indicates that only cEOF1 includes variability that significantly projects onto the mean zonally anomalous field.
+However, cEOF2's mean is almost zero, which indicates that only cEOF1 includes variability that significantly projects onto the mean zonally anomalous field.
 
@@
 Their phases coincide with the ones shown in Figure \@ref(fig:ceofs-1)a.1, with the positive centre of the Real cEOF1 located towards the dateline, and the one of the Imaginary cEOF1 located over East Antarctica.
-Instead of a wave-1 pattern, regressed anomalies describe a monopole with its centre displaced from the South Pole. The latter indicates an influence of cEOF1 in modulating the polar vortex.
+Instead of a wave-1 pattern, regressed anomalies describe a monopole with its centre displaced from the South Pole.
+The latter indicates an influence of cEOF1 in modulating the polar vortex.
 
@@
 
-(ref:sam-eof-vertical-cap) Coefficient of determination between the real and imaginary part of each cEOF and the SAM, Asymmetric SAM (A-SAM) and Symmetric SAM (S-SAM) indices computed at each level according to @@campitelli2022 for the 1979 -- 2019 period.
+(ref:sam-eof-vertical-cap) Coefficient of determination between the real and imaginary part of each cEOF and the SAM, Asymmetric SAM (A-SAM) and Symmetric SAM (S-SAM) indices computed at each level according to \@@campitelli2022 for the 1979 -- 2019 period.
 Thick lines represent estimates with p-value \< 0.01 corrected for False Detection Rate [@benjamini1995].
@@
 In the stratosphere, the Real cEOF1 is correlated with both A-SAM and S-SAM, while the Imaginary cEOF1 is highly correlated only with the A-SAM.
-These correlations are consistent with the regression patterns of geopotential height in Figure \@ref(fig:eof1-regr-gh) and their comparison with those obtained for SAM, A-SAM and S-SAM by @@campitelli2022.
+These correlations are consistent with the regression patterns of geopotential height in Figure \@ref(fig:eof1-regr-gh) and their comparison with those obtained for SAM, A-SAM and S-SAM by \@@campitelli2022.
 
@@
 
-The regression maps of cEOF1 with fields of Total Ozone Column (TOC) (Fig.
- \@ref(fig:o3-regr)) show zonal wave 1 pattens in TOC associated with both phases of cEOF1.
+The regression maps of cEOF1 with fields of Total Ozone Column (TOC) (Fig.  \@ref(fig:o3-regr)) show zonal wave 1 pattens in TOC associated with both phases of cEOF1.
 Climatologically, the springtime Ozone minimum is located off the South Pole and towards the Weddell Sea [@grytsai2011].
@@
 The correlation between the Real cEOF2 and ENSO is also not significant (`r enso_cor[cEOF == "cEOF2" & part == "Real"]$text`).
-Rows a and c in Fig. \@ref(fig:sst-psi-2) show that the intermediate phases are still associated with significant SST anomalies over the Pacific Ocean, but with with slightly different location.
+Rows a and c in Fig. 
+\@ref(fig:sst-psi-2) show that the intermediate phases are still associated with significant SST anomalies over the Pacific Ocean, but with with slightly different location.
 The 135º phase is associated with SST in the central Pacific, while the 45º phase is associated with SST in the eastern Pacific, which correspond roughly to the Eastern Pacific and Central Pacific "flavours" of ENSO (Citas).
@@
 Years with magnitude of cEOF2 greater or smaller than the 50th percentile are shown as orange diamonds and green circles and respectively.
-Black line is the fit ONI ~ sin(phase) computed by weighted OLS using the magnitude of the cEOF2 as weights.
+Black line is the fit ONI \~ sin(phase) computed by weighted OLS using the magnitude of the cEOF2 as weights.
 
@@
 
-<!-- To better understand the relationship between precipitation anomalies over eastern Australia and the cEOF2, Figure \@ref(fig:australia-pp-phase) plots precipitation anomalies averaged over eastern Australia (box in Figure \@ref(fig:pp-oceania)) as a function of cEOF2 phase.
+<!-- To better understand the relationship between precipitation anomalies over eastern Australia and the cEOF2, Figure @ref(fig:australia-pp-phase) plots precipitation anomalies averaged over eastern Australia (box in Figure @ref(fig:pp-oceania)) as a function of cEOF2 phase.
 -->
+
 <!-- The black line shows a sinusoidal fit, whose coefficients are fitted by weighted least squares using the magnitude of the EFO2 as weight. -->
+
 <!-- The $r^2$ corresponding to the fit is `r signif(summary(model)$r.squared, 2)` (p-value `r report_p(p_value)`). -->
+
 <!-- There is a strong relationship between the phase of the cEOF2 and precipitation anomalies in eastern Australia, but the Real and Imaginary phases we chose are not exactly aligned with the direction that maximises this relationship.
@@
 Over South America, the Imaginary cEOF has positive correlations with precipitation in South Eastern South America (SESA) and central Chile, and negative correlations in eastern Brazil.
-This again is a well-known springtime precipitation signature of ENSO [@cai2020a] and it is also similar to the precipitation anomalies associated with the A-SAM [@@campitelli2022].
+This again is a well-known springtime precipitation signature of ENSO [@cai2020a] and it is also similar to the precipitation anomalies associated with the A-SAM [\@ @campitelli2022].
 
@@
 There is some disagreement in the literature of whether the phase of the PSA pattern is affected by the location of tropical SST anomalies [@irving2016].
-We show not only that the cEOF2 tends to be in the imaginary phase (~PSA1) when the ENSO region is warm, but that central Pacific SST anomalies tend to move the cEOF towards the negative real phase and eastern Pacific SST anomalies tend to move it towards positive real phase.
+We show not only that the cEOF2 tends to be in the imaginary phase (\~PSA1) when the ENSO region is warm, but that central Pacific SST anomalies tend to move the cEOF towards the negative real phase and eastern Pacific SST anomalies tend to move it towards positive real phase.
 When ENSO is neutral, the cEOF2 is still active, but with no particular preferred phase.
@@
 
-* The Oceanic Niño Index is available via NOAA's Climate Prediction Center: <https://www.cpc.ncep.noaa.gov/products/analysis_monitoring/ensostuff/detrend.nino34.ascii.txt>.
-
-* The Dipole Mode Index is available via Global Climate Observing System Working Group on Surface Pressure: <https://psl.noaa.gov/gcos_wgsp/Timeseries/Data/dmi.had.long.data>
+-   The Oceanic Niño Index is available via NOAA's Climate Prediction Center: <https://www.cpc.ncep.noaa.gov/products/analysis_monitoring/ensostuff/detrend.nino34.ascii.txt>.
+
+-   The Dipole Mode Index is available via Global Climate Observing System Working Group on Surface Pressure: <https://psl.noaa.gov/gcos_wgsp/Timeseries/Data/dmi.had.long.data>
 
@@
 (ref:pp-temp-cap-1) Same as Figure \@ref(fig:pp-temp-2) but for cEOF1.
+
 (ref:pp-temp-cap-1) Same as Figure \@ref(fig:pp-temp-2) but for cEOF1.

### 2022-03-23 analysis/paper/shceof.Rmd
Revisión

@@
                        guide = guide_colorsteps_bottom(15, even.steps = TRUE), 
-                       limits = c(-1800, 1800)) 
+                       limits = c(-2000, 2000)) 
 ```
@@
 Figure \@ref(fig:eof1-regr-gh) shows regression maps of SON geopotential height anomalies upon cEOF1.
-At 50 hPa (Figure \@ref(fig:eof1-regr-gh) row a), both the Real and Imaginary cEOF1 are associated with planetary wave 1 patterns, that are 90º out of phase.
-Their phases coincide with the ones shown in Figure \@ref(fig:ceofs-1)a.1, with the positive centre of the Real cEOF1 located towards the dateline, and the one of the Imaginary cEOF1 located over East Antarctica.
-Instead of a wave-1 pattern, regressed anomalies describe a monopole with its centre displaced from the South Pole.
-The latter indicates an influence of cEOF1 in modulating the polar vortex.
-
-The regression maps for cEOF1 at 200 hPa (Figure \@ref(fig:eof1-regr-gh) row b) are similarly influenced by the zonally symmetric circulation.
-It is only possible to partially distinguish the wave 1 pattern in relation with the Real cEOF1 (Figure \@ref(fig:eof1-regr-gh)b.1).
-The Imaginary cEOF1 shows a much more zonally symmetrical pattern resembling the negative SAM phase.
-
-With the exception of the Imaginary cEOF1, it is clear that these patterns are very different than the fully zonally asymmetric versions (Fig. \@ref(fig:ceofs-1)), particularly at 200 hPa.
-Moreover, only in the stratosphere these patterns actually show a distinguishable wave 1 pattern shifted in phase by 90º, suggesting that using the cEOF method is artificially generating a wave structure at 200hPa.
+At 50 hPa (Figure \@ref(fig:eof1-regr-gh) row a), the Real cEOF1 is associated with a monopole centred over the Ross Sea.
+The Imaginary cEOF1 is associated with a clear wave 1 pattern with maximum over the coast of East Antarctica.
+
+At 200 hPa (Figure \@ref(fig:eof1-regr-gh) row b) the Real cEOF1 shows a single area of positive anomalies spanning West Antarctica surrounded by opposite anomalies in lower latitudes with is centre shifted slightly eastward compared with the upper level anomalies.
+The Imaginary cEOF2.
+The Imaginary cEOF1 shows a much more zonally symmetrical pattern resembling the negative SAM phase [@fogt2020].
+Both components of cEOF1 could be associated to zonal anomalies in the upper troposphere jet stream intensity.
+
 Therefore, the magnitude and phase of the cEOF1 are associated with the magnitude and phase of a zonal wave only in the stratosphere.
@@
 
-(ref:eof2-regr-gh-cap) Same as Figure \@ref(fig:eof1-regr-gh) but for the second cEOF.
+(ref:eof2-regr-gh-cap) Same as Figure \@ref(fig:eof1-regr-gh) but for cEOF2.
 
@@
 
-Figure \@ref(fig:eof2-regr-gh) shows the regression pattern of geopotential height anomalies upon the cEOF2.
-Unlike for cEOF1, in this case the regression patterns are similar to the fully zonally asymmetric patterns from Figure \@ref(fig:ceofs-1).
-Although there are some differences (particularly in 50 hPa), the wave trains identified before are well characterised and patterns associated with the Real cEOF2 are 90º out of phase with those associated with the Imaginary cEOF2.
-Zonal wave 3 dominates all fields, but only in the western hemisphere, over the Pacific and Atlantic Oceans.
-cEOF2 then represents an equivalent barotropic wave train that is very similar to the the Pacific South American Patterns [@mo2001].
-Comparing the location of the positive anomaly near 90ºW in column b of Figure \@ref(fig:eof2-regr-gh) with Figures 1.a and b from @mo2001, the Real cEOF2 can roughly be identified with PSA2, while the Imaginary cEOF2 resembles PSA1.
-
-```{r zonal-waves, fig.cap = "Amplitud de ondas zonales de la regresión con altura geopotencial (figuras anteriores)"}
-# regr %>%
-#   .[lat <= -20] %>%
-#   .[lev %in% c(200, 50)] %>%
-#   # .[cEOF == "cEOF1"] %>%
-#   .[, FitWave(estimate, 0:3), by = .(lev, cEOF, term, lat)] %>%
-#   # .[, pseudor2 := amplitudepseudor2^2/sum(amplitude^2), by = .(lev, cEOF, term, lat)] %>% 
-#   ggplot(aes(lat, amplitude)) +
-#   geom_line(aes(color = factor(k))) +
-#   ggh4x::facet_nested(cEOF + lev ~ term, scales = "free", labeller = labeller(lev = lab_lev)) +
-#   scale_color_brewer("Wavenumber", palette = "Dark2") +
-#   scale_y_continuous("Amplitude") +
-#   scale_x_latitude() +
-#   coord_flip() +
-#   panel_background
-```
+Figure \@ref(fig:eof2-regr-gh) shows the regression maps of geopotential height anomalies upon the cEOF2.
+Wave trains similar to those identified for cEOF2 patterns (Fig \@ref(fig:ceofs-1)) can be distinguished.
+Regressed anomalies associated with the Real cEOF2 are 90º out of phase with those associated with the Imaginary cEOF2.
+All fields have a dominant zonal wave 3 limited to the western hemisphere, over the Pacific and Atlantic Oceans.
+cEOF2 then represents an equivalent barotropic wave train that is very similar to the the PSA Patterns [@mo2001].
+Comparing the location of the positive anomaly near 90ºW in column b of Figure \@ref(fig:eof2-regr-gh) with Figures 1.a and b from @mo2001, the Real cEOF2 regression map could be identified with PSA2, while the Imaginary cEOF2 resembles PSA1.
 

### 2022-03-25 analysis/paper/shceof.Rmd
Mueve sección de PSA

@@
       latex_engine: "xelatex"
+      keep_tex: true
       # citation_package: "natbib"
@@
 
-### SAM
-
-(ref:sam-eof-vertical-cap) Coefficient of determination between the real and imaginary part of each cEOF and the SAM, Asymmetric SAM (A-SAM) and Symmetric SAM (S-SAM) indices computed at each level according to \@@campitelli2022 for the 1979 -- 2019 period.
-Thick lines represent estimates with p-value \< 0.01 corrected for False Detection Rate [@benjamini1995].
-
-```{r sam-eof-vertical, fig.cap = "(ref:sam-eof-vertical-cap)", fig.height=4, fig.width=width_column}
-time <- ceof[, eof[[1]]$left, by = .(season)] %>%
-  sep_ReIm() %>%
-  .[, hgt := scale(hgt), by = .(cEOF, part)]
-
-sam_r2 <- sams %>%
-  .[time, on = .NATURAL, allow.cartesian = TRUE] %>%
-  na.omit() %>% 
-  .[, correlate(hgt, estimate), by = .(lev, term, part, cEOF)]
-
-max_r2 <- sam_r2[part == "Imaginary" & term == "asym"] %>% 
-  .[, .SD[which.max(estimate^2)], by = .(cEOF)]
-
-sam_r2 %>%
-  .[, p.value_a := p.adjust(p.value, "fdr")] %>% 
-  ggplot(aes(lev, estimate^2)) +
-  geom_vline(xintercept = c(50, 200),  size = 0.2, color = "gray50") +
-  # geom_point(data = ~.x[p.adjust(p.value, "fdr") < 0.01], aes(color = term)) +
-  geom_line(data = ~copy(.x)[p.value_a > 0.01, estimate := NA], 
-            aes(colour = stage(term, after_scale = scales::alpha(colour, 0.5))), 
-            size = 2) +
-  geom_line(aes(color = term, group = term)) +
-
-  scale_x_level("Level (hPa)", minor_breaks = NULL, 
-                sec.axis = sa_height_axis(breaks = seq(0, 50, by = 10),
-                                          labels = function(x) paste0("~", x, " km"))) +
-  scale_y_continuous(r2, limits = c(0, 1), 
-                     guide = guide_axis(check.overlap = TRUE)) +
-  scale_size_manual("p-value", values = c("TRUE" = 1, 
-                                          "FALSE" = 0.5), 
-                    guide = "none") +
-  scale_color_brewer(NULL, palette = "Dark2", labels = c(full = "SAM",
-                                                         asym = "A-SAM",
-                                                         sym = "S-SAM")) +
-  facet_grid(part ~ cEOF) +
-  coord_flip() +
-  tag_facets("rc") +
-  panel_background +
-  theme(panel.spacing = grid::unit(1, "lines"), 
-        strip.placement = "outside")
-```
-
-To explore the relationship between SAM and the modes described from the cEOF, we compute the coefficient of determination between the cEOFs timeseries and the three SAM indices (SAM, A-SAM and S-SAM) at each vertical level (Fig. \@ref(fig:sam-eof-vertical)).
-The SAM index is statistically signficant correlated with the Real cEOF1 in all levels, and with the Imaginary cEOF1 and Imaginary cEOF2 in the troposphere.
-Correlations between SAM and the Real cEOF2 are non-significant.
-
-The relationship between the tropospheric SAM and cEOF1 is explained entirely by the zonally symmetric component of the SAM as shown by the low and statistically non-significant correlations between the A-SAM and either the Real or Imaginary cEOF1 and the high correlation with the S-SAM below 100 hPa.
-In the stratosphere, the Real cEOF1 is correlated with both A-SAM and S-SAM, while the Imaginary cEOF1 is highly correlated only with the A-SAM.
-These correlations are consistent with the regression patterns of geopotential height in Figure \@ref(fig:eof1-regr-gh) and their comparison with those obtained for SAM, A-SAM and S-SAM by \@@campitelli2022.
-
-```{r eof-filtered, eval=FALSE, fig.cap="Spatial pattern of the leading EOF of 200 hPa geopotential height with the variability of cEOF2 filtered out South of 20º.
-Arbitrary units.", fig.height=2.5, fig.width=2.5, include=FALSE}
-cEOF2 <- ceof$eof[[1]]$left[cEOF == "cEOF2"] %>% 
-  copy() %>% 
-  .[, cEOF := NULL] %>% 
-  .[]
-
-eof_filtered <- era5 %>% 
-  .[sep_ReIm(cEOF2, format = "wide")[, hgt := NULL], on = .NATURAL] %>% 
-  .[lev == 200] %>% 
-  .[lat <= -20] %>% 
-  .[season(time) == "SON"] %>% 
-  .[, hgt := resid(.lm.fit(cbind(1, Real, Imaginary), hgt)), .(lon, lat, lev, season(time))] %>%
-  .[, eof := Anomaly(hgt)*sqrt(cos(lat*pi/180)), by = .(lon, lat, lev, season(time))] %>% 
-  EOF(eof ~ lon + lat | time, n = 1, data = .)
-
-eof_filtered_cor <- eof_filtered$right[sams, on = "time"] %>% 
-  .[lev == 200] %>% 
-  .[, correlate(estimate, eof), by = term]
-
-eof_filtered %>%
-  .$left %>%
-  periodic(lon = c(0, 360)) %>%
-  ggplot(aes(lon, lat)) +
-  geom_contour_fill(aes(z = eof), breaks = AnchorBreaks(0, exclude = 0)) +
-  geom_contour_tanaka(aes(z = eof), breaks = AnchorBreaks(0, exclude = 0)) +
-  scale_fill_divergent(guide = "none") +
-  geom_qmap() +
-  scale_x_longitude() +
-  scale_y_latitude() +
-  geom_coords() +
-  coord_polar() +
-
-  theme(axis.text = element_blank())
-
-eof_filtered_cor %>% 
-  .[, .(term, text)] %>%
-  .[, term := factor(term, labels = c("SAM", "A-SAM", "S-SAM"))] %>%
-  knitr::kable(col.names = c("Índice SAM", "Correlación"))
-```
-
-In the case of cEOF2, the moderate correlation between SAM and Imaginary cEOF2 becomes extremely high only when the zonally asymmetric variability of the SAM is considered.
-The Imaginary cEOF2 explains up to `r scales::percent(max_r2[cEOF == "cEOF2"]$estimate^2, 2)` of A-SAM variance, reached the maximum value at `r max_r2[cEOF == "cEOF2"]$lev` hPa (Figure \@ref(fig:sam-eof-vertical).b2).
-Such extremely high correlation between the asymmetric SAM and the Imaginary cEOF2 suggests that these might be different ways of characterising the same phenomenon.
-
 ### PSA
 
-(ref:psa-eof2-cap) Correlations between the Real and Imaginary parts of cEOF2 and the PSA1 and PSA2 modes computed as the second and third EOFs of seasonal geopotential height anomalies [following @mo2001] for the 1979 -- 2019 period.
+(ref:psa-eof2-cap) Correlation coefficients (r) between cEOF2 components and the PSA1 and PSA2 modes computed as @mo2001 for the 1979 -- 2019 period.
 95% confidence intervals in parenthesis.
@@
   kable_classic() 
-
-# cEOF2[psa_time, on = "time"] %>% 
-#   .[PC != "SAM"] %>% 
-#   copy() %>% 
-#   # .[, hgt := rotate(hgt, psa_angle)] %>%
-#   sep_ReIm() %>% 
-#   ggplot(aes(value, hgt)) +
-#   geom_hline(yintercept = 0, size = 0.2, color = "gray50") +
-#   geom_vline(xintercept = 0, size = 0.2, color = "gray50") +
-#   geom_point() +
-#   geom_smooth(method = "lm") +
-#   geom_smooth(method = "lm") + 
-#   scale_y_continuous("EOF2") +
-#   scale_x_continuous("PSA") +
-#   facet_grid(part ~ PC) +
-#   panel_background +
-#   tag_facets("rc")
-```
-
-Due to the similarity between the PSA patterns and the cEOF2 spatial patterns (Fig. \@ref(fig:eof2-regr-gh)), we study the relationship between these modes.
-Table \@ref(tab:psa-eof2) shows the correlations between the two PSA indices and the timeseries for Real and Imaginary phase of cEOF2.
-As anticipated by Figure \@ref(fig:eof2-regr-gh), there is a strong correlation between PSA1 and Imaginary cEOF2, and between PSA2 and Real cEOF2.
-Conversely, there is no relationship between PSA1 and Real cEOF2, and between PSA2 and Imaginary cEOF2.
-As a result, cEOF2 represents well both the spatial structure and temporal evolution of the PSA modes, as well as it is possible to make a rather clean association between its two phases and the two PSA modes.
-So, It could be concluded that the same particular rotation of cEOF2 that maximises the association between cEOF2 parts and PSA modes, is the one which maximises the relationship between ENSO and Imaginary cEOF2.
-
-The reason the conventional EOF analysis arrives at the same separation than our particular cEOF rotation is probably the fact that not all phases are equally frequent.
-To visualize that, Figure \@ref(fig:phase-histogram) shows an histogram that counts the number of SON years in which the cEOF2 was close to each of the four particular phases (positive/negative Real/Imaginary), with the observations for each season marked as rugs on the horizontal axis.
-For instance, years with cEOF2 phase within 45º of 0º are nearest the "positive Real" phase.
-About two thirds of time cEOF2 has a phase similar to either the negative or positive Imaginary phase.
-It could be seen that Imaginary phase is the most common phase and It is also, by construction, the direction that has the maximum relationship with ENSO.
-Therefore, the Imaginary cEOF2 explains more variance than the Real cEOF2 and conventional EOF analysis will tend to separate the two.
-
-(ref:phase-histogram-cap) Histogram of phase distribution of cEOF2 for the 1979 -- 2019 period.
+```
+
+Given the similarity between the cEOF2 regression maps (Fig. \@ref(fig:eof2-regr-gh)) and documented PSAs we study the relationship between these modes.
+Table \@ref(tab:psa-eof2) shows the correlations between the two PSA indices and the time series for Real and Imaginary phases of cEOF2.
+As anticipated by Figure \@ref(fig:eof2-regr-gh), there is a strong positive correlation between PSA1 and Imaginary cEOF2, and also between PSA2 and Real cEOF2.
+On the other hand, there is no significant relationship between PSA1 and Real cEOF2, and between PSA2 and Imaginary cEOF2.
+As a result, cEOF2 represents well both the spatial structure and temporal evolution of the PSA modes, so it is possible to make an association between its two phases and the two PSA modes.
+The same phase election for cEOF2 that maximises the relationship between ENSO and Imaginary cEOF2, also maximises the association between cEOF2 components and PSA modes.
+
+(ref:phase-histogram-cap) Histogram of phase distribution of cEOF2 phase for the 1979 -- 2019 period.
 Bins are centred at 90º, 0º, -90º, -180º with a binwidth of 90º.
@@
 N <- nrow(cut(ceof$eof[[1]], 2)$left)
-cut(ceof$eof[[1]], 2)$left %>% 
+plot <- cut(ceof$eof[[1]], 2)$left %>% 
   .[, arg := Arg(hgt)] %>% 
@@
   no_grid
-```
-
-The result obtained here is similar to the one obtained by @irving2016.
-They show that the phase distribution of the PSA-like variability obtained by them is bimodal and isolate the PSA pattern from the rest of the PSA-like variability by selecting events that are near the peaks of the distribution (compare our Figure \@ref(fig:phase-histogram) with their Figure 6).
-
-The advantage of our method is that it is much simpler to implement, it provides magnitude and phase naturally, and it facilitates the description of this mode as a propagating wave instead of as standing oscillations.
-As a consequence, the cEOF2 offers an alternative way of representing the PSA which has several advantages over using the second and third EOFs.
+
+imaginary <- sum(ggplot_build(plot)$data[[1]][c(1, 3), ]$count)/N
+plot
+```
+
+Figure \@ref(fig:phase-histogram) shows an histogram that counts the number of SON seasons in which the cEOF2 phase was close to each of the four particular phases (positive/negative of Real/Imaginary component), with the observations for each season marked as rugs on the horizontal axis.
+In `r scales::percent(imaginary)` of seasons cEOF2 has a phase similar to either the negative or positive Imaginary phase.
+It could be seen that Imaginary phase is the most common phase and It is also, by construction, the direction that has the maximum relationship with ENSO.
+Therefore, the Imaginary cEOF2 explains more variance than the Real cEOF2 and conventional EOF analysis will tend to separate the two.
+
+This phase preferences is sin agreement with @irving2016, who found a bimodal distribution to PSA-like variability (compare our Figure \@ref(fig:phase-histogram) with their Figure 6).
+
+### SAM
+
+(ref:sam-eof-vertical-cap) Coefficient of determination between the real and imaginary part of each cEOF and the SAM, Asymmetric SAM (A-SAM) and Symmetric SAM (S-SAM) indices computed at each level according to \@@campitelli2022 for the 1979 -- 2019 period.
+Thick lines represent estimates with p-value \< 0.01 corrected for False Detection Rate [@benjamini1995].
+
+```{r sam-eof-vertical, fig.cap = "(ref:sam-eof-vertical-cap)", fig.height=4, fig.width=width_column}
+time <- ceof[, eof[[1]]$left, by = .(season)] %>%
+  sep_ReIm() %>%
+  .[, hgt := scale(hgt), by = .(cEOF, part)]
+
+sam_r2 <- sams %>%
+  .[time, on = .NATURAL, allow.cartesian = TRUE] %>%
+  na.omit() %>% 
+  .[, correlate(hgt, estimate), by = .(lev, term, part, cEOF)]
+
+max_r2 <- sam_r2[part == "Imaginary" & term == "asym"] %>% 
+  .[, .SD[which.max(estimate^2)], by = .(cEOF)]
+
+sam_r2 %>%
+  .[, p.value_a := p.adjust(p.value, "fdr")] %>% 
+  ggplot(aes(lev, estimate^2)) +
+  geom_vline(xintercept = c(50, 200),  size = 0.2, color = "gray50") +
+  # geom_point(data = ~.x[p.adjust(p.value, "fdr") < 0.01], aes(color = term)) +
+  geom_line(data = ~copy(.x)[p.value_a > 0.01, estimate := NA], 
+            aes(colour = stage(term, after_scale = scales::alpha(colour, 0.5))), 
+            size = 2) +
+  geom_line(aes(color = term, group = term)) +
+
+  scale_x_level("Level (hPa)", minor_breaks = NULL, 
+                sec.axis = sa_height_axis(breaks = seq(0, 50, by = 10),
+                                          labels = function(x) paste0("~", x, " km"))) +
+  scale_y_continuous(r2, limits = c(0, 1), 
+                     guide = guide_axis(check.overlap = TRUE)) +
+  scale_size_manual("p-value", values = c("TRUE" = 1, 
+                                          "FALSE" = 0.5), 
+                    guide = "none") +
+  scale_color_brewer(NULL, palette = "Dark2", labels = c(full = "SAM",
+                                                         asym = "A-SAM",
+                                                         sym = "S-SAM")) +
+  facet_grid(part ~ cEOF) +
+  coord_flip() +
+  tag_facets("rc") +
+  panel_background +
+  theme(panel.spacing = grid::unit(1, "lines"), 
+        strip.placement = "outside")
+```
+
+To explore the relationship between SAM and the modes described from the cEOF, we compute the coefficient of determination between the cEOFs timeseries and the three SAM indices (SAM, A-SAM and S-SAM) at each vertical level (Fig. \@ref(fig:sam-eof-vertical)).
+The SAM index is statistically signficant correlated with the Real cEOF1 in all levels, and with the Imaginary cEOF1 and Imaginary cEOF2 in the troposphere.
+Correlations between SAM and the Real cEOF2 are non-significant.
+
+The relationship between the tropospheric SAM and cEOF1 is explained entirely by the zonally symmetric component of the SAM as shown by the low and statistically non-significant correlations between the A-SAM and either the Real or Imaginary cEOF1 and the high correlation with the S-SAM below 100 hPa.
+In the stratosphere, the Real cEOF1 is correlated with both A-SAM and S-SAM, while the Imaginary cEOF1 is highly correlated only with the A-SAM.
+These correlations are consistent with the regression patterns of geopotential height in Figure \@ref(fig:eof1-regr-gh) and their comparison with those obtained for SAM, A-SAM and S-SAM by \@@campitelli2022.
+
+```{r eof-filtered, eval=FALSE, fig.cap="Spatial pattern of the leading EOF of 200 hPa geopotential height with the variability of cEOF2 filtered out South of 20º.
+Arbitrary units.", fig.height=2.5, fig.width=2.5, include=FALSE}
+cEOF2 <- ceof$eof[[1]]$left[cEOF == "cEOF2"] %>% 
+  copy() %>% 
+  .[, cEOF := NULL] %>% 
+  .[]
+
+eof_filtered <- era5 %>% 
+  .[sep_ReIm(cEOF2, format = "wide")[, hgt := NULL], on = .NATURAL] %>% 
+  .[lev == 200] %>% 
+  .[lat <= -20] %>% 
+  .[season(time) == "SON"] %>% 
+  .[, hgt := resid(.lm.fit(cbind(1, Real, Imaginary), hgt)), .(lon, lat, lev, season(time))] %>%
+  .[, eof := Anomaly(hgt)*sqrt(cos(lat*pi/180)), by = .(lon, lat, lev, season(time))] %>% 
+  EOF(eof ~ lon + lat | time, n = 1, data = .)
+
+eof_filtered_cor <- eof_filtered$right[sams, on = "time"] %>% 
+  .[lev == 200] %>% 
+  .[, correlate(estimate, eof), by = term]
+
+eof_filtered %>%
+  .$left %>%
+  periodic(lon = c(0, 360)) %>%
+  ggplot(aes(lon, lat)) +
+  geom_contour_fill(aes(z = eof), breaks = AnchorBreaks(0, exclude = 0)) +
+  geom_contour_tanaka(aes(z = eof), breaks = AnchorBreaks(0, exclude = 0)) +
+  scale_fill_divergent(guide = "none") +
+  geom_qmap() +
+  scale_x_longitude() +
+  scale_y_latitude() +
+  geom_coords() +
+  coord_polar() +
+
+  theme(axis.text = element_blank())
+
+eof_filtered_cor %>% 
+  .[, .(term, text)] %>%
+  .[, term := factor(term, labels = c("SAM", "A-SAM", "S-SAM"))] %>%
+  knitr::kable(col.names = c("Índice SAM", "Correlación"))
+```
+
+In the case of cEOF2, the moderate correlation between SAM and Imaginary cEOF2 becomes extremely high only when the zonally asymmetric variability of the SAM is considered.
+The Imaginary cEOF2 explains up to `r scales::percent(max_r2[cEOF == "cEOF2"]$estimate^2, 2)` of A-SAM variance, reached the maximum value at `r max_r2[cEOF == "cEOF2"]$lev` hPa (Figure \@ref(fig:sam-eof-vertical).b2).
+Such extremely high correlation between the asymmetric SAM and the Imaginary cEOF2 suggests that these might be different ways of characterising the same phenomenon.
 

### 2022-03-25 analysis/paper/shceof.Rmd
Revisión

@@
 
-Figure \@ref(fig:phase-histogram) shows an histogram that counts the number of SON seasons in which the cEOF2 phase was close to each of the four particular phases (positive/negative of Real/Imaginary component), with the observations for each season marked as rugs on the horizontal axis.
-In `r scales::percent(imaginary)` of seasons cEOF2 has a phase similar to either the negative or positive Imaginary phase.
-It could be seen that Imaginary phase is the most common phase and It is also, by construction, the direction that has the maximum relationship with ENSO.
+Figure \@ref(fig:phase-histogram) shows an histogram that counts the number of SON seasons in which the cEOF2 phase was close to each of the four particular phases (positive/negative of real/imaginary component), with the observations for each season marked as rugs on the horizontal axis.
+In `r scales::percent(imaginary)` of seasons cEOF2 has a phase similar to either the negative or positive imaginary phase, making the imaginary phase the most common phase. 
+This is also, by construction, the phase that is most correlated with ENSO.
 Therefore, the Imaginary cEOF2 explains more variance than the Real cEOF2 and conventional EOF analysis will tend to separate the two.
 
-This phase preferences is sin agreement with @irving2016, who found a bimodal distribution to PSA-like variability (compare our Figure \@ref(fig:phase-histogram) with their Figure 6).
+This phase preferences is in agreement with @irving2016, who found a bimodal distribution to PSA-like variability (compare our Figure \@ref(fig:phase-histogram) with their Figure 6).
 
@@
 
-(ref:sam-eof-vertical-cap) Coefficient of determination between the real and imaginary part of each cEOF and the SAM, Asymmetric SAM (A-SAM) and Symmetric SAM (S-SAM) indices computed at each level according to \@@campitelli2022 for the 1979 -- 2019 period.
-Thick lines represent estimates with p-value \< 0.01 corrected for False Detection Rate [@benjamini1995].
+(ref:sam-eof-vertical-cap) Coefficient of determination ($r^2$) between each component of cEOFs and the SAM, Asymmetric SAM (A-SAM) and Symmetric SAM (S-SAM) indices computed at each level for the 1979 -- 2019 period.
+Thick lines represent estimates with p-value < 0.01 corrected for False Detection Rate [@benjamini1995].
 
@@
 
-max_r2 <- sam_r2[part == "Imaginary" & term == "asym"] %>% 
-  .[, .SD[which.max(estimate^2)], by = .(cEOF)]
+max_r2 <- sam_r2[part == "Imaginary"] %>% 
+  .[, .SD[which.max(estimate^2)], by = .(cEOF, term)]
 
@@
 
-To explore the relationship between SAM and the modes described from the cEOF, we compute the coefficient of determination between the cEOFs timeseries and the three SAM indices (SAM, A-SAM and S-SAM) at each vertical level (Fig. \@ref(fig:sam-eof-vertical)).
-The SAM index is statistically signficant correlated with the Real cEOF1 in all levels, and with the Imaginary cEOF1 and Imaginary cEOF2 in the troposphere.
-Correlations between SAM and the Real cEOF2 are non-significant.
-
-The relationship between the tropospheric SAM and cEOF1 is explained entirely by the zonally symmetric component of the SAM as shown by the low and statistically non-significant correlations between the A-SAM and either the Real or Imaginary cEOF1 and the high correlation with the S-SAM below 100 hPa.
+We now explore the relationship between SAM and the cEOFs motivated by the resemblance between cEOFs regression maps and SAM patterns shown in the previous section.
+To that end, we computed the coefficient of determination between the cEOFs time series and the three SAM indices (SAM, A-SAM and S-SAM) defined by @campitelli2022 at each vertical level (Fig.\ \@ref(fig:sam-eof-vertical)).
+The SAM index is statistically significant correlated with the Real cEOF1 in all levels, and with the Imaginary cEOF1 and Imaginary cEOF2 in the troposphere.
+On the other hand, correlations between SAM and the Real cEOF2 are non-significant.
+
+The relationship between the SAM and cEOF1 in the troposphere is explained entirely by the zonally symmetric component of the SAM as shown by the high correlation with the S-SAM below 100 hPa and the low and statistically non-significant correlations between the A-SAM and either the Real or Imaginary cEOF1.
 In the stratosphere, the Real cEOF1 is correlated with both A-SAM and S-SAM, while the Imaginary cEOF1 is highly correlated only with the A-SAM. 
-These correlations are consistent with the regression patterns of geopotential height in Figure \@ref(fig:eof1-regr-gh) and their comparison with those obtained for SAM, A-SAM and S-SAM by \@@campitelli2022.
-
-```{r eof-filtered, eval=FALSE, fig.cap="Spatial pattern of the leading EOF of 200 hPa geopotential height with the variability of cEOF2 filtered out South of 20º.
-Arbitrary units.", fig.height=2.5, fig.width=2.5, include=FALSE}
-cEOF2 <- ceof$eof[[1]]$left[cEOF == "cEOF2"] %>% 
-  copy() %>% 
-  .[, cEOF := NULL] %>% 
-  .[]
-
-eof_filtered <- era5 %>% 
-  .[sep_ReIm(cEOF2, format = "wide")[, hgt := NULL], on = .NATURAL] %>% 
-  .[lev == 200] %>% 
-  .[lat <= -20] %>% 
-  .[season(time) == "SON"] %>% 
-  .[, hgt := resid(.lm.fit(cbind(1, Real, Imaginary), hgt)), .(lon, lat, lev, season(time))] %>%
-  .[, eof := Anomaly(hgt)*sqrt(cos(lat*pi/180)), by = .(lon, lat, lev, season(time))] %>% 
-  EOF(eof ~ lon + lat | time, n = 1, data = .)
-
-eof_filtered_cor <- eof_filtered$right[sams, on = "time"] %>% 
-  .[lev == 200] %>% 
-  .[, correlate(estimate, eof), by = term]
-
-eof_filtered %>%
-  .$left %>%
-  periodic(lon = c(0, 360)) %>%
-  ggplot(aes(lon, lat)) +
-  geom_contour_fill(aes(z = eof), breaks = AnchorBreaks(0, exclude = 0)) +
-  geom_contour_tanaka(aes(z = eof), breaks = AnchorBreaks(0, exclude = 0)) +
-  scale_fill_divergent(guide = "none") +
-  geom_qmap() +
-  scale_x_longitude() +
-  scale_y_latitude() +
-  geom_coords() +
-  coord_polar() +
-
-  theme(axis.text = element_blank())
-
-eof_filtered_cor %>% 
-  .[, .(term, text)] %>%
-  .[, term := factor(term, labels = c("SAM", "A-SAM", "S-SAM"))] %>%
-  knitr::kable(col.names = c("Índice SAM", "Correlación"))
-```
-
-In the case of cEOF2, the moderate correlation between SAM and Imaginary cEOF2 becomes extremely high only when the zonally asymmetric variability of the SAM is considered.
-The Imaginary cEOF2 explains up to `r scales::percent(max_r2[cEOF == "cEOF2"]$estimate^2, 2)` of A-SAM variance, reached the maximum value at `r max_r2[cEOF == "cEOF2"]$lev` hPa (Figure \@ref(fig:sam-eof-vertical).b2).
-Such extremely high correlation between the asymmetric SAM and the Imaginary cEOF2 suggests that these might be different ways of characterising the same phenomenon.
+These correlations are consistent with the regression maps of geopotential height in Figure\ \@ref(fig:eof1-regr-gh) and their comparison with those obtained for SAM, A-SAM and S-SAM by @campitelli2022.
+
+In the case of Imaginary cEOF2, the correlation between SAM and Imaginary cEOF2 for the troposphere is associated to the asymmetric variability of the SAM. 
+Indeed, the Imaginary cEOF2 shares up to `r scales::percent(max_r2[cEOF == "cEOF2" & term == "asym"]$estimate^2, 2)` variance with the A-SAM and only `r scales::percent(max_r2[cEOF == "cEOF2" & term == "sym"]$estimate^2, 2)` at most with the S-SAM (Figure \@ref(fig:sam-eof-vertical).b2).
+Such extremely high correlation between A-SAM and Imaginary cEOF2 suggests that the methodology proposed in this work is able to characterise the zonally asymmetric component of the SAM described previously by @campitelli2022.
 
@@
 
-(ref:t-vertical-cap) Regression of SON anomalies of mean air temperature (shaded, Kelvin) and ozone mixing ratio (contours, negative contours with dashed lines, labels in parts per billion by mass) averaged between `r knitr::combine_words(LatLabel(o3wave_lats))` with the (a) real and (b) imaginary parts of the first cEOF for the 1979 -- 2019 period.
+(ref:t-vertical-cap) Regression of SON zonal anomalies  averaged between `r knitr::combine_words(LatLabel(o3wave_lats))` of mean air temperature (shaded, Kelvin) and ozone mixing ratio (contours, negative contours with dashed lines, labels in parts per billion by mass) with the (a) Real and (b) Imaginary components of the cEOF1 for the 1979 -- 2019 period.
 
@@
 
-The relation between cEOFs and air temperature was also evaluated.
-Figure \@ref(fig:eof1-regr-t) shows regression patterns of air temperature at 50hPa and 200hPa onto cEOF1.
+The influence of cEOFs variability on air temperature was also evaluated.
+Figure\ \@ref(fig:eof1-regr-t) shows regression maps of air temperature anomalies at 50 hPa and 200 hPa upon cEOF1. 
 In both levels, the Real cEOF1 is associated with a positive monopole over the South Pole with its centre moved slightly towards 150ºE (Fig. \@ref(fig:eof1-regr-t) column 1).
@@
 
-Figure \@ref(fig:t-vertical) shows the vertical distribution of the regression coefficients on cEOF1 from zonal anomalies of air temperature and zonal anomalies of ozone mixing ratio averaged between `r knitr::combine_words(LatLabel(o3wave_lats))`.
-Temperature zonal anomalies associated with cEOF1 show a clear wave 1 pattern for both real and imaginary components throughout the atmosphere above 250 hPa with a change in sign above 10 hPa.
-Following hydrostatic balance, this is the level in which the geopotential anomaly have maximum amplitude (not shown).
-
-The maximum ozone anomalies are co-located with the minimum temperature anomalies above 10 hPa and with the maximum temperature anomalies below 10 hPa (Fig. \@ref(fig:t-vertical)).
+Figure\ \@ref(fig:t-vertical) shows the vertical distribution of the regression coefficients on cEOF1 from zonal anomalies averaged between `r knitr::combine_words(LatLabel(o3wave_lats))` of air temperature and of ozone mixing ratio.
+Temperature zonal anomalies associated with cEOF1 show a clear wave 1 pattern for both real and imaginary components throughout the atmosphere above 250 hPa with a sign reversal above 10 hPa.
+As a result of the hydrostatic balance, this is the level in which the geopotential anomaly have maximum amplitude (not shown).
+
+The maximum ozone regressed anomalies match with the minimum temperature anomalies above 10 hPa and with the maximum temperature anomalies below 10 hPa (Fig. \@ref(fig:t-vertical)).
 Therefore, the ozone zonal wave 1 is anticorrelated with the temperature zonal wave 1 in the upper stratosphere, and directly correlated in the upper stratosphere. 
-This change in phase is observed in ozone anomalies forced by planetary waves that reach the stratosphere.
-In the photochemically-dominated upper stratosphere, cold temperatures inhibit the destruction of ozone.
-On the other hand, in the advectively-dominated lower stratosphere, ozone anomalies are 90º out of phase with horizontal and vertical transport, which are in adittion 90º out of phase with temperature anomalies [@hartmann1979; @wirth1993; @smith1995].
-
-```{r o3-t-phase}
-# t %>% 
-#   .[lat %between% c(-80, -35)] %>% 
-#   .[, .(o3 = weighted.mean(o3, cos(lat*pi/180)),
-#         t = weighted.mean(t, cos(lat*pi/180))), by = .(lev, time, lon)] %>%  
-#   melt(id.vars = c("lev", "time", "lon")) %>% 
-#   # .[, value := Anomaly(value), by = .(time, lev, variable)] %>%
-#   .[cut(ceof$eof[[1]], 1)$left, on = "time"] %>% 
-#   sep_ReIm(format = "wider") %>% 
-#   .[, ":="(Real = as.numeric(scale(Real)), Imaginary = as.numeric(scale(Imaginary))), by = .(cEOF)] %>% 
-#   .[, FitLm(value, Real, Imaginary, se = TRUE), by = .(lev, lon, variable)] %>% 
-#   rm_intercept() %>% 
-#   .[, term := factor_ReIm(term)] %>% 
-#   .[, FitWave(estimate, 1), by = .(lev, variable, term)] %>% 
-#   dcast(lev + term ~ variable, value.var = c("phase", "amplitude")) %>%
-#   ggplot(aes(lev, acos(cos(phase_t - phase_o3)))) +
-#   geom_line(aes(color = term)) +
-#   scale_x_level() +
-#   scale_colour_brewer(NULL, palette = "Dark2") +
-#   scale_y_continuous("Wave-1 phase difference\nbetween Temperature and Ozone",
-#                      breaks = seq(-pi, pi, by = 30*pi/180),
-#                      labels = function(x) paste0(round(x*180/pi, 5), "º"))  +
-#   coord_flip()
-```
+This change in phase is observed in ozone anomalies forced by planetary waves that reach the stratosphere [@hartmann1979; @wirth1993; @smith1995]. 
+In the photochemically-dominated upper stratosphere, cold temperatures inhibit the destruction of ozone, explaining the opposite behaviour for both variables.
+On the other hand, in the advectively-dominated lower stratosphere, ozone anomalies are 90º out of phase with horizontal and vertical transport, which are in addition 90º out of phase with temperature anomalies, resulting in same sign anomalies for the response of both variables
 
@@
 
-(ref:o3-regr-cap) Regression of SON mean Total Ozone Column anomalies (shaded, Dobson Units) with the (a) real and (b) imaginary parts of the cEOF1 for the 1979 -- 2019 period.
+(ref:o3-regr-cap) Regression of SON mean Total Ozone Column anomalies (shaded, Dobson Units) with the (a) real and (b) imaginary components of the cEOF1 for the 1979 -- 2019 period.
 On contours, the mean zonal anomaly of Total Ozone Column (negative contours in dashed lines, Dobson Units).
@@
 
-The regression maps of cEOF1 with fields of Total Ozone Column (TOC) (Fig.  \@ref(fig:o3-regr)) show zonal wave 1 pattens in TOC associated with both phases of cEOF1.
-Climatologically, the springtime Ozone minimum is located off the South Pole and towards the Weddell Sea [@grytsai2011].
-Thus, the Real cEOF1 regression pattern (Figure \@ref(fig:o3-regr)a) coincides with the climatological position of the ozone hole while the one for the Imaginary cEOF1 is shifted by 90º.
-
-(ref:wave1-o3-cap) Relationship between cEOF1 amplitude and phase with amplitude and phase of the zonal wave 1 in Total Ozone Column averaged between `r knitr::combine_words(LatLabel(o3wave_lats))` for the period 1979 -- 2019.
-
 ```{r wave1-o3, fig.cap = "(ref:wave1-o3-cap)", fig.width=width_column, fig.height=width_column*.7, include = FALSE}
@@
 
-# temp %>% 
-#   copy() %>% 
-#   na.omit() %>% 
-#   sep_ReIm() %>% 
-#   .[, hgt := scale(hgt), by  = .(part)] %>% 
-#   ggplot(aes(amplitude, hgt)) +
-#   geom_point() +
-#   geom_smooth(method = "lm", color = "black") +
-#   facet_grid(cEOF ~ part) +
-#   panel_background
 wave1_cor <- rbind(temp %>%
@@
 
-There is a close spatial relationship between amplitudes and phases of the cEOF1 and TOC planetary wave 1 between `r knitr::combine_words(LatLabel(o3wave_lats))`.
-The correlation between the amplitude of both indices is `r wave1_cor[with == "amplitude"]$text` and `r wave1_cor[with == "phase"]$text` between their phases.
+The regression maps of  TOC anomalies upon cEOF1 (Fig.\ \@ref(fig:o3-regr)) show zonal wave 1 patterns  associated with both components of cEOF1.
+The climatological position of the springtime Ozone minimum (ozone hole) is  outside the South Pole and towards the Weddell Sea [e.g. @grytsai2011].
+Thus, the Real cEOF1 regression field (Figure\ \@ref(fig:o3-regr)a) coincides with the climatological position of the ozone hole, while it is 90° out of phase for the Imaginary cEOF1.
+The temporal correlation between the amplitudes of TOC planetary wave 1 and cEOF1 is `r wave1_cor[with == "amplitude"]$text`, while the correlation between their phases is `r wave1_cor[with == "phase"]$text`.
+Consequently, cEOF1 is strongly related with the SH ozone variability.
+
+(ref:wave1-o3-cap) Relationship between cEOF1 amplitude and phase with amplitude and phase of the zonal wave 1 in Total Ozone Column averaged between `r knitr::combine_words(LatLabel(o3wave_lats))` for the period 1979 -- 2019.
 
@@
 
-(ref:sst-psi-2-cap) Regression maps of cEOF2 with SST (K, column 1) and streamfunction zonal anomalies ($m^2/s\times10^-7$, shaded) with their corresponding activity wave flux (vectors) (column 2) for the 1979 -- 2019 period and different phases of the cEOF (illustrated in the lower-left arrow).
+(ref:sst-psi-2-cap) Regression of SST (K, left column ) and streamfunction zonal anomalies ($m^2/s\times10^-7$, shaded) with their corresponding activity wave flux (vectors) (right column) upon cEOF2 different phases (illustrated in the lower-left arrow) for the 1979 -- 2019 period.
 Areas marked with dots have p-values smaller than 0.01 adjusted for FDR.
 
-The connections between cEOFs and Tropical sources are also assessed.
-Figure \@ref(fig:sst-psi-2) shows the regression maps of Sea Surface Temperatures (SST) and streamfunction anomalies at 200 hPa respectively upon standardised cEOF2, which can be interpreted as the anomalies associated with a 1 standard deviation increase of each index.
-Besides showing the regression of the real and imaginary parts, we also show regressions in two intermediate directions.
-The Imaginary cEOF2 is associated with strong positive SST anomalies on the Central Pacific and negative anomalies over an area across the North of Australia and New Zealand, the South Pacific Convergence Zone (SPCZ) (Figure \@ref(fig:sst-psi-2).b1).
-This pattern is almost canonically positive ENSO [@bamston1997] and indeed, the correlation between the Imaginary cEOF2 and the Oceanic Niño Index [@bamston1997] is significant and very high `r enso_cor[cEOF == "cEOF2" & part == "Imaginary"]$text`.
-On top of this ENSO-like pattern, there are positive anomalies in the western Indian Ocean and negative values in the eastern Indian Ocean, resembling a positive Indian Ocean Dipole.
+The connections between cEOFs and Tropical sources are also assessed, to explore plausible forcings for the described cEOFs patterns.
+Figure \@ref(fig:sst-psi-2) shows the regression maps of Sea Surface Temperatures (SST) and streamfunction anomalies at 200 hPa respectively upon standardised cEOF2,.
+Besides showing the regression maps for the Real and Imaginary components, we including the corresponding regressions for two intermediate directions (corresponding to 45° and 135°).
+
+The Imaginary cEOF2 (second row) is associated with strong positive SST anomalies on the Central Pacific and negative anomalies over an area across northern Australia,  New Zealand the South Pacific Convergence Zone (SPCZ) (Figure \@ref(fig:sst-psi-2).b1).
+The regression field of SST anomalies bears a strong resemblance with canonically positive ENSO [@bamston1997].
+Indeed, there is a significant and very high correlation (`r enso_cor[cEOF == "cEOF2" & part == "Imaginary"]$text`) between the ONI and the Imaginary cEOF2 time series.
+Besides the Pacific ENSO-like pattern, there are positive anomalies in the western Indian Ocean and negative values in the eastern Indian Ocean, resembling a positive Indian Ocean Dipole [@saji1999].
 Unsurprisingly, the correlation between the Imaginary cEOF2 and the DMI is `r dmi_cor[cEOF == "cEOF2" & part == "Imaginary"]$text`.
 
-Fittingly, the Imaginary cEOF2 is also associated with strong wave-like streamfunction anomalies emanating from the tropics (Figure \@ref(fig:sst-psi-2).b2), both from the Central Pacific sector and the Incean Ocean.
-These are consistent with the effect of ENSO and the Indian Ocean Dipole on the extratropics: the SST anomalies initiate anomalous tropical convection that excites Rossby waves that propagate meridionally towards higher latitudes [@mo2000; @cai2011; @nuncio2015].
-
-Not all phases of the cEOF2 are associated with the same tropical SSTs, however.
-Figure \@ref(fig:sst-psi-2).d1 and d2 show that the Real cEOF2 is not associated either with any significant SST nor streamfunction anomalies in the tropics.
-The correlation between the Real cEOF2 and ENSO is also not significant (`r enso_cor[cEOF == "cEOF2" & part == "Real"]$text`).
-Rows a and c in Fig. 
-\@ref(fig:sst-psi-2) show that the intermediate phases are still associated with significant SST anomalies over the Pacific Ocean, but with with slightly different location.
-The 135º phase is associated with SST in the central Pacific, while the 45º phase is associated with SST in the eastern Pacific, which correspond roughly to the Eastern Pacific and Central Pacific "flavours" of ENSO (Citas).
-
-(ref:enso-phase-cap) ONI plotted against cEOF2 phase for th 1979 -- 2019 period.
-Years with magnitude of cEOF2 greater or smaller than the 50th percentile are shown as orange diamonds and green circles and respectively.
+The Imaginary cEOF2 is associated with strong wave-like streamfunction anomalies emanating from the tropics (Figure\ \@ref(fig:sst-psi-2).b2), both from the Central Pacific sector and the Indian Ocean.
+These atmospheric response is consistent with the effect of ENSO and the Indian Ocean Dipole on the extratropics: the SST anomalies initiate anomalous tropical convection that excites Rossby waves that propagate meridionally towards higher latitudes  [@mo2000; @cai2011; @nuncio2015].
+
+The cEOF2 is not associated with the same tropical SSTs for all their phases. 
+Figure\ \@ref(fig:sst-psi-2).d1 and d2 show that the Real cEOF2 is not associated either with any significant SST nor streamfunction anomalies in the tropics. 
+As a result, the correlation between the Real cEOF2 and ENSO is also not significant  (`r enso_cor[cEOF == "cEOF2" & part == "Real"]$text`).
+Meanwhile, Rows a and c in Fig.\ \@ref(fig:sst-psi-2) show that the intermediate phases are still associated with significant SST anomalies over the Pacific Ocean, but with with slightly different location.
+The 135º phase is associated with SST in the central Pacific (Fig.\ \@ref(fig:sst-psi-2)a.1), while the 45º phase is associated with SST in the eastern Pacific (Fig.\ \@ref(fig:sst-psi-2)c.1), which correspond roughly to the Central Pacific and Eastern Pacific and "flavours" of ENSO, respectively [@kao2009].
+Both of these phases could be also associated to wave trains generated in the region surrounding Australia and propagates toward the extra-tropics.
+
+(ref:enso-phase-cap) SON ONI values plotted against cEOF2 phase for the 1979 -- 2019 period.
+Years with magnitude of cEOF2 greater (smaller) than the 50th percentile are shown as orange diamonds (green circles).
 Black line is the fit ONI \~ sin(phase) computed by weighted OLS using the magnitude of the cEOF2 as weights.
@@
 
-To better explore the relationship between tropical forcing and phase of the cEOF2, Figure \@ref(fig:enso-phase) shows the ONI index against the cEOF2 phase for each SON trimester between 1979 and 2019, highlighting years in which the magnitude of cEOF2 is above the median.
-In years with positive ONI, the phase of the cEOF2 is mostly around +90º (corresponding with positive imaginary part) and vice versa.
-In years with near neutral ENSO, the phase of the cEOF2 is much more variable.
-The black line in Figure \@ref(fig:enso-phase) is a sinusoidal fit of the relationship between ONI and cEOF2 phase.
+To better explore the relationship between tropical forcing and phases of the cEOF2, Figure\ \@ref(fig:enso-phase) shows the ONI index plotted against the cEOF2 phase for each SON trimester between 1979 and 2019, highlighting years in which the magnitude of cEOF2 is above the median.
+In years with positive (negative) ONI, the cEOF2 phase is mostly around 90º (-90°),  corresponding with positive (negative) Imaginary component.
+In the neutral ENSO seasons, the cEOF2 phase is much more variable.
+The black line in Figure\ \@ref(fig:enso-phase) is a sinusoidal fit of the relationship between ONI and cEOF2 phase.
 The $r^2$ corresponding to the fit is `r signif(summary(cos_model)$r.squared, 2)`, statistically significant with p-value `r report_p(coef(summary(cos_model))[4])`, indicating a quasi-sinusoidal relation between these two variables.
 
-The correlation between the absolute magnitude of the ONI and the magnitude of the cEOF2 is `r cor_enso_mag$text`.
-This relationship, however, appears to be driven only by the three years with strongest ENSO events in the period (`r knitr::combine_words(year(gdata[order(-oni)][1:3]$time))`) which also coincide with the three years with strongest cEOF2 magnitude.
+The correlation between the absolute magnitude of the ONI index and the  cEOF2 amplitude is `r cor_enso_mag$text`. 
+However, this relationship is mostly driven by the three years with strongest ENSO events in the period (`r knitr::combine_words(year(gdata[order(-oni)][1:3]$time))`) which coincide with the three years with strongest cEOF2 magnitude (not shown).
 If those years are removed, the correlation becomes non-significant (`r cor_enso_mag_outlier$text`).
 Furthermore, even when using all years, the Spearman correlation --which is robust to outliers-- is also non-significant (`r signif(cor_enso_spearman$estimate, 2)`, p-value`r report_p(cor_enso_spearman$p.value)`).
-Therefore, although the location of tropical SST seem to have an effect in defining the phase of the cEOF2, the relationship between the magnitude of the cEOF2 train wave and ONI remains uncertain.
+Therefore, although the location of tropical SST seem to have an effect in defining the phase of the cEOF2, the relationship between the magnitude of cEOF2 and ONI remains uncertain.
 
@@
 However, when cEOF2 is excited by tropical SST variability, it tends to remain locked to the imaginary phase.
-This explains the relative over-abundance of years with cEOF2 near positive and negative imaginary phase in Figure \@ref(fig:phase-histogram).
-
-Unlike the cEOF2 case, there is no significant pattern of SST anomalies associated with either the Real or Imaginary cEOF1 (Figure \@ref(fig:sst-psi-1)).
+This explains the relative over-abundance of years with cEOF2 near positive and negative imaginary phase in Figure\ \@ref(fig:phase-histogram).
+
+Unlike the cEOF2 case, there are no significant  SST regressed anomalies associated with either the Real or Imaginary cEOF1 (Sup. Figure\ \@ref(fig:sst-psi-1)). 
 Consistently, streamfunction anomalies do not show any tropical influence. 
-Instead, the real and Imaginary cEOF1 are associated with zonally wave activity fluxes in the extra-tropics around 60ºS, except for an equatorward flow from the coast of Antarctica around 150ºE in the real component.
-This suggests that its variability is driven primary by the internal variability of the extra-tropics.
+Instead, the real and Imaginary cEOF1 are associated with zonally wave activity fluxes in the extra-tropics around 60ºS, except for an equatorward flow from the coast of Antarctica around 150ºE in the Real component.
+This suggests that the variability of cEOF1 is driven primary by the internal variability of the extra-tropics.
 
@@
 
-(ref:cap-pp-t2m-r2) Explained variance ($r^2$) of 2-metre temperature (row a) and precipitation (row b) with cEOF1 (column 1) and cEOF2 (column 2).
-
-The influence of cEOFs variability in continental rainfall and 2-metre temperature in the Southern Hemisphere is also explored.
-Figure \@ref(fig:pp-t2m-r2) shows the $r^2$ between the cEOF1 and cEOF2 and 2-metre temperature and precipitation.
-For the cEOF1 there are no significant signals in precipitation and most of the temperature signal is low and restricted to the eastern coast of Antarctica.
-
-There is a strong signal between both variables and cEOF2, with explained variance greater than 50% in some regions.
-For 2-metre temperature, there are high values in the tropical Pacific and the SPCZ, as well as the region following an arc between New Zealand and the South Atlantic.
-Over the continents, moderate values of about 30% variance explained can be seen in southern Australia, some regions of South America and the Antarctic Peninsula.
-For precipitation, there are high values over the tropics.
-At higher latitudes, moderate values are observed over eastern Australia and some regions of southern South America.
-
-Since the cEOF1 has a relatively weak signal, we will focus on the cEOF2.
+(ref:cap-pp-t2m-r2) Explained variance ($r^2$ as percentage) of 2-metre temperature (row a) and precipitation (row b) anomalies by the regression upon cEOF1 (column 1) and cEOF2 (column 2).
+
+Figure\ \@ref(fig:eof1-regr-t) shows the variance of the 2-meter temperature or precipitation anomalies explained variance by the mluitple linear model of Real cEOF1 and Imaginary cEOF1 (column 1), and Real cEOF2 and Imaginary cEOF2 (column 2).
+The variance explained by cEOF1 for precipitation anomalies is extremely low and also for most of the temperature anomalies, except for the eastern western of Antarctica (Fig.\  \@ref(fig:eof1-regr-t)a.1).
+
+On the other hand, the cEOF2  explained variance is greater than 50% in some regions for both variables (Fig.\ \@ref(fig:eof1-regr-t) column 2). 
+For 2-metre temperature, there are high values in the tropical Pacific and the SPCZ, as well as the region following an arc between New Zealand and the South Atlantic, with higher values in the Southern Ocean.
+Over the continents, there are moderate values of about 30% variance explained  in southern Australia, Southern South America and the Antarctic Peninsula.
+For precipitation, there are high values over the tropics. At higher latitudes, moderate values are observed over eastern Australia and some regions of southern South America.
+
+Since the cEOF1 has a relatively weak signal in the surface variables explored here, we will only focus on the cEOF2 influence.
 
@@
 
-Figure \@ref(fig:pp-temp-2) shows regression maps of 2-meter temperature (column 1) and precipitation (column 2) upon standardised cEOF2 which, again, can be interpreted as the anomalies associates with a 1 standard deviation increase.
-
-(ref:pp-temp-cap) Regression of SON mean 2-meter temperature (K, shaded) and 850hPa geopotential height (m, contours) (column 1), and precipitation (correlation, column 2) with different phases of cEOF2.
+Figure\ \@ref(fig:pp-temp-2) shows regression maps of 2-metre temperature (column 1) and precipitation (column 2) anomalies upon different phases of standardised cEOF2.
+
+(ref:pp-temp-cap) Regression of SON mean 2-meter temperature (K, shaded) and 850 hPa geopotential height (m, contours) (column 1), and precipitation (correlation, column 2) upon different phases of cEOF2.
 For the 1979 -- 2019.
@@
 
-Temperature anomalies associated with the Imaginary cEOF (Fig.
-\@ref(fig:pp-temp-2).b1) show positive values in the tropical Pacific, consistent with SSTs anomalies associated with the same phase (Fig.
-\@ref(fig:sst-psi-2).b1).
-
-At higher latitudes there is a wave-like pattern of positive and negative values that coincide with the nodes of the 850hPa geopotential height regression patterns.
+Temperature anomalies associated with the Imaginary cEOF (Fig.\ \@ref(fig:pp-temp-2).b1) show positive values in the tropical Pacific, consistent with SSTs anomalies associated with the same phase (Fig.\ \@ref(fig:sst-psi-2).b1).
+
+At higher latitudes there is a wave-like pattern of positive and negative values that coincide with the nodes of the 850 hPa geopotential height regression patterns.
 This is consistent with temperature anomalies produced by meridional advection of temperature by the meridional winds arising from geostrophic balance.
 
-Over the continents, the Imaginary cEOF2 (Fig.
-\@ref(fig:pp-temp-2)b.1) is associated with positive temperatures in southern Australia and negative temperature in southern South America and the Antarctic Peninsula.
-
-The temperatures anomalies associated with the Real cEOF2 (Fig. \@ref(fig:pp-temp-2)d.1) are less extensive.
-There is no significant signal over the continents except for positive anomalies near the Antarctic Peninsula.
-
-Tropical precipitation anomalies associated with the Imaginary cEOF2 are strong, with positive anomalies in the central Pacific and western Indian, an negative anomalies in the eastern Pacific.
-This pattern is consistent with the SST anomalies shown before as the increased SST enhances tropical convection and vice versa.
-
-On the extra-tropics, the Imaginary cEOFs is correlated with drier conditions over eastern Australia; a similar signal as the one associated with ENSO @cai2011.
+Over the continents, the Imaginary cEOF2 (Fig.\ \@ref(fig:pp-temp-2)b.1) is associated with positive temperatures in southern Australia and negative temperature in southern South America and the Antarctic Peninsula, that are a result of the wave train described before.
+
+The temperatures anomalies associated with the Real cEOF2 (Fig.\ \@ref(fig:pp-temp-2)d.1) are less extensive and restricted to mid and high latitudes.   
+Over the continents, the temperature anomalies regressions are non significant, except for positive anomalies near the Antarctic Peninsula.
+
+Tropical precipitation anomalies associated with the Imaginary cEOF2 are strong, with positive anomalies in the central Pacific and western Indian, and negative anomalies in the eastern Pacific (Fig.\ \@ref(fig:pp-temp-2)b.2).
+This field is consistent with the SST regression map (Fig.\ \@ref(fig:pp-temp-2)b.1) as the positive SST anomalies enhance tropical convection and the negative SST anomalies inhibits it.
+
+In the extra-tropics, the positive Imaginary cEOF2 is related to drier conditions over eastern Australia and the surrounding ocean, that it is similar signal as the one associated with ENSO [@cai2011].
 However, the Imaginary cEOF2 is not the direction most correlated with precipitation in that area. 
-The 135º direction (positive Imaginary + negative Real cEOF2) is associated with stronger and more extensive correlations with precipitation over Australia and New Zealand.
-This Australian signal is most likely there due to the more direct impacts of ENSO rather than on the interconnection pattern represented by the cEOF2.
-
-(ref:australia-pp-phase-cap) Same as Figure \@ref(fig:enso-phase) but for precipitation anomalies averaged between `r knitr::combine_words(LonLabel(box_australia$lon))` and `r knitr::combine_words(LatLabel(box_australia$lat))` (box shown in Figure \@ref(fig:pp-oceania) column b) in $\frac{\mathrm{mm}}{\mathrm{day}}$.
-Black line is the fit precipitation \~ sin(phase) + cos(phase) computed by weighted OLS using the magnitude of the cEOF2 as weights.
-
-```{r australia-pp-phase, fig.cap = "(ref:australia-pp-phase-cap)", fig.width = width_column, fig.height=3, include = FALSE}
-gdata <- cmap %>% 
-  .[lon %between% box_australia$lon  & lat %between% box_australia$lat] %>% 
-  .[, .(pp = weighted.mean(pp, cos(lat*pi/180))), by = time] %>% 
-  .[, pp := Anomaly(pp), by = season(time)] %>% 
-  .[ceof$eof[[1]]$left, on  = "time"] %>% 
-  .[cEOF == "cEOF2"] %>% 
-  .[, arg := Arg(hgt)] %>% 
-  .[, mag := Mod(hgt)] %>% 
-  .[, q := ifelse(mag >= quantile(mag, 0.5), "|cEOF2| > 50%", "|cEOF2| < 50%")]
-
-model <- lm(pp ~ I(sin(arg)) + I(cos(arg)) - 1, weights = mag, data = gdata)
-fstatistic <- summary(model)$fstatistic
-p_value <- pf(fstatistic["value"], fstatistic["numdf"], fstatistic["dendf"], 
-              lower.tail = FALSE)
-
-equation <- latex2exp::TeX(paste0("$\\mathrm{Precipitation}_{\\mathrm{East Australia}} = ", 
-                                  signif(coef(model)[1], 2), "\\sin{(\\mathrm{phase})} + ", 
-                                  signif(coef(model)[2], 2), "\\cos{(\\mathrm{phase})}$"))
-
-gdata %>% 
-  ggplot(aes(Arg(hgt), pp)) +
-  geom_hline(yintercept = 0, size = 0.2, color = "gray50") +
-  geom_vline(xintercept = 0, size = 0.2, color = "gray50") +
-  annotate(shadowtext:::GeomShadowText,
-           label = c("- Real", "- Imaginary", "+ Real", "+ Imaginary", "- Real"),
-           color = "gray60", bg.colour = "white", bg.r = 0.2,
-           size = 3,
-           y = -Inf, x = c(-pi, -pi/2, 0, pi/2, pi), angle = 90, hjust = 0) +
-
-  geom_smooth(data = ~qwrap(.x, arg = c(-pi, pi) ~ c(-2*pi, 2*pi)),
-              method = "lm", aes(weight = mag),
-              formula = y ~ I(sin(x)) + I(cos(x)) -1,
-              se = FALSE, color = "black", size = 0.5, 
-              fullrange = TRUE) +
-  geom_point(aes(colour = q,
-                 shape  = q),
-             size = 3) +
-
-  # annotate("text", x = 90*pi/180, y = 1, label = equation, parse = TRUE) +
-
-  scale_color_brewer(NULL, palette = "Dark2") +
-  scale_shape_manual(NULL, values = c(20, 18)) +
-  scale_y_continuous("Precipitation anomaly") +
-  scale_x_continuous("Phase",
-                     breaks = seq(-pi, pi, by = 30*pi/180),
-                     labels = function(x) paste0(round(x*180/pi, 5), "º")) +
-  coord_cartesian(xlim = c(-pi, pi)) +
-  labs(subtitle = equation) +
-  theme(plot.subtitle = element_text(size = 8)) +
-  panel_background
-```
-
-<!-- To better understand the relationship between precipitation anomalies over eastern Australia and the cEOF2, Figure @ref(fig:australia-pp-phase) plots precipitation anomalies averaged over eastern Australia (box in Figure @ref(fig:pp-oceania)) as a function of cEOF2 phase.
--->
-
-<!-- The black line shows a sinusoidal fit, whose coefficients are fitted by weighted least squares using the magnitude of the EFO2 as weight. -->
-
-<!-- The $r^2$ corresponding to the fit is `r signif(summary(model)$r.squared, 2)` (p-value `r report_p(p_value)`). -->
-
-<!-- There is a strong relationship between the phase of the cEOF2 and precipitation anomalies in eastern Australia, but the Real and Imaginary phases we chose are not exactly aligned with the direction that maximises this relationship.
--->
-
-Over South America, the Imaginary cEOF has positive correlations with precipitation in South Eastern South America (SESA) and central Chile, and negative correlations in eastern Brazil.
-This again is a well-known springtime precipitation signature of ENSO [@cai2020a] and it is also similar to the precipitation anomalies associated with the A-SAM [\@ @campitelli2022].
-
-(ref:pp-america-cap) As Figure \@ref(fig:pp-oceania) but for South America
-
-```{r pp-america, fig.cap = "(ref:pp-america-cap)", fig.width=width_column, fig.height=width_column*1.1}
-# plot_pp(data = regr_pp, which_cEOF = c("cEOF1", "cEOF2"), which_continent = "america")  +
-#   scale_fill_divergent_discretised("Precipitation (mm/day)", 
-#                                    limits = c(-1, 1),
-#                                    # oob = scales::squish,
-#                                    low = "#8E521C",
-#                                    high = "#00665E",
-#                                    guide = guide_colorsteps_bottom(15)) 
-# annotate("rect", xmin = min(sesa$lon), xmax = max(sesa$lon), 
-#          ymin = min(sesa$lat), ymax = max(sesa$lat),
-#          fill = NA, color = "gray50", size = 0.3)
-#          
-```
-
-```{r sinusoidal-sesa}
-gdata <- cmap %>% 
-  .[lon %between% sesa$lon  & lat %between% sesa$lat] %>% 
-  .[, .(pp = weighted.mean(pp, cos(lat*pi/180))), by = time] %>% 
-  .[, pp := Anomaly(pp), by = season(time)] %>% 
-  .[ceof$eof[[1]]$left, on  = "time"] %>% 
-  .[cEOF == "cEOF2"] %>% 
-  .[, arg := Arg(hgt)] %>% 
-  .[, mag := Mod(hgt)]
-
-model <- lm(pp ~ I(sin(arg)) + I(cos(arg)) - 1, data = gdata)
-
-fstatistic <- summary(model)$fstatistic
-p_value <- pf(fstatistic["value"], fstatistic["numdf"], fstatistic["dendf"], 
-              lower.tail = FALSE)
-```
-
-This is not surprising considering the close relationship between the ONI, the A-SAM index and the Imaginary cEOF2 shown previously, but further consolidates the identification of this mode with the PSA pattern.
-Resembling the relationship between ONI and the phase of cEOF2 (Fig. \@ref(fig:enso-phase)), there is a cEOF2 phase dependence of the precipitation anomalies in SESA (not shown).
-These variables could be related with a significant sinusoidal fit that has a statistically significant coefficient of determination of `r round(summary(model)$r.squared, 2)`.
-
-The precipitation signal associated with the Real cEOF2 is less strong.
+The 135º phase (an intermediate between positive Imaginary and negative Real cEOF2) component is associated with stronger and more extensive temporal correlations  with precipitation over Australia and New Zealand.
+The influence of cEOF2 in Australian precipitation could be more related to the direct impacts of SST anomalies in the surrounded oceans rather than on the interconnection pattern represented by the cEOF2.
+
+Over South America, the Imaginary cEOF2 has positive correlations with precipitation in South Eastern South America (SESA) and central Chile, and negative correlations in eastern Brazil.
+This correlation field  matches the springtime precipitation signature of ENSO [e.g.
+@cai2020a] and it is also similar to the precipitation anomalies associated with the A-SAM [@@campitelli2022].
+
+This result is not surprising considering the close relationship of the Imaginary cEOF2 with both ONI and A-SAM index, which was shown previously.  
+Furthermore, it consolidates the identification of the cEOF2 with the PSA pattern.
+Resembling the relationship between ONI and the phase of cEOF2 (Fig.\  \@ref(fig:enso-phase)), there is a cEOF2 phase dependence of the precipitation anomalies in SESA (not shown).
+
+The correlation coefficients between precipitation anomalies and the Real cEOF2 (Fig.\ \@ref(fig:pp-temp-2)d.2) are weaker than for Imaginary cEOF2.
 There is a residual positive correlation in the equatorial eastern Pacific and small, not statistically significant positive correlations over eastern Australia and negative ones over New Zealand.
@@
 
-(ref:pp-temp-cap-1) Same as Figure \@ref(fig:pp-temp-2) but for cEOF1.
-
 ```{r pp-temp-1, fig.cap="(ref:pp-temp-cap-1)", fig.width=6.6, fig.height=5.5, dependson="plot_pp"}

### 2022-03-25 analysis/paper/shceof.tex
Render

@@
 Addressing this limitation, the Fourier technique can be generalized to integrate all planetary wave amplitude regardless of wave number, by computing the wave envelope (Irving and Simmonds 2015).
-The latter makes possible to represent planetary waves with different amplitude at different longitudes, but it removes all information about phase and wave number.
+The latter makes it possible to represent planetary waves with different amplitude at different longitudes, but it removes all information about phase and wave number.
 With this approach, Irving and Simmonds (2015) showed that planetary wave amplitude in general is associated to Antarctic sea-ice concentration and temperature, as well as to precipitation anomalies in regions of significant topography in SH mid-latitudes and Antarctica.
@@
 An alternative methodology that has been proposed to study travelling and standing waves is complex Empirical Orthogonal Functions (Horel 1984).
-This method extends EOF analysis to capture oscillations with varying amplitude and phase and has been applied to the time domain on water level to study tide-river interactions (Laurel-Castillo and Valle-Levinson 2020) and temperature and precipitation anomalies (e.g.
-Krokhin and Luxemburg 2007; Gelbrecht, Boers, and Jürgen Kurths 2018).
+This method extends EOF analysis to capture oscillations with varying amplitude and phase and has been applied to the time domain.
+For instance, Krokhin and Luxemburg (2007) applied cEOF to station-based monthly precipitation anomalies and monthly temperature anomalies in the Eastern Siberia and the Far East region to characterise the main modes of variability and their connection to teleconnection indices.
+Similarly, Gelbrecht, Boers, and Jürgen Kurths (2018) used cEOF applied to daily precipitation from reanalysis to study the propagating characteristics of the South American Monsoon.
 To our knowledge, complex EOF has not been applied in the spatial domain to capture the phase-varying nature of planetary waves.
@@
 
-In Section 2 we describe the methods.
-In Section 3.1, we analyse the spatial patterns of each complex EOF.
-In Section 3.2 we study the spatial regressions with geopotential height and temperature anomalies.
-In Section 3.3, we analyse the relationship between cEOF1 and Ozone, and between cEOF2 with the SAM and the PSA.
-In Section 3.4 we study tropical forcings that explain the variability of each cEOF.
-Finally, in Section 3.5 we show the relationship between these modes of variability and precipitation anomalies in South America and Oceania.
+In Section \ref{methods} we describe the methods.
+In Section \ref{spatial} we analyse the spatial patterns of each complex EOF.
+In Section \ref{regressions} we study the spatial regressions with geopotential height and temperature anomalies.
+In Section \ref{other-variables} we analyse the relationship between cEOF1 and Ozone, and between cEOF2 with the SAM and the PSA.
+In Section \ref{tropical} we study tropical forcings that explain the variability of each cEOF.
+Finally, in Section \ref{precipitation} we show the relationship between these modes of variability and precipitation anomalies in South America and Oceania.
 
@@
 
-We used monthly geopotential height, air temperature, ozone mixing ratio, and total ozone column (TOC) at 2.5\degree~longitude by 2.5\degree~latitude of horizontal resolution and 37 vertical isobaric levels from the European Centre for Medium-Range Weather Forecasts Reanalysis version 5 {[}ERA; Hersbach et al.
-(2019){]} for the period 1979-2019.
+We used monthly geopotential height, air temperature, ozone mixing ratio, and total ozone column (TOC) at 2.5\degree longitude by 2.5\degree latitude of horizontal resolution and 37 vertical isobaric levels from the European Centre for Medium-Range Weather Forecasts Reanalysis version 5 {[}ERA; Hersbach et al.
+(2019){]} for the period 1979 -- 2019.
 Most of our analysis is restricted to the post-satellite era to avoid any confounding factors arising from the incorporation of satellite observations, but we also used the preliminary back extension of ERA5 from 1950 to 1978 (Bell et al.
@@
 2017) and precipitation monthly data from the CPC Merged Analysis of Precipitation (P.
-Xie and Arkin 1997), with a 2.5\degree~resolution in latitude and longitude.
+Xie and Arkin 1997), with a 2.5\degree resolution in latitude and longitude.
 The rainfall gridded dataset is based on information from different sources such as rain gauge observations, satellite inferred estimations and the NCEP-NCAR reanalysis, and it is available since 1979 to present.
+
+The Oceanic Niño Index (ONI, Bamston, Chelliah, and Goldenberg 1997) comes from NOAA's Climate Prediction Center and the Dipole Mode Index (DMI, Saji and Yamagata 2003) from Global Climate Observing System Working Group on Surface Pressure.
 
@@
 
-We computed the level-dependent SAM index as the leading EOF of year-round monthly geopotential height anomalies south of 20ºS at each level for the whole period according (Baldwin and Thompson 2009).
+We computed the level-dependent SAM index as the leading EOF of year-round monthly geopotential height anomalies south of 20ºS at each level for the whole period (Baldwin and Thompson 2009).
 We further split the SAM into its zonally symmetric and zonally asymmetric components (S-SAM and A-SAM indices respectively).
 These indices were obtained by projecting the zonally asymmetric and zonally symmetric part of the SAM spatial pattern onto monthly geopotential height fields, as proposed by Campitelli, Díaz, and Vera (2022).
-Seasonal indices of the Pacific South American patterns (PSA1 and PSA2) were calculated -- in agreement with Mo and Paegle (2001) -- as the third and fourth leading EOF of seasonal mean anomalies for SH 500 hPa geopotential height.
+Seasonal indices of the Pacific South American patterns (PSA1 and PSA2) were calculated, in agreement with Mo and Paegle (2001), as the third and fourth leading EOF of seasonal mean anomalies for 500-hPa geopotential heights at SH.
 
@@
 
-In standard way of applying EOF analysis zonal waves may appear as pairs of EOFs, that could be degenerated representing similar patterns but shifted in phase (Horel 1984).
-Figure~\ref{fig:eof-naive} shows the four eading EOFs of SON geopotential height zonal anomalies at 50 hPa south of 20\degree S.
-It is clear that the first two EOFs represent the same zonal wave 1 pattern and the last two represent a same zonal wave pattern with shorter wavenumber and four centers of action shifted by 1/4 wavelength.
+In the standard EOF analysis, zonal waves may appear as pairs of (possibly degenerated) EOFs representing similar patterns but shifted in phase (Horel 1984).
+Figure \ref{fig:eof-naive} shows the four leading EOFs of SON geopotential height zonal anomalies at 50 hPa south of 20\degree S.
+It is clear that the first two EOFs represent a single phase-varying zonal wave 1 pattern and the last two represent a similarly phase-varying pattern with shorter wavenumber and four centres of action shifted by 1/4 wavelength.
 A similar EOF structure can be seen in 200 hPa (not shown).
-Since each pair of EOFs seem to represent the same phase-varying structure, it would be desirable to combine them into a single index with amplitude and phase.
-
-ON the other hand, complex Empirical Orthogonal Functions (cEOF) is a useful method to characterise zonal waves, associated with phase-varying structures (Horel 1984).
+Since each pair of EOFs seems to represent the same phase-varying structure, it would be desirable to combine them into a single pattern described by amplitude and phase.
+
+Complex Empirical Orthogonal Functions (cEOF) are a useful method for characterising zonal waves, associated with phase-varying structures (Horel 1984).
 This method involves computing EOF on the analytic representation of the original field.
@@
 
-The phase of principal components is defined up to an additive constant.
-For real principal components, this constant can be either 0 or \(\pi\), corresponding to a change in sign.
-For complex principal components, it can be any number between 0 and \(2\pi\) (Horel 1984).
-Since any choice is arbitrary and equally valid, we chose the phase of each cEOF so that the real and imaginary parts are aligned with meaningful phases in our analysis.
-This procedure does not create a spurious correlation, it only takes a known relationship and aligns it with a specific phase.
-For the first complex principal component, the phase was chosen so that the time series corresponding to the real part has the maximum correlation with the zonal wave 1 of Total Ozone Column between 75°S and 45°S.
-This also nearly minimises the correlation with the imaginary part.
-For the second complex principal component, the phase was chosen so that the coefficient of determination between the Oceanic Niño Index (Bamston, Chelliah, and Goldenberg 1997) and the real part was minimized, which also nearly maximizes the correlation with the imaginary part.
+The phase of cEOFs is defined up to an additive constant.
+For real EOFs, this constant can be either 0 or \(\pi\), corresponding to a change in sign.
+For cEOFs, it can be any real number between 0 and \(2\pi\) (Horel 1984), corresponding to rotations in the complex plane.
+Since any choice is arbitrary and equally valid, we chose the phase of each cEOF so that the real or imaginary parts are aligned with meaningful phases in our analysis.
+
+Preliminary analysis showed that the first cEOF was closely related to the the zonal wave 1 of Total Ozone Column and the second cEOF was closely related to ENSO.
+Therefore, to aid in the interpretation, we chose the phases of each cEOF so that these relationships are aligned with one of the complex components.
+
+We chose the phase of cEOF1 so that the time series corresponding to the real part has the maximum correlation with the zonal wave 1 of Total Ozone Column between 75°S and 45°S.
+And we chose the phase of cEOF2 so that the coefficient of determination between the Oceanic Niño Index (Bamston, Chelliah, and Goldenberg 1997) and the real part is minimized, which also nearly maximizes the correlation with the imaginary part.
+
+This procedure does not create spurious correlations, it only takes an existing relationship and aligns it with a specific phase.
 
@@
 \[
-X(\lambda, \phi, t) = \alpha(\lambda, \phi) \operatorname{Im}(\operatorname{cEOF}) + \beta(\lambda, \phi) \operatorname{Re}(\operatorname{cEOF}) + X_0(\lambda, \phi) +  \epsilon(\lambda, \phi, t)
+X(\lambda, \phi, t) = \alpha(\lambda, \phi) \operatorname{Im}(\mathrm{cEOF}) + \beta(\lambda, \phi) \mathrm{Re}(\mathrm{cEOF}) + X_0(\lambda, \phi) + \epsilon(\lambda, \phi, t)
 \]
@@
 All graphics are made using ggplot2 (Wickham 2009).
-We downloaded data from reanalysis using the ecmwfr package (Hufkens 2020) and indices of the ENSO with the rsoi package (Albers and Campitelli 2020).
-The paper was rendered using knitr and rmarkdown (Y.
-Xie 2015; Allaire et al.
-2020).
+We downloaded data from reanalysis using the ecmwfr package (Hufkens 2020) and indices of ENSO and Indian Ocean Dipole with the rsoi package (Albers and Campitelli 2020).
+The paper was rendered using knitr and rmarkdown (Y. Xie 2015; Allaire et al. 2020).
 
@@
 \begin{figure*}
-\includegraphics{../figures/extended-series-1} \caption{Time series extended using ERA5 back extended preliminary edition (period 1950 -- 1978) and ERA5 (period 1979 -- 2019).
+\includegraphics{../figures/extended-series-1} \caption{Time series of the two leading cEOFs of SON zonal anomalies of geopotential height at 50 hPa and 200 hPa.
+cEOF1 (row a) and cEOF2 (row b) separated in their real (column 1) and imaginary (column 2) components.
 Dark straight line is the linear trend.
@@
 To describe the variability of the circulation zonal anomalies, the spatial and temporal parts of the first two leading cEOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa, computed jointly at both levels, are shown in Figures \ref{fig:ceofs-1} and \ref{fig:extended-series}.
-The first mode (cEOF1) explains 82\% of the variance, while the second mode (cEOF2) explains a smaller fraction (7\%).
+The first mode (cEOF1) explains 82\% of the variance of the zonally anomalous fields, while the second mode (cEOF2) explains a smaller fraction (7\%).
 In the spatial patterns (Fig.~\ref{fig:ceofs-1}), the real and the imaginary components are in quadrature by construction, so that each cEOF describe a single wave-like pattern whose amplitude and position (i.e.~phase) is controlled by the magnitude and phase of the temporal cEOF.
@@
 In particular, the dominant structure at both levels is a wave 3 but with larger amplitude in the pacific sector.
-This modulated amplitude is especially apparent at 200 hPa.
-There is no apparent phase shift but the amplitude of the pattern is greatly reduced in the stratosphere, suggesting that this barotropic mode represents mainly tropospheric variability.
-
-All cEOFs have non zero mean (Fig. \ref{fig:extended-series}) because the geopotential fields that enter into the cEOFs algorithm are anomalies with respect to the zonal mean instead of the time mean.
+This modulated amplitude more evident at 200 hPa.
+There is no apparent phase shift with height but the amplitude of the pattern is greatly reduced in the stratosphere, suggesting that this barotropic mode represents mainly tropospheric variability.
+
+There is no significant simultaneous correlation between cEOFs time series.
+Both cEOFs show year-to-year variability but show no evidence of decadal variability (Fig. \ref{fig:extended-series}).
+Because the geopotential fields that enter into the cEOFs algorithm are anomalies with respect to the zonal mean instead of the time mean, the cEOFs have non zero temporal mean.
 However, cEOF2's mean is almost zero, which indicates that only cEOF1 includes variability that significantly projects onto the mean zonally anomalous field.
-There is no significant simultaneous correlation between the time series.
-Both cEOFs show year-to-year variability but show no evidence of decadal variability.
 
 A significant upward trend in the real component of cEOF1 is evident (Fig.~\ref{fig:extended-series}a.1, p-value~= 0.0037) while there is no significant trend in any of the complex components of cEOF2.
-The positive trend in the Real cEOF1 translates into a positive trend in cEOF1 magnitude, but not in the phase (not shown).
+The positive trend in the Real cEOF1 translates into a positive trend in cEOF1 magnitude, but not systematic change in phase (not shown).
 This long-term change indicates an increase in the magnitude of the high latitude zonal wave 1.
@@
 \hypertarget{regressions}{%
-\subsection{cEOFs Regressions}\label{regressions}}
+\subsection{cEOFs Regression maps}\label{regressions}}
 
@@
 Figure~\ref{fig:eof1-regr-gh} shows regression maps of SON geopotential height anomalies upon cEOF1.
-At 50 hPa (Figure~\ref{fig:eof1-regr-gh} row a), both the Real and Imaginary cEOF1 are associated with planetary wave 1 patterns, that are 90º out of phase.
-Their phases coincide with the ones shown in Figure~\ref{fig:ceofs-1}a.1, with the positive centre of the Real cEOF1 located towards the dateline, and the one of the Imaginary cEOF1 located over East Antarctica.
-Instead of a wave-1 pattern, regressed anomalies describe a monopole with its centre displaced from the South Pole. The latter indicates an influence of cEOF1 in modulating the polar vortex.
-
-The regression maps for cEOF1 at 200 hPa (Figure~\ref{fig:eof1-regr-gh} row b) are similarly influenced by the zonally symmetric circulation.
-It is only possible to partially distinguish the wave 1 pattern in relation with the Real cEOF1 (Figure~\ref{fig:eof1-regr-gh}b.1).
-The Imaginary cEOF1 shows a much more zonally symmetrical pattern resembling the negative SAM phase.
-
-With the exception of the Imaginary cEOF1, it is clear that these patterns are very different than the fully zonally asymmetric versions (Fig.~\ref{fig:ceofs-1}), particularly at 200 hPa.
-Moreover, only in the stratosphere these patterns actually show a distinguishable wave 1 pattern shifted in phase by 90º, suggesting that using the cEOF method is artificially generating a wave structure at 200hPa.
+At 50 hPa (Figure~\ref{fig:eof1-regr-gh} row a), the Real cEOF1 is associated with a monopole centred over the Ross Sea.
+The Imaginary cEOF1 is associated with a clear wave 1 pattern with maximum over the coast of East Antarctica.
+
+At 200 hPa (Figure~\ref{fig:eof1-regr-gh} row b) the Real cEOF1 shows a single area of positive anomalies spanning West Antarctica surrounded by opposite anomalies in lower latitudes with is centre shifted slightly eastward compared with the upper level anomalies.
+The Imaginary cEOF2.
+The Imaginary cEOF1 shows a much more zonally symmetrical pattern resembling the negative SAM phase (Fogt and Marshall 2020).
+Both components of cEOF1 could be associated to zonal anomalies in the upper troposphere jet stream intensity.
+
 Therefore, the magnitude and phase of the cEOF1 are associated with the magnitude and phase of a zonal wave only in the stratosphere.
@@
 \includegraphics{../figures/eof2-regr-gh-1.pdf}
-\caption{\label{fig:eof2-regr-gh}Same as Figure~\ref{fig:eof1-regr-gh} but for the second cEOF.}
-\end{figure}
-
-Figure~\ref{fig:eof2-regr-gh} shows the regression pattern of geopotential height anomalies upon the cEOF2.
-Unlike for cEOF1, in this case the regression patterns are similar to the fully zonally asymmetric patterns from Figure~\ref{fig:ceofs-1}.
-Although there are some differences (particularly in 50 hPa), the wave trains identified before are well characterised and patterns associated with the Real cEOF2 are 90º out of phase with those associated with the Imaginary cEOF2.
-Zonal wave 3 dominates all fields, but only in the western hemisphere, over the Pacific and Atlantic Oceans.
-cEOF2 then represents an equivalent barotropic wave train that is very similar to the the Pacific South American Patterns (Mo and Paegle 2001).
-Comparing the location of the positive anomaly near 90ºW in column b of Figure~\ref{fig:eof2-regr-gh} with Figures 1.a and b from Mo and Paegle (2001), the Real cEOF2 can roughly be identified with PSA2, while the Imaginary cEOF2 resembles PSA1.
+\caption{\label{fig:eof2-regr-gh}Same as Figure~\ref{fig:eof1-regr-gh} but for cEOF2.}
+\end{figure}
+
+Figure~\ref{fig:eof2-regr-gh} shows the regression maps of geopotential height anomalies upon the cEOF2.
+Wave trains similar to those identified for cEOF2 patterns (Fig~\ref{fig:ceofs-1}) can be distinguished.
+Regressed anomalies associated with the Real cEOF2 are 90º out of phase with those associated with the Imaginary cEOF2.
+All fields have a dominant zonal wave 3 limited to the western hemisphere, over the Pacific and Atlantic Oceans.
+cEOF2 then represents an equivalent barotropic wave train that is very similar to the the PSA Patterns (Mo and Paegle 2001).
+Comparing the location of the positive anomaly near 90ºW in column b of Figure~\ref{fig:eof2-regr-gh} with Figures 1.a and b from Mo and Paegle (2001), the Real cEOF2 regression map could be identified with PSA2, while the Imaginary cEOF2 resembles PSA1.
 
@@
 
-\hypertarget{sam}{%
-\subsubsection{SAM}\label{sam}}
-
-\begin{figure}
-\centering
-\includegraphics{../figures/sam-eof-vertical-1.pdf}
-\caption{\label{fig:sam-eof-vertical}Coefficient of determination between the real and imaginary part of each cEOF and the SAM, Asymmetric SAM (A-SAM) and Symmetric SAM (S-SAM) indices computed at each level according to @Campitelli, Díaz, and Vera (2022) for the 1979 -- 2019 period.
-Thick lines represent estimates with p-value \textless{} 0.01 corrected for False Detection Rate (Benjamini and Hochberg 1995).}
-\end{figure}
-
-To explore the relationship between SAM and the modes described from the cEOF, we compute the coefficient of determination between the cEOFs timeseries and the three SAM indices (SAM, A-SAM and S-SAM) at each vertical level (Fig.~\ref{fig:sam-eof-vertical}).
-The SAM index is statistically signficant correlated with the Real cEOF1 in all levels, and with the Imaginary cEOF1 and Imaginary cEOF2 in the troposphere.
-Correlations between SAM and the Real cEOF2 are non-significant.
-
-The relationship between the tropospheric SAM and cEOF1 is explained entirely by the zonally symmetric component of the SAM as shown by the low and statistically non-significant correlations between the A-SAM and either the Real or Imaginary cEOF1 and the high correlation with the S-SAM below 100 hPa.
-In the stratosphere, the Real cEOF1 is correlated with both A-SAM and S-SAM, while the Imaginary cEOF1 is highly correlated only with the A-SAM.
-These correlations are consistent with the regression patterns of geopotential height in Figure~\ref{fig:eof1-regr-gh} and their comparison with those obtained for SAM, A-SAM and S-SAM by @Campitelli, Díaz, and Vera (2022).
-
-In the case of cEOF2, the moderate correlation between SAM and Imaginary cEOF2 becomes extremely high only when the zonally asymmetric variability of the SAM is considered.
-The Imaginary cEOF2 explains up to 92\% of A-SAM variance, reached the maximum value at 225 hPa (Figure~\ref{fig:sam-eof-vertical}.b2).
-Such extremely high correlation between the asymmetric SAM and the Imaginary cEOF2 suggests that these might be different ways of characterising the same phenomenon.
-
 \hypertarget{psa}{%
@@
 
-\caption{\label{tab:psa-eof2}Correlations between the Real and Imaginary parts of cEOF2 and the PSA1 and PSA2 modes computed as the second and third EOFs of seasonal geopotential height anomalies (following Mo and Paegle 2001) for the 1979 -- 2019 period.
+\caption{\label{tab:psa-eof2}Correlation coefficients (r) between cEOF2 components and the PSA1 and PSA2 modes computed as Mo and Paegle (2001) for the 1979 -- 2019 period.
 95\% confidence intervals in parenthesis.
@@
 
-Due to the similarity between the PSA patterns and the cEOF2 spatial patterns (Fig. \ref{fig:eof2-regr-gh}), we study the relationship between these modes.
-Table \ref{tab:psa-eof2} shows the correlations between the two PSA indices and the timeseries for Real and Imaginary phase of cEOF2.
-As anticipated by Figure~\ref{fig:eof2-regr-gh}, there is a strong correlation between PSA1 and Imaginary cEOF2, and between PSA2 and Real cEOF2.
-Conversely, there is no relationship between PSA1 and Real cEOF2, and between PSA2 and Imaginary cEOF2.
-As a result, cEOF2 represents well both the spatial structure and temporal evolution of the PSA modes, as well as it is possible to make a rather clean association between its two phases and the two PSA modes.
-So, It could be concluded that the same particular rotation of cEOF2 that maximises the association between cEOF2 parts and PSA modes, is the one which maximises the relationship between ENSO and Imaginary cEOF2.
-
-The reason the conventional EOF analysis arrives at the same separation than our particular cEOF rotation is probably the fact that not all phases are equally frequent.
-To visualize that, Figure~\ref{fig:phase-histogram} shows an histogram that counts the number of SON years in which the cEOF2 was close to each of the four particular phases (positive/negative Real/Imaginary), with the observations for each season marked as rugs on the horizontal axis.
-For instance, years with cEOF2 phase within 45º of 0º are nearest the ``positive Real'' phase.
-About two thirds of time cEOF2 has a phase similar to either the negative or positive Imaginary phase.
-It could be seen that Imaginary phase is the most common phase and It is also, by construction, the direction that has the maximum relationship with ENSO.
-Therefore, the Imaginary cEOF2 explains more variance than the Real cEOF2 and conventional EOF analysis will tend to separate the two.
+Given the similarity between the cEOF2 regression maps (Fig.~\ref{fig:eof2-regr-gh}) and documented PSAs we study the relationship between these modes.
+Table~\ref{tab:psa-eof2} shows the correlations between the two PSA indices and the time series for Real and Imaginary phases of cEOF2.
+As anticipated by Figure~\ref{fig:eof2-regr-gh}, there is a strong positive correlation between PSA1 and Imaginary cEOF2, and also between PSA2 and Real cEOF2.
+On the other hand, there is no significant relationship between PSA1 and Real cEOF2, and between PSA2 and Imaginary cEOF2.
+As a result, cEOF2 represents well both the spatial structure and temporal evolution of the PSA modes, so it is possible to make an association between its two phases and the two PSA modes.
+The same phase election for cEOF2 that maximises the relationship between ENSO and Imaginary cEOF2, also maximises the association between cEOF2 components and PSA modes.
 
@@
 \includegraphics{../figures/phase-histogram-1.pdf}
-\caption{\label{fig:phase-histogram}Histogram of phase distribution of cEOF2 for the 1979 -- 2019 period.
+\caption{\label{fig:phase-histogram}Histogram of phase distribution of cEOF2 phase for the 1979 -- 2019 period.
 Bins are centred at 90º, 0º, -90º, -180º with a binwidth of 90º.
@@
 
-The result obtained here is similar to the one obtained by Irving and Simmonds (2016).
-They show that the phase distribution of the PSA-like variability obtained by them is bimodal and isolate the PSA pattern from the rest of the PSA-like variability by selecting events that are near the peaks of the distribution (compare our Figure~\ref{fig:phase-histogram} with their Figure 6).
-
-The advantage of our method is that it is much simpler to implement, it provides magnitude and phase naturally, and it facilitates the description of this mode as a propagating wave instead of as standing oscillations.
-As a consequence, the cEOF2 offers an alternative way of representing the PSA which has several advantages over using the second and third EOFs.
+Figure~\ref{fig:phase-histogram} shows an histogram that counts the number of SON seasons in which the cEOF2 phase was close to each of the four particular phases (positive/negative of real/imaginary component), with the observations for each season marked as rugs on the horizontal axis.
+In 62\% of seasons cEOF2 has a phase similar to either the negative or positive imaginary phase, making the imaginary phase the most common phase.
+This is also, by construction, the phase that is most correlated with ENSO.
+Therefore, the Imaginary cEOF2 explains more variance than the Real cEOF2 and conventional EOF analysis will tend to separate the two.
+
+This phase preferences is in agreement with Irving and Simmonds (2016), who found a bimodal distribution to PSA-like variability (compare our Figure~\ref{fig:phase-histogram} with their Figure 6).
+
+\hypertarget{sam}{%
+\subsubsection{SAM}\label{sam}}
+
+\begin{figure}
+\centering
+\includegraphics{../figures/sam-eof-vertical-1.pdf}
+\caption{\label{fig:sam-eof-vertical}Coefficient of determination (\(r^2\)) between each component of cEOFs and the SAM, Asymmetric SAM (A-SAM) and Symmetric SAM (S-SAM) indices computed at each level for the 1979 -- 2019 period.
+Thick lines represent estimates with p-value \textless{} 0.01 corrected for False Detection Rate (Benjamini and Hochberg 1995).}
+\end{figure}
+
+We now explore the relationship between SAM and the cEOFs motivated by the resemblance between cEOFs regression maps and SAM patterns shown in the previous section.
+To that end, we computed the coefficient of determination between the cEOFs time series and the three SAM indices (SAM, A-SAM and S-SAM) defined by Campitelli, Díaz, and Vera (2022) at each vertical level (Fig.~\ref{fig:sam-eof-vertical}).
+The SAM index is statistically significant correlated with the Real cEOF1 in all levels, and with the Imaginary cEOF1 and Imaginary cEOF2 in the troposphere.
+On the other hand, correlations between SAM and the Real cEOF2 are non-significant.
+
+The relationship between the SAM and cEOF1 in the troposphere is explained entirely by the zonally symmetric component of the SAM as shown by the high correlation with the S-SAM below 100 hPa and the low and statistically non-significant correlations between the A-SAM and either the Real or Imaginary cEOF1.
+In the stratosphere, the Real cEOF1 is correlated with both A-SAM and S-SAM, while the Imaginary cEOF1 is highly correlated only with the A-SAM.
+These correlations are consistent with the regression maps of geopotential height in Figure~\ref{fig:eof1-regr-gh} and their comparison with those obtained for SAM, A-SAM and S-SAM by Campitelli, Díaz, and Vera (2022).
+
+In the case of Imaginary cEOF2, the correlation between SAM and Imaginary cEOF2 for the troposphere is associated to the asymmetric variability of the SAM.
+Indeed, the Imaginary cEOF2 shares up to 92\% variance with the A-SAM and only 12\% at most with the S-SAM (Figure~\ref{fig:sam-eof-vertical}.b2).
+Such extremely high correlation between A-SAM and Imaginary cEOF2 suggests that the methodology proposed in this work is able to characterise the zonally asymmetric component of the SAM described previously by Campitelli, Díaz, and Vera (2022).
 
@@
 \includegraphics{../figures/t-vertical-1.pdf}
-\caption{\label{fig:t-vertical}Regression of SON anomalies of mean air temperature (shaded, Kelvin) and ozone mixing ratio (contours, negative contours with dashed lines, labels in parts per billion by mass) averaged between 75°S and 45°S with the (a) real and (b) imaginary parts of the first cEOF for the 1979 -- 2019 period.}
-\end{figure}
-
-The relation between cEOFs and air temperature was also evaluated.
-Figure~\ref{fig:eof1-regr-t} shows regression patterns of air temperature at 50hPa and 200hPa onto cEOF1.
+\caption{\label{fig:t-vertical}Regression of SON zonal anomalies averaged between 75°S and 45°S of mean air temperature (shaded, Kelvin) and ozone mixing ratio (contours, negative contours with dashed lines, labels in parts per billion by mass) with the (a) Real and (b) Imaginary components of the cEOF1 for the 1979 -- 2019 period.}
+\end{figure}
+
+The influence of cEOFs variability on air temperature was also evaluated.
+Figure~\ref{fig:eof1-regr-t} shows regression maps of air temperature anomalies at 50 hPa and 200 hPa upon cEOF1.
 In both levels, the Real cEOF1 is associated with a positive monopole over the South Pole with its centre moved slightly towards 150ºE (Fig.~\ref{fig:eof1-regr-t} column 1).
@@
 
-Figure \ref{fig:t-vertical} shows the vertical distribution of the regression coefficients on cEOF1 from zonal anomalies of air temperature and zonal anomalies of ozone mixing ratio averaged between 75°S and 45°S.
-Temperature zonal anomalies associated with cEOF1 show a clear wave 1 pattern for both real and imaginary components throughout the atmosphere above 250 hPa with a change in sign above 10 hPa.
-Following hydrostatic balance, this is the level in which the geopotential anomaly have maximum amplitude (not shown).
-
-The maximum ozone anomalies are co-located with the minimum temperature anomalies above 10 hPa and with the maximum temperature anomalies below 10 hPa (Fig.~\ref{fig:t-vertical}).
+Figure~\ref{fig:t-vertical} shows the vertical distribution of the regression coefficients on cEOF1 from zonal anomalies averaged between 75°S and 45°S of air temperature and of ozone mixing ratio.
+Temperature zonal anomalies associated with cEOF1 show a clear wave 1 pattern for both real and imaginary components throughout the atmosphere above 250 hPa with a sign reversal above 10 hPa.
+As a result of the hydrostatic balance, this is the level in which the geopotential anomaly have maximum amplitude (not shown).
+
+The maximum ozone regressed anomalies match with the minimum temperature anomalies above 10 hPa and with the maximum temperature anomalies below 10 hPa (Fig.~\ref{fig:t-vertical}).
 Therefore, the ozone zonal wave 1 is anticorrelated with the temperature zonal wave 1 in the upper stratosphere, and directly correlated in the upper stratosphere.
-This change in phase is observed in ozone anomalies forced by planetary waves that reach the stratosphere.
-In the photochemically-dominated upper stratosphere, cold temperatures inhibit the destruction of ozone.
-On the other hand, in the advectively-dominated lower stratosphere, ozone anomalies are 90º out of phase with horizontal and vertical transport, which are in adittion 90º out of phase with temperature anomalies (Hartmann and Garcia 1979; Wirth 1993; Smith 1995).
+This change in phase is observed in ozone anomalies forced by planetary waves that reach the stratosphere (Hartmann and Garcia 1979; Wirth 1993; Smith 1995).
+In the photochemically-dominated upper stratosphere, cold temperatures inhibit the destruction of ozone, explaining the opposite behaviour for both variables.
+On the other hand, in the advectively-dominated lower stratosphere, ozone anomalies are 90º out of phase with horizontal and vertical transport, which are in addition 90º out of phase with temperature anomalies, resulting in same sign anomalies for the response of both variables
 
@@
 \includegraphics{../figures/o3-regr-1.pdf}
-\caption{\label{fig:o3-regr}Regression of SON mean Total Ozone Column anomalies (shaded, Dobson Units) with the (a) real and (b) imaginary parts of the cEOF1 for the 1979 -- 2019 period.
+\caption{\label{fig:o3-regr}Regression of SON mean Total Ozone Column anomalies (shaded, Dobson Units) with the (a) real and (b) imaginary components of the cEOF1 for the 1979 -- 2019 period.
 On contours, the mean zonal anomaly of Total Ozone Column (negative contours in dashed lines, Dobson Units).
@@
 
-The regression maps of cEOF1 with fields of Total Ozone Column (TOC) (Fig.
-~\ref{fig:o3-regr}) show zonal wave 1 pattens in TOC associated with both phases of cEOF1.
-Climatologically, the springtime Ozone minimum is located off the South Pole and towards the Weddell Sea (Grytsai 2011).
-Thus, the Real cEOF1 regression pattern (Figure~\ref{fig:o3-regr}a) coincides with the climatological position of the ozone hole while the one for the Imaginary cEOF1 is shifted by 90º.
-
-There is a close spatial relationship between amplitudes and phases of the cEOF1 and TOC planetary wave 1 between 75°S and 45°S.
-The correlation between the amplitude of both indices is 0.79 (CI: 0.63 -- 0.88) and -0.85 (CI: -0.92 -- -0.74) between their phases.
+The regression maps of TOC anomalies upon cEOF1 (Fig.~\ref{fig:o3-regr}) show zonal wave 1 patterns associated with both components of cEOF1.
+The climatological position of the springtime Ozone minimum (ozone hole) is outside the South Pole and towards the Weddell Sea (e.g. Grytsai 2011).
+Thus, the Real cEOF1 regression field (Figure~\ref{fig:o3-regr}a) coincides with the climatological position of the ozone hole, while it is 90° out of phase for the Imaginary cEOF1.
+The temporal correlation between the amplitudes of TOC planetary wave 1 and cEOF1 is 0.79 (CI: 0.63 -- 0.88), while the correlation between their phases is -0.85 (CI: -0.92 -- -0.74).
+Consequently, cEOF1 is strongly related with the SH ozone variability.
 
@@
 \begin{figure*}
-\includegraphics{../figures/sst-psi-2-1} \caption{Regression maps of cEOF2 with SST (K, column 1) and streamfunction zonal anomalies (\(m^2/s\times10^-7\), shaded) with their corresponding activity wave flux (vectors) (column 2) for the 1979 -- 2019 period and different phases of the cEOF (illustrated in the lower-left arrow).
+\includegraphics{../figures/sst-psi-2-1} \caption{Regression of SST (K, left column ) and streamfunction zonal anomalies (\(m^2/s\times10^-7\), shaded) with their corresponding activity wave flux (vectors) (right column) upon cEOF2 different phases (illustrated in the lower-left arrow) for the 1979 -- 2019 period.
 Areas marked with dots have p-values smaller than 0.01 adjusted for FDR.}\label{fig:sst-psi-2}
@@
 
-The connections between cEOFs and Tropical sources are also assessed.
-Figure~\ref{fig:sst-psi-2} shows the regression maps of Sea Surface Temperatures (SST) and streamfunction anomalies at 200 hPa respectively upon standardised cEOF2, which can be interpreted as the anomalies associated with a 1 standard deviation increase of each index.
-Besides showing the regression of the real and imaginary parts, we also show regressions in two intermediate directions.
-The Imaginary cEOF2 is associated with strong positive SST anomalies on the Central Pacific and negative anomalies over an area across the North of Australia and New Zealand, the South Pacific Convergence Zone (SPCZ) (Figure~\ref{fig:sst-psi-2}.b1).
-This pattern is almost canonically positive ENSO (Bamston, Chelliah, and Goldenberg 1997) and indeed, the correlation between the Imaginary cEOF2 and the Oceanic Niño Index (Bamston, Chelliah, and Goldenberg 1997) is significant and very high 0.76 (CI: 0.6 -- 0.87).
-On top of this ENSO-like pattern, there are positive anomalies in the western Indian Ocean and negative values in the eastern Indian Ocean, resembling a positive Indian Ocean Dipole.
-
-Fittingly, the Imaginary cEOF2 is also associated with strong wave-like streamfunction anomalies emanating from the tropics (Figure~\ref{fig:sst-psi-2}.b2), both from the Central Pacific sector and the Incean Ocean.
-These are consistent with the effect of ENSO and the Indian Ocean Dipole on the extratropics: the SST anomalies initiate anomalous tropical convection that excites Rossby waves that propagate meridionally towards higher latitudes (Mo 2000; Cai et al.
+The connections between cEOFs and Tropical sources are also assessed, to explore plausible forcings for the described cEOFs patterns.
+Figure~\ref{fig:sst-psi-2} shows the regression maps of Sea Surface Temperatures (SST) and streamfunction anomalies at 200 hPa respectively upon standardised cEOF2,.
+Besides showing the regression maps for the Real and Imaginary components, we including the corresponding regressions for two intermediate directions (corresponding to 45° and 135°).
+
+The Imaginary cEOF2 (second row) is associated with strong positive SST anomalies on the Central Pacific and negative anomalies over an area across northern Australia, New Zealand the South Pacific Convergence Zone (SPCZ) (Figure~\ref{fig:sst-psi-2}.b1).
+The regression field of SST anomalies bears a strong resemblance with canonically positive ENSO (Bamston, Chelliah, and Goldenberg 1997).
+Indeed, there is a significant and very high correlation (0.76 (CI: 0.6 -- 0.87)) between the ONI and the Imaginary cEOF2 time series.
+Besides the Pacific ENSO-like pattern, there are positive anomalies in the western Indian Ocean and negative values in the eastern Indian Ocean, resembling a positive Indian Ocean Dipole (Saji et al.
+1999).
+Unsurprisingly, the correlation between the Imaginary cEOF2 and the DMI is 0.62 (CI: 0.38 -- 0.78).
+
+The Imaginary cEOF2 is associated with strong wave-like streamfunction anomalies emanating from the tropics (Figure~\ref{fig:sst-psi-2}.b2), both from the Central Pacific sector and the Indian Ocean.
+These atmospheric response is consistent with the effect of ENSO and the Indian Ocean Dipole on the extratropics: the SST anomalies initiate anomalous tropical convection that excites Rossby waves that propagate meridionally towards higher latitudes (Mo 2000; Cai et al.
 2011; Nuncio and Yuan 2015).
 
-Not all phases of the cEOF2 are associated with the same tropical SSTs, however.
+The cEOF2 is not associated with the same tropical SSTs for all their phases.
 Figure~\ref{fig:sst-psi-2}.d1 and d2 show that the Real cEOF2 is not associated either with any significant SST nor streamfunction anomalies in the tropics.
-The correlation between the Real cEOF2 and ENSO is also not significant (0 (CI: -0.3 -- 0.3)).
-Rows a and c in Fig.~\ref{fig:sst-psi-2} show that the intermediate phases are still associated with significant SST anomalies over the Pacific Ocean, but with with slightly different location.
-The 135º phase is associated with SST in the central Pacific, while the 45º phase is associated with SST in the eastern Pacific, which correspond roughly to the Eastern Pacific and Central Pacific ``flavours'' of ENSO (Citas).
+As a result, the correlation between the Real cEOF2 and ENSO is also not significant (0 (CI: -0.3 -- 0.3)).
+Meanwhile, Rows a and c in Fig.~\ref{fig:sst-psi-2} show that the intermediate phases are still associated with significant SST anomalies over the Pacific Ocean, but with with slightly different location.
+The 135º phase is associated with SST in the central Pacific (Fig.~\ref{fig:sst-psi-2}a.1), while the 45º phase is associated with SST in the eastern Pacific (Fig.~\ref{fig:sst-psi-2}c.1), which correspond roughly to the Central Pacific and Eastern Pacific and ``flavours'' of ENSO, respectively (Kao and Yu 2009).
+Both of these phases could be also associated to wave trains generated in the region surrounding Australia and propagates toward the extra-tropics.
 
@@
 \includegraphics{../figures/enso-phase-1.pdf}
-\caption{\label{fig:enso-phase}ONI plotted against cEOF2 phase for th 1979 -- 2019 period.
-Years with magnitude of cEOF2 greater or smaller than the 50th percentile are shown as orange diamonds and green circles and respectively.
+\caption{\label{fig:enso-phase}SON ONI values plotted against cEOF2 phase for the 1979 -- 2019 period.
+Years with magnitude of cEOF2 greater (smaller) than the 50th percentile are shown as orange diamonds (green circles).
 Black line is the fit ONI \textasciitilde{} sin(phase) computed by weighted OLS using the magnitude of the cEOF2 as weights.}
@@
 
-To better explore the relationship between tropical forcing and phase of the cEOF2, Figure~\ref{fig:enso-phase} shows the ONI index against the cEOF2 phase for each SON trimester between 1979 and 2019, highlighting years in which the magnitude of cEOF2 is above the median.
-In years with positive ONI, the phase of the cEOF2 is mostly around +90º (corresponding with positive imaginary part) and vice versa.
-In years with near neutral ENSO, the phase of the cEOF2 is much more variable.
+To better explore the relationship between tropical forcing and phases of the cEOF2, Figure~\ref{fig:enso-phase} shows the ONI index plotted against the cEOF2 phase for each SON trimester between 1979 and 2019, highlighting years in which the magnitude of cEOF2 is above the median.
+In years with positive (negative) ONI, the cEOF2 phase is mostly around 90º (-90°), corresponding with positive (negative) Imaginary component.
+In the neutral ENSO seasons, the cEOF2 phase is much more variable.
 The black line in Figure~\ref{fig:enso-phase} is a sinusoidal fit of the relationship between ONI and cEOF2 phase.
@@
 
-The correlation between the absolute magnitude of the ONI and the magnitude of the cEOF2 is 0.45 (CI: 0.17 -- 0.66).
-This relationship, however, appears to be driven only by the three years with strongest ENSO events in the period (2015, 1997, and 1982) which also coincide with the three years with strongest cEOF2 magnitude.
+The correlation between the absolute magnitude of the ONI index and the cEOF2 amplitude is 0.45 (CI: 0.17 -- 0.66).
+However, this relationship is mostly driven by the three years with strongest ENSO events in the period (2015, 1997, and 1982) which coincide with the three years with strongest cEOF2 magnitude (not shown).
 If those years are removed, the correlation becomes non-significant (0.04 (CI: -0.28 -- 0.35)).
 Furthermore, even when using all years, the Spearman correlation --which is robust to outliers-- is also non-significant (0.2, p-value= 0.21).
-Therefore, although the location of tropical SST seem to have an effect in defining the phase of the cEOF2, the relationship between the magnitude of the cEOF2 train wave and ONI remains uncertain.
+Therefore, although the location of tropical SST seem to have an effect in defining the phase of the cEOF2, the relationship between the magnitude of cEOF2 and ONI remains uncertain.
 
@@
 
-Unlike the cEOF2 case, there is no significant pattern of SST anomalies associated with either the Real or Imaginary cEOF1 (Figure~\ref{fig:sst-psi-1}).
+Unlike the cEOF2 case, there are no significant SST regressed anomalies associated with either the Real or Imaginary cEOF1 (Sup. Figure~\ref{fig:sst-psi-1}).
 Consistently, streamfunction anomalies do not show any tropical influence.
-Instead, the real and Imaginary cEOF1 are associated with zonally wave activity fluxes in the extra-tropics around 60ºS, except for an equatorward flow from the coast of Antarctica around 150ºE in the real component.
-This suggests that its variability is driven primary by the internal variability of the extra-tropics.
+Instead, the real and Imaginary cEOF1 are associated with zonally wave activity fluxes in the extra-tropics around 60ºS, except for an equatorward flow from the coast of Antarctica around 150ºE in the Real component.
+This suggests that the variability of cEOF1 is driven primary by the internal variability of the extra-tropics.
 
@@
 \includegraphics{../figures/pp-t2m-r2-1.pdf}
-\caption{\label{fig:pp-t2m-r2}Explained variance (\(r^2\)) of 2-metre temperature (row a) and precipitation (row b) with cEOF1 (column 1) and cEOF2 (column 2).}
-\end{figure}
-
-The influence of cEOFs variability in continental rainfall and 2-metre temperature in the Southern Hemisphere is also explored.
-Figure \ref{fig:pp-t2m-r2} shows the \(r^2\) between the cEOF1 and cEOF2 and 2-metre temperature and precipitation.
-For the cEOF1 there are no significant signals in precipitation and most of the temperature signal is low and restricted to the eastern coast of Antarctica.
-
-There is a strong signal between both variables and cEOF2, with explained variance greater than 50\% in some regions.
-For 2-metre temperature, there are high values in the tropical Pacific and the SPCZ, as well as the region following an arc between New Zealand and the South Atlantic.
-Over the continents, moderate values of about 30\% variance explained can be seen in southern Australia, some regions of South America and the Antarctic Peninsula.
-For precipitation, there are high values over the tropics.
-At higher latitudes, moderate values are observed over eastern Australia and some regions of southern South America.
-
-Since the cEOF1 has a relatively weak signal, we will focus on the cEOF2.
-
-Figure \ref{fig:pp-temp-2} shows regression maps of 2-meter temperature (column 1) and precipitation (column 2) upon standardised cEOF2 which, again, can be interpreted as the anomalies associates with a 1 standard deviation increase.
+\caption{\label{fig:pp-t2m-r2}Explained variance (\(r^2\) as percentage) of 2-metre temperature (row a) and precipitation (row b) anomalies by the regression upon cEOF1 (column 1) and cEOF2 (column 2).}
+\end{figure}
+
+Figure~\ref{fig:eof1-regr-t} shows the variance of the 2-meter temperature or precipitation anomalies explained variance by the mluitple linear model of Real cEOF1 and Imaginary cEOF1 (column 1), and Real cEOF2 and Imaginary cEOF2 (column 2).
+The variance explained by cEOF1 for precipitation anomalies is extremely low and also for most of the temperature anomalies, except for the eastern western of Antarctica (Fig.~ \ref{fig:eof1-regr-t}a.1).
+
+On the other hand, the cEOF2 explained variance is greater than 50\% in some regions for both variables (Fig.~\ref{fig:eof1-regr-t} column 2).
+For 2-metre temperature, there are high values in the tropical Pacific and the SPCZ, as well as the region following an arc between New Zealand and the South Atlantic, with higher values in the Southern Ocean.
+Over the continents, there are moderate values of about 30\% variance explained in southern Australia, Southern South America and the Antarctic Peninsula.
+For precipitation, there are high values over the tropics. At higher latitudes, moderate values are observed over eastern Australia and some regions of southern South America.
+
+Since the cEOF1 has a relatively weak signal in the surface variables explored here, we will only focus on the cEOF2 influence.
+
+Figure~\ref{fig:pp-temp-2} shows regression maps of 2-metre temperature (column 1) and precipitation (column 2) anomalies upon different phases of standardised cEOF2.
 
@@
 \includegraphics{../figures/pp-temp-2-1.pdf}
-\caption{\label{fig:pp-temp-2}Regression of SON mean 2-meter temperature (K, shaded) and 850hPa geopotential height (m, contours) (column 1), and precipitation (correlation, column 2) with different phases of cEOF2.
+\caption{\label{fig:pp-temp-2}Regression of SON mean 2-meter temperature (K, shaded) and 850 hPa geopotential height (m, contours) (column 1), and precipitation (correlation, column 2) upon different phases of cEOF2.
 For the 1979 -- 2019.
@@
 
-Temperature anomalies associated with the Imaginary cEOF (Fig.
-\ref{fig:pp-temp-2}.b1) show positive values in the tropical Pacific, consistent with SSTs anomalies associated with the same phase (Fig.
-\ref{fig:sst-psi-2}.b1).
-
-At higher latitudes there is a wave-like pattern of positive and negative values that coincide with the nodes of the 850hPa geopotential height regression patterns.
+Temperature anomalies associated with the Imaginary cEOF (Fig.~\ref{fig:pp-temp-2}.b1) show positive values in the tropical Pacific, consistent with SSTs anomalies associated with the same phase (Fig.~\ref{fig:sst-psi-2}.b1).
+
+At higher latitudes there is a wave-like pattern of positive and negative values that coincide with the nodes of the 850 hPa geopotential height regression patterns.
 This is consistent with temperature anomalies produced by meridional advection of temperature by the meridional winds arising from geostrophic balance.
 
-Over the continents, the Imaginary cEOF2 (Fig.
-\ref{fig:pp-temp-2}b.1) is associated with positive temperatures in southern Australia and negative temperature in southern South America and the Antarctic Peninsula.
-
-The temperatures anomalies associated with the Real cEOF2 (Fig. \ref{fig:pp-temp-2}d.1) are less extensive.
-There is no significant signal over the continents except for positive anomalies near the Antarctic Peninsula.
-
-Tropical precipitation anomalies associated with the Imaginary cEOF2 are strong, with positive anomalies in the central Pacific and western Indian, an negative anomalies in the eastern Pacific.
-This pattern is consistent with the SST anomalies shown before as the increased SST enhances tropical convection and vice versa.
-
-On the extra-tropics, the Imaginary cEOFs is correlated with drier conditions over eastern Australia; a similar signal as the one associated with ENSO Cai et al. (2011).
+Over the continents, the Imaginary cEOF2 (Fig.~\ref{fig:pp-temp-2}b.1) is associated with positive temperatures in southern Australia and negative temperature in southern South America and the Antarctic Peninsula, that are a result of the wave train described before.
+
+The temperatures anomalies associated with the Real cEOF2 (Fig.~\ref{fig:pp-temp-2}d.1) are less extensive and restricted to mid and high latitudes.\\
+Over the continents, the temperature anomalies regressions are non significant, except for positive anomalies near the Antarctic Peninsula.
+
+Tropical precipitation anomalies associated with the Imaginary cEOF2 are strong, with positive anomalies in the central Pacific and western Indian, and negative anomalies in the eastern Pacific (Fig.~\ref{fig:pp-temp-2}b.2).
+This field is consistent with the SST regression map (Fig.~\ref{fig:pp-temp-2}b.1) as the positive SST anomalies enhance tropical convection and the negative SST anomalies inhibits it.
+
+In the extra-tropics, the positive Imaginary cEOF2 is related to drier conditions over eastern Australia and the surrounding ocean, that it is similar signal as the one associated with ENSO (Cai et al.
+2011).
 However, the Imaginary cEOF2 is not the direction most correlated with precipitation in that area.
-The 135º direction (positive Imaginary + negative Real cEOF2) is associated with stronger and more extensive correlations with precipitation over Australia and New Zealand.
-This Australian signal is most likely there due to the more direct impacts of ENSO rather than on the interconnection pattern represented by the cEOF2.
-
-Over South America, the Imaginary cEOF has positive correlations with precipitation in South Eastern South America (SESA) and central Chile, and negative correlations in eastern Brazil.
-This again is a well-known springtime precipitation signature of ENSO (Cai et al.
+The 135º phase (an intermediate between positive Imaginary and negative Real cEOF2) component is associated with stronger and more extensive temporal correlations with precipitation over Australia and New Zealand.
+The influence of cEOF2 in Australian precipitation could be more related to the direct impacts of SST anomalies in the surrounded oceans rather than on the interconnection pattern represented by the cEOF2.
+
+Over South America, the Imaginary cEOF2 has positive correlations with precipitation in South Eastern South America (SESA) and central Chile, and negative correlations in eastern Brazil.
+This correlation field matches the springtime precipitation signature of ENSO (e.g.
+Cai et al.
 2020) and it is also similar to the precipitation anomalies associated with the A-SAM (@ Campitelli, Díaz, and Vera 2022).
 
-This is not surprising considering the close relationship between the ONI, the A-SAM index and the Imaginary cEOF2 shown previously, but further consolidates the identification of this mode with the PSA pattern.
-Resembling the relationship between ONI and the phase of cEOF2 (Fig.~\ref{fig:enso-phase}), there is a cEOF2 phase dependence of the precipitation anomalies in SESA (not shown).
-These variables could be related with a significant sinusoidal fit that has a statistically significant coefficient of determination of~0.3.
-
-The precipitation signal associated with the Real cEOF2 is less strong.
+This result is not surprising considering the close relationship of the Imaginary cEOF2 with both ONI and A-SAM index, which was shown previously.\\
+Furthermore, it consolidates the identification of the cEOF2 with the PSA pattern.
+Resembling the relationship between ONI and the phase of cEOF2 (Fig.~~\ref{fig:enso-phase}), there is a cEOF2 phase dependence of the precipitation anomalies in SESA (not shown).
+
+The correlation coefficients between precipitation anomalies and the Real cEOF2 (Fig.~\ref{fig:pp-temp-2}d.2) are weaker than for Imaginary cEOF2.
 There is a residual positive correlation in the equatorial eastern Pacific and small, not statistically significant positive correlations over eastern Australia and negative ones over New Zealand.
-
-Lastly, the cEOF1 doesn't have a significant signal associated with 2-metre temperature or precipitacion (Fig. \ref{fig:pp-temp-cap-1}).
 
@@
 
-In this study we assessed the Austral Spring Extratropical Southern Hemisphere zonally assymetric circulation.
+In this study we assessed the Austral Spring Extratropical Southern Hemisphere zonally asymmetric circulation.
 For this purpose, two complex indices were derived using Complex Empirical Orthogonal Functions.
@@
 
+\begin{itemize}
+\item
+  The Oceanic Niño Index is available via NOAA's Climate Prediction Center: \url{https://www.cpc.ncep.noaa.gov/products/analysis_monitoring/ensostuff/detrend.nino34.ascii.txt}.
+\item
+  The Dipole Mode Index is available via Global Climate Observing System Working Group on Surface Pressure: \url{https://psl.noaa.gov/gcos_wgsp/Timeseries/Data/dmi.had.long.data}
+\end{itemize}
+
 \appendix
@@
 \includegraphics{../figures/pp-temp-1-1.pdf}
-\caption{\label{fig:pp-temp-1}Same as Figure~\ref{fig:pp-temp-2} but for cEOF1.}
+\caption{\label{fig:pp-temp-1}Same as Figure \ref{fig:pp-temp-2} but for cEOF1.}
 \end{figure}
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-kao2009}{}}%
+Kao, Hsun-Ying, and Jin-Yi Yu. 2009. {``Contrasting {Eastern-Pacific} and {Central-Pacific Types} of {ENSO}.''} \emph{Journal of Climate} 22 (3): 615--32. \url{https://doi.org/10.1175/2008JCLI2309.1}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-katz1991}{}}%
@@
 \url{https://doi.org/10.5194/hess-11-1831-2007}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-laurel-castillo2020}{}}%
-Laurel-Castillo, J.
-A., and A.
-Valle-Levinson.
-2020.
-{``Tidal and {Subtidal Variations} in {Water Level Produced} by {Ocean-River Interactions} in a {Subtropical Estuary}.''} \emph{Journal of Geophysical Research: Oceans} 125 (7): e2018JC014116.
-\url{https://doi.org/10.1029/2018JC014116}.
 
@@
 2015.
-{``The {Influence} of the {Indian Ocean Dipole} on {Antarctic Sea Ice}*.''} \emph{Journal of Climate} 28 (7): 2682--90.
+{``The {Influence} of the {Indian Ocean Dipole} on {Antarctic Sea Ice}.''} \emph{Journal of Climate} 28 (7): 2682--90.
 \url{https://doi.org/10.1175/JCLI-D-14-00390.1}.
@@
 \url{https://doi.org/10.1175/1520-0442(2003)016\%3C2915:RLCITE\%3E2.0.CO;2}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-saji1999}{}}%
+Saji, N.
+H., B.
+N.
+Goswami, P.
+N.
+Vinayachandran, and T.
+Yamagata.
+1999.
+{``A Dipole Mode in the Tropical {Indian Ocean}.''} \emph{Nature} 401 (6751, 6751): 360--63.
+\url{https://doi.org/10.1038/43854}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-saji2003}{}}%
+Saji, N. H., and T. Yamagata. 2003. {``Possible Impacts of {Indian Ocean Dipole} Mode Events on Global Climate.''} \emph{Climate Research} 25 (2): 151--69. \url{https://doi.org/10.3354/cr025151}.
 

### 2022-03-28 analysis/paper/shceof.Rmd
Auto-cite packages
Separa la conclusión en dos secciones

@@
 The paper was rendered using knitr and rmarkdown [@xie2015; @allaire2020].
+
+```{r}
+# Hacer citas automáticamente. Manejarlo manual con zotero en una colección 
+# global no sirve porque pueden cambiar las versiones.
+
+# knitr::write_bib(c("data.table",
+#                    "base",
+#                    "metR",
+#                    "ggplot",
+#                    "ecmwfr",
+#                    "rsoi",
+#                    "knitr",
+#                    "rmarkdown"), file = "autopackages.bib")
+```
 
@@
 
-# Conclusions
+# Discussion
+
+There is some disagreement in the literature of whether the phase of the PSA pattern is affected by the location of tropical SST anomalies [@irving2016].
+We show not only that the cEOF2 tends to be in the imaginary phase (\~PSA1) when the ENSO region is warm, but that central Pacific SST anomalies tend to move the cEOF towards the negative real phase and eastern Pacific SST anomalies tend to move it towards positive real phase.
+When ENSO is neutral, the cEOF2 is still as active, but with no preferred phase.
+This mirrors the results of @cai2002, who showed that the CSIRO Model can create PSA-like variability even in the absence of ENSO forcing (with a climatological run), but that the variability of one of the PSA modes was enhanced when adding the ENSO signal.
+
+The sensitivity of the phase of the cEOF2 to the location of the tropical SST anomalies was also seen by @ciasto2015, who detected similar Rossby wave patterns associated with central Pacific and eastern Pacific SST anomalies but with a change in phase.
+
+Some aspects of our work contrast with @irving2016 analysis of the PSA.
+They didn't find neither a strong association between PSA-like variability and ENSO, nor a relationship between it and Australian precipitation.
+Part of the disagreement could be due to the fact that we focus only on SON, which is when the teleconnection pattern is strongest, instead of the whole year.
+@irving2016 use of ERA5 forectasted precipitation instead of the more observationally constrained CMAP dataset might also be an issue.
+But in light of the previous discussion on the importance of the phase of the cEOF2 in its relationship with ENSO and with precipitation impacts, a big difference could be fact that @irving2016's methodology is insensitive to phase changes of the PSA-like variability.
+
+To the extent that the cEOF2 can be identified with the PSA, what is the advantage of the technique proposed here?
+The spatial fields that the cEOF2 index describes are in quadrature by construction, which makes it possible to derive a proper amplitude and phase.
+Furthermore, because of the way they are constructed, they are not forced to be orthogonal to the SAM.
+This allows us to show that Imaginary cEOF2 is closely related to the Southern Annular Mode in the troposphere.
+In fact, it is so highly correlated to the the zonally asymmetric portion of the SAM that we speculate that that the asymmetric component of the SAM might actually a statistical contamination of the PSA mode.
+
+# Conclusion
 
@@
 
-There is some disagreement in the literature of whether the phase of the PSA pattern is affected by the location of tropical SST anomalies [@irving2016].
-We show not only that the cEOF2 tends to be in the imaginary phase (\~PSA1) when the ENSO region is warm, but that central Pacific SST anomalies tend to move the cEOF towards the negative real phase and eastern Pacific SST anomalies tend to move it towards positive real phase.
-When ENSO is neutral, the cEOF2 is still active, but with no particular preferred phase.
-This mirrors the results of @cai2002, who showed that the CSIRO Model can create PSA-like variability even in the absence of ENSO forcing (with a climatological run), but the variability of one of the PSA modes was enhanced when adding the ENSO signal.
-The magnitude of the cEOF2 appears not to be related to the magnitude of ENSO, except for the three strongest ENSO events in the period which coincide with the three strongest cEOF2 years.
-
-The sensitivity of the phase of the cEOF2 to the location of the tropical SST anomalies was also seen by @ciasto2015, who detected similar Rossby wave patterns associated with central Pacific and eastern Pacific SST anomalies but with a change in phase.
-
-We further show that the Imaginary cEOF2 is closely related to the Southern Annular Mode in the troposphere.
-In fact, it has a very close resemblance to the zonally asymmetric portion of the SAM.
-This raises the possibility that the asymmetric component of the SAM is actually a statistical contamination of the PSA mode.
+The phase of the cEOF2 is related to the the variability of ENSO, but its magnitude appears not to be.
+Except for the three strongest ENSO events in the period which coincide with the three strongest cEOF2 years.
 
 Precipitation anomalies in South America associated with the Imaginary cEOF2 show a clear ENSO-like impact, with positive anomalies in South-Eastern South America, negative anomalies in Southern Brazil and positive anomalies in central Chile.
-Precipitation anomalies associated with the Real cEOF2 are low and not statistically significant, showing that the Imaginary phase is optimally aligned with the direction of maximum precipitation impacts.
+Precipitation anomalies associated with the Real cEOF2 are low and not statistically significant in this region, showing that the Imaginary phase is optimally aligned with the direction of maximum precipitation impacts there.
 
 On the other hand, over Australia, both the Real and Imaginary phase are associated with significant precipitation anomalies, and we further show that the direction of maximum impact is not aligned with our chosen rotation of cEOF2.
-However, this underscores the benefit of using complex EOF, since it would be trivial to rotate it.
-That is, look for the rotation of complex EOF that maximises the relationship and use that as the relevant index for study or forecast.
-
-Some aspects of our work are in direct contrast with @irving2016 analysis of the PSA.
-They didn't find neither a strong association between PSA-like variability and ENSO, nor a relationship between it and Australian precipitation.
-Part of the disagreement could be due to the fact that we focus data only for SON, which is when the teleconnection pattern is strongest, instead of the whole year.
-@irving2016 use of ERA5 forectasted precipitation instead of the more observationally constrained CMAP dataset might also be an issue.
-But in light of the previous discussion on the importance of the phase of the cEOF2 in its relationship with ENSO and with precipitation impacts, a big difference could be fact that @irving2016's methodology is insensitive to phase changes of the PSA-like variability.
-
-To the extent that the cEOF2 can be identified with the PSA, what is the advantage of the technique proposed here?
-The spatial fields that the cEOF2 index describes are in quadrature by construction, which makes it possible to derive a proper amplitude and phase.
-Furthermore, because of the way they are constructed, they are not forced to be orthogonal to the SAM.
-This makes it possible to study their relationship with the SAM free of the statistical contamination that comes with the usual definition of the PSA as the second and third EOFs.
+An analysis of the precipitation impacts on this regions could benefit for a differnet chosen rotation of the cEOF.
 

### 2022-03-28 analysis/paper/shceof.Rmd
Render

@@
 output: 
-  # bookdown::word_document2: default
+  bookdown::word_document2: default
   bookdown::pdf_book: 

### 2022-03-28 analysis/paper/shceof.tex
Render

@@
 
-\hypertarget{conclusions}{%
-\section{Conclusions}\label{conclusions}}
+\hypertarget{discussion}{%
+\section{Discussion}\label{discussion}}
+
+There is some disagreement in the literature of whether the phase of the PSA pattern is affected by the location of tropical SST anomalies (Irving and Simmonds 2016).
+We show not only that the cEOF2 tends to be in the imaginary phase (\textasciitilde PSA1) when the ENSO region is warm, but that central Pacific SST anomalies tend to move the cEOF towards the negative real phase and eastern Pacific SST anomalies tend to move it towards positive real phase.
+When ENSO is neutral, the cEOF2 is still as active, but with no preferred phase.
+This mirrors the results of Cai and Watterson (2002), who showed that the CSIRO Model can create PSA-like variability even in the absence of ENSO forcing (with a climatological run), but that the variability of one of the PSA modes was enhanced when adding the ENSO signal.
+
+The sensitivity of the phase of the cEOF2 to the location of the tropical SST anomalies was also seen by Ciasto, Simpkins, and England (2015), who detected similar Rossby wave patterns associated with central Pacific and eastern Pacific SST anomalies but with a change in phase.
+
+Some aspects of our work contrast with Irving and Simmonds (2016) analysis of the PSA.
+They didn't find neither a strong association between PSA-like variability and ENSO, nor a relationship between it and Australian precipitation.
+Part of the disagreement could be due to the fact that we focus only on SON, which is when the teleconnection pattern is strongest, instead of the whole year.
+Irving and Simmonds (2016) use of ERA5 forectasted precipitation instead of the more observationally constrained CMAP dataset might also be an issue.
+But in light of the previous discussion on the importance of the phase of the cEOF2 in its relationship with ENSO and with precipitation impacts, a big difference could be fact that Irving and Simmonds (2016)'s methodology is insensitive to phase changes of the PSA-like variability.
+
+To the extent that the cEOF2 can be identified with the PSA, what is the advantage of the technique proposed here?
+The spatial fields that the cEOF2 index describes are in quadrature by construction, which makes it possible to derive a proper amplitude and phase.
+Furthermore, because of the way they are constructed, they are not forced to be orthogonal to the SAM.
+This allows us to show that Imaginary cEOF2 is closely related to the Southern Annular Mode in the troposphere.
+In fact, it is so highly correlated to the the zonally asymmetric portion of the SAM that we speculate that that the asymmetric component of the SAM might actually a statistical contamination of the PSA mode.
+
+\hypertarget{conclusion}{%
+\section{Conclusion}\label{conclusion}}
 
@@
 
-There is some disagreement in the literature of whether the phase of the PSA pattern is affected by the location of tropical SST anomalies (Irving and Simmonds 2016).
-We show not only that the cEOF2 tends to be in the imaginary phase (\textasciitilde PSA1) when the ENSO region is warm, but that central Pacific SST anomalies tend to move the cEOF towards the negative real phase and eastern Pacific SST anomalies tend to move it towards positive real phase.
-When ENSO is neutral, the cEOF2 is still active, but with no particular preferred phase.
-This mirrors the results of Cai and Watterson (2002), who showed that the CSIRO Model can create PSA-like variability even in the absence of ENSO forcing (with a climatological run), but the variability of one of the PSA modes was enhanced when adding the ENSO signal.
-The magnitude of the cEOF2 appears not to be related to the magnitude of ENSO, except for the three strongest ENSO events in the period which coincide with the three strongest cEOF2 years.
-
-The sensitivity of the phase of the cEOF2 to the location of the tropical SST anomalies was also seen by Ciasto, Simpkins, and England (2015), who detected similar Rossby wave patterns associated with central Pacific and eastern Pacific SST anomalies but with a change in phase.
-
-We further show that the Imaginary cEOF2 is closely related to the Southern Annular Mode in the troposphere.
-In fact, it has a very close resemblance to the zonally asymmetric portion of the SAM.
-This raises the possibility that the asymmetric component of the SAM is actually a statistical contamination of the PSA mode.
+The phase of the cEOF2 is related to the the variability of ENSO, but its magnitude appears not to be.
+Except for the three strongest ENSO events in the period which coincide with the three strongest cEOF2 years.
 
 Precipitation anomalies in South America associated with the Imaginary cEOF2 show a clear ENSO-like impact, with positive anomalies in South-Eastern South America, negative anomalies in Southern Brazil and positive anomalies in central Chile.
-Precipitation anomalies associated with the Real cEOF2 are low and not statistically significant, showing that the Imaginary phase is optimally aligned with the direction of maximum precipitation impacts.
+Precipitation anomalies associated with the Real cEOF2 are low and not statistically significant in this region, showing that the Imaginary phase is optimally aligned with the direction of maximum precipitation impacts there.
 
 On the other hand, over Australia, both the Real and Imaginary phase are associated with significant precipitation anomalies, and we further show that the direction of maximum impact is not aligned with our chosen rotation of cEOF2.
-However, this underscores the benefit of using complex EOF, since it would be trivial to rotate it.
-That is, look for the rotation of complex EOF that maximises the relationship and use that as the relevant index for study or forecast.
-
-Some aspects of our work are in direct contrast with Irving and Simmonds (2016) analysis of the PSA.
-They didn't find neither a strong association between PSA-like variability and ENSO, nor a relationship between it and Australian precipitation.
-Part of the disagreement could be due to the fact that we focus data only for SON, which is when the teleconnection pattern is strongest, instead of the whole year.
-Irving and Simmonds (2016) use of ERA5 forectasted precipitation instead of the more observationally constrained CMAP dataset might also be an issue.
-But in light of the previous discussion on the importance of the phase of the cEOF2 in its relationship with ENSO and with precipitation impacts, a big difference could be fact that Irving and Simmonds (2016)'s methodology is insensitive to phase changes of the PSA-like variability.
-
-To the extent that the cEOF2 can be identified with the PSA, what is the advantage of the technique proposed here?
-The spatial fields that the cEOF2 index describes are in quadrature by construction, which makes it possible to derive a proper amplitude and phase.
-Furthermore, because of the way they are constructed, they are not forced to be orthogonal to the SAM.
-This makes it possible to study their relationship with the SAM free of the statistical contamination that comes with the usual definition of the PSA as the second and third EOFs.
+An analysis of the precipitation impacts on this regions could benefit for a differnet chosen rotation of the cEOF.
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-albers2020}{}}%
-Albers, Sam, and Elio Campitelli. 2020. {``Rsoi: {Import Various Northern} and {Southern Hemisphere Climate Indices}.''}
+Albers, Sam, and Elio Campitelli. 2020. \emph{Rsoi: {Import Various Northern} and {Southern Hemisphere Climate Indices}} (version 0.5.4). \url{https://CRAN.R-project.org/package=rsoi}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-allaire2020}{}}%
-Allaire, J. J., Yihui Xie {[}aut, cre, Jonathan McPherson, Javier Luraschi, Kevin Ushey, Aron Atkins, et al. 2020. {``Rmarkdown: {Dynamic Documents} for {R}.''}
+Allaire, J.
+J., Yihui Xie {[}aut, cre, Jonathan McPherson, Javier Luraschi, Kevin Ushey, Aron Atkins, et al.
+2020.
+\emph{Rmarkdown: {Dynamic Documents} for {R}} (version 2.4).
+\url{https://CRAN.R-project.org/package=rmarkdown}.
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-campitelli2020}{}}%
-Campitelli, Elio. 2020. {``{metR}: {Tools} for {Easier Analysis} of {Meteorological Fields}.''}
+Campitelli, Elio. 2020. \emph{{metR}: {Tools} for {Easier Analysis} of {Meteorological Fields}} (version 0.7.0). \url{https://CRAN.R-project.org/package=metR}.
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-dowle2020}{}}%
-Dowle, Matt, and Arun Srinivasan. 2020. {``Data.table: {Extension} of 'Data.frame'.''}
+Dowle, Matt, and Arun Srinivasan. 2020. \emph{Data.table: {Extension} of 'Data.frame'} (version 1.13.0). \url{https://CRAN.R-project.org/package=data.table}.
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-hufkens2020}{}}%
-Hufkens, Koen. 2020. {``Ecmwfr: {Programmatic} Interface to the Two {European Centre} for {Medium-Range Weather Forecasts API} Services.''}
+Hufkens, Koen.
+2020.
+\emph{Ecmwfr: {Programmatic} Interface to the Two {European Centre} for {Medium-Range Weather Forecasts API} Services} (version 1.3.0).
+\url{http://doi.org/10.5281/zenodo.2647541}.
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-rcoreteam2020}{}}%
-R Core Team. 2020. \emph{R: {A} Language and Environment for Statistical Computing}. Manual. {Vienna, Austria}: {R Foundation for Statistical Computing}.
+R Core Team. 2020. \emph{R: {A} Language and Environment for Statistical Computing}. Manual. {Vienna, Austria}: {R Foundation for Statistical Computing}. \url{https://www.R-project.org/}.
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-xie2015}{}}%
-Xie, Yihui. 2015. \emph{Dynamic Documents with {R} and Knitr}. Second. {Boca Raton, Florida}: {Chapman and Hall/CRC}.
+Xie, Yihui. 2015. \emph{Dynamic Documents with {R} and Knitr}. 2nd ed. {Boca Raton, Florida}: {Chapman and Hall/CRC}. \url{https://yihui.org/knitr/}.
 

### 2022-03-28 analysis/paper/shceof.Rmd
Render

--- 2022-03-28 analysis/paper/shceof.Rmd
@@
 ```{r read-indices}
-enso <- rsoi::download_oni(TRUE, data_path("raw", "oni.csv")) %>%
-  as.data.table() %>%
+enso <- ENSO() %>%
+  fread() %>% 
   .[, .(time = lubridate::as_datetime(Date), oni = ONI)] %>% 
@@
 
-dmi <- rsoi::download_dmi(TRUE, data_path("raw", "dmi.csv")) %>%
-  as.data.table() %>%
+dmi <- DMI() %>%
+  fread() %>% 
   .[, .(time = lubridate::as_datetime(Date), dmi = DMI)] %>% 

### 2022-04-05 analysis/paper/shceof.Rmd
Agrega comentarios de Leandro 2

@@
 These patterns influence precipitation anomalies in South America [@mo2001].
-Although these patterns are usually derived by applying EOF to temporal anomalies, [@raphael2003] also applied EOF methods specifically to zonal anomalies.
-More recently, [@irving2016] proposed a novel methodology for objectively identifying the PSA pattern using Fourier decomposition.
+Although these patterns are usually derived by applying EOF to temporal anomalies, @raphael2003 also applied EOF methods specifically to zonal anomalies.
+More recently, @irving2016 proposed a novel methodology for objectively identifying the PSA pattern using Fourier decomposition.
 
@@
 
-A third methodology commonly used consists on identifying particular features of interest and creating indices using simple methods such as averages and differences.
-Examples of this methodology are the SAM Index of [@gong1999], the SH wave 3 activity index defined by @raphael2004 and the SH zonally asymmetric circulation index from @hobbs2010.
+A third methodology commonly used to describe circulation anomalies consists on identifying particular features of interest and creating indices using simple methods such as averages and differences.
+Examples of this methodology are the SAM Index of @gong1999, the SH wave 3 activity index defined by @raphael2004 and the SH zonally asymmetric circulation index from @hobbs2010.
 These derived methods are grounded on other methods such as Fourier decomposition or EOF to identify the centres of action for the described phenomena and can be useful to characterise features that are not readily apparent with these methods.
@@
 
-The general goal of this study is to improve the description and understanding of the zonally asymmetric extratropical SH circulation using complex Empirical Orthogonal Functions, which allow to describe phase varying planetary waves with variable amplitudes along a latitude circle.
+The general goal of this study is to improve the description and understanding of the zonally asymmetric extratropical SH circulation using cEOF, which allow to describe phase varying planetary waves with variable amplitudes along a latitude circle.
 In addition, it is proposed to expand the knowledge of the simultaneous behaviour of SH asymmetric circulation in the troposphere and the stratosphere.
 
-We restrict this paper to the September-October-November (SON) trimester.
-We chose this season because it showed the strongest and more clearly interpretable results in a preliminary analysis of the four season.
-This is is consistent with previous research that shows that this season experiences a maximum in tropical teleconnections [@cazes-boezio2003a] and stratosphere-troposphere interactions [@lim2018].
-
-In Section \@ref(methods) we describe the methods.
-In Section \@ref(spatial) we analyse the spatial patterns of each complex EOF.
-In Section \@ref(regressions) we study the spatial regressions with geopotential height and temperature anomalies.
-In Section \@ref(other-variables) we analyse the relationship between cEOF1 and Ozone, and between cEOF2 with the SAM and the PSA.
-In Section \@ref(tropical) we study tropical forcings that explain the variability of each cEOF.
-Finally, in Section \@ref(precipitation) we show the relationship between these modes of variability and precipitation anomalies in South America and Oceania.
+We restrict this paper to the spring September-October-November (SON) trimester.
+This season experiences a maximum in tropical teleconnections [@cazes-boezio2003a] and stratosphere-troposphere interactions [@lim2018].
+
+In Section\ \@ref(methods) we describe the methods.
+In Section\ \@ref(spatial) we analyse the spatial patterns of each complex EOF.
+In Section\ \@ref(regressions) we study the spatial regressions with geopotential height, temperature and ozone anomalies.
+In Section\ \@ref(psa) and \@ref(sam) we analyse the relationship between cEOF2 the PSA and SAM modes.
+In Section\ \@ref(tropical) we study tropical forcings that explain the variability of each cEOF.
+Finally, in Section \@ref(precipitation) we show the relationship between these modes of variability and precipitation and surface temperature anomalies in South America and Oceania.
 
@@
 We derived streamfunction at 200 hPa from ERA5 vorticity using the FORTRAN subroutine FISHPACK [@fishpack] and we computed horizontal wave activity fluxes following @plumb1985.
-Sea Surface Temperature (SST) monthly fields are from Extended Reconstructed Sea Surface Temperature (ERSST) v5 [@huang2017] and precipitation monthly data from the CPC Merged Analysis of Precipitation [@cmap], with a 2.5\degree resolution in latitude and longitude.
+Sea Surface Temperature (SST) monthly fields are from Extended Reconstructed Sea Surface Temperature (ERSST) v5 [@huang2017] and precipitation monthly data from the CPC Merged Analysis of Precipitation [@cmap], with a 2\degree and 2.5\degree resolution in latitude and longitude respectively.
 The rainfall gridded dataset is based on information from different sources such as rain gauge observations, satellite inferred estimations and the NCEP-NCAR reanalysis, and it is available since 1979 to present.
@@
 Since any choice is arbitrary and equally valid, we chose the phase of each cEOF so that the real or imaginary parts are aligned with meaningful phases in our analysis.
-
-Preliminary analysis showed that the first cEOF was closely related to the the zonal wave 1 of Total Ozone Column and the second cEOF was closely related to ENSO.
-Therefore, to aid in the interpretation, we chose the phases of each cEOF so that these relationships are aligned with one of the complex components.
-
-We chose the phase of cEOF1 so that the time series corresponding to the real part has the maximum correlation with the zonal wave 1 of Total Ozone Column between `r knitr::combine_words(LatLabel(o3wave_lats))`.
-And we chose the phase of cEOF2 so that the coefficient of determination between the Oceanic Niño Index [@bamston1997] and the real part is minimized, which also nearly maximizes the correlation with the imaginary part.
-
 This procedure does not create spurious correlations, it only takes an existing relationship and aligns it with a specific phase.
+
+Preliminary analysis showed that the first cEOF was closely related to the zonal wave 1 of Total Ozone Column and the second cEOF was closely related to ENSO.
+Therefore, to aid in the interpretation we chose the phase of cEOF1 so that the time series corresponding to the real part has the maximum correlation with the zonal wave 1 of Total Ozone Column between `r knitr::combine_words(LatLabel(o3wave_lats))`.
+Similarly, we chose the phase of cEOF2 so that the coefficient of determination between the Oceanic Niño Index [@bamston1997] and the real part is minimized, which also nearly maximizes the correlation with the imaginary part.
 
@@
 All graphics are made using ggplot2 [@wickham2009].
-We downloaded data from reanalysis using the ecmwfr package [@hufkens2020] and indices of ENSO and Indian Ocean Dipole with the rsoi package [@albers2020].
+We downloaded data from reanalysis using the ecmwfr package [@hufkens2020] and indices of ENSO and Indian Ocean Dipole (IOD) with the rsoi package [@albers2020].
 The paper was rendered using knitr and rmarkdown [@xie2015; @allaire2020].
@@
 The shading (contours) corresponds to real (imaginary) part.
-Arbitrary units.", fig.width=width_column, fig.height=width_column*1.2}
+Arbitrary units.
+In parenthesis, the proportion of variance (with respect to the zonal mean) explained.", fig.width=width_column, fig.height=width_column*1.2}
 
@@
 phase) is controlled by the magnitude and phase of the temporal cEOF.
-The wave patterns described by these cEOFs match the patterns seen in the traditional EOFs of Figure \@ref(fig:eof-naive).
+The wave patterns described by these cEOFs match the patterns seen in the standard EOFs of Figure \@ref(fig:eof-naive).
 
@@
 In particular, the dominant structure at both levels is a wave 3 but with larger amplitude in the pacific sector.
-This modulated amplitude more evident at 200 hPa.
+This modulated amplitude is more evident at 200 hPa.
 There is no apparent phase shift with height but the amplitude of the pattern is greatly reduced in the stratosphere, suggesting that this barotropic mode represents mainly tropospheric variability.
@@
 Both cEOFs show year-to-year variability but show no evidence of decadal variability (Fig. \@ref(fig:extended-series)).
-Because the geopotential fields that enter into the cEOFs algorithm are anomalies with respect to the zonal mean instead of the time mean, the cEOFs have non zero temporal mean.
-However, cEOF2's mean is almost zero, which indicates that only cEOF1 includes variability that significantly projects onto the mean zonally anomalous field.
+Because the geopotential fields that enter into the cEOFs algorithm are anomalies with respect to the zonal mean instead of the time mean, the cEOFs time series have non zero temporal mean. 
+However, cEOF2 temporal mean is almost zero, which indicates that only cEOF1 includes variability that significantly projects onto the mean zonally anomalous field.
 
@@
 ## cEOFs Regression maps {#regressions}
+
+### Geopotential
 
@@
 At 200 hPa (Figure \@ref(fig:eof1-regr-gh) row b) the Real cEOF1 shows a single area of positive anomalies spanning West Antarctica surrounded by opposite anomalies in lower latitudes with is centre shifted slightly eastward compared with the upper level anomalies.
-The Imaginary cEOF2.
-The Imaginary cEOF1 shows a much more zonally symmetrical pattern resembling the negative SAM phase [@fogt2020].
+The Imaginary cEOF1 shows a much more zonally symmetrical pattern resembling the negative SAM phase [e.g. @fogt2020].
 Both components of cEOF1 could be associated to zonal anomalies in the upper troposphere jet stream intensity.
@@
 
-## cEOFs relationship with known modes of variability {#other-variables}
-
-### PSA
-
-(ref:psa-eof2-cap) Correlation coefficients (r) between cEOF2 components and the PSA1 and PSA2 modes computed as @mo2001 for the 1979 -- 2019 period.
-95% confidence intervals in parenthesis.
-p-values lower than 0.01 in bold.
-
-```{r psa-eof2, fig.cap = "(ref:psa-eof2-cap)", fig.width=7, fig.height=3, fig.env="figure*"}
-psa2 <- psa$left[PC == "PSA2"][season(time) == "SON"]
-cEOF2 <- ceof$eof[[1]]$left[cEOF == "cEOF2"] %>% 
-  copy() %>% 
-  .[, cEOF := NULL] %>% 
-  .[]
-
-rotations_psa <- lapply(angles, function(a) {
-  cEOF2 %>% 
-    copy() %>% 
-    .[, hgt := rotate(hgt, a)] %>% 
-    .[psa2, on = "time"] %>% 
-    .[, cor(value, Re(hgt))] 
-})
-
-psa_angle <- angles[which.min(unlist(rotations_psa))]
-
-make_numeric <- function(x) {
-  x <- strsplit(x, "\ ")
-  vapply(x, function(x) as.numeric(x[[1]]), numeric(1))
-}
-
-background_palette <- grDevices::colorRampPalette(c(scales::muted("blue"),
-                                                    "white",
-                                                    scales::muted("red")),
-                                                  space = "Lab")
-
-table <- cEOF2[psa_time, on = "time"] %>% 
-  .[PC != "SAM"] %>% 
-  copy() %>% 
-  sep_ReIm() %>% 
-  .[, correlate(value, hgt), by = .(PC, part)]
-
-p.val <- table %>% 
-  dcast(PC ~ part, value.var = "p.value")
-
-table %>% 
-  dcast(PC ~ part, value.var = "text") %>% 
-  kbl2(booktabs = TRUE, caption = "(ref:psa-eof2-cap)") %>%
-  column_spec2(2:3, bold = p.val[[.col]] < 0.01) %>%
-  add_header_above(c("", "cEOF2" = 2)) %>%
-  kable_classic() 
-```
-
-Given the similarity between the cEOF2 regression maps (Fig. \@ref(fig:eof2-regr-gh)) and documented PSAs we study the relationship between these modes.
-Table \@ref(tab:psa-eof2) shows the correlations between the two PSA indices and the time series for Real and Imaginary phases of cEOF2.
-As anticipated by Figure \@ref(fig:eof2-regr-gh), there is a strong positive correlation between PSA1 and Imaginary cEOF2, and also between PSA2 and Real cEOF2.
-On the other hand, there is no significant relationship between PSA1 and Real cEOF2, and between PSA2 and Imaginary cEOF2.
-As a result, cEOF2 represents well both the spatial structure and temporal evolution of the PSA modes, so it is possible to make an association between its two phases and the two PSA modes.
-The same phase election for cEOF2 that maximises the relationship between ENSO and Imaginary cEOF2, also maximises the association between cEOF2 components and PSA modes.
-
-(ref:phase-histogram-cap) Histogram of phase distribution of cEOF2 phase for the 1979 -- 2019 period.
-Bins are centred at 90º, 0º, -90º, -180º with a binwidth of 90º.
-The small vertical lines near the horizontal axis mark the observations.
-
-```{r phase-histogram, fig.cap = "(ref:phase-histogram-cap)", fig.width=width_column*.8, fig.height=2.5}
-N <- nrow(cut(ceof$eof[[1]], 2)$left)
-plot <- cut(ceof$eof[[1]], 2)$left %>% 
-  .[, arg := Arg(hgt)] %>% 
-  .[, w := Mod(hgt)/mean(Mod(hgt))] %>% 
-  qwrap(arg = c(-pi, pi) ~  c(-pi - 1/4*pi, pi - 1/4*pi)) %>% 
-  ggplot(aes(arg)) +
-  geom_histogram(binwidth = 90*pi/180, center = 0, size = 0.2,
-                 color = "black", fill = "gray70") + 
-  annotate("text",
-           label = c("- Real", "- Imaginary", "+ Real", "+ Imaginary"),
-           color = "black", size = 3,
-           y = 0.5, x = c(-pi, -pi/2, 0, pi/2), angle = 90, hjust = 0) +
-  geom_rug() +
-  scale_y_continuous("Number of years", 
-                     sec.axis = sec_axis(~.x/N, 
-                                         name = "Proportion of years", 
-                                         labels = scales::percent)) +
-  scale_x_continuous("Phase", trans = scales::reverse_trans(),
-                     breaks = seq(-pi - 1/4*pi, pi - 1/4*pi, by = 45*pi/180),
-                     labels = function(x) paste0(round(x*180/pi, 5), "º")) +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
-  no_grid
-
-imaginary <- sum(ggplot_build(plot)$data[[1]][c(1, 3), ]$count)/N
-plot
-```
-
-Figure \@ref(fig:phase-histogram) shows an histogram that counts the number of SON seasons in which the cEOF2 phase was close to each of the four particular phases (positive/negative of real/imaginary component), with the observations for each season marked as rugs on the horizontal axis.
-In `r scales::percent(imaginary)` of seasons cEOF2 has a phase similar to either the negative or positive imaginary phase, making the imaginary phase the most common phase. 
-This is also, by construction, the phase that is most correlated with ENSO.
-Therefore, the Imaginary cEOF2 explains more variance than the Real cEOF2 and conventional EOF analysis will tend to separate the two.
-
-This phase preferences is in agreement with @irving2016, who found a bimodal distribution to PSA-like variability (compare our Figure \@ref(fig:phase-histogram) with their Figure 6).
-
-### SAM
-
-(ref:sam-eof-vertical-cap) Coefficient of determination ($r^2$) between each component of cEOFs and the SAM, Asymmetric SAM (A-SAM) and Symmetric SAM (S-SAM) indices computed at each level for the 1979 -- 2019 period.
-Thick lines represent estimates with p-value < 0.01 corrected for False Detection Rate [@benjamini1995].
-
-```{r sam-eof-vertical, fig.cap = "(ref:sam-eof-vertical-cap)", fig.height=4, fig.width=width_column}
-time <- ceof[, eof[[1]]$left, by = .(season)] %>%
-  sep_ReIm() %>%
-  .[, hgt := scale(hgt), by = .(cEOF, part)]
-
-sam_r2 <- sams %>%
-  .[time, on = .NATURAL, allow.cartesian = TRUE] %>%
-  na.omit() %>% 
-  .[, correlate(hgt, estimate), by = .(lev, term, part, cEOF)]
-
-max_r2 <- sam_r2[part == "Imaginary"] %>% 
-  .[, .SD[which.max(estimate^2)], by = .(cEOF, term)]
-
-sam_r2 %>%
-  .[, p.value_a := p.adjust(p.value, "fdr")] %>% 
-  ggplot(aes(lev, estimate^2)) +
-  geom_vline(xintercept = c(50, 200),  size = 0.2, color = "gray50") +
-  # geom_point(data = ~.x[p.adjust(p.value, "fdr") < 0.01], aes(color = term)) +
-  geom_line(data = ~copy(.x)[p.value_a > 0.01, estimate := NA], 
-            aes(colour = stage(term, after_scale = scales::alpha(colour, 0.5))), 
-            size = 2) +
-  geom_line(aes(color = term, group = term)) +
-
-  scale_x_level("Level (hPa)", minor_breaks = NULL, 
-                sec.axis = sa_height_axis(breaks = seq(0, 50, by = 10),
-                                          labels = function(x) paste0("~", x, " km"))) +
-  scale_y_continuous(r2, limits = c(0, 1), 
-                     guide = guide_axis(check.overlap = TRUE)) +
-  scale_size_manual("p-value", values = c("TRUE" = 1, 
-                                          "FALSE" = 0.5), 
-                    guide = "none") +
-  scale_color_brewer(NULL, palette = "Dark2", labels = c(full = "SAM",
-                                                         asym = "A-SAM",
-                                                         sym = "S-SAM")) +
-  facet_grid(part ~ cEOF) +
-  coord_flip() +
-  tag_facets("rc") +
-  panel_background +
-  theme(panel.spacing = grid::unit(1, "lines"), 
-        strip.placement = "outside")
-```
-
-We now explore the relationship between SAM and the cEOFs motivated by the resemblance between cEOFs regression maps and SAM patterns shown in the previous section.
-To that end, we computed the coefficient of determination between the cEOFs time series and the three SAM indices (SAM, A-SAM and S-SAM) defined by @campitelli2022 at each vertical level (Fig.\ \@ref(fig:sam-eof-vertical)).
-The SAM index is statistically significant correlated with the Real cEOF1 in all levels, and with the Imaginary cEOF1 and Imaginary cEOF2 in the troposphere.
-On the other hand, correlations between SAM and the Real cEOF2 are non-significant.
-
-The relationship between the SAM and cEOF1 in the troposphere is explained entirely by the zonally symmetric component of the SAM as shown by the high correlation with the S-SAM below 100 hPa and the low and statistically non-significant correlations between the A-SAM and either the Real or Imaginary cEOF1.
-In the stratosphere, the Real cEOF1 is correlated with both A-SAM and S-SAM, while the Imaginary cEOF1 is highly correlated only with the A-SAM. 
-These correlations are consistent with the regression maps of geopotential height in Figure\ \@ref(fig:eof1-regr-gh) and their comparison with those obtained for SAM, A-SAM and S-SAM by @campitelli2022.
-
-In the case of Imaginary cEOF2, the correlation between SAM and Imaginary cEOF2 for the troposphere is associated to the asymmetric variability of the SAM. 
-Indeed, the Imaginary cEOF2 shares up to `r scales::percent(max_r2[cEOF == "cEOF2" & term == "asym"]$estimate^2, 2)` variance with the A-SAM and only `r scales::percent(max_r2[cEOF == "cEOF2" & term == "sym"]$estimate^2, 2)` at most with the S-SAM (Figure \@ref(fig:sam-eof-vertical).b2).
-Such extremely high correlation between A-SAM and Imaginary cEOF2 suggests that the methodology proposed in this work is able to characterise the zonally asymmetric component of the SAM described previously by @campitelli2022.
-
-### Temperature and Ozone
+### Temperature and Ozone {#temp-ozone}
 
@@
 In the photochemically-dominated upper stratosphere, cold temperatures inhibit the destruction of ozone, explaining the opposite behaviour for both variables.
-On the other hand, in the advectively-dominated lower stratosphere, ozone anomalies are 90º out of phase with horizontal and vertical transport, which are in addition 90º out of phase with temperature anomalies, resulting in same sign anomalies for the response of both variables
+On the other hand, in the advectively-dominated lower stratosphere, ozone anomalies are 90º out of phase with horizontal and vertical transport, which are in addition 90º out of phase with temperature anomalies, resulting in same sign anomalies for the response of both variables [@hartmann1979; @wirth1993; @smith1995].
 
@@
 
+(ref:wave1-o3-cap) Relationship between cEOF1 amplitude and phase with amplitude and phase of the zonal wave 1 in Total Ozone Column averaged between `r knitr::combine_words(LatLabel(o3wave_lats))` for the period 1979 -- 2019.
+
 The regression maps of  TOC anomalies upon cEOF1 (Fig.\ \@ref(fig:o3-regr)) show zonal wave 1 patterns  associated with both components of cEOF1.
@@
 
-(ref:wave1-o3-cap) Relationship between cEOF1 amplitude and phase with amplitude and phase of the zonal wave 1 in Total Ozone Column averaged between `r knitr::combine_words(LatLabel(o3wave_lats))` for the period 1979 -- 2019.
-
-## Tropical sources {#tropical}
+## PSA  {#psa}
+
+(ref:psa-eof2-cap) Correlation coefficients (r) between cEOF2 components and the PSA1 and PSA2 modes computed as @mo2001 for the 1979 -- 2019 period.
+95% confidence intervals in parenthesis.
+p-values lower than 0.01 in bold.
+
+```{r psa-eof2, fig.cap = "(ref:psa-eof2-cap)", fig.width=7, fig.height=3, fig.env="figure*"}
+psa2 <- psa$left[PC == "PSA2"][season(time) == "SON"]
+cEOF2 <- ceof$eof[[1]]$left[cEOF == "cEOF2"] %>% 
+  copy() %>% 
+  .[, cEOF := NULL] %>% 
+  .[]
+
+rotations_psa <- lapply(angles, function(a) {
+  cEOF2 %>% 
+    copy() %>% 
+    .[, hgt := rotate(hgt, a)] %>% 
+    .[psa2, on = "time"] %>% 
+    .[, cor(value, Re(hgt))] 
+})
+
+psa_angle <- angles[which.max(unlist(rotations_psa))]
+
+make_numeric <- function(x) {
+  x <- strsplit(x, "\ ")
+  vapply(x, function(x) as.numeric(x[[1]]), numeric(1))
+}
+
+background_palette <- grDevices::colorRampPalette(c(scales::muted("blue"),
+                                                    "white",
+                                                    scales::muted("red")),
+                                                  space = "Lab")
+
+table <- cEOF2[psa_time, on = "time"] %>% 
+  .[PC != "SAM"] %>% 
+  copy() %>% 
+  sep_ReIm() %>% 
+  .[, correlate(value, hgt), by = .(PC, part)]
+
+p.val <- table %>% 
+  dcast(PC ~ part, value.var = "p.value")
+
+table %>% 
+  dcast(PC ~ part, value.var = "text") %>% 
+  kbl2(booktabs = TRUE, caption = "(ref:psa-eof2-cap)") %>%
+  column_spec2(2:3, bold = p.val[[.col]] < 0.01) %>%
+  add_header_above(c("", "cEOF2" = 2)) %>%
+  kable_classic() 
+```
+
+Given the similarity between the cEOF2 regression maps (Fig. \@ref(fig:eof2-regr-gh)) and documented PSAs we study the relationship between these modes.
+Table \@ref(tab:psa-eof2) shows the correlations between the two PSA indices and the time series for Real and Imaginary phases of cEOF2.
+As anticipated by Figure \@ref(fig:eof2-regr-gh), there is a strong positive correlation between PSA1 and Imaginary cEOF2, and also between PSA2 and Real cEOF2.
+On the other hand, there is no significant relationship between PSA1 and Real cEOF2, and between PSA2 and Imaginary cEOF2.
+As a result, cEOF2 represents well both the spatial structure and temporal evolution of the PSA modes, so it is possible to make an association between its two phases and the two PSA modes.
+The same phase election for cEOF2 that maximises the relationship between ENSO and Imaginary cEOF2, also maximises the association between cEOF2 components and PSA modes (not shown).
+
+(ref:phase-histogram-cap) Histogram of phase distribution of cEOF2 phase for the 1979 -- 2019 period.
+Bins are centred at 90º, 0º, -90º, -180º with a binwidth of 90º.
+The small vertical lines near the horizontal axis mark the observations.
+
+```{r phase-histogram, fig.cap = "(ref:phase-histogram-cap)", fig.width=width_column*.8, fig.height=2.5}
+N <- nrow(cut(ceof$eof[[1]], 2)$left)
+plot <- cut(ceof$eof[[1]], 2)$left %>% 
+  .[, arg := Arg(hgt)] %>% 
+  .[, w := Mod(hgt)/mean(Mod(hgt))] %>% 
+  qwrap(arg = c(-pi, pi) ~  c(-pi - 1/4*pi, pi - 1/4*pi)) %>% 
+  ggplot(aes(arg)) +
+  geom_histogram(binwidth = 90*pi/180, center = 0, size = 0.2,
+                 color = "black", fill = "gray70") + 
+  annotate("text",
+           label = c("- Real", "- Imaginary", "+ Real", "+ Imaginary"),
+           color = "black", size = 3,
+           y = 0.5, x = c(-pi, -pi/2, 0, pi/2), angle = 90, hjust = 0) +
+  geom_rug() +
+  scale_y_continuous("Number of years", 
+                     sec.axis = sec_axis(~.x/N, 
+                                         name = "Proportion of years", 
+                                         labels = scales::percent)) +
+  scale_x_continuous("Phase", trans = scales::reverse_trans(),
+                     breaks = seq(-pi - 1/4*pi, pi - 1/4*pi, by = 45*pi/180),
+                     labels = function(x) paste0(round(x*180/pi, 5), "º")) +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
+  no_grid
+
+imaginary <- sum(ggplot_build(plot)$data[[1]][c(1, 3), ]$count)/N
+plot
+```
+
+Figure \@ref(fig:phase-histogram) shows an histogram that counts the number of SON seasons in which the cEOF2 phase was close to each of the four particular phases (positive/negative of real/imaginary component), with the observations for each season marked as rugs on the horizontal axis.
+In `r scales::percent(imaginary)` of seasons cEOF2 has a phase similar to either the negative or positive imaginary phase, making the imaginary phase the most common phase. 
+This is also, by construction, the phase that is most correlated with ENSO.
+Therefore, the Imaginary cEOF2 explains more variance than the Real cEOF2 and conventional EOF analysis will tend to separate the two.
+
+This phase preferences is in agreement with @irving2016, who found a bimodal distribution to PSA-like variability (compare our Figure \@ref(fig:phase-histogram) with their Figure 6).
+
+## SAM {#sam}
+
+(ref:sam-eof-vertical-cap) Coefficient of determination ($r^2$) between each component of cEOFs and the SAM, Asymmetric SAM (A-SAM) and Symmetric SAM (S-SAM) indices computed at each level for the 1979 -- 2019 period.
+Thick lines represent estimates with p-value < 0.01 corrected for False Detection Rate [@benjamini1995].
+
+```{r sam-eof-vertical, fig.cap = "(ref:sam-eof-vertical-cap)", fig.height=4, fig.width=width_column}
+time <- ceof[, eof[[1]]$left, by = .(season)] %>%
+  sep_ReIm() %>%
+  .[, hgt := scale(hgt), by = .(cEOF, part)]
+
+sam_r2 <- sams %>%
+  .[time, on = .NATURAL, allow.cartesian = TRUE] %>%
+  na.omit() %>% 
+  .[, correlate(hgt, estimate), by = .(lev, term, part, cEOF)]
+
+max_r2 <- sam_r2[part == "Imaginary"] %>% 
+  .[, .SD[which.max(estimate^2)], by = .(cEOF, term)]
+
+sam_r2 %>%
+  .[, p.value_a := p.adjust(p.value, "fdr")] %>% 
+  ggplot(aes(lev, estimate^2)) +
+  geom_vline(xintercept = c(50, 200),  size = 0.2, color = "gray50") +
+  # geom_point(data = ~.x[p.adjust(p.value, "fdr") < 0.01], aes(color = term)) +
+  geom_line(data = ~copy(.x)[p.value_a > 0.01, estimate := NA], 
+            aes(colour = stage(term, after_scale = scales::alpha(colour, 0.5))), 
+            size = 2) +
+  geom_line(aes(color = term, group = term)) +
+
+  scale_x_level("Level (hPa)", minor_breaks = NULL, 
+                sec.axis = sa_height_axis(breaks = seq(0, 50, by = 10),
+                                          labels = function(x) paste0("~", x, " km"))) +
+  scale_y_continuous(r2, limits = c(0, 1), 
+                     guide = guide_axis(check.overlap = TRUE)) +
+  scale_size_manual("p-value", values = c("TRUE" = 1, 
+                                          "FALSE" = 0.5), 
+                    guide = "none") +
+  scale_color_brewer(NULL, palette = "Dark2", labels = c(full = "SAM",
+                                                         asym = "A-SAM",
+                                                         sym = "S-SAM")) +
+  facet_grid(part ~ cEOF) +
+  coord_flip() +
+  tag_facets("rc") +
+  panel_background +
+  theme(panel.spacing = grid::unit(1, "lines"), 
+        strip.placement = "outside")
+```
+
+We now explore the relationship between SAM and the cEOFs motivated by the resemblance between cEOFs regression maps and SAM patterns shown in Section\ \@ref(regressions).
+To that end, we computed the coefficient of determination between the cEOFs time series and the three SAM indices (SAM, A-SAM and S-SAM) defined by @campitelli2022 at each vertical level (Fig.\ \@ref(fig:sam-eof-vertical)).
+The SAM index is statistically significant correlated with the Real cEOF1 in all levels, and with the Imaginary cEOF1 and Imaginary cEOF2 in the troposphere.
+On the other hand, correlations between SAM and the Real cEOF2 are non-significant.
+
+The relationship between the SAM and cEOF1 in the troposphere is explained entirely by the zonally symmetric component of the SAM as shown by the high correlation with the S-SAM below 100 hPa and the low and statistically non-significant correlations between the A-SAM and either the Real or Imaginary cEOF1.
+In the stratosphere, the Real cEOF1 is correlated with both A-SAM and S-SAM, while the Imaginary cEOF1 is highly correlated only with the A-SAM. 
+These correlations are consistent with the regression maps of geopotential height in Figure\ \@ref(fig:eof1-regr-gh) and their comparison with those obtained for SAM, A-SAM and S-SAM by @campitelli2022.
+
+In the case of Imaginary cEOF2, the correlation between SAM and Imaginary cEOF2 for the troposphere is associated to the asymmetric variability of the SAM. 
+Indeed, the Imaginary cEOF2 shares up to `r scales::percent(max_r2[cEOF == "cEOF2" & term == "asym"]$estimate^2, 2)` variance with the A-SAM and only `r scales::percent(max_r2[cEOF == "cEOF2" & term == "sym"]$estimate^2, 2)` at most with the S-SAM (Figure \@ref(fig:sam-eof-vertical).b2).
+Such extremely high correlation between A-SAM and Imaginary cEOF2 suggests that the methodology proposed in this work is able to characterise the zonally asymmetric component of the SAM described previously by @campitelli2022.
+
+## Tropical sources of cEOFs variabitliy {#tropical}
 
@@
 The connections between cEOFs and Tropical sources are also assessed, to explore plausible forcings for the described cEOFs patterns.
-Figure \@ref(fig:sst-psi-2) shows the regression maps of Sea Surface Temperatures (SST) and streamfunction anomalies at 200 hPa respectively upon standardised cEOF2,.
-Besides showing the regression maps for the Real and Imaginary components, we including the corresponding regressions for two intermediate directions (corresponding to 45° and 135°).
-
-The Imaginary cEOF2 (second row) is associated with strong positive SST anomalies on the Central Pacific and negative anomalies over an area across northern Australia,  New Zealand the South Pacific Convergence Zone (SPCZ) (Figure \@ref(fig:sst-psi-2).b1).
+Figure \@ref(fig:sst-psi-2) shows the regression maps of Sea Surface Temperatures (SST) and streamfunction anomalies at 200 hPa respectively upon standardised cEOF2.
+Besides the regression maps for the Real and Imaginary components, we include the corresponding regressions for two intermediate directions (corresponding to 45° and 135°).
+
+Furthermore, the Imaginary cEOF2 (second row) is associated with strong positive SST anomalies on the Central Pacific and negative anomalies over an area across northern Australia,  New Zealand the South Pacific Convergence Zone (SPCZ) (Figure \@ref(fig:sst-psi-2).b1).
 The regression field of SST anomalies bears a strong resemblance with canonically positive ENSO [@bamston1997].
 Indeed, there is a significant and very high correlation (`r enso_cor[cEOF == "cEOF2" & part == "Imaginary"]$text`) between the ONI and the Imaginary cEOF2 time series.
-Besides the Pacific ENSO-like pattern, there are positive anomalies in the western Indian Ocean and negative values in the eastern Indian Ocean, resembling a positive Indian Ocean Dipole [@saji1999].
-Unsurprisingly, the correlation between the Imaginary cEOF2 and the DMI is `r dmi_cor[cEOF == "cEOF2" & part == "Imaginary"]$text`.
+Besides the Pacific ENSO-like pattern, there are positive anomalies in the western Indian Ocean and negative values in the eastern Indian Ocean, resembling a positive IOD [@saji1999].
+Consistently, the correlation between the Imaginary cEOF2 and the DMI is `r dmi_cor[cEOF == "cEOF2" & part == "Imaginary"]$text`.
 
 The Imaginary cEOF2 is associated with strong wave-like streamfunction anomalies emanating from the tropics (Figure\ \@ref(fig:sst-psi-2).b2), both from the Central Pacific sector and the Indian Ocean.
-These atmospheric response is consistent with the effect of ENSO and the Indian Ocean Dipole on the extratropics: the SST anomalies initiate anomalous tropical convection that excites Rossby waves that propagate meridionally towards higher latitudes  [@mo2000; @cai2011; @nuncio2015].
+These atmospheric response is consistent with the effect of ENSO and the IOD on the extratropics: the SST anomalies initiate anomalous tropical convection that excites Rossby waves that propagate meridionally towards higher latitudes  [@mo2000; @cai2011; @nuncio2015].
 
@@
 Meanwhile, Rows a and c in Fig.\ \@ref(fig:sst-psi-2) show that the intermediate phases are still associated with significant SST anomalies over the Pacific Ocean, but with with slightly different location.
-The 135º phase is associated with SST in the central Pacific (Fig.\ \@ref(fig:sst-psi-2)a.1), while the 45º phase is associated with SST in the eastern Pacific (Fig.\ \@ref(fig:sst-psi-2)c.1), which correspond roughly to the Central Pacific and Eastern Pacific and "flavours" of ENSO, respectively [@kao2009].
+The 135º phase is associated with SST in the central Pacific (Fig.\ \@ref(fig:sst-psi-2)a.1), while the 45º phase is associated with SST in the eastern Pacific (Fig.\ \@ref(fig:sst-psi-2)c.1), which correspond roughly to the Central Pacific and Eastern Pacific "flavours" of ENSO, respectively [@kao2009].
 Both of these phases could be also associated to wave trains generated in the region surrounding Australia and propagates toward the extra-tropics.
@@
 
-## Precipitation and 2-metre temperature {#precipitation}
+## cEOFs surface impacts {#precipitation}
 
@@
 
-Figure\ \@ref(fig:eof1-regr-t) shows the variance of the 2-meter temperature or precipitation anomalies explained variance by the mluitple linear model of Real cEOF1 and Imaginary cEOF1 (column 1), and Real cEOF2 and Imaginary cEOF2 (column 2).
-The variance explained by cEOF1 for precipitation anomalies is extremely low and also for most of the temperature anomalies, except for the eastern western of Antarctica (Fig.\  \@ref(fig:eof1-regr-t)a.1).
-
-On the other hand, the cEOF2  explained variance is greater than 50% in some regions for both variables (Fig.\ \@ref(fig:eof1-regr-t) column 2). 
+The influence of cEOFs variability in and 2-metre temperature and temperature in the Southern Hemisphere is also explored.
+Figure\ \@ref(fig:pp-t2m-r2) shows the 2-meter temperature or precipitation anomalies explained variance by the multiple linear model of both Real and Imaginary cEOF1 components (column 1), and both Real  and Imaginary cEOF2 components (column 2).
+The variance explained by cEOF1 for precipitation anomalies is extremely low and also for most of the temperature anomalies, except for the eastern western of Antarctica (Fig.\  \@ref(fig:pp-t2m-r2)a.1).
+
+On the other hand, the cEOF2  explained variance is greater than 50% in some regions for both variables (Fig.\ \@ref(fig:pp-t2m-r2) column 2). 
 For 2-metre temperature, there are high values in the tropical Pacific and the SPCZ, as well as the region following an arc between New Zealand and the South Atlantic, with higher values in the Southern Ocean.
@@
 
-There is some disagreement in the literature of whether the phase of the PSA pattern is affected by the location of tropical SST anomalies [@irving2016].
-We show not only that the cEOF2 tends to be in the imaginary phase (\~PSA1) when the ENSO region is warm, but that central Pacific SST anomalies tend to move the cEOF towards the negative real phase and eastern Pacific SST anomalies tend to move it towards positive real phase.
+In previous sections, it has been shown that variability patterns that arise from cEOF methodology allow to describe the zonally asymmetric springtime extratropical SH circulation, reproducing previous features such as the variability related to PSAs or A-SAM.
+However, it could be questioned what is the advantage of using the methodology proposed here.
+Firstly, the spatial fields obtained from both components of cEOF2, that resembles PSA patterns, are in quadrature by construction.
+Thus, it is possible to derive a joint PSA index from the amplitude and phase obtained when applying cEOF methodology.
+Secondly, because of how these components are constructed, they are not forced to be orthogonal to the SAM as in the standard EOF methodology.
+This allows us to show that Imaginary cEOF2, that resembles PSA1 variability, is closely related to the SAM in the troposphere.
+Moreover, the high correlation of this component to the the zonally asymmetric portion of the SAM allow to speculate that the A-SAM might actually be a statistical contamination of the PSA mode.
+
+@irving2016 argued that there is some disagreement in the literature of whether the phase of the PSA pattern is affected by the location of tropical SST anomalies.
+With the methodology used in this study, we were able to show not only that the cEOF2 tends to be in the positive or negative imaginary phase (\~PSA1) when the ENSO region is warm or cold, respectively, but that central Pacific SST anomalies tend to align the cEOF towards the negative real phase and eastern Pacific SST anomalies tend to align it towards positive real phase.
 When ENSO is neutral, the cEOF2 is still as active, but with no preferred phase.
-This mirrors the results of @cai2002, who showed that the CSIRO Model can create PSA-like variability even in the absence of ENSO forcing (with a climatological run), but that the variability of one of the PSA modes was enhanced when adding the ENSO signal.
-
-The sensitivity of the phase of the cEOF2 to the location of the tropical SST anomalies was also seen by @ciasto2015, who detected similar Rossby wave patterns associated with central Pacific and eastern Pacific SST anomalies but with a change in phase.
-
-Some aspects of our work contrast with @irving2016 analysis of the PSA.
-They didn't find neither a strong association between PSA-like variability and ENSO, nor a relationship between it and Australian precipitation.
-Part of the disagreement could be due to the fact that we focus only on SON, which is when the teleconnection pattern is strongest, instead of the whole year.
-@irving2016 use of ERA5 forectasted precipitation instead of the more observationally constrained CMAP dataset might also be an issue.
-But in light of the previous discussion on the importance of the phase of the cEOF2 in its relationship with ENSO and with precipitation impacts, a big difference could be fact that @irving2016's methodology is insensitive to phase changes of the PSA-like variability.
-
-To the extent that the cEOF2 can be identified with the PSA, what is the advantage of the technique proposed here?
-The spatial fields that the cEOF2 index describes are in quadrature by construction, which makes it possible to derive a proper amplitude and phase.
-Furthermore, because of the way they are constructed, they are not forced to be orthogonal to the SAM.
-This allows us to show that Imaginary cEOF2 is closely related to the Southern Annular Mode in the troposphere.
-In fact, it is so highly correlated to the the zonally asymmetric portion of the SAM that we speculate that that the asymmetric component of the SAM might actually a statistical contamination of the PSA mode.
+This mirrors the results of @cai2002, who showed that the CSIRO Model can create PSA-like variability even in the absence of ENSO forcing (i.e.
+with a climatological run), but that the variability of one of the PSA modes was enhanced when adding the ENSO signal.
+
+The sensitivity of the phase of the PSAs to the location of the tropical SST anomalies was also seen by @ciasto2015, who detected similar Rossby wave patterns associated with central Pacific and eastern Pacific SST anomalies but with a change in phase.
+
+Some results of our work contrast with those obtained by @irving2016 when analysing the PSA.
+They did not find neither a strong association between PSA-like variability and ENSO, nor a relationship between this variability and Australian precipitation.
+Part of the disagreement could be due to the fact that we focus only on SON, which is the season when the teleconnection pattern is strongest, instead of the whole year.
+Moreover, @irving2016 use of ERA5 forectasted precipitation instead of the more observationally constrained CMAP dataset might also be an issue.
+Additionally, the disagreement could be related to the fact that @irving2016 show weak relationship between ENSO and their PSA pattern, which explains partly the relationship between Australian rainfall and PSA-like variability.
 
@@
 
-The first complex EOF represents the variability of the zonal wave 1 in the stratosphere, but represents a more zonally symmetric monopole in the troposphere.
+The cEOF1 represents the variability of the zonal wave 1 in the stratosphere, but represents a more zonally symmetric monopole in the troposphere.
 There is a statistically positive trend in the magnitude of this cEOF, which is consistent with previous studies that showed secular changes in springtime wave-1-like patterns [e.g. @raphael2003].
@@
 
-The second complex EOF represents a wave-3 pattern with maximum magnitude in the Pacific sector.
-Essentially, it is an alternate representation of the PSA1 and PSA2 patterns [@mo2001].
-We show that the Imaginary cEOF2 can be roughly identified with the PSA1 and the Real cEOF2 with the PSA2.
-
-The phase of the cEOF2 is related to the the variability of ENSO, but its magnitude appears not to be.
-Except for the three strongest ENSO events in the period which coincide with the three strongest cEOF2 years.
-
-Precipitation anomalies in South America associated with the Imaginary cEOF2 show a clear ENSO-like impact, with positive anomalies in South-Eastern South America, negative anomalies in Southern Brazil and positive anomalies in central Chile.
+The cEOF2 represents a wave-3 pattern with maximum magnitude in the Pacific sector, that is an alternative representation of the PSA1 and PSA2 patterns [@mo2001].
+The Imaginary cEOF2 can be  identified with the PSA1 and the Real cEOF2 with the PSA2.
+
+The phase of the cEOF2 is related to the the variability of ENSO, but its magnitude appears not to be, with the exception of the three strongest ENSO events in the period which coincide with the three strongest cEOF2 years.
+
+The cEOF2 variability could be related to surface impacts, while the cEOF1 influence is almost negligible. 
+For instance, precipitation anomalies in South America associated with the Imaginary cEOF2 show a clear ENSO-like impact, with positive anomalies in South-Eastern South America, negative anomalies in Southern Brazil and positive anomalies in central Chile.
 Precipitation anomalies associated with the Real cEOF2 are low and not statistically significant in this region, showing that the Imaginary phase is optimally aligned with the direction of maximum precipitation impacts there.
@@
 
+The methodology proposed in this study allow a deeper understanding of  the zonally asymmetric springtime extratropical SH circulation such as a better description of PSA like variability using a unique complex index and the understanding of relationship between PSAs and ENSO or SAM variability.
 Further work should extend this analysis to other seasons and further study the relationship between the cEOF2 and the SAM.
-In particular, it is not clear whether the close identification between the Imaginary cEOF2 and the Asymmetric SAM arises from physical processes or statistical contamination.
 

### 2022-04-05 analysis/paper/shceof.tex
render pdf

@@
 These patterns influence precipitation anomalies in South America (Mo and Paegle 2001).
-Although these patterns are usually derived by applying EOF to temporal anomalies, (M. Raphael 2003) also applied EOF methods specifically to zonal anomalies.
-More recently, (Irving and Simmonds 2016) proposed a novel methodology for objectively identifying the PSA pattern using Fourier decomposition.
+Although these patterns are usually derived by applying EOF to temporal anomalies, M. Raphael (2003) also applied EOF methods specifically to zonal anomalies.
+More recently, Irving and Simmonds (2016) proposed a novel methodology for objectively identifying the PSA pattern using Fourier decomposition.
 
@@
 
-A third methodology commonly used consists on identifying particular features of interest and creating indices using simple methods such as averages and differences.
-Examples of this methodology are the SAM Index of (Gong and Wang 1999), the SH wave 3 activity index defined by M.
+A third methodology commonly used to describe circulation anomalies consists on identifying particular features of interest and creating indices using simple methods such as averages and differences.
+Examples of this methodology are the SAM Index of Gong and Wang (1999), the SH wave 3 activity index defined by M.
 N.
@@
 
-The general goal of this study is to improve the description and understanding of the zonally asymmetric extratropical SH circulation using complex Empirical Orthogonal Functions, which allow to describe phase varying planetary waves with variable amplitudes along a latitude circle.
+The general goal of this study is to improve the description and understanding of the zonally asymmetric extratropical SH circulation using cEOF, which allow to describe phase varying planetary waves with variable amplitudes along a latitude circle.
 In addition, it is proposed to expand the knowledge of the simultaneous behaviour of SH asymmetric circulation in the troposphere and the stratosphere.
 
-We restrict this paper to the September-October-November (SON) trimester.
-We chose this season because it showed the strongest and more clearly interpretable results in a preliminary analysis of the four season.
-This is is consistent with previous research that shows that this season experiences a maximum in tropical teleconnections (Cazes-Boezio, Robertson, and Mechoso 2003) and stratosphere-troposphere interactions (Lim, Hendon, and Thompson 2018).
-
-In Section \ref{methods} we describe the methods.
-In Section \ref{spatial} we analyse the spatial patterns of each complex EOF.
-In Section \ref{regressions} we study the spatial regressions with geopotential height and temperature anomalies.
-In Section \ref{other-variables} we analyse the relationship between cEOF1 and Ozone, and between cEOF2 with the SAM and the PSA.
-In Section \ref{tropical} we study tropical forcings that explain the variability of each cEOF.
-Finally, in Section \ref{precipitation} we show the relationship between these modes of variability and precipitation anomalies in South America and Oceania.
+We restrict this paper to the spring September-October-November (SON) trimester.
+This season experiences a maximum in tropical teleconnections (Cazes-Boezio, Robertson, and Mechoso 2003) and stratosphere-troposphere interactions (Lim, Hendon, and Thompson 2018).
+
+In Section~\ref{methods} we describe the methods.
+In Section~\ref{spatial} we analyse the spatial patterns of each complex EOF.
+In Section~\ref{regressions} we study the spatial regressions with geopotential height, temperature and ozone anomalies.
+In Section~\ref{psa} and \ref{sam} we analyse the relationship between cEOF2 the PSA and SAM modes.
+In Section~\ref{tropical} we study tropical forcings that explain the variability of each cEOF.
+Finally, in Section \ref{precipitation} we show the relationship between these modes of variability and precipitation and surface temperature anomalies in South America and Oceania.
 
@@
 2017) and precipitation monthly data from the CPC Merged Analysis of Precipitation (P.
-Xie and Arkin 1997), with a 2.5\degree resolution in latitude and longitude.
+Xie and Arkin 1997), with a 2\degree and 2.5\degree resolution in latitude and longitude respectively.
 The rainfall gridded dataset is based on information from different sources such as rain gauge observations, satellite inferred estimations and the NCEP-NCAR reanalysis, and it is available since 1979 to present.
@@
 Since any choice is arbitrary and equally valid, we chose the phase of each cEOF so that the real or imaginary parts are aligned with meaningful phases in our analysis.
-
-Preliminary analysis showed that the first cEOF was closely related to the the zonal wave 1 of Total Ozone Column and the second cEOF was closely related to ENSO.
-Therefore, to aid in the interpretation, we chose the phases of each cEOF so that these relationships are aligned with one of the complex components.
-
-We chose the phase of cEOF1 so that the time series corresponding to the real part has the maximum correlation with the zonal wave 1 of Total Ozone Column between 75°S and 45°S.
-And we chose the phase of cEOF2 so that the coefficient of determination between the Oceanic Niño Index (Bamston, Chelliah, and Goldenberg 1997) and the real part is minimized, which also nearly maximizes the correlation with the imaginary part.
-
 This procedure does not create spurious correlations, it only takes an existing relationship and aligns it with a specific phase.
+
+Preliminary analysis showed that the first cEOF was closely related to the zonal wave 1 of Total Ozone Column and the second cEOF was closely related to ENSO.
+Therefore, to aid in the interpretation we chose the phase of cEOF1 so that the time series corresponding to the real part has the maximum correlation with the zonal wave 1 of Total Ozone Column between 75°S and 45°S.
+Similarly, we chose the phase of cEOF2 so that the coefficient of determination between the Oceanic Niño Index (Bamston, Chelliah, and Goldenberg 1997) and the real part is minimized, which also nearly maximizes the correlation with the imaginary part.
 
@@
 All graphics are made using ggplot2 (Wickham 2009).
-We downloaded data from reanalysis using the ecmwfr package (Hufkens 2020) and indices of ENSO and Indian Ocean Dipole with the rsoi package (Albers and Campitelli 2020).
+We downloaded data from reanalysis using the ecmwfr package (Hufkens 2020) and indices of ENSO and Indian Ocean Dipole (IOD) with the rsoi package (Albers and Campitelli 2020).
 The paper was rendered using knitr and rmarkdown (Y. Xie 2015; Allaire et al. 2020).
@@
 The shading (contours) corresponds to real (imaginary) part.
-Arbitrary units.}
+Arbitrary units.
+In parenthesis, the proportion of variance (with respect to the zonal mean) explained.}
 \end{figure}
@@
 In the spatial patterns (Fig.~\ref{fig:ceofs-1}), the real and the imaginary components are in quadrature by construction, so that each cEOF describe a single wave-like pattern whose amplitude and position (i.e.~phase) is controlled by the magnitude and phase of the temporal cEOF.
-The wave patterns described by these cEOFs match the patterns seen in the traditional EOFs of Figure~\ref{fig:eof-naive}.
+The wave patterns described by these cEOFs match the patterns seen in the standard EOFs of Figure~\ref{fig:eof-naive}.
 
@@
 Both cEOFs show year-to-year variability but show no evidence of decadal variability (Fig. \ref{fig:extended-series}).
-Because the geopotential fields that enter into the cEOFs algorithm are anomalies with respect to the zonal mean instead of the time mean, the cEOFs have non zero temporal mean.
-However, cEOF2's mean is almost zero, which indicates that only cEOF1 includes variability that significantly projects onto the mean zonally anomalous field.
+Because the geopotential fields that enter into the cEOFs algorithm are anomalies with respect to the zonal mean instead of the time mean, the cEOFs time series have non zero temporal mean.
+However, cEOF2 temporal mean is almost zero, which indicates that only cEOF1 includes variability that significantly projects onto the mean zonally anomalous field.
 
@@
 
+\hypertarget{geopotential}{%
+\subsubsection{Geopotential}\label{geopotential}}
+
 In the previous section, cEOFs were applied to zonal anomalies, that is anomalies derived by removing the zonally mean values in order to isolate the main characteristics of the main zonal waves characterizing the circulation in the SH.
@@
 At 200 hPa (Figure~\ref{fig:eof1-regr-gh} row b) the Real cEOF1 shows a single area of positive anomalies spanning West Antarctica surrounded by opposite anomalies in lower latitudes with is centre shifted slightly eastward compared with the upper level anomalies.
-The Imaginary cEOF2.
-The Imaginary cEOF1 shows a much more zonally symmetrical pattern resembling the negative SAM phase (Fogt and Marshall 2020).
+The Imaginary cEOF1 shows a much more zonally symmetrical pattern resembling the negative SAM phase (e.g. Fogt and Marshall 2020).
 Both components of cEOF1 could be associated to zonal anomalies in the upper troposphere jet stream intensity.
@@
 
-\hypertarget{other-variables}{%
-\subsection{cEOFs relationship with known modes of variability}\label{other-variables}}
+\hypertarget{temp-ozone}{%
+\subsubsection{Temperature and Ozone}\label{temp-ozone}}
+
+\begin{figure}
+\centering
+\includegraphics{../figures/eof1-regr-t-1.pdf}
+\caption{\label{fig:eof1-regr-t}Same as Figure~\ref{fig:eof1-regr-gh} but for air temperature (K).}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics{../figures/t-vertical-1.pdf}
+\caption{\label{fig:t-vertical}Regression of SON zonal anomalies averaged between 75°S and 45°S of mean air temperature (shaded, Kelvin) and ozone mixing ratio (contours, negative contours with dashed lines, labels in parts per billion by mass) with the (a) Real and (b) Imaginary components of the cEOF1 for the 1979 -- 2019 period.}
+\end{figure}
+
+The influence of cEOFs variability on air temperature was also evaluated.
+Figure~\ref{fig:eof1-regr-t} shows regression maps of air temperature anomalies at 50 hPa and 200 hPa upon cEOF1.
+In both levels, the Real cEOF1 is associated with a positive monopole over the South Pole with its centre moved slightly towards 150ºE (Fig.~\ref{fig:eof1-regr-t} column 1).
+On the other hand, the regression maps on the Imaginary cEOF1 show a more clear wave 1 pattern with its maximum around 60ºE.
+The distribution of temperature regression coefficients at 50 hPa and at 200 hPa mirror the geopotential height regression maps at 50 hPa (Fig.~\ref{fig:eof1-regr-gh}).
+
+Figure~\ref{fig:t-vertical} shows the vertical distribution of the regression coefficients on cEOF1 from zonal anomalies averaged between 75°S and 45°S of air temperature and of ozone mixing ratio.
+Temperature zonal anomalies associated with cEOF1 show a clear wave 1 pattern for both real and imaginary components throughout the atmosphere above 250 hPa with a sign reversal above 10 hPa.
+As a result of the hydrostatic balance, this is the level in which the geopotential anomaly have maximum amplitude (not shown).
+
+The maximum ozone regressed anomalies match with the minimum temperature anomalies above 10 hPa and with the maximum temperature anomalies below 10 hPa (Fig.~\ref{fig:t-vertical}).
+Therefore, the ozone zonal wave 1 is anticorrelated with the temperature zonal wave 1 in the upper stratosphere, and directly correlated in the upper stratosphere.
+This change in phase is observed in ozone anomalies forced by planetary waves that reach the stratosphere (Hartmann and Garcia 1979; Wirth 1993; Smith 1995).
+In the photochemically-dominated upper stratosphere, cold temperatures inhibit the destruction of ozone, explaining the opposite behaviour for both variables.
+On the other hand, in the advectively-dominated lower stratosphere, ozone anomalies are 90º out of phase with horizontal and vertical transport, which are in addition 90º out of phase with temperature anomalies, resulting in same sign anomalies for the response of both variables (Hartmann and Garcia 1979; Wirth 1993; Smith 1995).
+
+\begin{figure}
+\centering
+\includegraphics{../figures/o3-regr-1.pdf}
+\caption{\label{fig:o3-regr}Regression of SON mean Total Ozone Column anomalies (shaded, Dobson Units) with the (a) real and (b) imaginary components of the cEOF1 for the 1979 -- 2019 period.
+On contours, the mean zonal anomaly of Total Ozone Column (negative contours in dashed lines, Dobson Units).
+Areas marked with dots have p-values smaller than 0.01 adjusted for False Detection Rate.}
+\end{figure}
+
+The regression maps of TOC anomalies upon cEOF1 (Fig.~\ref{fig:o3-regr}) show zonal wave 1 patterns associated with both components of cEOF1.
+The climatological position of the springtime Ozone minimum (ozone hole) is outside the South Pole and towards the Weddell Sea (e.g. Grytsai 2011).
+Thus, the Real cEOF1 regression field (Figure~\ref{fig:o3-regr}a) coincides with the climatological position of the ozone hole, while it is 90° out of phase for the Imaginary cEOF1.
+The temporal correlation between the amplitudes of TOC planetary wave 1 and cEOF1 is 0.79 (CI: 0.63 -- 0.88), while the correlation between their phases is -0.85 (CI: -0.92 -- -0.74).
+Consequently, cEOF1 is strongly related with the SH ozone variability.
 
 \hypertarget{psa}{%
-\subsubsection{PSA}\label{psa}}
+\subsection{PSA}\label{psa}}
 
@@
 As a result, cEOF2 represents well both the spatial structure and temporal evolution of the PSA modes, so it is possible to make an association between its two phases and the two PSA modes.
-The same phase election for cEOF2 that maximises the relationship between ENSO and Imaginary cEOF2, also maximises the association between cEOF2 components and PSA modes.
+The same phase election for cEOF2 that maximises the relationship between ENSO and Imaginary cEOF2, also maximises the association between cEOF2 components and PSA modes (not shown).
 
@@
 \hypertarget{sam}{%
-\subsubsection{SAM}\label{sam}}
+\subsection{SAM}\label{sam}}
 
@@
 
-We now explore the relationship between SAM and the cEOFs motivated by the resemblance between cEOFs regression maps and SAM patterns shown in the previous section.
+We now explore the relationship between SAM and the cEOFs motivated by the resemblance between cEOFs regression maps and SAM patterns shown in Section~\ref{regressions}.
 To that end, we computed the coefficient of determination between the cEOFs time series and the three SAM indices (SAM, A-SAM and S-SAM) defined by Campitelli, Díaz, and Vera (2022) at each vertical level (Fig.~\ref{fig:sam-eof-vertical}).
@@
 
-\hypertarget{temperature-and-ozone}{%
-\subsubsection{Temperature and Ozone}\label{temperature-and-ozone}}
-
-\begin{figure}
-\centering
-\includegraphics{../figures/eof1-regr-t-1.pdf}
-\caption{\label{fig:eof1-regr-t}Same as Figure~\ref{fig:eof1-regr-gh} but for air temperature (K).}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics{../figures/t-vertical-1.pdf}
-\caption{\label{fig:t-vertical}Regression of SON zonal anomalies averaged between 75°S and 45°S of mean air temperature (shaded, Kelvin) and ozone mixing ratio (contours, negative contours with dashed lines, labels in parts per billion by mass) with the (a) Real and (b) Imaginary components of the cEOF1 for the 1979 -- 2019 period.}
-\end{figure}
-
-The influence of cEOFs variability on air temperature was also evaluated.
-Figure~\ref{fig:eof1-regr-t} shows regression maps of air temperature anomalies at 50 hPa and 200 hPa upon cEOF1.
-In both levels, the Real cEOF1 is associated with a positive monopole over the South Pole with its centre moved slightly towards 150ºE (Fig.~\ref{fig:eof1-regr-t} column 1).
-On the other hand, the regression maps on the Imaginary cEOF1 show a more clear wave 1 pattern with its maximum around 60ºE.
-The distribution of temperature regression coefficients at 50 hPa and at 200 hPa mirror the geopotential height regression maps at 50 hPa (Fig.~\ref{fig:eof1-regr-gh}).
-
-Figure~\ref{fig:t-vertical} shows the vertical distribution of the regression coefficients on cEOF1 from zonal anomalies averaged between 75°S and 45°S of air temperature and of ozone mixing ratio.
-Temperature zonal anomalies associated with cEOF1 show a clear wave 1 pattern for both real and imaginary components throughout the atmosphere above 250 hPa with a sign reversal above 10 hPa.
-As a result of the hydrostatic balance, this is the level in which the geopotential anomaly have maximum amplitude (not shown).
-
-The maximum ozone regressed anomalies match with the minimum temperature anomalies above 10 hPa and with the maximum temperature anomalies below 10 hPa (Fig.~\ref{fig:t-vertical}).
-Therefore, the ozone zonal wave 1 is anticorrelated with the temperature zonal wave 1 in the upper stratosphere, and directly correlated in the upper stratosphere.
-This change in phase is observed in ozone anomalies forced by planetary waves that reach the stratosphere (Hartmann and Garcia 1979; Wirth 1993; Smith 1995).
-In the photochemically-dominated upper stratosphere, cold temperatures inhibit the destruction of ozone, explaining the opposite behaviour for both variables.
-On the other hand, in the advectively-dominated lower stratosphere, ozone anomalies are 90º out of phase with horizontal and vertical transport, which are in addition 90º out of phase with temperature anomalies, resulting in same sign anomalies for the response of both variables
-
-\begin{figure}
-\centering
-\includegraphics{../figures/o3-regr-1.pdf}
-\caption{\label{fig:o3-regr}Regression of SON mean Total Ozone Column anomalies (shaded, Dobson Units) with the (a) real and (b) imaginary components of the cEOF1 for the 1979 -- 2019 period.
-On contours, the mean zonal anomaly of Total Ozone Column (negative contours in dashed lines, Dobson Units).
-Areas marked with dots have p-values smaller than 0.01 adjusted for False Detection Rate.}
-\end{figure}
-
-The regression maps of TOC anomalies upon cEOF1 (Fig.~\ref{fig:o3-regr}) show zonal wave 1 patterns associated with both components of cEOF1.
-The climatological position of the springtime Ozone minimum (ozone hole) is outside the South Pole and towards the Weddell Sea (e.g. Grytsai 2011).
-Thus, the Real cEOF1 regression field (Figure~\ref{fig:o3-regr}a) coincides with the climatological position of the ozone hole, while it is 90° out of phase for the Imaginary cEOF1.
-The temporal correlation between the amplitudes of TOC planetary wave 1 and cEOF1 is 0.79 (CI: 0.63 -- 0.88), while the correlation between their phases is -0.85 (CI: -0.92 -- -0.74).
-Consequently, cEOF1 is strongly related with the SH ozone variability.
-
 \hypertarget{tropical}{%
-\subsection{Tropical sources}\label{tropical}}
+\subsection{Tropical sources of cEOFs variabitliy}\label{tropical}}
 
@@
 The connections between cEOFs and Tropical sources are also assessed, to explore plausible forcings for the described cEOFs patterns.
-Figure~\ref{fig:sst-psi-2} shows the regression maps of Sea Surface Temperatures (SST) and streamfunction anomalies at 200 hPa respectively upon standardised cEOF2,.
-Besides showing the regression maps for the Real and Imaginary components, we including the corresponding regressions for two intermediate directions (corresponding to 45° and 135°).
-
-The Imaginary cEOF2 (second row) is associated with strong positive SST anomalies on the Central Pacific and negative anomalies over an area across northern Australia, New Zealand the South Pacific Convergence Zone (SPCZ) (Figure~\ref{fig:sst-psi-2}.b1).
+Figure~\ref{fig:sst-psi-2} shows the regression maps of Sea Surface Temperatures (SST) and streamfunction anomalies at 200 hPa respectively upon standardised cEOF2.
+Besides the regression maps for the Real and Imaginary components, we include the corresponding regressions for two intermediate directions (corresponding to 45° and 135°).
+
+Furthermore, the Imaginary cEOF2 (second row) is associated with strong positive SST anomalies on the Central Pacific and negative anomalies over an area across northern Australia, New Zealand the South Pacific Convergence Zone (SPCZ) (Figure~\ref{fig:sst-psi-2}.b1).
 The regression field of SST anomalies bears a strong resemblance with canonically positive ENSO (Bamston, Chelliah, and Goldenberg 1997).
 Indeed, there is a significant and very high correlation (0.76 (CI: 0.6 -- 0.87)) between the ONI and the Imaginary cEOF2 time series.
-Besides the Pacific ENSO-like pattern, there are positive anomalies in the western Indian Ocean and negative values in the eastern Indian Ocean, resembling a positive Indian Ocean Dipole (Saji et al.
-1999).
-Unsurprisingly, the correlation between the Imaginary cEOF2 and the DMI is 0.62 (CI: 0.38 -- 0.78).
+Besides the Pacific ENSO-like pattern, there are positive anomalies in the western Indian Ocean and negative values in the eastern Indian Ocean, resembling a positive IOD (Saji et al. 1999).
+Consistently, the correlation between the Imaginary cEOF2 and the DMI is 0.62 (CI: 0.38 -- 0.78).
 
 The Imaginary cEOF2 is associated with strong wave-like streamfunction anomalies emanating from the tropics (Figure~\ref{fig:sst-psi-2}.b2), both from the Central Pacific sector and the Indian Ocean.
-These atmospheric response is consistent with the effect of ENSO and the Indian Ocean Dipole on the extratropics: the SST anomalies initiate anomalous tropical convection that excites Rossby waves that propagate meridionally towards higher latitudes (Mo 2000; Cai et al.
+These atmospheric response is consistent with the effect of ENSO and the IOD on the extratropics: the SST anomalies initiate anomalous tropical convection that excites Rossby waves that propagate meridionally towards higher latitudes (Mo 2000; Cai et al.
 2011; Nuncio and Yuan 2015).
@@
 Meanwhile, Rows a and c in Fig.~\ref{fig:sst-psi-2} show that the intermediate phases are still associated with significant SST anomalies over the Pacific Ocean, but with with slightly different location.
-The 135º phase is associated with SST in the central Pacific (Fig.~\ref{fig:sst-psi-2}a.1), while the 45º phase is associated with SST in the eastern Pacific (Fig.~\ref{fig:sst-psi-2}c.1), which correspond roughly to the Central Pacific and Eastern Pacific and ``flavours'' of ENSO, respectively (Kao and Yu 2009).
+The 135º phase is associated with SST in the central Pacific (Fig.~\ref{fig:sst-psi-2}a.1), while the 45º phase is associated with SST in the eastern Pacific (Fig.~\ref{fig:sst-psi-2}c.1), which correspond roughly to the Central Pacific and Eastern Pacific ``flavours'' of ENSO, respectively (Kao and Yu 2009).
 Both of these phases could be also associated to wave trains generated in the region surrounding Australia and propagates toward the extra-tropics.
@@
 \hypertarget{precipitation}{%
-\subsection{Precipitation and 2-metre temperature}\label{precipitation}}
+\subsection{cEOFs surface impacts}\label{precipitation}}
 
@@
 
-Figure~\ref{fig:eof1-regr-t} shows the variance of the 2-meter temperature or precipitation anomalies explained variance by the mluitple linear model of Real cEOF1 and Imaginary cEOF1 (column 1), and Real cEOF2 and Imaginary cEOF2 (column 2).
-The variance explained by cEOF1 for precipitation anomalies is extremely low and also for most of the temperature anomalies, except for the eastern western of Antarctica (Fig.~ \ref{fig:eof1-regr-t}a.1).
-
-On the other hand, the cEOF2 explained variance is greater than 50\% in some regions for both variables (Fig.~\ref{fig:eof1-regr-t} column 2).
+The influence of cEOFs variability in and 2-metre temperature and temperature in the Southern Hemisphere is also explored.
+Figure~\ref{fig:pp-t2m-r2} shows the 2-meter temperature or precipitation anomalies explained variance by the multiple linear model of both Real and Imaginary cEOF1 components (column 1), and both Real and Imaginary cEOF2 components (column 2).
+The variance explained by cEOF1 for precipitation anomalies is extremely low and also for most of the temperature anomalies, except for the eastern western of Antarctica (Fig.~ \ref{fig:pp-t2m-r2}a.1).
+
+On the other hand, the cEOF2 explained variance is greater than 50\% in some regions for both variables (Fig.~\ref{fig:pp-t2m-r2} column 2).
 For 2-metre temperature, there are high values in the tropical Pacific and the SPCZ, as well as the region following an arc between New Zealand and the South Atlantic, with higher values in the Southern Ocean.
@@
 
-There is some disagreement in the literature of whether the phase of the PSA pattern is affected by the location of tropical SST anomalies (Irving and Simmonds 2016).
-We show not only that the cEOF2 tends to be in the imaginary phase (\textasciitilde PSA1) when the ENSO region is warm, but that central Pacific SST anomalies tend to move the cEOF towards the negative real phase and eastern Pacific SST anomalies tend to move it towards positive real phase.
+In previous sections, it has been shown that variability patterns that arise from cEOF methodology allow to describe the zonally asymmetric springtime extratropical SH circulation, reproducing previous features such as the variability related to PSAs or A-SAM.
+However, it could be questioned what is the advantage of using the methodology proposed here.
+Firstly, the spatial fields obtained from both components of cEOF2, that resembles PSA patterns, are in quadrature by construction.
+Thus, it is possible to derive a joint PSA index from the amplitude and phase obtained when applying cEOF methodology.
+Secondly, because of how these components are constructed, they are not forced to be orthogonal to the SAM as in the standard EOF methodology.
+This allows us to show that Imaginary cEOF2, that resembles PSA1 variability, is closely related to the SAM in the troposphere.
+Moreover, the high correlation of this component to the the zonally asymmetric portion of the SAM allow to speculate that the A-SAM might actually be a statistical contamination of the PSA mode.
+
+Irving and Simmonds (2016) argued that there is some disagreement in the literature of whether the phase of the PSA pattern is affected by the location of tropical SST anomalies.
+With the methodology used in this study, we were able to show not only that the cEOF2 tends to be in the positive or negative imaginary phase (\textasciitilde PSA1) when the ENSO region is warm or cold, respectively, but that central Pacific SST anomalies tend to align the cEOF towards the negative real phase and eastern Pacific SST anomalies tend to align it towards positive real phase.
 When ENSO is neutral, the cEOF2 is still as active, but with no preferred phase.
-This mirrors the results of Cai and Watterson (2002), who showed that the CSIRO Model can create PSA-like variability even in the absence of ENSO forcing (with a climatological run), but that the variability of one of the PSA modes was enhanced when adding the ENSO signal.
-
-The sensitivity of the phase of the cEOF2 to the location of the tropical SST anomalies was also seen by Ciasto, Simpkins, and England (2015), who detected similar Rossby wave patterns associated with central Pacific and eastern Pacific SST anomalies but with a change in phase.
-
-Some aspects of our work contrast with Irving and Simmonds (2016) analysis of the PSA.
-They didn't find neither a strong association between PSA-like variability and ENSO, nor a relationship between it and Australian precipitation.
-Part of the disagreement could be due to the fact that we focus only on SON, which is when the teleconnection pattern is strongest, instead of the whole year.
-Irving and Simmonds (2016) use of ERA5 forectasted precipitation instead of the more observationally constrained CMAP dataset might also be an issue.
-But in light of the previous discussion on the importance of the phase of the cEOF2 in its relationship with ENSO and with precipitation impacts, a big difference could be fact that Irving and Simmonds (2016)'s methodology is insensitive to phase changes of the PSA-like variability.
-
-To the extent that the cEOF2 can be identified with the PSA, what is the advantage of the technique proposed here?
-The spatial fields that the cEOF2 index describes are in quadrature by construction, which makes it possible to derive a proper amplitude and phase.
-Furthermore, because of the way they are constructed, they are not forced to be orthogonal to the SAM.
-This allows us to show that Imaginary cEOF2 is closely related to the Southern Annular Mode in the troposphere.
-In fact, it is so highly correlated to the the zonally asymmetric portion of the SAM that we speculate that that the asymmetric component of the SAM might actually a statistical contamination of the PSA mode.
+This mirrors the results of Cai and Watterson (2002), who showed that the CSIRO Model can create PSA-like variability even in the absence of ENSO forcing (i.e.~with a climatological run), but that the variability of one of the PSA modes was enhanced when adding the ENSO signal.
+
+The sensitivity of the phase of the PSAs to the location of the tropical SST anomalies was also seen by Ciasto, Simpkins, and England (2015), who detected similar Rossby wave patterns associated with central Pacific and eastern Pacific SST anomalies but with a change in phase.
+
+Some results of our work contrast with those obtained by Irving and Simmonds (2016) when analysing the PSA.
+They did not find neither a strong association between PSA-like variability and ENSO, nor a relationship between this variability and Australian precipitation.
+Part of the disagreement could be due to the fact that we focus only on SON, which is the season when the teleconnection pattern is strongest, instead of the whole year.
+Moreover, Irving and Simmonds (2016) use of ERA5 forectasted precipitation instead of the more observationally constrained CMAP dataset might also be an issue.
+Additionally, the disagreement could be related to the fact that Irving and Simmonds (2016) show weak relationship between ENSO and their PSA pattern, which explains partly the relationship between Australian rainfall and PSA-like variability.
 
@@
 
-The first complex EOF represents the variability of the zonal wave 1 in the stratosphere, but represents a more zonally symmetric monopole in the troposphere.
+The cEOF1 represents the variability of the zonal wave 1 in the stratosphere, but represents a more zonally symmetric monopole in the troposphere.
 There is a statistically positive trend in the magnitude of this cEOF, which is consistent with previous studies that showed secular changes in springtime wave-1-like patterns (e.g. M. Raphael 2003).
@@
 
-The second complex EOF represents a wave-3 pattern with maximum magnitude in the Pacific sector.
-Essentially, it is an alternate representation of the PSA1 and PSA2 patterns (Mo and Paegle 2001).
-We show that the Imaginary cEOF2 can be roughly identified with the PSA1 and the Real cEOF2 with the PSA2.
-
-The phase of the cEOF2 is related to the the variability of ENSO, but its magnitude appears not to be.
-Except for the three strongest ENSO events in the period which coincide with the three strongest cEOF2 years.
-
-Precipitation anomalies in South America associated with the Imaginary cEOF2 show a clear ENSO-like impact, with positive anomalies in South-Eastern South America, negative anomalies in Southern Brazil and positive anomalies in central Chile.
+The cEOF2 represents a wave-3 pattern with maximum magnitude in the Pacific sector, that is an alternative representation of the PSA1 and PSA2 patterns (Mo and Paegle 2001).
+The Imaginary cEOF2 can be identified with the PSA1 and the Real cEOF2 with the PSA2.
+
+The phase of the cEOF2 is related to the the variability of ENSO, but its magnitude appears not to be, with the exception of the three strongest ENSO events in the period which coincide with the three strongest cEOF2 years.
+
+The cEOF2 variability could be related to surface impacts, while the cEOF1 influence is almost negligible.
+For instance, precipitation anomalies in South America associated with the Imaginary cEOF2 show a clear ENSO-like impact, with positive anomalies in South-Eastern South America, negative anomalies in Southern Brazil and positive anomalies in central Chile.
 Precipitation anomalies associated with the Real cEOF2 are low and not statistically significant in this region, showing that the Imaginary phase is optimally aligned with the direction of maximum precipitation impacts there.
@@
 
+The methodology proposed in this study allow a deeper understanding of the zonally asymmetric springtime extratropical SH circulation such as a better description of PSA like variability using a unique complex index and the understanding of relationship between PSAs and ENSO or SAM variability.
 Further work should extend this analysis to other seasons and further study the relationship between the cEOF2 and the SAM.
-In particular, it is not clear whether the close identification between the Imaginary cEOF2 and the Asymmetric SAM arises from physical processes or statistical contamination.
 

### 2022-04-20 analysis/paper/shceof.Rmd
Agrega comentarios de Leandro

@@
   comment = "#>",
-  cache.extra = 42,
+  cache.extra = 43,
   cache.path = "../cache/paper/",
@@
 
-We restrict this paper to the spring September-October-November (SON) trimester.
-This season experiences a maximum in tropical teleconnections [@cazes-boezio2003a] and stratosphere-troposphere interactions [@lim2018].
+We restrict this paper to the spring September-October-November (SON) trimester because this season experiences a maximum in tropical teleconnections over South America [@cazes-boezio2003] and Southern Hemisphere stratosphere-troposphere interactions [@lim2018].
 
@@
 In Section\ \@ref(tropical) we study tropical forcings that explain the variability of each cEOF.
-Finally, in Section \@ref(precipitation) we show the relationship between these modes of variability and precipitation and surface temperature anomalies in South America and Oceania.
+In Section\ \@ref(precipitation) we show the relationship between these modes of variability and precipitation and surface temperature anomalies in South America and Oceania.
+In Section\ \@ref(discussion) we compare our results with previous studies and discuss the benefits of our methodology.
+In Section\ \@ref(conclusion) we summarise the conclusions.
 
@@
 We derived streamfunction at 200 hPa from ERA5 vorticity using the FORTRAN subroutine FISHPACK [@fishpack] and we computed horizontal wave activity fluxes following @plumb1985.
-Sea Surface Temperature (SST) monthly fields are from Extended Reconstructed Sea Surface Temperature (ERSST) v5 [@huang2017] and precipitation monthly data from the CPC Merged Analysis of Precipitation [@cmap], with a 2\degree and 2.5\degree resolution in latitude and longitude respectively.
+Sea Surface Temperature (SST) monthly fields are from Extended Reconstructed Sea Surface Temperature (ERSST) v5 [@huang2017] and precipitation monthly data from the CPC Merged Analysis of Precipitation [@cmap], with a 2\degree and 2.5\degree horizontal resolution respectively.
 The rainfall gridded dataset is based on information from different sources such as rain gauge observations, satellite inferred estimations and the NCEP-NCAR reanalysis, and it is available since 1979 to present.
@@
 These indices were obtained by projecting the zonally asymmetric and zonally symmetric part of the SAM spatial pattern onto monthly geopotential height fields, as proposed by @campitelli2022.
-Seasonal indices of the Pacific South American patterns (PSA1 and PSA2) were calculated, in agreement with @mo2001, as the third and fourth leading EOF of seasonal mean anomalies for 500-hPa geopotential heights at SH.
+Seasonal indices of the PSA patterns (PSA1 and PSA2) were calculated, in agreement with @mo2001, as the third and fourth leading EOF of seasonal mean anomalies for 500-hPa geopotential heights at SH.
 
@@
 Arbitrary units.
-In parenthesis, the proportion of variance (with respect to the zonal mean) explained.", fig.width=width_column, fig.height=width_column*1.2}
-
+The proportion of variance explained for each mode with respect to the zonal mean is indicated in parenthesis.", fig.width=width_column, fig.height=width_column*1.2}
 gdata <- ceof[, eof[[1]]$right, by = .(season)] %>% 
@@
 
-Given the similarity between the cEOF2 regression maps (Fig. \@ref(fig:eof2-regr-gh)) and documented PSAs we study the relationship between these modes.
+Given the similarity between the cEOF2 regression maps (Fig. \@ref(fig:eof2-regr-gh)) and documented PSA patterns we study the relationship between these modes.
 Table \@ref(tab:psa-eof2) shows the correlations between the two PSA indices and the time series for Real and Imaginary phases of cEOF2.
@@
 Indeed, the Imaginary cEOF2 shares up to `r scales::percent(max_r2[cEOF == "cEOF2" & term == "asym"]$estimate^2, 2)` variance with the A-SAM and only `r scales::percent(max_r2[cEOF == "cEOF2" & term == "sym"]$estimate^2, 2)` at most with the S-SAM (Figure \@ref(fig:sam-eof-vertical).b2).
-Such extremely high correlation between A-SAM and Imaginary cEOF2 suggests that the methodology proposed in this work is able to characterise the zonally asymmetric component of the SAM described previously by @campitelli2022.
+Such extremely high correlation between A-SAM and Imaginary cEOF2 suggests that the modes obtained in this work are able to characterise the zonally asymmetric component of the SAM described previously by @campitelli2022.
 
@@
 As a result, the correlation between the Real cEOF2 and ENSO is also not significant  (`r enso_cor[cEOF == "cEOF2" & part == "Real"]$text`).
-Meanwhile, Rows a and c in Fig.\ \@ref(fig:sst-psi-2) show that the intermediate phases are still associated with significant SST anomalies over the Pacific Ocean, but with with slightly different location.
+Meanwhile, Rows a and c in Fig.\ \@ref(fig:sst-psi-2) show that the intermediate phases are still associated with significant SST regressed anomalies over the Pacific Ocean, but with with slightly different location.
 The 135º phase is associated with SST in the central Pacific (Fig.\ \@ref(fig:sst-psi-2)a.1), while the 45º phase is associated with SST in the eastern Pacific (Fig.\ \@ref(fig:sst-psi-2)c.1), which correspond roughly to the Central Pacific and Eastern Pacific "flavours" of ENSO, respectively [@kao2009].
@@
 
-equation <- latex2exp::TeX(paste0("$\\mathrm{ONI} = ", 
-                                  signif(coef(cos_model)[1], 2), "\\sin{(\\mathrm{phase})}$"))
+equation <- latex2exp::TeX(paste0("$ONI = ", 
+                                  signif(coef(cos_model)[1], 2), 
+                                  "sin(Phase)$"))
 
@@
 
-Over the continents, the Imaginary cEOF2 (Fig.\ \@ref(fig:pp-temp-2)b.1) is associated with positive temperatures in southern Australia and negative temperature in southern South America and the Antarctic Peninsula, that are a result of the wave train described before.
+Over the continents, the Imaginary cEOF2 (Fig.\ \@ref(fig:pp-temp-2)b.1) is associated with positive regressed temperature anomalies in southern Australia and negative regressed anomalies in southern South America and the Antarctic Peninsula, that are a result of the wave train described before.
 
@@
 They did not find neither a strong association between PSA-like variability and ENSO, nor a relationship between this variability and Australian precipitation.
-Part of the disagreement could be due to the fact that we focus only on SON, which is the season when the teleconnection pattern is strongest, instead of the whole year.
-Moreover, @irving2016 use of ERA5 forectasted precipitation instead of the more observationally constrained CMAP dataset might also be an issue.
-Additionally, the disagreement could be related to the fact that @irving2016 show weak relationship between ENSO and their PSA pattern, which explains partly the relationship between Australian rainfall and PSA-like variability.
+Part of the disagreement could be due to the fact that we focus our analysis only on SON, which is the season when the teleconnection patterns are strongest, instead of the whole year.
+Moreover, @irving2016 use of ERA5 forectasted precipitation instead of the more observationally constrained CMAP dataset might be another source of disagreement.
+Additionally, the discrepancy could be related to the fact that @irving2016 show a weak relationship between ENSO and their PSA pattern, which explains partly the relationship between Australian rainfall and PSA-like variability.
 
@@
 The cEOF2 variability could be related to surface impacts, while the cEOF1 influence is almost negligible. 
-For instance, precipitation anomalies in South America associated with the Imaginary cEOF2 show a clear ENSO-like impact, with positive anomalies in South-Eastern South America, negative anomalies in Southern Brazil and positive anomalies in central Chile.
+For instance, precipitation anomalies in South America associated with the Imaginary cEOF2 show a clear ENSO-like impact, with positive anomalies in South-Eastern South America, negative anomalies in Southern Brazil and positive anomalies in central Chile for positive Imaginary cEOF2 phase.
 Precipitation anomalies associated with the Real cEOF2 are low and not statistically significant in this region, showing that the Imaginary phase is optimally aligned with the direction of maximum precipitation impacts there.
@@
 On the other hand, over Australia, both the Real and Imaginary phase are associated with significant precipitation anomalies, and we further show that the direction of maximum impact is not aligned with our chosen rotation of cEOF2.
-An analysis of the precipitation impacts on this regions could benefit for a differnet chosen rotation of the cEOF.
+An analysis of the precipitation impacts on this regions could benefit for a different chosen rotation of the cEOF.
 

### 2022-04-20 analysis/paper/shceof.Rmd
Agrega abstract to meta

@@
 keywords:
-
-abstract: |
-  abstract
+  - Southern Hemisphere circulation
+  - Teleconnections
+  - Pacific South American Mode
+  - Southern Annular Mode
+  - Stratosphere
 
@@
 output: 
-  bookdown::word_document2: default
   bookdown::pdf_book: 
@@
       # citation_package: "natbib"
-      # pandoc_args: ["--lua-filter=abstract-to-meta.lua"]
+      pandoc_args: ["--lua-filter=abstract-to-meta.lua"]
 always_allow_html: true
@@
 ```
+# Abstract
+
+asdfsdf
 

### 2022-04-20 analysis/paper/shceof.Rmd
Quita patchwork y purrr

@@
 library(ggperiodic)
-library(patchwork)
+# library(patchwork)
 library(kableExtra)

### 2022-04-20 R/maps.R

@@ -26,8 +26,6 @@ geom_qmap <- function(subset = identity,
     data <- sf::st_crop(data, bbox)
   }
 
-
-  subset <- purrr::as_mapper(subset)
   data <- subset(data)
 
   ggplot2::geom_sf(data = data,

### 2022-04-20 analysis/paper/shceof.Rmd
Quita ggthemes y gg4hx

--- 2022-04-20 analysis/paper/shceof.Rmd
@@
 library(ggperiodic)
-library(patchwork)
 library(kableExtra)
@@
              aes(yintercept = V1), size = 0.2, color = "gray30") +
-  ggthemes::geom_rangeframe(sides = "r", size = 0.2, color = "gray30") +
+  geom_segment(data = ~.x[, .(min = min(hgt)*1e3,
+                              max = max(hgt)*1e3), by  = .(cEOF, part)],
+               aes(x = structure(Inf, class = "Date"),
+                   xend = structure(Inf, class = "Date"),
+                   y = min,
+                   yend = max), size = 0.2, color = "gray30") +
   geom_line(aes(color = part)) +
@@
   if (length(which_cEOF) > 1) {
-    facet <- ggh4x::facet_nested(cEOF ~  term)
-  } else {
-    facet <- ggh4x::facet_nested(lev ~ cEOF +  term, 
-                                 labeller = labeller(lev = lab_lev))
+    stop("More than one ceof")
   }
@@
     grid_y +
-    facet +
+    facet_grid(lev ~ term, 
+               labeller = labeller(lev = lab_lev)) +
     geom_coords() +
     coord_polar() +
-    tag_facets("rc")
+    tag_facets("rc") +
+    labs(title = which_cEOF) +
+    theme(plot.title = element_text(hjust = 0.5, size = theme_get()$strip.text$size))
 }

### 2022-04-20 R/jacobi.R

@@ -1,241 +1,0 @@
-if (FALSE) {
-lapl <- function(y, div)   {
-
-  ddx <- t(apply(y, 1, function(x) metR:::.derv(x, matrix$coldims$lat,
-                                                order = 2, cyclical = FALSE, fill = TRUE)))
-  ddy <- apply(y, 2, function(x) metR:::.derv(x, matrix$rowdims$lon, order = 2, cyclical = TRUE))
-  lap <- ddx + ddy + div
-
-
-  return(dY)
-}
-library(metR)
-data <- "~/DATOS/NCEP Reanalysis/psi.mon.mean.nc" %>%
-  ReadNetCDF(subset = list(level = 0.1682,
-                           time = c("2020-01-01"))) %>%
-  .[, psi := Anomaly(psi), by = lat]
-
-
-
-
-laplacian <- function(x, lon, lat) {
-  xbk <- x
-  x <- x$matrix
-  prev <- function(N) c(N, seq_len(N)[-N])
-  nex <- function(N) c(seq_len(N)[-1], 1)
-
-  lon <- lon*pi/180
-  lat <- (lat+90)*pi/180
-
-  dlon <- diff(lon)[1]
-  dlat <- diff(lat)[1]
-
-  Nrow <- nrow(x)
-  Ncol <- ncol(x)
-
-  sinlat <- sin(lat)
-  sinplus <- sin(lat + dlat/2)
-  sinminus <- sin(lat - dlat/2)
-
-  lap <- 1/dlat^2*(sinplus*(x[nex(Nrow), ] - x) - sinminus*(x - x[prev(Nrow), ])) +
-    1/(dlon^2*sinlat^2)*(-2*x + x[, nex(Ncol)] + x[, prev(Ncol)])
-
-
-  # lap[c(1, Nrow), ] <- lap[c(2, Nrow-1), ]
-  xbk$matrix <- lap
-  xbk
-}
-
-jacob_laplacian <- function(x, lon, lat, forcing) {
-  xbk <- x
-  x <- x$matrix
-  prev <- function(N) c(N, seq_len(N)[-N])
-  nex <- function(N) c(seq_len(N)[-1], 1)
-
-  lon <- lon*pi/180
-  lat <- (lat+90)*pi/180
-
-  dlon <- diff(lon)[1]
-  dlat <- diff(lat)[1]
-
-  Nrow <- nrow(x)
-  Ncol <- ncol(x)
-
[…]

### 2022-04-20 analysis/paper/shceof.Rmd
Quita latex2expr

@@
 
+r2 <- expression(`$r^2$` = paste("", "r", phantom()^{
+  paste("2")
+}, "", ""))
 ```
@@
 ceof <- ceof[rotation == "rotated"][, rotation := NULL][]
-```
-
-(ref:rotations-cap) a) Correlation of the Real and Imaginary cEOF1 with wave 1 amplitude of Total Ozone Column averaged between `r knitr::combine_words(LatLabel(o3wave_lats))` and b) $R^2$ between the Real and Imaginary cEOF2 and the ONI index for different values of rotation.
-The vertical line marks the selected rotation, which maximises de correlation.
-
-```{r rotations, fig.cap="(ref:rotations-cap)", fig.height=2.5, fig.width=width_column, include=FALSE}
-r2 <- expression(`$r^2$` = paste("", "r", phantom()^{
-  paste("2")
-}, "", ""))
-
-rotations_o3 %>%
-  ggplot(aes(rotation, cor)) +
-  geom_vline(xintercept = best_rotation_cEOF1) +
-  geom_hline(yintercept = 0,  size = 0.2, color = "gray50") +
-  geom_line(aes(color = part)) +
-  scale_x_continuous("Rotation", breaks = seq(-pi, pi, by = 90*pi/180),
-                     labels = function(x) paste0(round(x*180/pi, 5), "º")) +
-  scale_y_continuous("Correlation") +
-  scale_color_brewer(NULL, palette = "Dark2", labels = lab_cplx) +
-
-  rotations_cEOF2 %>%
-  melt(id.vars = "rotation") %>%
-  ggplot(aes(rotation, value^2)) +
-  geom_hline(yintercept = 0,  size = 0.2, color = "gray50") +
-  geom_vline(xintercept = best_rotation_cEOF2) +
-  geom_line(aes(color = variable)) +
-  scale_x_continuous("Rotation", breaks = seq(-pi, pi, by = 90*pi/180),
-                     labels = function(x) paste0(round(x*180/pi, 5), "º")) +
-  scale_y_continuous(r2) +
-  scale_color_brewer(NULL, palette = "Dark2", labels = lab_cplx) +
-
-  plot_layout(guides = "collect") +
-  plot_annotation(tag_levels = "a", tag_suffix = ")")
-
 ```
@@
 
-mgp <- latex2exp::TeX('Geopotential height ($m^2s^{-1}$)')
+# From latex2exp::TeX('Geopotential height ($m^2s^{-1}$)')
+mpg <- expression('Geopotential height ('*m^{2}*s^{phantom() - 1}*')')
 ```
@@
 plot_regr(regr , "cEOF1", c(200, 50), "hgt") +
-  scale_fill_divergent(mgp,
+  scale_fill_divergent(mpg,
                        breaks = AnchorBreaks(0, exclude = 0),
@@
 plot_regr(regr, "cEOF2", c(200, 50), "hgt") +
-  scale_fill_divergent(mgp, 
+  scale_fill_divergent(mpg, 
                        breaks = AnchorBreaks(0, 200, exclude = 0),
@@
 
-temp %>% 
-  ggplot(aes(Mod(hgt), amplitude)) +
-  scale_x_continuous("EOF1 Amplitude") +
-  scale_y_continuous("O3 Wave-1\nAmplitude") +
-
-  temp %>% 
-  copy() %>%
-  na.omit() %>% 
-  qwrap(phase = c(0, 2*pi) ~ c(-pi, pi)) %>% 
-  ggplot(aes(Arg(hgt), phase)) +
-  scale_x_continuous("EOF1 Phase",
-                     breaks = seq(-pi, pi, by = 15*pi/180),
-                     labels = function(x) paste0(round(x*180/pi, 5), "º")) +
-  scale_y_continuous("O3 Wave-1\nPhase",
-                     breaks = seq(-pi, pi, by = 30*pi/180),
-                     labels = function(x) paste0(round(x*180/pi, 5), "º"))  +
-  plot_layout(ncol  = 1) &
-  geom_point()  &
-  geom_smooth(method = "lm") & 
-  panel_background
+# temp %>% 
+#   ggplot(aes(Mod(hgt), amplitude)) +
+#   scale_x_continuous("EOF1 Amplitude") +
+#   scale_y_continuous("O3 Wave-1\nAmplitude") +
+#   
+#   temp %>% 
+#   copy() %>%
+#   na.omit() %>% 
+#   qwrap(phase = c(0, 2*pi) ~ c(-pi, pi)) %>% 
+#   ggplot(aes(Arg(hgt), phase)) +
+#   scale_x_continuous("EOF1 Phase",
+#                      breaks = seq(-pi, pi, by = 15*pi/180),
+#                      labels = function(x) paste0(round(x*180/pi, 5), "º")) +
+#   scale_y_continuous("O3 Wave-1\nPhase",
+#                      breaks = seq(-pi, pi, by = 30*pi/180),
+#                      labels = function(x) paste0(round(x*180/pi, 5), "º"))  +
+#   plot_layout(ncol  = 1) &
+#   geom_point()  &
+#   geom_smooth(method = "lm") & 
+#   panel_background
 
@@
 
+  #from latex2exp::TeX("Streamfunction ($m^2s^{-1}\\times 1e^{-7}$)")
+  psi_expr <- expression('Streamfunction ('*m^{2}*s^{phantom() - 1} %*% 1*e^{phantom() - 7}*')')
   # browser()
@@
     geom_qmap(keep = .015) +
-    scale_fill_divergent_discretised(latex2exp::TeX("Streamfunction ($m^2s^{-1}\\times 1e^{-7}$)"), 
+    scale_fill_divergent_discretised(psi_expr, 
                                      labels = one_less,
@@
 
-```{r enso-phase, fig.cap = "(ref:enso-phase-cap)", fig.width = width_column, fig.height=3, include = TRUE}
+```{r enso-phase, fig.cap = "(ref:enso-phase-cap)", fig.width = width_column, fig.height=3}
 gdata <- ceof[, eof[[1]]$left, by = .(season)] %>%
@@
 
-equation <- latex2exp::TeX(paste0("$ONI = ", 
-                                  signif(coef(cos_model)[1], 2), 
-                                  "sin(Phase)$"))
+equation <- bquote(ONI  * {phantom() == phantom()} * .(signif(coef(cos_model)[1], 2)) * sin(Phase) )
 
@@
   ggrepel::geom_text_repel(data = ~.x[order(-oni)][1:3],
-                           aes(label = year(time))) +
+                           aes(label = year(time)), size = 2) +
 

### 2022-04-20 analysis/paper/shceof.Rmd
Elimina rmapshaper

@@
     scale_mag(max_size = .5, guide = "none") +
-    geom_qmap(keep = .015) +
+    geom_qmap() +
     scale_fill_divergent_discretised(psi_expr, 
@@
   #                   skip = 1, stroke = 0.15, stroke.colour = "white") +
-  geom_qmap(keep = .015) +
+  geom_qmap() +
   scale_fill_divergent_discretised(name = "Explained variance", midpoint = 0,

### 2022-04-20 R/maps.R

@@ -3,18 +3,15 @@
 #'
 #' @param subset a function that will be applied to the map data.
 #' @param color,size,fill parameters passed to [ggplot2::geom_polygon()]
-#' @param wrap vector of length 2 interpreted as the longitude range of a global map (see [maps::map()])
-#' @param keep proportions of point to retain (passed to [rmapshaper::ms_simplify()])
 #' @param ... other arguments passed to [ggplot2::geom_polygon()]
 #'
 #' @export
 geom_qmap <- function(subset = identity,
                       crop = NULL,
                       color = "gray50", size = 0.3,
-                      fill = NA, wrap = c(0, 360), weighting = 0.7,
-                      keep = 0.015, ...) {
+                      fill = NA, ...) {
   lon <- lat <- group <- NULL
-  data <- map_simple(wrap = wrap, keep  = keep, weighting = weighting)
+  data <- simplemap
 
   if (!is.null(crop)) {
     bbox <- sf::st_bbox(data)
@@ -37,23 +34,11 @@ geom_qmap <- function(subset = identity,
 
 }
 
-map_simple <- function(wrap = c(0, 360), keep = 0.015, weighting = 0.7) {
-  map <- maps::map("world", fill = TRUE,
-                   col = "transparent", plot = FALSE, wrap = wrap)
-  map <- sf::st_as_sf(map)
-  if (keep != 1) {
-    map <- rmapshaper::ms_simplify(map, keep = keep, weighting = weighting)
-  }
-
-
-  map
-}
 
 fortify <- ggplot2::fortify
 
 # zzz.R
 .onLoad <- function(lib, pkg) {
-  map_simple <<- memoise::memoise(map_simple)
   fortify <<- memoise::memoise(fortify)
 }
 

### 2022-04-20 data-raw/simplemap.R

@@ -1,0 +1,8 @@
+## code to prepare `simplemap` dataset goes here
+
+simplemap <- maps::map("world", fill = TRUE,
+                 col = "transparent", plot = FALSE, wrap = c(0, 360))
+simplemap <- sf::st_as_sf(simplemap)
+simplemap <- rmapshaper::ms_simplify(simplemap, keep = 0.015, weighting = 0.7)
+
+usethis::use_data(simplemap, overwrite = TRUE, internal = TRUE)

### 2022-04-20 analysis/paper/shceof.tex
Render

@@
 \begin{abstract}
-abstract
+asdfsdf
 \\
 \keywords{
-         \and
+        Southern Hemisphere circulation \and
+        Teleconnections \and
+        Pacific South American Mode \and
+        Southern Annular Mode \and
+        Stratosphere \and
     }
@@
 
-We restrict this paper to the spring September-October-November (SON) trimester.
-This season experiences a maximum in tropical teleconnections (Cazes-Boezio, Robertson, and Mechoso 2003) and stratosphere-troposphere interactions (Lim, Hendon, and Thompson 2018).
+We restrict this paper to the spring September-October-November (SON) trimester because this season experiences a maximum in tropical teleconnections over South America (Cazes-Boezio, Robertson, and Mechoso 2003) and Southern Hemisphere stratosphere-troposphere interactions (Lim, Hendon, and Thompson 2018).
 
@@
 In Section~\ref{tropical} we study tropical forcings that explain the variability of each cEOF.
-Finally, in Section \ref{precipitation} we show the relationship between these modes of variability and precipitation and surface temperature anomalies in South America and Oceania.
+In Section~\ref{precipitation} we show the relationship between these modes of variability and precipitation and surface temperature anomalies in South America and Oceania.
+In Section~\ref{discussion} we compare our results with previous studies and discuss the benefits of our methodology.
+In Section~\ref{conclusion} we summarise the conclusions.
 
@@
 2017) and precipitation monthly data from the CPC Merged Analysis of Precipitation (P.
-Xie and Arkin 1997), with a 2\degree and 2.5\degree resolution in latitude and longitude respectively.
+Xie and Arkin 1997), with a 2\degree and 2.5\degree horizontal resolution respectively.
 The rainfall gridded dataset is based on information from different sources such as rain gauge observations, satellite inferred estimations and the NCEP-NCAR reanalysis, and it is available since 1979 to present.
@@
 These indices were obtained by projecting the zonally asymmetric and zonally symmetric part of the SAM spatial pattern onto monthly geopotential height fields, as proposed by Campitelli, Díaz, and Vera (2022).
-Seasonal indices of the Pacific South American patterns (PSA1 and PSA2) were calculated, in agreement with Mo and Paegle (2001), as the third and fourth leading EOF of seasonal mean anomalies for 500-hPa geopotential heights at SH.
+Seasonal indices of the PSA patterns (PSA1 and PSA2) were calculated, in agreement with Mo and Paegle (2001), as the third and fourth leading EOF of seasonal mean anomalies for 500-hPa geopotential heights at SH.
 
@@
 Arbitrary units.
-In parenthesis, the proportion of variance (with respect to the zonal mean) explained.}
+The proportion of variance explained for each mode with respect to the zonal mean is indicated in parenthesis.}
 \end{figure}
@@
 
-Given the similarity between the cEOF2 regression maps (Fig.~\ref{fig:eof2-regr-gh}) and documented PSAs we study the relationship between these modes.
+Given the similarity between the cEOF2 regression maps (Fig.~\ref{fig:eof2-regr-gh}) and documented PSA patterns we study the relationship between these modes.
 Table~\ref{tab:psa-eof2} shows the correlations between the two PSA indices and the time series for Real and Imaginary phases of cEOF2.
@@
 Indeed, the Imaginary cEOF2 shares up to 92\% variance with the A-SAM and only 12\% at most with the S-SAM (Figure~\ref{fig:sam-eof-vertical}.b2).
-Such extremely high correlation between A-SAM and Imaginary cEOF2 suggests that the methodology proposed in this work is able to characterise the zonally asymmetric component of the SAM described previously by Campitelli, Díaz, and Vera (2022).
+Such extremely high correlation between A-SAM and Imaginary cEOF2 suggests that the modes obtained in this work are able to characterise the zonally asymmetric component of the SAM described previously by Campitelli, Díaz, and Vera (2022).
 
@@
 As a result, the correlation between the Real cEOF2 and ENSO is also not significant (0 (CI: -0.3 -- 0.3)).
-Meanwhile, Rows a and c in Fig.~\ref{fig:sst-psi-2} show that the intermediate phases are still associated with significant SST anomalies over the Pacific Ocean, but with with slightly different location.
+Meanwhile, Rows a and c in Fig.~\ref{fig:sst-psi-2} show that the intermediate phases are still associated with significant SST regressed anomalies over the Pacific Ocean, but with with slightly different location.
 The 135º phase is associated with SST in the central Pacific (Fig.~\ref{fig:sst-psi-2}a.1), while the 45º phase is associated with SST in the eastern Pacific (Fig.~\ref{fig:sst-psi-2}c.1), which correspond roughly to the Central Pacific and Eastern Pacific ``flavours'' of ENSO, respectively (Kao and Yu 2009).
@@
 
-Over the continents, the Imaginary cEOF2 (Fig.~\ref{fig:pp-temp-2}b.1) is associated with positive temperatures in southern Australia and negative temperature in southern South America and the Antarctic Peninsula, that are a result of the wave train described before.
+Over the continents, the Imaginary cEOF2 (Fig.~\ref{fig:pp-temp-2}b.1) is associated with positive regressed temperature anomalies in southern Australia and negative regressed anomalies in southern South America and the Antarctic Peninsula, that are a result of the wave train described before.
 
@@
 They did not find neither a strong association between PSA-like variability and ENSO, nor a relationship between this variability and Australian precipitation.
-Part of the disagreement could be due to the fact that we focus only on SON, which is the season when the teleconnection pattern is strongest, instead of the whole year.
-Moreover, Irving and Simmonds (2016) use of ERA5 forectasted precipitation instead of the more observationally constrained CMAP dataset might also be an issue.
-Additionally, the disagreement could be related to the fact that Irving and Simmonds (2016) show weak relationship between ENSO and their PSA pattern, which explains partly the relationship between Australian rainfall and PSA-like variability.
+Part of the disagreement could be due to the fact that we focus our analysis only on SON, which is the season when the teleconnection patterns are strongest, instead of the whole year.
+Moreover, Irving and Simmonds (2016) use of ERA5 forectasted precipitation instead of the more observationally constrained CMAP dataset might be another source of disagreement.
+Additionally, the discrepancy could be related to the fact that Irving and Simmonds (2016) show a weak relationship between ENSO and their PSA pattern, which explains partly the relationship between Australian rainfall and PSA-like variability.
 
@@
 The cEOF2 variability could be related to surface impacts, while the cEOF1 influence is almost negligible.
-For instance, precipitation anomalies in South America associated with the Imaginary cEOF2 show a clear ENSO-like impact, with positive anomalies in South-Eastern South America, negative anomalies in Southern Brazil and positive anomalies in central Chile.
+For instance, precipitation anomalies in South America associated with the Imaginary cEOF2 show a clear ENSO-like impact, with positive anomalies in South-Eastern South America, negative anomalies in Southern Brazil and positive anomalies in central Chile for positive Imaginary cEOF2 phase.
 Precipitation anomalies associated with the Real cEOF2 are low and not statistically significant in this region, showing that the Imaginary phase is optimally aligned with the direction of maximum precipitation impacts there.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-cazes-boezio2003a}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-cazes-boezio2003}{}}%
 Cazes-Boezio, Gabriel, Andrew W.

### 2022-04-20 analysis/paper/shceof.Rmd
Render

@@
   comment = "#>",
-  cache.extra = 41,
+  cache.extra = 46,
   cache.path = "../cache/paper/",

### 2022-04-20 R/0-download.R

@@ -48,7 +48,7 @@ download_ozone <- function(request) {
                                 path = tempdir())
 
 
-    files <- unzip(file_zip, exdir = tempdir())
+    files <- utils::unzip(file_zip, exdir = tempdir())
 
     data <- do.call(rbind, lapply(files, metR::ReadNetCDF))
     data <- normalise_coords(data)
@@ -206,7 +206,7 @@ request_era5_2mt <- list(
 simple_download <- function(url) {
   force(url)
   function(file){
-    download.file(url, destfile = file)
+    utils::download.file(url, destfile = file)
     file
   }
 }
@@ -223,7 +223,7 @@ download_ersst <- function(file) {
 
   files <- vapply(urls, function(url) {
     file <- tempfile(fileext = ".nc")
-    download.file(url, file)
+    utils::download.file(url, file)
     file
   }, character(1))
 

### 2022-04-20 R/WaveFlux.R

@@ -1,12 +1,18 @@
+#' Computes wave flux
+#'
+#' @param psi data.table with columns lon, lat and psi.z
+#' @param p pressure level in hPa
+#' @param a radius of the Earth in meters
+#'
 #' @export
 WaveFlux <- function(psi, p = 250, a = 6371000) {
   k <- p*100/(a^2*2000)
-  psi <- copy(psi)
-  psi[, c("psi.dlon", "psi.dlat") := Derivate(psi.z ~ lon + lat,
-                                              cyclical = c(TRUE, FALSE))] %>%
-    .[, psi.ddlon := Derivate(psi.z ~ lon, cyclical = TRUE, order = 2),
+  psi <- data.table::copy(psi)
+  psi[, c("psi.dlon", "psi.dlat") := metR::Derivate(psi.z ~ lon + lat,
+                                                    cyclical = c(TRUE, FALSE))] %>%
+    .[, psi.ddlon := metR::Derivate(psi.z ~ lon, cyclical = TRUE, order = 2),
       by = lat] %>%
-    .[, psi.dlondlat := Derivate(psi.dlon ~ lat),
+    .[, psi.dlondlat := metR::Derivate(psi.dlon ~ lat),
       by = lon] %>%
     .[, `:=`(f.lon = k/cos(lat*pi/180)*(psi.dlon^2 - psi.z*psi.ddlon),
              f.lat = k*(psi.dlon*psi.dlat - psi.z*psi.dlondlat))]

### 2022-04-20 R/compute-ceof.R

@@ -1,8 +1,9 @@
 #' Computes ceof
 #'
-#' @param dt data.table.
+#' @param hgt vector of geopotential height observations
+#' @param lon,lat.lev vectors of locations of geopotential height observations
 #' @param temporal Logical whether to compute temporal anomalies.
-#' @param lats.eof Range of latitudes.
+#' @param lats.eof Range of latitudes to use.
 #' @param n Numeric vector of EOFs to compute.
 #'
 #' @export
@@ -14,37 +15,19 @@ compute_ceof <- function(hgt, lon, lat, lev, temporal = FALSE, lats.eof = c(-90,
   }
 
   dt <- dt[lev %in% c(50, 200) & lat %between% lats.eof] %>%
-    .[, hgt := Anomaly(hgt),
+    .[, hgt := metR::Anomaly(hgt),
       by  = .(lat, time, lev)]
 
   if (temporal) {
-    dt <- dt[, hgt := Anomaly(hgt), by = .(lon, lat, lev)]
+    dt <- dt[, hgt := metR::Anomaly(hgt), by = .(lon, lat, lev)]
   }
 
   dt %>%
-    .[, hgt := hgt/sd(hgt), by = .(lev)] %>%
+    .[, hgt := hgt/stats::sd(hgt), by = .(lev)] %>%
     .[, hgt := hgt*sqrt(cos(lat*pi/180))] %>%
     .[, hgt.cpx := spectral::analyticFunction(hgt),
       by = .(lat, time, lev)] %>%
     .[, hgt := hgt.cpx] %>%
-    EOF(hgt ~ time | lon + lat + lev, n = n, suffix = "cEOF", data = .)
+    metR::EOF(hgt ~ time | lon + lat + lev, n = n, suffix = "cEOF", data = .)
 }
 
-# compute_ceof <- function(dt, temporal = FALSE, lats = c(-80, -20), n = 1:2) {
-#   # browser()
-#   dt <- dt[lev %in% c(50, 200) & lat %between% lats] %>%
-#     .[, hgt := hgt - mean(hgt),
-#       by  = .(lat, year, lev)]
-#
-#   if (temporal) {
-#     dt <- dt[, hgt := Anomaly(hgt), by = .(lon, lat, lev)]
-#   }
-#
-#   dt %>%
-#     .[, hgt := hgt/sd(hgt), by = .(lev)] %>%
[…]

### 2022-04-20 R/correlate.R

@@ -10,7 +10,7 @@
 #'
 #' @export
 correlate <- function(x, y, signif = 2, sep = "\ ") {
-  correlation <- cor.test(x, y)
+  correlation <- stats::cor.test(x, y)
   out <- with(correlation,
               list(estimate = estimate,
                    p.value = p.value,

### 2022-04-20 R/ditto.R

@@ -1,81 +1,0 @@
-#' Propagate mapping and parameters
-#'
-#' @param expr A ggplot-like expression consisting on
-#'
-#' @export
-ditto <- function(expr) {
-  plus_fun <- function(e1, e2) {
-    if (is.list(e1)) {
-      c(e1, list(e2))
-    } else {
-      list(e1, e2)
-    }
-
-  }
-
-  geom_list <- eval(substitute(expr), envir = list("+.gg" = plus_fun), enclos = parent.frame())
-  e1 <- geom_list[[1]]
-  for (i in seq_along(geom_list)[-1]) {
-    e2 <- geom_list[[i]]
-
-    if (inherits(e2, "Scale")) {
-      geom_list[[i]] <- stomp(e2, c(e1$geom_params, e1$aes_parama,  e1$stat_params))
-    } else {
-      geom_list[[i]]$geom_params <- stomp(e2$geom_params,
-                                          e1$geom_params)
-      geom_list[[i]]$aes_params <- stomp(e2$aes_params,
-                                         e1$aes_params)
-      geom_list[[i]]$stat_params <- stomp(e2$stat_params,
-                                          e1$stat_params)
-      geom_list[[i]]$mapping <- e1$mapping
-
-    }
-
-
-  }
-
-  geom_list
-}
-
-#' @export
-ditto_params <- function(expr) {
-  plus_fun <- function(e1, e2) {
-    if (is.list(e1)) {
-      c(e1, list(e2))
-    } else {
-      list(e1, e2)
-    }
-
-  }
-
-  geom_list <- eval(substitute(expr), envir = list("+.gg" = plus_fun), enclos = parent.frame())
-  e1 <- geom_list[[1]]
-  for (i in seq_along(geom_list)[-1]) {
-    e2 <- geom_list[[i]]
-
-    if (inherits(e2, "Scale")) {
-      geom_list[[i]] <- stomp(e2, c(e1$geom_params, e1$aes_parama,  e1$stat_params))
-    } else {
-      geom_list[[i]]$geom_params <- stomp(e2$geom_params,
-                                          e1$geom_params)
-      geom_list[[i]]$aes_params <- stomp(e2$aes_params,
-                                         e1$aes_params)
-      geom_list[[i]]$stat_params <- stomp(e2$stat_params,
-                                          e1$stat_params)
-    }
-
-
-  }
-
-  geom_list
-
-
-}
-
-stomp <- function(original, new) {
[…]

### 2022-04-20 R/eof.R

@@ -6,7 +6,7 @@
 #'
 #' @export
 flip_ceof <- function(eof, n = NULL) {
-  eof <- copy(eof)
+  eof <- data.table::copy(eof)
 
   pc_col <- attr(eof, "suffix", TRUE)
   val_col <- attr(eof, "value.var", TRUE)

### 2022-04-20 R/filter.R

@@ -1,9 +1,0 @@
-#' @export
-is.sa <- function(data, lons = c(265, 325), lats = c(-80, -10)) {
-  data[, lon %between% lons & lat %between% lats]
-}
-
-#' @export
-filter_sa <- function(data, lons = c(265, 325), lats = c(-60, -10)) {
-  data[is.sa(data, lons = lons, lats = lats)]
-}

### 2022-04-20 R/invert-laplacian.R

@@ -1,33 +1,0 @@
-#' @export
-invert_laplacian <- function(v, x, y) {
-  a <- 6371000
-  field <- data.table::data.table(x, y, v)
-
-  field[, c("k", "l", "v.hat") := fft2d(v, x, y)]
-  field[, lp := l*2*pi/a]
-  field[, kp := k*2*pi/(a*cos(y*pi/180))]
-
-  field[, v.hat := -v.hat/(k^2 + l^2)]
-  # field[, v.hat := -v.hat/(kp^2 + lp^2)]
-  field[, v.hat := c(0, v.hat[2:.N])]
-
-  field[, c("k", "l", "v.inv") := fft2d(v.hat, k, l, inverse = TRUE)]
-
-  return(Re(field$v.inv))
-}
-
-fft2d <- function(v, x, y, inverse = FALSE) {
-
-  field <- as.matrix(reshape2::dcast(data.table::data.table(v, x, y),
-                                     y ~ x, value.var = "v"))
-
-  field.fft <- fft(field[, -1], inverse = inverse)
-
-  dimnames(field.fft) <- list(
-    k = 1:nrow(field.fft) - 1,
-    l = 1:ncol(field.fft) - 1)
-
-  field.fft <- setDT(melt(field.fft, value.name = "f"))
-
-  with(field.fft, list(k, l, f))
-}

### 2022-04-20 R/lm_lite.R

@@ -36,15 +36,6 @@ var.wt <- function(x, w, na.rm = FALSE) {
   (sum(w*x^2) * sum.w - sum(w*x)^2) / (sum.w^2 - sum(w^2))
 }
 
-#' @export
-Detrend <- function(y, x) {
-   nas <- is.na(y)
-   m <- mean(y, na.rm = TRUE)
-   if (!hasArg(x)) x <- seq_along(y)
-   y[!nas] <- .lm.fit(cbind(1, x[!nas]), y[!nas])$residuals
-   return(y + m)
-}
-
 #' Compute p-values
 #'
 #' Computes p-values from the result of [metR::FitLm()] (or any assumed student-t-distributed

### 2022-04-20 R/maps.R

@@ -2,6 +2,7 @@
 #'
 #'
 #' @param subset a function that will be applied to the map data.
+#' @param crop optional bounding box to crop the map.
 #' @param color,size,fill parameters passed to [ggplot2::geom_polygon()]
 #' @param ... other arguments passed to [ggplot2::geom_polygon()]
 #'
@@ -35,17 +36,9 @@ geom_qmap <- function(subset = identity,
 }
 
 
-fortify <- ggplot2::fortify
-
-# zzz.R
-.onLoad <- function(lib, pkg) {
-  fortify <<- memoise::memoise(fortify)
-}
-
-
 
 `%||%` <- function (x, y) {
-  if (is_null(x))
+  if (is.null(x))
     y
   else x
 }

### 2022-04-20 R/on_unique.R

@@ -1,6 +1,6 @@
 #' Applies a function to a vector only once for each value
 #'
-#' In cases when you have a large vector with a lot of repreated values
+#' In cases when you have a large vector with a lot of repeated values
 #' (for example, the dimension columns of a spatial field in a  data.table)
 #' applying a function to each unique value of a vector and then repeating
 #' the result is faster and more memory efficient than applying the function
@@ -9,7 +9,7 @@
 #' @param x, vector
 #' @param f, function to apply. It needs to be a function that takes 1 value and
 #' returns 1 value.
-#' @param ..., aditional arguments passed to `f`
+#' @param ..., additional arguments passed to `f`
 #'
 #' @returns
 #' a vector the same length as `x`

### 2022-04-20 R/season_weight.R

@@ -16,10 +16,10 @@
 #'
 #'
 #' @export
-season_weights <- function(time, groups = seasonally(time)) {
+season_weights <- function(time, groups = metR::seasonally(time)) {
   w <- as.numeric(on_unique(time, lubridate::days_in_month))
 
-  setDT(list(w = w, groups = groups))[, w := w/mean(w), by = groups]$w
+  data.table::setDT(list(w = w, groups = groups))[, w := w/mean(w), by = groups]$w
 }
 
 

### 2022-04-20 R/solve-poisson.R

@@ -1,3 +1,8 @@
+#' Solve the poisson equation
+#'
+#' @param forcing numeric vector with the forcing (the vorticity)
+#' @param lon,lat numeric vectors with the locations of the forcing.
+#'
 #' @useDynLib shceof hwsssp_
 #' @export
 solve_poisson <- function(forcing, lon, lat) {

### 2022-04-20 R/spectrum.R

@@ -1,57 +1,0 @@
-#' FFT spectrum
-#'
-#' A light wrapper around [stats::spec.pgram()] and [stats::ar()]. Returns the spectrum of
-#' the signal and the spectrum of the fitted autoregressiv model.
-#'
-#' @param x numeric vector
-#' @param B number of bootstrap samples.
-#' @param spans vector of odd integers giving the widths of modified Daniell
-#' smoothers to be used to smooth the periodogram
-#' @param ... other arguments passed to [stats::spec.pgram()]
-#'
-#' @export
-fftspectrum <- function(x, spans = NULL, B = 10000, ..., probs = 0.95) {
-  mtm <- spec.pgram(x, spans = spans, ..., plot = FALSE)
-
-  out <- as.data.table(mtm[c("freq", "spec")])
-
-  ar <- ar(ts(x))
-  # rho <- a$ar
-  # var <- a$var.pred
-  out[, ar_spectrum := arspectrum(mtm$freq, ar$ar, ar$var.pred)]
-  out[, c(scales::percent(probs)) := null_ar_spectrum(B = B, length(x), ar, spans = spans, ..., probs = probs)]
-
-  return(out[])
-}
-
-arspectrum <- function(freq, rho, var) {
-  k <- seq_along(rho)
-  e <- vapply(freq, function(f) sum(rho * exp(-2*1i*pi*f*k)), complex(1))
-  var / (Mod(1 - e))^2
-}
-
-
-
-null_ar_spectrum_ <- function(B = 100, n, ar, spans = NULL, ..., probs = 0.95) {
-  y <- as.vector(arima.sim(model = list(ar = ar$ar), n = n))*sqrt(ar$var.pred)
-  nfreq <- length(spec.pgram(y, spans = spans, ..., plot = FALSE)$spec)
-  boots <- vapply(seq_len(B), function(b) {
-    y <- as.vector(arima.sim(model = list(ar = ar$ar), n = n))*sqrt(ar$var.pred)
-    spec.pgram(y, spans = spans, plot = FALSE)$spec
-
-  }, numeric(nfreq))
-  data.table::as.data.table(t(apply(boots, 1, quantile, probs = probs)))
-}
-
-null_ar_spectrum <- memoise::memoise(null_ar_spectrum_)
-
-null_spec <- memoise::memoise(function(x, spans, B = 1000, ..., probs = 0.95) {
-
-  b <- boot::boot(x, function(d, i) spec.pgram(d[i], spans = spans,
[…]

### 2022-04-20 analysis/paper/shceof.Rmd
Agrega abstract

@@
 title: "Revisiting the Austral Spring Extratropical Southern Hemisphere zonally asymmetric circulation using complex Empirical Orthogonal Functions"
-titlerunning:  Title
+titlerunning:  SH zonally asymmetric circulation
 authors: 
@@
 ```
-# Abstract
-
-asdfsdf
 
@@
 Sea Surface Temperature (SST) monthly fields are from Extended Reconstructed Sea Surface Temperature (ERSST) v5 [@huang2017] and precipitation monthly data from the CPC Merged Analysis of Precipitation [@cmap], with a 2\degree and 2.5\degree horizontal resolution respectively.
-The rainfall gridded dataset is based on information from different sources such as rain gauge observations, satellite inferred estimations and the NCEP-NCAR reanalysis, and it is available since 1979 to present.
+The rainfall gridded dataset is based on information from different sources such as rain gauge observations, satellite inferred estimations and the NCEP-NCAR reanalysis, and it is available since 1979 to the present.
 
@@
 
-The methodology proposed in this study allow a deeper understanding of  the zonally asymmetric springtime extratropical SH circulation such as a better description of PSA like variability using a unique complex index and the understanding of relationship between PSAs and ENSO or SAM variability.
+The methodology proposed in this study allows for a deeper understanding of the zonally asymmetric springtime extratropical SH circulation such as a better description of PSA like variability using a unique complex index and the understanding of relationship between PSAs and ENSO or SAM variability.
 Further work should extend this analysis to other seasons and further study the relationship between the cEOF2 and the SAM.
+
+# Abstract
+
+We analyse the joint variabilty of the zonally asymmetric springtime stratospheric and tropospheric circulation using Complex Empirical Orthogonal Functions (cEOF) to characterise planetary waves of varying amplitude and phase.
+The leading cEOF represents variabilty of a zonal wave 1 in the stratosphere and a monopole in the troposphere that correlates slightly with the Symmetric Southern Annular Mode (S-SAM).
+The second cEOF (cEOF2) is an alternative representation of the Pacific-South American modes. 
+One phase of this cEOF is also extremely correlated with the Asymmetric SAM (A-SAM) in the troposphere. 
+Springs with an active ENSO tend to phase-lock the cEOF2, but have no consistent impact on its magnitude.
+Furthermore, we find indications that the location of Pacific Sea Surface Temperature anomalies affect the precise phase of the cEOF2.
+
+----
 

### 2022-04-20 analysis/paper/shceof.tex
Agrega abstract

@@
 
-    \titlerunning{Title}
+    \titlerunning{SH zonally asymmetric circulation}
 
@@
 \begin{abstract}
-asdfsdf
+We analyse the joint variabilty of the zonally asymmetric springtime stratospheric and tropospheric circulation using Complex Empirical Orthogonal Functions (cEOF) to characterise planetary waves of varying amplitude and phase.
+The leading cEOF represents variabilty of a zonal wave 1 in the stratosphere and a monopole in the troposphere that correlates slightly with the Symmetric Southern Annular Mode (S-SAM).
+The second cEOF (cEOF2) is an alternative representation of the Pacific-South American modes.
+One phase of this cEOF is also extremely correlated with the Asymmetric SAM (A-SAM) in the troposphere.
+Springs with an active ENSO tend to phase-lock the cEOF2, but have no consistent impact on its magnitude.
+Furthermore, we find indications that the location of Pacific Sea Surface Temperature anomalies affect the precise phase of the cEOF2.
 \\
@@
 Xie and Arkin 1997), with a 2\degree and 2.5\degree horizontal resolution respectively.
-The rainfall gridded dataset is based on information from different sources such as rain gauge observations, satellite inferred estimations and the NCEP-NCAR reanalysis, and it is available since 1979 to present.
+The rainfall gridded dataset is based on information from different sources such as rain gauge observations, satellite inferred estimations and the NCEP-NCAR reanalysis, and it is available since 1979 to the present.
 
@@
 
-The methodology proposed in this study allow a deeper understanding of the zonally asymmetric springtime extratropical SH circulation such as a better description of PSA like variability using a unique complex index and the understanding of relationship between PSAs and ENSO or SAM variability.
+The methodology proposed in this study allows for a deeper understanding of the zonally asymmetric springtime extratropical SH circulation such as a better description of PSA like variability using a unique complex index and the understanding of relationship between PSAs and ENSO or SAM variability.
 Further work should extend this analysis to other seasons and further study the relationship between the cEOF2 and the SAM.

### 2022-04-20 analysis/paper/shceof.Rmd
Mueve el abstract al principio
(por el word output)
Render PDF

@@
 output: 
+  # bookdown::word_document2: default    
   bookdown::pdf_book: 
@@
 ```
+
+# Abstract
+
+We analyse the joint variabilty of the zonally asymmetric springtime stratospheric and tropospheric circulation using Complex Empirical Orthogonal Functions (cEOF) to characterise planetary waves of varying amplitude and phase.
+The leading cEOF represents variabilty of a zonal wave 1 in the stratosphere and a monopole in the troposphere that correlates slightly with the Symmetric Southern Annular Mode (S-SAM).
+The second cEOF (cEOF2) is an alternative representation of the Pacific-South American modes. 
+One phase of this cEOF is also extremely correlated with the Asymmetric SAM (A-SAM) in the troposphere. 
+Springs with an active ENSO tend to phase-lock the cEOF2, but have no consistent impact on its magnitude.
+Furthermore, we find indications that the location of Pacific Sea Surface Temperature anomalies affect the precise phase of the cEOF2.
 
@@
 
-# Abstract
-
-We analyse the joint variabilty of the zonally asymmetric springtime stratospheric and tropospheric circulation using Complex Empirical Orthogonal Functions (cEOF) to characterise planetary waves of varying amplitude and phase.
-The leading cEOF represents variabilty of a zonal wave 1 in the stratosphere and a monopole in the troposphere that correlates slightly with the Symmetric Southern Annular Mode (S-SAM).
-The second cEOF (cEOF2) is an alternative representation of the Pacific-South American modes. 
-One phase of this cEOF is also extremely correlated with the Asymmetric SAM (A-SAM) in the troposphere. 
-Springs with an active ENSO tend to phase-lock the cEOF2, but have no consistent impact on its magnitude.
-Furthermore, we find indications that the location of Pacific Sea Surface Temperature anomalies affect the precise phase of the cEOF2.
-
-----
-
 ## Code availability {.unnumbered}

### 2022-05-27 analysis/paper/shceof.Rmd
Cambia a british 🧐

@@
 
-Another extensively used approach to characterize the SH tropospheric circulation anomalies, is by computing Empirical Orthogonal Functions (EOF, also known as Principal Component Analysis).
+Another extensively used approach to characterise the SH tropospheric circulation anomalies, is by computing Empirical Orthogonal Functions (EOF, also known as Principal Component Analysis).
 Within the EOF framework, the Southern Annular Mode (SAM) appears as the leading mode of variability of the SH circulation [@fogt2020].
@@
 Therefore, to aid in the interpretation we chose the phase of cEOF1 so that the time series corresponding to the real part has the maximum correlation with the zonal wave 1 of Total Ozone Column between `r knitr::combine_words(LatLabel(o3wave_lats))`.
-Similarly, we chose the phase of cEOF2 so that the coefficient of determination between the Oceanic Niño Index [@bamston1997] and the real part is minimized, which also nearly maximizes the correlation with the imaginary part.
+Similarly, we chose the phase of cEOF2 so that the coefficient of determination between the Oceanic Niño Index [@bamston1997] and the real part is minimised, which also nearly maximises the correlation with the imaginary part.
 
@@
 This correlation field  matches the springtime precipitation signature of ENSO [e.g.
-@cai2020a] and it is also similar to the precipitation anomalies associated with the A-SAM [@@campitelli2022].
+@cai2020a] and it is also similar to the precipitation anomalies associated with the A-SAM [@campitelli2022].
 

### 2022-05-27 analysis/paper/shceof.Rmd
Pasa cambios rápidos de Carolina

@@
 Springs with an active ENSO tend to phase-lock the cEOF2, but have no consistent impact on its magnitude.
-Furthermore, we find indications that the location of Pacific Sea Surface Temperature anomalies affect the precise phase of the cEOF2.
+Furthermore, we find indications that the location of Pacific Sea Surface Temperature anomalies affect the phase of the cEOF2.
 
@@
 
+Some of the zonally asymmetric patterns of the SH circulation variability described previously, appear to have experienced secular changes.
+For instance, @raphael2003 found that the amplitude of the zonal wave 1 experienced a large increase and that the zonal wave 3 experienced changes in its annual cycle between 1958 and 1996.
+However, little is known yet about variability and trends of these patterns.
+
 Patterns resulting from EOF analyses are more flexible than Fourier decomposition derived modes in the sense that they can capture oscillation patterns that cannot be characterised by purely sinusoidal waves with constant amplitude.
-Nonetheless, they are restricted to standing oscillation modes and could not represent properly propagating or phase-varying features such as zonal waves.
+Nonetheless, they are restricted to standing oscillation modes and could not properly represent propagating or phase-varying features such as zonal waves.
 A single EOF can also represent a mixture of two or more physical modes.
@@
 
-An alternative methodology that has been proposed to study travelling and standing waves is complex Empirical Orthogonal Functions [@horel1984].
+An alternative methodology that has been proposed to study travelling and standing waves is complex Empirical Orthogonal Functions [cEOF; @horel1984].
 This method extends EOF analysis to capture oscillations with varying amplitude and phase and has been applied to the time domain.
@@
 Similarly, @gelbrecht2018 used cEOF applied to daily precipitation from reanalysis to study the propagating characteristics of the South American Monsoon.
-To our knowledge, complex EOF has not been applied in the spatial domain to capture the phase-varying nature of planetary waves.
-
-Some of the zonally asymmetric patterns of the SH circulation variability described previously, appear to have experienced secular changes.
-For instance, @raphael2003 found that the amplitude of the zonal wave 1 experienced a large increase and that the zonal wave 3 experienced changes in its annual cycle between 1958 and 1996.
-However, little is known yet about variability and trends of these patterns.
+To our knowledge, cEOF analysis has not been applied in the spatial domain to capture the phase-varying nature of planetary waves in the atmosphere.
 
@@
 
-We restrict this paper to the spring September-October-November (SON) trimester because this season experiences a maximum in tropical teleconnections over South America [@cazes-boezio2003] and Southern Hemisphere stratosphere-troposphere interactions [@lim2018].
+We restrict this paper to the spring September-October-November (SON) trimester because this season experiences a maximum in tropical teleconnections over South America [@cazes-boezio2003] and SH stratosphere-troposphere interactions [@lim2018].
 
@@
 
-A significant upward trend in the real component of cEOF1 is evident (Fig. \@ref(fig:extended-series)a.1, p-value `r trends[part == "Real" & cEOF == "cEOF1"]$p_report`) while there is no significant trend in any of the complex components of cEOF2.
+A significant positive trend in the real component of cEOF1 is evident (Fig. \@ref(fig:extended-series)a.1, p-value `r trends[part == "Real" & cEOF == "cEOF1"]$p_report`) while there is no significant trend in any of the complex components of cEOF2.
 The positive trend in the Real cEOF1 translates into a positive trend in cEOF1 magnitude, but not systematic change in phase (not shown).
@@
 
-In the previous section, cEOFs were applied to zonal anomalies, that is anomalies derived by removing the zonally mean values in order to isolate the main characteristics of the main zonal waves characterizing the circulation in the SH.
-In this section, regression fields were computed using the full fields of the variables in order to describe the influence of the cEOFs on the temporal anomalies.
+In the previous section, cEOFs analysis was applied to zonal anomalies, that is anomalies derived by removing the zonally mean values in order to isolate the main characteristics of the main zonal waves characterizing the circulation in the SH.
+In addition, regression fields were computed using the full fields of the variables in order to describe the influence of the cEOFs on the temporal anomalies.
 
@@
 Figure \@ref(fig:eof1-regr-gh) shows regression maps of SON geopotential height anomalies upon cEOF1.
-At 50 hPa (Figure \@ref(fig:eof1-regr-gh) row a), the Real cEOF1 is associated with a monopole centred over the Ross Sea.
-The Imaginary cEOF1 is associated with a clear wave 1 pattern with maximum over the coast of East Antarctica.
-
-At 200 hPa (Figure \@ref(fig:eof1-regr-gh) row b) the Real cEOF1 shows a single area of positive anomalies spanning West Antarctica surrounded by opposite anomalies in lower latitudes with is centre shifted slightly eastward compared with the upper level anomalies.
+At 50 hPa (Figure \@ref(fig:eof1-regr-gh) row a), the Real cEOF1 is associated with a centre located over the Ross Sea.
+The Imaginary cEOF1 is associated with a distinctive wave 1 pattern with maximum over the coast of East Antarctica.
+
+At 200 hPa (Figure \@ref(fig:eof1-regr-gh) row b) the Real cEOF1 shows a single centre of positive anomalies spanning West Antarctica surrounded by opposite anomalies in lower latitudes, with is centre shifted slightly eastward compared with the upper-level anomalies.
 The Imaginary cEOF1 shows a much more zonally symmetrical pattern resembling the negative SAM phase [e.g. @fogt2020].
-Both components of cEOF1 could be associated to zonal anomalies in the upper troposphere jet stream intensity.
-
-Therefore, the magnitude and phase of the cEOF1 are associated with the magnitude and phase of a zonal wave only in the stratosphere.
-While in the troposphere, they are associated with slightly off-centre monopoles.
+
+Therefore, the magnitude and phase of the cEOF1 are associated with the magnitude and phase of a zonal wave mainly in the stratosphere.
 
@@
 
-The influence of cEOFs variability on air temperature was also evaluated.
+The signature of cEOFs variability on air temperature was also evaluated.
 Figure\ \@ref(fig:eof1-regr-t) shows regression maps of air temperature anomalies at 50 hPa and 200 hPa upon cEOF1. 
-In both levels, the Real cEOF1 is associated with a positive monopole over the South Pole with its centre moved slightly towards 150ºE (Fig. \@ref(fig:eof1-regr-t) column 1).
+The distribution of temperature regression coefficients at 50 hPa and at 200 hPa mirror the geopotential height regression maps at 50 hPa (Fig. \@ref(fig:eof1-regr-gh)).
+In both levels, the Real cEOF1 is associated with a positive centre over the South Pole with its centre moved slightly towards 150ºE (Fig. \@ref(fig:eof1-regr-t) column 1).
 On the other hand, the regression maps on the Imaginary cEOF1 show a more clear wave 1 pattern with its maximum around 60ºE.
-The distribution of temperature regression coefficients at 50 hPa and at 200 hPa mirror the geopotential height regression maps at 50 hPa (Fig. \@ref(fig:eof1-regr-gh)).
 
@@
 
-Given the similarity between the cEOF2 regression maps (Fig. \@ref(fig:eof2-regr-gh)) and documented PSA patterns we study the relationship between these modes.
+Given the similarity between the cEOF2 related anomaly structures (Fig. \@ref(fig:eof2-regr-gh)) and documented PSA patterns we study the relationship between them.
 Table \@ref(tab:psa-eof2) shows the correlations between the two PSA indices and the time series for Real and Imaginary phases of cEOF2.
-As anticipated by Figure \@ref(fig:eof2-regr-gh), there is a strong positive correlation between PSA1 and Imaginary cEOF2, and also between PSA2 and Real cEOF2.
+As visually anticipated by Figure \@ref(fig:eof2-regr-gh), there is a large positive correlation between PSA1 and Imaginary cEOF2, and between PSA2 and Real cEOF2.
 On the other hand, there is no significant relationship between PSA1 and Real cEOF2, and between PSA2 and Imaginary cEOF2.
 As a result, cEOF2 represents well both the spatial structure and temporal evolution of the PSA modes, so it is possible to make an association between its two phases and the two PSA modes.
-The same phase election for cEOF2 that maximises the relationship between ENSO and Imaginary cEOF2, also maximises the association between cEOF2 components and PSA modes (not shown).
+That is, the phase election for cEOF2 that maximises the relationship between ENSO and Imaginary cEOF2, also maximises the association between cEOF2 components and PSA modes (not shown).
 
@@
 We now explore the relationship between SAM and the cEOFs motivated by the resemblance between cEOFs regression maps and SAM patterns shown in Section\ \@ref(regressions).
-To that end, we computed the coefficient of determination between the cEOFs time series and the three SAM indices (SAM, A-SAM and S-SAM) defined by @campitelli2022 at each vertical level (Fig.\ \@ref(fig:sam-eof-vertical)).
+We then computed the coefficient of determination between the cEOFs time series and the three SAM indices (SAM, A-SAM and S-SAM) defined by @campitelli2022 at each vertical level (Fig.\ \@ref(fig:sam-eof-vertical)).
 The SAM index is statistically significant correlated with the Real cEOF1 in all levels, and with the Imaginary cEOF1 and Imaginary cEOF2 in the troposphere.
@@
 
-The connections between cEOFs and Tropical sources are also assessed, to explore plausible forcings for the described cEOFs patterns.
+The connections between cEOFs and Tropical sources of variability were also assessed.
 Figure \@ref(fig:sst-psi-2) shows the regression maps of Sea Surface Temperatures (SST) and streamfunction anomalies at 200 hPa respectively upon standardised cEOF2.
@@
 
-Furthermore, the Imaginary cEOF2 (second row) is associated with strong positive SST anomalies on the Central Pacific and negative anomalies over an area across northern Australia,  New Zealand the South Pacific Convergence Zone (SPCZ) (Figure \@ref(fig:sst-psi-2).b1).
+The Imaginary cEOF2 (second row) is associated with strong positive SST anomalies on the Central Pacific and negative anomalies over an area across northern Australia,  New Zealand the South Pacific Convergence Zone (SPCZ) (Figure \@ref(fig:sst-psi-2).b1).
 The regression field of SST anomalies bears a strong resemblance with canonically positive ENSO [@bamston1997].
@@
 The Imaginary cEOF2 is associated with strong wave-like streamfunction anomalies emanating from the tropics (Figure\ \@ref(fig:sst-psi-2).b2), both from the Central Pacific sector and the Indian Ocean.
-These atmospheric response is consistent with the effect of ENSO and the IOD on the extratropics: the SST anomalies initiate anomalous tropical convection that excites Rossby waves that propagate meridionally towards higher latitudes  [@mo2000; @cai2011; @nuncio2015].
-
-The cEOF2 is not associated with the same tropical SSTs for all their phases. 
+The atmospheric response associated with Imaginary cEOF2 is then consistent with the combined effect of ENSO and the IOD on the extratropics: with SST anomalies inducing anomalous tropical convection that in turn excite Rossby waves propagating meridionally towards higher latitudes [@mo2000; @cai2011; @nuncio2015].
+
+However, the cEOF2 is not associated with the same tropical SST anomaly patterns at all their phases
 Figure\ \@ref(fig:sst-psi-2).d1 and d2 show that the Real cEOF2 is not associated either with any significant SST nor streamfunction anomalies in the tropics. 
 As a result, the correlation between the Real cEOF2 and ENSO is also not significant  (`r enso_cor[cEOF == "cEOF2" & part == "Real"]$text`).
-Meanwhile, Rows a and c in Fig.\ \@ref(fig:sst-psi-2) show that the intermediate phases are still associated with significant SST regressed anomalies over the Pacific Ocean, but with with slightly different location.
-The 135º phase is associated with SST in the central Pacific (Fig.\ \@ref(fig:sst-psi-2)a.1), while the 45º phase is associated with SST in the eastern Pacific (Fig.\ \@ref(fig:sst-psi-2)c.1), which correspond roughly to the Central Pacific and Eastern Pacific "flavours" of ENSO, respectively [@kao2009].
-Both of these phases could be also associated to wave trains generated in the region surrounding Australia and propagates toward the extra-tropics.
+Meanwhile, Rows a and c in Fig.\ \@ref(fig:sst-psi-2) show that the intermediate phases are still associated with significant SST regressed anomalies over the Pacific Ocean, but at slightly different locations.
+The 135º phase is associated with SST anomalies in the central Pacific (Fig.\ \@ref(fig:sst-psi-2)a.1), while the 45º phase is associated with SST anomalies (Fig.\ \@ref(fig:sst-psi-2)c.1) that correspond roughly to the Central Pacific and Eastern Pacific "flavours" of ENSO, respectively [@kao2009].
+Both phases could be also associated to wave trains generated in the region surrounding Australia and propagates toward the extra-tropics.
 
@@
 
-To better explore the relationship between tropical forcing and phases of the cEOF2, Figure\ \@ref(fig:enso-phase) shows the ONI index plotted against the cEOF2 phase for each SON trimester between 1979 and 2019, highlighting years in which the magnitude of cEOF2 is above the median.
+To further explore the relationship between tropical forcing and phases of the cEOF2, Figure\ \@ref(fig:enso-phase) shows the ONI index plotted against the cEOF2 phase for each SON trimester between 1979 and 2019, highlighting years in which the magnitude of cEOF2 is above the median.
 In years with positive (negative) ONI, the cEOF2 phase is mostly around 90º (-90°),  corresponding with positive (negative) Imaginary component.
@@
 Furthermore, even when using all years, the Spearman correlation --which is robust to outliers-- is also non-significant (`r signif(cor_enso_spearman$estimate, 2)`, p-value`r report_p(cor_enso_spearman$p.value)`).
-Therefore, although the location of tropical SST seem to have an effect in defining the phase of the cEOF2, the relationship between the magnitude of cEOF2 and ONI remains uncertain.
+Therefore, although the location of tropical SST anomalies seem to have an effect in defining the phase of the cEOF2, the relationship between the magnitude of cEOF2 and ONI remains uncertain.
 
@@
 
-The influence of cEOFs variability in and 2-metre temperature and temperature in the Southern Hemisphere is also explored.
+The influence of cEOFs variability in the anomalies of both 2-metre air temperature and precipitation in the Southern Hemisphere is also explored.
 Figure\ \@ref(fig:pp-t2m-r2) shows the 2-meter temperature or precipitation anomalies explained variance by the multiple linear model of both Real and Imaginary cEOF1 components (column 1), and both Real  and Imaginary cEOF2 components (column 2).
@@
 In previous sections, it has been shown that variability patterns that arise from cEOF methodology allow to describe the zonally asymmetric springtime extratropical SH circulation, reproducing previous features such as the variability related to PSAs or A-SAM.
-However, it could be questioned what is the advantage of using the methodology proposed here.
+However the advantage of using the methodology proposed here could be questioned.
 Firstly, the spatial fields obtained from both components of cEOF2, that resembles PSA patterns, are in quadrature by construction.
-Thus, it is possible to derive a joint PSA index from the amplitude and phase obtained when applying cEOF methodology.
-Secondly, because of how these components are constructed, they are not forced to be orthogonal to the SAM as in the standard EOF methodology.
-This allows us to show that Imaginary cEOF2, that resembles PSA1 variability, is closely related to the SAM in the troposphere.
-Moreover, the high correlation of this component to the the zonally asymmetric portion of the SAM allow to speculate that the A-SAM might actually be a statistical contamination of the PSA mode.
+Thus, the cEOF methodology allows to derive a joint PSA index from the resulted amplitude and phase.
+Secondly, because of how these components are constructed, they are not forced to be orthogonal to each other as in the standard EOF methodology. 
+This allows us to show for example, that Imaginary cEOF2, that resembles PSA1 variability, is closely related to the SAM in the troposphere. 
+Moreover, the high correlation of this component to the zonally asymmetric portion of the SAM allow to speculate that the A-SAM might actually be a statistical contamination of the PSA mode.
 
 @irving2016 argued that there is some disagreement in the literature of whether the phase of the PSA pattern is affected by the location of tropical SST anomalies.
-With the methodology used in this study, we were able to show not only that the cEOF2 tends to be in the positive or negative imaginary phase (\~PSA1) when the ENSO region is warm or cold, respectively, but that central Pacific SST anomalies tend to align the cEOF towards the negative real phase and eastern Pacific SST anomalies tend to align it towards positive real phase.
-When ENSO is neutral, the cEOF2 is still as active, but with no preferred phase.
-This mirrors the results of @cai2002, who showed that the CSIRO Model can create PSA-like variability even in the absence of ENSO forcing (i.e.
+With the methodology used in this study, we were able to show not only that the cEOF2 tends to be in the positive or negative imaginary phase (\~PSA1) when the ENSO region is warm or cold, respectively, but also that central Pacific SST anomalies tend to align the cEOF towards the negative real phase and eastern Pacific SST anomalies tend to align it towards positive real phase.
+When ENSO phase is neutral, the cEOF2 is still as active, but with no preferred phase.
+The latter agrees with the results of @cai2002, who showed that the CSIRO Model can develop PSA-like variability even in the absence of ENSO forcing (i.e.
 with a climatological run), but that the variability of one of the PSA modes was enhanced when adding the ENSO signal.
@@
 
-In this study we assessed the Austral Spring Extratropical Southern Hemisphere zonally asymmetric circulation.
-For this purpose, two complex indices were derived using Complex Empirical Orthogonal Functions.
-This allowed us to characterise both the amplitude and phase of planetary waves that are not perfectly organised as sinusoidal waves with constant amplitude at each latitude circle.
+In this study we assessed extratropical Southern Hemisphere zonally asymmetric circulation in austral spring.
+For this purpose, two complex indices were derived using Complex Empirical Orthogonal Functions and used to characterise both  amplitude and phase of planetary waves.
 
@@
 
-The phase of the cEOF2 is related to the the variability of ENSO, but its magnitude appears not to be, with the exception of the three strongest ENSO events in the period which coincide with the three strongest cEOF2 years.
+The phase of the cEOF2 is related to the variability of ENSO, but its magnitude appears not to be, except for the three strongest ENSO events in the period which coincide with the three strongest cEOF2 years.
 

### 2022-05-27 analysis/paper/shceof.Rmd
Agrega descripcíon del monopolo en la estrasótfera del cEOF2

@@
 Figure \@ref(fig:eof2-regr-gh) shows the regression maps of geopotential height anomalies upon the cEOF2.
-Wave trains similar to those identified for cEOF2 patterns (Fig \@ref(fig:ceofs-1)) can be distinguished.
+In the troposphere (Fig. \@ref(fig:eof2-regr-gh) column b), wave trains similar to those identified for cEOF2 patterns (Fig \@ref(fig:ceofs-1)) can be distinguished.
 Regressed anomalies associated with the Real cEOF2 are 90º out of phase with those associated with the Imaginary cEOF2.
@@
 Comparing the location of the positive anomaly near 90ºW in column b of Figure \@ref(fig:eof2-regr-gh) with Figures 1.a and b from @mo2001, the Real cEOF2 regression map could be identified with PSA2, while the Imaginary cEOF2 resembles PSA1.
+
+These wave patterns are also present in the stratosphere (Fig. \@ref(fig:eof2-regr-gh) column a) supporting their  equivalent barotropic nature. 
+But also present is a monopole over the pole with negative sign associated with the Real cEOF2 and positive sign associated with the Imaginary cEOF2. 
+This monopole might indicate strengthening of the polar vortex associated with positive values of the Real cEOF2 and weakening associated with positive Imaginary cEOF2.
 

### 2022-05-27 analysis/paper/shceof.Rmd
Agrega aclaración de eventos ENSO

@@
 Furthermore, even when using all years, the Spearman correlation --which is robust to outliers-- is also non-significant (`r signif(cor_enso_spearman$estimate, 2)`, p-value`r report_p(cor_enso_spearman$p.value)`).
-Therefore, although the location of tropical SST anomalies seem to have an effect in defining the phase of the cEOF2, the relationship between the magnitude of cEOF2 and ONI remains uncertain.
+Therefore, although the location of tropical SST anomalies seem to have an effect in defining the phase of the cEOF2, the relationship between the magnitude of cEOF2 and ONI remains uncertain might be evident only in particularly intense ENSO events, for which there are few data.
 

### 2022-05-27 analysis/paper/shceof.Rmd
Reescribe la influencia de cEOF1 en antaŕtida

@@
 Figure\ \@ref(fig:pp-t2m-r2) shows the 2-meter temperature or precipitation anomalies explained variance by the multiple linear model of both Real and Imaginary cEOF1 components (column 1), and both Real  and Imaginary cEOF2 components (column 2).
-The variance explained by cEOF1 for precipitation anomalies is extremely low and also for most of the temperature anomalies, except for the eastern western of Antarctica (Fig.\  \@ref(fig:pp-t2m-r2)a.1).
+The variance explained by cEOF1 for precipitation anomalies is extremely low and also for most of the temperature anomalies, except for the northern tip of the Antarctic Peninsula, northern Weddell Sea and the Ross Sea coast (Fig.\  \@ref(fig:pp-t2m-r2)a.1).
 

### 2022-05-27 analysis/paper/shceof.Rmd
Borra comparación con Simmonds (2016)

@@
 
-Some results of our work contrast with those obtained by @irving2016 when analysing the PSA.
-They did not find neither a strong association between PSA-like variability and ENSO, nor a relationship between this variability and Australian precipitation.
-Part of the disagreement could be due to the fact that we focus our analysis only on SON, which is the season when the teleconnection patterns are strongest, instead of the whole year.
-Moreover, @irving2016 use of ERA5 forectasted precipitation instead of the more observationally constrained CMAP dataset might be another source of disagreement.
-Additionally, the discrepancy could be related to the fact that @irving2016 show a weak relationship between ENSO and their PSA pattern, which explains partly the relationship between Australian rainfall and PSA-like variability.
-
 # Conclusion

### 2022-05-31 analysis/paper/shceof.Rmd
Borra mención de monopolo
Aclara frase del PSA

@@
 This is also, by construction, the phase that is most correlated with ENSO.
-Therefore, the Imaginary cEOF2 explains more variance than the Real cEOF2 and conventional EOF analysis will tend to separate the two.
+
+Therefore, by virtue of being the most common phase, the Imaginary cEOF2 explains more variance than the Real cEOF2.
+Conventional EOF analysis will therefore tend to separate them relatively cleanly, with the EOF representing the Imaginary cEOF2 always leading the one representing the Real cEOF2.
 
@@
 
-The cEOF1 represents the variability of the zonal wave 1 in the stratosphere, but represents a more zonally symmetric monopole in the troposphere.
+The cEOF1 represents the variability of the zonal wave 1 in the stratosphere.
 There is a statistically positive trend in the magnitude of this cEOF, which is consistent with previous studies that showed secular changes in springtime wave-1-like patterns [e.g. @raphael2003].

### 2022-05-31 analysis/paper/shceof.Rmd
Agrega cita a goyal y unifica dicusión y conclusion

@@
 Although these patterns are usually derived by applying EOF to temporal anomalies, @raphael2003 also applied EOF methods specifically to zonal anomalies.
-More recently, @irving2016 proposed a novel methodology for objectively identifying the PSA pattern using Fourier decomposition.
+@irving2016 proposed a novel methodology for objectively identifying the PSA pattern using Fourier decomposition. 
+More recently @goyal2022 created an index of amplitude and phase of zonal wave 3-like variability by combining the two leading EOFs of meridional wind anomalies.
 
@@
 
-# Discussion
-
-In previous sections, it has been shown that variability patterns that arise from cEOF methodology allow to describe the zonally asymmetric springtime extratropical SH circulation, reproducing previous features such as the variability related to PSAs or A-SAM.
+# Discussion and conclusions
+
+In previous sections, it has been shown that variability patterns that arise from cEOF methodology describe the zonally asymmetric springtime extratropical SH circulation, reproducing previous features such as the variability related to PSAs or A-SAM.
+
 However the advantage of using the methodology proposed here could be questioned.
@@
 Thus, the cEOF methodology allows to derive a joint PSA index from the resulted amplitude and phase.
-Secondly, because of how these components are constructed, they are not forced to be orthogonal to each other as in the standard EOF methodology. 
-This allows us to show for example, that Imaginary cEOF2, that resembles PSA1 variability, is closely related to the SAM in the troposphere. 
-Moreover, the high correlation of this component to the zonally asymmetric portion of the SAM allow to speculate that the A-SAM might actually be a statistical contamination of the PSA mode.
+Secondly, because of how these components are constructed, they are not forced to be orthogonal to other modes of circulation, as in the standard EOF methodology. 
+This allows us to show for example, that the Imaginary cEOF2, that resembles PSA1 variability, is closely associated to the SAM in the troposphere. 
+Moreover, the high correlation of this component to the zonally asymmetric portion of the SAM allow us to speculate that the A-SAM might actually be a statistical contamination of the PSA mode on the symmetric SAM.
 
@@
 with a climatological run), but that the variability of one of the PSA modes was enhanced when adding the ENSO signal.
-
 The sensitivity of the phase of the PSAs to the location of the tropical SST anomalies was also seen by @ciasto2015, who detected similar Rossby wave patterns associated with central Pacific and eastern Pacific SST anomalies but with a change in phase.
 
-# Conclusion
-
-In this study we assessed extratropical Southern Hemisphere zonally asymmetric circulation in austral spring.
-For this purpose, two complex indices were derived using Complex Empirical Orthogonal Functions and used to characterise both amplitude and phase of planetary waves.
-
-The cEOF1 represents the variability of the zonal wave 1 in the stratosphere.
-There is a statistically positive trend in the magnitude of this cEOF, which is consistent with previous studies that showed secular changes in springtime wave-1-like patterns [e.g. @raphael2003].
-This mode is closely related to stratospheric variability such as anomalies in Total Ozone Column.
-Otherwise, this complex EOF is not related with SST variability and continental precipitation in the Southern Hemisphere.
-
-The cEOF2 represents a wave-3 pattern with maximum magnitude in the Pacific sector, that is an alternative representation of the PSA1 and PSA2 patterns [@mo2001].
-The Imaginary cEOF2 can be  identified with the PSA1 and the Real cEOF2 with the PSA2.
-
-The phase of the cEOF2 is related to the variability of ENSO, but its magnitude appears not to be, except for the three strongest ENSO events in the period which coincide with the three strongest cEOF2 years.
-
-The cEOF2 variability could be related to surface impacts, while the cEOF1 influence is almost negligible. 
-For instance, precipitation anomalies in South America associated with the Imaginary cEOF2 show a clear ENSO-like impact, with positive anomalies in South-Eastern South America, negative anomalies in Southern Brazil and positive anomalies in central Chile for positive Imaginary cEOF2 phase.
-Precipitation anomalies associated with the Real cEOF2 are low and not statistically significant in this region, showing that the Imaginary phase is optimally aligned with the direction of maximum precipitation impacts there.
-
-On the other hand, over Australia, both the Real and Imaginary phase are associated with significant precipitation anomalies, and we further show that the direction of maximum impact is not aligned with our chosen rotation of cEOF2.
-An analysis of the precipitation impacts on this regions could benefit for a different chosen rotation of the cEOF.
+Our method is similar to @goyal2022 as they also derive an index with amplitude and phase of wave-like activity in the SH with very similar patterns to our cEOF2. 
+Both indices are likely very similar (a detailed comparison is out of scope for this paper) but we believe that cEOF have the advantage of constructing the indices based on patterns that are exactly in quadrature by construction.
 

### 2022-05-31 analysis/paper/shceof.tex
REnderiza

@@
 Springs with an active ENSO tend to phase-lock the cEOF2, but have no consistent impact on its magnitude.
-Furthermore, we find indications that the location of Pacific Sea Surface Temperature anomalies affect the precise phase of the cEOF2.
+Furthermore, we find indications that the location of Pacific Sea Surface Temperature anomalies affect the phase of the cEOF2.
 \\
@@
 
-Another extensively used approach to characterize the SH tropospheric circulation anomalies, is by computing Empirical Orthogonal Functions (EOF, also known as Principal Component Analysis).
+Another extensively used approach to characterise the SH tropospheric circulation anomalies, is by computing Empirical Orthogonal Functions (EOF, also known as Principal Component Analysis).
 Within the EOF framework, the Southern Annular Mode (SAM) appears as the leading mode of variability of the SH circulation (Fogt and Marshall 2020).
@@
 Although these patterns are usually derived by applying EOF to temporal anomalies, M. Raphael (2003) also applied EOF methods specifically to zonal anomalies.
-More recently, Irving and Simmonds (2016) proposed a novel methodology for objectively identifying the PSA pattern using Fourier decomposition.
+Irving and Simmonds (2016) proposed a novel methodology for objectively identifying the PSA pattern using Fourier decomposition.
+More recently Goyal et al. (2022) created an index of amplitude and phase of zonal wave 3-like variability by combining the two leading EOFs of meridional wind anomalies.
+
+Some of the zonally asymmetric patterns of the SH circulation variability described previously, appear to have experienced secular changes.
+For instance, M. Raphael (2003) found that the amplitude of the zonal wave 1 experienced a large increase and that the zonal wave 3 experienced changes in its annual cycle between 1958 and 1996.
+However, little is known yet about variability and trends of these patterns.
 
 Patterns resulting from EOF analyses are more flexible than Fourier decomposition derived modes in the sense that they can capture oscillation patterns that cannot be characterised by purely sinusoidal waves with constant amplitude.
-Nonetheless, they are restricted to standing oscillation modes and could not represent properly propagating or phase-varying features such as zonal waves.
+Nonetheless, they are restricted to standing oscillation modes and could not properly represent propagating or phase-varying features such as zonal waves.
 A single EOF can also represent a mixture of two or more physical modes.
@@
 
-An alternative methodology that has been proposed to study travelling and standing waves is complex Empirical Orthogonal Functions (Horel 1984).
+An alternative methodology that has been proposed to study travelling and standing waves is complex Empirical Orthogonal Functions {[}cEOF; Horel (1984){]}.
 This method extends EOF analysis to capture oscillations with varying amplitude and phase and has been applied to the time domain.
@@
 Similarly, Gelbrecht, Boers, and Jürgen Kurths (2018) used cEOF applied to daily precipitation from reanalysis to study the propagating characteristics of the South American Monsoon.
-To our knowledge, complex EOF has not been applied in the spatial domain to capture the phase-varying nature of planetary waves.
-
-Some of the zonally asymmetric patterns of the SH circulation variability described previously, appear to have experienced secular changes.
-For instance, M. Raphael (2003) found that the amplitude of the zonal wave 1 experienced a large increase and that the zonal wave 3 experienced changes in its annual cycle between 1958 and 1996.
-However, little is known yet about variability and trends of these patterns.
+To our knowledge, cEOF analysis has not been applied in the spatial domain to capture the phase-varying nature of planetary waves in the atmosphere.
 
@@
 
-We restrict this paper to the spring September-October-November (SON) trimester because this season experiences a maximum in tropical teleconnections over South America (Cazes-Boezio, Robertson, and Mechoso 2003) and Southern Hemisphere stratosphere-troposphere interactions (Lim, Hendon, and Thompson 2018).
+We restrict this paper to the spring September-October-November (SON) trimester because this season experiences a maximum in tropical teleconnections over South America (Cazes-Boezio, Robertson, and Mechoso 2003) and SH stratosphere-troposphere interactions (Lim, Hendon, and Thompson 2018).
 
@@
 In Section~\ref{discussion} we compare our results with previous studies and discuss the benefits of our methodology.
-In Section~\ref{conclusion} we summarise the conclusions.
 
@@
 Therefore, to aid in the interpretation we chose the phase of cEOF1 so that the time series corresponding to the real part has the maximum correlation with the zonal wave 1 of Total Ozone Column between 75°S and 45°S.
-Similarly, we chose the phase of cEOF2 so that the coefficient of determination between the Oceanic Niño Index (Bamston, Chelliah, and Goldenberg 1997) and the real part is minimized, which also nearly maximizes the correlation with the imaginary part.
+Similarly, we chose the phase of cEOF2 so that the coefficient of determination between the Oceanic Niño Index (Bamston, Chelliah, and Goldenberg 1997) and the real part is minimised, which also nearly maximises the correlation with the imaginary part.
 
@@
 
-A significant upward trend in the real component of cEOF1 is evident (Fig.~\ref{fig:extended-series}a.1, p-value~= 0.0037) while there is no significant trend in any of the complex components of cEOF2.
+A significant positive trend in the real component of cEOF1 is evident (Fig.~\ref{fig:extended-series}a.1, p-value~= 0.0037) while there is no significant trend in any of the complex components of cEOF2.
 The positive trend in the Real cEOF1 translates into a positive trend in cEOF1 magnitude, but not systematic change in phase (not shown).
@@
 Figure~\ref{fig:eof1-regr-gh} shows regression maps of SON geopotential height anomalies upon cEOF1.
-At 50 hPa (Figure~\ref{fig:eof1-regr-gh} row a), the Real cEOF1 is associated with a monopole centred over the Ross Sea.
-The Imaginary cEOF1 is associated with a clear wave 1 pattern with maximum over the coast of East Antarctica.
-
-At 200 hPa (Figure~\ref{fig:eof1-regr-gh} row b) the Real cEOF1 shows a single area of positive anomalies spanning West Antarctica surrounded by opposite anomalies in lower latitudes with is centre shifted slightly eastward compared with the upper level anomalies.
+At 50 hPa (Figure~\ref{fig:eof1-regr-gh} row a), the Real cEOF1 is associated with a centre located over the Ross Sea.
+The Imaginary cEOF1 is associated with a distinctive wave 1 pattern with maximum over the coast of East Antarctica.
+
+At 200 hPa (Figure~\ref{fig:eof1-regr-gh} row b) the Real cEOF1 shows a single centre of positive anomalies spanning West Antarctica surrounded by opposite anomalies in lower latitudes, with is centre shifted slightly eastward compared with the upper-level anomalies.
 The Imaginary cEOF1 shows a much more zonally symmetrical pattern resembling the negative SAM phase (e.g. Fogt and Marshall 2020).
-Both components of cEOF1 could be associated to zonal anomalies in the upper troposphere jet stream intensity.
-
-Therefore, the magnitude and phase of the cEOF1 are associated with the magnitude and phase of a zonal wave only in the stratosphere.
-While in the troposphere, they are associated with slightly off-centre monopoles.
+
+Therefore, the magnitude and phase of the cEOF1 are associated with the magnitude and phase of a zonal wave mainly in the stratosphere.
 
@@
 Figure~\ref{fig:eof2-regr-gh} shows the regression maps of geopotential height anomalies upon the cEOF2.
-Wave trains similar to those identified for cEOF2 patterns (Fig~\ref{fig:ceofs-1}) can be distinguished.
+In the troposphere (Fig.~\ref{fig:eof2-regr-gh} column b), wave trains similar to those identified for cEOF2 patterns (Fig~\ref{fig:ceofs-1}) can be distinguished.
 Regressed anomalies associated with the Real cEOF2 are 90º out of phase with those associated with the Imaginary cEOF2.
@@
 
+These wave patterns are also present in the stratosphere (Fig.~\ref{fig:eof2-regr-gh} column a) supporting their equivalent barotropic nature.
+But also present is a monopole over the pole with negative sign associated with the Real cEOF2 and positive sign associated with the Imaginary cEOF2.
+This monopole might indicate strengthening of the polar vortex associated with positive values of the Real cEOF2 and weakening associated with positive Imaginary cEOF2.
+
 \hypertarget{temp-ozone}{%
@@
 
-The influence of cEOFs variability on air temperature was also evaluated.
+The signature of cEOFs variability on air temperature was also evaluated.
 Figure~\ref{fig:eof1-regr-t} shows regression maps of air temperature anomalies at 50 hPa and 200 hPa upon cEOF1.
-In both levels, the Real cEOF1 is associated with a positive monopole over the South Pole with its centre moved slightly towards 150ºE (Fig.~\ref{fig:eof1-regr-t} column 1).
+The distribution of temperature regression coefficients at 50 hPa and at 200 hPa mirror the geopotential height regression maps at 50 hPa (Fig.~\ref{fig:eof1-regr-gh}).
+In both levels, the Real cEOF1 is associated with a positive centre over the South Pole with its centre moved slightly towards 150ºE (Fig.~\ref{fig:eof1-regr-t} column 1).
 On the other hand, the regression maps on the Imaginary cEOF1 show a more clear wave 1 pattern with its maximum around 60ºE.
-The distribution of temperature regression coefficients at 50 hPa and at 200 hPa mirror the geopotential height regression maps at 50 hPa (Fig.~\ref{fig:eof1-regr-gh}).
 
@@
 
-Given the similarity between the cEOF2 regression maps (Fig.~\ref{fig:eof2-regr-gh}) and documented PSA patterns we study the relationship between these modes.
+Given the similarity between the cEOF2 related anomaly structures (Fig.~\ref{fig:eof2-regr-gh}) and documented PSA patterns we study the relationship between them.
 Table~\ref{tab:psa-eof2} shows the correlations between the two PSA indices and the time series for Real and Imaginary phases of cEOF2.
-As anticipated by Figure~\ref{fig:eof2-regr-gh}, there is a strong positive correlation between PSA1 and Imaginary cEOF2, and also between PSA2 and Real cEOF2.
+As visually anticipated by Figure~\ref{fig:eof2-regr-gh}, there is a large positive correlation between PSA1 and Imaginary cEOF2, and between PSA2 and Real cEOF2.
 On the other hand, there is no significant relationship between PSA1 and Real cEOF2, and between PSA2 and Imaginary cEOF2.
 As a result, cEOF2 represents well both the spatial structure and temporal evolution of the PSA modes, so it is possible to make an association between its two phases and the two PSA modes.
-The same phase election for cEOF2 that maximises the relationship between ENSO and Imaginary cEOF2, also maximises the association between cEOF2 components and PSA modes (not shown).
+That is, the phase election for cEOF2 that maximises the relationship between ENSO and Imaginary cEOF2, also maximises the association between cEOF2 components and PSA modes (not shown).
 
@@
 We now explore the relationship between SAM and the cEOFs motivated by the resemblance between cEOFs regression maps and SAM patterns shown in Section~\ref{regressions}.
-To that end, we computed the coefficient of determination between the cEOFs time series and the three SAM indices (SAM, A-SAM and S-SAM) defined by Campitelli, Díaz, and Vera (2022) at each vertical level (Fig.~\ref{fig:sam-eof-vertical}).
+We then computed the coefficient of determination between the cEOFs time series and the three SAM indices (SAM, A-SAM and S-SAM) defined by Campitelli, Díaz, and Vera (2022) at each vertical level (Fig.~\ref{fig:sam-eof-vertical}).
 The SAM index is statistically significant correlated with the Real cEOF1 in all levels, and with the Imaginary cEOF1 and Imaginary cEOF2 in the troposphere.
@@
 
-The connections between cEOFs and Tropical sources are also assessed, to explore plausible forcings for the described cEOFs patterns.
+The connections between cEOFs and Tropical sources of variability were also assessed.
 Figure~\ref{fig:sst-psi-2} shows the regression maps of Sea Surface Temperatures (SST) and streamfunction anomalies at 200 hPa respectively upon standardised cEOF2.
@@
 
-Furthermore, the Imaginary cEOF2 (second row) is associated with strong positive SST anomalies on the Central Pacific and negative anomalies over an area across northern Australia, New Zealand the South Pacific Convergence Zone (SPCZ) (Figure~\ref{fig:sst-psi-2}.b1).
+The Imaginary cEOF2 (second row) is associated with strong positive SST anomalies on the Central Pacific and negative anomalies over an area across northern Australia, New Zealand the South Pacific Convergence Zone (SPCZ) (Figure~\ref{fig:sst-psi-2}.b1).
 The regression field of SST anomalies bears a strong resemblance with canonically positive ENSO (Bamston, Chelliah, and Goldenberg 1997).
@@
 The Imaginary cEOF2 is associated with strong wave-like streamfunction anomalies emanating from the tropics (Figure~\ref{fig:sst-psi-2}.b2), both from the Central Pacific sector and the Indian Ocean.
-These atmospheric response is consistent with the effect of ENSO and the IOD on the extratropics: the SST anomalies initiate anomalous tropical convection that excites Rossby waves that propagate meridionally towards higher latitudes (Mo 2000; Cai et al.
+The atmospheric response associated with Imaginary cEOF2 is then consistent with the combined effect of ENSO and the IOD on the extratropics: with SST anomalies inducing anomalous tropical convection that in turn excite Rossby waves propagating meridionally towards higher latitudes (Mo 2000; Cai et al.
 2011; Nuncio and Yuan 2015).
 
-The cEOF2 is not associated with the same tropical SSTs for all their phases.
+However, the cEOF2 is not associated with the same tropical SST anomaly patterns at all their phases
 Figure~\ref{fig:sst-psi-2}.d1 and d2 show that the Real cEOF2 is not associated either with any significant SST nor streamfunction anomalies in the tropics.
 As a result, the correlation between the Real cEOF2 and ENSO is also not significant (0 (CI: -0.3 -- 0.3)).
-Meanwhile, Rows a and c in Fig.~\ref{fig:sst-psi-2} show that the intermediate phases are still associated with significant SST regressed anomalies over the Pacific Ocean, but with with slightly different location.
-The 135º phase is associated with SST in the central Pacific (Fig.~\ref{fig:sst-psi-2}a.1), while the 45º phase is associated with SST in the eastern Pacific (Fig.~\ref{fig:sst-psi-2}c.1), which correspond roughly to the Central Pacific and Eastern Pacific ``flavours'' of ENSO, respectively (Kao and Yu 2009).
-Both of these phases could be also associated to wave trains generated in the region surrounding Australia and propagates toward the extra-tropics.
+Meanwhile, Rows a and c in Fig.~\ref{fig:sst-psi-2} show that the intermediate phases are still associated with significant SST regressed anomalies over the Pacific Ocean, but at slightly different locations.
+The 135º phase is associated with SST anomalies in the central Pacific (Fig.~\ref{fig:sst-psi-2}a.1), while the 45º phase is associated with SST anomalies (Fig.~\ref{fig:sst-psi-2}c.1) that correspond roughly to the Central Pacific and Eastern Pacific ``flavours'' of ENSO, respectively (Kao and Yu 2009).
+Both phases could be also associated to wave trains generated in the region surrounding Australia and propagates toward the extra-tropics.
 
@@
 
-To better explore the relationship between tropical forcing and phases of the cEOF2, Figure~\ref{fig:enso-phase} shows the ONI index plotted against the cEOF2 phase for each SON trimester between 1979 and 2019, highlighting years in which the magnitude of cEOF2 is above the median.
+To further explore the relationship between tropical forcing and phases of the cEOF2, Figure~\ref{fig:enso-phase} shows the ONI index plotted against the cEOF2 phase for each SON trimester between 1979 and 2019, highlighting years in which the magnitude of cEOF2 is above the median.
 In years with positive (negative) ONI, the cEOF2 phase is mostly around 90º (-90°), corresponding with positive (negative) Imaginary component.
@@
 Furthermore, even when using all years, the Spearman correlation --which is robust to outliers-- is also non-significant (0.2, p-value= 0.21).
-Therefore, although the location of tropical SST seem to have an effect in defining the phase of the cEOF2, the relationship between the magnitude of cEOF2 and ONI remains uncertain.
+Therefore, although the location of tropical SST anomalies seem to have an effect in defining the phase of the cEOF2, the relationship between the magnitude of cEOF2 and ONI remains uncertain might be evident only in particularly intense ENSO events, for which there are few data.
 
@@
 
-The influence of cEOFs variability in and 2-metre temperature and temperature in the Southern Hemisphere is also explored.
+The influence of cEOFs variability in the anomalies of both 2-metre air temperature and precipitation in the Southern Hemisphere is also explored.
 Figure~\ref{fig:pp-t2m-r2} shows the 2-meter temperature or precipitation anomalies explained variance by the multiple linear model of both Real and Imaginary cEOF1 components (column 1), and both Real and Imaginary cEOF2 components (column 2).
-The variance explained by cEOF1 for precipitation anomalies is extremely low and also for most of the temperature anomalies, except for the eastern western of Antarctica (Fig.~ \ref{fig:pp-t2m-r2}a.1).
+The variance explained by cEOF1 for precipitation anomalies is extremely low and also for most of the temperature anomalies, except for the northern tip of the Antarctic Peninsula, northern Weddell Sea and the Ross Sea coast (Fig.~ \ref{fig:pp-t2m-r2}a.1).
 
@@
 Cai et al.
-2020) and it is also similar to the precipitation anomalies associated with the A-SAM (@ Campitelli, Díaz, and Vera 2022).
+2020) and it is also similar to the precipitation anomalies associated with the A-SAM (Campitelli, Díaz, and Vera 2022).
 
@@
 \hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
-
-In previous sections, it has been shown that variability patterns that arise from cEOF methodology allow to describe the zonally asymmetric springtime extratropical SH circulation, reproducing previous features such as the variability related to PSAs or A-SAM.
-However, it could be questioned what is the advantage of using the methodology proposed here.
+\section{Discussion and conclusions}\label{discussion}}
+
+In previous sections, it has been shown that variability patterns that arise from cEOF methodology describe the zonally asymmetric springtime extratropical SH circulation, reproducing previous features such as the variability related to PSAs or A-SAM.
+
+However the advantage of using the methodology proposed here could be questioned.
 Firstly, the spatial fields obtained from both components of cEOF2, that resembles PSA patterns, are in quadrature by construction.
-Thus, it is possible to derive a joint PSA index from the amplitude and phase obtained when applying cEOF methodology.
-Secondly, because of how these components are constructed, they are not forced to be orthogonal to the SAM as in the standard EOF methodology.
-This allows us to show that Imaginary cEOF2, that resembles PSA1 variability, is closely related to the SAM in the troposphere.
-Moreover, the high correlation of this component to the the zonally asymmetric portion of the SAM allow to speculate that the A-SAM might actually be a statistical contamination of the PSA mode.
+Thus, the cEOF methodology allows to derive a joint PSA index from the resulted amplitude and phase.
+Secondly, because of how these components are constructed, they are not forced to be orthogonal to other modes of circulation, as in the standard EOF methodology.
+This allows us to show for example, that the Imaginary cEOF2, that resembles PSA1 variability, is closely associated to the SAM in the troposphere.
+Moreover, the high correlation of this component to the zonally asymmetric portion of the SAM allow us to speculate that the A-SAM might actually be a statistical contamination of the PSA mode on the symmetric SAM.
 
 Irving and Simmonds (2016) argued that there is some disagreement in the literature of whether the phase of the PSA pattern is affected by the location of tropical SST anomalies.
-With the methodology used in this study, we were able to show not only that the cEOF2 tends to be in the positive or negative imaginary phase (\textasciitilde PSA1) when the ENSO region is warm or cold, respectively, but that central Pacific SST anomalies tend to align the cEOF towards the negative real phase and eastern Pacific SST anomalies tend to align it towards positive real phase.
-When ENSO is neutral, the cEOF2 is still as active, but with no preferred phase.
-This mirrors the results of Cai and Watterson (2002), who showed that the CSIRO Model can create PSA-like variability even in the absence of ENSO forcing (i.e.~with a climatological run), but that the variability of one of the PSA modes was enhanced when adding the ENSO signal.
-
+With the methodology used in this study, we were able to show not only that the cEOF2 tends to be in the positive or negative imaginary phase (\textasciitilde PSA1) when the ENSO region is warm or cold, respectively, but also that central Pacific SST anomalies tend to align the cEOF towards the negative real phase and eastern Pacific SST anomalies tend to align it towards positive real phase.
+When ENSO phase is neutral, the cEOF2 is still as active, but with no preferred phase.
+The latter agrees with the results of Cai and Watterson (2002), who showed that the CSIRO Model can develop PSA-like variability even in the absence of ENSO forcing (i.e.~with a climatological run), but that the variability of one of the PSA modes was enhanced when adding the ENSO signal.
 The sensitivity of the phase of the PSAs to the location of the tropical SST anomalies was also seen by Ciasto, Simpkins, and England (2015), who detected similar Rossby wave patterns associated with central Pacific and eastern Pacific SST anomalies but with a change in phase.
 
-Some results of our work contrast with those obtained by Irving and Simmonds (2016) when analysing the PSA.
-They did not find neither a strong association between PSA-like variability and ENSO, nor a relationship between this variability and Australian precipitation.
-Part of the disagreement could be due to the fact that we focus our analysis only on SON, which is the season when the teleconnection patterns are strongest, instead of the whole year.
-Moreover, Irving and Simmonds (2016) use of ERA5 forectasted precipitation instead of the more observationally constrained CMAP dataset might be another source of disagreement.
-Additionally, the discrepancy could be related to the fact that Irving and Simmonds (2016) show a weak relationship between ENSO and their PSA pattern, which explains partly the relationship between Australian rainfall and PSA-like variability.
-
-\hypertarget{conclusion}{%
-\section{Conclusion}\label{conclusion}}
-
-In this study we assessed the Austral Spring Extratropical Southern Hemisphere zonally asymmetric circulation.
-For this purpose, two complex indices were derived using Complex Empirical Orthogonal Functions.
-This allowed us to characterise both the amplitude and phase of planetary waves that are not perfectly organised as sinusoidal waves with constant amplitude at each latitude circle.
-
-The cEOF1 represents the variability of the zonal wave 1 in the stratosphere, but represents a more zonally symmetric monopole in the troposphere.
-There is a statistically positive trend in the magnitude of this cEOF, which is consistent with previous studies that showed secular changes in springtime wave-1-like patterns (e.g. M. Raphael 2003).
-This mode is closely related to stratospheric variability such as anomalies in Total Ozone Column.
-Otherwise, this complex EOF is not related with SST variability and continental precipitation in the Southern Hemisphere.
-
-The cEOF2 represents a wave-3 pattern with maximum magnitude in the Pacific sector, that is an alternative representation of the PSA1 and PSA2 patterns (Mo and Paegle 2001).
-The Imaginary cEOF2 can be identified with the PSA1 and the Real cEOF2 with the PSA2.
-
-The phase of the cEOF2 is related to the the variability of ENSO, but its magnitude appears not to be, with the exception of the three strongest ENSO events in the period which coincide with the three strongest cEOF2 years.
-
-The cEOF2 variability could be related to surface impacts, while the cEOF1 influence is almost negligible.
-For instance, precipitation anomalies in South America associated with the Imaginary cEOF2 show a clear ENSO-like impact, with positive anomalies in South-Eastern South America, negative anomalies in Southern Brazil and positive anomalies in central Chile for positive Imaginary cEOF2 phase.
-Precipitation anomalies associated with the Real cEOF2 are low and not statistically significant in this region, showing that the Imaginary phase is optimally aligned with the direction of maximum precipitation impacts there.
-
-On the other hand, over Australia, both the Real and Imaginary phase are associated with significant precipitation anomalies, and we further show that the direction of maximum impact is not aligned with our chosen rotation of cEOF2.
-An analysis of the precipitation impacts on this regions could benefit for a different chosen rotation of the cEOF.
+Our method is similar to Goyal et al. (2022) as they also derive an index with amplitude and phase of wave-like activity in the SH with very similar patterns to our cEOF2.
+Both indices are likely very similar (a detailed comparison is out of scope for this paper) but we believe that cEOF have the advantage of constructing the indices based on patterns that are exactly in quadrature by construction.
 
@@
 Gong, Daoyi, and Shaowu Wang. 1999. {``Definition of {Antarctic Oscillation} Index.''} \emph{Geophysical Research Letters} 26 (4): 459--62. \url{https://doi.org/10.1029/1999GL900003}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-goyal2022}{}}%
+Goyal, Rishav, Martin Jucker, Alex Sen Gupta, and Matthew H.
+England.
+2022.
+{``A New Zonal Wave 3 Index for the {Southern Hemisphere}.''} \emph{Journal of Climate} -1 (May): 1--25.
+\url{https://doi.org/10.1175/JCLI-D-21-0927.1}.
 

### 2022-05-31 analysis/paper/shceof.Rmd
Arregla cosas de referencias
Quita lo del monopolo en el abstract

@@
 We analyse the joint variabilty of the zonally asymmetric springtime stratospheric and tropospheric circulation using Complex Empirical Orthogonal Functions (cEOF) to characterise planetary waves of varying amplitude and phase.
-The leading cEOF represents variabilty of a zonal wave 1 in the stratosphere and a monopole in the troposphere that correlates slightly with the Symmetric Southern Annular Mode (S-SAM).
+The leading cEOF represents variabilty of a zonal wave 1 troposphere that correlates slightly with the Symmetric Southern Annular Mode (S-SAM).
 The second cEOF (cEOF2) is an alternative representation of the Pacific-South American modes. 
@@
 In Section\ \@ref(discussion) we compare our results with previous studies and discuss the benefits of our methodology.
-In Section\ \@ref(conclusion) we summarise the conclusions.
 
@@
 
-# Discussion and conclusions
+# Discussion and conclusions {#discussion}
 

### 2022-05-31 analysis/paper/shceof.tex
Render

@@
 We analyse the joint variabilty of the zonally asymmetric springtime stratospheric and tropospheric circulation using Complex Empirical Orthogonal Functions (cEOF) to characterise planetary waves of varying amplitude and phase.
-The leading cEOF represents variabilty of a zonal wave 1 in the stratosphere and a monopole in the troposphere that correlates slightly with the Symmetric Southern Annular Mode (S-SAM).
+The leading cEOF represents variabilty of a zonal wave 1 troposphere that correlates slightly with the Symmetric Southern Annular Mode (S-SAM).
 The second cEOF (cEOF2) is an alternative representation of the Pacific-South American modes.

### 2022-06-03 analysis/paper/shceof.Rmd
Agerga comentarios Leandro

@@
 
-We analyse the joint variabilty of the zonally asymmetric springtime stratospheric and tropospheric circulation using Complex Empirical Orthogonal Functions (cEOF) to characterise planetary waves of varying amplitude and phase.
-The leading cEOF represents variabilty of a zonal wave 1 troposphere that correlates slightly with the Symmetric Southern Annular Mode (S-SAM).
+The large-scale extratropical circulation in the Southern Hemisphere is strongly zonally symmetric, but its zonal departures, albeit highly relevant for regional impacts, have been less studied.
+In this study we analyse the joint variability of the zonally asymmetric springtime stratospheric and tropospheric circulation using Complex Empirical Orthogonal Functions (cEOF) to characterise planetary waves of varying amplitude and phase.
+The leading cEOF represents variability of a zonal wave 1 in the stratosphere that correlates slightly with the Symmetric Southern Annular Mode (S-SAM).
 The second cEOF (cEOF2) is an alternative representation of the Pacific-South American modes. 
@@
 Furthermore, we find indications that the location of Pacific Sea Surface Temperature anomalies affect the phase of the cEOF2. 
+As a result, the methodology proposed in this study provides a deeper understanding of the zonally asymmetric springtime extratropical SH circulation.
 
@@
 
-Both the spatial pattern similarity and the high temporal correlation of cOEFs computed at 50 hPa and 200 hPa suggest that these are, to a large extent, modes of joint variability.
+Both the spatial pattern similarity and the high temporal correlation of cEOFs computed at 50 hPa and 200 hPa suggest that these are, to a large extent, modes of joint variability.
 This motivates the decision of performing complex EOF jointly between levels.
@@
 The Imaginary cEOF1 is associated with a distinctive wave 1 pattern with maximum over the coast of East Antarctica. 
-
 At 200 hPa (Figure \@ref(fig:eof1-regr-gh) row b) the Real cEOF1 shows a single centre of positive anomalies spanning West Antarctica surrounded by opposite anomalies in lower latitudes, with is centre shifted slightly eastward compared with the upper-level anomalies.
 The Imaginary cEOF1 shows a much more zonally symmetrical pattern resembling the negative SAM phase [e.g. @fogt2020].
-
 Therefore, the magnitude and phase of the cEOF1 are associated with the magnitude and phase of a zonal wave mainly in the stratosphere.
@@
 Conventional EOF analysis will therefore tend to separate them relatively cleanly, with the EOF representing the Imaginary cEOF2 always leading the one representing the Real cEOF2.
-
 This phase preferences is in agreement with @irving2016, who found a bimodal distribution to PSA-like variability (compare our Figure \@ref(fig:phase-histogram) with their Figure 6).
@@
 
-In the case of Imaginary cEOF2, the correlation between SAM and Imaginary cEOF2 for the troposphere is associated to the asymmetric variability of the SAM. 
+In the case of Imaginary cEOF2, its correlation with the SAM for the troposphere is associated to the asymmetric variability of the SAM. 
 Indeed, the Imaginary cEOF2 shares up to `r scales::percent(max_r2[cEOF == "cEOF2" & term == "asym"]$estimate^2, 2)` variance with the A-SAM and only `r scales::percent(max_r2[cEOF == "cEOF2" & term == "sym"]$estimate^2, 2)` at most with the S-SAM (Figure \@ref(fig:sam-eof-vertical).b2).
@@
 Furthermore, even when using all years, the Spearman correlation --which is robust to outliers-- is also non-significant (`r signif(cor_enso_spearman$estimate, 2)`, p-value`r report_p(cor_enso_spearman$p.value)`).
-Therefore, although the location of tropical SST anomalies seem to have an effect in defining the phase of the cEOF2, the relationship between the magnitude of cEOF2 and ONI remains uncertain might be evident only in particularly intense ENSO events, for which there are few data.
+Therefore, although the location of tropical SST anomalies seem to have an effect in defining the phase of the cEOF2, the relationship between the magnitude of cEOF2 and ONI remains uncertain and might be only evident in very strong ENSO events, that are scarce in the historical observational record.
 
@@
 
-The influence of cEOFs variability in the anomalies of both 2-metre air temperature and precipitation in the Southern Hemisphere is also explored.
+The influence of cEOFs variability in the anomalies of both 2-metre air temperature and precipitation in the SH was also explored.
 Figure\ \@ref(fig:pp-t2m-r2) shows the 2-meter temperature or precipitation anomalies explained variance by the multiple linear model of both Real and Imaginary cEOF1 components (column 1), and both Real  and Imaginary cEOF2 components (column 2).
@@
 
-In previous sections, it has been shown that variability patterns that arise from cEOF methodology describe the zonally asymmetric springtime extratropical SH circulation, reproducing previous features such as the variability related to PSAs or A-SAM.
+In this study we assessed extratropical Southern Hemisphere zonally asymmetric circulation in austral spring.
+For this purpose, two complex indices were derived using Complex Empirical Orthogonal Functions and used to characterise both amplitude and phase of planetary waves.
+
+The cEOF1 represents the variability of the zonal wave 1 in the stratosphere and is closely related to stratospheric variability such as anomalies in Total Ozone Column.
+Otherwise, this complex EOF is not related with SST variability and continental precipitation in the Southern Hemisphere.
+On the other hand, the cEOF2 represents a wave-3 pattern with maximum magnitude in the Pacific sector, that is an alternative representation of the PSA1 and PSA2 patterns (Mo and Paegle 2001).
+The Imaginary cEOF2 can be identified with the PSA1 and the Real cEOF2 with the PSA2.
+
+While the cEOF2 variability could be related to surface impacts, the cEOF1 influence is almost negligible.
+For instance, precipitation anomalies in South America associated with the Imaginary cEOF2 show a clear ENSO-like impact, with positive anomalies in South-Eastern South America, negative anomalies in Southern Brazil and positive anomalies in central Chile for positive Imaginary cEOF2 phase.
+
+Variability patterns that arise from cEOF methodology describe the zonally asymmetric springtime extratropical SH circulation, reproducing previous features such as the variability related to PSAs or A-SAM.
 
@@
 Our method is similar to @goyal2022 as they also derive an index with amplitude and phase of wave-like activity in the SH with very similar patterns to our cEOF2. 
-Both indices are likely very similar (a detailed comparison is out of scope for this paper) but we believe that cEOF have the advantage of constructing the indices based on patterns that are exactly in quadrature by construction.
+Both indices are likely very similar (a detailed comparison is out of scope for this paper) but cEOF have the advantage of constructing the indices based on patterns that are exactly in quadrature by construction.
 

### 2022-06-03 analysis/paper/shceof.Rmd
Renderiza

@@
   comment = "#>",
-  cache.extra = 46,
+  cache.extra = 42,
   cache.path = "../cache/paper/",

### 2022-06-03 analysis/paper/shceof.tex
Renderiza

@@
 \begin{abstract}
-We analyse the joint variabilty of the zonally asymmetric springtime stratospheric and tropospheric circulation using Complex Empirical Orthogonal Functions (cEOF) to characterise planetary waves of varying amplitude and phase.
-The leading cEOF represents variabilty of a zonal wave 1 troposphere that correlates slightly with the Symmetric Southern Annular Mode (S-SAM).
+The large-scale extratropical circulation in the Southern Hemisphere is strongly zonally symmetric, but its zonal departures, albeit highly relevant for regional impacts, have been less studied.
+In this study we analyse the joint variability of the zonally asymmetric springtime stratospheric and tropospheric circulation using Complex Empirical Orthogonal Functions (cEOF) to characterise planetary waves of varying amplitude and phase.
+The leading cEOF represents variability of a zonal wave 1 in the stratosphere that correlates slightly with the Symmetric Southern Annular Mode (S-SAM).
 The second cEOF (cEOF2) is an alternative representation of the Pacific-South American modes.
@@
 Furthermore, we find indications that the location of Pacific Sea Surface Temperature anomalies affect the phase of the cEOF2.
+As a result, the methodology proposed in this study provides a deeper understanding of the zonally asymmetric springtime extratropical SH circulation.
 \\
@@
 The Imaginary cEOF1 is associated with a distinctive wave 1 pattern with maximum over the coast of East Antarctica.
-
 At 200 hPa (Figure~\ref{fig:eof1-regr-gh} row b) the Real cEOF1 shows a single centre of positive anomalies spanning West Antarctica surrounded by opposite anomalies in lower latitudes, with is centre shifted slightly eastward compared with the upper-level anomalies.
 The Imaginary cEOF1 shows a much more zonally symmetrical pattern resembling the negative SAM phase (e.g. Fogt and Marshall 2020).
-
 Therefore, the magnitude and phase of the cEOF1 are associated with the magnitude and phase of a zonal wave mainly in the stratosphere.
@@
 Conventional EOF analysis will therefore tend to separate them relatively cleanly, with the EOF representing the Imaginary cEOF2 always leading the one representing the Real cEOF2.
-
 This phase preferences is in agreement with Irving and Simmonds (2016), who found a bimodal distribution to PSA-like variability (compare our Figure~\ref{fig:phase-histogram} with their Figure 6).
@@
 
-In the case of Imaginary cEOF2, the correlation between SAM and Imaginary cEOF2 for the troposphere is associated to the asymmetric variability of the SAM.
+In the case of Imaginary cEOF2, its correlation with the SAM for the troposphere is associated to the asymmetric variability of the SAM.
 Indeed, the Imaginary cEOF2 shares up to 92\% variance with the A-SAM and only 12\% at most with the S-SAM (Figure~\ref{fig:sam-eof-vertical}.b2).
@@
 Furthermore, even when using all years, the Spearman correlation --which is robust to outliers-- is also non-significant (0.2, p-value= 0.21).
-Therefore, although the location of tropical SST anomalies seem to have an effect in defining the phase of the cEOF2, the relationship between the magnitude of cEOF2 and ONI remains uncertain might be evident only in particularly intense ENSO events, for which there are few data.
+Therefore, although the location of tropical SST anomalies seem to have an effect in defining the phase of the cEOF2, the relationship between the magnitude of cEOF2 and ONI remains uncertain and might be only evident in very strong ENSO events, that are scarce in the historical observational record.
 
@@
 
-The influence of cEOFs variability in the anomalies of both 2-metre air temperature and precipitation in the Southern Hemisphere is also explored.
+The influence of cEOFs variability in the anomalies of both 2-metre air temperature and precipitation in the SH was also explored.
 Figure~\ref{fig:pp-t2m-r2} shows the 2-meter temperature or precipitation anomalies explained variance by the multiple linear model of both Real and Imaginary cEOF1 components (column 1), and both Real and Imaginary cEOF2 components (column 2).
@@
 Our method is similar to Goyal et al. (2022) as they also derive an index with amplitude and phase of wave-like activity in the SH with very similar patterns to our cEOF2.
-Both indices are likely very similar (a detailed comparison is out of scope for this paper) but we believe that cEOF have the advantage of constructing the indices based on patterns that are exactly in quadrature by construction.
+Both indices are likely very similar (a detailed comparison is out of scope for this paper) but cEOF have the advantage of constructing the indices based on patterns that are exactly in quadrature by construction.
 
@@
 {``A New Zonal Wave 3 Index for the {Southern Hemisphere}.''} \emph{Journal of Climate} -1 (May): 1--25.
-\url{https://doi.org/10.1175/JCLI-D-21-0927.1}.
+\url{https://journals.ametsoc.org/view/journals/clim/aop/JCLI-D-21-0927.1/JCLI-D-21-0927.1.xml}.
 

### 2022-06-03 analysis/paper/shceof.Rmd
Cambia código para usar datos de zenodo finales

@@
 
-sams <- readRDS(data_path("derived", "sam.Rds")) %>%
+sams <- SAM() %>% 
+  fread() %>%
   .[season(time) == "SON"] %>%

### 2022-06-03 R/data_locations.R

@@ -5,7 +5,8 @@
 #'
 #'
 #' @param type character indicating the type of data. The resulting base will be type_data
-#' @param ... characters to be contactenadted with [file.path()]
+#' @param ... characters to be concatenated with [file.path()]
+#'
 #' @export
 #' @rdname data_locations
 data_path <- function(type = c("raw", "derived"), ...) {
@@ -47,20 +48,22 @@ data_function <- function(name,
   )
 
   function(from_source = FALSE) {
-    if (from_source) {
-      message("Downloading dataset ", data$name, " from remote source.")
-      download_from_source(data)
-    }
+    will_download <- from_source == TRUE || !file.exists(data$file)
 
-    if (!file.exists(data$file)) {
-      message("Downloading dataset ", data$name, " from zenodo repository.")
-      download_from_zenodo(data)
+    if (will_download) {
+      if (from_source || is.null(data$zenodo_url)) {
+        message("Downloading dataset ", data$name, " from remote source.")
+        download_from_source(data)
+      } else {
+        message("Downloading dataset ", data$name, " from zenodo repository.")
+        download_from_zenodo(data)
+      }
     }
 
     if (!is.null(data$zenodo_md5)) {
       md5 <- digest::digest(data$file, "md5", file = TRUE)
       if (data$zenodo_md5 != md5) {
-        stop("File for dataset ", data$name, " has incorrect checksum.")
+        warning("File for dataset ", data$name, " has incorrect checksum compared to the Zenodo reference.")
       }
     }
 
@@ -73,13 +76,24 @@ data_function <- function(name,
 #' @param from_source Logical indicating whether to force download of data from the
 #' remote source instead of the Zeonodo repository.
 #'
+#' @details
+#' Each dataset has two potential sources, either the original source or the
+#' Zenodo repository associated with this paper. The default behaviour is to
+#' download the file from Zenodo if the file does not exist. This will
+#' get the exact files that where used for the paper.
[…]

### 2022-06-03 analysis/scripts/01-compute_uv.R

@@ -1,25 +1,0 @@
-subset = list(lat = -90:10)
-levels = c(50, 200, 500)
-
-years <- seq(1979, 2018)
-
-uv <- data.table::rbindlist(lapply(years, function(y) {
-  cat("Processing year ", y, "\r")
-  ufile <- paste0("datos/NCEP Reanalysis/daily/uwnd.", y, ".nc")
-  vfile <- paste0("datos/NCEP Reanalysis/daily/vwnd.", y, ".nc")
-  
-  wnd <- metR::ReadNetCDF(ufile, "uwnd", subset = subset)
-  wnd[, vwnd := metR::ReadNetCDF(vfile, "vwnd", subset = subset, out = "vector")[[1]]]
-  wnd <- wnd[level %in% levels]
-  
-  wnd[, year := year(time[1]), by = time]
-  wnd[, season := metR::season(time[1]), by = time]
-  
-  wnd[, .(uv = cov(uwnd, vwnd)), by = .(level, lon, lat, year, season)]
-}))
-
-data.table::setnames(uv, "level", "lev")
-uv[, dataset := "ncep"]
-cat("Saving data.")
-saveRDS(uv, "datos/uv.Rds")
-

### 2022-06-03 analysis/scripts/01-get-data.R

@@ -1,26 +1,0 @@
-library(shceof)
-
-force <- FALSE
-
-# Download CMAP -----------------------------------------------------------
-CMAP(force)
-
-# Download HadSST -----------------------------------------------------------
-HADSST(force)
-
-# Download Streamfunction NCEP ----------------------------------------------------
-NCEP_PSI(force)
-
-# Download Vorticity NCEP ----------------------------------------------------
-NCEP_VOR(force)
-
-# Download ERA5 -----------------------------------------------------------
-ERA5(force)
-ERA5_BE(force)
-ERSST(force)
-
-# Download ERA20C -----------------------------------------------------------
-ERA20C(force)
-
-# Download O3 --------------------------------
-O3(force)

### 2022-06-03 analysis/scripts/01-preprocess_data.R

@@ -1,54 +1,0 @@
-library(metR)
-library(data.table)
-library(magrittr)
-library(here)
-
-
-setnames <- function(x, ...) {
-  names <- c(...)
-  # print(names)
-  data.table::setnames(x, names(names), unname(names))
-}
-
-
-subset_ncep <- list(lat = -90:0,
-                    time = c("1979-01-01", "2018-12-31"))
-subset_era <- list(latitude = -90:0,
-                   time = c("1979-01-01", "2018-12-31"))
-
-ncep <- ReadNetCDF(here("datos", "NCEP Reanalysis", "hgt.mon.mean.nc"),
-                   subset = subset_ncep) %>%
-  .[, dataset := "ncep"] %>% 
-  setnames(., level = "lev") %>% 
-  .[lev %in% c(200, 50)]
-
-era <- ReadNetCDF(here("datos", "ERA-Interim", "hgt.mon.mean.nc"),
-                  c(hgt = "z"),
-                  subset = subset_era) %>%
-  .[, hgt := hgt/9.8] %>%
-  .[, dataset := "era"] %>%
-  setnames(longitude = "lon", latitude = "lat", level = "lev") %>% 
-  .[lev %in% c(200, 50)]
-
-era20 <- rbind(ReadNetCDF(here("datos", "ERA-20C", "hgt.mon.mean.nc"),
-                          c(hgt = "z"),
-                          subset = list(latitude = -90:0, 
-                                        level = 50)),
-               ReadNetCDF(here("datos", "ERA-20C", "hgt.mon.mean.nc"),
-                          c(hgt = "z"),
-                          subset = list(latitude = -90:0, 
-                                        level = 200))) %>% 
-  .[, hgt := hgt/9.8] %>%
-  .[, dataset := "era20"] %>%
-  setnames(longitude = "lon", latitude = "lat", level = "lev")
-
-datos <- rbind(era20, era, ncep)
-# datos <- ncep
-# remove(ncep)
-remove(era, ncep, era20)
-
-datos <- datos[, .(hgt = mean(hgt)), 
-               by = .(lev, lon, lat, year(time), season(time), dataset)]
-
-
-saveRDS(datos, here("datos", "datos.Rds"))

### 2022-06-03 analysis/scripts/02-compute-sam.R

@@ -1,26 +1,0 @@
-sam_file <- data_path("derived", "sam.Rds")
-
-if (file.exists(sam_file)) {
-  message("Sam file already exists. Exiting.")
-} else {
-  check_asymsam <- requireNamespace("asymsam", quietly = TRUE)
-  if (!check_asymsam) {
-    stop("Computing the Asymmetric and Symmetric SAM requires package asymsam.\n",
-         "Install it with remotes::install_github(\"eliocamp/asymsam\").")
-  }
-
-  library(shceof)
-  library(metR)
-
-  sams <- ERA5_geopotential_all() %>%
-    ReadNetCDF(vars = c("hgt" = "z"),
-               subset = list(latitude = c(-90, -20))) %>%
-    normalise_coords() %>%
-    .[, hgt := Anomaly(hgt), by = .(lon, lat, lev, month(time))] %>%
-    .[, asymsam::eof_asym(hgt, lon, lat, time), by = lev] %>%
-    rm_singleton()
-
-  saveRDS(sams, sam_file)
-}
-
-

### 2022-06-03 analysis/scripts/facet_nested.R

@@ -1,417 +1,0 @@
-library(grid)
-library(gtable)
-#' @title Layout panels in a grid with nested strips
-#'
-#' @description \code{facet_nest()} form a matrix of panels defined by row and
-#'   column faceting variables and nests grouped facets.
-#'
-#' @param
-#' rows,cols,scales,space,shrink,labeller,as.table,switch,drop,margins,facets
-#' see \code{\link[ggplot2]{facet_grid}}
-#' @param nest_line a \code{logical} vector of length 1, indicating whether to
-#'   draw a nesting line to indicate the nesting of variables.
-#' @param resect  a \code{unit} vector of length 1, indicating how much the
-#'   nesting line should be shortened.
-#' @param bleed a \code{logical} vector of length 1, indicating wether merging
-#'   of lower-level variables is allowed when the higher-level variables are
-#'   seperate. See details.
-#'
-#' @details Unlike \code{facet_grid()}, this function only automatically expands
-#'   missing variables when they have no variables in that direction, to allow
-#'   for unnested variables. It still requires at least one layer to have all
-#'   faceting variables.
-#'
-#'   Hierarchies are inferred from the order of variables supplied to
-#'   \code{rows} or \code{cols}. The first variable is interpreted to be the
-#'   outermost variable, while the last variable is interpreted to be the
-#'   innermost variable. They display order is always such that the outermost
-#'   variable is placed the furthest away from the panels. Strips are
-#'   automatically grouped when they span a nested variable.
-#'
-#'   The \code{bleed} argument controls wether lower-level variables are allowed
-#'   to be merged when higher-level are different, i.e. they can bleed over
-#'   hierarchies. Suppose the \code{facet_grid()} behaviour would be the
-#'   following:
-#'
-#'   \code{[_1_][_2_][_2_]} \cr \code{[_3_][_3_][_4_]}
-#'
-#'   In such case, the default \code{bleed = FALSE} argument would result in the
-#'   following:
-#'
-#'   \code{[_1_][___2____]} \cr \code{[_3_][_3_][_4_]}
-#'
-#'   Whereas \code{bleed = TRUE} would allow the following:
-#'
-#'   \code{[_1_][___2____]} \cr \code{[___3____][_4_]}
-#'
-#' @export
-#'
[…]

### 2022-06-03 analysis/scripts/helperfun.R

@@ -1,1689 +1,0 @@
-## Archivo con funciones
-
-#Librerías
-# library(ggplot2)
-# library(ggforce)
-# library(stringi)
-# # library(ggthemes)
-# library(magrittr)
-# library(data.table)
-# library(lubridate)
-# library(akima)
-
-# library(RColorBrewer)
-library(compiler)
-enableJIT(0)
-
-# source(here::here("scripts/eof_methods.R"))
-
-`%b%` <- function(x, y) data.table::`%between%`(x, y) 
-
-# Mapa
-BuildMap <- function(res = 1, smooth = 1, pm = 180,
-                     countries = FALSE, ...) {
-   # Caraga datos de mapas y cambia el meridiano principal, suaviza y cambia
-   # resolución para plotear más rápido.
-   # Entra:
-   #   res: resolución (cuantos puntos se eliminan entre los que quedan)
-   #   smooth: factor de suavifzado (ventana del promedio corrido)
-   #   pm: longitud del meridiano central
-   #   countries: ¿datos a nivel país?
-   # Sale:
-   #   un data.table con las coordenadas de cada polígono y su grupo
-   library(data.table)
-   if (countries) {
-      library(rworldxtra)
-      data("countriesLow")
-      m <- countriesLow
-   } else {
-      library(rworldmap)
-      data(coastsCoarse)
-      m <- coastsCoarse
-   }
-   
-   m <- as.data.table(fortify(m))
-   m[, group := as.numeric(group)]
-   m[, id := as.numeric(id)]
-   
-   # Cambio el prime meridian.
-   m2 <- copy(m)
-   m2[, long := long + 360]
-   m2[, group := group + max(group) + 1]
-   m <- rbind(m, m2)
-   m <- m[long >= pm - 180 & long <= 180 + pm]
-   
-   m[, MO := max(order), by = group]
-   
-   # Suavizo.
-   if (smooth > 1) {
-      cut <- max(smooth, res)
-      notsmooth <- m[, .N, by = group][N < cut + 4, group]
-      m <- m[!(group %in% notsmooth)]    # saco los grupos muy chicos
-      
-      m[order != 1 & order != MO,
-        `:=`(long = zoo::rollmean(long, smooth, fill = "extend"),
-             lat = zoo::rollmean(lat, smooth, fill = "extend")),
-        by = group]
-   }
-   # Bajo la resolución.
[…]

### 2022-06-03 analysis/scripts/preprocess_data.R

@@ -1,47 +1,0 @@
-library(metR)
-library(data.table)
-library(magrittr)
-library(here)
-
-
-setnames <- function(x, ...) {
-  names <- c(...)
-  # print(names)
-  data.table::setnames(x, names(names), unname(names))
-}
-
-
-subset_ncep <- list(lat = -90:0,
-                    time = c("1979-01-01", "2018-12-31"))
-subset_era <- list(latitude = -90:0,
-                   time = c("1979-01-01", "2018-12-31"))
-
-ncep <- ReadNetCDF(here("datos", "NCEP Reanalysis", "hgt.mon.mean.nc"),
-                   subset = subset_ncep) %>% 
-  .[, u := ReadNetCDF(here("datos", "NCEP Reanalysis", "uwnd.mon.mean.nc"),
-                      out = "vector", 
-                      subset = subset_ncep)] %>% 
-  .[, dataset := "ncep"] %>% 
-  setnames(., level = "lev")
-
-era <- ReadNetCDF(here("datos", "ERA-Interim", "hgt.mon.mean.nc"),
-                  c(hgt = "z"),
-                  subset = subset_era) %>%
-  .[, hgt := hgt/9.8] %>%
-  .[, u := ReadNetCDF(here("datos", "ERA-Interim", "uwnd.mon.mean.nc"),
-                      c(u = "u"),
-                      out = "vector",
-                      subset = subset_era)] %>%
-  .[, dataset := "era"] %>%
-  setnames(longitude = "lon", latitude = "lat", level = "lev")
-
-datos <- rbind(era, ncep)
-# datos <- ncep
-# remove(ncep)
-remove(era, ncep)
-
-datos <- datos[, .(u = mean(u), hgt = mean(hgt)), 
-               by = .(lev, lon, lat, year(time), season(time), dataset)]
-
-
-saveRDS(datos, "datos/datos.Rda")

### 2022-06-03 analysis/paper/shceof.Rmd
Agerga data availabilty y modifica goyal

@@
 In Section\ \@ref(discussion) we compare our results with previous studies and discuss the benefits of our methodology.
-
-<!-- The large-scale extratropical circulation in the Southern Hemisphere (SH) is strongly zonally symmetric, but its zonal departures are highly relevant for regional impacts and could even be related to the occurrence of high-impact climate extremes (REFERENCE).
--->
-
-<!-- Zonal asymmetries of extratropical circulation in the Southern Hemisphere strongly modulate weather systems and regional climate through promoting longitudinally different latitudinal transport of heat, humidity, and momentum [@trenberth1980].
--->
-
-<!-- Hablar simplemente de cómo estos fenómenos varían en longitud.  -->
-
-<!-- @hill2009 ->  -->
-
-<!-- Comparan eventos ENSO del 97 y de 2002 -->
-
-<!-- La posición de las anomalías de circulación (PSA) son importantes para determinar el impacto del ENSO en sudamérica.
-Sugiere que la localización de la anomalía de SST cambia la localización del PSA.
--->
-
-<!-- > The meridional winds from the idealised experiments based on simplified and localised SST warming showed a significant difference between experiments forced by eastern versus central-west Pacific SST anomalies, with a clear phase shift of the PSA apparent.
-This phase shift in the idealised experiments is possibly the main factor responsible for the different responses in South American atmospheric circulation between the 1997/1998 and 2002/2003 El Niño events.
-The phase shift over southeastern South America was also seen in the low-level 850 hPa winds in the EN97 and EN02 experiments, suggesting that the PSA phase shift is responsible for the shift in the moisture transport over southeastern Brazil between the two El Niño events.
-In summary, the location of the maximum SST anomaly appears to be an important factor in controlling the amount of moisture transported into the SACZ region during El Niño events.
--->
-
-<!-- @hobbs2010 -> muestra la importancia de la longitud de ciclones -->
-
-<!-- ciasto2015 -> no investiga impactos -->
-
-<!-- sun2013 -->
-
-<!-- wilson2014 -> -->
-
-<!-- Observaciones y modelos enso canonico y modoki. diferentes patrones de circulacion.  -->
-
-<!-- wang2021 ->  -->
-
-<!-- > However, the spatial pattern of the PSA wave train excited by ENSO seems to be -->
-
-<!-- > non-stationary. Fogt and Bromwich (2006) indicated that the PSA mode moved -->
-
-<!-- > eastwards by 20° longitude in the 1990s owing to the eastward shift of the tropical -->
-
-<!-- > convective position. In addition, Wilson et al. (2014, 2016) and Ciasto et al. (2015) -->
-
-<!-- > found that the different types of ENSO can contribute to the wave shift. Specifically, -->
-
-<!-- > Rossby waves excited by central Pacific (CP)–type El Niño shift westwards owing to -->
-
-<!-- > the westward movement of the tropical diabatic heating. -->
-
-<!-- Diferencia en la localización de los centros en LN y EN -> diferencias en los impactos en la península antártica.  -->
-
-<!-- Lo más importante es la localización de la salida del jet, no la perturbación de sst.  -->
-
-<!-- fogt2006 ->  -->
-
-<!-- cambios en la posición de las teleccionexiones. Pero es todo con composiciones y comparando décadas. No mira impactos. -->
-
-<!-- lou2021 ->  -->
-
-<!-- > In combination, they produce the single propagating PSA mode. -->
-
-<!-- Complicado y no entiendo mucho, pero cosas que puedo ver es que PSA forza el océano subtropical que luego es precursora de ENSO. No sé si eso es súper  -->
 
@@
 
-Our method is similar to @goyal2022 as they also derive an index with amplitude and phase of wave-like activity in the SH with very similar patterns to our cEOF2. 
-Both indices are likely very similar (a detailed comparison is out of scope for this paper) but cEOF have the advantage of constructing the indices based on patterns that are exactly in quadrature by construction.
+The method used in this study have similarities to the one used by @goyal2022 as they construct an index of amplitude and phase of zonal wave 3-like variability by combining the two leading EOFs of meridional wind anomalies.
+The patterns obtained by them bear high resemblance with cEOF2. 
+Although a detailed comparison is out of scope for this paper, the cEOF analysis has the advantage of constructing the indices based on patterns that are exactly in quadrature by construction.
 
@@
 
+## Availability of data and material {.unnumbered}
+
+All data used in this paper available in a version-controlled repository [@campitelli2022a]. 
+It is also freely available from their respective sources:
+
+- ERA5 data can be obtained via the Copernicus Climate Data Store (<https://cds.climate.copernicus.eu/cdsapp\#!/dataset/reanalysis-era5-pressure-levels-monthly-means\>).
+
+- ERSSTv5 can be obtained via NOAA's NCEI websiste at <https://www.ncei.noaa.gov/access/metadata/landing-page/bin/iso?id=gov.noaa.ncdc:C00927>
+
+- CMAP Precipitation data provided by the NOAA/OAR/ESRL PSL, Boulder, Colorado, USA, from their Web site at <https://psl.noaa.gov/data/gridded/data.cmap.html>.
+
+- The Oceanic Niño Index is available via NOAA's Climate Prediction Center: <https://www.cpc.ncep.noaa.gov/products/analysis_monitoring/ensostuff/detrend.nino34.ascii.txt>.
+
+-   The Oceanic Niño Index is available via NOAA's Climate Prediction Center: <https://www.cpc.ncep.noaa.gov/products/analysis_monitoring/ensostuff/detrend.nino34.ascii.txt>.
+
+-   The Dipole Mode Index is available via Global Climate Observing System Working Group on Surface Pressure: <https://psl.noaa.gov/gcos_wgsp/Timeseries/Data/dmi.had.long.data>
+
 ## Code availability {.unnumbered}
@@
 A version-controlled repository of the code used to create this analysis, including the code used to download the data can be found at <https://github.com/eliocamp/shceof>.
-
--   The Oceanic Niño Index is available via NOAA's Climate Prediction Center: <https://www.cpc.ncep.noaa.gov/products/analysis_monitoring/ensostuff/detrend.nino34.ascii.txt>.
-
--   The Dipole Mode Index is available via Global Climate Observing System Working Group on Surface Pressure: <https://psl.noaa.gov/gcos_wgsp/Timeseries/Data/dmi.had.long.data>
 

### 2022-06-03 analysis/paper/shceof.tex
Render

@@
 We further split the SAM into its zonally symmetric and zonally asymmetric components (S-SAM and A-SAM indices respectively).
-These indices were obtained by projecting the zonally asymmetric and zonally symmetric part of the SAM spatial pattern onto monthly geopotential height fields, as proposed by Campitelli, Díaz, and Vera (2022).
+These indices were obtained by projecting the zonally asymmetric and zonally symmetric part of the SAM spatial pattern onto monthly geopotential height fields, as proposed by Campitelli, Díaz, and Vera (2022a).
 Seasonal indices of the PSA patterns (PSA1 and PSA2) were calculated, in agreement with Mo and Paegle (2001), as the third and fourth leading EOF of seasonal mean anomalies for 500-hPa geopotential heights at SH.
@@
 We now explore the relationship between SAM and the cEOFs motivated by the resemblance between cEOFs regression maps and SAM patterns shown in Section~\ref{regressions}.
-We then computed the coefficient of determination between the cEOFs time series and the three SAM indices (SAM, A-SAM and S-SAM) defined by Campitelli, Díaz, and Vera (2022) at each vertical level (Fig.~\ref{fig:sam-eof-vertical}).
+We then computed the coefficient of determination between the cEOFs time series and the three SAM indices (SAM, A-SAM and S-SAM) defined by Campitelli, Díaz, and Vera (2022a) at each vertical level (Fig.~\ref{fig:sam-eof-vertical}).
 The SAM index is statistically significant correlated with the Real cEOF1 in all levels, and with the Imaginary cEOF1 and Imaginary cEOF2 in the troposphere.
@@
 In the stratosphere, the Real cEOF1 is correlated with both A-SAM and S-SAM, while the Imaginary cEOF1 is highly correlated only with the A-SAM.
-These correlations are consistent with the regression maps of geopotential height in Figure~\ref{fig:eof1-regr-gh} and their comparison with those obtained for SAM, A-SAM and S-SAM by Campitelli, Díaz, and Vera (2022).
+These correlations are consistent with the regression maps of geopotential height in Figure~\ref{fig:eof1-regr-gh} and their comparison with those obtained for SAM, A-SAM and S-SAM by Campitelli, Díaz, and Vera (2022a).
 
@@
 Indeed, the Imaginary cEOF2 shares up to 92\% variance with the A-SAM and only 12\% at most with the S-SAM (Figure~\ref{fig:sam-eof-vertical}.b2).
-Such extremely high correlation between A-SAM and Imaginary cEOF2 suggests that the modes obtained in this work are able to characterise the zonally asymmetric component of the SAM described previously by Campitelli, Díaz, and Vera (2022).
+Such extremely high correlation between A-SAM and Imaginary cEOF2 suggests that the modes obtained in this work are able to characterise the zonally asymmetric component of the SAM described previously by Campitelli, Díaz, and Vera (2022a).
 
@@
 Cai et al.
-2020) and it is also similar to the precipitation anomalies associated with the A-SAM (Campitelli, Díaz, and Vera 2022).
+2020) and it is also similar to the precipitation anomalies associated with the A-SAM (Campitelli, Díaz, and Vera 2022a).
 
@@
 
-Our method is similar to Goyal et al. (2022) as they also derive an index with amplitude and phase of wave-like activity in the SH with very similar patterns to our cEOF2.
-Both indices are likely very similar (a detailed comparison is out of scope for this paper) but cEOF have the advantage of constructing the indices based on patterns that are exactly in quadrature by construction.
+The method used in this study have similarities to the one used by Goyal et al.
+(2022) as they construct an index of amplitude and phase of zonal wave 3-like variability by combining the two leading EOFs of meridional wind anomalies.
+The patterns obtained by them bear high resemblance with cEOF2.
+Although a detailed comparison is out of scope for this paper, the cEOF analysis has the advantage of constructing the indices based on patterns that are exactly in quadrature by construction.
 
@@
 
-\hypertarget{code-availability}{%
-\subsection*{Code availability}\label{code-availability}}
-\addcontentsline{toc}{subsection}{Code availability}
-
-A version-controlled repository of the code used to create this analysis, including the code used to download the data can be found at \url{https://github.com/eliocamp/shceof}.
+\hypertarget{availability-of-data-and-material}{%
+\subsection*{Availability of data and material}\label{availability-of-data-and-material}}
+\addcontentsline{toc}{subsection}{Availability of data and material}
+
+All data used in this paper available in a version-controlled repository (Campitelli, Díaz, and Vera 2022b).
+It is also freely available from their respective sources:
 
 \begin{itemize}
+\item
+  ERA5 data can be obtained via the Copernicus Climate Data Store (\url{https://cds.climate.copernicus.eu/cdsapp/\#!/dataset/reanalysis-era5-pressure-levels-monthly-means/}).
+\item
+  ERSSTv5 can be obtained via NOAA's NCEI websiste at \url{https://www.ncei.noaa.gov/access/metadata/landing-page/bin/iso?id=gov.noaa.ncdc:C00927}
+\item
+  CMAP Precipitation data provided by the NOAA/OAR/ESRL PSL, Boulder, Colorado, USA, from their Web site at \url{https://psl.noaa.gov/data/gridded/data.cmap.html}.
+\item
+  The Oceanic Niño Index is available via NOAA's Climate Prediction Center: \url{https://www.cpc.ncep.noaa.gov/products/analysis_monitoring/ensostuff/detrend.nino34.ascii.txt}.
 \item
@@
 \end{itemize}
+
+\hypertarget{code-availability}{%
+\subsection*{Code availability}\label{code-availability}}
+\addcontentsline{toc}{subsection}{Code availability}
+
+A version-controlled repository of the code used to create this analysis, including the code used to download the data can be found at \url{https://github.com/eliocamp/shceof}.
 
@@
 Díaz, and Carolina Vera.
-2022.
+2022a.
 {``Assessment of Zonally Symmetric and Asymmetric Components of the {Southern Annular Mode} Using a Novel Approach.''} \emph{Climate Dynamics} 58 (1): 161--78.
 \url{https://doi.org/10.1007/s00382-021-05896-5}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-campitelli2022a}{}}%
+Campitelli, Elio, Leandro Díaz, and Carolina Vera.
+2022b.
+{``Data for "{Revisiting} the {Austral Spring Extratropical Southern Hemisphere} Zonally Asymmetric Circulation Using Complex {Empirical Orthogonal Functions}".''} {Zenodo}.
+\url{https://doi.org/10.5281/zenodo.6612324}.
 
@@
 2019.
-{``{Era5} Monthly Averaged Data on Pressure Levels from 1979 to Present.''} \emph{{Copernicus Climate Change Service} ({C3s}) {Climate Data Store} ({CDS})} (Accessed on {\(<\)}07-09-2021{\(>\)}).
+{``{Era5} Monthly Averaged Data on Pressure Levels from 1979 to Present.''} \emph{{Copernicus Climate Change Service} ({C3s}) {Climate Data Store} ({CDS})} (Accessed on {\(<\)}03-06-2022{\(>\)}).
 \url{https://doi.org/10.24381/cds.6860a573}.

### 2022-06-03 analysis/paper/shceof.Rmd
Arregla leer ERSST

@@
 sst <- ERSST() %>%
-  readRDS() %>% 
+  fread() %>% 
   .[season(time) == "SON"] %>%

### 2022-06-03 analysis/paper/shceof.tex
Renderiza

@@
 2019.
-{``{Era5} Monthly Averaged Data on Pressure Levels from 1979 to Present.''} \emph{{Copernicus Climate Change Service} ({C3s}) {Climate Data Store} ({CDS})} (Accessed on {\(<\)}03-06-2022{\(>\)}).
+{``{Era5} Monthly Averaged Data on Pressure Levels from 1979 to Present.''} \emph{{Copernicus Climate Change Service} ({C3s}) {Climate Data Store} ({CDS})} (Accessed on {\(<\)}07-09-2021{\(>\)}).
 \url{https://doi.org/10.24381/cds.6860a573}.

### 2022-06-07 analysis/paper/shceof.Rmd
Pasa todo al nuevo template

@@
 ---
+classoptions: 
+  - sn-basic
+year: 2022
 title:  "Revisiting the Austral Spring Extratropical Southern Hemisphere zonally asymmetric circulation using complex Empirical Orthogonal Functions"
-titlerunning:  SH zonally asymmetric circulation
+titlerunning: SH zonally asymmetric circulation with cEOF
+
 authors: 
-- name: Elio Campitelli
-  address: |
-    Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos. Buenos Aires, Argentina.  
-    CONICET – Universidad de Buenos Aires. Centro de Investigaciones del Mar y la Atmósfera (CIMA). Buenos Aires, Argentina.  
-    CNRS – IRD – CONICET – UBA. Instituto Franco-Argentino para el Estudio del Clima y sus Impactos (IRL 3351 IFAECI). Buenos Aires, Argentina.
+  - firstname: Elio
+    lastname: Campitelli
     email: elio.campitelli@cima.fcen.uba.ar
-
-- name:  Leandro B. Díaz
-  address: |
-    Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos. Buenos Aires, Argentina.
-    CONICET – Universidad de Buenos Aires. Centro de Investigaciones del Mar y la Atmósfera (CIMA). Buenos Aires, Argentina.
-    CNRS – IRD – CONICET – UBA. Instituto Franco-Argentino para el Estudio del Clima y sus Impactos (IRL 3351 IFAECI). Buenos Aires, Argentina.
-
-- name:  Carolina Vera
-  address: |
-    Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos. Buenos Aires, Argentina.
-    CONICET – Universidad de Buenos Aires. Centro de Investigaciones del Mar y la Atmósfera (CIMA). Buenos Aires, Argentina.
-    CNRS – IRD – CONICET – UBA. Instituto Franco-Argentino para el Estudio del Clima y sus Impactos (IRL 3351 IFAECI). Buenos Aires, Argentina.
-
-thanks: | 
-  The research was supported by UBACyT20020170100428BA and the CLIMAX Project funded by Belmont Forum/ANR-15-JCL/-0002-01. Elio Campitelli was supported by a PhD grant from CONICET, Argentina.
+    affiliation: [1,2,3]
+    corresponding: TRUE
+  - firstname: Leandro B.
+    lastname: Díaz
+    affiliation: [1,2,3]
+  - firstname: Carolina
+    lastname: Vera
+    affiliation: [1,2,3]
+
+affiliations:
+  - number: 1
+    corresponding: TRUE
+    info:
+      orgname: Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos
+    address:
+        city: Buenos Aires
+        country: Argentina
+  - number: 2
+    corresponding: TRUE
+    info:
+      orgname: CONICET – Universidad de Buenos Aires. Centro de Investigaciones del Mar y la Atmósfera (CIMA)
+    address:
+        city: Buenos Aires
+        country: Argentina  
+  - number: 3
+    corresponding: TRUE
+    info:
+      orgname: CNRS – IRD – CONICET – UBA. Instituto Franco-Argentino para el Estudio del Clima y sus Impactos (IRL 3351 IFAECI)
+    address:
+        city: Buenos Aires
+        country: Argentina
 
@@
 
-date: "`r format(Sys.time(), '%d %B, %Y')`"
-
 bibliography: 
@@
 
-bibstyle: spbasic
-# bibstyle options spbasic(default), spphys, spmpsci
+# bibstyle: spbasic
+
 output: 
-  # bookdown::word_document2: default    
-  bookdown::pdf_book: 
-      base_format: rticles::springer_article
+  bookdown::pdf_document2:
       latex_engine: "xelatex"
       keep_tex: true
-      # citation_package: "natbib"
-      pandoc_args: ["--lua-filter=abstract-to-meta.lua"]
+      citation_package: "natbib"
+      template: "template.tex"
+      pandoc_args: ["--lua-filter=abstract-to-meta.lua", "--wrap=preserve"]
 always_allow_html: true
@@
  - \usepackage{chngcntr}
- - \usepackage{natbib}
- - \usepackage{lineno}
- - \linenumbers 
+ # - \usepackage{natbib}
+ # - \usepackage{lineno}
+ # - \linenumbers 
 ---
@@
 
-```{r extended-series, fig.cap = "(ref:extended-series-cap)", fig.height=3, fig.width = 6, fig.env = "figure*"}
+```{r extended-series, fig.cap = "(ref:extended-series-cap)", fig.height=3, fig.width = 5, fig.env = "figure*"}
 # Check that the extended index coincides with the
@@
               show.legend = FALSE, size = 0.8) +
-  scale_x_date(NULL, date_labels = "%Y",
+  scale_x_date(NULL, 
                breaks = seq(as.Date("1950-01-01"),
                             as.Date("2020-01-01"), by = "10 years"),
+               labels = function(x) JumpBy(format(x, "%Y"), 2, fill = " "),
                minor_breaks = seq(as.Date("1950-01-01"),
@@
              default_crs = "+proj=longlat +lon_wrap=180 +over") + 
-    theme(legend.box = "horizontal",
+    theme(legend.box = "vertical",
           legend.text = element_text(size = rel(0.7)), 
+          legend.spacing = grid::unit(0, "lines"),
           legend.box.spacing = grid::unit(0, "lines") ) +
@@
 
-```{r  sst-psi-2,  fig.cap = "(ref:sst-psi-2-cap)", fig.width = 7, fig.height=5.6, dependson="plot_sst_psi", fig.env="figure*"}
+```{r sst-psi-2,  fig.cap = "(ref:sst-psi-2-cap)", fig.width = 5, fig.height=6, dependson="plot_sst_psi", fig.env="figure*"}
 plot_sst_psi(sst_regr[cEOF == "cEOF2"], psi_regr[cEOF == "cEOF2"]) 
@@
 
-```{r pp-t2m-r2, fig.cap = "(ref:cap-pp-t2m-r2)", fig.width=3.3*2, fig.height=3.3}
+```{r pp-t2m-r2, fig.cap = "(ref:cap-pp-t2m-r2)", fig.width=5, fig.height=3}
 breaks <- seq(.2, 1, by = .1)
@@
     theme(panel.spacing = grid::unit(1, "lines")) +
-    theme(legend.box = "horizontal",
+    theme(legend.box = "vertical",
+          legend.spacing = grid::unit(0, "lines"),
           legend.text = element_text(size = rel(0.7)), 
@@
 
-```{r pp-temp-2, fig.cap="(ref:pp-temp-cap)", fig.width=6.6, fig.height=5.5, dependson="plot_pp"}
+```{r pp-temp-2, fig.cap="(ref:pp-temp-cap)", fig.width=5, fig.height=6, dependson="plot_pp"}
 plot_pp(pp_regr[cEOF == "cEOF2"], t2m_regr[cEOF == "cEOF2"], hgt850_regr[cEOF == "cEOF2"])
@@
 
+\backmatter
+
+\bmhead{Acknowledgments}
+
+The research was supported by UBACyT20020170100428BA and the CLIMAX Project funded by Belmont Forum/ANR-15-JCL/-0002-01. Elio Campitelli was supported by a PhD grant from CONICET, Argentina.
+
 \appendix
@@
 
-```{r sst-psi-1, fig.cap = "(ref:sst-psi-1-cap)", fig.width = 7, fig.height=5.6, dependson="plot_sst_psi", fig.env = "figure*"}
+```{r sst-psi-1, fig.cap = "(ref:sst-psi-1-cap)", fig.width = 5, fig.height=6, dependson="plot_sst_psi", fig.env = "figure*"}
 plot_sst_psi(sst_regr[cEOF == "cEOF1"], psi_regr[cEOF == "cEOF1"]) 

### 2022-06-07 analysis/paper/shceof.tex
Render

@@
-% !TeX program = pdfLaTeX
-\documentclass[smallextended]{svjour3}       % onecolumn (second format)
-%\documentclass[twocolumn]{svjour3}          % twocolumn
-%
-\smartqed  % flush right qed marks, e.g. at end of proof
-%
-\usepackage{amsmath}
-\usepackage{graphicx}
-\usepackage[utf8]{inputenc}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                                                 %%
+%% Please do not use \input{...} to include other tex files.       %%
+%% Submit your LaTeX manuscript as one .tex document.              %%
+%%                                                                 %%
+%% All additional figures and files should be attached             %%
+%% separately and not embedded in the \TeX\ document itself.       %%
+%%                                                                 %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\usepackage[hyphens]{url} % not crucial - just used below for the URL
-\usepackage{hyperref}
+%%\documentclass[referee,sn-basic]{sn-jnl}% referee option is meant for double line spacing
 
-%
-% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
-%
-% insert here the call for the packages your document requires
-%\usepackage{latexsym}
-% etc.
-%
-% please place your own definitions here and don't use \def but
-% \newcommand{}{}
-%
-% Insert the name of "your journal" with
-% \journalname{myjournal}
-%
+%%=======================================================%%
+%% to print line numbers in the margin use lineno option %%
+%%=======================================================%%
+
+%%\documentclass[lineno,sn-basic]{sn-jnl}% Basic Springer Nature Reference Style/Chemistry Reference Style
+
+%%======================================================%%
+%% to compile with pdflatex/xelatex use pdflatex option %%
+%%======================================================%%
+
+%%\documentclass[pdflatex,sn-basic]{sn-jnl}% Basic Springer Nature Reference Style/Chemistry Reference Style
+
+\documentclass[pdflatex,sn-basic]{sn-jnl}
+%%\documentclass[sn-basic]{sn-jnl}% Basic Springer Nature Reference Style/Chemistry Reference Style
+%%\documentclass[pdflatex,sn-mathphys]{sn-jnl}% Math and Physical Sciences Reference Style
+%%\documentclass[sn-aps]{sn-jnl}% American Physical Society (APS) Reference Style
+%%\documentclass[sn-vancouver]{sn-jnl}% Vancouver Reference Style
+%%\documentclass[sn-apa]{sn-jnl}% APA Reference Style
+%%\documentclass[sn-chicago]{sn-jnl}% Chicago-based Humanities Reference Style
+%%\documentclass[sn-standardnature]{sn-jnl}% Standard Nature Portfolio Reference Style
+%%\documentclass[default]{sn-jnl}% Default
+%%\documentclass[default,iicol]{sn-jnl}% Default with double column layout
+
+%%%% Standard Packages
+%%<additional latex packages if required can be included here>
+%%%%
+
+%%%%%=============================================================================%%%%
+%%%%  Remarks: This template is provided to aid authors with the preparation
+%%%%  of original research articles intended for submission to journals published
+%%%%  by Springer Nature. The guidance has been prepared in partnership with
+%%%%  production teams to conform to Springer Nature technical requirements.
+%%%%  Editorial and presentation requirements differ among journal portfolios and
+%%%%  research disciplines. You may find sections in this template are irrelevant
+%%%%  to your work and are empowered to omit any such section if allowed by the
+%%%%  journal you intend to submit to. The submission guidelines and policies
+%%%%  of the journal take precedence. A detailed User Manual is available in the
+%%%%  template package for technical guidance.
+%%%%%=============================================================================%%%%
+
+\jyear{2022}%
+
+%% as per the requirement new theorem styles can be included as shown below
+\theoremstyle{thmstyleone}%
+\newtheorem{theorem}{Theorem}%  meant for continuous numbers
+%%\newtheorem{theorem}{Theorem}[section]% meant for sectionwise numbers
+%% optional argument [theorem] produces theorem numbering sequence instead of independent numbers for Proposition
+\newtheorem{proposition}[theorem]{Proposition}%
+%%\newtheorem{proposition}{Proposition}% to get separate numbers for theorem and proposition etc.
+
+\theoremstyle{thmstyletwo}%
+\newtheorem{example}{Example}%
+\newtheorem{remark}{Remark}%
+
+\theoremstyle{thmstylethree}%
+\newtheorem{definition}{Definition}%
+
+\raggedbottom
+%%\unnumbered% uncomment this for unnumbered level heads
 
@@
 
-% Pandoc citation processing
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newlength{\cslentryspacingunit} % times entry-spacing
-\setlength{\cslentryspacingunit}{\parskip}
-% for Pandoc 2.8 to 2.10.1
-\newenvironment{cslreferences}%
-  {}%
-  {\par}
-% For Pandoc 2.11+
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1
-  \let\oldpar\par
-  \def\par{\hangindent=\cslhangindent\oldpar}
-  \fi
-  % set entry spacing
-  \setlength{\parskip}{#2\cslentryspacingunit}
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
-
 \usepackage{gensymb}
@@
 \usepackage{chngcntr}
-\usepackage{natbib}
-\usepackage{lineno}
-\linenumbers
 \usepackage{booktabs}
@@
 \usepackage{xcolor}
+
 \begin{document}
 
-\title{Revisiting the Austral Spring Extratropical Southern Hemisphere zonally asymmetric circulation using complex Empirical Orthogonal Functions \thanks{The research was supported by UBACyT20020170100428BA and the CLIMAX Project funded by Belmont Forum/ANR-15-JCL/-0002-01.
-Elio Campitelli was supported by a PhD grant from CONICET, Argentina.} }
+\title[SH zonally asymmetric circulation with cEOF]{Revisiting the Austral Spring Extratropical Southern Hemisphere zonally asymmetric circulation using complex Empirical Orthogonal Functions}
 
-    \titlerunning{SH zonally asymmetric circulation}
+%%=============================================================%%
+%% Prefix	-> \pfx{Dr}
+%% GivenName	-> \fnm{Joergen W.}
+%% Particle	-> \spfx{van der} -> surname prefix
+%% FamilyName	-> \sur{Ploeg}
+%% Suffix	-> \sfx{IV}
+%% NatureName	-> \tanm{Poet Laureate} -> Title after name
+%% Degrees	-> \dgr{MSc, PhD}
+%% \author*[1,2]{\pfx{Dr} \fnm{Joergen W.} \spfx{van der} \sur{Ploeg} \sfx{IV} \tanm{Poet Laureate}
+%%                 \dgr{MSc, PhD}}\email{iauthor@gmail.com}
+%%=============================================================%%
 
-\author{  Elio Campitelli \and  Leandro B. Díaz \and  Carolina Vera \and  }
+\author*[1,2,3]{\fnm{Elio} \sur{Campitelli} }\email{\href{mailto:elio.campitelli@cima.fcen.uba.ar}{\nolinkurl{elio.campitelli@cima.fcen.uba.ar}}}
 
-\institute{
-        Elio Campitelli \at
-     Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos. Buenos Aires, Argentina.\\
-CONICET -- Universidad de Buenos Aires. Centro de Investigaciones del Mar y la Atmósfera (CIMA). Buenos Aires, Argentina.\\
-CNRS -- IRD -- CONICET -- UBA. Instituto Franco-Argentino para el Estudio del Clima y sus Impactos (IRL 3351 IFAECI). Buenos Aires, Argentina. \\
-     \email{\href{mailto:elio.campitelli@cima.fcen.uba.ar}{\nolinkurl{elio.campitelli@cima.fcen.uba.ar}}}  %  \\
-%             \emph{Present address:} of F. Author  %  if needed
-    \and
-        Leandro B. Díaz \at
-     Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos. Buenos Aires, Argentina.
-CONICET -- Universidad de Buenos Aires. Centro de Investigaciones del Mar y la Atmósfera (CIMA). Buenos Aires, Argentina.
-CNRS -- IRD -- CONICET -- UBA. Instituto Franco-Argentino para el Estudio del Clima y sus Impactos (IRL 3351 IFAECI). Buenos Aires, Argentina. \\
-     %  \\
-%             \emph{Present address:} of F. Author  %  if needed
-    \and
-        Carolina Vera \at
-     Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos. Buenos Aires, Argentina.
-CONICET -- Universidad de Buenos Aires. Centro de Investigaciones del Mar y la Atmósfera (CIMA). Buenos Aires, Argentina.
-CNRS -- IRD -- CONICET -- UBA. Instituto Franco-Argentino para el Estudio del Clima y sus Impactos (IRL 3351 IFAECI). Buenos Aires, Argentina. \\
-     %  \\
-%             \emph{Present address:} of F. Author  %  if needed
-    \and
-    }
+\author[1,2,3]{\fnm{Leandro B.} \sur{Díaz} }
 
-\date{Received: date / Accepted: date}
-% The correct dates will be entered by the editor
+\author[1,2,3]{\fnm{Carolina} \sur{Vera} }
 
-\maketitle
+  \affil*[1]{\orgname{Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos}, \orgaddress{\city{Buenos Aires}, \country{Argentina}}}
+  \affil*[2]{\orgname{CONICET -- Universidad de Buenos Aires. Centro de Investigaciones del Mar y la Atmósfera (CIMA)}, \orgaddress{\city{Buenos Aires}, \country{Argentina}}}
+  \affil*[3]{\orgname{CNRS -- IRD -- CONICET -- UBA. Instituto Franco-Argentino para el Estudio del Clima y sus Impactos (IRL 3351 IFAECI)}, \orgaddress{\city{Buenos Aires}, \country{Argentina}}}
 
-\begin{abstract}
-The large-scale extratropical circulation in the Southern Hemisphere is strongly zonally symmetric, but its zonal departures, albeit highly relevant for regional impacts, have been less studied.
+\abstract{The large-scale extratropical circulation in the Southern Hemisphere is strongly zonally symmetric, but its zonal departures, albeit highly relevant for regional impacts, have been less studied.
 In this study we analyse the joint variability of the zonally asymmetric springtime stratospheric and tropospheric circulation using Complex Empirical Orthogonal Functions (cEOF) to characterise planetary waves of varying amplitude and phase.
@@
 Furthermore, we find indications that the location of Pacific Sea Surface Temperature anomalies affect the phase of the cEOF2.
-As a result, the methodology proposed in this study provides a deeper understanding of the zonally asymmetric springtime extratropical SH circulation.
-\\
-\keywords{
-        Southern Hemisphere circulation \and
-        Teleconnections \and
-        Pacific South American Mode \and
-        Southern Annular Mode \and
-        Stratosphere \and
-    }
+As a result, the methodology proposed in this study provides a deeper understanding of the zonally asymmetric springtime extratropical SH circulation.}
 
-\end{abstract}
+\keywords{Southern Hemisphere circulation, Teleconnections, Pacific South American Mode, Southern Annular Mode, Stratosphere}
 
-\def\spacingset#1{\renewcommand{\baselinestretch}%
-{#1}\small\normalsize} \spacingset{1}
+\maketitle
 
@@
 
-The large-scale extratropical circulation in the Southern Hemisphere (SH) is strongly zonally symmetric, but its zonal departures are highly relevant for regional impacts (e.g.
-Hoskins and Hodges 2005).
-They strongly modulate weather systems and regional climate through promoting longitudinally different latitudinal transport of heat, humidity, and momentum (K.
-E.
-Trenberth 1980; M.
-N.
-Raphael 2007) and could even be related to the occurrence of high-impact climate extremes (Pezza, Rashid, and Simmonds 2012).
+The large-scale extratropical circulation in the Southern Hemisphere (SH) is strongly zonally symmetric, but its zonal departures are highly relevant for regional impacts \citep[e.g.][]{hoskins2005}.
+They strongly modulate weather systems and regional climate through promoting longitudinally different latitudinal transport of heat, humidity, and momentum \citep{trenberth1980a, raphael2007} and could even be related to the occurrence of high-impact climate extremes \citep{pezza2012}.
 
-Zonally asymmetric circulation is typically described by the amplitude and phase of zonal waves obtained by Fourier decomposition of geopotential heights or sea-level pressures at each latitude (e.g.
-van Loon and Jenne 1972; K.
-E.
-Trenberth 1980; Turner et al.
-2017).
-This approach suggests that zonal waves 1 and 3 explain almost 99\% of the total variance in the annual mean 500-hPa pattern at 50ºS (van Loon and Jenne 1972).
-K. F. Trenberth and Mo (1985) concluded that wave 3 plays a role in the development of blocking events.
+Zonally asymmetric circulation is typically described by the amplitude and phase of zonal waves obtained by Fourier decomposition of geopotential heights or sea-level pressures at each latitude \citep[e.g.][]{vanloon1972, trenberth1980a, turner2017}.
+This approach suggests that zonal waves 1 and 3 explain almost 99\% of the total variance in the annual mean 500-hPa pattern at 50ºS \citep{vanloon1972}.
+\citet{trenberth1985} concluded that wave 3 plays a role in the development of blocking events.
 In addition, previous works have identified at extratropical and subpolar latitudes, wave-like patterns with dominant wavenumbers 3-4, also exerting distinctive regional impacts.
-M. N. Raphael (2007) shows that variability in the planetary wave 3 projected onto its climatological location is associated with anomalies in the Antarctic sea-ice concentration.
+\citet{raphael2007} shows that variability in the planetary wave 3 projected onto its climatological location is associated with anomalies in the Antarctic sea-ice concentration.
 
@@
 However, this is not valid for meridionally propagating waves or zonal waves with localised amplitudes.
-Addressing this limitation, the Fourier technique can be generalized to integrate all planetary wave amplitude regardless of wave number, by computing the wave envelope (Irving and Simmonds 2015).
+Addressing this limitation, the Fourier technique can be generalized to integrate all planetary wave amplitude regardless of wave number, by computing the wave envelope \citep{irving2015}.
 The latter makes it possible to represent planetary waves with different amplitude at different longitudes, but it removes all information about phase and wave number.
-With this approach, Irving and Simmonds (2015) showed that planetary wave amplitude in general is associated to Antarctic sea-ice concentration and temperature, as well as to precipitation anomalies in regions of significant topography in SH mid-latitudes and Antarctica.
+With this approach, \citet{irving2015} showed that planetary wave amplitude in general is associated to Antarctic sea-ice concentration and temperature, as well as to precipitation anomalies in regions of significant topography in SH mid-latitudes and Antarctica.
 
 Another extensively used approach to characterise the SH tropospheric circulation anomalies, is by computing Empirical Orthogonal Functions (EOF, also known as Principal Component Analysis).
-Within the EOF framework, the Southern Annular Mode (SAM) appears as the leading mode of variability of the SH circulation (Fogt and Marshall 2020).
+Within the EOF framework, the Southern Annular Mode (SAM) appears as the leading mode of variability of the SH circulation \citep{fogt2020}.
 SAM represents a relatively zonally symmetric pattern of alternating low pressures in polar latitude and a ring of high pressures in high latitudes with an embedded wave 3 pattern that is more prominent in the Pacific sector.
-The 2nd and 3rd EOFs, usually known as Pacific--South American Patterns (PSA) 1 and PSA2 patterns, respectively, describe meridionally propagating wave trains that originate in the eastern equatorial Pacific and Australian-Indian Ocean sector, and travel towards the South Atlantic following a great-circle arch along the Antarctic coast (Mo and Paegle 2001).
-These patterns influence precipitation anomalies in South America (Mo and Paegle 2001).
-Although these patterns are usually derived by applying EOF to temporal anomalies, M. Raphael (2003) also applied EOF methods specifically to zonal anomalies.
-Irving and Simmonds (2016) proposed a novel methodology for objectively identifying the PSA pattern using Fourier decomposition.
-More recently Goyal et al. (2022) created an index of amplitude and phase of zonal wave 3-like variability by combining the two leading EOFs of meridional wind anomalies.
+The 2nd and 3rd EOFs, usually known as Pacific--South American Patterns (PSA) 1 and PSA2 patterns, respectively, describe meridionally propagating wave trains that originate in the eastern equatorial Pacific and Australian-Indian Ocean sector, and travel towards the South Atlantic following a great-circle arch along the Antarctic coast \citep{mo2001}.
+These patterns influence precipitation anomalies in South America \citep{mo2001}.
+Although these patterns are usually derived by applying EOF to temporal anomalies, \citet{raphael2003} also applied EOF methods specifically to zonal anomalies.
+\citet{irving2016} proposed a novel methodology for objectively identifying the PSA pattern using Fourier decomposition.
+More recently \citet{goyal2022} created an index of amplitude and phase of zonal wave 3-like variability by combining the two leading EOFs of meridional wind anomalies.
 
 Some of the zonally asymmetric patterns of the SH circulation variability described previously, appear to have experienced secular changes.
-For instance, M. Raphael (2003) found that the amplitude of the zonal wave 1 experienced a large increase and that the zonal wave 3 experienced changes in its annual cycle between 1958 and 1996.
+For instance, \citet{raphael2003} found that the amplitude of the zonal wave 1 experienced a large increase and that the zonal wave 3 experienced changes in its annual cycle between 1958 and 1996.
 However, little is known yet about variability and trends of these patterns.
@@
 A third methodology commonly used to describe circulation anomalies consists on identifying particular features of interest and creating indices using simple methods such as averages and differences.
-Examples of this methodology are the SAM Index of Gong and Wang (1999), the SH wave 3 activity index defined by M.
-N.
-Raphael (2004) and the SH zonally asymmetric circulation index from Hobbs and Raphael (2010).
+Examples of this methodology are the SAM Index of \citet{gong1999}, the SH wave 3 activity index defined by \citet{raphael2004} and the SH zonally asymmetric circulation index from \citet{hobbs2010}.
 These derived methods are grounded on other methods such as Fourier decomposition or EOF to identify the centres of action for the described phenomena and can be useful to characterise features that are not readily apparent with these methods.
@@
 
-An alternative methodology that has been proposed to study travelling and standing waves is complex Empirical Orthogonal Functions {[}cEOF; Horel (1984){]}.
+An alternative methodology that has been proposed to study travelling and standing waves is complex Empirical Orthogonal Functions {[}cEOF; \citet{horel1984}{]}.
 This method extends EOF analysis to capture oscillations with varying amplitude and phase and has been applied to the time domain.
-For instance, Krokhin and Luxemburg (2007) applied cEOF to station-based monthly precipitation anomalies and monthly temperature anomalies in the Eastern Siberia and the Far East region to characterise the main modes of variability and their connection to teleconnection indices.
-Similarly, Gelbrecht, Boers, and Jürgen Kurths (2018) used cEOF applied to daily precipitation from reanalysis to study the propagating characteristics of the South American Monsoon.
+For instance, \citet{krokhin2007} applied cEOF to station-based monthly precipitation anomalies and monthly temperature anomalies in the Eastern Siberia and the Far East region to characterise the main modes of variability and their connection to teleconnection indices.
+Similarly, \citet{gelbrecht2018} used cEOF applied to daily precipitation from reanalysis to study the propagating characteristics of the South American Monsoon.
 To our knowledge, cEOF analysis has not been applied in the spatial domain to capture the phase-varying nature of planetary waves in the atmosphere.
@@
 
-We restrict this paper to the spring September-October-November (SON) trimester because this season experiences a maximum in tropical teleconnections over South America (Cazes-Boezio, Robertson, and Mechoso 2003) and SH stratosphere-troposphere interactions (Lim, Hendon, and Thompson 2018).
+We restrict this paper to the spring September-October-November (SON) trimester because this season experiences a maximum in tropical teleconnections over South America \citep{cazes-boezio2003} and SH stratosphere-troposphere interactions \citep{lim2018}.
 
@@
 
-We used monthly geopotential height, air temperature, ozone mixing ratio, and total ozone column (TOC) at 2.5\degree longitude by 2.5\degree latitude of horizontal resolution and 37 vertical isobaric levels from the European Centre for Medium-Range Weather Forecasts Reanalysis version 5 {[}ERA; Hersbach et al.
-(2019){]} for the period 1979 -- 2019.
-Most of our analysis is restricted to the post-satellite era to avoid any confounding factors arising from the incorporation of satellite observations, but we also used the preliminary back extension of ERA5 from 1950 to 1978 (Bell et al.
-2020) to describe long-term trends.
-We derived streamfunction at 200 hPa from ERA5 vorticity using the FORTRAN subroutine FISHPACK (Adams, Swartztrauber, and Sweet 1999) and we computed horizontal wave activity fluxes following Plumb (1985).
-Sea Surface Temperature (SST) monthly fields are from Extended Reconstructed Sea Surface Temperature (ERSST) v5 (Huang et al.
-2017) and precipitation monthly data from the CPC Merged Analysis of Precipitation (P.
-Xie and Arkin 1997), with a 2\degree and 2.5\degree horizontal resolution respectively.
+We used monthly geopotential height, air temperature, ozone mixing ratio, and total ozone column (TOC) at 2.5\degree longitude by 2.5\degree latitude of horizontal resolution and 37 vertical isobaric levels from the European Centre for Medium-Range Weather Forecasts Reanalysis version 5 {[}ERA; \citet{era5}{]} for the period 1979 -- 2019.
+Most of our analysis is restricted to the post-satellite era to avoid any confounding factors arising from the incorporation of satellite observations, but we also used the preliminary back extension of ERA5 from 1950 to 1978 \citep{era5be} to describe long-term trends.
+We derived streamfunction at 200 hPa from ERA5 vorticity using the FORTRAN subroutine FISHPACK \citep{fishpack} and we computed horizontal wave activity fluxes following \citet{plumb1985}.
+Sea Surface Temperature (SST) monthly fields are from Extended Reconstructed Sea Surface Temperature (ERSST) v5 \citep{huang2017} and precipitation monthly data from the CPC Merged Analysis of Precipitation \citep{cmap}, with a 2\degree and 2.5\degree horizontal resolution respectively.
 The rainfall gridded dataset is based on information from different sources such as rain gauge observations, satellite inferred estimations and the NCEP-NCAR reanalysis, and it is available since 1979 to the present.
 
-The Oceanic Niño Index (ONI, Bamston, Chelliah, and Goldenberg 1997) comes from NOAA's Climate Prediction Center and the Dipole Mode Index (DMI, Saji and Yamagata 2003) from Global Climate Observing System Working Group on Surface Pressure.
+The Oceanic Niño Index \citep[ONI,][]{bamston1997} comes from NOAA's Climate Prediction Center and the Dipole Mode Index \citep[DMI,][]{saji2003} from Global Climate Observing System Working Group on Surface Pressure.
 
@@
 
-We computed the level-dependent SAM index as the leading EOF of year-round monthly geopotential height anomalies south of 20ºS at each level for the whole period (Baldwin and Thompson 2009).
+We computed the level-dependent SAM index as the leading EOF of year-round monthly geopotential height anomalies south of 20ºS at each level for the whole period \citep{baldwin2009}.
 We further split the SAM into its zonally symmetric and zonally asymmetric components (S-SAM and A-SAM indices respectively).
-These indices were obtained by projecting the zonally asymmetric and zonally symmetric part of the SAM spatial pattern onto monthly geopotential height fields, as proposed by Campitelli, Díaz, and Vera (2022a).
-Seasonal indices of the PSA patterns (PSA1 and PSA2) were calculated, in agreement with Mo and Paegle (2001), as the third and fourth leading EOF of seasonal mean anomalies for 500-hPa geopotential heights at SH.
+These indices were obtained by projecting the zonally asymmetric and zonally symmetric part of the SAM spatial pattern onto monthly geopotential height fields, as proposed by \citet{campitelli2022}.
+Seasonal indices of the PSA patterns (PSA1 and PSA2) were calculated, in agreement with \citet{mo2001}, as the third and fourth leading EOF of seasonal mean anomalies for 500-hPa geopotential heights at SH.
 
-Linear trends were computed by Ordinary Least Squares (OLS) and the 95\% confidence interval was computed assuming a t-distribution with the appropriate residual degrees of freedom (D. Wilks 2011).
+Linear trends were computed by Ordinary Least Squares (OLS) and the 95\% confidence interval was computed assuming a t-distribution with the appropriate residual degrees of freedom \citep{wilks2011}.
 
@@
 
-In the standard EOF analysis, zonal waves may appear as pairs of (possibly degenerated) EOFs representing similar patterns but shifted in phase (Horel 1984).
+In the standard EOF analysis, zonal waves may appear as pairs of (possibly degenerated) EOFs representing similar patterns but shifted in phase \citep{horel1984}.
 Figure \ref{fig:eof-naive} shows the four leading EOFs of SON geopotential height zonal anomalies at 50 hPa south of 20\degree S.
@@
 
-Complex Empirical Orthogonal Functions (cEOF) are a useful method for characterising zonal waves, associated with phase-varying structures (Horel 1984).
+Complex Empirical Orthogonal Functions (cEOF) are a useful method for characterising zonal waves, associated with phase-varying structures \citep{horel1984}.
 This method involves computing EOF on the analytic representation of the original field.
@@
 For real EOFs, this constant can be either 0 or \(\pi\), corresponding to a change in sign.
-For cEOFs, it can be any real number between 0 and \(2\pi\) (Horel 1984), corresponding to rotations in the complex plane.
+For cEOFs, it can be any real number between 0 and \(2\pi\) \citep{horel1984}, corresponding to rotations in the complex plane.
 Since any choice is arbitrary and equally valid, we chose the phase of each cEOF so that the real or imaginary parts are aligned with meaningful phases in our analysis.
@@
 Therefore, to aid in the interpretation we chose the phase of cEOF1 so that the time series corresponding to the real part has the maximum correlation with the zonal wave 1 of Total Ozone Column between 75°S and 45°S.
-Similarly, we chose the phase of cEOF2 so that the coefficient of determination between the Oceanic Niño Index (Bamston, Chelliah, and Goldenberg 1997) and the real part is minimised, which also nearly maximises the correlation with the imaginary part.
+Similarly, we chose the phase of cEOF2 so that the coefficient of determination between the Oceanic Niño Index \citep{bamston1997} and the real part is minimised, which also nearly maximises the correlation with the imaginary part.
 
@@
 
-We evaluated statistical significance using a two-sided t-test and, in the case of regression maps, p-values were adjusted by controlling for the False Discovery Rate (Benjamini and Hochberg 1995; D.
-S.
-Wilks 2016) to avoid misleading results from the high number of regressions (Walker 1914; Katz and Brown 1991).
+We evaluated statistical significance using a two-sided t-test and, in the case of regression maps, p-values were adjusted by controlling for the False Discovery Rate \citep{benjamini1995, wilks2016} to avoid misleading results from the high number of regressions \citep{walker1914, katz1991}.
 
@@
 
-We performed all analysis in this paper using the R programming language (R Core Team 2020), using data.table (Dowle and Srinivasan 2020) and metR (Campitelli 2020) packages.
-All graphics are made using ggplot2 (Wickham 2009).
-We downloaded data from reanalysis using the ecmwfr package (Hufkens 2020) and indices of ENSO and Indian Ocean Dipole (IOD) with the rsoi package (Albers and Campitelli 2020).
-The paper was rendered using knitr and rmarkdown (Y. Xie 2015; Allaire et al. 2020).
+We performed all analysis in this paper using the R programming language \citep{rcoreteam2020}, using data.table \citep{dowle2020} and metR \citep{campitelli2020} packages.
+All graphics are made using ggplot2 \citep{wickham2009}.
+We downloaded data from reanalysis using the ecmwfr package \citep{hufkens2020} and indices of ENSO and Indian Ocean Dipole (IOD) with the rsoi package \citep{albers2020}.
+The paper was rendered using knitr and rmarkdown \citep{xie2015, allaire2020}.
 
@@
 This long-term change indicates an increase in the magnitude of the high latitude zonal wave 1.
-In agreement, M.
-Raphael (2003) detected a step change after around 1975 in the temporal evolution of leading EOF computed from August-September-October 500 hPa zonal geopotential height anomalies, which is similar to the cEOF1.
+In agreement, \citet{raphael2003} detected a step change after around 1975 in the temporal evolution of leading EOF computed from August-September-October 500 hPa zonal geopotential height anomalies, which is similar to the cEOF1.
 
@@
 At 200 hPa (Figure~\ref{fig:eof1-regr-gh} row b) the Real cEOF1 shows a single centre of positive anomalies spanning West Antarctica surrounded by opposite anomalies in lower latitudes, with is centre shifted slightly eastward compared with the upper-level anomalies.
-The Imaginary cEOF1 shows a much more zonally symmetrical pattern resembling the negative SAM phase (e.g. Fogt and Marshall 2020).
+The Imaginary cEOF1 shows a much more zonally symmetrical pattern resembling the negative SAM phase \citep[e.g.][]{fogt2020}.
 Therefore, the magnitude and phase of the cEOF1 are associated with the magnitude and phase of a zonal wave mainly in the stratosphere.
@@
 All fields have a dominant zonal wave 3 limited to the western hemisphere, over the Pacific and Atlantic Oceans.
-cEOF2 then represents an equivalent barotropic wave train that is very similar to the the PSA Patterns (Mo and Paegle 2001).
-Comparing the location of the positive anomaly near 90ºW in column b of Figure~\ref{fig:eof2-regr-gh} with Figures 1.a and b from Mo and Paegle (2001), the Real cEOF2 regression map could be identified with PSA2, while the Imaginary cEOF2 resembles PSA1.
+cEOF2 then represents an equivalent barotropic wave train that is very similar to the the PSA Patterns \citep{mo2001}.
+Comparing the location of the positive anomaly near 90ºW in column b of Figure~\ref{fig:eof2-regr-gh} with Figures 1.a and b from \citet{mo2001}, the Real cEOF2 regression map could be identified with PSA2, while the Imaginary cEOF2 resembles PSA1.
 
@@
 Therefore, the ozone zonal wave 1 is anticorrelated with the temperature zonal wave 1 in the upper stratosphere, and directly correlated in the upper stratosphere.
-This change in phase is observed in ozone anomalies forced by planetary waves that reach the stratosphere (Hartmann and Garcia 1979; Wirth 1993; Smith 1995).
+This change in phase is observed in ozone anomalies forced by planetary waves that reach the stratosphere \citep{hartmann1979, wirth1993, smith1995}.
 In the photochemically-dominated upper stratosphere, cold temperatures inhibit the destruction of ozone, explaining the opposite behaviour for both variables.
-On the other hand, in the advectively-dominated lower stratosphere, ozone anomalies are 90º out of phase with horizontal and vertical transport, which are in addition 90º out of phase with temperature anomalies, resulting in same sign anomalies for the response of both variables (Hartmann and Garcia 1979; Wirth 1993; Smith 1995).
+On the other hand, in the advectively-dominated lower stratosphere, ozone anomalies are 90º out of phase with horizontal and vertical transport, which are in addition 90º out of phase with temperature anomalies, resulting in same sign anomalies for the response of both variables \citep{hartmann1979, wirth1993, smith1995}.
 
@@
 The regression maps of TOC anomalies upon cEOF1 (Fig.~\ref{fig:o3-regr}) show zonal wave 1 patterns associated with both components of cEOF1.
-The climatological position of the springtime Ozone minimum (ozone hole) is outside the South Pole and towards the Weddell Sea (e.g. Grytsai 2011).
+The climatological position of the springtime Ozone minimum (ozone hole) is outside the South Pole and towards the Weddell Sea \citep[e.g.][]{grytsai2011}.
 Thus, the Real cEOF1 regression field (Figure~\ref{fig:o3-regr}a) coincides with the climatological position of the ozone hole, while it is 90° out of phase for the Imaginary cEOF1.
@@
 
-\caption{\label{tab:psa-eof2}Correlation coefficients (r) between cEOF2 components and the PSA1 and PSA2 modes computed as Mo and Paegle (2001) for the 1979 -- 2019 period.
+\caption{\label{tab:psa-eof2}Correlation coefficients (r) between cEOF2 components and the PSA1 and PSA2 modes computed as \citet{mo2001} for the 1979 -- 2019 period.
 95\% confidence intervals in parenthesis.
@@
 Conventional EOF analysis will therefore tend to separate them relatively cleanly, with the EOF representing the Imaginary cEOF2 always leading the one representing the Real cEOF2.
-This phase preferences is in agreement with Irving and Simmonds (2016), who found a bimodal distribution to PSA-like variability (compare our Figure~\ref{fig:phase-histogram} with their Figure 6).
+This phase preferences is in agreement with \citet{irving2016}, who found a bimodal distribution to PSA-like variability (compare our Figure~\ref{fig:phase-histogram} with their Figure 6).
 
@@
 \caption{\label{fig:sam-eof-vertical}Coefficient of determination (\(r^2\)) between each component of cEOFs and the SAM, Asymmetric SAM (A-SAM) and Symmetric SAM (S-SAM) indices computed at each level for the 1979 -- 2019 period.
-Thick lines represent estimates with p-value \textless{} 0.01 corrected for False Detection Rate (Benjamini and Hochberg 1995).}
+Thick lines represent estimates with p-value \textless{} 0.01 corrected for False Detection Rate \citep{benjamini1995}.}
 \end{figure}
@@
 We now explore the relationship between SAM and the cEOFs motivated by the resemblance between cEOFs regression maps and SAM patterns shown in Section~\ref{regressions}.
-We then computed the coefficient of determination between the cEOFs time series and the three SAM indices (SAM, A-SAM and S-SAM) defined by Campitelli, Díaz, and Vera (2022a) at each vertical level (Fig.~\ref{fig:sam-eof-vertical}).
+We then computed the coefficient of determination between the cEOFs time series and the three SAM indices (SAM, A-SAM and S-SAM) defined by \citet{campitelli2022} at each vertical level (Fig.~\ref{fig:sam-eof-vertical}).
 The SAM index is statistically significant correlated with the Real cEOF1 in all levels, and with the Imaginary cEOF1 and Imaginary cEOF2 in the troposphere.
@@
 In the stratosphere, the Real cEOF1 is correlated with both A-SAM and S-SAM, while the Imaginary cEOF1 is highly correlated only with the A-SAM.
-These correlations are consistent with the regression maps of geopotential height in Figure~\ref{fig:eof1-regr-gh} and their comparison with those obtained for SAM, A-SAM and S-SAM by Campitelli, Díaz, and Vera (2022a).
+These correlations are consistent with the regression maps of geopotential height in Figure~\ref{fig:eof1-regr-gh} and their comparison with those obtained for SAM, A-SAM and S-SAM by \citet{campitelli2022}.
 
@@
 Indeed, the Imaginary cEOF2 shares up to 92\% variance with the A-SAM and only 12\% at most with the S-SAM (Figure~\ref{fig:sam-eof-vertical}.b2).
-Such extremely high correlation between A-SAM and Imaginary cEOF2 suggests that the modes obtained in this work are able to characterise the zonally asymmetric component of the SAM described previously by Campitelli, Díaz, and Vera (2022a).
+Such extremely high correlation between A-SAM and Imaginary cEOF2 suggests that the modes obtained in this work are able to characterise the zonally asymmetric component of the SAM described previously by \citet{campitelli2022}.
 
@@
 The Imaginary cEOF2 (second row) is associated with strong positive SST anomalies on the Central Pacific and negative anomalies over an area across northern Australia, New Zealand the South Pacific Convergence Zone (SPCZ) (Figure~\ref{fig:sst-psi-2}.b1).
-The regression field of SST anomalies bears a strong resemblance with canonically positive ENSO (Bamston, Chelliah, and Goldenberg 1997).
+The regression field of SST anomalies bears a strong resemblance with canonically positive ENSO \citep{bamston1997}.
 Indeed, there is a significant and very high correlation (0.76 (CI: 0.6 -- 0.87)) between the ONI and the Imaginary cEOF2 time series.
-Besides the Pacific ENSO-like pattern, there are positive anomalies in the western Indian Ocean and negative values in the eastern Indian Ocean, resembling a positive IOD (Saji et al. 1999).
+Besides the Pacific ENSO-like pattern, there are positive anomalies in the western Indian Ocean and negative values in the eastern Indian Ocean, resembling a positive IOD \citep{saji1999}.
 Consistently, the correlation between the Imaginary cEOF2 and the DMI is 0.62 (CI: 0.38 -- 0.78).
@@
 The Imaginary cEOF2 is associated with strong wave-like streamfunction anomalies emanating from the tropics (Figure~\ref{fig:sst-psi-2}.b2), both from the Central Pacific sector and the Indian Ocean.
-The atmospheric response associated with Imaginary cEOF2 is then consistent with the combined effect of ENSO and the IOD on the extratropics: with SST anomalies inducing anomalous tropical convection that in turn excite Rossby waves propagating meridionally towards higher latitudes (Mo 2000; Cai et al.
-2011; Nuncio and Yuan 2015).
+The atmospheric response associated with Imaginary cEOF2 is then consistent with the combined effect of ENSO and the IOD on the extratropics: with SST anomalies inducing anomalous tropical convection that in turn excite Rossby waves propagating meridionally towards higher latitudes \citep{mo2000, cai2011, nuncio2015}.
 
@@
 Meanwhile, Rows a and c in Fig.~\ref{fig:sst-psi-2} show that the intermediate phases are still associated with significant SST regressed anomalies over the Pacific Ocean, but at slightly different locations.
-The 135º phase is associated with SST anomalies in the central Pacific (Fig.~\ref{fig:sst-psi-2}a.1), while the 45º phase is associated with SST anomalies (Fig.~\ref{fig:sst-psi-2}c.1) that correspond roughly to the Central Pacific and Eastern Pacific ``flavours'' of ENSO, respectively (Kao and Yu 2009).
+The 135º phase is associated with SST anomalies in the central Pacific (Fig.~\ref{fig:sst-psi-2}a.1), while the 45º phase is associated with SST anomalies (Fig.~\ref{fig:sst-psi-2}c.1) that correspond roughly to the Central Pacific and Eastern Pacific ``flavours'' of ENSO, respectively \citep{kao2009}.
 Both phases could be also associated to wave trains generated in the region surrounding Australia and propagates toward the extra-tropics.
@@
 
-In the extra-tropics, the positive Imaginary cEOF2 is related to drier conditions over eastern Australia and the surrounding ocean, that it is similar signal as the one associated with ENSO (Cai et al.
-2011).
+In the extra-tropics, the positive Imaginary cEOF2 is related to drier conditions over eastern Australia and the surrounding ocean, that it is similar signal as the one associated with ENSO \citep{cai2011}.
 However, the Imaginary cEOF2 is not the direction most correlated with precipitation in that area.
@@
 Over South America, the Imaginary cEOF2 has positive correlations with precipitation in South Eastern South America (SESA) and central Chile, and negative correlations in eastern Brazil.
-This correlation field matches the springtime precipitation signature of ENSO (e.g.
-Cai et al.
-2020) and it is also similar to the precipitation anomalies associated with the A-SAM (Campitelli, Díaz, and Vera 2022a).
+This correlation field matches the springtime precipitation signature of ENSO \citep[e.g.][]{cai2020a} and it is also similar to the precipitation anomalies associated with the A-SAM \citep{campitelli2022}.
 
@@
 
-Irving and Simmonds (2016) argued that there is some disagreement in the literature of whether the phase of the PSA pattern is affected by the location of tropical SST anomalies.
+\citet{irving2016} argued that there is some disagreement in the literature of whether the phase of the PSA pattern is affected by the location of tropical SST anomalies.
 With the methodology used in this study, we were able to show not only that the cEOF2 tends to be in the positive or negative imaginary phase (\textasciitilde PSA1) when the ENSO region is warm or cold, respectively, but also that central Pacific SST anomalies tend to align the cEOF towards the negative real phase and eastern Pacific SST anomalies tend to align it towards positive real phase.
 When ENSO phase is neutral, the cEOF2 is still as active, but with no preferred phase.
-The latter agrees with the results of Cai and Watterson (2002), who showed that the CSIRO Model can develop PSA-like variability even in the absence of ENSO forcing (i.e.~with a climatological run), but that the variability of one of the PSA modes was enhanced when adding the ENSO signal.
-The sensitivity of the phase of the PSAs to the location of the tropical SST anomalies was also seen by Ciasto, Simpkins, and England (2015), who detected similar Rossby wave patterns associated with central Pacific and eastern Pacific SST anomalies but with a change in phase.
+The latter agrees with the results of \citet{cai2002}, who showed that the CSIRO Model can develop PSA-like variability even in the absence of ENSO forcing (i.e.~with a climatological run), but that the variability of one of the PSA modes was enhanced when adding the ENSO signal.
+The sensitivity of the phase of the PSAs to the location of the tropical SST anomalies was also seen by \citet{ciasto2015}, who detected similar Rossby wave patterns associated with central Pacific and eastern Pacific SST anomalies but with a change in phase.
 
-The method used in this study have similarities to the one used by Goyal et al.
-(2022) as they construct an index of amplitude and phase of zonal wave 3-like variability by combining the two leading EOFs of meridional wind anomalies.
+The method used in this study have similarities to the one used by \citet{goyal2022} as they construct an index of amplitude and phase of zonal wave 3-like variability by combining the two leading EOFs of meridional wind anomalies.
 The patterns obtained by them bear high resemblance with cEOF2.
@@
 
-All data used in this paper available in a version-controlled repository (Campitelli, Díaz, and Vera 2022b).
+All data used in this paper available in a version-controlled repository \citep{campitelli2022a}.
 It is also freely available from their respective sources:
@@
 
-\hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-fishpack}{}}%
-Adams, John C., Paul N.
-Swartztrauber, and Roland Sweet.
-1999.
-{``{FISHPACK}, a Package of {Fortran} Subprograms for the Solution of Separable Elliptic Partial Differential Equations.''} 1999.
-\url{https://www2.cisl.ucar.edu/resources/legacy/fishpack}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-albers2020}{}}%
-Albers, Sam, and Elio Campitelli. 2020. \emph{Rsoi: {Import Various Northern} and {Southern Hemisphere Climate Indices}} (version 0.5.4). \url{https://CRAN.R-project.org/package=rsoi}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-allaire2020}{}}%
-Allaire, J.
-J., Yihui Xie {[}aut, cre, Jonathan McPherson, Javier Luraschi, Kevin Ushey, Aron Atkins, et al.
-2020.
-\emph{Rmarkdown: {Dynamic Documents} for {R}} (version 2.4).
-\url{https://CRAN.R-project.org/package=rmarkdown}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-baldwin2009}{}}%
-Baldwin, Mark P., and David W.
-J.
-Thompson.
-2009.
-{``A Critical Comparison of Stratosphere--Troposphere Coupling Indices.''} \emph{Quarterly Journal of the Royal Meteorological Society} 135 (644): 1661--72.
-\url{https://doi.org/10.1002/qj.479}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-bamston1997}{}}%
-Bamston, Anthony G., Muthuvel Chelliah, and Stanley B.
-Goldenberg.
-1997.
-{``Documentation of a Highly {ENSO}‐related Sst Region in the Equatorial Pacific: {Research} Note.''} \emph{Atmosphere-Ocean} 35 (3): 367--83.
-\url{https://doi.org/10.1080/07055900.1997.9649597}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-era5be}{}}%
-Bell, B., H.
-Hersbach, P.
-Berrisford, P.
-Dahlgren, A.
-Horányi, J.
-Muñoz Sabater, J.
-Nicolas, et al.
-2020.
-{``{Era5} Monthly Averaged Data on Pressure Levels from 1950 to 1978 (Preliminary Version).''} \emph{{Copernicus Climate Change Service} ({C3s}) {Climate Data Store} ({CDS})} (Accessed on {\(<\)}26-08-2021{\(>\)}), https://cds.climate.copernicus.eu/cdsapp\#!/dataset/reanalysis-era5-pressure-levels-monthly-means-preliminary-back-extension?tab=overview.
-
-\leavevmode\vadjust pre{\hypertarget{ref-benjamini1995}{}}%
-Benjamini, Yoav, and Yosef Hochberg.
-1995.
-{``Controlling the {False Discovery Rate}: {A Practical} and {Powerful Approach} to {Multiple Testing}.''} \emph{Journal of the Royal Statistical Society: Series B (Methodological)} 57 (1): 289--300.
-\url{https://doi.org/10.1111/j.2517-6161.1995.tb02031.x}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-cai2020a}{}}%
-Cai, Wenju, Michael J.
-McPhaden, Alice M.
-Grimm, Regina R.
-Rodrigues, Andréa S.
-Taschetto, René D.
-Garreaud, Boris Dewitte, et al.
-2020.
-{``Climate Impacts of the {El Niño}--{Southern Oscillation} on {South America}.''} \emph{Nature Reviews Earth \& Environment} 1 (4, 4): 215--31.
-\url{https://doi.org/10.1038/s43017-020-0040-3}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-cai2011}{}}%
-Cai, Wenju, Peter van Rensch, Tim Cowan, and Harry H.
-Hendon.
-2011.
-{``Teleconnection {Pathways} of {ENSO} and the {IOD} and the {Mechanisms} for {Impacts} on {Australian Rainfall}.''} \emph{Journal of Climate} 24 (15): 3910--23.
-\url{https://doi.org/10.1175/2011JCLI4129.1}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-cai2002}{}}%
-Cai, Wenju, and Ian G.
-Watterson.
-2002.
-{``Modes of {Interannual Variability} of the {Southern Hemisphere Circulation Simulated} by the {CSIRO Climate Model}.''} \emph{Journal of Climate} 15 (10): 1159--74.
-\url{https://doi.org/10.1175/1520-0442(2002)015\%3C1159:MOIVOT\%3E2.0.CO;2}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-campitelli2020}{}}%
-Campitelli, Elio. 2020. \emph{{metR}: {Tools} for {Easier Analysis} of {Meteorological Fields}} (version 0.7.0). \url{https://CRAN.R-project.org/package=metR}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-campitelli2022}{}}%
-Campitelli, Elio, Leandro B.
-Díaz, and Carolina Vera.
-2022a.
-{``Assessment of Zonally Symmetric and Asymmetric Components of the {Southern Annular Mode} Using a Novel Approach.''} \emph{Climate Dynamics} 58 (1): 161--78.
-\url{https://doi.org/10.1007/s00382-021-05896-5}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-campitelli2022a}{}}%
-Campitelli, Elio, Leandro Díaz, and Carolina Vera.
-2022b.
-{``Data for "{Revisiting} the {Austral Spring Extratropical Southern Hemisphere} Zonally Asymmetric Circulation Using Complex {Empirical Orthogonal Functions}".''} {Zenodo}.
-\url{https://doi.org/10.5281/zenodo.6612324}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-cazes-boezio2003}{}}%
-Cazes-Boezio, Gabriel, Andrew W.
-Robertson, and Carlos R.
-Mechoso.
-2003.
-{``Seasonal {Dependence} of {ENSO Teleconnections} over {South America} and {Relationships} with {Precipitation} in {Uruguay}.''} \emph{Journal of Climate} 16 (8): 1159--76.
-\url{https://doi.org/10.1175/1520-0442(2003)16\%3C1159:SDOETO\%3E2.0.CO;2}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-ciasto2015}{}}%
-Ciasto, Laura M., Graham R.
-Simpkins, and Matthew H.
-England.
-2015.
-{``Teleconnections Between {Tropical Pacific SST Anomalies} and {Extratropical Southern Hemisphere Climate}.''} \emph{Journal of Climate} 28 (1): 56--65.
-\url{https://doi.org/10.1175/JCLI-D-14-00438.1}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-dowle2020}{}}%
-Dowle, Matt, and Arun Srinivasan. 2020. \emph{Data.table: {Extension} of 'Data.frame'} (version 1.13.0). \url{https://CRAN.R-project.org/package=data.table}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-fogt2020}{}}%
-Fogt, Ryan L., and Gareth J.
-Marshall.
-2020.
-{``The {Southern Annular Mode}: {Variability}, Trends, and Climate Impacts Across the {Southern Hemisphere}.''} \emph{WIREs Climate Change} 11 (4): e652.
-\url{https://doi.org/10.1002/wcc.652}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-gelbrecht2018}{}}%
-Gelbrecht, Maximilian, Niklas Boers, and Jürgen Kurths.
-2018.
-{``Phase Coherence Between Precipitation in {South America} and {Rossby} Waves.''} \emph{Science Advances} 4 (12): eaau3191.
-\url{https://doi.org/10.1126/sciadv.aau3191}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-gong1999}{}}%
-Gong, Daoyi, and Shaowu Wang. 1999. {``Definition of {Antarctic Oscillation} Index.''} \emph{Geophysical Research Letters} 26 (4): 459--62. \url{https://doi.org/10.1029/1999GL900003}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-goyal2022}{}}%
-Goyal, Rishav, Martin Jucker, Alex Sen Gupta, and Matthew H.
-England.
-2022.
-{``A New Zonal Wave 3 Index for the {Southern Hemisphere}.''} \emph{Journal of Climate} -1 (May): 1--25.
-\url{https://journals.ametsoc.org/view/journals/clim/aop/JCLI-D-21-0927.1/JCLI-D-21-0927.1.xml}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-grytsai2011}{}}%
-Grytsai, A.
-2011.
-{``Planetary Wave Peculiarities in {Antarctic} Ozone Distribution During 1979--2008.''} \emph{International Journal of Remote Sensing} 32 (11): 3139--51.
-\url{https://doi.org/10.1080/01431161.2010.541518}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-hartmann1979}{}}%
-Hartmann, Dennis L., and Rolando R.
-Garcia.
-1979.
-{``A {Mechanistic Model} of {Ozone Transport} by {Planetary Waves} in the {Stratosphere}.''} \emph{Journal of the Atmospheric Sciences} 36 (2): 350--64.
-\url{https://doi.org/10.1175/1520-0469(1979)036\%3C0350:AMMOOT\%3E2.0.CO;2}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-era5}{}}%
-Hersbach, H., B.
-Bell, P.
-Berrisford, G.
-Biavati, A.
-Horányi, J.
-Muñoz Sabater, J.
-Nicolas, et al.
-2019.
-{``{Era5} Monthly Averaged Data on Pressure Levels from 1979 to Present.''} \emph{{Copernicus Climate Change Service} ({C3s}) {Climate Data Store} ({CDS})} (Accessed on {\(<\)}07-09-2021{\(>\)}).
-\url{https://doi.org/10.24381/cds.6860a573}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-hobbs2010}{}}%
-Hobbs, William R., and Marilyn N.
-Raphael.
-2010.
-{``Characterizing the Zonally Asymmetric Component of the {SH} Circulation.''} \emph{Climate Dynamics} 35 (5): 859--73.
-\url{https://doi.org/10.1007/s00382-009-0663-z}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-horel1984}{}}%
-Horel, J.
-D.
-1984.
-{``Complex {Principal Component Analysis}: {Theory} and {Examples}.''} \emph{Journal of Applied Meteorology and Climatology} 23 (12): 1660--73.
-\url{https://doi.org/10.1175/1520-0450(1984)023\%3C1660:CPCATA\%3E2.0.CO;2}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-hoskins2005}{}}%
-Hoskins, B. J., and K. I. Hodges. 2005. {``A {New Perspective} on {Southern Hemisphere Storm Tracks}.''} \emph{Journal of Climate} 18 (20): 4108--29. \url{https://doi.org/10.1175/JCLI3570.1}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-huang2017}{}}%
-Huang, Boyin, Peter W.
-Thorne, Viva F.
-Banzon, Tim Boyer, Gennady Chepurin, Jay H.
-Lawrimore, Matthew J.
-Menne, Thomas M.
-Smith, Russell S.
-Vose, and Huai-Min Zhang.
-2017.
-{``Extended {Reconstructed Sea Surface Temperature}, {Version} 5 ({ERSSTv5}): {Upgrades}, {Validations}, and {Intercomparisons}.''} \emph{Journal of Climate} 30 (20): 8179--8205.
-\url{https://doi.org/10.1175/JCLI-D-16-0836.1}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-hufkens2020}{}}%
-Hufkens, Koen.
-2020.
-\emph{Ecmwfr: {Programmatic} Interface to the Two {European Centre} for {Medium-Range Weather Forecasts API} Services} (version 1.3.0).
-\url{http://doi.org/10.5281/zenodo.2647541}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-irving2015}{}}%
-Irving, Damien, and Ian Simmonds.
-2015.
-{``A {Novel Approach} to {Diagnosing Southern Hemisphere Planetary Wave Activity} and {Its Influence} on {Regional Climate Variability}.''} \emph{Journal of Climate} 28 (23): 9041--57.
-\url{https://doi.org/10.1175/JCLI-D-15-0287.1}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-irving2016}{}}%
----------.
-2016.
-{``A {New Method} for {Identifying} the {Pacific}--{South American Pattern} and {Its Influence} on {Regional Climate Variability}.''} \emph{Journal of Climate} 29 (17): 6109--25.
-\url{https://doi.org/10.1175/JCLI-D-15-0843.1}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-kao2009}{}}%
-Kao, Hsun-Ying, and Jin-Yi Yu. 2009. {``Contrasting {Eastern-Pacific} and {Central-Pacific Types} of {ENSO}.''} \emph{Journal of Climate} 22 (3): 615--32. \url{https://doi.org/10.1175/2008JCLI2309.1}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-katz1991}{}}%
-Katz, Richard W., and Barbara G.
-Brown.
-1991.
-{``The Problem of Multiplicity in Research on Teleconnections.''} \emph{International Journal of Climatology} 11 (5): 505--13.
-\url{https://doi.org/10.1002/joc.3370110504}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-krokhin2007}{}}%
-Krokhin, V.
-V., and W.
-M.
-J.
-Luxemburg.
-2007.
-{``Temperatures and Precipitation Totals over the {Russian Far East} and {Eastern Siberia}: Long-Term Variability and Its Links to Teleconnection Indices.''} \emph{Hydrology and Earth System Sciences} 11 (6): 1831--41.
-\url{https://doi.org/10.5194/hess-11-1831-2007}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-lim2018}{}}%
-Lim, E.-P., H.
-H.
-Hendon, and D.
-W.
-J.
-Thompson.
-2018.
-{``Seasonal {Evolution} of {Stratosphere-Troposphere Coupling} in the {Southern Hemisphere} and {Implications} for the {Predictability} of {Surface Climate}.''} \emph{Journal of Geophysical Research: Atmospheres} 123 (21): 12, 002--12, 016.
-\url{https://doi.org/10.1029/2018JD029321}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-vanloon1972}{}}%
-Loon, Harry van, and Roy L.
-Jenne.
-1972.
-{``The Zonal Harmonic Standing Waves in the Southern Hemisphere.''} \emph{Journal of Geophysical Research} 77 (6): 992--1003.
-\url{https://doi.org/10.1029/JC077i006p00992}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-mo2000}{}}%
-Mo, Kingtse C.
-2000.
-{``Relationships Between {Low-Frequency Variability} in the {Southern Hemisphere} and {Sea Surface Temperature Anomalies}.''} \emph{Journal of Climate} 13 (20): 3599--3610.
-\url{https://doi.org/10.1175/1520-0442(2000)013\%3C3599:RBLFVI\%3E2.0.CO;2}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-mo2001}{}}%
-Mo, Kingtse C., and Julia N.
-Paegle.
-2001.
-{``The {Pacific}--{South American} Modes and Their Downstream Effects.''} \emph{International Journal of Climatology} 21 (10): 1211--29.
-\url{https://doi.org/10.1002/joc.685}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-nuncio2015}{}}%
-Nuncio, M., and Xiaojun Yuan.
-2015.
-{``The {Influence} of the {Indian Ocean Dipole} on {Antarctic Sea Ice}.''} \emph{Journal of Climate} 28 (7): 2682--90.
-\url{https://doi.org/10.1175/JCLI-D-14-00390.1}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-pezza2012}{}}%
-Pezza, Alexandre Bernardes, Harun A.
-Rashid, and Ian Simmonds.
-2012.
-{``Climate Links and Recent Extremes in Antarctic Sea Ice, High-Latitude Cyclones, {Southern Annular Mode} and {ENSO}.''} \emph{Climate Dynamics} 38 (1): 57--73.
-\url{https://doi.org/10.1007/s00382-011-1044-y}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-plumb1985}{}}%
-Plumb, R.
-Alan.
-1985.
-{``On the {Three-Dimensional Propagation} of {Stationary Waves}.''} \emph{Journal of the Atmospheric Sciences} 42 (3): 217--29.
-\url{https://doi.org/10.1175/1520-0469(1985)042\%3C0217:OTTDPO\%3E2.0.CO;2}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-rcoreteam2020}{}}%
-R Core Team. 2020. \emph{R: {A} Language and Environment for Statistical Computing}. Manual. {Vienna, Austria}: {R Foundation for Statistical Computing}. \url{https://www.R-project.org/}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-raphael2004}{}}%
-Raphael, M. N. 2004. {``A Zonal Wave 3 Index for the {Southern Hemisphere}.''} \emph{Geophysical Research Letters} 31 (23). \url{https://doi.org/10.1029/2004GL020365}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-raphael2007}{}}%
----------.
-2007.
-{``The Influence of Atmospheric Zonal Wave Three on {Antarctic} Sea Ice Variability.''} \emph{Journal of Geophysical Research: Atmospheres} 112 (D12).
-\url{https://doi.org/10.1029/2006JD007852}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-raphael2003}{}}%
-Raphael, Marilyn.
-2003.
-{``Recent, {Large-Scale Changes} in the {Extratropical Southern Hemisphere Atmospheric Circulation}.''} \emph{Journal of Climate} 16 (17): 2915--24.
-\url{https://doi.org/10.1175/1520-0442(2003)016\%3C2915:RLCITE\%3E2.0.CO;2}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-saji1999}{}}%
-Saji, N.
-H., B.
-N.
-Goswami, P.
-N.
-Vinayachandran, and T.
-Yamagata.
-1999.
-{``A Dipole Mode in the Tropical {Indian Ocean}.''} \emph{Nature} 401 (6751, 6751): 360--63.
-\url{https://doi.org/10.1038/43854}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-saji2003}{}}%
-Saji, N. H., and T. Yamagata. 2003. {``Possible Impacts of {Indian Ocean Dipole} Mode Events on Global Climate.''} \emph{Climate Research} 25 (2): 151--69. \url{https://doi.org/10.3354/cr025151}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-smith1995}{}}%
-Smith, Anne K.
-1995.
-{``Numerical Simulation of Global Variations of Temperature, Ozone, and Trace Species in the Stratosphere.''} \emph{Journal of Geophysical Research: Atmospheres} 100 (D1): 1253--69.
-\url{https://doi.org/10.1029/94JD02395}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-trenberth1980a}{}}%
-Trenberth, Kevin E.
-1980.
-{``Planetary {Waves} at 500 Mb in the {Southern Hemisphere}.''} \emph{Monthly Weather Review} 108 (9): 1378--89.
-\url{https://doi.org/10.1175/1520-0493(1980)108\%3C1378:PWAMIT\%3E2.0.CO;2}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-trenberth1985}{}}%
-Trenberth, Kevin F., and K.
-C.
-Mo.
-1985.
-{``Blocking in the {Southern Hemisphere}.''} \emph{Monthly Weather Review} 113 (1): 3--21.
-\url{https://doi.org/10.1175/1520-0493(1985)113\%3C0003:BITSH\%3E2.0.CO;2}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-turner2017}{}}%
-Turner, John, J.
-Scott Hosking, Thomas J.
-Bracegirdle, Tony Phillips, and Gareth J.
-Marshall.
-2017.
-{``Variability and Trends in the {Southern Hemisphere} High Latitude, Quasi-Stationary Planetary Waves.''} \emph{International Journal of Climatology} 37 (5): 2325--36.
-\url{https://doi.org/10.1002/joc.4848}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-walker1914}{}}%
-Walker, Sir Gilbert Thomas.
-1914.
-\emph{Correlation in {Seasonal Variations} of {Weather}, {III}: {On} the {Criterion} for the {Reality} of {Relationships Or Periodicities}}.
-{Meteorological Office}.
-\url{https://books.google.com?id=cuGtuAAACAAJ}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-wickham2009}{}}%
-Wickham, Hadley. 2009. \emph{Ggplot2: {Elegant Graphics} for {Data Analysis}}. Use {R}! {New York}: {Springer-Verlag}. \url{https://doi.org/10.1007/978-0-387-98141-3}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-wilks2016}{}}%
-Wilks, D.
-S.
-2016.
-{``{`{The Stippling Shows Statistically Significant Grid Points}'}: {How Research Results} Are {Routinely Overstated} and {Overinterpreted}, and {What} to {Do} about {It}.''} \emph{Bulletin of the American Meteorological Society} 97 (12): 2263--73.
-\url{https://doi.org/10.1175/BAMS-D-15-00267.1}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-wilks2011}{}}%
-Wilks, Daniel. 2011. \emph{Statistical {Methods} in the {Atmospheric Sciences}}. Vol. 100. {Elsevier}. \url{https://doi.org/10.1016/B978-0-12-385022-5.00022-1}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-wirth1993}{}}%
-Wirth, Volkmar.
-1993.
-{``Quasi-Stationary Planetary Waves in Total Ozone and Their Correlation with Lower Stratospheric Temperature.''} \emph{Journal of Geophysical Research: Atmospheres} 98 (D5): 8873--82.
-\url{https://doi.org/10.1029/92JD02820}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-cmap}{}}%
-Xie, Pingping, and Phillip A.
-Arkin.
-1997.
-{``Global {Precipitation}: {A} 17-{Year Monthly Analysis Based} on {Gauge Observations}, {Satellite Estimates}, and {Numerical Model Outputs}.''} \emph{Bull.
-Amer.
-Meteor.
-Soc.} 78 (Accessed on {\(<\)}26-08-2021{\(>\)}) (11): 2539--58.
-\url{https://doi.org/10.1175/1520-0477(1997)078\%3C2539:GPAYMA\%3E2.0.CO;2}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-xie2015}{}}%
-Xie, Yihui. 2015. \emph{Dynamic Documents with {R} and Knitr}. 2nd ed. {Boca Raton, Florida}: {Chapman and Hall/CRC}. \url{https://yihui.org/knitr/}.
-
-\end{CSLReferences}
-
-\bibliographystyle{spbasic}
 \bibliography{references.bib,data.bib,packages.bib}

### 2022-12-21 analysis/paper/shceof.Rmd
Actualiza

@@
 lab_lev <- AddSuffix(" hPa")
-lab_cplx <- c(R = "Real", 
-              I = "Imaginary")
+lab_cplx <- c("Real" = "0º", 
+              "Imaginary" = "90º")
 
@@
 These derived methods are grounded on other methods such as Fourier decomposition or EOF to identify the centres of action for the described phenomena and can be useful to characterise features that are not readily apparent with these methods.
-It is commonly easy the computation of these kind of indices, but they could be unable to capture non-stationary patterns.
+These kinds of indices are generally easy to compute, but they usually capture non-stationary patterns.
 
@@
 
-We restrict this paper to the spring September-October-November (SON) trimester because this season experiences a maximum in tropical teleconnections over South America [@cazes-boezio2003] and SH stratosphere-troposphere interactions [@lim2018].
+We restrict this paper to the spring September-October-November (SON) trimester because during this season both the tropical teleconnections over South America [@cazes-boezio2003] and the SH stratosphere-troposphere interactions [@lim2018] are at their maximum amplitude.
 
@@
 In Section\ \@ref(regressions) we study the spatial regressions with geopotential height, temperature and ozone anomalies.
-In Section\ \@ref(psa) and \@ref(sam) we analyse the relationship between cEOF2 the PSA and SAM modes.
+In Section\ \@ref(psa) and \@ref(sam) we analyse the relationship between cEOF2, the PSA and SAM modes.
 In Section\ \@ref(tropical) we study tropical forcings that explain the variability of each cEOF.
@@
 
-We used monthly geopotential height, air temperature, ozone mixing ratio, and total ozone column (TOC) at 2.5\degree longitude by 2.5\degree latitude of horizontal resolution and 37 vertical isobaric levels from the European Centre for Medium-Range Weather Forecasts Reanalysis version 5 [ERA; @era5] for the period 1979 -- 2019.
+We used monthly geopotential height, air temperature, ozone mixing ratio, and total column ozone (TOC) at 2.5\degree longitude by 2.5\degree latitude of horizontal resolution and 37 vertical isobaric levels from the European Centre for Medium-Range Weather Forecasts Reanalysis version 5 [ERA; @era5] for the period 1979 -- 2019.
 Most of our analysis is restricted to the post-satellite era to avoid any confounding factors arising from the incorporation of satellite observations, but we also used the preliminary back extension of ERA5 from 1950 to 1978 [@era5be] to describe long-term trends.
@@
 We further split the SAM into its zonally symmetric and zonally asymmetric components (S-SAM and A-SAM indices respectively).
-These indices were obtained by projecting the zonally asymmetric and zonally symmetric part of the SAM spatial pattern onto monthly geopotential height fields, as proposed by @campitelli2022.
-Seasonal indices of the PSA patterns (PSA1 and PSA2) were calculated, in agreement with @mo2001, as the third and fourth leading EOF of seasonal mean anomalies for 500-hPa geopotential heights at SH.
+These indices were obtained by projecting monthly geopotential height fields onto the zonally asymmetric and zonally symmetric parts of the SAM spatial pattern, as proposed by @campitelli2022.
+Seasonal indices of the PSA patterns (PSA1 and PSA2) were calculated, in agreement with @mo2001, as the third and fourth EOFs of seasonal mean anomalies for 500-hPa geopotential heights at SH.
 
@@
   tagger::tag_facets("rc")
-
-```
-
-In the standard EOF analysis, zonal waves may appear as pairs of (possibly degenerated) EOFs representing similar patterns but shifted in phase [@horel1984].
+```
+
+In the standard EOF analysis, zonal waves may appear as pairs of (possibly degenerate) EOFs representing similar patterns but shifted in phase [@horel1984].
 Figure \@ref(fig:eof-naive) shows the four leading EOFs of SON geopotential height zonal anomalies at 50 hPa south of 20\degree S.
-It is clear that the first two EOFs represent a single phase-varying zonal wave 1 pattern and the last two represent a similarly phase-varying pattern with shorter wavenumber and four centres of action shifted by 1/4 wavelength.
-A similar EOF structure can be seen in 200 hPa (not shown).
-Since each pair of EOFs seems to represent the same phase-varying structure, it would be desirable to combine them into a single pattern described by amplitude and phase.
-
-Complex Empirical Orthogonal Functions (cEOF) are a useful method for characterising zonal waves, associated with phase-varying structures [@horel1984].
-This method involves computing EOF on the analytic representation of the original field.
-That representation is a complex field in which the real part is the original data and the imaginary part is the original data shifted by 90\degree at each spectral frequency -- i.e.
-its Hilbert transform.
+It is clear that the first two EOFs represent a single phase-varying zonal wave 1 pattern and the last two represent a similarly phase-varying pattern with higher wavenumber and three centres of action shifted by 1/4 wavelength (90\degree in frequency space).
+
+To describe the phase-varying nature of these two waves one could combine each pair into indices of amplitude of phase. 
+So, for instance, the amplitude of the wave 1 EOF could be measured as $\sqrt{\mathrm{EOF1}^2 + \mathrm{EOF2}^2}$ and it's phase as $\tan^{-1} \left ( \frac{\mathrm{EOF2}}{\mathrm{EOF1}} \right )$ (where $\mathrm{EOF1}$ and $\mathrm{EOF2}$ are the time series associated with each EOF).
+However, this rests upon visual inspection of the spatial patterns and only works if both phases appear cleanly in different EOFs, which is not guaranteed. 
+In particular, this doesn't work with the wave 1 pattern observed as the leading EOF in 200hPa geopotential heihgt zonal anomalies (not shown).
+
+In Complex Empirical Orthogonal Functions (cEOF) analysis, each cEOF is represented by two spatial patterns that are shifted by 1/4 wavelength by construction. 
+The actual field reconstructed by each cEOF is the linear combination of the two spatial fields weighted by its respective time series. 
+This is analogous to how any sine wave with any amplitude and phase can be constructed by the sum of a sine wave and cosine wave with different amplitude but constant phase. 
+This means that cEOFs naturally represent phase-varying wave-like patterns that change location as well as amplitude.
+
+For example, Figure \@ref(fig:ceofs-1) a.1 shows the two spatial patterns of a cEOF, one with shaded  contours (which we call the 0º phase) and the other with black contours (which we call the 90º phase).
+This cEOF is very similar to the two leading EOFs shown in Figure \@ref(fig:eof-naive) and represents a zonal wave 1 pattern. 
+When the phase of the wave matches the 0º phase, then the 0º phase time series will be positive and the 90º phase time series will be zero. 
+Similarly, when the phase of the wave matches the 90º phase, the 90º phase time series will be positive and the 0º phase time series will be zero.  
+Cases will have non zero values in both time series will represent intermediate phases.
+
+These two phases are not special and the same cEOF could be represented by any two linear combinations of the phases shown in Figure \@ref(fig:ceofs-1) a.1. 
+This parallels traditional EOFs, which are defined up to an additive constant.
+For real EOFs, this constant can be either 0 or $\pi$, corresponding to a change in sign.
+For cEOFs, it can be any real number between 0 and $2\pi$ [@horel1984], corresponding to rotations in the complex plane.
+
+Compute cEOFs are computed the same way as traditional EOFs except that the data is first augmented by computing its analytic signal. 
+This is a complex number whose real part is the original series and the imaginary part is the original data shifted by 90\degree at each spectral frequency -- i.e. its Hilbert transform.
+Each cEOF is a set of complex-valued spatial patterns and time series in which the real and imaginary parts are the two orthogonal phases of the wave-like pattern.
+
+In this paper we use the term 0º cEOF and 90º cEOF to refer to each part of the whole cEOF.
+
 The Hilbert transform is usually understood in terms of time-varying signal.
@@
 Since each latitude circle is a periodic domain, this procedure does not suffer from edge effects.
-
-The result of the cEOF methodology is a set of complex spatial patterns and complex time series.
-The real and imaginary part of each spatial pattern represent the two phases of wave-like spatial patterns that are in quadrature.
-The magnitude and argument of each complex time series represent the amplitude and phase of each zonal wave.
 
@@
 
-(ref:corr-ceof-splitted-cap) Coefficient of determination ($R^2$) between the absolute magnitude of complex EOFs computed separately at 200 hPa and 50 hPa (p-values lower than 0.01 in bold).
+(ref:corr-ceof-splitted-cap) Coefficient of determination ($r^2$) between the time series of the absolute magnitude of complex EOFs computed separately at 200 hPa and 50 hPa (p-values lower than 0.01 in bold).
 
@@
 
-The cEOF methodology is applied to SON geopotential height zonal anomalies south of 20\degree S separately at 50 and 200 hPa.
+The cEOF methodology is applied to zonal anomalies of SON geopotential height south of 20\degree S separately at 50 and 200 hPa.
 Table \@ref(tab:corr-ceof-splitted) shows the coefficient of determination between time series of the amplitude of each cEOF across levels.
@@
 
-The phase of cEOFs is defined up to an additive constant.
-For real EOFs, this constant can be either 0 or $\pi$, corresponding to a change in sign.
-For cEOFs, it can be any real number between 0 and $2\pi$ [@horel1984], corresponding to rotations in the complex plane.
-Since any choice is arbitrary and equally valid, we chose the phase of each cEOF so that the real or imaginary parts are aligned with meaningful phases in our analysis.
+As mentioned before, the choice of phases is arbitrary and equally valid. 
+But to make the interpretation easier, we chose the phase of each cEOF so that the either the º0 cEOF or the 90º cEOF is aligned with meaningful variables in our analysis.
 This procedure does not create spurious correlations, it only takes an existing relationship and aligns it with a specific phase.
 
-Preliminary analysis showed that the first cEOF was closely related to the zonal wave 1 of Total Ozone Column and the second cEOF was closely related to ENSO.
-Therefore, to aid in the interpretation we chose the phase of cEOF1 so that the time series corresponding to the real part has the maximum correlation with the zonal wave 1 of Total Ozone Column between `r knitr::combine_words(LatLabel(o3wave_lats))`.
-Similarly, we chose the phase of cEOF2 so that the coefficient of determination between the Oceanic Niño Index [@bamston1997] and the real part is minimised, which also nearly maximises the correlation with the imaginary part.
+In Section \@ref(precipitation) we show regressions of precipitation and temperature associated with intermediate phases.
+For those plots, we rotated the cEOFs by 1/4 wavelength by multiplying the complex time series by $\cos(\pi/4) + i\sin(\pi/4)$ and computing the regression on those rotated timeseries.
+
+Preliminary analysis showed that the first cEOF was closely related to the zonal wave 1 of Total Column Ozone and the second cEOF was closely related to ENSO.
+Therefore, we chose the phase of cEOF1 so that the time series corresponding to the 0ª cEOF1 has the maximum correlation with the zonal wave 1 of Total Column Ozone between `r knitr::combine_words(LatLabel(o3wave_lats))`.
+Similarly, we chose the phase of cEOF2 so that the coefficient of determination between the Oceanic Niño Index [@bamston1997] and the º0 cEOF2 is minimised, which also nearly maximises the correlation with the 90º cEOF2.
 
@@
 angles <- seq(-pi, pi, by = .5*pi/180)
-
-# rotations_cEOF1 <- lapply(angles, function(a) {
-#   ceof$eof[[1]]$left[cEOF == "cEOF1"] %>%
-#     copy() %>%
-#     .[, hgt := rotate(hgt, a)] %>%
-#     sep_ReIm() %>%
-#     .[, .(mean = mean(hgt)), by = part] %>%
-#     .[, rotation := a]
-# }) %>%
-#   rbindlist()
-# 
-# 
-# best_rotation_cEOF1 <- rotations_cEOF1[part == "Real"][which.min(mean)]$rotation
 
@@
 We performed linear regressions to quantify the association between the cEOFs and other variables (e.g. geopotential height, temperature, precipitation, and others).
-For each cEOF, we computed regression maps by fitting a multiple linear model involving both the real and the imaginary part.
-To obtain the linear coefficients of a variable $X$ with the imaginary and real parts of each cEOF we fit the equation
+For each cEOF, we computed regression maps by fitting a multiple linear model involving both the 0º and the 90º phases.
+To obtain the linear coefficients of a variable $X$ with the 0º and 90º phase of each cEOF we fit the equation
 
 $$
-X(\lambda, \phi, t) = \alpha(\lambda, \phi) \operatorname{Im}(\mathrm{cEOF}) + \beta(\lambda, \phi) \mathrm{Re}(\mathrm{cEOF}) + X_0(\lambda, \phi) + \epsilon(\lambda, \phi, t)
+X(\lambda, \phi, t) = \alpha(\lambda, \phi) \operatorname{cEOF_{0º}} + \beta(\lambda, \phi) \operatorname{cEOF_{90º}} + X_0(\lambda, \phi) + \epsilon(\lambda, \phi, t)
 $$
 
-where $\lambda$ and $\phi$ are the longitude and latitude, $t$ is the time, $\alpha$ and $\beta$ are the linear regression coefficients for imaginary and real parts respectively, $X_0$ and $\epsilon$ are the constant and error terms respectively.
+where $\lambda$ and $\phi$ are the longitude and latitude, $t$ is the time, $\alpha$ and $\beta$ are the linear regression coefficients for 0º and 90º phases respectively, $X_0$ and $\epsilon$ are the constant and error terms respectively.
 
@@
 ```{r ceofs-1, fig.cap = "Spatial patterns for the two leading cEOFs of SON zonal anomalies of geopotential height at 50 hPa and 200 hPa for the 1979 -- 2019 period.
-The shading (contours) corresponds to real (imaginary) part.
+The shading (contours) corresponds to 0º (90º) phase.
 Arbitrary units.
@@
 (ref:extended-series-cap) Time series of the two leading cEOFs of SON zonal anomalies of geopotential height at 50 hPa and 200 hPa.
-cEOF1 (row a) and cEOF2 (row b) separated in their real (column 1) and imaginary (column 2) components.
+cEOF1 (row a) and cEOF2 (row b) separated in their 0º (column 1) and 90º (column 2) phase.
 Dark straight line is the linear trend.
@@
   geom_line(aes(color = part)) +
+  # geom_line(data = ceof[, eof[[1]]$left] %>% 
+  #             sep_ReIm()) + 
   geom_point(aes(color = part), size = 0.25) +
@@
                      guide = guide_legend(override.aes = list(fill = NA))) +
-  facet_grid(cEOF ~ part) +
+  facet_grid(cEOF ~ part, labeller = labeller(part = lab_cplx)) +
   theme(panel.spacing = grid::unit(1.5, "lines"),
@@
 The first mode (cEOF1) explains 82% of the variance of the zonally anomalous fields, while the second mode (cEOF2) explains a smaller fraction (7%).
-In the spatial patterns (Fig. \@ref(fig:ceofs-1)), the real and the imaginary components are in quadrature by construction, so that each cEOF describe a single wave-like pattern whose amplitude and position (i.e.
+In the spatial patterns (Fig. \@ref(fig:ceofs-1)), the 0º and the 90º phases are in quadrature by construction, so that each cEOF describe a single wave-like pattern whose amplitude and position (i.e.
 phase) is controlled by the magnitude and phase of the temporal cEOF.
@@
 The cEOF1 (Fig. \@ref(fig:ceofs-1) column a) is a hemispheric wave 1 pattern with maximum amplitude at high latitudes.
-At 50 hPa the Real cEOF1 has the maximum of the wave 1 at 150ºE and at 200 hPa, the maximum is located at around 175ºE indicating a westerly shift in phase.
+At 50 hPa the 0º cEOF1 has the maximum of the wave 1 at 150ºE and at 200 hPa, the maximum is located at around 175ºE indicating a westward shift with height.
 The cEOF2 (Fig. \@ref(fig:eof-naive) column b) shows also a zonal wave-like structure with maximum amplitude at high latitudes, but with shorter spatial scales.
 In particular, the dominant structure at both levels is a wave 3 but with larger amplitude in the pacific sector.
-This modulated amplitude is more evident at 200 hPa.
+This modulated amplitude is more evident at 200 hPa, which is consistent with the cEOF2 computed separatedly for 200 hPa explains a bit more variance than the cEOF2 computed separatedly for 50 hPa (`r scales::percent(ceof_splitted[, eof[[1]]$sdev, by = levs][cEOF == "cEOF2", r2][1])` vs.
+`r scales::percent(ceof_splitted[, eof[[1]]$sdev, by = levs][cEOF == "cEOF2", r2][2])`, respectively).
 There is no apparent phase shift with height but the amplitude of the pattern is greatly reduced in the stratosphere, suggesting that this barotropic mode represents mainly tropospheric variability.
+
+```{r}
+mean <- era5 %>% 
+  copy() %>% 
+  .[, hgt := metR::Anomaly(hgt),
+    by  = .(lat, time, lev)] %>% 
+  .[, hgt := hgt/sd(hgt), by = .(lev)] %>%
+  # .[, hgt := hgt*sqrt(cos(lat*pi/180))] %>% 
+  .[, .(hgt_m = mean(hgt)), by = .(lev, lon, lat)]
+
+ceof_mean_r2 <- ceof[, eof[[1]]$right] %>% 
+  sep_ReIm(format = "wider") %>%
+  .[, hgt := NULL] %>% 
+  .[] %>% 
+  .[mean, on = .NATURAL] %>% 
+  .[lat <= -20] %>% 
+  .[, FitLm(hgt_m, Real, Imaginary, r2 = TRUE), by = .(cEOF)] %>% 
+  .[, .SD[1], by = .(cEOF)]
+```
 
@@
 However, cEOF2 temporal mean is almost zero, which indicates that only cEOF1 includes variability that significantly projects onto the mean zonally anomalous field.
-
-A significant positive trend in the real component of cEOF1 is evident (Fig. \@ref(fig:extended-series)a.1, p-value `r trends[part == "Real" & cEOF == "cEOF1"]$p_report`) while there is no significant trend in any of the complex components of cEOF2.
-The positive trend in the Real cEOF1 translates into a positive trend in cEOF1 magnitude, but not systematic change in phase (not shown).
+This is consistent with the fact that the mean zonally anomalous field of geopotential height is very similar to the cEOF1 ($r^2$ = `r scales::percent(ceof_mean_r2[["r.squared"]][1])`) and not similar to the cEOF2 ($r^2$ = `r scales::percent(ceof_mean_r2[["r.squared"]][2])`).
+
+A significant positive trend in the 0º phase of cEOF1 is evident (Fig. \@ref(fig:extended-series)a.1, p-value `r trends[part == "Real" & cEOF == "cEOF1"]$p_report`) while there is no significant trend in any of the phases of cEOF2.
+The positive trend in the 0º cEOF1 translates into a positive trend in cEOF1 magnitude, but not systematic change in phase (not shown).
 This long-term change indicates an increase in the magnitude of the high latitude zonal wave 1.
-In agreement, @raphael2003 detected a step change after around 1975 in the temporal evolution of leading EOF computed from August-September-October 500 hPa zonal geopotential height anomalies, which is similar to the cEOF1.
 
@@
 
-In the previous section, cEOFs analysis was applied to zonal anomalies, that is anomalies derived by removing the zonally mean values in order to isolate the main characteristics of the main zonal waves characterizing the circulation in the SH.
+In the previous section, cEOFs analysis was applied to zonal anomalies derived by removing the zonally mean values in order to isolate the main characteristics of the main zonal waves characterizing the circulation in the SH.
 In addition, regression fields were computed using the full fields of the variables in order to describe the influence of the cEOFs on the temporal anomalies.
@@
     facet_grid(lev ~ term, 
-               labeller = labeller(lev = lab_lev)) +
+               labeller = labeller(lev = lab_lev, term = lab_cplx)) +
     geom_coords() +
@@
 
-(ref:eof1-regr-gh-cap) Regression of SON geopotential height anomalies ($m^2s^{-1}$) with the (column 1) real and (column 2) imaginary parts of the first cEOF for the 1979 -- 2019 period at (row a) 50 hPa and (row b) 200 hPa.
-These coefficients come from multiple linear regression involving the real and imaginary parts.
+(ref:eof1-regr-gh-cap) Regression of SON geopotential height anomalies ($m^2s^{-1}$) with the (column 1) 0º and (column 2) 90º phases of the first cEOF for the 1979 -- 2019 period at (row a) 50 hPa and (row b) 200 hPa.
+These coefficients come from multiple linear regression involving the 0º and 90º phases.
 Areas marked with dots have p-values smaller than 0.01 adjusted for False Detection Rate.
@@
 Figure \@ref(fig:eof1-regr-gh) shows regression maps of SON geopotential height anomalies upon cEOF1.
-At 50 hPa (Figure \@ref(fig:eof1-regr-gh) row a), the Real cEOF1 is associated with a centre located over the Ross Sea.
-The Imaginary cEOF1 is associated with a distinctive wave 1 pattern with maximum over the coast of East Antarctica. 
-At 200 hPa (Figure \@ref(fig:eof1-regr-gh) row b) the Real cEOF1 shows a single centre of positive anomalies spanning West Antarctica surrounded by opposite anomalies in lower latitudes, with is centre shifted slightly eastward compared with the upper-level anomalies.
-The Imaginary cEOF1 shows a much more zonally symmetrical pattern resembling the negative SAM phase [e.g. @fogt2020].
+At 50 hPa (Figure \@ref(fig:eof1-regr-gh) row a), the 0º cEOF1 is associated with a centre located over the Ross Sea.
+The 90º cEOF1 is associated with a distinctive wave 1 pattern with maximum over the coast of East Antarctica. 
+At 200 hPa (Figure \@ref(fig:eof1-regr-gh) row b) the 0º cEOF1 shows a single centre of positive anomalies spanning West Antarctica surrounded by opposite anomalies in lower latitudes, with its centre shifted slightly eastward compared with the upper-level anomalies.
+The 90º cEOF1 shows a much more zonally symmetrical pattern resembling the negative SAM phase [e.g. @fogt2020].
 Therefore, the magnitude and phase of the cEOF1 are associated with the magnitude and phase of a zonal wave mainly in the stratosphere.
@@
 Figure \@ref(fig:eof2-regr-gh) shows the regression maps of geopotential height anomalies upon the cEOF2.
-In the troposphere (Fig. \@ref(fig:eof2-regr-gh) column b), wave trains similar to those identified for cEOF2 patterns (Fig \@ref(fig:ceofs-1)) can be distinguished.
-Regressed anomalies associated with the Real cEOF2 are 90º out of phase with those associated with the Imaginary cEOF2.
+In the troposphere (Fig. \@ref(fig:eof2-regr-gh) row b) the regression maps show wave trains similar to those identified for cEOF2 patterns (Fig \@ref(fig:ceofs-1)).
+Regressed anomalies associated with the 0º cEOF2 are 1/4 wavelength out of phase with those associated with the 90º cEOF2.
 All fields have a dominant zonal wave 3 limited to the western hemisphere, over the Pacific and Atlantic Oceans.
 cEOF2 then represents an equivalent barotropic wave train that is very similar to the the PSA Patterns [@mo2001].
-Comparing the location of the positive anomaly near 90ºW in column b of Figure \@ref(fig:eof2-regr-gh) with Figures 1.a and b from @mo2001, the Real cEOF2 regression map could be identified with PSA2, while the Imaginary cEOF2 resembles PSA1.
-
-These wave patterns are also present in the stratosphere (Fig. \@ref(fig:eof2-regr-gh) column a) supporting their  equivalent barotropic nature. 
-But also present is a monopole over the pole with negative sign associated with the Real cEOF2 and positive sign associated with the Imaginary cEOF2. 
-This monopole might indicate strengthening of the polar vortex associated with positive values of the Real cEOF2 and weakening associated with positive Imaginary cEOF2.
+Comparing the location of the positive anomaly near 90ºW in column b of Figure \@ref(fig:eof2-regr-gh) with Figures 1.a and b from @mo2001, the 0º cEOF2 regression map could be identified with PSA2, while the 90º cEOF2 resembles PSA1.
+
+These wave patterns are also present in the stratosphere (Fig. \@ref(fig:eof2-regr-gh) row a) supporting their  equivalent barotropic nature. 
+But also present is a monopole over the pole with negative sign associated with the 0º cEOF2 and positive sign associated with the 90º cEOF2. 
+This monopole might indicate strengthening of the polar vortex associated with positive values of the 0º cEOF2 and weakening associated with negative values of 0º cEOF2.
 
@@
 
-(ref:t-vertical-cap) Regression of SON zonal anomalies  averaged between `r knitr::combine_words(LatLabel(o3wave_lats))` of mean air temperature (shaded, Kelvin) and ozone mixing ratio (contours, negative contours with dashed lines, labels in parts per billion by mass) with the (a) Real and (b) Imaginary components of the cEOF1 for the 1979 -- 2019 period.
+(ref:t-vertical-cap) Regression of SON zonal anomalies  averaged between `r knitr::combine_words(LatLabel(o3wave_lats))` of mean air temperature (shaded, Kelvin) and ozone mixing ratio (contours, negative contours with dashed lines, labels in parts per billion by mass) with the (a) 0º and (b) 90º phase of the cEOF1 for the 1979 -- 2019 period.
 
@@
   scale_x_longitude(ticks = 60) +
-  facet_grid(~term) +
+  facet_grid(~term, labeller = labeller(term = lab_cplx)) +
   axis_labs_smol +
@@
 The distribution of temperature regression coefficients at 50 hPa and at 200 hPa mirror the geopotential height regression maps at 50 hPa (Fig. \@ref(fig:eof1-regr-gh)).
-In both levels, the Real cEOF1 is associated with a positive centre over the South Pole with its centre moved slightly towards 150ºE (Fig. \@ref(fig:eof1-regr-t) column 1).
-On the other hand, the regression maps on the Imaginary cEOF1 show a more clear wave 1 pattern with its maximum around 60ºE.
+In both levels, the 0º cEOF1 is associated with a positive centre over the South Pole with its centre moved slightly towards 150ºE (Fig. \@ref(fig:eof1-regr-t) column 1).
+On the other hand, the regression maps on the 90º cEOF1 show a more clear wave 1 pattern with its maximum around 60ºE.
 
 Figure\ \@ref(fig:t-vertical) shows the vertical distribution of the regression coefficients on cEOF1 from zonal anomalies averaged between `r knitr::combine_words(LatLabel(o3wave_lats))` of air temperature and of ozone mixing ratio.
-Temperature zonal anomalies associated with cEOF1 show a clear wave 1 pattern for both real and imaginary components throughout the atmosphere above 250 hPa with a sign reversal above 10 hPa.
+Temperature zonal anomalies associated with cEOF1 show a clear wave 1 pattern for both 0º and 90º phases throughout the atmosphere above 250 hPa with a sign reversal above 10 hPa.
 As a result of the hydrostatic balance, this is the level in which the geopotential anomaly have maximum amplitude (not shown).
@@
 The maximum ozone regressed anomalies match with the minimum temperature anomalies above 10 hPa and with the maximum temperature anomalies below 10 hPa (Fig. \@ref(fig:t-vertical)).
-Therefore, the ozone zonal wave 1 is anticorrelated with the temperature zonal wave 1 in the upper stratosphere, and directly correlated in the upper stratosphere. 
+Therefore, the ozone zonal wave 1 is negatively correlated with the temperature zonal wave 1 in the upper stratosphere, and positively correlated in the upper stratosphere. 
 This change in phase is observed in ozone anomalies forced by planetary waves that reach the stratosphere [@hartmann1979; @wirth1993; @smith1995]. 
 In the photochemically-dominated upper stratosphere, cold temperatures inhibit the destruction of ozone, explaining the opposite behaviour for both variables.
-On the other hand, in the advectively-dominated lower stratosphere, ozone anomalies are 90º out of phase with horizontal and vertical transport, which are in addition 90º out of phase with temperature anomalies, resulting in same sign anomalies for the response of both variables [@hartmann1979; @wirth1993; @smith1995].
+On the other hand, in the advectively-dominated lower stratosphere, ozone anomalies are 1/4 wavelength out of phase with horizontal and vertical transport, which are in addition 1/4 wavelength out of phase with temperature anomalies, resulting in same sign anomalies for the response of both variables [@hartmann1979; @wirth1993; @smith1995].
 
@@
 
-(ref:o3-regr-cap) Regression of SON mean Total Ozone Column anomalies (shaded, Dobson Units) with the (a) real and (b) imaginary components of the cEOF1 for the 1979 -- 2019 period.
-On contours, the mean zonal anomaly of Total Ozone Column (negative contours in dashed lines, Dobson Units).
+(ref:o3-regr-cap) Regression of SON mean Total Column Ozone anomalies (shaded, Dobson Units) with the (a) 0º and (b) 90º phases of the cEOF1 for the 1979 -- 2019 period.
+On contours, the mean zonal anomaly of Total Column Ozone (negative contours in dashed lines, Dobson Units).
 Areas marked with dots have p-values smaller than 0.01 adjusted for False Detection Rate.
@@
   geom_qmap() +
-  scale_fill_divergent_discretised("Total Ozone Column (DU)", 
+  scale_fill_divergent_discretised("Total Column Ozone (DU)", 
                                    limits = c(-30, 30), 
@@
   coord_polar() +
-  facet_grid(cEOF~term) +
+  facet_grid(cEOF~term, labeller = labeller(term = lab_cplx)) +
   tag_facets("panel")
@@
 
-(ref:wave1-o3-cap) Relationship between cEOF1 amplitude and phase with amplitude and phase of the zonal wave 1 in Total Ozone Column averaged between `r knitr::combine_words(LatLabel(o3wave_lats))` for the period 1979 -- 2019.
+(ref:wave1-o3-cap) Relationship between cEOF1 amplitude and phase with amplitude and phase of the zonal wave 1 in Total Column Ozone averaged between `r knitr::combine_words(LatLabel(o3wave_lats))` for the period 1979 -- 2019.
 
@@
 The climatological position of the springtime Ozone minimum (ozone hole) is  outside the South Pole and towards the Weddell Sea [e.g. @grytsai2011].
-Thus, the Real cEOF1 regression field (Figure\ \@ref(fig:o3-regr)a) coincides with the climatological position of the ozone hole, while it is 90° out of phase for the Imaginary cEOF1.
+Thus, the 0º cEOF1 regression field (Figure\ \@ref(fig:o3-regr)a) coincides with the climatological position of the ozone hole, while it is 90° out of phase for the 90º cEOF1.
 The temporal correlation between the amplitudes of TOC planetary wave 1 and cEOF1 is `r wave1_cor[with == "amplitude"]$text`, while the correlation between their phases is `r wave1_cor[with == "phase"]$text`.
@@
 Given the similarity between the cEOF2 related anomaly structures (Fig. \@ref(fig:eof2-regr-gh)) and documented PSA patterns we study the relationship between them.
-Table \@ref(tab:psa-eof2) shows the correlations between the two PSA indices and the time series for Real and Imaginary phases of cEOF2.
-As visually anticipated by Figure \@ref(fig:eof2-regr-gh), there is a large positive correlation between PSA1 and Imaginary cEOF2, and between PSA2 and Real cEOF2.
-On the other hand, there is no significant relationship between PSA1 and Real cEOF2, and between PSA2 and Imaginary cEOF2.
+Table \@ref(tab:psa-eof2) shows the correlations between the two PSA indices and the time series for 0º and 90º phases of cEOF2.
+As visually anticipated by Figure \@ref(fig:eof2-regr-gh), there is a large positive correlation between PSA1 and 90º cEOF2, and between PSA2 and 0º cEOF2.
+On the other hand, there is no significant relationship between PSA1 and 0º cEOF2, and between PSA2 and 90º cEOF2.
 As a result, cEOF2 represents well both the spatial structure and temporal evolution of the PSA modes, so it is possible to make an association between its two phases and the two PSA modes.
-That is, the phase election for cEOF2 that maximises the relationship between ENSO and Imaginary cEOF2, also maximises the association between cEOF2 components and PSA modes (not shown).
+That is, the phase election for cEOF2 that maximises the relationship between ENSO and 90º cEOF2, also maximises the association between cEOF2 components and PSA modes (not shown).
 
@@
                  color = "black", fill = "gray70") + 
-  annotate("text",
-           label = c("- Real", "- Imaginary", "+ Real", "+ Imaginary"),
-           color = "black", size = 3,
-           y = 0.5, x = c(-pi, -pi/2, 0, pi/2), angle = 90, hjust = 0) +
+  # annotate("text",
+  #          label = c("- Real", "- Imaginary", "+ Real", "+ Imaginary"),
+  #          color = "black", size = 3,
+  #          y = 0.5, x = c(-pi, -pi/2, 0, pi/2), angle = 90, hjust = 0) +
   geom_rug() +
@@
 
-Figure \@ref(fig:phase-histogram) shows an histogram that counts the number of SON seasons in which the cEOF2 phase was close to each of the four particular phases (positive/negative of real/imaginary component), with the observations for each season marked as rugs on the horizontal axis.
-In `r scales::percent(imaginary)` of seasons cEOF2 has a phase similar to either the negative or positive imaginary phase, making the imaginary phase the most common phase. 
+Figure \@ref(fig:phase-histogram) shows an histogram that counts the number of SON seasons in which the cEOF2 phase was close to each of the four particular phases (positive/negative of 0º and 90º phases), with the observations for each season marked as rugs on the horizontal axis.
+In `r scales::percent(imaginary)` of seasons cEOF2 has a phase similar to either the negative or positive 90º phase, making the 90º phase the most common phase. 
 This is also, by construction, the phase that is most correlated with ENSO.
 
-Therefore, by virtue of being the most common phase, the Imaginary cEOF2 explains more variance than the Real cEOF2.
-Conventional EOF analysis will therefore tend to separate them relatively cleanly, with the EOF representing the Imaginary cEOF2 always leading the one representing the Real cEOF2.
+Therefore, by virtue of being the most common phase, the 90º cEOF2 explains more variance than the 0º cEOF2.
+Conventional EOF analysis will therefore tend to separate them relatively cleanly, with the EOF representing the 90º cEOF2 always leading the one representing the 0º cEOF2.
 This phase preferences is in agreement with @irving2016, who found a bimodal distribution to PSA-like variability (compare our Figure \@ref(fig:phase-histogram) with their Figure 6).
@@
                                                          sym = "S-SAM")) +
-  facet_grid(part ~ cEOF) +
+  facet_grid(part ~ cEOF, labeller = labeller(part = lab_cplx)) +
   coord_flip() +
@@
 We then computed the coefficient of determination between the cEOFs time series and the three SAM indices (SAM, A-SAM and S-SAM) defined by @campitelli2022 at each vertical level (Fig.\ \@ref(fig:sam-eof-vertical)).
-The SAM index is statistically significant correlated with the Real cEOF1 in all levels, and with the Imaginary cEOF1 and Imaginary cEOF2 in the troposphere.
-On the other hand, correlations between SAM and the Real cEOF2 are non-significant.
-
-The relationship between the SAM and cEOF1 in the troposphere is explained entirely by the zonally symmetric component of the SAM as shown by the high correlation with the S-SAM below 100 hPa and the low and statistically non-significant correlations between the A-SAM and either the Real or Imaginary cEOF1.
-In the stratosphere, the Real cEOF1 is correlated with both A-SAM and S-SAM, while the Imaginary cEOF1 is highly correlated only with the A-SAM. 
+The SAM index is statistically significantly correlated with the 0º cEOF1 in all levels, and with the 90º cEOF1 and 90º cEOF2 in the troposphere.
+On the other hand, correlations between SAM and the 0º cEOF2 are non-significant.
+
+The relationship between the SAM and cEOF1 in the troposphere is explained entirely by the zonally symmetric component of the SAM as shown by the high correlation with the S-SAM below 100 hPa and the low and statistically non-significant correlations between the A-SAM and either the 0º or 90º cEOF1.
+In the stratosphere, the 0º cEOF1 is correlated with both A-SAM and S-SAM, while the 90º cEOF1 is highly correlated only with the A-SAM. 
 These correlations are consistent with the regression maps of geopotential height in Figure\ \@ref(fig:eof1-regr-gh) and their comparison with those obtained for SAM, A-SAM and S-SAM by @campitelli2022.
 
-In the case of Imaginary cEOF2, its correlation with the SAM for the troposphere is associated to the asymmetric variability of the SAM. 
-Indeed, the Imaginary cEOF2 shares up to `r scales::percent(max_r2[cEOF == "cEOF2" & term == "asym"]$estimate^2, 2)` variance with the A-SAM and only `r scales::percent(max_r2[cEOF == "cEOF2" & term == "sym"]$estimate^2, 2)` at most with the S-SAM (Figure \@ref(fig:sam-eof-vertical).b2).
-Such extremely high correlation between A-SAM and Imaginary cEOF2 suggests that the modes obtained in this work are able to characterise the zonally asymmetric component of the SAM described previously by @campitelli2022.
+In the case of 90º cEOF2, its correlation with the SAM for the troposphere is associated to the asymmetric variability of the SAM. 
+Indeed, the 90º cEOF2 shares up to `r scales::percent(max_r2[cEOF == "cEOF2" & term == "asym"]$estimate^2, 2)` variance with the A-SAM and only `r scales::percent(max_r2[cEOF == "cEOF2" & term == "sym"]$estimate^2, 2)` at most with the S-SAM (Figure \@ref(fig:sam-eof-vertical).b2).
+Such extremely high correlation between A-SAM and 90º cEOF2 suggests that the modes obtained in this work are able to characterise the zonally asymmetric component of the SAM described previously by @campitelli2022.
 
@@
 angles_regr <- -c(0, 45, 90, 45+90)*pi/180
-angles_labs <- setNames(c("Real", "45º", "Imaginary", "135º"), angles_regr*180/pi)
+angles_labs <- setNames(c("0º", "45º", "90º", "135º"), angles_regr*180/pi)
 
@@
 ```{r plot_sst_psi}
-plot_sst_psi <- function(sst_regr, psi_regr) {
+plot_sst_psi <- function(sst_regr, psi_regr, min.mag = 0.002) {
   sst_gdata <- sst_regr %>%
@@
     #                 arrow.angle = 14, arrow.length = 0.2) +
-    geom_vector(aes(dx = fx, dy = fy), skip = 3, min.mag = 0.002, 
+    geom_vector(aes(dx = fx, dy = fy), skip = 3, min.mag = min.mag, 
                 size = 0.1) +
@@
 
-The connections between cEOFs and Tropical sources of variability were also assessed.
+The connections between cEOFs and tropical sources of variability were also assessed.
 Figure \@ref(fig:sst-psi-2) shows the regression maps of Sea Surface Temperatures (SST) and streamfunction anomalies at 200 hPa respectively upon standardised cEOF2.
-Besides the regression maps for the Real and Imaginary components, we include the corresponding regressions for two intermediate directions (corresponding to 45° and 135°).
-
-The Imaginary cEOF2 (second row) is associated with strong positive SST anomalies on the Central Pacific and negative anomalies over an area across northern Australia,  New Zealand the South Pacific Convergence Zone (SPCZ) (Figure \@ref(fig:sst-psi-2).b1).
+Besides the regression maps for the 0º and 90º phases, we include the corresponding regressions for two intermediate directions (corresponding to 45° and 135°).
+
+The 90º cEOF2 (second row) is associated with strong positive SST anomalies on the Central to Eastern Pacific and negative anomalies over an area across northern Australia,  New Zealand the South Pacific Convergence Zone (SPCZ) (Figure \@ref(fig:sst-psi-2).b1).
 The regression field of SST anomalies bears a strong resemblance with canonically positive ENSO [@bamston1997].
-Indeed, there is a significant and very high correlation (`r enso_cor[cEOF == "cEOF2" & part == "Imaginary"]$text`) between the ONI and the Imaginary cEOF2 time series.
+Indeed, there is a significant and very high correlation (`r enso_cor[cEOF == "cEOF2" & part == "Imaginary"]$text`) between the ONI and the 90º cEOF2 time series.
 Besides the Pacific ENSO-like pattern, there are positive anomalies in the western Indian Ocean and negative values in the eastern Indian Ocean, resembling a positive IOD [@saji1999].
-Consistently, the correlation between the Imaginary cEOF2 and the DMI is `r dmi_cor[cEOF == "cEOF2" & part == "Imaginary"]$text`.
-
-The Imaginary cEOF2 is associated with strong wave-like streamfunction anomalies emanating from the tropics (Figure\ \@ref(fig:sst-psi-2).b2), both from the Central Pacific sector and the Indian Ocean.
-The atmospheric response associated with Imaginary cEOF2 is then consistent with the combined effect of ENSO and the IOD on the extratropics: with SST anomalies inducing anomalous tropical convection that in turn excite Rossby waves propagating meridionally towards higher latitudes [@mo2000; @cai2011; @nuncio2015].
+Consistently, the correlation between the 90º cEOF2 and the DMI is `r dmi_cor[cEOF == "cEOF2" & part == "Imaginary"]$text`.
+
+The 90º cEOF2 is associated with strong wave-like streamfunction anomalies emanating from the tropics (Figure\ \@ref(fig:sst-psi-2).b2), both from the Central Pacific sector and the Indian Ocean. 
+The atmospheric response associated with 90º cEOF2 is then consistent with the combined effect of ENSO and the IOD on the extratropics: with SST anomalies inducing anomalous tropical convection that in turn excite Rossby waves propagating meridionally towards higher latitudes [@mo2000; @cai2011; @nuncio2015].
 
 However, the cEOF2 is not associated with the same tropical SST anomaly patterns at all their phases
-Figure\ \@ref(fig:sst-psi-2).d1 and d2 show that the Real cEOF2 is not associated either with any significant SST nor streamfunction anomalies in the tropics. 
-As a result, the correlation between the Real cEOF2 and ENSO is also not significant  (`r enso_cor[cEOF == "cEOF2" & part == "Real"]$text`).
+Figure\ \@ref(fig:sst-psi-2).d1 and d2 show that the 0º cEOF2 is not associated either with any significant SST nor streamfunction anomalies in the tropics. 
+As a result, the correlation between the 0º cEOF2 and ENSO is also not significant  (`r enso_cor[cEOF == "cEOF2" & part == "Real"]$text`).
 Meanwhile, Rows a and c in Fig.\ \@ref(fig:sst-psi-2) show that the intermediate phases are still associated with significant SST regressed anomalies over the Pacific Ocean, but at slightly different locations.
@@
   geom_vline(xintercept = 0, size = 0.2, color = "gray50") +
-  annotate(shadowtext:::GeomShadowText,
-           label = c("- Real", "- Imaginary", "+ Real", "+ Imaginary", "- Real"),
-           color = "gray60", bg.colour = "white", bg.r = 0.2,
-           size = 3,
-           y = -Inf, x = c(-pi, -pi/2, 0, pi/2, pi), angle = 90, hjust = 0) +
+  # annotate(shadowtext:::GeomShadowText,
+  #          label = c("- Real", "- Imaginary", "+ Real", "+ Imaginary", "- Real"),
+  #          color = "gray60", bg.colour = "white", bg.r = 0.2,
+  #          size = 3,
+  #          y = -Inf, x = c(-pi, -pi/2, 0, pi/2, pi), angle = 90, hjust = 0) +
 
@@
 
-To further explore the relationship between tropical forcing and phases of the cEOF2, Figure\ \@ref(fig:enso-phase) shows the ONI index plotted against the cEOF2 phase for each SON trimester between 1979 and 2019, highlighting years in which the magnitude of cEOF2 is above the median.
-In years with positive (negative) ONI, the cEOF2 phase is mostly around 90º (-90°),  corresponding with positive (negative) Imaginary component.
+To further explore the relationship between tropical forcing and phases of the cEOF2, Figure\ \@ref(fig:enso-phase) shows the ONI plotted against the cEOF2 phase for each SON trimester between 1979 and 2019, highlighting years in which the magnitude of cEOF2 is above the median.
+In years with positive (negative) ONI, the cEOF2 phase is mostly around 90º (-90°).
 In the neutral ENSO seasons, the cEOF2 phase is much more variable.
@@
 
-The correlation between the absolute magnitude of the ONI index and the  cEOF2 amplitude is `r cor_enso_mag$text`. 
+The correlation between the absolute magnitude of the ONI and the  cEOF2 amplitude is `r cor_enso_mag$text`. 
 However, this relationship is mostly driven by the three years with strongest ENSO events in the period (`r knitr::combine_words(year(gdata[order(-oni)][1:3]$time))`) which coincide with the three years with strongest cEOF2 magnitude (not shown).
@@
 In the former case, the wave train has little phase preference.
-However, when cEOF2 is excited by tropical SST variability, it tends to remain locked to the imaginary phase.
-This explains the relative over-abundance of years with cEOF2 near positive and negative imaginary phase in Figure\ \@ref(fig:phase-histogram).
-
-Unlike the cEOF2 case, there are no significant  SST regressed anomalies associated with either the Real or Imaginary cEOF1 (Sup. Figure\ \@ref(fig:sst-psi-1)). 
+However, when cEOF2 is excited by tropical SST variability, it tends to remain locked to the 90º phase.
+This explains the relative over-abundance of years with cEOF2 near positive and negative 90º phase in Figure\ \@ref(fig:phase-histogram).
+
+Unlike the cEOF2 case, there are no significant SST regressed anomalies associated with either the 0º or 90º cEOF1 (Sup. Figure\ \@ref(fig:sst-psi-1)). 
 Consistently, streamfunction anomalies do not show any tropical influence. 
-Instead, the real and Imaginary cEOF1 are associated with zonally wave activity fluxes in the extra-tropics around 60ºS, except for an equatorward flow from the coast of Antarctica around 150ºE in the Real component.
+Instead, the 0º and 90º cEOF1 are associated with zonally propagating wave activity fluxes in the extra-tropics around 60ºS, except for an equatorward flow from the coast of Antarctica around 150ºE in the 0º phase.
 This suggests that the variability of cEOF1 is driven primary by the internal variability of the extra-tropics.
@@
 ```{r pp-t2m-r2, fig.cap = "(ref:cap-pp-t2m-r2)", fig.width=5, fig.height=3}
-breaks <- seq(.2, 1, by = .1)
+breaks <- seq(.1, 1, by = .1)
 
@@
 The influence of cEOFs variability in the anomalies of both 2-metre air temperature and precipitation in the SH was also explored.
-Figure\ \@ref(fig:pp-t2m-r2) shows the 2-meter temperature or precipitation anomalies explained variance by the multiple linear model of both Real and Imaginary cEOF1 components (column 1), and both Real  and Imaginary cEOF2 components (column 2).
+Figure\ \@ref(fig:pp-t2m-r2) shows the 2-meter temperature or precipitation anomalies explained variance by the multiple linear model of both 0º and 90º cEOF1 (column 1), and both 0º and 90º cEOF2 (column 2).
 The variance explained by cEOF1 for precipitation anomalies is extremely low and also for most of the temperature anomalies, except for the northern tip of the Antarctic Peninsula, northern Weddell Sea and the Ross Sea coast (Fig.\  \@ref(fig:pp-t2m-r2)a.1).
+
+This lack of strong relationship between the cEOF1 and temperature and precipitation might be surprising considering the correlation between the cEOF1 and the SAM (Fig.
+\@ref(fig:sam-eof-vertical) column 1) and the correlation between SAM and temperature east and west of the Antarctic Peninsula, and with precipitation in western Australia (BUSCAR CITA).
+There a two main reasons for this. 
+First, the correlation between cEOF1 and the SAM in the troposphere is modest, with less than 50% of shared variance (Fig.
+\@ref(fig:sam-eof-vertical) column 1), so one need not expect these to indices to be equivalent.
+Second, on a previous paper [@campitelli2022] we showed that the strong relationship between the SAM and temperature anomalies around the Antarctic Peninsula is mainly due to the asymmetric part of the SAM.
+Meanwhile, the cEOF1 is significantly correlated only with the symmetric part of the SAM (Fig.
+\@ref(fig:sam-eof-vertical) column 1), which by itself is not significantly correlated with surface temperatures in that area.
 
@@
 
-Temperature anomalies associated with the Imaginary cEOF (Fig.\ \@ref(fig:pp-temp-2).b1) show positive values in the tropical Pacific, consistent with SSTs anomalies associated with the same phase (Fig.\ \@ref(fig:sst-psi-2).b1).
+Temperature anomalies associated with the 90º cEOF (Fig.\ \@ref(fig:pp-temp-2).b1) show positive values in the tropical Pacific, consistent with SSTs anomalies associated with the same phase (Fig.\ \@ref(fig:sst-psi-2).b1).
 
@@
 
-Over the continents, the Imaginary cEOF2 (Fig.\ \@ref(fig:pp-temp-2)b.1) is associated with positive regressed temperature anomalies in southern Australia and negative regressed anomalies in southern South America and the Antarctic Peninsula, that are a result of the wave train described before.
-
-The temperatures anomalies associated with the Real cEOF2 (Fig.\ \@ref(fig:pp-temp-2)d.1) are less extensive and restricted to mid and high latitudes.   
+Over the continents, the 90º cEOF2 (Fig.\ \@ref(fig:pp-temp-2)b.1) is associated with positive regressed temperature anomalies in southern Australia and negative regressed anomalies in southern South America and the Antarctic Peninsula, that are a result of the wave train described before.
+
+The temperatures anomalies associated with the 0º cEOF2 (Fig.\ \@ref(fig:pp-temp-2)d.1) are less extensive and restricted to mid and high latitudes.   
 Over the continents, the temperature anomalies regressions are non significant, except for positive anomalies near the Antarctic Peninsula.
 
-Tropical precipitation anomalies associated with the Imaginary cEOF2 are strong, with positive anomalies in the central Pacific and western Indian, and negative anomalies in the eastern Pacific (Fig.\ \@ref(fig:pp-temp-2)b.2).
+Tropical precipitation anomalies associated with the 90º cEOF2 are strong, with positive anomalies in the central Pacific and western Indian, and negative anomalies in the eastern Pacific (Fig.\ \@ref(fig:pp-temp-2)b.2).
 This field is consistent with the SST regression map (Fig.\ \@ref(fig:pp-temp-2)b.1) as the positive SST anomalies enhance tropical convection and the negative SST anomalies inhibits it.
 
-In the extra-tropics, the positive Imaginary cEOF2 is related to drier conditions over eastern Australia and the surrounding ocean, that it is similar signal as the one associated with ENSO [@cai2011].
-However, the Imaginary cEOF2 is not the direction most correlated with precipitation in that area. 
-The 135º phase (an intermediate between positive Imaginary and negative Real cEOF2) component is associated with stronger and more extensive temporal correlations  with precipitation over Australia and New Zealand.
+In the extra-tropics, the positive 90º cEOF2 is related to drier conditions over eastern Australia and the surrounding ocean, that it is similar signal as the one associated with ENSO [@cai2011].
+However, the 90º cEOF2 is not the direction most correlated with precipitation in that area. 
+The 135º phase (an intermediate between positive 90º and 180º cEOF2) component is associated with stronger and more extensive temporal correlations  with precipitation over Australia and New Zealand. 
 The influence of cEOF2 in Australian precipitation could be more related to the direct impacts of SST anomalies in the surrounded oceans rather than on the interconnection pattern represented by the cEOF2.
 
-Over South America, the Imaginary cEOF2 has positive correlations with precipitation in South Eastern South America (SESA) and central Chile, and negative correlations in eastern Brazil.
+Over South America, the 90º cEOF2 has positive correlations with precipitation in South Eastern South America (SESA) and central Chile, and negative correlations in eastern Brazil.
 This correlation field  matches the springtime precipitation signature of ENSO [e.g.
@@
 
-This result is not surprising considering the close relationship of the Imaginary cEOF2 with both ONI and A-SAM index, which was shown previously.  
+This result is not surprising considering the close relationship of the 90º cEOF2 with both ONI and A-SAM index, which was shown previously.  
 Furthermore, it consolidates the identification of the cEOF2 with the PSA pattern.
@@
 
-The correlation coefficients between precipitation anomalies and the Real cEOF2 (Fig.\ \@ref(fig:pp-temp-2)d.2) are weaker than for Imaginary cEOF2.
+The correlation coefficients between precipitation anomalies and the 0º cEOF2 (Fig.\ \@ref(fig:pp-temp-2)d.2) are weaker than for 90º cEOF2.
 There is a residual positive correlation in the equatorial eastern Pacific and small, not statistically significant positive correlations over eastern Australia and negative ones over New Zealand.
@@
 
-The cEOF1 represents the variability of the zonal wave 1 in the stratosphere and is closely related to stratospheric variability such as anomalies in Total Ozone Column.
+The cEOF1 represents the variability of the zonal wave 1 in the stratosphere and is closely related to stratospheric variability such as anomalies in Total Column Ozone.
 Otherwise, this complex EOF is not related with SST variability and continental precipitation in the Southern Hemisphere.
 On the other hand, the cEOF2 represents a wave-3 pattern with maximum magnitude in the Pacific sector, that is an alternative representation of the PSA1 and PSA2 patterns (Mo and Paegle 2001).
-The Imaginary cEOF2 can be identified with the PSA1 and the Real cEOF2 with the PSA2.
+The 90º cEOF2 can be identified with the PSA1 and the 0º cEOF2 with the PSA2.
 
 While the cEOF2 variability could be related to surface impacts, the cEOF1 influence is almost negligible.
-For instance, precipitation anomalies in South America associated with the Imaginary cEOF2 show a clear ENSO-like impact, with positive anomalies in South-Eastern South America, negative anomalies in Southern Brazil and positive anomalies in central Chile for positive Imaginary cEOF2 phase.
+For instance, precipitation anomalies in South America associated with the 90º cEOF2 show a clear ENSO-like impact, with positive anomalies in South-Eastern South America, negative anomalies in Southern Brazil and positive anomalies in central Chile for positive 90º cEOF2 phase.
 
@@
 Secondly, because of how these components are constructed, they are not forced to be orthogonal to other modes of circulation, as in the standard EOF methodology. 
-This allows us to show for example, that the Imaginary cEOF2, that resembles PSA1 variability, is closely associated to the SAM in the troposphere. 
+This allows us to show for example, that the 90º cEOF2, that resembles PSA1 variability, is closely associated to the SAM in the troposphere. 
 Moreover, the high correlation of this component to the zonally asymmetric portion of the SAM allow us to speculate that the A-SAM might actually be a statistical contamination of the PSA mode on the symmetric SAM.
@@
 @irving2016 argued that there is some disagreement in the literature of whether the phase of the PSA pattern is affected by the location of tropical SST anomalies.
-With the methodology used in this study, we were able to show not only that the cEOF2 tends to be in the positive or negative imaginary phase (\~PSA1) when the ENSO region is warm or cold, respectively, but also that central Pacific SST anomalies tend to align the cEOF towards the negative real phase and eastern Pacific SST anomalies tend to align it towards positive real phase.
+With the methodology used in this study, we were able to show not only that the cEOF2 tends to be in the positive or negative 90º phase (\~PSA1) when the ENSO region is warm or cold, respectively, but also that central Pacific SST anomalies tend to align the cEOF towards the negative 0º phase and eastern Pacific SST anomalies tend to align it towards positive 0º phase.
 When ENSO phase is neutral, the cEOF2 is still as active, but with no preferred phase.
@@
 
-## Availability of data and material {.unnumbered}
+\backmatter
+
+# Declarations {.unnumbered}
+
+## Funding {.unnumbered}
+
+The research was supported by UBACyT20020170100428BA and the CLIMAX Project funded by Belmont Forum/ANR-15-JCL/-0002-01. Elio Campitelli was supported by a PhD grant from CONICET, Argentina.
+
+## Conflict of interest/Competing interests  {.unnumbered}
+
+The authors have no relevant financial or non-financial interests to disclose.
+
+## Ethics approval {.unnumbered}
+
+Not applicable.
+
+## Consent to participate {.unnumbered}
+
+Not applicable.
+
+## Consent for publication {.unnumbered}
+
+Not applicable.
+
+## Availability of data and materials {.unnumbered}
 
@@
 
-\backmatter
-
-\bmhead{Acknowledgments}
-
-The research was supported by UBACyT20020170100428BA and the CLIMAX Project funded by Belmont Forum/ANR-15-JCL/-0002-01. Elio Campitelli was supported by a PhD grant from CONICET, Argentina.
+##  Authors' contributions {.unnumbered}
+
+E.C. and L.D. wrote the main manuscript text. E.C. prepared the figures. All authors reviewed the manuscript.
 
@@
 ```{r sst-psi-1, fig.cap = "(ref:sst-psi-1-cap)", fig.width = 5, fig.height=6, dependson="plot_sst_psi", fig.env = "figure*"}
-plot_sst_psi(sst_regr[cEOF == "cEOF1"], psi_regr[cEOF == "cEOF1"]) 
+plot_sst_psi(sst_regr[cEOF == "cEOF1"], psi_regr[cEOF == "cEOF1"], min.mag = 0.0005) 
 ```

### 2022-12-21 analysis/paper/shceof.tex
Actualiza

@@
 
-We restrict this paper to the spring September-October-November (SON) trimester because this season experiences a maximum in tropical teleconnections over South America \citep{cazes-boezio2003} and SH stratosphere-troposphere interactions \citep{lim2018}.
+We restrict this paper to the spring September-October-November (SON) trimester because during this season both the tropical teleconnections over South America \citep{cazes-boezio2003} and the SH stratosphere-troposphere interactions \citep{lim2018} are at their maximum amplitude.
 
@@
 In Section~\ref{regressions} we study the spatial regressions with geopotential height, temperature and ozone anomalies.
-In Section~\ref{psa} and \ref{sam} we analyse the relationship between cEOF2 the PSA and SAM modes.
+In Section~\ref{psa} and \ref{sam} we analyse the relationship between cEOF2, the PSA and SAM modes.
 In Section~\ref{tropical} we study tropical forcings that explain the variability of each cEOF.
@@
 
-We used monthly geopotential height, air temperature, ozone mixing ratio, and total ozone column (TOC) at 2.5\degree longitude by 2.5\degree latitude of horizontal resolution and 37 vertical isobaric levels from the European Centre for Medium-Range Weather Forecasts Reanalysis version 5 {[}ERA; \citet{era5}{]} for the period 1979 -- 2019.
+We used monthly geopotential height, air temperature, ozone mixing ratio, and total column ozone (TOC) at 2.5\degree longitude by 2.5\degree latitude of horizontal resolution and 37 vertical isobaric levels from the European Centre for Medium-Range Weather Forecasts Reanalysis version 5 {[}ERA; \citet{era5}{]} for the period 1979 -- 2019.
 Most of our analysis is restricted to the post-satellite era to avoid any confounding factors arising from the incorporation of satellite observations, but we also used the preliminary back extension of ERA5 from 1950 to 1978 \citep{era5be} to describe long-term trends.
@@
 We further split the SAM into its zonally symmetric and zonally asymmetric components (S-SAM and A-SAM indices respectively).
-These indices were obtained by projecting the zonally asymmetric and zonally symmetric part of the SAM spatial pattern onto monthly geopotential height fields, as proposed by \citet{campitelli2022}.
-Seasonal indices of the PSA patterns (PSA1 and PSA2) were calculated, in agreement with \citet{mo2001}, as the third and fourth leading EOF of seasonal mean anomalies for 500-hPa geopotential heights at SH.
+These indices were obtained by projecting monthly geopotential height fields onto the zonally asymmetric and zonally symmetric parts of the SAM spatial pattern, as proposed by \citet{campitelli2022}.
+Seasonal indices of the PSA patterns (PSA1 and PSA2) were calculated, in agreement with \citet{mo2001}, as the third and fourth EOFs of seasonal mean anomalies for 500-hPa geopotential heights at SH.
 
@@
 
-In the standard EOF analysis, zonal waves may appear as pairs of (possibly degenerated) EOFs representing similar patterns but shifted in phase \citep{horel1984}.
+In the standard EOF analysis, zonal waves may appear as pairs of (possibly degenerate) EOFs representing similar patterns but shifted in phase \citep{horel1984}.
 Figure \ref{fig:eof-naive} shows the four leading EOFs of SON geopotential height zonal anomalies at 50 hPa south of 20\degree S.
-It is clear that the first two EOFs represent a single phase-varying zonal wave 1 pattern and the last two represent a similarly phase-varying pattern with shorter wavenumber and four centres of action shifted by 1/4 wavelength.
-A similar EOF structure can be seen in 200 hPa (not shown).
-Since each pair of EOFs seems to represent the same phase-varying structure, it would be desirable to combine them into a single pattern described by amplitude and phase.
+It is clear that the first two EOFs represent a single phase-varying zonal wave 1 pattern and the last two represent a similarly phase-varying pattern with higher wavenumber and three centres of action shifted by 1/4 wavelength (90\degree in frequency space).
 
-Complex Empirical Orthogonal Functions (cEOF) are a useful method for characterising zonal waves, associated with phase-varying structures \citep{horel1984}.
-This method involves computing EOF on the analytic representation of the original field.
-That representation is a complex field in which the real part is the original data and the imaginary part is the original data shifted by 90\degree at each spectral frequency -- i.e.~its Hilbert transform.
+To describe the phase-varying nature of these two waves one could combine each pair into indices of amplitude of phase.
+So, for instance, the amplitude of the wave 1 EOF could be measured as \(\sqrt{\mathrm{EOF1}^2 + \mathrm{EOF2}^2}\) and it's phase as \(\tan^{-1} \left ( \frac{\mathrm{EOF2}}{\mathrm{EOF1}} \right )\) (where \(\mathrm{EOF1}\) and \(\mathrm{EOF2}\) are the time series associated with each EOF).
+However, this rests upon visual inspection of the spatial patterns and only works if both phases appear cleanly in different EOFs, which is not guaranteed.
+In particular, this doesn't work with the wave 1 pattern observed as the leading EOF in 200hPa geopotential heihgt zonal anomalies (not shown).
+
+In Complex Empirical Orthogonal Functions (cEOF) analysis, each cEOF is represented by two spatial patterns that are shifted by 1/4 wavelength by construction.
+The actual field reconstructed by each cEOF is the linear combination of the two spatial fields weighted by its respective time series.
+This is analogous to how any sine wave with any amplitude and phase can be constructed by the sum of a sine wave and cosine wave with different amplitude but constant phase.
+This means that cEOFs naturally represent phase-varying wave-like patterns that change location as well as amplitude.
+
+For example, Figure \ref{fig:ceofs-1} a.1 shows the two spatial patterns of a cEOF, one with shaded contours (which we call the 0º phase) and the other with black contours (which we call the 90º phase).
+This cEOF is very similar to the two leading EOFs shown in Figure \ref{fig:eof-naive} and represents a zonal wave 1 pattern.
+When the phase of the wave matches the 0º phase, then the 0º phase time series will be positive and the 90º phase time series will be zero.
+Similarly, when the phase of the wave matches the 90º phase, the 90º phase time series will be positive and the 0º phase time series will be zero.\\
+Cases will have non zero values in both time series will represent intermediate phases.
+
+These two phases are not special and the same cEOF could be represented by any two linear combinations of the phases shown in Figure \ref{fig:ceofs-1} a.1.
+This parallels traditional EOFs, which are defined up to an additive constant.
+For real EOFs, this constant can be either 0 or \(\pi\), corresponding to a change in sign.
+For cEOFs, it can be any real number between 0 and \(2\pi\) \citep{horel1984}, corresponding to rotations in the complex plane.
+
+Compute cEOFs are computed the same way as traditional EOFs except that the data is first augmented by computing its analytic signal.
+This is a complex number whose real part is the original series and the imaginary part is the original data shifted by 90\degree at each spectral frequency -- i.e.~its Hilbert transform.
+Each cEOF is a set of complex-valued spatial patterns and time series in which the real and imaginary parts are the two orthogonal phases of the wave-like pattern.
+
+In this paper we use the term 0º cEOF and 90º cEOF to refer to each part of the whole cEOF.
+
 The Hilbert transform is usually understood in terms of time-varying signal.
@@
 
-The result of the cEOF methodology is a set of complex spatial patterns and complex time series.
-The real and imaginary part of each spatial pattern represent the two phases of wave-like spatial patterns that are in quadrature.
-The magnitude and argument of each complex time series represent the amplitude and phase of each zonal wave.
-
 \begin{table}
 
-\caption{\label{tab:corr-ceof-splitted}Coefficient of determination (\(R^2\)) between the absolute magnitude of complex EOFs computed separately at 200 hPa and 50 hPa (p-values lower than 0.01 in bold).}
+\caption{\label{tab:corr-ceof-splitted}Coefficient of determination (\(r^2\)) between the time series of the absolute magnitude of complex EOFs computed separately at 200 hPa and 50 hPa (p-values lower than 0.01 in bold).}
 \centering
@@
 
-The cEOF methodology is applied to SON geopotential height zonal anomalies south of 20\degree S separately at 50 and 200 hPa.
+The cEOF methodology is applied to zonal anomalies of SON geopotential height south of 20\degree S separately at 50 and 200 hPa.
 Table \ref{tab:corr-ceof-splitted} shows the coefficient of determination between time series of the amplitude of each cEOF across levels.
@@
 
-The phase of cEOFs is defined up to an additive constant.
-For real EOFs, this constant can be either 0 or \(\pi\), corresponding to a change in sign.
-For cEOFs, it can be any real number between 0 and \(2\pi\) \citep{horel1984}, corresponding to rotations in the complex plane.
-Since any choice is arbitrary and equally valid, we chose the phase of each cEOF so that the real or imaginary parts are aligned with meaningful phases in our analysis.
+As mentioned before, the choice of phases is arbitrary and equally valid.
+But to make the interpretation easier, we chose the phase of each cEOF so that the either the º0 cEOF or the 90º cEOF is aligned with meaningful variables in our analysis.
 This procedure does not create spurious correlations, it only takes an existing relationship and aligns it with a specific phase.
 
-Preliminary analysis showed that the first cEOF was closely related to the zonal wave 1 of Total Ozone Column and the second cEOF was closely related to ENSO.
-Therefore, to aid in the interpretation we chose the phase of cEOF1 so that the time series corresponding to the real part has the maximum correlation with the zonal wave 1 of Total Ozone Column between 75°S and 45°S.
-Similarly, we chose the phase of cEOF2 so that the coefficient of determination between the Oceanic Niño Index \citep{bamston1997} and the real part is minimised, which also nearly maximises the correlation with the imaginary part.
+In Section \ref{precipitation} we show regressions of precipitation and temperature associated with intermediate phases.
+For those plots, we rotated the cEOFs by 1/4 wavelength by multiplying the complex time series by \(\cos(\pi/4) + i\sin(\pi/4)\) and computing the regression on those rotated timeseries.
+
+Preliminary analysis showed that the first cEOF was closely related to the zonal wave 1 of Total Column Ozone and the second cEOF was closely related to ENSO.
+Therefore, we chose the phase of cEOF1 so that the time series corresponding to the 0ª cEOF1 has the maximum correlation with the zonal wave 1 of Total Column Ozone between 75°S and 45°S.
+Similarly, we chose the phase of cEOF2 so that the coefficient of determination between the Oceanic Niño Index \citep{bamston1997} and the º0 cEOF2 is minimised, which also nearly maximises the correlation with the 90º cEOF2.
 
@@
 We performed linear regressions to quantify the association between the cEOFs and other variables (e.g.~geopotential height, temperature, precipitation, and others).
-For each cEOF, we computed regression maps by fitting a multiple linear model involving both the real and the imaginary part.
-To obtain the linear coefficients of a variable \(X\) with the imaginary and real parts of each cEOF we fit the equation
+For each cEOF, we computed regression maps by fitting a multiple linear model involving both the 0º and the 90º phases.
+To obtain the linear coefficients of a variable \(X\) with the 0º and 90º phase of each cEOF we fit the equation
 
 \[
-X(\lambda, \phi, t) = \alpha(\lambda, \phi) \operatorname{Im}(\mathrm{cEOF}) + \beta(\lambda, \phi) \mathrm{Re}(\mathrm{cEOF}) + X_0(\lambda, \phi) + \epsilon(\lambda, \phi, t)
+X(\lambda, \phi, t) = \alpha(\lambda, \phi) \operatorname{cEOF_{0º}} + \beta(\lambda, \phi) \operatorname{cEOF_{90º}} + X_0(\lambda, \phi) + \epsilon(\lambda, \phi, t)
 \]
 
-where \(\lambda\) and \(\phi\) are the longitude and latitude, \(t\) is the time, \(\alpha\) and \(\beta\) are the linear regression coefficients for imaginary and real parts respectively, \(X_0\) and \(\epsilon\) are the constant and error terms respectively.
+where \(\lambda\) and \(\phi\) are the longitude and latitude, \(t\) is the time, \(\alpha\) and \(\beta\) are the linear regression coefficients for 0º and 90º phases respectively, \(X_0\) and \(\epsilon\) are the constant and error terms respectively.
 
@@
 \caption{\label{fig:ceofs-1}Spatial patterns for the two leading cEOFs of SON zonal anomalies of geopotential height at 50 hPa and 200 hPa for the 1979 -- 2019 period.
-The shading (contours) corresponds to real (imaginary) part.
+The shading (contours) corresponds to 0º (90º) phase.
 Arbitrary units.
@@
 \includegraphics{../figures/extended-series-1} \caption{Time series of the two leading cEOFs of SON zonal anomalies of geopotential height at 50 hPa and 200 hPa.
-cEOF1 (row a) and cEOF2 (row b) separated in their real (column 1) and imaginary (column 2) components.
+cEOF1 (row a) and cEOF2 (row b) separated in their 0º (column 1) and 90º (column 2) phase.
 Dark straight line is the linear trend.
@@
 The first mode (cEOF1) explains 82\% of the variance of the zonally anomalous fields, while the second mode (cEOF2) explains a smaller fraction (7\%).
-In the spatial patterns (Fig.~\ref{fig:ceofs-1}), the real and the imaginary components are in quadrature by construction, so that each cEOF describe a single wave-like pattern whose amplitude and position (i.e.~phase) is controlled by the magnitude and phase of the temporal cEOF.
+In the spatial patterns (Fig.~\ref{fig:ceofs-1}), the 0º and the 90º phases are in quadrature by construction, so that each cEOF describe a single wave-like pattern whose amplitude and position (i.e.~phase) is controlled by the magnitude and phase of the temporal cEOF.
 The wave patterns described by these cEOFs match the patterns seen in the standard EOFs of Figure~\ref{fig:eof-naive}.
@@
 The cEOF1 (Fig.~\ref{fig:ceofs-1} column~a) is a hemispheric wave 1 pattern with maximum amplitude at high latitudes.
-At 50 hPa the Real cEOF1 has the maximum of the wave 1 at 150ºE and at 200 hPa, the maximum is located at around 175ºE indicating a westerly shift in phase.
+At 50 hPa the 0º cEOF1 has the maximum of the wave 1 at 150ºE and at 200 hPa, the maximum is located at around 175ºE indicating a westward shift with height.
 The cEOF2 (Fig.~\ref{fig:eof-naive} column~b) shows also a zonal wave-like structure with maximum amplitude at high latitudes, but with shorter spatial scales.
 In particular, the dominant structure at both levels is a wave 3 but with larger amplitude in the pacific sector.
-This modulated amplitude is more evident at 200 hPa.
+This modulated amplitude is more evident at 200 hPa, which is consistent with the cEOF2 computed separatedly for 200 hPa explains a bit more variance than the cEOF2 computed separatedly for 50 hPa (11\% vs.~3\%, respectively).
 There is no apparent phase shift with height but the amplitude of the pattern is greatly reduced in the stratosphere, suggesting that this barotropic mode represents mainly tropospheric variability.
@@
 However, cEOF2 temporal mean is almost zero, which indicates that only cEOF1 includes variability that significantly projects onto the mean zonally anomalous field.
+This is consistent with the fact that the mean zonally anomalous field of geopotential height is very similar to the cEOF1 (\(r^2\) = 98\%) and not similar to the cEOF2 (\(r^2\) = 0\%).
 
-A significant positive trend in the real component of cEOF1 is evident (Fig.~\ref{fig:extended-series}a.1, p-value~= 0.0037) while there is no significant trend in any of the complex components of cEOF2.
-The positive trend in the Real cEOF1 translates into a positive trend in cEOF1 magnitude, but not systematic change in phase (not shown).
+A significant positive trend in the 0º phase of cEOF1 is evident (Fig.~\ref{fig:extended-series}a.1, p-value~= 0.0037) while there is no significant trend in any of the phases of cEOF2.
+The positive trend in the 0º cEOF1 translates into a positive trend in cEOF1 magnitude, but not systematic change in phase (not shown).
 This long-term change indicates an increase in the magnitude of the high latitude zonal wave 1.
-In agreement, \citet{raphael2003} detected a step change after around 1975 in the temporal evolution of leading EOF computed from August-September-October 500 hPa zonal geopotential height anomalies, which is similar to the cEOF1.
 
@@
 \includegraphics{../figures/eof1-regr-gh-1.pdf}
-\caption{\label{fig:eof1-regr-gh}Regression of SON geopotential height anomalies (\(m^2s^{-1}\)) with the (column 1) real and (column 2) imaginary parts of the first cEOF for the 1979 -- 2019 period at (row a) 50 hPa and (row b) 200 hPa.
-These coefficients come from multiple linear regression involving the real and imaginary parts.
+\caption{\label{fig:eof1-regr-gh}Regression of SON geopotential height anomalies (\(m^2s^{-1}\)) with the (column 1) 0º and (column 2) 90º phases of the first cEOF for the 1979 -- 2019 period at (row a) 50 hPa and (row b) 200 hPa.
+These coefficients come from multiple linear regression involving the 0º and 90º phases.
 Areas marked with dots have p-values smaller than 0.01 adjusted for False Detection Rate.}
@@
 Figure~\ref{fig:eof1-regr-gh} shows regression maps of SON geopotential height anomalies upon cEOF1.
-At 50 hPa (Figure~\ref{fig:eof1-regr-gh} row a), the Real cEOF1 is associated with a centre located over the Ross Sea.
-The Imaginary cEOF1 is associated with a distinctive wave 1 pattern with maximum over the coast of East Antarctica.
-At 200 hPa (Figure~\ref{fig:eof1-regr-gh} row b) the Real cEOF1 shows a single centre of positive anomalies spanning West Antarctica surrounded by opposite anomalies in lower latitudes, with is centre shifted slightly eastward compared with the upper-level anomalies.
-The Imaginary cEOF1 shows a much more zonally symmetrical pattern resembling the negative SAM phase \citep[e.g.][]{fogt2020}.
+At 50 hPa (Figure~\ref{fig:eof1-regr-gh} row a), the 0º cEOF1 is associated with a centre located over the Ross Sea.
+The 90º cEOF1 is associated with a distinctive wave 1 pattern with maximum over the coast of East Antarctica.
+At 200 hPa (Figure~\ref{fig:eof1-regr-gh} row b) the 0º cEOF1 shows a single centre of positive anomalies spanning West Antarctica surrounded by opposite anomalies in lower latitudes, with its centre shifted slightly eastward compared with the upper-level anomalies.
+The 90º cEOF1 shows a much more zonally symmetrical pattern resembling the negative SAM phase \citep[e.g.][]{fogt2020}.
 Therefore, the magnitude and phase of the cEOF1 are associated with the magnitude and phase of a zonal wave mainly in the stratosphere.
@@
 Figure~\ref{fig:eof2-regr-gh} shows the regression maps of geopotential height anomalies upon the cEOF2.
-In the troposphere (Fig.~\ref{fig:eof2-regr-gh} column b), wave trains similar to those identified for cEOF2 patterns (Fig~\ref{fig:ceofs-1}) can be distinguished.
-Regressed anomalies associated with the Real cEOF2 are 90º out of phase with those associated with the Imaginary cEOF2.
+In the troposphere (Fig.~\ref{fig:eof2-regr-gh} row b) the regression maps show wave trains similar to those identified for cEOF2 patterns (Fig~\ref{fig:ceofs-1}).
+Regressed anomalies associated with the 0º cEOF2 are 1/4 wavelength out of phase with those associated with the 90º cEOF2.
 All fields have a dominant zonal wave 3 limited to the western hemisphere, over the Pacific and Atlantic Oceans.
 cEOF2 then represents an equivalent barotropic wave train that is very similar to the the PSA Patterns \citep{mo2001}.
-Comparing the location of the positive anomaly near 90ºW in column b of Figure~\ref{fig:eof2-regr-gh} with Figures 1.a and b from \citet{mo2001}, the Real cEOF2 regression map could be identified with PSA2, while the Imaginary cEOF2 resembles PSA1.
+Comparing the location of the positive anomaly near 90ºW in column b of Figure~\ref{fig:eof2-regr-gh} with Figures 1.a and b from \citet{mo2001}, the 0º cEOF2 regression map could be identified with PSA2, while the 90º cEOF2 resembles PSA1.
 
-These wave patterns are also present in the stratosphere (Fig.~\ref{fig:eof2-regr-gh} column a) supporting their equivalent barotropic nature.
-But also present is a monopole over the pole with negative sign associated with the Real cEOF2 and positive sign associated with the Imaginary cEOF2.
-This monopole might indicate strengthening of the polar vortex associated with positive values of the Real cEOF2 and weakening associated with positive Imaginary cEOF2.
+These wave patterns are also present in the stratosphere (Fig.~\ref{fig:eof2-regr-gh} row a) supporting their equivalent barotropic nature.
+But also present is a monopole over the pole with negative sign associated with the 0º cEOF2 and positive sign associated with the 90º cEOF2.
+This monopole might indicate strengthening of the polar vortex associated with positive values of the 0º cEOF2 and weakening associated with negative values of 0º cEOF2.
 
@@
 \includegraphics{../figures/t-vertical-1.pdf}
-\caption{\label{fig:t-vertical}Regression of SON zonal anomalies averaged between 75°S and 45°S of mean air temperature (shaded, Kelvin) and ozone mixing ratio (contours, negative contours with dashed lines, labels in parts per billion by mass) with the (a) Real and (b) Imaginary components of the cEOF1 for the 1979 -- 2019 period.}
+\caption{\label{fig:t-vertical}Regression of SON zonal anomalies averaged between 75°S and 45°S of mean air temperature (shaded, Kelvin) and ozone mixing ratio (contours, negative contours with dashed lines, labels in parts per billion by mass) with the (a) 0º and (b) 90º phase of the cEOF1 for the 1979 -- 2019 period.}
 \end{figure}
@@
 The distribution of temperature regression coefficients at 50 hPa and at 200 hPa mirror the geopotential height regression maps at 50 hPa (Fig.~\ref{fig:eof1-regr-gh}).
-In both levels, the Real cEOF1 is associated with a positive centre over the South Pole with its centre moved slightly towards 150ºE (Fig.~\ref{fig:eof1-regr-t} column 1).
-On the other hand, the regression maps on the Imaginary cEOF1 show a more clear wave 1 pattern with its maximum around 60ºE.
+In both levels, the 0º cEOF1 is associated with a positive centre over the South Pole with its centre moved slightly towards 150ºE (Fig.~\ref{fig:eof1-regr-t} column 1).
+On the other hand, the regression maps on the 90º cEOF1 show a more clear wave 1 pattern with its maximum around 60ºE.
 
 Figure~\ref{fig:t-vertical} shows the vertical distribution of the regression coefficients on cEOF1 from zonal anomalies averaged between 75°S and 45°S of air temperature and of ozone mixing ratio.
-Temperature zonal anomalies associated with cEOF1 show a clear wave 1 pattern for both real and imaginary components throughout the atmosphere above 250 hPa with a sign reversal above 10 hPa.
+Temperature zonal anomalies associated with cEOF1 show a clear wave 1 pattern for both 0º and 90º phases throughout the atmosphere above 250 hPa with a sign reversal above 10 hPa.
 As a result of the hydrostatic balance, this is the level in which the geopotential anomaly have maximum amplitude (not shown).
@@
 The maximum ozone regressed anomalies match with the minimum temperature anomalies above 10 hPa and with the maximum temperature anomalies below 10 hPa (Fig.~\ref{fig:t-vertical}).
-Therefore, the ozone zonal wave 1 is anticorrelated with the temperature zonal wave 1 in the upper stratosphere, and directly correlated in the upper stratosphere.
+Therefore, the ozone zonal wave 1 is negatively correlated with the temperature zonal wave 1 in the upper stratosphere, and positively correlated in the upper stratosphere.
 This change in phase is observed in ozone anomalies forced by planetary waves that reach the stratosphere \citep{hartmann1979, wirth1993, smith1995}.
 In the photochemically-dominated upper stratosphere, cold temperatures inhibit the destruction of ozone, explaining the opposite behaviour for both variables.
-On the other hand, in the advectively-dominated lower stratosphere, ozone anomalies are 90º out of phase with horizontal and vertical transport, which are in addition 90º out of phase with temperature anomalies, resulting in same sign anomalies for the response of both variables \citep{hartmann1979, wirth1993, smith1995}.
+On the other hand, in the advectively-dominated lower stratosphere, ozone anomalies are 1/4 wavelength out of phase with horizontal and vertical transport, which are in addition 1/4 wavelength out of phase with temperature anomalies, resulting in same sign anomalies for the response of both variables \citep{hartmann1979, wirth1993, smith1995}.
 
@@
 \includegraphics{../figures/o3-regr-1.pdf}
-\caption{\label{fig:o3-regr}Regression of SON mean Total Ozone Column anomalies (shaded, Dobson Units) with the (a) real and (b) imaginary components of the cEOF1 for the 1979 -- 2019 period.
-On contours, the mean zonal anomaly of Total Ozone Column (negative contours in dashed lines, Dobson Units).
+\caption{\label{fig:o3-regr}Regression of SON mean Total Column Ozone anomalies (shaded, Dobson Units) with the (a) 0º and (b) 90º phases of the cEOF1 for the 1979 -- 2019 period.
+On contours, the mean zonal anomaly of Total Column Ozone (negative contours in dashed lines, Dobson Units).
 Areas marked with dots have p-values smaller than 0.01 adjusted for False Detection Rate.}
@@
 The climatological position of the springtime Ozone minimum (ozone hole) is outside the South Pole and towards the Weddell Sea \citep[e.g.][]{grytsai2011}.
-Thus, the Real cEOF1 regression field (Figure~\ref{fig:o3-regr}a) coincides with the climatological position of the ozone hole, while it is 90° out of phase for the Imaginary cEOF1.
+Thus, the 0º cEOF1 regression field (Figure~\ref{fig:o3-regr}a) coincides with the climatological position of the ozone hole, while it is 90° out of phase for the 90º cEOF1.
 The temporal correlation between the amplitudes of TOC planetary wave 1 and cEOF1 is 0.79 (CI: 0.63 -- 0.88), while the correlation between their phases is -0.85 (CI: -0.92 -- -0.74).
@@
 Given the similarity between the cEOF2 related anomaly structures (Fig.~\ref{fig:eof2-regr-gh}) and documented PSA patterns we study the relationship between them.
-Table~\ref{tab:psa-eof2} shows the correlations between the two PSA indices and the time series for Real and Imaginary phases of cEOF2.
-As visually anticipated by Figure~\ref{fig:eof2-regr-gh}, there is a large positive correlation between PSA1 and Imaginary cEOF2, and between PSA2 and Real cEOF2.
-On the other hand, there is no significant relationship between PSA1 and Real cEOF2, and between PSA2 and Imaginary cEOF2.
+Table~\ref{tab:psa-eof2} shows the correlations between the two PSA indices and the time series for 0º and 90º phases of cEOF2.
+As visually anticipated by Figure~\ref{fig:eof2-regr-gh}, there is a large positive correlation between PSA1 and 90º cEOF2, and between PSA2 and 0º cEOF2.
+On the other hand, there is no significant relationship between PSA1 and 0º cEOF2, and between PSA2 and 90º cEOF2.
 As a result, cEOF2 represents well both the spatial structure and temporal evolution of the PSA modes, so it is possible to make an association between its two phases and the two PSA modes.
-That is, the phase election for cEOF2 that maximises the relationship between ENSO and Imaginary cEOF2, also maximises the association between cEOF2 components and PSA modes (not shown).
+That is, the phase election for cEOF2 that maximises the relationship between ENSO and 90º cEOF2, also maximises the association between cEOF2 components and PSA modes (not shown).
 
@@
 
-Figure~\ref{fig:phase-histogram} shows an histogram that counts the number of SON seasons in which the cEOF2 phase was close to each of the four particular phases (positive/negative of real/imaginary component), with the observations for each season marked as rugs on the horizontal axis.
-In 62\% of seasons cEOF2 has a phase similar to either the negative or positive imaginary phase, making the imaginary phase the most common phase.
+Figure~\ref{fig:phase-histogram} shows an histogram that counts the number of SON seasons in which the cEOF2 phase was close to each of the four particular phases (positive/negative of 0º and 90º phases), with the observations for each season marked as rugs on the horizontal axis.
+In 62\% of seasons cEOF2 has a phase similar to either the negative or positive 90º phase, making the 90º phase the most common phase.
 This is also, by construction, the phase that is most correlated with ENSO.
 
-Therefore, by virtue of being the most common phase, the Imaginary cEOF2 explains more variance than the Real cEOF2.
-Conventional EOF analysis will therefore tend to separate them relatively cleanly, with the EOF representing the Imaginary cEOF2 always leading the one representing the Real cEOF2.
+Therefore, by virtue of being the most common phase, the 90º cEOF2 explains more variance than the 0º cEOF2.
+Conventional EOF analysis will therefore tend to separate them relatively cleanly, with the EOF representing the 90º cEOF2 always leading the one representing the 0º cEOF2.
 This phase preferences is in agreement with \citet{irving2016}, who found a bimodal distribution to PSA-like variability (compare our Figure~\ref{fig:phase-histogram} with their Figure 6).
@@
 We then computed the coefficient of determination between the cEOFs time series and the three SAM indices (SAM, A-SAM and S-SAM) defined by \citet{campitelli2022} at each vertical level (Fig.~\ref{fig:sam-eof-vertical}).
-The SAM index is statistically significant correlated with the Real cEOF1 in all levels, and with the Imaginary cEOF1 and Imaginary cEOF2 in the troposphere.
-On the other hand, correlations between SAM and the Real cEOF2 are non-significant.
+The SAM index is statistically significantly correlated with the 0º cEOF1 in all levels, and with the 90º cEOF1 and 90º cEOF2 in the troposphere.
+On the other hand, correlations between SAM and the 0º cEOF2 are non-significant.
 
-The relationship between the SAM and cEOF1 in the troposphere is explained entirely by the zonally symmetric component of the SAM as shown by the high correlation with the S-SAM below 100 hPa and the low and statistically non-significant correlations between the A-SAM and either the Real or Imaginary cEOF1.
-In the stratosphere, the Real cEOF1 is correlated with both A-SAM and S-SAM, while the Imaginary cEOF1 is highly correlated only with the A-SAM.
+The relationship between the SAM and cEOF1 in the troposphere is explained entirely by the zonally symmetric component of the SAM as shown by the high correlation with the S-SAM below 100 hPa and the low and statistically non-significant correlations between the A-SAM and either the 0º or 90º cEOF1.
+In the stratosphere, the 0º cEOF1 is correlated with both A-SAM and S-SAM, while the 90º cEOF1 is highly correlated only with the A-SAM.
 These correlations are consistent with the regression maps of geopotential height in Figure~\ref{fig:eof1-regr-gh} and their comparison with those obtained for SAM, A-SAM and S-SAM by \citet{campitelli2022}.
 
-In the case of Imaginary cEOF2, its correlation with the SAM for the troposphere is associated to the asymmetric variability of the SAM.
-Indeed, the Imaginary cEOF2 shares up to 92\% variance with the A-SAM and only 12\% at most with the S-SAM (Figure~\ref{fig:sam-eof-vertical}.b2).
-Such extremely high correlation between A-SAM and Imaginary cEOF2 suggests that the modes obtained in this work are able to characterise the zonally asymmetric component of the SAM described previously by \citet{campitelli2022}.
+In the case of 90º cEOF2, its correlation with the SAM for the troposphere is associated to the asymmetric variability of the SAM.
+Indeed, the 90º cEOF2 shares up to 92\% variance with the A-SAM and only 12\% at most with the S-SAM (Figure~\ref{fig:sam-eof-vertical}.b2).
+Such extremely high correlation between A-SAM and 90º cEOF2 suggests that the modes obtained in this work are able to characterise the zonally asymmetric component of the SAM described previously by \citet{campitelli2022}.
 
@@
 
-The connections between cEOFs and Tropical sources of variability were also assessed.
+The connections between cEOFs and tropical sources of variability were also assessed.
 Figure~\ref{fig:sst-psi-2} shows the regression maps of Sea Surface Temperatures (SST) and streamfunction anomalies at 200 hPa respectively upon standardised cEOF2.
-Besides the regression maps for the Real and Imaginary components, we include the corresponding regressions for two intermediate directions (corresponding to 45° and 135°).
+Besides the regression maps for the 0º and 90º phases, we include the corresponding regressions for two intermediate directions (corresponding to 45° and 135°).
 
-The Imaginary cEOF2 (second row) is associated with strong positive SST anomalies on the Central Pacific and negative anomalies over an area across northern Australia, New Zealand the South Pacific Convergence Zone (SPCZ) (Figure~\ref{fig:sst-psi-2}.b1).
+The 90º cEOF2 (second row) is associated with strong positive SST anomalies on the Central to Eastern Pacific and negative anomalies over an area across northern Australia, New Zealand the South Pacific Convergence Zone (SPCZ) (Figure~\ref{fig:sst-psi-2}.b1).
 The regression field of SST anomalies bears a strong resemblance with canonically positive ENSO \citep{bamston1997}.
-Indeed, there is a significant and very high correlation (0.76 (CI: 0.6 -- 0.87)) between the ONI and the Imaginary cEOF2 time series.
+Indeed, there is a significant and very high correlation (0.76 (CI: 0.6 -- 0.87)) between the ONI and the 90º cEOF2 time series.
 Besides the Pacific ENSO-like pattern, there are positive anomalies in the western Indian Ocean and negative values in the eastern Indian Ocean, resembling a positive IOD \citep{saji1999}.
-Consistently, the correlation between the Imaginary cEOF2 and the DMI is 0.62 (CI: 0.38 -- 0.78).
+Consistently, the correlation between the 90º cEOF2 and the DMI is 0.62 (CI: 0.38 -- 0.78).
 
-The Imaginary cEOF2 is associated with strong wave-like streamfunction anomalies emanating from the tropics (Figure~\ref{fig:sst-psi-2}.b2), both from the Central Pacific sector and the Indian Ocean.
-The atmospheric response associated with Imaginary cEOF2 is then consistent with the combined effect of ENSO and the IOD on the extratropics: with SST anomalies inducing anomalous tropical convection that in turn excite Rossby waves propagating meridionally towards higher latitudes \citep{mo2000, cai2011, nuncio2015}.
+The 90º cEOF2 is associated with strong wave-like streamfunction anomalies emanating from the tropics (Figure~\ref{fig:sst-psi-2}.b2), both from the Central Pacific sector and the Indian Ocean.
+The atmospheric response associated with 90º cEOF2 is then consistent with the combined effect of ENSO and the IOD on the extratropics: with SST anomalies inducing anomalous tropical convection that in turn excite Rossby waves propagating meridionally towards higher latitudes \citep{mo2000, cai2011, nuncio2015}.
 
 However, the cEOF2 is not associated with the same tropical SST anomaly patterns at all their phases
-Figure~\ref{fig:sst-psi-2}.d1 and d2 show that the Real cEOF2 is not associated either with any significant SST nor streamfunction anomalies in the tropics.
-As a result, the correlation between the Real cEOF2 and ENSO is also not significant (0 (CI: -0.3 -- 0.3)).
+Figure~\ref{fig:sst-psi-2}.d1 and d2 show that the 0º cEOF2 is not associated either with any significant SST nor streamfunction anomalies in the tropics.
+As a result, the correlation between the 0º cEOF2 and ENSO is also not significant (0 (CI: -0.3 -- 0.3)).
 Meanwhile, Rows a and c in Fig.~\ref{fig:sst-psi-2} show that the intermediate phases are still associated with significant SST regressed anomalies over the Pacific Ocean, but at slightly different locations.
@@
 
-To further explore the relationship between tropical forcing and phases of the cEOF2, Figure~\ref{fig:enso-phase} shows the ONI index plotted against the cEOF2 phase for each SON trimester between 1979 and 2019, highlighting years in which the magnitude of cEOF2 is above the median.
-In years with positive (negative) ONI, the cEOF2 phase is mostly around 90º (-90°), corresponding with positive (negative) Imaginary component.
+To further explore the relationship between tropical forcing and phases of the cEOF2, Figure~\ref{fig:enso-phase} shows the ONI plotted against the cEOF2 phase for each SON trimester between 1979 and 2019, highlighting years in which the magnitude of cEOF2 is above the median.
+In years with positive (negative) ONI, the cEOF2 phase is mostly around 90º (-90°).
 In the neutral ENSO seasons, the cEOF2 phase is much more variable.
@@
 
-The correlation between the absolute magnitude of the ONI index and the cEOF2 amplitude is 0.45 (CI: 0.17 -- 0.66).
+The correlation between the absolute magnitude of the ONI and the cEOF2 amplitude is 0.45 (CI: 0.17 -- 0.66).
 However, this relationship is mostly driven by the three years with strongest ENSO events in the period (2015, 1997, and 1982) which coincide with the three years with strongest cEOF2 magnitude (not shown).
@@
 In the former case, the wave train has little phase preference.
-However, when cEOF2 is excited by tropical SST variability, it tends to remain locked to the imaginary phase.
-This explains the relative over-abundance of years with cEOF2 near positive and negative imaginary phase in Figure~\ref{fig:phase-histogram}.
+However, when cEOF2 is excited by tropical SST variability, it tends to remain locked to the 90º phase.
+This explains the relative over-abundance of years with cEOF2 near positive and negative 90º phase in Figure~\ref{fig:phase-histogram}.
 
-Unlike the cEOF2 case, there are no significant SST regressed anomalies associated with either the Real or Imaginary cEOF1 (Sup. Figure~\ref{fig:sst-psi-1}).
+Unlike the cEOF2 case, there are no significant SST regressed anomalies associated with either the 0º or 90º cEOF1 (Sup. Figure~\ref{fig:sst-psi-1}).
 Consistently, streamfunction anomalies do not show any tropical influence.
-Instead, the real and Imaginary cEOF1 are associated with zonally wave activity fluxes in the extra-tropics around 60ºS, except for an equatorward flow from the coast of Antarctica around 150ºE in the Real component.
+Instead, the 0º and 90º cEOF1 are associated with zonally propagating wave activity fluxes in the extra-tropics around 60ºS, except for an equatorward flow from the coast of Antarctica around 150ºE in the 0º phase.
 This suggests that the variability of cEOF1 is driven primary by the internal variability of the extra-tropics.
@@
 The influence of cEOFs variability in the anomalies of both 2-metre air temperature and precipitation in the SH was also explored.
-Figure~\ref{fig:pp-t2m-r2} shows the 2-meter temperature or precipitation anomalies explained variance by the multiple linear model of both Real and Imaginary cEOF1 components (column 1), and both Real and Imaginary cEOF2 components (column 2).
+Figure~\ref{fig:pp-t2m-r2} shows the 2-meter temperature or precipitation anomalies explained variance by the multiple linear model of both 0º and 90º cEOF1 (column 1), and both 0º and 90º cEOF2 (column 2).
 The variance explained by cEOF1 for precipitation anomalies is extremely low and also for most of the temperature anomalies, except for the northern tip of the Antarctic Peninsula, northern Weddell Sea and the Ross Sea coast (Fig.~ \ref{fig:pp-t2m-r2}a.1).
+
+This lack of strong relationship between the cEOF1 and temperature and precipitation might be surprising considering the correlation between the cEOF1 and the SAM (Fig.
+\ref{fig:sam-eof-vertical} column 1) and the correlation between SAM and temperature east and west of the Antarctic Peninsula, and with precipitation in western Australia (BUSCAR CITA).
+There a two main reasons for this.
+First, the correlation between cEOF1 and the SAM in the troposphere is modest, with less than 50\% of shared variance (Fig.
+\ref{fig:sam-eof-vertical} column 1), so one need not expect these to indices to be equivalent.
+Second, on a previous paper \citep{campitelli2022} we showed that the strong relationship between the SAM and temperature anomalies around the Antarctic Peninsula is mainly due to the asymmetric part of the SAM.
+Meanwhile, the cEOF1 is significantly correlated only with the symmetric part of the SAM (Fig.
+\ref{fig:sam-eof-vertical} column 1), which by itself is not significantly correlated with surface temperatures in that area.
 
@@
 
-Temperature anomalies associated with the Imaginary cEOF (Fig.~\ref{fig:pp-temp-2}.b1) show positive values in the tropical Pacific, consistent with SSTs anomalies associated with the same phase (Fig.~\ref{fig:sst-psi-2}.b1).
+Temperature anomalies associated with the 90º cEOF (Fig.~\ref{fig:pp-temp-2}.b1) show positive values in the tropical Pacific, consistent with SSTs anomalies associated with the same phase (Fig.~\ref{fig:sst-psi-2}.b1).
 
@@
 
-Over the continents, the Imaginary cEOF2 (Fig.~\ref{fig:pp-temp-2}b.1) is associated with positive regressed temperature anomalies in southern Australia and negative regressed anomalies in southern South America and the Antarctic Peninsula, that are a result of the wave train described before.
+Over the continents, the 90º cEOF2 (Fig.~\ref{fig:pp-temp-2}b.1) is associated with positive regressed temperature anomalies in southern Australia and negative regressed anomalies in southern South America and the Antarctic Peninsula, that are a result of the wave train described before.
 
-The temperatures anomalies associated with the Real cEOF2 (Fig.~\ref{fig:pp-temp-2}d.1) are less extensive and restricted to mid and high latitudes.\\
+The temperatures anomalies associated with the 0º cEOF2 (Fig.~\ref{fig:pp-temp-2}d.1) are less extensive and restricted to mid and high latitudes.\\
 Over the continents, the temperature anomalies regressions are non significant, except for positive anomalies near the Antarctic Peninsula.
 
-Tropical precipitation anomalies associated with the Imaginary cEOF2 are strong, with positive anomalies in the central Pacific and western Indian, and negative anomalies in the eastern Pacific (Fig.~\ref{fig:pp-temp-2}b.2).
+Tropical precipitation anomalies associated with the 90º cEOF2 are strong, with positive anomalies in the central Pacific and western Indian, and negative anomalies in the eastern Pacific (Fig.~\ref{fig:pp-temp-2}b.2).
 This field is consistent with the SST regression map (Fig.~\ref{fig:pp-temp-2}b.1) as the positive SST anomalies enhance tropical convection and the negative SST anomalies inhibits it.
 
-In the extra-tropics, the positive Imaginary cEOF2 is related to drier conditions over eastern Australia and the surrounding ocean, that it is similar signal as the one associated with ENSO \citep{cai2011}.
-However, the Imaginary cEOF2 is not the direction most correlated with precipitation in that area.
-The 135º phase (an intermediate between positive Imaginary and negative Real cEOF2) component is associated with stronger and more extensive temporal correlations with precipitation over Australia and New Zealand.
+In the extra-tropics, the positive 90º cEOF2 is related to drier conditions over eastern Australia and the surrounding ocean, that it is similar signal as the one associated with ENSO \citep{cai2011}.
+However, the 90º cEOF2 is not the direction most correlated with precipitation in that area.
+The 135º phase (an intermediate between positive 90º and 180º cEOF2) component is associated with stronger and more extensive temporal correlations with precipitation over Australia and New Zealand.
 The influence of cEOF2 in Australian precipitation could be more related to the direct impacts of SST anomalies in the surrounded oceans rather than on the interconnection pattern represented by the cEOF2.
 
-Over South America, the Imaginary cEOF2 has positive correlations with precipitation in South Eastern South America (SESA) and central Chile, and negative correlations in eastern Brazil.
+Over South America, the 90º cEOF2 has positive correlations with precipitation in South Eastern South America (SESA) and central Chile, and negative correlations in eastern Brazil.
 This correlation field matches the springtime precipitation signature of ENSO \citep[e.g.][]{cai2020a} and it is also similar to the precipitation anomalies associated with the A-SAM \citep{campitelli2022}.
 
-This result is not surprising considering the close relationship of the Imaginary cEOF2 with both ONI and A-SAM index, which was shown previously.\\
+This result is not surprising considering the close relationship of the 90º cEOF2 with both ONI and A-SAM index, which was shown previously.\\
 Furthermore, it consolidates the identification of the cEOF2 with the PSA pattern.
@@
 
-The correlation coefficients between precipitation anomalies and the Real cEOF2 (Fig.~\ref{fig:pp-temp-2}d.2) are weaker than for Imaginary cEOF2.
+The correlation coefficients between precipitation anomalies and the 0º cEOF2 (Fig.~\ref{fig:pp-temp-2}d.2) are weaker than for 90º cEOF2.
 There is a residual positive correlation in the equatorial eastern Pacific and small, not statistically significant positive correlations over eastern Australia and negative ones over New Zealand.
@@
 Secondly, because of how these components are constructed, they are not forced to be orthogonal to other modes of circulation, as in the standard EOF methodology.
-This allows us to show for example, that the Imaginary cEOF2, that resembles PSA1 variability, is closely associated to the SAM in the troposphere.
+This allows us to show for example, that the 90º cEOF2, that resembles PSA1 variability, is closely associated to the SAM in the troposphere.
 Moreover, the high correlation of this component to the zonally asymmetric portion of the SAM allow us to speculate that the A-SAM might actually be a statistical contamination of the PSA mode on the symmetric SAM.
@@
 \citet{irving2016} argued that there is some disagreement in the literature of whether the phase of the PSA pattern is affected by the location of tropical SST anomalies.
-With the methodology used in this study, we were able to show not only that the cEOF2 tends to be in the positive or negative imaginary phase (\textasciitilde PSA1) when the ENSO region is warm or cold, respectively, but also that central Pacific SST anomalies tend to align the cEOF towards the negative real phase and eastern Pacific SST anomalies tend to align it towards positive real phase.
+With the methodology used in this study, we were able to show not only that the cEOF2 tends to be in the positive or negative 90º phase (\textasciitilde PSA1) when the ENSO region is warm or cold, respectively, but also that central Pacific SST anomalies tend to align the cEOF towards the negative 0º phase and eastern Pacific SST anomalies tend to align it towards positive 0º phase.
 When ENSO phase is neutral, the cEOF2 is still as active, but with no preferred phase.
@@
 
-\hypertarget{availability-of-data-and-material}{%
-\subsection*{Availability of data and material}\label{availability-of-data-and-material}}
-\addcontentsline{toc}{subsection}{Availability of data and material}
+\backmatter
+
+\hypertarget{declarations}{%
+\section*{Declarations}\label{declarations}}
+\addcontentsline{toc}{section}{Declarations}
+
+\hypertarget{funding}{%
+\subsection*{Funding}\label{funding}}
+\addcontentsline{toc}{subsection}{Funding}
+
+The research was supported by UBACyT20020170100428BA and the CLIMAX Project funded by Belmont Forum/ANR-15-JCL/-0002-01. Elio Campitelli was supported by a PhD grant from CONICET, Argentina.
+
+\hypertarget{conflict-of-interestcompeting-interests}{%
+\subsection*{Conflict of interest/Competing interests}\label{conflict-of-interestcompeting-interests}}
+\addcontentsline{toc}{subsection}{Conflict of interest/Competing interests}
+
+The authors have no relevant financial or non-financial interests to disclose.
+
+\hypertarget{ethics-approval}{%
+\subsection*{Ethics approval}\label{ethics-approval}}
+\addcontentsline{toc}{subsection}{Ethics approval}
+
+Not applicable.
+
+\hypertarget{consent-to-participate}{%
+\subsection*{Consent to participate}\label{consent-to-participate}}
+\addcontentsline{toc}{subsection}{Consent to participate}
+
+Not applicable.
+
+\hypertarget{consent-for-publication}{%
+\subsection*{Consent for publication}\label{consent-for-publication}}
+\addcontentsline{toc}{subsection}{Consent for publication}
+
+Not applicable.
+
+\hypertarget{availability-of-data-and-materials}{%
+\subsection*{Availability of data and materials}\label{availability-of-data-and-materials}}
+\addcontentsline{toc}{subsection}{Availability of data and materials}
 
@@
 
-\backmatter
+\hypertarget{authors-contributions}{%
+\subsection*{Authors' contributions}\label{authors-contributions}}
+\addcontentsline{toc}{subsection}{Authors' contributions}
 
-\bmhead{Acknowledgments}
-
-The research was supported by UBACyT20020170100428BA and the CLIMAX Project funded by Belmont Forum/ANR-15-JCL/-0002-01. Elio Campitelli was supported by a PhD grant from CONICET, Argentina.
+E.C. and L.D. wrote the main manuscript text. E.C. prepared the figures. All authors reviewed the manuscript.
 

### 2022-12-21 analysis/poster-congremet/figures.R

@@ -1,0 +1,557 @@
+base_dir <- here::here("analysis", "poster-congremet")
+
+source(file.path(base_dir, "data.R"))
+
+library(kableExtra)
+library(ggplot2)
+library(ggperiodic)
+library(tagger)
+sf::sf_use_s2(FALSE)
+# mgp <- latex2exp::TeX('Geopotential height ($m^2s^{-1}$)')
+mpg <- 'Altura geopotencial ($m^2s^{-1}$)'
+
+theme_set(theme_shceof(base_size = 20) +
+            theme(legend.box.spacing = grid::unit(-.5, "lines")) +
+            theme(plot.margin = margin(0, 0, 0, 0, "cm")))
+panel_background <- theme(panel.background = element_rect(fill = "#fbfbfb", color = NA),
+                          panel.ontop = FALSE,
+                          tagger.panel.tag.background = ggplot2::element_rect(color = NA,
+                                                                              fill = "#fbfbfb"))
+one_less <- function(x) {
+  n <- length(x)
+  if (n < 4) return(x)
+  x <- round(x, 5)
+
+  c( rev(JumpBy(rev(x[x < 0]), 2, fill = "")),
+     JumpBy(x[x > 0], 2, fill = ""))
+}
+no_grid <- theme(panel.grid = element_blank())
+guide_colorstrip_bottom <- function(width = 15, height = 0.5,
+                                    title.position = "top",...) {
+  guide_colorstrip(title.position = title.position, title.hjust = 0.5,
+                   barheight = height,
+                   barwidth = width, ...)
+}
+
+f <- formals(geom_contour_tanaka)
+f$range <-  c(0.01, 0.2)
+formals(geom_contour_tanaka) <- f
+
+guide_colorsteps_bottom <- function(width = 15, height = 0.5, even.steps = FALSE,
+                                    title.position = "top", show.limits = TRUE, ticks = TRUE,
+                                    ...) {
+  guide_colorsteps(show.limits = show.limits, even.steps = even.steps,
+                   title.position = title.position, title.hjust = 0.5,
+                   barheight = height, ticks = ticks,
+                   ticks.colour = "black",
+                   barwidth = width, frame.colour = "black", ...)
+}
+
+grid_y <- theme(panel.grid.major.y = element_line(color = "gray60", size = 0.1))
+
+geom_coords <- function() {
+
+  list(
[…]

### 2022-12-21 analysis/poster-icshmo/data.R

@@ -14,7 +14,7 @@ enso <- rsoi::download_oni(TRUE, data_path("raw", "oni.csv")) %>%
 
 
 sst <- ERSST() %>%
-  readRDS() %>%
+  fread() %>%
   .[season(time) == "SON"] %>%
   normalise_coords() %>%
   .[, w := season_weights(time)] %>%

### 2022-12-21 analysis/poster-sparc7/data.R

@@ -1,0 +1,365 @@
+library(metR)
+library(data.table)
+library(magrittr)
+library(shceof)
+
+sams <- SAM() %>%
+  fread() %>%
+  .[season(time) == "SON"] %>%
+  .[, estimate := estimate*season_weights(time)] %>%
+  .[, .(estimate = mean(estimate)),
+    by = .(lev, term, time = seasonally(time))]
+
+enso <- rsoi::download_oni(TRUE, data_path("raw", "oni.csv")) %>%
+  as.data.table() %>%
+  .[, .(time = lubridate::as_datetime(Date), oni = ONI)] %>%
+  na.omit() %>%
+  .[season(time) == "SON"] %>%
+  .[, oni := oni*season_weights(time)] %>%
+  .[, .(oni = mean(oni)), by = .(time = seasonally(time))]
+
+
+sst <- ERSST() %>%
+  fread() %>%
+  .[season(time) == "SON"] %>%
+  normalise_coords() %>%
+  .[, w := season_weights(time)] %>%
+  .[, .(t = mean(t*w)), by = .(lon, lat, time = seasonally(time))]
+
+x <- "irlba"
+has_irlba <- requireNamespace(x, quietly = TRUE)
+if (has_irlba) {
+  stop("If installed, metR::EOF uses irlba, but that package has a bug with complex matrices\n(see: https://github.com/bwlewis/irlba/issues/55). Uninstall irlba with renv::remove(\"irlba\") and try again.")
+}
+
+
+cmap <- ReadNetCDF(CMAP(), vars = c(pp = "precip"),
+                   subset = list(lat = -60:10)) %>%
+  normalise_coords() %>%
+  .[, w := season_weights(time)] %>%
+  .[, .(pp = mean(pp*w)), by = .(lon, lat, time = seasonally(time))]
+
+t2m <- ERA5_T2M() %>%
+  ReadNetCDF() %>%
+  normalise_coords() %>%
+  .[, w := season_weights(time)] %>%
+  .[, .(t2m = mean(t2m*w)), by = .(lon, lat, time = seasonally(time))]
+
+era5 <- ERA5_geopotential() %>%
+  ReadNetCDF(vars = c(hgt = "z"),
+             subset = list(level = list(200, 50),
+                           latitude = -90:10)) %>%
[…]

### 2022-12-21 analysis/poster-sparc7/figures.R

@@ -1,0 +1,485 @@
+base_dir <- here::here("analysis", "poster-sparc7")
+
+source(file.path(base_dir, "data.R"))
+
+library(kableExtra)
+library(ggplot2)
+library(ggperiodic)
+library(tagger)
+sf::sf_use_s2(FALSE)
+# mgp <- latex2exp::TeX('Geopotential height ($m^2s^{-1}$)')
+mpg <- 'Geopotential height ($m^2s^{-1}$)'
+
+theme_set(theme_shceof(base_size = 20) +
+            theme(legend.box.spacing = grid::unit(-.5, "lines")))
+panel_background <- theme(panel.background = element_rect(fill = "#fbfbfb", color = NA),
+                          panel.ontop = FALSE,
+                          tagger.panel.tag.background = ggplot2::element_rect(color = NA,
+                                                                              fill = "#fbfbfb"))
+one_less <- function(x) {
+  n <- length(x)
+  if (n < 4) return(x)
+  x <- round(x, 5)
+
+  c( rev(JumpBy(rev(x[x < 0]), 2, fill = "")),
+     JumpBy(x[x > 0], 2, fill = ""))
+}
+no_grid <- theme(panel.grid = element_blank())
+guide_colorstrip_bottom <- function(width = 15, height = 0.5,
+                                    title.position = "top",...) {
+  guide_colorstrip(title.position = title.position, title.hjust = 0.5,
+                   barheight = height,
+                   barwidth = width, ...)
+}
+
+f <- formals(geom_contour_tanaka)
+f$range <-  c(0.01, 0.2)
+formals(geom_contour_tanaka) <- f
+
+guide_colorsteps_bottom <- function(width = 15, height = 0.5, even.steps = FALSE,
+                                    title.position = "top", show.limits = TRUE, ticks = TRUE,
+                                    ...) {
+  guide_colorsteps(show.limits = show.limits, even.steps = even.steps,
+                   title.position = title.position, title.hjust = 0.5,
+                   barheight = height, ticks = ticks,
+                   ticks.colour = "black",
+                   barwidth = width, frame.colour = "black", ...)
+}
+
+grid_y <- theme(panel.grid.major.y = element_line(color = "gray60", size = 0.1))
+
+geom_coords <- function() {
+
+  list(
+    no_grid,
+    annotate("segment",
+             y = seq(-90, 0, by = 15),
[…]

### 2023-01-02 analysis/paper/shceof.Rmd
Más respuestas

@@
 output: 
+  # bookdown::word_document2:
+  #     pandoc_args: ["--lua-filter=abstract-to-meta.lua", "--wrap=preserve"]
   bookdown::pdf_document2:
@@
 
-To describe the phase-varying nature of these two waves one could combine each pair into indices of amplitude of phase. 
-So, for instance, the amplitude of the wave 1 EOF could be measured as $\sqrt{\mathrm{EOF1}^2 + \mathrm{EOF2}^2}$ and it's phase as $\tan^{-1} \left ( \frac{\mathrm{EOF2}}{\mathrm{EOF1}} \right )$ (where $\mathrm{EOF1}$ and $\mathrm{EOF2}$ are the time series associated with each EOF).
-However, this rests upon visual inspection of the spatial patterns and only works if both phases appear cleanly in different EOFs, which is not guaranteed. 
+To describe the phase-varying nature of these two waves, the simplest solution is to combine each pair of EOFs into indices of amplitude and phase.
+So, for instance, the amplitude of the wave 1-like EOF could be measured as $\sqrt{\mathrm{PC1}^2 + \mathrm{PC2}^2}$ and its phase as $\tan^{-1} \left ( \frac{\mathrm{PC2}}{\mathrm{PC1}} \right )$ (where $\mathrm{PC1}$ and $\mathrm{PC2}$ are the time series associated with each EOF).
+However, this rests upon visual inspection of the spatial patterns and only works properly if both phases appear cleanly in different EOFs, which is not guaranteed by construction.
 In particular, this doesn't work with the wave 1 pattern observed as the leading EOF in 200hPa geopotential heihgt zonal anomalies (not shown).
 
-In Complex Empirical Orthogonal Functions (cEOF) analysis, each cEOF is represented by two spatial patterns that are shifted by 1/4 wavelength by construction. 
+A better alternative for describing phase-varying waves is to use Complex Empirical Orthogonal Functions (cEOF) analysis (Horel, 1984). 
+Each cEOF is represented by two spatial patterns that are shifted by 1/4 wavelength by construction.
 The actual field reconstructed by each cEOF is the linear combination of the two spatial fields weighted by its respective time series. 
-This is analogous to how any sine wave with any amplitude and phase can be constructed by the sum of a sine wave and cosine wave with different amplitude but constant phase. 
+This is analogous to how any sine wave can be constructed by the sum of a sine wave and cosine wave with different amplitude but constant phase. 
 This means that cEOFs naturally represent phase-varying wave-like patterns that change location as well as amplitude.
 
-For example, Figure \@ref(fig:ceofs-1) a.1 shows the two spatial patterns of a cEOF, one with shaded  contours (which we call the 0º phase) and the other with black contours (which we call the 90º phase).
+Figure \@ref(fig:ceofs-1) a.1 shows the two spatial patterns of the cEOF corresponding to XXX. 
+For each subplot one of the cEOFs is plotted with shaded contours (which represents the 0º phase) and the other with black contours (which represents the 90º phase). 
 This cEOF is very similar to the two leading EOFs shown in Figure \@ref(fig:eof-naive) and represents a zonal wave 1 pattern. 
-When the phase of the wave matches the 0º phase, then the 0º phase time series will be positive and the 90º phase time series will be zero. 
+For instance, when the phase of the wave matches the 0º phase, then the 0º phase time series will be positive and the 90º phase time series will be zero.
 Similarly, when the phase of the wave matches the 90º phase, the 90º phase time series will be positive and the 0º phase time series will be zero.
-Cases will have non zero values in both time series will represent intermediate phases.
-
-These two phases are not special and the same cEOF could be represented by any two linear combinations of the phases shown in Figure \@ref(fig:ceofs-1) a.1. 
+The intermediate phases will have non zero values in both time series.
+The two phases described are not special and the same cEOF could be represented by any linear combinations of the phases shown in Figure \@ref(fig:ceofs-1) a.1. 
 This parallels traditional EOFs, which are defined up to an additive constant. 
 For real EOFs, this constant can be either 0 or $\pi$, corresponding to a change in sign.
-For cEOFs, it can be any real number between 0 and $2\pi$ [@horel1984], corresponding to rotations in the complex plane.
-
-Compute cEOFs are computed the same way as traditional EOFs except that the data is first augmented by computing its analytic signal. 
-This is a complex number whose real part is the original series and the imaginary part is the original data shifted by 90\degree at each spectral frequency -- i.e. its Hilbert transform.
-Each cEOF is a set of complex-valued spatial patterns and time series in which the real and imaginary parts are the two orthogonal phases of the wave-like pattern.
-
+On the other hand, for cEOFs, it can be any real number between 0 and $2\pi$ [@horel1984], corresponding to rotations in the complex plane.
+
+cEOFs are computed in the same way as traditional EOFs except that the data is first augmented by computing its analytic signal. 
+This is a complex number whose real part is the original series and the imaginary part is the original data shifted by 90º at each spectral frequency -- i.e. 
+its Hilbert transform.
+Each cEOF is a set of complex-valued spatial patterns and time series in which the real (or 0º phase) and imaginary (or 90º phase) parts are the two orthogonal phases of the wave-like pattern.
 In this paper we use the term 0º cEOF and 90º cEOF to refer to each part of the whole cEOF.
@@
 This monopole might indicate strengthening of the polar vortex associated with positive values of the 0º cEOF2 and weakening associated with negative values of 0º cEOF2.
+However, since these anomalies are not statistically significant, this feature should not be overinterpreted.
 
@@
 Therefore, the ozone zonal wave 1 is negatively correlated with the temperature zonal wave 1 in the upper stratosphere, and positively correlated in the upper stratosphere. 
-This change in phase is observed in ozone anomalies forced by planetary waves that reach the stratosphere [@hartmann1979; @wirth1993; @smith1995]. 
-In the photochemically-dominated upper stratosphere, cold temperatures inhibit the destruction of ozone, explaining the opposite behaviour for both variables.
+This change in phase is observed in ozone anomalies forced by planetary waves that reach the stratosphere.
+In the photochemically-dominated upper stratosphere, cold temperatures inhibit the destruction of ozone, explaining the opposite behaviour for both variables as were elucidated with dynamical chemical models [@hartmann1979; @wirth1993; @smith1995].
 On the other hand, in the advectively-dominated lower stratosphere, ozone anomalies are 1/4 wavelength out of phase with horizontal and vertical transport, which are in addition 1/4 wavelength out of phase with temperature anomalies, resulting in same sign anomalies for the response of both variables [@hartmann1979; @wirth1993; @smith1995].
@@
 In `r scales::percent(imaginary)` of seasons cEOF2 has a phase similar to either the negative or positive 90º phase, making the 90º phase the most common phase. 
-This is also, by construction, the phase that is most correlated with ENSO.
+This is also the phase that is most correlated with ENSO by the definition of the 0° phase as described in Section \@ref(methods).
 
@@
 
-This lack of strong relationship between the cEOF1 and temperature and precipitation might be surprising considering the correlation between the cEOF1 and the SAM (Fig.
-\@ref(fig:sam-eof-vertical) column 1) and the correlation between SAM and temperature east and west of the Antarctic Peninsula, and with precipitation in western Australia (BUSCAR CITA).
-There a two main reasons for this. 
+This lack of strong relationship between the cEOF1 and SST, temperature and precipitation might be surprising considering the correlation between the cEOF1 and the SAM (Fig.
+\@ref(fig:sam-eof-vertical) column 1) and the correlation between SAM and Central Pacific SST, temperature east and west of the Antarctic Peninsula, and with precipitation in western Australia [@fogt2020].
+There are two main reasons for this.
 First, the correlation between cEOF1 and the SAM in the troposphere is modest, with less than 50% of shared variance (Fig.
-\@ref(fig:sam-eof-vertical) column 1), so one need not expect these to indices to be equivalent.
-Second, on a previous paper [@campitelli2022] we showed that the strong relationship between the SAM and temperature anomalies around the Antarctic Peninsula is mainly due to the asymmetric part of the SAM.
+\@ref(fig:sam-eof-vertical) column 1), so one need not expect these indices to be equivalent.
+Second, @campitelli2022 showed that the strong relationship between the SAM and Pacific SSTs and temperature anomalies around the Antarctic Peninsula is mainly due to the asymmetric part of the SAM.
 Meanwhile, the cEOF1 is significantly correlated only with the symmetric part of the SAM (Fig.
@@
 In this study we assessed extratropical Southern Hemisphere zonally asymmetric circulation in austral spring. 
-For this purpose, two complex indices were derived using Complex Empirical Orthogonal Functions and used to characterise both amplitude and phase of planetary waves.
+For this purpose, were derived two complex indices using Complex Empirical Orthogonal Functions and used to characterise both amplitude and phase of planetary waves.
 
@@
 
-However the advantage of using the methodology proposed here could be questioned.
-Firstly, the spatial fields obtained from both components of cEOF2, that resembles PSA patterns, are in quadrature by construction.
-Thus, the cEOF methodology allows to derive a joint PSA index from the resulted amplitude and phase.
-Secondly, because of how these components are constructed, they are not forced to be orthogonal to other modes of circulation, as in the standard EOF methodology. 
-This allows us to show for example, that the 90º cEOF2, that resembles PSA1 variability, is closely associated to the SAM in the troposphere. 
-Moreover, the high correlation of this component to the zonally asymmetric portion of the SAM allow us to speculate that the A-SAM might actually be a statistical contamination of the PSA mode on the symmetric SAM.
-
-@irving2016 argued that there is some disagreement in the literature of whether the phase of the PSA pattern is affected by the location of tropical SST anomalies.
-With the methodology used in this study, we were able to show not only that the cEOF2 tends to be in the positive or negative 90º phase (\~PSA1) when the ENSO region is warm or cold, respectively, but also that central Pacific SST anomalies tend to align the cEOF towards the negative 0º phase and eastern Pacific SST anomalies tend to align it towards positive 0º phase.
-When ENSO phase is neutral, the cEOF2 is still as active, but with no preferred phase.
-The latter agrees with the results of @cai2002, who showed that the CSIRO Model can develop PSA-like variability even in the absence of ENSO forcing (i.e.
-with a climatological run), but that the variability of one of the PSA modes was enhanced when adding the ENSO signal.
+Since the spatial fields obtained from both components of cEOF2, which resemble PSA patterns, are in quadrature by construction, the cEOF methodology allows to derive, for the first time to our knowledge, a joint PSA index from the resulted amplitude and phase.
+These patterns are not forced to be orthogonal to other modes of circulation, like they are in standard EOF methodology. 
+This allows us to show for example, that the 90º cEOF2, corresponding to PSA1 variability, is closely associated to the SAM in the troposphere. 
+Previous research in the SAM--PSA relationship had the issue that the SAM and the PSA patterns are not independently derived and so the correlation between these indices had to be zero by construction.
+
+```{r partial_cor_enso}
+partial_cor <- function(y, x1, x2) {
+
+  estimate1 <- ppcor::pcor.test(y, x1, x2)
+  estimate2 <- ppcor::pcor.test(y, x2, x1)
+
+  # rho_x1x2 <- stats::cov.wt(cbind(x1, x2), wt = weights, cor = TRUE)$cor[1, 2]
+  # rho_x1y <- stats::cov.wt(cbind(x1, y), wt = weights, cor = TRUE)$cor[1, 2]
+  # rho_x2y <- stats::cov.wt(cbind(x2, y), wt = weights, cor = TRUE)$cor[1, 2]
+  #
+  # estimate1 <- (rho_x1y - rho_x2y*rho_x1x2 ) / (sqrt(1 - rho_x2y^2) * sqrt(1 - rho_x1x2^2))
+  # estimate2 <- (rho_x2y - rho_x1y*rho_x1x2 ) / (sqrt(1 - rho_x1y^2) * sqrt(1 - rho_x1x2^2))
+
+  list(term = c(deparse(substitute(x1)),
+                deparse(substitute(x2))),
+       partial_correlation = c(estimate1$estimate, estimate2$estimate),
+       p.value = c(estimate1$p.value, estimate2$p.value)
+  )
+}
+
+fullsam <- sams[term == "full" & lev == 200] %>% 
+  .[, sam := scale(estimate)]
+
+partial_cor_enso <- ceof[, eof[[1]]$left, by = .(season)] %>%
+  .[cEOF == "cEOF2"] %>% 
+  sep_ReIm(format = "wider") %>% 
+  .[, `:=`(Real = scale(Real), 
+           Imaginary = scale(Imaginary))] %>% 
+  fullsam[., on = "time"] %>% 
+  na.omit() %>% 
+  enso[., on = "time"] %>% 
+  .[, oni := scale(oni)] %>%
+  .[, partial_cor(oni, sam, Imaginary)]
+
+```
+
+Most studies rely on correlations between an ENSO index and the SAM index [e.g.
+@lheureux2006, @cai2011a] or between the SAM index and other variables associated with tropical convection, such as OLR or tropical SSTS [e.g.
+@carvalho2005].
+@campitelli2022 showed that the correlation between ENSO and SAM is almost completely explained by the asymmetric component of the SAM. 
+In this work which we show that the asymmetric component of the SAM can be identified with the PSA1. 
+Therefore, the correlation between ENSO and SAM is predominantly the correlation between ENSO and PSA1 (the partial correlation between the SAM index and the ONI controlling for the 90º cEOF2 is `r partial_cor_enso[term == "sam", scales::number(partial_correlation, accuracy = 0.01)]`, p-value = `r partial_cor_enso[term == "sam", scales::pvalue(p.value, accuracy = 0.01)]`).
+This sheds new light into the previous literature, as one cannot assume that a correlation between ENSO and SAM indicates a relationship between ENSO and a zonally symmetric component.
+
+Further investigation is necessary to determine the connection between the symmetric component of the SAM and the PSA. 
+It is possible that the PSA may force a zonally symmetric response (or vice versa) via wave-zonal mean flow interactions, or that this correlation is simply a statistical artefact resulting from the EOF methodology used to define the SAM.
+
+@irving2016 argued that there is some disagreement in the literature of whether the phase of the PSA pattern is affected by the location of tropical SS   
 The sensitivity of the phase of the PSAs to the location of the tropical SST anomalies was also seen by @ciasto2015, who detected similar Rossby wave patterns associated with central Pacific and eastern Pacific SST anomalies but with a change in phase.

### 2023-01-02 analysis/paper/shceof.tex
Más respuestas

@@
 
-To describe the phase-varying nature of these two waves one could combine each pair into indices of amplitude of phase.
-So, for instance, the amplitude of the wave 1 EOF could be measured as \(\sqrt{\mathrm{EOF1}^2 + \mathrm{EOF2}^2}\) and it's phase as \(\tan^{-1} \left ( \frac{\mathrm{EOF2}}{\mathrm{EOF1}} \right )\) (where \(\mathrm{EOF1}\) and \(\mathrm{EOF2}\) are the time series associated with each EOF).
-However, this rests upon visual inspection of the spatial patterns and only works if both phases appear cleanly in different EOFs, which is not guaranteed.
+To describe the phase-varying nature of these two waves, the simplest solution is to combine each pair of EOFs into indices of amplitude and phase.
+So, for instance, the amplitude of the wave 1-like EOF could be measured as \(\sqrt{\mathrm{PC1}^2 + \mathrm{PC2}^2}\) and its phase as \(\tan^{-1} \left ( \frac{\mathrm{PC2}}{\mathrm{PC1}} \right )\) (where \(\mathrm{PC1}\) and \(\mathrm{PC2}\) are the time series associated with each EOF).
+However, this rests upon visual inspection of the spatial patterns and only works properly if both phases appear cleanly in different EOFs, which is not guaranteed by construction.
 In particular, this doesn't work with the wave 1 pattern observed as the leading EOF in 200hPa geopotential heihgt zonal anomalies (not shown).
 
-In Complex Empirical Orthogonal Functions (cEOF) analysis, each cEOF is represented by two spatial patterns that are shifted by 1/4 wavelength by construction.
+A better alternative for describing phase-varying waves is to use Complex Empirical Orthogonal Functions (cEOF) analysis (Horel, 1984).
+Each cEOF is represented by two spatial patterns that are shifted by 1/4 wavelength by construction.
 The actual field reconstructed by each cEOF is the linear combination of the two spatial fields weighted by its respective time series.
-This is analogous to how any sine wave with any amplitude and phase can be constructed by the sum of a sine wave and cosine wave with different amplitude but constant phase.
+This is analogous to how any sine wave can be constructed by the sum of a sine wave and cosine wave with different amplitude but constant phase.
 This means that cEOFs naturally represent phase-varying wave-like patterns that change location as well as amplitude.
 
-For example, Figure \ref{fig:ceofs-1} a.1 shows the two spatial patterns of a cEOF, one with shaded contours (which we call the 0º phase) and the other with black contours (which we call the 90º phase).
+Figure \ref{fig:ceofs-1} a.1 shows the two spatial patterns of the cEOF corresponding to XXX.
+For each subplot one of the cEOFs is plotted with shaded contours (which represents the 0º phase) and the other with black contours (which represents the 90º phase).
 This cEOF is very similar to the two leading EOFs shown in Figure \ref{fig:eof-naive} and represents a zonal wave 1 pattern.
-When the phase of the wave matches the 0º phase, then the 0º phase time series will be positive and the 90º phase time series will be zero.
-Similarly, when the phase of the wave matches the 90º phase, the 90º phase time series will be positive and the 0º phase time series will be zero.\\
-Cases will have non zero values in both time series will represent intermediate phases.
-
-These two phases are not special and the same cEOF could be represented by any two linear combinations of the phases shown in Figure \ref{fig:ceofs-1} a.1.
+For instance, when the phase of the wave matches the 0º phase, then the 0º phase time series will be positive and the 90º phase time series will be zero.
+Similarly, when the phase of the wave matches the 90º phase, the 90º phase time series will be positive and the 0º phase time series will be zero.
+The intermediate phases will have non zero values in both time series.
+The two phases described are not special and the same cEOF could be represented by any linear combinations of the phases shown in Figure \ref{fig:ceofs-1} a.1.
 This parallels traditional EOFs, which are defined up to an additive constant.
 For real EOFs, this constant can be either 0 or \(\pi\), corresponding to a change in sign.
-For cEOFs, it can be any real number between 0 and \(2\pi\) \citep{horel1984}, corresponding to rotations in the complex plane.
+On the other hand, for cEOFs, it can be any real number between 0 and \(2\pi\) \citep{horel1984}, corresponding to rotations in the complex plane.
 
-Compute cEOFs are computed the same way as traditional EOFs except that the data is first augmented by computing its analytic signal.
-This is a complex number whose real part is the original series and the imaginary part is the original data shifted by 90\degree at each spectral frequency -- i.e.~its Hilbert transform.
-Each cEOF is a set of complex-valued spatial patterns and time series in which the real and imaginary parts are the two orthogonal phases of the wave-like pattern.
-
+cEOFs are computed in the same way as traditional EOFs except that the data is first augmented by computing its analytic signal.
+This is a complex number whose real part is the original series and the imaginary part is the original data shifted by 90º at each spectral frequency -- i.e.~
+its Hilbert transform.
+Each cEOF is a set of complex-valued spatial patterns and time series in which the real (or 0º phase) and imaginary (or 90º phase) parts are the two orthogonal phases of the wave-like pattern.
 In this paper we use the term 0º cEOF and 90º cEOF to refer to each part of the whole cEOF.
@@
 Therefore, the ozone zonal wave 1 is negatively correlated with the temperature zonal wave 1 in the upper stratosphere, and positively correlated in the upper stratosphere.
-This change in phase is observed in ozone anomalies forced by planetary waves that reach the stratosphere \citep{hartmann1979, wirth1993, smith1995}.
-In the photochemically-dominated upper stratosphere, cold temperatures inhibit the destruction of ozone, explaining the opposite behaviour for both variables.
+This change in phase is observed in ozone anomalies forced by planetary waves that reach the stratosphere.
+In the photochemically-dominated upper stratosphere, cold temperatures inhibit the destruction of ozone, explaining the opposite behaviour for both variables as were elucidated with dynamical chemical models \citep{hartmann1979, wirth1993, smith1995}.
 On the other hand, in the advectively-dominated lower stratosphere, ozone anomalies are 1/4 wavelength out of phase with horizontal and vertical transport, which are in addition 1/4 wavelength out of phase with temperature anomalies, resulting in same sign anomalies for the response of both variables \citep{hartmann1979, wirth1993, smith1995}.
@@
 In 62\% of seasons cEOF2 has a phase similar to either the negative or positive 90º phase, making the 90º phase the most common phase.
-This is also, by construction, the phase that is most correlated with ENSO.
+This is also the phase that is most correlated with ENSO by the definition of the 0° phase as described in Section \ref{methods}.~
 
@@
 
-This lack of strong relationship between the cEOF1 and temperature and precipitation might be surprising considering the correlation between the cEOF1 and the SAM (Fig.
-\ref{fig:sam-eof-vertical} column 1) and the correlation between SAM and temperature east and west of the Antarctic Peninsula, and with precipitation in western Australia (BUSCAR CITA).
-There a two main reasons for this.
+This lack of strong relationship between the cEOF1 and SST, temperature and precipitation might be surprising considering the correlation between the cEOF1 and the SAM (Fig.
+\ref{fig:sam-eof-vertical} column 1) and the correlation between SAM and Central Pacific SST, temperature east and west of the Antarctic Peninsula, and with precipitation in western Australia \citep{fogt2020}.
+There are two main reasons for this.
 First, the correlation between cEOF1 and the SAM in the troposphere is modest, with less than 50\% of shared variance (Fig.
-\ref{fig:sam-eof-vertical} column 1), so one need not expect these to indices to be equivalent.
-Second, on a previous paper \citep{campitelli2022} we showed that the strong relationship between the SAM and temperature anomalies around the Antarctic Peninsula is mainly due to the asymmetric part of the SAM.
+\ref{fig:sam-eof-vertical} column 1), so one need not expect these indices to be equivalent.
+Second, \citet{campitelli2022} showed that the strong relationship between the SAM and Pacific SSTs and temperature anomalies around the Antarctic Peninsula is mainly due to the asymmetric part of the SAM.
 Meanwhile, the cEOF1 is significantly correlated only with the symmetric part of the SAM (Fig.
@@
 In this study we assessed extratropical Southern Hemisphere zonally asymmetric circulation in austral spring.
-For this purpose, two complex indices were derived using Complex Empirical Orthogonal Functions and used to characterise both amplitude and phase of planetary waves.
+For this purpose, were derived two complex indices using Complex Empirical Orthogonal Functions and used to characterise both amplitude and phase of planetary waves.
 
@@
 
-However the advantage of using the methodology proposed here could be questioned.
-Firstly, the spatial fields obtained from both components of cEOF2, that resembles PSA patterns, are in quadrature by construction.
-Thus, the cEOF methodology allows to derive a joint PSA index from the resulted amplitude and phase.
-Secondly, because of how these components are constructed, they are not forced to be orthogonal to other modes of circulation, as in the standard EOF methodology.
-This allows us to show for example, that the 90º cEOF2, that resembles PSA1 variability, is closely associated to the SAM in the troposphere.
-Moreover, the high correlation of this component to the zonally asymmetric portion of the SAM allow us to speculate that the A-SAM might actually be a statistical contamination of the PSA mode on the symmetric SAM.
+Since the spatial fields obtained from both components of cEOF2, which resemble PSA patterns, are in quadrature by construction, the cEOF methodology allows to derive, for the first time to our knowledge, a joint PSA index from the resulted amplitude and phase.
+These patterns are not forced to be orthogonal to other modes of circulation, like they are in standard EOF methodology.
+This allows us to show for example, that the 90º cEOF2, corresponding to PSA1 variability, is closely associated to the SAM in the troposphere.
+Previous research in the SAM--PSA relationship had the issue that the SAM and the PSA patterns are not independently derived and so the correlation between these indices had to be zero by construction.
 
-\citet{irving2016} argued that there is some disagreement in the literature of whether the phase of the PSA pattern is affected by the location of tropical SST anomalies.
-With the methodology used in this study, we were able to show not only that the cEOF2 tends to be in the positive or negative 90º phase (\textasciitilde PSA1) when the ENSO region is warm or cold, respectively, but also that central Pacific SST anomalies tend to align the cEOF towards the negative 0º phase and eastern Pacific SST anomalies tend to align it towards positive 0º phase.
-When ENSO phase is neutral, the cEOF2 is still as active, but with no preferred phase.
-The latter agrees with the results of \citet{cai2002}, who showed that the CSIRO Model can develop PSA-like variability even in the absence of ENSO forcing (i.e.~with a climatological run), but that the variability of one of the PSA modes was enhanced when adding the ENSO signal.
+Most studies rely on correlations between an ENSO index and the SAM index \citep[e.g.][\citet{cai2011a}]{lheureux2006} or between the SAM index and other variables associated with tropical convection, such as OLR or tropical SSTS \citep[e.g.][]{carvalho2005}.
+\citet{campitelli2022} showed that the correlation between ENSO and SAM is almost completely explained by the asymmetric component of the SAM.
+In this work which we show that the asymmetric component of the SAM can be identified with the PSA1.
+Therefore, the correlation between ENSO and SAM is predominantly the correlation between ENSO and PSA1 (the partial correlation between the SAM index and the ONI controlling for the 90º cEOF2 is 0.20, p-value = 0.23).
+This sheds new light into the previous literature, as one cannot assume that a correlation between ENSO and SAM indicates a relationship between ENSO and a zonally symmetric component.
+
+Further investigation is necessary to determine the connection between the symmetric component of the SAM and the PSA.
+It is possible that the PSA may force a zonally symmetric response (or vice versa) via wave-zonal mean flow interactions, or that this correlation is simply a statistical artefact resulting from the EOF methodology used to define the SAM.
+
+\citet{irving2016} argued that there is some disagreement in the literature of whether the phase of the PSA pattern is affected by the location of tropical SS\\
 The sensitivity of the phase of the PSAs to the location of the tropical SST anomalies was also seen by \citet{ciasto2015}, who detected similar Rossby wave patterns associated with central Pacific and eastern Pacific SST anomalies but with a change in phase.

### 2023-01-04 analysis/paper/shceof.Rmd
Todas las respuestas

@@
 output: 
-  # bookdown::word_document2:
-  #     pandoc_args: ["--lua-filter=abstract-to-meta.lua", "--wrap=preserve"]
+  bookdown::word_document2:
+      pandoc_args: ["--lua-filter=abstract-to-meta.lua", "--wrap=preserve"]
   bookdown::pdf_document2:
@@
   comment = "#>",
-  cache.extra = 42,
+  cache.extra = 44,
   cache.path = "../cache/paper/",
@@
 
-The large-scale extratropical circulation in the Southern Hemisphere is strongly zonally symmetric, but its zonal departures, albeit highly relevant for regional impacts, have been less studied.
+The large-scale extratropical circulation in the Southern Hemisphere is much more zonally symmetric than that of the Northern Hemisphere, but its zonal departures, albeit highly relevant for regional impacts, have been less studied.
 In this study we analyse the joint variability of the zonally asymmetric springtime stratospheric and tropospheric circulation using Complex Empirical Orthogonal Functions (cEOF) to characterise planetary waves of varying amplitude and phase.
@@
 
-The large-scale extratropical circulation in the Southern Hemisphere (SH) is strongly zonally symmetric, but its zonal departures are highly relevant for regional impacts [e.g. @hoskins2005].
-They strongly modulate weather systems and regional climate through promoting longitudinally different latitudinal transport of heat, humidity, and momentum [@trenberth1980a; @raphael2007] and could even be related to the occurrence of high-impact climate extremes [@pezza2012].
-
-Zonally asymmetric circulation is typically described by the amplitude and phase of zonal waves obtained by Fourier decomposition of geopotential heights or sea-level pressures at each latitude [e.g.
+The large-scale extratropical circulation in the Southern Hemisphere (SH) is much more zonally symmetric than that of the Northern Hemisphere, but departures from the zonal mean are associated with regional impacts [e.g.
+@hoskins2005].
+They strongly modulate weather systems and regional climate through promoting longitudinally varying meridional transport of heat, humidity, and momentum [@trenberth1980a; @raphael2007] and could even be related to the occurrence of high-impact climate extremes [@pezza2012].
+
+The zonally asymmetric circulation is typically described by the amplitude and phase of zonal waves obtained by Fourier decomposition of geopotential height or sea-level pressure at each latitude [e.g.
 @vanloon1972; @trenberth1980a; @turner2017].
-This approach suggests that zonal waves 1 and 3 explain almost 99% of the total variance in the annual mean 500-hPa pattern at 50ºS [@vanloon1972].
+This approach suggests that zonal waves 1 and 3 explain almost 99% of the total variance in the annual mean 500 hPa geopotential height zonal anomalies at 50ºS [@vanloon1972].
 @trenberth1985 concluded that wave 3 plays a role in the development of blocking events.
-In addition, previous works have identified at extratropical and subpolar latitudes, wave-like patterns with dominant wavenumbers 3-4, also exerting distinctive regional impacts.
-@raphael2007 shows that variability in the planetary wave 3 projected onto its climatological location is associated with anomalies in the Antarctic sea-ice concentration.
-
-The Fourier decomposition relies on the assumption that the circulation can be meaningfully described in terms of zonal waves of constant amplitude along a latitude circle.
+In addition, previous works have identified wave-like patterns with dominant wavenumbers 3-4 at extratropical and subpolar latitudes with distinctive regional impacts.
+@raphael2007 showed that variability in the planetary wave 3 projected onto its climatological location is associated with anomalies in the Antarctic sea-ice concentration.
+
+Fourier decomposition relies on the assumption that the circulation can be meaningfully described in terms of zonal waves of constant amplitude along a latitude circle.
 However, this is not valid for meridionally propagating waves or zonal waves with localised amplitudes.
-Addressing this limitation, the Fourier technique can be generalized to integrate all planetary wave amplitude regardless of wave number, by computing the wave envelope [@irving2015].
-The latter makes it possible to represent planetary waves with different amplitude at different longitudes, but it removes all information about phase and wave number.
-With this approach, @irving2015 showed that planetary wave amplitude in general is associated to Antarctic sea-ice concentration and temperature, as well as to precipitation anomalies in regions of significant topography in SH mid-latitudes and Antarctica.
-
-Another extensively used approach to characterise the SH tropospheric circulation anomalies, is by computing Empirical Orthogonal Functions (EOF, also known as Principal Component Analysis).
+Addressing this limitation, the Fourier technique can be generalized to integrate all planetary wave amplitude regardless of wave number by computing the wave envelope [@irving2015].
+The wave envelope can represent planetary waves with different amplitude at different longitudes, but lacks information about phase and wave number.
+Using this method, @irving2015 showed that planetary wave amplitude in general is associated to Antarctic sea-ice concentration and temperature, as well as to precipitation anomalies in regions of significant topography in SH mid-latitudes and Antarctica.
+
+Another extensively-used approach to characterise the SH tropospheric circulation anomalies, is computing Empirical Orthogonal Functions (EOF, also known as Principal Component Analysis).
 Within the EOF framework, the Southern Annular Mode (SAM) appears as the leading mode of variability of the SH circulation [@fogt2020].
-SAM represents a relatively zonally symmetric pattern of alternating low pressures in polar latitude and a ring of high pressures in high latitudes with an embedded wave 3 pattern that is more prominent in the Pacific sector.
+The SAM represents a relatively zonally symmetric pattern of alternating low pressures in polar latitudes and a ring of high pressures in high latitudes with an embedded wave 3 pattern that is more prominent in the Pacific sector.
 The 2nd and 3rd EOFs, usually known as Pacific--South American Patterns (PSA) 1 and PSA2 patterns, respectively, describe meridionally propagating wave trains that originate in the eastern equatorial Pacific and Australian-Indian Ocean sector, and travel towards the South Atlantic following a great-circle arch along the Antarctic coast [@mo2001].
@@
 
-Some of the zonally asymmetric patterns of the SH circulation variability described previously, appear to have experienced secular changes.
-For instance, @raphael2003 found that the amplitude of the zonal wave 1 experienced a large increase and that the zonal wave 3 experienced changes in its annual cycle between 1958 and 1996.
+Some of the zonally asymmetric patterns of the SH circulation variability described previously appear to have experienced secular changes.
+For instance, @raphael2003 suggests that the amplitude of the zonal wave 1 experienced a large increase and that the zonal wave 3 experienced changes in its annual cycle between 1958 and 1996.
 However, little is known yet about variability and trends of these patterns.
 
-Patterns resulting from EOF analyses are more flexible than Fourier decomposition derived modes in the sense that they can capture oscillation patterns that cannot be characterised by purely sinusoidal waves with constant amplitude.
-Nonetheless, they are restricted to standing oscillation modes and could not properly represent propagating or phase-varying features such as zonal waves.
+Patterns resulting from EOF analysis are more flexible than Fourier decomposition-derived modes in that they can capture oscillation patterns that cannot be characterised by purely sinusoidal waves with constant amplitude.
+Nonetheless, they are restricted to standing oscillation modes and cannot properly represent propagating or phase-varying features such as zonal waves.
 A single EOF can also represent a mixture of two or more physical modes.
@@
 These derived methods are grounded on other methods such as Fourier decomposition or EOF to identify the centres of action for the described phenomena and can be useful to characterise features that are not readily apparent with these methods.
-These kinds of indices are generally easy to compute, but they usually capture non-stationary patterns.
-
-An alternative methodology that has been proposed to study travelling and standing waves is complex Empirical Orthogonal Functions [cEOF; @horel1984].
+These kinds of indices are generally easy to compute, but they usually don't capture non-stationary patterns.
+
+An alternative methodology that has been proposed to study travelling and standing waves is complex Empirical Orthogonal Functions (cEOF; @horel1984).
 This method extends EOF analysis to capture oscillations with varying amplitude and phase and has been applied to the time domain.
-For instance, @krokhin2007 applied cEOF to station-based monthly precipitation anomalies and monthly temperature anomalies in the Eastern Siberia and the Far East region to characterise the main modes of variability and their connection to teleconnection indices.
-Similarly, @gelbrecht2018 used cEOF applied to daily precipitation from reanalysis to study the propagating characteristics of the South American Monsoon.
+For instance, @krokhin2007 applied cEOF to station-based monthly precipitation anomalies and monthly temperature anomalies in the Eastern Siberia and the Far East region to characterise the main modes of variability and their relationship with teleconnection indices.
+Similarly, @gelbrecht2018 applied cEOF to daily precipitation from reanalysis to study the propagating characteristics of the South American Monsoon.
 To our knowledge, cEOF analysis has not been applied in the spatial domain to capture the phase-varying nature of planetary waves in the atmosphere.
 
-The general goal of this study is to improve the description and understanding of the zonally asymmetric extratropical SH circulation using cEOF, which allow to describe phase varying planetary waves with variable amplitudes along a latitude circle.
-In addition, it is proposed to expand the knowledge of the simultaneous behaviour of SH asymmetric circulation in the troposphere and the stratosphere.
-
-We restrict this paper to the spring September-October-November (SON) trimester because during this season both the tropical teleconnections over South America [@cazes-boezio2003] and the SH stratosphere-troposphere interactions [@lim2018] are at their maximum amplitude.
+The general goal of this study is to improve the description and understanding of the zonally asymmetric extratropical SH circulation using cEOF, which can describe phase varying planetary waves with variable amplitude along a latitude circle.
+In addition, we try to expand the knowledge of the simultaneous behaviour of SH asymmetric circulation in the troposphere and the stratosphere.
+
+We restrict this paper to the September-October-November (SON) trimester because during this season both the tropical teleconnections over South America [@cazes-boezio2003] and the SH stratosphere-troposphere interactions [@lim2018] are at their maximum amplitude.
 
@@
 
-We used monthly geopotential height, air temperature, ozone mixing ratio, and total column ozone (TOC) at 2.5\degree longitude by 2.5\degree latitude of horizontal resolution and 37 vertical isobaric levels from the European Centre for Medium-Range Weather Forecasts Reanalysis version 5 [ERA; @era5] for the period 1979 -- 2019.
-Most of our analysis is restricted to the post-satellite era to avoid any confounding factors arising from the incorporation of satellite observations, but we also used the preliminary back extension of ERA5 from 1950 to 1978 [@era5be] to describe long-term trends.
+We used monthly geopotential height, air temperature, ozone mixing ratio, and total column ozone (TOC) at 2.5º longitude by 2.5º latitude of horizontal resolution and 37 vertical isobaric levels from the European Centre for Medium-Range Weather Forecasts Reanalysis version 5 [ERA; @era5] for the period 1979 -- 2019.
+Most of our analysis is restricted to the post-satellite era to avoid confounding factors arising from the incorporation of satellite observations, but we also used the preliminary back extension of ERA5 from 1950 to 1978 [@era5be] to describe long-term trends.
 We derived streamfunction at 200 hPa from ERA5 vorticity using the FORTRAN subroutine FISHPACK [@fishpack] and we computed horizontal wave activity fluxes following @plumb1985.
-Sea Surface Temperature (SST) monthly fields are from Extended Reconstructed Sea Surface Temperature (ERSST) v5 [@huang2017] and precipitation monthly data from the CPC Merged Analysis of Precipitation [@cmap], with a 2\degree and 2.5\degree horizontal resolution respectively.
-The rainfall gridded dataset is based on information from different sources such as rain gauge observations, satellite inferred estimations and the NCEP-NCAR reanalysis, and it is available since 1979 to the present.
+Sea Surface Temperature (SST) monthly fields are from Extended Reconstructed Sea Surface Temperature (ERSST) v5 [@huang2017] and precipitation monthly data from the CPC Merged Analysis of Precipitation [@cmap], with a 2º and 2.5º horizontal resolution, respectively.
+The rainfall gridded dataset is based on information from different sources such as rain gauge observations, satellite inferred estimations and the NCEP-NCAR reanalysis, and it is available from 1979 to the present.
 
@@
 
-The amplitude of the zonal waves was obtained through the Fourier transform of the spatial field at each latitude circle.
-For the analysis of wave 1, we computed its amplitude and phase by averaging (area-weighted) the data for each variable and each SON between `r knitr::combine_words(LatLabel(o3wave_lats))`, and then extracting the wave-1 component of the Fourier spectrum.
+We computed the amplitude and phase of the TOC wave 1 by averaging (area-weighted) the data of each SON between `r knitr::combine_words(LatLabel(o3wave_lats))`, and then extracting the wave-1 component of the Fourier spectrum.
 We chose this latitude band because it is wide enough to capture most of the relevant anomalies of SH mid-latitudes.
@@
 
-(ref:eof-naive-cap) Spatial patterns of the four leading EOFs of SON zonal anomalies of geopotential height at 50 hPa south of 20\degree S for the 1979 -- 2019 period (arbitrary units).
+(ref:eof-naive-cap) Spatial patterns of the four leading EOFs of SON geopotential height zonal anomalies of at 50 hPa south of 20º S for the 1979 -- 2019 period (arbitrary units).
 
@@
 In the standard EOF analysis, zonal waves may appear as pairs of (possibly degenerate) EOFs representing similar patterns but shifted in phase [@horel1984].
-Figure \@ref(fig:eof-naive) shows the four leading EOFs of SON geopotential height zonal anomalies at 50 hPa south of 20\degree S.
-It is clear that the first two EOFs represent a single phase-varying zonal wave 1 pattern and the last two represent a similarly phase-varying pattern with higher wavenumber and three centres of action shifted by 1/4 wavelength (90\degree in frequency space).
+Figure \@ref(fig:eof-naive) shows the four leading EOFs of SON geopotential height zonal anomalies at 50 hPa south of 20º S.
+It is clear that the first two EOFs represent a single phase-varying zonal wave 1 pattern and the last two represent a similarly phase-varying pattern with higher wavenumber and three centres of action shifted by 1/4 wavelength (90º in frequency space).
 
@@
 So, for instance, the amplitude of the wave 1-like EOF could be measured as $\sqrt{\mathrm{PC1}^2 + \mathrm{PC2}^2}$ and its phase as $\tan^{-1} \left ( \frac{\mathrm{PC2}}{\mathrm{PC1}} \right )$ (where $\mathrm{PC1}$ and $\mathrm{PC2}$ are the time series associated with each EOF).
-However, this rests upon visual inspection of the spatial patterns and only works properly if both phases appear cleanly in different EOFs, which is not guaranteed by construction.
-In particular, this doesn't work with the wave 1 pattern observed as the leading EOF in 200hPa geopotential heihgt zonal anomalies (not shown).
-
-A better alternative for describing phase-varying waves is to use Complex Empirical Orthogonal Functions (cEOF) analysis (Horel, 1984). 
+However, this rests on visual inspection of the spatial patterns and only works properly if both phases appear cleanly in different EOFs, which is not guaranteed by construction.
+In particular, this doesn't work with the wave 1 pattern observed as the leading EOF in 200 hPa geopotential height zonal anomalies (not shown).
+
+A better alternative for describing phase-varying waves is to use Complex Empirical Orthogonal Functions (cEOF) analysis [@horel1984]. 
 Each cEOF is represented by two spatial patterns that are shifted by 1/4 wavelength by construction.
@@
 
-Figure \@ref(fig:ceofs-1) a.1 shows the two spatial patterns of the cEOF corresponding to XXX. 
-For each subplot one of the cEOFs is plotted with shaded contours (which represents the 0º phase) and the other with black contours (which represents the 90º phase). 
-This cEOF is very similar to the two leading EOFs shown in Figure \@ref(fig:eof-naive) and represents a zonal wave 1 pattern. 
+Figure \@ref(fig:ceofs-1) a.1 shows the two spatial patterns of the cEOF of geopotential height zonal anomalies. 
+One of the cEOFs is plotted with shaded contours (which represents the 0º phase) and the other with black contours (which represents the 90º phase). 
+This cEOF is very similar to the two leading EOFs shown in Figure \@ref(fig:eof-naive) and represents a zonal wave 1 pattern; the 0º phase is roughtly the EOF1 and the 90º phase is roughtly the EOF2).
 For instance, when the phase of the wave matches the 0º phase, then the 0º phase time series will be positive and the 90º phase time series will be zero.
@@
 The two phases described are not special and the same cEOF could be represented by any linear combinations of the phases shown in Figure \@ref(fig:ceofs-1) a.1. 
-This parallels traditional EOFs, which are defined up to an additive constant. 
-For real EOFs, this constant can be either 0 or $\pi$, corresponding to a change in sign.
-On the other hand, for cEOFs, it can be any real number between 0 and $2\pi$ [@horel1984], corresponding to rotations in the complex plane.
+This parallels traditional EOFs, which are defined up to sign, which corresponds to a rotation in the complex plane of either 0 or $\pi$.
+On the other hand cEOFs are defined up to a rotation in the complex plane of any value between 0 and $2\pi$ [@horel1984].
 
 cEOFs are computed in the same way as traditional EOFs except that the data is first augmented by computing its analytic signal. 
-This is a complex number whose real part is the original series and the imaginary part is the original data shifted by 90º at each spectral frequency -- i.e. 
+This is a complex number whose real part is the original series and whose imaginary part is the original data shifted by 90º at each spectral frequency -- i.e. 
 its Hilbert transform.
@@
 The Hilbert transform is usually understood in terms of time-varying signal.
-However, in this work we apply the Hilbert transform at each latitude circle and at each considered time.
+However, in this work we apply the Hilbert transform at each latitude circle and at each considered time (e.g. the signal only depends on longitude).
 Since each latitude circle is a periodic domain, this procedure does not suffer from edge effects.
@@
 
-The cEOF methodology is applied to zonal anomalies of SON geopotential height south of 20\degree S separately at 50 and 200 hPa.
+First, we applied the cEOF methodology is applied to zonal anomalies of SON geopotential height south of 20º S separately at 50 and 200 hPa.
 Table \@ref(tab:corr-ceof-splitted) shows the coefficient of determination between time series of the amplitude of each cEOF across levels.
@@
 Both the spatial pattern similarity and the high temporal correlation of cEOFs computed at 50 hPa and 200 hPa suggest that these are, to a large extent, modes of joint variability.
-This motivates the decision of performing complex EOF jointly between levels.
+This motivates the decision of performing cEOF jointly between levels.
 Therefore cEOFs were computed using data from both levels at the same time.
@@
 As mentioned before, the choice of phases is arbitrary and equally valid. 
-But to make the interpretation easier, we chose the phase of each cEOF so that the either the º0 cEOF or the 90º cEOF is aligned with meaningful variables in our analysis.
+But to make the interpretation easier, we chose the phase of each cEOF so that the either the 0º cEOF or the 90º cEOF is aligned with meaningful variables in our analysis.
 This procedure does not create spurious correlations, it only takes an existing relationship and aligns it with a specific phase.
 
-In Section \@ref(precipitation) we show regressions of precipitation and temperature associated with intermediate phases.
-For those plots, we rotated the cEOFs by 1/4 wavelength by multiplying the complex time series by $\cos(\pi/4) + i\sin(\pi/4)$ and computing the regression on those rotated timeseries.
-
 Preliminary analysis showed that the first cEOF was closely related to the zonal wave 1 of Total Column Ozone and the second cEOF was closely related to ENSO.
-Therefore, we chose the phase of cEOF1 so that the time series corresponding to the 0ª cEOF1 has the maximum correlation with the zonal wave 1 of Total Column Ozone between `r knitr::combine_words(LatLabel(o3wave_lats))`.
-Similarly, we chose the phase of cEOF2 so that the coefficient of determination between the Oceanic Niño Index [@bamston1997] and the º0 cEOF2 is minimised, which also nearly maximises the correlation with the 90º cEOF2.
+Therefore, we chose the phase of cEOF1 so that the time series corresponding to the 0º cEOF1 has the maximum correlation with the zonal wave 1 of Total Column Ozone between `r knitr::combine_words(LatLabel(o3wave_lats))`.
+Similarly, we chose the phase of cEOF2 so that the coefficient of determination between the Oceanic Niño Index [@bamston1997] and the 0º cEOF2 is minimised, which also nearly maximises the correlation with the 90º cEOF2.
 
@@
 
+In Section \@ref(precipitation) we show regressions of precipitation and temperature associated with intermediate phases.
+For those plots, we rotated the cEOFs by 1/4 wavelength by multiplying the complex time series by $\cos(\pi/4) + i\sin(\pi/4)$ and computing the regression on those rotated timeseries.
+
 While we compute these complex principal components using data from 1979 to 2019, we extended the complex time series back to the 1950 -- 1978 period by projecting monthly geopotential height zonal anomalies standardised by level south of 20ºS onto the corresponding spatial patterns.
@@
 
-```{r ceofs-1, fig.cap = "Spatial patterns for the two leading cEOFs of SON zonal anomalies of geopotential height at 50 hPa and 200 hPa for the 1979 -- 2019 period.
+(ref:ceofs-1-cap) Spatial patterns for the two leading cEOFs of SON geopotential height zonal anomalies at 50 hPa and 200 hPa for the 1979 -- 2019 period.
 The shading (contours) corresponds to 0º (90º) phase.
 Arbitrary units.
-The proportion of variance explained for each mode with respect to the zonal mean is indicated in parenthesis.", fig.width=width_column, fig.height=width_column*1.2}
+The proportion of variance explained for each mode with respect to the zonal mean is indicated in parenthesis.
+
+```{r ceofs-1, fig.cap = "(ref:ceofs-1-cap)", fig.width=width_column, fig.height=width_column*1.2}
 gdata <- ceof[, eof[[1]]$right, by = .(season)] %>% 
@@
 
-  tag_facets(tag = "cr")
+  tag_facets(tag = "rc")
 
@@
 
-(ref:extended-series-cap) Time series of the two leading cEOFs of SON zonal anomalies of geopotential height at 50 hPa and 200 hPa.
+(ref:extended-series-cap) Time series of the two leading cEOFs of SON geopotential height zonal anomalies at 50 hPa and 200 hPa.
 cEOF1 (row a) and cEOF2 (row b) separated in their 0º (column 1) and 90º (column 2) phase.
@@
 
-The cEOF1 (Fig. \@ref(fig:ceofs-1) column a) is a hemispheric wave 1 pattern with maximum amplitude at high latitudes.
+The cEOF1 (Fig. \@ref(fig:ceofs-1) column 1) is a hemispheric wave 1 pattern with maximum amplitude at high latitudes.
 At 50 hPa the 0º cEOF1 has the maximum of the wave 1 at 150ºE and at 200 hPa, the maximum is located at around 175ºE indicating a westward shift with height.
-The cEOF2 (Fig. \@ref(fig:eof-naive) column b) shows also a zonal wave-like structure with maximum amplitude at high latitudes, but with shorter spatial scales.
+The cEOF2 (Fig. \@ref(fig:eof-naive) column 2) shows also a zonal wave-like structure with maximum amplitude at high latitudes, but with shorter spatial scales.
 In particular, the dominant structure at both levels is a wave 3 but with larger amplitude in the pacific sector.
-This modulated amplitude is more evident at 200 hPa, which is consistent with the cEOF2 computed separatedly for 200 hPa explains a bit more variance than the cEOF2 computed separatedly for 50 hPa (`r scales::percent(ceof_splitted[, eof[[1]]$sdev, by = levs][cEOF == "cEOF2", r2][1])` vs.
+There is no apparent phase shift with height but the amplitude of the pattern is greatly reduced in the stratosphere, which is consistent with the the fact that the cEOF2 computed separately for 200 hPa explains a bit more variance than the cEOF2 computed separately for 50 hPa (`r scales::percent(ceof_splitted[, eof[[1]]$sdev, by = levs][cEOF == "cEOF2", r2][1])` vs.
 `r scales::percent(ceof_splitted[, eof[[1]]$sdev, by = levs][cEOF == "cEOF2", r2][2])`, respectively).
-There is no apparent phase shift with height but the amplitude of the pattern is greatly reduced in the stratosphere, suggesting that this barotropic mode represents mainly tropospheric variability.
+This suggest that this barotropic mode represents mainly tropospheric variability.
 
@@
 
-In the previous section, cEOFs analysis was applied to zonal anomalies derived by removing the zonally mean values in order to isolate the main characteristics of the main zonal waves characterizing the circulation in the SH.
-In addition, regression fields were computed using the full fields of the variables in order to describe the influence of the cEOFs on the temporal anomalies.
+In the previous section, cEOF analysis was applied to zonal anomalies derived by removing the zonally mean values in order to isolate the main characteristics of the main zonal waves characterizing the circulation in the SH.
+In this section we compute regression fields using the full fields of the variables in order to describe the influence of the cEOFs on the temporal anomalies.
 
@@
 Figure \@ref(fig:eof1-regr-gh) shows regression maps of SON geopotential height anomalies upon cEOF1.
-At 50 hPa (Figure \@ref(fig:eof1-regr-gh) row a), the 0º cEOF1 is associated with a centre located over the Ross Sea.
+At 50 hPa (Figure \@ref(fig:eof1-regr-gh) row a), the 0º cEOF1 is associated with a positive centre located over the Ross Sea.
 The 90º cEOF1 is associated with a distinctive wave 1 pattern with maximum over the coast of East Antarctica. 
@@
 Figure \@ref(fig:eof2-regr-gh) shows the regression maps of geopotential height anomalies upon the cEOF2.
-In the troposphere (Fig. \@ref(fig:eof2-regr-gh) row b) the regression maps show wave trains similar to those identified for cEOF2 patterns (Fig \@ref(fig:ceofs-1)).
+In the troposphere (Fig. \@ref(fig:eof2-regr-gh) row a) the regression maps show wave trains similar to those identified for cEOF2 patterns (Fig \@ref(fig:ceofs-1)).
 Regressed anomalies associated with the 0º cEOF2 are 1/4 wavelength out of phase with those associated with the 90º cEOF2.
@@
 cEOF2 then represents an equivalent barotropic wave train that is very similar to the the PSA Patterns [@mo2001].
-Comparing the location of the positive anomaly near 90ºW in column b of Figure \@ref(fig:eof2-regr-gh) with Figures 1.a and b from @mo2001, the 0º cEOF2 regression map could be identified with PSA2, while the 90º cEOF2 resembles PSA1.
+Comparing the location of the positive anomaly near 90ºW in column 2 of Figure \@ref(fig:eof2-regr-gh) with Figures 1.a and b from @mo2001, the 0º cEOF2 regression map could be identified with PSA2, while the 90º cEOF2 resembles PSA1.
 
@@
   coord_cartesian(clip = "off") +
-  tag_facets() +
+  tag_facets(tag = "panel") +
   NULL
@@
 In both levels, the 0º cEOF1 is associated with a positive centre over the South Pole with its centre moved slightly towards 150ºE (Fig. \@ref(fig:eof1-regr-t) column 1).
-On the other hand, the regression maps on the 90º cEOF1 show a more clear wave 1 pattern with its maximum around 60ºE.
-
-Figure\ \@ref(fig:t-vertical) shows the vertical distribution of the regression coefficients on cEOF1 from zonal anomalies averaged between `r knitr::combine_words(LatLabel(o3wave_lats))` of air temperature and of ozone mixing ratio.
+On the other hand, the regression maps on the 90º cEOF1 show a clearer wave 1 pattern with its maximum around 60ºE.
+
+Figure\ \@ref(fig:t-vertical) shows the vertical distribution of the regression coefficients on cEOF1 from zonal anomalies of air temperature and of ozone mixing ratio averaged between `r knitr::combine_words(LatLabel(o3wave_lats))`.
 Temperature zonal anomalies associated with cEOF1 show a clear wave 1 pattern for both 0º and 90º phases throughout the atmosphere above 250 hPa with a sign reversal above 10 hPa.
@@
 
-The maximum ozone regressed anomalies match with the minimum temperature anomalies above 10 hPa and with the maximum temperature anomalies below 10 hPa (Fig. \@ref(fig:t-vertical)).
+The maximum ozone regressed anomalies coincide with the minimum temperature anomalies above 10 hPa and with the maximum temperature anomalies below 10 hPa (Fig. \@ref(fig:t-vertical)).
 Therefore, the ozone zonal wave 1 is negatively correlated with the temperature zonal wave 1 in the upper stratosphere, and positively correlated in the upper stratosphere. 
@@
 Thus, the 0º cEOF1 regression field (Figure\ \@ref(fig:o3-regr)a) coincides with the climatological position of the ozone hole, while it is 90° out of phase for the 90º cEOF1.
-The temporal correlation between the amplitudes of TOC planetary wave 1 and cEOF1 is `r wave1_cor[with == "amplitude"]$text`, while the correlation between their phases is `r wave1_cor[with == "phase"]$text`.
+The temporal correlation between the amplitudes of TOC planetary wave 1 and the amplitude of cEOF1 is `r wave1_cor[with == "amplitude"]$text`, while the correlation between their phases is `r wave1_cor[with == "phase"]$text`.
 Consequently, cEOF1 is strongly related with the SH ozone variability.
@@
 
-Given the similarity between the cEOF2 related anomaly structures (Fig. \@ref(fig:eof2-regr-gh)) and documented PSA patterns we study the relationship between them.
+Given the similarity between the cEOF2 related-associated structures (Fig. \@ref(fig:eof2-regr-gh)) and documented PSA patterns, we study the relationship between them.
 Table \@ref(tab:psa-eof2) shows the correlations between the two PSA indices and the time series for 0º and 90º phases of cEOF2.
@@
 Conventional EOF analysis will therefore tend to separate them relatively cleanly, with the EOF representing the 90º cEOF2 always leading the one representing the 0º cEOF2.
-This phase preferences is in agreement with @irving2016, who found a bimodal distribution to PSA-like variability (compare our Figure \@ref(fig:phase-histogram) with their Figure 6).
+This phase preference is in agreement with @irving2016, who found a bimodal distribution to PSA-like variability (compare our Figure \@ref(fig:phase-histogram) with their Figure 6).
 
@@
 We now explore the relationship between SAM and the cEOFs motivated by the resemblance between cEOFs regression maps and SAM patterns shown in Section\ \@ref(regressions).
-We then computed the coefficient of determination between the cEOFs time series and the three SAM indices (SAM, A-SAM and S-SAM) defined by @campitelli2022 at each vertical level (Fig.\ \@ref(fig:sam-eof-vertical)).
+We computed the coefficient of determination between the cEOFs time series and the three SAM indices (SAM, A-SAM and S-SAM) defined by @campitelli2022 at each vertical level (Fig.\ \@ref(fig:sam-eof-vertical)).
 The SAM index is statistically significantly correlated with the 0º cEOF1 in all levels, and with the 90º cEOF1 and 90º cEOF2 in the troposphere.
@@
 
-(ref:sst-psi-2-cap) Regression of SST (K, left column ) and streamfunction zonal anomalies ($m^2/s\times10^-7$, shaded) with their corresponding activity wave flux (vectors) (right column) upon cEOF2 different phases (illustrated in the lower-left arrow) for the 1979 -- 2019 period.
+(ref:sst-psi-2-cap) Regression of SST (K, left column) and streamfunction zonal anomalies ($m^2/s\times10^-7$, shaded) with their corresponding activity wave flux (vectors) (right column) upon cEOF2 different phases (illustrated in the lower-left arrow) for the 1979 -- 2019 period.
 Areas marked with dots have p-values smaller than 0.01 adjusted for FDR.
@@
 The connections between cEOFs and tropical sources of variability were also assessed.
-Figure \@ref(fig:sst-psi-2) shows the regression maps of Sea Surface Temperatures (SST) and streamfunction anomalies at 200 hPa respectively upon standardised cEOF2.
-Besides the regression maps for the 0º and 90º phases, we include the corresponding regressions for two intermediate directions (corresponding to 45° and 135°).
+Figure \@ref(fig:sst-psi-2) shows the regression maps of Sea Surface Temperature (SST) and streamfunction anomalies at 200 hPa upon standardised cEOF2.
+As well as regression maps for the 0º and 90º phases, we include corresponding regressions for two intermediate directions (corresponding to 45° and 135°).
 
@@
 Indeed, there is a significant and very high correlation (`r enso_cor[cEOF == "cEOF2" & part == "Imaginary"]$text`) between the ONI and the 90º cEOF2 time series.
-Besides the Pacific ENSO-like pattern, there are positive anomalies in the western Indian Ocean and negative values in the eastern Indian Ocean, resembling a positive IOD [@saji1999].
+In addition to the Pacific ENSO-like pattern, there are also positive anomalies in the western Indian Ocean and negative values in the eastern Indian Ocean, resembling a positive IOD [@saji1999].
 Consistently, the correlation between the 90º cEOF2 and the DMI is `r dmi_cor[cEOF == "cEOF2" & part == "Imaginary"]$text`.
@@
 Figure\ \@ref(fig:sst-psi-2).d1 and d2 show that the 0º cEOF2 is not associated either with any significant SST nor streamfunction anomalies in the tropics. 
-As a result, the correlation between the 0º cEOF2 and ENSO is also not significant  (`r enso_cor[cEOF == "cEOF2" & part == "Real"]$text`).
+As a result, the correlation between the 0º cEOF2 and ENSO is not significant  (`r enso_cor[cEOF == "cEOF2" & part == "Real"]$text`).
 Meanwhile, Rows a and c in Fig.\ \@ref(fig:sst-psi-2) show that the intermediate phases are still associated with significant SST regressed anomalies over the Pacific Ocean, but at slightly different locations.
-The 135º phase is associated with SST anomalies in the central Pacific (Fig.\ \@ref(fig:sst-psi-2)a.1), while the 45º phase is associated with SST anomalies (Fig.\ \@ref(fig:sst-psi-2)c.1) that correspond roughly to the Central Pacific and Eastern Pacific "flavours" of ENSO, respectively [@kao2009].
-Both phases could be also associated to wave trains generated in the region surrounding Australia and propagates toward the extra-tropics.
+The 135º phase is associated with SST anomalies in the central Pacific (Fig.\ \@ref(fig:sst-psi-2)a.1), while the 45º phase is associated with SST anomalies that correspond roughly to the Central Pacific and Eastern Pacific "flavours" of ENSO, respectively (Fig.\ \@ref(fig:sst-psi-2)c.1)  [@kao2009].
+Both phases are also associated to wave trains generated in the region surrounding Australia and propagates toward the extra-tropics, although less intense than the ones assocaited with the 90º phase.
 
@@
 If those years are removed, the correlation becomes non-significant (`r cor_enso_mag_outlier$text`).
-Furthermore, even when using all years, the Spearman correlation --which is robust to outliers-- is also non-significant (`r signif(cor_enso_spearman$estimate, 2)`, p-value`r report_p(cor_enso_spearman$p.value)`).
+Furthermore, even when using all years, the Spearman correlation --which is robust to outliers-- is also non-significant (`r signif(cor_enso_spearman$estimate, 2)`, p-value `r report_p(cor_enso_spearman$p.value)`).
 Therefore, although the location of tropical SST anomalies seem to have an effect in defining the phase of the cEOF2, the relationship between the magnitude of cEOF2 and ONI remains uncertain and might be only evident in very strong ENSO events, that are scarce in the historical observational record.
 
-It could be concluded that the wave train represented by cEOF2 can be both part of the internal variability of the extratropical atmosphere or forced by tropical SSTs.
+We believe that the wave train represented by cEOF2 can be both part of the internal variability of the extratropical atmosphere and forced by tropical SSTs.
 In the former case, the wave train has little phase preference.
@@
 The influence of cEOFs variability in the anomalies of both 2-metre air temperature and precipitation in the SH was also explored.
-Figure\ \@ref(fig:pp-t2m-r2) shows the 2-meter temperature or precipitation anomalies explained variance by the multiple linear model of both 0º and 90º cEOF1 (column 1), and both 0º and 90º cEOF2 (column 2).
-The variance explained by cEOF1 for precipitation anomalies is extremely low and also for most of the temperature anomalies, except for the northern tip of the Antarctic Peninsula, northern Weddell Sea and the Ross Sea coast (Fig.\  \@ref(fig:pp-t2m-r2)a.1).
+Figure\ \@ref(fig:pp-t2m-r2) shows the 2-meter temperature and precipitation anomalies explained variance by the multiple linear model of both 0º and 90º cEOF1 (column 1), and both 0º and 90º cEOF2 (column 2).
+The variance explained by cEOF1 for precipitation anomalies and temperature anomalies in most regions is extremely low, except for the northern tip of the Antarctic Peninsula, northern Weddell Sea and the Ross Sea coast (Fig.\  \@ref(fig:pp-t2m-r2)a.1).
 
@@
 
-While the cEOF2 variability could be related to surface impacts, the cEOF1 influence is almost negligible.
+While the cEOF2 variability is related to surface impacts, the cEOF1 influence is almost negligible.
 For instance, precipitation anomalies in South America associated with the 90º cEOF2 show a clear ENSO-like impact, with positive anomalies in South-Eastern South America, negative anomalies in Southern Brazil and positive anomalies in central Chile for positive 90º cEOF2 phase.
@@
 
-Most studies rely on correlations between an ENSO index and the SAM index [e.g.
+Most studies on the relationship between ENSO and SAM rely on correlations between an ENSO index and the SAM index [e.g.
 @lheureux2006, @cai2011a] or between the SAM index and other variables associated with tropical convection, such as OLR or tropical SSTS [e.g.
@@
 @campitelli2022 showed that the correlation between ENSO and SAM is almost completely explained by the asymmetric component of the SAM. 
-In this work which we show that the asymmetric component of the SAM can be identified with the PSA1. 
-Therefore, the correlation between ENSO and SAM is predominantly the correlation between ENSO and PSA1 (the partial correlation between the SAM index and the ONI controlling for the 90º cEOF2 is `r partial_cor_enso[term == "sam", scales::number(partial_correlation, accuracy = 0.01)]`, p-value = `r partial_cor_enso[term == "sam", scales::pvalue(p.value, accuracy = 0.01)]`).
-This sheds new light into the previous literature, as one cannot assume that a correlation between ENSO and SAM indicates a relationship between ENSO and a zonally symmetric component.
+In this work we show that the asymmetric component of the SAM can be identified with the PSA1. 
+Therefore, the correlation between ENSO and SAM is predominantly the correlation between ENSO and PSA1 (the partial correlation between the SAM index and the ONI, controlling for the 90º cEOF2 is non significant at `r partial_cor_enso[term == "sam", scales::number(partial_correlation, accuracy = 0.01)]`, p-value = `r partial_cor_enso[term == "sam", scales::pvalue(p.value, accuracy = 0.01)]`).
+This sheds new light into the previous literature, as one cannot assume that a high correlation between ENSO and SAM indexes indicates a relationship between ENSO and zonally symmetric variability.
 
 Further investigation is necessary to determine the connection between the symmetric component of the SAM and the PSA. 
-It is possible that the PSA may force a zonally symmetric response (or vice versa) via wave-zonal mean flow interactions, or that this correlation is simply a statistical artefact resulting from the EOF methodology used to define the SAM.
-
-@irving2016 argued that there is some disagreement in the literature of whether the phase of the PSA pattern is affected by the location of tropical SS   
+It is possible that the PSA may force a zonally symmetric response (or vice versa) via wave-zonal mean flow interactions, or that this correlation is simply a statistical artefact resulting from the EOF methodology used to define the SAM and the fact that the spatial structure of the PSA projects onto the spatial structure of the symmetric SAM.
+
+@irving2016 argued that there is some disagreement in the literature of whether the phase of the PSA pattern is affected by the location of tropical SST anomalies.
+With the methodology used in this study, we were able to show not only that the cEOF2 tends to be in the positive or negative 90º phase (\~PSA1) when the ENSO region is warm or cold, respectively, but also that central Pacific SST anomalies tend to align the cEOF towards the negative 0º phase and eastern Pacific SST anomalies tend to align it towards positive 0º phase.
+When ENSO phase is neutral, the cEOF2 is still as active, but with no preferred phase.
+The latter agrees with the results of @cai2002, who showed that the CSIRO Model can develop PSA-like variability even in the absence of ENSO forcing (i.e.
+with a climatological run), but that the variability of one of the PSA modes was enhanced when adding the ENSO signal.
 The sensitivity of the phase of the PSAs to the location of the tropical SST anomalies was also seen by @ciasto2015, who detected similar Rossby wave patterns associated with central Pacific and eastern Pacific SST anomalies but with a change in phase.
 
-The method used in this study have similarities to the one used by @goyal2022 as they construct an index of amplitude and phase of zonal wave 3-like variability by combining the two leading EOFs of meridional wind anomalies.
+The method used in this study has similarities to the one used by @goyal2022 as they construct an index of amplitude and phase of zonal wave 3-like variability by combining the two leading EOFs of meridional wind anomalies.
 The patterns obtained by them bear high resemblance with cEOF2. 
@@
 
-All data used in this paper available in a version-controlled repository [@campitelli2022a]. 
+All data used in this paper available in a Zenodo repository [@campitelli2022a] (<https://zenodo.org/record/6612429>).
+Indices updated monthly and daily will be made available at http://www.cima.fcen.uba.ar/~elio.campitelli/shceof/.
+
 It is also freely available from their respective sources:

### 2023-01-04 analysis/paper/shceof.tex
Todas las respuestas

@@
 
-\abstract{The large-scale extratropical circulation in the Southern Hemisphere is strongly zonally symmetric, but its zonal departures, albeit highly relevant for regional impacts, have been less studied.
+\abstract{The large-scale extratropical circulation in the Southern Hemisphere is much more zonally symmetric than that of the Northern Hemisphere, but its zonal departures, albeit highly relevant for regional impacts, have been less studied.
 In this study we analyse the joint variability of the zonally asymmetric springtime stratospheric and tropospheric circulation using Complex Empirical Orthogonal Functions (cEOF) to characterise planetary waves of varying amplitude and phase.
@@
 
-The large-scale extratropical circulation in the Southern Hemisphere (SH) is strongly zonally symmetric, but its zonal departures are highly relevant for regional impacts \citep[e.g.][]{hoskins2005}.
-They strongly modulate weather systems and regional climate through promoting longitudinally different latitudinal transport of heat, humidity, and momentum \citep{trenberth1980a, raphael2007} and could even be related to the occurrence of high-impact climate extremes \citep{pezza2012}.
+The large-scale extratropical circulation in the Southern Hemisphere (SH) is much more zonally symmetric than that of the Northern Hemisphere, but departures from the zonal mean are associated with regional impacts \citep[e.g.][]{hoskins2005}.
+They strongly modulate weather systems and regional climate through promoting longitudinally varying meridional transport of heat, humidity, and momentum \citep{trenberth1980a, raphael2007} and could even be related to the occurrence of high-impact climate extremes \citep{pezza2012}.
 
-Zonally asymmetric circulation is typically described by the amplitude and phase of zonal waves obtained by Fourier decomposition of geopotential heights or sea-level pressures at each latitude \citep[e.g.][]{vanloon1972, trenberth1980a, turner2017}.
-This approach suggests that zonal waves 1 and 3 explain almost 99\% of the total variance in the annual mean 500-hPa pattern at 50ºS \citep{vanloon1972}.
+The zonally asymmetric circulation is typically described by the amplitude and phase of zonal waves obtained by Fourier decomposition of geopotential height or sea-level pressure at each latitude \citep[e.g.][]{vanloon1972, trenberth1980a, turner2017}.
+This approach suggests that zonal waves 1 and 3 explain almost 99\% of the total variance in the annual mean 500 hPa geopotential height zonal anomalies at 50ºS \citep{vanloon1972}.
 \citet{trenberth1985} concluded that wave 3 plays a role in the development of blocking events.
-In addition, previous works have identified at extratropical and subpolar latitudes, wave-like patterns with dominant wavenumbers 3-4, also exerting distinctive regional impacts.
-\citet{raphael2007} shows that variability in the planetary wave 3 projected onto its climatological location is associated with anomalies in the Antarctic sea-ice concentration.
+In addition, previous works have identified wave-like patterns with dominant wavenumbers 3-4 at extratropical and subpolar latitudes with distinctive regional impacts.
+\citet{raphael2007} showed that variability in the planetary wave 3 projected onto its climatological location is associated with anomalies in the Antarctic sea-ice concentration.
 
-The Fourier decomposition relies on the assumption that the circulation can be meaningfully described in terms of zonal waves of constant amplitude along a latitude circle.
+Fourier decomposition relies on the assumption that the circulation can be meaningfully described in terms of zonal waves of constant amplitude along a latitude circle.
 However, this is not valid for meridionally propagating waves or zonal waves with localised amplitudes.
-Addressing this limitation, the Fourier technique can be generalized to integrate all planetary wave amplitude regardless of wave number, by computing the wave envelope \citep{irving2015}.
-The latter makes it possible to represent planetary waves with different amplitude at different longitudes, but it removes all information about phase and wave number.
-With this approach, \citet{irving2015} showed that planetary wave amplitude in general is associated to Antarctic sea-ice concentration and temperature, as well as to precipitation anomalies in regions of significant topography in SH mid-latitudes and Antarctica.
+Addressing this limitation, the Fourier technique can be generalized to integrate all planetary wave amplitude regardless of wave number by computing the wave envelope \citep{irving2015}.
+The wave envelope can represent planetary waves with different amplitude at different longitudes, but lacks information about phase and wave number.
+Using this method, \citet{irving2015} showed that planetary wave amplitude in general is associated to Antarctic sea-ice concentration and temperature, as well as to precipitation anomalies in regions of significant topography in SH mid-latitudes and Antarctica.
 
-Another extensively used approach to characterise the SH tropospheric circulation anomalies, is by computing Empirical Orthogonal Functions (EOF, also known as Principal Component Analysis).
+Another extensively-used approach to characterise the SH tropospheric circulation anomalies, is computing Empirical Orthogonal Functions (EOF, also known as Principal Component Analysis).
 Within the EOF framework, the Southern Annular Mode (SAM) appears as the leading mode of variability of the SH circulation \citep{fogt2020}.
-SAM represents a relatively zonally symmetric pattern of alternating low pressures in polar latitude and a ring of high pressures in high latitudes with an embedded wave 3 pattern that is more prominent in the Pacific sector.
+The SAM represents a relatively zonally symmetric pattern of alternating low pressures in polar latitudes and a ring of high pressures in high latitudes with an embedded wave 3 pattern that is more prominent in the Pacific sector.
 The 2nd and 3rd EOFs, usually known as Pacific--South American Patterns (PSA) 1 and PSA2 patterns, respectively, describe meridionally propagating wave trains that originate in the eastern equatorial Pacific and Australian-Indian Ocean sector, and travel towards the South Atlantic following a great-circle arch along the Antarctic coast \citep{mo2001}.
@@
 
-Some of the zonally asymmetric patterns of the SH circulation variability described previously, appear to have experienced secular changes.
-For instance, \citet{raphael2003} found that the amplitude of the zonal wave 1 experienced a large increase and that the zonal wave 3 experienced changes in its annual cycle between 1958 and 1996.
+Some of the zonally asymmetric patterns of the SH circulation variability described previously appear to have experienced secular changes.
+For instance, \citet{raphael2003} suggests that the amplitude of the zonal wave 1 experienced a large increase and that the zonal wave 3 experienced changes in its annual cycle between 1958 and 1996.
 However, little is known yet about variability and trends of these patterns.
 
-Patterns resulting from EOF analyses are more flexible than Fourier decomposition derived modes in the sense that they can capture oscillation patterns that cannot be characterised by purely sinusoidal waves with constant amplitude.
-Nonetheless, they are restricted to standing oscillation modes and could not properly represent propagating or phase-varying features such as zonal waves.
+Patterns resulting from EOF analysis are more flexible than Fourier decomposition-derived modes in that they can capture oscillation patterns that cannot be characterised by purely sinusoidal waves with constant amplitude.
+Nonetheless, they are restricted to standing oscillation modes and cannot properly represent propagating or phase-varying features such as zonal waves.
 A single EOF can also represent a mixture of two or more physical modes.
@@
 These derived methods are grounded on other methods such as Fourier decomposition or EOF to identify the centres of action for the described phenomena and can be useful to characterise features that are not readily apparent with these methods.
-These kinds of indices are generally easy to compute, but they usually capture non-stationary patterns.
+These kinds of indices are generally easy to compute, but they usually don't capture non-stationary patterns.
 
-An alternative methodology that has been proposed to study travelling and standing waves is complex Empirical Orthogonal Functions {[}cEOF; \citet{horel1984}{]}.
+An alternative methodology that has been proposed to study travelling and standing waves is complex Empirical Orthogonal Functions (cEOF; \citet{horel1984}).
 This method extends EOF analysis to capture oscillations with varying amplitude and phase and has been applied to the time domain.
-For instance, \citet{krokhin2007} applied cEOF to station-based monthly precipitation anomalies and monthly temperature anomalies in the Eastern Siberia and the Far East region to characterise the main modes of variability and their connection to teleconnection indices.
-Similarly, \citet{gelbrecht2018} used cEOF applied to daily precipitation from reanalysis to study the propagating characteristics of the South American Monsoon.
+For instance, \citet{krokhin2007} applied cEOF to station-based monthly precipitation anomalies and monthly temperature anomalies in the Eastern Siberia and the Far East region to characterise the main modes of variability and their relationship with teleconnection indices.
+Similarly, \citet{gelbrecht2018} applied cEOF to daily precipitation from reanalysis to study the propagating characteristics of the South American Monsoon.
 To our knowledge, cEOF analysis has not been applied in the spatial domain to capture the phase-varying nature of planetary waves in the atmosphere.
 
-The general goal of this study is to improve the description and understanding of the zonally asymmetric extratropical SH circulation using cEOF, which allow to describe phase varying planetary waves with variable amplitudes along a latitude circle.
-In addition, it is proposed to expand the knowledge of the simultaneous behaviour of SH asymmetric circulation in the troposphere and the stratosphere.
+The general goal of this study is to improve the description and understanding of the zonally asymmetric extratropical SH circulation using cEOF, which can describe phase varying planetary waves with variable amplitude along a latitude circle.
+In addition, we try to expand the knowledge of the simultaneous behaviour of SH asymmetric circulation in the troposphere and the stratosphere.
 
-We restrict this paper to the spring September-October-November (SON) trimester because during this season both the tropical teleconnections over South America \citep{cazes-boezio2003} and the SH stratosphere-troposphere interactions \citep{lim2018} are at their maximum amplitude.
+We restrict this paper to the September-October-November (SON) trimester because during this season both the tropical teleconnections over South America \citep{cazes-boezio2003} and the SH stratosphere-troposphere interactions \citep{lim2018} are at their maximum amplitude.
 
@@
 
-We used monthly geopotential height, air temperature, ozone mixing ratio, and total column ozone (TOC) at 2.5\degree longitude by 2.5\degree latitude of horizontal resolution and 37 vertical isobaric levels from the European Centre for Medium-Range Weather Forecasts Reanalysis version 5 {[}ERA; \citet{era5}{]} for the period 1979 -- 2019.
-Most of our analysis is restricted to the post-satellite era to avoid any confounding factors arising from the incorporation of satellite observations, but we also used the preliminary back extension of ERA5 from 1950 to 1978 \citep{era5be} to describe long-term trends.
+We used monthly geopotential height, air temperature, ozone mixing ratio, and total column ozone (TOC) at 2.5º longitude by 2.5º latitude of horizontal resolution and 37 vertical isobaric levels from the European Centre for Medium-Range Weather Forecasts Reanalysis version 5 {[}ERA; \citet{era5}{]} for the period 1979 -- 2019.
+Most of our analysis is restricted to the post-satellite era to avoid confounding factors arising from the incorporation of satellite observations, but we also used the preliminary back extension of ERA5 from 1950 to 1978 \citep{era5be} to describe long-term trends.
 We derived streamfunction at 200 hPa from ERA5 vorticity using the FORTRAN subroutine FISHPACK \citep{fishpack} and we computed horizontal wave activity fluxes following \citet{plumb1985}.
-Sea Surface Temperature (SST) monthly fields are from Extended Reconstructed Sea Surface Temperature (ERSST) v5 \citep{huang2017} and precipitation monthly data from the CPC Merged Analysis of Precipitation \citep{cmap}, with a 2\degree and 2.5\degree horizontal resolution respectively.
-The rainfall gridded dataset is based on information from different sources such as rain gauge observations, satellite inferred estimations and the NCEP-NCAR reanalysis, and it is available since 1979 to the present.
+Sea Surface Temperature (SST) monthly fields are from Extended Reconstructed Sea Surface Temperature (ERSST) v5 \citep{huang2017} and precipitation monthly data from the CPC Merged Analysis of Precipitation \citep{cmap}, with a 2º and 2.5º horizontal resolution, respectively.
+The rainfall gridded dataset is based on information from different sources such as rain gauge observations, satellite inferred estimations and the NCEP-NCAR reanalysis, and it is available from 1979 to the present.
 
@@
 
-The amplitude of the zonal waves was obtained through the Fourier transform of the spatial field at each latitude circle.
-For the analysis of wave 1, we computed its amplitude and phase by averaging (area-weighted) the data for each variable and each SON between 75°S and 45°S, and then extracting the wave-1 component of the Fourier spectrum.
+We computed the amplitude and phase of the TOC wave 1 by averaging (area-weighted) the data of each SON between 75°S and 45°S, and then extracting the wave-1 component of the Fourier spectrum.
 We chose this latitude band because it is wide enough to capture most of the relevant anomalies of SH mid-latitudes.
@@
 \includegraphics{../figures/eof-naive-1.pdf}
-\caption{\label{fig:eof-naive}Spatial patterns of the four leading EOFs of SON zonal anomalies of geopotential height at 50 hPa south of 20\degree S for the 1979 -- 2019 period (arbitrary units).}
+\caption{\label{fig:eof-naive}Spatial patterns of the four leading EOFs of SON geopotential height zonal anomalies of at 50 hPa south of 20º S for the 1979 -- 2019 period (arbitrary units).}
 \end{figure}
@@
 In the standard EOF analysis, zonal waves may appear as pairs of (possibly degenerate) EOFs representing similar patterns but shifted in phase \citep{horel1984}.
-Figure \ref{fig:eof-naive} shows the four leading EOFs of SON geopotential height zonal anomalies at 50 hPa south of 20\degree S.
-It is clear that the first two EOFs represent a single phase-varying zonal wave 1 pattern and the last two represent a similarly phase-varying pattern with higher wavenumber and three centres of action shifted by 1/4 wavelength (90\degree in frequency space).
+Figure \ref{fig:eof-naive} shows the four leading EOFs of SON geopotential height zonal anomalies at 50 hPa south of 20º S.
+It is clear that the first two EOFs represent a single phase-varying zonal wave 1 pattern and the last two represent a similarly phase-varying pattern with higher wavenumber and three centres of action shifted by 1/4 wavelength (90º in frequency space).
 
@@
 So, for instance, the amplitude of the wave 1-like EOF could be measured as \(\sqrt{\mathrm{PC1}^2 + \mathrm{PC2}^2}\) and its phase as \(\tan^{-1} \left ( \frac{\mathrm{PC2}}{\mathrm{PC1}} \right )\) (where \(\mathrm{PC1}\) and \(\mathrm{PC2}\) are the time series associated with each EOF).
-However, this rests upon visual inspection of the spatial patterns and only works properly if both phases appear cleanly in different EOFs, which is not guaranteed by construction.
-In particular, this doesn't work with the wave 1 pattern observed as the leading EOF in 200hPa geopotential heihgt zonal anomalies (not shown).
+However, this rests on visual inspection of the spatial patterns and only works properly if both phases appear cleanly in different EOFs, which is not guaranteed by construction.
+In particular, this doesn't work with the wave 1 pattern observed as the leading EOF in 200 hPa geopotential height zonal anomalies (not shown).
 
-A better alternative for describing phase-varying waves is to use Complex Empirical Orthogonal Functions (cEOF) analysis (Horel, 1984).
+A better alternative for describing phase-varying waves is to use Complex Empirical Orthogonal Functions (cEOF) analysis \citep{horel1984}.
 Each cEOF is represented by two spatial patterns that are shifted by 1/4 wavelength by construction.
@@
 
-Figure \ref{fig:ceofs-1} a.1 shows the two spatial patterns of the cEOF corresponding to XXX.
-For each subplot one of the cEOFs is plotted with shaded contours (which represents the 0º phase) and the other with black contours (which represents the 90º phase).
-This cEOF is very similar to the two leading EOFs shown in Figure \ref{fig:eof-naive} and represents a zonal wave 1 pattern.
+Figure \ref{fig:ceofs-1} a.1 shows the two spatial patterns of the cEOF of geopotential height zonal anomalies.
+One of the cEOFs is plotted with shaded contours (which represents the 0º phase) and the other with black contours (which represents the 90º phase).
+This cEOF is very similar to the two leading EOFs shown in Figure \ref{fig:eof-naive} and represents a zonal wave 1 pattern; the 0º phase is roughtly the EOF1 and the 90º phase is roughtly the EOF2).
 For instance, when the phase of the wave matches the 0º phase, then the 0º phase time series will be positive and the 90º phase time series will be zero.
@@
 The two phases described are not special and the same cEOF could be represented by any linear combinations of the phases shown in Figure \ref{fig:ceofs-1} a.1.
-This parallels traditional EOFs, which are defined up to an additive constant.
-For real EOFs, this constant can be either 0 or \(\pi\), corresponding to a change in sign.
-On the other hand, for cEOFs, it can be any real number between 0 and \(2\pi\) \citep{horel1984}, corresponding to rotations in the complex plane.
+This parallels traditional EOFs, which are defined up to sign, which corresponds to a rotation in the complex plane of either 0 or \(\pi\).
+On the other hand cEOFs are defined up to a rotation in the complex plane of any value between 0 and \(2\pi\) \citep{horel1984}.
 
 cEOFs are computed in the same way as traditional EOFs except that the data is first augmented by computing its analytic signal.
-This is a complex number whose real part is the original series and the imaginary part is the original data shifted by 90º at each spectral frequency -- i.e.~
+This is a complex number whose real part is the original series and whose imaginary part is the original data shifted by 90º at each spectral frequency -- i.e.~
 its Hilbert transform.
@@
 The Hilbert transform is usually understood in terms of time-varying signal.
-However, in this work we apply the Hilbert transform at each latitude circle and at each considered time.
+However, in this work we apply the Hilbert transform at each latitude circle and at each considered time (e.g.~the signal only depends on longitude).
 Since each latitude circle is a periodic domain, this procedure does not suffer from edge effects.
@@
 
-The cEOF methodology is applied to zonal anomalies of SON geopotential height south of 20\degree S separately at 50 and 200 hPa.
+First, we applied the cEOF methodology is applied to zonal anomalies of SON geopotential height south of 20º S separately at 50 and 200 hPa.
 Table \ref{tab:corr-ceof-splitted} shows the coefficient of determination between time series of the amplitude of each cEOF across levels.
@@
 As mentioned before, the choice of phases is arbitrary and equally valid.
-But to make the interpretation easier, we chose the phase of each cEOF so that the either the º0 cEOF or the 90º cEOF is aligned with meaningful variables in our analysis.
+But to make the interpretation easier, we chose the phase of each cEOF so that the either the 0º cEOF or the 90º cEOF is aligned with meaningful variables in our analysis.
 This procedure does not create spurious correlations, it only takes an existing relationship and aligns it with a specific phase.
+
+Preliminary analysis showed that the first cEOF was closely related to the zonal wave 1 of Total Column Ozone and the second cEOF was closely related to ENSO.
+Therefore, we chose the phase of cEOF1 so that the time series corresponding to the 0º cEOF1 has the maximum correlation with the zonal wave 1 of Total Column Ozone between 75°S and 45°S.
+Similarly, we chose the phase of cEOF2 so that the coefficient of determination between the Oceanic Niño Index \citep{bamston1997} and the 0º cEOF2 is minimised, which also nearly maximises the correlation with the 90º cEOF2.
 
@@
 For those plots, we rotated the cEOFs by 1/4 wavelength by multiplying the complex time series by \(\cos(\pi/4) + i\sin(\pi/4)\) and computing the regression on those rotated timeseries.
-
-Preliminary analysis showed that the first cEOF was closely related to the zonal wave 1 of Total Column Ozone and the second cEOF was closely related to ENSO.
-Therefore, we chose the phase of cEOF1 so that the time series corresponding to the 0ª cEOF1 has the maximum correlation with the zonal wave 1 of Total Column Ozone between 75°S and 45°S.
-Similarly, we chose the phase of cEOF2 so that the coefficient of determination between the Oceanic Niño Index \citep{bamston1997} and the º0 cEOF2 is minimised, which also nearly maximises the correlation with the 90º cEOF2.
 
@@
 \includegraphics{../figures/ceofs-1-1.pdf}
-\caption{\label{fig:ceofs-1}Spatial patterns for the two leading cEOFs of SON zonal anomalies of geopotential height at 50 hPa and 200 hPa for the 1979 -- 2019 period.
+\caption{\label{fig:ceofs-1}Spatial patterns for the two leading cEOFs of SON geopotential height zonal anomalies at 50 hPa and 200 hPa for the 1979 -- 2019 period.
 The shading (contours) corresponds to 0º (90º) phase.
@@
 \begin{figure*}
-\includegraphics{../figures/extended-series-1} \caption{Time series of the two leading cEOFs of SON zonal anomalies of geopotential height at 50 hPa and 200 hPa.
+\includegraphics{../figures/extended-series-1} \caption{Time series of the two leading cEOFs of SON geopotential height zonal anomalies at 50 hPa and 200 hPa.
 cEOF1 (row a) and cEOF2 (row b) separated in their 0º (column 1) and 90º (column 2) phase.
@@
 
-The cEOF1 (Fig.~\ref{fig:ceofs-1} column~a) is a hemispheric wave 1 pattern with maximum amplitude at high latitudes.
+The cEOF1 (Fig.~\ref{fig:ceofs-1} column~1) is a hemispheric wave 1 pattern with maximum amplitude at high latitudes.
 At 50 hPa the 0º cEOF1 has the maximum of the wave 1 at 150ºE and at 200 hPa, the maximum is located at around 175ºE indicating a westward shift with height.
-The cEOF2 (Fig.~\ref{fig:eof-naive} column~b) shows also a zonal wave-like structure with maximum amplitude at high latitudes, but with shorter spatial scales.
+The cEOF2 (Fig.~\ref{fig:eof-naive} column~2) shows also a zonal wave-like structure with maximum amplitude at high latitudes, but with shorter spatial scales.
 In particular, the dominant structure at both levels is a wave 3 but with larger amplitude in the pacific sector.
-This modulated amplitude is more evident at 200 hPa, which is consistent with the cEOF2 computed separatedly for 200 hPa explains a bit more variance than the cEOF2 computed separatedly for 50 hPa (11\% vs.~3\%, respectively).
-There is no apparent phase shift with height but the amplitude of the pattern is greatly reduced in the stratosphere, suggesting that this barotropic mode represents mainly tropospheric variability.
+There is no apparent phase shift with height but the amplitude of the pattern is greatly reduced in the stratosphere, which is consistent with the the fact that the cEOF2 computed separately for 200 hPa explains a bit more variance than the cEOF2 computed separately for 50 hPa (11\% vs.~3\%, respectively).
+This suggest that this barotropic mode represents mainly tropospheric variability.
 
@@
 Figure~\ref{fig:eof1-regr-gh} shows regression maps of SON geopotential height anomalies upon cEOF1.
-At 50 hPa (Figure~\ref{fig:eof1-regr-gh} row a), the 0º cEOF1 is associated with a centre located over the Ross Sea.
+At 50 hPa (Figure~\ref{fig:eof1-regr-gh} row a), the 0º cEOF1 is associated with a positive centre located over the Ross Sea.
 The 90º cEOF1 is associated with a distinctive wave 1 pattern with maximum over the coast of East Antarctica.
@@
 Figure~\ref{fig:eof2-regr-gh} shows the regression maps of geopotential height anomalies upon the cEOF2.
-In the troposphere (Fig.~\ref{fig:eof2-regr-gh} row b) the regression maps show wave trains similar to those identified for cEOF2 patterns (Fig~\ref{fig:ceofs-1}).
+In the troposphere (Fig.~\ref{fig:eof2-regr-gh} row a) the regression maps show wave trains similar to those identified for cEOF2 patterns (Fig~\ref{fig:ceofs-1}).
 Regressed anomalies associated with the 0º cEOF2 are 1/4 wavelength out of phase with those associated with the 90º cEOF2.
@@
 cEOF2 then represents an equivalent barotropic wave train that is very similar to the the PSA Patterns \citep{mo2001}.
-Comparing the location of the positive anomaly near 90ºW in column b of Figure~\ref{fig:eof2-regr-gh} with Figures 1.a and b from \citet{mo2001}, the 0º cEOF2 regression map could be identified with PSA2, while the 90º cEOF2 resembles PSA1.
+Comparing the location of the positive anomaly near 90ºW in column 2 of Figure~\ref{fig:eof2-regr-gh} with Figures 1.a and b from \citet{mo2001}, the 0º cEOF2 regression map could be identified with PSA2, while the 90º cEOF2 resembles PSA1.
 
@@
 In both levels, the 0º cEOF1 is associated with a positive centre over the South Pole with its centre moved slightly towards 150ºE (Fig.~\ref{fig:eof1-regr-t} column 1).
-On the other hand, the regression maps on the 90º cEOF1 show a more clear wave 1 pattern with its maximum around 60ºE.
+On the other hand, the regression maps on the 90º cEOF1 show a clearer wave 1 pattern with its maximum around 60ºE.
 
-Figure~\ref{fig:t-vertical} shows the vertical distribution of the regression coefficients on cEOF1 from zonal anomalies averaged between 75°S and 45°S of air temperature and of ozone mixing ratio.
+Figure~\ref{fig:t-vertical} shows the vertical distribution of the regression coefficients on cEOF1 from zonal anomalies of air temperature and of ozone mixing ratio averaged between 75°S and 45°S.
 Temperature zonal anomalies associated with cEOF1 show a clear wave 1 pattern for both 0º and 90º phases throughout the atmosphere above 250 hPa with a sign reversal above 10 hPa.
@@
 
-The maximum ozone regressed anomalies match with the minimum temperature anomalies above 10 hPa and with the maximum temperature anomalies below 10 hPa (Fig.~\ref{fig:t-vertical}).
+The maximum ozone regressed anomalies coincide with the minimum temperature anomalies above 10 hPa and with the maximum temperature anomalies below 10 hPa (Fig.~\ref{fig:t-vertical}).
 Therefore, the ozone zonal wave 1 is negatively correlated with the temperature zonal wave 1 in the upper stratosphere, and positively correlated in the upper stratosphere.
@@
 Thus, the 0º cEOF1 regression field (Figure~\ref{fig:o3-regr}a) coincides with the climatological position of the ozone hole, while it is 90° out of phase for the 90º cEOF1.
-The temporal correlation between the amplitudes of TOC planetary wave 1 and cEOF1 is 0.79 (CI: 0.63 -- 0.88), while the correlation between their phases is -0.85 (CI: -0.92 -- -0.74).
+The temporal correlation between the amplitudes of TOC planetary wave 1 and the amplitude of cEOF1 is 0.79 (CI: 0.63 -- 0.88), while the correlation between their phases is -0.85 (CI: -0.92 -- -0.74).
 Consequently, cEOF1 is strongly related with the SH ozone variability.
@@
 
-Given the similarity between the cEOF2 related anomaly structures (Fig.~\ref{fig:eof2-regr-gh}) and documented PSA patterns we study the relationship between them.
+Given the similarity between the cEOF2 related-associated structures (Fig.~\ref{fig:eof2-regr-gh}) and documented PSA patterns, we study the relationship between them.
 Table~\ref{tab:psa-eof2} shows the correlations between the two PSA indices and the time series for 0º and 90º phases of cEOF2.
@@
 Conventional EOF analysis will therefore tend to separate them relatively cleanly, with the EOF representing the 90º cEOF2 always leading the one representing the 0º cEOF2.
-This phase preferences is in agreement with \citet{irving2016}, who found a bimodal distribution to PSA-like variability (compare our Figure~\ref{fig:phase-histogram} with their Figure 6).
+This phase preference is in agreement with \citet{irving2016}, who found a bimodal distribution to PSA-like variability (compare our Figure~\ref{fig:phase-histogram} with their Figure 6).
 
@@
 We now explore the relationship between SAM and the cEOFs motivated by the resemblance between cEOFs regression maps and SAM patterns shown in Section~\ref{regressions}.
-We then computed the coefficient of determination between the cEOFs time series and the three SAM indices (SAM, A-SAM and S-SAM) defined by \citet{campitelli2022} at each vertical level (Fig.~\ref{fig:sam-eof-vertical}).
+We computed the coefficient of determination between the cEOFs time series and the three SAM indices (SAM, A-SAM and S-SAM) defined by \citet{campitelli2022} at each vertical level (Fig.~\ref{fig:sam-eof-vertical}).
 The SAM index is statistically significantly correlated with the 0º cEOF1 in all levels, and with the 90º cEOF1 and 90º cEOF2 in the troposphere.
@@
 \begin{figure*}
-\includegraphics{../figures/sst-psi-2-1} \caption{Regression of SST (K, left column ) and streamfunction zonal anomalies (\(m^2/s\times10^-7\), shaded) with their corresponding activity wave flux (vectors) (right column) upon cEOF2 different phases (illustrated in the lower-left arrow) for the 1979 -- 2019 period.
+\includegraphics{../figures/sst-psi-2-1} \caption{Regression of SST (K, left column) and streamfunction zonal anomalies (\(m^2/s\times10^-7\), shaded) with their corresponding activity wave flux (vectors) (right column) upon cEOF2 different phases (illustrated in the lower-left arrow) for the 1979 -- 2019 period.
 Areas marked with dots have p-values smaller than 0.01 adjusted for FDR.}\label{fig:sst-psi-2}
@@
 The connections between cEOFs and tropical sources of variability were also assessed.
-Figure~\ref{fig:sst-psi-2} shows the regression maps of Sea Surface Temperatures (SST) and streamfunction anomalies at 200 hPa respectively upon standardised cEOF2.
-Besides the regression maps for the 0º and 90º phases, we include the corresponding regressions for two intermediate directions (corresponding to 45° and 135°).
+Figure~\ref{fig:sst-psi-2} shows the regression maps of Sea Surface Temperature (SST) and streamfunction anomalies at 200 hPa upon standardised cEOF2.
+As well as regression maps for the 0º and 90º phases, we include corresponding regressions for two intermediate directions (corresponding to 45° and 135°).
 
@@
 Indeed, there is a significant and very high correlation (0.76 (CI: 0.6 -- 0.87)) between the ONI and the 90º cEOF2 time series.
-Besides the Pacific ENSO-like pattern, there are positive anomalies in the western Indian Ocean and negative values in the eastern Indian Ocean, resembling a positive IOD \citep{saji1999}.
+In addition to the Pacific ENSO-like pattern, there are also positive anomalies in the western Indian Ocean and negative values in the eastern Indian Ocean, resembling a positive IOD \citep{saji1999}.
 Consistently, the correlation between the 90º cEOF2 and the DMI is 0.62 (CI: 0.38 -- 0.78).
@@
 Figure~\ref{fig:sst-psi-2}.d1 and d2 show that the 0º cEOF2 is not associated either with any significant SST nor streamfunction anomalies in the tropics.
-As a result, the correlation between the 0º cEOF2 and ENSO is also not significant (0 (CI: -0.3 -- 0.3)).
+As a result, the correlation between the 0º cEOF2 and ENSO is not significant (0 (CI: -0.3 -- 0.3)).
 Meanwhile, Rows a and c in Fig.~\ref{fig:sst-psi-2} show that the intermediate phases are still associated with significant SST regressed anomalies over the Pacific Ocean, but at slightly different locations.
-The 135º phase is associated with SST anomalies in the central Pacific (Fig.~\ref{fig:sst-psi-2}a.1), while the 45º phase is associated with SST anomalies (Fig.~\ref{fig:sst-psi-2}c.1) that correspond roughly to the Central Pacific and Eastern Pacific ``flavours'' of ENSO, respectively \citep{kao2009}.
-Both phases could be also associated to wave trains generated in the region surrounding Australia and propagates toward the extra-tropics.
+The 135º phase is associated with SST anomalies in the central Pacific (Fig.~\ref{fig:sst-psi-2}a.1), while the 45º phase is associated with SST anomalies that correspond roughly to the Central Pacific and Eastern Pacific ``flavours'' of ENSO, respectively (Fig.~\ref{fig:sst-psi-2}c.1) \citep{kao2009}.
+Both phases are also associated to wave trains generated in the region surrounding Australia and propagates toward the extra-tropics, although less intense than the ones assocaited with the 90º phase.
 
@@
 If those years are removed, the correlation becomes non-significant (0.04 (CI: -0.28 -- 0.35)).
-Furthermore, even when using all years, the Spearman correlation --which is robust to outliers-- is also non-significant (0.2, p-value= 0.21).
+Furthermore, even when using all years, the Spearman correlation --which is robust to outliers-- is also non-significant (0.2, p-value = 0.21).
 Therefore, although the location of tropical SST anomalies seem to have an effect in defining the phase of the cEOF2, the relationship between the magnitude of cEOF2 and ONI remains uncertain and might be only evident in very strong ENSO events, that are scarce in the historical observational record.
 
-It could be concluded that the wave train represented by cEOF2 can be both part of the internal variability of the extratropical atmosphere or forced by tropical SSTs.
+We believe that the wave train represented by cEOF2 can be both part of the internal variability of the extratropical atmosphere and forced by tropical SSTs.
 In the former case, the wave train has little phase preference.
@@
 The influence of cEOFs variability in the anomalies of both 2-metre air temperature and precipitation in the SH was also explored.
-Figure~\ref{fig:pp-t2m-r2} shows the 2-meter temperature or precipitation anomalies explained variance by the multiple linear model of both 0º and 90º cEOF1 (column 1), and both 0º and 90º cEOF2 (column 2).
-The variance explained by cEOF1 for precipitation anomalies is extremely low and also for most of the temperature anomalies, except for the northern tip of the Antarctic Peninsula, northern Weddell Sea and the Ross Sea coast (Fig.~ \ref{fig:pp-t2m-r2}a.1).
+Figure~\ref{fig:pp-t2m-r2} shows the 2-meter temperature and precipitation anomalies explained variance by the multiple linear model of both 0º and 90º cEOF1 (column 1), and both 0º and 90º cEOF2 (column 2).
+The variance explained by cEOF1 for precipitation anomalies and temperature anomalies in most regions is extremely low, except for the northern tip of the Antarctic Peninsula, northern Weddell Sea and the Ross Sea coast (Fig.~ \ref{fig:pp-t2m-r2}a.1).
 
@@
 
-Most studies rely on correlations between an ENSO index and the SAM index \citep[e.g.][\citet{cai2011a}]{lheureux2006} or between the SAM index and other variables associated with tropical convection, such as OLR or tropical SSTS \citep[e.g.][]{carvalho2005}.
+Most studies on the relationship between ENSO and SAM rely on correlations between an ENSO index and the SAM index \citep[e.g.][\citet{cai2011a}]{lheureux2006} or between the SAM index and other variables associated with tropical convection, such as OLR or tropical SSTS \citep[e.g.][]{carvalho2005}.
 \citet{campitelli2022} showed that the correlation between ENSO and SAM is almost completely explained by the asymmetric component of the SAM.
-In this work which we show that the asymmetric component of the SAM can be identified with the PSA1.
-Therefore, the correlation between ENSO and SAM is predominantly the correlation between ENSO and PSA1 (the partial correlation between the SAM index and the ONI controlling for the 90º cEOF2 is 0.20, p-value = 0.23).
-This sheds new light into the previous literature, as one cannot assume that a correlation between ENSO and SAM indicates a relationship between ENSO and a zonally symmetric component.
+In this work we show that the asymmetric component of the SAM can be identified with the PSA1.
+Therefore, the correlation between ENSO and SAM is predominantly the correlation between ENSO and PSA1 (the partial correlation between the SAM index and the ONI, controlling for the 90º cEOF2 is non significant at 0.20, p-value = 0.23).
+This sheds new light into the previous literature, as one cannot assume that a high correlation between ENSO and SAM indexes indicates a relationship between ENSO and zonally symmetric variability.
 
 Further investigation is necessary to determine the connection between the symmetric component of the SAM and the PSA.
-It is possible that the PSA may force a zonally symmetric response (or vice versa) via wave-zonal mean flow interactions, or that this correlation is simply a statistical artefact resulting from the EOF methodology used to define the SAM.
+It is possible that the PSA may force a zonally symmetric response (or vice versa) via wave-zonal mean flow interactions, or that this correlation is simply a statistical artefact resulting from the EOF methodology used to define the SAM and the fact that the spatial structure of the PSA projects onto the spatial structure of the symmetric SAM.
 
-\citet{irving2016} argued that there is some disagreement in the literature of whether the phase of the PSA pattern is affected by the location of tropical SS\\
+\citet{irving2016} argued that there is some disagreement in the literature of whether the phase of the PSA pattern is affected by the location of tropical SST anomalies.
+With the methodology used in this study, we were able to show not only that the cEOF2 tends to be in the positive or negative 90º phase (\textasciitilde PSA1) when the ENSO region is warm or cold, respectively, but also that central Pacific SST anomalies tend to align the cEOF towards the negative 0º phase and eastern Pacific SST anomalies tend to align it towards positive 0º phase.
+When ENSO phase is neutral, the cEOF2 is still as active, but with no preferred phase.
+The latter agrees with the results of \citet{cai2002}, who showed that the CSIRO Model can develop PSA-like variability even in the absence of ENSO forcing (i.e.~with a climatological run), but that the variability of one of the PSA modes was enhanced when adding the ENSO signal.
 The sensitivity of the phase of the PSAs to the location of the tropical SST anomalies was also seen by \citet{ciasto2015}, who detected similar Rossby wave patterns associated with central Pacific and eastern Pacific SST anomalies but with a change in phase.
 
-The method used in this study have similarities to the one used by \citet{goyal2022} as they construct an index of amplitude and phase of zonal wave 3-like variability by combining the two leading EOFs of meridional wind anomalies.
+The method used in this study has similarities to the one used by \citet{goyal2022} as they construct an index of amplitude and phase of zonal wave 3-like variability by combining the two leading EOFs of meridional wind anomalies.
 The patterns obtained by them bear high resemblance with cEOF2.
@@
 
-All data used in this paper available in a version-controlled repository \citep{campitelli2022a}.
+All data used in this paper available in a Zenodo repository \citep{campitelli2022a} (\url{https://zenodo.org/record/6612429}).
+Indices updated monthly and daily will be made available at \url{http://www.cima.fcen.uba.ar/~elio.campitelli/shceof/}.
+
 It is also freely available from their respective sources:

### 2023-01-10 analysis/paper/shceof.Rmd
Revisión

@@
 classoptions: 
+  - lineno
   - sn-basic
@@
 output: 
-  bookdown::word_document2:
-      pandoc_args: ["--lua-filter=abstract-to-meta.lua", "--wrap=preserve"]
+  # bookdown::word_document2:
+  #     pandoc_args: ["--lua-filter=abstract-to-meta.lua", "--wrap=preserve"]
   bookdown::pdf_document2:
@@
   cache.path = "../cache/paper/",
-  fig.path = "../figures/",
+  # fig.path = "../figures/",
   dpi = 300,
@@
 These derived methods are grounded on other methods such as Fourier decomposition or EOF to identify the centres of action for the described phenomena and can be useful to characterise features that are not readily apparent with these methods.
-These kinds of indices are generally easy to compute, but they usually don't capture non-stationary patterns.
+These kinds of indices are generally easy to compute, but they usually do not capture non-stationary patterns.
 
@@
 
-We restrict this paper to the September-October-November (SON) trimester because during this season both the tropical teleconnections over South America [@cazes-boezio2003] and the SH stratosphere-troposphere interactions [@lim2018] are at their maximum amplitude.
+We restrict this work to the September-October-November (SON) trimester because during this season both the tropical teleconnections over South America [@cazes-boezio2003] and the SH stratosphere-troposphere interactions [@lim2018] are at their maximum amplitude.
 
@@
 We derived streamfunction at 200 hPa from ERA5 vorticity using the FORTRAN subroutine FISHPACK [@fishpack] and we computed horizontal wave activity fluxes following @plumb1985.
-Sea Surface Temperature (SST) monthly fields are from Extended Reconstructed Sea Surface Temperature (ERSST) v5 [@huang2017] and precipitation monthly data from the CPC Merged Analysis of Precipitation [@cmap], with a 2º and 2.5º horizontal resolution, respectively.
+Sea Surface Temperature (SST) monthly fields are from Extended Reconstructed Sea Surface Temperature (ERSST) v5 [@huang2017] and precipitation monthly data from the CPC Merged Analysis of Precipitation [CMAP, @cmap], with a 2º and 2.5º horizontal resolution, respectively.
 The rainfall gridded dataset is based on information from different sources such as rain gauge observations, satellite inferred estimations and the NCEP-NCAR reanalysis, and it is available from 1979 to the present.
@@
 
-(ref:eof-naive-cap) Spatial patterns of the four leading EOFs of SON geopotential height zonal anomalies of at 50 hPa south of 20º S for the 1979 -- 2019 period (arbitrary units).
+(ref:eof-naive-cap) Spatial patterns of the four leading EOFs of SON geopotential height zonal anomalies at 50 hPa south of 20º S for the 1979 -- 2019 period (arbitrary units).
 
@@
 
-To describe the phase-varying nature of these two waves, the simplest solution is to combine each pair of EOFs into indices of amplitude and phase.
+To describe the phase-varying nature of these two wave patterns, one way is to combine each pair of EOFs into indices of amplitude and phase.
 So, for instance, the amplitude of the wave 1-like EOF could be measured as $\sqrt{\mathrm{PC1}^2 + \mathrm{PC2}^2}$ and its phase as $\tan^{-1} \left ( \frac{\mathrm{PC2}}{\mathrm{PC1}} \right )$ (where $\mathrm{PC1}$ and $\mathrm{PC2}$ are the time series associated with each EOF).
-However, this rests on visual inspection of the spatial patterns and only works properly if both phases appear cleanly in different EOFs, which is not guaranteed by construction.
-In particular, this doesn't work with the wave 1 pattern observed as the leading EOF in 200 hPa geopotential height zonal anomalies (not shown).
-
-A better alternative for describing phase-varying waves is to use Complex Empirical Orthogonal Functions (cEOF) analysis [@horel1984]. 
-Each cEOF is represented by two spatial patterns that are shifted by 1/4 wavelength by construction.
-The actual field reconstructed by each cEOF is the linear combination of the two spatial fields weighted by its respective time series. 
+However, this rests on visual inspection of the spatial patterns and only works properly if both phases appear clearly in different EOFs, which is not guaranteed by construction.
+In particular, this does not work with the wave 1 pattern depicted as the leading EOF in 200 hPa geopotential height zonal anomalies (not shown).
+
+On the other hand, a better alternative for describing phase-varying waves is to use Complex Empirical Orthogonal Functions (cEOF) analysis [@horel1984]. 
+Each cEOF is a set of complex-valued spatial patterns and time series. 
+The real and imaginary components of the complex spatial pattern can be though of as representing two spatial patterns that are shifted by 1/4 wavelength by construction, similar to EOF1 and EOF2 in Figure \@ref(fig:eof-naive).
+In this paper we use the term 0º cEOF and 90º cEOF to refer to each part of the whole cEOF.
+The actual field reconstructed by each cEOF is then the linear combination of the two spatial fields weighted by its respective time series. 
 This is analogous to how any sine wave can be constructed by the sum of a sine wave and cosine wave with different amplitude but constant phase. 
@@
 
-Figure \@ref(fig:ceofs-1) a.1 shows the two spatial patterns of the cEOF of geopotential height zonal anomalies. 
-One of the cEOFs is plotted with shaded contours (which represents the 0º phase) and the other with black contours (which represents the 90º phase). 
-This cEOF is very similar to the two leading EOFs shown in Figure \@ref(fig:eof-naive) and represents a zonal wave 1 pattern; the 0º phase is roughtly the EOF1 and the 90º phase is roughtly the EOF2).
-For instance, when the phase of the wave matches the 0º phase, then the 0º phase time series will be positive and the 90º phase time series will be zero.
-Similarly, when the phase of the wave matches the 90º phase, the 90º phase time series will be positive and the 0º phase time series will be zero.
-The intermediate phases will have non zero values in both time series.
-The two phases described are not special and the same cEOF could be represented by any linear combinations of the phases shown in Figure \@ref(fig:ceofs-1) a.1. 
-This parallels traditional EOFs, which are defined up to sign, which corresponds to a rotation in the complex plane of either 0 or $\pi$.
-On the other hand cEOFs are defined up to a rotation in the complex plane of any value between 0 and $2\pi$ [@horel1984].
+For instance, when the phase of the wave matches the 0º phase, then the 0º phase time series is positive and the 90º phase time series is zero.
+Similarly, when the phase of the wave matches the 90º phase, the 90º phase time series is positive and the 0º phase time series is zero.
+The intermediate phases have non-zero values in both time series.
+
+Traditional EOFs are not unique, and instead are defined up to sign, which corresponds to a rotation in the complex plane of either 0 or $\pi$.
+In the same way, cEOFs are defined up to a rotation in the complex plane of any value between 0 and $2\pi$ [@horel1984].
 
 cEOFs are computed in the same way as traditional EOFs except that the data is first augmented by computing its analytic signal. 
-This is a complex number whose real part is the original series and whose imaginary part is the original data shifted by 90º at each spectral frequency -- i.e. 
-its Hilbert transform.
-Each cEOF is a set of complex-valued spatial patterns and time series in which the real (or 0º phase) and imaginary (or 90º phase) parts are the two orthogonal phases of the wave-like pattern.
-In this paper we use the term 0º cEOF and 90º cEOF to refer to each part of the whole cEOF.
-
+This is a complex number whose real part is the original series and whose imaginary part is the original data shifted by 90º at each spectral frequency -- i.e. its Hilbert transform. 
 The Hilbert transform is usually understood in terms of time-varying signal.
-However, in this work we apply the Hilbert transform at each latitude circle and at each considered time (e.g. the signal only depends on longitude).
+However, in this work we apply the Hilbert transform at each latitude circle and at each considered time (i.e. the signal only depends on longitude).
 Since each latitude circle is a periodic domain, this procedure does not suffer from edge effects.
+
+We first applied cEOF analysis to geopotential height zonal anomalies south of 20ºS at 50 and at 200 hPa. 
+Figure \@ref(fig:ceofs-1) a.1 shows the spatial patterns of the two leading cEOF. 
+The 0º phase is plotted with shaded contours and the 90º phase, with black contours.
+The two phases of the leading cEOF are very similar to the two leading EOFs shown in Figure \@ref(fig:eof-naive) and represent a zonal wave 1 pattern; the 0º phase is roughly the EOF1 and the 90º phase is roughly the EOF2).
 
@@
 
-First, we applied the cEOF methodology is applied to zonal anomalies of SON geopotential height south of 20º S separately at 50 and 200 hPa.
 Table \@ref(tab:corr-ceof-splitted) shows the coefficient of determination between time series of the amplitude of each cEOF across levels.
@@
 As mentioned before, the choice of phases is arbitrary and equally valid. 
-But to make the interpretation easier, we chose the phase of each cEOF so that the either the 0º cEOF or the 90º cEOF is aligned with meaningful variables in our analysis.
+But to make the interpretation easier, we chose the phase of each cEOF so that either the 0º cEOF or the 90º cEOF is aligned with meaningful variables in our analysis.
 This procedure does not create spurious correlations, it only takes an existing relationship and aligns it with a specific phase.
 
-Preliminary analysis showed that the first cEOF was closely related to the zonal wave 1 of Total Column Ozone and the second cEOF was closely related to ENSO.
-Therefore, we chose the phase of cEOF1 so that the time series corresponding to the 0º cEOF1 has the maximum correlation with the zonal wave 1 of Total Column Ozone between `r knitr::combine_words(LatLabel(o3wave_lats))`.
-Similarly, we chose the phase of cEOF2 so that the coefficient of determination between the Oceanic Niño Index [@bamston1997] and the 0º cEOF2 is minimised, which also nearly maximises the correlation with the 90º cEOF2.
+Preliminary analysis showed that the first cEOF was closely related to the zonal wave 1 of TOC and the second cEOF was closely related to ENSO.
+Therefore, we chose the phase of cEOF1 so that the time series corresponding to the 0º cEOF1 has the maximum correlation with the zonal wave 1 of TOC between `r knitr::combine_words(LatLabel(o3wave_lats))`.
+Similarly, we chose the phase of cEOF2 so that the coefficient of determination between the ONI and the 0º cEOF2 is minimised, which also nearly maximises the correlation with the 90º cEOF2.
 
@@
 The 135º phase is associated with SST anomalies in the central Pacific (Fig.\ \@ref(fig:sst-psi-2)a.1), while the 45º phase is associated with SST anomalies that correspond roughly to the Central Pacific and Eastern Pacific "flavours" of ENSO, respectively (Fig.\ \@ref(fig:sst-psi-2)c.1)  [@kao2009].
-Both phases are also associated to wave trains generated in the region surrounding Australia and propagates toward the extra-tropics, although less intense than the ones assocaited with the 90º phase.
+Both phases are also associated to wave trains generated in the region surrounding Australia and propagates toward the extra-tropics, although less intense than the ones associated with the 90º phase.
 
@@
 
-We believe that the wave train represented by cEOF2 can be both part of the internal variability of the extratropical atmosphere and forced by tropical SSTs.
+We conclude that the wave train represented by cEOF2 can be both part of the internal variability of the extratropical atmosphere and forced by tropical SSTs.
 In the former case, the wave train has little phase preference.
@@
 First, the correlation between cEOF1 and the SAM in the troposphere is modest, with less than 50% of shared variance (Fig.
-\@ref(fig:sam-eof-vertical) column 1), so one need not expect these indices to be equivalent.
+\@ref(fig:sam-eof-vertical) column 1), so these indices are not expected to be equivalent.
 Second, @campitelli2022 showed that the strong relationship between the SAM and Pacific SSTs and temperature anomalies around the Antarctic Peninsula is mainly due to the asymmetric part of the SAM.
@@
 For precipitation, there are high values over the tropics. At higher latitudes, moderate values are observed over eastern Australia and some regions of southern South America.
-
-Since the cEOF1 has a relatively weak signal in the surface variables explored here, we will only focus on the cEOF2 influence.
 
@@
 
+Since the cEOF1 has a relatively weak signal in the surface variables explored here, we will only focus on the cEOF2 influence.
 Figure\ \@ref(fig:pp-temp-2) shows regression maps of 2-metre temperature (column 1) and precipitation (column 2) anomalies upon different phases of standardised cEOF2.
@@
 Temperature anomalies associated with the 90º cEOF (Fig.\ \@ref(fig:pp-temp-2).b1) show positive values in the tropical Pacific, consistent with SSTs anomalies associated with the same phase (Fig.\ \@ref(fig:sst-psi-2).b1).
-
 At higher latitudes there is a wave-like pattern of positive and negative values that coincide with the nodes of the 850 hPa geopotential height regression patterns.
 This is consistent with temperature anomalies produced by meridional advection of temperature by the meridional winds arising from geostrophic balance.
-
 Over the continents, the 90º cEOF2 (Fig.\ \@ref(fig:pp-temp-2)b.1) is associated with positive regressed temperature anomalies in southern Australia and negative regressed anomalies in southern South America and the Antarctic Peninsula, that are a result of the wave train described before.
@@
 In the extra-tropics, the positive 90º cEOF2 is related to drier conditions over eastern Australia and the surrounding ocean, that it is similar signal as the one associated with ENSO [@cai2011].
-However, the 90º cEOF2 is not the direction most correlated with precipitation in that area. 
+However, the 90º cEOF2 is not the phase most correlated with precipitation in that area. 
 The 135º phase (an intermediate between positive 90º and 180º cEOF2) component is associated with stronger and more extensive temporal correlations  with precipitation over Australia and New Zealand. 
@@
 Over South America, the 90º cEOF2 has positive correlations with precipitation in South Eastern South America (SESA) and central Chile, and negative correlations in eastern Brazil. 
-This correlation field  matches the springtime precipitation signature of ENSO [e.g.
-@cai2020a] and it is also similar to the precipitation anomalies associated with the A-SAM [@campitelli2022].
-
+This correlation field  matches the springtime precipitation signature of ENSO [e.g. @cai2020a] and it is also similar to the precipitation anomalies associated with the A-SAM [@campitelli2022].
 This result is not surprising considering the close relationship of the 90º cEOF2 with both ONI and A-SAM index, which was shown previously.
@@
 In this study we assessed extratropical Southern Hemisphere zonally asymmetric circulation in austral spring. 
-For this purpose, were derived two complex indices using Complex Empirical Orthogonal Functions and used to characterise both amplitude and phase of planetary waves.
+For this purpose, we derived two complex indices using Complex Empirical Orthogonal Functions and used to characterise both amplitude and phase of planetary waves.
 
@@
 The 90º cEOF2 can be identified with the PSA1 and the 0º cEOF2 with the PSA2. 
-
-While the cEOF2 variability is related to surface impacts, the cEOF1 influence is almost negligible.
-For instance, precipitation anomalies in South America associated with the 90º cEOF2 show a clear ENSO-like impact, with positive anomalies in South-Eastern South America, negative anomalies in Southern Brazil and positive anomalies in central Chile for positive 90º cEOF2 phase.
+While the cEOF2 variability is related to surface impacts, the cEOF1 surface influence is almost negligible.
+For instance, precipitation anomalies in South America associated with the 90º cEOF2 show a clear ENSO-like impact, with positive anomalies in South-eastern South America, negative anomalies in Southern Brazil and positive anomalies in central Chile for positive 90º cEOF2 phase.
 
@@
 @carvalho2005].
-@campitelli2022 showed that the correlation between ENSO and SAM is almost completely explained by the asymmetric component of the SAM. 
+However, @campitelli2022 showed that the correlation between ENSO and SAM is almost completely explained by the asymmetric component of the SAM. 
 In this work we show that the asymmetric component of the SAM can be identified with the PSA1. 
-Therefore, the correlation between ENSO and SAM is predominantly the correlation between ENSO and PSA1 (the partial correlation between the SAM index and the ONI, controlling for the 90º cEOF2 is non significant at `r partial_cor_enso[term == "sam", scales::number(partial_correlation, accuracy = 0.01)]`, p-value = `r partial_cor_enso[term == "sam", scales::pvalue(p.value, accuracy = 0.01)]`).
-This sheds new light into the previous literature, as one cannot assume that a high correlation between ENSO and SAM indexes indicates a relationship between ENSO and zonally symmetric variability.
+Therefore, the correlation between ENSO and SAM is predominantly the correlation between ENSO and PSA1.
+This sheds new light into the previous literature, as it cannot be assumed that a high correlation between ENSO and SAM indexes indicates a relationship between ENSO and zonally symmetric variability.
 
@@
 
-The research was supported by UBACyT20020170100428BA and the CLIMAX Project funded by Belmont Forum/ANR-15-JCL/-0002-01. Elio Campitelli was supported by a PhD grant from CONICET, Argentina.
+The research was supported by UBACyT20020170100428BA, Consejo Nacional de Investigaciones Científicas y Técnicas (CONICET) PIP 11220200102038CO, PICT-2020-SERIEA-I-INVI-00540, and the CLIMAX Project funded by Belmont Forum/ANR-15-JCL/-0002-01.
+Elio Campitelli was supported by a PhD grant from CONICET, Argentina.
 
@@
 
-E.C. and L.D. wrote the main manuscript text. E.C. prepared the figures. All authors reviewed the manuscript.
+E.C. made the data curation, formal analysis and prepared all the figures. E.C. and L.D. wrote the main manuscript text.
+All authors reviewed the manuscript.
 

### 2023-01-10 analysis/paper/shceof.tex
Revisión

@@
 
-\documentclass[pdflatex,sn-basic]{sn-jnl}
+\documentclass[pdflatex,lineno,sn-basic]{sn-jnl}
 %%\documentclass[sn-basic]{sn-jnl}% Basic Springer Nature Reference Style/Chemistry Reference Style
@@
 
-We restrict this paper to the September-October-November (SON) trimester because during this season both the tropical teleconnections over South America \citep{cazes-boezio2003} and the SH stratosphere-troposphere interactions \citep{lim2018} are at their maximum amplitude.
+We restrict this work to the September-October-November (SON) trimester because during this season both the tropical teleconnections over South America \citep{cazes-boezio2003} and the SH stratosphere-troposphere interactions \citep{lim2018} are at their maximum amplitude.
 
@@
 We derived streamfunction at 200 hPa from ERA5 vorticity using the FORTRAN subroutine FISHPACK \citep{fishpack} and we computed horizontal wave activity fluxes following \citet{plumb1985}.
-Sea Surface Temperature (SST) monthly fields are from Extended Reconstructed Sea Surface Temperature (ERSST) v5 \citep{huang2017} and precipitation monthly data from the CPC Merged Analysis of Precipitation \citep{cmap}, with a 2º and 2.5º horizontal resolution, respectively.
+Sea Surface Temperature (SST) monthly fields are from Extended Reconstructed Sea Surface Temperature (ERSST) v5 \citep{huang2017} and precipitation monthly data from the CPC Merged Analysis of Precipitation \citep[CMAP,][]{cmap}, with a 2º and 2.5º horizontal resolution, respectively.
 The rainfall gridded dataset is based on information from different sources such as rain gauge observations, satellite inferred estimations and the NCEP-NCAR reanalysis, and it is available from 1979 to the present.
@@
 \centering
-\includegraphics{../figures/eof-naive-1.pdf}
-\caption{\label{fig:eof-naive}Spatial patterns of the four leading EOFs of SON geopotential height zonal anomalies of at 50 hPa south of 20º S for the 1979 -- 2019 period (arbitrary units).}
+\includegraphics{shceof_files/figure-latex/eof-naive-1.pdf}
+\caption{\label{fig:eof-naive}Spatial patterns of the four leading EOFs of SON geopotential height zonal anomalies at 50 hPa south of 20º S for the 1979 -- 2019 period (arbitrary units).}
 \end{figure}
@@
 
-To describe the phase-varying nature of these two waves, the simplest solution is to combine each pair of EOFs into indices of amplitude and phase.
+To describe the phase-varying nature of these two wave patterns, one way is to combine each pair of EOFs into indices of amplitude and phase.
 So, for instance, the amplitude of the wave 1-like EOF could be measured as \(\sqrt{\mathrm{PC1}^2 + \mathrm{PC2}^2}\) and its phase as \(\tan^{-1} \left ( \frac{\mathrm{PC2}}{\mathrm{PC1}} \right )\) (where \(\mathrm{PC1}\) and \(\mathrm{PC2}\) are the time series associated with each EOF).
-However, this rests on visual inspection of the spatial patterns and only works properly if both phases appear cleanly in different EOFs, which is not guaranteed by construction.
-In particular, this doesn't work with the wave 1 pattern observed as the leading EOF in 200 hPa geopotential height zonal anomalies (not shown).
+However, this rests on visual inspection of the spatial patterns and only works properly if both phases appear clearly in different EOFs, which is not guaranteed by construction.
+In particular, this does not work with the wave 1 pattern depicted as the leading EOF in 200 hPa geopotential height zonal anomalies (not shown).
 
-A better alternative for describing phase-varying waves is to use Complex Empirical Orthogonal Functions (cEOF) analysis \citep{horel1984}.
-Each cEOF is represented by two spatial patterns that are shifted by 1/4 wavelength by construction.
-The actual field reconstructed by each cEOF is the linear combination of the two spatial fields weighted by its respective time series.
+On the other hand, a better alternative for describing phase-varying waves is to use Complex Empirical Orthogonal Functions (cEOF) analysis \citep{horel1984}.
+Each cEOF is a set of complex-valued spatial patterns and time series.
+The real and imaginary components of the complex spatial pattern can be though of as representing two spatial patterns that are shifted by 1/4 wavelength by construction, similar to EOF1 and EOF2 in Figure \ref{fig:eof-naive}.
+In this paper we use the term 0º cEOF and 90º cEOF to refer to each part of the whole cEOF.
+The actual field reconstructed by each cEOF is then the linear combination of the two spatial fields weighted by its respective time series.
 This is analogous to how any sine wave can be constructed by the sum of a sine wave and cosine wave with different amplitude but constant phase.
@@
 
-Figure \ref{fig:ceofs-1} a.1 shows the two spatial patterns of the cEOF of geopotential height zonal anomalies.
-One of the cEOFs is plotted with shaded contours (which represents the 0º phase) and the other with black contours (which represents the 90º phase).
-This cEOF is very similar to the two leading EOFs shown in Figure \ref{fig:eof-naive} and represents a zonal wave 1 pattern; the 0º phase is roughtly the EOF1 and the 90º phase is roughtly the EOF2).
-For instance, when the phase of the wave matches the 0º phase, then the 0º phase time series will be positive and the 90º phase time series will be zero.
-Similarly, when the phase of the wave matches the 90º phase, the 90º phase time series will be positive and the 0º phase time series will be zero.
-The intermediate phases will have non zero values in both time series.
-The two phases described are not special and the same cEOF could be represented by any linear combinations of the phases shown in Figure \ref{fig:ceofs-1} a.1.
-This parallels traditional EOFs, which are defined up to sign, which corresponds to a rotation in the complex plane of either 0 or \(\pi\).
-On the other hand cEOFs are defined up to a rotation in the complex plane of any value between 0 and \(2\pi\) \citep{horel1984}.
+For instance, when the phase of the wave matches the 0º phase, then the 0º phase time series is positive and the 90º phase time series is zero.
+Similarly, when the phase of the wave matches the 90º phase, the 90º phase time series is positive and the 0º phase time series is zero.
+The intermediate phases have non-zero values in both time series.
+
+Traditional EOFs are not unique, and instead are defined up to sign, which corresponds to a rotation in the complex plane of either 0 or \(\pi\).
+In the same way, cEOFs are defined up to a rotation in the complex plane of any value between 0 and \(2\pi\) \citep{horel1984}.
 
 cEOFs are computed in the same way as traditional EOFs except that the data is first augmented by computing its analytic signal.
-This is a complex number whose real part is the original series and whose imaginary part is the original data shifted by 90º at each spectral frequency -- i.e.~
-its Hilbert transform.
-Each cEOF is a set of complex-valued spatial patterns and time series in which the real (or 0º phase) and imaginary (or 90º phase) parts are the two orthogonal phases of the wave-like pattern.
-In this paper we use the term 0º cEOF and 90º cEOF to refer to each part of the whole cEOF.
+This is a complex number whose real part is the original series and whose imaginary part is the original data shifted by 90º at each spectral frequency -- i.e.~its Hilbert transform.
+The Hilbert transform is usually understood in terms of time-varying signal.
+However, in this work we apply the Hilbert transform at each latitude circle and at each considered time (i.e.~the signal only depends on longitude).
+Since each latitude circle is a periodic domain, this procedure does not suffer from edge effects.
 
-The Hilbert transform is usually understood in terms of time-varying signal.
-However, in this work we apply the Hilbert transform at each latitude circle and at each considered time (e.g.~the signal only depends on longitude).
-Since each latitude circle is a periodic domain, this procedure does not suffer from edge effects.
+We first applied cEOF analysis to geopotential height zonal anomalies south of 20ºS at 50 and at 200 hPa.
+Figure \ref{fig:ceofs-1} a.1 shows the spatial patterns of the two leading cEOF.
+The 0º phase is plotted with shaded contours and the 90º phase, with black contours.
+The two phases of the leading cEOF are very similar to the two leading EOFs shown in Figure \ref{fig:eof-naive} and represent a zonal wave 1 pattern; the 0º phase is roughly the EOF1 and the 90º phase is roughly the EOF2).
 
@@
 
-First, we applied the cEOF methodology is applied to zonal anomalies of SON geopotential height south of 20º S separately at 50 and 200 hPa.
 Table \ref{tab:corr-ceof-splitted} shows the coefficient of determination between time series of the amplitude of each cEOF across levels.
@@
 As mentioned before, the choice of phases is arbitrary and equally valid.
-But to make the interpretation easier, we chose the phase of each cEOF so that the either the 0º cEOF or the 90º cEOF is aligned with meaningful variables in our analysis.
+But to make the interpretation easier, we chose the phase of each cEOF so that either the 0º cEOF or the 90º cEOF is aligned with meaningful variables in our analysis.
 This procedure does not create spurious correlations, it only takes an existing relationship and aligns it with a specific phase.
 
-Preliminary analysis showed that the first cEOF was closely related to the zonal wave 1 of Total Column Ozone and the second cEOF was closely related to ENSO.
-Therefore, we chose the phase of cEOF1 so that the time series corresponding to the 0º cEOF1 has the maximum correlation with the zonal wave 1 of Total Column Ozone between 75°S and 45°S.
-Similarly, we chose the phase of cEOF2 so that the coefficient of determination between the Oceanic Niño Index \citep{bamston1997} and the 0º cEOF2 is minimised, which also nearly maximises the correlation with the 90º cEOF2.
+Preliminary analysis showed that the first cEOF was closely related to the zonal wave 1 of TOC and the second cEOF was closely related to ENSO.
+Therefore, we chose the phase of cEOF1 so that the time series corresponding to the 0º cEOF1 has the maximum correlation with the zonal wave 1 of TOC between 75°S and 45°S.
+Similarly, we chose the phase of cEOF2 so that the coefficient of determination between the ONI and the 0º cEOF2 is minimised, which also nearly maximises the correlation with the 90º cEOF2.
 
@@
 \centering
-\includegraphics{../figures/ceofs-1-1.pdf}
+\includegraphics{shceof_files/figure-latex/ceofs-1-1.pdf}
 \caption{\label{fig:ceofs-1}Spatial patterns for the two leading cEOFs of SON geopotential height zonal anomalies at 50 hPa and 200 hPa for the 1979 -- 2019 period.
@@
 \begin{figure*}
-\includegraphics{../figures/extended-series-1} \caption{Time series of the two leading cEOFs of SON geopotential height zonal anomalies at 50 hPa and 200 hPa.
+\includegraphics{shceof_files/figure-latex/extended-series-1} \caption{Time series of the two leading cEOFs of SON geopotential height zonal anomalies at 50 hPa and 200 hPa.
 cEOF1 (row a) and cEOF2 (row b) separated in their 0º (column 1) and 90º (column 2) phase.
@@
 \centering
-\includegraphics{../figures/eof1-regr-gh-1.pdf}
+\includegraphics{shceof_files/figure-latex/eof1-regr-gh-1.pdf}
 \caption{\label{fig:eof1-regr-gh}Regression of SON geopotential height anomalies (\(m^2s^{-1}\)) with the (column 1) 0º and (column 2) 90º phases of the first cEOF for the 1979 -- 2019 period at (row a) 50 hPa and (row b) 200 hPa.
@@
 \centering
-\includegraphics{../figures/eof2-regr-gh-1.pdf}
+\includegraphics{shceof_files/figure-latex/eof2-regr-gh-1.pdf}
 \caption{\label{fig:eof2-regr-gh}Same as Figure~\ref{fig:eof1-regr-gh} but for cEOF2.}
@@
 \centering
-\includegraphics{../figures/eof1-regr-t-1.pdf}
+\includegraphics{shceof_files/figure-latex/eof1-regr-t-1.pdf}
 \caption{\label{fig:eof1-regr-t}Same as Figure~\ref{fig:eof1-regr-gh} but for air temperature (K).}
@@
 \centering
-\includegraphics{../figures/t-vertical-1.pdf}
+\includegraphics{shceof_files/figure-latex/t-vertical-1.pdf}
 \caption{\label{fig:t-vertical}Regression of SON zonal anomalies averaged between 75°S and 45°S of mean air temperature (shaded, Kelvin) and ozone mixing ratio (contours, negative contours with dashed lines, labels in parts per billion by mass) with the (a) 0º and (b) 90º phase of the cEOF1 for the 1979 -- 2019 period.}
@@
 \centering
-\includegraphics{../figures/o3-regr-1.pdf}
+\includegraphics{shceof_files/figure-latex/o3-regr-1.pdf}
 \caption{\label{fig:o3-regr}Regression of SON mean Total Column Ozone anomalies (shaded, Dobson Units) with the (a) 0º and (b) 90º phases of the cEOF1 for the 1979 -- 2019 period.
@@
 \centering
-\includegraphics{../figures/phase-histogram-1.pdf}
+\includegraphics{shceof_files/figure-latex/phase-histogram-1.pdf}
 \caption{\label{fig:phase-histogram}Histogram of phase distribution of cEOF2 phase for the 1979 -- 2019 period.
@@
 \centering
-\includegraphics{../figures/sam-eof-vertical-1.pdf}
+\includegraphics{shceof_files/figure-latex/sam-eof-vertical-1.pdf}
 \caption{\label{fig:sam-eof-vertical}Coefficient of determination (\(r^2\)) between each component of cEOFs and the SAM, Asymmetric SAM (A-SAM) and Symmetric SAM (S-SAM) indices computed at each level for the 1979 -- 2019 period.
@@
 \begin{figure*}
-\includegraphics{../figures/sst-psi-2-1} \caption{Regression of SST (K, left column) and streamfunction zonal anomalies (\(m^2/s\times10^-7\), shaded) with their corresponding activity wave flux (vectors) (right column) upon cEOF2 different phases (illustrated in the lower-left arrow) for the 1979 -- 2019 period.
+\includegraphics{shceof_files/figure-latex/sst-psi-2-1} \caption{Regression of SST (K, left column) and streamfunction zonal anomalies (\(m^2/s\times10^-7\), shaded) with their corresponding activity wave flux (vectors) (right column) upon cEOF2 different phases (illustrated in the lower-left arrow) for the 1979 -- 2019 period.
 Areas marked with dots have p-values smaller than 0.01 adjusted for FDR.}\label{fig:sst-psi-2}
@@
 The 135º phase is associated with SST anomalies in the central Pacific (Fig.~\ref{fig:sst-psi-2}a.1), while the 45º phase is associated with SST anomalies that correspond roughly to the Central Pacific and Eastern Pacific ``flavours'' of ENSO, respectively (Fig.~\ref{fig:sst-psi-2}c.1) \citep{kao2009}.
-Both phases are also associated to wave trains generated in the region surrounding Australia and propagates toward the extra-tropics, although less intense than the ones assocaited with the 90º phase.
+Both phases are also associated to wave trains generated in the region surrounding Australia and propagates toward the extra-tropics, although less intense than the ones associated with the 90º phase.
 
@@
 \centering
-\includegraphics{../figures/enso-phase-1.pdf}
+\includegraphics{shceof_files/figure-latex/enso-phase-1.pdf}
 \caption{\label{fig:enso-phase}SON ONI values plotted against cEOF2 phase for the 1979 -- 2019 period.
@@
 \centering
-\includegraphics{../figures/pp-t2m-r2-1.pdf}
+\includegraphics{shceof_files/figure-latex/pp-t2m-r2-1.pdf}
 \caption{\label{fig:pp-t2m-r2}Explained variance (\(r^2\) as percentage) of 2-metre temperature (row a) and precipitation (row b) anomalies by the regression upon cEOF1 (column 1) and cEOF2 (column 2).}
@@
 First, the correlation between cEOF1 and the SAM in the troposphere is modest, with less than 50\% of shared variance (Fig.
-\ref{fig:sam-eof-vertical} column 1), so one need not expect these indices to be equivalent.
+\ref{fig:sam-eof-vertical} column 1), so these indices are not expected to be equivalent.
 Second, \citet{campitelli2022} showed that the strong relationship between the SAM and Pacific SSTs and temperature anomalies around the Antarctic Peninsula is mainly due to the asymmetric part of the SAM.
@@
 Since the cEOF1 has a relatively weak signal in the surface variables explored here, we will only focus on the cEOF2 influence.
-
 Figure~\ref{fig:pp-temp-2} shows regression maps of 2-metre temperature (column 1) and precipitation (column 2) anomalies upon different phases of standardised cEOF2.
@@
 \centering
-\includegraphics{../figures/pp-temp-2-1.pdf}
+\includegraphics{shceof_files/figure-latex/pp-temp-2-1.pdf}
 \caption{\label{fig:pp-temp-2}Regression of SON mean 2-meter temperature (K, shaded) and 850 hPa geopotential height (m, contours) (column 1), and precipitation (correlation, column 2) upon different phases of cEOF2.
@@
 Temperature anomalies associated with the 90º cEOF (Fig.~\ref{fig:pp-temp-2}.b1) show positive values in the tropical Pacific, consistent with SSTs anomalies associated with the same phase (Fig.~\ref{fig:sst-psi-2}.b1).
-
 At higher latitudes there is a wave-like pattern of positive and negative values that coincide with the nodes of the 850 hPa geopotential height regression patterns.
 This is consistent with temperature anomalies produced by meridional advection of temperature by the meridional winds arising from geostrophic balance.
-
 Over the continents, the 90º cEOF2 (Fig.~\ref{fig:pp-temp-2}b.1) is associated with positive regressed temperature anomalies in southern Australia and negative regressed anomalies in southern South America and the Antarctic Peninsula, that are a result of the wave train described before.
@@
 In the extra-tropics, the positive 90º cEOF2 is related to drier conditions over eastern Australia and the surrounding ocean, that it is similar signal as the one associated with ENSO \citep{cai2011}.
-However, the 90º cEOF2 is not the direction most correlated with precipitation in that area.
+However, the 90º cEOF2 is not the phase most correlated with precipitation in that area.
 The 135º phase (an intermediate between positive 90º and 180º cEOF2) component is associated with stronger and more extensive temporal correlations with precipitation over Australia and New Zealand.
@@
 This correlation field matches the springtime precipitation signature of ENSO \citep[e.g.][]{cai2020a} and it is also similar to the precipitation anomalies associated with the A-SAM \citep{campitelli2022}.
-
-This result is not surprising considering the close relationship of the 90º cEOF2 with both ONI and A-SAM index, which was shown previously.\\
+This result is not surprising considering the close relationship of the 90º cEOF2 with both ONI and A-SAM index, which was shown previously.
 Furthermore, it consolidates the identification of the cEOF2 with the PSA pattern.
@@
 In this study we assessed extratropical Southern Hemisphere zonally asymmetric circulation in austral spring.
-For this purpose, were derived two complex indices using Complex Empirical Orthogonal Functions and used to characterise both amplitude and phase of planetary waves.
+For this purpose, we derived two complex indices using Complex Empirical Orthogonal Functions and used to characterise both amplitude and phase of planetary waves.
 
@@
 The 90º cEOF2 can be identified with the PSA1 and the 0º cEOF2 with the PSA2.
-
-While the cEOF2 variability is related to surface impacts, the cEOF1 influence is almost negligible.
-For instance, precipitation anomalies in South America associated with the 90º cEOF2 show a clear ENSO-like impact, with positive anomalies in South-Eastern South America, negative anomalies in Southern Brazil and positive anomalies in central Chile for positive 90º cEOF2 phase.
+While the cEOF2 variability is related to surface impacts, the cEOF1 surface influence is almost negligible.
+For instance, precipitation anomalies in South America associated with the 90º cEOF2 show a clear ENSO-like impact, with positive anomalies in South-eastern South America, negative anomalies in Southern Brazil and positive anomalies in central Chile for positive 90º cEOF2 phase.
 
@@
 Most studies on the relationship between ENSO and SAM rely on correlations between an ENSO index and the SAM index \citep[e.g.][\citet{cai2011a}]{lheureux2006} or between the SAM index and other variables associated with tropical convection, such as OLR or tropical SSTS \citep[e.g.][]{carvalho2005}.
-\citet{campitelli2022} showed that the correlation between ENSO and SAM is almost completely explained by the asymmetric component of the SAM.
+However, \citet{campitelli2022} showed that the correlation between ENSO and SAM is almost completely explained by the asymmetric component of the SAM.
 In this work we show that the asymmetric component of the SAM can be identified with the PSA1.
-Therefore, the correlation between ENSO and SAM is predominantly the correlation between ENSO and PSA1 (the partial correlation between the SAM index and the ONI, controlling for the 90º cEOF2 is non significant at 0.20, p-value = 0.23).
-This sheds new light into the previous literature, as one cannot assume that a high correlation between ENSO and SAM indexes indicates a relationship between ENSO and zonally symmetric variability.
+Therefore, the correlation between ENSO and SAM is predominantly the correlation between ENSO and PSA1.
+This sheds new light into the previous literature, as it cannot be assumed that a high correlation between ENSO and SAM indexes indicates a relationship between ENSO and zonally symmetric variability.
 
@@
 \begin{figure*}
-\includegraphics{../figures/sst-psi-1-1} \caption{Same as Figure~\ref{fig:sst-psi-2} but for cEOF1.}\label{fig:sst-psi-1}
+\includegraphics{shceof_files/figure-latex/sst-psi-1-1} \caption{Same as Figure~\ref{fig:sst-psi-2} but for cEOF1.}\label{fig:sst-psi-1}
 \end{figure*}
@@
 \centering
-\includegraphics{../figures/pp-temp-1-1.pdf}
+\includegraphics{shceof_files/figure-latex/pp-temp-1-1.pdf}
 \caption{\label{fig:pp-temp-1}Same as Figure \ref{fig:pp-temp-2} but for cEOF1.}

### 2023-03-13 analysis/paper/shceof.Rmd
Segunda revisión

@@
 year: 2022
-title:  "Revisiting the Austral Spring Extratropical Southern Hemisphere zonally asymmetric circulation using complex Empirical Orthogonal Functions"
+title:  "Revisiting the zonally asymmetric extratropical circulation of the Southern Hemisphere spring using complex empirical orthogonal functions"
 titlerunning: SH zonally asymmetric circulation with cEOF
@@
   comment = "#>",
-  cache.extra = 44,
+  cache.extra = 41,
   cache.path = "../cache/paper/",
@@
 The second cEOF (cEOF2) is an alternative representation of the Pacific-South American modes. 
-One phase of this cEOF is also extremely correlated with the Asymmetric SAM (A-SAM) in the troposphere. 
-Springs with an active ENSO tend to phase-lock the cEOF2, but have no consistent impact on its magnitude.
+One phase of this cEOF is also very highly correlated with the Asymmetric SAM (A-SAM) in the troposphere. 
+Springs with an active ENSO tend to lock the cEOF2 to a specific phase, but have no consistent impact on its magnitude.
 Furthermore, we find indications that the location of Pacific Sea Surface Temperature anomalies affect the phase of the cEOF2. 
@@
 The wave envelope can represent planetary waves with different amplitude at different longitudes, but lacks information about phase and wave number.
-Using this method, @irving2015 showed that planetary wave amplitude in general is associated to Antarctic sea-ice concentration and temperature, as well as to precipitation anomalies in regions of significant topography in SH mid-latitudes and Antarctica.
+Using this method, @irving2015 showed that planetary wave amplitude in general is associated with Antarctic sea-ice concentration and temperature, as well as to precipitation anomalies in regions of significant topography in SH mid-latitudes and Antarctica.
 
@@
 Within the EOF framework, the Southern Annular Mode (SAM) appears as the leading mode of variability of the SH circulation [@fogt2020].
-The SAM represents a relatively zonally symmetric pattern of alternating low pressures in polar latitudes and a ring of high pressures in high latitudes with an embedded wave 3 pattern that is more prominent in the Pacific sector.
+Spatially, the SAM is characterised by a centre of geopotential anomalies over Antarctica surrounded by anomalies of the opposite sign at middle latitudes. 
+Embedded in this zonally symmetric structure is a wave 3 pattern that is more prominent in the Pacific sector.
 The 2nd and 3rd EOFs, usually known as Pacific--South American Patterns (PSA) 1 and PSA2 patterns, respectively, describe meridionally propagating wave trains that originate in the eastern equatorial Pacific and Australian-Indian Ocean sector, and travel towards the South Atlantic following a great-circle arch along the Antarctic coast [@mo2001].
@@
 
-We restrict this work to the September-October-November (SON) trimester because during this season both the tropical teleconnections over South America [@cazes-boezio2003] and the SH stratosphere-troposphere interactions [@lim2018] are at their maximum amplitude.
+We restrict this work to the September-October-November (SON) trimester.
+During this season the tropical teleconnections over South America are maximised [@cazes-boezio2003], and the SH zonal winds associated with the stratospheric polar vortex increase to peak in October and extend downward after that [@lim2018].
 
@@
 In Section\ \@ref(spatial) we analyse the spatial patterns of each complex EOF.
-In Section\ \@ref(regressions) we study the spatial regressions with geopotential height, temperature and ozone anomalies.
+In Section\ \@ref(regressions) we study the spatial regressions with geopotential height, temperature, and ozone anomalies.
 In Section\ \@ref(psa) and \@ref(sam) we analyse the relationship between cEOF2, the PSA and SAM modes.
@@
 
-We used monthly geopotential height, air temperature, ozone mixing ratio, and total column ozone (TOC) at 2.5º longitude by 2.5º latitude of horizontal resolution and 37 vertical isobaric levels from the European Centre for Medium-Range Weather Forecasts Reanalysis version 5 [ERA; @era5] for the period 1979 -- 2019.
+We used monthly geopotential height, air temperature, ozone mixing ratio, and total column ozone (TCO) at 2.5º longitude by 2.5º latitude of horizontal resolution and 37 vertical isobaric levels from the European Centre for Medium-Range Weather Forecasts Reanalysis version 5 [ERA; @era5] for the period 1979 -- 2019.
 Most of our analysis is restricted to the post-satellite era to avoid confounding factors arising from the incorporation of satellite observations, but we also used the preliminary back extension of ERA5 from 1950 to 1978 [@era5be] to describe long-term trends.
@@
 We compute seasonal means for the different variables, averaging monthly values weighted by the number of days in each month.
-We use the 200 hPa level to represent the high troposphere and 50 hPa to represent the lower stratosphere.
-
-We computed the amplitude and phase of the TOC wave 1 by averaging (area-weighted) the data of each SON between `r knitr::combine_words(LatLabel(o3wave_lats))`, and then extracting the wave-1 component of the Fourier spectrum.
+We use the 200 hPa level to represent the upper troposphere and 50 hPa to represent the lower stratosphere.
+
+We computed the amplitude and phase of the TCO wave 1 by averaging (area-weighted) the data of each SON between `r knitr::combine_words(LatLabel(o3wave_lats))`, and then extracting the wave-1 component of the Fourier spectrum.
 We chose this latitude band because it is wide enough to capture most of the relevant anomalies of SH mid-latitudes.
@@
 We computed the level-dependent SAM index as the leading EOF of year-round monthly geopotential height anomalies south of 20ºS at each level for the whole period [@baldwin2009].
-We further split the SAM into its zonally symmetric and zonally asymmetric components (S-SAM and A-SAM indices respectively).
-These indices were obtained by projecting monthly geopotential height fields onto the zonally asymmetric and zonally symmetric parts of the SAM spatial pattern, as proposed by @campitelli2022.
+We further split the SAM into its zonally symmetric and zonally asymmetric components (S-SAM and A-SAM indices respectively) following  @campitelli2022.
+The method consists in first computing the leading EOF of monthly geopotential height anomalies at each level and then computing the zonal mean and the zonal anomalies from its spatial pattern. 
+We then project each level's monthly geopotential height fields onto the corresponding EOF field, the zonally symmetric field and the zonally asymmetric fields to obtain time series corresponding to the full SAM, the symmetric SAM and the asymmetric SAM, respectively.
+
 Seasonal indices of the PSA patterns (PSA1 and PSA2) were calculated, in agreement with @mo2001, as the third and fourth EOFs of seasonal mean anomalies for 500-hPa geopotential heights at SH.
@@
 
-## Complex Empirical Orthogonal Functions (cEOF)
+## Complex Empirical Orthogonal Functions (cEOF) {#ceof-methods}
 
@@
 Each cEOF is a set of complex-valued spatial patterns and time series. 
-The real and imaginary components of the complex spatial pattern can be though of as representing two spatial patterns that are shifted by 1/4 wavelength by construction, similar to EOF1 and EOF2 in Figure \@ref(fig:eof-naive).
+The real and imaginary components of the complex spatial pattern can be thought of as representing two spatial patterns that are shifted by 1/4 wavelength by construction, similar to EOF1 and EOF2 in Figure \@ref(fig:eof-naive).
 In this paper we use the term 0º cEOF and 90º cEOF to refer to each part of the whole cEOF.
@@
 
-For instance, when the phase of the wave matches the 0º phase, then the 0º phase time series is positive and the 90º phase time series is zero.
-Similarly, when the phase of the wave matches the 90º phase, the 90º phase time series is positive and the 0º phase time series is zero.
+For instance, when the phase of the wave matches the 0º phase, then the 0º phase time series is positive, and the 90º phase time series is zero.
+Similarly, when the phase of the wave matches the 90º phase, the 90º phase time series is positive, and the 0º phase time series is zero.
 The intermediate phases have non-zero values in both time series.
 
-Traditional EOFs are not unique, and instead are defined up to sign, which corresponds to a rotation in the complex plane of either 0 or $\pi$.
+In traditional EOFs, the resulting modes are not unique, and instead are defined up to sign, which corresponds to a rotation in the complex plane of either 0 or $\pi$. 
 In the same way, cEOFs are defined up to a rotation in the complex plane of any value between 0 and $2\pi$ [@horel1984].
@@
 The Hilbert transform is usually understood in terms of time-varying signal.
-However, in this work we apply the Hilbert transform at each latitude circle and at each considered time (i.e. the signal only depends on longitude).
+However, in this work we apply the Hilbert transform at each latitude circle, level, and at each SON mean (i.e. the signal only depends on longitude).
 Since each latitude circle is a periodic domain, this procedure does not suffer from edge effects.
@@
 
+Because we are computing the cEOFs of zonal anomalies and not temporal anomalies, the cEOFs need to account for the time-mean zonal anomalies. 
+These will tend to be represented by the leading cEOF, which therefore will have a non-zero temporal mean.
+
 As mentioned before, the choice of phases is arbitrary and equally valid. 
@@
 
-Preliminary analysis showed that the first cEOF was closely related to the zonal wave 1 of TOC and the second cEOF was closely related to ENSO.
-Therefore, we chose the phase of cEOF1 so that the time series corresponding to the 0º cEOF1 has the maximum correlation with the zonal wave 1 of TOC between `r knitr::combine_words(LatLabel(o3wave_lats))`.
+Preliminary analysis showed that the first cEOF was closely related to the zonal wave 1 of TCO and the second cEOF was closely related to ENSO.
+Therefore, we chose the phase of cEOF1 so that the time series corresponding to the 0º cEOF1 has the maximum correlation with the zonal wave 1 of TCO between `r knitr::combine_words(LatLabel(o3wave_lats))`.
 Similarly, we chose the phase of cEOF2 so that the coefficient of determination between the ONI and the 0º cEOF2 is minimised, which also nearly maximises the correlation with the 90º cEOF2.
@@
 $$
-X(\lambda, \phi, t) = \alpha(\lambda, \phi) \operatorname{cEOF_{0º}} + \beta(\lambda, \phi) \operatorname{cEOF_{90º}} + X_0(\lambda, \phi) + \epsilon(\lambda, \phi, t)
+X(\lambda, \phi, t) = \alpha(\lambda, \phi) \operatorname{cEOF_{0^\circ}} + \beta(\lambda, \phi) \operatorname{cEOF_{90^\circ}} + X_0(\lambda, \phi) + \epsilon(\lambda, \phi, t)
 $$
@@
 Both cEOFs show year-to-year variability but show no evidence of decadal variability (Fig. \@ref(fig:extended-series)).
-Because the geopotential fields that enter into the cEOFs algorithm are anomalies with respect to the zonal mean instead of the time mean, the cEOFs time series have non zero temporal mean. 
-However, cEOF2 temporal mean is almost zero, which indicates that only cEOF1 includes variability that significantly projects onto the mean zonally anomalous field.
-This is consistent with the fact that the mean zonally anomalous field of geopotential height is very similar to the cEOF1 ($r^2$ = `r scales::percent(ceof_mean_r2[["r.squared"]][1])`) and not similar to the cEOF2 ($r^2$ = `r scales::percent(ceof_mean_r2[["r.squared"]][2])`).
+The 0º cEOF has a non-zero temporal mean which, as discussed in Section \@ref(ceof-methods), is due to the fact that the temporal mean of zonal anomalies need to be captured by the cEOFs. 
+The other indices have almost zero temporal mean, which indicates that only cEOF1 includes variability that significantly projects onto the mean zonal anomalies. 
+This is consistent with the fact that the mean zonal anomalies of geopotential height are very similar to the cEOF1 ($r^2$ = `r scales::percent(ceof_mean_r2[["r.squared"]][1])`) and not similar to the cEOF2 ($r^2$ = `r scales::percent(ceof_mean_r2[["r.squared"]][2])`).
 
@@
 
+```{r jet_correlation}
+jet <- ERA5_jet() %>% 
+  ReadNetCDF() %>% 
+  normalise_coords() %>% 
+  .[lat == -60] %>% 
+  .[, .(jet = mean(u)), by = .(time)] %>% 
+  .[, .(jet = mean(jet)), by = .(time = seasonally(time))]
+
+jet_correlation <- ceof[, eof[[1]]$left, by = season] %>% 
+  .[jet, on = 'time'] %>% 
+  sep_ReIm() %>% 
+  .[, correlate(jet, Re(hgt)), by = .(cEOF, part)] 
+```
+
 Figure \@ref(fig:eof1-regr-gh) shows regression maps of SON geopotential height anomalies upon cEOF1.
 At 50 hPa (Figure \@ref(fig:eof1-regr-gh) row a), the 0º cEOF1 is associated with a positive centre located over the Ross Sea.
+The correlation between the 0º cEOF1 and the zonal mean zonal wind at 60ºS and 10hPa is `r jet_correlation[cEOF == "cEOF1" & part == "Real"]$text`, indicating a moderate relationship with the SON stratospheric jet.
 The 90º cEOF1 is associated with a distinctive wave 1 pattern with maximum over the coast of East Antarctica. 
@@
 
-The regression maps of  TOC anomalies upon cEOF1 (Fig.\ \@ref(fig:o3-regr)) show zonal wave 1 patterns  associated with both components of cEOF1.
+The regression maps of TCO anomalies upon cEOF1 (Fig.\ \@ref(fig:o3-regr)) show zonal wave 1 patterns  associated with both components of cEOF1.
 The climatological position of the springtime Ozone minimum (ozone hole) is  outside the South Pole and towards the Weddell Sea [e.g. @grytsai2011].
 Thus, the 0º cEOF1 regression field (Figure\ \@ref(fig:o3-regr)a) coincides with the climatological position of the ozone hole, while it is 90° out of phase for the 90º cEOF1.
-The temporal correlation between the amplitudes of TOC planetary wave 1 and the amplitude of cEOF1 is `r wave1_cor[with == "amplitude"]$text`, while the correlation between their phases is `r wave1_cor[with == "phase"]$text`.
+The temporal correlation between the amplitudes of TCO planetary wave 1 and the amplitude of cEOF1 is `r wave1_cor[with == "amplitude"]$text`, while the correlation between their phases is `r wave1_cor[with == "phase"]$text`.
 Consequently, cEOF1 is strongly related with the SH ozone variability.
@@
 
-In the case of 90º cEOF2, its correlation with the SAM for the troposphere is associated to the asymmetric variability of the SAM. 
+In the case of 90º cEOF2, its correlation with the SAM for the troposphere is associated with the asymmetric variability of the SAM. 
 Indeed, the 90º cEOF2 shares up to `r scales::percent(max_r2[cEOF == "cEOF2" & term == "asym"]$estimate^2, 2)` variance with the A-SAM and only `r scales::percent(max_r2[cEOF == "cEOF2" & term == "sym"]$estimate^2, 2)` at most with the S-SAM (Figure \@ref(fig:sam-eof-vertical).b2).
@@
 Meanwhile, Rows a and c in Fig.\ \@ref(fig:sst-psi-2) show that the intermediate phases are still associated with significant SST regressed anomalies over the Pacific Ocean, but at slightly different locations.
-The 135º phase is associated with SST anomalies in the central Pacific (Fig.\ \@ref(fig:sst-psi-2)a.1), while the 45º phase is associated with SST anomalies that correspond roughly to the Central Pacific and Eastern Pacific "flavours" of ENSO, respectively (Fig.\ \@ref(fig:sst-psi-2)c.1)  [@kao2009].
-Both phases are also associated to wave trains generated in the region surrounding Australia and propagates toward the extra-tropics, although less intense than the ones associated with the 90º phase.
+The 135º phase is associated with SST anomalies in the central Pacific (Fig.\ \@ref(fig:sst-psi-2)a.1), while the 45º phase is associated with SST anomalies in the eastern Pacific, which correspond roughly to the Central Pacific and Eastern Pacific "flavours" of ENSO, respectively (Fig.\ \@ref(fig:sst-psi-2)c.1)  [@kao2009].
+Both phases are also associated with wave trains generated in the region surrounding Australia and propagates toward the extra-tropics, although less intense than the ones associated with the 90º phase.
 
@@
 First, the correlation between cEOF1 and the SAM in the troposphere is modest, with less than 50% of shared variance (Fig.
-\@ref(fig:sam-eof-vertical) column 1), so these indices are not expected to be equivalent.
+\@ref(fig:sam-eof-vertical) column 1), so these indices are not expected to be equivalent to each other.
 Second, @campitelli2022 showed that the strong relationship between the SAM and Pacific SSTs and temperature anomalies around the Antarctic Peninsula is mainly due to the asymmetric part of the SAM.
@@
 However, the 90º cEOF2 is not the phase most correlated with precipitation in that area. 
-The 135º phase (an intermediate between positive 90º and 180º cEOF2) component is associated with stronger and more extensive temporal correlations  with precipitation over Australia and New Zealand. 
-The influence of cEOF2 in Australian precipitation could be more related to the direct impacts of SST anomalies in the surrounded oceans rather than on the interconnection pattern represented by the cEOF2.
+The 135º phase (an intermediate between positive 90º and 180º cEOF2) component is associated with stronger and more extensive temporal correlations with precipitation over Australia and New Zealand, consistent with the effect of ENSO on precipitation in that region.
+These anomalies are probably related to the direct effect of vertical anomalies of the Southern Oscillation and less so to the large-scale circulation anomalies described by cEOF2 [@cai2011b].
+Circulation anomalies associated with the Indian Ocean Dipole could also play a role [@cai2011b].
 
@@
 These patterns are not forced to be orthogonal to other modes of circulation, like they are in standard EOF methodology. 
-This allows us to show for example, that the 90º cEOF2, corresponding to PSA1 variability, is closely associated to the SAM in the troposphere. 
-Previous research in the SAM--PSA relationship had the issue that the SAM and the PSA patterns are not independently derived and so the correlation between these indices had to be zero by construction.
+This allows us to show for example, that the 90º cEOF2, corresponding to PSA1 variability, is closely associated with the SAM in the troposphere. 
+Previous research in the SAM--PSA relationship had the issue that the SAM and the PSA patterns are not independently derived and so the correlation between these indices had to be zero by construction [e.g.
+@yu2015].
 
@@
 In this work we show that the asymmetric component of the SAM can be identified with the PSA1.
-Therefore, the correlation between ENSO and SAM is predominantly the correlation between ENSO and PSA1.
+Therefore, the correlation between ENSO and SAM in SON is predominantly the correlation between ENSO and PSA1, at least in SON. 
 This sheds new light into the previous literature, as it cannot be assumed that a high correlation between ENSO and SAM indexes indicates a relationship between ENSO and zonally symmetric variability.
@@
 Further investigation is necessary to determine the connection between the symmetric component of the SAM and the PSA. 
-It is possible that the PSA may force a zonally symmetric response (or vice versa) via wave-zonal mean flow interactions, or that this correlation is simply a statistical artefact resulting from the EOF methodology used to define the SAM and the fact that the spatial structure of the PSA projects onto the spatial structure of the symmetric SAM.
+It is possible that the PSA may force a zonally symmetric response (or vice versa) via wave-zonal mean flow interactions [@kim2004], or that this correlation is simply a statistical artefact resulting from the EOF methodology used to define the SAM and the fact that the spatial structure of the PSA projects onto the spatial structure of the symmetric SAM.
 

### 2023-03-13 analysis/paper/shceof.tex
Segunda revisión

@@
 
-\title[SH zonally asymmetric circulation with cEOF]{Revisiting the Austral Spring Extratropical Southern Hemisphere zonally asymmetric circulation using complex Empirical Orthogonal Functions}
+\title[SH zonally asymmetric circulation with cEOF]{Revisiting the zonally asymmetric extratropical circulation of the Southern Hemisphere spring using complex empirical orthogonal functions}
 
@@
 The second cEOF (cEOF2) is an alternative representation of the Pacific-South American modes.
-One phase of this cEOF is also extremely correlated with the Asymmetric SAM (A-SAM) in the troposphere.
-Springs with an active ENSO tend to phase-lock the cEOF2, but have no consistent impact on its magnitude.
+One phase of this cEOF is also very highly correlated with the Asymmetric SAM (A-SAM) in the troposphere.
+Springs with an active ENSO tend to lock the cEOF2 to a specific phase, but have no consistent impact on its magnitude.
 Furthermore, we find indications that the location of Pacific Sea Surface Temperature anomalies affect the phase of the cEOF2.
@@
 The wave envelope can represent planetary waves with different amplitude at different longitudes, but lacks information about phase and wave number.
-Using this method, \citet{irving2015} showed that planetary wave amplitude in general is associated to Antarctic sea-ice concentration and temperature, as well as to precipitation anomalies in regions of significant topography in SH mid-latitudes and Antarctica.
+Using this method, \citet{irving2015} showed that planetary wave amplitude in general is associated with Antarctic sea-ice concentration and temperature, as well as to precipitation anomalies in regions of significant topography in SH mid-latitudes and Antarctica.
 
@@
 Within the EOF framework, the Southern Annular Mode (SAM) appears as the leading mode of variability of the SH circulation \citep{fogt2020}.
-The SAM represents a relatively zonally symmetric pattern of alternating low pressures in polar latitudes and a ring of high pressures in high latitudes with an embedded wave 3 pattern that is more prominent in the Pacific sector.
+Spatially, the SAM is characterised by a centre of geopotential anomalies over Antarctica surrounded by anomalies of the opposite sign at middle latitudes.
+Embedded in this zonally symmetric structure is a wave 3 pattern that is more prominent in the Pacific sector.
 The 2nd and 3rd EOFs, usually known as Pacific--South American Patterns (PSA) 1 and PSA2 patterns, respectively, describe meridionally propagating wave trains that originate in the eastern equatorial Pacific and Australian-Indian Ocean sector, and travel towards the South Atlantic following a great-circle arch along the Antarctic coast \citep{mo2001}.
@@
 
-We restrict this work to the September-October-November (SON) trimester because during this season both the tropical teleconnections over South America \citep{cazes-boezio2003} and the SH stratosphere-troposphere interactions \citep{lim2018} are at their maximum amplitude.
+We restrict this work to the September-October-November (SON) trimester.
+During this season the tropical teleconnections over South America are maximised \citep{cazes-boezio2003}, and the SH zonal winds associated with the stratospheric polar vortex increase to peak in October and extend downward after that \citep{lim2018}.
 
@@
 In Section~\ref{spatial} we analyse the spatial patterns of each complex EOF.
-In Section~\ref{regressions} we study the spatial regressions with geopotential height, temperature and ozone anomalies.
+In Section~\ref{regressions} we study the spatial regressions with geopotential height, temperature, and ozone anomalies.
 In Section~\ref{psa} and \ref{sam} we analyse the relationship between cEOF2, the PSA and SAM modes.
@@
 
-We used monthly geopotential height, air temperature, ozone mixing ratio, and total column ozone (TOC) at 2.5º longitude by 2.5º latitude of horizontal resolution and 37 vertical isobaric levels from the European Centre for Medium-Range Weather Forecasts Reanalysis version 5 {[}ERA; \citet{era5}{]} for the period 1979 -- 2019.
+We used monthly geopotential height, air temperature, ozone mixing ratio, and total column ozone (TCO) at 2.5º longitude by 2.5º latitude of horizontal resolution and 37 vertical isobaric levels from the European Centre for Medium-Range Weather Forecasts Reanalysis version 5 {[}ERA; \citet{era5}{]} for the period 1979 -- 2019.
 Most of our analysis is restricted to the post-satellite era to avoid confounding factors arising from the incorporation of satellite observations, but we also used the preliminary back extension of ERA5 from 1950 to 1978 \citep{era5be} to describe long-term trends.
@@
 We compute seasonal means for the different variables, averaging monthly values weighted by the number of days in each month.
-We use the 200 hPa level to represent the high troposphere and 50 hPa to represent the lower stratosphere.
+We use the 200 hPa level to represent the upper troposphere and 50 hPa to represent the lower stratosphere.
 
-We computed the amplitude and phase of the TOC wave 1 by averaging (area-weighted) the data of each SON between 75°S and 45°S, and then extracting the wave-1 component of the Fourier spectrum.
+We computed the amplitude and phase of the TCO wave 1 by averaging (area-weighted) the data of each SON between 75°S and 45°S, and then extracting the wave-1 component of the Fourier spectrum.
 We chose this latitude band because it is wide enough to capture most of the relevant anomalies of SH mid-latitudes.
@@
 We computed the level-dependent SAM index as the leading EOF of year-round monthly geopotential height anomalies south of 20ºS at each level for the whole period \citep{baldwin2009}.
-We further split the SAM into its zonally symmetric and zonally asymmetric components (S-SAM and A-SAM indices respectively).
-These indices were obtained by projecting monthly geopotential height fields onto the zonally asymmetric and zonally symmetric parts of the SAM spatial pattern, as proposed by \citet{campitelli2022}.
+We further split the SAM into its zonally symmetric and zonally asymmetric components (S-SAM and A-SAM indices respectively) following \citet{campitelli2022}.
+The method consists in first computing the leading EOF of monthly geopotential height anomalies at each level and then computing the zonal mean and the zonal anomalies from its spatial pattern.
+We then project each level's monthly geopotential height fields onto the corresponding EOF field, the zonally symmetric field and the zonally asymmetric fields to obtain time series corresponding to the full SAM, the symmetric SAM and the asymmetric SAM, respectively.
+
 Seasonal indices of the PSA patterns (PSA1 and PSA2) were calculated, in agreement with \citet{mo2001}, as the third and fourth EOFs of seasonal mean anomalies for 500-hPa geopotential heights at SH.
@@
 
-\hypertarget{complex-empirical-orthogonal-functions-ceof}{%
-\subsection{Complex Empirical Orthogonal Functions (cEOF)}\label{complex-empirical-orthogonal-functions-ceof}}
+\hypertarget{ceof-methods}{%
+\subsection{Complex Empirical Orthogonal Functions (cEOF)}\label{ceof-methods}}
 
@@
 Each cEOF is a set of complex-valued spatial patterns and time series.
-The real and imaginary components of the complex spatial pattern can be though of as representing two spatial patterns that are shifted by 1/4 wavelength by construction, similar to EOF1 and EOF2 in Figure \ref{fig:eof-naive}.
+The real and imaginary components of the complex spatial pattern can be thought of as representing two spatial patterns that are shifted by 1/4 wavelength by construction, similar to EOF1 and EOF2 in Figure \ref{fig:eof-naive}.
 In this paper we use the term 0º cEOF and 90º cEOF to refer to each part of the whole cEOF.
@@
 
-For instance, when the phase of the wave matches the 0º phase, then the 0º phase time series is positive and the 90º phase time series is zero.
-Similarly, when the phase of the wave matches the 90º phase, the 90º phase time series is positive and the 0º phase time series is zero.
+For instance, when the phase of the wave matches the 0º phase, then the 0º phase time series is positive, and the 90º phase time series is zero.
+Similarly, when the phase of the wave matches the 90º phase, the 90º phase time series is positive, and the 0º phase time series is zero.
 The intermediate phases have non-zero values in both time series.
 
-Traditional EOFs are not unique, and instead are defined up to sign, which corresponds to a rotation in the complex plane of either 0 or \(\pi\).
+In traditional EOFs, the resulting modes are not unique, and instead are defined up to sign, which corresponds to a rotation in the complex plane of either 0 or \(\pi\).
 In the same way, cEOFs are defined up to a rotation in the complex plane of any value between 0 and \(2\pi\) \citep{horel1984}.
@@
 The Hilbert transform is usually understood in terms of time-varying signal.
-However, in this work we apply the Hilbert transform at each latitude circle and at each considered time (i.e.~the signal only depends on longitude).
+However, in this work we apply the Hilbert transform at each latitude circle, level, and at each SON mean (i.e.~the signal only depends on longitude).
 Since each latitude circle is a periodic domain, this procedure does not suffer from edge effects.
@@
 
+Because we are computing the cEOFs of zonal anomalies and not temporal anomalies, the cEOFs need to account for the time-mean zonal anomalies.
+These will tend to be represented by the leading cEOF, which therefore will have a non-zero temporal mean.
+
 As mentioned before, the choice of phases is arbitrary and equally valid.
@@
 
-Preliminary analysis showed that the first cEOF was closely related to the zonal wave 1 of TOC and the second cEOF was closely related to ENSO.
-Therefore, we chose the phase of cEOF1 so that the time series corresponding to the 0º cEOF1 has the maximum correlation with the zonal wave 1 of TOC between 75°S and 45°S.
+Preliminary analysis showed that the first cEOF was closely related to the zonal wave 1 of TCO and the second cEOF was closely related to ENSO.
+Therefore, we chose the phase of cEOF1 so that the time series corresponding to the 0º cEOF1 has the maximum correlation with the zonal wave 1 of TCO between 75°S and 45°S.
 Similarly, we chose the phase of cEOF2 so that the coefficient of determination between the ONI and the 0º cEOF2 is minimised, which also nearly maximises the correlation with the 90º cEOF2.
@@
 \[
-X(\lambda, \phi, t) = \alpha(\lambda, \phi) \operatorname{cEOF_{0º}} + \beta(\lambda, \phi) \operatorname{cEOF_{90º}} + X_0(\lambda, \phi) + \epsilon(\lambda, \phi, t)
+X(\lambda, \phi, t) = \alpha(\lambda, \phi) \operatorname{cEOF_{0^\circ}} + \beta(\lambda, \phi) \operatorname{cEOF_{90^\circ}} + X_0(\lambda, \phi) + \epsilon(\lambda, \phi, t)
 \]
@@
 Both cEOFs show year-to-year variability but show no evidence of decadal variability (Fig. \ref{fig:extended-series}).
-Because the geopotential fields that enter into the cEOFs algorithm are anomalies with respect to the zonal mean instead of the time mean, the cEOFs time series have non zero temporal mean.
-However, cEOF2 temporal mean is almost zero, which indicates that only cEOF1 includes variability that significantly projects onto the mean zonally anomalous field.
-This is consistent with the fact that the mean zonally anomalous field of geopotential height is very similar to the cEOF1 (\(r^2\) = 98\%) and not similar to the cEOF2 (\(r^2\) = 0\%).
+The 0º cEOF has a non-zero temporal mean which, as discussed in Section \ref{ceof-methods}, is due to the fact that the temporal mean of zonal anomalies need to be captured by the cEOFs.
+The other indices have almost zero temporal mean, which indicates that only cEOF1 includes variability that significantly projects onto the mean zonal anomalies.
+This is consistent with the fact that the mean zonal anomalies of geopotential height are very similar to the cEOF1 (\(r^2\) = 98\%) and not similar to the cEOF2 (\(r^2\) = 0\%).
 
@@
 At 50 hPa (Figure~\ref{fig:eof1-regr-gh} row a), the 0º cEOF1 is associated with a positive centre located over the Ross Sea.
+The correlation between the 0º cEOF1 and the zonal mean zonal wind at 60ºS and 10hPa is -0.59 (CI: -0.76 -- -0.35), indicating a moderate relationship with the SON stratospheric jet.
 The 90º cEOF1 is associated with a distinctive wave 1 pattern with maximum over the coast of East Antarctica.
@@
 
-The regression maps of TOC anomalies upon cEOF1 (Fig.~\ref{fig:o3-regr}) show zonal wave 1 patterns associated with both components of cEOF1.
+The regression maps of TCO anomalies upon cEOF1 (Fig.~\ref{fig:o3-regr}) show zonal wave 1 patterns associated with both components of cEOF1.
 The climatological position of the springtime Ozone minimum (ozone hole) is outside the South Pole and towards the Weddell Sea \citep[e.g.][]{grytsai2011}.
 Thus, the 0º cEOF1 regression field (Figure~\ref{fig:o3-regr}a) coincides with the climatological position of the ozone hole, while it is 90° out of phase for the 90º cEOF1.
-The temporal correlation between the amplitudes of TOC planetary wave 1 and the amplitude of cEOF1 is 0.79 (CI: 0.63 -- 0.88), while the correlation between their phases is -0.85 (CI: -0.92 -- -0.74).
+The temporal correlation between the amplitudes of TCO planetary wave 1 and the amplitude of cEOF1 is 0.79 (CI: 0.63 -- 0.88), while the correlation between their phases is -0.85 (CI: -0.92 -- -0.74).
 Consequently, cEOF1 is strongly related with the SH ozone variability.
@@
 
-In the case of 90º cEOF2, its correlation with the SAM for the troposphere is associated to the asymmetric variability of the SAM.
+In the case of 90º cEOF2, its correlation with the SAM for the troposphere is associated with the asymmetric variability of the SAM.
 Indeed, the 90º cEOF2 shares up to 92\% variance with the A-SAM and only 12\% at most with the S-SAM (Figure~\ref{fig:sam-eof-vertical}.b2).
@@
 Meanwhile, Rows a and c in Fig.~\ref{fig:sst-psi-2} show that the intermediate phases are still associated with significant SST regressed anomalies over the Pacific Ocean, but at slightly different locations.
-The 135º phase is associated with SST anomalies in the central Pacific (Fig.~\ref{fig:sst-psi-2}a.1), while the 45º phase is associated with SST anomalies that correspond roughly to the Central Pacific and Eastern Pacific ``flavours'' of ENSO, respectively (Fig.~\ref{fig:sst-psi-2}c.1) \citep{kao2009}.
-Both phases are also associated to wave trains generated in the region surrounding Australia and propagates toward the extra-tropics, although less intense than the ones associated with the 90º phase.
+The 135º phase is associated with SST anomalies in the central Pacific (Fig.~\ref{fig:sst-psi-2}a.1), while the 45º phase is associated with SST anomalies in the eastern Pacific, which correspond roughly to the Central Pacific and Eastern Pacific ``flavours'' of ENSO, respectively (Fig.~\ref{fig:sst-psi-2}c.1) \citep{kao2009}.
+Both phases are also associated with wave trains generated in the region surrounding Australia and propagates toward the extra-tropics, although less intense than the ones associated with the 90º phase.
 
@@
 First, the correlation between cEOF1 and the SAM in the troposphere is modest, with less than 50\% of shared variance (Fig.
-\ref{fig:sam-eof-vertical} column 1), so these indices are not expected to be equivalent.
+\ref{fig:sam-eof-vertical} column 1), so these indices are not expected to be equivalent to each other.
 Second, \citet{campitelli2022} showed that the strong relationship between the SAM and Pacific SSTs and temperature anomalies around the Antarctic Peninsula is mainly due to the asymmetric part of the SAM.
@@
 However, the 90º cEOF2 is not the phase most correlated with precipitation in that area.
-The 135º phase (an intermediate between positive 90º and 180º cEOF2) component is associated with stronger and more extensive temporal correlations with precipitation over Australia and New Zealand.
-The influence of cEOF2 in Australian precipitation could be more related to the direct impacts of SST anomalies in the surrounded oceans rather than on the interconnection pattern represented by the cEOF2.
+The 135º phase (an intermediate between positive 90º and 180º cEOF2) component is associated with stronger and more extensive temporal correlations with precipitation over Australia and New Zealand, consistent with the effect of ENSO on precipitation in that region.
+These anomalies are probably related to the direct effect of vertical anomalies of the Southern Oscillation and less so to the large-scale circulation anomalies described by cEOF2 \citep{cai2011b}.
+Circulation anomalies associated with the Indian Ocean Dipole could also play a role \citep{cai2011b}.
 
@@
 These patterns are not forced to be orthogonal to other modes of circulation, like they are in standard EOF methodology.
-This allows us to show for example, that the 90º cEOF2, corresponding to PSA1 variability, is closely associated to the SAM in the troposphere.
-Previous research in the SAM--PSA relationship had the issue that the SAM and the PSA patterns are not independently derived and so the correlation between these indices had to be zero by construction.
+This allows us to show for example, that the 90º cEOF2, corresponding to PSA1 variability, is closely associated with the SAM in the troposphere.
+Previous research in the SAM--PSA relationship had the issue that the SAM and the PSA patterns are not independently derived and so the correlation between these indices had to be zero by construction \citep[e.g.][]{yu2015}.
 
@@
 In this work we show that the asymmetric component of the SAM can be identified with the PSA1.
-Therefore, the correlation between ENSO and SAM is predominantly the correlation between ENSO and PSA1.
+Therefore, the correlation between ENSO and SAM in SON is predominantly the correlation between ENSO and PSA1, at least in SON.
 This sheds new light into the previous literature, as it cannot be assumed that a high correlation between ENSO and SAM indexes indicates a relationship between ENSO and zonally symmetric variability.
@@
 Further investigation is necessary to determine the connection between the symmetric component of the SAM and the PSA.
-It is possible that the PSA may force a zonally symmetric response (or vice versa) via wave-zonal mean flow interactions, or that this correlation is simply a statistical artefact resulting from the EOF methodology used to define the SAM and the fact that the spatial structure of the PSA projects onto the spatial structure of the symmetric SAM.
+It is possible that the PSA may force a zonally symmetric response (or vice versa) via wave-zonal mean flow interactions \citep{kim2004}, or that this correlation is simply a statistical artefact resulting from the EOF methodology used to define the SAM and the fact that the spatial structure of the PSA projects onto the spatial structure of the symmetric SAM.
 

### 2023-03-13 R/0-download.R

@@ -213,6 +213,20 @@ request_era5_2mt <- list(
 )
 
 
+request_jet <- list(
+  format = "netcdf",
+  product_type = "monthly_averaged_reanalysis",
+  variable = c("u_component_of_wind"),
+  pressure_level = c("10"),
+  year = as.character(1979:2020),
+  month = formatC(9:11, width = 2, flag = 0),   # Need all months to compute PSA
+  time = "00:00",
+  grid = c("2.5", "2.5"),
+  area = southern_hemisphere,
+  dataset_short_name = "reanalysis-era5-pressure-levels-monthly-means"
+)
+
+
 
 simple_download <- function(url) {
   force(url)

### 2023-03-13 R/data_locations.R

@@ -28,7 +28,7 @@ download_from_zenodo <- function(data) {
 }
 
 
-zenodo_record <- "https://zenodo.org/record/6612429/"
+zenodo_record <- "https://zenodo.org/record/7730671/"
 url_from_base <- function(file) {
   paste0(zenodo_record, "/files/", basename(file), "?download=1")
 }
@@ -127,7 +127,16 @@ ERA5_geopotential_all <- data_function(
   citation = citation_era5
 )
 
-
+#' @export
+#' @rdname data_locations
+ERA5_jet <- data_function(
+  name = "era5-jet",
+  file = data_path("raw", "era5-uwind.mon.mean.nc"),
+  source = download_cds(request_jet),
+  zenodo_url = url_from_base,
+  zenodo_md5 = "bba785ef40fb6e7c16cbec2dc23cf1a3",
+  citation = citation_era5
+)
 
 #' @export
 #' @rdname data_locations

### 2023-04-23 analysis/paper/shceof.Rmd
Adds support for HTML version

@@
 output: 
+  bookdown::html_document2:
+      code_folding: "hide"
   # bookdown::word_document2:
@@
   dpi = 300,
-  dev = c("pdf", "png")
+  dev = if (knitr::is_html_output()) c("png") else c("pdf", "png")
 )

### 2023-06-08 analysis/paper/shceof.Rmd
Adds gitbook

--- 2023-03-13 analysis/paper/shceof.Rmd
@@
 
+link-citations: true
 # bibstyle: spbasic
-
 output: 
+  bookdown::gitbook:
+      global_numbering: true
+
+  bookdown::html_document2:
+      code_folding: "hide"
   # bookdown::word_document2:
@@
 ```{r setup, echo = FALSE, warning=FALSE, message=FALSE}
+format <- knitr::pandoc_to()
 knitr::opts_chunk$set(
@@
   comment = "#>",
-  cache.extra = 41,
-  cache.path = "../cache/paper/",
+  cache.extra = 42,
+  cache.path = paste0("../cache/paper/", format, "/"),
   # fig.path = "../figures/",
   dpi = 300,
-  dev = c("pdf", "png")
+  dev = if (knitr::is_html_output()) c("png") else c("pdf", "png")
 )
